### 2019-09-03 ThesisProposal.Rmd
Created Markdown file for (draft) Thesis Proposal

---
title: "Draft Thesis Proposal"
output: 
  pdf_document: 
    fig_caption: yes
author: "Hanne Oberman"
---

# Introduction

Research goals...

develop a valid method to investigate the plausibility of multiply imputed data based on:

- models (imputation and non-response)

- data features (cross-tabs, point estimates, aggregate statistics, etc.)

- assumptions

- algorithmic convergence

# Literature review

Van Buuren, 2018...

# Approach

R Shiny...

ShinyMICE

- single measure to assess whether algorithm converged (based on Gelman-Rubin statistic?)

- data visualizations pre and post imputation (scatterplots, densities, cross-tabs)

- statistical evaluation of relations between variables pre and post imputation (chi square or t-tests)

### 2019-09-05 ThesisProposal.Rmd
Linked reference manager (using Better BibLaTex)
Added original thesis proposal and some lit

@@
 title: "Draft Thesis Proposal"
+author: "Hanne Oberman"
+date: "`r Sys.Date()`"
 output: 
@@
     fig_caption: yes
-author: "Hanne Oberman"
+bibliography: ThesisProposal.bib
 ---
@@
 
-Van Buuren, 2018...
+@buur18
+
+@abay08
+
+@bart15
+
+@li91
+
+@rubin87
+
+@rubin96
+
+@vinknd
+
+@mice11
+
+@shiny17
 
@@
 - statistical evaluation of relations between variables pre and post imputation (chi square or t-tests)
+
+# Initial proposal
+
+ShinyMice: an evaluation suite for multiple imputation
+Supervisors: Prof.dr. Stef van Buuren, Dr. Gerko Vink
+
+The MICE package in R is a world-leading software package for multiple imputation.
+When using the mice function to solve missing data, vast amounts of information are calculated and stored.
+However, the MICE package currently lacks a user-friendly means of assessing this information.
+This project focuses on developing and programming novel means of evaluating, presenting and organising this information in a web-browser based local app that allows for 1) comparing the imputed data to the observations, 2) inspecting the algorithmic convergence, 3) inspecting multivariate distributions, 4) the plausibility of the imputed data, and so on.
+During this project you will work closely with the developers of MICE in R and the coding components in this project will focus on R and Shiny.
+
+The objective of the research is to 1) create a Shiny app to investigate and evaluate multiply imputed data sets, 2) implement it in MICE in R, 3) write an instructional vignette, 4) write a technical paper on the workings and usage of the software aimed at e.g.
+the R Journal or Journal of Statistical Software.
+The expected output is a state-of-the-art shiny app that can find its way into MICE.
+
+# References

### 2019-09-09 ThesisProposal.Rmd
more lit with some quotes

@@
 
-@buur18
+“There is no clear-cut method for determining when the MICE algorithm has converged” (@buur18, \textsection 6.5).
+
+“Convergence is diagnosed when the variance between different sequences is no larger than the variance within each individual sequence” (@buur18, \textsection 6.5).
+
+“Several expository reviews are available that assess convergence diagnostics for MCMC methods (Cowles and Carlin 1996; Brooks and Gelman 1998; El Adlouni, Favre, and Bobée 2006).
+Cowles and Carlin (1996) conclude that “automated convergence monitoring (as by a machine) is unsafe and should be avoided.” No method works best in all circumstances.
+The consensus is to assess convergence with a combination of tools.
+The added value of using a combination of convergence diagnostics for missing data imputation has not yet been systematically studied” (@buur18, \textsection 6.5).
 
@@
 @shiny17
+
+@scha02
+
+@cowl96
+
+"As MICE is an iterative procedure, it is important that convergence is achieved.
+This may be checked by computing, at each cycle, the means of imputed values and/or the values of regression coefficients, and seeing if they are stable" (@whit11, p.
+394).
 

### 2019-09-10 ThesisProposal.Rmd
added references expressing need for convergence eval

@@
 
+@zhu15
+
 "As MICE is an iterative procedure, it is important that convergence is achieved.
@@
 394).
+
+"A common diagnostic tool is to plot one or more parameters against the iteration number and assess convergence by how different the variance between different sequences is relative to the variance within each individual sequence, similar to the Gelman-Rubin statistic (Gelman and Rubin, 1992) used in Markov chain Monte Carlo (MCMC) diagnostics" (@li14, \textsection 4.3).
+
+"R.hat: The value of the Rˆ statistic used as a convergence criterion. The default is 1.1 (Gelman and Rubin 1992; Gelman, Carlin, Stern, and Rubin 2004)" (@su11 p. 4).
+
+"Our mi offers two ways to check the convergence of the multiple imputation procedure.
+By default, mi() monitors the mixing of each variable by the variance of its mean and standard deviation within and between different chains of the imputation.
+If the Rˆ statistic is smaller than 1.1, (i.e., the difference of the within and between variance is trivial), the imputation is considered converged (Gelman, Carlin, Stern, and Rubin 2004).
+Additionally, by specifying mi(data, check.coef.convergence = TRUE, ...), users can check the convergence of the parameters of the conditional models" (@su11, p.
+13).
+
+"the convergence properties of FCS are currently under debate due to possible incompatibility (Li, Yu, and Rubin 2012; Zhu and Raghunathan 2015)" (@taka17, p. )
+
+"The convergence properties of FCS in general settings is still mostly an open question.
+The behavior of FCS algorithms under non- or quasi-Bayesian imputation procedures like PMM is entirely an open question" (@murr18, p.19)
+
+Bayes course on convergence:
+```{r}
+# E. Assessing convergence
+# -----------------------------------------------------------------------------
+# History plot, autocorrelation plot
+
+# History plots show the sampled parameters over the iterations (excluding
+# the burn-in).
+# The development/pattern in these plots gives an indication of convergence.
+# When the history plot is stable (a fat catterpillar), convergence is reached.
+# Autocorrelation can be measured at many lags.
+# High autocorrelation indicates slow mixing of the random path.
+
+mcmcplot(mcmcout = samples)
+# #sigma seems fine, but b doesn't:
+# #even at lag 5 there is still quite some autocorrelation: therefore we need to center the predictors!
+# #otherwise we could use a million iterations to deminish this effect
+
+# Gelman-Rubin diagnostic
+
+# The Gelman and Rubin statistic requires you run
+# the sampler/algorithm at least twice: These runs are
+# referred to as multiple chains.
+# It compares the variance between chains to the variance
+# within chains (G-R statistic = T/W = (pooled within chain
+# var + between chain var)/pooled within chain var.
+# It's the red line in the plot, and should be near 1.
+# See Gibbs sampler presentation (week 2)
+# https://drive.google.com/file/d/1ABHm8ala3c_puVvf32zF8h6TOZ3898uB/view
+# pdf p. 41, slide nr 29.
+
+gelman.plot(samples)
+# #this appears to be okay with a really small deviation from 1
+
+# MC error (OPTIONAL?)
+
+# MC error = SD/sqrt(number of iterations)
+# SD represents the variation across iterations
+# MC error thus represents how much the means differ w.r.t. the iterations
+# MC error decreases as number of iterations increases.
+# It should not be larger than 5% of the sample standard deviation
+
+# History and density plot (OPTIONAL)
+# plot(samples)
+
+# Autocorrelation plots (OPTIONAL)
+# autocorr.plot(samples)
+
+# If parameters did not converge, you may:
+# . Use (many) more iterations
+# . Use a different parametrization (e.g., center predictors)
+# . Use different priors (e.g., multivariate normal prior (i.e.,
+#   dmnorm(,)) for parameters which are correlated)
+# . Use other initial values
+```
 

### 2019-09-10 ThesisProposal.Rmd
added proposal requirements and title page

@@
 ---
-title: "Draft Thesis Proposal"
+title: "Thesis Proposal"
+subtitle: "Methodology and Statistics for the Behavioural, Biomedical and Social Sciences"
 author: "Hanne Oberman"
@@
 ---
+
+\centering
+
+# ShinyMICE: an Evaluation Suite for Multiple Imputation
+
+Supervised by Prof. Dr. Stef van Buuren, & Dr. Gerko Vink
+
+Department of Methodology and Statistics
+
+Utrecht University
+
+\raggedright
+
+Requirements: max 750 words (excl. references)
+
+In general the proposal should focus on the relevancy of the project, the gap in the scientific literature, and the feasibility to finish the project within 8 months.
+Possibly specify a 'step 2' to pursue after the main objective is reached.
+
+\newpage
+
+# Proposal by supervisors
+
+The MICE package in R is a world-leading software package for multiple imputation.
+When using the mice function to solve missing data, vast amounts of information are calculated and stored.
+However, the MICE package currently lacks a user-friendly means of assessing this information.
+This project focuses on developing and programming novel means of evaluating, presenting and organising this information in a web-browser based local app that allows for 1) comparing the imputed data to the observations, 2) inspecting the algorithmic convergence, 3) inspecting multivariate distributions, 4) the plausibility of the imputed data, and so on.
+During this project you will work closely with the developers of MICE in R and the coding components in this project will focus on R and Shiny.
+
+The objective of the research is to 1) create a Shiny app to investigate and evaluate multiply imputed data sets, 2) implement it in MICE in R, 3) write an instructional vignette, 4) write a technical paper on the workings and usage of the software aimed at e.g.
+the R Journal or Journal of Statistical Software.
+The expected output is a state-of-the-art shiny app that can find its way into MICE.
 
@@
 
+"Imputers who do choose to use FCS should use flexible univariate models wherever possible and take care to assess apparent convergence of the algorithm, for example by computing traces of pooled estimates or other statistics and using standard MCMC diagnostics (Gelman et al., 2013, Chpater 11).
+It may also be helpful to examine the results of many independent runs of the algorithm with different initializations and to use random scans over the p variables to try to identify any convergence issues and mitigate possible order dependence" (@murr18, p.
+19).
+
+# Approach
+
+R Shiny...
+
+ShinyMICE
+
+- single measure to assess whether algorithm converged (based on Gelman-Rubin statistic?)
+
+- data visualizations pre and post imputation (scatterplots, densities, cross-tabs)
+
+- statistical evaluation of relations between variables pre and post imputation (chi square or t-tests)
+
+# Additional resources
+
 Bayes course on convergence:
@@
 # History plot, autocorrelation plot
-
 # History plots show the sampled parameters over the iterations (excluding
@@
 
-mcmcplot(mcmcout = samples)
+# mcmcplot(mcmcout = samples)
 # #sigma seems fine, but b doesn't:
@@
 # Gelman-Rubin diagnostic
-
 # The Gelman and Rubin statistic requires you run
@@
 # pdf p. 41, slide nr 29.
-
-gelman.plot(samples)
+# gelman.plot(samples)
 # #this appears to be okay with a really small deviation from 1
@@
 # MC error (OPTIONAL?)
-
 # MC error = SD/sqrt(number of iterations)
@@
 
-# Approach
-
-R Shiny...
-
-ShinyMICE
-
-- single measure to assess whether algorithm converged (based on Gelman-Rubin statistic?)
-
-- data visualizations pre and post imputation (scatterplots, densities, cross-tabs)
-
-- statistical evaluation of relations between variables pre and post imputation (chi square or t-tests)
-
-# Initial proposal
-
-ShinyMice: an evaluation suite for multiple imputation
-Supervisors: Prof.dr. Stef van Buuren, Dr. Gerko Vink
-
-The MICE package in R is a world-leading software package for multiple imputation.
-When using the mice function to solve missing data, vast amounts of information are calculated and stored.
-However, the MICE package currently lacks a user-friendly means of assessing this information.
-This project focuses on developing and programming novel means of evaluating, presenting and organising this information in a web-browser based local app that allows for 1) comparing the imputed data to the observations, 2) inspecting the algorithmic convergence, 3) inspecting multivariate distributions, 4) the plausibility of the imputed data, and so on.
-During this project you will work closely with the developers of MICE in R and the coding components in this project will focus on R and Shiny.
-
-The objective of the research is to 1) create a Shiny app to investigate and evaluate multiply imputed data sets, 2) implement it in MICE in R, 3) write an instructional vignette, 4) write a technical paper on the workings and usage of the software aimed at e.g.
-the R Journal or Journal of Statistical Software.
-The expected output is a state-of-the-art shiny app that can find its way into MICE.
-
 # References

### 2019-09-10 ThesisProposal.Rmd
paraphrased some quotes for lit. review

@@
 subtitle: "Methodology and Statistics for the Behavioural, Biomedical and Social Sciences"
-author: "Hanne Oberman"
+author: "Hanne Oberman (4216318)"
 date: "`r Sys.Date()`"
@@
 
+Word count:
+
 \raggedright
@@
 However, the MICE package currently lacks a user-friendly means of assessing this information.
-This project focuses on developing and programming novel means of evaluating, presenting and organising this information in a web-browser based local app that allows for 1) comparing the imputed data to the observations, 2) inspecting the algorithmic convergence, 3) inspecting multivariate distributions, 4) the plausibility of the imputed data, and so on.
+This project focuses on developing and programming novel means of evaluating, presenting and organizing this information in a web-browser based local app that allows for 1) comparing the imputed data to the observations, 2) inspecting the algorithmic convergence, 3) inspecting multivariate distributions, 4) the plausibility of the imputed data, and so on.
 During this project you will work closely with the developers of MICE in R and the coding components in this project will focus on R and Shiny.
@@
 
-# Introduction
+# Introduction/context and research question/goals
 
-Research goals...
+Almost all (social) scientific research has to cope with missingness.
+Ignoring missing data yields invalid inferences (@buur18).
+Multiple imputation provides a solution and is growing in popularity.
+The incomplete data is completed several times.
+The variability between these imputations en-captures the uncertainty of the inferences due to missingness (@rubin87).
 
-develop a valid method to investigate the plausibility of multiply imputed data based on:
-
-- models (imputation and non-response)
-
-- data features (cross-tabs, point estimates, aggregate statistics, etc.)
-
-- assumptions
-
-- algorithmic convergence
+The MICE package in R is a world-leading software package for multiple imputation (Multiple Imputation using Chained Equations; @mice11).
+However, the package currently does not provide user-friendly means to evaluate the multiply imputed data.
+Therefore, the goal of this project is develop a MICE evaluation suite, featuring existing modules (like plots to compare the incomplete and completed data sets), and novel assessment tools (like measures to quantify algorithmic convergence).
+The practical relevance of this research project is in facilitating applied researchers in making valid inferences despite of missingness.
+Moreover, it contributes to the scientific literature because there is not yet a clear-cut way of diagnosing convergence for MICE algorithms.
 
@@
 
-“There is no clear-cut method for determining when the MICE algorithm has converged” (@buur18, \textsection 6.5).
+That is, “there is no clear-cut method for determining when the MICE algorithm has converged” (@buur18, \textsection 6.5).
+Currently, users have to rely on visual inspection of the algorithm's iterations.
+Convergence is said to be reached when the parameters (like means of completed variables or regression coefficients) are stable over iterations (@whit11).
+<!-- "As MICE is an iterative procedure, it is important that convergence is achieved.
+This may be checked by computing, at each cycle, the means of imputed values and/or the values of regression coefficients, and seeing if they are stable" (@whit11, p.
+394).
+-->
+And additionally, when the variance between the several imputation chains is no larger than the variances within each chain (@buur18). 
+<!-- “Convergence is diagnosed when the variance between different sequences is no larger than the variance within each individual sequence” (@buur18, \textsection 6.5).  -->
+According to @li14 the latter part of this procedure is conceptually similar to the Gelman-Rubin statistic (@gelm92). 
+<!-- "A common diagnostic tool is to plot one or more parameters against the iteration number and assess convergence by how different the variance between different sequences is relative to the variance within each individual sequence, similar to the Gelman-Rubin statistic (Gelman and Rubin, 1992) used in Markov chain Monte Carlo (MCMC) diagnostics" (@li14, \textsection 4.3).
+-->
 
-“Convergence is diagnosed when the variance between different sequences is no larger than the variance within each individual sequence” (@buur18, \textsection 6.5).
-
-“Several expository reviews are available that assess convergence diagnostics for MCMC methods (Cowles and Carlin 1996; Brooks and Gelman 1998; El Adlouni, Favre, and Bobée 2006).
-Cowles and Carlin (1996) conclude that “automated convergence monitoring (as by a machine) is unsafe and should be avoided.” No method works best in all circumstances.
-The consensus is to assess convergence with a combination of tools.
-The added value of using a combination of convergence diagnostics for missing data imputation has not yet been systematically studied” (@buur18, \textsection 6.5).
+Since the G-R statistic assumes independence between ..., it cannot be directly applied as convergence criterion with 1.1 as conventional cut-off (see @su11).
+<!-- "R.hat: The value of the $\hat{R}$ statistic used as a convergence criterion. The default is 1.1 (Gelman and Rubin 1992; Gelman, Carlin, Stern, and Rubin 2004)" (@su11 p. 4). -->
+<!-- "Our mi offers two ways to check the convergence of the multiple imputation procedure.
+By default, mi() monitors the mixing of each variable by the variance of its mean and standard deviation within and between different chains of the imputation.
+If the $\hat{R}$ statistic is smaller than 1.1, (i.e., the difference of the within and between variance is trivial), the imputation is considered converged (Gelman, Carlin, Stern, and Rubin 2004).
+Additionally, by specifying mi(data, check.coef.convergence = TRUE, ...), users can check the convergence of the parameters of the conditional models" (@su11, p.
+13).
+-->
 
@@
 
-@mice11
-
-@shiny17
-
 @scha02
@@
 
-"As MICE is an iterative procedure, it is important that convergence is achieved.
-This may be checked by computing, at each cycle, the means of imputed values and/or the values of regression coefficients, and seeing if they are stable" (@whit11, p.
-394).
-
-"A common diagnostic tool is to plot one or more parameters against the iteration number and assess convergence by how different the variance between different sequences is relative to the variance within each individual sequence, similar to the Gelman-Rubin statistic (Gelman and Rubin, 1992) used in Markov chain Monte Carlo (MCMC) diagnostics" (@li14, \textsection 4.3).
-
-"R.hat: The value of the Rˆ statistic used as a convergence criterion. The default is 1.1 (Gelman and Rubin 1992; Gelman, Carlin, Stern, and Rubin 2004)" (@su11 p. 4).
-
-"Our mi offers two ways to check the convergence of the multiple imputation procedure.
-By default, mi() monitors the mixing of each variable by the variance of its mean and standard deviation within and between different chains of the imputation.
-If the Rˆ statistic is smaller than 1.1, (i.e., the difference of the within and between variance is trivial), the imputation is considered converged (Gelman, Carlin, Stern, and Rubin 2004).
-Additionally, by specifying mi(data, check.coef.convergence = TRUE, ...), users can check the convergence of the parameters of the conditional models" (@su11, p.
-13).
-
 "the convergence properties of FCS are currently under debate due to possible incompatibility (Li, Yu, and Rubin 2012; Zhu and Raghunathan 2015)" (@taka17, p. )
@@
 
-"Imputers who do choose to use FCS should use flexible univariate models wherever possible and take care to assess apparent convergence of the algorithm, for example by computing traces of pooled estimates or other statistics and using standard MCMC diagnostics (Gelman et al., 2013, Chpater 11).
+"Imputers who do choose to use FCS should use flexible univariate models wherever possible and take care to assess apparent convergence of the algorithm, for example by computing traces of pooled estimates or other statistics and using standard MCMC diagnostics (Gelman et al., 2013, Chpater [*sic*] 11).
 It may also be helpful to examine the results of many independent runs of the algorithm with different initializations and to use random scans over the p variables to try to identify any convergence issues and mitigate possible order dependence" (@murr18, p.
@@
 
-# Approach
+“Several expository reviews are available that assess convergence diagnostics for MCMC methods (Cowles and Carlin 1996; Brooks and Gelman 1998; El Adlouni, Favre, and Bobée 2006).
+Cowles and Carlin (1996) conclude that “automated convergence monitoring (as by a machine) is unsafe and should be avoided.” No method works best in all circumstances.
+The consensus is to assess convergence with a combination of tools.
+The added value of using a combination of convergence diagnostics for missing data imputation has not yet been systematically studied” (@buur18, \textsection 6.5).
 
-R Shiny...
+# Approach that will be used to answer the research question
 
-ShinyMICE
+Work directly with developers of the MICE R package.
+Use R Shiny (@shiny17) to program ShinyMICE.
+First, create/develop.
+Then, implement in existing MICE R package.
+Then write technical paper on the workings and usage of the software (thesis).
+Then write vingette for applied researchers.
 
-- single measure to assess whether algorithm converged (based on Gelman-Rubin statistic?)
+Develop a valid method to investigate the plausibility of multiply imputed data based on:
+
+- models (imputation and non-response)
+
+- data features (cross-tabs, point estimates, aggregate statistics, etc.)
+
+- assumptions
+
+- algorithmic convergence (single measure to assess whether algorithm converged (potentially based on Gelman-Rubin statistic))
 
@@
 - statistical evaluation of relations between variables pre and post imputation (chi square or t-tests)
+
+Possible journal for publication: the R Journal or Journal of Statistical Software.
 

### 2019-09-12 ThesisProposal.Rmd
first draft thesis proposal

@@
 
-# Proposal by supervisors
-
-The MICE package in R is a world-leading software package for multiple imputation.
-When using the mice function to solve missing data, vast amounts of information are calculated and stored.
-However, the MICE package currently lacks a user-friendly means of assessing this information.
-This project focuses on developing and programming novel means of evaluating, presenting and organizing this information in a web-browser based local app that allows for 1) comparing the imputed data to the observations, 2) inspecting the algorithmic convergence, 3) inspecting multivariate distributions, 4) the plausibility of the imputed data, and so on.
-During this project you will work closely with the developers of MICE in R and the coding components in this project will focus on R and Shiny.
-
-The objective of the research is to 1) create a Shiny app to investigate and evaluate multiply imputed data sets, 2) implement it in MICE in R, 3) write an instructional vignette, 4) write a technical paper on the workings and usage of the software aimed at e.g.
-the R Journal or Journal of Statistical Software.
-The expected output is a state-of-the-art shiny app that can find its way into MICE.
-
 # Introduction/context and research question/goals
 
-Almost all (social) scientific research has to cope with missingness.
-Ignoring missing data yields invalid inferences (@buur18).
-Multiple imputation provides a solution and is growing in popularity.
-The incomplete data is completed several times.
-The variability between these imputations en-captures the uncertainty of the inferences due to missingness (@rubin87).
+Almost all (social) scientific research copes with missingness (source?).
+Ignoring missing data can yield wildly invalid inferences (@buur18).
+A solution is provided by multiple imputation (MI; @rubin87).
+This technique -- that is growing in popularity (@buur18) -- entails completing a data set with missingness several times.
+The variability between these completions en-captures the uncertainty of the inferences due to missingness (@rubin87).
 
-The MICE package in R is a world-leading software package for multiple imputation (Multiple Imputation using Chained Equations; @mice11).
+The R package MICE (Multiple Imputation using Chained Equations; @mice11) is the world-leading software for multiple imputation (@buur18).
 However, the package currently does not provide user-friendly means to evaluate the multiply imputed data.
-Therefore, the goal of this project is develop a MICE evaluation suite, featuring existing modules (like plots to compare the incomplete and completed data sets), and novel assessment tools (like measures to quantify algorithmic convergence).
+Therefore, the goal of this project is develop a MICE evaluation suite, featuring existing modules (like plots to compare the incomplete and completed data sets), and novel assessment tools (like a measure to assess algorithmic convergence).
 The practical relevance of this research project is in facilitating applied researchers in making valid inferences despite of missingness.
@@
 
-That is, “there is no clear-cut method for determining when the MICE algorithm has converged” (@buur18, \textsection 6.5).
-Currently, users have to rely on visual inspection of the algorithm's iterations.
+Currently, “there is no clear-cut method for determining when the MICE algorithm has converged” (@buur18, \textsection 6.5).
+Instead, users have to rely on visual inspection of the algorithm's iterations.
 Convergence is said to be reached when the parameters (like means of completed variables or regression coefficients) are stable over iterations (@whit11).
@@
 -->
-And additionally, when the variance between the several imputation chains is no larger than the variances within each chain (@buur18). 
+Additionally, convergence requires the variance between the several imputation chains to be no larger than the variances within each chain (@buur18). 
 <!-- “Convergence is diagnosed when the variance between different sequences is no larger than the variance within each individual sequence” (@buur18, \textsection 6.5).  -->
-According to @li14 the latter part of this procedure is conceptually similar to the Gelman-Rubin statistic (@gelm92). 
+According to @li14 the latter is conceptually similar to the Gelman-Rubin statistic. 
 <!-- "A common diagnostic tool is to plot one or more parameters against the iteration number and assess convergence by how different the variance between different sequences is relative to the variance within each individual sequence, similar to the Gelman-Rubin statistic (Gelman and Rubin, 1992) used in Markov chain Monte Carlo (MCMC) diagnostics" (@li14, \textsection 4.3).
@@
 
-Since the G-R statistic assumes independence between ..., it cannot be directly applied as convergence criterion with 1.1 as conventional cut-off (see @su11).
+The Gelman-Rubin statistic (@gelm92), however, assumes independence between chains. Therefore, it cannot be directly applied to compare imputations that share the same incomplete data (source??).
+
+Notwithstandlingly, it seems like @su11 did implement the Gelman-Rubin statistic as convergence criterion into their R package 'mi', including the conventional cut-off of 1.1.
 <!-- "R.hat: The value of the $\hat{R}$ statistic used as a convergence criterion. The default is 1.1 (Gelman and Rubin 1992; Gelman, Carlin, Stern, and Rubin 2004)" (@su11 p. 4). -->
@@
 -->
+It is worth investigating its validity in the context of MICE's convergence, and the broader framework of fully conditional specification (FSC) algorithms.
+
+The most recent literature on the topic states that convergence properties of FCS algorithms are still under debate (@taka17), with the specific case of a MICE procedure like predictive mean matching posing an entirely open question (@murr18).
+<!-- "the convergence properties of FCS are currently under debate due to possible incompatibility (Li, Yu, and Rubin 2012; Zhu and Raghunathan 2015)" (@taka17, p. ) -->
+<!-- "The convergence properties of FCS in general settings is still mostly an open question.
+The behavior of FCS algorithms under non- or quasi-Bayesian imputation procedures like PMM is entirely an open question" (@murr18, p.19) -->
+@buur18 summarizes this problem as follows: "No method works best in all circumstances.
+The consensus is to assess convergence with a combination of tools.
+The added value of using a combination of convergence diagnostics for missing data imputation has not yet been systematically studied” (@buur18, \textsection 6.5).
+<!-- “Several expository reviews are available that assess convergence diagnostics for MCMC methods (Cowles and Carlin 1996; Brooks and Gelman 1998; El Adlouni, Favre, and Bobée 2006).
+Cowles and Carlin (1996) conclude that “automated convergence monitoring (as by a machine) is unsafe and should be avoided.” No method works best in all circumstances.
+The consensus is to assess convergence with a combination of tools.
+The added value of using a combination of convergence diagnostics for missing data imputation has not yet been systematically studied” (@buur18, \textsection 6.5).
+-->
+And that is exactly the gap in the scientific literature that this thesis will contribute to.
+
+# Approach that will be used to answer the research question
+
+During this research project I will work directly with developers of the MICE R package.
+I will use R Shiny (@shiny17) to program a web-browser based local app for the evaluation of multiply imputed data: 'ShinyMICE'.
+
+<!-- Develop a valid method to investigate the plausibility of multiply imputed data based on: -->
+
+<!-- - models (imputation and non-response) -->
+
+<!-- - data features (cross-tabs, point estimates, aggregate statistics, etc.) -->
+
+<!-- - assumptions -->
+
+The research report that is to be handed in December 13th 2019 will consist of an investigation into algorithmic convergence of MICE.
+Ideally, this will result in a single indicator to flag non-convergence (potentially based on the Gelman-Rubin statistic, autocorrelation, or MC error).
+
+The second stage of the research project is of more practical nature: programming ShinyMICE.
+The application will at least consist of the following: 1) one or more measures to assess algorithmic convergence; 2) data visualizations (e.g., scatterplots, densities, and cross-tabulations of the data pre- and post-imputation); and 3) statistical evaluation of relations between variables pre- versus post-imputation (i.e., chi square tests or t-tests).
+This part of the thesis also entails research into user interface (UI) design, local versus cloud-based data storage, and optimizing the application's efficiency.
+
+A working beta version of ShinyMICE will be considered sufficient to proceed to the next step: writing a technical paper on the workings and usage of the software.
+Possible journals for publication of this manuscript are *The R Journal* and *Journal of Statistical Software*.
+
+Finally, the shinyMICE package will be integrated into the existing MICE environment, and a vingette for applied researchers will be written.
+
+# Additional resources
+
+"Imputers who do choose to use FCS should use flexible univariate models wherever possible and take care to assess apparent convergence of the algorithm, for example by computing traces of pooled estimates or other statistics and using standard MCMC diagnostics (Gelman et al., 2013, Chpater [*sic*] 11).
+It may also be helpful to examine the results of many independent runs of the algorithm with different initializations and to use random scans over the p variables to try to identify any convergence issues and mitigate possible order dependence" (@murr18, p.
+19).
 
@@
 
-"the convergence properties of FCS are currently under debate due to possible incompatibility (Li, Yu, and Rubin 2012; Zhu and Raghunathan 2015)" (@taka17, p. )
+@gelm92's convergence criterion is defined as the ratio between the sum of the pooled within chain variance plus the between chain variance, and the pooled within chain variance.
 
-"The convergence properties of FCS in general settings is still mostly an open question.
-The behavior of FCS algorithms under non- or quasi-Bayesian imputation procedures like PMM is entirely an open question" (@murr18, p.19)
+Proposal by supervisors:
 
-"Imputers who do choose to use FCS should use flexible univariate models wherever possible and take care to assess apparent convergence of the algorithm, for example by computing traces of pooled estimates or other statistics and using standard MCMC diagnostics (Gelman et al., 2013, Chpater [*sic*] 11).
-It may also be helpful to examine the results of many independent runs of the algorithm with different initializations and to use random scans over the p variables to try to identify any convergence issues and mitigate possible order dependence" (@murr18, p.
-19).
+"The MICE package in R is a world-leading software package for multiple imputation.
+When using the mice function to solve missing data, vast amounts of information are calculated and stored.
+However, the MICE package currently lacks a user-friendly means of assessing this information.
+This project focuses on developing and programming novel means of evaluating, presenting and organizing this information in a web-browser based local app that allows for 1) comparing the imputed data to the observations, 2) inspecting the algorithmic convergence, 3) inspecting multivariate distributions, 4) the plausibility of the imputed data, and so on.
+During this project you will work closely with the developers of MICE in R and the coding components in this project will focus on R and Shiny.
 
-“Several expository reviews are available that assess convergence diagnostics for MCMC methods (Cowles and Carlin 1996; Brooks and Gelman 1998; El Adlouni, Favre, and Bobée 2006).
-Cowles and Carlin (1996) conclude that “automated convergence monitoring (as by a machine) is unsafe and should be avoided.” No method works best in all circumstances.
-The consensus is to assess convergence with a combination of tools.
-The added value of using a combination of convergence diagnostics for missing data imputation has not yet been systematically studied” (@buur18, \textsection 6.5).
-
-# Approach that will be used to answer the research question
-
-Work directly with developers of the MICE R package.
-Use R Shiny (@shiny17) to program ShinyMICE.
-First, create/develop.
-Then, implement in existing MICE R package.
-Then write technical paper on the workings and usage of the software (thesis).
-Then write vingette for applied researchers.
-
-Develop a valid method to investigate the plausibility of multiply imputed data based on:
-
-- models (imputation and non-response)
-
-- data features (cross-tabs, point estimates, aggregate statistics, etc.)
-
-- assumptions
-
-- algorithmic convergence (single measure to assess whether algorithm converged (potentially based on Gelman-Rubin statistic))
-
-- data visualizations pre and post imputation (scatterplots, densities, cross-tabs)
-
-- statistical evaluation of relations between variables pre and post imputation (chi square or t-tests)
-
-Possible journal for publication: the R Journal or Journal of Statistical Software.
-
-# Additional resources
+The objective of the research is to 1) create a Shiny app to investigate and evaluate multiply imputed data sets, 2) implement it in MICE in R, 3) write an instructional vignette, 4) write a technical paper on the workings and usage of the software aimed at e.g.
+the R Journal or Journal of Statistical Software.
+The expected output is a state-of-the-art shiny app that can find its way into MICE".
 

### 2019-09-18 ThesisProposal.Rmd
Update references and draft presentation

@@
 
-Almost all (social) scientific research copes with missingness (source?).
-Ignoring missing data can yield wildly invalid inferences (@buur18).
-A solution is provided by multiple imputation (MI; @rubin87).
-This technique -- that is growing in popularity (@buur18) -- entails completing a data set with missingness several times.
-The variability between these completions en-captures the uncertainty of the inferences due to missingness (@rubin87).
+At some point, any (social) scientist conducting statistical analyses will run into a missing data problem (@alli02).
+Simply ignoring the missingness or using ad hoc solutions can yield wildly invalid inferences (@buur18).
+Multiple imputation (MI; @rubin87) provides a framework to circumvent the problem of missing information.
+This technique -- that is growing in popularity (@buur18) -- entails completing the missing values in an incomplete data set several times.
+The variability between the completed data sets captures the uncertainty in the inferences due to missingness (@rubin87).
 
@@
 However, the package currently does not provide user-friendly means to evaluate the multiply imputed data.
-Therefore, the goal of this project is develop a MICE evaluation suite, featuring existing modules (like plots to compare the incomplete and completed data sets), and novel assessment tools (like a measure to assess algorithmic convergence).
+Therefore, the goal of this project is develop a MICE evaluation suite, featuring existing modules (like plots to compare the incomplete and completed data sets), and novel assessment tools (like a measure to flag algorithmic non-convergence).
 The practical relevance of this research project is in facilitating applied researchers in making valid inferences despite of missingness.
-Moreover, it contributes to the scientific literature because there is not yet a clear-cut way of diagnosing convergence for MICE algorithms.
+Simultaniously it contributes to the scientific literature because there is not yet a clear-cut way of diagnosing convergence for MI algorithms.
 
@@
 I will use R Shiny (@shiny17) to program a web-browser based local app for the evaluation of multiply imputed data: 'ShinyMICE'.
+This approach does not require the use of any (personal) data.
+Therefore, the project will not be reviewed by the ethical committee.
 
@@
 A working beta version of ShinyMICE will be considered sufficient to proceed to the next step: writing a technical paper on the workings and usage of the software.
-Possible journals for publication of this manuscript are *The R Journal* and *Journal of Statistical Software*.
+The preferred journal for publication of this manuscript is *Journal of Statistical Software* -- alternatively, publication in *The R Journal* will be attempted.
 
@@
 
-# Additional resources
+# Additional resources (not used)
 
@@
 
-@abay08
+<!-- @abay08 -->
 
-@bart15
+<!-- @bart15 -->
 
-@li91
+<!-- @li91 -->
 
-@rubin87
+<!-- @rubin87 -->
 
-@rubin96
+<!-- @rubin96 -->
 
-@vinknd
+<!-- @vinknd -->
 
-@scha02
+<!-- @scha02 -->
 
-@cowl96
+<!-- @cowl96 -->
 
-@zhu15
+<!-- @zhu15 -->
 
@@
 
-Proposal by supervisors:
+**Proposal by supervisors:**
 
@@
 
-Bayes course on convergence:
+**Bayes course on convergence:**
 ```{r}

### 2019-09-19 ResearchReport.Rmd
Update after feedback Gerko on draft thesis proposal

---
title: "Research Report"
subtitle: "Methodology and Statistics for the Behavioural, Biomedical and Social Sciences"
author: "Hanne Oberman (4216318)"
date: "`r Sys.Date()`"
output: 
  pdf_document: 
    fig_caption: yes
bibliography: ThesisProposal.bib
---

\centering

# ShinyMICE: an Evaluation Suite for Multiple Imputation

Supervised by prof. dr. Stef van Buuren, & dr. Gerko Vink

Department of Methodology and Statistics

Utrecht University

\raggedright

Goals:

- find an answer to a research question in the field of Methods and Statistics

- get work experience in the field of Methods and Statistics

- a first step in writing a scientific paper: perform research and report the findings in a short research report

Grading:

- The amount of work done

- Work independency

- Clarity and logic of the research report (clear and correct formulations, a logical structure of sections and of paragraphs within sections, no typos or errors in grammar, and flow of sections, like key questions from literature or conclusions from results).

- Nice looking and consistent lay-out of the report (conciseness in presentation of tables, figures, and formulas (if any)).

- General comments

Format (either/or):

- It is written as a (mini) thesis, with an introduction, methods section, some results (i.e., preliminary analyses, or pilot simulations), and a discussion of results. The length of the research report should be maximally 2500 words of text (without references list and/or tables and figures). Please do not include appendices, and no more than 6 tables/figures. Table and Figure captions do not count towards the word limit. An abstract may be included, but is not necessary.

- It is the first ‘half’ of the thesis, i.e., there are no results included yet, but the report contains a full introduction including a literature review and a methods section that contains details about the data, instruments and/or statistical procedures. There is no (absolute) word limit but the length of the report should comply with recommendations/expectations of the proposed journal. Note that most journals have word limits or rather strong opinions on (too) lengthy manuscripts.

\newpage

# Introduction

- MI (and ad hoc solutions?)

# Theoretical Background

- Terminology (MCAR, MAR, MNAR)

- Rubin's rules

- FCS vs. JM

# Methods

-

# References

### 2019-09-19 ThesisProposal.Rmd
Update after feedback Gerko on draft thesis proposal

@@
 
-Supervised by Prof. Dr. Stef van Buuren, & Dr. Gerko Vink
+Supervised by prof. dr. Stef van Buuren, & dr. Gerko Vink
 
@@
 
-# Introduction/context and research question/goals
+# Introduction
 
@@
 Simply ignoring the missingness or using ad hoc solutions can yield wildly invalid inferences (@buur18).
-Multiple imputation (MI; @rubin87) provides a framework to circumvent the problem of missing information.
-This technique -- that is growing in popularity (@buur18) -- entails completing the missing values in an incomplete data set several times.
-The variability between the completed data sets captures the uncertainty in the inferences due to missingness (@rubin87).
+Multiple imputation (MI; @rubin87) provides a framework to circumvent the *ubiquitous* problem of missing information.
+This technique -- that is growing in popularity (@buur18) -- entails 'guessing' the missing values in an incomplete data set several times.
+The variability between the resulting completed data sets represents how much uncertainty in the inferences is due to missingness (@rubin87).
 
 The R package MICE (Multiple Imputation using Chained Equations; @mice11) is the world-leading software for multiple imputation (@buur18).
-However, the package currently does not provide user-friendly means to evaluate the multiply imputed data.
+However, the package currently does not provide user-friendly means to evaluate multiply imputed data.
+*This is vital because of the amount of assumptions in MI algorithms.
+Thanks to technological developments in computing speed and data storage it has become possible to check assumptions where that was unfeasible before.
+The very first assumption to check is algorithmic convergence; without convergence, all 'deeper' assumptions and subsequent inferences are invalid.*
+
 Therefore, the goal of this project is develop a MICE evaluation suite, featuring existing modules (like plots to compare the incomplete and completed data sets), and novel assessment tools (like a measure to flag algorithmic non-convergence).
 The practical relevance of this research project is in facilitating applied researchers in making valid inferences despite of missingness.
-Simultaniously it contributes to the scientific literature because there is not yet a clear-cut way of diagnosing convergence for MI algorithms.
+Simultaneously it contributes to the scientific literature because there is not yet a clear-cut way of diagnosing convergence for MI algorithms.
+*The research question central to this project is: 'How can empirical researchers be facilitated in evaluating the  validity of multiply imputed data?'.*
 
@@
 Instead, users have to rely on visual inspection of the algorithm's iterations.
-Convergence is said to be reached when the parameters (like means of completed variables or regression coefficients) are stable over iterations (@whit11).
+Convergence is said to be reached when parameters (like means of completed variables or regression coefficients) are stable across iterations (@whit11).
 <!-- "As MICE is an iterative procedure, it is important that convergence is achieved.
@@
 -->
-Additionally, convergence requires the variance between the several imputation chains to be no larger than the variances within each chain (@buur18). 
+And additionally, the variation between imputations should be no larger than the variation across iterations of each individual chain (@buur18). 
 <!-- “Convergence is diagnosed when the variance between different sequences is no larger than the variance within each individual sequence” (@buur18, \textsection 6.5).  -->
-According to @li14 the latter is conceptually similar to the Gelman-Rubin statistic. 
+*Conceptually, this visual evaluation procedure is similar to the 'Gelman-Rubin statistic' (@li14).
+Practically, however, Gelman and Rubin's convergence diagnostic $\widehat{R}$ ('potential scale reduction'; 1992) cannot be directly applied to assess convergence of MICE algorithms because the measure is famously prone to produce false positives when applied to MI data.*
 <!-- "A common diagnostic tool is to plot one or more parameters against the iteration number and assess convergence by how different the variance between different sequences is relative to the variance within each individual sequence, similar to the Gelman-Rubin statistic (Gelman and Rubin, 1992) used in Markov chain Monte Carlo (MCMC) diagnostics" (@li14, \textsection 4.3).
 -->
+*That is, the overdispersed initial state of the MICE algorithm does not conform to $\widehat{R}$'s assumption of equal starting points (source!).*
 
-The Gelman-Rubin statistic (@gelm92), however, assumes independence between chains. Therefore, it cannot be directly applied to compare imputations that share the same incomplete data (source??).
-
-Notwithstandlingly, it seems like @su11 did implement the Gelman-Rubin statistic as convergence criterion into their R package 'mi', including the conventional cut-off of 1.1.
+Notwithstandingly, it seems like @su11 did implement $\widehat{R}$ as convergence criterion into their R package 'mi', including the conventional cut-off of 1.1.
 <!-- "R.hat: The value of the $\hat{R}$ statistic used as a convergence criterion. The default is 1.1 (Gelman and Rubin 1992; Gelman, Carlin, Stern, and Rubin 2004)" (@su11 p. 4). -->
@@
 -->
-It is worth investigating its validity in the context of MICE's convergence, and the broader framework of fully conditional specification (FSC) algorithms.
+It is worth investigating the validity of the convergence statistic in the context of MICE, and the broader framework of fully conditional specification (FSC) algorithms.
 
@@
 
+**Potentially add something about updated (2019) version of $\widehat{R}$ and the new threshold of 1.01, see @veht19. Or leave it for the Research Report.**
+
 # Approach that will be used to answer the research question
@@
 During this research project I will work directly with developers of the MICE R package.
-I will use R Shiny (@shiny17) to program a web-browser based local app for the evaluation of multiply imputed data: 'ShinyMICE'.
-This approach does not require the use of any (personal) data.
-Therefore, the project will not be reviewed by the ethical committee.
+I will use R Shiny (@shiny17) to program an interactive evaluation device for the evaluation of multiply imputed data: 'ShinyMICE'.
+This approach does not require the use of any empirical data -- personal or otherwise.
+Therefore, the project does not need to be reviewed/approved by the faculty's ethical committee.
 
@@
 
-Finally, the shinyMICE package will be integrated into the existing MICE environment, and a vingette for applied researchers will be written.
+Finally, the shinyMICE package will be integrated into the existing MICE environment, and a vignette for applied researchers will be written.
 
@@
 
-"Imputers who do choose to use FCS should use flexible univariate models wherever possible and take care to assess apparent convergence of the algorithm, for example by computing traces of pooled estimates or other statistics and using standard MCMC diagnostics (Gelman et al., 2013, Chpater [*sic*] 11).
+"For models based on multiple imputed data sets, this [R hat > 1.1] is often a false positive: Chains of different submodels may not overlay each other exactly, since there were fitted to different data" (see https://cran.r-project.org/web/packages/brms/vignettes/brms_missings.html).
+
+"Imputers who do choose to use FCS should use flexible univariate models wherever possible and take care to assess apparent convergence of the algorithm, for example by computing traces of pooled estimates or other statistics and using standard MCMC diagnostics (Gelman et al., 2013, Chapter 11).
 It may also be helpful to examine the results of many independent runs of the algorithm with different initializations and to use random scans over the p variables to try to identify any convergence issues and mitigate possible order dependence" (@murr18, p.

### 2019-09-20 ThesisProposal.Rmd
Same but with spell check

@@
 -->
-*That is, the overdispersed initial state of the MICE algorithm does not conform to $\widehat{R}$'s assumption of equal starting points (source!).*
+*That is, the over-dispersed initial state of the MICE algorithm does not conform to $\widehat{R}$'s assumption of equal starting points (source?).*
 
-Notwithstandingly, it seems like @su11 did implement $\widehat{R}$ as convergence criterion into their R package 'mi', including the conventional cut-off of 1.1.
+Notwithstanding, it seems like @su11 did implement $\widehat{R}$ as convergence criterion into their R package 'mi', including the conventional cut-off of 1.1.
 <!-- "R.hat: The value of the $\hat{R}$ statistic used as a convergence criterion. The default is 1.1 (Gelman and Rubin 1992; Gelman, Carlin, Stern, and Rubin 2004)" (@su11 p. 4). -->
@@
 The research report that is to be handed in December 13th 2019 will consist of an investigation into algorithmic convergence of MICE.
-Ideally, this will result in a single indicator to flag non-convergence (potentially based on the Gelman-Rubin statistic, autocorrelation, or MC error).
+Ideally, this will result in a single indicator to flag non-convergence (potentially based on the Gelman-Rubin statistic, auto-correlation, or MC error).
 
 The second stage of the research project is of more practical nature: programming ShinyMICE.
-The application will at least consist of the following: 1) one or more measures to assess algorithmic convergence; 2) data visualizations (e.g., scatterplots, densities, and cross-tabulations of the data pre- and post-imputation); and 3) statistical evaluation of relations between variables pre- versus post-imputation (i.e., chi square tests or t-tests).
+The application will at least consist of the following: 1) one or more measures to assess algorithmic convergence; 2) data visualizations (e.g., scatter-plots, densities, and cross-tabulations of the data pre- and post-imputation); and 3) statistical evaluation of relations between variables pre- versus post-imputation (i.e., chi square tests or t-tests).
 This part of the thesis also entails research into user interface (UI) design, local versus cloud-based data storage, and optimizing the application's efficiency.
@@
 @gelm92's convergence criterion is defined as the ratio between the sum of the pooled within chain variance plus the between chain variance, and the pooled within chain variance.
+
+"Why can the number of iterations in MICE be so low?
+First of all, realize that the imputed data Ymis form the only memory in the the MICE algorithm.
+Chapter 3 explained that imputed data can have a considerable amount of random noise, depending on the strength of the relations between the variables.
+Applications of MICE with lowly correlated data therefore inject a lot of noise into the system.
+Hence, the auto-correlation over t will be low, and convergence will be rapid, and in fact immediate if all variables are independent.
+Thus, the incorporation of noise into the imputed data has pleasant side-effect of speeding up convergence" (@buur18, par.
+4.5).
 

### 2019-09-25 ResearchReport.Rmd
Thesis proposal for feedback research seminar

@@
 
--
+- Potential convergence measures:
+
+*why do people say the GR stat will give false positives?
+perhaps the window (nr of it) is too small.
+if you start from dispersed, you'll end up with convergence (GR = 1) --> the bad thing is to stop too early.
+this is a false negative, not positive!
+possible approach: replicate and build a decision rule to solve the problem.
+*
+
+*in mice algorithms, iterating is very fast compared to regular MCMC.
+therefore, we could use the average over the last 5 iterations you'll take the 'majority vote' to indicate conv.
+so compute GR at each iteration: band and threshold.
+another approach is to compute autocorrelation  (which is decreasing over itarations because we add in noise.
+schafer (1997, p.
+129) wrote on worst linear stat and calculate the autocorr of that stat we know that the alg did converge.
+mice has noise so little autocorr.
+but we have very rich data so we're not estimating everything from scratch.
+not: we're talking about missing data only, not the combined data (that autocorr.
+is very high).
+autocorr is high if we can predict the miss value very well.
+the higher r sq the higher the autocorr.
+so the harder to converge.
+watch out for the problem with texp that will converge at once because you can deductively impute the value per class.
+possibly use slope of mean too to see whether there is trending.
+maybe also compute first eigenvalue of the matrix, not just mean and sd.
+also in pca we have several measures for nr of components, why not do something similar for mice?
+however, now we use mean and sd of missing values only, and that's not possible fr correlations because we'll have different ns for different correls.
+but of we take the compl.
+data that should not be a problem because that's constant.
+maybe look into allpying pca on the imputed data and if that stays the same we know that the means and variances are stable too.
+patrick ...
+and ian white work on stata proc mi also work on convergence diagnostics --> autocorr function plot in sas of worst linear function, and likelihood ratio that you could save.*
+
+- Bayes course on convergence:
+```{r}
+# E. Assessing convergence
+# -----------------------------------------------------------------------------
+# History plot, autocorrelation plot
+# History plots show the sampled parameters over the iterations (excluding
+# the burn-in).
+# The development/pattern in these plots gives an indication of convergence.
+# When the history plot is stable (a fat catterpillar), convergence is reached.
+# Autocorrelation can be measured at many lags.
+# High autocorrelation indicates slow mixing of the random path.
+
+# mcmcplot(mcmcout = samples)
+# #sigma seems fine, but b doesn't:
+# #even at lag 5 there is still quite some autocorrelation: therefore we need to center the predictors!
+# #otherwise we could use a million iterations to deminish this effect
+
+# Gelman-Rubin diagnostic
+# The Gelman and Rubin statistic requires you run
+# the sampler/algorithm at least twice: These runs are
+# referred to as multiple chains.
+# It compares the variance between chains to the variance
+# within chains (G-R statistic = T/W = (pooled within chain
+# var + between chain var)/pooled within chain var.
+# It's the red line in the plot, and should be near 1.
+# See Gibbs sampler presentation (week 2)
+# https://drive.google.com/file/d/1ABHm8ala3c_puVvf32zF8h6TOZ3898uB/view
+# pdf p. 41, slide nr 29.
+# gelman.plot(samples)
+# #this appears to be okay with a really small deviation from 1
+
+# MC error (OPTIONAL?)
+# MC error = SD/sqrt(number of iterations)
+# SD represents the variation across iterations
+# MC error thus represents how much the means differ w.r.t. the iterations
+# MC error decreases as number of iterations increases.
+# It should not be larger than 5% of the sample standard deviation
+
+# History and density plot (OPTIONAL)
+# plot(samples)
+
+# Autocorrelation plots (OPTIONAL)
+# autocorr.plot(samples)
+
+# If parameters did not converge, you may:
+# . Use (many) more iterations
+# . Use a different parametrization (e.g., center predictors)
+# . Use different priors (e.g., multivariate normal prior (i.e.,
+#   dmnorm(,)) for parameters which are correlated)
+# . Use other initial values
+```
+
+# Additional resources (not used)
+
+**Potentially add something about updated (2019) version of $\widehat{R}$ and the new threshold of 1.01, see @veht19. Or leave it for the Research Report.**
+
+"For models based on multiple imputed data sets, this [R hat > 1.1] is often a false positive: Chains of different submodels may not overlay each other exactly, since there were fitted to different data" (see https://cran.r-project.org/web/packages/brms/vignettes/brms_missings.html).
+
+"Imputers who do choose to use FCS should use flexible univariate models wherever possible and take care to assess apparent convergence of the algorithm, for example by computing traces of pooled estimates or other statistics and using standard MCMC diagnostics (Gelman et al., 2013, Chapter 11).
+It may also be helpful to examine the results of many independent runs of the algorithm with different initializations and to use random scans over the p variables to try to identify any convergence issues and mitigate possible order dependence" (@murr18, p.
+19).
+
+@gelm92's convergence criterion is defined as the ratio between the sum of the pooled within chain variance plus the between chain variance, and the pooled within chain variance.
+
+"Why can the number of iterations in MICE be so low?
+First of all, realize that the imputed data Ymis form the only memory in the the MICE algorithm.
+Chapter 3 explained that imputed data can have a considerable amount of random noise, depending on the strength of the relations between the variables.
+Applications of MICE with lowly correlated data therefore inject a lot of noise into the system.
+Hence, the auto-correlation over t will be low, and convergence will be rapid, and in fact immediate if all variables are independent.
+Thus, the incorporation of noise into the imputed data has pleasant side-effect of speeding up convergence" (@buur18, par.
+4.5).
 

### 2019-09-25 ThesisProposal.Rmd
Thesis proposal for feedback research seminar

--- 2019-09-18 ThesisProposal.Rmd
@@
 title: "Thesis Proposal"
-subtitle: "Methodology and Statistics for the Behavioural, Biomedical and Social Sciences"
 author: "Hanne Oberman (4216318)"
@@
 output:
+  #word_document: default
   pdf_document:
     fig_caption: yes
+subtitle: Methodology and Statistics for the Behavioural, Biomedical and Social Sciences
 bibliography: ThesisProposal.bib
@@
 
-Word count:
+Word count: 821
 
@@
 
-Requirements: max 750 words (excl. references)
+<!-- Requirements: max 750 words (excl. references) -->
 
-In general the proposal should focus on the relevancy of the project, the gap in the scientific literature, and the feasibility to finish the project within 8 months.
+<!-- In general the proposal should focus on the relevancy of the project, the gap in the scientific literature, and the feasibility to finish the project within 8 months.
 Possibly specify a 'step 2' to pursue after the main objective is reached.
+-->
 
@@
 
-# Introduction/context and research question/goals
+# Introduction
 
-At some point, any (social) scientist conducting statistical analyses will run into a missing data problem (@alli02).
-Simply ignoring the missingness or using ad hoc solutions can yield wildly invalid inferences (@buur18).
-Multiple imputation (MI; @rubin87) provides a framework to circumvent the problem of missing information.
-This technique -- that is growing in popularity (@buur18) -- entails completing the missing values in an incomplete data set several times.
-The variability between the completed data sets captures the uncertainty in the inferences due to missingness (@rubin87).
+At some point, any (social) scientist conducting statistical analyses will run into a missing data problem [@alli02].
+Simply ignoring the missingness or using ad hoc solutions can yield wildly invalid inferences [@buur18].
+Multiple imputation (MI; [@rubin87]) provides a framework to circumvent the *ubiquitous* problem of missing information.
+This technique -- that is growing in popularity [@buur18] -- entails 'guessing' the missing values in an incomplete data set several times.
+The variability between the resulting completed data sets represents how much uncertainty in the inferences is due to missingness [@rubin87].
 
-The R package MICE (Multiple Imputation using Chained Equations; @mice11) is the world-leading software for multiple imputation (@buur18).
-However, the package currently does not provide user-friendly means to evaluate the multiply imputed data.
-Therefore, the goal of this project is develop a MICE evaluation suite, featuring existing modules (like plots to compare the incomplete and completed data sets), and novel assessment tools (like a measure to flag algorithmic non-convergence).
+The R package MICE (Multiple Imputation using Chained Equations; [@mice11]) is the world-leading software for multiple imputation [@buur18].
+However, the package currently does not provide user-friendly means to evaluate multiply imputed data.
+*This is vital because of the amount of assumptions in MICE algorithms.
+Thanks to technological developments in computing speed and data storage it has become possible to check assumptions where that was unfeasible before.
+The very first assumption to check is algorithmic convergence; without convergence, all 'deeper' assumptions and subsequent inferences are invalid.*
+
+*The research question central to this project is: 'How can empirical researchers be facilitated in evaluating the  validity of multiply imputed data?'.* Therefore, the goal is to develop a MICE evaluation suite, featuring interactive adaptations of existing modules (like plots to compare the incomplete and completed data sets), and novel assessment tools (like a measure to flag algorithmic non-convergence).
 The practical relevance of this research project is in facilitating applied researchers in making valid inferences despite of missingness.
-Simultaniously it contributes to the scientific literature because there is not yet a clear-cut way of diagnosing convergence for MI algorithms.
+Simultaneously, it contributes to the scientific literature because there is no definitive way of diagnosing non-convergence for MI algorithms.
 
-# Literature review
+# Literature Review
 
-Currently, “there is no clear-cut method for determining when the MICE algorithm has converged” (@buur18, \textsection 6.5).
-Instead, users have to rely on visual inspection of the algorithm's iterations.
-Convergence is said to be reached when the parameters (like means of completed variables or regression coefficients) are stable over iterations (@whit11).
+Since “there is no clear-cut method for determining when the MICE algorithm has converged” (@buur18, \textsection 6.5), users have to rely on visual inspection of the algorithm's iterations.
+Convergence is said to be reached when parameters (e.g., means of completed variables) are stable across iterations [@whit11].
 <!-- "As MICE is an iterative procedure, it is important that convergence is achieved.
@@
 -->
-Additionally, convergence requires the variance between the several imputation chains to be no larger than the variances within each chain (@buur18). 
+And additionally, the variation between imputation chains should be no larger than the variation within each individual chain [@buur18]. 
 <!-- “Convergence is diagnosed when the variance between different sequences is no larger than the variance within each individual sequence” (@buur18, \textsection 6.5).  -->
-According to @li14 the latter is conceptually similar to the Gelman-Rubin statistic. 
+*Conceptually, this visual evaluation procedure is similar to the 'Gelman-Rubin statistic' [@li14].
+Practically, however, Gelman and Rubin's convergence diagnostic $\widehat{R}$ [-@gelm92] cannot be applied directly to MI data.
+The measure produces too many 'false alarms': chains that converged within ten iterations are still flagged as non-converged at iteration fifty [@lace07].*
 <!-- "A common diagnostic tool is to plot one or more parameters against the iteration number and assess convergence by how different the variance between different sequences is relative to the variance within each individual sequence, similar to the Gelman-Rubin statistic (Gelman and Rubin, 1992) used in Markov chain Monte Carlo (MCMC) diagnostics" (@li14, \textsection 4.3).
@@
 
-The Gelman-Rubin statistic (@gelm92), however, assumes independence between chains. Therefore, it cannot be directly applied to compare imputations that share the same incomplete data (source??).
-
-Notwithstandlingly, it seems like @su11 did implement the Gelman-Rubin statistic as convergence criterion into their R package 'mi', including the conventional cut-off of 1.1.
+Notwithstandingly, it seems like @su11 did implement $\widehat{R}$ as convergence criterion into their R package 'mi', including the conventional cut-off.
 <!-- "R.hat: The value of the $\hat{R}$ statistic used as a convergence criterion. The default is 1.1 (Gelman and Rubin 1992; Gelman, Carlin, Stern, and Rubin 2004)" (@su11 p. 4). -->
@@
 -->
-It is worth investigating its validity in the context of MICE's convergence, and the broader framework of fully conditional specification (FSC) algorithms.
+So it is worth investigating the validity of this convergence statistic in the context of multiple imputation.
 
-The most recent literature on the topic states that convergence properties of FCS algorithms are still under debate (@taka17), with the specific case of a MICE procedure like predictive mean matching posing an entirely open question (@murr18).
+The most recent literature on the topic states that convergence properties of MICE algorithms in general are still under debate [@taka17]--with specific procedures like 'predictive mean matching' posing entirely open questions [@murr18].
 <!-- "the convergence properties of FCS are currently under debate due to possible incompatibility (Li, Yu, and Rubin 2012; Zhu and Raghunathan 2015)" (@taka17, p. ) -->
@@
 
-# Approach that will be used to answer the research question
+# Methods
 
 During this research project I will work directly with developers of the MICE R package.
-I will use R Shiny (@shiny17) to program a web-browser based local app for the evaluation of multiply imputed data: 'ShinyMICE'.
-This approach does not require the use of any (personal) data.
-Therefore, the project will not be reviewed by the ethical committee.
+I will use R Shiny [@shiny17] to program an interactive evaluation device for the evaluation of multiply imputed data: 'ShinyMICE'.
+*This approach does not require the use of any empirical data.
+Therefore, the project does not need to be reviewed/approved by the faculty's ethical committee.*
 
@@
 The research report that is to be handed in December 13th 2019 will consist of an investigation into algorithmic convergence of MICE.
-Ideally, this will result in a single indicator to flag non-convergence (potentially based on the Gelman-Rubin statistic, autocorrelation, or MC error).
+Ideally, this will result in a single indicator to flag non-convergence (potentially based on $\widehat{R}$, auto-correlation, or MC error).
 
 The second stage of the research project is of more practical nature: programming ShinyMICE.
-The application will at least consist of the following: 1) one or more measures to assess algorithmic convergence; 2) data visualizations (e.g., scatterplots, densities, and cross-tabulations of the data pre- and post-imputation); and 3) statistical evaluation of relations between variables pre- versus post-imputation (i.e., chi square tests or t-tests).
-This part of the thesis also entails research into user interface (UI) design, local versus cloud-based data storage, and optimizing the application's efficiency.
+The application will at least consist of the following: 1) one or more measures to assess algorithmic convergence; 2) data visualizations (e.g., scatter-plots, densities, and cross-tabulations of the data pre- and post-imputation); and 3) statistical evaluation of relations between variables pre- versus post-imputation (i.e., $\chi^2$-tests or t-tests).
+*This part of the thesis also entails research into user interface design, local versus cloud-based data storage, and optimizing the application's efficiency.*
+<!-- 1) comparing the imputed data to the observations, 2) inspecting the algorithmic convergence, 3) inspecting multivariate distributions, 4) the plausibility of the imputed data -->
 
@@
 
-Finally, the shinyMICE package will be integrated into the existing MICE environment, and a vingette for applied researchers will be written.
-
-# Additional resources (not used)
-
-"Imputers who do choose to use FCS should use flexible univariate models wherever possible and take care to assess apparent convergence of the algorithm, for example by computing traces of pooled estimates or other statistics and using standard MCMC diagnostics (Gelman et al., 2013, Chpater [*sic*] 11).
-It may also be helpful to examine the results of many independent runs of the algorithm with different initializations and to use random scans over the p variables to try to identify any convergence issues and mitigate possible order dependence" (@murr18, p.
-19).
-
-<!-- @abay08 -->
-
-<!-- @bart15 -->
-
-<!-- @li91 -->
-
-<!-- @rubin87 -->
-
-<!-- @rubin96 -->
-
-<!-- @vinknd -->
-
-<!-- @scha02 -->
-
-<!-- @cowl96 -->
-
-<!-- @zhu15 -->
-
-@gelm92's convergence criterion is defined as the ratio between the sum of the pooled within chain variance plus the between chain variance, and the pooled within chain variance.
-
-**Proposal by supervisors:**
-
-"The MICE package in R is a world-leading software package for multiple imputation.
-When using the mice function to solve missing data, vast amounts of information are calculated and stored.
-However, the MICE package currently lacks a user-friendly means of assessing this information.
-This project focuses on developing and programming novel means of evaluating, presenting and organizing this information in a web-browser based local app that allows for 1) comparing the imputed data to the observations, 2) inspecting the algorithmic convergence, 3) inspecting multivariate distributions, 4) the plausibility of the imputed data, and so on.
-During this project you will work closely with the developers of MICE in R and the coding components in this project will focus on R and Shiny.
-
-The objective of the research is to 1) create a Shiny app to investigate and evaluate multiply imputed data sets, 2) implement it in MICE in R, 3) write an instructional vignette, 4) write a technical paper on the workings and usage of the software aimed at e.g.
-the R Journal or Journal of Statistical Software.
-The expected output is a state-of-the-art shiny app that can find its way into MICE".
-
-**Bayes course on convergence:**
-```{r}
-# E. Assessing convergence
-# -----------------------------------------------------------------------------
-# History plot, autocorrelation plot
-# History plots show the sampled parameters over the iterations (excluding
-# the burn-in).
-# The development/pattern in these plots gives an indication of convergence.
-# When the history plot is stable (a fat catterpillar), convergence is reached.
-# Autocorrelation can be measured at many lags.
-# High autocorrelation indicates slow mixing of the random path.
-
-# mcmcplot(mcmcout = samples)
-# #sigma seems fine, but b doesn't:
-# #even at lag 5 there is still quite some autocorrelation: therefore we need to center the predictors!
-# #otherwise we could use a million iterations to deminish this effect
-
-# Gelman-Rubin diagnostic
-# The Gelman and Rubin statistic requires you run
-# the sampler/algorithm at least twice: These runs are
-# referred to as multiple chains.
-# It compares the variance between chains to the variance
-# within chains (G-R statistic = T/W = (pooled within chain
-# var + between chain var)/pooled within chain var.
-# It's the red line in the plot, and should be near 1.
-# See Gibbs sampler presentation (week 2)
-# https://drive.google.com/file/d/1ABHm8ala3c_puVvf32zF8h6TOZ3898uB/view
-# pdf p. 41, slide nr 29.
-# gelman.plot(samples)
-# #this appears to be okay with a really small deviation from 1
-
-# MC error (OPTIONAL?)
-# MC error = SD/sqrt(number of iterations)
-# SD represents the variation across iterations
-# MC error thus represents how much the means differ w.r.t. the iterations
-# MC error decreases as number of iterations increases.
-# It should not be larger than 5% of the sample standard deviation
-
-# History and density plot (OPTIONAL)
-# plot(samples)
-
-# Autocorrelation plots (OPTIONAL)
-# autocorr.plot(samples)
-
-# If parameters did not converge, you may:
-# . Use (many) more iterations
-# . Use a different parametrization (e.g., center predictors)
-# . Use different priors (e.g., multivariate normal prior (i.e.,
-#   dmnorm(,)) for parameters which are correlated)
-# . Use other initial values
-```
+Finally, the ShinyMICE package will be integrated into the existing MICE environment, and a vignette for applied researchers will be written.
 

### 2019-09-27 ThesisProposal.Rmd
Draft Proposal after feedback Gerko

@@
 
-Word count: 821
+Word count: 736
 
@@
 
-The R package MICE (Multiple Imputation using Chained Equations; [@mice11]) is the world-leading software for multiple imputation [@buur18].
-However, the package currently does not provide user-friendly means to evaluate multiply imputed data.
-*This is vital because of the amount of assumptions in MICE algorithms.
-Thanks to technological developments in computing speed and data storage it has become possible to check assumptions where that was unfeasible before.
-The very first assumption to check is algorithmic convergence; without convergence, all 'deeper' assumptions and subsequent inferences are invalid.*
+With MI, many assumptions are made about the nature of the observed and missing parts of the data, and their relation to the 'true' data generating model.
+Without proper evaluation of the imputations and the underlying assumptions, any drawn inference may erroneously be deemed valid (**source?**).
+Such evaluation measures are currently missing or under-developed in MI software.
+Therefore, I aim to answer the following question: 'Which measures are vital for evaluating the validity of multiply imputed data?'.
 
-*The research question central to this project is: 'How can empirical researchers be facilitated in evaluating the  validity of multiply imputed data?'.* Therefore, the goal is to develop a MICE evaluation suite, featuring interactive adaptations of existing modules (like plots to compare the incomplete and completed data sets), and novel assessment tools (like a measure to flag algorithmic non-convergence).
-The practical relevance of this research project is in facilitating applied researchers in making valid inferences despite of missingness.
-Simultaneously, it contributes to the scientific literature because there is no definitive way of diagnosing non-convergence for MI algorithms.
+The goal is to develop an MI evaluation suite for the world leading R package `MICE` [@mice11].
+It will feature novel assessment tools (e.g., a measure to flag algorithmic non-convergence), and  interactive adaptations of (partially) implemented modules (e.g., plots to compare the incomplete and completed data sets).
+This research project will aid applied researchers in drawing valid inference from incomplete data sets.
+Simultaneously, a contribution to the scientific literature is made by developing novel methodology and guidelines for evaluating MI data methods.
 
@@
 
-Since “there is no clear-cut method for determining when the MICE algorithm has converged” (@buur18, \textsection 6.5), users have to rely on visual inspection of the algorithm's iterations.
+**Elaborate: The most vital state to be evaluated is the convergence of the algorithm. Without convergence, any 'deeper' assumption and resulting inference is invalid.**
+
+Convergence properties of iterative MI algorithms are still under debate [@taka17]--with specific procedures like *predictive mean matching* posing an open question entirely [@murr18].
+<!-- "the convergence properties of FCS are currently under debate due to possible incompatibility (Li, Yu, and Rubin 2012; Zhu and Raghunathan 2015)" (@taka17, p. ) -->
+<!-- "The convergence properties of FCS in general settings is still mostly an open question.
+The behavior of FCS algorithms under non- or quasi-Bayesian imputation procedures like PMM is entirely an open question" (@murr18, p.19) -->
+@buur18 summarizes the state of the art as follows: "No method works best in all circumstances.
+The consensus is to assess convergence with a combination of tools.
+The added value of using a combination of convergence diagnostics for missing data imputation has not yet been systematically studied” (@buur18, \textsection 6.5).
+<!-- “Several expository reviews are available that assess convergence diagnostics for MCMC methods (Cowles and Carlin 1996; Brooks and Gelman 1998; El Adlouni, Favre, and Bobée 2006).
+Cowles and Carlin (1996) conclude that “automated convergence monitoring (as by a machine) is unsafe and should be avoided.” No method works best in all circumstances.
+The consensus is to assess convergence with a combination of tools.
+The added value of using a combination of convergence diagnostics for missing data imputation has not yet been systematically studied” (@buur18, \textsection 6.5).
+-->
+
+Currently, applied researchers have to rely on visual inspection of the algorithm's iterations [@buur18].
 Convergence is said to be reached when parameters (e.g., means of completed variables) are stable across iterations [@whit11].
@@
 <!-- “Convergence is diagnosed when the variance between different sequences is no larger than the variance within each individual sequence” (@buur18, \textsection 6.5).  -->
-*Conceptually, this visual evaluation procedure is similar to the 'Gelman-Rubin statistic' [@li14].
-Practically, however, Gelman and Rubin's convergence diagnostic $\widehat{R}$ [-@gelm92] cannot be applied directly to MI data.
-The measure produces too many 'false alarms': chains that converged within ten iterations are still flagged as non-converged at iteration fifty [@lace07].*
+Conceptually, the visual inspection procedure is equivalent to  Gelman, & Rubin's [-@gelm92] convergence statistic $\widehat{R}$ [@li14].
+In practice, however, $\widehat{R}$ cannot be applied directly to MI data, and is prone to producing false negatives [@lace07].
 <!-- "A common diagnostic tool is to plot one or more parameters against the iteration number and assess convergence by how different the variance between different sequences is relative to the variance within each individual sequence, similar to the Gelman-Rubin statistic (Gelman and Rubin, 1992) used in Markov chain Monte Carlo (MCMC) diagnostics" (@li14, \textsection 4.3).
@@
 
-Notwithstandingly, it seems like @su11 did implement $\widehat{R}$ as convergence criterion into their R package 'mi', including the conventional cut-off.
+<!-- Notwithstandingly, it seems like @su11 did implement $\widehat{R}$ as convergence criterion into their R package 'mi', including the conventional cut-off. -->
 <!-- "R.hat: The value of the $\hat{R}$ statistic used as a convergence criterion. The default is 1.1 (Gelman and Rubin 1992; Gelman, Carlin, Stern, and Rubin 2004)" (@su11 p. 4). -->
@@
 -->
-So it is worth investigating the validity of this convergence statistic in the context of multiple imputation.
+<!-- So it is worth investigating the validity of this convergence statistic in the context of multiple imputation.  -->
 
-The most recent literature on the topic states that convergence properties of MICE algorithms in general are still under debate [@taka17]--with specific procedures like 'predictive mean matching' posing entirely open questions [@murr18].
-<!-- "the convergence properties of FCS are currently under debate due to possible incompatibility (Li, Yu, and Rubin 2012; Zhu and Raghunathan 2015)" (@taka17, p. ) -->
-<!-- "The convergence properties of FCS in general settings is still mostly an open question.
-The behavior of FCS algorithms under non- or quasi-Bayesian imputation procedures like PMM is entirely an open question" (@murr18, p.19) -->
-@buur18 summarizes this problem as follows: "No method works best in all circumstances.
-The consensus is to assess convergence with a combination of tools.
-The added value of using a combination of convergence diagnostics for missing data imputation has not yet been systematically studied” (@buur18, \textsection 6.5).
-<!-- “Several expository reviews are available that assess convergence diagnostics for MCMC methods (Cowles and Carlin 1996; Brooks and Gelman 1998; El Adlouni, Favre, and Bobée 2006).
-Cowles and Carlin (1996) conclude that “automated convergence monitoring (as by a machine) is unsafe and should be avoided.” No method works best in all circumstances.
-The consensus is to assess convergence with a combination of tools.
-The added value of using a combination of convergence diagnostics for missing data imputation has not yet been systematically studied” (@buur18, \textsection 6.5).
--->
-And that is exactly the gap in the scientific literature that this thesis will contribute to.
+**Add: which other assumptions could be checked?
+The only available paper on evaluating MI data is @abay08.
+Ideally, we would want to study all possible combinations of variables: univariate, bivariate, etc.
+And to include both plots and stats.**
 
@@
 
-During this research project I will work directly with developers of the MICE R package.
-I will use R Shiny [@shiny17] to program an interactive evaluation device for the evaluation of multiply imputed data: 'ShinyMICE'.
-*This approach does not require the use of any empirical data.
-Therefore, the project does not need to be reviewed/approved by the faculty's ethical committee.*
+This research project will be supervised by the `MICE` developers.
+I will develop novel  methodology for evaluating MI data, and implement these methods in an interactive evaluation device in R Shiny [@shiny17]: '`ShinyMICE`'.
+This approach does not require the use of any empirical data.
+Therefore, the project is not subject to the school's ethical approval process.
 
@@
 
-The research report that is to be handed in December 13th 2019 will consist of an investigation into algorithmic convergence of MICE.
-Ideally, this will result in a single indicator to flag non-convergence (potentially based on $\widehat{R}$, auto-correlation, or MC error).
+The research report will consist of an investigation into algorithmic convergence of MI algorithms.
+Ideally, this will result in a single summary indicator to flag non-convergence (potentially based on $\widehat{R}$, auto-correlation, or MC error).
 
-The second stage of the research project is of more practical nature: programming ShinyMICE.
+In the second stage of the research project I will focus on other evaluation measures to implement in `ShinyMICE`.
 The application will at least consist of the following: 1) one or more measures to assess algorithmic convergence; 2) data visualizations (e.g., scatter-plots, densities, and cross-tabulations of the data pre- and post-imputation); and 3) statistical evaluation of relations between variables pre- versus post-imputation (i.e., $\chi^2$-tests or t-tests).
@@
 
-A working beta version of ShinyMICE will be considered sufficient to proceed to the next step: writing a technical paper on the workings and usage of the software.
-The preferred journal for publication of this manuscript is *Journal of Statistical Software* -- alternatively, publication in *The R Journal* will be attempted.
-
-Finally, the ShinyMICE package will be integrated into the existing MICE environment, and a vignette for applied researchers will be written.
+A working beta version of ShinyMICE will be considered a sufficient milestone to proceed with writing a technical paper on the methodology and the software.
+I aim to submit this for publication in *Journal of Statistical Software*.
+Finally, ShinyMICE will be integrated into the existing MICE environment, and a vignette for applied researchers will be written.
 

### 2019-09-27 ThesisProposal.Rmd
Update word count and docs file

@@
 
-Word count: 736
+Word count: 775
 
@@
 Simply ignoring the missingness or using ad hoc solutions can yield wildly invalid inferences [@buur18].
-Multiple imputation (MI; [@rubin87]) provides a framework to circumvent the *ubiquitous* problem of missing information.
+A framework to circumvent the *ubiquitous* problem of missing information is multiple imputation (MI; [@rubin87]).
 This technique -- that is growing in popularity [@buur18] -- entails 'guessing' the missing values in an incomplete data set several times.
-The variability between the resulting completed data sets represents how much uncertainty in the inferences is due to missingness [@rubin87].
+The variability between the completed data sets represents how much uncertainty in the inferences is due to missingness [@rubin87].
 
@@
 
-**Elaborate: The most vital state to be evaluated is the convergence of the algorithm. Without convergence, any 'deeper' assumption and resulting inference is invalid.**
+The numerous assumptions underlying MI algorithms can principally not be verified from the incomplete observed data (**source?**).
+Instead, imputers are  designated to evaluate the (plausibility of) multiply imputed data, and the robustness of results to the assumptions.
+The latter can be assessed by performing sensitivity analyses, see e.g.
+@nguy17 for practical guidelines.
+Methodology for evaluating the MI data however, is still largely missing.
+In the only available review of diagnostics, @abay08 overlook the most vital state to be evaluated: convergence of the algorithm.
+Without convergence, any 'deeper' assumption and resulting inference is invalid (**source?**).
+<!-- The weighting approach (a selection based method) [25,26], and the pattern-mixture approach [25,27] are two methods that have been proposed in the literature for conducting sensitivity analyses within the MI framework.
+(https://bmcmedresmethodol.biomedcentral.com/articles/10.1186/s12874-015-0022-1) -->
 
-Convergence properties of iterative MI algorithms are still under debate [@taka17]--with specific procedures like *predictive mean matching* posing an open question entirely [@murr18].
+While the convergence properties of iterative MI algorithms are still under debate [@taka17], specific procedures like *predictive mean matching* pose entirely open questions [@murr18].
 <!-- "the convergence properties of FCS are currently under debate due to possible incompatibility (Li, Yu, and Rubin 2012; Zhu and Raghunathan 2015)" (@taka17, p. ) -->
@@
 The behavior of FCS algorithms under non- or quasi-Bayesian imputation procedures like PMM is entirely an open question" (@murr18, p.19) -->
-@buur18 summarizes the state of the art as follows: "No method works best in all circumstances.
+@buur18 summarizes the issues with diagnosing convergence as follows: "No method works best in all circumstances.
 The consensus is to assess convergence with a combination of tools.
@@
 -->
-
-Currently, applied researchers have to rely on visual inspection of the algorithm's iterations [@buur18].
+Meanwhile, applied researchers have to rely on visual inspection of the algorithm's iterations [@buur18].
 Convergence is said to be reached when parameters (e.g., means of completed variables) are stable across iterations [@whit11].
@@
 -->
-
+The limitations in assessing convergence of MI algorithms may be overcome by combining $\widehat{R}$ with other diagnostics like auto-correlation and MC error.
 <!-- Notwithstandingly, it seems like @su11 did implement $\widehat{R}$ as convergence criterion into their R package 'mi', including the conventional cut-off. -->
@@
 **Add: which other assumptions could be checked?
-The only available paper on evaluating MI data is @abay08.
+The assumption of MAR.
 Ideally, we would want to study all possible combinations of variables: univariate, bivariate, etc.
@@
 
-The research report will consist of an investigation into algorithmic convergence of MI algorithms.
-Ideally, this will result in a single summary indicator to flag non-convergence (potentially based on $\widehat{R}$, auto-correlation, or MC error).
+The research report will consist of an investigation into algorithmic convergence of MI algorithms. Ideally, this will result in a single summary indicator to flag non-convergence.
 
@@
 The application will at least consist of the following: 1) one or more measures to assess algorithmic convergence; 2) data visualizations (e.g., scatter-plots, densities, and cross-tabulations of the data pre- and post-imputation); and 3) statistical evaluation of relations between variables pre- versus post-imputation (i.e., $\chi^2$-tests or t-tests).
-*This part of the thesis also entails research into user interface design, local versus cloud-based data storage, and optimizing the application's efficiency.*
 <!-- 1) comparing the imputed data to the observations, 2) inspecting the algorithmic convergence, 3) inspecting multivariate distributions, 4) the plausibility of the imputed data -->

### 2019-09-30 ThesisProposal_GV.Rmd
add suggestions

--- 2019-09-27 ThesisProposal.Rmd
@@
 
-The numerous assumptions underlying MI algorithms can principally not be verified from the incomplete observed data (**source?**).
+The numerous assumptions underlying MI algorithms can principally not be verified from the incomplete observed data **No source needed - we'd need the true data to verify**.
 Instead, imputers are  designated to evaluate the (plausibility of) multiply imputed data, and the robustness of results to the assumptions.
@@
 In the only available review of diagnostics, @abay08 overlook the most vital state to be evaluated: convergence of the algorithm.
-Without convergence, any 'deeper' assumption and resulting inference is invalid (**source?**).
+Without convergence, any 'deeper' assumption and resulting inference may be (anders erg sterk) invalid.
 <!-- The weighting approach (a selection based method) [25,26], and the pattern-mixture approach [25,27] are two methods that have been proposed in the literature for conducting sensitivity analyses within the MI framework.

### 2019-09-30 compare files.R

@@ -1,0 +1,5 @@
+modified_file <- system.file("ThesisProposal_GV.Rmd", package = "reviewer")
+reference_file <- system.file("ThesisProposal.Rmd", package = "reviewer")
+
+library(reviewer)
+result <- diff_rmd(modified_file, reference_file)

### 2019-09-30 ThesisProposal.Rmd
Notes MICE meeting

@@
 Simultaneously, a contribution to the scientific literature is made by developing novel methodology and guidelines for evaluating MI data methods.
+
+**conv te veel nadruk, weinig op diagnostiek, dat is veel sprekender.diag is ook wetenschap, meer modeling dan algorithmisch.
+in hoeverre duik ik de piepte om conv maat te maken.
+en aan de andere kant hoe grbruikers helpen om de diagn te maken in mice?
+bijv kijken of je overimp kunt gebruiken om te zien hoe goed het imp model is.
+--> double robustness/sensit.
+analys.**
 
@@
 I will develop novel  methodology for evaluating MI data, and implement these methods in an interactive evaluation device in R Shiny [@shiny17]: '`ShinyMICE`'.
-This approach does not require the use of any empirical data.
+This approach does not require the use of any unpublished empirical data.
 Therefore, the project is not subject to the school's ethical approval process.
+**or only published data/zonder bewerking van empirische data geen FETC/vrijelijk toegankelijke open access data bewerken vereist ook al FETC misschien?**
 

### 2019-10-01 ThesisProposal.Rmd
New format Lit. Rev. and integrated comments from Gerko

--- 2019-09-27 ThesisProposal.Rmd
@@
 At some point, any (social) scientist conducting statistical analyses will run into a missing data problem [@alli02].
-Simply ignoring the missingness or using ad hoc solutions can yield wildly invalid inferences [@buur18].
-A framework to circumvent the *ubiquitous* problem of missing information is multiple imputation (MI; [@rubin87]).
-This technique -- that is growing in popularity [@buur18] -- entails 'guessing' the missing values in an incomplete data set several times.
-The variability between the completed data sets represents how much uncertainty in the inferences is due to missingness [@rubin87].
+Missingness is problematic because statistical inference cannot be performed on incomplete data, and  ad hoc solutions can yield wildly invalid results [@buur18].
+To circumvent the *ubiquitous* problem of missing information, @rubin87 proposed the framework of multiple imputation (MI).
+MI is an iterative algorithmic procedure in which missing data points are 'guessed' several times.
+The variability between these 'guesses' validly reflects how much uncertainty in the inference is due to missing information--that is, if all statistical assumptions are met.
 
-With MI, many assumptions are made about the nature of the observed and missing parts of the data, and their relation to the 'true' data generating model.
-Without proper evaluation of the imputations and the underlying assumptions, any drawn inference may erroneously be deemed valid (**source?**).
+With MI, many assumptions are made about the nature of the observed and missing parts of the data, and their relation to the 'true' data generating model [@buur18].
+Without proper evaluation of the imputations and the underlying assumptions, any drawn inference may erroneously be deemed valid.
 Such evaluation measures are currently missing or under-developed in MI software.
@@
 
+<!-- **conv te veel nadruk, weinig op diagnostiek, dat is veel sprekender.diag is ook wetenschap, meer modeling dan algorithmisch.
+in hoeverre duik ik de piepte om conv maat te maken.
+en aan de andere kant hoe grbruikers helpen om de diagn te maken in mice?
+bijv kijken of je overimp kunt gebruiken om te zien hoe goed het imp model is.
+see double robustness or sensit.
+analys.** -->
+
 # Literature Review
 
-The numerous assumptions underlying MI algorithms can principally not be verified from the incomplete observed data (**source?**).
-Instead, imputers are  designated to evaluate the (plausibility of) multiply imputed data, and the robustness of results to the assumptions.
+By definition, the numerous assumptions inherent to MI methods cannot be verified from the data that is missing.
+
+- 1) assumption about the nature of the missingness (the missing data mechanism): that the 'cause' of the missingness is not missing, i.e.
+we can model the missingness from observed data [@rubin87].
+Robustness of the results to violations of this assumption can be assessed using sensitivity analyses.
+See @nguy17 for practical guidelines.
+
+What is missing? Implement in MICE with guidelines to interpret.
+
+- 2) assumption about the missing data model: that we can 'guess' missing values based on the distribution of the observed values and covariates [@buur18].
+This can be evaluated by evaluating plots of the distributions of the observed and imputed data.
+See review of diagnostics by @abay08.
+
+What is missing?
+To develop an interactive interface to perform analyses and plot the (in)complete data, use statistical tests to assess relations between variables, and use over-imputation or 'double robustness' to evaluate model fit.
+
+- 3) assumptions about the algorithm: that it has converged to a stable distribution [@buur18].
+This is a gap in the scientific literature.
+There is no consensus on the convergence properties of MI algorithms [@taka17].
+For some MI techniques it is not known whether a stable distribution can be reached at all [@murr18].
+It is vital therefore, to assess convergence after running an MI algorithm.
+Multiply imputed data does not conform to the assumptions of conventional measures to diagnose convergence--e.g., Gelman and Rubin's [-@gelm92] statistic $\widehat{R}$ [@lace07].
+Empirical researchers have to rely on a visual inspection procedure that is theoretically equivalant to $\widehat{R}$ [@whit11].
+In practice, however, visually assessing convergence is subjective, and difficult for the untrained eye.
+
+What is missing? A systematic study of convergence diagnostics for MI methods is missing [@buur18].
+
+<!-- Instead, imputers are  designated to evaluate the  -->
+<!-- The numerous assumptions underlying MI algorithms cannot be verified from incomplete observed data (**source!**). Because the As a proxy for assumption tests on the missing part of the data,  -->
+
+<!-- (plausibility of) multiply imputed data, and the robustness of results to the assumptions.
 The latter can be assessed by performing sensitivity analyses, see e.g.
@@
 In the only available review of diagnostics, @abay08 overlook the most vital state to be evaluated: convergence of the algorithm.
-Without convergence, any 'deeper' assumption and resulting inference is invalid (**source?**).
-<!-- The weighting approach (a selection based method) [25,26], and the pattern-mixture approach [25,27] are two methods that have been proposed in the literature for conducting sensitivity analyses within the MI framework.
-(https://bmcmedresmethodol.biomedcentral.com/articles/10.1186/s12874-015-0022-1) -->
+Without convergence, any 'deeper' assumption and resulting inference may be invalid.
+-->
+<!-- <!-- The weighting approach (a selection based method) [25,26], and the pattern-mixture approach [25,27] are two methods that have been proposed in the literature for conducting sensitivity analyses within the MI framework.
+(https://bmcmedresmethodol.biomedcentral.com/articles/10.1186/s12874-015-0022-1) --> -->
 
-While the convergence properties of iterative MI algorithms are still under debate [@taka17], specific procedures like *predictive mean matching* pose entirely open questions [@murr18].
-<!-- "the convergence properties of FCS are currently under debate due to possible incompatibility (Li, Yu, and Rubin 2012; Zhu and Raghunathan 2015)" (@taka17, p. ) -->
-<!-- "The convergence properties of FCS in general settings is still mostly an open question.
-The behavior of FCS algorithms under non- or quasi-Bayesian imputation procedures like PMM is entirely an open question" (@murr18, p.19) -->
-@buur18 summarizes the issues with diagnosing convergence as follows: "No method works best in all circumstances.
+<!-- While the convergence properties of iterative MI algorithms are still under debate [@taka17], specific procedures like *predictive mean matching* pose entirely open questions [@murr18]. -->
+<!-- <!-- "the convergence properties of FCS are currently under debate due to possible incompatibility (Li, Yu, and Rubin 2012; Zhu and Raghunathan 2015)" (@taka17, p. ) --> -->
+<!-- <!-- "The convergence properties of FCS in general settings is still mostly an open question.
+The behavior of FCS algorithms under non- or quasi-Bayesian imputation procedures like PMM is entirely an open question" (@murr18, p.19) --> -->
+<!-- @buur18 summarizes the issues with diagnosing convergence as follows: "No method works best in all circumstances.
+The consensus is to assess convergence with a combination of tools.
+The added value of using a combination of convergence diagnostics for missing data imputation has not yet been systematically studied (@buur18, \textsection 6.5).
+-->
+<!-- <!-- Several expository reviews are available that assess convergence diagnostics for MCMC methods (Cowles and Carlin 1996; Brooks and Gelman 1998; El Adlouni, Favre, and Bobée 2006).
+Cowles and Carlin (1996) conclude that automated convergence monitoring (as by a machine) is unsafe and should be avoided.
+No method works best in all circumstances.
 The consensus is to assess convergence with a combination of tools.
 The added value of using a combination of convergence diagnostics for missing data imputation has not yet been systematically studied” (@buur18, \textsection 6.5).
-<!-- “Several expository reviews are available that assess convergence diagnostics for MCMC methods (Cowles and Carlin 1996; Brooks and Gelman 1998; El Adlouni, Favre, and Bobée 2006).
-Cowles and Carlin (1996) conclude that “automated convergence monitoring (as by a machine) is unsafe and should be avoided.” No method works best in all circumstances.
-The consensus is to assess convergence with a combination of tools.
-The added value of using a combination of convergence diagnostics for missing data imputation has not yet been systematically studied” (@buur18, \textsection 6.5).
+--> -->
+<!-- Meanwhile, applied researchers have to rely on visual inspection of the algorithm's iterations [@buur18].
+Convergence is said to be reached when parameters (e.g., means of completed variables) are stable across iterations [@whit11].
 -->
-Meanwhile, applied researchers have to rely on visual inspection of the algorithm's iterations [@buur18].
-Convergence is said to be reached when parameters (e.g., means of completed variables) are stable across iterations [@whit11].
-<!-- "As MICE is an iterative procedure, it is important that convergence is achieved.
+<!-- <!-- "As MICE is an iterative procedure, it is important that convergence is achieved.
 This may be checked by computing, at each cycle, the means of imputed values and/or the values of regression coefficients, and seeing if they are stable" (@whit11, p.
 394).
+--> -->
+<!-- And additionally, the variation between imputation chains should be no larger than the variation within each individual chain [@buur18].  -->
+<!-- <!-- “Convergence is diagnosed when the variance between different sequences is no larger than the variance within each individual sequence” (@buur18, \textsection 6.5).  --> -->
+<!-- Conceptually, the visual inspection procedure is equivalent to  Gelman, & Rubin's [-@gelm92] convergence statistic $\widehat{R}$ [@li14].
+In practice, however, $\widehat{R}$ cannot be applied directly to MI data, and is prone to producing false negatives [@lace07].
 -->
-And additionally, the variation between imputation chains should be no larger than the variation within each individual chain [@buur18]. 
-<!-- “Convergence is diagnosed when the variance between different sequences is no larger than the variance within each individual sequence” (@buur18, \textsection 6.5).  -->
-Conceptually, the visual inspection procedure is equivalent to  Gelman, & Rubin's [-@gelm92] convergence statistic $\widehat{R}$ [@li14].
-In practice, however, $\widehat{R}$ cannot be applied directly to MI data, and is prone to producing false negatives [@lace07].
-<!-- "A common diagnostic tool is to plot one or more parameters against the iteration number and assess convergence by how different the variance between different sequences is relative to the variance within each individual sequence, similar to the Gelman-Rubin statistic (Gelman and Rubin, 1992) used in Markov chain Monte Carlo (MCMC) diagnostics" (@li14, \textsection 4.3).
--->
-The limitations in assessing convergence of MI algorithms may be overcome by combining $\widehat{R}$ with other diagnostics like auto-correlation and MC error.
-<!-- Notwithstandingly, it seems like @su11 did implement $\widehat{R}$ as convergence criterion into their R package 'mi', including the conventional cut-off. -->
-<!-- "R.hat: The value of the $\hat{R}$ statistic used as a convergence criterion. The default is 1.1 (Gelman and Rubin 1992; Gelman, Carlin, Stern, and Rubin 2004)" (@su11 p. 4). -->
-<!-- "Our mi offers two ways to check the convergence of the multiple imputation procedure.
+<!-- <!-- "A common diagnostic tool is to plot one or more parameters against the iteration number and assess convergence by how different the variance between different sequences is relative to the variance within each individual sequence, similar to the Gelman-Rubin statistic (Gelman and Rubin, 1992) used in Markov chain Monte Carlo (MCMC) diagnostics" (@li14, \textsection 4.3).
+--> -->
+<!-- The limitations in assessing convergence of MI algorithms may be overcome by combining $\widehat{R}$ with other diagnostics like auto-correlation and MC error. -->
+<!-- <!-- Notwithstandingly, it seems like @su11 did implement $\widehat{R}$ as convergence criterion into their R package 'mi', including the conventional cut-off. --> -->
+<!-- <!-- "R.hat: The value of the $\hat{R}$ statistic used as a convergence criterion. The default is 1.1 (Gelman and Rubin 1992; Gelman, Carlin, Stern, and Rubin 2004)" (@su11 p. 4). --> -->
+<!-- <!-- "Our mi offers two ways to check the convergence of the multiple imputation procedure.
 By default, mi() monitors the mixing of each variable by the variance of its mean and standard deviation within and between different chains of the imputation.
@@
 13).
--->
-<!-- So it is worth investigating the validity of this convergence statistic in the context of multiple imputation.  -->
+--> -->
+<!-- <!-- So it is worth investigating the validity of this convergence statistic in the context of multiple imputation.  --> -->
 
-**Add: which other assumptions could be checked?
+<!-- **Add: which other assumptions could be checked?
 The assumption of MAR.
 Ideally, we would want to study all possible combinations of variables: univariate, bivariate, etc.
-And to include both plots and stats.**
+And to include both plots and stats.** -->
 
@@
 This research project will be supervised by the `MICE` developers.
-I will develop novel  methodology for evaluating MI data, and implement these methods in an interactive evaluation device in R Shiny [@shiny17]: '`ShinyMICE`'.
-This approach does not require the use of any empirical data.
-Therefore, the project is not subject to the school's ethical approval process.
+I will develop novel methodology for evaluating MI data, and implement these methods using R Shiny [@shiny17].
+The endproduct is an interactive evaluation device: '`ShinyMICE`'.
+The R code and documentation will be open source (available on Github).
+Since this project does not require the use of unpublished empirical data, I expect that the FETC will grant this project the label 'exempt'.
 
@@
 
-The research report will consist of an investigation into algorithmic convergence of MI algorithms. Ideally, this will result in a single summary indicator to flag non-convergence.
+The research report will consist of an investigation into algorithmic convergence of MI algorithms.
+Ideally, this will result in a single summary indicator to flag non-convergence (potentially based on $\widehat{R}$, autocorrelation, or simulation error).
 

### 2019-10-02 ThesisProposal.Rmd
Lit. rev. rewritten based on MI assumptions

--- 2019-09-27 ThesisProposal.Rmd
@@
 author: "Hanne Oberman (4216318)"
-date: "`r Sys.Date()`"
+date: "`r format(Sys.time(), '%d %B %Y')`"
 output:
@@
 
-Word count: 775
+Word count: 711
 
@@
 
-At some point, any (social) scientist conducting statistical analyses will run into a missing data problem [@alli02].
-Simply ignoring the missingness or using ad hoc solutions can yield wildly invalid inferences [@buur18].
-A framework to circumvent the *ubiquitous* problem of missing information is multiple imputation (MI; [@rubin87]).
-This technique -- that is growing in popularity [@buur18] -- entails 'guessing' the missing values in an incomplete data set several times.
-The variability between the completed data sets represents how much uncertainty in the inferences is due to missingness [@rubin87].
+At some point, any (social) scientist conducting statistical analyses will run into a missing data problem ^[@alli02].
+Missingness is problematic because statistical inference cannot be performed on incomplete data, and  ad hoc solutions can yield wildly invalid results ^[@buur18].
+To circumvent the *ubiquitous* problem of missing information, @rubin87 proposed the framework of multiple imputation (MI).
+MI is an iterative algorithmic procedure in which missing data points are 'guessed' several times.
+The variability between these 'guesses' validly reflects how much uncertainty in the inference is due to missing information--that is, if all statistical assumptions are met ^[@rubin87].
 
-With MI, many assumptions are made about the nature of the observed and missing parts of the data, and their relation to the 'true' data generating model.
-Without proper evaluation of the imputations and the underlying assumptions, any drawn inference may erroneously be deemed valid (**source?**).
+With MI, many assumptions are made about the nature of the observed and missing parts of the data, and their relation to the 'true' data generating model ^[@buur18].
+Without proper evaluation of the imputations and the underlying assumptions, any drawn inference may erroneously be deemed valid.
 Such evaluation measures are currently missing or under-developed in MI software.
@@
 
-The goal is to develop an MI evaluation suite for the world leading R package `MICE` [@mice11].
-It will feature novel assessment tools (e.g., a measure to flag algorithmic non-convergence), and  interactive adaptations of (partially) implemented modules (e.g., plots to compare the incomplete and completed data sets).
+The goal is to develop an MI evaluation suite for the world leading R package `MICE` ^[@mice11].
 This research project will aid applied researchers in drawing valid inference from incomplete data sets.
@@
 
-The numerous assumptions underlying MI algorithms can principally not be verified from the incomplete observed data (**source?**).
-Instead, imputers are  designated to evaluate the (plausibility of) multiply imputed data, and the robustness of results to the assumptions.
-The latter can be assessed by performing sensitivity analyses, see e.g.
-@nguy17 for practical guidelines.
-Methodology for evaluating the MI data however, is still largely missing.
-In the only available review of diagnostics, @abay08 overlook the most vital state to be evaluated: convergence of the algorithm.
-Without convergence, any 'deeper' assumption and resulting inference is invalid (**source?**).
-<!-- The weighting approach (a selection based method) [25,26], and the pattern-mixture approach [25,27] are two methods that have been proposed in the literature for conducting sensitivity analyses within the MI framework.
-(https://bmcmedresmethodol.biomedcentral.com/articles/10.1186/s12874-015-0022-1) -->
+The numerous assumptions inherent to MI methods, can be roughly divided into assumptions about: 1) the nature of the missingness (*missingness mechanism*); 2)  the relation between observed and missing parts of the data (*missing data model*); and 3) the distribution of the missing values.
+By definition, assumptions cannot be verified on missing data.
+Therefore, proxy measures are used.
 
-While the convergence properties of iterative MI algorithms are still under debate [@taka17], specific procedures like *predictive mean matching* pose entirely open questions [@murr18].
-<!-- "the convergence properties of FCS are currently under debate due to possible incompatibility (Li, Yu, and Rubin 2012; Zhu and Raghunathan 2015)" (@taka17, p. ) -->
-<!-- "The convergence properties of FCS in general settings is still mostly an open question.
-The behavior of FCS algorithms under non- or quasi-Bayesian imputation procedures like PMM is entirely an open question" (@murr18, p.19) -->
-@buur18 summarizes the issues with diagnosing convergence as follows: "No method works best in all circumstances.
-The consensus is to assess convergence with a combination of tools.
-The added value of using a combination of convergence diagnostics for missing data imputation has not yet been systematically studied” (@buur18, \textsection 6.5).
-<!-- “Several expository reviews are available that assess convergence diagnostics for MCMC methods (Cowles and Carlin 1996; Brooks and Gelman 1998; El Adlouni, Favre, and Bobée 2006).
-Cowles and Carlin (1996) conclude that “automated convergence monitoring (as by a machine) is unsafe and should be avoided.” No method works best in all circumstances.
-The consensus is to assess convergence with a combination of tools.
-The added value of using a combination of convergence diagnostics for missing data imputation has not yet been systematically studied” (@buur18, \textsection 6.5).
--->
-Meanwhile, applied researchers have to rely on visual inspection of the algorithm's iterations [@buur18].
-Convergence is said to be reached when parameters (e.g., means of completed variables) are stable across iterations [@whit11].
-<!-- "As MICE is an iterative procedure, it is important that convergence is achieved.
-This may be checked by computing, at each cycle, the means of imputed values and/or the values of regression coefficients, and seeing if they are stable" (@whit11, p.
-394).
--->
-And additionally, the variation between imputation chains should be no larger than the variation within each individual chain [@buur18]. 
-<!-- “Convergence is diagnosed when the variance between different sequences is no larger than the variance within each individual sequence” (@buur18, \textsection 6.5).  -->
-Conceptually, the visual inspection procedure is equivalent to  Gelman, & Rubin's [-@gelm92] convergence statistic $\widehat{R}$ [@li14].
-In practice, however, $\widehat{R}$ cannot be applied directly to MI data, and is prone to producing false negatives [@lace07].
-<!-- "A common diagnostic tool is to plot one or more parameters against the iteration number and assess convergence by how different the variance between different sequences is relative to the variance within each individual sequence, similar to the Gelman-Rubin statistic (Gelman and Rubin, 1992) used in Markov chain Monte Carlo (MCMC) diagnostics" (@li14, \textsection 4.3).
--->
-The limitations in assessing convergence of MI algorithms may be overcome by combining $\widehat{R}$ with other diagnostics like auto-correlation and MC error.
-<!-- Notwithstandingly, it seems like @su11 did implement $\widehat{R}$ as convergence criterion into their R package 'mi', including the conventional cut-off. -->
-<!-- "R.hat: The value of the $\hat{R}$ statistic used as a convergence criterion. The default is 1.1 (Gelman and Rubin 1992; Gelman, Carlin, Stern, and Rubin 2004)" (@su11 p. 4). -->
-<!-- "Our mi offers two ways to check the convergence of the multiple imputation procedure.
-By default, mi() monitors the mixing of each variable by the variance of its mean and standard deviation within and between different chains of the imputation.
-If the $\hat{R}$ statistic is smaller than 1.1, (i.e., the difference of the within and between variance is trivial), the imputation is considered converged (Gelman, Carlin, Stern, and Rubin 2004).
-Additionally, by specifying mi(data, check.coef.convergence = TRUE, ...), users can check the convergence of the parameters of the conditional models" (@su11, p.
-13).
--->
-<!-- So it is worth investigating the validity of this convergence statistic in the context of multiple imputation.  -->
+The mechanism is assumed to be *missing at random*--i.e., the 'cause' of the missingness does not depend on the missing data itself ^[@rubin87].
+Evaluation is done via sensitivity analyses to assess the robustness of inferences to assumption violations.
+Some practical guidelines exist ^[e.g., @nguy17], but current MI software does provide empirical researchers with this methodology.
 
-**Add: which other assumptions could be checked?
-The assumption of MAR.
-Ideally, we would want to study all possible combinations of variables: univariate, bivariate, etc.
-And to include both plots and stats.**
+The *missing data model* assumption entails that missing values are distributed similar to observed values, after correcting for covariates ^[@buur18].
+Differences between observed and imputed data can be evaluated by plotting conditional distributions ^[@abay08].
+This procedure currently lacks an interactive interface and statistical tests to quantify the relations with covariates.
+Moreover, model fit could be assessed using *over-imputation* ^[@buur18] or *double robustness* ^[bang05].
+
+The final assumption is met when the algorithm has converged to a stable distribution of MI values ^[@buur18].
+Conventional measures to diagnose convergence--e.g., Gelman and Rubin's [-@gelm92] statistic $\widehat{R}$--do not suit multiply imputed data ^[@lace07].
+Therefore, empirical researchers have to rely on a visual inspection procedure that is theoretically equivalent to $\widehat{R}$ ^[@whit11].
+Not only is visually assessing convergence difficult for the untrained eye, it might also be futile.
+The convergence properties of MI algorithms lack scientific consensus ^[@taka17], and some default `MICE` techniques might not converge to stable distributions at all ^[@murr18].
+Convergence diagnostics for MI methods have not been systematically studied ^[@buur18].
 
@@
 This research project will be supervised by the `MICE` developers.
-I will develop novel  methodology for evaluating MI data, and implement these methods in an interactive evaluation device in R Shiny [@shiny17]: '`ShinyMICE`'.
-This approach does not require the use of any empirical data.
-Therefore, the project is not subject to the school's ethical approval process.
+I will develop novel methodology for evaluating MI data, and implement these methods using R Shiny ^[@shiny17].
+The endproduct is an interactive evaluation device: '`ShinyMICE`'.
+The R code and documentation will be open source (available on Github).
+Since this project does not require the use of unpublished empirical data, I expect that the FETC will grant this project the label 'exempt'.
 
-<!-- Develop a valid method to investigate the plausibility of multiply imputed data based on: -->
+The research report will consist of an investigation into algorithmic convergence of MI algorithms.
+I will replicate @lace07's simulation study on $\widehat{R}$, and potentially combine this convergence measure with auto-correlation or simulation error diagnostics.
+Ideally, I will develop a single summary indicator to flag non-convergence.
 
-<!-- - models (imputation and non-response) -->
+Subsequently, I will implement this and other evaluation measures in `ShinyMICE`.
+The application will at least contain modules for: sensitivity analyses; data visualizations (e.g., scatter-plots, densities, cross-tabulations); and statistical evaluations of relations between variables pre- versus post-imputation (i.e., $\chi^2$-tests or t-tests).
 
-<!-- - data features (cross-tabs, point estimates, aggregate statistics, etc.) -->
-
-<!-- - assumptions -->
-
-The research report will consist of an investigation into algorithmic convergence of MI algorithms. Ideally, this will result in a single summary indicator to flag non-convergence.
-
-In the second stage of the research project I will focus on other evaluation measures to implement in `ShinyMICE`.
-The application will at least consist of the following: 1) one or more measures to assess algorithmic convergence; 2) data visualizations (e.g., scatter-plots, densities, and cross-tabulations of the data pre- and post-imputation); and 3) statistical evaluation of relations between variables pre- versus post-imputation (i.e., $\chi^2$-tests or t-tests).
-<!-- 1) comparing the imputed data to the observations, 2) inspecting the algorithmic convergence, 3) inspecting multivariate distributions, 4) the plausibility of the imputed data -->
-
-A working beta version of ShinyMICE will be considered a sufficient milestone to proceed with writing a technical paper on the methodology and the software.
+A working beta version of `ShinyMICE` will be considered a sufficient milestone to proceed with writing a technical paper on the methodology and the software.
 I aim to submit this for publication in *Journal of Statistical Software*.

### 2019-10-02 ThesisProposal.Rmd
Added figure: Impression of ShinyMICE

@@
 bibliography: ThesisProposal.bib
+biblio-style: "apalike"
 ---
@@
 This procedure currently lacks an interactive interface and statistical tests to quantify the relations with covariates.
-Moreover, model fit could be assessed using *over-imputation* ^[@buur18] or *double robustness* ^[bang05].
+Moreover, model fit could be assessed using *over-imputation* ^[@buur18] or *double robustness* ^[@bang05].
 
@@
 
-Subsequently, I will implement this and other evaluation measures in `ShinyMICE`.
-The application will at least contain modules for: sensitivity analyses; data visualizations (e.g., scatter-plots, densities, cross-tabulations); and statistical evaluations of relations between variables pre- versus post-imputation (i.e., $\chi^2$-tests or t-tests).
+Subsequently, I will implement this and other evaluation measures in `ShinyMICE`, see Figure 1.
+The application will at least contain methodology for: sensitivity analyses; data visualizations (e.g., scatter-plots, densities, cross-tabulations); and statistical evaluations of relations between variables pre- versus post-imputation (i.e., $\chi^2$-tests or t-tests).
+
+![Preliminary impression of the interactive `ShinyMICE` user interface](Figures/Impression.png)
 

### 2019-10-03 ThesisProposal.Rmd
After feedback Gerko on lit. rev.

@@
 bibliography: ThesisProposal.bib
-biblio-style: "apalike"
+csl: acm.csl
+
 ---
@@
 
-At some point, any (social) scientist conducting statistical analyses will run into a missing data problem ^[@alli02].
-Missingness is problematic because statistical inference cannot be performed on incomplete data, and  ad hoc solutions can yield wildly invalid results ^[@buur18].
-To circumvent the *ubiquitous* problem of missing information, @rubin87 proposed the framework of multiple imputation (MI).
+At some point, any scientist conducting statistical analyses will run into a missing data problem [@alli02].
+Missingness is problematic because statistical inference cannot be performed on incomplete data, and  ad hoc solutions can yield wildly invalid results [@buur18].
+To circumvent the ubiquitous problem of missing information, Rubin [@rubin87] proposed the framework of multiple imputation (MI).
 MI is an iterative algorithmic procedure in which missing data points are 'guessed' several times.
-The variability between these 'guesses' validly reflects how much uncertainty in the inference is due to missing information--that is, if all statistical assumptions are met ^[@rubin87].
+The variability between these 'guesses' validly reflects how much uncertainty in the inference is due to missing information--that is, if all statistical assumptions are met [@rubin87].
 
-With MI, many assumptions are made about the nature of the observed and missing parts of the data, and their relation to the 'true' data generating model ^[@buur18].
+With MI, many assumptions are made about the nature of the observed and missing parts of the data, and their relation to the 'true' data generating model [@buur18].
 Without proper evaluation of the imputations and the underlying assumptions, any drawn inference may erroneously be deemed valid.
@@
 
-The goal is to develop an MI evaluation suite for the world leading R package `MICE` ^[@mice11].
+The goal is to develop an MI evaluation suite for the world leading R package `MICE` [@mice11].
 This research project will aid applied researchers in drawing valid inference from incomplete data sets.
@@
 
-The numerous assumptions inherent to MI methods, can be roughly divided into assumptions about: 1) the nature of the missingness (*missingness mechanism*); 2)  the relation between observed and missing parts of the data (*missing data model*); and 3) the distribution of the missing values.
+The numerous assumptions inherent to MI methods, can be roughly divided into assumptions about: 1) the *missingness mechanism*; 2)  the relation between observed and missing parts of the data (*imputation model*); and 3) the *non-response model*.
 By definition, assumptions cannot be verified on missing data.
@@
 
-The mechanism is assumed to be *missing at random*--i.e., the 'cause' of the missingness does not depend on the missing data itself ^[@rubin87].
-Evaluation is done via sensitivity analyses to assess the robustness of inferences to assumption violations.
-Some practical guidelines exist ^[e.g., @nguy17], but current MI software does provide empirical researchers with this methodology.
+<!-- **miss mech is model hoe obs and miss samenhangen, draait om ignorability =  op moment dat je obs kunt gebruiken om miss data te schatten.
+hoeft niet hetzelfde te zijn.
+relatie kunnen we uit de data halen, ondanks dat we de ontbrekende waaren zelf kunnen negeren, alleen het mech dat die waarden heeft gegenereerd.
+** -->
 
-The *missing data model* assumption entails that missing values are distributed similar to observed values, after correcting for covariates ^[@buur18].
-Differences between observed and imputed data can be evaluated by plotting conditional distributions ^[@abay08].
+The mechanism is assumed to be *missing at random*--i.e., the 'cause' of the missingness does not depend on the missing data itself [@rubin87].
+Evaluation is done via sensitivity analyses to assess the robustness of inferences to violations.
+Some practical guidelines exist (e.g., [@nguy17]), but current MI software does not provide empirical researchers with this methodology.
+
+The relation between observed and missing parts of the data can be evaluated by plotting conditional distributions [@abay08].
 This procedure currently lacks an interactive interface and statistical tests to quantify the relations with covariates.
-Moreover, model fit could be assessed using *over-imputation* ^[@buur18] or *double robustness* ^[@bang05].
+Moreover, model fit could be assessed using *over-imputation* [@buur18] or *double robustness* [@bang05].
 
-The final assumption is met when the algorithm has converged to a stable distribution of MI values ^[@buur18].
-Conventional measures to diagnose convergence--e.g., Gelman and Rubin's [-@gelm92] statistic $\widehat{R}$--do not suit multiply imputed data ^[@lace07].
-Therefore, empirical researchers have to rely on a visual inspection procedure that is theoretically equivalent to $\widehat{R}$ ^[@whit11].
+<!-- **er zijn 3 modellen: analyse, imputatie, non-response model (of missing data model).
+in mice wordt iedere var voorspeld uit andere vars.
+eerst analysemodel, is scientific mod of int.
+als je dat niet meeneemt dan mis je ruimte in alg om datgene dat je probeert te vinden mee te nemen (non-comp).
+non-resp is verklaren waarom missings in een var vaker voorkomen dan andere vars.
+meng's 1994 paper on compatibility and congeneality.** -->
+
+The final assumption is met when the algorithm has converged to a stable distribution of MI values [@buur18].
+Conventional measures to diagnose convergence--e.g., Gelman and Rubin's [-@gelm92] statistic $\widehat{R}$--do not suit multiply imputed data [@lace07].
+Therefore, empirical researchers have to rely on a visual inspection procedure that is theoretically equivalent to $\widehat{R}$ [@whit11].
 Not only is visually assessing convergence difficult for the untrained eye, it might also be futile.
-The convergence properties of MI algorithms lack scientific consensus ^[@taka17], and some default `MICE` techniques might not converge to stable distributions at all ^[@murr18].
-Convergence diagnostics for MI methods have not been systematically studied ^[@buur18].
+The convergence properties of MI algorithms lack scientific consensus [@taka17], and some default `MICE` techniques might not converge to stable distributions at all [@murr18].
+Convergence diagnostics for MI methods have not been systematically studied [@buur18].
 
@@
 This research project will be supervised by the `MICE` developers.
-I will develop novel methodology for evaluating MI data, and implement these methods using R Shiny ^[@shiny17].
+I will develop novel methodology for evaluating MI data, and implement these methods using R Shiny [@shiny17].
 The endproduct is an interactive evaluation device: '`ShinyMICE`'.
@@
 The research report will consist of an investigation into algorithmic convergence of MI algorithms.
-I will replicate @lace07's simulation study on $\widehat{R}$, and potentially combine this convergence measure with auto-correlation or simulation error diagnostics.
+I will replicate Lacerda et al.'s simulation study on $\widehat{R}$ [@lace07], and potentially combine this convergence measure with auto-correlation or simulation error diagnostics.
 Ideally, I will develop a single summary indicator to flag non-convergence.
@@
 Subsequently, I will implement this and other evaluation measures in `ShinyMICE`, see Figure 1.
-The application will at least contain methodology for: sensitivity analyses; data visualizations (e.g., scatter-plots, densities, cross-tabulations); and statistical evaluations of relations between variables pre- versus post-imputation (i.e., $\chi^2$-tests or t-tests).
+The application will at least contain methodology for: sensitivity analyses; data visualizations (e.g., scatter-plots, densities, cross-tabulations); and statistical evaluations of relations between variables pre- versus post-imputation (i.e., $\chi^2$-tests or $t$-tests).
 
-![Preliminary impression of the interactive `ShinyMICE` user interface](Figures/Impression.png)
+![Preliminary impression of the interactive `ShinyMICE` user interface.](Figures/Impression.png)
 

### 2019-10-07 ThesisProposal.Rmd
Redid the lit rev again to incorporate meng and rubin on assumptions (before research seminar feedback).

@@
 
-Word count: 711
+Word count: 739
 
@@
 
-The numerous assumptions inherent to MI methods, can be roughly divided into assumptions about: 1) the *missingness mechanism*; 2)  the relation between observed and missing parts of the data (*imputation model*); and 3) the *non-response model*.
-By definition, assumptions cannot be verified on missing data.
-Therefore, proxy measures are used.
+The validity of MI data depends on numerous assumptions that--by definition--cannot be verified from the incomplete data.
+Therefore, existing evaluation methods rely on proxy measures.
+For the following assumtions, no reliable proxy measures have been proposed and/or implemented in `MICE`: 1) *ignorability* of the *missingness mechanism*; 2) suitability of the *imputation models*; and 3) *compatibility* with the MI algorithm.
+<!-- Rubin (1987b, 160-66) subdivided the work needed to create imputations into three tasks.
+The modeling task chooses a specific model for the data, the estimation task formulates the posterior parameters distribution given the model and the imputation task takes a random draws for the missing data by drawing successively from parameter and data distributions @buur18 par 4.5.1 -->
+
+<!-- account for the process that created the missing data, -->
+<!-- preserve the relations in the data, and -->
+<!-- preserve the uncertainty about these relations. -->
+
+<!-- The modeling task chooses a specific model for the data -> assumes ignorability. -->
+
+<!-- The estimation task computes the posterior distribution of the model parameters of the assumes correct specification of the model. ->
+
+<!-- The imputation task draws plausible values from the conditional posterior distribution of the missing data given the estimated parameters -> assumes convergence in distribution -->
+
+1.
+The missingness mechanism is said to be ignorable when the 'cause' of the missingness does not depend on the missing data itself [@rubin87].
+Evaluation is usually done with sensitivity analyses to assess the robustness of inferences to violations.
+Some practical guidelines exist (e.g., [@nguy17]), but current MI software does not facilitate this methodology for empirical researchers.
 
@@
 
-The mechanism is assumed to be *missing at random*--i.e., the 'cause' of the missingness does not depend on the missing data itself [@rubin87].
-Evaluation is done via sensitivity analyses to assess the robustness of inferences to violations.
-Some practical guidelines exist (e.g., [@nguy17]), but current MI software does not provide empirical researchers with this methodology.
-
-The relation between observed and missing parts of the data can be evaluated by plotting conditional distributions [@abay08].
-This procedure currently lacks an interactive interface and statistical tests to quantify the relations with covariates.
-Moreover, model fit could be assessed using *over-imputation* [@buur18] or *double robustness* [@bang05].
+2.
+The imputation models should capture the relations between observed and missing parts of the data.
+These relations can be evaluated by plotting conditional distributions [@abay08].
+Such visualizations are available in `MICE`, but subsequent statistical tests to quantify the relations with covariates are not yet provided.
+Additionally, there is potential to assess model fit by means of *over-imputation* [@buur18] or *double robustness* [@bang05].
 
@@
 
-The final assumption is met when the algorithm has converged to a stable distribution of MI values [@buur18].
+3.
+The third assumption is met when the MI algorithm converges to a stable distribution.
 Conventional measures to diagnose convergence--e.g., Gelman and Rubin's [-@gelm92] statistic $\widehat{R}$--do not suit multiply imputed data [@lace07].
@@
 The convergence properties of MI algorithms lack scientific consensus [@taka17], and some default `MICE` techniques might not converge to stable distributions at all [@murr18].
-Convergence diagnostics for MI methods have not been systematically studied [@buur18].
+Moreover, convergence diagnostics for MI methods have not been systematically studied [@buur18].
+
+In short, the existing literature provides both possibilities and limitations to evaluating the validity of multiply imputed data. My aim is to (further) develop and implement the methodology.
 
@@
 I aim to submit this for publication in *Journal of Statistical Software*.
-Finally, ShinyMICE will be integrated into the existing MICE environment, and a vignette for applied researchers will be written.
+Finally, `ShinyMICE` will be integrated into the existing `MICE` environment, and a vignette for applied researchers will be written.
 

### 2019-10-08 ThesisProposal.Rmd
After feedback research seminar mentors and peers (reorganized the intro and approach sections, removed some typos)

@@
 To circumvent the ubiquitous problem of missing information, Rubin [@rubin87] proposed the framework of multiple imputation (MI).
-MI is an iterative algorithmic procedure in which missing data points are 'guessed' several times.
-The variability between these 'guesses' validly reflects how much uncertainty in the inference is due to missing information--that is, if all statistical assumptions are met [@rubin87].
+MI is an iterative algorithmic procedure in which missing data points are 'guessed' (i.e.
+imputed) several times.
+The variability between the imputations validly reflects how much uncertainty in the inference is due to missing information--that is, if all statistical assumptions are met [@rubin87].
 
-With MI, many assumptions are made about the nature of the observed and missing parts of the data, and their relation to the 'true' data generating model [@buur18].
+With MI, many assumptions are made about the nature of the observed and missing parts of the data **[remove "and their relation to the 'true' data generating mode"?l** [@buur18].
 Without proper evaluation of the imputations and the underlying assumptions, any drawn inference may erroneously be deemed valid.
-Such evaluation measures are currently missing or under-developed in MI software.
-Therefore, I aim to answer the following question: 'Which measures are vital for evaluating the validity of multiply imputed data?'.
-
-The goal is to develop an MI evaluation suite for the world leading R package `MICE` [@mice11].
-This research project will aid applied researchers in drawing valid inference from incomplete data sets.
-Simultaneously, a contribution to the scientific literature is made by developing novel methodology and guidelines for evaluating MI data methods.
+Such evaluation measures are currently missing or under-developed in MI software, like the world leading `R` package `MICE` [@mice11].
+Therefore, I will answer the following question: 'Which measures are vital for evaluating the validity of multiply imputed data?'.
 
@@
 The validity of MI data depends on numerous assumptions that--by definition--cannot be verified from the incomplete data.
-Therefore, existing evaluation methods rely on proxy measures.
-For the following assumtions, no reliable proxy measures have been proposed and/or implemented in `MICE`: 1) *ignorability* of the *missingness mechanism*; 2) suitability of the *imputation models*; and 3) *compatibility* with the MI algorithm.
+Consequently, existing evaluation methods rely on proxy measures.
+For the following assumptions, no reliable proxy measures have been proposed and/or implemented: 1) *ignorability* of the *missingness mechanism*; 2) suitability of the *imputation models*; and 3) *compatibility* with the MI algorithm.
 <!-- Rubin (1987b, 160-66) subdivided the work needed to create imputations into three tasks.
@@
 1.
-The missingness mechanism is said to be ignorable when the 'cause' of the missingness does not depend on the missing data itself [@rubin87].
-Evaluation is usually done with sensitivity analyses to assess the robustness of inferences to violations.
+A missingness mechanism is said to be ignorable when the 'cause' of the missingness does not depend on the missing data itself [@rubin87].
+Evaluation of this assumption is usually done with sensitivity analyses to assess the robustness of inferences to violations.
 Some practical guidelines exist (e.g., [@nguy17]), but current MI software does not facilitate this methodology for empirical researchers.
@@
 2.
-The imputation models should capture the relations between observed and missing parts of the data.
+Suitable imputation models capture the relations between observed and missing parts of the data.
 These relations can be evaluated by plotting conditional distributions [@abay08].
@@
 The third assumption is met when the MI algorithm converges to a stable distribution.
-Conventional measures to diagnose convergence--e.g., Gelman and Rubin's [-@gelm92] statistic $\widehat{R}$--do not suit multiply imputed data [@lace07].
+However, conventional measures to diagnose convergence--e.g., Gelman and Rubin's [-@gelm92] statistic $\widehat{R}$--do not suit multiply imputed data [@lace07].
 Therefore, empirical researchers have to rely on a visual inspection procedure that is theoretically equivalent to $\widehat{R}$ [@whit11].
@@
 
-In short, the existing literature provides both possibilities and limitations to evaluating the validity of multiply imputed data. My aim is to (further) develop and implement the methodology.
+In short, the existing literature provides both possibilities and limitations to evaluating the validity of multiply imputed data.
+The goal of this research project is to develop novel methodology and guidelines for evaluating MI methods, and implement these in an interactive evaluation device for multiple imputation.
+This tool will aid applied researchers in drawing valid inference from incomplete datasets.
 
-# Methods
+# Approach
 
-This research project will be supervised by the `MICE` developers.
-I will develop novel methodology for evaluating MI data, and implement these methods using R Shiny [@shiny17].
-The endproduct is an interactive evaluation device: '`ShinyMICE`'.
-The R code and documentation will be open source (available on Github).
-Since this project does not require the use of unpublished empirical data, I expect that the FETC will grant this project the label 'exempt'.
+Initially, the research project will consist of an investigation into algorithmic convergence of MI algorithms.
+I will replicate Lacerda et al.'s simulation study on $\widehat{R}$ [@lace07], and develop novel guidelines for assessing convergence.
+Ideally, I will integrate several diagnostics (e.g., $\widehat{R}$, *auto-correlation*, and *simulation error*) into a single summary indicator to flag non-convergence.
 
-The research report will consist of an investigation into algorithmic convergence of MI algorithms.
-I will replicate Lacerda et al.'s simulation study on $\widehat{R}$ [@lace07], and potentially combine this convergence measure with auto-correlation or simulation error diagnostics.
-Ideally, I will develop a single summary indicator to flag non-convergence.
-
-Subsequently, I will implement this and other evaluation measures in `ShinyMICE`, see Figure 1.
+Subsequently, I will use `R Shiny` [@shiny17] to implement the convergence indicator and other evaluation measures in `ShinyMICE`, see Figure 1.
 The application will at least contain methodology for: sensitivity analyses; data visualizations (e.g., scatter-plots, densities, cross-tabulations); and statistical evaluations of relations between variables pre- versus post-imputation (i.e., $\chi^2$-tests or $t$-tests).
@@
 A working beta version of `ShinyMICE` will be considered a sufficient milestone to proceed with writing a technical paper on the methodology and the software.
-I aim to submit this for publication in *Journal of Statistical Software*.
+I will submit the paper for publication in *Journal of Statistical Software*.
 Finally, `ShinyMICE` will be integrated into the existing `MICE` environment, and a vignette for applied researchers will be written.
 
+The R code and documentation of this project will be open source (available on Github).
+Since the study does not require the use of unpublished empirical data, I expect that the FETC will grant the label 'exempt'.
+
 # References

### 2019-10-10 ThesisProposal.Rmd
Revised after feedback Gerko on lit rev

@@
 output:
-  #word_document: default
+  #word_document: 
   pdf_document: 
@@
 
-Word count: 739
+Word count: 713
 
@@
 
-With MI, many assumptions are made about the nature of the observed and missing parts of the data **[remove "and their relation to the 'true' data generating mode"?l** [@buur18].
+With MI, many assumptions are made about the nature of the observed and missing parts of the data and their relation to the 'true' data generating model [@buur18].
 Without proper evaluation of the imputations and the underlying assumptions, any drawn inference may erroneously be deemed valid.
@@
 
-The validity of MI data depends on numerous assumptions that--by definition--cannot be verified from the incomplete data.
-Consequently, existing evaluation methods rely on proxy measures.
-For the following assumptions, no reliable proxy measures have been proposed and/or implemented: 1) *ignorability* of the *missingness mechanism*; 2) suitability of the *imputation models*; and 3) *compatibility* with the MI algorithm.
+The validity of the MI solution depends on numerous assumptions that cannot be verified from the observed data alone.
+Consequently, existing evaluation methods rely on **[rephrase: proxy measures]**.
+For the following assumptions, no reliable *proxy measures* have been proposed and/or implemented: 1) *ignorability* of the *missingness mechanism*; 2) congruity of the *imputation models*; and 3) *compatibility* of the MI modeling procedure.
 <!-- Rubin (1987b, 160-66) subdivided the work needed to create imputations into three tasks.
@@
 1.
-A missingness mechanism is said to be ignorable when the 'cause' of the missingness does not depend on the missing data itself [@rubin87].
+A missingness mechanism is said to be ignorable when the sprobability to be missing does not depend on the missing data itself [@rubin87].
 Evaluation of this assumption is usually done with sensitivity analyses to assess the robustness of inferences to violations.
@@
 2.
-Suitable imputation models capture the relations between observed and missing parts of the data.
-These relations can be evaluated by plotting conditional distributions [@abay08].
-Such visualizations are available in `MICE`, but subsequent statistical tests to quantify the relations with covariates are not yet provided.
+Congruent imputation models capture all required relations between observed and missing parts of the data.
+The exent to which this has been successful can be evaluated by plotting conditional distributions [@abay08].
+Such visualizations are available in `MICE`, but subsequent statistical tests to quantify the relations with covariates are not provided.
 Additionally, there is potential to assess model fit by means of *over-imputation* [@buur18] or *double robustness* [@bang05].
@@
 The third assumption is met when the MI algorithm converges to a stable distribution.
-However, conventional measures to diagnose convergence--e.g., Gelman and Rubin's [-@gelm92] statistic $\widehat{R}$--do not suit multiply imputed data [@lace07].
-Therefore, empirical researchers have to rely on a visual inspection procedure that is theoretically equivalent to $\widehat{R}$ [@whit11].
-Not only is visually assessing convergence difficult for the untrained eye, it might also be futile.
+However, conventional measures to diagnose convergence-- e.g., Gelman and Rubin's [-@gelm92] statistic $\widehat{R}$ --are not applicable on multiply imputed data [@lace07].
+Therefore, empirical researchers have to rely on visual inspection procedures that are theoretically equivalent to $\widehat{R}$ [@whit11].
+Visually assessing convergence is not only difficult to the untrained eye, it might also be futile.
 The convergence properties of MI algorithms lack scientific consensus [@taka17], and some default `MICE` techniques might not converge to stable distributions at all [@murr18].
@@
 In short, the existing literature provides both possibilities and limitations to evaluating the validity of multiply imputed data.
-The goal of this research project is to develop novel methodology and guidelines for evaluating MI methods, and implement these in an interactive evaluation device for multiple imputation.
-This tool will aid applied researchers in drawing valid inference from incomplete datasets.
+The goal of this research project is to develop novel methodology and guidelines for evaluating MI methods, and implement these in an interactive evaluation framework for multiple imputation.
+This framework will aid applied researchers in drawing valid inference from incomplete datasets.
 

### 2019-10-16 ResearchReport.Rmd
Final comments Gerko&Stef. Included logo on front page.
Final version thesis proposal

@@
 
-- It is the first ‘half’ of the thesis, i.e., there are no results included yet, but the report contains a full introduction including a literature review and a methods section that contains details about the data, instruments and/or statistical procedures.
+- It is the first half of the thesis, i.e., there are no results included yet, but the report contains a full introduction including a literature review and a methods section that contains details about the data, instruments and/or statistical procedures.
 There is no (absolute) word limit but the length of the report should comply with recommendations/expectations of the proposed journal.
@@
 
-- Potential convergence measures:
+Replicate @lace07: chains that converged within ten iterations are still flagged as non-converged at iteration fifty [@lace07].
 
-*why do people say the GR stat will give false positives?
+"The monitoring statistic was computed for mean monthly earnings at each iteration in chains of length k = 200.
+Since calculation of the statistic requires M parallel sequences, m = 5 such chains were constructed.
+This value of m was informed by the preferred choice given in the literature on multiple imputation.
+The monitoring statistic computed at each iteration is presented in Figures 15 to 17 for each of the three missingness mechanisms.
+The red vertical line denotes ten iterations." [@lace07, p.
+49]
+
+![](Figures/lace07.png)
+
+- Notes MICE meetings:
+
+Why do people say the GR stat will give false positives?
 perhaps the window (nr of it) is too small.
@@
 possible approach: replicate and build a decision rule to solve the problem.
-*
 
-*in mice algorithms, iterating is very fast compared to regular MCMC.
+In mice algorithms, iterating is very fast compared to regular MCMC.
 therefore, we could use the average over the last 5 iterations you'll take the 'majority vote' to indicate conv.
@@
 patrick ...
-and ian white work on stata proc mi also work on convergence diagnostics --> autocorr function plot in sas of worst linear function, and likelihood ratio that you could save.*
+and ian white work on stata proc mi also work on convergence diagnostics --> autocorr function plot in sas of worst linear function, and likelihood ratio that you could save.
 
@@
 
+**quick win: xyplot and hist have all lattice functions (all sorts of scatterplots!)**
+
 **Potentially add something about updated (2019) version of $\widehat{R}$ and the new threshold of 1.01, see @veht19. Or leave it for the Research Report.**
 
-"For models based on multiple imputed data sets, this [R hat > 1.1] is often a false positive: Chains of different submodels may not overlay each other exactly, since there were fitted to different data" (see https://cran.r-project.org/web/packages/brms/vignettes/brms_missings.html).
+"For models based on multiple imputed data sets, this [R hat \> 1.1] is often a false positive: Chains of different submodels may not overlay each other exactly, since there were fitted to different data" (see https://cran.r-project.org/web/packages/brms/vignettes/brms_missings.html).
 

### 2019-10-21 ResearchReport.Rmd
Start Research Report, and JSS style

@@
 author: "Hanne Oberman (4216318)"
-date: "`r Sys.Date()`"
+date: "`r format(Sys.time(), '%d %B %Y')`"
 output: 
@@
 bibliography: ThesisProposal.bib
+
 ---
@@
 \centering
+
+![](Figures/logo.png)
 
@@
 
+\newpage
+
 \raggedright
 
-Goals:
+# Requirements
+
+## Goals:
 
@@
 
-Grading:
+## Grading:
 
@@
 
-Format (either/or):
+## Format (either/or):
 
@@
 
+**From thesis proposal**:
+
+At some point, any scientist conducting statistical analyses will run into a missing data problem [@alli02].
+Missingness is problematic because statistical inference cannot be performed on incomplete data, and  ad hoc solutions can yield wildly invalid results [@buur18].
+To circumvent the ubiquitous problem of missing information, Rubin [@rubin87] proposed the framework of multiple imputation (MI).
+MI is an iterative algorithmic procedure in which missing data points are 'guessed' (i.e.
+imputed) several times.
+The variability between the imputations validly reflects how much uncertainty in the inference is due to missing information--that is, if all statistical assumptions are met [@rubin87].
+
+With MI, many assumptions are made about the nature of the observed and missing parts of the data and their relation to the 'true' *data generating model* [@buur18].
+Without proper evaluation of the imputations and the underlying assumptions, any drawn inference may erroneously be deemed valid.
+Such evaluation measures are currently missing or under-developed in MI software, like the world leading `R` package `MICE` [@mice11].
+Therefore, I will answer the following question: 'Which measures are vital for evaluating the validity of multiply imputed data?'.
+
 # Theoretical Background
@@
 
+**From thesis proposal**:
+
+The validity of the MI solution depends on numerous assumptions that cannot be verified from the observed data alone.
+So instead of statistical tests for assumptions, evaluation procedures have been developed.
+For the following assumptions, no reliable procedure has been proposed and/or implemented: 1) *ignorability* of the *missingness mechanism* [@rubin87]; 2) *congeneality* of the imputation models [@meng94]; and 3) *compatibility* of the MI modeling procedure [@rubin96].
+
+1.
+A missingness mechanism is said to be ignorable when the probability to be missing does not depend on the missing data itself.
+Violation of this assumption can gravely affect inferences.
+Robustness of inferences to varying degrees of violation can be assessed with sensitivity analyses.
+Some practical guidelines exist (e.g., [@nguy17]), but current MI software does not facilitate this methodology for empirical researchers.
+
+2.
+Congeneal imputation models capture all required relations between observed and missing parts of the data.
+The extent to which this has been successful can be evaluated by plotting conditional distributions [@abay08].
+Such visualizations are available in `MICE`, but subsequent statistical tests to quantify the relations with covariates are not provided.
+^[Additionally, there is potential to assess model fit by means of *over-imputation* [@buur18] or *double robustness* [@bang05]-- topics that are only persued if time permits.]
+
+3.
+The third assumption is met when the MI algorithm converges to a stable distribution.
+However, conventional measures to diagnose convergence-- e.g., Gelman and Rubin's [-@gelm92] statistic $\widehat{R}$ --are not applicable on multiply imputed data [@lace07].
+Therefore, empirical researchers have to rely on visual inspection procedures that are theoretically equivalent to $\widehat{R}$ [@whit11].
+Visually assessing convergence is not only difficult to the untrained eye, it might also be futile.
+The convergence properties of MI algorithms lack scientific consensus [@taka17], and some default `MICE` techniques might not converge to stable distributions at all [@murr18].
+Moreover, convergence diagnostics for MI methods have not been systematically studied [@buur18].
+
+In short, the existing literature provides both possibilities and limitations to evaluating the validity of multiply imputed data.
+The goal of this research project is to develop novel methodology and guidelines for evaluating MI methods, and implement these in an interactive evaluation framework for multiple imputation.
+This framework will aid applied researchers in drawing valid inference from incomplete datasets.
+
 # Methods
+
+**From thesis proposal**:
+
+Initially, the research project will consist of an investigation into algorithmic convergence of MI algorithms.
+I will replicate Lacerda et al.'s simulation study on $\widehat{R}$ [@lace07], and develop novel guidelines for assessing convergence.
+Ideally, I will integrate several diagnostics (e.g., $\widehat{R}$, *auto-correlation*, and *simulation error*) into a single summary indicator to flag non-convergence.
+
+Subsequently, I will use `R Shiny` [@shiny17] to implement the convergence indicator and existing evaluation measures in `ShinyMICE`, see Figure 1.
+The application will at least contain methodology for: sensitivity analyses; data visualizations (e.g., scatter-plots, densities, cross-tabulations); and statistical evaluation of relations between variables pre- versus post-imputation (i.e., $\chi^2$-tests or $t$-tests).
+![Preliminary impression of the interactive `ShinyMICE` user interface.](Figures/Impression.png)
+
+A working beta version of `ShinyMICE` will be considered a sufficient milestone to proceed with writing a technical paper on the methodology and the software.
+I will submit the paper for publication in *Journal of Statistical Software*.
+Finally, `ShinyMICE` will be integrated into the existing `MICE` environment, and a vignette for applied researchers will be written.
+
+The R code and documentation of this project will be open source (available on Github).
+Since the study does not require the use of unpublished empirical data, I expect that the FETC will grant the label 'exempt'.
 
@@
 but we have very rich data so we're not estimating everything from scratch.
-not: we're talking about missing data only, not the combined data (that autocorr.
+note: we're talking about missing data only, not the combined data (that autocorr.
 is very high).
@@
 data that should not be a problem because that's constant.
-maybe look into allpying pca on the imputed data and if that stays the same we know that the means and variances are stable too.
+maybe look into applying pca on the imputed data and if that stays the same we know that the means and variances are stable too.
 patrick ...

### 2019-10-24 Research Report/article.Rnw
New folder structure for research report

\documentclass[article]{jss}

%% -- LaTeX packages and custom commands ---------------------------------------

%% recommended packages
\usepackage{thumbpdf,lmodern}

%% another package (only for this demo article)
% \usepackage{framed}

%% new custom commands
\newcommand{\class}[1]{`\code{#1}'}
\newcommand{\fct}[1]{\code{#1()}}

%% For Sweave-based articles about R packages:
%% need no \usepackage{Sweave}
\SweaveOpts{engine=R, eps=FALSE, keep.source = TRUE}
<<preliminaries, echo=FALSE, results=hide>>=
options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
library("MASS")
@

%% -- Article metainformation (author, title, ...) -----------------------------

%% - \author{} with primary affiliation
%% - \Plainauthor{} without affiliations
%% - Separate authors by \And or \AND (in \author) or by comma (in \Plainauthor).
%% - \AND starts a new line, \And does not.
\author{Hanne Oberman\\Utrecht University}
   % \And Second Author\\Plus Affiliation}
\Plainauthor{Hanne Oberman}

%% - \title{} in title case
%% - \Plaintitle{} without LaTeX markup (if any)
%% - \Shorttitle{} with LaTeX markup (if any), used as running title
\title{\pkg{ShinyMICE}: an Evaluation Suite for Multiple Imputation}
\Plaintitle{ShinyMICE: an Evaluation Suite for Multiple Imputation}
\Shorttitle{\pkg{ShinyMICE}: an Evaluation Suite for Multiple Imputation}

%% - \Abstract{} almost as usual
\Abstract{
  This Research Report contains the Introduction and Methods section of the technical paper that will be submitted for publication in \emph{Journal of Statistical Software}. ("There is no page limit, nor a limit on the number of figures or tables", see \url{https://www.jstatsoft.org/pages/view/authors}.) I have chosen to use the second format from the Research Report guidelines: "It is the first half of the thesis, i.e., there are no results included yet, but the report contains a full introduction including a literature review and a methods section that contains details about the data, instruments and or statistical procedures". The goal was to develop novel methodology and guidelines for evaluating multiple imputation methods, and implement these in an interactive evaluation framework for multiple imputation: \pkg{ShinyMICE}.
  }

%% - \Keywords{} with LaTeX markup, at least one required
%% - \Plainkeywords{} without LaTeX markup (if necessary)
%% - Should be comma-separated and in sentence case.
\Keywords{multiple imputation, evaluation methodology, \pkg{ShinyMICE}, \pkg{mice}, \proglang{R}}
\Plainkeywords{multiple imputation, evaluation methodology, ShinyMICE, mice, R}

%% - \Address{} of at least one author
%% - May contain multiple affiliations for each author
%%   (in extra lines, separated by \emph{and}\\).
%% - May contain multiple authors for the same affiliation
%%   (in the same first line, separated by comma).
\Address{
  Hanne Ida Oberman, BSc.\\
  Methodology and Statistics for the Behavioural, Biomedical and Social Sciences\\
  Department of Methodology and Statistics\\
  Faculty of Social and Behavioral Sciences\\
  Utrecht University\\
  Heidelberglaan~15\\
  3500 Utrecht, The Netherlands\\
  E-mail: \email{h.i.oberman@uu.nl} % \\
  % URL: \url{https://eeecon.uibk.ac.at/~zeileis/}
}

\begin{document}

%% -- Introduction -------------------------------------------------------------

%% - In principle "as usual".
%% - But should typically have some discussion of both _software_ and _methods_.
%% - Use \proglang{}, \pkg{}, \fct{}, and \code{} markup throughout the manuscript.
%% - If such markup is in (sub)section titles, a plain text version has to be
%%   added as well.
%% - All software mentioned should be properly \cite-d.
%% - All abbreviations should be introduced.
%% - Unless the expansions of abbreviations are proper names (like "Journal
%%   of Statistical Software" above) they should be in sentence case (like
%%   "generalized linear models" below).

\section[Introduction: Multiple Imputation Methodology]{Introduction: Multiple Imputation Methodology} \label{sec:intro} % curly brackets can handle typesetting

% \begin{leftbar}
% The introduction is in principle ``as usual''. However, it should usually embed
% both the implemented \emph{methods} and the \emph{software} into the respective
% relevant literature. For the latter both competing and complementary software
% should be discussed (within the same software environment and beyond), bringing
% out relative (dis)advantages. All software mentioned should be properly
% \verb|\cite{}|d. (See also Appendix~\ref{app:bibtex} for more details on
% \textsc{Bib}{\TeX}.) Example: % \citep[][Chapter~7.4]{Venables+Ripley:2002} 
% 
% For writing about software JSS requires authors to use the markup
% \verb|\proglang{}| (programming languages and large programmable systems),
% \verb|\pkg{}| (software packages), \verb|\code{}| (functions, commands,
% arguments, etc.). If there is such markup in (sub)section titles (as above), a
% plain text version has to be provided in the {\LaTeX} command as well. Below we
% also illustrate how abbrevations should be introduced and citation commands can
% be employed. See the {\LaTeX} code for more details.
% \end{leftbar}

At some point, any scientist conducting statistical analyses will run into a missing data problem \citep{alli02}. Missingness is problematic because statistical inference cannot be performed on incomplete data, and  ad hoc solutions can yield wildly invalid results \citep{buur18}. To circumvent the ubiquitous problem of missing information, \cite{rubin87} proposed the framework of multiple imputation (MI). MI is an iterative algorithmic procedure in which missing data points are 'guessed' (i.e. imputed) several times. The variability between the imputations validly reflects how much uncertainty in the inference is due to missing information--that is, if all statistical assumptions are met \cite{rubin87}.

With MI, many assumptions are made about the nature of the observed and missing parts of the data and their relation to the 'true' \emph{data generating model} \citep{buur18}. Without proper evaluation of the imputations and the underlying assumptions, any drawn inference may erroneously be deemed valid. Such evaluation measures are currently missing or under-developed in MI software, like the world leading \proglang{R} package \pkg{mice} \citep{mice11}. Therefore, I will answer the following question: 'Which measures are vital for evaluating the validity of multiply imputed data?'.

%% -- Features ---------------------------------------------------------------
\subsection{Features} \label{sec:features}

All programming code used in this paper is available in the file XYZ.R along with the manuscript and on Github repository XYZ.

"The intended audience of this paper consists of applied researchers who want to address problems caused by missing data by multiple imputation. The text assumes basic familiarity with R. The document contains hands-on analysis using the mice package. We do not discuss problems of incomplete data in general. We refer to the excellent books by Little and Rubin (2002) and Schafer (1997). Theory and applications of multiple imputation have been developed in Rubin (1987) and Rubin (1996). van Buuren (2012) introduces multiple imputation from an applied perspective" \cite[p.~4]{mice}.

%% -- Notation ---------------------------------------------------------------
\subsection{Notation and conventions?} \label{sec:notation}

Blue points are observed, the red points are imputed.

%% -- Theoretical Background ---------------------------------------------------------------
\section{Theoretical Background} \label{sec:background}

- Terminology (MCAR, MAR, MNAR)

- Rubin's rules

- FCS vs. JM?

The validity of the MI solution depends on numerous assumptions that cannot be verified from the observed data alone. So instead of statistical tests for assumptions, evaluation procedures have been developed. For the following assumptions, no reliable procedure has been proposed and/or implemented: 1) \emph{ignorability} of the \emph{missingness mechanism} \citep{rubin87}; 2) \emph{congeneality} of the imputation models \citep{meng94}; and 3) \emph{compatibility} of the MI modeling procedure \citep{rubin96}.

\begin{enumerate}
\item A missingness mechanism is said to be ignorable when the probability to be missing does not depend on the missing data itself. Violation of this assumption can gravely affect inferences. Robustness of inferences to varying degrees of violation can be assessed with sensitivity analyses. Some practical guidelines exist (e.g., \citep{nguy17}), but current MI software does not facilitate this methodology for empirical researchers. 
%
\item Congeneal imputation models capture all required relations between observed and missing parts of the data. The extent to which this has been successful can be evaluated by plotting conditional distributions \citep{abay08}. Such visualizations are available in MICE, but subsequent statistical tests to quantify the relations with covariates are not provided. 
% ^[Additionally, there is potential to assess model fit by means of \emph{over-imputation} \citep{buur18} or *double robustness* \citep{bang05}-- topics that are only persued if time permits.] 
%
\item The third assumption is met when the MI algorithm converges to a stable distribution. However, conventional measures to diagnose convergence-- e.g., Gelman and Rubin's \citeyear{gelm92} statistic $\widehat{R}$ --are not applicable on multiply imputed data \citep{lace07}. Therefore, empirical researchers have to rely on visual inspection procedures that are theoretically equivalent to $\widehat{R}$ \citep{whit11}. Visually assessing convergence is not only difficult to the untrained eye, it might also be futile. The convergence properties of MI algorithms lack scientific consensus \citep{taka17}, and some default MICE techniques might not converge to stable distributions at all \citep{murr18}. Moreover, convergence diagnostics for MI methods have not been systematically studied \citep{buur18}.
\end{enumerate}

In short, the existing literature provides both possibilities and limitations to evaluating the validity of multiply imputed data. The goal of this research project is to develop novel methodology and guidelines for evaluating MI methods, and implement these in an interactive evaluation framework for multiple imputation. This framework will aid applied researchers in drawing valid inference from incomplete datasets.

%% -- Methods ---------------------------------------------------------------

%%
%%
%%
%%

\section{Methods} \label{sec:methods}

Initially, the research project will consist of an investigation into algorithmic convergence of MI algorithms. I will replicate Lacerda et al.'s simulation study on $\widehat{R}$ \citep{lace07}, and develop novel guidelines for assessing convergence. Ideally, I will integrate several diagnostics (e.g., $\widehat{R}$, \emph{auto-correlation}, and \emph{simulation error}) into a single summary indicator to flag non-convergence.

Subsequently, I will use\proglang{R}Shiny \citep{shiny} to implement the convergence indicator and existing evaluation measures in \pkg{ShinyMICE}, see Figure 1. The application will at least contain methodology for: sensitivity analyses; data visualizations (e.g., scatter-plots, densities, cross-tabulations); and statistical evaluation of relations between variables pre- versus post-imputation (i.e., $\chi^2$~tests or $t$~tests).

A working beta version of \pkg{ShinyMICE} will be considered a sufficient milestone to proceed with writing a technical paper on the methodology and the software. I will submit the paper for publication in \emph{Journal of Statistical Software}. Finally, \pkg{ShinyMICE} will be integrated into the existing MICE environment, and a vignette for applied researchers will be written.

The\proglang{R}code and documentation of this project will be open source (available on Github). Since the study does not require the use of unpublished empirical data, I expect that the FETC will grant the label exempt.

% \includegraphics[height=80mm,keepaspectratio]{logo} % this is not correct yet, it needs a figure label and stuff

%% -- Methods: Sensitivity Analyses ---------------------------------------------------------------
\subsection{Sensitivity Analyses} \label{sec:sensitivity}

- Robustness of inferences to varying degrees of violation can be assessed with sensitivity analyses. See \citep{nguy17}.

- Forked MICE with MNAR sensitivity analyses, see \url{https://thestatsgeek.com/2018/06/07/missing-not-at-random-sensitivity-analysis-with-fcs-multiple-imputation/}.

"The MAR assumption can never be tested from the observed data. One can however check whether the imputations created by MICE algorithm are plausible" \cite[p.~42]{mice}.

"Plot densities of both the observed and imputed values of all variables to see whether the imputations are reasonable. Differences in the densities between the observed and imputed values may suggest a problem that needs to be further checked" \cite[p.~43]{mice}.

%% -- Methods: Visualizations ---------------------------------------------------------------
\subsection{Data Visualization} \label{sec:visualization}

- Implement existing plot functions in \pkg{ShinyMICE}, add subsequent statistical tests to quantify the relations with covariates.

- Relations between observed and missing parts of the data: use \pkg{lattice} package functionalities to improve the \pkg{mice} functions \fct{hist} and \fct{xyplot}.

"An important step in multiple imputation is to assess whether imputations are plausible. Imputations should be values that could have been obtained had they not been missing. Imputations should be close to the data. Data values that are clearly impossible (e.g., negative counts, pregnant fathers) should not occur in the imputed data. Imputations should respect relations between variables, and reflect the appropriate amount of uncertainty about their 'true' values. Diagnostic checks on the imputed data provide a way to check the plausibility of the imputations" \cite[p.~11]{mice}.

"It is often useful to inspect the distributions of original and the imputed data. One way of doing this is to use the function \fct{stripplot} \dots  The differences between the
red points represents our uncertainty about the true (but unknown) values. \dots Under MCAR, univariate distributions of the observed and imputed data are expected to be identical. Under MAR, they can be different, both in location and spread, but their multivariate distribution is assumed to be identical." \cite[p.~12]{mice}. Create scatterplots via \fct{xyplot}.

%% -- Methods: Convergence ---------------------------------------------------------------
\subsection{Convergence Diagnostic} \label{sec:convergence}

- Convergence issues: "Imputers who do choose to use FCS should use flexible univariate models wherever possible and take care to assess apparent convergence of the algorithm, for example by computing traces of pooled estimates or other statistics and using standard MCMC diagnostics (Gelman et al., 2013, Chapter 11). It may also be helpful to examine the results of many independent runs of the algorithm with different initializations and to use random scans over the p variables to try to identify any convergence issues and mitigate possible order dependence" \cite[p.~19]{murr18}.

- Replicate simulation study and build a decision rule to solve the problem: "The monitoring statistic was computed for mean monthly earnings at each iteration in chains of length k = 200. Since calculation of the statistic requires M parallel sequences, m = 5 such chains were constructed. This value of m was informed by the preferred choice given in the literature on multiple imputation. The monitoring statistic computed at each iteration is presented in Figures 15 to 17 for each of the three missingness mechanisms. The red vertical line denotes ten iterations" \cite[p.~49]{lace07}. % also, use '\citep[see][for further details]{...}' for '(see Author Year for further details)'

- Potentially add something about updated (2019) version of $\widehat{R}$ and the new threshold of 1.01, see \citep{veht19}.

- Auto-correlation: "Applications of MICE with lowly correlated data therefore inject a lot of noise into the system. Hence, the auto-correlation over t will be low, and convergence will be rapid, and in fact immediate if all variables are independent. Thus, the incorporation of noise into the imputed data has pleasant side-effect of speeding up convergence" \citep{buur18}, par. 4.5. Also, schafer (1997, p. 129) wrote on worst linear statistic. We could calculate the autocorrelation of that statistic to know that the algorithm converged elsewhere too. See autocorr function plot in sas of worst linear function. Note: we're talking about missing data only, not the combined data (that autocorrelation is very high, as is the autocorrelation of deductively imputed values, like in the texp example). Worst linear function in SAS see \url{https://support.sas.com/documentation/cdl/en/statug/63033/HTML/default/viewer.htm#statug_mi_sect027.htm}.

- Stability of the solution: possibly use the slope of means over iterations too to see whether there is trending. Or apply PCA on the imputed data and if that (the eigenvalues?) stays the same we know that the means and variances are stable as well. Or look at MC error: MC error = SD/sqrt(number of iterations), where SD represents the variation across iterations. The MC error thus represents how much the means differ w.r.t. the iterations. MC error decreases as number of iterations increases. It should not be larger than 5\% of the sample standard deviation.

"In general, you cannot know for sure if your chain has converged. But sometimes you can know if your chain has not converged, so we at least check for this latter possibility" \cite[p.~101]{hoff09}

%% -- Manuscript ---------------------------------------------------------------

%% - In principle "as usual" again.
%% - When using equations (e.g., {equation}, {eqnarray}, {align}, etc.
%%   avoid empty lines before and after the equation (which would signal a new
%%   paragraph.
%% - When describing longer chunks of code that are _not_ meant for execution
%%   (e.g., a function synopsis or list of arguments), the environment {Code}
%%   is recommended. Alternatively, a plain {verbatim} can also be used.
%%   (For executed code see the next section.)

\section{Models and software} \label{sec:models}

\emph{This is a placeholder.}

% A count response variable $y_i$ ($i = 1, \dots, n$) by assuming a Poisson distribution
% $y_i \sim \mathrm{Pois}(\mu_i)$. The dependence of the conditional mean
% $\E[y_i \, | \, x_i] = \mu_i$ on the regressors $x_i$ is then specified via a
% log link and a linear predictor
% %
% \begin{equation} \label{eq:mean}
% \log(\mu_i) \quad = \quad x_i^\top \beta,
% \end{equation}
% %
% where the regression coefficients $\beta$ are estimated by maximum likelihood
% (ML) using the iterative weighted least squares (IWLS) algorithm.

% \begin{leftbar}
% Note that around the \verb|{equation}| above there should be no spaces (avoided
% in the {\LaTeX} code by \verb|%| lines) so that ``normal'' spacing is used and
% not a new paragraph started.
% \end{leftbar}

% Its most important arguments are
% \begin{Code}
% glm(formula, data, subset, na.action, weights, offset,
%   family = gaussian, start = NULL, control = glm.control(...),
%   model = TRUE, y = TRUE, x = FALSE, ...)
% \end{Code}
% where \code{formula} plus \code{data} is the now standard way of specifying
% regression relationships in \proglang{R}/\proglang{S} introduced in
% \cite{Chambers+Hastie:1992}. The remaining arguments in the first line
% (\code{subset}, \code{na.action}, \code{weights}, and \code{offset}) are also
% standard  for setting up formula-based regression models in
% \proglang{R}/\proglang{S}. The arguments in the second line control aspects
% specific to GLMs while the arguments in the last line specify which components
% are returned in the fitted model object (of class \class{glm} which inherits
% from \class{lm}). For further arguments to \fct{glm} (including alternative
% specifications of starting values) see \code{?glm}. For estimating a Poisson
% model \code{family = poisson} has to be specified.

% \begin{leftbar}
% As the synopsis above is a code listing that is not meant to be executed,
% one can use either the dedicated \verb|{Code}| environment or a simple
% \verb|{verbatim}| environment for this. Again, spaces before and after should be
% avoided.
% 
% Finally, there might be a reference to a \verb|{table}| such as
% Table~\ref{tab:overview}. Usually, these are placed at the top of the page
% (\verb|[t!]|), centered (\verb|\centering|), with a caption below the table,
% column headers and captions in sentence style, and if possible avoiding vertical
% lines.
% \end{leftbar}

% \begin{table}[t!]
% \centering
% \begin{tabular}{lllp{7.4cm}}
% \hline
% Type           & Distribution & Method   & Description \\ \hline
% GLM            & Poisson      & ML       & Poisson regression: classical GLM,
%                                            estimated by maximum likelihood (ML) \\
%                &              & Quasi    & ``Quasi-Poisson regression'':
%                                            same mean function, estimated by
%                                            quasi-ML (QML) or equivalently
%                                            generalized estimating equations (GEE),
%                                            inference adjustment via estimated
%                                            dispersion parameter \\
%                &              & Adjusted & ``Adjusted Poisson regression'':
%                                            same mean function, estimated by
%                                            QML/GEE, inference adjustment via
%                                            sandwich covariances\\
%                & NB           & ML       & NB regression: extended GLM,
%                                            estimated by ML including additional
%                                            shape parameter \\ \hline
% Zero-augmented & Poisson      & ML       & Zero-inflated Poisson (ZIP),
%                                            hurdle Poisson \\
%                & NB           & ML       & Zero-inflated NB (ZINB),
%                                            hurdle NB \\ \hline
% \end{tabular}
% \caption{\label{tab:overview} Overview of various count regression models. The
% table is usually placed at the top of the page (\texttt{[t!]}), centered
% (\texttt{centering}), has a caption below the table, column headers and captions
% are in sentence style, and if possible vertical lines should be avoided.}
% \end{table}

%% -- Illustrations ------------------------------------------------------------

%% - Virtually all JSS manuscripts list source code along with the generated
%%   output. The style files provide dedicated environments for this.
%% - In R, the environments {Sinput} and {Soutput} - as produced by Sweave() or
%%   or knitr using the render_sweave() hook - are used (without the need to
%%   load Sweave.sty).
%% - Equivalently, {CodeInput} and {CodeOutput} can be used.
%% - The code input should use "the usual" command prompt in the respective
%%   software system.
%% - For R code, the prompt "R> " should be used with "+  " as the
%%   continuation prompt.
%% - Comments within the code chunks should be avoided - these should be made
%%   within the regular LaTeX text.

\section{Illustrations} \label{sec:illustrations}

\emph{This is a placeholder.}

% For a simple illustration of basic Poisson and NB count regression the
% \code{quine} data from the \pkg{MASS} package is used. This provides the number
% of \code{Days} that children were absent from school in Australia in a
% particular year, along with several covariates that can be employed as regressors.
% The data can be loaded by
%
% <<data>>=
% data("quine", package = "MASS")
@
%
% and a basic frequency distribution of the response variable is displayed in
% Figure~\ref{fig:quine}.

% \begin{leftbar}
% For code input and output, the style files provide dedicated environments.
% Either the ``agnostic'' \verb|{CodeInput}| and \verb|{CodeOutput}| can be used
% or, equivalently, the environments \verb|{Sinput}| and \verb|{Soutput}| as
% produced by \fct{Sweave} or \pkg{knitr} when using the \code{render_sweave()}
% hook. Please make sure that all code is properly spaced, e.g., using
% \code{y = a + b * x} and \emph{not} \code{y=a+b*x}. Moreover, code input should
% use ``the usual'' command prompt in the respective software system. For
% \proglang{R} code, the prompt \code{"R> "} should be used with \code{"+  "} as
% the continuation prompt. Generally, comments within the code chunks should be
% avoided -- and made in the regular {\LaTeX} text instead. Finally, empty lines
% before and after code input/output should be avoided (see above).
% \end{leftbar}

% \begin{figure}[t!]
% \centering
% <<visualization, echo=FALSE, fig=TRUE, height=5.2, width=7>>=
% par(mar = c(4, 4, 1, 1))
% plot(table(quine$Days), xlab = "Days", ylab = "Frequency", axes = FALSE)
% axis(2)
% axis(1, at = 0:16 * 5, labels = FALSE)
% axis(1, at = 0:8 * 10)
% box()
@
% \caption{\label{fig:quine} Frequency distribution for number of days absent
% from school.}
% \end{figure}

% As a first model for the \code{quine} data, we fit the basic Poisson regression
% model. (Note that JSS prefers when the second line of code is indented by two
% spaces.)
%
% <<poisson>>=
% m_pois <- glm(Days ~ (Eth + Sex + Age + Lrn)^2, data = quine,
%   family = poisson)
@
%
% To account for potential overdispersion we also consider a negative binomial
% GLM.
%
% <<negbin>>=
% library("MASS")
% m_nbin <- glm.nb(Days ~ (Eth + Sex + Age + Lrn)^2, data = quine)
@
%
% In a comparison with the BIC the latter model is clearly preferred.
% %
% <<comparison>>=
% BIC(m_pois, m_nbin)
@
% %
% Hence, the full summary of that model is shown below.
% %
% <<summary>>=
% summary(m_nbin)
@

%% -- Summary/conclusions/discussion -------------------------------------------

\section{Summary and discussion} \label{sec:summary}

\emph{This is a placeholder.}

% \begin{leftbar}
% As usual \dots
% \end{leftbar}

%% -- Optional special unnumbered sections -------------------------------------

\section*{Computational details}

% \begin{leftbar}
% If necessary or useful, information about certain computational details
% such as version numbers, operating systems, or compilers could be included
% in an unnumbered section. Also, auxiliary packages (say, for visualizations,
% maps, tables, \dots) that are not cited in the main text can be credited here.
% \end{leftbar}

The results in this paper were obtained using \proglang{R}~\Sexpr{paste(R.Version()[6:7], collapse = ".")} with the \pkg{mice}~\Sexpr{packageVersion("mice")} package. \proglang{R} itself and all packages used are available from the Comprehensive \proglang{R} Archive Network (CRAN) at \url{https://CRAN.R-project.org/}.

\section*{Acknowledgments}

% \begin{leftbar}
% All acknowledgments (note the AE spelling) should be collected in this
% unnumbered section before the references. It may contain the usual information
% about funding and feedback from colleagues/reviewers/etc. Furthermore,
% information such as relative contributions of the authors may be added here
% (if any).
% \end{leftbar}

This technical paper is written by the sole author (Hanne Oberman, BSc.), with guidance from Master thesis supervisors prof. dr. Stef van Buuren, and dr. Gerko Vink.

%% -- Bibliography -------------------------------------------------------------
%% - References need to be provided in a .bib BibTeX database.
%% - All references should be made with \cite, \citet, \citep, \citealp etc.
%%   (and never hard-coded). See the FAQ for details.
%% - JSS-specific markup (\proglang, \pkg, \code) should be used in the .bib.
%% - Titles in the .bib should be in title case.
%% - DOIs should be included where available.

\bibliography{ShinyMICE}

%% -- Appendix (if any) --------------------------------------------------------
%% - After the bibliography with page break.
%% - With proper section titles and _not_ just "Appendix".

\newpage

\begin{appendix}

\section{More technical details} \label{app:technical}

% \begin{leftbar}
% Appendices can be included after the bibliography (with a page break). Each
% section within the appendix should have a proper section title (rather than
% just \emph{Appendix}).
% 
% For more technical style details, please check out JSS's style FAQ at
% \url{https://www.jstatsoft.org/pages/view/style#frequently-asked-questions}
% which includes the following topics:
% \begin{itemize}
%   \item Title vs.\ sentence case.
%   \item Graphics formatting.
%   \item Naming conventions.
%   \item Turning JSS manuscripts into \proglang{R} package vignettes.
%   \item Trouble shooting.
%   \item Many other potentially helpful details\dots
% \end{itemize}
% \end{leftbar}

\section[Using BibTeX]{Using \textsc{Bib}{\TeX}} \label{app:bibtex}

% \begin{leftbar}
% References need to be provided in a \textsc{Bib}{\TeX} file (\code{.bib}). All
% references should be made with \verb|\cite|, \verb|\citet|, \verb|\citep|,
% \verb|\citealp| etc.\ (and never hard-coded). This commands yield different
% formats of author-year citations and allow to include additional details (e.g.,
% pages, chapters, \dots) in brackets. In case you are not familiar with these
% commands see the JSS style FAQ for details.
% 
% JSS requires the following format.
% \begin{itemize}
%   \item JSS-specific markup (\verb|\proglang|, \verb|\pkg|, \verb|\code|) should
%     be used in the references.
%   \item Titles should be in title case.
%   \item Journal titles should not be abbreviated and in title case.
%   \item DOIs should be included where available.
%   \item Software should be properly cited as well. For \proglang{R} packages
%     \code{citation("pkgname")} typically provides a good starting point.
% \end{itemize}
% \end{leftbar}

\end{appendix}

%% -----------------------------------------------------------------------------

\end{document}

### 2019-10-24 ResearchReport/article.tex
rename folders
Added a simulation folder too

--- 2019-10-24 Research Report/article.Rnw
@@
 %% need no \usepackage{Sweave}
-\SweaveOpts{engine=R, eps=FALSE, keep.source = TRUE}
-<<preliminaries, echo=FALSE, results=hide>>=
-options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
-library("MASS")
-@
 
@@
 % data("quine", package = "MASS")
-@
 %
@@
 % box()
-@
 % \caption{\label{fig:quine} Frequency distribution for number of days absent
@@
 %   family = poisson)
-@
 %
@@
 % m_nbin <- glm.nb(Days ~ (Eth + Sex + Age + Lrn)^2, data = quine)
-@
 %
@@
 % BIC(m_pois, m_nbin)
-@
 % %
@@
 % summary(m_nbin)
-@
 
@@
 
-The results in this paper were obtained using \proglang{R}~\Sexpr{paste(R.Version()[6:7], collapse = ".")} with the \pkg{mice}~\Sexpr{packageVersion("mice")} package.
+The results in this paper were obtained using \proglang{R}~3.6.1 with the \pkg{mice}~3.6.0.9000 package.
 \proglang{R} itself and all packages used are available from the Comprehensive \proglang{R} Archive Network (CRAN) at \url{https://CRAN.R-project.org/}.

### 2019-10-24 ThesisProposal/ThesisProposal.Rmd
rename folders
Added a simulation folder too

--- 2019-10-10 ThesisProposal.Rmd
@@
 
+![](Figures/logo.png)
+
 # ShinyMICE: an Evaluation Suite for Multiple Imputation
@@
 
-Word count: 713
+Word count: 709 words
 
@@
 
-With MI, many assumptions are made about the nature of the observed and missing parts of the data and their relation to the 'true' data generating model [@buur18].
+With MI, many assumptions are made about the nature of the observed and missing parts of the data and their relation to the 'true' *data generating model* [@buur18].
 Without proper evaluation of the imputations and the underlying assumptions, any drawn inference may erroneously be deemed valid.
@@
 The validity of the MI solution depends on numerous assumptions that cannot be verified from the observed data alone.
-Consequently, existing evaluation methods rely on **[rephrase: proxy measures]**.
-For the following assumptions, no reliable *proxy measures* have been proposed and/or implemented: 1) *ignorability* of the *missingness mechanism*; 2) congruity of the *imputation models*; and 3) *compatibility* of the MI modeling procedure.
+So instead of statistical tests for assumptions, evaluation procedures have been developed.
+For the following assumptions, no reliable procedure has been proposed and/or implemented: 1) *ignorability* of the *missingness mechanism* [@rubin87]; 2) *congeneality* of the imputation models [@meng94]; and 3) *compatibility* of the MI modeling procedure [@rubin96].
 <!-- Rubin (1987b, 160-66) subdivided the work needed to create imputations into three tasks.
@@
 1.
-A missingness mechanism is said to be ignorable when the sprobability to be missing does not depend on the missing data itself [@rubin87].
-Evaluation of this assumption is usually done with sensitivity analyses to assess the robustness of inferences to violations.
+A missingness mechanism is said to be ignorable when the probability to be missing does not depend on the missing data itself.
+Violation of this assumption can gravely affect inferences.
+Robustness of inferences to varying degrees of violation can be assessed with sensitivity analyses.
 Some practical guidelines exist (e.g., [@nguy17]), but current MI software does not facilitate this methodology for empirical researchers.
@@
 2.
-Congruent imputation models capture all required relations between observed and missing parts of the data.
-The exent to which this has been successful can be evaluated by plotting conditional distributions [@abay08].
+Congeneal imputation models capture all required relations between observed and missing parts of the data.
+The extent to which this has been successful can be evaluated by plotting conditional distributions [@abay08].
 Such visualizations are available in `MICE`, but subsequent statistical tests to quantify the relations with covariates are not provided.
-Additionally, there is potential to assess model fit by means of *over-imputation* [@buur18] or *double robustness* [@bang05].
+^[Additionally, there is potential to assess model fit by means of *over-imputation* [@buur18] or *double robustness* [@bang05]-- topics that are only persued if time permits.]
 
@@
 
+&nbsp;
+
 # Approach
@@
 
-Subsequently, I will use `R Shiny` [@shiny17] to implement the convergence indicator and other evaluation measures in `ShinyMICE`, see Figure 1.
-The application will at least contain methodology for: sensitivity analyses; data visualizations (e.g., scatter-plots, densities, cross-tabulations); and statistical evaluations of relations between variables pre- versus post-imputation (i.e., $\chi^2$-tests or $t$-tests).
+Subsequently, I will use `R Shiny` [@shiny17] to implement the convergence indicator and existing evaluation measures in `ShinyMICE`, see Figure 1.
+The application will at least contain methodology for: sensitivity analyses; data visualizations (e.g., scatter-plots, densities, cross-tabulations); and statistical evaluation of relations between variables pre- versus post-imputation (i.e., $\chi^2$-tests or $t$-tests).
 
@@
 
+\newpage
+
 # References

### 2019-10-30 ResearchReport/article.Rnw
rename folders
Added a simulation folder too
Update research report with quotes from FIMD

--- 2019-10-24 Research Report/article.Rnw
@@
 
+Notation in this paper is copied from \cite{buur18}, and thus based on \cite{rubin87}.
+
+% "Let Y denote the n × p matrix containing the data values on p variables for all n units in the sample.
+We define the response indicator R as an n × p 0-1 matrix.
+The elements of Y and R are denoted by y i j and r i j , respectively, where i = 1 , ...
+, n and j = 1 , ...
+, p .
+If y i j is observed, then r i j = 1 , and if y i j is missing, then r i j = 0 ." ...
+The observed data are collectively denoted by Y o b s .
+The missing data are collectively denoted as Y m i s , and contain all elements y i j where r i j = 0 .
+When taken together Y = ( Y o b s , Y m i s ) contain the hypothetically complete data." \cite[par.~2.2]{buur18}
+
+Terminology (MCAR, MAR, MNAR)
+
+% The data are said to be MCAR if Pr ( R = 0 | Y o b s , Y m i s , ??
+) = Pr ( R = 0 | ??
+) (2.1) so the probability of being missing depends only on some parameters ??
+, the overall probability of being missing.
+The data are said to be MAR if Pr ( R = 0 | Y o b s , Y m i s , ??
+) = Pr ( R = 0 | Y o b s , ??
+) (2.2) so the missingness probability may depend on observed information, including any design factors.
+Finally, the data are MNAR if Pr ( R = 0 | Y o b s , Y m i s , ??
+) (2.3) does not simplify, so here the probability to be missing also depends on unobserved information, including Y m i s itself.
+\cite[par.~2.2]{buur18}
+
 Blue points are observed, the red points are imputed.
@@
 
-- Terminology (MCAR, MAR, MNAR)
+- missingness mechanims, ignorability
+
+"The practical importance of the distinction between MCAR, MAR and MNAR is that it clarifies the conditions under which we can accurately estimate the scientifically interesting parameters without the need to know $\psi$" \cite[par.~2.2]{buur18}.
 
@@
 
-The\proglang{R}code and documentation of this project will be open source (available on Github).
+The \proglang{R} code and documentation of this project will be open source (available on Github).
 Since the study does not require the use of unpublished empirical data, I expect that the FETC will grant the label exempt.
@@
 - Forked MICE with MNAR sensitivity analyses, see \url{https://thestatsgeek.com/2018/06/07/missing-not-at-random-sensitivity-analysis-with-fcs-multiple-imputation/}.
+
+See https://stefvanbuuren.name/fimd/sec-sensitivity.html
 
@@
 
+Create scatterplots via \fct{xyplot} and use lattice functionalities to adapt.
+
 "An important step in multiple imputation is to assess whether imputations are plausible.
@@
 
+"The mice package contains several graphic functions that can be used to gain insight into the correspondence of the observed and imputed data: \fct{bwplot}, \fct{stripplot}, \fct{densityplot} and \fct{xyplot}." \cite[par.~6.6]{buur18}
+
+see https://onlinelibrary.wiley.com/doi/full/10.1111/1467-9574.00219
+
 "It is often useful to inspect the distributions of original and the imputed data. One way of doing this is to use the function \fct{stripplot} \dots  The differences between the
@@
 Under MAR, they can be different, both in location and spread, but their multivariate distribution is assumed to be identical." \cite[p.~12]{mice}.
-Create scatterplots via \fct{xyplot}.
+
+"Bondarenko and Raghunathan (2016) proposed a more refined diagnostic tool that aims to compare the distributions of observed and imputed data conditional on the missingness probability.
+The idea is that under MAR the conditional distributions should be similar if the assumed model for creating multiple imputations has a good fit.
+An example is created as ..."  \cite[par.~6.6]{buur18}
+
+see https://www.ncbi.nlm.nih.gov/pubmed/26952693
+
+% fit <- with(imp, glm(ici(imp) ~ age + bmi + hyp + chl,
+%                      family = binomial))
+% ps <- rep(rowMeans(sapply(fit$analyses, fitted.values)),
+%           imp$m + 1)
+% xyplot(imp, bmi ~ ps | as.factor(.imp),
+%        xlab = "Probability that record is incomplete",
+%        ylab = "BMI", pch = c(1, 19), col = mdc(1:2))
 

### 2019-10-31 ResearchReport/article.Rnw
Archive answers to questions about simulation

--- 2019-10-24 Research Report/article.Rnw
@@
 
-All programming code used in this paper is available in the file XYZ.R along with the manuscript and on Github repository XYZ.
+The aim of this paper is to provide applied researchers an introduction to the interactive evaluation device \pkg{ShinyMICE}.
+All programming code used in this paper is available in the file XYZ.R along with the manuscript, and on Github repository XYZ.
 
@@
 
+Notation in this paper is copied from \cite{buur18}, and thus based on \cite{rubin87}.
+Let $Y$ denote an $n \times p$ matrix containing the data values on $p$ variables for all $n$ units in a sample.
+The collection of observed data values is denoted as $Y_{obs}$; the missing part of $Y$ is refered to as $Y_{mis}$.
+Response indicator $R$ shows whether a data value in $Y$ is missing or observed.
+The relation between $R$, $Y_{obs}$, and $Y_{mis}$ determines the missingness mechanism.
+
+% "Let Y denote the n x p matrix containing the data values on p variables for all n units in the sample.
+We define the response indicator R as an n x p 0-1 matrix.
+The elements of Y and R are denoted by y i j and r i j , respectively, where i = 1 , ...
+, n and j = 1 , ...
+, p .
+If y i j is observed, then r i j = 1 , and if y i j is missing, then r i j = 0 ." ...
+The observed data are collectively denoted by Y o b s .
+The missing data are collectively denoted as Y m i s , and contain all elements y i j where r i j = 0 .
+When taken together Y = ( Y o b s , Y m i s ) contain the hypothetically complete data." \cite[par.~2.2]{buur18}
+
+% The data are said to be MCAR if Pr ( R = 0 | Y o b s , Y m i s , $\psi$ ) = Pr ( R = 0 | $\psi$ ) (2.1) so the probability of being missing depends only on some parameters $\psi$ , the overall probability of being missing.
+The data are said to be MAR if Pr ( R = 0 | Y o b s , Y m i s , $\psi$ ) = Pr ( R = 0 | Y o b s , $\psi$ ) (2.2) so the missingness probability may depend on observed information, including any design factors.
+Finally, the data are MNAR if Pr ( R = 0 | Y o b s , Y m i s , $\psi$ ) (2.3) does not simplify, so here the probability to be missing also depends on unobserved information, including Y m i s itself.
+\cite[par.~2.2]{buur18}
+
+Terminology (MCAR, MAR, MNAR).
+
 Blue points are observed, the red points are imputed.
@@
 - FCS vs. JM?
+
+\subsection{Theoretical Background from Thesis Proposal}
 
@@
 Some practical guidelines exist (e.g., \citep{nguy17}), but current MI software does not facilitate this methodology for empirical researchers.
+% use '\citep[see][for further details]{...}' for '(see Author Year for further details)'
 %
@@
 Such visualizations are available in MICE, but subsequent statistical tests to quantify the relations with covariates are not provided.
-% ^[Additionally, there is potential to assess model fit by means of \emph{over-imputation} \citep{buur18} or *double robustness* \citep{bang05}-- topics that are only persued if time permits.] 
+% Additionally, there is potential to assess model fit by means of \emph{over-imputation} \citep{buur18} or *double robustness* \citep{bang05}-- topics that are only persued if time permits.
 %
@@
 
+\subsection{Approach from Thesis Proposal}
+
 Initially, the research project will consist of an investigation into algorithmic convergence of MI algorithms.
@@
 
-The\proglang{R}code and documentation of this project will be open source (available on Github).
+The \proglang{R} code and documentation of this project will be open source (available on Github).
 Since the study does not require the use of unpublished empirical data, I expect that the FETC will grant the label exempt.
 
-% \includegraphics[height=80mm,keepaspectratio]{logo} % this is not correct yet, it needs a figure label and stuff
+% \includegraphics[height=80mm,keepaspectratio]{impression} % this is not correct yet, it needs a figure label and stuff
 
@@
 - Forked MICE with MNAR sensitivity analyses, see \url{https://thestatsgeek.com/2018/06/07/missing-not-at-random-sensitivity-analysis-with-fcs-multiple-imputation/}.
+
+- FIMD on sensitivity analyses, see https://stefvanbuuren.name/fimd/sec-sensitivity.html
 
@@
 
+- Evaluation of MI methods in SAS, see https://onlinelibrary.wiley.com/doi/full/10.1111/1467-9574.00219
+
 "An important step in multiple imputation is to assess whether imputations are plausible.
@@
 
-"It is often useful to inspect the distributions of original and the imputed data. One way of doing this is to use the function \fct{stripplot} \dots  The differences between the
-red points represents our uncertainty about the true (but unknown) values.
+"The mice package contains several graphic functions that can be used to gain insight into the correspondence of the observed and imputed data: \fct{bwplot}, \fct{stripplot}, \fct{densityplot} and \fct{xyplot}." \cite[par.~6.6]{buur18}
+
+"It is often useful to inspect the distributions of original and the imputed data.
+One way of doing this is to use the function \fct{stripplot} \dots  The differences between the red points represents our uncertainty about the true (but unknown) values.
 \dots Under MCAR, univariate distributions of the observed and imputed data are expected to be identical.
 Under MAR, they can be different, both in location and spread, but their multivariate distribution is assumed to be identical." \cite[p.~12]{mice}.
-Create scatterplots via \fct{xyplot}.
+
+"Bondarenko and Raghunathan (2016) [see https://www.ncbi.nlm.nih.gov/pubmed/26952693] proposed a more refined diagnostic tool that aims to compare the distributions of observed and imputed data conditional on the missingness probability.
+The idea is that under MAR the conditional distributions should be similar if the assumed model for creating multiple imputations has a good fit.
+An example is created as ..."  \cite[par.~6.6]{buur18}
+
+% fit <- with(imp, glm(ici(imp) ~ age + bmi + hyp + chl,
+%                      family = binomial))
+% ps <- rep(rowMeans(sapply(fit$analyses, fitted.values)),
+%           imp$m + 1)
+% xyplot(imp, bmi ~ ps | as.factor(.imp),
+%        xlab = "Probability that record is incomplete",
+%        ylab = "BMI", pch = c(1, 19), col = mdc(1:2))
 
@@
 The red vertical line denotes ten iterations" \cite[p.~49]{lace07}.
-% also, use '\citep[see][for further details]{...}' for '(see Author Year for further details)'
 

### 2019-11-10 ResearchReport/article.Rnw
New folder structure of simulation study

@@
 
+"In order to converge to a stationary distribution, a Markov chain needs to satisfy three important properties (Roberts 1996; Tierney 1996): irreducible, the chain must be able to reach all interesting parts of the state space; aperiodic, the chain should not oscillate between different states; recurrence, all interesting parts can be reached infinitely often, at least from almost all starting points.
+Do these properties hold for the MICE algorithm?
+Irreducibility is generally not a problem since the user has large control over the interesting parts of the state space.
+This flexibility is actually the main rationale for FCS instead of a joint model.
+Periodicity is a potential problem, and can arise in the situation where imputation models are clearly inconsistent.
+A rather artificial example of an oscillatory behavior occurs when Y 1 is imputed by Y 2 $\beta$ + $\epsilon$  1 and Y 2 is imputed by - Y 1 $\beta$ + $\epsilon$  2 for some fixed, nonzero $\beta$ .
+The sampler will oscillate between two qualitatively different states, so the correlation between Y 1 and Y 2 after imputing Y 1 will differ from that after imputing Y 2 .
+In general, we would like the statistical inferences to be independent of the stopping point.
+A way to diagnose the ping-pong problem, or order effect, is to stop the chain at different points.
+The stopping point should not affect the statistical inferences.
+The addition of noise to create imputations is a safeguard against periodicity, and allows the sampler to "break out" more easily.
+Non-recurrence may also be a potential difficulty, manifesting itself as explosive or non-stationary behavior.
+For example, if imputations are made by deterministic functions, the Markov chain may lock up.
+Such cases can sometimes be diagnosed from the trace lines of the sampler.
+See Section 6.5.2 for an example.
+As long as the parameters of imputation models are estimated from the data, non-recurrence is mild or absent.
+The required properties of the MCMC method can be translated into conditions on the eigenvalues of the matrix of transition probabilities (MacKay 2003, 372-73).
+The development of practical tools that put these conditions to work for multiple imputation is still an ongoing research problem." \cite[par.~4.5]{buur18}
+
+\section{Simulation study}
+
+\subsection{Simulation: Introduction}
+
+R hat not working on MI data?
+"if over-dispersion does not hold, $var_plus$ can be too low, which can lead to falsely diagnosing convergence.)" \cite[p~437]{broo98}.
+Aim to replicate Lacerda et al.
+that R hat will not be smaller than 1.1 before iteration nr.
+50.
+
+As recommended by ... Rhat is computed as .... to be able to detect ... in the tails of the distribution.
+
+R hat is not appropriate because it assumes overdispersed initial values, which means that the initial values of the m imputation chains should be 'far away' from the distribution that the chains are converging to (the modus of the distribution).
+With MI procedures, initial values are chosen such that ...
+or maybe even estimated from the observed data.
+This means that at most one initial value is necessary, but probably none at al.
+I should look this up.
+
+The potential scale reduction factor tells us how much variance could be shrunken down by running the chains infinitely long.
+That tells us something about how dependent the chains are on the starting values.
+If there is no dependence on the initial values anymore, the chains have converged (in the mixing sense of the word).
+
+Convergence can also be interpreted as stable over iterations, or non-recurring.
+Non-recurrence can be evaluated with autocorrelation.
+Autocorrelation shows how dependent subsequent draws of an imputation chain are on the previous value.
+If there is a lot of dependence, draws at e.g.
+iteration five are significantly correlated with the value of the first draw.
+Ideally, we want the chains to intermingle nicely, without trends witihin chains.
+Autocorrelation above the confidence level indicate dependence within chains.
+
+\subsection{Simulation: Methodology}
+
+Explain diagnostics here (see below).
+
+We want to evaluate whether the imputation procedure has converged.
+We can apply GR statistic Rhat to the first two moments of the variables of interest (mean and variance).
+The primary research interest is in determining whether Rhat is an appropriate convergence diagnostic, and if so, which level of stringency suits MI data.
+The simulation diagnostics are as recommended by Stef van Buuren \cite{buur18}.
+That is, absolute bias confidence interval width, and empirical coverage rate (coverage probability?) across simulations.
+
+Hypothesis based on \cite{lace07} is that the conventional acceptable level of Rhat is too strict for MI data.
+We expect that the simulation diagnostics will indicate proper imputations before Rhat will.
+
+\subsection{Simulation: Simulation Approach}
+
+Set-up: simulate data, impute missing data points with varying number of iterations.
+Then evaluate the regular diagnostics of MI simulations, see Vink, n.d., and the convergence.
+Diagnostics include absolute bias of the estimated regression coefficient, confidence interval width, and empirical coverage rate (coverage probability?) across simulations.
+Convergence is evaluated with $\hat{R}$, or potential scale reduction factor.
+Proposed by Gelman and Rubin (1992), but since developed into ...
+
+In this study, 1000 MCMC simulations are performed.
+
+\begin{Code}
+# pseudocode of similation 
+simulate data with missingness
+for (number of iterations from 1 to 100)
+  impute the missingness
+  compute convergence diagnostic
+  perform analysis
+  pool results
+  compute simulation diagnostics
+save diagnostics
+\end{Code}
+
+Simulation code is available in online appendix XYZ on Github.
+
+\subsection{Simulation: Results}
+
+Rhat is not perfect for MI data.
+The value shoudl theoretically not be smaller than one, yet it happened several times in this study.
+How could Rhat smaller than 1 occur?
+The number of simulations is smaller than in 'regular' MCMC processes.
+Therefore, the '$n-1/n)$' correction factor can influence the estimated potential scale reduction facotr.
+This downwards bias is in the opposite direction than expected.
+
+\subsection{Simulation Discussion}
+
+"The mixture-of-sequences variance, V, should stabilize as a function of n.
+(Before convergence, we expect $\sigma^2$ to decrease with n, only increasing if the sequences explore a new area of parameter space, which would imply that the original sequences were not overdispersed for the particular scalar summary being monitored.)"
+% \cite[p~438]{broo98}
+
+We would like to have convergence measures for multivariable statistics (scalars?) of interest.
+This is, however, dependent of the complete data model.
+The eigenvector decomposition methof proposed by May 9?
+should be implemented.
+I could not find any resources to apply this method and it is outside the scope of this thesis to investigate how this apprach could be implemented.
+
 %% -- Manuscript ---------------------------------------------------------------

### 2019-11-10 ResearchReport/article.tex
New folder structure of simulation study

@@
 
+"In order to converge to a stationary distribution, a Markov chain needs to satisfy three important properties (Roberts 1996; Tierney 1996): irreducible, the chain must be able to reach all interesting parts of the state space; aperiodic, the chain should not oscillate between different states; recurrence, all interesting parts can be reached infinitely often, at least from almost all starting points.
+Do these properties hold for the MICE algorithm?
+Irreducibility is generally not a problem since the user has large control over the interesting parts of the state space.
+This flexibility is actually the main rationale for FCS instead of a joint model.
+Periodicity is a potential problem, and can arise in the situation where imputation models are clearly inconsistent.
+A rather artificial example of an oscillatory behavior occurs when Y 1 is imputed by Y 2 $\beta$ + $\epsilon$  1 and Y 2 is imputed by - Y 1 $\beta$ + $\epsilon$  2 for some fixed, nonzero $\beta$ .
+The sampler will oscillate between two qualitatively different states, so the correlation between Y 1 and Y 2 after imputing Y 1 will differ from that after imputing Y 2 .
+In general, we would like the statistical inferences to be independent of the stopping point.
+A way to diagnose the ping-pong problem, or order effect, is to stop the chain at different points.
+The stopping point should not affect the statistical inferences.
+The addition of noise to create imputations is a safeguard against periodicity, and allows the sampler to "break out" more easily.
+Non-recurrence may also be a potential difficulty, manifesting itself as explosive or non-stationary behavior.
+For example, if imputations are made by deterministic functions, the Markov chain may lock up.
+Such cases can sometimes be diagnosed from the trace lines of the sampler.
+See Section 6.5.2 for an example.
+As long as the parameters of imputation models are estimated from the data, non-recurrence is mild or absent.
+The required properties of the MCMC method can be translated into conditions on the eigenvalues of the matrix of transition probabilities (MacKay 2003, 372-73).
+The development of practical tools that put these conditions to work for multiple imputation is still an ongoing research problem." \cite[par.~4.5]{buur18}
+
+\section{Simulation study}
+
+\subsection{Simulation: Introduction}
+
+R hat not working on MI data?
+"if over-dispersion does not hold, $var_plus$ can be too low, which can lead to falsely diagnosing convergence.)" \cite[p~437]{broo98}.
+Aim to replicate Lacerda et al.
+that R hat will not be smaller than 1.1 before iteration nr.
+50.
+
+As recommended by ... Rhat is computed as .... to be able to detect ... in the tails of the distribution.
+
+R hat is not appropriate because it assumes overdispersed initial values, which means that the initial values of the m imputation chains should be 'far away' from the distribution that the chains are converging to (the modus of the distribution).
+With MI procedures, initial values are chosen such that ...
+or maybe even estimated from the observed data.
+This means that at most one initial value is necessary, but probably none at al.
+I should look this up.
+
+The potential scale reduction factor tells us how much variance could be shrunken down by running the chains infinitely long.
+That tells us something about how dependent the chains are on the starting values.
+If there is no dependence on the initial values anymore, the chains have converged (in the mixing sense of the word).
+
+Convergence can also be interpreted as stable over iterations, or non-recurring.
+Non-recurrence can be evaluated with autocorrelation.
+Autocorrelation shows how dependent subsequent draws of an imputation chain are on the previous value.
+If there is a lot of dependence, draws at e.g.
+iteration five are significantly correlated with the value of the first draw.
+Ideally, we want the chains to intermingle nicely, without trends witihin chains.
+Autocorrelation above the confidence level indicate dependence within chains.
+
+\subsection{Simulation: Methodology}
+
+Explain diagnostics here (see below).
+
+We want to evaluate whether the imputation procedure has converged.
+We can apply GR statistic Rhat to the first two moments of the variables of interest (mean and variance).
+The primary research interest is in determining whether Rhat is an appropriate convergence diagnostic, and if so, which level of stringency suits MI data.
+The simulation diagnostics are as recommended by Stef van Buuren \cite{buur18}.
+That is, absolute bias confidence interval width, and empirical coverage rate (coverage probability?) across simulations.
+
+Hypothesis based on \cite{lace07} is that the conventional acceptable level of Rhat is too strict for MI data.
+We expect that the simulation diagnostics will indicate proper imputations before Rhat will.
+
+\subsection{Simulation: Simulation Approach}
+
+Set-up: simulate data, impute missing data points with varying number of iterations.
+Then evaluate the regular diagnostics of MI simulations, see Vink, n.d., and the convergence.
+Diagnostics include absolute bias of the estimated regression coefficient, confidence interval width, and empirical coverage rate (coverage probability?) across simulations.
+Convergence is evaluated with $\hat{R}$, or potential scale reduction factor.
+Proposed by Gelman and Rubin (1992), but since developed into ...
+
+In this study, 1000 MCMC simulations are performed.
+
+\begin{Code}
+# pseudocode of similation 
+simulate data with missingness
+for (number of iterations from 1 to 100)
+  impute the missingness
+  compute convergence and simulation diagnostics
+save diagnostics
+\end{Code}
+
+Simulation code is available in appendix A and on Github.
+
+\subsection{Simulation: Results}
+
+Rhat is not perfect for MI data.
+The value shoudl theoretically not be smaller than one, yet it happened several times in this study.
+How could Rhat smaller than 1 occur?
+The number of simulations is smaller than in 'regular' MCMC processes.
+Therefore, the '$n-1/n)$' correction factor can influence the estimated potential scale reduction facotr.
+This downwards bias is in the opposite direction than expected.
+
+\subsection{Simulation Discussion}
+
+"The mixture-of-sequences variance, V, should stabilize as a function of n.
+(Before convergence, we expect $\sigma^2$ to decrease with n, only increasing if the sequences explore a new area of parameter space, which would imply that the original sequences were not overdispersed for the particular scalar summary being monitored.)"
+% \cite[p~438]{broo98}
+
+We would like to have convergence measures for multivariable statistics (scalars?) of interest.
+This is, however, dependent of the complete data model.
+The eigenvector decomposition methof proposed by May 9?
+should be implemented.
+I could not find any resources to apply this method and it is outside the scope of this thesis to investigate how this apprach could be implemented.
+
 %% -- Manuscript ---------------------------------------------------------------

### 2019-11-10 Simulation/Functions/Evaluate.R

@@ -1,85 +1,15 @@
-# evaluation function
-# requires packages 'miceadds' and 'dplyr'
+# Extract aggregated simulation results by averaging over runs
 
-########## Part 1 ############
-# extract R hat values
-convergence.diag <- function(sims) {
-  # create objects to store output in
-  r.hat <- matrix(NA, nrow = n.iter, ncol = n.sim)
-  convergence <- matrix(NA, n.iter, ncol = 4)
+evaluate.sim <- function(sims, true = 2) {
   
-  # R hat
-  for (i in 2:n.iter) {
-    for (j in 1:n.sim) {
-      # extract R hat
-      r.hat[i, j] <-
-        max(as.numeric(Rhat.mice(sims[[i]][[j]])$Rhat.M.imp))
-    }
-  }
-  
-  # store the diagnostics in convergence matrix
-  convergence[, 1] <- mean.r.hat <- rowMeans(r.hat, na.rm = T)
-  convergence[, 2] <-
-    conv.rate.broad <- rowMeans(r.hat < 1.2, na.rm = T)
-  convergence[, 3] <-
-    conv.rate.middle <- rowMeans(r.hat < 1.1, na.rm = T)
-  convergence[, 4] <-
-    conv.rate.narrow <- rowMeans(r.hat < 1.01, na.rm = T)
-  
-  #output
-  return(convergence)
-}
+  # compute average over simulations
+  RB <- rowMeans(res[, , 1]) - true
+  PB <- 100 * abs((rowMeans(res[, , 1]) - true) / true)
+  CR <- rowMeans(res[, , 2] < true & true < res[, , 3])
+  AW <- rowMeans(res[, , 3] - res[, , 2])
+  Rh <- rowMeans(res[, , 4])
+  AC <- rowMeans(res[, , 5])
 
-########## Part 2 ############
-# apply lm() on each sim, on each maxit value
-my.lm <-
-  function(OUT)
-    summary(pool(lm.mids(V1 ~ V2, OUT)), conf.int = T)
-
-# testing
-test1 <- mice(data)
-test2 <- lm.mids(Y ~ X + Z1 + Z2, test)
-test3 <- pool(test2)
-# use t of finite pop.
-imp_est <- test3$pooled$estimate[[2]]
-imp_var <- test3$pooled$b[[2]] + (test3$pooled$b[[2]]/test3$m)
-imp_se <- sqrt(imp_var)
-z_value <- qnorm(.025,lower.tail = F)
-ci_low <- imp_est - z_value * imp_se
[…]

### 2019-11-11 ResearchReport/article.Rnw
Cleaned up article file, solved mistake in calculation CI.

@@
 Without proper evaluation of the imputations and the underlying assumptions, any drawn inference may erroneously be deemed valid.
-Such evaluation measures are currently missing or under-developed in MI software, like the world leading \proglang{R} package \pkg{mice} \citep{mice11}.
+Such evaluation measures are currently missing or under-developed in MI software, like the world leading \proglang{R} package \pkg{mice} \citep{mice}.
 Therefore, I will answer the following question: 'Which measures are vital for evaluating the validity of multiply imputed data?'.
@@
 The aim of this paper is to provide applied researchers an introduction to the interactive evaluation device \pkg{ShinyMICE}.
-All programming code used in this paper is available in the file XYZ.R along with the manuscript, and on Github repository XYZ.
+The intended audience consists of empirical researchers and statisticians who (want to) use multiple imputation to solve missing data problems.
+Basic familiarity with multiple imputation methodology is assumed.
+For an accessible and comprehensive introduction to MI from an applied perspective, see \cite{buur18}.
+For the theoretical foundation of MI, see \cite{rubin87}.
+All programming code used in this paper is available in the file 'XYZ.R' along with the manuscript, and on Github repository 'XYZ'.
 
-"The intended audience of this paper consists of applied researchers who want to address problems caused by missing data by multiple imputation.
+% "The intended audience of this paper consists of applied researchers who want to address problems caused by missing data by multiple imputation.
 The text assumes basic familiarity with R.
@@
 %% -- Notation ---------------------------------------------------------------
-\subsection{Notation and conventions?} \label{sec:notation}
+\subsection{Notation} \label{sec:notation}
 
-Notation in this paper is copied from \cite{buur18}, and thus based on \cite{rubin87}.
+The \proglang{R} \pkg{Shiny} application introduced in this paper is developed with the aim to integrate it into the \pkg{mice} environment.
+\pkg{ShinyMICE} therefore follows notation and conventions of \cite{mice}.
+For an overview of the deviations from the 'original' notation by \cite{rubin87}, see \cite{buur18}.
+
 Let $Y$ denote an $n \times p$ matrix containing the data values on $p$ variables for all $n$ units in a sample.
@@
 
-- Robustness of inferences to varying degrees of violation can be assessed with sensitivity analyses. See \citep{nguy17}.
+% In general: "The MAR assumption can never be tested from the observed data. One can however check whether the imputations created by MICE algorithm are plausible" \cite[p.~42]{mice}.
 
-- Forked MICE with MNAR sensitivity analyses, see \url{https://thestatsgeek.com/2018/06/07/missing-not-at-random-sensitivity-analysis-with-fcs-multiple-imputation/}.
+\subsubsection{What is already implemented?}
 
-- FIMD on sensitivity analyses, see https://stefvanbuuren.name/fimd/sec-sensitivity.html
+- Current guideine: "Plot densities of both the observed and imputed values of all variables to see whether the imputations are reasonable.
+Differences in the densities between the observed and imputed values may suggest a problem that needs to be further checked" \cite[p.~43]{mice}.
+% FIMD on sensitivity analyses, see https://stefvanbuuren.name/fimd/sec-sensitivity.html
 
-"The MAR assumption can never be tested from the observed data. One can however check whether the imputations created by MICE algorithm are plausible" \cite[p.~42]{mice}.
+\subsubsection{What is not yet implemented, but exists?}
 
-"Plot densities of both the observed and imputed values of all variables to see whether the imputations are reasonable.
-Differences in the densities between the observed and imputed values may suggest a problem that needs to be further checked" \cite[p.~43]{mice}.
+- Robustness of inferences to varying degrees of violation can be assessed with sensitivity analyses. See \citep{nguy17} for guidelines.
+
+- Forked MICE with MNAR sensitivity analyses
+% see \url{https://thestatsgeek.com/2018/06/07/missing-not-at-random-sensitivity-analysis-with-fcs-multiple-imputation/}.
+
+\subsubsection{What is not implemented, and NA?}
+
+- Nothing within the scope of this thesis.
 
@@
 
-- Implement existing plot functions in \pkg{ShinyMICE}, add subsequent statistical tests to quantify the relations with covariates.
-
-- Relations between observed and missing parts of the data: use \pkg{lattice} package functionalities to improve the \pkg{mice} functions \fct{hist} and \fct{xyplot}.
-
-- Evaluation of MI methods in SAS, see https://onlinelibrary.wiley.com/doi/full/10.1111/1467-9574.00219
-
-"An important step in multiple imputation is to assess whether imputations are plausible.
+% In general: "An important step in multiple imputation is to assess whether imputations are plausible.
 Imputations should be values that could have been obtained had they not been missing.
@@
 
-"The mice package contains several graphic functions that can be used to gain insight into the correspondence of the observed and imputed data: \fct{bwplot}, \fct{stripplot}, \fct{densityplot} and \fct{xyplot}." \cite[par.~6.6]{buur18}
+\subsubsection{What is already implemented?}
 
-"It is often useful to inspect the distributions of original and the imputed data.
+- Existing plot functions to evaluate relations between observed and missing parts of the data (\pkg{mice} functions \fct{hist}, \fct{xyplot}, and \fct{stripplot}). 
+% "The mice package contains several graphic functions that can be used to gain insight into the correspondence of the observed and imputed data: \fct{bwplot}, \fct{stripplot}, \fct{densityplot} and \fct{xyplot}." \cite[par.~6.6]{buur18}
+
+% "It is often useful to inspect the distributions of original and the imputed data.
 One way of doing this is to use the function \fct{stripplot} \dots  The differences between the red points represents our uncertainty about the true (but unknown) values.
@@
 
-"Bondarenko and Raghunathan (2016) [see https://www.ncbi.nlm.nih.gov/pubmed/26952693] proposed a more refined diagnostic tool that aims to compare the distributions of observed and imputed data conditional on the missingness probability.
+\subsubsection{What is not yet implemented, but exists?}
+
+- Improve plots using \pkg{lattice} package functionalities.
+
+- Subsequent statistical tests to quantify the relations with covariates.
+% "Bondarenko and Raghunathan (2016) [see https://www.ncbi.nlm.nih.gov/pubmed/26952693] proposed a more refined diagnostic tool that aims to compare the distributions of observed and imputed data conditional on the missingness probability.
 The idea is that under MAR the conditional distributions should be similar if the assumed model for creating multiple imputations has a good fit.
 An example is created as ..."  \cite[par.~6.6]{buur18}
-
+% 
 % fit <- with(imp, glm(ici(imp) ~ age + bmi + hyp + chl,
@@
 
+- Use evaluation of MI methods in SAS as example
+% see https://onlinelibrary.wiley.com/doi/full/10.1111/1467-9574.00219
+
+\subsubsection{What is not implemented, and NA?}
+
+- Nothing within the scope of this thesis.
+
 %% -- Methods: Convergence ---------------------------------------------------------------
@@
 
-- Convergence issues: "Imputers who do choose to use FCS should use flexible univariate models wherever possible and take care to assess apparent convergence of the algorithm, for example by computing traces of pooled estimates or other statistics and using standard MCMC diagnostics (Gelman et al., 2013, Chapter 11).
-It may also be helpful to examine the results of many independent runs of the algorithm with different initializations and to use random scans over the p variables to try to identify any convergence issues and mitigate possible order dependence" \cite[p.~19]{murr18}.
+% In general: "In general, you cannot know for sure if your chain has converged.
+But sometimes you can know if your chain has not converged, so we at least check for this latter possibility" \cite[p.~101]{hoff09}
 
-- Replicate simulation study and build a decision rule to solve the problem: "The monitoring statistic was computed for mean monthly earnings at each iteration in chains of length k = 200.
-Since calculation of the statistic requires M parallel sequences, m = 5 such chains were constructed.
-This value of m was informed by the preferred choice given in the literature on multiple imputation.
-The monitoring statistic computed at each iteration is presented in Figures 15 to 17 for each of the three missingness mechanisms.
-The red vertical line denotes ten iterations" \cite[p.~49]{lace07}.
-
-- Potentially add something about updated (2019) version of $\widehat{R}$ and the new threshold of 1.01, see \citep{veht19}.
-
-- Auto-correlation: "Applications of MICE with lowly correlated data therefore inject a lot of noise into the system.
-Hence, the auto-correlation over t will be low, and convergence will be rapid, and in fact immediate if all variables are independent.
-Thus, the incorporation of noise into the imputed data has pleasant side-effect of speeding up convergence" \citep{buur18}, par.
-4.5.
-Also, schafer (1997, p.
-129) wrote on worst linear statistic.
-We could calculate the autocorrelation of that statistic to know that the algorithm converged elsewhere too.
-See autocorr function plot in sas of worst linear function.
-Note: we're talking about missing data only, not the combined data (that autocorrelation is very high, as is the autocorrelation of deductively imputed values, like in the texp example).
-Worst linear function in SAS see \url{https://support.sas.com/documentation/cdl/en/statug/63033/HTML/default/viewer.htm#statug_mi_sect027.htm}.
-
-- Stability of the solution: possibly use the slope of means over iterations too to see whether there is trending.
-Or apply PCA on the imputed data and if that (the eigenvalues?) stays the same we know that the means and variances are stable as well.
-Or look at MC error: MC error = SD/sqrt(number of iterations), where SD represents the variation across iterations.
-The MC error thus represents how much the means differ w.r.t.
-the iterations.
-MC error decreases as number of iterations increases.
-It should not be larger than 5\% of the sample standard deviation.
-
-"In general, you cannot know for sure if your chain has converged. But sometimes you can know if your chain has not converged, so we at least check for this latter possibility" \cite[p.~101]{hoff09}
-
-"In order to converge to a stationary distribution, a Markov chain needs to satisfy three important properties (Roberts 1996; Tierney 1996): irreducible, the chain must be able to reach all interesting parts of the state space; aperiodic, the chain should not oscillate between different states; recurrence, all interesting parts can be reached infinitely often, at least from almost all starting points.
+% "In order to converge to a stationary distribution, a Markov chain needs to satisfy three important properties (Roberts 1996; Tierney 1996): irreducible, the chain must be able to reach all interesting parts of the state space; aperiodic, the chain should not oscillate between different states; recurrence, all interesting parts can be reached infinitely often, at least from almost all starting points.
 Do these properties hold for the MICE algorithm?
@@
 
+\subsubsection{What is already implemented?}
+
+- Traceplots
+
+\subsubsection{What is not yet implemented, but exists?}
+
+- $\widehat{R}$, but too stringent (new) threshold, and assumption of overdispersed initial values of imputation chains not met.
+% Potentially add something about updated (2019) version of $\widehat{R}$ and the new threshold of 1.01, see \citep{veht19}.
+
+- Autocorrelation.
+Schafer (1997, p.
+129) wrote on worst linear statistic.
+We could calculate the autocorrelation of that statistic to know that the algorithm converged elsewhere too.
+See autocorr function plot in sas of worst linear function.
+% Worst linear function in SAS see \url{https://support.sas.com/documentation/cdl/en/statug/63033/HTML/default/viewer.htm#statug_mi_sect027.htm}.
+% Note: we're talking about missing data only, not the combined data (that autocorrelation is very high, as is the autocorrelation of deductively imputed values, like in the texp example). 
+% "Applications of MICE with lowly correlated data therefore inject a lot of noise into the system.
+Hence, the auto-correlation over t will be low, and convergence will be rapid, and in fact immediate if all variables are independent.
+Thus, the incorporation of noise into the imputed data has pleasant side-effect of speeding up convergence" \citep{buur18}, par.
+4.5.
+
+- Sensitivity analysis: Run algorithm several times and compare results. 
+% "Imputers who do choose to use FCS should use flexible univariate models wherever possible and take care to assess apparent convergence of the algorithm, for example by computing traces of pooled estimates or other statistics and using standard MCMC diagnostics (Gelman et al., 2013, Chapter 11).
+It may also be helpful to examine the results of many independent runs of the algorithm with different initializations and to use random scans over the p variables to try to identify any convergence issues and mitigate possible order dependence" \cite[p.~19]{murr18}.
+
+\subsubsection{What is not implemented, and NA?}
+
+- $\hat{R}$ threshold: Replicate simulation study and build a decision rule to solve the problem with $\hat{R}$.
+% "The monitoring statistic was computed for mean monthly earnings at each iteration in chains of length k = 200.
+Since calculation of the statistic requires M parallel sequences, m = 5 such chains were constructed.
+This value of m was informed by the preferred choice given in the literature on multiple imputation.
+The monitoring statistic computed at each iteration is presented in Figures 15 to 17 for each of the three missingness mechanisms.
+The red vertical line denotes ten iterations" \cite[p.~49]{lace07}.
+
+- Stability of the solution: Possibly use the slope of means over iterations too to see whether there is trending.
+Or apply PCA on the imputed data and if that (the eigenvalues?) stays the same we know that the means and variances are stable as well, see McKay (?).
+
+- MC error: MC error = SD/sqrt(number of iterations), where SD represents the variation across iterations.
+The MC error thus represents how much the means differ w.r.t.
+the iterations.
+MC error decreases as number of iterations increases.
+It should not be larger than 5\% of the sample standard deviation.
+
 \section{Simulation study}
@@
 
+\emph{It is written as a (mini) thesis, with an introduction, methods section, some results (i.e., preliminary analyses, or pilot simulations), and a discussion of results.
+The length of the research report should be maximally 2500 words of text (without references list and or tables and figures).
+Please do not include appendices, and no more than 6 tables or figures.
+Table and Figure captions do not count towards the word limit.
+An abstract may be included, but is not necessary.}
+
 R hat not working on MI data?
-"if over-dispersion does not hold, $var_plus$ can be too low, which can lead to falsely diagnosing convergence.)" \cite[p~437]{broo98}.
+"if over-dispersion does not hold, $\sigma_+^2$ can be too low, which can lead to falsely diagnosing convergence." \cite[p~437]{broo98}.
 Aim to replicate Lacerda et al.
@@
 
+Evaluate: bias, variance, and cov rate to find problems that you might miss otherwise.
+There is no tradeoff between them.
+Also look at distributional char., and plausibility of imp.
+values, see \cite{vinknd}.
+
+% Distributional characteristics: In practice, the distribution of the incomplete data may differ greatly from the observed data.
+Under anything but the MCAR assumption, this can be expected.
+When evaluating imputations, the distributional shapes should be checked and diagnostic evaluations should be performed (see Abayomi et al., 2008, for an detailed overview of diagnostic evaluation for multivariate imputations).
+When anomalies are found, and if the imputation method is valid, there should be an explanation, especially in the controlled environment of a properly executed simulation study.
+
+% Plausibility of the imputed values: Plausible imputations - imputations that could be real values if they had been observed - are not a necessary condition for obtaining valid inference.
+However, in practice, especially when the imputer and the analyst are different persons, plausibility of imputations may be a desired property.
+When evaluating imputation routines, the evaluator should mention whether the routine is prone to deliver implausible value
+
 \subsection{Simulation: Simulation Approach}
@@
 Proposed by Gelman and Rubin (1992), but since developed into ...
+% Missingness mechanism: "With MCAR missingness mechanisms, the probability to be missing is the same for all cases.
+This is a necessary simulation condition for evaluating the performance of imputation procedures.
+If an imputation method is not able to solve the problem (i.e.
+yield valid inference) under MCAR, the statistical properties of the procedure are not sound." (Vink, n.d., p.
+4)
 
@@
 "The mixture-of-sequences variance, V, should stabilize as a function of n.
-(Before convergence, we expect $\sigma^2$ to decrease with n, only increasing if the sequences explore a new area of parameter space, which would imply that the original sequences were not overdispersed for the particular scalar summary being monitored.)"
-% \cite[p~438]{broo98}
+(Before convergence, we expect $\sigma^2$ to decrease with n, only increasing if the sequences explore a new area of parameter space, which would imply that the original sequences were not overdispersed for the particular scalar summary being monitored.)" \cite[p~438]{broo98}.
 
@@
 This is, however, dependent of the complete data model.
-The eigenvector decomposition methof proposed by May 9?
-should be implemented.
+The eigenvector decomposition methof proposed by McKay (?) should be implemented.
 I could not find any resources to apply this method and it is outside the scope of this thesis to investigate how this apprach could be implemented.
@@
 
-\section{Models and software} \label{sec:models}
-
-\emph{This is a placeholder.}
+% \section{Models and software} \label{sec:models}
+% 
+% \emph{This is a placeholder.}
 
@@
 
-\section{Illustrations} \label{sec:illustrations}
-
-\emph{This is a placeholder.}
+% \section{Illustrations} \label{sec:illustrations}
+% 
+% \emph{This is a placeholder.}
 
@@
 
-\section{Summary and discussion} \label{sec:summary}
-
-\emph{This is a placeholder.}
+% \section{Summary and discussion} \label{sec:summary}
+% 
+% \emph{This is a placeholder.}
 
@@
 
-The results in this paper were obtained using \proglang{R}~\Sexpr{paste(R.Version()[6:7], collapse = ".")} with the \pkg{mice}~\Sexpr{packageVersion("mice")} package.
+The results in this paper were obtained using \proglang{R}~\Sexpr{paste(R.Version()[6:7], collapse = ".")} with the \pkg{mice}~\Sexpr{paste(packageVersion("mice"), collapse = ".")} package.
 \proglang{R} itself and all packages used are available from the Comprehensive \proglang{R} Archive Network (CRAN) at \url{https://CRAN.R-project.org/}.
@@
 
-This technical paper is written by the sole author (Hanne Oberman, BSc.), with guidance from Master thesis supervisors prof. dr. Stef van Buuren, and dr. Gerko Vink.
+This paper is written by the sole author (Hanne Oberman, BSc.), with guidance from Master thesis supervisors prof. dr. Stef van Buuren, and dr. Gerko Vink.
 
@@
 
-\section{More technical details} \label{app:technical}
+% \section{More technical details} \label{app:technical}
 
@@
 
-\section[Using BibTeX]{Using \textsc{Bib}{\TeX}} \label{app:bibtex}
+% \section[Using BibTeX]{Using \textsc{Bib}{\TeX}} \label{app:bibtex}
 

### 2019-11-11 ResearchReport/article.tex
Cleaned up article file, solved mistake in calculation CI.

@@
 
+bias, variance, cov rate: evaluate problems that you might miss otherwise.
+there is no tradeoff between them.
+then your inference is valid.
+generate data, create miss, usually there is a new method that you want to investigate and then compare to cca.
+
+% Missingness mechanism: "With MCAR missingness mechanisms, the probability to be missing is the same for all cases.
+This is a necessary simulation condition for evaluating the performance of imputation procedures.
+If an imputation method is not able to solve the problem (i.e.
+yield valid inference) under MCAR, the statistical properties of the procedure are not sound." (Vink, n.d., p.
+4)
+
 \subsection{Simulation: Simulation Approach}
@@
   impute the missingness
-  compute convergence and simulation diagnostics
+  compute convergence diagnostic
+  perform analysis
+  pool results
+  compute simulation diagnostics
 save diagnostics
@@
 
-Simulation code is available in appendix A and on Github.
+Simulation code is available in online appendix XYZ on Github.
 

### 2019-11-11 Simulation/1.Execute.R

@@ -18,12 +18,12 @@ source("Functions/Evaluate.R")
 
 # simulation parameters
 populationsize <- 1000 #n of simulated dataset
-n.iter <- 15 #nr of iterations (varying 1:n.iter)
-n.sim <- 10 #nr of simulations per iteration value
+n.iter <- 50 #nr of iterations (varying 1:n.iter)
+n.sim <- 100 #nr of simulations per iteration value
 true_effect <- 2 #regression coefficient to be estimated
 
 # start simulation study
-set.seed(128)
+set.seed(1111)
 
 ###
 
@@ -47,7 +47,7 @@ simulate <- function(runs = 10, n.iter = 5) {
 }
 
 # simulate
-res <- simulate(runs = n.sims, n.iter = n.iter)
+res <- simulate(runs = n.sim, n.iter = n.iter)
 
 ###
 
@@ -61,4 +61,4 @@ res <- simulate(runs = n.sims, n.iter = n.iter)
 ###
 
 # save for future reference
-save.Rdata(out, name = "results.Rdata", path = "Results")
+save.Rdata(out, name = "results3.Rdata", path = "Results")

### 2019-11-11 Simulation/Functions/CreateData.R

@@ -16,7 +16,7 @@ data.simulation <- function(n = populationsize, true_effect = 2) {
   
   # compute  outcome variable from predictors
   simdata$Y <-
-    1 + true_effect * simdata$X + 0.5 * simdata$Z1 - 1 * simdata$Z2 + rnorm(n = populationsize, sd = 10)
+    true_effect * simdata$X + 0.5 * simdata$Z1 - 1 * simdata$Z2 + rnorm(n = populationsize, sd = 10)
   
   # remove values at random with 50 percent probability to be missing
   ampdata <- ampute(simdata, prop = 0.5, mech = "MCAR")$amp

### 2019-11-11 Simulation/Functions/Evaluate.R

@@ -5,7 +5,7 @@ evaluate.sim <- function(sims, true = 2) {
   # compute average over simulations
   RB <- rowMeans(res[, , 1]) - true
   PB <- 100 * abs((rowMeans(res[, , 1]) - true) / true)
-  CR <- rowMeans(res[, , 2] < true & true < res[, , 3])
+  CR <- rowMeans(res[, , 2] <= true & true <= res[, , 3])
   AW <- rowMeans(res[, , 3] - res[, , 2])
   Rh <- rowMeans(res[, , 4])
   AC <- rowMeans(res[, , 5])

### 2019-11-12 Simulation/Writeup/preprint.Rnw
Decided to split the Research Report and the final article (Thesis). RR is in Writeup folder of Simulation now.

--- 2019-11-11 ResearchReport/article.Rnw
@@
 %% recommended packages
-\usepackage{thumbpdf,lmodern}
+\usepackage{thumbpdf,lmodern, amsmath}
 
@@
 %% - \Shorttitle{} with LaTeX markup (if any), used as running title
-\title{\pkg{ShinyMICE}: an Evaluation Suite for Multiple Imputation}
-\Plaintitle{ShinyMICE: an Evaluation Suite for Multiple Imputation}
-\Shorttitle{\pkg{ShinyMICE}: an Evaluation Suite for Multiple Imputation}
+\title{A Note on Convergence Diagnostics for Multiple Imputation using Chained Equations (MICE)}
+\Plaintitle{A Note on Convergence Diagnostics for Multiple Imputation using Chained Equations (MICE)}
+\Shorttitle{Convergence Diagnostics for MICE}
 
@@
 \Abstract{
-  This Research Report contains the Introduction and Methods section of the technical paper that will be submitted for publication in \emph{Journal of Statistical Software}.
-("There is no page limit, nor a limit on the number of figures or tables", see \url{https://www.jstatsoft.org/pages/view/authors}.) I have chosen to use the second format from the Research Report guidelines: "It is the first half of the thesis, i.e., there are no results included yet, but the report contains a full introduction including a literature review and a methods section that contains details about the data, instruments and or statistical procedures".
+  This Research Report contains a simulation study that serves as the basis of the technical paper that will be submitted for publication in \emph{Journal of Statistical Software}.
+I have chosen to use the first format from the Research Report guidelines: \emph{It is written as a (mini) thesis, with an introduction, methods section, some results (i.e., preliminary analyses, or pilot simulations), and a discussion of results.
+The length of the research report should be maximally 2500 words of text (without references list and or tables and figures).
+Please do not include appendices, and no more than 6 tables or figures.
+Table and Figure captions do not count towards the word limit.
+An abstract may be included, but is not necessary.}
+%"It is the first half of the thesis, i.e., there are no results included yet, but the report contains a full introduction including a literature review and a methods section that contains details about the data, instruments and or statistical procedures".
 The goal was to develop novel methodology and guidelines for evaluating multiple imputation methods, and implement these in an interactive evaluation framework for multiple imputation: \pkg{ShinyMICE}.
@@
 %% - Should be comma-separated and in sentence case.
-\Keywords{multiple imputation, evaluation methodology, \pkg{ShinyMICE}, \pkg{mice}, \proglang{R}}
-\Plainkeywords{multiple imputation, evaluation methodology, ShinyMICE, mice, R}
+\Keywords{multiple imputation, convergence, \pkg{mice}, \proglang{R}}
+\Plainkeywords{multiple imputation, convergence, mice, R}
 
@@
 
-%% - In principle "as usual".
-%% - But should typically have some discussion of both _software_ and _methods_.
-%% - Use \proglang{}, \pkg{}, \fct{}, and \code{} markup throughout the manuscript.
-%% - If such markup is in (sub)section titles, a plain text version has to be
-%%   added as well.
-%% - All software mentioned should be properly \cite-d.
-%% - All abbreviations should be introduced.
-%% - Unless the expansions of abbreviations are proper names (like "Journal
-%%   of Statistical Software" above) they should be in sentence case (like
-%%   "generalized linear models" below).
-
-\section[Introduction: Multiple Imputation Methodology]{Introduction: Multiple Imputation Methodology} \label{sec:intro} % curly brackets can handle typesetting
-
-% \begin{leftbar}
-% The introduction is in principle ``as usual''. However, it should usually embed
-% both the implemented \emph{methods} and the \emph{software} into the respective
-% relevant literature. For the latter both competing and complementary software
-% should be discussed (within the same software environment and beyond), bringing
-% out relative (dis)advantages. All software mentioned should be properly
-% \verb|\cite{}|d. (See also Appendix~\ref{app:bibtex} for more details on
-% \textsc{Bib}{\TeX}.) Example: % \citep[][Chapter~7.4]{Venables+Ripley:2002} 
-% 
-% For writing about software JSS requires authors to use the markup
-% \verb|\proglang{}| (programming languages and large programmable systems),
-% \verb|\pkg{}| (software packages), \verb|\code{}| (functions, commands,
-% arguments, etc.). If there is such markup in (sub)section titles (as above), a
-% plain text version has to be provided in the {\LaTeX} command as well. Below we
-% also illustrate how abbrevations should be introduced and citation commands can
-% be employed. See the {\LaTeX} code for more details.
-% \end{leftbar}
+\section{Introduction} \label{sec:intro} % curly brackets can handle typesetting
 
@@
 
-With MI, many assumptions are made about the nature of the observed and missing parts of the data and their relation to the 'true' \emph{data generating model} \citep{buur18}.
+Missing data is ubiquitous. MICE can solve stuff.
+
+Validity of inference MCMC algorithms is threatened by non-convergence.
+But it is not known whether conventional convergence diagnostics work for MI data.
+Aim of this paper is to investigate convergence properties of MI algorithms by means of simulation.
+Results of simulation study can be used as guideline for applied researchers.
+%I will replicate Lacerda et al.'s simulation study on $\widehat{R}$ \citep{lace07}, and develop novel guidelines for assessing convergence.
+Ideally, I will integrate several diagnostics (e.g., $\widehat{R}$, \emph{auto-correlation}, and \emph{simulation error}) into a single summary indicator to flag non-convergence.
+
+% With MI, many assumptions are made about the nature of the observed and missing parts of the data and their relation to the 'true' \emph{data generating model} \citep{buur18}.
 Without proper evaluation of the imputations and the underlying assumptions, any drawn inference may erroneously be deemed valid.
@@
 
-The aim of this paper is to provide applied researchers an introduction to the interactive evaluation device \pkg{ShinyMICE}.
-The intended audience consists of empirical researchers and statisticians who (want to) use multiple imputation to solve missing data problems.
+The aim of this paper is to provide applied researchers a tool to assess convergence of MICE algorithms.
+The intended audience consists of empirical researchers and statisticians who use multiple imputation to solve missing data problems.
 Basic familiarity with multiple imputation methodology is assumed.
@@
 
-The \proglang{R} \pkg{Shiny} application introduced in this paper is developed with the aim to integrate it into the \pkg{mice} environment.
-\pkg{ShinyMICE} therefore follows notation and conventions of \cite{mice}.
+The convergence diagnostic introduced in this paper is developed with the aim to integrate it into the \proglang{R} package \pkg{mice} environment.
+It therefore follows notation and conventions of \cite{mice}.
 For an overview of the deviations from the 'original' notation by \cite{rubin87}, see \cite{buur18}.
@@
 Let $Y$ denote an $n \times p$ matrix containing the data values on $p$ variables for all $n$ units in a sample.
-The collection of observed data values is denoted as $Y_{obs}$; the missing part of $Y$ is refered to as $Y_{mis}$.
+The collection of observed data values is denoted as $Y_{obs}$; the missing part of $Y$ is referred to as $Y_{mis}$.
 Response indicator $R$ shows whether a data value in $Y$ is missing or observed.
@@
 When taken together Y = ( Y o b s , Y m i s ) contain the hypothetically complete data." \cite[par.~2.2]{buur18}
-
 % The data are said to be MCAR if Pr ( R = 0 | Y o b s , Y m i s , $\psi$ ) = Pr ( R = 0 | $\psi$ ) (2.1) so the probability of being missing depends only on some parameters $\psi$ , the overall probability of being missing.
@@
 
-Terminology (MCAR, MAR, MNAR).
-
-Blue points are observed, the red points are imputed.
+% Also introduce: 
+% 
+% - Terminology (MCAR, MAR, MNAR)?
+% 
+% - Blue points are observed, the red points are imputed?
 
@@
 
-- missingness mechanims, ignorability
+Non-convergence has two interpretations [look up source!!!]: mixing between chains and stability over iterations within chains.
+Diagnose non-convergence: $\widehat{R}$, auto-correlation, MC error, Geweke, the other one.
+However, existing methodology may not work on MI data.
+Focus on $\widehat{R}$.
 
-"The practical importance of the distinction between MCAR, MAR and MNAR is that it clarifies the conditions under which we can accurately estimate the scientifically interesting parameters without the need to know $\psi$" \cite[par.~2.2]{buur18}.
+The potential scale reduction factor tells us how much variance could be shrunken down by running the chains infinitely long.
+That tells us something about how dependent the chains are on the starting values.
+If there is no dependence on the initial values anymore, the chains have converged (in the mixing sense of the word).
 
-- Rubin's rules
+As recommended by \cite{veth19} $\widehat{R}$ is computed as .... to be able to detect ... in the tails of the distribution.
 
-- FCS vs. JM?
+Why is R hat not applicable on MI data?
+R hat is not appropriate because it assumes over-dispersed initial values, which means that the initial values of the m imputation chains should be 'far away' from the distribution that the chains are converging to (the mode of the distribution).
+With MI procedures, initial values are chosen such that ...
+or maybe even estimated from the observed data.
+This means that at most one initial value is necessary, but probably none at al.
+I should look this up.
+And, "if over-dispersion does not hold, $\sigma_+^2$ can be too low, which can lead to falsely diagnosing convergence." \cite[p~437]{broo98}.
+Empirical finding suggests otherwise: that R hat will not be smaller than 1.1 before iteration number.
+50.
+Therefore, the aim is to replicate empirical finding Lacerda et al.
 
-\subsection{Theoretical Background from Thesis Proposal}
+Convergence can also be interpreted as stable over iterations, or non-recurring.
+Non-recurrence can be evaluated with auto-correlation.
+Auto-correlation shows how dependent subsequent draws of an imputation chain are on the previous value.
+If there is a lot of dependence, draws at e.g.
+iteration five are significantly correlated with the value of the first draw.
+Ideally, we want the chains to intermingle nicely, without trends within chains.
+Auto-correlation above the confidence level indicate dependence within chains.
 
-The validity of the MI solution depends on numerous assumptions that cannot be verified from the observed data alone.
-So instead of statistical tests for assumptions, evaluation procedures have been developed.
-For the following assumptions, no reliable procedure has been proposed and/or implemented: 1) \emph{ignorability} of the \emph{missingness mechanism} \citep{rubin87}; 2) \emph{congeneality} of the imputation models \citep{meng94}; and 3) \emph{compatibility} of the MI modeling procedure \citep{rubin96}.
-
-\begin{enumerate}
-\item A missingness mechanism is said to be ignorable when the probability to be missing does not depend on the missing data itself.
-Violation of this assumption can gravely affect inferences.
-Robustness of inferences to varying degrees of violation can be assessed with sensitivity analyses.
-Some practical guidelines exist (e.g., \citep{nguy17}), but current MI software does not facilitate this methodology for empirical researchers.
-% use '\citep[see][for further details]{...}' for '(see Author Year for further details)'
+% Also introduce:
 % 
-\item Congeneal imputation models capture all required relations between observed and missing parts of the data.
-The extent to which this has been successful can be evaluated by plotting conditional distributions \citep{abay08}.
-Such visualizations are available in MICE, but subsequent statistical tests to quantify the relations with covariates are not provided.
-% Additionally, there is potential to assess model fit by means of \emph{over-imputation} \citep{buur18} or *double robustness* \citep{bang05}-- topics that are only persued if time permits.
+% - missingness mechanims, ignorability?
+% % "The practical importance of the distinction between MCAR, MAR and MNAR is that it clarifies the conditions under which we can accurately estimate the scientifically interesting parameters without the need to know $\psi$" \cite[par.~2.2]{buur18}.
 % 
-\item The third assumption is met when the MI algorithm converges to a stable distribution.
-However, conventional measures to diagnose convergence-- e.g., Gelman and Rubin's \citeyear{gelm92} statistic $\widehat{R}$ --are not applicable on multiply imputed data \citep{lace07}.
-Therefore, empirical researchers have to rely on visual inspection procedures that are theoretically equivalent to $\widehat{R}$ \citep{whit11}.
-Visually assessing convergence is not only difficult to the untrained eye, it might also be futile.
-The convergence properties of MI algorithms lack scientific consensus \citep{taka17}, and some default MICE techniques might not converge to stable distributions at all \citep{murr18}.
-Moreover, convergence diagnostics for MI methods have not been systematically studied \citep{buur18}.
-\end{enumerate}
-
-In short, the existing literature provides both possibilities and limitations to evaluating the validity of multiply imputed data.
-The goal of this research project is to develop novel methodology and guidelines for evaluating MI methods, and implement these in an interactive evaluation framework for multiple imputation.
-This framework will aid applied researchers in drawing valid inference from incomplete datasets.
+% - Rubin's rules?
+% 
+% - FCS vs. JM?
+%
 
@@
 
-\subsection{Approach from Thesis Proposal}
+The validity of the MI solution depends on numerous assumptions that cannot be verified from the observed data alone.
+So instead of statistical tests for assumptions, evaluation procedures have been developed.
+For the following assumptions, no reliable procedure has been proposed and/or implemented: 1) \emph{ignorability} of the \emph{missingness mechanism} \citep{rubin87}; 2) \emph{congeniality} of the imputation models \citep{meng94}; and 3) \emph{compatibility} of the MI modeling procedure \citep{rubin96}.
 
-Initially, the research project will consist of an investigation into algorithmic convergence of MI algorithms.
-I will replicate Lacerda et al.'s simulation study on $\widehat{R}$ \citep{lace07}, and develop novel guidelines for assessing convergence.
-Ideally, I will integrate several diagnostics (e.g., $\widehat{R}$, \emph{auto-correlation}, and \emph{simulation error}) into a single summary indicator to flag non-convergence.
-
-Subsequently, I will use\proglang{R}Shiny \citep{shiny} to implement the convergence indicator and existing evaluation measures in \pkg{ShinyMICE}, see Figure 1.
-The application will at least contain methodology for: sensitivity analyses; data visualizations (e.g., scatter-plots, densities, cross-tabulations); and statistical evaluation of relations between variables pre- versus post-imputation (i.e., $\chi^2$~tests or $t$~tests).
-
-A working beta version of \pkg{ShinyMICE} will be considered a sufficient milestone to proceed with writing a technical paper on the methodology and the software.
-I will submit the paper for publication in \emph{Journal of Statistical Software}.
-Finally, \pkg{ShinyMICE} will be integrated into the existing MICE environment, and a vignette for applied researchers will be written.
-
-The \proglang{R} code and documentation of this project will be open source (available on Github).
-Since the study does not require the use of unpublished empirical data, I expect that the FETC will grant the label exempt.
-
-% \includegraphics[height=80mm,keepaspectratio]{impression} % this is not correct yet, it needs a figure label and stuff
-
-%% -- Methods: Sensitivity Analyses ---------------------------------------------------------------
-\subsection{Sensitivity Analyses} \label{sec:sensitivity}
-
-% In general: "The MAR assumption can never be tested from the observed data. One can however check whether the imputations created by MICE algorithm are plausible" \cite[p.~42]{mice}.
-
-\subsubsection{What is already implemented?}
-
-- Current guideine: "Plot densities of both the observed and imputed values of all variables to see whether the imputations are reasonable.
-Differences in the densities between the observed and imputed values may suggest a problem that needs to be further checked" \cite[p.~43]{mice}.
-% FIMD on sensitivity analyses, see https://stefvanbuuren.name/fimd/sec-sensitivity.html
-
-\subsubsection{What is not yet implemented, but exists?}
-
-- Robustness of inferences to varying degrees of violation can be assessed with sensitivity analyses. See \citep{nguy17} for guidelines.
-
-- Forked MICE with MNAR sensitivity analyses
-% see \url{https://thestatsgeek.com/2018/06/07/missing-not-at-random-sensitivity-analysis-with-fcs-multiple-imputation/}.
-
-\subsubsection{What is not implemented, and NA?}
-
-- Nothing within the scope of this thesis.
-
-%% -- Methods: Visualizations ---------------------------------------------------------------
-\subsection{Data Visualization} \label{sec:visualization}
-
-% In general: "An important step in multiple imputation is to assess whether imputations are plausible.
-Imputations should be values that could have been obtained had they not been missing.
-Imputations should be close to the data.
-Data values that are clearly impossible (e.g., negative counts, pregnant fathers) should not occur in the imputed data.
-Imputations should respect relations between variables, and reflect the appropriate amount of uncertainty about their 'true' values.
-Diagnostic checks on the imputed data provide a way to check the plausibility of the imputations" \cite[p.~11]{mice}.
-
-\subsubsection{What is already implemented?}
-
-- Existing plot functions to evaluate relations between observed and missing parts of the data (\pkg{mice} functions \fct{hist}, \fct{xyplot}, and \fct{stripplot}). 
-% "The mice package contains several graphic functions that can be used to gain insight into the correspondence of the observed and imputed data: \fct{bwplot}, \fct{stripplot}, \fct{densityplot} and \fct{xyplot}." \cite[par.~6.6]{buur18}
-
-% "It is often useful to inspect the distributions of original and the imputed data.
-One way of doing this is to use the function \fct{stripplot} \dots  The differences between the red points represents our uncertainty about the true (but unknown) values.
-\dots Under MCAR, univariate distributions of the observed and imputed data are expected to be identical.
-Under MAR, they can be different, both in location and spread, but their multivariate distribution is assumed to be identical." \cite[p.~12]{mice}.
-
-\subsubsection{What is not yet implemented, but exists?}
-
-- Improve plots using \pkg{lattice} package functionalities.
-
-- Subsequent statistical tests to quantify the relations with covariates.
-% "Bondarenko and Raghunathan (2016) [see https://www.ncbi.nlm.nih.gov/pubmed/26952693] proposed a more refined diagnostic tool that aims to compare the distributions of observed and imputed data conditional on the missingness probability.
-The idea is that under MAR the conditional distributions should be similar if the assumed model for creating multiple imputations has a good fit.
-An example is created as ..."  \cite[par.~6.6]{buur18}
-% 
-% fit <- with(imp, glm(ici(imp) ~ age + bmi + hyp + chl,
-%                      family = binomial))
-% ps <- rep(rowMeans(sapply(fit$analyses, fitted.values)),
-%           imp$m + 1)
-% xyplot(imp, bmi ~ ps | as.factor(.imp),
-%        xlab = "Probability that record is incomplete",
-%        ylab = "BMI", pch = c(1, 19), col = mdc(1:2))
-
-- Use evaluation of MI methods in SAS as example
-% see https://onlinelibrary.wiley.com/doi/full/10.1111/1467-9574.00219
-
-\subsubsection{What is not implemented, and NA?}
-
-- Nothing within the scope of this thesis.
-
-%% -- Methods: Convergence ---------------------------------------------------------------
-\subsection{Convergence Diagnostic} \label{sec:convergence}
+% \begin{enumerate}
+% \item A missingness mechanism is said to be ignorable when the probability to be missing does not depend on the missing data itself.
+Violation of this assumption can gravely affect inferences.
+Robustness of inferences to varying degrees of violation can be assessed with sensitivity analyses.
+Some practical guidelines exist (e.g., \citep{nguy17}), but current MI software does not facilitate this methodology for empirical researchers.
+% use '\citep[see][for further details]{...}' for '(see Author Year for further details)'
+% %
+% \item Congeneal imputation models capture all required relations between observed and missing parts of the data.
+The extent to which this has been successful can be evaluated by plotting conditional distributions \citep{abay08}.
+Such visualizations are available in MICE, but subsequent statistical tests to quantify the relations with covariates are not provided.
+% % Additionally, there is potential to assess model fit by means of \emph{over-imputation} \citep{buur18} or *double robustness* \citep{bang05}-- topics that are only persued if time permits.
+% %
+%\item 
+The third assumption is met when the MI algorithm converges to a stable distribution.
+However, conventional measures to diagnose convergence-- e.g., Gelman and Rubin's \citeyear{gelm92} statistic $\widehat{R}$ --are not applicable on multiply imputed data \citep{lace07}.
+Therefore, empirical researchers have to rely on visual inspection procedures that are theoretically equivalent to $\widehat{R}$ \citep{whit11}.
+Visually assessing convergence is not only difficult to the untrained eye, it might also be futile.
+The convergence properties of MI algorithms lack scientific consensus \citep{taka17}, and some default MICE techniques might not converge to stable distributions at all \citep{murr18}.
+Moreover, convergence diagnostics for MI methods have not been systematically studied \citep{buur18}.
+%\end{enumerate}
 
@@
 
-- Traceplots
+- Trace-plots
 
@@
 
-- $\widehat{R}$, but too stringent (new) threshold, and assumption of overdispersed initial values of imputation chains not met.
+- $\widehat{R}$, but too stringent (new) threshold, and assumption of over-dispersed initial values of imputation chains not met.
 % Potentially add something about updated (2019) version of $\widehat{R}$ and the new threshold of 1.01, see \citep{veht19}.
 
-- Autocorrelation.
+- Auto-correlation.
 Schafer (1997, p.
 129) wrote on worst linear statistic.
-We could calculate the autocorrelation of that statistic to know that the algorithm converged elsewhere too.
-See autocorr function plot in sas of worst linear function.
+We could calculate the auto-correlation of that statistic to know that the algorithm converged elsewhere too.
+See autocorr function plot in SAS of worst linear function.
 % Worst linear function in SAS see \url{https://support.sas.com/documentation/cdl/en/statug/63033/HTML/default/viewer.htm#statug_mi_sect027.htm}.
@@
 
-- $\hat{R}$ threshold: Replicate simulation study and build a decision rule to solve the problem with $\hat{R}$.
+- $\widehat{R}$ threshold: Replicate simulation study and build a decision rule to solve the problem with $\widehat{R}$.
 % "The monitoring statistic was computed for mean monthly earnings at each iteration in chains of length k = 200.
@@
 
-\section{Simulation study}
+In short, the existing literature provides both possibilities and limitations to evaluating the validity of multiply imputed data.
+The goal of this research project is to develop novel methodology and guidelines for evaluating MI methods, and implement these in an interactive evaluation framework for multiple imputation.
+This framework will aid applied researchers in drawing valid inference from incomplete datasets.
 
-\subsection{Simulation: Introduction}
+This research project consists of an investigation into algorithmic convergence of MI algorithms.
+I will replicate Lacerda et al.'s simulation study on $\widehat{R}$ \citep{lace07}, and develop novel guidelines for assessing convergence.
+Ideally, I will integrate several diagnostics (e.g., $\widehat{R}$, \emph{auto-correlation}, and \emph{simulation error}) into a single summary indicator to flag non-convergence.
 
-\emph{It is written as a (mini) thesis, with an introduction, methods section, some results (i.e., preliminary analyses, or pilot simulations), and a discussion of results.
-The length of the research report should be maximally 2500 words of text (without references list and or tables and figures).
-Please do not include appendices, and no more than 6 tables or figures.
-Table and Figure captions do not count towards the word limit.
-An abstract may be included, but is not necessary.}
-
-R hat not working on MI data?
-"if over-dispersion does not hold, $\sigma_+^2$ can be too low, which can lead to falsely diagnosing convergence." \cite[p~437]{broo98}.
-Aim to replicate Lacerda et al.
-that R hat will not be smaller than 1.1 before iteration nr.
-50.
-
-As recommended by ... Rhat is computed as .... to be able to detect ... in the tails of the distribution.
-
-R hat is not appropriate because it assumes overdispersed initial values, which means that the initial values of the m imputation chains should be 'far away' from the distribution that the chains are converging to (the modus of the distribution).
-With MI procedures, initial values are chosen such that ...
-or maybe even estimated from the observed data.
-This means that at most one initial value is necessary, but probably none at al.
-I should look this up.
-
-The potential scale reduction factor tells us how much variance could be shrunken down by running the chains infinitely long.
-That tells us something about how dependent the chains are on the starting values.
-If there is no dependence on the initial values anymore, the chains have converged (in the mixing sense of the word).
-
-Convergence can also be interpreted as stable over iterations, or non-recurring.
-Non-recurrence can be evaluated with autocorrelation.
-Autocorrelation shows how dependent subsequent draws of an imputation chain are on the previous value.
-If there is a lot of dependence, draws at e.g.
-iteration five are significantly correlated with the value of the first draw.
-Ideally, we want the chains to intermingle nicely, without trends witihin chains.
-Autocorrelation above the confidence level indicate dependence within chains.
-
-\subsection{Simulation: Methodology}
-
-Explain diagnostics here (see below).
-
-We want to evaluate whether the imputation procedure has converged.
-We can apply GR statistic Rhat to the first two moments of the variables of interest (mean and variance).
-The primary research interest is in determining whether Rhat is an appropriate convergence diagnostic, and if so, which level of stringency suits MI data.
+The aim is to evaluate whether the imputation procedure has converged.
+The primary research interest is in determining whether GR statistic $\widehat{R}$ is an appropriate convergence diagnostic, and if so, which level of stringency suits MI data.
+We can apply $\widehat{R}$ to the mean (or to the first two moments) of the variables of interest.
 The simulation diagnostics are as recommended by Stef van Buuren \cite{buur18}.
-That is, absolute bias confidence interval width, and empirical coverage rate (coverage probability?) across simulations.
-
-Hypothesis based on \cite{lace07} is that the conventional acceptable level of Rhat is too strict for MI data.
-We expect that the simulation diagnostics will indicate proper imputations before Rhat will.
-
-Evaluate: bias, variance, and cov rate to find problems that you might miss otherwise.
-There is no tradeoff between them.
-Also look at distributional char., and plausibility of imp.
-values, see \cite{vinknd}.
+That is, average bias, average confidence interval width, and empirical coverage rate (coverage probability) across simulations.
+Also look at distributional characteristics, and plausibility of imputed values, see \cite{vinknd}.
+% Or use: Then evaluate the regular diagnostics of MI simulations, see Vink, n.d., and  convergence.
+Diagnostics include absolute bias of the estimated regression coefficient, confidence interval width, and empirical coverage rate across simulations.
+Convergence is evaluated with $\widehat{R}$, or potential scale reduction factor.
 
@@
 
-\subsection{Simulation: Simulation Approach}
+Hypothesis based on \cite{lace07} is that the conventional acceptable level of $\widehat{R}$ is too strict for MI data.
+We expect that the simulation diagnostics will indicate proper imputations before $\widehat{R}$ will.
 
-Set-up: simulate data, impute missing data points with varying number of iterations.
-Then evaluate the regular diagnostics of MI simulations, see Vink, n.d., and the convergence.
-Diagnostics include absolute bias of the estimated regression coefficient, confidence interval width, and empirical coverage rate (coverage probability?) across simulations.
-Convergence is evaluated with $\hat{R}$, or potential scale reduction factor.
-Proposed by Gelman and Rubin (1992), but since developed into ...
+%% -- Simulation Approach ---------------------------------------------------------------
+
+%%
+%%
+%%
+%%
+
+\section{Simulation Approach}
+
+In this study, 1000 MCMC simulations are performed. The simulation set-up consists of several steps, see pseudo-code below. Simulation code is available in online appendix XYZ on Github.
+
+\begin{Code}
+# pseudo-code of simulation 
+simulate data with missingness
+for (number of simulation runs from 1 to 1000)
+  for (number of iterations from 1 to 100)
+    impute the missingness
+    compute convergence diagnostic
+    perform analysis
+    pool results
+    compute simulation diagnostics
+save convergence and simulation diagnostics
+\end{Code}
+
+The simulated data is a finite population of $N=1000$.
+The variables are dependent variable $Y$, independent variable $X$, and covariates $Z_1$ and $Z_2$.
+The quantity of scientific interest is the estimated regression coefficient of $X$ on $Y$ in linear regression model $Y \sim X+Z_1+Z_2$.
+The data generating model of the predictors is a multivariate normal distribution with means structure $\mu$, and variance-covariance matrix $\Sigma$.
+Outcome variable $Y$ is deduced from the predictor variables as follows:
+
+\begin{align*}
+\begin{pmatrix}X\\
+Z_{1}\\
+Z_{2}\\
+\epsilon
+\end{pmatrix} &\sim  N
+\begin{bmatrix}
+\begin{pmatrix}
+12\\
+3\\
+0.5\\
+0
+\end{pmatrix}\!\!,&
+\begin{pmatrix}
+4 & 4 & 1.8 & 0\\
+4 & 16 & 4.8 & 0\\
+1.8 & 4.8 & 9 & 0\\
+0 & 0 & 0 & 10
+\end{pmatrix}
+\end{bmatrix}\\[2\jot]
+Y &=  2 \times X + 0.5 \times Z_1 - 1 \times Z_2 + \epsilon
+\end{align*}
+
+After data generation, the complete data is 'amputed'.
+That is, the \pkg{mice} function \fct{ampute} is used to impose an MCAR missingness mechanism upon the data.
+The probability to be missing is the same for all cells, namely XYZ\%.
+Table XYZ shows a summary of the generated complete data, and the amputed data.
+The amputed data is the starting point of each of the 1000 simulations per simulation condition.
+
+Missing data points are imputed with \pkg{mice} in \proglang{R}.
+All simulations are performed with imputation method 'norm' (Bayesian linear regression imputation), and five imputation chains.
+The number of iterations is varied over simulations ('maxit' argument between 1 and 100).
+For each number of iterations, simulation and convergence diagnostics are aggregated over 1000 MCMC simulations.
 % Missingness mechanism: "With MCAR missingness mechanisms, the probability to be missing is the same for all cases.
@@
 
-In this study, 1000 MCMC simulations are performed.
+%% -- Results ---------------------------------------------------------------
 
-\begin{Code}
-# pseudocode of similation 
-simulate data with missingness
-for (number of iterations from 1 to 100)
-  impute the missingness
-  compute convergence diagnostic
-  perform analysis
-  pool results
-  compute simulation diagnostics
-save diagnostics
-\end{Code}
+%%
+%%
+%%
+%%
 
-Simulation code is available in online appendix XYZ on Github.
+\section{Results}
 
-\subsection{Simulation: Results}
+Figure \ref{resultssim} shows the results over 1000 simulations. A subset of simulation conditions is presented in Table \ref{subsetresults}.
 
-Rhat is not perfect for MI data.
-The value shoudl theoretically not be smaller than one, yet it happened several times in this study.
-How could Rhat smaller than 1 occur?
+\begin{figure}[h]
+  \resizebox{\textwidth}{!}{ %notice the \resizebox{} command
+        \includegraphics{res3.pdf}
+  }      
+  \caption{Simulation and convergence diagnostics over 1000 MCMC simulations.}
+    \label{resultssim}
+\end{figure}
+
+From the simulation diagnostics (...) it appears that as little as three iterations is sufficient to draw valid inference.
+Convergence diagnostics $\widehat{R}$ and auto-correlation, however, indicate 20-30 iterations.
+Apparently, they are not perfect for MI data.
+
+Moreover, $\widehat{R}$ could  theoretically not be smaller than one, yet it happened several times in this study (see online appendix XYZ).
+How could $\widehat{R}$ smaller than 1 occur?
 The number of simulations is smaller than in 'regular' MCMC processes.
-Therefore, the '$n-1/n)$' correction factor can influence the estimated potential scale reduction facotr.
-This downwards bias is in the opposite direction than expected.
+Therefore, the '$(n-1/n)$' correction factor can influence the estimated potential scale reduction factor.
+This downwards bias is in the opposite direction than expected: "The mixture-of-sequences variance, V, should stabilize as a function of n.
+(Before convergence, we expect $\sigma^2$ to decrease with n, only increasing if the sequences explore a new area of parameter space, which would imply that the original sequences were not overdispersed for the particular scalar summary being monitored.)" \cite[p~438]{broo98}.
 
-\subsection{Simulation Discussion}
+% latex table generated in R 3.6.1 by xtable 1.8-4 package
+% Tue Nov 12 17:43:11 2019
+\begin{table}[ht]
+\centering
+\begin{tabular}{lrrrrrr}
+  \hline
+It. & Bias & Emp. SE & CI width & Cov. rate & $\widehat{R}$ & Auto-corr. \\ 
+  \hline
+  1 & -0.135 & 0.115 & 0.637 & 0.930 &  &   \\ 
+  2 & -0.088 & 0.125 & 0.691 & 0.990 & 1.499 & -0.500 \\ 
+  3 & -0.090 & 0.112 & 0.623 & 0.950 & 1.200 & -0.658 \\ 
+  4 & -0.093 & 0.116 & 0.645 & 0.980 & 1.146 & -0.738 \\ 
+  5 & -0.087 & 0.108 & 0.599 & 0.980 & 1.098 & -0.711 \\ 
+  6 & -0.092 & 0.116 & 0.644 & 0.950 & 1.090 & -0.674 \\ 
+  7 & -0.095 & 0.122 & 0.675 & 1.000 & 1.063 & -0.588 \\ 
+  8 & -0.099 & 0.113 & 0.626 & 0.970 & 1.058 & -0.523 \\ 
+  9 & -0.095 & 0.115 & 0.639 & 1.000 & 1.044 & -0.519 \\ 
+  10 & -0.090 & 0.111 & 0.618 & 0.990 & 1.048 & -0.414 \\ 
+  15 & -0.093 & 0.109 & 0.604 & 0.950 & 1.020 & -0.339 \\ 
+  20 & -0.098 & 0.109 & 0.606 & 0.980 & 1.023 & -0.234 \\ 
+  25 & -0.100 & 0.113 & 0.626 & 0.990 & 1.016 & -0.091 \\ 
+  30 & -0.091 & 0.118 & 0.654 & 0.960 & 1.014 & -0.153 \\ 
+  40 & -0.088 & 0.117 & 0.650 & 0.990 & 1.009 & -0.070 \\ 
+  50 & -0.087 & 0.119 & 0.662 & 0.990 & 1.009 & -0.029 \\ 
+  % 11 & -0.100 & 0.127 & 0.704 & 0.970 & 1.046 & -0.360 \\ 
+  % 12 & -0.096 & 0.115 & 0.641 & 0.980 & 1.032 & -0.354 \\ 
+  % 13 & -0.098 & 0.120 & 0.664 & 0.960 & 1.034 & -0.328 \\ 
+  % 14 & -0.094 & 0.113 & 0.630 & 0.990 & 1.030 & -0.353 \\ 
+  % 15 & -0.093 & 0.109 & 0.604 & 0.950 & 1.020 & -0.339 \\ 
+  % 16 & -0.094 & 0.114 & 0.630 & 0.980 & 1.025 & -0.272 \\ 
+  % 17 & -0.091 & 0.116 & 0.645 & 0.980 & 1.028 & -0.243 \\ 
+  % 18 & -0.091 & 0.116 & 0.645 & 0.990 & 1.023 & -0.179 \\ 
+  % 19 & -0.087 & 0.114 & 0.630 & 0.980 & 1.024 & -0.281 \\ 
+  % 20 & -0.098 & 0.109 & 0.606 & 0.980 & 1.023 & -0.234 \\ 
+  % 21 & -0.102 & 0.116 & 0.646 & 0.960 & 1.017 & -0.192 \\ 
+  % 22 & -0.095 & 0.114 & 0.633 & 0.980 & 1.016 & -0.119 \\ 
+  % 23 & -0.088 & 0.107 & 0.595 & 0.980 & 1.017 & -0.246 \\ 
+  % 24 & -0.097 & 0.112 & 0.621 & 0.980 & 1.016 & -0.148 \\ 
+  % 25 & -0.100 & 0.113 & 0.626 & 0.990 & 1.016 & -0.091 \\ 
+  % 26 & -0.096 & 0.113 & 0.628 & 0.980 & 1.017 & -0.177 \\ 
+  % 27 & -0.096 & 0.117 & 0.650 & 0.980 & 1.014 & -0.167 \\ 
+  % 28 & -0.102 & 0.116 & 0.645 & 0.960 & 1.015 & -0.038 \\ 
+  % 29 & -0.098 & 0.114 & 0.632 & 0.990 & 1.016 & -0.137 \\ 
+  % 30 & -0.091 & 0.118 & 0.654 & 0.960 & 1.014 & -0.153 \\ 
+  % 31 & -0.098 & 0.115 & 0.639 & 0.980 & 1.012 & -0.101 \\ 
+  % 32 & -0.096 & 0.113 & 0.626 & 0.940 & 1.011 & -0.095 \\ 
+  % 33 & -0.089 & 0.117 & 0.652 & 0.970 & 1.011 & -0.088 \\ 
+  % 34 & -0.093 & 0.115 & 0.639 & 0.960 & 1.013 & -0.061 \\ 
+  % 35 & -0.103 & 0.115 & 0.638 & 0.970 & 1.011 & -0.126 \\ 
+  % 36 & -0.090 & 0.113 & 0.630 & 0.980 & 1.010 & -0.161 \\ 
+  % 37 & -0.098 & 0.122 & 0.675 & 0.990 & 1.011 & -0.070 \\ 
+  % 38 & -0.095 & 0.118 & 0.656 & 0.970 & 1.011 & -0.119 \\ 
+  % 39 & -0.091 & 0.121 & 0.670 & 0.980 & 1.009 & -0.100 \\ 
+  % 40 & -0.088 & 0.117 & 0.650 & 0.990 & 1.009 & -0.070 \\ 
+  % 41 & -0.095 & 0.117 & 0.648 & 0.960 & 1.010 & -0.148 \\ 
+  % 42 & -0.095 & 0.118 & 0.657 & 0.990 & 1.009 & -0.117 \\ 
+  % 43 & -0.095 & 0.116 & 0.645 & 0.980 & 1.009 & -0.055 \\ 
+  % 44 & -0.101 & 0.125 & 0.692 & 0.980 & 1.009 & -0.068 \\ 
+  % 45 & -0.085 & 0.117 & 0.651 & 0.980 & 1.008 & -0.086 \\ 
+  % 46 & -0.092 & 0.117 & 0.649 & 0.970 & 1.008 & -0.048 \\ 
+  % 47 & -0.095 & 0.115 & 0.638 & 0.980 & 1.007 & -0.069 \\ 
+  % 48 & -0.096 & 0.112 & 0.620 & 0.970 & 1.009 & -0.076 \\ 
+  % 49 & -0.090 & 0.108 & 0.601 & 0.980 & 1.009 & -0.083 \\ 
+  % 50 & -0.087 & 0.119 & 0.662 & 0.990 & 1.009 & -0.029 \\ 
+   \hline
+\end{tabular}
+\label{subsetresults}
+\end{table}
 
-"The mixture-of-sequences variance, V, should stabilize as a function of n.
-(Before convergence, we expect $\sigma^2$ to decrease with n, only increasing if the sequences explore a new area of parameter space, which would imply that the original sequences were not overdispersed for the particular scalar summary being monitored.)" \cite[p~438]{broo98}.
+%% -- Manuscript ---------------------------------------------------------------
+
+%% -- Illustrations ------------------------------------------------------------
+
+%% -- Summary/conclusions/discussion -------------------------------------------
+%%
+%%
+%%
+%%
+
+\section{Summary and discussion} \label{sec:summary}
+
+This note illustrates that conventional convergence diagnostics behave differently on MI data as compared to the output of 'regular' MCMC algorithms.
+The recommended threshold for $\widehat{R}$ is too stringent for MI data.
 
@@
 This is, however, dependent of the complete data model.
-The eigenvector decomposition methof proposed by McKay (?) should be implemented.
-I could not find any resources to apply this method and it is outside the scope of this thesis to investigate how this apprach could be implemented.
-
-%% -- Manuscript ---------------------------------------------------------------
-
-%% - In principle "as usual" again.
-%% - When using equations (e.g., {equation}, {eqnarray}, {align}, etc.
-%%   avoid empty lines before and after the equation (which would signal a new
-%%   paragraph.
-%% - When describing longer chunks of code that are _not_ meant for execution
-%%   (e.g., a function synopsis or list of arguments), the environment {Code}
-%%   is recommended. Alternatively, a plain {verbatim} can also be used.
-%%   (For executed code see the next section.)
-
-% \section{Models and software} \label{sec:models}
-% 
-% \emph{This is a placeholder.}
-
-% A count response variable $y_i$ ($i = 1, \dots, n$) by assuming a Poisson distribution
-% $y_i \sim \mathrm{Pois}(\mu_i)$. The dependence of the conditional mean
-% $\E[y_i \, | \, x_i] = \mu_i$ on the regressors $x_i$ is then specified via a
-% log link and a linear predictor
-% %
-% \begin{equation} \label{eq:mean}
-% \log(\mu_i) \quad = \quad x_i^\top \beta,
-% \end{equation}
-% %
-% where the regression coefficients $\beta$ are estimated by maximum likelihood
-% (ML) using the iterative weighted least squares (IWLS) algorithm.
-
-% \begin{leftbar}
-% Note that around the \verb|{equation}| above there should be no spaces (avoided
-% in the {\LaTeX} code by \verb|%| lines) so that ``normal'' spacing is used and
-% not a new paragraph started.
-% \end{leftbar}
-
-% Its most important arguments are
-% \begin{Code}
-% glm(formula, data, subset, na.action, weights, offset,
-%   family = gaussian, start = NULL, control = glm.control(...),
-%   model = TRUE, y = TRUE, x = FALSE, ...)
-% \end{Code}
-% where \code{formula} plus \code{data} is the now standard way of specifying
-% regression relationships in \proglang{R}/\proglang{S} introduced in
-% \cite{Chambers+Hastie:1992}. The remaining arguments in the first line
-% (\code{subset}, \code{na.action}, \code{weights}, and \code{offset}) are also
-% standard  for setting up formula-based regression models in
-% \proglang{R}/\proglang{S}. The arguments in the second line control aspects
-% specific to GLMs while the arguments in the last line specify which components
-% are returned in the fitted model object (of class \class{glm} which inherits
-% from \class{lm}). For further arguments to \fct{glm} (including alternative
-% specifications of starting values) see \code{?glm}. For estimating a Poisson
-% model \code{family = poisson} has to be specified.
-
-% \begin{leftbar}
-% As the synopsis above is a code listing that is not meant to be executed,
-% one can use either the dedicated \verb|{Code}| environment or a simple
-% \verb|{verbatim}| environment for this. Again, spaces before and after should be
-% avoided.
-% 
-% Finally, there might be a reference to a \verb|{table}| such as
-% Table~\ref{tab:overview}. Usually, these are placed at the top of the page
-% (\verb|[t!]|), centered (\verb|\centering|), with a caption below the table,
-% column headers and captions in sentence style, and if possible avoiding vertical
-% lines.
-% \end{leftbar}
-
-% \begin{table}[t!]
-% \centering
-% \begin{tabular}{lllp{7.4cm}}
-% \hline
-% Type           & Distribution & Method   & Description \\ \hline
-% GLM            & Poisson      & ML       & Poisson regression: classical GLM,
-%                                            estimated by maximum likelihood (ML) \\
-%                &              & Quasi    & ``Quasi-Poisson regression'':
-%                                            same mean function, estimated by
-%                                            quasi-ML (QML) or equivalently
-%                                            generalized estimating equations (GEE),
-%                                            inference adjustment via estimated
-%                                            dispersion parameter \\
-%                &              & Adjusted & ``Adjusted Poisson regression'':
-%                                            same mean function, estimated by
-%                                            QML/GEE, inference adjustment via
-%                                            sandwich covariances\\
-%                & NB           & ML       & NB regression: extended GLM,
-%                                            estimated by ML including additional
-%                                            shape parameter \\ \hline
-% Zero-augmented & Poisson      & ML       & Zero-inflated Poisson (ZIP),
-%                                            hurdle Poisson \\
-%                & NB           & ML       & Zero-inflated NB (ZINB),
-%                                            hurdle NB \\ \hline
-% \end{tabular}
-% \caption{\label{tab:overview} Overview of various count regression models. The
-% table is usually placed at the top of the page (\texttt{[t!]}), centered
-% (\texttt{centering}), has a caption below the table, column headers and captions
-% are in sentence style, and if possible vertical lines should be avoided.}
-% \end{table}
-
-%% -- Illustrations ------------------------------------------------------------
-
-%% - Virtually all JSS manuscripts list source code along with the generated
-%%   output. The style files provide dedicated environments for this.
-%% - In R, the environments {Sinput} and {Soutput} - as produced by Sweave() or
-%%   or knitr using the render_sweave() hook - are used (without the need to
-%%   load Sweave.sty).
-%% - Equivalently, {CodeInput} and {CodeOutput} can be used.
-%% - The code input should use "the usual" command prompt in the respective
-%%   software system.
-%% - For R code, the prompt "R> " should be used with "+  " as the
-%%   continuation prompt.
-%% - Comments within the code chunks should be avoided - these should be made
-%%   within the regular LaTeX text.
-
-% \section{Illustrations} \label{sec:illustrations}
-% 
-% \emph{This is a placeholder.}
-
-% For a simple illustration of basic Poisson and NB count regression the
-% \code{quine} data from the \pkg{MASS} package is used. This provides the number
-% of \code{Days} that children were absent from school in Australia in a
-% particular year, along with several covariates that can be employed as regressors.
-% The data can be loaded by
-%
-% <<data>>=
-% data("quine", package = "MASS")
-@
-%
-% and a basic frequency distribution of the response variable is displayed in
-% Figure~\ref{fig:quine}.
-
-% \begin{leftbar}
-% For code input and output, the style files provide dedicated environments.
-% Either the ``agnostic'' \verb|{CodeInput}| and \verb|{CodeOutput}| can be used
-% or, equivalently, the environments \verb|{Sinput}| and \verb|{Soutput}| as
-% produced by \fct{Sweave} or \pkg{knitr} when using the \code{render_sweave()}
-% hook. Please make sure that all code is properly spaced, e.g., using
-% \code{y = a + b * x} and \emph{not} \code{y=a+b*x}. Moreover, code input should
-% use ``the usual'' command prompt in the respective software system. For
-% \proglang{R} code, the prompt \code{"R> "} should be used with \code{"+  "} as
-% the continuation prompt. Generally, comments within the code chunks should be
-% avoided -- and made in the regular {\LaTeX} text instead. Finally, empty lines
-% before and after code input/output should be avoided (see above).
-% \end{leftbar}
-
-% \begin{figure}[t!]
-% \centering
-% <<visualization, echo=FALSE, fig=TRUE, height=5.2, width=7>>=
-% par(mar = c(4, 4, 1, 1))
-% plot(table(quine$Days), xlab = "Days", ylab = "Frequency", axes = FALSE)
-% axis(2)
-% axis(1, at = 0:16 * 5, labels = FALSE)
-% axis(1, at = 0:8 * 10)
-% box()
-@
-% \caption{\label{fig:quine} Frequency distribution for number of days absent
-% from school.}
-% \end{figure}
-
-% As a first model for the \code{quine} data, we fit the basic Poisson regression
-% model. (Note that JSS prefers when the second line of code is indented by two
-% spaces.)
-%
-% <<poisson>>=
-% m_pois <- glm(Days ~ (Eth + Sex + Age + Lrn)^2, data = quine,
-%   family = poisson)
-@
-%
-% To account for potential overdispersion we also consider a negative binomial
-% GLM.
-%
-% <<negbin>>=
-% library("MASS")
-% m_nbin <- glm.nb(Days ~ (Eth + Sex + Age + Lrn)^2, data = quine)
-@
-%
-% In a comparison with the BIC the latter model is clearly preferred.
-% %
-% <<comparison>>=
-% BIC(m_pois, m_nbin)
-@
-% %
-% Hence, the full summary of that model is shown below.
-% %
-% <<summary>>=
-% summary(m_nbin)
-@
-
-%% -- Summary/conclusions/discussion -------------------------------------------
-
-% \section{Summary and discussion} \label{sec:summary}
-% 
-% \emph{This is a placeholder.}
-
-% \begin{leftbar}
-% As usual \dots
-% \end{leftbar}
+The eigenvector decomposition method proposed by McKay (?) should be implemented.
+I could not find any resources to apply this method and it is outside the scope of this thesis to investigate how this approach could be implemented.
 
@@
 
-% \begin{leftbar}
-% If necessary or useful, information about certain computational details
-% such as version numbers, operating systems, or compilers could be included
-% in an unnumbered section. Also, auxiliary packages (say, for visualizations,
-% maps, tables, \dots) that are not cited in the main text can be credited here.
-% \end{leftbar}
-
 The results in this paper were obtained using \proglang{R}~\Sexpr{paste(R.Version()[6:7], collapse = ".")} with the \pkg{mice}~\Sexpr{paste(packageVersion("mice"), collapse = ".")} package.
@@
 \section*{Acknowledgments}
-
-% \begin{leftbar}
-% All acknowledgments (note the AE spelling) should be collected in this
-% unnumbered section before the references. It may contain the usual information
-% about funding and feedback from colleagues/reviewers/etc. Furthermore,
-% information such as relative contributions of the authors may be added here
-% (if any).
-% \end{leftbar}
 
@@
 
-% \section{More technical details} \label{app:technical}
-
-% \begin{leftbar}
-% Appendices can be included after the bibliography (with a page break). Each
-% section within the appendix should have a proper section title (rather than
-% just \emph{Appendix}).
-% 
-% For more technical style details, please check out JSS's style FAQ at
-% \url{https://www.jstatsoft.org/pages/view/style#frequently-asked-questions}
-% which includes the following topics:
-% \begin{itemize}
-%   \item Title vs.\ sentence case.
-%   \item Graphics formatting.
-%   \item Naming conventions.
-%   \item Turning JSS manuscripts into \proglang{R} package vignettes.
-%   \item Trouble shooting.
-%   \item Many other potentially helpful details\dots
-% \end{itemize}
-% \end{leftbar}
-
-% \section[Using BibTeX]{Using \textsc{Bib}{\TeX}} \label{app:bibtex}
-
-% \begin{leftbar}
-% References need to be provided in a \textsc{Bib}{\TeX} file (\code{.bib}). All
-% references should be made with \verb|\cite|, \verb|\citet|, \verb|\citep|,
-% \verb|\citealp| etc.\ (and never hard-coded). This commands yield different
-% formats of author-year citations and allow to include additional details (e.g.,
-% pages, chapters, \dots) in brackets. In case you are not familiar with these
-% commands see the JSS style FAQ for details.
-% 
-% JSS requires the following format.
-% \begin{itemize}
-%   \item JSS-specific markup (\verb|\proglang|, \verb|\pkg|, \verb|\code|) should
-%     be used in the references.
-%   \item Titles should be in title case.
-%   \item Journal titles should not be abbreviated and in title case.
-%   \item DOIs should be included where available.
-%   \item Software should be properly cited as well. For \proglang{R} packages
-%     \code{citation("pkgname")} typically provides a good starting point.
-% \end{itemize}
-% \end{leftbar}
-
 \end{appendix}

### 2019-11-12 Simulation/Writeup/preprint.tex
Decided to split the Research Report and the final article (Thesis). RR is in Writeup folder of Simulation now.

--- 2019-11-12 Simulation/Writeup/preprint.Rnw
@@
 Therefore, the '$(n-1/n)$' correction factor can influence the estimated potential scale reduction factor.
-This downwards bias is in the opposite direction than expected: "The mixture-of-sequences variance, V, should stabilize as a function of n.
-(Before convergence, we expect $\sigma^2$ to decrease with n, only increasing if the sequences explore a new area of parameter space, which would imply that the original sequences were not overdispersed for the particular scalar summary being monitored.)" \cite[p~438]{broo98}.
+This downwards bias is in the opposite direction than expected.
 
@@
 
-This note illustrates that conventional convergence diagnostics behave differently on MI data as compared to the output of 'regular' MCMC algorithms.
-The recommended threshold for $\widehat{R}$ is too stringent for MI data.
+"The mixture-of-sequences variance, V, should stabilize as a function of n.
+(Before convergence, we expect $\sigma^2$ to decrease with n, only increasing if the sequences explore a new area of parameter space, which would imply that the original sequences were not overdispersed for the particular scalar summary being monitored.)" \cite[p~438]{broo98}.
 
@@
 
-The results in this paper were obtained using \proglang{R}~\Sexpr{paste(R.Version()[6:7], collapse = ".")} with the \pkg{mice}~\Sexpr{paste(packageVersion("mice"), collapse = ".")} package.
+The results in this paper were obtained using \proglang{R}~3.6.1 with the \pkg{mice}~3.6.0.9000 package.
 \proglang{R} itself and all packages used are available from the Comprehensive \proglang{R} Archive Network (CRAN) at \url{https://CRAN.R-project.org/}.

### 2019-11-12 Simulation/1.Execute.R

@@ -8,6 +8,7 @@ library(mice)
 library(mvtnorm)
 library(miceadds)
 library(dplyr)
+library(ggplot2)
 
 # load simulation/evaluation functions
 source("Functions/CreateData.R")
@@ -31,7 +32,7 @@ set.seed(1111)
 simulate <- function(runs = 10, n.iter = 5) {
   pb <- txtProgressBar(min = 0, max = runs, style = 3)
   # object for output
-  res <- array(NA, dim = c(n.iter, runs, 5))
+  res <- array(NA, dim = c(n.iter, runs, 6))
   # simulate data once
   data <- data.simulation(n = poulationsize)
   # repeat mi procedure 'runs'  times for each nr of iterations
@@ -57,8 +58,11 @@ res <- simulate(runs = n.sim, n.iter = n.iter)
 ###
 
 # plot
+# load("Results/results2.Rdata.Rdata")
+# out <- dat
+plot.ts(out, main = "", xlab = "Number of iterations")
 
 ###
 
 # save for future reference
-save.Rdata(out, name = "results3.Rdata", path = "Results")
+save.Rdata(out, name = "results4.Rdata", path = "Results")

### 2019-11-12 Simulation/Functions/CreateData.R

@@ -19,7 +19,7 @@ data.simulation <- function(n = populationsize, true_effect = 2) {
     true_effect * simdata$X + 0.5 * simdata$Z1 - 1 * simdata$Z2 + rnorm(n = populationsize, sd = 10)
   
   # remove values at random with 50 percent probability to be missing
-  ampdata <- ampute(simdata, prop = 0.5, mech = "MCAR")$amp
+  ampdata <- ampute(simdata, prop = 0.8, mech = "MCAR")$amp
 
   # output
   return(ampdata)

### 2019-11-12 Simulation/Writeup/preprint.tex
Added equations for $\widehat{R}$

--- 2019-11-12 Simulation/Writeup/preprint.Rnw
@@
 If there is no dependence on the initial values anymore, the chains have converged (in the mixing sense of the word).
-
-As recommended by \cite{veth19} $\widehat{R}$ is computed as .... to be able to detect ... in the tails of the distribution.
 
@@
 
+As recommended by \cite{veth19} $\widehat{R}$ is computed as .... to be able to detect ... in the tails of the distribution.
+
+\begin{equation}
+B=\frac{N}{M-1} \sum_{m=1}^{M}\left(\bar{\theta}^{(\cdot m)}-\bar{\theta}^{(\cdot \cdot)}\right)^{2}, \quad \text { where } \quad \bar{\theta}^{(\cdot m)}=\frac{1}{N} \sum_{n=1}^{N} \theta^{(n m)}, \quad \bar{\theta}^{(\cdot \cdot)}=\frac{1}{M} \sum_{m=1}^{M} \bar{\theta}^{(\cdot m)}
+\end{equation}
+
+\begin{equation}
+W=\frac{1}{M} \sum_{m=1}^{M} s_{j}^{2}, \quad \text { where } \quad s_{m}^{2}=\frac{1}{N-1} \sum_{n=1}^{N}\left(\theta^{(n m)}-\bar{\theta}^{(\cdot m)}\right)^{2}
+\end{equation}
+
+\begin{equation}
+\widehat{\operatorname{var}}^{+}(\theta | y)=\frac{N-1}{N} W+\frac{1}{N} B
+\end{equation}
+
+\begin{equation}
+\widehat{R}=\sqrt{\frac{\widehat{\operatorname{var}}^{+}(\theta | y)}{W}}
+\end{equation}
+
 Hypothesis based on \cite{lace07} is that the conventional acceptable level of $\widehat{R}$ is too strict for MI data.
@@
 1.8 & 4.8 & 9 & 0\\
-0 & 0 & 0 & 10
+0 & 0 & 0 & 100
 \end{pmatrix}
@@
 4)
+
+Add "post-processing steps (e.g.
+aggregation computations, summary statistics, regressions on the simulation results) used to transform simulation outputs to reported results".
+%, see \url{http://jsterman.scripts.mit.edu/docs/Rahmandad-Sterman%20Simulation%20Reporting%20Standards3.30%20update.pdf}.
 
@@
 \end{tabular}
+\caption{Simulation and convergence diagnostics over 1000 MCMC simulations.}
 \label{subsetresults}
 \end{table}
+
+Add simulation time (computational cost) to table? Add empirical SE of diagnostics?
 

### 2019-11-13 Simulation/Writeup/preprint.Rnw
Added equations for $\widehat{R}$
Draft research report for peer feedback research seminar

@@
 %% recommended packages
-\usepackage{thumbpdf,lmodern, amsmath}
+\usepackage{thumbpdf,lmodern}
 
@@
 % \usepackage{framed}
+\usepackage{multirow, amsmath}
 
@@
   This Research Report contains a simulation study that serves as the basis of the technical paper that will be submitted for publication in \emph{Journal of Statistical Software}.
-I have chosen to use the first format from the Research Report guidelines: \emph{It is written as a (mini) thesis, with an introduction, methods section, some results (i.e., preliminary analyses, or pilot simulations), and a discussion of results.
+I have chosen to use the first format from the Research Report guidelines: ``\emph{It is written as a (mini) thesis, with an introduction, methods section, some results (i.e., preliminary analyses, or pilot simulations), and a discussion of results.
 The length of the research report should be maximally 2500 words of text (without references list and or tables and figures).
@@
 Table and Figure captions do not count towards the word limit.
-An abstract may be included, but is not necessary.}
+An abstract may be included, but is not necessary}''.
+I aim to publish a pre-print of this research report on ArXiv (\url{https://arxiv.org/}).
+That way, I can refer to the simulation study described here, without 'bulking up' the technical paper that I want to submit for publication in JSS.
+Another option would be to attach this note as online appendix to the technical paper on ShinyMICE.
 %"It is the first half of the thesis, i.e., there are no results included yet, but the report contains a full introduction including a literature review and a methods section that contains details about the data, instruments and or statistical procedures".
@@
 imputed) several times.
-The variability between the imputations validly reflects how much uncertainty in the inference is due to missing information--that is, if all statistical assumptions are met \cite{rubin87}.
+[Add some stuff here about MICE.] The variability between the imputations validly reflects how much uncertainty in the inference is due to missing information--that is, if all statistical assumptions are met \cite{rubin87}.
+%Missing data is ubiquitous. MICE can solve stuff.
 
-Missing data is ubiquitous. MICE can solve stuff.
+With MI, many assumptions are made about the nature of the observed and missing parts of the data and their relation to the `true' \emph{data generating model} \citep{buur18}.
+Without proper evaluation of the imputations and the underlying assumptions, any drawn inference may erroneously be deemed valid.
+Such evaluation measures are currently missing or under-developed in MI software, like the world leading \proglang{R} package \pkg{mice} \citep{mice}.
+%Therefore, I will answer the following question: 'Which measures are vital for evaluating the validity of multiply imputed data?'.
 
-Validity of inference MCMC algorithms is threatened by non-convergence.
-But it is not known whether conventional convergence diagnostics work for MI data.
-Aim of this paper is to investigate convergence properties of MI algorithms by means of simulation.
-Results of simulation study can be used as guideline for applied researchers.
+A fundamental assumption of MICE is convergence of the algorithm.
+MICE is a sort of Gibbs sampler, a type of MCMC algorithm.
+In general, the validity of inference resulting from MCMC algorithms is threatened by non-convergence.
+So we use convergence diagnostics to flag non-convergence.
+But it is not known whether conventional convergence diagnostics for MCMC methods work on MI data.
+
+Conventional thresholds to diagnose non-convergence-- e.g., Gelman and Rubin's \citeyear{gelm92} statistic $\widehat{R} < 1.1$ --are not applicable on multiply imputed data \citep{lace07}.
+Therefore, empirical researchers have to rely on visual inspection procedures that are theoretically equivalent to $\widehat{R}$ \citep{whit11}.
+Visually assessing convergence is not only difficult to the untrained eye, it might also be futile.
+The convergence properties of MI algorithms lack scientific consensus \citep{taka17}, and some default MICE techniques might not converge to stable distributions at all \citep{murr18}.
+Moreover, convergence diagnostics for MI methods have not been systematically studied \citep{buur18}.
+
+This paper investigates convergence properties of MI algorithms by means of model-based simulation in \proglang{R} \cite{R}.
+The results of this simulation study are guidelines for applied researchers to evaluate convergence of MI algorithms.
 %I will replicate Lacerda et al.'s simulation study on $\widehat{R}$ \citep{lace07}, and develop novel guidelines for assessing convergence.
 Ideally, I will integrate several diagnostics (e.g., $\widehat{R}$, \emph{auto-correlation}, and \emph{simulation error}) into a single summary indicator to flag non-convergence.
-
-% With MI, many assumptions are made about the nature of the observed and missing parts of the data and their relation to the 'true' \emph{data generating model} \citep{buur18}.
-Without proper evaluation of the imputations and the underlying assumptions, any drawn inference may erroneously be deemed valid.
-Such evaluation measures are currently missing or under-developed in MI software, like the world leading \proglang{R} package \pkg{mice} \citep{mice}.
-Therefore, I will answer the following question: 'Which measures are vital for evaluating the validity of multiply imputed data?'.
 
@@
 All programming code used in this paper is available in the file 'XYZ.R' along with the manuscript, and on Github repository 'XYZ'.
-
 % "The intended audience of this paper consists of applied researchers who want to address problems caused by missing data by multiple imputation.
@@
 
-The convergence diagnostic introduced in this paper is developed with the aim to integrate it into the \proglang{R} package \pkg{mice} environment.
+The convergence guidelines introduced in this paper are developed to be integrated into the \pkg{mice} environment \cite{mice} in \proglang{R} \cite{R}.
 It therefore follows notation and conventions of \cite{mice}.
@@
 The relation between $R$, $Y_{obs}$, and $Y_{mis}$ determines the missingness mechanism.
+This paper only considers a 'missing completely at random' mechanism, where the probability to be missing is equal for all $n \times p$ cells in $Y$.
+
+With multiple imputation, $Y_{mis}$ is 'filled in' (i.e.
+imputed) $m$ times.
+The $m$ imputed data sets are analyzed separately, and their results pooled.
+A distinction is made between the imputation step (employing $p$ 'imputation models'), and the analysis step (using one 'complete-data model' for each of the $m$ imputated data sets).
+The quantity of scientific interest is the result of the analysis step, e.g., a regression coefficient, and denoted with $Q$.
+That is, $Q$ is estimated on each imputed data set, resulting in $m$ separate $\hat{Q}$ values.
+These are combined into pooled estimate $\bar{Q}$.
+% "Let Q denote the quantity of scientific interest (e.g., a regression coefficient)." \cite[p.~4]{mice}.
 
@@
 
-Non-convergence has two interpretations [look up source!!!]: mixing between chains and stability over iterations within chains.
-Diagnose non-convergence: $\widehat{R}$, auto-correlation, MC error, Geweke, the other one.
-However, existing methodology may not work on MI data.
-Focus on $\widehat{R}$.
+Convergence has two interpretations [look up source!!!]: mixing between chains and stability over iterations within chains.
+Ideally, we want the chains to intermingle nicely, without trends within chains.
+% "In order to converge to a stationary distribution, a Markov chain needs to satisfy three important properties (Roberts 1996; Tierney 1996): irreducible, the chain must be able to reach all interesting parts of the state space; aperiodic, the chain should not oscillate between different states; recurrence, all interesting parts can be reached infinitely often, at least from almost all starting points.
+Do these properties hold for the MICE algorithm?
+Irreducibility is generally not a problem since the user has large control over the interesting parts of the state space.
+This flexibility is actually the main rationale for FCS instead of a joint model.
+Periodicity is a potential problem, and can arise in the situation where imputation models are clearly inconsistent.
+A rather artificial example of an oscillatory behavior occurs when Y 1 is imputed by Y 2 $\beta$ + $\epsilon$  1 and Y 2 is imputed by - Y 1 $\beta$ + $\epsilon$  2 for some fixed, nonzero $\beta$ .
+The sampler will oscillate between two qualitatively different states, so the correlation between Y 1 and Y 2 after imputing Y 1 will differ from that after imputing Y 2 .
+In general, we would like the statistical inferences to be independent of the stopping point.
+A way to diagnose the ping-pong problem, or order effect, is to stop the chain at different points.
+The stopping point should not affect the statistical inferences.
+The addition of noise to create imputations is a safeguard against periodicity, and allows the sampler to "break out" more easily.
+Non-recurrence may also be a potential difficulty, manifesting itself as explosive or non-stationary behavior.
+For example, if imputations are made by deterministic functions, the Markov chain may lock up.
+Such cases can sometimes be diagnosed from the trace lines of the sampler.
+See Section 6.5.2 for an example.
+As long as the parameters of imputation models are estimated from the data, non-recurrence is mild or absent.
+The required properties of the MCMC method can be translated into conditions on the eigenvalues of the matrix of transition probabilities (MacKay 2003, 372-73).
+The development of practical tools that put these conditions to work for multiple imputation is still an ongoing research problem." \cite[par.~4.5]{buur18}
+We can never be certain of convergence, therefore convergence diagnostics evaluate signs of non-convergence \cite{hoff09}.
+% In general: "In general, you cannot know for sure if your chain has converged.
+But sometimes you can know if your chain has not converged, so we at least check for this latter possibility" \cite[p.~101]{hoff09}
+Diagnose non-convergence in the mixing sense: $\widehat{R}$. Diagnose non-convergence in the stability sense: auto-correlation.
 
-The potential scale reduction factor tells us how much variance could be shrunken down by running the chains infinitely long.
-That tells us something about how dependent the chains are on the starting values.
-If there is no dependence on the initial values anymore, the chains have converged (in the mixing sense of the word).
-
-As recommended by \cite{veth19} $\widehat{R}$ is computed as .... to be able to detect ... in the tails of the distribution.
-
-Why is R hat not applicable on MI data?
-R hat is not appropriate because it assumes over-dispersed initial values, which means that the initial values of the m imputation chains should be 'far away' from the distribution that the chains are converging to (the mode of the distribution).
-With MI procedures, initial values are chosen such that ...
-or maybe even estimated from the observed data.
-This means that at most one initial value is necessary, but probably none at al.
-I should look this up.
-And, "if over-dispersion does not hold, $\sigma_+^2$ can be too low, which can lead to falsely diagnosing convergence." \cite[p~437]{broo98}.
-Empirical finding suggests otherwise: that R hat will not be smaller than 1.1 before iteration number.
-50.
-Therefore, the aim is to replicate empirical finding Lacerda et al.
-
-Convergence can also be interpreted as stable over iterations, or non-recurring.
+In the latter interpretation of convergence, convergence is interpreted as stability over iterations, or non-recurring.
 Non-recurrence can be evaluated with auto-correlation.
@@
 iteration five are significantly correlated with the value of the first draw.
-Ideally, we want the chains to intermingle nicely, without trends within chains.
 Auto-correlation above the confidence level indicate dependence within chains.
+Also, possible to use MC error, Geweke, and the other one.
+But outside of scope of this study.
+Here, the focus is mainly on $\widehat{R}$.
+
+The potential scale reduction factor $\widehat{R}$ tells us how much variance of an estimate could be shrunken down by running the chains infinitely long.
+That then tells us something about how dependent the chains are on the starting values.
+If there is no dependence on the initial values anymore, the chains have converged (in the mixing sense of the word).
+$\widehat{R}$ is then equal to one.
+Conventional threshold was $\widehat{R} < 1.2$ is acceptable.
+Most recent guideline by \cite{veht19} is more stringent: $\widehat{R} < 1.01$.
+
+Why is $\widehat{R}$ not applicable on MI data?
+$\widehat{R}$ is not appropriate because it assumes over-dispersed initial values, which means that the initial values of the $m$ imputation chains are 'far away' from the distribution that the chains are converging to.
+With MI procedures, initial values are chosen such that ...
+[or maybe even estimated from the observed data??].
+This means that at most one initial value is necessary, but probably none at all.
+[I should look this up.] Without over-dispersed initial states, $\widehat{R}$ can falsely diagnose convergence \cite{broo98}.
+%would not be sensitive enough to flag non-convergence of MI algorithms: "if over-dispersion does not hold, $\sigma_+^2$ can be too low, which can lead to falsely diagnosing convergence." \cite[p~437]{broo98}.
+This suggests that $\widehat{R}$ would not be sensitive enough to flag non-convergence of MI algorithms.
+Empirical finding, however, show that the opposite may be true: \cite{lace07} report $\widehat{R}$ values above the threshold of $\widehat{R} < 1.1$ after fifty iterations.
+%that $\widehat{R}$ will not be smaller than 1.1 before iteration number 50.
+Therefore, the aim is to replicate empirical finding Lacerda et al.
+Hypothesis of this simulation study is that $\widehat{R}$ will over-estimate non-convergence.
+[Or use:] Hypothesis based on \cite{lace07} is that the conventional acceptable level of $\widehat{R}$ is too strict for MI data.
+We expect that the simulation diagnostics will indicate valid inference before $\widehat{R}$ will.
 
@@
 
-The validity of the MI solution depends on numerous assumptions that cannot be verified from the observed data alone.
-So instead of statistical tests for assumptions, evaluation procedures have been developed.
-For the following assumptions, no reliable procedure has been proposed and/or implemented: 1) \emph{ignorability} of the \emph{missingness mechanism} \citep{rubin87}; 2) \emph{congeniality} of the imputation models \citep{meng94}; and 3) \emph{compatibility} of the MI modeling procedure \citep{rubin96}.
-
-% \begin{enumerate}
-% \item A missingness mechanism is said to be ignorable when the probability to be missing does not depend on the missing data itself.
-Violation of this assumption can gravely affect inferences.
-Robustness of inferences to varying degrees of violation can be assessed with sensitivity analyses.
-Some practical guidelines exist (e.g., \citep{nguy17}), but current MI software does not facilitate this methodology for empirical researchers.
-% use '\citep[see][for further details]{...}' for '(see Author Year for further details)'
-% %
-% \item Congeneal imputation models capture all required relations between observed and missing parts of the data.
-The extent to which this has been successful can be evaluated by plotting conditional distributions \citep{abay08}.
-Such visualizations are available in MICE, but subsequent statistical tests to quantify the relations with covariates are not provided.
-% % Additionally, there is potential to assess model fit by means of \emph{over-imputation} \citep{buur18} or *double robustness* \citep{bang05}-- topics that are only persued if time permits.
-% %
-%\item 
-The third assumption is met when the MI algorithm converges to a stable distribution.
-However, conventional measures to diagnose convergence-- e.g., Gelman and Rubin's \citeyear{gelm92} statistic $\widehat{R}$ --are not applicable on multiply imputed data \citep{lace07}.
-Therefore, empirical researchers have to rely on visual inspection procedures that are theoretically equivalent to $\widehat{R}$ \citep{whit11}.
-Visually assessing convergence is not only difficult to the untrained eye, it might also be futile.
-The convergence properties of MI algorithms lack scientific consensus \citep{taka17}, and some default MICE techniques might not converge to stable distributions at all \citep{murr18}.
-Moreover, convergence diagnostics for MI methods have not been systematically studied \citep{buur18}.
-%\end{enumerate}
-
-% In general: "In general, you cannot know for sure if your chain has converged.
-But sometimes you can know if your chain has not converged, so we at least check for this latter possibility" \cite[p.~101]{hoff09}
-
-% "In order to converge to a stationary distribution, a Markov chain needs to satisfy three important properties (Roberts 1996; Tierney 1996): irreducible, the chain must be able to reach all interesting parts of the state space; aperiodic, the chain should not oscillate between different states; recurrence, all interesting parts can be reached infinitely often, at least from almost all starting points.
-Do these properties hold for the MICE algorithm?
-Irreducibility is generally not a problem since the user has large control over the interesting parts of the state space.
-This flexibility is actually the main rationale for FCS instead of a joint model.
-Periodicity is a potential problem, and can arise in the situation where imputation models are clearly inconsistent.
-A rather artificial example of an oscillatory behavior occurs when Y 1 is imputed by Y 2 $\beta$ + $\epsilon$  1 and Y 2 is imputed by - Y 1 $\beta$ + $\epsilon$  2 for some fixed, nonzero $\beta$ .
-The sampler will oscillate between two qualitatively different states, so the correlation between Y 1 and Y 2 after imputing Y 1 will differ from that after imputing Y 2 .
-In general, we would like the statistical inferences to be independent of the stopping point.
-A way to diagnose the ping-pong problem, or order effect, is to stop the chain at different points.
-The stopping point should not affect the statistical inferences.
-The addition of noise to create imputations is a safeguard against periodicity, and allows the sampler to "break out" more easily.
-Non-recurrence may also be a potential difficulty, manifesting itself as explosive or non-stationary behavior.
-For example, if imputations are made by deterministic functions, the Markov chain may lock up.
-Such cases can sometimes be diagnosed from the trace lines of the sampler.
-See Section 6.5.2 for an example.
-As long as the parameters of imputation models are estimated from the data, non-recurrence is mild or absent.
-The required properties of the MCMC method can be translated into conditions on the eigenvalues of the matrix of transition probabilities (MacKay 2003, 372-73).
-The development of practical tools that put these conditions to work for multiple imputation is still an ongoing research problem." \cite[par.~4.5]{buur18}
-
-\subsubsection{What is already implemented?}
-
-- Trace-plots
-
-\subsubsection{What is not yet implemented, but exists?}
-
-- $\widehat{R}$, but too stringent (new) threshold, and assumption of over-dispersed initial values of imputation chains not met.
-% Potentially add something about updated (2019) version of $\widehat{R}$ and the new threshold of 1.01, see \citep{veht19}.
-
-- Auto-correlation.
-Schafer (1997, p.
-129) wrote on worst linear statistic.
-We could calculate the auto-correlation of that statistic to know that the algorithm converged elsewhere too.
-See autocorr function plot in SAS of worst linear function.
-% Worst linear function in SAS see \url{https://support.sas.com/documentation/cdl/en/statug/63033/HTML/default/viewer.htm#statug_mi_sect027.htm}.
-% Note: we're talking about missing data only, not the combined data (that autocorrelation is very high, as is the autocorrelation of deductively imputed values, like in the texp example). 
-% "Applications of MICE with lowly correlated data therefore inject a lot of noise into the system.
-Hence, the auto-correlation over t will be low, and convergence will be rapid, and in fact immediate if all variables are independent.
-Thus, the incorporation of noise into the imputed data has pleasant side-effect of speeding up convergence" \citep{buur18}, par.
-4.5.
-
-- Sensitivity analysis: Run algorithm several times and compare results. 
-% "Imputers who do choose to use FCS should use flexible univariate models wherever possible and take care to assess apparent convergence of the algorithm, for example by computing traces of pooled estimates or other statistics and using standard MCMC diagnostics (Gelman et al., 2013, Chapter 11).
-It may also be helpful to examine the results of many independent runs of the algorithm with different initializations and to use random scans over the p variables to try to identify any convergence issues and mitigate possible order dependence" \cite[p.~19]{murr18}.
-
-\subsubsection{What is not implemented, and NA?}
-
-- $\widehat{R}$ threshold: Replicate simulation study and build a decision rule to solve the problem with $\widehat{R}$.
-% "The monitoring statistic was computed for mean monthly earnings at each iteration in chains of length k = 200.
-Since calculation of the statistic requires M parallel sequences, m = 5 such chains were constructed.
-This value of m was informed by the preferred choice given in the literature on multiple imputation.
-The monitoring statistic computed at each iteration is presented in Figures 15 to 17 for each of the three missingness mechanisms.
-The red vertical line denotes ten iterations" \cite[p.~49]{lace07}.
-
-- Stability of the solution: Possibly use the slope of means over iterations too to see whether there is trending.
-Or apply PCA on the imputed data and if that (the eigenvalues?) stays the same we know that the means and variances are stable as well, see McKay (?).
-
-- MC error: MC error = SD/sqrt(number of iterations), where SD represents the variation across iterations.
-The MC error thus represents how much the means differ w.r.t.
-the iterations.
-MC error decreases as number of iterations increases.
-It should not be larger than 5\% of the sample standard deviation.
-
-In short, the existing literature provides both possibilities and limitations to evaluating the validity of multiply imputed data.
-The goal of this research project is to develop novel methodology and guidelines for evaluating MI methods, and implement these in an interactive evaluation framework for multiple imputation.
-This framework will aid applied researchers in drawing valid inference from incomplete datasets.
-
 This research project consists of an investigation into algorithmic convergence of MI algorithms.
@@
 The aim is to evaluate whether the imputation procedure has converged.
-The primary research interest is in determining whether GR statistic $\widehat{R}$ is an appropriate convergence diagnostic, and if so, which level of stringency suits MI data.
+The primary research interest is in determining whether $\widehat{R}$ is an appropriate convergence diagnostic, and if so, which level of stringency suits MI data.
 We can apply $\widehat{R}$ to the mean (or to the first two moments) of the variables of interest.
-The simulation diagnostics are as recommended by Stef van Buuren \cite{buur18}.
-That is, average bias, average confidence interval width, and empirical coverage rate (coverage probability) across simulations.
+The simulation diagnostics are as recommended by \cite{buur18}.
+Namely, average bias, average confidence interval width, and empirical coverage rate (coverage probability) across simulations.
+%"We expect that in accordance with Neyman (1934), at least 95 percent of the confidence intervals should contain the true population value, although some room for simulation error should be taken into account to counteract the finite nature of our simulations"\cite{schou18}.
 Also look at distributional characteristics, and plausibility of imputed values, see \cite{vinknd}.
@@
 
-Hypothesis based on \cite{lace07} is that the conventional acceptable level of $\widehat{R}$ is too strict for MI data.
-We expect that the simulation diagnostics will indicate proper imputations before $\widehat{R}$ will.
+Convergence diagnostics are $\widehat{R}$ and auto-correlation at lag one.
+$\widehat{R}$ is computed as follows (notation from \cite[p.~5]{veht19}).
+The scalar summary of interest in this study is the average of the imputed values per variable.
+%As recommended by \cite{veht19} $\widehat{R}$ is computed as ....
+to be able to detect ...
+in the tails of the distribution.
+\cite[p.~5]{veth19}
+
+\begin{quotation}
+``In the equations below, $N$ is the number of draws per chain, $M$ is the number of chains, and $S = MN$ is the total number of draws from all chains.
+For each scalar summary of interest $\theta$, we compute $B$ and $W$, the between- and within-chain variances:
+%
+\begin{align*}
+B=\frac{N}{M-1} \sum_{m=1}^{M}\left(\bar{\theta}^{(\cdot m)}-\bar{\theta}^{(\cdot \cdot)}\right)^{2}, \text { where } \bar{\theta}^{(\cdot m)}=\frac{1}{N} \sum_{n=1}^{N} \theta^{(n m)}, \quad \bar{\theta}^{(\cdot \cdot)}=\frac{1}{M} \sum_{m=1}^{M} \bar{\theta}^{(\cdot m)} \\
+W=\frac{1}{M} \sum_{m=1}^{M} s_{j}^{2},  \text { where } s_{m}^{2}=\frac{1}{N-1} \sum_{n=1}^{N}\left(\theta^{(n m)}-\bar{\theta}^{(\cdot m)}\right)^{2}". \text{ \cite[p.~5]{veht19}} 
+\end{align*}
+%
+\end{quotation}
+
+The proportion of within chain variance $W$ against the weighted average of $B$ and $W$, $\widehat{\operatorname{var}}^{+}$ tells us how much variance of $\theta$ could be shrunken down if $N\to\infty$.
+$\widehat{\operatorname{var}}^{+}$ is computed as follows: %over-estimates the marginal posterior variance of the estimand, $\theta$.
+
+\begin{equation*}
+\widehat{\operatorname{var}}^{+}(\theta | y)=\frac{N-1}{N} W+\frac{1}{N} B.
+\end{equation*}
+
+The potential scale reduction factor $\widehat{R}$ is obtained as:
+
+\begin{equation*}
+\widehat{R}=\sqrt{\frac{\widehat{\operatorname{var}}^{+}(\theta | y)}{W}}.
+\end{equation*}
+
+In this study, $\widehat{R}$ is computed for each variable, in each imputation chain, over all simulation conditions and MCMC simulations.
+For each simulation in each simulation condition, the maximum value of $\widehat{R}$ is reported.
+
+% \subsubsection{What is already implemented?}
+% 
+% - Trace-plots
+% 
+% \subsubsection{What is not yet implemented, but exists?}
+% 
+% - $\widehat{R}$, but too stringent (new) threshold, and assumption of over-dispersed initial values of imputation chains not met.
+% % Potentially add something about updated (2019) version of $\widehat{R}$ and the new threshold of 1.01, see \citep{veht19}.
+% 
+% - Auto-correlation.
+Schafer (1997, p.
+129) wrote on worst linear statistic.
+We could calculate the auto-correlation of that statistic to know that the algorithm converged elsewhere too.
+See autocorr function plot in SAS of worst linear function.
+% % Worst linear function in SAS see \url{https://support.sas.com/documentation/cdl/en/statug/63033/HTML/default/viewer.htm#statug_mi_sect027.htm}.
+% % Note: we're talking about missing data only, not the combined data (that autocorrelation is very high, as is the autocorrelation of deductively imputed values, like in the texp example). 
+% % "Applications of MICE with lowly correlated data therefore inject a lot of noise into the system.
+Hence, the auto-correlation over t will be low, and convergence will be rapid, and in fact immediate if all variables are independent.
+Thus, the incorporation of noise into the imputed data has pleasant side-effect of speeding up convergence" \citep{buur18}, par.
+4.5.
+% 
+% - Sensitivity analysis: Run algorithm several times and compare results. 
+% % "Imputers who do choose to use FCS should use flexible univariate models wherever possible and take care to assess apparent convergence of the algorithm, for example by computing traces of pooled estimates or other statistics and using standard MCMC diagnostics (Gelman et al., 2013, Chapter 11).
+It may also be helpful to examine the results of many independent runs of the algorithm with different initializations and to use random scans over the p variables to try to identify any convergence issues and mitigate possible order dependence" \cite[p.~19]{murr18}.
+% 
+% \subsubsection{What is not implemented, and NA?}
+% 
+% - $\widehat{R}$ threshold: Replicate simulation study and build a decision rule to solve the problem with $\widehat{R}$.
+% % "The monitoring statistic was computed for mean monthly earnings at each iteration in chains of length k = 200.
+Since calculation of the statistic requires M parallel sequences, m = 5 such chains were constructed.
+This value of m was informed by the preferred choice given in the literature on multiple imputation.
+The monitoring statistic computed at each iteration is presented in Figures 15 to 17 for each of the three missingness mechanisms.
+The red vertical line denotes ten iterations" \cite[p.~49]{lace07}.
+% 
+% - Stability of the solution: Possibly use the slope of means over iterations too to see whether there is trending.
+Or apply PCA on the imputed data and if that (the eigenvalues?) stays the same we know that the means and variances are stable as well, see McKay (?).
+% 
+% - MC error: MC error = SD/sqrt(number of iterations), where SD represents the variation across iterations.
+The MC error thus represents how much the means differ w.r.t.
+the iterations.
+MC error decreases as number of iterations increases.
+It should not be larger than 5\% of the sample standard deviation.
 
@@
 The probability to be missing is the same for all cells, namely XYZ\%.
-Table XYZ shows a summary of the generated complete data, and the amputed data.
-The amputed data is the starting point of each of the 1000 simulations per simulation condition.
+%"Furthermore, we generate three kinds of missingness proportions: 0.1, 0.5, and 0.9.
+Note that these values indicate the sampled proportion of incomplete cases in Y" \cite{schou18}.
+Table XYZ shows a summary of the generated complete data, and the amputed data. The amputed data is the starting point of each of the 1000 simulations per simulation condition.
 
@@
 4)
+% "combine the m completed data inferences into a single inference following Rubin's (1987) rules (pp. 76-77)" \cite{schou18}.
+
+Add ``post-processing steps (e.g.
+aggregation computations, summary statistics, regressions on the simulation results) used to transform simulation outputs to reported results''.
+%, see \url{http://jsterman.scripts.mit.edu/docs/Rahmandad-Sterman%20Simulation%20Reporting%20Standards3.30%20update.pdf}.
 
@@
 
-From the simulation diagnostics (...) it appears that as little as three iterations is sufficient to draw valid inference.
-Convergence diagnostics $\widehat{R}$ and auto-correlation, however, indicate 20-30 iterations.
-Apparently, they are not perfect for MI data.
+Evaluate simulation diagnostics one by one: bias, empirical SE, confidence interval width, and coverage rate. Then evaluate convergence diagnostics $\widehat{R}$ and auto-correlation.
 
-Moreover, $\widehat{R}$ could  theoretically not be smaller than one, yet it happened several times in this study (see online appendix XYZ).
-How could $\widehat{R}$ smaller than 1 occur?
-The number of simulations is smaller than in 'regular' MCMC processes.
-Therefore, the '$(n-1/n)$' correction factor can influence the estimated potential scale reduction factor.
-This downwards bias is in the opposite direction than expected: "The mixture-of-sequences variance, V, should stabilize as a function of n.
-(Before convergence, we expect $\sigma^2$ to decrease with n, only increasing if the sequences explore a new area of parameter space, which would imply that the original sequences were not overdispersed for the particular scalar summary being monitored.)" \cite[p~438]{broo98}.
+%% --------------------------------
+\subsection{Simulation diagnostics}
+Bias. After 1 or 2 iterations bias fluctuates within a narrow range of the average bias across all simulation conditions.
+
+Empirical standard error (SE).
+
+Confidence interval width. Does not show a trend across the simulation conditions. [Is related to empirical SE??]
+
+Coverage rate. Is more or less stable after one iteration. But higher percentages then the expected 95\% \cite{neym43}.
+
+%% --------------------------------
+\subsection{Convergence diagnostics}
+
+$\widehat{R}$.
+Steep drop down up-to iteration 30-40, then stable.
+Conventional threshold $\widehat{R} < 1.2$ reached after three iterations, $\widehat{R} < 1.1$ after four iterations, $\widehat{R} < 1.01$ after 38 iterations (see online appendix).
+
+Auto-correlation. Auto-correlation is dangerous when positive. These auto-correlations are negative. Still, we want the iterations to be stable. Low auto-correlations
 
@@
 \centering
-\begin{tabular}{lrrrrrr}
+\begin{tabular}{lrrrrrrr}
   \hline
-It. & Bias & Emp. SE & CI width & Cov. rate & $\widehat{R}$ & Auto-corr. \\ 
+  &\multicolumn{4}{c}{Simulation Diagnostics}& & \multicolumn{2}{c}{Convergence Diagn.} \\
+  \cline{2-5} \cline{7-8}
+It. & Bias & Emp. SE & CI width & Cov. rate & & $\widehat{R}$ & Auto-corr. \\ 
   \hline
-  1 & -0.135 & 0.115 & 0.637 & 0.930 &  &   \\ 
-  2 & -0.088 & 0.125 & 0.691 & 0.990 & 1.499 & -0.500 \\ 
-  3 & -0.090 & 0.112 & 0.623 & 0.950 & 1.200 & -0.658 \\ 
-  4 & -0.093 & 0.116 & 0.645 & 0.980 & 1.146 & -0.738 \\ 
-  5 & -0.087 & 0.108 & 0.599 & 0.980 & 1.098 & -0.711 \\ 
-  6 & -0.092 & 0.116 & 0.644 & 0.950 & 1.090 & -0.674 \\ 
-  7 & -0.095 & 0.122 & 0.675 & 1.000 & 1.063 & -0.588 \\ 
-  8 & -0.099 & 0.113 & 0.626 & 0.970 & 1.058 & -0.523 \\ 
-  9 & -0.095 & 0.115 & 0.639 & 1.000 & 1.044 & -0.519 \\ 
-  10 & -0.090 & 0.111 & 0.618 & 0.990 & 1.048 & -0.414 \\ 
-  15 & -0.093 & 0.109 & 0.604 & 0.950 & 1.020 & -0.339 \\ 
-  20 & -0.098 & 0.109 & 0.606 & 0.980 & 1.023 & -0.234 \\ 
-  25 & -0.100 & 0.113 & 0.626 & 0.990 & 1.016 & -0.091 \\ 
-  30 & -0.091 & 0.118 & 0.654 & 0.960 & 1.014 & -0.153 \\ 
-  40 & -0.088 & 0.117 & 0.650 & 0.990 & 1.009 & -0.070 \\ 
-  50 & -0.087 & 0.119 & 0.662 & 0.990 & 1.009 & -0.029 \\ 
+  1 & -0.135 & 0.115 & 0.637 & 0.930 &  & & \\ 
+  2 & -0.088 & 0.125 & 0.691 & 0.990 & & 1.499 & -0.500 \\ 
+  3 & -0.090 & 0.112 & 0.623 & 0.950 & &  1.200 & -0.658 \\ 
+  4 & -0.093 & 0.116 & 0.645 & 0.980 &  & 1.146 & -0.738 \\ 
+  5 & -0.087 & 0.108 & 0.599 & 0.980 &  & 1.098 & -0.711 \\ 
+  6 & -0.092 & 0.116 & 0.644 & 0.950 & &  1.090 & -0.674 \\ 
+  7 & -0.095 & 0.122 & 0.675 & 1.000 & &  1.063 & -0.588 \\ 
+  8 & -0.099 & 0.113 & 0.626 & 0.970 & &  1.058 & -0.523 \\ 
+  9 & -0.095 & 0.115 & 0.639 & 1.000 & &  1.044 & -0.519 \\ 
+  10 & -0.090 & 0.111 & 0.618 & 0.990 & &  1.048 & -0.414 \\ 
+  15 & -0.093 & 0.109 & 0.604 & 0.950 & &  1.020 & -0.339 \\ 
+  20 & -0.098 & 0.109 & 0.606 & 0.980 & &  1.023 & -0.234 \\ 
+  25 & -0.100 & 0.113 & 0.626 & 0.990 &  & 1.016 & -0.091 \\ 
+  30 & -0.091 & 0.118 & 0.654 & 0.960 & &  1.014 & -0.153 \\ 
+  40 & -0.088 & 0.117 & 0.650 & 0.990 &  & 1.009 & -0.070 \\ 
+  50 & -0.087 & 0.119 & 0.662 & 0.990 &  & 1.009 & -0.029 \\ 
   % 11 & -0.100 & 0.127 & 0.704 & 0.970 & 1.046 & -0.360 \\ 
@@
 \end{tabular}
+\caption{Simulation and convergence diagnostics over 1000 MCMC simulations.}
 \label{subsetresults}
 \end{table}
+
+From the simulation diagnostics it appears that as little as three iterations is sufficient to draw valid inference.
+Convergence diagnostics $\widehat{R}$ and auto-correlation, however, indicate 20-30 iterations.
+Apparently, they are not perfect for MI data [move interpretation to Discussion!].
+
+Add simulation time (computational cost) to table? Add empirical SE of all diagnostics? Sample some random simulations and check distribution and plausibility of imputations?
 
@@
 The recommended threshold for $\widehat{R}$ is too stringent for MI data.
+
+Moreover, $\widehat{R}$ could theoretically not be smaller than one, yet it happened several times in this study (see online appendix XYZ).
+How could $\widehat{R}$ smaller than 1 occur?
+The number of simulations is smaller than in `regular' MCMC processes.
+Therefore, the '$(n-1/n)$' correction factor can influence the estimated potential scale reduction factor.
+This downwards bias is in the opposite direction than expected: ``The mixture-of-sequences variance, V, should stabilize as a function of n.
+(Before convergence, we expect $\sigma^2$ to decrease with n, only increasing if the sequences explore a new area of parameter space, which would imply that the original sequences were not overdispersed for the particular scalar summary being monitored.)'' \cite[p~438]{broo98}.
 

### 2019-11-13 Simulation/Writeup/preprint.tex
Draft research report for peer feedback research seminar

--- 2019-11-12 Simulation/Writeup/preprint.tex
@@
 
-From the simulation diagnostics (...) it appears that as little as three iterations is sufficient to draw valid inference.
-Convergence diagnostics $\widehat{R}$ and auto-correlation, however, indicate 20-30 iterations.
-Apparently, they are not perfect for MI data.
+Evaluate simulation diagnostics one by one: bias, empirical SE, confidence interval width, and coverage rate. Then evaluate convergence diagnostics $\widehat{R}$ and auto-correlation.
 
-Moreover, $\widehat{R}$ could  theoretically not be smaller than one, yet it happened several times in this study (see online appendix XYZ).
-How could $\widehat{R}$ smaller than 1 occur?
-The number of simulations is smaller than in 'regular' MCMC processes.
-Therefore, the '$(n-1/n)$' correction factor can influence the estimated potential scale reduction factor.
-This downwards bias is in the opposite direction than expected.
+%% --------------------------------
+\subsection{Simulation diagnostics}
+Bias. After 1 or 2 iterations bias fluctuates within a narrow range of the average bias across all simulation conditions.
+
+Empirical standard error (SE).
+
+Confidence interval width. Does not show a trend across the simulation conditions. [Is related to empirical SE??]
+
+Coverage rate. Is more or less stable after one iteration. But higher percentages then the expected 95\% \cite{neym43}.
+
+%% --------------------------------
+\subsection{Convergence diagnostics}
+
+$\widehat{R}$.
+Steep drop down up-to iteration 30-40, then stable.
+Conventional threshold $\widehat{R} < 1.2$ reached after three iterations, $\widehat{R} < 1.1$ after four iterations, $\widehat{R} < 1.01$ after 38 iterations (see online appendix).
+
+Auto-correlation. Auto-correlation is dangerous when positive. These auto-correlations are negative. Still, we want the iterations to be stable. Low auto-correlations
 
@@
 
-"The mixture-of-sequences variance, V, should stabilize as a function of n.
-(Before convergence, we expect $\sigma^2$ to decrease with n, only increasing if the sequences explore a new area of parameter space, which would imply that the original sequences were not overdispersed for the particular scalar summary being monitored.)" \cite[p~438]{broo98}.
+This note illustrates that conventional convergence diagnostics behave differently on MI data as compared to the output of 'regular' MCMC algorithms.
+The recommended threshold for $\widehat{R}$ is too stringent for MI data.
+
+Moreover, $\widehat{R}$ could theoretically not be smaller than one, yet it happened several times in this study (see online appendix XYZ).
+How could $\widehat{R}$ smaller than 1 occur?
+The number of simulations is smaller than in `regular' MCMC processes.
+Therefore, the '$(n-1/n)$' correction factor can influence the estimated potential scale reduction factor.
+This downwards bias is in the opposite direction than expected: ``The mixture-of-sequences variance, V, should stabilize as a function of n.
+(Before convergence, we expect $\sigma^2$ to decrease with n, only increasing if the sequences explore a new area of parameter space, which would imply that the original sequences were not overdispersed for the particular scalar summary being monitored.)'' \cite[p~438]{broo98}.
 

### 2019-11-13 Simulation/Functions/Convergence.R

@@ -15,6 +15,38 @@ get.rhat <- function(imp, maxit, n.var = 4) {
       sqrt((var_between[v] / var_within[v] + maxit - 1) / maxit)
   }
   
+  # # possibly integrate rstan functions for folded, spplit-half rank-normalized Rhat
+  # z_scale <- function(x) {
+  #   S <- length(x)
+  #   r <- rank(x, ties.method = 'average')
+  #   z <- qnorm((r - 1 / 2) / S)
+  #   if (!is.null(dim(x))) {
+  #     # output should have the input dimension
+  #     z <- array(z, dim = dim(x), dimnames = dimnames(x))
+  #   }
+  #   z
+  # }
+  
+  # split_chains <- function(sims) {
+  #   # split Markov chains
+  #   # Args:
+  #   #   sims: a 2D array of samples (# iter * # chains)
+  #   if (is.vector(sims)) {
+  #     dim(sims) <- c(length(sims), 1)
+  #   }
+  #   niter <- dim(sims)[1]
+  #   if (niter == 1L) return(sims)
+  #   half <- niter / 2
+  #   cbind(sims[1:floor(half), ], sims[ceiling(half + 1):niter, ])
+  # }
+  
+  # rhat <- function(sims) {
+  #   bulk_rhat <- rhat_rfun(z_scale(split_chains(sims)))
+  #   sims_folded <- abs(sims - median(sims))
+  #   tail_rhat <- rhat_rfun(z_scale(split_chains(sims_folded)))
+  #   max(bulk_rhat, tail_rhat)
+  # }
+  
   # output
   max(rhat)
 }

### 2019-11-14 Simulation/Writeup/preprint.Rnw
1000 sims results

@@
 imputed) several times.
-[Add some stuff here about MICE.] The variability between the imputations validly reflects how much uncertainty in the inference is due to missing information--that is, if all statistical assumptions are met \cite{rubin87}.
+[Add some stuff here about MICE.] The variability between the imputations validly reflects how much uncertainty in the inference is due to missing information--that is, if all statistical assumptions are met \citep{rubin87}.
 %Missing data is ubiquitous. MICE can solve stuff.
@@
 
-This paper investigates convergence properties of MI algorithms by means of model-based simulation in \proglang{R} \cite{R}.
+This paper investigates convergence properties of MI algorithms by means of model-based simulation in \proglang{R} \citep{R}.
 The results of this simulation study are guidelines for applied researchers to evaluate convergence of MI algorithms.
@@
 
-The convergence guidelines introduced in this paper are developed to be integrated into the \pkg{mice} environment \cite{mice} in \proglang{R} \cite{R}.
+The convergence guidelines introduced in this paper are developed to be integrated into the \pkg{mice} environment \citep{mice} in \proglang{R} \citep{R}.
 It therefore follows notation and conventions of \cite{mice}.
@@
 The $m$ imputed data sets are analyzed separately, and their results pooled.
-A distinction is made between the imputation step (employing $p$ 'imputation models'), and the analysis step (using one 'complete-data model' for each of the $m$ imputated data sets).
+A distinction is made between the imputation step (employing $p$ 'imputation models'), and the analysis step (using one 'complete-data model' for each of the $m$ imputed data sets).
 The quantity of scientific interest is the result of the analysis step, e.g., a regression coefficient, and denoted with $Q$.
@@
 % "Let Q denote the quantity of scientific interest (e.g., a regression coefficient)." \cite[p.~4]{mice}.
+
+[Explain the use of `imputation chain' (the iterations of an imputation, only the final imputed values are used), `simulation condition' (the number of iterations the algorithm has before imputing), ...?]
 
@@
 The development of practical tools that put these conditions to work for multiple imputation is still an ongoing research problem." \cite[par.~4.5]{buur18}
-We can never be certain of convergence, therefore convergence diagnostics evaluate signs of non-convergence \cite{hoff09}.
+We can never be certain of convergence, therefore convergence diagnostics evaluate signs of non-convergence \citep{hoff09}.
 % In general: "In general, you cannot know for sure if your chain has converged.
@@
 This means that at most one initial value is necessary, but probably none at all.
-[I should look this up.] Without over-dispersed initial states, $\widehat{R}$ can falsely diagnose convergence \cite{broo98}.
+[I should look this up.] Without over-dispersed initial states, $\widehat{R}$ can falsely diagnose convergence \citep{broo98}.
 %would not be sensitive enough to flag non-convergence of MI algorithms: "if over-dispersion does not hold, $\sigma_+^2$ can be too low, which can lead to falsely diagnosing convergence." \cite[p~437]{broo98}.
@@
 
-Convergence diagnostics are $\widehat{R}$ and auto-correlation at lag one.
-$\widehat{R}$ is computed as follows (notation from \cite[p.~5]{veht19}).
-The scalar summary of interest in this study is the average of the imputed values per variable.
+Convergence diagnostics are $\widehat{R}$ and auto-correlation (at lag one).
+$\widehat{R}$ is computed as follows: % The scalar summary of interest in this study is the average of the imputed values per variable.
 %As recommended by \cite{veht19} $\widehat{R}$ is computed as ....
@@
 
-In this study, $\widehat{R}$ is computed for each variable, in each imputation chain, over all simulation conditions and MCMC simulations.
-For each simulation in each simulation condition, the maximum value of $\widehat{R}$ is reported.
+In this study, $\widehat{R}$ is computed for the mean of each variable, in each imputation chain, over all simulation conditions and MCMC simulations.
+%For each simulation in each simulation condition, the maximum value of $\widehat{R}$ is reported.
 
@@
     compute simulation diagnostics
-save convergence and simulation diagnostics
+aggregate convergence and simulation diagnostics
 \end{Code}
@@
 Missing data points are imputed with \pkg{mice} in \proglang{R}.
-All simulations are performed with imputation method 'norm' (Bayesian linear regression imputation), and five imputation chains.
+All simulations are performed with imputation method 'norm' (Bayesian linear regression imputation), and five imputation chains ($m = 5$).
 The number of iterations is varied over simulations ('maxit' argument between 1 and 100).
-For each number of iterations, simulation and convergence diagnostics are aggregated over 1000 MCMC simulations.
+Each simulation condition (i.e., each number of iterations) is simulated 1000 times.
+Simulation and convergence diagnostics are aggregated over the 1000 MCMC simulations.
 % Missingness mechanism: "With MCAR missingness mechanisms, the probability to be missing is the same for all cases.
@@
 %, see \url{http://jsterman.scripts.mit.edu/docs/Rahmandad-Sterman%20Simulation%20Reporting%20Standards3.30%20update.pdf}.
+
+Post-processing of simulation diagnostics:
+
+Bias is computed as the estimated regression coefficient after MI minus the true regression coefficient. Bias is averaged over all simulations with the same simulation condition.
+
+Confidence interval width (CIW) is computed as the diffence between the lower and upper bound of the 95\% confidence interval (CI95\%).
+The CI95\% bounds are computed as the estimated regression coefficient plus or minus (respectively) the pooled SE across imputations times 2.66 (the quantile of a t distribution with m-1 degrees of freedom).
+CIW is averaged over all simulations with the same simulation condition.
+
+Coverage rate is computed as the proportion of simulations (with the same simulation condition) in which the true regression coefficient is between the bounds of the CI95\%.
+
+Post-processing of convergence diagnostics:
+
+$\widehat{R}$ is computed within imputation chains: for each variable, for each simulation condition, for each simulation.
+The maximum value across variables within the same simulation is reported.
+Maximum $\widehat{R}$ values per simulation condition are aggregated across simulations.
+
+Autocorrelation (AC) is computed within imputation chains, as the correlation between $\theta$ at the $i^{th}$ iteration and $\theta$ at the $i+1^{th}$ iteration.
+For now, we only consider the chain mean as scalar of interest $\theta$.
+The AC of the variable with the highest absolute AC value is reported per simulation.
+These values are then averaged per simulation condition.
 
@@
 
-Add simulation time (computational cost) to table? Add empirical SE of all diagnostics? Sample some random simulations and check distribution and plausibility of imputations?
+[Add simulation time (computational cost) to table? Add empirical SE of all diagnostics? Sample some random simulations and check distribution and plausibility of imputations?]
 

### 2019-11-14 Simulation/Writeup/preprint.tex
1000 sims results

@@
 % "Let Q denote the quantity of scientific interest (e.g., a regression coefficient)." \cite[p.~4]{mice}.
+
+Explain the use of `imputation chain' (the iterations of an imputation, only the final imputed values are used), `simulation condition' (the number of iterations the algorithm has before imputing), ...?
 
@@
 %, see \url{http://jsterman.scripts.mit.edu/docs/Rahmandad-Sterman%20Simulation%20Reporting%20Standards3.30%20update.pdf}.
+
+Post-processing of simulation diagnostics:
+
+Bias is computed as the estimated regression coefficient after MI minus the true regression coefficient. Bias is averaged over all simulations with the same simulation condition.
+Confidence interval width (CIW) is computed as the diffence between the lower and upper bound of the 95\% confidence interval (CI95\%).
+The CI95\% bounds are computed as the estimated regression coefficient plus or minus (respectively) the pooled SE across imputations times 2.66 (the quantile of a t distribution with m-1 degrees of freedom).
+CIW is averaged over all simulations with the same simulation condition.
+Coverage rate is computed as the proportion of simulations (with the same simulation condition) in which the true regression coefficient is between the bounds of the CI95\%.
+
+Post-processing of convergence diagnostics:
+
+$\widehat{R}$ is computed within imputation chains: for each variable, for each simulation condition, for each simulation.
+The maximum value across variables within the same simulation is reported.
+Maximum $\widehat{R}$ values per simulation condition are aggregated across simulations.
+
+Autocorrelation (AC) is computed within imputation chains, as the correlation between $\theta$ at the $i^{th}$ iteration and $\theta$ at the $i+1^{th}$ iteration.
+For now, we only consider the chain mean as scalar of interest $\theta$.
+The AC of the variable with the highest absolute AC value is reported per simulation.
+These values are then averaged per simulation condition.
 

### 2019-11-14 Simulation/Stepbystep.Rmd
Step-by-step computation of Rhat in Rmd

---
title: "Convergence of MI algorithms"
output: html_notebook
---

This file describes how convergence diagnostic $\widehat{R}$ is computed in the simulation study I ran for my Research Report. $\widehat{R}$ is also called 'potential scale reduction factor'. It tells us how much the variance of an estimate could be shrunken down if we would have infinitely many iterations.

# Step 1: Impute some data

It's easiest to show how $\widehat{R}$ is computed by applying it on a simple example. Load the nhanes data and impute the variables with missingness.
```{r echo=TRUE, warning=FALSE}
# load required packages/functions
library(mice)
library(magrittr)
source("Functions/Convergence.R")
source("Functions/Convergence_supplement.R")

# set parameters
maxit <- 10
n.var <- 3

# run imputation
imp <- mice(nhanes[,-1], maxit = maxit, print = F)

```

# Step 2: Evaluate the functions to use

We are going to compute $\widehat{R}$ values per imputed variable. As recommended by Vehtari et al. (2019), we compute split-half, rank-normalized $\widehat{R}$ values for the bulk and the tails of the distribution. For the latter, we need to fold the imputation chains before splitting them in half and rank-normalizing them. Then, $\widehat{R}$ is computed for both (bulk and tails). The highest of these two $\widehat{R}$ values is extracted. This is performed for each of the imputed variables, and only the maximum $\widehat{R}$ value across variables is reported.

The rhat_function function is the wrapper that performs all of this. Within the function, we call 4 other functions (explaination below): fold_sims(), split_chains(), z_scale(), and get.rhat().

```{r}
# show the contents of the function
rhat_function

```

The first function inside rhat_function is split_chains(). This 'chops' the chains in two, to assess trending within chains. 
```{r warning=FALSE}
# show the contents of the function
split_chains

```

Then the halved chains are rank-normalized.
```{r}
# show the contents of the function
z_scale

```

To detect non-convergence in the tails, $\widehat{R}$ is also computed on the folded chains (which are also split and rank-normalized).
```{r}
# show the contents of the function
fold_sims

```

Only then do we get to the actual $\widehat{R}$ function. In the get.rhat function, the following equations are applied:

"In the equations below, $N$ is the number of draws per chain, $M$ is the number of chains, and $S = MN$ is the total number of draws from all chains. For each scalar summary of interest $\theta$, we compute $B$ and $W$, the between- and within-chain variances:

\begin{align*}
B=\frac{N}{M-1} \sum_{m=1}^{M}\left(\bar{\theta}^{(\cdot m)}-\bar{\theta}^{(\cdot \cdot)}\right)^{2}, \text { where } \bar{\theta}^{(\cdot m)}=\frac{1}{N} \sum_{n=1}^{N} \theta^{(n m)}, \quad \bar{\theta}^{(\cdot \cdot)}=\frac{1}{M} \sum_{m=1}^{M} \bar{\theta}^{(\cdot m)} \\
W=\frac{1}{M} \sum_{m=1}^{M} s_{j}^{2},  \text { where } s_{m}^{2}=\frac{1}{N-1} \sum_{n=1}^{N}\left(\theta^{(n m)}-\bar{\theta}^{(\cdot m)}\right)^{2}". \text{ Vehtari et al., 2019, p. 5} 
\end{align*}

The between and within chain variances are computed from the chain means of the mids object, that underwent the previous transformations (folding, splitting, rank-normalizing). The average of the chain means per imputation $m$ is $\bar{\theta}^{(\cdot m)}$. The average of these $m$ values is $\bar{\theta}^{(\cdot \cdot)}$. $N$ is the number of iterations ('maxit'). So we compute $B$ as `maxit * var(apply(sims, 2, mean))`. And $W$ is computed as `mean(apply(sims, 2, var))`.

To get $\widehat{R}$ we need to compute the weighted average of $B$ and $W$, $\widehat{\operatorname{var}}^{+}$, and evaluate this against the within chain variance W. The weighted average is is computed as follows:

\begin{equation*}
\widehat{\operatorname{var}}^{+}(\theta | y)=\frac{N-1}{N} W+\frac{1}{N} B,
\end{equation*}

Then finally, potential scale reduction factor $\widehat{R}$ is obtained as:

\begin{equation*}
\widehat{R}=\sqrt{\frac{\widehat{\operatorname{var}}^{+}(\theta | y)}{W}}.
\end{equation*}

$\widehat{R}$ is computed for each variable with the function 'get.rhat()'. 
```{r warning=FALSE}
# show the contents of the function
get.rhat

```

# Step 3: Apply these functions to get $\widehat{R}$ values

Now we know what's happening, let's apply it.
```{r}
rhat_function(imp, maxit, n.var = n.var)

```

This $\widehat{R}$ value is nice and high, but I also got values below one. Which is theoretically impossible.

# So my question to you is: How is it possible that we can obtain  $\widehat{R}$ values < 1?

### 2019-11-14 Simulation/Writeup/preprint.Rnw
Step-by-step computation of Rhat in Rmd

@@
     impute the missingness
-    compute convergence diagnostic
+    compute convergence diagnostics
     perform analysis
@@
 After data generation, the complete data is 'amputed'.
-That is, the \pkg{mice} function \fct{ampute} is used to impose an MCAR missingness mechanism upon the data.
+That is, the \pkg{mice} function \fct{ampute} in \proglang{R} is used to impose an MCAR missingness mechanism upon the data.
 The probability to be missing is the same for all cells, namely XYZ\%.
@@
 
-[Add simulation time (computational cost) to table? Add empirical SE of all diagnostics? Sample some random simulations and check distribution and plausibility of imputations?]
+[Add simulation time (computational cost) to table?
+Add empirical SE of all diagnostics?
+Sample some random simulations and check distribution and plausibility of imputations?
+Use chain variance as scalar of interest for rhat and autocorrelation?]
 

### 2019-11-14 Simulation/Functions/Convergence.R

@@ -1,51 +1,21 @@
 # Extract convergence diagnostic Rhat from mids object
 # Used within imputation function
+# Requires package 'magrittr'/'dplyr', and the functions in Convergence_supplement.R
 
-get.rhat <- function(imp, maxit, n.var = 4) {
+rhat_function <- function(imp, maxit, n.var = 4) {
+  # extract convergence diagnostic Rhat from mids object
   
-  # objects with variables in columns
-  var_between <- var_within <- rhat <- matrix(NA, nrow = 1, ncol = n.var)
+  # object to store output for loop
+  rhat <- matrix(NA, nrow = 1, ncol = n.var)
   
-  # compute converegnce diagnostic per variable
+  # compute converegnce diagnostic per variable v
   for (v in 1:n.var) {
-    var_between[v] <-
-      maxit * var(apply(imp$chainMean[v, , ], 2, mean))
-    var_within[v] <- mean(apply(imp$chainMean[v, , ], 2, var))
-    rhat[v] <-
-      sqrt((var_between[v] / var_within[v] + maxit - 1) / maxit)
+    rhat_bulk <-
+      imp$chainMean[v, , ] %>% split_chains() %>% z_scale() %>% get.rhat(maxit = maxit)
+    rhat_tail <-
+      imp$chainMean[v, , ] %>% fold_sims() %>% split_chains() %>% z_scale() %>% get.rhat(maxit = maxit)
+    rhat[v] <- max(rhat_bulk, rhat_tail)
   }
-  
-  # # possibly integrate rstan functions for folded, spplit-half rank-normalized Rhat
-  # z_scale <- function(x) {
-  #   S <- length(x)
-  #   r <- rank(x, ties.method = 'average')
-  #   z <- qnorm((r - 1 / 2) / S)
-  #   if (!is.null(dim(x))) {
-  #     # output should have the input dimension
-  #     z <- array(z, dim = dim(x), dimnames = dimnames(x))
-  #   }
-  #   z
-  # }
-  
-  # split_chains <- function(sims) {
-  #   # split Markov chains
-  #   # Args:
-  #   #   sims: a 2D array of samples (# iter * # chains)
-  #   if (is.vector(sims)) {
-  #     dim(sims) <- c(length(sims), 1)
-  #   }
-  #   niter <- dim(sims)[1]
-  #   if (niter == 1L) return(sims)
-  #   half <- niter / 2
[…]

### 2019-11-15 Simulation/1.Execute.R
Cleaned up functions to run simulation by Gerko

@@ -11,17 +11,17 @@ library(dplyr)
 library(ggplot2)
 
 # load simulation/evaluation functions
-source("Functions/CreateData.R")
-source("Functions/Convergence.R")
-source("Functions/Convergence_supplement.R")
-source("Functions/Autocorrelation.R")
-source("Functions/Impute.R")
-source("Functions/Evaluate.R")
+source("Simulation/Functions/CreateData.R")
+source("Simulation/Functions/Convergence.R")
+source("Simulation/Functions/Convergence_supplement.R")
+source("Simulation/Functions/Autocorrelation.R")
+source("Simulation/Functions/Impute.R")
+source("Simulation/Functions/Evaluate.R")
 
 # simulation parameters
 populationsize <- 1000 #n of simulated dataset
-n.iter <- 15 #nr of iterations (varying 1:n.iter)
-n.sim <- 10 #nr of simulations per iteration value
+n.iter <- 100 #nr of iterations (varying 1:n.iter)
+n.sim <- 1000 #nr of simulations per iteration value
 true_effect <- 2 #regression coefficient to be estimated
 
 # start simulation study
@@ -30,42 +30,42 @@ set.seed(1111)
 ###
 
 # combine separate functions into wrapper
-simulate <- function(runs = 10, n.iter = 5) {
+simulate <- function(runs = 10, n.iter = 5, populationsize = 1000, true_effect = 2) {
   pb <- txtProgressBar(min = 0, max = runs, style = 3)
   # object for output
   res <- array(NA, dim = c(n.iter, runs, 6))
   # simulate data once
-  data <- data.simulation(n = poulationsize)
+  data <- data.simulation(n = poulationsize, true_effect)
   # repeat mi procedure 'runs'  times for each nr of iterations
   for (run in 1:runs) {
     for (i in 1:n.iter) {
-      res[i, run, ] <- test.impute(data, method = "norm",
-                                   maxit = i)
+      res[i, run, ] <- test.impute(data, maxit = i)
     }
     setTxtProgressBar(pb, run)
   }
   close(pb)
+  # output
   res
 }
 
 # simulate
-res <- simulate(runs = n.sim, n.iter = n.iter)
+res <- simulate(runs = n.sim, n.iter = n.iter, populationsize = populationsize, true_effect = true_effect)
 
 ###
 
 # evaluate
-(out <- evaluate.sim(res))
[…]

### 2019-11-19 Simulation/Writeup/NoteConvergence.Rnw
Feedback Research Seminar

--- 2019-11-14 Simulation/Writeup/preprint.Rnw
@@
 \section{Introduction} \label{sec:intro} % curly brackets can handle typesetting
+
+Feedback research seminar: order is not intuitive, mentioned, explained later.
+short sentences with a lot of content.
+implicitly a lot happens.
+expand sentence or it's obvious and i should delete it.
+Explain 'true data generating model'.
+follow important concepts with explanation or examples in next part.
+Look at abbreviations.
+for example r hat is just metnioned as 'e.g.' not clear in intro how important it is.
+also, look at distinction between mi and mice, they look like synonyms now.
+look at imputing vs analysing because they're called steps somewhere, and something else somewhere else.
+explain what a chain is.
+make clear when i talk about my study or lit.
+look at equations vehtari, because there is no real need for it.
+try to rephrase in my own words because of text flow.
+make clear that i only generate 1 dataset (not simulated data).
+in figure, r hat looks stable after 20 it.
+in discussion, explain that less it is advantage of mi, not a disadv.
+compared to MCMC.
+then final par is too negative, explain scope in intro instead.
 

### 2019-11-21 Simulation/Writeup/NoteConvergence.Rnw
put feedback at right places, ran improved simulation with nsim=100

--- 2019-11-14 Simulation/Writeup/preprint.Rnw
@@
 % \usepackage{framed}
-\usepackage{multirow, amsmath}
+\usepackage{multirow, amsmath, tikz}
+\usetikzlibrary{fit, positioning}
 
@@
   E-mail: \email{h.i.oberman@uu.nl} % \\
-  % URL: \url{https://eeecon.uibk.ac.at/~zeileis/}
+  URL: \url{https://hanneoberman.github.io/}
 }
@@
 At some point, any scientist conducting statistical analyses will run into a missing data problem \citep{alli02}.
-Missingness is problematic because statistical inference cannot be performed on incomplete data, and ad hoc solutions can yield wildly invalid results \citep{buur18}.
+Missingness is problematic because statistical inference cannot be performed on incomplete data, and ad hoc solutions (e.g., listwise deletion) can yield wildly invalid results \citep{buur18}.
 To circumvent the ubiquitous problem of missing information, \cite{rubin87} proposed the framework of multiple imputation (MI).
@@
 imputed) several times.
-[Add some stuff here about MICE.] The variability between the imputations validly reflects how much uncertainty in the inference is due to missing information--that is, if all statistical assumptions are met \citep{rubin87}.
+[Add some stuff here about MICE.] The variability between the imputations validly reflects how much uncertainty in the inference is due to missing information--that is, if all statistical assumptions are met \citep{rubin87} [adjust this sentence to be more clear!].
+
+[Feedback: The order of the introduction is not intuitive.
+some terms are mentioned first and only explained later.
+There are some short sentences with a lot of content, so implicitly a lot happens.
+Expand these sentences or delete (because the thing would be too obvious?).
+Explain 'true data generating model'.
+Follow up other important concepts with explanation or examples too.]
 %Missing data is ubiquitous. MICE can solve stuff.
 
-With MI, many assumptions are made about the nature of the observed and missing parts of the data and their relation to the `true' \emph{data generating model} \citep{buur18}.
+With MI, many assumptions are made about the nature of the observed and missing parts of the data and their relation to the `true' \emph{data generating model} \citep{buur18} [explain what this is!].
 Without proper evaluation of the imputations and the underlying assumptions, any drawn inference may erroneously be deemed valid.
@@
 
+[Feedback: Look at abbreviations. For example r hat is just metnioned as 'e.g.'. It's not clear in the introduction section how important it is later.]
+
 A fundamental assumption of MICE is convergence of the algorithm.
-MICE is a sort of Gibbs sampler, a type of MCMC algorithm.
+MICE is a sort of Gibbs sampler, a type of Markov chain Monte Carlo (MCMC) algorithm.
 In general, the validity of inference resulting from MCMC algorithms is threatened by non-convergence.
-So we use convergence diagnostics to flag non-convergence.
+Hence we use convergence diagnostics to flag non-convergence.
 But it is not known whether conventional convergence diagnostics for MCMC methods work on MI data.
 
-Conventional thresholds to diagnose non-convergence-- e.g., Gelman and Rubin's \citeyear{gelm92} statistic $\widehat{R} < 1.1$ --are not applicable on multiply imputed data \citep{lace07}.
-Therefore, empirical researchers have to rely on visual inspection procedures that are theoretically equivalent to $\widehat{R}$ \citep{whit11}.
+Conventional thresholds to diagnose non-convergence-- e.g., Gelman and Rubin's \citeyear{gelm92} statistic $\widehat{R} < 1.1$ --are not applicable on multiply imputed data \citep{lace07} [Expand or remove Rhat here!
+Refer to future section?].
+Therefore, empirical researchers have to rely on visual inspection procedures that are theoretically equivalent to $\widehat{R}$ \citep{whit11} [explain why this is appropriate, and numerical is not].
 Visually assessing convergence is not only difficult to the untrained eye, it might also be futile.
@@
 This paper investigates convergence properties of MI algorithms by means of model-based simulation in \proglang{R} \citep{R}.
+All programming code used in this note is available in the file 'XYZ.R' along with the manuscript, and on Github repository 'XYZ'.
 The results of this simulation study are guidelines for applied researchers to evaluate convergence of MI algorithms.
@@
 %% -- Features ---------------------------------------------------------------
-\subsection{Features} \label{sec:features}
+\subsection{Terminology} \label{sec:terms}
 
-The aim of this paper is to provide applied researchers a tool to assess convergence of MICE algorithms.
+The aim of this paper is to provide applied researchers a tool to assess convergence of multiple imputation algorithms.
 The intended audience consists of empirical researchers and statisticians who use multiple imputation to solve missing data problems.
@@
 For the theoretical foundation of MI, see \cite{rubin87}.
-All programming code used in this paper is available in the file 'XYZ.R' along with the manuscript, and on Github repository 'XYZ'.
-% "The intended audience of this paper consists of applied researchers who want to address problems caused by missing data by multiple imputation.
-The text assumes basic familiarity with R.
-The document contains hands-on analysis using the mice package.
-We do not discuss problems of incomplete data in general.
-We refer to the excellent books by Little and Rubin (2002) and Schafer (1997).
-Theory and applications of multiple imputation have been developed in Rubin (1987) and Rubin (1996).
-van Buuren (2012) introduces multiple imputation from an applied perspective" \cite[p.~4]{mice}.
+
+[explain difference MI, MICE, mice]
+
+[Explain the use of `imputation chain' (the iterations of an imputation, only the final imputed values are used), starting values of chains, `simulation condition' (the number of iterations the algorithm has before imputing), ...?]
+
+[look at imputing vs analysing because they're called steps somewhere, and something else somewhere else. explain what a chain is.]
+
+\begin{figure}
+\centering
+	\large{\begin{tikzpicture}
+	\tikzstyle{main}=[circle, minimum size = 10mm, thick, draw =black!80, node distance = 20mm]
+\tikzstyle{connect}=[-latex, thick]
+\node[main, fill = white!100] (data) [label=left:Incomplete data] { };
+\node[main] (mids) [right=of data] { };
+\node[main] (mids2) [above=of mids] { };
+\node[main] (mids3) [below=of mids,label=below:Imputed data] { };
+\node[main] (mira) [right=of mids] {};
+\node[main] (mira2) [above=of mira] { };
+\node[main] (mira3) [below=of mira,label=below:Analysis results] { };
+\node[main, fill = white!100] (mipo) [right=of mira,label=right:Pooled result] { };
+\path (data) edge [connect] (mids)
+      (data) edge [connect] (mids2)
+      (data) edge [connect] (mids3)
+      (mids) edge [connect] (mira)
+      (mids2) edge [connect] (mira2)
+      (mids3) edge [connect] (mira3)
+		  (mira) edge [connect] (mipo)
+		  (mira2) edge [connect] (mipo)
+		  (mira3) edge [connect] (mipo);
+\end{tikzpicture}}
+\caption{Scheme of main steps in multiple imputation (Adapted from \cite[\S~1.4.1]{buur18})}
+\end{figure}
+
+% \begin{figure}[h]
+%   \resizebox{\textwidth}{!}{ %notice the \resizebox{} command
+%         \includegraphics{steps.png}
+%   }
+%   \caption{Scheme of main steps in multiple imputation (Adapted from \cite[par.~1.4.1]{buur18})}
+%     \label{resultssim}
+% \end{figure}
 
@@
 Let $Y$ denote an $n \times p$ matrix containing the data values on $p$ variables for all $n$ units in a sample.
-The collection of observed data values is denoted as $Y_{obs}$; the missing part of $Y$ is referred to as $Y_{mis}$.
+The collection of observed data values in $Y$ is denoted as $Y_{obs}$; the missing part of $Y$ is referred to as $Y_{mis}$.
 Response indicator $R$ shows whether a data value in $Y$ is missing or observed.
 The relation between $R$, $Y_{obs}$, and $Y_{mis}$ determines the missingness mechanism.
-This paper only considers a 'missing completely at random' mechanism, where the probability to be missing is equal for all $n \times p$ cells in $Y$.
+This paper only considers a 'missing completely at random' (MCAR) mechanism, where the probability to be missing is equal for all $n \times p$ cells in $Y$.
 
@@
 imputed) $m$ times.
-The $m$ imputed data sets are analyzed separately, and their results pooled.
+The $m$ imputed data sets are analyzed separately, and their results pooled [Make this more clear!].
 A distinction is made between the imputation step (employing $p$ 'imputation models'), and the analysis step (using one 'complete-data model' for each of the $m$ imputed data sets).
@@
 % "Let Q denote the quantity of scientific interest (e.g., a regression coefficient)." \cite[p.~4]{mice}.
-
-[Explain the use of `imputation chain' (the iterations of an imputation, only the final imputed values are used), `simulation condition' (the number of iterations the algorithm has before imputing), ...?]
 
@@
 iteration five are significantly correlated with the value of the first draw.
-Auto-correlation above the confidence level indicate dependence within chains.
+Auto-correlation above the confidence level [explain that this is a critical value for correlation based on alpha .05] indicate dependence within chains.
 Also, possible to use MC error, Geweke, and the other one.
@@
 
-The potential scale reduction factor $\widehat{R}$ tells us how much variance of an estimate could be shrunken down by running the chains infinitely long.
+The potential scale reduction factor $\widehat{R}$ tells us how much variance of an estimate could be shrunken down by running the chains infinitely long [Add reference, gelman something?].
 That then tells us something about how dependent the chains are on the starting values.
@@
 
+[make clear when i talk about my study or lit. look at equations vehtari, because there is no real need for it. try to rephrase in my own words because of text flow.]
+
 In this study, $\widehat{R}$ is computed for the mean of each variable, in each imputation chain, over all simulation conditions and MCMC simulations.
@@
 
-The simulated data is a finite population of $N=1000$.
-The variables are dependent variable $Y$, independent variable $X$, and covariates $Z_1$ and $Z_2$.
-The quantity of scientific interest is the estimated regression coefficient of $X$ on $Y$ in linear regression model $Y \sim X+Z_1+Z_2$.
+The simulated dataset is a finite population of $N=1000$.
+The data are simulated to solve a multiple linear regression complete data problem.
+The quantity of scientific interest is the estimated regression coefficient of predictor $X$ on outcome variable $Y$.
+The linear regression model is $Y \sim X+Z_1+Z_2$, where $Y$ is the dependent variable, $X$ is an independent variable, and $Z_1$ and $Z_2$ are covariates.
 The data generating model of the predictors is a multivariate normal distribution with means structure $\mu$, and variance-covariance matrix $\Sigma$.
@@
 That is, the \pkg{mice} function \fct{ampute} in \proglang{R} is used to impose an MCAR missingness mechanism upon the data.
-The probability to be missing is the same for all cells, namely XYZ\%.
+The missingness is univariate, and the probability to be missing is the same for all variables, namely 20\%.
+This leaves 20\% of the rows completely observed.
 %"Furthermore, we generate three kinds of missingness proportions: 0.1, 0.5, and 0.9.
 Note that these values indicate the sampled proportion of incomplete cases in Y" \cite{schou18}.
-Table XYZ shows a summary of the generated complete data, and the amputed data. The amputed data is the starting point of each of the 1000 simulations per simulation condition.
+Table XYZ shows a summary of the generated complete data, and the amputed data. The amputed dataset is the starting point of each of the 1000 simulations per simulation condition.
 
@@
   \resizebox{\textwidth}{!}{ %notice the \resizebox{} command
-        \includegraphics{res3.pdf}
+        \includegraphics{resultsA4.pdf}
   }      
@@
 \subsection{Simulation diagnostics}
-Bias. After 1 or 2 iterations bias fluctuates within a narrow range of the average bias across all simulation conditions.
+Bias.
+After 1 or 2 iterations bias fluctuates within a narrow range of the average bias across all simulation conditions.
+Average bias is not zero because there is only one incomplete dataset to be imputed.
+Bias is a sample effect.
 
@@
 
-Coverage rate. Is more or less stable after one iteration. But higher percentages then the expected 95\% \cite{neym43}.
+Coverage rate. Is more or less stable from two iterations upwards. But higher percentages then the expected 95\% \cite{neym43}. This is due to pooled SE??
 
@@
 $\widehat{R}$.
-Steep drop down up-to iteration 30-40, then stable.
+Steep drop down up-to iteration 20-30, then stable.
 Conventional threshold $\widehat{R} < 1.2$ reached after three iterations, $\widehat{R} < 1.1$ after four iterations, $\widehat{R} < 1.01$ after 38 iterations (see online appendix).
 
-Auto-correlation. Auto-correlation is dangerous when positive. These auto-correlations are negative. Still, we want the iterations to be stable. Low auto-correlations
+Auto-correlation.
+Auto-correlation is dangerous when positive.
+These auto-correlations are negative.
+Still, we want the iterations to be stable and independent.
+Default maxit is five iterations now, should this be different?
+Are the 'waves' most pronounced at iteration 5?
+But why the dip??
 
@@
 How could $\widehat{R}$ smaller than 1 occur?
-The number of simulations is smaller than in `regular' MCMC processes.
-Therefore, the '$(n-1/n)$' correction factor can influence the estimated potential scale reduction factor.
+The number of simulations is smaller than in `regular' MCMC processes[explain why this is positive!].
+Therefore, the `$(n-1/n)$' [add equation number] correction factor can influence the estimated potential scale reduction factor.
 This downwards bias is in the opposite direction than expected: ``The mixture-of-sequences variance, V, should stabilize as a function of n.
@@
 
-We would like to have convergence measures for multivariable statistics (scalars?) of interest.
+% \url{https://discourse.mc-stan.org/t/rhat-1-as-low-as-9-94e-01-why/9252/3}
+
+Future research: We would like to have convergence measures for multivariable statistics (scalars?) of interest.
 This is, however, dependent of the complete data model.
@@
 I could not find any resources to apply this method and it is outside the scope of this thesis to investigate how this approach could be implemented.
+
+[in discussion, explain that less it is advantage of mi, not a disadv. compared to MCMC. then final par is too negative, explain scope in intro instead.]
 

### 2019-11-22 Simulation/Functions/Autocorrelation.R
convergence of chain variances added

@@ -1,16 +1,23 @@
 # Compute the maximum auto-correlation at lag 1
 # Used within imputation function
 
-autocorr_function <- function(imp, maxit, m = 5, n.var = 4){
+autocorr_function <- function(imp, maxit, m = 5, n.var = 4, moment = "mean"){
   
   # object for for loop
   ac <- matrix(NA, nrow = m, ncol = n.var)
+  
+  # make suitable for convergence of mean and variance
+  if (moment == "mean"){
+    sims <- imp$chainMean
+  } else if (moment == "variance"){
+    sims <- imp$chainVar
+  }
   
   # compute convergence diagnostic per variable
   for (v in 1:n.var) {
     for (chain in 1:m) {
       ac[chain, v] <-
-        acf(imp$chainMean[v, , chain],
+        acf(sims[v, , chain],
           lag.max = 1,
           plot = F)$acf[-1]
   }

### 2019-11-24 Simulation/Writeup/NoteConvergence.Rnw
convergence of chain variances added
updated table and figure

@@
 This means that at most one initial value is necessary, but probably none at all.
-[I should look this up.] Without over-dispersed initial states, $\widehat{R}$ can falsely diagnose convergence \citep{broo98}.
+[I should look this up.]
+
+Starting values of MICE chains: "For each j, fill in starting imputations $Y^0_j$ by random draws from $Y^{obs}_j$." \cite[\S~4.5.2]{buur18} before for loops over variables and imputations, see \url{https://stefvanbuuren.name/fimd/sec-FCS.html#def:mice}
+
+Without over-dispersed initial states, $\widehat{R}$ can falsely diagnose convergence \citep{broo98}.
 %would not be sensitive enough to flag non-convergence of MI algorithms: "if over-dispersion does not hold, $\sigma_+^2$ can be too low, which can lead to falsely diagnosing convergence." \cite[p~437]{broo98}.
@@
   \resizebox{\textwidth}{!}{ %notice the \resizebox{} command
-        \includegraphics{resultsA4.pdf}
+        \includegraphics{Figures/8plotresults.pdf}
   }      
@@
 % latex table generated in R 3.6.1 by xtable 1.8-4 package
-% Tue Nov 12 17:43:11 2019
+% Sun Nov 24 19:18:40 2019
 \begin{table}[ht]
 \centering
-\begin{tabular}{lrrrrrrr}
+\caption{Simulation and convergence diagnostics over 1000 MCMC simulations.} 
+\label{results}
+\begin{tabular}{lrrrrrrrr}
   \hline
-  &\multicolumn{4}{c}{Simulation Diagnostics}& & \multicolumn{2}{c}{Convergence Diagn.} \\
-  \cline{2-5} \cline{7-8}
-It. & Bias & Emp. SE & CI width & Cov. rate & & $\widehat{R}$ & Auto-corr. \\ 
+ & Bias & Emp.SE & CI.width & Cov.rate & R.hat & R.var & Autocorr & AC.var \\ 
   \hline
-  1 & -0.135 & 0.115 & 0.637 & 0.930 &  & & \\ 
-  2 & -0.088 & 0.125 & 0.691 & 0.990 & & 1.499 & -0.500 \\ 
-  3 & -0.090 & 0.112 & 0.623 & 0.950 & &  1.200 & -0.658 \\ 
-  4 & -0.093 & 0.116 & 0.645 & 0.980 &  & 1.146 & -0.738 \\ 
-  5 & -0.087 & 0.108 & 0.599 & 0.980 &  & 1.098 & -0.711 \\ 
-  6 & -0.092 & 0.116 & 0.644 & 0.950 & &  1.090 & -0.674 \\ 
-  7 & -0.095 & 0.122 & 0.675 & 1.000 & &  1.063 & -0.588 \\ 
-  8 & -0.099 & 0.113 & 0.626 & 0.970 & &  1.058 & -0.523 \\ 
-  9 & -0.095 & 0.115 & 0.639 & 1.000 & &  1.044 & -0.519 \\ 
-  10 & -0.090 & 0.111 & 0.618 & 0.990 & &  1.048 & -0.414 \\ 
-  15 & -0.093 & 0.109 & 0.604 & 0.950 & &  1.020 & -0.339 \\ 
-  20 & -0.098 & 0.109 & 0.606 & 0.980 & &  1.023 & -0.234 \\ 
-  25 & -0.100 & 0.113 & 0.626 & 0.990 &  & 1.016 & -0.091 \\ 
-  30 & -0.091 & 0.118 & 0.654 & 0.960 & &  1.014 & -0.153 \\ 
-  40 & -0.088 & 0.117 & 0.650 & 0.990 &  & 1.009 & -0.070 \\ 
-  50 & -0.087 & 0.119 & 0.662 & 0.990 &  & 1.009 & -0.029 \\ 
-  % 11 & -0.100 & 0.127 & 0.704 & 0.970 & 1.046 & -0.360 \\ 
-  % 12 & -0.096 & 0.115 & 0.641 & 0.980 & 1.032 & -0.354 \\ 
-  % 13 & -0.098 & 0.120 & 0.664 & 0.960 & 1.034 & -0.328 \\ 
-  % 14 & -0.094 & 0.113 & 0.630 & 0.990 & 1.030 & -0.353 \\ 
-  % 15 & -0.093 & 0.109 & 0.604 & 0.950 & 1.020 & -0.339 \\ 
-  % 16 & -0.094 & 0.114 & 0.630 & 0.980 & 1.025 & -0.272 \\ 
-  % 17 & -0.091 & 0.116 & 0.645 & 0.980 & 1.028 & -0.243 \\ 
-  % 18 & -0.091 & 0.116 & 0.645 & 0.990 & 1.023 & -0.179 \\ 
-  % 19 & -0.087 & 0.114 & 0.630 & 0.980 & 1.024 & -0.281 \\ 
-  % 20 & -0.098 & 0.109 & 0.606 & 0.980 & 1.023 & -0.234 \\ 
-  % 21 & -0.102 & 0.116 & 0.646 & 0.960 & 1.017 & -0.192 \\ 
-  % 22 & -0.095 & 0.114 & 0.633 & 0.980 & 1.016 & -0.119 \\ 
-  % 23 & -0.088 & 0.107 & 0.595 & 0.980 & 1.017 & -0.246 \\ 
-  % 24 & -0.097 & 0.112 & 0.621 & 0.980 & 1.016 & -0.148 \\ 
-  % 25 & -0.100 & 0.113 & 0.626 & 0.990 & 1.016 & -0.091 \\ 
-  % 26 & -0.096 & 0.113 & 0.628 & 0.980 & 1.017 & -0.177 \\ 
-  % 27 & -0.096 & 0.117 & 0.650 & 0.980 & 1.014 & -0.167 \\ 
-  % 28 & -0.102 & 0.116 & 0.645 & 0.960 & 1.015 & -0.038 \\ 
-  % 29 & -0.098 & 0.114 & 0.632 & 0.990 & 1.016 & -0.137 \\ 
-  % 30 & -0.091 & 0.118 & 0.654 & 0.960 & 1.014 & -0.153 \\ 
-  % 31 & -0.098 & 0.115 & 0.639 & 0.980 & 1.012 & -0.101 \\ 
-  % 32 & -0.096 & 0.113 & 0.626 & 0.940 & 1.011 & -0.095 \\ 
-  % 33 & -0.089 & 0.117 & 0.652 & 0.970 & 1.011 & -0.088 \\ 
-  % 34 & -0.093 & 0.115 & 0.639 & 0.960 & 1.013 & -0.061 \\ 
-  % 35 & -0.103 & 0.115 & 0.638 & 0.970 & 1.011 & -0.126 \\ 
-  % 36 & -0.090 & 0.113 & 0.630 & 0.980 & 1.010 & -0.161 \\ 
-  % 37 & -0.098 & 0.122 & 0.675 & 0.990 & 1.011 & -0.070 \\ 
-  % 38 & -0.095 & 0.118 & 0.656 & 0.970 & 1.011 & -0.119 \\ 
-  % 39 & -0.091 & 0.121 & 0.670 & 0.980 & 1.009 & -0.100 \\ 
-  % 40 & -0.088 & 0.117 & 0.650 & 0.990 & 1.009 & -0.070 \\ 
-  % 41 & -0.095 & 0.117 & 0.648 & 0.960 & 1.010 & -0.148 \\ 
-  % 42 & -0.095 & 0.118 & 0.657 & 0.990 & 1.009 & -0.117 \\ 
-  % 43 & -0.095 & 0.116 & 0.645 & 0.980 & 1.009 & -0.055 \\ 
-  % 44 & -0.101 & 0.125 & 0.692 & 0.980 & 1.009 & -0.068 \\ 
-  % 45 & -0.085 & 0.117 & 0.651 & 0.980 & 1.008 & -0.086 \\ 
-  % 46 & -0.092 & 0.117 & 0.649 & 0.970 & 1.008 & -0.048 \\ 
-  % 47 & -0.095 & 0.115 & 0.638 & 0.980 & 1.007 & -0.069 \\ 
-  % 48 & -0.096 & 0.112 & 0.620 & 0.970 & 1.009 & -0.076 \\ 
-  % 49 & -0.090 & 0.108 & 0.601 & 0.980 & 1.009 & -0.083 \\ 
-  % 50 & -0.087 & 0.119 & 0.662 & 0.990 & 1.009 & -0.029 \\ 
-   \hline
+1 & -0.279 & 0.072 & 0.905 & 0.851 & \textsuperscript{a} & \textsuperscript{a} & \textsuperscript{a} & \textsuperscript{a} \\ 
+  2 & -0.171 & 0.070 & 0.924 & 0.947 & 1.665 & 1.635 & -0.500 & -0.500 \\ 
+  3 & -0.162 & 0.068 & 0.901 & 0.954 & 1.322 & 1.306 & -0.660 & -0.660 \\ 
+  4 & -0.158 & 0.072 & 0.899 & 0.961 & 1.459 & 1.449 & -0.738 & -0.737 \\ 
+  5 & -0.162 & 0.071 & 0.907 & 0.951 & 1.469 & 1.471 & -0.706 & -0.706 \\ 
+  6 & -0.160 & 0.072 & 0.910 & 0.962 & 1.252 & 1.256 & -0.652 & -0.658 \\ 
+  7 & -0.160 & 0.072 & 0.921 & 0.967 & 1.265 & 1.259 & -0.572 & -0.582 \\ 
+  8 & -0.158 & 0.069 & 0.914 & 0.963 & 1.181 & 1.178 & -0.496 & -0.510 \\ 
+  9 & -0.162 & 0.072 & 0.923 & 0.963 & 1.188 & 1.182 & -0.434 & -0.472 \\ 
+  10 & -0.161 & 0.072 & 0.911 & 0.968 & 1.139 & 1.138 & -0.396 & -0.452 \\ 
+  15 & -0.161 & 0.073 & 0.923 & 0.960 & 1.097 & 1.101 & -0.273 & -0.301 \\ 
+  20 & -0.158 & 0.070 & 0.916 & 0.971 & 1.067 & 1.066 & -0.187 & -0.212 \\ 
+  25 & -0.159 & 0.071 & 0.885 & 0.961 & 1.056 & 1.055 & -0.122 & -0.166 \\ 
+  50 & -0.159 & 0.070 & 0.906 & 0.950 & 1.026 & 1.026 & -0.032 & -0.071 \\ 
+  100 & -0.157 & 0.069 & 0.921 & 0.973 & 1.013 & 1.013 & 0.012 & -0.013 \\ 
+   \hline %{\footnotesize{Note: \textsuperscript{a}  signifies that no convergence diagnostic could be computed.}
 \end{tabular}
-\caption{Simulation and convergence diagnostics over 1000 MCMC simulations.}
-\label{subsetresults}
 \end{table}
 
-From the simulation diagnostics it appears that as little as three iterations is sufficient to draw valid inference.
+From \textsuperscript{a} the simulation diagnostics it appears that as little as three iterations is sufficient to draw valid inference.
 Convergence diagnostics $\widehat{R}$ and auto-correlation, however, indicate 20-30 iterations.
@@
 How could $\widehat{R}$ smaller than 1 occur?
-The number of simulations is smaller than in `regular' MCMC processes[explain why this is positive!].
+The number of simulations is smaller than in `regular' MCMC processes [explain that less iterations is an advantage of MI, not a disadvantage compared to MCMC].
 Therefore, the `$(n-1/n)$' [add equation number] correction factor can influence the estimated potential scale reduction factor.
@@
 
-[in discussion, explain that less it is advantage of mi, not a disadv. compared to MCMC. then final par is too negative, explain scope in intro instead.]
+[final par is too negative, explain scope in intro instead.]
 

### 2019-11-24 Simulation/Writeup/NoteConvergence.tex
convergence of chain variances added
updated table and figure

@@
 
+% % latex table generated in R 3.6.1 by xtable 1.8-4 package
+% % Tue Nov 12 17:43:11 2019
+% \begin{table}[ht]
+% \centering
+% \begin{tabular}{lrrrrrrr}
+%   \hline
+%   &\multicolumn{4}{c}{Simulation Diagnostics}& & \multicolumn{2}{c}{Convergence Diagn.} \\
+%   \cline{2-5} \cline{7-8}
+% It. & Bias & Emp. SE & CI width & Cov. rate & & $\widehat{R}$ & Auto-corr. \\ 
+%   \hline
+% ...
+% \end{tabular}
+% \caption{Simulation and convergence diagnostics over 1000 MCMC simulations.}
+% \label{subsetresults}
+% \end{table}
+
 % latex table generated in R 3.6.1 by xtable 1.8-4 package
-% Tue Nov 12 17:43:11 2019
+% Sun Nov 24 19:18:40 2019
 \begin{table}[ht]
 \centering
-\begin{tabular}{lrrrrrrr}
+\caption{Simulation and convergence diagnostics over 1000 MCMC simulations.} 
+\label{results}
+\begin{tabular}{lrrrrrrrr}
   \hline
-  &\multicolumn{4}{c}{Simulation Diagnostics}& & \multicolumn{2}{c}{Convergence Diagn.} \\
-  \cline{2-5} \cline{7-8}
-It. & Bias & Emp. SE & CI width & Cov. rate & & $\widehat{R}$ & Auto-corr. \\ 
+ & Bias & Emp.SE & CI.width & Cov.rate & R.hat & R.var & Autocorr & AC.var \\ 
   \hline
-  1 & -0.135 & 0.115 & 0.637 & 0.930 &  & & \\ 
-  2 & -0.088 & 0.125 & 0.691 & 0.990 & & 1.499 & -0.500 \\ 
-  3 & -0.090 & 0.112 & 0.623 & 0.950 & &  1.200 & -0.658 \\ 
-  4 & -0.093 & 0.116 & 0.645 & 0.980 &  & 1.146 & -0.738 \\ 
-  5 & -0.087 & 0.108 & 0.599 & 0.980 &  & 1.098 & -0.711 \\ 
-  6 & -0.092 & 0.116 & 0.644 & 0.950 & &  1.090 & -0.674 \\ 
-  7 & -0.095 & 0.122 & 0.675 & 1.000 & &  1.063 & -0.588 \\ 
-  8 & -0.099 & 0.113 & 0.626 & 0.970 & &  1.058 & -0.523 \\ 
-  9 & -0.095 & 0.115 & 0.639 & 1.000 & &  1.044 & -0.519 \\ 
-  10 & -0.090 & 0.111 & 0.618 & 0.990 & &  1.048 & -0.414 \\ 
-  15 & -0.093 & 0.109 & 0.604 & 0.950 & &  1.020 & -0.339 \\ 
-  20 & -0.098 & 0.109 & 0.606 & 0.980 & &  1.023 & -0.234 \\ 
-  25 & -0.100 & 0.113 & 0.626 & 0.990 &  & 1.016 & -0.091 \\ 
-  30 & -0.091 & 0.118 & 0.654 & 0.960 & &  1.014 & -0.153 \\ 
-  40 & -0.088 & 0.117 & 0.650 & 0.990 &  & 1.009 & -0.070 \\ 
-  50 & -0.087 & 0.119 & 0.662 & 0.990 &  & 1.009 & -0.029 \\ 
-  % 11 & -0.100 & 0.127 & 0.704 & 0.970 & 1.046 & -0.360 \\ 
-  % 12 & -0.096 & 0.115 & 0.641 & 0.980 & 1.032 & -0.354 \\ 
-  % 13 & -0.098 & 0.120 & 0.664 & 0.960 & 1.034 & -0.328 \\ 
-  % 14 & -0.094 & 0.113 & 0.630 & 0.990 & 1.030 & -0.353 \\ 
-  % 15 & -0.093 & 0.109 & 0.604 & 0.950 & 1.020 & -0.339 \\ 
-  % 16 & -0.094 & 0.114 & 0.630 & 0.980 & 1.025 & -0.272 \\ 
-  % 17 & -0.091 & 0.116 & 0.645 & 0.980 & 1.028 & -0.243 \\ 
-  % 18 & -0.091 & 0.116 & 0.645 & 0.990 & 1.023 & -0.179 \\ 
-  % 19 & -0.087 & 0.114 & 0.630 & 0.980 & 1.024 & -0.281 \\ 
-  % 20 & -0.098 & 0.109 & 0.606 & 0.980 & 1.023 & -0.234 \\ 
-  % 21 & -0.102 & 0.116 & 0.646 & 0.960 & 1.017 & -0.192 \\ 
-  % 22 & -0.095 & 0.114 & 0.633 & 0.980 & 1.016 & -0.119 \\ 
-  % 23 & -0.088 & 0.107 & 0.595 & 0.980 & 1.017 & -0.246 \\ 
-  % 24 & -0.097 & 0.112 & 0.621 & 0.980 & 1.016 & -0.148 \\ 
-  % 25 & -0.100 & 0.113 & 0.626 & 0.990 & 1.016 & -0.091 \\ 
-  % 26 & -0.096 & 0.113 & 0.628 & 0.980 & 1.017 & -0.177 \\ 
-  % 27 & -0.096 & 0.117 & 0.650 & 0.980 & 1.014 & -0.167 \\ 
-  % 28 & -0.102 & 0.116 & 0.645 & 0.960 & 1.015 & -0.038 \\ 
-  % 29 & -0.098 & 0.114 & 0.632 & 0.990 & 1.016 & -0.137 \\ 
-  % 30 & -0.091 & 0.118 & 0.654 & 0.960 & 1.014 & -0.153 \\ 
-  % 31 & -0.098 & 0.115 & 0.639 & 0.980 & 1.012 & -0.101 \\ 
-  % 32 & -0.096 & 0.113 & 0.626 & 0.940 & 1.011 & -0.095 \\ 
-  % 33 & -0.089 & 0.117 & 0.652 & 0.970 & 1.011 & -0.088 \\ 
-  % 34 & -0.093 & 0.115 & 0.639 & 0.960 & 1.013 & -0.061 \\ 
-  % 35 & -0.103 & 0.115 & 0.638 & 0.970 & 1.011 & -0.126 \\ 
-  % 36 & -0.090 & 0.113 & 0.630 & 0.980 & 1.010 & -0.161 \\ 
-  % 37 & -0.098 & 0.122 & 0.675 & 0.990 & 1.011 & -0.070 \\ 
-  % 38 & -0.095 & 0.118 & 0.656 & 0.970 & 1.011 & -0.119 \\ 
-  % 39 & -0.091 & 0.121 & 0.670 & 0.980 & 1.009 & -0.100 \\ 
-  % 40 & -0.088 & 0.117 & 0.650 & 0.990 & 1.009 & -0.070 \\ 
-  % 41 & -0.095 & 0.117 & 0.648 & 0.960 & 1.010 & -0.148 \\ 
-  % 42 & -0.095 & 0.118 & 0.657 & 0.990 & 1.009 & -0.117 \\ 
-  % 43 & -0.095 & 0.116 & 0.645 & 0.980 & 1.009 & -0.055 \\ 
-  % 44 & -0.101 & 0.125 & 0.692 & 0.980 & 1.009 & -0.068 \\ 
-  % 45 & -0.085 & 0.117 & 0.651 & 0.980 & 1.008 & -0.086 \\ 
-  % 46 & -0.092 & 0.117 & 0.649 & 0.970 & 1.008 & -0.048 \\ 
-  % 47 & -0.095 & 0.115 & 0.638 & 0.980 & 1.007 & -0.069 \\ 
-  % 48 & -0.096 & 0.112 & 0.620 & 0.970 & 1.009 & -0.076 \\ 
-  % 49 & -0.090 & 0.108 & 0.601 & 0.980 & 1.009 & -0.083 \\ 
-  % 50 & -0.087 & 0.119 & 0.662 & 0.990 & 1.009 & -0.029 \\ 
-   \hline
+1 & -0.279 & 0.072 & 0.905 & 0.851 & \textsuperscript{a} & \textsuperscript{a} & \textsuperscript{a} & \textsuperscript{a} \\ 
+  2 & -0.171 & 0.070 & 0.924 & 0.947 & 1.665 & 1.635 & -0.500 & -0.500 \\ 
+  3 & -0.162 & 0.068 & 0.901 & 0.954 & 1.322 & 1.306 & -0.660 & -0.660 \\ 
+  4 & -0.158 & 0.072 & 0.899 & 0.961 & 1.459 & 1.449 & -0.738 & -0.737 \\ 
+  5 & -0.162 & 0.071 & 0.907 & 0.951 & 1.469 & 1.471 & -0.706 & -0.706 \\ 
+  6 & -0.160 & 0.072 & 0.910 & 0.962 & 1.252 & 1.256 & -0.652 & -0.658 \\ 
+  7 & -0.160 & 0.072 & 0.921 & 0.967 & 1.265 & 1.259 & -0.572 & -0.582 \\ 
+  8 & -0.158 & 0.069 & 0.914 & 0.963 & 1.181 & 1.178 & -0.496 & -0.510 \\ 
+  9 & -0.162 & 0.072 & 0.923 & 0.963 & 1.188 & 1.182 & -0.434 & -0.472 \\ 
+  10 & -0.161 & 0.072 & 0.911 & 0.968 & 1.139 & 1.138 & -0.396 & -0.452 \\ 
+  15 & -0.161 & 0.073 & 0.923 & 0.960 & 1.097 & 1.101 & -0.273 & -0.301 \\ 
+  20 & -0.158 & 0.070 & 0.916 & 0.971 & 1.067 & 1.066 & -0.187 & -0.212 \\ 
+  25 & -0.159 & 0.071 & 0.885 & 0.961 & 1.056 & 1.055 & -0.122 & -0.166 \\ 
+  50 & -0.159 & 0.070 & 0.906 & 0.950 & 1.026 & 1.026 & -0.032 & -0.071 \\ 
+  100 & -0.157 & 0.069 & 0.921 & 0.973 & 1.013 & 1.013 & 0.012 & -0.013 \\ 
+   \hline %{\footnotesize{Note: \textsuperscript{a}  signifies that no convergence diagnostic could be computed.}
 \end{tabular}
-\caption{Simulation and convergence diagnostics over 1000 MCMC simulations.}
-\label{subsetresults}
 \end{table}
 
-From the simulation diagnostics it appears that as little as three iterations is sufficient to draw valid inference.
+From \textsuperscript{a} the simulation diagnostics it appears that as little as three iterations is sufficient to draw valid inference.
 Convergence diagnostics $\widehat{R}$ and auto-correlation, however, indicate 20-30 iterations.

### 2019-11-29 Simulation/Writeup/NoteConvergence.Rnw
Remarks meeting with Gerko on overcoverage
Adjust table with results

@@
 % latex table generated in R 3.6.1 by xtable 1.8-4 package
-% Sun Nov 24 19:18:40 2019
+% Fri Nov 29 11:38:27 2019
 \begin{table}[ht]
@@
 \label{results}
-\begin{tabular}{lrrrrrrrr}
+\begin{tabular}{lrrrrrrr}
   \hline
- & Bias & Emp.SE & CI.width & Cov.rate & R.hat & R.var & Autocorr & AC.var \\ 
+It. & Bias & CI width & Cov. rate & $\widehat{R}$ & $\widehat{R}$ & Auto-corr. & Auto-corr. \\ 
   \hline
-1 & -0.279 & 0.072 & 0.905 & 0.851 & \textsuperscript{a} & \textsuperscript{a} & \textsuperscript{a} & \textsuperscript{a} \\ 
-  2 & -0.171 & 0.070 & 0.924 & 0.947 & 1.665 & 1.635 & -0.500 & -0.500 \\ 
-  3 & -0.162 & 0.068 & 0.901 & 0.954 & 1.322 & 1.306 & -0.660 & -0.660 \\ 
-  4 & -0.158 & 0.072 & 0.899 & 0.961 & 1.459 & 1.449 & -0.738 & -0.737 \\ 
-  5 & -0.162 & 0.071 & 0.907 & 0.951 & 1.469 & 1.471 & -0.706 & -0.706 \\ 
-  6 & -0.160 & 0.072 & 0.910 & 0.962 & 1.252 & 1.256 & -0.652 & -0.658 \\ 
-  7 & -0.160 & 0.072 & 0.921 & 0.967 & 1.265 & 1.259 & -0.572 & -0.582 \\ 
-  8 & -0.158 & 0.069 & 0.914 & 0.963 & 1.181 & 1.178 & -0.496 & -0.510 \\ 
-  9 & -0.162 & 0.072 & 0.923 & 0.963 & 1.188 & 1.182 & -0.434 & -0.472 \\ 
-  10 & -0.161 & 0.072 & 0.911 & 0.968 & 1.139 & 1.138 & -0.396 & -0.452 \\ 
-  15 & -0.161 & 0.073 & 0.923 & 0.960 & 1.097 & 1.101 & -0.273 & -0.301 \\ 
-  20 & -0.158 & 0.070 & 0.916 & 0.971 & 1.067 & 1.066 & -0.187 & -0.212 \\ 
-  25 & -0.159 & 0.071 & 0.885 & 0.961 & 1.056 & 1.055 & -0.122 & -0.166 \\ 
-  50 & -0.159 & 0.070 & 0.906 & 0.950 & 1.026 & 1.026 & -0.032 & -0.071 \\ 
-  100 & -0.157 & 0.069 & 0.921 & 0.973 & 1.013 & 1.013 & 0.012 & -0.013 \\ 
-   \hline %{\footnotesize{Note: \textsuperscript{a}  signifies that no convergence diagnostic could be computed.}
+   1 & -0.128 & 0.922 & 0.930 & NA & NA & NA & NA \\ 
+     2 & -0.002 & 0.915 & 0.940 & 1.686 & 1.709 & -0.500 & -0.500 \\ 
+     3 & 0.012 & 0.923 & 0.960 & 1.317 & 1.326 & -0.661 & -0.659 \\ 
+     4 & 0.019 & 0.892 & 0.970 & 1.461 & 1.464 & -0.735 & -0.746 \\ 
+     5 & 0.015 & 0.906 & 0.910 & 1.486 & 1.472 & -0.706 & -0.705 \\ 
+     6 & 0.010 & 0.928 & 0.950 & 1.252 & 1.261 & -0.651 & -0.667 \\ 
+     7 & 0.006 & 0.908 & 0.910 & 1.259 & 1.267 & -0.619 & -0.584 \\ 
+     8 & 0.018 & 0.920 & 0.940 & 1.178 & 1.181 & -0.516 & -0.549 \\ 
+     9 & 0.011 & 0.994 & 0.940 & 1.190 & 1.186 & -0.506 & -0.488 \\ 
+    10 & 0.020 & 0.940 & 0.900 & 1.142 & 1.145 & -0.448 & -0.420 \\ 
+    15 & 0.009 & 0.917 & 0.960 & 1.098 & 1.109 & -0.256 & -0.289 \\ 
+    20 & 0.020 & 0.848 & 0.930 & 1.069 & 1.065 & -0.179 & -0.214 \\ 
+    25 & 0.026 & 0.897 & 0.970 & 1.054 & 1.054 & -0.146 & -0.131 \\ 
+    50 & 0.008 & 0.946 & 0.970 & 1.026 & 1.026 & -0.029 & -0.096 \\ 
+   100 & 0.011 & 0.967 & 0.930 & 1.013 & 1.013 & 0.054 & -0.020 \\ 
+   \hline
 \end{tabular}
@@
 
-From \textsuperscript{a} the simulation diagnostics it appears that as little as three iterations is sufficient to draw valid inference.
+From the simulation diagnostics it appears that as little as three iterations is sufficient to draw valid inference.
 Convergence diagnostics $\widehat{R}$ and auto-correlation, however, indicate 20-30 iterations.

### 2019-11-29 Simulation/Writeup/NoteConvergence.tex
Remarks meeting with Gerko on overcoverage
Adjust table with results

@@
 
-% % latex table generated in R 3.6.1 by xtable 1.8-4 package
-% % Tue Nov 12 17:43:11 2019
-% \begin{table}[ht]
-% \centering
-% \begin{tabular}{lrrrrrrr}
-%   \hline
-%   &\multicolumn{4}{c}{Simulation Diagnostics}& & \multicolumn{2}{c}{Convergence Diagn.} \\
-%   \cline{2-5} \cline{7-8}
-% It. & Bias & Emp. SE & CI width & Cov. rate & & $\widehat{R}$ & Auto-corr. \\ 
-%   \hline
-% ...
-% \end{tabular}
-% \caption{Simulation and convergence diagnostics over 1000 MCMC simulations.}
-% \label{subsetresults}
-% \end{table}
-
 % latex table generated in R 3.6.1 by xtable 1.8-4 package
-% Sun Nov 24 19:18:40 2019
+% Fri Nov 29 11:38:27 2019
 \begin{table}[ht]

### 2019-12-04 Simulation/Writeup/NoteConvergence.Rnw
Finally integrated comments, also created nicer graph and worked on better flow in introduction section.

@@
 An abstract may be included, but is not necessary}''.
-I aim to publish a pre-print of this research report on ArXiv (\url{https://arxiv.org/}).
+I aim to publish a pre-print of this research report on \href{https://arxiv.org/}{ArXiv}.
 That way, I can refer to the simulation study described here, without 'bulking up' the technical paper that I want to submit for publication in JSS.
 Another option would be to attach this note as online appendix to the technical paper on ShinyMICE.
+\textbf{TO ADD: appendix with extra results tables with different ampute and or very high or very low correlations between predictors}.
 %"It is the first half of the thesis, i.e., there are no results included yet, but the report contains a full introduction including a literature review and a methods section that contains details about the data, instruments and or statistical procedures".
@@
 At some point, any scientist conducting statistical analyses will run into a missing data problem \citep{alli02}.
-Missingness is problematic because statistical inference cannot be performed on incomplete data, and ad hoc solutions (e.g., listwise deletion) can yield wildly invalid results \citep{buur18}.
-To circumvent the ubiquitous problem of missing information, \cite{rubin87} proposed the framework of multiple imputation (MI).
-MI is an iterative algorithmic procedure in which missing data points are 'guessed' (i.e.
+Missingness is problematic because statistical inference cannot be performed on incomplete data without employing \emph{ad hoc} solutions (e.g., listwise deletion), which can yield wildly invalid results \citep{buur18}.
+A popular answer to the ubiquitous problem of missing information is to use the framework of multiple imputation (MI), proposed by \cite{rubin87}.
+MI is an iterative algorithmic procedure in which each missing data point is `guessed' (i.e.
 imputed) several times.
-[Add some stuff here about MICE.] The variability between the imputations validly reflects how much uncertainty in the inference is due to missing information--that is, if all statistical assumptions are met \citep{rubin87} [adjust this sentence to be more clear!].
+The variablity between imputations is used to reflect how much uncertainty in the inference is introduced by the missingness.
+Therefore, MI can provide valid inferences despite of missing information.
 
-[Feedback: The order of the introduction is not intuitive.
-some terms are mentioned first and only explained later.
-There are some short sentences with a lot of content, so implicitly a lot happens.
-Expand these sentences or delete (because the thing would be too obvious?).
-Explain 'true data generating model'.
-Follow up other important concepts with explanation or examples too.]
-%Missing data is ubiquitous. MICE can solve stuff.
-
-With MI, many assumptions are made about the nature of the observed and missing parts of the data and their relation to the `true' \emph{data generating model} \citep{buur18} [explain what this is!].
-Without proper evaluation of the imputations and the underlying assumptions, any drawn inference may erroneously be deemed valid.
+\textbf{ADD some references in this paragraph!} To obtain valid inferences with MI, the variability between imputations should be correct.
+If the variance between imputations is under-estimated, confidence intervals around estimates will be too narrow, which can yield spurious results.
+On the other hand, over-estimation of the variance between imputations results in unnecessarily wide confidence intervals, which can be costly because it requires more observations than strictly required.
+Since both of these situations are undesirable, imputations and their variability should be properly evaluated.
 Such evaluation measures are currently missing or under-developed in MI software, like the world leading \proglang{R} package \pkg{mice} \citep{mice}.
 %Therefore, I will answer the following question: 'Which measures are vital for evaluating the validity of multiply imputed data?'.
+The aim of this note is to provide guidelines for the evaulation of imputations to applied researchers who use MI methods.
+More specifically, this note adresses the question: `How to diagnose convergence of the MI algorithm implemented in \pkg{mice}?'.
 
-[Feedback: Look at abbreviations. For example r hat is just metnioned as 'e.g.'. It's not clear in the introduction section how important it is later.]
-
-A fundamental assumption of MICE is convergence of the algorithm.
-MICE is a sort of Gibbs sampler, a type of Markov chain Monte Carlo (MCMC) algorithm.
-In general, the validity of inference resulting from MCMC algorithms is threatened by non-convergence.
-Hence we use convergence diagnostics to flag non-convergence.
-But it is not known whether conventional convergence diagnostics for MCMC methods work on MI data.
-
-Conventional thresholds to diagnose non-convergence-- e.g., Gelman and Rubin's \citeyear{gelm92} statistic $\widehat{R} < 1.1$ --are not applicable on multiply imputed data \citep{lace07} [Expand or remove Rhat here!
-Refer to future section?].
-Therefore, empirical researchers have to rely on visual inspection procedures that are theoretically equivalent to $\widehat{R}$ \citep{whit11} [explain why this is appropriate, and numerical is not].
-Visually assessing convergence is not only difficult to the untrained eye, it might also be futile.
-The convergence properties of MI algorithms lack scientific consensus \citep{taka17}, and some default MICE techniques might not converge to stable distributions at all \citep{murr18}.
-Moreover, convergence diagnostics for MI methods have not been systematically studied \citep{buur18}.
-
-This paper investigates convergence properties of MI algorithms by means of model-based simulation in \proglang{R} \citep{R}.
-All programming code used in this note is available in the file 'XYZ.R' along with the manuscript, and on Github repository 'XYZ'.
+The convergence properties of \pkg{mice} are investigated by means of model-based simulation in \proglang{R} \citep{R}.
+All programming code used in this note is available on Github repository \href{https://github.com/gerkovink/shinyMice}{`ShinyMICE'}.
 The results of this simulation study are guidelines for applied researchers to evaluate convergence of MI algorithms.
+These guidelines will be implemented in the interactive evaluation tool for \pkg{mice}, `ShinyMICE', which is currently under development.
 %I will replicate Lacerda et al.'s simulation study on $\widehat{R}$ \citep{lace07}, and develop novel guidelines for assessing convergence.
@@
 
-The aim of this paper is to provide applied researchers a tool to assess convergence of multiple imputation algorithms.
-The intended audience consists of empirical researchers and statisticians who use multiple imputation to solve missing data problems.
-Basic familiarity with multiple imputation methodology is assumed.
+The intended audience of this note consists of empirical researchers and statisticians who use multiple imputation to solve missing data problems.
+Basic familiarity with MI methodology is assumed.
+For the theoretical foundation of MI, see \cite{rubin87}.
 For an accessible and comprehensive introduction to MI from an applied perspective, see \cite{buur18}.
-For the theoretical foundation of MI, see \cite{rubin87}.
 
-[explain difference MI, MICE, mice]
+The convergence guidelines introduced in this paper are developed to be integrated into the \pkg{mice} environment \citep{mice} in \proglang{R} \citep{R}.
+This note therefore follows notation and conventions of \cite{mice}.
+Deviations from the 'original' notation by \cite{rubin87}, are described in \citep[\S~XYZ]{buur18}.
 
-[Explain the use of `imputation chain' (the iterations of an imputation, only the final imputed values are used), starting values of chains, `simulation condition' (the number of iterations the algorithm has before imputing), ...?]
+Let $Y$ denote an $n \times p$ matrix containing the data values on $p$ variables for all $n$ units in a sample.
+The collection of observed data values in $Y$ is denoted as $Y_{obs}$; the missing part of $Y$ is referred to as $Y_{mis}$.
+Response indicator $R$ shows whether a data value in $Y$ is missing or observed.
+The relation between $R$, $Y_{obs}$, and $Y_{mis}$ determines the missingness mechanism.
+This paper only considers a 'missing completely at random' (MCAR) mechanism, where the probability to be missing is equal for all $n \times p$ cells in $Y$.
 
-[look at imputing vs analysing because they're called steps somewhere, and something else somewhere else. explain what a chain is.]
+In this note, the terms `unobserved' and `missing' data are used interchangibly.
+Both refer to $Y_{mis}$, e.g.
+`$NA$' values in \proglang{R} or `999' in \proglang{SPSS}.
+The terms 'incomplete' or 'observed' data denote $Y_{obs}$.
+Incomplete data is the starting point of the multiple imputation procedure.
+Figure \ref{fig:steps} provides a schematic overview of the steps involved with MI.
 
 \begin{figure}
+\label{fig:steps}
 \centering
@@
 
+Missing data in $Y$ is `imputed' (i.e., filled in) $m$ times.
+The imputed data is combined with observed data $Y_{obs}$ to create $m$ completed data sets.
+On each completed data set, the analysis of scientific interest (or `complete data model') is performed.
+The quantity of scientific interest (e.g., a regression coefficient) is denoted with $Q$.
+Since $Q$ is estimated on each completed data set, $m$ separate $\hat{Q}$ values are obtained.
+These $m$ values are combined into a single pooled estimate $\bar{Q}$.
+
+This note is primarily concerned with the algorithmic properties of the imputation step.
+The algorithm employed within this step has an iterative nature.
+That is, before drawing $m$ imputed values for each missing data point in $Y_{mis}$, a `chain` of potential values is considered.
+Each of the $m$ chains starts with an initial value, drawn randomly from $Y_{obs}$.
+The chains are terminated after a predefined number of iterations.
+Only the ultimate value that a chain lands on is imputed, and hence used in the next steps (analysis and pooling).
+The collection of values (or summary statistics) across iterations of one of the $m$ imputations will be referred to as `imputation chain'.
+
+\textbf{ADD THIS? The process of deliberately creating missingness in an otherwise complete data set is referred to as `amputing' the data.}
+
+\textbf{[explain difference MI, FCS, mice]}
+
+\textbf{[Explain `simulation condition' (the number of iterations the algorithm has before imputing), ...?]}
+
 % \begin{figure}[h]
@@
 
-%% -- Notation ---------------------------------------------------------------
-\subsection{Notation} \label{sec:notation}
-
-The convergence guidelines introduced in this paper are developed to be integrated into the \pkg{mice} environment \citep{mice} in \proglang{R} \citep{R}.
-It therefore follows notation and conventions of \cite{mice}.
-For an overview of the deviations from the 'original' notation by \cite{rubin87}, see \cite{buur18}.
-
-Let $Y$ denote an $n \times p$ matrix containing the data values on $p$ variables for all $n$ units in a sample.
-The collection of observed data values in $Y$ is denoted as $Y_{obs}$; the missing part of $Y$ is referred to as $Y_{mis}$.
-Response indicator $R$ shows whether a data value in $Y$ is missing or observed.
-The relation between $R$, $Y_{obs}$, and $Y_{mis}$ determines the missingness mechanism.
-This paper only considers a 'missing completely at random' (MCAR) mechanism, where the probability to be missing is equal for all $n \times p$ cells in $Y$.
-
-With multiple imputation, $Y_{mis}$ is 'filled in' (i.e.
-imputed) $m$ times.
-The $m$ imputed data sets are analyzed separately, and their results pooled [Make this more clear!].
-A distinction is made between the imputation step (employing $p$ 'imputation models'), and the analysis step (using one 'complete-data model' for each of the $m$ imputed data sets).
-The quantity of scientific interest is the result of the analysis step, e.g., a regression coefficient, and denoted with $Q$.
-That is, $Q$ is estimated on each imputed data set, resulting in $m$ separate $\hat{Q}$ values.
-These are combined into pooled estimate $\bar{Q}$.
 % "Let Q denote the quantity of scientific interest (e.g., a regression coefficient)." \cite[p.~4]{mice}.
@@
 %% -- Theoretical Background ---------------------------------------------------------------
-\section{Theoretical Background} \label{sec:background}
+\subsection{Theoretical Background (NOG NIET AF VANAF HIER)} \label{sec:background}
 
-Convergence has two interpretations [look up source!!!]: mixing between chains and stability over iterations within chains.
+\textbf{Explain the measures that I'm investigating here. R hat is ... Auto-correlation is... How are they computed? What do they say? How are they computed on MI data? Looking at imputation chains. 
+Explain that there is no baseline to compare the measures with. Only current convergence check is to visually inspect trace plots. (But in the traceplots convergence seems immediate).}
+
+\textbf{What is convergence in general?}
+
+We can never be certain of convergence, therefore convergence diagnostics evaluate signs of non-convergence \citep{hoff09}.
+% In general: "In general, you cannot know for sure if your chain has converged.
+But sometimes you can know if your chain has not converged, so we at least check for this latter possibility" \cite[p.~101]{hoff09}
+Convergence has two interpretations \textbf{[look up source!!!]}: mixing between chains and stability over iterations within chains.
 Ideally, we want the chains to intermingle nicely, without trends within chains.
+Diagnose non-convergence in the mixing sense: $\widehat{R}$.
+Diagnose non-convergence in the stability sense: auto-correlation.
+
 % "In order to converge to a stationary distribution, a Markov chain needs to satisfy three important properties (Roberts 1996; Tierney 1996): irreducible, the chain must be able to reach all interesting parts of the state space; aperiodic, the chain should not oscillate between different states; recurrence, all interesting parts can be reached infinitely often, at least from almost all starting points.
@@
 The development of practical tools that put these conditions to work for multiple imputation is still an ongoing research problem." \cite[par.~4.5]{buur18}
-We can never be certain of convergence, therefore convergence diagnostics evaluate signs of non-convergence \citep{hoff09}.
-% In general: "In general, you cannot know for sure if your chain has converged.
-But sometimes you can know if your chain has not converged, so we at least check for this latter possibility" \cite[p.~101]{hoff09}
-Diagnose non-convergence in the mixing sense: $\widehat{R}$. Diagnose non-convergence in the stability sense: auto-correlation.
 
@@
 iteration five are significantly correlated with the value of the first draw.
-Auto-correlation above the confidence level [explain that this is a critical value for correlation based on alpha .05] indicate dependence within chains.
-Also, possible to use MC error, Geweke, and the other one.
+Auto-correlation above the confidence level \textbf{[explain that this is a critical value for correlation based on alpha .05]} indicate dependence within chains.
+Also, possible to use MC error, or Geweke statistic.
 But outside of scope of this study.
@@
 
-The potential scale reduction factor $\widehat{R}$ tells us how much variance of an estimate could be shrunken down by running the chains infinitely long [Add reference, gelman something?].
+The potential scale reduction factor $\widehat{R}$ tells us how much variance of an estimate could be shrunken down by running the chains infinitely long \cite{gelm92}.
 That then tells us something about how dependent the chains are on the starting values.
@@
 
-Why is $\widehat{R}$ not applicable on MI data?
-$\widehat{R}$ is not appropriate because it assumes over-dispersed initial values, which means that the initial values of the $m$ imputation chains are 'far away' from the distribution that the chains are converging to.
-With MI procedures, initial values are chosen such that ...
-[or maybe even estimated from the observed data??].
-This means that at most one initial value is necessary, but probably none at all.
-[I should look this up.]
-
-Starting values of MICE chains: "For each j, fill in starting imputations $Y^0_j$ by random draws from $Y^{obs}_j$." \cite[\S~4.5.2]{buur18} before for loops over variables and imputations, see \url{https://stefvanbuuren.name/fimd/sec-FCS.html#def:mice}
-
-Without over-dispersed initial states, $\widehat{R}$ can falsely diagnose convergence \citep{broo98}.
-%would not be sensitive enough to flag non-convergence of MI algorithms: "if over-dispersion does not hold, $\sigma_+^2$ can be too low, which can lead to falsely diagnosing convergence." \cite[p~437]{broo98}.
-This suggests that $\widehat{R}$ would not be sensitive enough to flag non-convergence of MI algorithms.
-Empirical finding, however, show that the opposite may be true: \cite{lace07} report $\widehat{R}$ values above the threshold of $\widehat{R} < 1.1$ after fifty iterations.
-%that $\widehat{R}$ will not be smaller than 1.1 before iteration number 50.
-Therefore, the aim is to replicate empirical finding Lacerda et al.
-Hypothesis of this simulation study is that $\widehat{R}$ will over-estimate non-convergence.
-[Or use:] Hypothesis based on \cite{lace07} is that the conventional acceptable level of $\widehat{R}$ is too strict for MI data.
-We expect that the simulation diagnostics will indicate valid inference before $\widehat{R}$ will.
-
-% Also introduce:
-% 
-% - missingness mechanims, ignorability?
-% % "The practical importance of the distinction between MCAR, MAR and MNAR is that it clarifies the conditions under which we can accurately estimate the scientifically interesting parameters without the need to know $\psi$" \cite[par.~2.2]{buur18}.
-% 
-% - Rubin's rules?
-% 
-% - FCS vs. JM?
-%
-
-%% -- Methods ---------------------------------------------------------------
-
-%%
-%%
-%%
-%%
-
-\section{Methods} \label{sec:methods}
-
-This research project consists of an investigation into algorithmic convergence of MI algorithms.
-I will replicate Lacerda et al.'s simulation study on $\widehat{R}$ \citep{lace07}, and develop novel guidelines for assessing convergence.
-Ideally, I will integrate several diagnostics (e.g., $\widehat{R}$, \emph{auto-correlation}, and \emph{simulation error}) into a single summary indicator to flag non-convergence.
-
-The aim is to evaluate whether the imputation procedure has converged.
-The primary research interest is in determining whether $\widehat{R}$ is an appropriate convergence diagnostic, and if so, which level of stringency suits MI data.
 We can apply $\widehat{R}$ to the mean (or to the first two moments) of the variables of interest.
-The simulation diagnostics are as recommended by \cite{buur18}.
-Namely, average bias, average confidence interval width, and empirical coverage rate (coverage probability) across simulations.
-%"We expect that in accordance with Neyman (1934), at least 95 percent of the confidence intervals should contain the true population value, although some room for simulation error should be taken into account to counteract the finite nature of our simulations"\cite{schou18}.
-Also look at distributional characteristics, and plausibility of imputed values, see \cite{vinknd}.
-% Or use: Then evaluate the regular diagnostics of MI simulations, see Vink, n.d., and  convergence.
-Diagnostics include absolute bias of the estimated regression coefficient, confidence interval width, and empirical coverage rate across simulations.
-Convergence is evaluated with $\widehat{R}$, or potential scale reduction factor.
-
-% Distributional characteristics: In practice, the distribution of the incomplete data may differ greatly from the observed data.
-Under anything but the MCAR assumption, this can be expected.
-When evaluating imputations, the distributional shapes should be checked and diagnostic evaluations should be performed (see Abayomi et al., 2008, for an detailed overview of diagnostic evaluation for multivariate imputations).
-When anomalies are found, and if the imputation method is valid, there should be an explanation, especially in the controlled environment of a properly executed simulation study.
-
-% Plausibility of the imputed values: Plausible imputations - imputations that could be real values if they had been observed - are not a necessary condition for obtaining valid inference.
-However, in practice, especially when the imputer and the analyst are different persons, plausibility of imputations may be a desired property.
-When evaluating imputation routines, the evaluator should mention whether the routine is prone to deliver implausible value
-
-Convergence diagnostics are $\widehat{R}$ and auto-correlation (at lag one).
 $\widehat{R}$ is computed as follows: % The scalar summary of interest in this study is the average of the imputed values per variable.
@@
 
-\begin{quotation}
-``In the equations below, $N$ is the number of draws per chain, $M$ is the number of chains, and $S = MN$ is the total number of draws from all chains.
-For each scalar summary of interest $\theta$, we compute $B$ and $W$, the between- and within-chain variances:
-%
+\textbf{REPHRASE: In the equations below, $N$ is the number of draws per chain, $M$ is the number of chains, and $S = MN$ is the total number of draws from all chains.
+For each scalar summary of interest $\theta$, we compute $B$ and $W$, the between- and within-chain variances:}
+
 \begin{align*}
-B=\frac{N}{M-1} \sum_{m=1}^{M}\left(\bar{\theta}^{(\cdot m)}-\bar{\theta}^{(\cdot \cdot)}\right)^{2}, \text { where } \bar{\theta}^{(\cdot m)}=\frac{1}{N} \sum_{n=1}^{N} \theta^{(n m)}, \quad \bar{\theta}^{(\cdot \cdot)}=\frac{1}{M} \sum_{m=1}^{M} \bar{\theta}^{(\cdot m)} \\
-W=\frac{1}{M} \sum_{m=1}^{M} s_{j}^{2},  \text { where } s_{m}^{2}=\frac{1}{N-1} \sum_{n=1}^{N}\left(\theta^{(n m)}-\bar{\theta}^{(\cdot m)}\right)^{2}". \text{ \cite[p.~5]{veht19}} 
+``B&=\frac{N}{M-1} \sum_{m=1}^{M}\left(\bar{\theta}^{(\cdot m)}-\bar{\theta}^{(\cdot \cdot)}\right)^{2}, \text { where } \bar{\theta}^{(\cdot m)}=\frac{1}{N} \sum_{n=1}^{N} \theta^{(n m)}, \quad \bar{\theta}^{(\cdot \cdot)}=\frac{1}{M} \sum_{m=1}^{M} \bar{\theta}^{(\cdot m)} \\
+W&=\frac{1}{M} \sum_{m=1}^{M} s_{j}^{2},  \text { where } s_{m}^{2}=\frac{1}{N-1} \sum_{n=1}^{N}\left(\theta^{(n m)}-\bar{\theta}^{(\cdot m)}\right)^{2}". \text{ \cite[p.~5]{veht19}} 
 \end{align*}
-%
-\end{quotation}
 
@@
 
-[make clear when i talk about my study or lit. look at equations vehtari, because there is no real need for it. try to rephrase in my own words because of text flow.]
+\textbf{Convergence of MI}
 
-In this study, $\widehat{R}$ is computed for the mean of each variable, in each imputation chain, over all simulation conditions and MCMC simulations.
-%For each simulation in each simulation condition, the maximum value of $\widehat{R}$ is reported.
+A fundamental assumption of MICE is convergence of the algorithm.
+MICE is a sort of Gibbs sampler, a type of Markov chain Monte Carlo (MCMC) algorithm.
+In general, the validity of inference resulting from MCMC algorithms is threatened by non-convergence.
+Hence we use convergence diagnostics to flag non-convergence.
+But it is not known whether conventional convergence diagnostics for MCMC methods work on MI data.
+
+Why is $\widehat{R}$ not applicable on MI data?
+$\widehat{R}$ is not appropriate because it assumes over-dispersed initial values, which means that the initial values of the $m$ imputation chains are 'far away' from the distribution that the chains are converging to.
+With MI procedures, initial values are chosen randomly from the observed data.
+This means that one initial value is necessary for each imputation chain.
+
+% Starting values of MICE chains: "For each j, fill in starting imputations $Y^0_j$ by random draws from $Y^{obs}_j$." \cite[\S~4.5.2]{buur18} before for loops over variables and imputations, see \url{https://stefvanbuuren.name/fimd/sec-FCS.html#def:mice}
+
+Without over-dispersed initial states, $\widehat{R}$ can falsely diagnose convergence \citep{broo98}.
+%would not be sensitive enough to flag non-convergence of MI algorithms: "if over-dispersion does not hold, $\sigma_+^2$ can be too low, which can lead to falsely diagnosing convergence." \cite[p~437]{broo98}.
+This suggests that $\widehat{R}$ would not be sensitive enough to flag non-convergence of MI algorithms.
+Empirical finding, however, show that the opposite may be true: \cite{lace07} report $\widehat{R}$ values above the threshold of $\widehat{R} < 1.1$ after fifty iterations.
+%that $\widehat{R}$ will not be smaller than 1.1 before iteration number 50.
+Therefore, the aim is to replicate empirical finding Lacerda et al.
+
+\textbf{How is convergence checked now (current practice)?}
+
+Conventional thresholds to diagnose non-convergence-- e.g., Gelman and Rubin's \citeyear{gelm92} statistic $\widehat{R} < 1.1$ --are not applicable on multiply imputed data \citep{lace07} \textbf{[Expand or remove Rhat here!
+Refer to future section?]}.
+Therefore, empirical researchers have to rely on visual inspection procedures that are theoretically equivalent to $\widehat{R}$ \citep{whit11} \textbf{[explain why this is appropriate, and numerical is not]}.
+Visually assessing convergence is not only difficult to the untrained eye, it might also be futile.
+The convergence properties of MI algorithms lack scientific consensus \citep{taka17}, and some default MICE techniques might not converge to stable distributions at all \citep{murr18}.
+Moreover, convergence diagnostics for MI methods have not been systematically studied \citep{buur18}.
+
+% This research project consists of an investigation into algorithmic convergence of MI algorithms.
+I will replicate Lacerda et al.'s simulation study on $\widehat{R}$ \citep{lace07}, and develop novel guidelines for assessing convergence.
+Ideally, I will integrate several diagnostics (e.g., $\widehat{R}$, \emph{auto-correlation}, and \emph{simulation error}) into a single summary indicator to flag non-convergence.
+
+\subsection{Simulation Hypothesis} \label{sec:hypothesis}
+
+The aim is to evaluate whether the imputation procedure has converged.
+The primary research interest is in determining whether $\widehat{R}$ is an appropriate convergence diagnostic, and if so, which level of stringency suits MI data.
+
+Hypothesis of this simulation study is that $\widehat{R}$ will over-estimate non-convergence.
+\textbf{[Or use:] Hypothesis based on \cite{lace07} is that the conventional acceptable level of $\widehat{R}$ is too strict for MI data.
+We expect that the simulation diagnostics will indicate valid inference before $\widehat{R}$ will.}
+
+% Also introduce:
+% 
+% - missingness mechanims, ignorability?
+% % "The practical importance of the distinction between MCAR, MAR and MNAR is that it clarifies the conditions under which we can accurately estimate the scientifically interesting parameters without the need to know $\psi$" \cite[par.~2.2]{buur18}.
+% 
+% - Rubin's rules?
+% 
+% - FCS vs. JM?
+%
 
@@
 
-%% -- Simulation Approach ---------------------------------------------------------------
+%% -- Methods ---------------------------------------------------------------
 
@@
 
-\section{Simulation Approach}
+\section{Methods} \label{sec:methods}
 
-In this study, 1000 MCMC simulations are performed. The simulation set-up consists of several steps, see pseudo-code below. Simulation code is available in online appendix XYZ on Github.
+In this study, 1000 simulations are performed for each of the 100 simulation conditions.
+The simulation set-up consists of several steps, summarized in the pseudo-code below.
+The complete \proglang{R} script of the simulation study is available on Github.
 
@@
 # pseudo-code of simulation 
-simulate data with missingness
+simulate data 
 for (number of simulation runs from 1 to 1000)
   for (number of iterations from 1 to 100)
+    create missingness
     impute the missingness
@@
 
+\subsection{Data}
 The simulated dataset is a finite population of $N=1000$.
@@
 
+\subsection{Amputation}
 After data generation, the complete data is 'amputed'.
@@
 
+\subsection{Imputation}
 Missing data points are imputed with \pkg{mice} in \proglang{R}.
@@
 
-Add ``post-processing steps (e.g.
+\subsection{Convergence Diagnostics}
+% Add ``post-processing steps (e.g.
 aggregation computations, summary statistics, regressions on the simulation results) used to transform simulation outputs to reported results''.
 %, see \url{http://jsterman.scripts.mit.edu/docs/Rahmandad-Sterman%20Simulation%20Reporting%20Standards3.30%20update.pdf}.
-
-Post-processing of simulation diagnostics:
-
-Bias is computed as the estimated regression coefficient after MI minus the true regression coefficient. Bias is averaged over all simulations with the same simulation condition.
-
-Confidence interval width (CIW) is computed as the diffence between the lower and upper bound of the 95\% confidence interval (CI95\%).
-The CI95\% bounds are computed as the estimated regression coefficient plus or minus (respectively) the pooled SE across imputations times 2.66 (the quantile of a t distribution with m-1 degrees of freedom).
-CIW is averaged over all simulations with the same simulation condition.
-
-Coverage rate is computed as the proportion of simulations (with the same simulation condition) in which the true regression coefficient is between the bounds of the CI95\%.
-
-Post-processing of convergence diagnostics:
 
@@
 
+\subsection{Analysis}
+
+Linear regression is performed using the \proglang{R} function \fct{lm}.
+The quantity of scientific interest $Q$ is the regression coefficient of $X$ on $Y$.
+$Q$ is estimated in each imputation, in each simulation condition, in each repetition.
+The $m = 5$ estimated regression coefficients $\hat{Q}$ are pooled according to Rubin's \citeyear{rubin87} rules with \pkg{mice} function \fct{pool}.
+
+\subsection{Simulation diagnostics}
+The simulation diagnostics are as recommended by \cite{buur18}.
+Namely, average bias, average confidence interval width, and empirical coverage rate (coverage probability) across simulations.
+%"We expect that in accordance with Neyman (1934), at least 95 percent of the confidence intervals should contain the true population value, although some room for simulation error should be taken into account to counteract the finite nature of our simulations"\cite{schou18}.
+We could also look at distributional characteristics, and plausibility of imputed values, see \cite{vinknd}.
+For now, this is outside of the scope of this study.
+% Or use: Then evaluate the regular diagnostics of MI simulations, see Vink, n.d., and  convergence.
+Diagnostics include absolute bias of the estimated regression coefficient, confidence interval width, and empirical coverage rate across simulations.
+Convergence is evaluated with $\widehat{R}$, or potential scale reduction factor.
+
+% Distributional characteristics: In practice, the distribution of the incomplete data may differ greatly from the observed data.
+Under anything but the MCAR assumption, this can be expected.
+When evaluating imputations, the distributional shapes should be checked and diagnostic evaluations should be performed (see Abayomi et al., 2008, for an detailed overview of diagnostic evaluation for multivariate imputations).
+When anomalies are found, and if the imputation method is valid, there should be an explanation, especially in the controlled environment of a properly executed simulation study.
+
+% Plausibility of the imputed values: Plausible imputations - imputations that could be real values if they had been observed - are not a necessary condition for obtaining valid inference.
+However, in practice, especially when the imputer and the analyst are different persons, plausibility of imputations may be a desired property.
+When evaluating imputation routines, the evaluator should mention whether the routine is prone to deliver implausible value
+
+Bias is computed as the estimated regression coefficient after MI minus the true regression coefficient. Bias is averaged over all simulations with the same simulation condition.
+
+Confidence interval width (CIW) is computed as the diffence between the lower and upper bound of the 95\% confidence interval (CI95\%).
+The CI95\% bounds are computed as the estimated regression coefficient plus or minus (respectively) the pooled SE across imputations times 2.66 (the quantile of a t distribution with m-1 degrees of freedom).
+CIW is averaged over all simulations with the same simulation condition.
+
+Coverage rate is computed as the proportion of simulations (with the same simulation condition) in which the true regression coefficient is between the bounds of the CI95\%.
+
 %% -- Results ---------------------------------------------------------------
@@
 
-Figure \ref{resultssim} shows the results over 1000 simulations. A subset of simulation conditions is presented in Table \ref{subsetresults}.
+Figure \ref{fig:results} shows the results over 1000 simulations. A subset of simulation conditions is presented in Table \ref{tab:results}.
 
-\begin{figure}[h]
-  \resizebox{\textwidth}{!}{ %notice the \resizebox{} command
-        \includegraphics{Figures/8plotresults.pdf}
-  }      
-  \caption{Simulation and convergence diagnostics over 1000 MCMC simulations.}
-    \label{resultssim}
-\end{figure}
+<<echo = FALSE, eval = TRUE>>=
+load("C:/Users/User/Desktop/shinyMice/Simulation/Results/results14.Rdata.Rdata")
+names(dat) <- c("It.", "Bias", "CI width", "Cov. rate", "R chain mean", "R chain var", "AC chain mean", "AC chain var")
+@
+
+<<fig = TRUE, echo = FALSE>>=
+plot.ts(dat[c(5:8, 2:4)], main = "", xlab = "Number of iterations")
+@
+
+\textbf{Try to put the chain means and vars together in one panel}
+
+% \begin{figure}[h]
+%   \resizebox{\textwidth}{!}{ %notice the \resizebox{} command
+%         \includegraphics{Figures/results.pdf}
+%   }
+%   \caption{Simulation and convergence diagnostics over 1000 MCMC simulations.}
+%     \label{resultssim}
+% \end{figure}
 
@@
 
-%% --------------------------------
-\subsection{Simulation diagnostics}
-Bias.
-After 1 or 2 iterations bias fluctuates within a narrow range of the average bias across all simulation conditions.
-Average bias is not zero because there is only one incomplete dataset to be imputed.
-Bias is a sample effect.
-
-Empirical standard error (SE).
-
-Confidence interval width. Does not show a trend across the simulation conditions. [Is related to empirical SE??]
-
-Coverage rate. Is more or less stable from two iterations upwards. But higher percentages then the expected 95\% \cite{neym43}. This is due to pooled SE??
+% latex table generated in R 3.6.1 by xtable 1.8-4 package
+% Wed Dec 04 18:16:11 2019
+\begin{table}[ht]
+\centering
+\caption{Simulation and convergence diagnostics over 1000 MCMC simulations.} 
+\label{tab:results}
+\begin{tabular}{lrrrrrrr}
+  \hline
+It. & Bias & CI width & Cov. rate & $\widehat{R}_{mean}$ & $\widehat{R}_{var}$ & $AC_{mean}$ & $AC_{var}$ \\ 
+  \hline
+   1 & -0.137 & 0.954 & 0.932 & NA & NA & NA & NA \\ 
+     2 & -0.006 & 0.932 & 0.953 & 1.650 & 1.632 & -0.500 & -0.500 \\ 
+     3 & 0.002 & 0.929 & 0.944 & 1.314 & 1.306 & -0.660 & -0.659 \\ 
+     4 & 0.003 & 0.933 & 0.957 & 1.461 & 1.457 & -0.733 & -0.735 \\ 
+     5 & 0.004 & 0.935 & 0.954 & 1.475 & 1.472 & -0.705 & -0.706 \\ 
+     6 & 0.001 & 0.934 & 0.956 & 1.258 & 1.256 & -0.656 & -0.646 \\ 
+     7 & 0.002 & 0.930 & 0.948 & 1.265 & 1.269 & -0.591 & -0.585 \\ 
+     8 & 0.004 & 0.930 & 0.954 & 1.181 & 1.178 & -0.495 & -0.516 \\ 
+     9 & 0.003 & 0.931 & 0.956 & 1.187 & 1.187 & -0.442 & -0.459 \\ 
+    10 & 0.003 & 0.952 & 0.943 & 1.141 & 1.140 & -0.403 & -0.423 \\ 
+    15 & 0.002 & 0.942 & 0.965 & 1.100 & 1.100 & -0.276 & -0.289 \\ 
+    25 & 0.005 & 0.934 & 0.955 & 1.057 & 1.057 & -0.143 & -0.159 \\ 
+    50 & -0.002 & 0.929 & 0.959 & 1.027 & 1.026 & -0.053 & -0.075 \\ 
+   100 & 0.000 & 0.920 & 0.946 & 1.013 & 1.013 & 0.022 & -0.017 \\ 
+   \hline
+\end{tabular}
+\end{table}
 
@@
 
-$\widehat{R}$.
-Steep drop down up-to iteration 20-30, then stable.
-Conventional threshold $\widehat{R} < 1.2$ reached after three iterations, $\widehat{R} < 1.1$ after four iterations, $\widehat{R} < 1.01$ after 38 iterations (see online appendix).
+\textbf{$\widehat{R}$}.
+Steep drop down up-to iteration twenty, then more gradual decrease upto iteration fourty, more or less stable after that.
+Conventional threshold $\widehat{R} < 1.2$ reached after seven or eight iterations, $\widehat{R} < 1.1$ after fourteen iterations for chain means and sixteen for chain variances, $\widehat{R} < 1.01$ not reached within the 100 iterations (see online appendix for full table of results).
 
-Auto-correlation.
-Auto-correlation is dangerous when positive.
-These auto-correlations are negative.
-Still, we want the iterations to be stable and independent.
-Default maxit is five iterations now, should this be different?
-Are the 'waves' most pronounced at iteration 5?
-But why the dip??
+\textbf{Auto-correlation.} Within the 100 iterations considered in this study, AC decreases steeply, then slowly increases.
+Lowest AC observed is \Sexpr{round(min(dat[-1, c("AC chain mean", "AC chain var")]), 3)} at four iterations.
+There appears to be a trend towards zero from four iterations upwards.
+Auto-correlation does not plateau completely.
+The average absolute maximum ACs are negative up to 77 iterations for chain means, and for chain variances all iterations considered here have negative ACs.
 
-% latex table generated in R 3.6.1 by xtable 1.8-4 package
-% Fri Nov 29 11:38:27 2019
-\begin{table}[ht]
-\centering
-\caption{Simulation and convergence diagnostics over 1000 MCMC simulations.} 
-\label{results}
-\begin{tabular}{lrrrrrrr}
-  \hline
-It. & Bias & CI width & Cov. rate & $\widehat{R}$ & $\widehat{R}$ & Auto-corr. & Auto-corr. \\ 
-  \hline
-   1 & -0.128 & 0.922 & 0.930 & NA & NA & NA & NA \\ 
-     2 & -0.002 & 0.915 & 0.940 & 1.686 & 1.709 & -0.500 & -0.500 \\ 
-     3 & 0.012 & 0.923 & 0.960 & 1.317 & 1.326 & -0.661 & -0.659 \\ 
-     4 & 0.019 & 0.892 & 0.970 & 1.461 & 1.464 & -0.735 & -0.746 \\ 
-     5 & 0.015 & 0.906 & 0.910 & 1.486 & 1.472 & -0.706 & -0.705 \\ 
-     6 & 0.010 & 0.928 & 0.950 & 1.252 & 1.261 & -0.651 & -0.667 \\ 
-     7 & 0.006 & 0.908 & 0.910 & 1.259 & 1.267 & -0.619 & -0.584 \\ 
-     8 & 0.018 & 0.920 & 0.940 & 1.178 & 1.181 & -0.516 & -0.549 \\ 
-     9 & 0.011 & 0.994 & 0.940 & 1.190 & 1.186 & -0.506 & -0.488 \\ 
-    10 & 0.020 & 0.940 & 0.900 & 1.142 & 1.145 & -0.448 & -0.420 \\ 
-    15 & 0.009 & 0.917 & 0.960 & 1.098 & 1.109 & -0.256 & -0.289 \\ 
-    20 & 0.020 & 0.848 & 0.930 & 1.069 & 1.065 & -0.179 & -0.214 \\ 
-    25 & 0.026 & 0.897 & 0.970 & 1.054 & 1.054 & -0.146 & -0.131 \\ 
-    50 & 0.008 & 0.946 & 0.970 & 1.026 & 1.026 & -0.029 & -0.096 \\ 
-   100 & 0.011 & 0.967 & 0.930 & 1.013 & 1.013 & 0.054 & -0.020 \\ 
-   \hline
-\end{tabular}
-\end{table}
-
-From the simulation diagnostics it appears that as little as three iterations is sufficient to draw valid inference.
-Convergence diagnostics $\widehat{R}$ and auto-correlation, however, indicate 20-30 iterations.
-Apparently, they are not perfect for MI data [move interpretation to Discussion!].
-
-[Add simulation time (computational cost) to table?
+% \textbf{[Add simulation time (computational cost) to table?
 Add empirical SE of all diagnostics?
 Sample some random simulations and check distribution and plausibility of imputations?
-Use chain variance as scalar of interest for rhat and autocorrelation?]
+Use chain variance as scalar of interest for rhat and autocorrelation?]}
 
-%% -- Manuscript ---------------------------------------------------------------
+%% --------------------------------
+\subsection{Simulation diagnostics}
+\textbf{Bias.} From two iterations upwards, average bias across repititions is stable. Across iterations two to one hundred, bias fluctuates within a narrow range around zero. 
+% Average bias is not zero because there is only one incomplete dataset to be imputed. Bias is a sample effect.
 
-%% -- Illustrations ------------------------------------------------------------
+\textbf{Confidence interval width.} CIW does not show a clear trend across iterations.
+One might argue that there is some downward trending upto fourty iterations.
+%After about twenty-five iterations the average CIW across all simulation conditions and repetitions is reached \Sexpr{round(mean(dat[, "CI width"]), 3)}.
+
+\textbf{Coverage rate.} Coverage rate is more or less stable from two iterations upwards.
+On average, the coverage rate is somewhat higher then the expected nominal coverage of 95\% \cite{neym43}, namely \Sexpr{round(mean(dat[, "Cov.
+rate"]), 3)}.
 
@@
 
-This note illustrates that conventional convergence diagnostics behave differently on MI data as compared to the output of 'regular' MCMC algorithms.
-The recommended threshold for $\widehat{R}$ is too stringent for MI data.
+\textbf{Summary.}
+This note illustrates that conventional convergence diagnostics behave differently on MI data than other MCMC methods like Gibbs samplers in Bayesian analyses.
+The most recent recommended threshold for $\widehat{R}$ ($\widehat{R} < 1.01$) is too stringent for MI data.
 
+From the simulation diagnostics it appears that as little as two iterations could be sufficient to draw valid inference.
+Convergence diagnostics $\widehat{R}$ and auto-correlation, however, indicate deviant behavior upto twenty or even fourty iterations.
+
+\textbf{R hat below 1.}
 Moreover, $\widehat{R}$ could theoretically not be smaller than one, yet it happened several times in this study (see online appendix XYZ).
@@
 
-Future research: We would like to have convergence measures for multivariable statistics (scalars?) of interest.
+\textbf{Negative ACs.} Auto-correlation is dangerous when positive.
+These auto-correlations are negative.
+Still, we want the iterations to be stable and independent.
+Default maxit is five iterations now, should this be different?
+Are the 'waves' most pronounced at iteration 5?
+But why the dip??
+
+Implication of dip in AC is that default maxit value of five iterations is the worst possible number of iterations.
+
+\textbf{Future research.} We would like to have convergence measures for multivariable statistics (scalars?) of interest.
 This is, however, dependent of the complete data model.
@@
 I could not find any resources to apply this method and it is outside the scope of this thesis to investigate how this approach could be implemented.
-
-[final par is too negative, explain scope in intro instead.]
+\textbf{[this paragraph is too negative, explain scope in intro instead.]}
 

### 2019-12-04 Simulation/Writeup/NoteConvergence.tex
Finally integrated comments, also created nicer graph and worked on better flow in introduction section.

@@
 
-Figure \ref{resultssim} shows the results over 1000 simulations. A subset of simulation conditions is presented in Table \ref{subsetresults}.
+Figure \ref{fig:results} shows the results over 1000 simulations. A subset of simulation conditions is presented in Table \ref{tab:results}.
 
-\begin{figure}[h]
-  \resizebox{\textwidth}{!}{ %notice the \resizebox{} command
-        \includegraphics{Figures/8plotresults.pdf}
-  }      
-  \caption{Simulation and convergence diagnostics over 1000 MCMC simulations.}
-    \label{resultssim}
-\end{figure}
+\includegraphics{NoteConvergence-003}
+
+\textbf{Try to put the chain means and vars together in one panel}
+
+% \begin{figure}[h]
+%   \resizebox{\textwidth}{!}{ %notice the \resizebox{} command
+%         \includegraphics{Figures/results.pdf}
+%   }
+%   \caption{Simulation and convergence diagnostics over 1000 MCMC simulations.}
+%     \label{resultssim}
+% \end{figure}
 
@@
 
-$\widehat{R}$.
-Steep drop down up-to iteration 20-30, then stable.
-Conventional threshold $\widehat{R} < 1.2$ reached after three iterations, $\widehat{R} < 1.1$ after four iterations, $\widehat{R} < 1.01$ after 38 iterations (see online appendix).
+\textbf{$\widehat{R}$}.
+Steep drop down up-to iteration twenty, then more gradual decrease upto iteration fourty, more or less stable after that.
+Conventional threshold $\widehat{R} < 1.2$ reached after seven or eight iterations, $\widehat{R} < 1.1$ after fourteen iterations for chain means and sixteen for chain variances, $\widehat{R} < 1.01$ not reached within the 100 iterations (see online appendix for full table of results).
 
-Auto-correlation.
-Auto-correlation is dangerous when positive.
-These auto-correlations are negative.
-Still, we want the iterations to be stable and independent.
-Default maxit is five iterations now, should this be different?
-Are the 'waves' most pronounced at iteration 5?
-But why the dip??
+\textbf{Auto-correlation.} Within the 100 iterations considered in this study, AC decreases steeply, then slowly increases.
+Lowest AC observed is -0.735 at four iterations.
+There appears to be a trend towards zero from four iterations upwards.
+Auto-correlation does not plateau completely.
+The average absolute maximum ACs are negative up to 77 iterations for chain means, and for chain variances all iterations considered here have negative ACs.
 
-% latex table generated in R 3.6.1 by xtable 1.8-4 package
-% Fri Nov 29 11:38:27 2019
-\begin{table}[ht]
-\centering
-\caption{Simulation and convergence diagnostics over 1000 MCMC simulations.} 
-\label{results}
-\begin{tabular}{lrrrrrrr}
-  \hline
-It. & Bias & CI width & Cov. rate & $\widehat{R}$ & $\widehat{R}$ & Auto-corr. & Auto-corr. \\ 
-  \hline
-   1 & -0.128 & 0.922 & 0.930 & NA & NA & NA & NA \\ 
-     2 & -0.002 & 0.915 & 0.940 & 1.686 & 1.709 & -0.500 & -0.500 \\ 
-     3 & 0.012 & 0.923 & 0.960 & 1.317 & 1.326 & -0.661 & -0.659 \\ 
-     4 & 0.019 & 0.892 & 0.970 & 1.461 & 1.464 & -0.735 & -0.746 \\ 
-     5 & 0.015 & 0.906 & 0.910 & 1.486 & 1.472 & -0.706 & -0.705 \\ 
-     6 & 0.010 & 0.928 & 0.950 & 1.252 & 1.261 & -0.651 & -0.667 \\ 
-     7 & 0.006 & 0.908 & 0.910 & 1.259 & 1.267 & -0.619 & -0.584 \\ 
-     8 & 0.018 & 0.920 & 0.940 & 1.178 & 1.181 & -0.516 & -0.549 \\ 
-     9 & 0.011 & 0.994 & 0.940 & 1.190 & 1.186 & -0.506 & -0.488 \\ 
-    10 & 0.020 & 0.940 & 0.900 & 1.142 & 1.145 & -0.448 & -0.420 \\ 
-    15 & 0.009 & 0.917 & 0.960 & 1.098 & 1.109 & -0.256 & -0.289 \\ 
-    20 & 0.020 & 0.848 & 0.930 & 1.069 & 1.065 & -0.179 & -0.214 \\ 
-    25 & 0.026 & 0.897 & 0.970 & 1.054 & 1.054 & -0.146 & -0.131 \\ 
-    50 & 0.008 & 0.946 & 0.970 & 1.026 & 1.026 & -0.029 & -0.096 \\ 
-   100 & 0.011 & 0.967 & 0.930 & 1.013 & 1.013 & 0.054 & -0.020 \\ 
-   \hline
-\end{tabular}
-\end{table}
-
-From the simulation diagnostics it appears that as little as three iterations is sufficient to draw valid inference.
-Convergence diagnostics $\widehat{R}$ and auto-correlation, however, indicate 20-30 iterations.
-Apparently, they are not perfect for MI data [move interpretation to Discussion!].
-
-[Add simulation time (computational cost) to table?
+% \textbf{[Add simulation time (computational cost) to table?
 Add empirical SE of all diagnostics?
 Sample some random simulations and check distribution and plausibility of imputations?
-Use chain variance as scalar of interest for rhat and autocorrelation?]
+Use chain variance as scalar of interest for rhat and autocorrelation?]}
 
-%% -- Manuscript ---------------------------------------------------------------
+%% --------------------------------
+\subsection{Simulation diagnostics}
+\textbf{Bias.} From two iterations upwards, average bias across repititions is stable. Across iterations two to one hundred, bias fluctuates within a narrow range around zero. 
+% Average bias is not zero because there is only one incomplete dataset to be imputed. Bias is a sample effect.
 
-%% -- Illustrations ------------------------------------------------------------
+\textbf{Confidence interval width.} CIW does not show a clear trend across iterations.
+One might argue that there is some downward trending upto fourty iterations.
+%After about twenty-five iterations the average CIW across all simulation conditions and repetitions is reached 0.928.
+
+\textbf{Coverage rate.} Coverage rate is more or less stable from two iterations upwards.
+On average, the coverage rate is somewhat higher then the expected nominal coverage of 95\% \cite{neym43}, namely 0.953.
 

### 2019-12-04 Simulation/2.CreateTable.R

@@ -1,6 +1,6 @@
 # create table to display results
 # requires the packages "dplyr", "xtable
-# and the object 'out' created with '1. Excecute'
+# and the object 'results' created with '1. Excecute'
 
 # load package
 library("dplyr")
@@ -9,12 +9,12 @@ library("xtable")
 # create table with only specific rows
 tab <- results %>% 
   as.tbl() %>% 
-  .[c(1:10, 15, 20, 25, 50, 100),] %>% 
-  xtable(caption = "Simulation and convergence diagnostics over 1000 MCMC simulations.", label = "results", digits = 3)
+  .[c(1:10, 15, 25, 50, 100),] %>% 
+  xtable(caption = "Simulation and convergence diagnostics over 1000 MCMC simulations.", label = "tab:results", digits = 3)
 
 # adjust table attributes
+names(tab) <- c("It.", "Bias", "CI width", "Cov. rate",  "$\\widehat{R}_{mean}$", "$\\widehat{R}_{var}$", "$AC_{mean}$", "$AC_{var}$")
 attr(tab, "align")[1] <- "l"
-names(tab) <- c("It.", "Bias", "CI width", "Cov. rate",  "$\\widehat{R}$", "$\\widehat{R}$", "Auto-corr.", "Auto-corr.")
 
 # # add footnote
 # comment <- list(pos = list(0), command = NULL)

### 2019-12-05 Simulation/Writeup/NoteConvergence.Rnw
Added new graphs to manuscript (badly)
some spelling errors

@@
 At some point, any scientist conducting statistical analyses will run into a missing data problem \citep{alli02}.
-Missingness is problematic because statistical inference cannot be performed on incomplete data without employing \emph{ad hoc} solutions (e.g., listwise deletion), which can yield wildly invalid results \citep{buur18}.
+Missingness is problematic because statistical inference cannot be performed on incomplete data without employing \emph{ad hoc} solutions (e.g., list-wise deletion), which can yield wildly invalid results \citep{buur18}.
 A popular answer to the ubiquitous problem of missing information is to use the framework of multiple imputation (MI), proposed by \cite{rubin87}.
@@
 imputed) several times.
-The variablity between imputations is used to reflect how much uncertainty in the inference is introduced by the missingness.
+The variability between imputations is used to reflect how much uncertainty in the inference is introduced by the missingness.
 Therefore, MI can provide valid inferences despite of missing information.
@@
 %Therefore, I will answer the following question: 'Which measures are vital for evaluating the validity of multiply imputed data?'.
-The aim of this note is to provide guidelines for the evaulation of imputations to applied researchers who use MI methods.
-More specifically, this note adresses the question: `How to diagnose convergence of the MI algorithm implemented in \pkg{mice}?'.
+The aim of this note is to provide guidelines for the evaluation of imputations to applied researchers and statisticians who use MI methods.
+More specifically, this note addresses the question: `How to diagnose convergence of the MI algorithm implemented in \pkg{mice}?'.
 
@@
 The relation between $R$, $Y_{obs}$, and $Y_{mis}$ determines the missingness mechanism.
-This paper only considers a 'missing completely at random' (MCAR) mechanism, where the probability to be missing is equal for all $n \times p$ cells in $Y$.
+This paper only considers a 'missing completely at random' (MCAR) mechanism, where the probability of being missing is equal for all $n \times p$ cells in $Y$.
 
-In this note, the terms `unobserved' and `missing' data are used interchangibly.
+In this note, the terms `unobserved' and `missing' data are used interchangeably.
 Both refer to $Y_{mis}$, e.g.
@@
 \textbf{Explain the measures that I'm investigating here. R hat is ... Auto-correlation is... How are they computed? What do they say? How are they computed on MI data? Looking at imputation chains. 
-Explain that there is no baseline to compare the measures with. Only current convergence check is to visually inspect trace plots. (But in the traceplots convergence seems immediate).}
+Explain that there is no baseline to compare the measures with. Only current convergence check is to visually inspect trace plots. (But in the trace-plots convergence seems immediate).}
 
@@
 
-Autocorrelation (AC) is computed within imputation chains, as the correlation between $\theta$ at the $i^{th}$ iteration and $\theta$ at the $i+1^{th}$ iteration.
+Auto-correlation (AC) is computed within imputation chains, as the correlation between $\theta$ at the $i^{th}$ iteration and $\theta$ at the $i+1^{th}$ iteration.
 For now, we only consider the chain mean as scalar of interest $\theta$.
@@
 
-Confidence interval width (CIW) is computed as the diffence between the lower and upper bound of the 95\% confidence interval (CI95\%).
+Confidence interval width (CIW) is computed as the difference between the lower and upper bound of the 95\% confidence interval (CI95\%).
 The CI95\% bounds are computed as the estimated regression coefficient plus or minus (respectively) the pooled SE across imputations times 2.66 (the quantile of a t distribution with m-1 degrees of freedom).
@@
 
-\textbf{Try to put the chain means and vars together in one panel}
+\textbf{Try to put the chain means and vars together in one panel, like \ref{fig:rhat} but better.}
 
-% \begin{figure}[h]
-%   \resizebox{\textwidth}{!}{ %notice the \resizebox{} command
-%         \includegraphics{Figures/results.pdf}
-%   }
-%   \caption{Simulation and convergence diagnostics over 1000 MCMC simulations.}
-%     \label{resultssim}
-% \end{figure}
+\begin{figure}[h]
+  \resizebox{\textwidth}{!}{ %notice the \resizebox{} command
+        \includegraphics{Figures/Rhat.pdf}
+  }
+  \caption{Simulation and convergence diagnostics over 1000 MCMC simulations.}
+    \label{fig:rhat}
+\end{figure}
+
+\begin{figure}[h]
+  \resizebox{\textwidth}{!}{ %notice the \resizebox{} command
+        \includegraphics{Figures/AC.pdf}
+  }
+  \caption{Simulation and convergence diagnostics over 1000 MCMC simulations.}
+    \label{fig:ac}
+\end{figure}
 
@@
 \textbf{$\widehat{R}$}.
-Steep drop down up-to iteration twenty, then more gradual decrease upto iteration fourty, more or less stable after that.
+Steep drop down up-to iteration twenty, then more gradual decrease up-to iteration forty, more or less stable after that.
 Conventional threshold $\widehat{R} < 1.2$ reached after seven or eight iterations, $\widehat{R} < 1.1$ after fourteen iterations for chain means and sixteen for chain variances, $\widehat{R} < 1.01$ not reached within the 100 iterations (see online appendix for full table of results).
@@
 \subsection{Simulation diagnostics}
-\textbf{Bias.} From two iterations upwards, average bias across repititions is stable. Across iterations two to one hundred, bias fluctuates within a narrow range around zero. 
+\textbf{Bias.} From two iterations upwards, average bias across repetitions is stable. Across iterations two to one hundred, bias fluctuates within a narrow range around zero. 
 % Average bias is not zero because there is only one incomplete dataset to be imputed. Bias is a sample effect.
@@
 \textbf{Confidence interval width.} CIW does not show a clear trend across iterations.
-One might argue that there is some downward trending upto fourty iterations.
+One might argue that there is some downward trending up-to forty iterations.
 %After about twenty-five iterations the average CIW across all simulation conditions and repetitions is reached \Sexpr{round(mean(dat[, "CI width"]), 3)}.
@@
 From the simulation diagnostics it appears that as little as two iterations could be sufficient to draw valid inference.
-Convergence diagnostics $\widehat{R}$ and auto-correlation, however, indicate deviant behavior upto twenty or even fourty iterations.
+Convergence diagnostics $\widehat{R}$ and auto-correlation, however, indicate deviant behavior up-to twenty or even forty iterations.
 

### 2019-12-05 Simulation/Writeup/NoteConvergence.tex
Added new graphs to manuscript (badly)
some spelling errors

@@
 \textbf{Confidence interval width.} CIW does not show a clear trend across iterations.
-One might argue that there is some downward trending upto fourty iterations.
+One might argue that there is some downward trending up-to forty iterations.
 %After about twenty-five iterations the average CIW across all simulation conditions and repetitions is reached 0.928.

### 2019-12-05 Simulation/3.CreateFigure.R

@@ -5,6 +5,8 @@
 # load package
 library("dplyr")
 library("ggplot2")
+library("ggpubr")
+
 
 # create table with only specific rows
 df <- results[-1,] 

### 2019-12-05 Simulation/Writeup/NoteConvergence.Rnw
Updated plots, and some more spell checking.

@@
 The variability between imputations is used to reflect how much uncertainty in the inference is introduced by the missingness.
-Therefore, MI can provide valid inferences despite of missing information.
+Therefore, MI can provide valid inferences despite missing information.
 
@@
 Since both of these situations are undesirable, imputations and their variability should be properly evaluated.
-Such evaluation measures are currently missing or under-developed in MI software, like the world leading \proglang{R} package \pkg{mice} \citep{mice}.
+Such evaluation measures are currently missing or under-developed in MI software, like the world-leading \proglang{R} package \pkg{mice} \citep{mice}.
 %Therefore, I will answer the following question: 'Which measures are vital for evaluating the validity of multiply imputed data?'.
@@
 iteration five are significantly correlated with the value of the first draw.
-Auto-correlation above the confidence level \textbf{[explain that this is a critical value for correlation based on alpha .05]} indicate dependence within chains.
+Auto-correlation above the confidence level \textbf{[explain that this is a critical value for correlation based on alpha .05]} indicates dependence within chains.
 Also, possible to use MC error, or Geweke statistic.
-But outside of scope of this study.
+But that is outside of the scope of this study.
 Here, the focus is mainly on $\widehat{R}$.
@@
 $\widehat{R}$ is then equal to one.
-Conventional threshold was $\widehat{R} < 1.2$ is acceptable.
+The conventional threshold was $\widehat{R} < 1.2$ is acceptable.
 Most recent guideline by \cite{veht19} is more stringent: $\widehat{R} < 1.01$.
@@
 
-Why is $\widehat{R}$ not applicable on MI data?
+Why is $\widehat{R}$ not an appropriate diagnostic to evaluate MI data?
 $\widehat{R}$ is not appropriate because it assumes over-dispersed initial values, which means that the initial values of the $m$ imputation chains are 'far away' from the distribution that the chains are converging to.
@@
 This suggests that $\widehat{R}$ would not be sensitive enough to flag non-convergence of MI algorithms.
-Empirical finding, however, show that the opposite may be true: \cite{lace07} report $\widehat{R}$ values above the threshold of $\widehat{R} < 1.1$ after fifty iterations.
+An empirical finding, however, shows that the opposite may be true: \cite{lace07} report $\widehat{R}$ values above the threshold of $\widehat{R} < 1.1$ after fifty iterations.
 %that $\widehat{R}$ will not be smaller than 1.1 before iteration number 50.
@@
 
-Hypothesis of this simulation study is that $\widehat{R}$ will over-estimate non-convergence.
+The hypothesis of this simulation study is that $\widehat{R}$ will over-estimate non-convergence.
 \textbf{[Or use:] Hypothesis based on \cite{lace07} is that the conventional acceptable level of $\widehat{R}$ is too strict for MI data.
@@
 \subsection{Amputation}
-After data generation, the complete data is 'amputed'.
+The complete data is the starting point for all simulations.
+In each repetition of the simulation study, the complete data is 'amputed' once.
 That is, the \pkg{mice} function \fct{ampute} in \proglang{R} is used to impose an MCAR missingness mechanism upon the data.
@@
 This leaves 20\% of the rows completely observed.
-%"Furthermore, we generate three kinds of missingness proportions: 0.1, 0.5, and 0.9.
+The resulting amputed data is equal for all simulation conditions in the same repetition.%"Furthermore, we generate three kinds of missingness proportions: 0.1, 0.5, and 0.9.
 Note that these values indicate the sampled proportion of incomplete cases in Y" \cite{schou18}.
-Table XYZ shows a summary of the generated complete data, and the amputed data. The amputed dataset is the starting point of each of the 1000 simulations per simulation condition.
 
@@
 Auto-correlation (AC) is computed within imputation chains, as the correlation between $\theta$ at the $i^{th}$ iteration and $\theta$ at the $i+1^{th}$ iteration.
-For now, we only consider the chain mean as scalar of interest $\theta$.
+For now, we only consider the chain means as scalar of interest $\theta$.
 The AC of the variable with the highest absolute AC value is reported per simulation.
@@
 
-Figure \ref{fig:results} shows the results over 1000 simulations. A subset of simulation conditions is presented in Table \ref{tab:results}.
-
 <<echo = FALSE, eval = TRUE>>=
@@
 
-<<fig = TRUE, echo = FALSE>>=
-plot.ts(dat[c(5:8, 2:4)], main = "", xlab = "Number of iterations")
-@
-
-\textbf{Try to put the chain means and vars together in one panel, like \ref{fig:rhat} but better.}
+Figures \ref{fig:conv} and \ref{fig:sim} shows the results over 1000 simulations. A subset of simulation conditions is presented in Table \ref{tab:results}.
 
@@
   \resizebox{\textwidth}{!}{ %notice the \resizebox{} command
-        \includegraphics{Figures/Rhat.pdf}
+        \includegraphics{Figures/convergence_diag.pdf}
   }
-  \caption{Simulation and convergence diagnostics over 1000 MCMC simulations.}
-    \label{fig:rhat}
+  \caption{Convergence diagnostics over 1000 MCMC simulations.}
+    \label{fig:conv}
 \end{figure}
@@
   \resizebox{\textwidth}{!}{ %notice the \resizebox{} command
-        \includegraphics{Figures/AC.pdf}
+        \includegraphics{Figures/Rplot.pdf}
   }
-  \caption{Simulation and convergence diagnostics over 1000 MCMC simulations.}
-    \label{fig:ac}
+  \caption{Simulation diagnostics over 1000 MCMC simulations.}
+    \label{fig:sim}
 \end{figure}
@@
 Steep drop down up-to iteration twenty, then more gradual decrease up-to iteration forty, more or less stable after that.
-Conventional threshold $\widehat{R} < 1.2$ reached after seven or eight iterations, $\widehat{R} < 1.1$ after fourteen iterations for chain means and sixteen for chain variances, $\widehat{R} < 1.01$ not reached within the 100 iterations (see online appendix for full table of results).
+Conventional threshold $\widehat{R} < 1.2$ reached after seven or eight iterations, $\widehat{R} < 1.1$ after fourteen iterations for chain means and sixteen for chain variances, $\widehat{R} < 1.01$ not reached within the 100 iterations (see online appendix for the full table of results).
 
@@
 \subsection{Simulation diagnostics}
-\textbf{Bias.} From two iterations upwards, average bias across repetitions is stable. Across iterations two to one hundred, bias fluctuates within a narrow range around zero. 
+\textbf{Bias.} From two iterations upwards, the average bias across repetitions is stable. Across iterations two to one hundred, bias fluctuates within a narrow range around zero. 
 % Average bias is not zero because there is only one incomplete dataset to be imputed. Bias is a sample effect.
@@
 
-\textbf{Coverage rate.} Coverage rate is more or less stable from two iterations upwards.
-On average, the coverage rate is somewhat higher then the expected nominal coverage of 95\% \cite{neym43}, namely \Sexpr{round(mean(dat[, "Cov.
+\textbf{Coverage rate.} The coverage rate is more or less stable from two iterations upwards.
+On average, the coverage rate is somewhat higher than the expected nominal coverage of 95\% \cite{neym43}, namely \Sexpr{round(mean(dat[, "Cov.
 rate"]), 3)}.
@@
 
-From the simulation diagnostics it appears that as little as two iterations could be sufficient to draw valid inference.
+From the simulation diagnostics, it appears that as little as two iterations could be sufficient to draw valid inferences.
 Convergence diagnostics $\widehat{R}$ and auto-correlation, however, indicate deviant behavior up-to twenty or even forty iterations.
@@
 How could $\widehat{R}$ smaller than 1 occur?
-The number of simulations is smaller than in `regular' MCMC processes [explain that less iterations is an advantage of MI, not a disadvantage compared to MCMC].
+The number of simulations is smaller than in `regular' MCMC processes [explain that fewer iterations is an advantage of MI, not a disadvantage compared to MCMC].
 Therefore, the `$(n-1/n)$' [add equation number] correction factor can influence the estimated potential scale reduction factor.
@@
 
-Implication of dip in AC is that default maxit value of five iterations is the worst possible number of iterations.
+An implication of the observed dip in AC is that default maxit value of five iterations is the worst possible number of iterations.
 
 \textbf{Future research.} We would like to have convergence measures for multivariable statistics (scalars?) of interest.
-This is, however, dependent of the complete data model.
+This is, however, dependent on the complete data model.
 The eigenvector decomposition method proposed by McKay (?) should be implemented.
@@
 This paper is written by the sole author (Hanne Oberman, BSc.), with guidance from Master thesis supervisors prof. dr. Stef van Buuren, and dr. Gerko Vink.
+
+\textbf{LEAVING THIS HERE FOR REFERENCE, PLEASE REMOVE BEFORE HANDING IN}
+
+<<fig = TRUE, echo = FALSE>>=
+plot.ts(dat[c(5:8, 2:4)], main = "", xlab = "Number of iterations")
+@
 

### 2019-12-05 Simulation/Writeup/NoteConvergence.tex
Updated plots, and some more spell checking.

@@
 
-Figure \ref{fig:results} shows the results over 1000 simulations. A subset of simulation conditions is presented in Table \ref{tab:results}.
-
-\includegraphics{NoteConvergence-003}
-
-\textbf{Try to put the chain means and vars together in one panel, like \ref{fig:rhat} but better.}
+Figures \ref{fig:conv} and \ref{fig:sim} shows the results over 1000 simulations. A subset of simulation conditions is presented in Table \ref{tab:results}.
 
@@
 
-\textbf{Coverage rate.} Coverage rate is more or less stable from two iterations upwards.
-On average, the coverage rate is somewhat higher then the expected nominal coverage of 95\% \cite{neym43}, namely 0.953.
+\textbf{Coverage rate.} The coverage rate is more or less stable from two iterations upwards.
+On average, the coverage rate is somewhat higher than the expected nominal coverage of 95\% \cite{neym43}, namely 0.953.
 
@@
 This paper is written by the sole author (Hanne Oberman, BSc.), with guidance from Master thesis supervisors prof. dr. Stef van Buuren, and dr. Gerko Vink.
+
+\textbf{LEAVING THIS HERE FOR REFERENCE, PLEASE REMOVE BEFORE HANDING IN}
+
+\includegraphics{NoteConvergence-003}
 

### 2019-12-05 Simulation/1.Execute.R

@@ -77,5 +77,5 @@ plot.ts(results[5:8], main = "Convergence Diagnostics", xlab = "Number of iterat
 ###
 
 # save for future reference
-save.Rdata(results, name = "results14.Rdata", path = "Simulation/Results")
-# save.image("environment.Rdata")
+save.Rdata(results, name = "results16.Rdata", path = "Simulation/Results")
+#save.image("environment.Rdata")

### 2019-12-06 Simulation/Writeup/NoteConvergence.Rnw
Update pdf to review
Feedback Gerko on Introduction.

@@
 At some point, any scientist conducting statistical analyses will run into a missing data problem \citep{alli02}.
-Missingness is problematic because statistical inference cannot be performed on incomplete data without employing \emph{ad hoc} solutions (e.g., list-wise deletion), which can yield wildly invalid results \citep{buur18}.
+Missingness is problematic because statistical inference cannot be performed on incomplete data without employing \emph{ad hoc} solutions (e.g., list-wise deletion), which may yield wildly invalid results \citep{buur18}.
 A popular answer to the ubiquitous problem of missing information is to use the framework of multiple imputation (MI), proposed by \cite{rubin87}.
-MI is an iterative algorithmic procedure in which each missing data point is `guessed' (i.e.
-imputed) several times.
+MI is an iterative algorithmic procedure in which each missing data point is `imputed' (i.e.
+filled in) several times.
 The variability between imputations is used to reflect how much uncertainty in the inference is introduced by the missingness.
@@
 
-\textbf{ADD some references in this paragraph!} To obtain valid inferences with MI, the variability between imputations should be correct.
+\textbf{ADD some references in this paragraph!} To obtain valid inferences with MI, the variability between imputations should be properly represented.
 If the variance between imputations is under-estimated, confidence intervals around estimates will be too narrow, which can yield spurious results.
-On the other hand, over-estimation of the variance between imputations results in unnecessarily wide confidence intervals, which can be costly because it requires more observations than strictly required.
+On the other hand, over-estimation of the variance between imputations results in unnecessarily wide confidence intervals, which can be costly because it lowers the statistical power.
 Since both of these situations are undesirable, imputations and their variability should be properly evaluated.
-Such evaluation measures are currently missing or under-developed in MI software, like the world-leading \proglang{R} package \pkg{mice} \citep{mice}.
+Such evaluation measures are currently missing or under-developed in MI software, like the world-leading \pkg{mice} package \citep{mice} in \proglang{R} \citep{R}.
 %Therefore, I will answer the following question: 'Which measures are vital for evaluating the validity of multiply imputed data?'.
-The aim of this note is to provide guidelines for the evaluation of imputations to applied researchers and statisticians who use MI methods.
-More specifically, this note addresses the question: `How to diagnose convergence of the MI algorithm implemented in \pkg{mice}?'.
+The aim of this research project is to investigate the convergence properties of the MI algorithm that is implemented in \pkg{mice}: `Multiple Imputation using Chained Equations' (MICE).
+This note addresses the question: `How to diagnose convergence of the multiple imputation algorithm implemented in \pkg{mice}?'.
 
-The convergence properties of \pkg{mice} are investigated by means of model-based simulation in \proglang{R} \citep{R}.
-All programming code used in this note is available on Github repository \href{https://github.com/gerkovink/shinyMice}{`ShinyMICE'}.
-The results of this simulation study are guidelines for applied researchers to evaluate convergence of MI algorithms.
+The convergence properties of the MICE algorithm are investigated by means of model-based simulation in \proglang{R} \citep{R}.
+The results of this simulation study are guidelines for applied researchers to evaluate convergence of MI methods.
 These guidelines will be implemented in the interactive evaluation tool for \pkg{mice}, `ShinyMICE', which is currently under development.
+All programming code used in this note is available from \href{https://github.com/gerkovink/shinyMice/simulation}{github.com/gerkovink/ShinyMICE/simulation}.
 %I will replicate Lacerda et al.'s simulation study on $\widehat{R}$ \citep{lace07}, and develop novel guidelines for assessing convergence.
@@
 This note therefore follows notation and conventions of \cite{mice}.
-Deviations from the 'original' notation by \cite{rubin87}, are described in \citep[\S~XYZ]{buur18}.
+Deviations from the `original' notation by \cite{rubin87} are described in \citep[\S~XYZ]{buur18}.
 
@@
 The relation between $R$, $Y_{obs}$, and $Y_{mis}$ determines the missingness mechanism.
-This paper only considers a 'missing completely at random' (MCAR) mechanism, where the probability of being missing is equal for all $n \times p$ cells in $Y$.
+This note only considers a `missing completely at random' (MCAR) mechanism, where the probability of being missing is equal for all $n \times p$ cells in $Y$.
 
@@
 Both refer to $Y_{mis}$, e.g.
-`$NA$' values in \proglang{R} or `999' in \proglang{SPSS}.
-The terms 'incomplete' or 'observed' data denote $Y_{obs}$.
+`\texttt{NA}' values in \proglang{R} or `\texttt{999}' in \proglang{SPSS}.
+The terms `incomplete' or `observed' data denote $Y_{obs}$.
 Incomplete data is the starting point of the multiple imputation procedure.
@@
 \node[main] (mids) [right=of data] { };
-\node[main] (mids2) [above=of mids] { };
-\node[main] (mids3) [below=of mids,label=below:Imputed data] { };
+\node[main] (mids2) [above=10mm of mids] { };
+\node[main] (mids3) [below=10mm of mids,label=below:Imputed data] { };
 \node[main] (mira) [right=of mids] {};
-\node[main] (mira2) [above=of mira] { };
-\node[main] (mira3) [below=of mira,label=below:Analysis results] { };
+\node[main] (mira2) [above=10mm of mira] { };
+\node[main] (mira3) [below=10mm of mira,label=below:Analysis results] { };
 \node[main, fill = white!100] (mipo) [right=of mira,label=right:Pooled result] { };
@@
 \end{tikzpicture}}
-\caption{Scheme of main steps in multiple imputation (Adapted from \cite[\S~1.4.1]{buur18})}
+\caption{Scheme of the main steps in multiple imputation ($m = 3$). Adapted from \citep[\S~1.4.1]{buur18}.}
 \end{figure}
@@
 
-This note is primarily concerned with the algorithmic properties of the imputation step.
+This note focusses on the algorithmic properties of the imputation step.
 The algorithm employed within this step has an iterative nature.
-That is, before drawing $m$ imputed values for each missing data point in $Y_{mis}$, a `chain` of potential values is considered.
+That is, before drawing $m$ imputed values for each missing data point in $Y_{mis}$, a `chain' of potential values is considered.
 Each of the $m$ chains starts with an initial value, drawn randomly from $Y_{obs}$.
@@
 Only the ultimate value that a chain lands on is imputed, and hence used in the next steps (analysis and pooling).
-The collection of values (or summary statistics) across iterations of one of the $m$ imputations will be referred to as `imputation chain'.
-
-\textbf{ADD THIS? The process of deliberately creating missingness in an otherwise complete data set is referred to as `amputing' the data.}
-
-\textbf{[explain difference MI, FCS, mice]}
-
-\textbf{[Explain `simulation condition' (the number of iterations the algorithm has before imputing), ...?]}
-
-% \begin{figure}[h]
-%   \resizebox{\textwidth}{!}{ %notice the \resizebox{} command
-%         \includegraphics{steps.png}
-%   }
-%   \caption{Scheme of main steps in multiple imputation (Adapted from \cite[par.~1.4.1]{buur18})}
-%     \label{resultssim}
-% \end{figure}
+The collection of values (or summary statistics) across iterations of each of the $m$ imputations will be referred to as an `imputation chain'.
 
 % "Let Q denote the quantity of scientific interest (e.g., a regression coefficient)." \cite[p.~4]{mice}.
-
 % "Let Y denote the n x p matrix containing the data values on p variables for all n units in the sample.
@@
 %% -- Theoretical Background ---------------------------------------------------------------
-\subsection{Theoretical Background (NOG NIET AF VANAF HIER)} \label{sec:background}
+\subsection{Theoretical Background} \label{sec:background}
+
+The method of multiple imputation The properties of multiple imputation procedures have been studied
 
@@
 But sometimes you can know if your chain has not converged, so we at least check for this latter possibility" \cite[p.~101]{hoff09}
-Convergence has two interpretations \textbf{[look up source!!!]}: mixing between chains and stability over iterations within chains.
-Ideally, we want the chains to intermingle nicely, without trends within chains.
+Convergence has two components \textbf{[look up source!!!]}: mixing between chains and stability over iterations within chains.
+Ideally, both components will be satisfied: there are no trends within chains, and the chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm.
 Diagnose non-convergence in the mixing sense: $\widehat{R}$.
-Diagnose non-convergence in the stability sense: auto-correlation.
+Diagnose non-convergence in the stability sense: auto-correlation (AC).
+Also, it is possible to compute MC errors, or the Geweke statistic \citep{XYZ}.
+All of these methods evaluate the convergence of univariate scalars of interest (e.g., chain mean or chain variance).
+The convergence of multivariable statistics (i.e.
+relations between such scalars) may be evaluated via the eigenvector decomposition method proposed by McKay \citep{XYZ}.
+These three methods, however, are outside of the scope of this study.
+Here, the focus is mainly on $\widehat{R}$.
 
@@
 
-In the latter interpretation of convergence, convergence is interpreted as stability over iterations, or non-recurring.
-Non-recurrence can be evaluated with auto-correlation.
-Auto-correlation shows how dependent subsequent draws of an imputation chain are on the previous value.
-If there is a lot of dependence, draws at e.g.
-iteration five are significantly correlated with the value of the first draw.
-Auto-correlation above the confidence level \textbf{[explain that this is a critical value for correlation based on alpha .05]} indicates dependence within chains.
-Also, possible to use MC error, or Geweke statistic.
-But that is outside of the scope of this study.
-Here, the focus is mainly on $\widehat{R}$.
-
-The potential scale reduction factor $\widehat{R}$ tells us how much variance of an estimate could be shrunken down by running the chains infinitely long \cite{gelm92}.
-That then tells us something about how dependent the chains are on the starting values.
+The potential scale reduction factor $\widehat{R}$ tells us by how much the variance of an estimate could be shrunken down, if the chains were to be infinitely long \cite{gelm92}.
+That then informs us about how dependent the chains are on the starting values.
 If there is no dependence on the initial values anymore, the chains have converged (in the mixing sense of the word).
 $\widehat{R}$ is then equal to one.
-The conventional threshold was $\widehat{R} < 1.2$ is acceptable.
-Most recent guideline by \cite{veht19} is more stringent: $\widehat{R} < 1.01$.
+The conventional acceptable threshold was $\widehat{R} < 1.2$.
+More recently, \cite{veht19} proposed a more stringent threshold: $\widehat{R} < 1.01$.
 
 We can apply $\widehat{R}$ to the mean (or to the first two moments) of the variables of interest.
-$\widehat{R}$ is computed as follows: % The scalar summary of interest in this study is the average of the imputed values per variable.
+$\widehat{R}$ is defined by \cite{veth19} as follows: % The scalar summary of interest in this study is the average of the imputed values per variable.
 %As recommended by \cite{veht19} $\widehat{R}$ is computed as ....
@@
 
+High $\widehat{R}$ would indicate .... low $\widehat{R}$ values indicate ....
+
+The second component of convergence ...
+interpreted as stability over iterations, or non-recurring.
+Non-recurrence can be evaluated with auto-correlation.
+Auto-correlation shows how dependent subsequent draws of an imputation chain are on the previous value.
+If there is a lot of dependence, draws at e.g.
+iteration five are significantly correlated with the value of the first draw.
+A high auto-correlation indicates dependence within chains.
+The magnitude of the AC can be interpreted qualitatively, or compared to the critical value of a \textbf{XYZ} distribution with \textbf{XYZ} degrees of freedom.
+
 \textbf{Convergence of MI}
@@
 A fundamental assumption of MICE is convergence of the algorithm.
-MICE is a sort of Gibbs sampler, a type of Markov chain Monte Carlo (MCMC) algorithm.
+MICE is a type of Markov chain Monte Carlo (MCMC) algorithm.
 In general, the validity of inference resulting from MCMC algorithms is threatened by non-convergence.
@@
 
-Why is $\widehat{R}$ not an appropriate diagnostic to evaluate MI data?
-$\widehat{R}$ is not appropriate because it assumes over-dispersed initial values, which means that the initial values of the $m$ imputation chains are 'far away' from the distribution that the chains are converging to.
-With MI procedures, initial values are chosen randomly from the observed data.
-This means that one initial value is necessary for each imputation chain.
+$\widehat{R}$ may not be an appropriate diagnostic to evaluate MI data because it assumes over-dispersed initial values.
+This means that the initial values of the $m$ imputation chains are `far away' from the distribution that the chains are converging to.
+In the \pkg{mice} algorithm, initial values are chosen randomly from the observed data.
+Therefore, we cannot be certain that the initial values are over-dispersed.
+%The initial parameter value is noodzakelijk.
+Daarom is het lastig om in MI that starten vanuit overdisp state.
 
@@
 
-Without over-dispersed initial states, $\widehat{R}$ can falsely diagnose convergence \citep{broo98}.
+Without over-dispersed initial states, $\widehat{R}$ may falsely diagnose convergence \citep{broo98}.
 %would not be sensitive enough to flag non-convergence of MI algorithms: "if over-dispersion does not hold, $\sigma_+^2$ can be too low, which can lead to falsely diagnosing convergence." \cite[p~437]{broo98}.
@@
 
-In this study, 1000 simulations are performed for each of the 100 simulation conditions.
+To investigate convergence properties of the \pkg{mice} algorithm, different lenghts of imputation chains are considered.
+this simulation study considers 100 simulation conditions.
+Each condition has a different maximum number of iterations, varying between one and one-hundred iterations.
+In this study, 1000 simulation runs are performed for each of the 100 simulation conditions.
 The simulation set-up consists of several steps, summarized in the pseudo-code below.
@@
 The quantity of scientific interest is the estimated regression coefficient of predictor $X$ on outcome variable $Y$.
-The linear regression model is $Y \sim X+Z_1+Z_2$, where $Y$ is the dependent variable, $X$ is an independent variable, and $Z_1$ and $Z_2$ are covariates.
+The linear regression model is $Y \sim \beta_1 X + \beta_2 Z_1 + \beta_3 Z_2$, where $Y$ is the dependent variable, $X$ is an independent variable, and $Z_1$ and $Z_2$ are covariates.
 The data generating model of the predictors is a multivariate normal distribution with means structure $\mu$, and variance-covariance matrix $\Sigma$.
-Outcome variable $Y$ is deduced from the predictor variables as follows:
+Outcome variable $Y$ is deduced from the predictor variables as $Y =  2 \times X + 0.5 \times Z_1 - 1 \times Z_2 + \epsilon$.
 
@@
 \epsilon
-\end{pmatrix} &\sim  N
+\end{pmatrix} \sim  N
 \begin{bmatrix}
@@
 0
-\end{pmatrix}\!\!,&
+\end{pmatrix}\!\!,
 \begin{pmatrix}
@@
 \end{bmatrix}\\[2\jot]
-Y &=  2 \times X + 0.5 \times Z_1 - 1 \times Z_2 + \epsilon
 \end{align*}
@@
 
-$\widehat{R}$ is computed within imputation chains: for each variable, for each simulation condition, for each simulation.
+$\widehat{R}$ is computed within imputation chains: for each variable, each simulation condition, and each simulation.
 The maximum value across variables within the same simulation is reported.
@@
 
-Auto-correlation (AC) is computed within imputation chains, as the correlation between $\theta$ at the $i^{th}$ iteration and $\theta$ at the $i+1^{th}$ iteration.
-For now, we only consider the chain means as scalar of interest $\theta$.
+Auto-correlation (AC) is computed within imputation chains, as the correlation between $\theta$ at the $i^{th}$ iteration and $\theta$ at the $i+1^{th}$ iteration, where $\theta$ is a scalar summary of interest (i.e., chain means and chain variances).
 The AC of the variable with the highest absolute AC value is reported per simulation.
@@
 The simulation diagnostics are as recommended by \cite{buur18}.
-Namely, average bias, average confidence interval width, and empirical coverage rate (coverage probability) across simulations.
+These comprise of average bias, average confidence interval width, and empirical coverage rate (coverage probability) across simulations.
 %"We expect that in accordance with Neyman (1934), at least 95 percent of the confidence intervals should contain the true population value, although some room for simulation error should be taken into account to counteract the finite nature of our simulations"\cite{schou18}.
@@
 
-Figures \ref{fig:conv} and \ref{fig:sim} shows the results over 1000 simulations. A subset of simulation conditions is presented in Table \ref{tab:results}.
+Figures \ref{fig:conv} and \ref{fig:sim} display results of the simulation study aggregated over 1000 repetitions.
+A subset of simulation conditions is presented in Table \ref{tab:results}.
+This section describes convergence diagnostics ($\widehat{R}$ and auto-correlation) and simulation diagnostics (bias, confidence interval width and coverage rate) separately.
 
@@
   \resizebox{\textwidth}{!}{ %notice the \resizebox{} command
-        \includegraphics{Figures/Rplot.pdf}
+        \includegraphics{Figures/simulation_diag.pdf}
   }
@@
 \end{figure}
-
-Evaluate simulation diagnostics one by one: bias, empirical SE, confidence interval width, and coverage rate. Then evaluate convergence diagnostics $\widehat{R}$ and auto-correlation.
 
@@
 \textbf{$\widehat{R}$}.
-Steep drop down up-to iteration twenty, then more gradual decrease up-to iteration forty, more or less stable after that.
-Conventional threshold $\widehat{R} < 1.2$ reached after seven or eight iterations, $\widehat{R} < 1.1$ after fourteen iterations for chain means and sixteen for chain variances, $\widehat{R} < 1.01$ not reached within the 100 iterations (see online appendix for the full table of results).
+Steep decline up-to iteration twenty, then more gradual decrease up-to iteration forty, more or less stable after that.
+Conventional threshold $\widehat{R} < 1.2$ reached after seven or eight iterations, $\widehat{R} < 1.1$ after fourteen iterations for chain means and sixteen for chain variances, $\widehat{R} < 1.01$ not reached within the 100 iterations considered in this simulation study (see online appendix for the full table of results).
 
@@
 \textbf{Confidence interval width.} CIW does not show a clear trend across iterations.
-One might argue that there is some downward trending up-to forty iterations.
+One might argue that there is some downward trending up-to thirthy to forty iterations.
 %After about twenty-five iterations the average CIW across all simulation conditions and repetitions is reached \Sexpr{round(mean(dat[, "CI width"]), 3)}.
@@
 \textbf{Coverage rate.} The coverage rate is more or less stable from two iterations upwards.
-On average, the coverage rate is somewhat higher than the expected nominal coverage of 95\% \cite{neym43}, namely \Sexpr{round(mean(dat[, "Cov.
-rate"]), 3)}.
+On average, the coverage rate is somewhat higher than the expected nominal coverage of 95\% \cite{neym34}, namely \Sexpr{100*round(mean(dat[, "Cov.
+rate"]), 3)}\%.
+
+\textbf{add loess line to plots to see when it becomes a flat line.}
 

### 2019-12-06 Simulation/Writeup/NoteConvergence.tex
Update pdf to review
Feedback Gerko on Introduction.

@@
 
-\textbf{ADD some references in this paragraph!} To obtain valid inferences with MI, the variability between imputations should be correct.
+\textbf{ADD some references in this paragraph!} To obtain valid inferences with MI, the variability between imputations should be properly represented.
 If the variance between imputations is under-estimated, confidence intervals around estimates will be too narrow, which can yield spurious results.
-On the other hand, over-estimation of the variance between imputations results in unnecessarily wide confidence intervals, which can be costly because it requires more observations than strictly required.
+On the other hand, over-estimation of the variance between imputations results in unnecessarily wide confidence intervals, which can be costly because it lowers the statistical power.
 Since both of these situations are undesirable, imputations and their variability should be properly evaluated.
-Such evaluation measures are currently missing or under-developed in MI software, like the world-leading \proglang{R} package \pkg{mice} \citep{mice}.
+Such evaluation measures are currently missing or under-developed in MI software, like the world-leading \pkg{mice} package \citep{mice} in \proglang{R} \citep{R}.
 %Therefore, I will answer the following question: 'Which measures are vital for evaluating the validity of multiply imputed data?'.
@@
 The convergence properties of \pkg{mice} are investigated by means of model-based simulation in \proglang{R} \citep{R}.
-All programming code used in this note is available on Github repository \href{https://github.com/gerkovink/shinyMice}{`ShinyMICE'}.
 The results of this simulation study are guidelines for applied researchers to evaluate convergence of MI algorithms.
 These guidelines will be implemented in the interactive evaluation tool for \pkg{mice}, `ShinyMICE', which is currently under development.
+All programming code used in this note is available from \href{https://github.com/gerkovink/shinyMice/simulation}{github.com/gerkovink/ShinyMICE/simulation}.
 %I will replicate Lacerda et al.'s simulation study on $\widehat{R}$ \citep{lace07}, and develop novel guidelines for assessing convergence.
@@
 The relation between $R$, $Y_{obs}$, and $Y_{mis}$ determines the missingness mechanism.
-This paper only considers a 'missing completely at random' (MCAR) mechanism, where the probability of being missing is equal for all $n \times p$ cells in $Y$.
+This paper only considers a `missing completely at random' (MCAR) mechanism, where the probability of being missing is equal for all $n \times p$ cells in $Y$.
 
@@
 Only the ultimate value that a chain lands on is imputed, and hence used in the next steps (analysis and pooling).
-The collection of values (or summary statistics) across iterations of one of the $m$ imputations will be referred to as `imputation chain'.
-
-\textbf{ADD THIS? The process of deliberately creating missingness in an otherwise complete data set is referred to as `amputing' the data.}
-
-\textbf{[explain difference MI, FCS, mice]}
+The collection of values (or summary statistics) across iterations of each of the $m$ imputations will be referred to as an `imputation chain'.
 
@@
 
-% \begin{figure}[h]
-%   \resizebox{\textwidth}{!}{ %notice the \resizebox{} command
-%         \includegraphics{steps.png}
-%   }
-%   \caption{Scheme of main steps in multiple imputation (Adapted from \cite[par.~1.4.1]{buur18})}
-%     \label{resultssim}
-% \end{figure}
-
 % "Let Q denote the quantity of scientific interest (e.g., a regression coefficient)." \cite[p.~4]{mice}.
-
 % "Let Y denote the n x p matrix containing the data values on p variables for all n units in the sample.
@@
 %% -- Theoretical Background ---------------------------------------------------------------
-\subsection{Theoretical Background (NOG NIET AF VANAF HIER)} \label{sec:background}
+\subsection{Theoretical Background} \label{sec:background}
 
@@
 
+High R hat would indicate .... low R hat values indicate ....
+
+The second component of convergence ...
+interpreted as stability over iterations, or non-recurring.
+Non-recurrence can be evaluated with auto-correlation.
+Auto-correlation shows how dependent subsequent draws of an imputation chain are on the previous value.
+If there is a lot of dependence, draws at e.g.
+iteration five are significantly correlated with the value of the first draw.
+A high auto-correlation indicates dependence within chains.
+The magnitude of the AC can be interpreted qualitatively, or compared to the critical value of a \textbf{XYZ} distribution with \textbf{XYZ} degrees of freedom.
+
 \textbf{Convergence of MI}
@@
 \textbf{Confidence interval width.} CIW does not show a clear trend across iterations.
-One might argue that there is some downward trending up-to forty iterations.
+One might argue that there is some downward trending up-to thirthy to forty iterations.
 %After about twenty-five iterations the average CIW across all simulation conditions and repetitions is reached 0.928.
@@
 \textbf{Coverage rate.} The coverage rate is more or less stable from two iterations upwards.
-On average, the coverage rate is somewhat higher than the expected nominal coverage of 95\% \cite{neym43}, namely 0.953.
+On average, the coverage rate is somewhat higher than the expected nominal coverage of 95\% \cite{neym34}, namely 95.3\%.
+
+\textbf{add loess line to plots to see when it becomes a flat line.}
 

### 2019-12-06 Simulation/1.Execute.R

@@ -37,16 +37,17 @@ set.seed(1111)
 
 # simulate data once
 data <- data.simulation(n = populationsize, true_effect)
+# run ampute once to get patterns object (and )to be able to adjust it for multivariate missingness)
+amp_patterns <- ampute(data)$patterns
+# for multivariable missingness, uncomment this
+amp_patterns[1:4, 1] <- 0
 
 # combine separate functions into wrapper
-simulate <- function(data, n.iter, true_effect) {
+simulate <- function(data, n.iter, true_effect, amp_patterns) {
   pb <- txtProgressBar(min = 0, max = n.iter, style = 3)
   
   # remove values at random with 20 percent probability to be missing
-  ampdata <- ampute(data, prop = 0.8, mech = "MCAR")
-  patterns <- ampdata$patterns
-  patterns[1:4, 1] <- 0
-  ampdata <- ampute(data = data, patterns = patterns, prop = 0.8, mech = "MCAR")$amp
+  ampdata <- ampute(data, patterns = amp_patterns, prop = 0.8, mech = "MCAR")$amp
   
   # object for output
   res <- list()

### 2019-12-06 Simulation/Writeup/NoteConvergence.Rnw
pdf update

@@
 The convergence properties of the MICE algorithm are investigated by means of model-based simulation in \proglang{R} \citep{R}.
-The results of this simulation study are guidelines for applied researchers to evaluate convergence of MI methods.
-These guidelines will be implemented in the interactive evaluation tool for \pkg{mice}, `ShinyMICE', which is currently under development.
+The results of this simulation study are guidelines for assessing convergence of this MI method.
+These guidelines will be implemented in an interactive evaluation tool for \pkg{mice}, `ShinyMICE', which is currently under development.
 All programming code used in this note is available from \href{https://github.com/gerkovink/shinyMice/simulation}{github.com/gerkovink/ShinyMICE/simulation}.

### 2019-12-06 Simulation/Writeup/NoteConvergence.tex
pdf update

@@
 %Therefore, I will answer the following question: 'Which measures are vital for evaluating the validity of multiply imputed data?'.
-The aim of this note is to provide guidelines for the evaluation of imputations to applied researchers and statisticians who use MI methods.
-More specifically, this note addresses the question: `How to diagnose convergence of the MI algorithm implemented in \pkg{mice}?'.
+The aim of this research project is to investigate the convergence properties of the MI algorithm that is implemented in \pkg{mice}: `Multiple Imputation using Chained Equations' (MICE).
+This note addresses the question: `How to diagnose convergence of the multiple imputation algorithm implemented in \pkg{mice}?'.
 
-The convergence properties of \pkg{mice} are investigated by means of model-based simulation in \proglang{R} \citep{R}.
-The results of this simulation study are guidelines for applied researchers to evaluate convergence of MI algorithms.
-These guidelines will be implemented in the interactive evaluation tool for \pkg{mice}, `ShinyMICE', which is currently under development.
+The convergence properties of the MICE algorithm are investigated by means of model-based simulation in \proglang{R} \citep{R}.
+The results of this simulation study are guidelines for assessing convergence of this MI method.
+These guidelines will be implemented in an interactive evaluation tool for \pkg{mice}, `ShinyMICE', which is currently under development.
 All programming code used in this note is available from \href{https://github.com/gerkovink/shinyMice/simulation}{github.com/gerkovink/ShinyMICE/simulation}.
@@
 The relation between $R$, $Y_{obs}$, and $Y_{mis}$ determines the missingness mechanism.
-This paper only considers a `missing completely at random' (MCAR) mechanism, where the probability of being missing is equal for all $n \times p$ cells in $Y$.
+This note only considers a `missing completely at random' (MCAR) mechanism, where the probability of being missing is equal for all $n \times p$ cells in $Y$.
 
@@
 
-\textbf{[Explain `simulation condition' (the number of iterations the algorithm has before imputing), ...?]}
-
 % "Let Q denote the quantity of scientific interest (e.g., a regression coefficient)." \cite[p.~4]{mice}.
@@
 \subsection{Theoretical Background} \label{sec:background}
+
+The method of multiple imputation The properties of multiple imputation procedures have been studied
 
@@
 
-High R hat would indicate .... low R hat values indicate ....
+High $\widehat{R}$ would indicate .... low $\widehat{R}$ values indicate ....
 

### 2019-12-07 Simulation/Writeup/NoteConvergence.Rnw
fiest paragraph of theoretical background

@@
 
-The convergence properties of the MICE algorithm are investigated by means of model-based simulation in \proglang{R} \citep{R}.
-The results of this simulation study are guidelines for assessing convergence of this MI method.
+The convergence properties of MICE are investigated by means of model-based simulation in \proglang{R} \citep{R}.
+The results of this simulation study are guidelines for assessing convergence of the MI algorithm.
 These guidelines will be implemented in an interactive evaluation tool for \pkg{mice}, `ShinyMICE', which is currently under development.
@@
 
-The method of multiple imputation The properties of multiple imputation procedures have been studied
+The methodology for diagnosing convergence of multiple imputation algorithms has not been systematically studied \cite{buur18}.
+Yet, reference manuals of promiment MI software do mention the importance of evaluating algorithmic convergence (e.g., \cite[p.XYZ]{mice}, XYZ, XYZ).
+This section provides a brief review of how convergence is defined in the context of iterative algorithmic procedures, what diagnostic tools are available, and which of these may be applicable to MI algorithms.
 
@@
 
+The convergence properties of multiple imputation algorithms lack scientific consensus \citep{taka17}.
+A review by \citep{murr18}  and some default MICE techniques might not converge to stable distributions at all \citep{murr18}.
+Convergence diagnostics for MI methods have not been systematically studied \citep{buur18}.
+Moreover,
 A fundamental assumption of MICE is convergence of the algorithm.
@@
 Visually assessing convergence is not only difficult to the untrained eye, it might also be futile.
-The convergence properties of MI algorithms lack scientific consensus \citep{taka17}, and some default MICE techniques might not converge to stable distributions at all \citep{murr18}.
-Moreover, convergence diagnostics for MI methods have not been systematically studied \citep{buur18}.
 

### 2019-12-09 Simulation/Writeup/NoteConvergence.Rnw

@@
   Utrecht University\\
-  Heidelberglaan~15\\
-  3500 Utrecht, The Netherlands\\
+  Heidelberglaan~15, 3500 Utrecht, The Netherlands\\
   E-mail: \email{h.i.oberman@uu.nl} % \\
@@
 Refer to future section?]}.
-Therefore, empirical researchers have to rely on visual inspection procedures that are theoretically equivalent to $\widehat{R}$ \citep{whit11} \textbf{[explain why this is appropriate, and numerical is not]}.
-Visually assessing convergence is not only difficult to the untrained eye, it might also be futile.
+Therefore, empirical researchers have to rely on visual inspection procedures that are theoretically equivalent to $\widehat{R}$ \citep{whit11} \textbf{[explain why this can still be appropriate, and numerical is not]}.
+Visually assessing convergence is not only challenging to the untrained eye, it might also be futile.
 
@@
 The primary research interest is in determining whether $\widehat{R}$ is an appropriate convergence diagnostic, and if so, which level of stringency suits MI data.
+\textbf{Or if there is a discrepancy between the visual and numeric inspection.}
 
@@
 To investigate convergence properties of the \pkg{mice} algorithm, different lenghts of imputation chains are considered.
-this simulation study considers 100 simulation conditions.
+This simulation study considers 100 simulation conditions.
 Each condition has a different maximum number of iterations, varying between one and one-hundred iterations.
 In this study, 1000 simulation runs are performed for each of the 100 simulation conditions.
-The simulation set-up consists of several steps, summarized in the pseudo-code below.
+\textbf{where for one through one hundred the Rhat and ...
+are calcuated.} The simulation set-up consists of several steps, summarized in the pseudo-code below.
 The complete \proglang{R} script of the simulation study is available on Github.
@@
 The data generating model of the predictors is a multivariate normal distribution with means structure $\mu$, and variance-covariance matrix $\Sigma$.
-Outcome variable $Y$ is deduced from the predictor variables as $Y =  2 \times X + 0.5 \times Z_1 - 1 \times Z_2 + \epsilon$.
 
@@
 
+Outcome variable $Y$ is subsequently calculated as $Y =  2X + .5Z_1 - Z_2 + \epsilon$.
+
 \subsection{Amputation}
-The complete data is the starting point for all simulations.
-In each repetition of the simulation study, the complete data is 'amputed' once.
+The complete data is the point of origin for all simulations.
+In each repetition of the simulation study, the complete data is `amputed' once.
 That is, the \pkg{mice} function \fct{ampute} in \proglang{R} is used to impose an MCAR missingness mechanism upon the data.
@@
 %"We expect that in accordance with Neyman (1934), at least 95 percent of the confidence intervals should contain the true population value, although some room for simulation error should be taken into account to counteract the finite nature of our simulations"\cite{schou18}.
-We could also look at distributional characteristics, and plausibility of imputed values, see \cite{vinknd}.
+We could also look at distributional characteristics, and plausibility of imputed values, see \cite{vinknd} (n.d.).
 For now, this is outside of the scope of this study.
@@
 Conventional threshold $\widehat{R} < 1.2$ reached after seven or eight iterations, $\widehat{R} < 1.1$ after fourteen iterations for chain means and sixteen for chain variances, $\widehat{R} < 1.01$ not reached within the 100 iterations considered in this simulation study (see online appendix for the full table of results).
+\textbf{[REWRITE AS STORY, NOT JUST DATA.]}
 
-\textbf{Auto-correlation.} Within the 100 iterations considered in this study, AC decreases steeply, then slowly increases.
-Lowest AC observed is \Sexpr{round(min(dat[-1, c("AC chain mean", "AC chain var")]), 3)} at four iterations.
+\textbf{Auto-correlation.} Within the 100 iterations considered in this study, the AC decreases steeply, before a slow increase.
+The lowest AC observed is \Sexpr{round(min(dat[-1, c("AC chain mean", "AC chain var")]), 3)} at four iterations.
 There appears to be a trend towards zero from four iterations upwards.
@@
 
-\textbf{add loess line to plots to see when it becomes a flat line.}
+\textbf{add loess line to plots to see when it becomes a flat line. also, give some implications for these measures.}
 
@@
 This note illustrates that conventional convergence diagnostics behave differently on MI data than other MCMC methods like Gibbs samplers in Bayesian analyses.
-The most recent recommended threshold for $\widehat{R}$ ($\widehat{R} < 1.01$) is too stringent for MI data.
+The most recent recommended threshold for $\widehat{R}$ ($\widehat{R} < 1.01$) may be too stringent for MI data.
 
-From the simulation diagnostics, it appears that as little as two iterations could be sufficient to draw valid inferences.
-Convergence diagnostics $\widehat{R}$ and auto-correlation, however, indicate deviant behavior up-to twenty or even forty iterations.
+From the simulation diagnostics, it appears that as little as two iterations could be sufficient to draw \textbf{[valid inferences?]}.
+Convergence diagnostics $\widehat{R}$ and auto-correlation, however, indicate that convergence may only be reached after twenty or even forty iterations.
 
 \textbf{R hat below 1.}
-Moreover, $\widehat{R}$ could theoretically not be smaller than one, yet it happened several times in this study (see online appendix XYZ).
-How could $\widehat{R}$ smaller than 1 occur?
-The number of simulations is smaller than in `regular' MCMC processes [explain that fewer iterations is an advantage of MI, not a disadvantage compared to MCMC].
+$\widehat{R}$ could theoretically not be smaller than one, yet it occurred several times in this study (see online appendix XYZ).
+This can happen when the number of simulations is smaller than in `regular' MCMC processes [explain that fewer iterations is an advantage of MI, not a disadvantage compared to MCMC].
 Therefore, the `$(n-1/n)$' [add equation number] correction factor can influence the estimated potential scale reduction factor.
-This downwards bias is in the opposite direction than expected: ``The mixture-of-sequences variance, V, should stabilize as a function of n.
+This downwards bias is in the opposite direction than expected: ``The mixture-of-sequences variance, $V$, should stabilize as a function of $n$.
 (Before convergence, we expect $\sigma^2$ to decrease with n, only increasing if the sequences explore a new area of parameter space, which would imply that the original sequences were not overdispersed for the particular scalar summary being monitored.)'' \cite[p~438]{broo98}.
@@
 
-The results in this paper were obtained using \proglang{R}~\Sexpr{paste(R.Version()[6:7], collapse = ".")} with the \pkg{mice}~\Sexpr{paste(packageVersion("mice"), collapse = ".")} package.
+The results in this paper were obtained using \proglang{R}~\Sexpr{paste(R.Version()[6:7], collapse = ".")} \cite{R} with the \pkg{mice}~\Sexpr{paste(packageVersion("mice"), collapse = ".")} package \cite{mice}.
 \proglang{R} itself and all packages used are available from the Comprehensive \proglang{R} Archive Network (CRAN) at \url{https://CRAN.R-project.org/}.

### 2019-12-09 Simulation/Writeup/NoteConvergence.tex

@@
 Refer to future section?]}.
-Therefore, empirical researchers have to rely on visual inspection procedures that are theoretically equivalent to $\widehat{R}$ \citep{whit11} \textbf{[explain why this is appropriate, and numerical is not]}.
-Visually assessing convergence is not only difficult to the untrained eye, it might also be futile.
-The convergence properties of MI algorithms lack scientific consensus \citep{taka17}, and some default MICE techniques might not converge to stable distributions at all \citep{murr18}.
-Moreover, convergence diagnostics for MI methods have not been systematically studied \citep{buur18}.
+Therefore, empirical researchers have to rely on visual inspection procedures that are theoretically equivalent to $\widehat{R}$ \citep{whit11} \textbf{[explain why this can still be appropriate, and numerical is not]}.
+Visually assessing convergence is not only challenging to the untrained eye, it might also be futile.
 
@@
 Conventional threshold $\widehat{R} < 1.2$ reached after seven or eight iterations, $\widehat{R} < 1.1$ after fourteen iterations for chain means and sixteen for chain variances, $\widehat{R} < 1.01$ not reached within the 100 iterations considered in this simulation study (see online appendix for the full table of results).
+\textbf{[REWRITE AS STORY, NOT JUST DATA.]}
 
-\textbf{Auto-correlation.} Within the 100 iterations considered in this study, AC decreases steeply, then slowly increases.
-Lowest AC observed is -0.735 at four iterations.
+\textbf{Auto-correlation.} Within the 100 iterations considered in this study, the AC decreases steeply, before a slow increase.
+The lowest AC observed is -0.735 at four iterations.
 There appears to be a trend towards zero from four iterations upwards.
@@
 
-The results in this paper were obtained using \proglang{R}~3.6.1 with the \pkg{mice}~3.6.0.9000 package.
+The results in this paper were obtained using \proglang{R}~3.6.1 \cite{R} with the \pkg{mice}~3.6.0.9000 package \cite{mice}.
 \proglang{R} itself and all packages used are available from the Comprehensive \proglang{R} Archive Network (CRAN) at \url{https://CRAN.R-project.org/}.

### 2019-12-11 Simulation/Writeup/NoteConvergence.Rnw
ran full simulation with convergence diagnostics per var instaed of max

@@
 %% - \AND starts a new line, \And does not.
-\author{Hanne Oberman\\Utrecht University}
+\author{\\Research Report (Preparation for Research Master Thesis, Research Seminar) \\ Methodology and Statistics for the Behavioural, Biomedical and Social Sciences \AND Hanne Oberman\\Utrecht University}
    % \And Second Author\\Plus Affiliation}
@@
 \Abstract{
-  This Research Report contains a simulation study that serves as the basis of the technical paper that will be submitted for publication in \emph{Journal of Statistical Software}.
+The ubiquitous problem of missing information may be be circumvented by implementing the framework of multiple imputation (MI).
+With MI, missing data points are `filled in' (i.e.
+imputed) several times, and statistical inference is performed on each imputed data set.
+The validity of inferences may be threathened by non-convergence of this iterative algorithmic procedure, but diagnostics to evaluate algorithmic convergence have not been implemented in MI software.
+This note investigates the performance of two convergence diagnostics ($\widehat{R}$ \citep{gelm92}, auto-correlation XYZ) by means of model-based simulation.
+To investigate convergence of the algorithm, the number of iterations is varied between one and one-hundred.
+Each of these 100 simulation conditions is replicated 1000 times.
+The convergence diagnostics are evaluated against three recommended simulation diagnostics (bias, confidence interval width, and coverage rate \citep{buur18}), since no benchmark diagnostic is available.
+This simulation study shows that three iterations may be sufficient to obtain ...
+\textbf{This Research Report contains a simulation study that serves as the basis of the technical paper that will be submitted for publication in \emph{Journal of Statistical Software}.
+I aim to publish a pre-print of this research report on \href{https://arxiv.org/}{arXiv}.
+That way, I can refer to the simulation study described here, without `bulking up' the technical paper that I want to submit for publication in JSS.
+Another option would be to attach this note as online appendix to the technical paper on ShinyMICE.}
+% This Research Report contains a simulation study that serves as the basis of the technical paper that will be submitted for publication in \emph{Journal of Statistical Software}.
 I have chosen to use the first format from the Research Report guidelines: ``\emph{It is written as a (mini) thesis, with an introduction, methods section, some results (i.e., preliminary analyses, or pilot simulations), and a discussion of results.
@@
 Another option would be to attach this note as online appendix to the technical paper on ShinyMICE.
-\textbf{TO ADD: appendix with extra results tables with different ampute and or very high or very low correlations between predictors}.
-%"It is the first half of the thesis, i.e., there are no results included yet, but the report contains a full introduction including a literature review and a methods section that contains details about the data, instruments and or statistical procedures".
-The goal was to develop novel methodology and guidelines for evaluating multiple imputation methods, and implement these in an interactive evaluation framework for multiple imputation: \pkg{ShinyMICE}.
   }
@@
 
-\textbf{ADD some references in this paragraph!} To obtain valid inferences with MI, the variability between imputations should be properly represented.
+To obtain valid inferences with MI, the variability between imputations should be properly represented \citep{rubin87, buur18}.
 If the variance between imputations is under-estimated, confidence intervals around estimates will be too narrow, which can yield spurious results.
 On the other hand, over-estimation of the variance between imputations results in unnecessarily wide confidence intervals, which can be costly because it lowers the statistical power.
-Since both of these situations are undesirable, imputations and their variability should be properly evaluated.
-Such evaluation measures are currently missing or under-developed in MI software, like the world-leading \pkg{mice} package \citep{mice} in \proglang{R} \citep{R}.
+Since both of these situations are undesirable, imputations and their variability should be evaluated.
+Evaluation measures, however, are currently missing or under-developed in MI software, like the world-leading \pkg{mice} package \citep{mice} in \proglang{R} \citep{R}.
+%\textbf{ADD sentence to link eval measures to convergence!}
 %Therefore, I will answer the following question: 'Which measures are vital for evaluating the validity of multiply imputed data?'.
-The aim of this research project is to investigate the convergence properties of the MI algorithm that is implemented in \pkg{mice}: `Multiple Imputation using Chained Equations' (MICE).
-This note addresses the question: `How to diagnose convergence of the multiple imputation algorithm implemented in \pkg{mice}?'.
+
+The goal of this research project is to develop novel methodology and guidelines for evaluating MI methods.
+These evaluation measures and guidelines will subsequently be implemented in an interactive evaluation framework for multiple imputation to aid applied researchers in drawing valid inference from incomplete datasets.
+This note provides the theoretical foundation for one of these evaluation measures that is vital in evaluting MI procedures: a diagnostic to assess convergence of the MI algorithm.
+More specifically, this note focusses on the MI algorithm that is implemented in the \proglang{R} package \pkg{mice}: `Multiple Imputation using Chained Equations' (MICE).
+The reseach question that will be adressed is: %.
+a measure that is vital in evaluating one vital measure for evaluating MI procedures--an algorithmic convergence diagnostic.
+This note specifically investigates one vital evaluation measure: how to diagnose convergence of a multiple imputation algorithm.
+%This note focusses on one measure that is vital for evaluating the validity of multiply imputed data: a convergence diagnostic. of the MI algorithm. 
+%Therefore, the aim of this research project is to investigate which measures are vital for .
+how the convergence properties of the MI algorithm that is implemented in \pkg{mice}: `Multiple Imputation using Chained Equations' (MICE).
+This note addresses the question:
+\emph{`How to diagnose convergence of the multiple imputation algorithm MICE?'.} \textbf{REMOVE THE :'s HERE!}
 
@@
 This note therefore follows notation and conventions of \cite{mice}.
-Deviations from the `original' notation by \cite{rubin87} are described in \citep[\S~XYZ]{buur18}.
+Deviations from the `original' notation by \cite{rubin87} are described in \citep[\S~2.2.3]{buur18}.
 
@@
 The algorithm employed within this step has an iterative nature.
-That is, before drawing $m$ imputed values for each missing data point in $Y_{mis}$, a `chain' of potential values is considered.
+That is, before drawing $m$ imputed values for each missing data point in $Y_{mis}$, a `chain' of potential values is sampled.
 Each of the $m$ chains starts with an initial value, drawn randomly from $Y_{obs}$.
 The chains are terminated after a predefined number of iterations.
-Only the ultimate value that a chain lands on is imputed, and hence used in the next steps (analysis and pooling).
-The collection of values (or summary statistics) across iterations of each of the $m$ imputations will be referred to as an `imputation chain'.
+Only the ultimate sample that a chain lands on is imputed, and subsequently used in the next steps (analysis and pooling).
+The collection of samples across iterations (between the initial value and the imputed value) for each of the $m$ imputations will be referred to as an `imputation chain'.
+The number of iterations will be denoted with $t$.
 
@@
 
-The methodology for diagnosing convergence of multiple imputation algorithms has not been systematically studied \cite{buur18}.
-Yet, reference manuals of promiment MI software do mention the importance of evaluating algorithmic convergence (e.g., \cite[p.XYZ]{mice}, XYZ, XYZ).
-This section provides a brief review of how convergence is defined in the context of iterative algorithmic procedures, what diagnostic tools are available, and which of these may be applicable to MI algorithms.
+The multiple imputation algorithm `MICE' is a special case of Markov chain Monte Carlo (MCMC) methods--the general framework of iterative algorithmic procedures.
+While methodology for diagnosing convergence of multiple imputation algorithms has not been systematically studied \citep{buur18}, several convergence diagnostics have been proposed for other types of MCMC methods (e.g., Gibbs samplers), or the framework in general.
+% For the framework of itrIt is not known whether convergence diagnostics that are available for the general framework that MICE is a special case of are appropriate for the evaluation of MI data.
+This is the framework of Markoc chain Monte Carlo %Yet, reference manuals of promiment MI software do mention the importance of evaluating algorithmic convergence (e.g., \cite[p.XYZ]{mice}, \textbf{XYZ, XYZ}).
+This section provides a brief review of: 1) how convergence is defined in the general context of iterative algorithmic procedures; 2) what diagnostic tools are available; and 3) which of these may be applicable to MI algorithms.
 
-\textbf{Explain the measures that I'm investigating here. R hat is ... Auto-correlation is... How are they computed? What do they say? How are they computed on MI data? Looking at imputation chains. 
-Explain that there is no baseline to compare the measures with. Only current convergence check is to visually inspect trace plots. (But in the trace-plots convergence seems immediate).}
+Losely speaking, the aim of MCMC algorithms (e.g., Gibbs samplers, or MICE) is to sample values from an unknown target distribution (e.g., the posterior distribution of a parameter in Bayesian inference, or a distribution of missing values, respectively).
+Convergence is reached once the algorithm samples exclusively from the target distribution \citep{cowl96}.
+Since this target distribution is unknown by definition, convergence can only be monitored by evaluating signs of non-convergence \citep{hoff09}.
+Therefore, in practice, the definition of convergence consists of two components: `mixing' of chains together, and `stationarity' of individual chains \citep[p.~284]{gelm13}.
+The mixing component is satisfied when chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm.
+Stationarity refers to the absence of trending within chains.
+\textbf{Figure XYZ shows a pathological case of non-convergence with respect to both components.}
+%The aim of MCMC procedures is to sample from an unknown (theoretical) distribution.
+The stability of the solution denpends on two components of convergence: should have converged, two components should be satisfied.
 
-\textbf{What is convergence in general?}
-
+% Assessing convergence is difficult.
 We can never be certain of convergence, therefore convergence diagnostics evaluate signs of non-convergence \citep{hoff09}.
@@
 But sometimes you can know if your chain has not converged, so we at least check for this latter possibility" \cite[p.~101]{hoff09}
-Convergence has two components \textbf{[look up source!!!]}: mixing between chains and stability over iterations within chains.
+% More importantly, in MCMC algorithms like MICE, "what is produced by the algorithm at convergence is not a single number or even a distribution, but rather a sample from a distribution" \citep[p.~883]{cowl96}.
+"automated convergence monitoring (as by a machine) is unsafe and should be avoided" \citep[p.~902]{cowl96}.
+%Convergence has two components: `mixing' and `stationarity' \citep[p.~284]{gelm13}.
+Mixing refers to the variation between chains of samples.
+Stationarity has to do with stability of samples within chains.
 Ideally, both components will be satisfied: there are no trends within chains, and the chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm.
-Diagnose non-convergence in the mixing sense: $\widehat{R}$.
+
+The two components of convergence may be evaluated with convergence diagnostics.
+Diagnostics that target the stationarity component are auto-correlation \citep{gelm13}, numeric standard error \citep[which is commonly called `MC error']{gewe92}, and Raftery and Lewis's \citeyear{raft92} procedure to determine the increase in the necessary number of iterations due to dependence within chains.
+The mixing component of convergence, on the other hand, is primarily assessed with the `Gelman-Rubin statistic'--potential scale reduction factor $\widehat{R}$ \citep{gelm92}.
+An adapted version of $\widehat{R}$ has been proposed by \cite{veht19}, which may also assess the stationarity component of convergence.
+Therefore, the focus of this study is mainly on $\widehat{R}$.
+Stationarity will also be evaluated with auto-correlation as diagnostic, as recommended by \cite[p.~898]{cowl96}.
+Other convergence diagnostics are outside the scope of this study.
+\footnotemark
+\footnotetext{All of these methods evaluate the convergence of univariate scalar summaries (e.g., chain means or variances).
+These convergence diagnostics cannot diagnose convergence of multivariable statistics (i.e., relations between scalar summaries).
+\cite{buur18} proposed to implement multivariable evaluation by means of eigenvalue decomposition \cite{mack03}.
+This method is outside of the scope of the current study.}%  he eigenvector decomposition method proposed by McKay \citep{XYZ}.} These four methods, however, are outside of the scope of this study.
+Here, the focus is mainly on $\widehat{R}$ and auto-correlation.
+
+%Diagnose non-convergence in the mixing sense: $\widehat{R}$.
 Diagnose non-convergence in the stability sense: auto-correlation (AC).
-Also, it is possible to compute MC errors, or the Geweke statistic \citep{XYZ}.
+Also, it is possible to compute MC errors, the Geweke statistic \textbf{\citep{XYZ}}, or to determine the appropriate number of iterations for convergence by employing \textbf{Raftery and Lewis's (1992)} method.
 All of these methods evaluate the convergence of univariate scalars of interest (e.g., chain mean or chain variance).
@@
 relations between such scalars) may be evaluated via the eigenvector decomposition method proposed by McKay \citep{XYZ}.
-These three methods, however, are outside of the scope of this study.
-Here, the focus is mainly on $\widehat{R}$.
+These four methods, however, are outside of the scope of this study.
+Here, the focus is mainly on $\widehat{R}$ and auto-correlation.}
+
+It is not known whether any convergence diagnostic is appropriate to diagnose convergence of the MICE algorithm.
+As recommended by \cite[\S~6.5.2]{buur18}, the current practice is to visually inspect imputation chains for signs of non-mixing and non-stationarity.
+Not only is visually assessing convergence challenging to the untrained eye, it might also be futile.
+That is, there is no scientific consensus on the convergence properties of multiple imputation algorithms \citep{taka17}, and in some situations the default techniques in \pkg{mice} might not converge to stable target distributions at all \citep{murr18}.
+%Convergence diagnostics for MI methods have not been systematically studied \citep{buur18}.
+Moreover,
+%A fundamental assumption of MICE is convergence of the algorithm.
+MICE is a type of Markov chain Monte Carlo (MCMC) algorithm.
+In general, the validity of inference resulting from MCMC algorithms is threatened by non-convergence.
+Hence we use convergence diagnostics to flag non-convergence.
+But it is not known whether conventional convergence diagnostics for MCMC methods work on MI data.
+Therefore, the need for convergence diagnostics is extra pressing in the context of the MICE algorithm.
 
@@
 
+\textbf{Explain the measures that I'm investigating here. R hat is ... Auto-correlation is... How are they computed? What do they say? How are they computed on MI data? Looking at imputation chains. 
+Explain that there is no baseline to compare the measures with. Only current convergence check is to visually inspect trace plots. (But in the trace-plots convergence seems immediate).}
+
 The potential scale reduction factor $\widehat{R}$ tells us by how much the variance of an estimate could be shrunken down, if the chains were to be infinitely long \cite{gelm92}.
@@
 
-We can apply $\widehat{R}$ to the mean (or to the first two moments) of the variables of interest.
-$\widehat{R}$ is defined by \cite{veth19} as follows: % The scalar summary of interest in this study is the average of the imputed values per variable.
+%We can apply $\widehat{R}$ to the mean (or to the first two moments) of the variables of interest. 
+$\widehat{R}$ is defined by \cite{veht19} as follows: % The scalar summary of interest in this study is the average of the imputed values per variable.
 %As recommended by \cite{veht19} $\widehat{R}$ is computed as ....
@@
 
-The second component of convergence ...
-interpreted as stability over iterations, or non-recurring.
+The second component of convergence stationarity requires chains to be stable across iterations, or non-recurring.
 Non-recurrence can be evaluated with auto-correlation.
@@
 
-\textbf{Convergence of MI}
-
-The convergence properties of multiple imputation algorithms lack scientific consensus \citep{taka17}.
-A review by \citep{murr18}  and some default MICE techniques might not converge to stable distributions at all \citep{murr18}.
-Convergence diagnostics for MI methods have not been systematically studied \citep{buur18}.
-Moreover,
-A fundamental assumption of MICE is convergence of the algorithm.
-MICE is a type of Markov chain Monte Carlo (MCMC) algorithm.
-In general, the validity of inference resulting from MCMC algorithms is threatened by non-convergence.
-Hence we use convergence diagnostics to flag non-convergence.
-But it is not known whether conventional convergence diagnostics for MCMC methods work on MI data.
+\textbf{Convergence diagnostics applied to MI}
 
@@
 
-\textbf{How is convergence checked now (current practice)?}
-
-Conventional thresholds to diagnose non-convergence-- e.g., Gelman and Rubin's \citeyear{gelm92} statistic $\widehat{R} < 1.1$ --are not applicable on multiply imputed data \citep{lace07} \textbf{[Expand or remove Rhat here!
+%Conventional thresholds to diagnose non-convergence-- e.g., Gelman and Rubin's \citeyear{gelm92} statistic $\widehat{R} < 1.1$ --are not applicable on multiply imputed data \citep{lace07} \textbf{[Expand or remove Rhat here!
 Refer to future section?]}.
@@
 
-\subsection{Data}
+\subsection{Data simulation}
 The simulated dataset is a finite population of $N=1000$.
@@
 \subsection{Simulation diagnostics}
-The simulation diagnostics are as recommended by \cite{buur18}.
-These comprise of average bias, average confidence interval width, and empirical coverage rate (coverage probability) across simulations.
+The simulation diagnostics are as recommended by \cite[\S~2.5.2]{buur18}.
+These evaluation criteria comprise of average bias, average confidence interval width, and empirical coverage rate across simulations.
 %"We expect that in accordance with Neyman (1934), at least 95 percent of the confidence intervals should contain the true population value, although some room for simulation error should be taken into account to counteract the finite nature of our simulations"\cite{schou18}.
@@
 A subset of simulation conditions is presented in Table \ref{tab:results}.
-This section describes convergence diagnostics ($\widehat{R}$ and auto-correlation) and simulation diagnostics (bias, confidence interval width and coverage rate) separately.
+This section describes convergence diagnostics ($\widehat{R}$ and auto-correlation) and simulation diagnostics (bias, confidence interval width and coverage rate) first separately, and then together.
 
@@
 \textbf{$\widehat{R}$}.
-Steep decline up-to iteration twenty, then more gradual decrease up-to iteration forty, more or less stable after that.
-Conventional threshold $\widehat{R} < 1.2$ reached after seven or eight iterations, $\widehat{R} < 1.1$ after fourteen iterations for chain means and sixteen for chain variances, $\widehat{R} < 1.01$ not reached within the 100 iterations considered in this simulation study (see online appendix for the full table of results).
-\textbf{[REWRITE AS STORY, NOT JUST DATA.]}
+The $\widehat{R}$ values for chain means and chain variances show equivalent trends, neither is there a clear difference between the four variables.
+Roughly speaking, we see that $\widehat{R}$ is higher for simulation conditions with a lower number of iterations, and approaches one for conditions with more iterations.
+Figure \ref{fig:conv} shows an initial dip, a small bump upwards, and then a steep decline.
+This decline \textbf{houdt aan} up-to iteration twenty, after which a more gradual decrease can be observed up-to iteration forty.
+The $\widehat{R}$ values are more or less stable after that.
+The conventional acceptable threshold for convergence, maximum $\widehat{R} < 1.2$, is reached in the simulation condition with three iterations, and conditions with six iterations or more.
+The widely used threshold $\widehat{R} < 1.1$ is met in conditions with ten or more iterations.
+The most recent recommended threshold is not reached within the 100 iterations considered in this simulation study.
+$\widehat{R} > 1.01$ for all simulation conditions (see online appendix for the full table of results).
+\textbf{Check is computing r ht is still correct!}
 
-\textbf{Auto-correlation.} Within the 100 iterations considered in this study, the AC decreases steeply, before a slow increase.
-The lowest AC observed is \Sexpr{round(min(dat[-1, c("AC chain mean", "AC chain var")]), 3)} at four iterations.
-There appears to be a trend towards zero from four iterations upwards.
-Auto-correlation does not plateau completely.
-The average absolute maximum ACs are negative up to 77 iterations for chain means, and for chain variances all iterations considered here have negative ACs.
+\textbf{Auto-correlation.} All observed auto-correlations (aggregated over the 1000 repetitions) are negative or close to zero.
+There are some differences between variables, and the ACs of chain variances seem lower than the ACs of chain means.
+The general trend in figure \ref{fig:conv} is that ACs decrease steeply as the number of iterations increase.
+\textbf{The lowest AC observed is %\Sexpr{round(min(dat[-1, c("AC chain mean", "AC chain var")]), 3)}
+at four iterations.
+Simulation conditions with a higher number of iterations have show a slow increase towards zero.
+Auto-correlations do not plateau completely within the range of simulation conditions considered in this study.
+It seems that ACs might become positive if more iterations would be allowed.
+The average absolute maximum ACs are negative up to 77 iterations for chain means, and for chain variances all iterations considered here have negative ACs.}
 
@@
 \subsection{Simulation diagnostics}
-\textbf{Bias.} From two iterations upwards, the average bias across repetitions is stable. Across iterations two to one hundred, bias fluctuates within a narrow range around zero. 
+\textbf{Bias.} Figure \ref{fig:sim} shows the average bias between the regression coefficient of $X$ on $Y$ in the complete dataset, and $\bar{Q}$, \textbf{the pooled estimated regression coefficient and the  is close to zero for simulation conditions with two iterations or more.
+The simulation condition with one iteration has a negative bias.
+From two iterations upwards, the average bias across repetitions is stable.
+Across iterations two to one hundred, bias fluctuates within a narrow range around zero.
+Loess line is flat around it = 20.}
 % Average bias is not zero because there is only one incomplete dataset to be imputed. Bias is a sample effect.
 
-\textbf{Confidence interval width.} CIW does not show a clear trend across iterations.
-One might argue that there is some downward trending up-to thirthy to forty iterations.
+\textbf{Confidence interval width.} CIW is only clearly divergent for the simulation condition with one iteration.
+Conditions with two or more iterations have similar CIWs.
+The magnitude of differences from the loess line is somewhat larger up-to the condition with four iterations.
+From four iterations on, CIW seems to be stable across iterations.
+\textbf{Loess line is never flat.}%One might argue that there is some downward trending up-to thirthy to forty iterations.
 %After about twenty-five iterations the average CIW across all simulation conditions and repetitions is reached \Sexpr{round(mean(dat[, "CI width"]), 3)}.
 
-\textbf{Coverage rate.} The coverage rate is more or less stable from two iterations upwards.
+\textbf{Coverage rate.} \textbf{The coverage rate is more or less stable from two iterations upwards.
 On average, the coverage rate is somewhat higher than the expected nominal coverage of 95\% \cite{neym34}, namely \Sexpr{100*round(mean(dat[, "Cov.
 rate"]), 3)}\%.
+Loess line is never flat.}
 
-\textbf{add loess line to plots to see when it becomes a flat line. also, give some implications for these measures.}
+\textbf{give some implications for these measures. connect convergence and simulation diagnostics.}
 
@@
 \section{Summary and discussion} \label{sec:summary}
+
+\textbf{Answer RQ.} How to diagnose non-convergence? It's complicated.
 
@@
 
-\textbf{Future research.} We would like to have convergence measures for multivariable statistics (scalars?) of interest.
+\textbf{Future research.} %We would like to have convergence measures for multivariable statistics (scalars?) of interest.
 This is, however, dependent on the complete data model.
@@
 This paper is written by the sole author (Hanne Oberman, BSc.), with guidance from Master thesis supervisors prof. dr. Stef van Buuren, and dr. Gerko Vink.
-
-\textbf{LEAVING THIS HERE FOR REFERENCE, PLEASE REMOVE BEFORE HANDING IN}
-
-<<fig = TRUE, echo = FALSE>>=
-plot.ts(dat[c(5:8, 2:4)], main = "", xlab = "Number of iterations")
-@
 

### 2019-12-11 Simulation/Writeup/NoteConvergence.tex
ran full simulation with convergence diagnostics per var instaed of max

@@
 \textbf{$\widehat{R}$}.
-Steep decline up-to iteration twenty, then more gradual decrease up-to iteration forty, more or less stable after that.
-Conventional threshold $\widehat{R} < 1.2$ reached after seven or eight iterations, $\widehat{R} < 1.1$ after fourteen iterations for chain means and sixteen for chain variances, $\widehat{R} < 1.01$ not reached within the 100 iterations considered in this simulation study (see online appendix for the full table of results).
-\textbf{[REWRITE AS STORY, NOT JUST DATA.]}
+The $\widehat{R}$ values for chain means and chain variances show equivalent trends, neither is there a clear difference between the four variables.
+Roughly speaking, we see that $\widehat{R}$ is higher for simulation conditions with a lower number of iterations, and approaches one for conditions with more iterations.
+Figure \ref{fig:conv} shows an initial dip, a small bump upwards, and then a steep decline.
+This decline \textbf{houdt aan} up-to iteration twenty, after which a more gradual decrease can be observed up-to iteration forty.
+The $\widehat{R}$ values are more or less stable after that.
+The conventional acceptable threshold for convergence, maximum $\widehat{R} < 1.2$, is reached in the simulation condition with three iterations, and conditions with six iterations or more.
+The widely used threshold $\widehat{R} < 1.1$ is met in conditions with ten or more iterations.
+The most recent recommended threshold is not reached within the 100 iterations considered in this simulation study.
+$\widehat{R} > 1.01$ for all simulation conditions (see online appendix for the full table of results).
+\textbf{Check is computing r ht is still correct!}
 
-\textbf{Auto-correlation.} Within the 100 iterations considered in this study, the AC decreases steeply, before a slow increase.
-The lowest AC observed is -0.735 at four iterations.
-There appears to be a trend towards zero from four iterations upwards.
-Auto-correlation does not plateau completely.
-The average absolute maximum ACs are negative up to 77 iterations for chain means, and for chain variances all iterations considered here have negative ACs.
+\textbf{Auto-correlation.} All observed auto-correlations (aggregated over the 1000 repetitions) are negative or close to zero.
+There are some differences between variables, and the ACs of chain variances seem lower than the ACs of chain means.
+The general trend in figure \ref{fig:conv} is that ACs decrease steeply as the number of iterations increase.
+\textbf{The lowest AC observed is %-0.735
+at four iterations.
+Simulation conditions with a higher number of iterations have show a slow increase towards zero.
+Auto-correlations do not plateau completely within the range of simulation conditions considered in this study.
+It seems that ACs might become positive if more iterations would be allowed.
+The average absolute maximum ACs are negative up to 77 iterations for chain means, and for chain variances all iterations considered here have negative ACs.}
 
@@
 \subsection{Simulation diagnostics}
-\textbf{Bias.} From two iterations upwards, the average bias across repetitions is stable. Across iterations two to one hundred, bias fluctuates within a narrow range around zero. 
+\textbf{Bias.} Figure \ref{fig:sim} shows the average bias between the regression coefficient of $X$ on $Y$ in the complete dataset, and $\bar{Q}$, \textbf{the pooled estimated regression coefficient and the  is close to zero for simulation conditions with two iterations or more.
+The simulation condition with one iteration has a negative bias.
+From two iterations upwards, the average bias across repetitions is stable.
+Across iterations two to one hundred, bias fluctuates within a narrow range around zero.
+Loess line is flat around it = 20.}
 % Average bias is not zero because there is only one incomplete dataset to be imputed. Bias is a sample effect.
 
-\textbf{Confidence interval width.} CIW does not show a clear trend across iterations.
-One might argue that there is some downward trending up-to thirthy to forty iterations.
+\textbf{Confidence interval width.} CIW is only clearly divergent for the simulation condition with one iteration.
+Conditions with two or more iterations have similar CIWs.
+The magnitude of differences from the loess line is somewhat larger up-to the condition with four iterations.
+From four iterations on, CIW seems to be stable across iterations.
+\textbf{Loess line is never flat.}%One might argue that there is some downward trending up-to thirthy to forty iterations.
 %After about twenty-five iterations the average CIW across all simulation conditions and repetitions is reached 0.928.
 
-\textbf{Coverage rate.} The coverage rate is more or less stable from two iterations upwards.
+\textbf{Coverage rate.} \textbf{The coverage rate is more or less stable from two iterations upwards.
 On average, the coverage rate is somewhat higher than the expected nominal coverage of 95\% \cite{neym34}, namely 95.3\%.
+Loess line is never flat.}
 
-\textbf{add loess line to plots to see when it becomes a flat line. also, give some implications for these measures.}
+\textbf{give some implications for these measures. connect convergence and simulation diagnostics.}
 
@@
 This paper is written by the sole author (Hanne Oberman, BSc.), with guidance from Master thesis supervisors prof. dr. Stef van Buuren, and dr. Gerko Vink.
-
-\textbf{LEAVING THIS HERE FOR REFERENCE, PLEASE REMOVE BEFORE HANDING IN}
-
-\includegraphics{NoteConvergence-003}
 

### 2019-12-11 Simulation/1.Execute.R

@@ -27,7 +27,7 @@ source("Simulation/Functions/Evaluate.R")
 # simulation parameters
 populationsize <- 1000 #n of simulated dataset
 n.iter <- 100 #nr of iterations (varying 1:n.iter)
-n.sim <- 3 #nr of simulations per iteration value
+n.sim <- 1000 #nr of simulations per iteration value
 true_effect <- 2 #regression coefficient to be estimated
 
 # start simulation study
@@ -81,5 +81,5 @@ results <- evaluate.sim(sims = out)
 ###
 
 # save for future reference
-save.Rdata(results, name = "results17.Rdata", path = "Simulation/Results")
-#save.image("environment.Rdata")
+save.Rdata(results, name = "final", path = "Simulation/Results")
+save.image("environment.Rdata")

### 2019-12-11 Simulation/Writeup/NoteConvergence.Rnw
Checked references

@@
 \Abstract{
-The ubiquitous problem of missing information may be be circumvented by implementing the framework of multiple imputation (MI).
+The ubiquitous problem of missing information may be circumvented by implementing the framework of multiple imputation (MI).
 With MI, missing data points are `filled in' (i.e.
 imputed) several times, and statistical inference is performed on each imputed data set.
-The validity of inferences may be threathened by non-convergence of this iterative algorithmic procedure, but diagnostics to evaluate algorithmic convergence have not been implemented in MI software.
-This note investigates the performance of two convergence diagnostics ($\widehat{R}$ \citep{gelm92}, auto-correlation XYZ) by means of model-based simulation.
+The validity of inferences may be threatened by non-convergence of this iterative algorithmic procedure, but diagnostics to evaluate algorithmic convergence have not been implemented in MI software.
+This note investigates the performance of two convergence diagnostics ($\widehat{R}$ \citep{gelm92}, auto-correlation \citep{scha97}) through model-based simulation.
 To investigate convergence of the algorithm, the number of iterations is varied between one and one-hundred.
 Each of these 100 simulation conditions is replicated 1000 times.
-The convergence diagnostics are evaluated against three recommended simulation diagnostics (bias, confidence interval width, and coverage rate \citep{buur18}), since no benchmark diagnostic is available.
+The convergence diagnostics are evaluated against three recommended simulation diagnostics (bias, confidence interval width, and coverage rate \citep{buur18}, since no benchmark diagnostic is available.
 This simulation study shows that three iterations may be sufficient to obtain ...
@@
 That way, I can refer to the simulation study described here, without `bulking up' the technical paper that I want to submit for publication in JSS.
-Another option would be to attach this note as online appendix to the technical paper on ShinyMICE.}
+Another option would be to attach this note as an online appendix to the technical paper on ShinyMICE.}
 % This Research Report contains a simulation study that serves as the basis of the technical paper that will be submitted for publication in \emph{Journal of Statistical Software}.
@@
 These evaluation measures and guidelines will subsequently be implemented in an interactive evaluation framework for multiple imputation to aid applied researchers in drawing valid inference from incomplete datasets.
-This note provides the theoretical foundation for one of these evaluation measures that is vital in evaluting MI procedures: a diagnostic to assess convergence of the MI algorithm.
-More specifically, this note focusses on the MI algorithm that is implemented in the \proglang{R} package \pkg{mice}: `Multiple Imputation using Chained Equations' (MICE).
-The reseach question that will be adressed is: %.
+This note provides the theoretical foundation for one of these evaluation measures that is vital in evaluating MI procedures: a diagnostic to assess convergence of the MI algorithm.
+More specifically, this note focuses on the MI algorithm that is implemented in the \proglang{R} package \pkg{mice}: `Multiple Imputation using Chained Equations' (MICE).
+The research question that will be addressed is: %.
 a measure that is vital in evaluating one vital measure for evaluating MI procedures--an algorithmic convergence diagnostic.
@@
 
-The convergence properties of MICE are investigated by means of model-based simulation in \proglang{R} \citep{R}.
+The convergence properties of MICE are investigated through model-based simulation in \proglang{R} \citep{R}.
 The results of this simulation study are guidelines for assessing convergence of the MI algorithm.
@@
 The convergence guidelines introduced in this paper are developed to be integrated into the \pkg{mice} environment \citep{mice} in \proglang{R} \citep{R}.
-This note therefore follows notation and conventions of \cite{mice}.
+This note, therefore, follows notation and conventions of \cite{mice}.
 Deviations from the `original' notation by \cite{rubin87} are described in \citep[\S~2.2.3]{buur18}.
@@
 
-This note focusses on the algorithmic properties of the imputation step.
+This note focuses on the algorithmic properties of the imputation step.
 The algorithm employed within this step has an iterative nature.
@@
 The chains are terminated after a predefined number of iterations.
-Only the ultimate sample that a chain lands on is imputed, and subsequently used in the next steps (analysis and pooling).
+Only the ultimate sample that a chain lands on is imputed and subsequently used in the analysis and pooling steps.
 The collection of samples across iterations (between the initial value and the imputed value) for each of the $m$ imputations will be referred to as an `imputation chain'.
-The number of iterations will be denoted with $t$.
+The number of iterations will be denoted with $t$ \textbf{[NOG TOEPASSEN]}.
 
@@
 This is the framework of Markoc chain Monte Carlo %Yet, reference manuals of promiment MI software do mention the importance of evaluating algorithmic convergence (e.g., \cite[p.XYZ]{mice}, \textbf{XYZ, XYZ}).
-This section provides a brief review of: 1) how convergence is defined in the general context of iterative algorithmic procedures; 2) what diagnostic tools are available; and 3) which of these may be applicable to MI algorithms.
+This section provides a brief review of: 1) how convergence is defined in the general context of iterative algorithmic procedures; 2) what diagnostic tools are available; and 3) which of these may apply to MI algorithms.
 
-Losely speaking, the aim of MCMC algorithms (e.g., Gibbs samplers, or MICE) is to sample values from an unknown target distribution (e.g., the posterior distribution of a parameter in Bayesian inference, or a distribution of missing values, respectively).
+Loosely speaking, the aim of MCMC algorithms (e.g., Gibbs samplers, or MICE) is to sample values from an unknown target distribution (e.g., the posterior distribution of a parameter in Bayesian inference, or a distribution of missing values, respectively).
 Convergence is reached once the algorithm samples exclusively from the target distribution \citep{cowl96}.
@@
 Stationarity refers to the absence of trending within chains.
-\textbf{Figure XYZ shows a pathological case of non-convergence with respect to both components.}
+\textbf{Figure XYZ shows a pathological case of non-convergence concerning both components.}
 %The aim of MCMC procedures is to sample from an unknown (theoretical) distribution.
@@
 The two components of convergence may be evaluated with convergence diagnostics.
-Diagnostics that target the stationarity component are auto-correlation \citep{gelm13}, numeric standard error \citep[which is commonly called `MC error']{gewe92}, and Raftery and Lewis's \citeyear{raft92} procedure to determine the increase in the necessary number of iterations due to dependence within chains.
+Diagnostics that target the stationarity component are auto-correlation \citep{scha97, gelm13}, numeric standard error \citep[which is commonly called `MC error']{gewe92}, and Raftery and Lewis's \citeyear{raft91} procedure to determine the increase in the necessary number of iterations due to dependence within chains.
 The mixing component of convergence, on the other hand, is primarily assessed with the `Gelman-Rubin statistic'--potential scale reduction factor $\widehat{R}$ \citep{gelm92}.
@@
 As recommended by \cite[\S~6.5.2]{buur18}, the current practice is to visually inspect imputation chains for signs of non-mixing and non-stationarity.
-Not only is visually assessing convergence challenging to the untrained eye, it might also be futile.
-That is, there is no scientific consensus on the convergence properties of multiple imputation algorithms \citep{taka17}, and in some situations the default techniques in \pkg{mice} might not converge to stable target distributions at all \citep{murr18}.
+Not only is visually assessing convergence challenging to the untrained eye--it might also be futile.
+That is, there is no scientific consensus on the convergence properties of multiple imputation algorithms \citep{taka17}, and in some situations, the default techniques in \pkg{mice} might not converge to stable target distributions at all \citep{murr18}.
 %Convergence diagnostics for MI methods have not been systematically studied \citep{buur18}.
@@
 \textbf{Explain the measures that I'm investigating here. R hat is ... Auto-correlation is... How are they computed? What do they say? How are they computed on MI data? Looking at imputation chains. 
-Explain that there is no baseline to compare the measures with. Only current convergence check is to visually inspect trace plots. (But in the trace-plots convergence seems immediate).}
+Explain that there is no baseline to compare the measures with. The only current convergence check is to visually inspect trace plots. (But in the trace-plots convergence seems immediate).}
 
-The potential scale reduction factor $\widehat{R}$ tells us by how much the variance of an estimate could be shrunken down, if the chains were to be infinitely long \cite{gelm92}.
+The potential scale reduction factor $\widehat{R}$ tells us by how much the variance of an estimate could be shrunken down if the chains were to be infinitely long \cite{gelm92}.
 That then informs us about how dependent the chains are on the starting values.
@@
 $\widehat{R}$ is then equal to one.
-The conventional acceptable threshold was $\widehat{R} < 1.2$.
+The conventionally acceptable threshold was $\widehat{R} < 1.2$.
 More recently, \cite{veht19} proposed a more stringent threshold: $\widehat{R} < 1.01$.
@@
 A high auto-correlation indicates dependence within chains.
-The magnitude of the AC can be interpreted qualitatively, or compared to the critical value of a \textbf{XYZ} distribution with \textbf{XYZ} degrees of freedom.
+The magnitude of the AC can be interpreted qualitatively or quantitatively.
+Quantitative evaluation of the AC entails comparing the observed ACs to the critical values of a two-tailed 95\% confidence interval, divided by the square root of the number of iterations.
+%.
+critical value corresponding to the $z-$~value standard normal distribution with \textbf{XYZ} degrees of freedom.
+%  AC_lim <- qnorm(.975) / sqrt(maxit)
 
@@
 
-The hypothesis of this simulation study is that $\widehat{R}$ will over-estimate non-convergence.
+This simulation study hypothesizes that $\widehat{R}$ will over-estimate non-convergence.
 \textbf{[Or use:] Hypothesis based on \cite{lace07} is that the conventional acceptable level of $\widehat{R}$ is too strict for MI data.
@@
 
-To investigate convergence properties of the \pkg{mice} algorithm, different lenghts of imputation chains are considered.
+To investigate the convergence properties of the \pkg{mice} algorithm, different lengths of imputation chains are considered.
 This simulation study considers 100 simulation conditions.
@@
 In this study, 1000 simulation runs are performed for each of the 100 simulation conditions.
-\textbf{where for one through one hundred the Rhat and ...
-are calcuated.} The simulation set-up consists of several steps, summarized in the pseudo-code below.
+\textbf{Estimands are convergence diagnostics.
+Performance measures are simulation diagnostics.
+where for one through one hundred the Rhat and ...
+are calculated.} The simulation set-up consists of several steps, summarized in the pseudo-code below.
 The complete \proglang{R} script of the simulation study is available on Github.
@@
 All simulations are performed with imputation method 'norm' (Bayesian linear regression imputation), and five imputation chains ($m = 5$).
-The number of iterations is varied over simulations ('maxit' argument between 1 and 100).
+The number of iterations is varied over simulations (\texttt{maxit} argument between 1 and 100).
 Each simulation condition (i.e., each number of iterations) is simulated 1000 times.
@@
 
-Auto-correlation (AC) is computed within imputation chains, as the correlation between $\theta$ at the $i^{th}$ iteration and $\theta$ at the $i+1^{th}$ iteration, where $\theta$ is a scalar summary of interest (i.e., chain means and chain variances).
+Auto-correlation (AC) is computed within imputation chains, as the correlation between $\theta$ at \textbf{the $i^{th}$ iteration and $\theta$ at the $i+1^{th}$ iteration [CHECK VARIABLE IN NOTATION]}, where $\theta$ is a scalar summary of interest (i.e., chain means and chain variances).
 The AC of the variable with the highest absolute AC value is reported per simulation.
@@
 The $\widehat{R}$ values are more or less stable after that.
-The conventional acceptable threshold for convergence, maximum $\widehat{R} < 1.2$, is reached in the simulation condition with three iterations, and conditions with six iterations or more.
+The \textbf{conventionally acceptable threshold for convergence, maximum $\widehat{R} < 1.2$,[CHECK IN INTRO]} is reached in the simulation condition with three iterations, and conditions with six iterations or more.
 The widely used threshold $\widehat{R} < 1.1$ is met in conditions with ten or more iterations.
@@
 $\widehat{R} > 1.01$ for all simulation conditions (see online appendix for the full table of results).
-\textbf{Check is computing r ht is still correct!}
+\textbf{Check is computing r ht is still correct!
+Also say something about Rh below 1.}
 
@@
 There are some differences between variables, and the ACs of chain variances seem lower than the ACs of chain means.
-The general trend in figure \ref{fig:conv} is that ACs decrease steeply as the number of iterations increase.
+The general trend in figure \ref{fig:conv} is that ACs decrease steeply as the number of iterations increases.
 \textbf{The lowest AC observed is %\Sexpr{round(min(dat[-1, c("AC chain mean", "AC chain var")]), 3)}
 at four iterations.
-Simulation conditions with a higher number of iterations have show a slow increase towards zero.
+Simulation conditions with a higher number of iterations show a slow increase towards zero.
 Auto-correlations do not plateau completely within the range of simulation conditions considered in this study.
@@
 \subsection{Simulation diagnostics}
-\textbf{Bias.} Figure \ref{fig:sim} shows the average bias between the regression coefficient of $X$ on $Y$ in the complete dataset, and $\bar{Q}$, \textbf{the pooled estimated regression coefficient and the  is close to zero for simulation conditions with two iterations or more.
+\textbf{Bias.} Figure \ref{fig:sim} shows the average bias between the regression coefficient of $X$ on $Y$ in the complete dataset, and $\bar{Q}$, \textbf{the pooled estimated regression coefficient and the ...
+Average bias is close to zero for simulation conditions with two iterations or more.
 The simulation condition with one iteration has a negative bias.
@@
 The most recent recommended threshold for $\widehat{R}$ ($\widehat{R} < 1.01$) may be too stringent for MI data.
+\textbf{However, using the 1.2 threshold may be to lenient because this was observed at three iterations and then again after six.
+If we would use 1.2, we might miss this increase after three iterations.
+The 1.1 threshold seems ok.
+Also, ten iterations are computationally not terrible.
+With increased performance and storage capacities, doubling the default is fine I guess.}
 
-From the simulation diagnostics, it appears that as little as two iterations could be sufficient to draw \textbf{[valid inferences?]}.
+From the simulation diagnostics, it appears that as little as three iterations might be sufficient to obtain unbiased, confidence valid estimates.
 Convergence diagnostics $\widehat{R}$ and auto-correlation, however, indicate that convergence may only be reached after twenty or even forty iterations.
@@
 
-\textbf{Future research.} %We would like to have convergence measures for multivariable statistics (scalars?) of interest.
+\textbf{Future research.} This study consdered only a MCAR missingness mechanism.
+Necessary but not sufficient to demonstrate the performance of convergence diagnostics.
+This is just a proof of concept.
+Further reseach is needed to investigate the performance under violation of convergence, e.g.
+dependency between predictors (very high correlations).
+%We would like to have convergence measures for multivariable statistics (scalars?) of interest.
 This is, however, dependent on the complete data model.

### 2019-12-11 Simulation/Writeup/NoteConvergence.tex
Checked references

@@
 The convergence guidelines introduced in this paper are developed to be integrated into the \pkg{mice} environment \citep{mice} in \proglang{R} \citep{R}.
-This note therefore follows notation and conventions of \cite{mice}.
-Deviations from the `original' notation by \cite{rubin87} are described in \citep[\S~XYZ]{buur18}.
+This note, therefore, follows notation and conventions of \cite{mice}.
+Deviations from the `original' notation by \cite{rubin87} are described in \citep[\S~2.2.3]{buur18}.
 
@@
 There are some differences between variables, and the ACs of chain variances seem lower than the ACs of chain means.
-The general trend in figure \ref{fig:conv} is that ACs decrease steeply as the number of iterations increase.
+The general trend in figure \ref{fig:conv} is that ACs decrease steeply as the number of iterations increases.
 \textbf{The lowest AC observed is %-0.735
 at four iterations.
-Simulation conditions with a higher number of iterations have show a slow increase towards zero.
+Simulation conditions with a higher number of iterations show a slow increase towards zero.
 Auto-correlations do not plateau completely within the range of simulation conditions considered in this study.

### 2019-12-11 Simulation/1.Execute.R

@@ -81,5 +81,5 @@ results <- evaluate.sim(sims = out)
 ###
 
 # save for future reference
-save.Rdata(results, name = "final", path = "Simulation/Results")
+save.Rdata(results, name = "multivar_miss", path = "Simulation/Results")
 save.image("environment.Rdata")

### 2019-12-11 Simulation/Writeup/NoteConvergence.Rnw
Some feedback from Gerko

@@
 These evaluation measures and guidelines will subsequently be implemented in an interactive evaluation framework for multiple imputation to aid applied researchers in drawing valid inference from incomplete datasets.
-This note provides the theoretical foundation for one of these evaluation measures that is vital in evaluating MI procedures: a diagnostic to assess convergence of the MI algorithm.
-More specifically, this note focuses on the MI algorithm that is implemented in the \proglang{R} package \pkg{mice}: `Multiple Imputation using Chained Equations' (MICE).
-The research question that will be addressed is: %.
+This note provides the theoretical foundation towards the diagnostic evaluation of the convergence of MI algorithms.
+% for diagnostics for evalua one of these evaluation measures that is vital in evaluating MI procedures: a diagnostic to assess convergence of the MI algorithm.
+For reasons of brevity, this note only focuses ion the MI algo implemented in mice. %The implementation limits itself to the MI algorithm 
+%More specifically, this note focuses on the MI algorithm that is implemented in the \proglang{R} package \pkg{mice}: `Multiple Imputation using Chained Equations' (MICE) \textbf{[can be removed, can be found in the mice ref]}.
+To this end we will evaluate how convergence of the imputation algo can be diagnosed.
+
+%The research question that will be addressed is: %.
 a measure that is vital in evaluating one vital measure for evaluating MI procedures--an algorithmic convergence diagnostic.
@@
 This note addresses the question:
-\emph{`How to diagnose convergence of the multiple imputation algorithm MICE?'.} \textbf{REMOVE THE :'s HERE!}
+%\emph{`How to diagnose convergence of the multiple imputation algorithm MICE?'.}
 
@@
 
-To investigate the convergence properties of the \pkg{mice} algorithm, different lengths of imputation chains are considered.
-This simulation study considers 100 simulation conditions.
-Each condition has a different maximum number of iterations, varying between one and one-hundred iterations.
-In this study, 1000 simulation runs are performed for each of the 100 simulation conditions.
+To investigate the convergence properties of the \pkg{mice} algorithm, we perform model-based simulation of the MI procedure with varying imputation chain lengths.
+Each simulation condition has a different maximum number of iterations, from one to one-hundred iterations.
+We compute $\widehat{R}$ and auto-correlation (`convergence diagnostics') and performance measures (`simulation diagnostics') for each condition.
+The number of simulation runs per condition is 1000.
+%All simulation conditions are replicated 1000 times.
+%For each of the 100 conditions 1000 simulation runs are performed.
+for each of the 100 simulation conditions.
 \textbf{Estimands are convergence diagnostics.
@@
 where for one through one hundred the Rhat and ...
-are calculated.} The simulation set-up consists of several steps, summarized in the pseudo-code below.
-The complete \proglang{R} script of the simulation study is available on Github.
+are calculated.}
+The simulation set-up consists of several steps, summarized in the pseudo-code below.\footnote{The complete \proglang{R} script of the simulation study is available on Github.}
 

### 2019-12-12 Simulation/Writeup/NoteConvergence.Rnw
definition rhat

@@
 %% - \Shorttitle{} with LaTeX markup (if any), used as running title
-\title{A Note on Convergence Diagnostics for Multiple Imputation using Chained Equations (MICE)}
-\Plaintitle{A Note on Convergence Diagnostics for Multiple Imputation using Chained Equations (MICE)}
-\Shorttitle{Convergence Diagnostics for MICE}
+\title{A Note on Convergence Diagnostics for Multiple Imputation Algorithms}
+\Plaintitle{A Note on Convergence Diagnostics for Multiple Imputation Algorithms}
+\Shorttitle{Convergence Diagnostics for MI}
 
@@
 The validity of inferences may be threatened by non-convergence of this iterative algorithmic procedure, but diagnostics to evaluate algorithmic convergence have not been implemented in MI software.
-This note investigates the performance of two convergence diagnostics ($\widehat{R}$ \citep{gelm92}, auto-correlation \citep{scha97}) through model-based simulation.
+This note investigates the performance of two convergence diagnostics (\citep[$\widehat{R}$,][]{gelm92}; \citep[auto-correlation,][]{scha97}) through model-based simulation.
 To investigate convergence of the algorithm, the number of iterations is varied between one and one-hundred.
@@
 % for diagnostics for evalua one of these evaluation measures that is vital in evaluating MI procedures: a diagnostic to assess convergence of the MI algorithm.
-For reasons of brevity, this note only focuses ion the MI algo implemented in mice. %The implementation limits itself to the MI algorithm 
+For reasons of brevity, this note only focuses on the MI algorithm implemented in \pkg{mice} \cite{mice}. %The implementation limits itself to the MI algorithm 
 %More specifically, this note focuses on the MI algorithm that is implemented in the \proglang{R} package \pkg{mice}: `Multiple Imputation using Chained Equations' (MICE) \textbf{[can be removed, can be found in the mice ref]}.
-To this end we will evaluate how convergence of the imputation algo can be diagnosed.
+To this end we will evaluate how convergence of the imputation algorithm can be diagnosed.
 
@@
 
-The convergence properties of MICE are investigated through model-based simulation in \proglang{R} \citep{R}.
-The results of this simulation study are guidelines for assessing convergence of the MI algorithm.
+The convergence properties of the MI algoritm in \pkg{mice} are investigated through model-based simulation.
+The results of this simulation study are guidelines for assessing convergence of MI algorithms.
 These guidelines will be implemented in an interactive evaluation tool for \pkg{mice}, `ShinyMICE', which is currently under development.
@@
 \end{tikzpicture}}
-\caption{Scheme of the main steps in multiple imputation ($m = 3$). Adapted from \citep[\S~1.4.1]{buur18}.}
+\caption{Scheme of the main steps in multiple imputation ($m = 3$). Adapted from \cite[\S~1.4.1]{buur18}.}
 \end{figure}
@@
 The collection of samples across iterations (between the initial value and the imputed value) for each of the $m$ imputations will be referred to as an `imputation chain'.
-The number of iterations will be denoted with $t$ \textbf{[NOG TOEPASSEN]}.
+The number of iterations will be denoted with $t$.
 
@@
 
-The multiple imputation algorithm `MICE' is a special case of Markov chain Monte Carlo (MCMC) methods--the general framework of iterative algorithmic procedures.
-While methodology for diagnosing convergence of multiple imputation algorithms has not been systematically studied \citep{buur18}, several convergence diagnostics have been proposed for other types of MCMC methods (e.g., Gibbs samplers), or the framework in general.
+The multiple imputation algorithm in \pkg{mice} is a special case of Markov chain Monte Carlo (MCMC) methods--a framework of iterative algorithmic procedures.
+While methodology for diagnosing convergence of MI algorithms has not been systematically studied \citep{buur18}, several convergence diagnostics have been proposed for other types of MCMC methods (e.g., Gibbs samplers) and the framework in general.
 % For the framework of itrIt is not known whether convergence diagnostics that are available for the general framework that MICE is a special case of are appropriate for the evaluation of MI data.
@@
 
-Loosely speaking, the aim of MCMC algorithms (e.g., Gibbs samplers, or MICE) is to sample values from an unknown target distribution (e.g., the posterior distribution of a parameter in Bayesian inference, or a distribution of missing values, respectively).
+Loosely speaking, the aim of MCMC algorithms (e.g., Gibbs samplers, or the MI algorithm in \pkg{mice}) is to sample values from an unknown target distribution (e.g., the posterior distribution of a parameter in Bayesian inference, or a distribution of missing values, respectively).
 Convergence is reached once the algorithm samples exclusively from the target distribution \citep{cowl96}.
 Since this target distribution is unknown by definition, convergence can only be monitored by evaluating signs of non-convergence \citep{hoff09}.
-Therefore, in practice, the definition of convergence consists of two components: `mixing' of chains together, and `stationarity' of individual chains \citep[p.~284]{gelm13}.
+In practice, the definition of convergence consists of two components: `mixing' of chains together, and `stationarity' of individual chains \citep[p.~284]{gelm13}.
 The mixing component is satisfied when chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm.
-Stationarity refers to the absence of trending within chains.
-\textbf{Figure XYZ shows a pathological case of non-convergence concerning both components.}
+Stationarity refers to the absence of trending across iterations within chains.
+%\textbf{Figure XYZ shows a pathological case of non-convergence concerning both components.}
 %The aim of MCMC procedures is to sample from an unknown (theoretical) distribution.
@@
 
-The two components of convergence may be evaluated with convergence diagnostics.
-Diagnostics that target the stationarity component are auto-correlation \citep{scha97, gelm13}, numeric standard error \citep[which is commonly called `MC error']{gewe92}, and Raftery and Lewis's \citeyear{raft91} procedure to determine the increase in the necessary number of iterations due to dependence within chains.
-The mixing component of convergence, on the other hand, is primarily assessed with the `Gelman-Rubin statistic'--potential scale reduction factor $\widehat{R}$ \citep{gelm92}.
-An adapted version of $\widehat{R}$ has been proposed by \cite{veht19}, which may also assess the stationarity component of convergence.
-Therefore, the focus of this study is mainly on $\widehat{R}$.
-Stationarity will also be evaluated with auto-correlation as diagnostic, as recommended by \cite[p.~898]{cowl96}.
-Other convergence diagnostics are outside the scope of this study.
-\footnotemark
+Most convergence diagnostics target either of the two components.
+The stationarity component of convergence may be evaluated with auto-correlation \citep{scha97, gelm13}, numeric standard error \citep[which is commonly called `MC error']{gewe92}, or Raftery and Lewis's (\citeyear{raft91}) procedure to determine how many iterations are necessary to deminish the effect of trending within chains.
+The mixing component can be assessed with the potential scale reduction factor $\widehat{R}$ \citep[widely referred to as `the Gelman-Rubin statistic']{gelm92}.
+With an adapted version of $\widehat{R}$, proposed by \cite{veht19}, we may also evaluate the stationarity component of convergence.
+This would make $\widehat{R}$ a general convergence diagnostic.
+The application of $\widehat{R}$ to assess stationarity has not been thourougly investigated.
+Therefore, this study employs both $\widehat{R}$ and auto-correlation to investigate algorithmic convergence, as recommended by \cite[p.~898]{cowl96}.
+%Therefore, the focus of this study is mainly on $\widehat{R}$. Stationarity will also be evaluated with auto-correlation as diagnostic, as recommended by \cite[p.~898]{cowl96}. 
+Other convergence diagnostics are outside the scope of this study.\footnotemark 
 \footnotetext{All of these methods evaluate the convergence of univariate scalar summaries (e.g., chain means or variances).
@@
 
-It is not known whether any convergence diagnostic is appropriate to diagnose convergence of the MICE algorithm.
-As recommended by \cite[\S~6.5.2]{buur18}, the current practice is to visually inspect imputation chains for signs of non-mixing and non-stationarity.
-Not only is visually assessing convergence challenging to the untrained eye--it might also be futile.
-That is, there is no scientific consensus on the convergence properties of multiple imputation algorithms \citep{taka17}, and in some situations, the default techniques in \pkg{mice} might not converge to stable target distributions at all \citep{murr18}.
+It is not known whether any convergence diagnostic is appropriate to diagnose convergence of \pkg{mice} algorithms.
+As by \cite[\S~6.5.2]{buur18}, the current practice is to visually inspect imputation chains for signs of non-mixing and non-stationarity.
+Visually assessing convergence, however, may be challenging to the untrained eye.
+\textbf{[This is problematic because]} There is no scientific consensus on the convergence properties of multiple imputation algorithms \citep{taka17}, and in some situations, the default techniques in \pkg{mice} might not converge to stable target distributions at all \citep{murr18}.
 %Convergence diagnostics for MI methods have not been systematically studied \citep{buur18}.
@@
 But it is not known whether conventional convergence diagnostics for MCMC methods work on MI data.
-Therefore, the need for convergence diagnostics is extra pressing in the context of the MICE algorithm.
+Therefore, convergence of the \pkg{mice} algorithms should be monitored carefully. %the need for convergence diagnostics is extra pressing in the context of the MICE algorithm.
 
@@
 
-\textbf{Explain the measures that I'm investigating here. R hat is ... Auto-correlation is... How are they computed? What do they say? How are they computed on MI data? Looking at imputation chains. 
-Explain that there is no baseline to compare the measures with. The only current convergence check is to visually inspect trace plots. (But in the trace-plots convergence seems immediate).}
-
-The potential scale reduction factor $\widehat{R}$ tells us by how much the variance of an estimate could be shrunken down if the chains were to be infinitely long \cite{gelm92}.
+The convergence diagnostics under consideration are $\widehat{R}$ and auto-correlation.
+Perfect convergence is reached when the variance between imputation chains is equivalent to the variance within chains ($\widehat{R} = 1$), and there is no dependency between subsequent samples within imputation chains (auto-correlation $= 0$).
+%
+%%%%%%%%%%%%%%%%%
+%As potential scale reduction factor, $\widehat{R}$ tells us by how much the variance of an estimate could be shrunken down if the chains were to be infinitely long \citep[$t \to \infty$;][]{gelm92}.
+If there is no added value compared to infinitely many iterations, the chains have converged.
+%%%%%%%%%%%%%%%%%%
+%
+%$\textbf{Explain the measures that I'm investigating here.
+R hat is ...
+Auto-correlation is...
+How are they computed?
+What do they say?
+How are they computed on MI data?
+Looking at imputation chains.
+%Explain that there is no baseline to compare the measures with. The only current convergence check is to visually inspect trace plots. (But in the trace-plots convergence seems immediate).}
+%
+%The potential scale reduction factor $\widehat{R}$ tells us by how much the variance of an estimate could be shrunken down if the chains were to be infinitely long \cite{gelm92}.
 That then informs us about how dependent the chains are on the starting values.
 If there is no dependence on the initial values anymore, the chains have converged (in the mixing sense of the word).
-$\widehat{R}$ is then equal to one.
-The conventionally acceptable threshold was $\widehat{R} < 1.2$.
-More recently, \cite{veht19} proposed a more stringent threshold: $\widehat{R} < 1.01$.
-
+%$\widehat{R}$ is then equal to one. 
+%
 %We can apply $\widehat{R}$ to the mean (or to the first two moments) of the variables of interest. 
-$\widehat{R}$ is defined by \cite{veht19} as follows: % The scalar summary of interest in this study is the average of the imputed values per variable.
+To define $\widehat{R}$, we follow notation by \citep[p.~5]{veht19}.
+% The scalar summary of interest in this study is the average of the imputed values per variable.
 %As recommended by \cite{veht19} $\widehat{R}$ is computed as ....
@@
 \cite[p.~5]{veth19}
-
-\textbf{REPHRASE: In the equations below, $N$ is the number of draws per chain, $M$ is the number of chains, and $S = MN$ is the total number of draws from all chains.
+Let $M$ be the number of chains, $N$ the number of iterations per chain, and $\theta$ the scalar summary of interest (e.g., chain mean or chain variance).
+We compute the variance within each chain of $\theta$ values ($W$) and between the $m$ chains ($B$) as follows:
+%\textbf{REPHRASE: In the equations below, $N$ is the number of draws per chain, $M$ is the number of chains, and $S = MN$ is the total number of draws from all chains.
 For each scalar summary of interest $\theta$, we compute $B$ and $W$, the between- and within-chain variances:}
@@
 \begin{align*}
-``B&=\frac{N}{M-1} \sum_{m=1}^{M}\left(\bar{\theta}^{(\cdot m)}-\bar{\theta}^{(\cdot \cdot)}\right)^{2}, \text { where } \bar{\theta}^{(\cdot m)}=\frac{1}{N} \sum_{n=1}^{N} \theta^{(n m)}, \quad \bar{\theta}^{(\cdot \cdot)}=\frac{1}{M} \sum_{m=1}^{M} \bar{\theta}^{(\cdot m)} \\
-W&=\frac{1}{M} \sum_{m=1}^{M} s_{j}^{2},  \text { where } s_{m}^{2}=\frac{1}{N-1} \sum_{n=1}^{N}\left(\theta^{(n m)}-\bar{\theta}^{(\cdot m)}\right)^{2}". \text{ \cite[p.~5]{veht19}} 
+B&=\frac{N}{M-1} \sum_{m=1}^{M}\left(\bar{\theta}^{(\cdot m)}-\bar{\theta}^{(\cdot \cdot)}\right)^{2}, \text { where } \bar{\theta}^{(\cdot m)}=\frac{1}{N} \sum_{n=1}^{N} \theta^{(n m)}, \quad \bar{\theta}^{(\cdot \cdot)}=\frac{1}{M} \sum_{m=1}^{M} \bar{\theta}^{(\cdot m)} \\
+W&=\frac{1}{M} \sum_{m=1}^{M} s_{j}^{2},  \text { where } s_{m}^{2}=\frac{1}{N-1} \sum_{n=1}^{N}\left(\theta^{(n m)}-\bar{\theta}^{(\cdot m)}\right)^{2}. %\text{ \cite[p.~5]{veht19}} 
 \end{align*}
 
-The proportion of within chain variance $W$ against the weighted average of $B$ and $W$, $\widehat{\operatorname{var}}^{+}$ tells us how much variance of $\theta$ could be shrunken down if $N\to\infty$.
+From the between and within chain variances we compute a weighted average, $\widehat{\operatorname{var}}^{+}$, which over-estimates the total variance of $\theta$.
+$\widehat{R}$ is then obtained as a ratio between the over-estimated total variance and the within chain variance:
+
+\begin{equation*}
+\widehat{R}=\sqrt{\frac{\widehat{\operatorname{var}}^{+}(\theta | y)}{W}},
+\text{ where } \widehat{\operatorname{var}}^{+}(\theta | y)=\frac{N-1}{N} W+\frac{1}{N} B.
+\end{equation*}
+
+%The proportion of within chain variance $W$ against the weighted average of $B$ and $W$, $\widehat{\operatorname{var}}^{+}$ tells us how much variance of $\theta$ could be shrunken down if $N\to\infty$.
 $\widehat{\operatorname{var}}^{+}$ is computed as follows: %over-estimates the marginal posterior variance of the estimand, $\theta$.
 
-\begin{equation*}
-\widehat{\operatorname{var}}^{+}(\theta | y)=\frac{N-1}{N} W+\frac{1}{N} B.
-\end{equation*}
+% \begin{equation*}
+% 
+% \end{equation*}
 
-The potential scale reduction factor $\widehat{R}$ is obtained as:
-
-\begin{equation*}
-\widehat{R}=\sqrt{\frac{\widehat{\operatorname{var}}^{+}(\theta | y)}{W}}.
-\end{equation*}
+%The potential scale reduction factor $\widehat{R}$ is obtained as:
 
 High $\widehat{R}$ would indicate .... low $\widehat{R}$ values indicate ....
+The conventionally acceptable threshold was $\widehat{R} < 1.2$ \cite{gelm92}. More recently, \cite{veht19} proposed a more stringent threshold: $\widehat{R} < 1.01$.
 
-The second component of convergence stationarity requires chains to be stable across iterations, or non-recurring.
 Non-recurrence can be evaluated with auto-correlation.
@@
 The primary research interest is in determining whether $\widehat{R}$ is an appropriate convergence diagnostic, and if so, which level of stringency suits MI data.
-\textbf{Or if there is a discrepancy between the visual and numeric inspection.}
+\textbf{Or if there is a discrepancy between the visual and numeric inspection.
+There is no baseline to compare the convergence diagnostics with.
+Therefore, we will only evaluate performance measures.}
 
@@
 
-This paper is written by the sole author (Hanne Oberman, BSc.), with guidance from Master thesis supervisors prof. dr. Stef van Buuren, and dr. Gerko Vink.
+This note is part of a Research Master's thesis project.
+It was written by the sole author (Hanne Oberman, BSc.), with guidance from Master thesis supervisors prof.
+dr.
+Stef van Buuren and dr.
+Gerko Vink, and research seminar mentor dr.
+Daniela Cianci.
 

### 2019-12-12 Simulation/Writeup/NoteConvergence.Rnw
theoretical background restructured

@@
 The validity of inferences may be threatened by non-convergence of this iterative algorithmic procedure, but diagnostics to evaluate algorithmic convergence have not been implemented in MI software.
-This note investigates the performance of two convergence diagnostics (\citep[$\widehat{R}$,][]{gelm92}; \citep[auto-correlation,][]{scha97}) through model-based simulation.
+This note investigates the performance of two convergence diagnostics (\citep[$\widehat{R}$,][]{gelm92}; \citep[autocorrelation,][]{scha97}) through model-based simulation.
 To investigate convergence of the algorithm, the number of iterations is varied between one and one-hundred.
@@
 %More specifically, this note focuses on the MI algorithm that is implemented in the \proglang{R} package \pkg{mice}: `Multiple Imputation using Chained Equations' (MICE) \textbf{[can be removed, can be found in the mice ref]}.
-To this end we will evaluate how convergence of the imputation algorithm can be diagnosed.
+To this end, we will evaluate how convergence of the imputation algorithm can be diagnosed.
 
@@
 
-The convergence properties of the MI algoritm in \pkg{mice} are investigated through model-based simulation.
+The convergence properties of the MI algorithm in \pkg{mice} are investigated through model-based simulation.
 The results of this simulation study are guidelines for assessing convergence of MI algorithms.
@@
 %I will replicate Lacerda et al.'s simulation study on $\widehat{R}$ \citep{lace07}, and develop novel guidelines for assessing convergence.
-Ideally, I will integrate several diagnostics (e.g., $\widehat{R}$, \emph{auto-correlation}, and \emph{simulation error}) into a single summary indicator to flag non-convergence.
+Ideally, I will integrate several diagnostics (e.g., $\widehat{R}$, \emph{autocorrelation}, and \emph{simulation error}) into a single summary indicator to flag non-convergence.
 
@@
 The multiple imputation algorithm in \pkg{mice} is a special case of Markov chain Monte Carlo (MCMC) methods--a framework of iterative algorithmic procedures.
-While methodology for diagnosing convergence of MI algorithms has not been systematically studied \citep{buur18}, several convergence diagnostics have been proposed for other types of MCMC methods (e.g., Gibbs samplers) and the framework in general.
+While methodology for diagnosing convergence of MI algorithms has not been systematically studied \citep{buur18}, several convergence diagnostics have been proposed for other types of MCMC methods and the framework in general.
 % For the framework of itrIt is not known whether convergence diagnostics that are available for the general framework that MICE is a special case of are appropriate for the evaluation of MI data.
@@
 
-Loosely speaking, the aim of MCMC algorithms (e.g., Gibbs samplers, or the MI algorithm in \pkg{mice}) is to sample values from an unknown target distribution (e.g., the posterior distribution of a parameter in Bayesian inference, or a distribution of missing values, respectively).
+Loosely speaking, the aim of MCMC algorithms is to sample values from an unknown target distribution (e.g., the posterior distribution of a parameter in Gibbs samplers, or the distribution of missing values in an MI algorithm).
 Convergence is reached once the algorithm samples exclusively from the target distribution \citep{cowl96}.
@@
 Most convergence diagnostics target either of the two components.
-The stationarity component of convergence may be evaluated with auto-correlation \citep{scha97, gelm13}, numeric standard error \citep[which is commonly called `MC error']{gewe92}, or Raftery and Lewis's (\citeyear{raft91}) procedure to determine how many iterations are necessary to deminish the effect of trending within chains.
+The stationarity component of convergence may be evaluated with autocorrelation \citep[$AC$][]{scha97, gelm13}, numeric standard error \citep[or `MC error'][]{gewe92}, and Raftery and Lewis's (\citeyear{raft91}) procedure to determine how many iterations are necessary to diminish the effect of trending within chains.
 The mixing component can be assessed with the potential scale reduction factor $\widehat{R}$ \citep[widely referred to as `the Gelman-Rubin statistic']{gelm92}.
@@
 This would make $\widehat{R}$ a general convergence diagnostic.
-The application of $\widehat{R}$ to assess stationarity has not been thourougly investigated.
-Therefore, this study employs both $\widehat{R}$ and auto-correlation to investigate algorithmic convergence, as recommended by \cite[p.~898]{cowl96}.
-%Therefore, the focus of this study is mainly on $\widehat{R}$. Stationarity will also be evaluated with auto-correlation as diagnostic, as recommended by \cite[p.~898]{cowl96}. 
+The application of $\widehat{R}$ to assess stationarity has not been thoroughly investigated.
+Therefore, this study employs both $\widehat{R}$ and autocorrelation to investigate algorithmic convergence, as recommended by \cite[p.~898]{cowl96}.
+%Therefore, the focus of this study is mainly on $\widehat{R}$. Stationarity will also be evaluated with autocorrelation as diagnostic, as recommended by \cite[p.~898]{cowl96}. 
 Other convergence diagnostics are outside the scope of this study.\footnotemark 
@@
 This method is outside of the scope of the current study.}%  he eigenvector decomposition method proposed by McKay \citep{XYZ}.} These four methods, however, are outside of the scope of this study.
-Here, the focus is mainly on $\widehat{R}$ and auto-correlation.
+Here, the focus is mainly on $\widehat{R}$ and autocorrelation.
 
 %Diagnose non-convergence in the mixing sense: $\widehat{R}$.
-Diagnose non-convergence in the stability sense: auto-correlation (AC).
+Diagnose non-convergence in the stability sense: autocorrelation (AC).
 Also, it is possible to compute MC errors, the Geweke statistic \textbf{\citep{XYZ}}, or to determine the appropriate number of iterations for convergence by employing \textbf{Raftery and Lewis's (1992)} method.
@@
 These four methods, however, are outside of the scope of this study.
-Here, the focus is mainly on $\widehat{R}$ and auto-correlation.}
-
-It is not known whether any convergence diagnostic is appropriate to diagnose convergence of \pkg{mice} algorithms.
-As by \cite[\S~6.5.2]{buur18}, the current practice is to visually inspect imputation chains for signs of non-mixing and non-stationarity.
-Visually assessing convergence, however, may be challenging to the untrained eye.
-\textbf{[This is problematic because]} There is no scientific consensus on the convergence properties of multiple imputation algorithms \citep{taka17}, and in some situations, the default techniques in \pkg{mice} might not converge to stable target distributions at all \citep{murr18}.
-%Convergence diagnostics for MI methods have not been systematically studied \citep{buur18}.
-Moreover,
-%A fundamental assumption of MICE is convergence of the algorithm.
-MICE is a type of Markov chain Monte Carlo (MCMC) algorithm.
-In general, the validity of inference resulting from MCMC algorithms is threatened by non-convergence.
-Hence we use convergence diagnostics to flag non-convergence.
-But it is not known whether conventional convergence diagnostics for MCMC methods work on MI data.
-Therefore, convergence of the \pkg{mice} algorithms should be monitored carefully. %the need for convergence diagnostics is extra pressing in the context of the MICE algorithm.
+Here, the focus is mainly on $\widehat{R}$ and autocorrelation.}
 
@@
 
-The convergence diagnostics under consideration are $\widehat{R}$ and auto-correlation.
-Perfect convergence is reached when the variance between imputation chains is equivalent to the variance within chains ($\widehat{R} = 1$), and there is no dependency between subsequent samples within imputation chains (auto-correlation $= 0$).
+The convergence diagnostics under consideration are $\widehat{R}$ and autocorrelation.
+Perfect convergence is reached when the variance between imputation chains is equivalent to the variance within chains ($\widehat{R} = 1$), and there is no dependency between subsequent samples within imputation chains ($AC = 0$).
 %
@@
 %
-%$\textbf{Explain the measures that I'm investigating here.
-R hat is ...
-Auto-correlation is...
-How are they computed?
-What do they say?
-How are they computed on MI data?
-Looking at imputation chains.
+%$\textbf{Explain the measures that I'm investigating here. R hat is ... autocorrelation is... How are they computed? What do they say? How are they computed on MI data? Looking at imputation chains. 
 %Explain that there is no baseline to compare the measures with. The only current convergence check is to visually inspect trace plots. (But in the trace-plots convergence seems immediate).}
@@
 %The potential scale reduction factor $\widehat{R}$ is obtained as:
-
-High $\widehat{R}$ would indicate .... low $\widehat{R}$ values indicate ....
+We can interpret $\widehat{R}$ as potential scale reduction factor, since it indicates by how much the variance of $\theta$ could be shrunken down if an infinite number of iterations per chain would be run \citep[$t \to \infty$;][]{gelm92}.
+High $\widehat{R}$-values would therefore indicate slow mixing of chains, e.g., because of dependence across chains on initial values.
+Once there is no dependence on initial values anymore, there is no added value of continuing iterations.
+The measure then diagnoses algorithmic convergence ($\widehat{R} = 1$).
+%If there is no added value compared to infinitely many iterations, the chains have converged. 
 The conventionally acceptable threshold was $\widehat{R} < 1.2$ \cite{gelm92}. More recently, \cite{veht19} proposed a more stringent threshold: $\widehat{R} < 1.01$.
 
-Non-recurrence can be evaluated with auto-correlation.
-Auto-correlation shows how dependent subsequent draws of an imputation chain are on the previous value.
+Following the same notation, autocorrelation is defined as the correlation between two subsequent $\theta$ values within the same chain.
+We only consider $AC$ at lag 1, i.e., between the $t^{th}$ and $t+1^{th}$ iteration of each chain:
+%
+\begin{equation*}
+AC = \left( \frac{T}{T-1} \right) \frac{\sum_{t=1}^{T-1}(\theta_t - \bar{\theta}^{(\cdot m)})(\theta_{t+1} - \bar{\theta}^{(\cdot m)})}{\sum_{t=1}^{T}(\theta_t - \bar{\theta}^{(\cdot m)})^2},
+\end{equation*}
+%
+where $T$ is the total number of iterations per chain \citep[p.~147]{lync07}.
+We can interpret the sign of $AC$-values as measure of dependence within chains ($-1 \leq AC \leq 1$).
+Positive $AC$-values would indicate recurrence: similar $\theta$-values for subsequent iterations, which may lead to trending.
+Negative $AC$-values also indicate dependence, but do not threaten the stationarity component of convergence.
+Instead of recurrence, there would be a `seesaw' effect within chains: subsequent $\theta$-values would deviate in opposite directions from the chain average $\theta^{(\cdot m)})$, increasing the variance of $\theta$.
+As convergence diagnostic, the interest is in positive $AC$-values.\footnotemark
+\footnotetext{The magnitude of $AC$-values can also be evaluated statistically, but that is outside of this note's scope.}
+
+%, where $\theta$ is a scalar summary of interest (i.e., chain means and chain variances).
+%Non-recurrence can be evaluated with autocorrelation.
+autocorrelation shows how dependent subsequent draws of an imputation chain are on the previous value.
 If there is a lot of dependence, draws at e.g.
 iteration five are significantly correlated with the value of the first draw.
-A high auto-correlation indicates dependence within chains.
+%A high autocorrelation indicates dependence within chains.
 The magnitude of the AC can be interpreted qualitatively or quantitatively.
@@
 
-\textbf{Convergence diagnostics applied to MI}
-
-$\widehat{R}$ may not be an appropriate diagnostic to evaluate MI data because it assumes over-dispersed initial values.
-This means that the initial values of the $m$ imputation chains are `far away' from the distribution that the chains are converging to.
-In the \pkg{mice} algorithm, initial values are chosen randomly from the observed data.
+It is not known whether $\widehat{R}$ and $AC$ are appropriate measures to diagnose convergence of MI algorithms.
+Assessing the stationarity component of convergence with $AC$ might be redundant, since high $AC$-values are implausible in MI procedures.
+That is, the randomness induced by the MI algorithm effectively mitigates the risk of dependency within chains.
+$AC$ would thus not be informative of the convergence of MI algorithms.
+$\widehat{R}$ may not be an appropriate diagnostic either, because it assumes over-dispersed initial values (i.e., the starting points of the chains are `far away' from the target distribution and each other).
+In \pkg{mice}, initial values of the algorithm are chosen randomly from the observed data.
 Therefore, we cannot be certain that the initial values are over-dispersed.
@@
 Daarom is het lastig om in MI that starten vanuit overdisp state.
-
 % Starting values of MICE chains: "For each j, fill in starting imputations $Y^0_j$ by random draws from $Y^{obs}_j$." \cite[\S~4.5.2]{buur18} before for loops over variables and imputations, see \url{https://stefvanbuuren.name/fimd/sec-FCS.html#def:mice}
-
 Without over-dispersed initial states, $\widehat{R}$ may falsely diagnose convergence \citep{broo98}.
@@
 This suggests that $\widehat{R}$ would not be sensitive enough to flag non-convergence of MI algorithms.
-An empirical finding, however, shows that the opposite may be true: \cite{lace07} report $\widehat{R}$ values above the threshold of $\widehat{R} < 1.1$ after fifty iterations.
+An empirical finding, on the other hand, shows that $\widehat{R}$ can indicate non-convergence at unconventionally long chain lengths.
+While the default number of iterations in \pkg{mice} is five, $\widehat{R} > 1.1$ was observed after 50 iterations, and $\widehat{R} > 1.01$ up-to 150 iterations.
+This suggests that $\widehat{R}$ would over-estimate non-convergence of MI algorithms.
+
+% the conventional threshold $\widehat{R} < 1.1$ might be too stringent. \cite[pp.~49-50]{lace07} report
+% 
+% $\widehat{R}$ is sensitive enough to indicate non-convergence can  reported $\widehat{R}$-values  
+% 
+% An empirical finding, however, shows that the opposite may be true: \cite{lace07} report $\widehat{R}$-values above the threshold of $\widehat{R} < 1.1$ after fifty iterations.
 %that $\widehat{R}$ will not be smaller than 1.1 before iteration number 50.
 Therefore, the aim is to replicate empirical finding Lacerda et al.
+%
 
@@
 I will replicate Lacerda et al.'s simulation study on $\widehat{R}$ \citep{lace07}, and develop novel guidelines for assessing convergence.
-Ideally, I will integrate several diagnostics (e.g., $\widehat{R}$, \emph{auto-correlation}, and \emph{simulation error}) into a single summary indicator to flag non-convergence.
+Ideally, I will integrate several diagnostics (e.g., $\widehat{R}$, \emph{autocorrelation}, and \emph{simulation error}) into a single summary indicator to flag non-convergence.
 
@@
 
-The aim is to evaluate whether the imputation procedure has converged.
-The primary research interest is in determining whether $\widehat{R}$ is an appropriate convergence diagnostic, and if so, which level of stringency suits MI data.
+This study evaluates whether $\widehat{R}$ and $AC$ can diagnose when the multiple imputation algorithm has converged.
+There is, however, no baseline measure available to evaluate their performance against.
+The current practice of assessing convergence is to visually inspect imputation chains for signs of non-mixing and non-stationarity \citep[\S~6.5.2]{buur18}.
+Visually assessing convergence, however, may be challenging to the untrained eye.
+\textbf{[This is problematic because.
+MAYBE REPLACE THIS PART!]} There is no scientific consensus on the convergence properties of multiple imputation algorithms \citep{taka17}, and in some situations, the default techniques in \pkg{mice} might not converge to stable target distributions at all \citep{murr18}.
+%Convergence diagnostics for MI methods have not been systematically studied \citep{buur18}.
+Moreover,
+%A fundamental assumption of MICE is convergence of the algorithm.
+MICE is a type of Markov chain Monte Carlo (MCMC) algorithm.
+In general, the validity of inference resulting from MCMC algorithms is threatened by non-convergence.
+Hence we use convergence diagnostics to flag non-convergence.
+But it is not known whether conventional convergence diagnostics for MCMC methods work on MI data.
+Therefore, convergence of the MI algorithms in \pkg{mice} should be monitored carefully. %the need for convergence diagnostics is extra pressing in the context of the MICE algorithm.
+
+%The primary research interest is in determining whether $\widehat{R}$ is an appropriate convergence diagnostic, and if so, which level of stringency suits MI data.
 \textbf{Or if there is a discrepancy between the visual and numeric inspection.
@@
 This simulation study hypothesizes that $\widehat{R}$ will over-estimate non-convergence.
-\textbf{[Or use:] Hypothesis based on \cite{lace07} is that the conventional acceptable level of $\widehat{R}$ is too strict for MI data.
+%\textbf{[Or use:] Hypothesis based on \cite{lace07} is that the conventional acceptable level of $\widehat{R}$ is too strict for MI data.
 We expect that the simulation diagnostics will indicate valid inference before $\widehat{R}$ will.}
@@
 % 
-% - Auto-correlation.
+% - autocorrelation.
 Schafer (1997, p.
 129) wrote on worst linear statistic.
-We could calculate the auto-correlation of that statistic to know that the algorithm converged elsewhere too.
+We could calculate the autocorrelation of that statistic to know that the algorithm converged elsewhere too.
 See autocorr function plot in SAS of worst linear function.
@@
 % % "Applications of MICE with lowly correlated data therefore inject a lot of noise into the system.
-Hence, the auto-correlation over t will be low, and convergence will be rapid, and in fact immediate if all variables are independent.
+Hence, the autocorrelation over t will be low, and convergence will be rapid, and in fact immediate if all variables are independent.
 Thus, the incorporation of noise into the imputed data has pleasant side-effect of speeding up convergence" \citep{buur18}, par.
@@
 Each simulation condition has a different maximum number of iterations, from one to one-hundred iterations.
-We compute $\widehat{R}$ and auto-correlation (`convergence diagnostics') and performance measures (`simulation diagnostics') for each condition.
+We compute $\widehat{R}$ and autocorrelation (`convergence diagnostics') and performance measures (`simulation diagnostics') for each condition.
 The number of simulation runs per condition is 1000.
@@
 The maximum value across variables within the same simulation is reported.
-Maximum $\widehat{R}$ values per simulation condition are aggregated across simulations.
+Maximum $\widehat{R}$-values per simulation condition are aggregated across simulations.
 
-Auto-correlation (AC) is computed within imputation chains, as the correlation between $\theta$ at \textbf{the $i^{th}$ iteration and $\theta$ at the $i+1^{th}$ iteration [CHECK VARIABLE IN NOTATION]}, where $\theta$ is a scalar summary of interest (i.e., chain means and chain variances).
-The AC of the variable with the highest absolute AC value is reported per simulation.
+autocorrelation (AC) is computed within imputation chains,  The AC of the variable with the highest absolute AC value is reported per simulation.
 These values are then averaged per simulation condition.
@@
 A subset of simulation conditions is presented in Table \ref{tab:results}.
-This section describes convergence diagnostics ($\widehat{R}$ and auto-correlation) and simulation diagnostics (bias, confidence interval width and coverage rate) first separately, and then together.
+This section describes convergence diagnostics ($\widehat{R}$ and autocorrelation) and simulation diagnostics (bias, confidence interval width and coverage rate) first separately, and then together.
 
@@
 \textbf{$\widehat{R}$}.
-The $\widehat{R}$ values for chain means and chain variances show equivalent trends, neither is there a clear difference between the four variables.
+The $\widehat{R}$-values for chain means and chain variances show equivalent trends, neither is there a clear difference between the four variables.
 Roughly speaking, we see that $\widehat{R}$ is higher for simulation conditions with a lower number of iterations, and approaches one for conditions with more iterations.
@@
 This decline \textbf{houdt aan} up-to iteration twenty, after which a more gradual decrease can be observed up-to iteration forty.
-The $\widehat{R}$ values are more or less stable after that.
+The $\widehat{R}$-values are more or less stable after that.
 The \textbf{conventionally acceptable threshold for convergence, maximum $\widehat{R} < 1.2$,[CHECK IN INTRO]} is reached in the simulation condition with three iterations, and conditions with six iterations or more.
@@
 
-\textbf{Auto-correlation.} All observed auto-correlations (aggregated over the 1000 repetitions) are negative or close to zero.
+\textbf{autocorrelation.} All observed autocorrelations (aggregated over the 1000 repetitions) are negative or close to zero.
 There are some differences between variables, and the ACs of chain variances seem lower than the ACs of chain means.
@@
 Simulation conditions with a higher number of iterations show a slow increase towards zero.
-Auto-correlations do not plateau completely within the range of simulation conditions considered in this study.
+autocorrelations do not plateau completely within the range of simulation conditions considered in this study.
 It seems that ACs might become positive if more iterations would be allowed.
@@
 If we would use 1.2, we might miss this increase after three iterations.
-The 1.1 threshold seems ok.
+The 1.1 threshold seems OK.
 Also, ten iterations are computationally not terrible.
@@
 From the simulation diagnostics, it appears that as little as three iterations might be sufficient to obtain unbiased, confidence valid estimates.
-Convergence diagnostics $\widehat{R}$ and auto-correlation, however, indicate that convergence may only be reached after twenty or even forty iterations.
+Convergence diagnostics $\widehat{R}$ and autocorrelation, however, indicate that convergence may only be reached after twenty or even forty iterations.
 
@@
 This can happen when the number of simulations is smaller than in `regular' MCMC processes [explain that fewer iterations is an advantage of MI, not a disadvantage compared to MCMC].
-Therefore, the `$(n-1/n)$' [add equation number] correction factor can influence the estimated potential scale reduction factor.
+Increasing Rhat values mean that the initial values were not appropriately over-dispersed \citep[p~438]{broo98}.
+%Therefore, the `$(n-1/n)$' [add equation number] correction factor can influence the estimated potential scale reduction factor.
 This downwards bias is in the opposite direction than expected: ``The mixture-of-sequences variance, $V$, should stabilize as a function of $n$.
@@
 
-\textbf{Negative ACs.} Auto-correlation is dangerous when positive.
-These auto-correlations are negative.
+\textbf{Negative ACs.} autocorrelation is dangerous when positive.
+These autocorrelations are negative.
 Still, we want the iterations to be stable and independent.
@@
 
-\textbf{Future research.} This study consdered only a MCAR missingness mechanism.
+\textbf{Future research.} This study considered only a MCAR missingness mechanism.
 Necessary but not sufficient to demonstrate the performance of convergence diagnostics.
 This is just a proof of concept.
-Further reseach is needed to investigate the performance under violation of convergence, e.g.
+Further research is needed to investigate the performance under violation of convergence, e.g.
 dependency between predictors (very high correlations).

### 2019-12-12 Simulation/Writeup/NoteConvergence.tex
theoretical background restructured

@@
 Most convergence diagnostics target either of the two components.
-The stationarity component of convergence may be evaluated with auto-correlation \citep{scha97, gelm13}, numeric standard error \citep[which is commonly called `MC error']{gewe92}, or Raftery and Lewis's (\citeyear{raft91}) procedure to determine how many iterations are necessary to deminish the effect of trending within chains.
+The stationarity component of convergence may be evaluated with autocorrelation \citep[$AC$][]{scha97, gelm13}, numeric standard error \citep[or `MC error'][]{gewe92}, and Raftery and Lewis's (\citeyear{raft91}) procedure to determine how many iterations are necessary to deminish the effect of trending within chains.
 The mixing component can be assessed with the potential scale reduction factor $\widehat{R}$ \citep[widely referred to as `the Gelman-Rubin statistic']{gelm92}.
@@
 The application of $\widehat{R}$ to assess stationarity has not been thourougly investigated.
-Therefore, this study employs both $\widehat{R}$ and auto-correlation to investigate algorithmic convergence, as recommended by \cite[p.~898]{cowl96}.
-%Therefore, the focus of this study is mainly on $\widehat{R}$. Stationarity will also be evaluated with auto-correlation as diagnostic, as recommended by \cite[p.~898]{cowl96}. 
+Therefore, this study employs both $\widehat{R}$ and autocorrelation to investigate algorithmic convergence, as recommended by \cite[p.~898]{cowl96}.
+%Therefore, the focus of this study is mainly on $\widehat{R}$. Stationarity will also be evaluated with autocorrelation as diagnostic, as recommended by \cite[p.~898]{cowl96}. 
 Other convergence diagnostics are outside the scope of this study.\footnotemark 
@@
 %The potential scale reduction factor $\widehat{R}$ is obtained as:
-
-High $\widehat{R}$ would indicate .... low $\widehat{R}$ values indicate ....
+We can interpret $\widehat{R}$ as potential scale reduction factor, since it indicates by how much the variance of $\theta$ could be shrunken down if an infinite number of iterations per chain would be run \citep[$t \to \infty$;][]{gelm92}.
+High $\widehat{R}$-values would therefore indicate slow mixing of chains, e.g., because of dependence across chains on initial values.
+Once there is no dependence on initial values anymore, there is no added value of continuing iterations.
+The measure then diagnoses algorithmic convergence ($\widehat{R} = 1$).
+%If there is no added value compared to infinitely many iterations, the chains have converged. 
 The conventionally acceptable threshold was $\widehat{R} < 1.2$ \cite{gelm92}. More recently, \cite{veht19} proposed a more stringent threshold: $\widehat{R} < 1.01$.
 
-Non-recurrence can be evaluated with auto-correlation.
-Auto-correlation shows how dependent subsequent draws of an imputation chain are on the previous value.
+Following the same notation, autocorrelation is defined as the correlation between two subsequent $\theta$ values within the same chain.
+We only consider $AC$ at lag 1, i.e., between the $t^{th}$ and $t+1^{th}$ iteration of each chain:
+%
+\begin{equation*}
+AC = \left( \frac{T}{T-1} \right) \frac{\sum_{t=1}^{T-1}(\theta_t - \bar{\theta}^{(\cdot m)})(\theta_{t+1} - \bar{\theta}^{(\cdot m)})}{\sum_{t=1}^{T}(\theta_t - \bar{\theta}^{(\cdot m)})^2},
+\end{equation*}
+%
+where $T$ is the total number of iterations per chain \citep[p.~147]{lync07}.
+We can interpret the sign of $AC$-values as measure of dependence within chains ($-1 \leq AC \leq 1$).
+Positive $AC$-values would indicate recurrence: similar $\theta$-values for subsequent iterations, which may lead to trending.
+Negative $AC$-values also indicate dependence, but do not threathen the stationarity component of convergence.
+Instead of recurrence, there would be a `seesaw' effect within chains: subsequent $\theta$-values would deviate in opposite directions from the chain average $\theta^{(\cdot m)})$, increasing the variance of $\theta$.
+As convergence diagnostic, the interest is in positive $AC$-values.\footnotemark
+\footnotetext{The magnitude of $AC$-values can also be evaluated statistically, but that is outside of this note's scope.}
+
+%, where $\theta$ is a scalar summary of interest (i.e., chain means and chain variances).
+%Non-recurrence can be evaluated with autocorrelation.
+autocorrelation shows how dependent subsequent draws of an imputation chain are on the previous value.
 If there is a lot of dependence, draws at e.g.
 iteration five are significantly correlated with the value of the first draw.
-A high auto-correlation indicates dependence within chains.
+%A high autocorrelation indicates dependence within chains.
 The magnitude of the AC can be interpreted qualitatively or quantitatively.
@@
 This suggests that $\widehat{R}$ would not be sensitive enough to flag non-convergence of MI algorithms.
-An empirical finding, however, shows that the opposite may be true: \cite{lace07} report $\widehat{R}$ values above the threshold of $\widehat{R} < 1.1$ after fifty iterations.
+An empirical finding, on the other hand, shows that $\widehat{R}$ can indicate non-convergence at unconventionally long chain lengths.
+While the default number of iterations in \pkg{mice} is five, $\widehat{R} > 1.1$ was observed after 50 iterations, and $\widehat{R} > 1.01$ upto 150 iterations.
+This suggests that $\widehat{R}$ would over-estimate non-convergence of MI algorithms.
+
+% the conventional threshold $\widehat{R} < 1.1$ might be too stringent. \cite[pp.~49-50]{lace07} report
+% 
+% $\widehat{R}$ is sensitive enough to indicate non-convergence can  reported $\widehat{R}$-values  
+% 
+% An empirical finding, however, shows that the opposite may be true: \cite{lace07} report $\widehat{R}$-values above the threshold of $\widehat{R} < 1.1$ after fifty iterations.
 %that $\widehat{R}$ will not be smaller than 1.1 before iteration number 50.
 Therefore, the aim is to replicate empirical finding Lacerda et al.
+%
 
@@
 
-The aim is to evaluate whether the imputation procedure has converged.
-The primary research interest is in determining whether $\widehat{R}$ is an appropriate convergence diagnostic, and if so, which level of stringency suits MI data.
+This study evaluates whether $\widehat{R}$ and $AC$ can diagnose when the multiple imputation algorithm has converged.
+There is, however, no baseline measure available to evaluatate their performance against.
+The current practice of assessing convergence is to visually inspect imputation chains for signs of non-mixing and non-stationarity \citep[\S~6.5.2]{buur18}.
+Visually assessing convergence, however, may be challenging to the untrained eye.
+\textbf{[This is problematic because.
+MAYBE REPLACE THIS PART!]} There is no scientific consensus on the convergence properties of multiple imputation algorithms \citep{taka17}, and in some situations, the default techniques in \pkg{mice} might not converge to stable target distributions at all \citep{murr18}.
+%Convergence diagnostics for MI methods have not been systematically studied \citep{buur18}.
+Moreover,
+%A fundamental assumption of MICE is convergence of the algorithm.
+MICE is a type of Markov chain Monte Carlo (MCMC) algorithm.
+In general, the validity of inference resulting from MCMC algorithms is threatened by non-convergence.
+Hence we use convergence diagnostics to flag non-convergence.
+But it is not known whether conventional convergence diagnostics for MCMC methods work on MI data.
+Therefore, convergence of the MI algorithms in \pkg{mice} should be monitored carefully. %the need for convergence diagnostics is extra pressing in the context of the MICE algorithm.
+
+%The primary research interest is in determining whether $\widehat{R}$ is an appropriate convergence diagnostic, and if so, which level of stringency suits MI data.
 \textbf{Or if there is a discrepancy between the visual and numeric inspection.

### 2019-12-12 Simulation/Writeup/NoteConvergence.Rnw
Comments Rosalie verwerkt, nog steeds te veel woorden...

@@
 To obtain valid inferences with MI, the variability between imputations should be properly represented \citep{rubin87, buur18}.
-If the variance between imputations is under-estimated, confidence intervals around estimates will be too narrow, which can yield spurious results.
-On the other hand, over-estimation of the variance between imputations results in unnecessarily wide confidence intervals, which can be costly because it lowers the statistical power.
+If this variability is under-estimated, confidence intervals around estimates will be too narrow, which can yield spurious results.
+However, over-estimation of the variance between imputations results in unnecessarily wide confidence intervals, which can be costly because it lowers the statistical power.
 Since both of these situations are undesirable, imputations and their variability should be evaluated.
@@
 The goal of this research project is to develop novel methodology and guidelines for evaluating MI methods.
-These evaluation measures and guidelines will subsequently be implemented in an interactive evaluation framework for multiple imputation to aid applied researchers in drawing valid inference from incomplete datasets.
+These tools will subsequently be implemented in an interactive evaluation framework for multiple imputation to aid applied researchers in drawing valid inference from incomplete datasets.
 This note provides the theoretical foundation towards the diagnostic evaluation of the convergence of MI algorithms.
@@
 The collection of observed data values in $Y$ is denoted as $Y_{obs}$; the missing part of $Y$ is referred to as $Y_{mis}$.
-Response indicator $R$ shows whether a data value in $Y$ is missing or observed.
-The relation between $R$, $Y_{obs}$, and $Y_{mis}$ determines the missingness mechanism.
+\textbf{REMOVE?
+Response indicator $R$ shows whether a data value in $Y$ is missing or observed.} The relation between $R$, $Y_{obs}$, and $Y_{mis}$ determines the missingness mechanism.
 This note only considers a `missing completely at random' (MCAR) mechanism, where the probability of being missing is equal for all $n \times p$ cells in $Y$.
 
-In this note, the terms `unobserved' and `missing' data are used interchangeably.
-Both refer to $Y_{mis}$, e.g.
-`\texttt{NA}' values in \proglang{R} or `\texttt{999}' in \proglang{SPSS}.
+In this note, the terms `unobserved' and `missing' data are used interchangeably to refer to $Y_{mis}$.
 The terms `incomplete' or `observed' data denote $Y_{obs}$.
 Incomplete data is the starting point of the multiple imputation procedure.
-Figure \ref{fig:steps} provides a schematic overview of the steps involved with MI.
+Figure \ref{fig:steps} provides an overview of the steps involved with MI.
 
@@
 The collection of samples across iterations (between the initial value and the imputed value) for each of the $m$ imputations will be referred to as an `imputation chain'.
-The number of iterations will be denoted with $t$.
+The total number of iterations per chain will be denoted with $T$, where $t$ varies over the integers $1, 2, \dots, T$.
 
@@
 
-The multiple imputation algorithm in \pkg{mice} is a special case of Markov chain Monte Carlo (MCMC) methods--a framework of iterative algorithmic procedures.
-While methodology for diagnosing convergence of MI algorithms has not been systematically studied \citep{buur18}, several convergence diagnostics have been proposed for other types of MCMC methods and the framework in general.
+There is no scientific consensus on the convergence properties of multiple imputation algorithms \citep{taka17}.
+Some default techniques in \pkg{mice} might not yield converged states at all \citep{murr18}.
+Algorithmic convergence should, therefore, be monitored carefully.
+In that respect is the current evaluation practice insuffient.
+That is, the  recommendation is to visually inspect imputation chains for signs of non-convergence, which may be challenging to the untrained eye \citep[\S~6.5.2]{buur18}.
+Quantitative evaluation of the convergence of MI algorithms would be preferred.
+
+The multiple imputation algorithm in \pkg{mice} is a special case of Markov chain Monte Carlo (MCMC) methods---a framework of iterative algorithmic procedures.
+Within this framework, several convergence diagnostics have been proposed, but the application of these measures on MI algorithms has not been systematically studied \citep{buur18}.
 % For the framework of itrIt is not known whether convergence diagnostics that are available for the general framework that MICE is a special case of are appropriate for the evaluation of MI data.
 This is the framework of Markoc chain Monte Carlo %Yet, reference manuals of promiment MI software do mention the importance of evaluating algorithmic convergence (e.g., \cite[p.XYZ]{mice}, \textbf{XYZ, XYZ}).
-This section provides a brief review of: 1) how convergence is defined in the general context of iterative algorithmic procedures; 2) what diagnostic tools are available; and 3) which of these may apply to MI algorithms.
+This section provides a brief review of:% briefly reviews?
+1) how convergence is defined in the general context of iterative algorithmic procedures; 2) what diagnostic tools are available; and 3) which of these may apply to MI algorithms.
 
@@
 
-Most convergence diagnostics target either of the two components.
-The stationarity component of convergence may be evaluated with autocorrelation \citep[$AC$][]{scha97, gelm13}, numeric standard error \citep[or `MC error'][]{gewe92}, and Raftery and Lewis's (\citeyear{raft91}) procedure to determine how many iterations are necessary to diminish the effect of trending within chains.
-The mixing component can be assessed with the potential scale reduction factor $\widehat{R}$ \citep[widely referred to as `the Gelman-Rubin statistic']{gelm92}.
-With an adapted version of $\widehat{R}$, proposed by \cite{veht19}, we may also evaluate the stationarity component of convergence.
+\textbf{[COULD BE SHORTER!]} Most convergence diagnostics target either of the two components.
+The stationarity component of convergence may be evaluated with autocorrelation \citep[$AC$; ][]{scha97, gelm13}, numeric standard error \citep[or `MC error'; ][]{gewe92}, and Raftery and Lewis's (\citeyear{raft91}) procedure to determine the effect of trending within chains.
+The mixing component can be assessed with the potential scale reduction factor $\widehat{R}$ \citep[a.k.a.
+`Gelman-Rubin statistic';][]{gelm92}.
+With an adapted version of $\widehat{R}$, proposed by \cite{veht19}, we might also evaluate the stationarity component of convergence.
 This would make $\widehat{R}$ a general convergence diagnostic.
 The application of $\widehat{R}$ to assess stationarity has not been thoroughly investigated.
-Therefore, this study employs both $\widehat{R}$ and autocorrelation to investigate algorithmic convergence, as recommended by \cite[p.~898]{cowl96}.
+Therefore, this study employs both $\widehat{R}$ and autocorrelation to investigate convergence, as recommended by \cite[p.~898]{cowl96}.
 %Therefore, the focus of this study is mainly on $\widehat{R}$. Stationarity will also be evaluated with autocorrelation as diagnostic, as recommended by \cite[p.~898]{cowl96}. 
@@
 Once there is no dependence on initial values anymore, there is no added value of continuing iterations.
-The measure then diagnoses algorithmic convergence ($\widehat{R} = 1$).
+\textbf{[MOGELIJK OVERBODIGE ZIN HIER]} The measure then diagnoses algorithmic convergence ($\widehat{R} = 1$).
 %If there is no added value compared to infinitely many iterations, the chains have converged. 
-The conventionally acceptable threshold was $\widehat{R} < 1.2$ \cite{gelm92}. More recently, \cite{veht19} proposed a more stringent threshold: $\widehat{R} < 1.01$.
+The conventionally acceptable threshold was $\widehat{R} < 1.2$ \cite{gelm92}. Recently, \cite{veht19} proposed a more stringent threshold: $\widehat{R} < 1.01$.
 
@@
 
-It is not known whether $\widehat{R}$ and $AC$ are appropriate measures to diagnose convergence of MI algorithms.
-Assessing the stationarity component of convergence with $AC$ might be redundant, since high $AC$-values are implausible in MI procedures.
-That is, the randomness induced by the MI algorithm effectively mitigates the risk of dependency within chains.
-$AC$ would thus not be informative of the convergence of MI algorithms.
-$\widehat{R}$ may not be an appropriate diagnostic either, because it assumes over-dispersed initial values (i.e., the starting points of the chains are `far away' from the target distribution and each other).
-In \pkg{mice}, initial values of the algorithm are chosen randomly from the observed data.
-Therefore, we cannot be certain that the initial values are over-dispersed.
-%The initial parameter value is noodzakelijk.
-Daarom is het lastig om in MI that starten vanuit overdisp state.
-% Starting values of MICE chains: "For each j, fill in starting imputations $Y^0_j$ by random draws from $Y^{obs}_j$." \cite[\S~4.5.2]{buur18} before for loops over variables and imputations, see \url{https://stefvanbuuren.name/fimd/sec-FCS.html#def:mice}
-Without over-dispersed initial states, $\widehat{R}$ may falsely diagnose convergence \citep{broo98}.
-%would not be sensitive enough to flag non-convergence of MI algorithms: "if over-dispersion does not hold, $\sigma_+^2$ can be too low, which can lead to falsely diagnosing convergence." \cite[p~437]{broo98}.
-This suggests that $\widehat{R}$ would not be sensitive enough to flag non-convergence of MI algorithms.
-An empirical finding, on the other hand, shows that $\widehat{R}$ can indicate non-convergence at unconventionally long chain lengths.
-While the default number of iterations in \pkg{mice} is five, $\widehat{R} > 1.1$ was observed after 50 iterations, and $\widehat{R} > 1.01$ up-to 150 iterations.
-This suggests that $\widehat{R}$ would over-estimate non-convergence of MI algorithms.
+%\textbf{HIER VALT TE SCHRAPPEN!} It is not known whether $\widehat{R}$ and $AC$ are appropriate measures to diagnose convergence of multiple imputation algorithms.
 
@@
 
-This study evaluates whether $\widehat{R}$ and $AC$ can diagnose when the multiple imputation algorithm has converged.
-There is, however, no baseline measure available to evaluate their performance against.
-The current practice of assessing convergence is to visually inspect imputation chains for signs of non-mixing and non-stationarity \citep[\S~6.5.2]{buur18}.
-Visually assessing convergence, however, may be challenging to the untrained eye.
-\textbf{[This is problematic because.
-MAYBE REPLACE THIS PART!]} There is no scientific consensus on the convergence properties of multiple imputation algorithms \citep{taka17}, and in some situations, the default techniques in \pkg{mice} might not converge to stable target distributions at all \citep{murr18}.
-%Convergence diagnostics for MI methods have not been systematically studied \citep{buur18}.
-Moreover,
+This study evaluates whether $\widehat{R}$ and $AC$ could diagnose convergence of multiple imputation algorithms.
+There is, however, no baseline measure available to evaluate their performance against.\footnote{Only severely pathological cases of non-convergence stand out when visually inspecting the imputation chains.} We will assess the appropriateness of the two convergence diagnostics by comparison with simulation quantities (e.g., bias in estimating $Q$).
+
+\textbf{We hypothesize that $\widehat{R}$ will over-estimate non-convergence. No hypothesis was formulated about AC, results are exploratory.
+%
+[MOVE THIS TO DISCUSSION?] Assessing the stationarity component of convergence with $AC$ might be redundant, since high $AC$-values are implausible in MI procedures.
+That is, the randomness induced by the MI algorithm effectively mitigates the risk of dependency within chains.
+$AC$ would thus not be informative of the convergence of MI algorithms.
+%
+Moreover, we hypothesize that $\widehat{R}$ may not be an appropriate diagnostic either, because it assumes over-dispersed initial values (i.e., the starting points of the chains are `far away' from the target distribution and each other).
+In \pkg{mice}, initial values of the algorithm are chosen randomly from the observed data.
+Therefore, we cannot be certain that the initial values are over-dispersed.
+%The initial parameter value is noodzakelijk.
+Daarom is het lastig om in MI that starten vanuit overdisp state.
+% Starting values of MICE chains: "For each j, fill in starting imputations $Y^0_j$ by random draws from $Y^{obs}_j$." \cite[\S~4.5.2]{buur18} before for loops over variables and imputations, see \url{https://stefvanbuuren.name/fimd/sec-FCS.html#def:mice}
+Without over-dispersed initial states, $\widehat{R}$ may falsely diagnose convergence \citep{broo98}.
+%would not be sensitive enough to flag non-convergence of MI algorithms: "if over-dispersion does not hold, $\sigma_+^2$ can be too low, which can lead to falsely diagnosing convergence." \cite[p~437]{broo98}.
+This suggests that $\widehat{R}$ would not be sensitive enough to flag non-convergence of MI algorithms.
+However, an empirical finding by \cite{lace07} shows that $\widehat{R}$ can indicate non-convergence at unconventionally long chain lengths.
+While the default number of iterations in \pkg{mice} is five, $\widehat{R} > 1.1$ was observed after 50 iterations, and $\widehat{R} > 1.01$ up-to 150 iterations.
+This suggests that $\widehat{R}$ would over-estimate non-convergence of MI algorithms.
+}
+
+%Convergence diagnostics for MI methods have not been systematically studied \citep{buur18}.  Moreover, 
 %A fundamental assumption of MICE is convergence of the algorithm.
@@
 But it is not known whether conventional convergence diagnostics for MCMC methods work on MI data.
-Therefore, convergence of the MI algorithms in \pkg{mice} should be monitored carefully. %the need for convergence diagnostics is extra pressing in the context of the MICE algorithm.
+%the need for convergence diagnostics is extra pressing in the context of the MICE algorithm.
 
@@
 
-This simulation study hypothesizes that $\widehat{R}$ will over-estimate non-convergence.
+%This simulation study hypothesizes that $\widehat{R}$ will over-estimate non-convergence.
 %\textbf{[Or use:] Hypothesis based on \cite{lace07} is that the conventional acceptable level of $\widehat{R}$ is too strict for MI data.
@@
 
-To investigate the convergence properties of the \pkg{mice} algorithm, we perform model-based simulation of the MI procedure with varying imputation chain lengths.
-Each simulation condition has a different maximum number of iterations, from one to one-hundred iterations.
-We compute $\widehat{R}$ and autocorrelation (`convergence diagnostics') and performance measures (`simulation diagnostics') for each condition.
+%To investigate the convergence properties of the multiple imputation algorithm in \pkg{mice}, w
+We use model-based simulation to perform multiple imputation under 100 simulation conditions.
+In each condition, the MI algorithm comprises a different imputation chain length ($T = 1, 2, \dots, 100$).
 The number of simulation runs per condition is 1000.
+Convergence diagnostics ($\widehat{R}$ and autocorrelation) and simulation diagnostics (bias, confidence interval width and coverage) are computed for each simulation condition, and aggregated across simulation runs.
+%We compute $\widehat{R}$ and autocorrelation (`convergence diagnostics') and performance measures (`simulation diagnostics') for each condition.
 %All simulation conditions are replicated 1000 times.
@@
 The simulated dataset is a finite population of $N=1000$.
-The data are simulated to solve a multiple linear regression complete data problem.
-The quantity of scientific interest is the estimated regression coefficient of predictor $X$ on outcome variable $Y$.
-The linear regression model is $Y \sim \beta_1 X + \beta_2 Z_1 + \beta_3 Z_2$, where $Y$ is the dependent variable, $X$ is an independent variable, and $Z_1$ and $Z_2$ are covariates.
-The data generating model of the predictors is a multivariate normal distribution with means structure $\mu$, and variance-covariance matrix $\Sigma$.
+The data are simulated to solve a multiple linear regression problem, where dependent variable $Y$ is regressed on independent variable $X$, and covariates $Z_1 \text{ and } Z_2$: $Y \sim \beta_1 X + \beta_2 Z_1 + \beta_3 Z_2$.
+The quantity of scientific interest is regression coefficient $\beta_1$.
+The data generating model is a multivariate normal distribution with means structure $\mu$, and variance-covariance matrix $\Sigma$.
 
@@
 The $m = 5$ estimated regression coefficients $\hat{Q}$ are pooled according to Rubin's \citeyear{rubin87} rules with \pkg{mice} function \fct{pool}.
+Variance is computed accroding to finite population pooling \textbf{Vink and Van Buuren, 2014??}.
 
 \subsection{Simulation diagnostics}
-The simulation diagnostics are as recommended by \cite[\S~2.5.2]{buur18}.
-These evaluation criteria comprise of average bias, average confidence interval width, and empirical coverage rate across simulations.
-%"We expect that in accordance with Neyman (1934), at least 95 percent of the confidence intervals should contain the true population value, although some room for simulation error should be taken into account to counteract the finite nature of our simulations"\cite{schou18}.
-We could also look at distributional characteristics, and plausibility of imputed values, see \cite{vinknd} (n.d.).
-For now, this is outside of the scope of this study.
-% Or use: Then evaluate the regular diagnostics of MI simulations, see Vink, n.d., and  convergence.
+The simulation diagnostics are as recommended by \citep[\S~2.5.2]{buur18}.
+These evaluation criteria comprise of average bias, average confidence interval width, and empirical coverage rate across simulations.% %"We expect that in accordance with Neyman (1934), at least 95 percent of the confidence intervals should contain the true population value, although some room for simulation error should be taken into account to counteract the finite nature of our simulations"\cite{schou18}.
+\footnote{We could also look at distributional characteristics, and plausibility of imputed values, see \cite{vinknd} (n.d.).
+For now, this is outside of the scope of this study.} % Or use: Then evaluate the regular diagnostics of MI simulations, see Vink, n.d., and  convergence.
 Diagnostics include absolute bias of the estimated regression coefficient, confidence interval width, and empirical coverage rate across simulations.

### 2019-12-12 Simulation/Writeup/NoteConvergence.tex
Comments Rosalie verwerkt, nog steeds te veel woorden...

@@
 
-Most convergence diagnostics target either of the two components.
-The stationarity component of convergence may be evaluated with autocorrelation \citep[$AC$][]{scha97, gelm13}, numeric standard error \citep[or `MC error'][]{gewe92}, and Raftery and Lewis's (\citeyear{raft91}) procedure to determine how many iterations are necessary to deminish the effect of trending within chains.
-The mixing component can be assessed with the potential scale reduction factor $\widehat{R}$ \citep[widely referred to as `the Gelman-Rubin statistic']{gelm92}.
-With an adapted version of $\widehat{R}$, proposed by \cite{veht19}, we may also evaluate the stationarity component of convergence.
+\textbf{[COULD BE SHORTER!]} Most convergence diagnostics target either of the two components.
+The stationarity component of convergence may be evaluated with autocorrelation \citep[$AC$; ][]{scha97, gelm13}, numeric standard error \citep[or `MC error'; ][]{gewe92}, and Raftery and Lewis's (\citeyear{raft91}) procedure to determine the effect of trending within chains.
+The mixing component can be assessed with the potential scale reduction factor $\widehat{R}$ \citep[a.k.a.
+`Gelman-Rubin statistic';][]{gelm92}.
+With an adapted version of $\widehat{R}$, proposed by \cite{veht19}, we might also evaluate the stationarity component of convergence.
 This would make $\widehat{R}$ a general convergence diagnostic.
-The application of $\widehat{R}$ to assess stationarity has not been thourougly investigated.
-Therefore, this study employs both $\widehat{R}$ and autocorrelation to investigate algorithmic convergence, as recommended by \cite[p.~898]{cowl96}.
+The application of $\widehat{R}$ to assess stationarity has not been thoroughly investigated.
+Therefore, this study employs both $\widehat{R}$ and autocorrelation to investigate convergence, as recommended by \cite[p.~898]{cowl96}.
 %Therefore, the focus of this study is mainly on $\widehat{R}$. Stationarity will also be evaluated with autocorrelation as diagnostic, as recommended by \cite[p.~898]{cowl96}. 
@@
 Positive $AC$-values would indicate recurrence: similar $\theta$-values for subsequent iterations, which may lead to trending.
-Negative $AC$-values also indicate dependence, but do not threathen the stationarity component of convergence.
+Negative $AC$-values also indicate dependence, but do not threaten the stationarity component of convergence.
 Instead of recurrence, there would be a `seesaw' effect within chains: subsequent $\theta$-values would deviate in opposite directions from the chain average $\theta^{(\cdot m)})$, increasing the variance of $\theta$.
@@
 
-It is not known whether $\widehat{R}$ and $AC$ are appropriate measures to diagnose convergence of MI algorithms.
-Assessing the stationarity component of convergence with $AC$ might be redundant, since high $AC$-values are implausible in MI procedures.
-That is, the randomness induced by the MI algorithm effectively mitigates the risk of dependency within chains.
-$AC$ would thus not be informative of the convergence of MI algorithms.
-$\widehat{R}$ may not be an appropriate diagnostic either, because it assumes over-dispersed initial values (i.e., the starting points of the chains are `far away' from the target distribution and each other).
-In \pkg{mice}, initial values of the algorithm are chosen randomly from the observed data.
-Therefore, we cannot be certain that the initial values are over-dispersed.
-%The initial parameter value is noodzakelijk.
-Daarom is het lastig om in MI that starten vanuit overdisp state.
-% Starting values of MICE chains: "For each j, fill in starting imputations $Y^0_j$ by random draws from $Y^{obs}_j$." \cite[\S~4.5.2]{buur18} before for loops over variables and imputations, see \url{https://stefvanbuuren.name/fimd/sec-FCS.html#def:mice}
-Without over-dispersed initial states, $\widehat{R}$ may falsely diagnose convergence \citep{broo98}.
-%would not be sensitive enough to flag non-convergence of MI algorithms: "if over-dispersion does not hold, $\sigma_+^2$ can be too low, which can lead to falsely diagnosing convergence." \cite[p~437]{broo98}.
-This suggests that $\widehat{R}$ would not be sensitive enough to flag non-convergence of MI algorithms.
-An empirical finding, on the other hand, shows that $\widehat{R}$ can indicate non-convergence at unconventionally long chain lengths.
-While the default number of iterations in \pkg{mice} is five, $\widehat{R} > 1.1$ was observed after 50 iterations, and $\widehat{R} > 1.01$ upto 150 iterations.
-This suggests that $\widehat{R}$ would over-estimate non-convergence of MI algorithms.
+%\textbf{HIER VALT TE SCHRAPPEN!} It is not known whether $\widehat{R}$ and $AC$ are appropriate measures to diagnose convergence of multiple imputation algorithms.
 
@@
 
-This study evaluates whether $\widehat{R}$ and $AC$ can diagnose when the multiple imputation algorithm has converged.
-There is, however, no baseline measure available to evaluatate their performance against.
-The current practice of assessing convergence is to visually inspect imputation chains for signs of non-mixing and non-stationarity \citep[\S~6.5.2]{buur18}.
-Visually assessing convergence, however, may be challenging to the untrained eye.
-\textbf{[This is problematic because.
-MAYBE REPLACE THIS PART!]} There is no scientific consensus on the convergence properties of multiple imputation algorithms \citep{taka17}, and in some situations, the default techniques in \pkg{mice} might not converge to stable target distributions at all \citep{murr18}.
-%Convergence diagnostics for MI methods have not been systematically studied \citep{buur18}.
-Moreover,
+This study evaluates whether $\widehat{R}$ and $AC$ could diagnose convergence of multiple imputation algorithms.
+There is, however, no baseline measure available to evaluate their performance against.\footnote{Only severely pathological cases of non-convergence stand out when visually inspecting the imputation chains.} We will assess the appropriateness of the two convergence diagnostics by comparison with simulation quantities (e.g., bias in estimating $Q$).
+
+\textbf{We hypothesize that $\widehat{R}$ will over-estimate non-convergence. No hypothesis was formulated about AC, results are exploratory.
+%
+[MOVE THIS TO DISCUSSION?] Assessing the stationarity component of convergence with $AC$ might be redundant, since high $AC$-values are implausible in MI procedures.
+That is, the randomness induced by the MI algorithm effectively mitigates the risk of dependency within chains.
+$AC$ would thus not be informative of the convergence of MI algorithms.
+%
+Moreover, we hypothesize that $\widehat{R}$ may not be an appropriate diagnostic either, because it assumes over-dispersed initial values (i.e., the starting points of the chains are `far away' from the target distribution and each other).
+In \pkg{mice}, initial values of the algorithm are chosen randomly from the observed data.
+Therefore, we cannot be certain that the initial values are over-dispersed.
+%The initial parameter value is noodzakelijk.
+Daarom is het lastig om in MI that starten vanuit overdisp state.
+% Starting values of MICE chains: "For each j, fill in starting imputations $Y^0_j$ by random draws from $Y^{obs}_j$." \cite[\S~4.5.2]{buur18} before for loops over variables and imputations, see \url{https://stefvanbuuren.name/fimd/sec-FCS.html#def:mice}
+Without over-dispersed initial states, $\widehat{R}$ may falsely diagnose convergence \citep{broo98}.
+%would not be sensitive enough to flag non-convergence of MI algorithms: "if over-dispersion does not hold, $\sigma_+^2$ can be too low, which can lead to falsely diagnosing convergence." \cite[p~437]{broo98}.
+This suggests that $\widehat{R}$ would not be sensitive enough to flag non-convergence of MI algorithms.
+However, an empirical finding by \cite{lace07} shows that $\widehat{R}$ can indicate non-convergence at unconventionally long chain lengths.
+While the default number of iterations in \pkg{mice} is five, $\widehat{R} > 1.1$ was observed after 50 iterations, and $\widehat{R} > 1.01$ up-to 150 iterations.
+This suggests that $\widehat{R}$ would over-estimate non-convergence of MI algorithms.
+}
+
+%Convergence diagnostics for MI methods have not been systematically studied \citep{buur18}.  Moreover, 
 %A fundamental assumption of MICE is convergence of the algorithm.

### 2019-12-13 Simulation/Writeup/NoteConvergence.Rnw
results

@@
 Algorithmic convergence should, therefore, be monitored carefully.
-In that respect is the current evaluation practice insuffient.
+In that respect is the current evaluation practice insufficient.
 That is, the  recommendation is to visually inspect imputation chains for signs of non-convergence, which may be challenging to the untrained eye \citep[\S~6.5.2]{buur18}.
@@
 
-Loosely speaking, the aim of MCMC algorithms is to sample values from an unknown target distribution (e.g., the posterior distribution of a parameter in Gibbs samplers, or the distribution of missing values in an MI algorithm).
+Loosely speaking, MCMC algorithms aim to sample values from an unknown target distribution (e.g., the posterior distribution of a parameter in Gibbs samplers, or the distribution of missing values in MI algorithms).
 Convergence is reached once the algorithm samples exclusively from the target distribution \citep{cowl96}.
@@
 These convergence diagnostics cannot diagnose convergence of multivariable statistics (i.e., relations between scalar summaries).
-\cite{buur18} proposed to implement multivariable evaluation by means of eigenvalue decomposition \cite{mack03}.
+\cite{buur18} proposed to implement multivariable evaluation through eigenvalue decomposition \cite{mack03}.
 This method is outside of the scope of the current study.}%  he eigenvector decomposition method proposed by McKay \citep{XYZ}.} These four methods, however, are outside of the scope of this study.
@@
 %The potential scale reduction factor $\widehat{R}$ is obtained as:
-We can interpret $\widehat{R}$ as potential scale reduction factor, since it indicates by how much the variance of $\theta$ could be shrunken down if an infinite number of iterations per chain would be run \citep[$t \to \infty$;][]{gelm92}.
-High $\widehat{R}$-values would therefore indicate slow mixing of chains, e.g., because of dependence across chains on initial values.
+We can interpret $\widehat{R}$ as potential scale reduction factor since it indicates by how much the variance of $\theta$ could be shrunken down if an infinite number of iterations per chain would be run \citep[$t \to \infty$;][]{gelm92}.
+High $\widehat{R}$-values, therefore, indicate slow mixing of chains, e.g., because of dependence across chains on initial values.
 Once there is no dependence on initial values anymore, there is no added value of continuing iterations.
@@
 where $T$ is the total number of iterations per chain \citep[p.~147]{lync07}.
-We can interpret the sign of $AC$-values as measure of dependence within chains ($-1 \leq AC \leq 1$).
-Positive $AC$-values would indicate recurrence: similar $\theta$-values for subsequent iterations, which may lead to trending.
-Negative $AC$-values also indicate dependence, but do not threaten the stationarity component of convergence.
+We can interpret the sign of $AC$-values as a measure of dependence within chains ($-1 \leq AC \leq 1$).
+Positive $AC$-values indicate recurrence: similar $\theta$-values for subsequent iterations, which may lead to trending.
+Negative $AC$-values also indicate dependence, but do not threaten the stationarity component of convergence \textbf{[REMOVE INTERPRETATION HERE?]}.
 Instead of recurrence, there would be a `seesaw' effect within chains: subsequent $\theta$-values would deviate in opposite directions from the chain average $\theta^{(\cdot m)})$, increasing the variance of $\theta$.
@@
 This study evaluates whether $\widehat{R}$ and $AC$ could diagnose convergence of multiple imputation algorithms.
-There is, however, no baseline measure available to evaluate their performance against.\footnote{Only severely pathological cases of non-convergence stand out when visually inspecting the imputation chains.} We will assess the appropriateness of the two convergence diagnostics by comparison with simulation quantities (e.g., bias in estimating $Q$).
+There is, however, no baseline measure available to evaluate their performance against.\footnote{Only severely pathological cases of non-convergence stand out when visually inspecting the imputation chains.} We will assess the appropriateness of the two convergence diagnostics by comparison with simulation quantities.
+The perfrormance diagnostics are the evaluation criteria  recommended by \citep[\S~2.5.2]{buur18}, whcih comprise of average bias, average confidence interval width, and empirical coverage rate across simulations.% %"We expect that in accordance with Neyman (1934), at least 95 percent of the confidence intervals should contain the true population value, although some room for simulation error should be taken into account to counteract the finite nature of our simulations"\cite{schou18}.
+\footnote{We could also look at distributional characteristics, and plausibility of imputed values, see \cite{vinknd} (n.d.).
+For now, this is outside of the scope of this study.} % Or use: Then evaluate the regular diagnostics of MI simulations, see Vink, n.d., and  convergence.
+Diagnostics include absolute bias of the estimated regression coefficient, confidence interval width, and empirical coverage rate across simulations.
+Convergence is evaluated with $\widehat{R}$, or potential scale reduction factor.
 
@@
 %To investigate the convergence properties of the multiple imputation algorithm in \pkg{mice}, w
-We use model-based simulation to perform multiple imputation under 100 simulation conditions.
+We use model-based simulation in \proglang{R} to perform multiple imputation under 100 simulation conditions.
 In each condition, the MI algorithm comprises a different imputation chain length ($T = 1, 2, \dots, 100$).
 The number of simulation runs per condition is 1000.
-Convergence diagnostics ($\widehat{R}$ and autocorrelation) and simulation diagnostics (bias, confidence interval width and coverage) are computed for each simulation condition, and aggregated across simulation runs.
+For each simulation condition (i.e., number of iterations) and each repetition (i.e., simulation run) we compute convergence diagnostics ($\widehat{R}$ and autocorrelation) and simulation diagnostics (bias, confidence interval width and coverage).
+These diagnostics are aggregated across repetitions to obtain averages per simulation condition.
 %We compute $\widehat{R}$ and autocorrelation (`convergence diagnostics') and performance measures (`simulation diagnostics') for each condition.
@@
 \subsection{Data simulation}
-The simulated dataset is a finite population of $N=1000$.
-The data are simulated to solve a multiple linear regression problem, where dependent variable $Y$ is regressed on independent variable $X$, and covariates $Z_1 \text{ and } Z_2$: $Y \sim \beta_1 X + \beta_2 Z_1 + \beta_3 Z_2$.
+The point of origin for all simulation conditions is a finite population of $N=1000$.
+The data are simulated to solve a multiple linear regression problem, where dependent variable $Y$ is regressed on independent variable $X$, and covariates $Z_1$ and $Z_2$: $Y \sim \beta_1 X + \beta_2 Z_1 + \beta_3 Z_2$.
 The quantity of scientific interest is regression coefficient $\beta_1$.
-The data generating model is a multivariate normal distribution with means structure $\mu$, and variance-covariance matrix $\Sigma$.
+The data generating model is a multivariate normal distribution with means structure $\mu$ and variance-covariance matrix $\Sigma$.
 
@@
 \subsection{Amputation}
-The complete data is the point of origin for all simulations.
-In each repetition of the simulation study, the complete data is `amputed' once.
-That is, the \pkg{mice} function \fct{ampute} in \proglang{R} is used to impose an MCAR missingness mechanism upon the data.
-The missingness is univariate, and the probability to be missing is the same for all variables, namely 20\%.
+The complete dataset is `amputed' once in each simulation repetition.
+That is, the \pkg{mice} function \fct{ampute} is used to impose a missingness mechanism upon the data.
+The missingness is univariate, and the probability to be missing is the same for all variables, namely 20\% (\texttt{prop = 0.8, mech = "MCAR"}).
 This leaves 20\% of the rows completely observed.
-The resulting amputed data is equal for all simulation conditions in the same repetition.%"Furthermore, we generate three kinds of missingness proportions: 0.1, 0.5, and 0.9.
+%The resulting amputed data is equal for all simulation conditions in the same repetition.
+%"Furthermore, we generate three kinds of missingness proportions: 0.1, 0.5, and 0.9.
 Note that these values indicate the sampled proportion of incomplete cases in Y" \cite{schou18}.
@@
 \subsection{Imputation}
-Missing data points are imputed with \pkg{mice} in \proglang{R}.
-All simulations are performed with imputation method 'norm' (Bayesian linear regression imputation), and five imputation chains ($m = 5$).
-The number of iterations is varied over simulations (\texttt{maxit} argument between 1 and 100).
-Each simulation condition (i.e., each number of iterations) is simulated 1000 times.
-Simulation and convergence diagnostics are aggregated over the 1000 MCMC simulations.
+Missing data points are imputed with the \pkg{mice} function \fct{mice}.
+All simulations are performed with Bayesian linear regression imputation (\texttt{method = "norm"}), and five imputation chains (\texttt{m = 5}).
+The number of iterations varies between simulation conditions (\texttt{maxit = 1, 2, \dots, 100}).
 % Missingness mechanism: "With MCAR missingness mechanisms, the probability to be missing is the same for all cases.
@@
 %, see \url{http://jsterman.scripts.mit.edu/docs/Rahmandad-Sterman%20Simulation%20Reporting%20Standards3.30%20update.pdf}.
+We investigate convergence by extracting not only the imputed values of each MI procedure (the $T^{th}$ iteration), but also information ($\theta$-values) from the intermediate iterations ($t = 1, 2, \dots, T$).
+The $\theta$s of interest are chain means and chain variances per variable.
+We compute convergence diagnostics separately for each of the four variables in the dataset, and report the maximum (absolute) value as the `worst linear predictor' of convergence \textbf{[Rephrase or footnote??]}.
+$\widehat{R}$ is computed by implementing the definition given in \textbf{EQUATION NR}.
+$AC$ is computed with \pkg{stats} function \fct{acf}.
 
-$\widehat{R}$ is computed within imputation chains: for each variable, each simulation condition, and each simulation.
-The maximum value across variables within the same simulation is reported.
-Maximum $\widehat{R}$-values per simulation condition are aggregated across simulations.
-
-autocorrelation (AC) is computed within imputation chains,  The AC of the variable with the highest absolute AC value is reported per simulation.
-These values are then averaged per simulation condition.
+% 
+% 
+% extract chain means and chain variances per iteration $t$. From the MI algorithm Convergence diagnostics are computed for chain means and chain variances.   
+% $\widehat{R}$ is computed within imputation chains: . The maximum value across variables within the same simulation is reported. 
+% 
+% autocorrelation (AC) is computed within imputation chains,  The AC of the variable with the highest absolute AC value is reported per simulation.
 
@@
 
-Linear regression is performed using the \proglang{R} function \fct{lm}.
-The quantity of scientific interest $Q$ is the regression coefficient of $X$ on $Y$.
-$Q$ is estimated in each imputation, in each simulation condition, in each repetition.
-The $m = 5$ estimated regression coefficients $\hat{Q}$ are pooled according to Rubin's \citeyear{rubin87} rules with \pkg{mice} function \fct{pool}.
-Variance is computed accroding to finite population pooling \textbf{Vink and Van Buuren, 2014??}.
+To estimate the quantity of scientific interest, $Q$, we perform multiple linear regression on each completed dataset with the \pkg{stats} function \fct{lm}.
+We obtain an estimated regression coefficient per imputation, which are pooled into a single estimate, $\bar{Q}$.
+We use the \pkg{mice} function \fct{pool} to get variance estimates according to Rubin's \citeyear{rubin87} rules, and utilize these to implement finite population pooling as per \citep{vink14}.
 
 \subsection{Simulation diagnostics}
-The simulation diagnostics are as recommended by \citep[\S~2.5.2]{buur18}.
-These evaluation criteria comprise of average bias, average confidence interval width, and empirical coverage rate across simulations.% %"We expect that in accordance with Neyman (1934), at least 95 percent of the confidence intervals should contain the true population value, although some room for simulation error should be taken into account to counteract the finite nature of our simulations"\cite{schou18}.
-\footnote{We could also look at distributional characteristics, and plausibility of imputed values, see \cite{vinknd} (n.d.).
-For now, this is outside of the scope of this study.} % Or use: Then evaluate the regular diagnostics of MI simulations, see Vink, n.d., and  convergence.
-Diagnostics include absolute bias of the estimated regression coefficient, confidence interval width, and empirical coverage rate across simulations.
-Convergence is evaluated with $\widehat{R}$, or potential scale reduction factor.
-
 % Distributional characteristics: In practice, the distribution of the incomplete data may differ greatly from the observed data.
@@
 When anomalies are found, and if the imputation method is valid, there should be an explanation, especially in the controlled environment of a properly executed simulation study.
-
 % Plausibility of the imputed values: Plausible imputations - imputations that could be real values if they had been observed - are not a necessary condition for obtaining valid inference.
@@
 
-Bias is computed as the estimated regression coefficient after MI minus the true regression coefficient. Bias is averaged over all simulations with the same simulation condition.
-
-Confidence interval width (CIW) is computed as the difference between the lower and upper bound of the 95\% confidence interval (CI95\%).
-The CI95\% bounds are computed as the estimated regression coefficient plus or minus (respectively) the pooled SE across imputations times 2.66 (the quantile of a t distribution with m-1 degrees of freedom).
-CIW is averaged over all simulations with the same simulation condition.
-
-Coverage rate is computed as the proportion of simulations (with the same simulation condition) in which the true regression coefficient is between the bounds of the CI95\%.
+We compute bias as the difference between $\bar{Q}$ and $Q$.
+Confidence interval width (CIW) is defined as the difference between the lower and upper bound of the 95\% confidence interval around $\bar{Q}$ (CI95\%).
+We compute the CI95\% bounds as $\bar{Q} \pm 2.66 \times SE_{\bar{Q}}$, where 2.66 is the quantile of a $t$-distribution with $m-1$ degrees of freedom, and $SE_{\bar{Q}}$ is the square root of the pooled variance estimate.
+From bias and CIW, we calculate empirical coverage rates.
+Coverage rate is the proportion of simulations in which $Q$ is between the bounds of the CI95\% around $\bar{Q}$.
 
@@
 
-Figures \ref{fig:conv} and \ref{fig:sim} display results of the simulation study aggregated over 1000 repetitions.
-A subset of simulation conditions is presented in Table \ref{tab:results}.
-This section describes convergence diagnostics ($\widehat{R}$ and autocorrelation) and simulation diagnostics (bias, confidence interval width and coverage rate) first separately, and then together.
+Figures \ref{fig:conv} and \ref{fig:sim} display results per simulation condition ($T = 1,2,\dots,100$). A subset of simulation conditions is presented in Table \ref{tab:results}.
 
@@
   \hline
-It. & Bias & CI width & Cov. rate & $\widehat{R}_{mean}$ & $\widehat{R}_{var}$ & $AC_{mean}$ & $AC_{var}$ \\ 
+T & Bias & CI width & Cov. rate & $\widehat{R}_{mean}$ & $\widehat{R}_{var}$ & $AC_{mean}$ & $AC_{var}$ \\ 
   \hline
@@
 \subsection{Convergence diagnostics}
+It is apparant that there is a relation between convergence diagnostics $\widehat{R}$ and $AC$, and the number of iterations per simulation condition($T$).
+Generally speaking, we see a trend towards convergence as $T$ increases with each simulation condition ($\widehat{R}$-values approache one, and $AC$-values approache zero).
+The convergence diagnostics show more or less equivalent trends for chain means versus chain variances.
+We, therefore, only discuss the $\widehat{R}$ and $AC$-values of chain means.
+%convergence diagnostics for the chain means.
+%do not differentiate in the interpretation of chain means .
 
-\textbf{$\widehat{R}$}.
-The $\widehat{R}$-values for chain means and chain variances show equivalent trends, neither is there a clear difference between the four variables.
+Figure \ref{fig:conv}A shows that $\widehat{R}$-values generally decrease with increasing imputation chain lengths.
+The decline stabilizes somewhere between the simulation conditions $T=30$ and $T=50$.
+The downward trend is most pronounced for $T=3$, and between $T = 5$ and $T = 10$.
+In the intervening conditions ($3 \leq T \leq 5$), however, we observe a steep increase in $\widehat{R}$-values.
+This increase implies that convergence cannot be diagnosed at $T=3$, as would be the case based on the conventional threshold $\widehat{R} < 1.2$.
+\textbf{This may be spurious.} According to the widely used threshold $\widehat{R} < 1.1$, convergence can be diagnosed for conditions where $T>9$.
+If we use the recently recommended threshold $\widehat{R} < 1.01$, we would conclude that we cannot diagnose convergence in any simulation condition.
+
+%The $\widehat{R}$-values for chain means and chain variances show equivalent trends, neither is there a clear difference between the four variables.
 Roughly speaking, we see that $\widehat{R}$ is higher for simulation conditions with a lower number of iterations, and approaches one for conditions with more iterations.
-Figure \ref{fig:conv} shows an initial dip, a small bump upwards, and then a steep decline.
-This decline \textbf{houdt aan} up-to iteration twenty, after which a more gradual decrease can be observed up-to iteration forty.
+
+%%%%
+%Figure \ref{fig:conv} shows an initial dip, a small bump upwards, and then a steep decline.
+This decline continues up-to iteration twenty, after which a more gradual decrease can be observed up-to iteration forty.
 The $\widehat{R}$-values are more or less stable after that.
-The \textbf{conventionally acceptable threshold for convergence, maximum $\widehat{R} < 1.2$,[CHECK IN INTRO]} is reached in the simulation condition with three iterations, and conditions with six iterations or more.
+%\textbf{[Remove stuff here]} %The conventionally acceptable threshold for convergence, maximum 
+%$\widehat{R} < 1.2$ is reached at $T =3$, and conditions with six iterations or more.
 The widely used threshold $\widehat{R} < 1.1$ is met in conditions with ten or more iterations.
@@
 $\widehat{R} > 1.01$ for all simulation conditions (see online appendix for the full table of results).
-\textbf{Check is computing r ht is still correct!
-Also say something about Rh below 1.}
+\textbf{say something about Rh below 1.}
+%%%%
 
-\textbf{autocorrelation.} All observed autocorrelations (aggregated over the 1000 repetitions) are negative or close to zero.
+The $AC$-values displayed in Figure \ref{fig:conv}B are almost uniformly increasing as a function of $T$.
+The only $AC$-value that deviates from this observation is for $T=2$.
+The lowest $AC$-value is obtained for $T=3$.
+At $T=5$, the $AC$-value reaches the level observed at $T=2$.
+Between $T=10$ and $T=30$ the increase plateaus off, and $AC$-values are stable around $T=70$.
+$AC$ is then indifferentiable from zero, indicating stationarity.
+%Simulation conditions where $T>5$ have $AC$-values greater than $T=2$.
+%We see an initial decrease between $T=2$ and $T=3$.
+According to this diagnostic, none of the simulation conditions show signs of non-convergence, since we only observe negative or zero $AC$-values.
+
+% All observed autocorrelations  are negative or close to zero.
 There are some differences between variables, and the ACs of chain variances seem lower than the ACs of chain means.
@@
 \textbf{The lowest AC observed is %\Sexpr{round(min(dat[-1, c("AC chain mean", "AC chain var")]), 3)}
-at four iterations.
+% at four iterations.
 Simulation conditions with a higher number of iterations show a slow increase towards zero.
@@
 
+Taken together, we see that convergence may be diagnosed in simulation conditions where $T>10$.
+
 %% --------------------------------
 \subsection{Simulation diagnostics}
-\textbf{Bias.} Figure \ref{fig:sim} shows the average bias between the regression coefficient of $X$ on $Y$ in the complete dataset, and $\bar{Q}$, \textbf{the pooled estimated regression coefficient and the ...
+We use average bias, average condidence interval width, and coverage rate as performance measures to evaluate $\widehat{R}$ and $AC$.
+We make the general observation that the simulation diagnostics behave as theorized for most simulation conditions (bias around zero, coverage rate around 95\%).
+
+Figure \ref{fig:sim}A shows that bias is fairly stable across simulation conditions.
+The condition $T=1$ clearly deviates from this trend with a negative bias.
+The bias for $T=2$ is below average, but within the range of fluctuations.
+Conditions where $T>3$ \textbf{are ok, Loess line is flat for $T>20$}.
+%the average bias between the regression coefficient of $X$ on $Y$ in the complete dataset, and $\bar{Q}$, \textbf{the pooled estimated regression coefficient and the ...
 Average bias is close to zero for simulation conditions with two iterations or more.
@@
 % Average bias is not zero because there is only one incomplete dataset to be imputed. Bias is a sample effect.
+%Figure \ref{fig:sim}A shows that bias is fairly stable across simulation conditions.
+The condition $T=1$ clearly deviates from this trend with a negative bias.
+The bias for $T=2$ is below average, but within the range of fluctuations.
+Conditions where $T>3$ \textbf{are ok, Loess line is flat for $T>20$}.
 
-\textbf{Confidence interval width.} CIW is only clearly divergent for the simulation condition with one iteration.
-Conditions with two or more iterations have similar CIWs.
-The magnitude of differences from the loess line is somewhat larger up-to the condition with four iterations.
-From four iterations on, CIW seems to be stable across iterations.
-\textbf{Loess line is never flat.}%One might argue that there is some downward trending up-to thirthy to forty iterations.
+CIW is only clearly divergent for the simulation condition $T=1$.
+However, we want to be on the safe side, and not under-estimate the variance.
+Otherwise, we might end up with spurious inferences.
+Conditions where $T>3$ have similar CIWs, which seems to be stable across iterations.
+\textbf{Loess line is never completely flat.}%One might argue that there is some downward trending up-to thirthy to forty iterations.
 %After about twenty-five iterations the average CIW across all simulation conditions and repetitions is reached \Sexpr{round(mean(dat[, "CI width"]), 3)}.
 
-\textbf{Coverage rate.} \textbf{The coverage rate is more or less stable from two iterations upwards.
+\textbf{Empirical coverage rate seems more or less stable for conditions where $T>1$.
+We see some over-coverage at $T=2$.
+We might over-estimate the variance, but better than under-estimating it.
 On average, the coverage rate is somewhat higher than the expected nominal coverage of 95\% \cite{neym34}, namely \Sexpr{100*round(mean(dat[, "Cov.
@@
 The most recent recommended threshold for $\widehat{R}$ ($\widehat{R} < 1.01$) may be too stringent for MI data.
-\textbf{However, using the 1.2 threshold may be to lenient because this was observed at three iterations and then again after six.
+\textbf{However, using the 1.2 threshold may be too lenient because this was observed at three iterations and then again after six.
 If we would use 1.2, we might miss this increase after three iterations.
@@
 
-An implication of the observed dip in AC is that default maxit value of five iterations is the worst possible number of iterations.
+The observed dip in AC implies that default maxit value of five iterations is the worst possible number of iterations.
 
-\textbf{Future research.} This study considered only a MCAR missingness mechanism.
+\textbf{Future research.} This study considered only an MCAR missingness mechanism.
 Necessary but not sufficient to demonstrate the performance of convergence diagnostics.

### 2019-12-13 Simulation/Writeup/NoteConvergence.tex
results

@@
 \subsection{Convergence diagnostics}
+It is apparant that there is a relation between convergence diagnostics $\widehat{R}$ and $AC$, and the number of iterations per simulation condition($T$).
+Generally speaking, we see a trend towards convergence as $T$ increases with each simulation condition ($\widehat{R}$-values approache one, and $AC$-values approache zero).
+The convergence diagnostics show more or less equivalent trends for chain means versus chain variances.
+We, therefore, only discuss the $\widehat{R}$ and $AC$-values of chain means.
+%convergence diagnostics for the chain means.
+%do not differentiate in the interpretation of chain means .
 
-\textbf{$\widehat{R}$}.
-The $\widehat{R}$-values for chain means and chain variances show equivalent trends, neither is there a clear difference between the four variables.
+Figure \ref{fig:conv}A shows that $\widehat{R}$-values generally decrease with increasing imputation chain lengths.
+The decline stabilizes somewhere between the simulation conditions $T=30$ and $T=50$.
+The downward trend is most pronounced for $T=3$, and between $T = 5$ and $T = 10$.
+In the intervening conditions ($3 \leq T \leq 5$), however, we observe a steep increase in $\widehat{R}$-values.
+This increase implies that convergence cannot be diagnosed at $T=3$, as would be the case based on the conventional threshold $\widehat{R} < 1.2$.
+According to the widely used threshold $\widehat{R} < 1.1$, convergence can be diagnosed for conditions where $T>9$.
+If we use the recently recommended threshold $\widehat{R} < 1.01$, we would conclude that we cannot diagnose convergence in any simulation condition.
+
+%The $\widehat{R}$-values for chain means and chain variances show equivalent trends, neither is there a clear difference between the four variables.
 Roughly speaking, we see that $\widehat{R}$ is higher for simulation conditions with a lower number of iterations, and approaches one for conditions with more iterations.
-Figure \ref{fig:conv} shows an initial dip, a small bump upwards, and then a steep decline.
-This decline \textbf{houdt aan} up-to iteration twenty, after which a more gradual decrease can be observed up-to iteration forty.
+
+%%%%
+%Figure \ref{fig:conv} shows an initial dip, a small bump upwards, and then a steep decline.
+This decline continues up-to iteration twenty, after which a more gradual decrease can be observed up-to iteration forty.
 The $\widehat{R}$-values are more or less stable after that.
-The \textbf{conventionally acceptable threshold for convergence, maximum $\widehat{R} < 1.2$,[CHECK IN INTRO]} is reached in the simulation condition with three iterations, and conditions with six iterations or more.
+%\textbf{[Remove stuff here]} %The conventionally acceptable threshold for convergence, maximum 
+%$\widehat{R} < 1.2$ is reached at $T =3$, and conditions with six iterations or more.
 The widely used threshold $\widehat{R} < 1.1$ is met in conditions with ten or more iterations.
@@
 $\widehat{R} > 1.01$ for all simulation conditions (see online appendix for the full table of results).
-\textbf{Check is computing r ht is still correct!
-Also say something about Rh below 1.}
+\textbf{say something about Rh below 1.}
+%%%%
 
-\textbf{autocorrelation.} All observed autocorrelations (aggregated over the 1000 repetitions) are negative or close to zero.
+The $AC$-values displayed in Figure \ref{fig:conv}B are almost uniformly increasing as a function of $T$.
+The only $AC$-value that deviates from this observation is for $T=2$.
+The lowest $AC$-value is obtained for $T=3$.
+At $T=5$, we reach the same $AC$-value as observed at $T=2$.
+$AC$-values plateau around $T=70$.
+$AC$ is then indifferentiable from zero, indicating stationarity.
+%Simulation conditions where $T>5$ have $AC$-values greater than $T=2$.
+%We see an initial decrease between $T=2$ and $T=3$.
+According to this diagnostic, none of the simulation conditions show signs of non-convergence, since we only observe negative or zero $AC$-values.
+
+% All observed autocorrelations  are negative or close to zero.
 There are some differences between variables, and the ACs of chain variances seem lower than the ACs of chain means.
@@
 \textbf{The lowest AC observed is %-0.735
-at four iterations.
+% at four iterations.
 Simulation conditions with a higher number of iterations show a slow increase towards zero.

### 2019-12-13 Simulation/Writeup/NoteConvergence.Rnw
prelim conclusion

@@
 If we use the recently recommended threshold $\widehat{R} < 1.01$, we would conclude that we cannot diagnose convergence in any simulation condition.
+\textbf{add something about Rh below 1.}
 
@@
 $\widehat{R} > 1.01$ for all simulation conditions (see online appendix for the full table of results).
-\textbf{say something about Rh below 1.}
 %%%%
@@
 
-\textbf{give some implications for these measures. connect convergence and simulation diagnostics.}
+\textbf{From the simulation diagnostics, we see that simulation conditions where $T>3$ are generally ok.
+The performance measures indicate that for the current data, MI procedure, and quantity of interest, the algorithm should have maxit argument of four or more.
+This is, however, a simple simulation problem, with MCAR missingness mechanism, univariate missingness, and moderate fmi's [REMOVE??
+COMPUTE??].
+harder problems require more iterations.} %give some implications for these measures.
+connect convergence and simulation diagnostics.}
+
+There is a discrepancy between what the convergence diagnostics and the performance measures indicate.
+While $T=4$ is ok with repsect to simulation quantities, $T=3$ is the worst possible value according to the convergence diagnostics, and $T=4$ provides marginally better convergence diagnostics.
 
@@
 \textbf{Answer RQ.} How to diagnose non-convergence? It's complicated.
+%We will evaluate how convergence of the imputation algorithm can be diagnosed.
 

### 2019-12-13 Simulation/Writeup/NoteConvergence.tex
prelim conclusion

@@
 This increase implies that convergence cannot be diagnosed at $T=3$, as would be the case based on the conventional threshold $\widehat{R} < 1.2$.
-According to the widely used threshold $\widehat{R} < 1.1$, convergence can be diagnosed for conditions where $T>9$.
+\textbf{This may be spurious.} According to the widely used threshold $\widehat{R} < 1.1$, convergence can be diagnosed for conditions where $T>9$.
 If we use the recently recommended threshold $\widehat{R} < 1.01$, we would conclude that we cannot diagnose convergence in any simulation condition.
+\textbf{add something about Rh below 1.}
 
@@
 The lowest $AC$-value is obtained for $T=3$.
-At $T=5$, we reach the same $AC$-value as observed at $T=2$.
-$AC$-values plateau around $T=70$.
+At $T=5$, the $AC$-value reaches the level observed at $T=2$.
+Between $T=10$ and $T=30$ the increase plateaus off, and $AC$-values are stable around $T=70$.
 $AC$ is then indifferentiable from zero, indicating stationarity.
@@
 % Average bias is not zero because there is only one incomplete dataset to be imputed. Bias is a sample effect.
+%Figure \ref{fig:sim}A shows that bias is fairly stable across simulation conditions.
+The condition $T=1$ clearly deviates from this trend with a negative bias.
+The bias for $T=2$ is below average, but within the range of fluctuations.
+Conditions where $T>3$ \textbf{are ok, Loess line is flat for $T>20$}.
 
-\textbf{Confidence interval width.} CIW is only clearly divergent for the simulation condition with one iteration.
-Conditions with two or more iterations have similar CIWs.
-The magnitude of differences from the loess line is somewhat larger up-to the condition with four iterations.
-From four iterations on, CIW seems to be stable across iterations.
-\textbf{Loess line is never flat.}%One might argue that there is some downward trending up-to thirthy to forty iterations.
+CIW is only clearly divergent for the simulation condition $T=1$.
+However, we want to be on the safe side, and not under-estimate the variance.
+Otherwise, we might end up with spurious inferences.
+Conditions where $T>3$ have similar CIWs, which seems to be stable across iterations.
+\textbf{Loess line is never completely flat.}%One might argue that there is some downward trending up-to thirthy to forty iterations.
 %After about twenty-five iterations the average CIW across all simulation conditions and repetitions is reached 0.928.
 
-\textbf{Coverage rate.} \textbf{The coverage rate is more or less stable from two iterations upwards.
+\textbf{Empirical coverage rate seems more or less stable for conditions where $T>1$.
+We see some over-coverage at $T=2$.
+We might over-estimate the variance, but better than under-estimating it.
 On average, the coverage rate is somewhat higher than the expected nominal coverage of 95\% \cite{neym34}, namely 95.3\%.

### 2019-12-13 Simulation/Writeup/NoteConvergence.Rnw
equations rhat and ac according to notation. simulation hypothesis rewritten

@@
 This is the framework of Markoc chain Monte Carlo %Yet, reference manuals of promiment MI software do mention the importance of evaluating algorithmic convergence (e.g., \cite[p.XYZ]{mice}, \textbf{XYZ, XYZ}).
-This section provides a brief review of:% briefly reviews?
+This section provides a brief review of: \textbf{[INMIDDELS INCORRECT]}% briefly reviews?
 1) how convergence is defined in the general context of iterative algorithmic procedures; 2) what diagnostic tools are available; and 3) which of these may apply to MI algorithms.
@@
 The convergence diagnostics under consideration are $\widehat{R}$ and autocorrelation.
-Perfect convergence is reached when the variance between imputation chains is equivalent to the variance within chains ($\widehat{R} = 1$), and there is no dependency between subsequent samples within imputation chains ($AC = 0$).
+Perfect convergence is reached when the variance between imputation chains is equivalent to the variance within chains ($\widehat{R} = 1$), and there is no dependency between subsequent iterations of imputation chains ($AC = 0$).
 %
@@
 %We can apply $\widehat{R}$ to the mean (or to the first two moments) of the variables of interest.
+
 To define $\widehat{R}$, we follow notation by \citep[p.~5]{veht19}.
@@
 \cite[p.~5]{veth19}
-Let $M$ be the number of chains, $N$ the number of iterations per chain, and $\theta$ the scalar summary of interest (e.g., chain mean or chain variance).
-We compute the variance within each chain of $\theta$ values ($W$) and between the $m$ chains ($B$) as follows:
+Let $M$ be the total number of chains, $T$ the number of iterations per chain, and $\theta$ the scalar summary of interest (e.g., chain mean or chain variance).
+For each chain ($m = 1, 2, \dots, M$), we estimate the variance of $\theta$, and average these to obtain within-chain variance $W$.
+
 %\textbf{REPHRASE: In the equations below, $N$ is the number of draws per chain, $M$ is the number of chains, and $S = MN$ is the total number of draws from all chains.
@@
 \begin{align*}
-B&=\frac{N}{M-1} \sum_{m=1}^{M}\left(\bar{\theta}^{(\cdot m)}-\bar{\theta}^{(\cdot \cdot)}\right)^{2}, \text { where } \bar{\theta}^{(\cdot m)}=\frac{1}{N} \sum_{n=1}^{N} \theta^{(n m)}, \quad \bar{\theta}^{(\cdot \cdot)}=\frac{1}{M} \sum_{m=1}^{M} \bar{\theta}^{(\cdot m)} \\
-W&=\frac{1}{M} \sum_{m=1}^{M} s_{j}^{2},  \text { where } s_{m}^{2}=\frac{1}{N-1} \sum_{n=1}^{N}\left(\theta^{(n m)}-\bar{\theta}^{(\cdot m)}\right)^{2}. %\text{ \cite[p.~5]{veht19}} 
+W&=\frac{1}{M} \sum_{m=1}^{M} s_{j}^{2},  \text { where } s_{m}^{2}=\frac{1}{T-1} \sum_{t=1}^{T}\left(\theta^{(t m)}-\bar{\theta}^{(\cdot m)}\right)^{2}. %\text{ \cite[p.~5]{veht19}} 
 \end{align*}
 
-From the between and within chain variances we compute a weighted average, $\widehat{\operatorname{var}}^{+}$, which over-estimates the total variance of $\theta$.
-$\widehat{R}$ is then obtained as a ratio between the over-estimated total variance and the within chain variance:
+We then estimate between-chain variance $B$ as the variance of the collection of average $\theta$ per chain.
+
+\begin{align*}
+B&=\frac{T}{M-1} \sum_{m=1}^{M}\left(\bar{\theta}^{(\cdot m)}-\bar{\theta}^{(\cdot \cdot)}\right)^{2}, \text { where } \bar{\theta}^{(\cdot m)}=\frac{1}{T} \sum_{t=1}^{T} \theta^{(t m)}, \quad \bar{\theta}^{(\cdot \cdot)}=\frac{1}{M} \sum_{m=1}^{M} \bar{\theta}^{(\cdot m)} \\
+ %\text{ \cite[p.~5]{veht19}} 
+\end{align*}
+
+From the between- and within-chain variances we compute a weighted average, $\widehat{\operatorname{var}}^{+}$, which over-estimates the total variance of $\theta$.
+$\widehat{R}$ is then obtained as a ratio between the over-estimated total variance and the within-chain variance:
 
@@
 %The potential scale reduction factor $\widehat{R}$ is obtained as:
-We can interpret $\widehat{R}$ as potential scale reduction factor since it indicates by how much the variance of $\theta$ could be shrunken down if an infinite number of iterations per chain would be run \citep[$t \to \infty$;][]{gelm92}.
-High $\widehat{R}$-values, therefore, indicate slow mixing of chains, e.g., because of dependence across chains on initial values.
+We can interpret $\widehat{R}$ as potential scale reduction factor since it indicates by how much the variance of $\theta$ could be shrunken down if an infinite number of iterations per chain would be run \citep{gelm92}.
+This interpretation assumes that chains are `over-dispersed' at $t=1$, and reach convergence as $T \to \infty$.
+Over-dispersion implies that the initial values of the chains are `far away' from the target distribution and each other.
+The iteration at which chains are independent of their initial values is the point where mixing occurs.
+This is diagnosed with $\widehat{R}$-values close to one.
+%If there is no added value
+High $\widehat{R}$-values thus indicate slow mixing of chains, which is a sign of non-convergence.
+%, e.g., because chains may depend on their initial value.
+%are of dependence across chains on initial values.
 Once there is no dependence on initial values anymore, there is no added value of continuing iterations.
@@
 %If there is no added value compared to infinitely many iterations, the chains have converged. 
-The conventionally acceptable threshold was $\widehat{R} < 1.2$ \cite{gelm92}. Recently, \cite{veht19} proposed a more stringent threshold: $\widehat{R} < 1.01$.
+The conventionally acceptable threshold for this convergence diagnostic was $\widehat{R} < 1.2$ \citep{gelm92}. Recently, \cite{veht19} proposed a more stringent threshold: $\widehat{R} < 1.01$.
 
-Following the same notation, autocorrelation is defined as the correlation between two subsequent $\theta$ values within the same chain.
-We only consider $AC$ at lag 1, i.e., between the $t^{th}$ and $t+1^{th}$ iteration of each chain:
+Following the same notation, we define autocorrelation as the correlation between two subsequent $\theta$-values within the same chain\footnotemark  \citep[p.~147]{lync07}. %. 
 %
 \begin{equation*}
-AC = \left( \frac{T}{T-1} \right) \frac{\sum_{t=1}^{T-1}(\theta_t - \bar{\theta}^{(\cdot m)})(\theta_{t+1} - \bar{\theta}^{(\cdot m)})}{\sum_{t=1}^{T}(\theta_t - \bar{\theta}^{(\cdot m)})^2},
+AC = \left( \frac{T}{T-1} \right) \frac{\sum_{t=1}^{T-1}(\theta_t - \bar{\theta}^{(\cdot m)})(\theta_{t+1} - \bar{\theta}^{(\cdot m)})}{\sum_{t=1}^{T}(\theta_t - \bar{\theta}^{(\cdot m)})^2}.
 \end{equation*}
 %
-where $T$ is the total number of iterations per chain \citep[p.~147]{lync07}.
-We can interpret the sign of $AC$-values as a measure of dependence within chains ($-1 \leq AC \leq 1$).
-Positive $AC$-values indicate recurrence: similar $\theta$-values for subsequent iterations, which may lead to trending.
-Negative $AC$-values also indicate dependence, but do not threaten the stationarity component of convergence \textbf{[REMOVE INTERPRETATION HERE?]}.
+\footnotetext{In this study we only consider $AC$ at lag 1, i.e., the correlation between the $t^{th}$ and $t+1^{th}$ iteration of the same chain.}
+
+We can interpret %the sign and magnitude of ... Any non-zero $AC$-value ($-1 \leq AC \leq 1$) . $AC = 0$ indicates independence between $\theta$-values.
+$AC$-values as a measure of dependence within chains.
+Positive $AC$-values indicate recurrence (i.e., similarity between $\theta$-values of subsequent iterations), which may lead to trending.
+Negative $AC$-values occur when the $\theta$-values of subsequent iterations diverge from oneanother.
+Dependence in this sense does not threaten the stationarity component of convergence.
+%that we aim to evaluate with $AC$.
+On the contrary even---it may increase the variance of $\theta$.
+\textbf{speed up convergence???}  %have also indicate dependence, but do not threaten the stationarity component of convergence \textbf{[REMOVE INTERPRETATION HERE?]}.
 Instead of recurrence, there would be a `seesaw' effect within chains: subsequent $\theta$-values would deviate in opposite directions from the chain average $\theta^{(\cdot m)})$, increasing the variance of $\theta$.
-As convergence diagnostic, the interest is in positive $AC$-values.\footnotemark
-\footnotetext{The magnitude of $AC$-values can also be evaluated statistically, but that is outside of this note's scope.}
+As convergence diagnostic, the interest is in positive $AC$-values.%\footnotemark 
+\footnote{The magnitude of $AC$-values can also be evaluated statistically, but that is outside of this note's scope.}
 
@@
 This study evaluates whether $\widehat{R}$ and $AC$ could diagnose convergence of multiple imputation algorithms.
-There is, however, no baseline measure available to evaluate their performance against.\footnote{Only severely pathological cases of non-convergence stand out when visually inspecting the imputation chains.} We will assess the appropriateness of the two convergence diagnostics by comparison with simulation quantities.
-The perfrormance diagnostics are the evaluation criteria  recommended by \citep[\S~2.5.2]{buur18}, whcih comprise of average bias, average confidence interval width, and empirical coverage rate across simulations.% %"We expect that in accordance with Neyman (1934), at least 95 percent of the confidence intervals should contain the true population value, although some room for simulation error should be taken into account to counteract the finite nature of our simulations"\cite{schou18}.
+There is, however, no baseline measure available to evaluate their performance against.
+\footnote{The current practice of visually inspecting imputation chains for signs of non-mixing or non-stationarity can only diagnose severely pathological cases of non-convergence.} 
+We will assess the appropriateness of the two convergence diagnostics in comparison to evaluation criteria recommended by \citep[\S~2.5.2]{buur18}.
+The performance diagnostics comprise of average bias, average confidence interval width, and empirical coverage rate across simulations.% %"We expect that in accordance with Neyman (1934), at least 95 percent of the confidence intervals should contain the true population value, although some room for simulation error should be taken into account to counteract the finite nature of our simulations"\cite{schou18}.
 \footnote{We could also look at distributional characteristics, and plausibility of imputed values, see \cite{vinknd} (n.d.).
@@
 
-\textbf{We hypothesize that $\widehat{R}$ will over-estimate non-convergence. No hypothesis was formulated about AC, results are exploratory.
-%
-[MOVE THIS TO DISCUSSION?] Assessing the stationarity component of convergence with $AC$ might be redundant, since high $AC$-values are implausible in MI procedures.
-That is, the randomness induced by the MI algorithm effectively mitigates the risk of dependency within chains.
-$AC$ would thus not be informative of the convergence of MI algorithms.
-%
-Moreover, we hypothesize that $\widehat{R}$ may not be an appropriate diagnostic either, because it assumes over-dispersed initial values (i.e., the starting points of the chains are `far away' from the target distribution and each other).
+\textbf{[HIER NOG ZOOI]} We hypothesize that $\widehat{R}$ may not be an appropriate diagnostic, because it assumes over-dispersed initial values.
 In \pkg{mice}, initial values of the algorithm are chosen randomly from the observed data.
 Therefore, we cannot be certain that the initial values are over-dispersed.
-%The initial parameter value is noodzakelijk.
-Daarom is het lastig om in MI that starten vanuit overdisp state.
+%The initial parameter value is noodzakelijk. Daarom is het lastig om in MI that starten vanuit overdisp state. 
 % Starting values of MICE chains: "For each j, fill in starting imputations $Y^0_j$ by random draws from $Y^{obs}_j$." \cite[\S~4.5.2]{buur18} before for loops over variables and imputations, see \url{https://stefvanbuuren.name/fimd/sec-FCS.html#def:mice}
@@
 This suggests that $\widehat{R}$ would over-estimate non-convergence of MI algorithms.
-}
+
+\textbf{We hypothesize that $\widehat{R}$ will over-estimate non-convergence. No hypothesis was formulated about AC, results are exploratory.}
+%
+%
 
@@
 
-\textbf{Answer RQ.} How to diagnose non-convergence? It's complicated.
+This note shows that the convergence of multiple imputation algorithms may be evaluated with convergence diagnostics $\widehat{R}$ and $AC$.
+These measures, however, indicate non-convergence at imputation chain lengths that produce acceptable simulation diagnostics.
+This may be due to the quantity of scientific interest chosen.
+More `complicated' $Q$s (e.g., higher order effects or variance components) might show bias, under- or over-coverage at higher $T$.
+
+At least $\widehat{R}$ does not under-estimate non-convergence.
+It's better to be too strict than too lenient.
+According to this simulation study, the recently proposed threshold $\widehat{R}$ is too stringent for MI algorithms, and the conventionally acceptable $\widehat{R} <1.2$ is too lenient.
+
+%\textbf{Answer RQ.} How to diagnose non-convergence? It's complicated.
 %We will evaluate how convergence of the imputation algorithm can be diagnosed.
@@
 
-From the simulation diagnostics, it appears that as little as three iterations might be sufficient to obtain unbiased, confidence valid estimates.
+From the simulation diagnostics, it appears that as little as four iterations might be sufficient to obtain unbiased, confidence valid estimates.
 Convergence diagnostics $\widehat{R}$ and autocorrelation, however, indicate that convergence may only be reached after twenty or even forty iterations.
@@
 This can happen when the number of simulations is smaller than in `regular' MCMC processes [explain that fewer iterations is an advantage of MI, not a disadvantage compared to MCMC].
-Increasing Rhat values mean that the initial values were not appropriately over-dispersed \citep[p~438]{broo98}.
+Increasing Rhat values can happen if the initial values were not appropriately over-dispersed \citep[p~438]{broo98}.
 %Therefore, the `$(n-1/n)$' [add equation number] correction factor can influence the estimated potential scale reduction factor.
@@
 But why the dip??
+
+Assessing the stationarity component of convergence with $AC$ might be redundant, since high $AC$-values are implausible in MI procedures.
+That is, the randomness induced by the MI algorithm effectively mitigates the risk of dependency within chains.
+$AC$ would thus not be informative of the convergence of MI algorithms.
 

### 2019-12-14 Simulation/Writeup/NoteConvergence.Rnw
shorter abstract

@@
 \Abstract{
-The ubiquitous problem of missing information may be circumvented by implementing the framework of multiple imputation (MI).
+This note investigates how algorithmic convergence of multiple imputation procedures---methodology to circumvent the ubiquitous problem of missing data---may be diagnosed.
+We conclude that potential scale reduction factor $\widehat{R}$ and autocorrelation may be appropriate convergence diagnostics, but their performance differs from conventional applications to iterative algorithmic procedures.
+These results will be implemented into an interactive framework for the evaluation of multiple imputation methods that is under development as research Master thesis project.
+%%%%%%%%%%%%%
+%The ubiquitous problem of missing information may be circumvented by implementing the framework of multiple imputation (MI).
 With MI, missing data points are `filled in' (i.e.
@@
 Another option would be to attach this note as an online appendix to the technical paper on ShinyMICE.}
+%%%%%%%%%%%%
 % This Research Report contains a simulation study that serves as the basis of the technical paper that will be submitted for publication in \emph{Journal of Statistical Software}.
@@
 
-The intended audience of this note consists of empirical researchers and statisticians who use multiple imputation to solve missing data problems.
-Basic familiarity with MI methodology is assumed.
-For the theoretical foundation of MI, see \cite{rubin87}.
-For an accessible and comprehensive introduction to MI from an applied perspective, see \cite{buur18}.
-
-The convergence guidelines introduced in this paper are developed to be integrated into the \pkg{mice} environment \citep{mice} in \proglang{R} \citep{R}.
-This note, therefore, follows notation and conventions of \cite{mice}.
-Deviations from the `original' notation by \cite{rubin87} are described in \citep[\S~2.2.3]{buur18}.
+The intended audience of this note consists of anyone who uses multiple imputation to solve missing data problems.
+Basic familiarity with MI methodology is assumed.\footnote{For the theoretical foundation of MI, see \cite{rubin87}.
+For an accessible and comprehensive introduction to MI from an applied perspective, see \cite{buur18}.}
+This note follows notation and conventions of \cite{mice}. % to suit the \pkg{mice} environment. %Deviations from the `original' notation by \cite{rubin87} are described in \citep[\S~2.2.3]{buur18}.
 
 Let $Y$ denote an $n \times p$ matrix containing the data values on $p$ variables for all $n$ units in a sample.
-The collection of observed data values in $Y$ is denoted as $Y_{obs}$; the missing part of $Y$ is referred to as $Y_{mis}$.
-\textbf{REMOVE?
+The collection of observed data values in $Y$ is denoted by $Y_{obs}$, and will be referred to as `incomplete' or `observed' data interchangeably.
+The missing part of $Y$ is denoted by $Y_{mis}$.
+Which parts of $Y$ are missing is determined by the `missingness mechanism'.
+%The relation between $Y_{obs}$ and $Y_{mis}$ \textbf{REMOVE?
 Response indicator $R$ shows whether a data value in $Y$ is missing or observed.} The relation between $R$, $Y_{obs}$, and $Y_{mis}$ determines the missingness mechanism.
@@
 
-In this note, the terms `unobserved' and `missing' data are used interchangeably to refer to $Y_{mis}$.
+%In this note, the terms `unobserved' and `missing' data are used interchangeably to refer to $Y_{mis}$.
 The terms `incomplete' or `observed' data denote $Y_{obs}$.
 Incomplete data is the starting point of the multiple imputation procedure.
+
 Figure \ref{fig:steps} provides an overview of the steps involved with MI.
+Missing data in $Y$ is `imputed' (i.e., filled in) $m$ times.
+The imputed data is combined with observed data $Y_{obs}$ to create $m$ completed data sets.
+On each completed data set, the analysis of scientific interest (or `complete data model') is performed.
+The quantity of scientific interest (e.g., a regression coefficient) is denoted with $Q$.
+Since $Q$ is estimated on each completed data set, $m$ separate $\hat{Q}$ values are obtained.
+These $m$ values are combined into a single pooled estimate $\bar{Q}$.
 
@@
 
-Missing data in $Y$ is `imputed' (i.e., filled in) $m$ times.
-The imputed data is combined with observed data $Y_{obs}$ to create $m$ completed data sets.
-On each completed data set, the analysis of scientific interest (or `complete data model') is performed.
-The quantity of scientific interest (e.g., a regression coefficient) is denoted with $Q$.
-Since $Q$ is estimated on each completed data set, $m$ separate $\hat{Q}$ values are obtained.
-These $m$ values are combined into a single pooled estimate $\bar{Q}$.
-
 This note focuses on the algorithmic properties of the imputation step.
@@
 The chains are terminated after a predefined number of iterations.
-Only the ultimate sample that a chain lands on is imputed and subsequently used in the analysis and pooling steps.
+Only the ultimate sample that a chain lands on is imputed, and subsequently used in the analysis and pooling steps.
 The collection of samples across iterations (between the initial value and the imputed value) for each of the $m$ imputations will be referred to as an `imputation chain'.
@@
 Positive $AC$-values indicate recurrence (i.e., similarity between $\theta$-values of subsequent iterations), which may lead to trending.
-Negative $AC$-values occur when the $\theta$-values of subsequent iterations diverge from oneanother.
+Negative $AC$-values occur when the $\theta$-values of subsequent iterations diverge from one-another.
 Dependence in this sense does not threaten the stationarity component of convergence.
@@
 \subsection{Convergence diagnostics}
-It is apparant that there is a relation between convergence diagnostics $\widehat{R}$ and $AC$, and the number of iterations per simulation condition($T$).
-Generally speaking, we see a trend towards convergence as $T$ increases with each simulation condition ($\widehat{R}$-values approache one, and $AC$-values approache zero).
+It is apparent that there is a relation between convergence diagnostics $\widehat{R}$ and $AC$, and the number of iterations per simulation condition($T$).
+Generally speaking, we see a trend towards convergence as $T$ increases with each simulation condition ($\widehat{R}$-values approach one, and $AC$-values approach zero).
 The convergence diagnostics show more or less equivalent trends for chain means versus chain variances.
@@
 This increase implies that convergence cannot be diagnosed at $T=3$, as would be the case based on the conventional threshold $\widehat{R} < 1.2$.
-\textbf{This may be spurious.} According to the widely used threshold $\widehat{R} < 1.1$, convergence can be diagnosed for conditions where $T>9$.
+\textbf{[This may be spurious??]} According to the widely used threshold $\widehat{R} < 1.1$, convergence can be diagnosed for conditions where $T>9$.
 If we use the recently recommended threshold $\widehat{R} < 1.01$, we would conclude that we cannot diagnose convergence in any simulation condition.
-\textbf{add something about Rh below 1.}
+\textbf{[Add something about $\widehat{R} <1$?]}
 
@@
 \subsection{Simulation diagnostics}
-We use average bias, average condidence interval width, and coverage rate as performance measures to evaluate $\widehat{R}$ and $AC$.
+We use average bias, average confidence interval width, and coverage rate as performance measures to evaluate $\widehat{R}$ and $AC$.
 We make the general observation that the simulation diagnostics behave as theorized for most simulation conditions (bias around zero, coverage rate around 95\%).
@@
 The bias for $T=2$ is below average, but within the range of fluctuations.
-Conditions where $T>3$ \textbf{are ok, Loess line is flat for $T>20$}.
+Conditions where $T>3$ \textbf{are OK, Loess line is flat for $T>20$}.
 %the average bias between the regression coefficient of $X$ on $Y$ in the complete dataset, and $\bar{Q}$, \textbf{the pooled estimated regression coefficient and the ...
@@
 
-\textbf{From the simulation diagnostics, we see that simulation conditions where $T>3$ are generally ok.
+\textbf{From the simulation diagnostics, we see that simulation conditions where $T>3$ are generally OK.
 The performance measures indicate that for the current data, MI procedure, and quantity of interest, the algorithm should have maxit argument of four or more.
@@
 There is a discrepancy between what the convergence diagnostics and the performance measures indicate.
-While $T=4$ is ok with repsect to simulation quantities, $T=3$ is the worst possible value according to the convergence diagnostics, and $T=4$ provides marginally better convergence diagnostics.
+While $T=4$ is OK with respect to simulation quantities, $T=3$ is the worst possible value according to the convergence diagnostics, and $T=4$ provides marginally better convergence diagnostics.
 

### 2019-12-14 Simulation/Writeup/NoteConvergence.Rnw
additions to discussion (future research topics)

@@
 % for diagnostics for evalua one of these evaluation measures that is vital in evaluating MI procedures: a diagnostic to assess convergence of the MI algorithm.
-For reasons of brevity, this note only focuses on the MI algorithm implemented in \pkg{mice} \cite{mice}. %The implementation limits itself to the MI algorithm 
+For reasons of brevity, this note only focuses on the MI algorithm implemented in \pkg{mice} \citep{mice}. %The implementation limits itself to the MI algorithm 
 %More specifically, this note focuses on the MI algorithm that is implemented in the \proglang{R} package \pkg{mice}: `Multiple Imputation using Chained Equations' (MICE) \textbf{[can be removed, can be found in the mice ref]}.
@@
 For an accessible and comprehensive introduction to MI from an applied perspective, see \cite{buur18}.}
-This note follows notation and conventions of \cite{mice}. % to suit the \pkg{mice} environment. %Deviations from the `original' notation by \cite{rubin87} are described in \citep[\S~2.2.3]{buur18}.
+This note follows notation and conventions of \pkg{mice} \citep{mice}.
+% to suit the \pkg{mice} environment.
+%Deviations from the `original' notation by \cite{rubin87} are described in \citep[\S~2.2.3]{buur18}.
 
@@
 
-To define $\widehat{R}$, we follow notation by \citep[p.~5]{veht19}.
+To define $\widehat{R}$, we follow notation by \cite[p.~5]{veht19}.
 % The scalar summary of interest in this study is the average of the imputed values per variable.
@@
 \footnote{The current practice of visually inspecting imputation chains for signs of non-mixing or non-stationarity can only diagnose severely pathological cases of non-convergence.} 
-We will assess the appropriateness of the two convergence diagnostics in comparison to evaluation criteria recommended by \citep[\S~2.5.2]{buur18}.
+We will assess the appropriateness of the two convergence diagnostics in comparison to evaluation criteria recommended by \cite[\S~2.5.2]{buur18}.
 The performance diagnostics comprise of average bias, average confidence interval width, and empirical coverage rate across simulations.% %"We expect that in accordance with Neyman (1934), at least 95 percent of the confidence intervals should contain the true population value, although some room for simulation error should be taken into account to counteract the finite nature of our simulations"\cite{schou18}.
@@
 The point of origin for all simulation conditions is a finite population of $N=1000$.
-The data are simulated to solve a multiple linear regression problem, where dependent variable $Y$ is regressed on independent variable $X$, and covariates $Z_1$ and $Z_2$: $Y \sim \beta_1 X + \beta_2 Z_1 + \beta_3 Z_2$.
+The data are simulated to solve a multiple linear regression problem, where dependent variable $Y$ is regressed on independent variable $X$, and covariates $Z_1$ and $Z_2$:
+
+$$Y \sim \beta_1 X + \beta_2 Z_1 + \beta_3 Z_2.$$
+
 The quantity of scientific interest is regression coefficient $\beta_1$.
@@
 
-Outcome variable $Y$ is subsequently calculated as $Y =  2X + .5Z_1 - Z_2 + \epsilon$.
+Outcome variable $Y$ is subsequently calculated as
+
+$$Y =  2X + .5Z_1 - Z_2 + \epsilon .$$
 
@@
 We compute convergence diagnostics separately for each of the four variables in the dataset, and report the maximum (absolute) value as the `worst linear predictor' of convergence \textbf{[Rephrase or footnote??]}.
-$\widehat{R}$ is computed by implementing the definition given in \textbf{EQUATION NR}.
+$\widehat{R}$ is computed by implementing Vehtari et al.'s \citeyear{veht19} recommendations.
 $AC$ is computed with \pkg{stats} function \fct{acf}.
@@
 Confidence interval width (CIW) is defined as the difference between the lower and upper bound of the 95\% confidence interval around $\bar{Q}$ (CI95\%).
-We compute the CI95\% bounds as $\bar{Q} \pm 2.66 \times SE_{\bar{Q}}$, where 2.66 is the quantile of a $t$-distribution with $m-1$ degrees of freedom, and $SE_{\bar{Q}}$ is the square root of the pooled variance estimate.
+We compute the CI95\% bounds as
+
+$$\bar{Q} \pm 2.66 \times SE_{\bar{Q}},$$
+
+where 2.66 is the quantile of a $t$-distribution with $m-1$ degrees of freedom, and $SE_{\bar{Q}}$ is the square root of the pooled variance estimate.
 From bias and CIW, we calculate empirical coverage rates.
@@
 We might over-estimate the variance, but better than under-estimating it.
-On average, the coverage rate is somewhat higher than the expected nominal coverage of 95\% \cite{neym34}, namely \Sexpr{100*round(mean(dat[, "Cov.
+On average, the coverage rate is somewhat higher than the expected nominal coverage of 95\% \citep{neym34}, namely \Sexpr{100*round(mean(dat[, "Cov.
 rate"]), 3)}\%.
@@
 
+\textbf{Does this mean that conv is not nec to obtain correct inferences?}
+
 \textbf{R hat below 1.}
@@
 
+Also, conv diag for substantive model? wald test for AC = 0? 
 %% -- Optional special unnumbered sections -------------------------------------

### 2019-12-14 Simulation/Writeup/NoteConvergence.tex
added empirical CIs to graphs

@@
 We might over-estimate the variance, but better than under-estimating it.
-On average, the coverage rate is somewhat higher than the expected nominal coverage of 95\% \cite{neym34}, namely 95.3\%.
+On average, the coverage rate is somewhat higher than the expected nominal coverage of 95\% \citep{neym34}, namely 95.3\%.
 Loess line is never flat.}

### 2019-12-14 Simulation/Functions/Evaluate.R

@@ -1,18 +1,57 @@
 # Extract aggregated simulation results by averaging over runs
+# requires the packages "purrr" and "data.table"
 
-evaluate.sim <- function(sims) {
-  # organize output for evaluation
-  dt <- map(sims, as.data.table) %>% rbindlist(fill = TRUE)
-  d <- dt[, lapply(.SD, mean)] %>% as.data.frame
-  tab <- 1:n.iter %>% as.data.frame()
-  vars <- c("bias", "CIW", "cov", "R_mean_X", "R_mean_Z1", "R_mean_Z2", "R_mean_Y", "R_var_X", "R_var_Z1", "R_var_Z2", "R_var_Y", "AC_mean_X", "AC_mean_Z1", "AC_mean_Z2", "AC_mean_Y", "AC_var_X", "AC_var_Z1", "AC_var_Z2", "AC_var_Y")
-  for (var in vars) {
-    tab <- cbind(tab, d[,grep(var, names(d))] %>% t) 
+evaluate.sim <-
+  function(sims,
+           n.iter = n.iter,
+           mean_or_SE = "mean"
+           ) {
+    # organize output for evaluation
+    dt <- map(sims, as.data.table) %>% rbindlist(fill = TRUE)
+    # determine whether we want averages per simulation condition or MCMC SEs
+    if (mean_or_SE == "mean") {
+      function_to_apply <- mean
+    } else if (mean_or_SE == "se") {
+      function_to_apply <- sd # or use: function(x) sqrt((var(x)))
+    } else if (mean_or_SE == "lower") {
+      function_to_apply <- function(x) quantile(x, .025, na.rm = TRUE)
+    } else if (mean_or_SE == "upper") {
+      function_to_apply <- function(x) quantile(x, .975, na.rm = TRUE)
+    }
+    
+    # apply function to aggregate simulation runs per simulation condition
+    d <- dt[, lapply(.SD, function_to_apply)] %>% as.data.frame
+    # create object for output
+    output <- 1:n.iter %>% as.data.frame()
+    # attatch diagnostics
+    vars <-
+      c(
+        "bias",
+        "CIW",
+        "cov",
+        "R_mean_X",
+        "R_mean_Z1",
+        "R_mean_Z2",
+        "R_mean_Y",
+        "R_var_X",
+        "R_var_Z1",
+        "R_var_Z2",
+        "R_var_Y",
+        "AC_mean_X",
[…]

### 2019-12-14 Simulation/Writeup/NoteConvergence.Rnw
Removed A LOT in intro

@@
 These results will be implemented into an interactive framework for the evaluation of multiple imputation methods that is under development as research Master thesis project.
+\textbf{TO DO still: check ref list, add legend to sim diags, check table and fig captions (abbrevations), references to tables, figs, appendix}
 %%%%%%%%%%%%%
 %The ubiquitous problem of missing information may be circumvented by implementing the framework of multiple imputation (MI).
-With MI, missing data points are `filled in' (i.e.
+With MI, missing datapoints are `filled in' (i.e.
 imputed) several times, and statistical inference is performed on each imputed data set.
@@
 A popular answer to the ubiquitous problem of missing information is to use the framework of multiple imputation (MI), proposed by \cite{rubin87}.
-MI is an iterative algorithmic procedure in which each missing data point is `imputed' (i.e.
+MI is an iterative algorithmic procedure in which each missing datapoint is `imputed' (i.e.
 filled in) several times.
@@
 %Therefore, I will answer the following question: 'Which measures are vital for evaluating the validity of multiply imputed data?'.
+The goal of this research project is to develop novel methodology and guidelines for evaluating MI methods.
+These tools will subsequently be implemented in an interactive evaluation framework for multiple imputation, which will aid applied researchers in drawing valid inference from incomplete datasets.
 
-The goal of this research project is to develop novel methodology and guidelines for evaluating MI methods.
-These tools will subsequently be implemented in an interactive evaluation framework for multiple imputation to aid applied researchers in drawing valid inference from incomplete datasets.
-This note provides the theoretical foundation towards the diagnostic evaluation of the convergence of MI algorithms.
+This note provides the theoretical foundation towards the diagnostic evaluation of %the convergence of 
+multiple imputation algorithms. For reasons of brevity, we only focus on the MI algorithm implemented in \pkg{mice} \citep{mice}. 
+The convergence properties of the MI algorithm in \pkg{mice} are investigated through model-based simulation.%The results of this simulation study are guidelines for assessing convergence of MI algorithms.
+These guidelines will be implemented in an interactive evaluation tool for \pkg{mice}, `ShinyMICE', which is currently under development.
+\footnote{All programming code used in this note is available from \href{https://github.com/gerkovink/shinyMice/simulation}{github.com/gerkovink/ShinyMICE/simulation}.} 
+%I will replicate Lacerda et al.'s simulation study on $\widehat{R}$ \citep{lace07}, and develop novel guidelines for assessing convergence.
+Ideally, I will integrate several diagnostics (e.g., $\widehat{R}$, \emph{autocorrelation}, and \emph{simulation error}) into a single summary indicator to flag non-convergence.
+We will evaluate how convergence of the MI algorithm can be diagnosed.
+
 % for diagnostics for evalua one of these evaluation measures that is vital in evaluating MI procedures: a diagnostic to assess convergence of the MI algorithm.
-For reasons of brevity, this note only focuses on the MI algorithm implemented in \pkg{mice} \citep{mice}. %The implementation limits itself to the MI algorithm 
+%The implementation limits itself to the MI algorithm
 %More specifically, this note focuses on the MI algorithm that is implemented in the \proglang{R} package \pkg{mice}: `Multiple Imputation using Chained Equations' (MICE) \textbf{[can be removed, can be found in the mice ref]}.
-To this end, we will evaluate how convergence of the imputation algorithm can be diagnosed.
 
@@
 
-The convergence properties of the MI algorithm in \pkg{mice} are investigated through model-based simulation.
-The results of this simulation study are guidelines for assessing convergence of MI algorithms.
-These guidelines will be implemented in an interactive evaluation tool for \pkg{mice}, `ShinyMICE', which is currently under development.
-All programming code used in this note is available from \href{https://github.com/gerkovink/shinyMice/simulation}{github.com/gerkovink/ShinyMICE/simulation}.
-%I will replicate Lacerda et al.'s simulation study on $\widehat{R}$ \citep{lace07}, and develop novel guidelines for assessing convergence.
-Ideally, I will integrate several diagnostics (e.g., $\widehat{R}$, \emph{autocorrelation}, and \emph{simulation error}) into a single summary indicator to flag non-convergence.
-
 %% -- Features ---------------------------------------------------------------
@@
 
-The intended audience of this note consists of anyone who uses multiple imputation to solve missing data problems.
-Basic familiarity with MI methodology is assumed.\footnote{For the theoretical foundation of MI, see \cite{rubin87}.
-For an accessible and comprehensive introduction to MI from an applied perspective, see \cite{buur18}.}
+%The intended audience of this note consists of anyone who uses multiple imputation to solve missing data problems.
+
 This note follows notation and conventions of \pkg{mice} \citep{mice}.
@@
 %Deviations from the `original' notation by \cite{rubin87} are described in \citep[\S~2.2.3]{buur18}.
+Basic familiarity with MI methodology is assumed.\footnote{For the theoretical foundation of MI, see \cite{rubin87}.
+For an accessible and comprehensive introduction to MI from an applied perspective, see \cite{buur18}.}
 
@@
 Response indicator $R$ shows whether a data value in $Y$ is missing or observed.} The relation between $R$, $Y_{obs}$, and $Y_{mis}$ determines the missingness mechanism.
-This note only considers a `missing completely at random' (MCAR) mechanism, where the probability of being missing is equal for all $n \times p$ cells in $Y$.
+This note only considers a `missing completely at random' (MCAR) mechanism, where the probability of being missing is equal for all $n \times p$ cells in $Y$ \citep{rubin87}.
 
@@
 
-Figure \ref{fig:steps} provides an overview of the steps involved with MI.
-Missing data in $Y$ is `imputed' (i.e., filled in) $m$ times.
-The imputed data is combined with observed data $Y_{obs}$ to create $m$ completed data sets.
-On each completed data set, the analysis of scientific interest (or `complete data model') is performed.
-The quantity of scientific interest (e.g., a regression coefficient) is denoted with $Q$.
-Since $Q$ is estimated on each completed data set, $m$ separate $\hat{Q}$ values are obtained.
-These $m$ values are combined into a single pooled estimate $\bar{Q}$.
+Figure \ref{fig:steps} provides an overview of the steps involved with MI---from incomplete data, to $m$ completed datasets, to $m$ estimated quantities of interest ($\hat{Q}$s), to a single pooled estimate $\bar{Q}$.
+This note focuses on the algorithmic properties of the imputation step.
+The MI algorithm has an iterative nature.
+For each missing datapoint in $Y_{mis}$, $m$ `chains' of potential values are sampled.
+Only the ultimate sample that each chain lands on is imputed.
+The number of iterations per chain will be denoted with $T$, where $t$ varies over the integers $1, 2, \dots, T$.
+%This is repeated $m$ times.
+Imputed values are the ultimate samples of a `chain' of For each imputed value ($t = T$), a s for each missing datapoint in $Y_{mis}$, .
+Each of the $m$ chains starts with an initial value, drawn randomly from $Y_{obs}$.
+The chains are terminated after a predefined number of iterations.
+, and subsequently used in the analysis and pooling steps.
+The collection of samples between the initial value (at $t=1$) and the imputed value (at $t=T$) will be referred to as an `imputation chain'.
 
@@
 \end{tikzpicture}}
-\caption{Scheme of the main steps in multiple imputation ($m = 3$). Adapted from \cite[\S~1.4.1]{buur18}.}
+\caption{Scheme of the main steps in multiple imputation \citep[$m = 3$; adapted from][\S~1.4.1]{buur18}.
+{\footnotesize Missing data in dataset $Y$ is `imputed' (i.e., filled in) $m$ times.
+The imputed data is combined with the observed data ($Y_{obs}$) to create $m$ completed datasets.
+On each completed dataset, the analysis of scientific interest (or `substantive model') is performed.
+The quantity of scientific interest (e.g., a regression coefficient) is denoted with $Q$.
+Since $Q$ is estimated on each completed dataset, $m$ separate $\hat{Q}$-values are obtained.
+These $m$ values are combined into a single pooled estimate $\bar{Q}$.}} %\footnotemark}
 \end{figure}
 
-This note focuses on the algorithmic properties of the imputation step.
-The algorithm employed within this step has an iterative nature.
-That is, before drawing $m$ imputed values for each missing data point in $Y_{mis}$, a `chain' of potential values is sampled.
-Each of the $m$ chains starts with an initial value, drawn randomly from $Y_{obs}$.
-The chains are terminated after a predefined number of iterations.
-Only the ultimate sample that a chain lands on is imputed, and subsequently used in the analysis and pooling steps.
-The collection of samples across iterations (between the initial value and the imputed value) for each of the $m$ imputations will be referred to as an `imputation chain'.
-The total number of iterations per chain will be denoted with $T$, where $t$ varies over the integers $1, 2, \dots, T$.
+%\footnotetext{Figure .}
 
@@
 This is the framework of Markoc chain Monte Carlo %Yet, reference manuals of promiment MI software do mention the importance of evaluating algorithmic convergence (e.g., \cite[p.XYZ]{mice}, \textbf{XYZ, XYZ}).
-This section provides a brief review of: \textbf{[INMIDDELS INCORRECT]}% briefly reviews?
-1) how convergence is defined in the general context of iterative algorithmic procedures; 2) what diagnostic tools are available; and 3) which of these may apply to MI algorithms.
+%%%%%%%
+%This section provides a brief review of: \textbf{[INMIDDELS INCORRECT]}% briefly reviews?
+%1) how convergence is defined in the general context of iterative algorithmic procedures; 2) what diagnostic tools are available; and 3) which of these may apply to MI algorithms.
 
@@
 
-\textbf{[COULD BE SHORTER!]} Most convergence diagnostics target either of the two components.
+\textbf{[COULD BE SHORTER!
+Maybe in footnote??]} Most convergence diagnostics target either of the two components.
 The stationarity component of convergence may be evaluated with autocorrelation \citep[$AC$; ][]{scha97, gelm13}, numeric standard error \citep[or `MC error'; ][]{gewe92}, and Raftery and Lewis's (\citeyear{raft91}) procedure to determine the effect of trending within chains.
@@
 are calculated.}
-The simulation set-up consists of several steps, summarized in the pseudo-code below.\footnote{The complete \proglang{R} script of the simulation study is available on Github.}
+The simulation set-up is summarized in the pseudo-code below.\footnote{The complete \proglang{R} script of the simulation study is available from \href{https://github.com/gerkovink/shinyMice/simulation}{github.com/gerkovink/ShinyMICE/simulation}.}
 
@@
 \subsection{Amputation}
-The complete dataset is `amputed' once in each simulation repetition.
+The complete data is `amputed' once in each simulation repetition.
 That is, the \pkg{mice} function \fct{ampute} is used to impose a missingness mechanism upon the data.
@@
 \subsection{Imputation}
-Missing data points are imputed with the \pkg{mice} function \fct{mice}.
+Missing datapoints are imputed with the \pkg{mice} function \fct{mice}.
 All simulations are performed with Bayesian linear regression imputation (\texttt{method = "norm"}), and five imputation chains (\texttt{m = 5}).
@@
 %, see \url{http://jsterman.scripts.mit.edu/docs/Rahmandad-Sterman%20Simulation%20Reporting%20Standards3.30%20update.pdf}.
-We investigate convergence by extracting not only the imputed values of each MI procedure (the $T^{th}$ iteration), but also information ($\theta$-values) from the intermediate iterations ($t = 1, 2, \dots, T$).
-The $\theta$s of interest are chain means and chain variances per variable.
-We compute convergence diagnostics separately for each of the four variables in the dataset, and report the maximum (absolute) value as the `worst linear predictor' of convergence \textbf{[Rephrase or footnote??]}.
+We investigate convergence by extracting not only the imputed values of each MI procedure (where $t = T$), but also information from the intermediate iterations ($t = 1, 2, \dots, T$).
+The quantities of interest ($\theta$-values) are chain means and chain variances.
+We compute convergence diagnostics separately for each of the four variables in the dataset.
+\textbf{We report the maximum (absolute) value as the `worst linear predictor' of convergence [Rephrase or footnote??]}.
 $\widehat{R}$ is computed by implementing Vehtari et al.'s \citeyear{veht19} recommendations.

### 2019-12-14 Simulation/Writeup/NoteConvergence.Rnw
safe plots with ses

@@
 %, see \url{http://jsterman.scripts.mit.edu/docs/Rahmandad-Sterman%20Simulation%20Reporting%20Standards3.30%20update.pdf}.
-We investigate convergence by extracting not only the imputed values of each MI procedure (where $t = T$), but also information from the intermediate iterations ($t = 1, 2, \dots, T$).
+We extract not only the imputed values of each MI procedure (where $t = T$), but also information from the intermediate iterations ($t = 1, 2, \dots, T$).
 The quantities of interest ($\theta$-values) are chain means and chain variances.
@@
 $AC$ is computed with \pkg{stats} function \fct{acf}.
-
+%%%%%%%%%%%%%%%%
+%From each imputation chain, we extract imputed values, chain means and chain variances ($\theta$-values).
+$\theta$-values are information from the intermediate iterations ($t = 1, 2, \dots, T$).
+The quantities of interest ($\theta$-values) are .
+We compute convergence diagnostics separately for each of the four variables in the dataset.
+\textbf{We report the maximum (absolute) value as the `worst linear predictor' of convergence [Rephrase or footnote??]}.
+$\widehat{R}$ is computed by implementing Vehtari et al.'s \citeyear{veht19} recommendations.
+$AC$ is computed with \pkg{stats} function \fct{acf}.
+%%%%%%%%%%%%%
 % 
@@
 
-$$\bar{Q} \pm 2.66 \times SE_{\bar{Q}},$$
+$$\bar{Q} \pm t_{(m-1)} \times SE_{\bar{Q}}, \text{ where } t_{(m-1)} = 2.66 \text{ if } \alpha = .05. $$
 
-where 2.66 is the quantile of a $t$-distribution with $m-1$ degrees of freedom, and $SE_{\bar{Q}}$ is the square root of the pooled variance estimate.
+$t_{(m-1)}$ is the quantile of a $t$-distribution with $m-1$ degrees of freedom, and $SE_{\bar{Q}}$ is the square root of the pooled variance estimate.
 From bias and CIW, we calculate empirical coverage rates.
@@
   \resizebox{\textwidth}{!}{ %notice the \resizebox{} command
+        \includegraphics{Figures/conv_diag_SEs.pdf}
+  }
+  \caption{Convergence diagnostics over 1000 MCMC simulations.}
+    \label{fig:conv2}
+\end{figure}
+
+\begin{figure}[h]
+  \resizebox{\textwidth}{!}{ %notice the \resizebox{} command
         \includegraphics{Figures/simulation_diag.pdf}
@@
     \label{fig:sim}
+\end{figure}
+
+\begin{figure}[h]
+  \resizebox{\textwidth}{!}{ %notice the \resizebox{} command
+        \includegraphics{Figures/sim_diag_SEs.pdf}
+  }
+  \caption{Simulation diagnostics over 1000 MCMC simulations.}
+    \label{fig:sim2}
 \end{figure}

### 2019-12-15 Presentation/Presentation.Rmd
Presentation added

---
title: "Developing ShinyMICE"
subtitle: "A Note on Convergence Diagnostics for Multiple Imputation Algorithms"
author: "Hanne Oberman"
date: "December 16, 2019"
output: 
    ioslides_presentation:
        logo: figures/shinymice.png
        transition: 0
        smaller: TRUE
bibliography: ShinyMICE.bib

---

```{r setup, include=FALSE}
library("mice")
library("magrittr")
library("DT")
library("plotly")
library("dplyr")
library("ggplot2")
library("ggpubr")

load("C:/Users/User/Desktop/shinyMice/Simulation/Results/final.Rdata")
```

<!-- ## Disclaimer -->

<!-- This presentation only serves the illustrative purpose of showing that I am able to create a presentation in R Markdown. -->

## This presentation | and the entire research archive is available on:

<center>
[github.com/gerkovink/ShinyMICE/](https://github.com/gerkovink/ShinyMICE/)
</center>
<br><br><br>
You can find all related materials, links and references there.

## Content of the Presentation

<div class="columns-2">
<img src="figures/shinymice.png" alt="HTML5 Icon" width = 110%>

<br><br><br>

- Introduction
- Methodology
- Simulation Results
- Discussion
</div>

# Introduction

## Multiple imputation
- Missing data is ubiquitous [@alli02]. 
- *Ad hoc* solutions may yield invalid inferences [@buur18].
- @rubin87 proposed the framework of MI.

## Multiple imputation
<br><br>

<center>
<img src="figures/steps.png" alt="HTML5 Icon" width = 100%>
</center>

## Algorithmic convergence
<br><br>

<center>
<img src="figures/traceplot.png" alt="HTML5 Icon" width = 100%>
<br>

<!-- Iteration number -->
</center>

# Methodology

## Mock code
```{r eval=FALSE}
simulation <- function() {
  mvtnorm(X,Z1,Z2) %>% 
  mutate(Y~X+Z1+Z2) %>% 
  for (max_iterations in 1:100) {
    ampute() %>% 
    impute() %T% 
    convergence_diagnostics %>% 
    lm(Y~X+Z1+Z2) %>% 
    pool %>% 
    simulation_diagnostics %>% 
    c(., convergence_diagnostics)
  }
}

replicate(simulation, n = 1000)
```

# Simulation Results

## Results

<center>
<img src="figures/results.png" alt="HTML5 Icon" width = 100%>
<br>

</center>

## Results | simulation diagnostics versus convergence diagnostics

```{r echo=FALSE}
dat %>% select("It.", "bias", "CIW", "cov", "R_mean_X", "AC_mean_X") %>% round(3) %>% datatable(options = list(pageLength = 5), rownames = FALSE)
```

## Inspect the autocorrelation dip
```{r echo=FALSE}
dat %>% round(3) %$%  plot_ly(x = It., y = AC_mean_X, 
        text = paste("R hat: ", R_mean_X),
        type = "scatter", mode = "lines") %>% 
  layout(#title = "Auto-correlation across iterations",
         xaxis = list(title = "Nr. of iterations"),
         yaxis = list(title = "Auto-correlation (at lag 1)"))
```

# Discussion

## Discussion
<center>
<img src="figures/shinymice.png" alt="HTML5 Icon" width = 71%>
</center>

<center>
<img src="figures/empty.png" alt="HTML5 Icon" width = 1%>
</center>

## Impression of ShinyMICE
<center>
<br>

<img src="figures/shiny.png" alt="HTML5 Icon" width = 90%>
</center>

## Inspect $\widehat{R}$
```{r echo=FALSE}
dat %>% round(3) %$%  plot_ly(x = It., y = R_mean_X, 
        text = paste("AC: ", AC_mean_X),
        type = "scatter", mode = "lines") %>% 
  layout(#title = "R across iterations",
         xaxis = list(title = "Nr. of iterations"),
         yaxis = list(title = expression(paste(widehat(R)))))

```

## Inspect autocorrelation
```{r echo=FALSE}
dat %>% round(3) %$%  plot_ly(x = It., y = AC_mean_X, 
        text = paste("R hat: ", R_mean_X),
        type = "scatter", mode = "lines") %>% 
  layout(#title = "Auto-correlation across iterations",
         xaxis = list(title = "Nr. of iterations"),
         yaxis = list(title = "Auto-correlation (at lag 1)"))
```

## Equations $\widehat{R}$
\[\begin{align}
B&=\frac{T}{M-1} \sum_{m=1}^{M}\left(\bar{\theta}^{(\cdot m)}-\bar{\theta}^{(\cdot \cdot)}\right)^{2}, \quad \text{where} \quad \bar{\theta}^{(\cdot m)}=\frac{1}{T} \sum_{t=1}^{T} \theta^{(t m)}, \\
&\quad \text{ and } \quad \bar{\theta}^{(\cdot \cdot)}=\frac{1}{M} \sum_{m=1}^{M} \bar{\theta}^{(\cdot m)} \\
W&=\frac{1}{M} \sum_{m=1}^{M} s_{j}^{2},  \quad \text{where} \quad s_{m}^{2}=\frac{1}{T-1} \sum_{t=1}^{T}\left(\theta^{(t m)}-\bar{\theta}^{(\cdot m)}\right)^{2}  \\

\widehat{R}&=\sqrt{\frac{\widehat{\operatorname{var}}^{+}(\theta | y)}{W}},
\quad \text{where} \quad \widehat{\operatorname{var}}^{+}(\theta | y)=\frac{N-1}{N} W+\frac{1}{N} B.
\end{align}\]

<center>
<img src="figures/empty.png" alt="HTML5 Icon" width = 1%>
</center>

## References

### 2019-12-15 Simulation/Writeup/NoteConvergence.Rnw
Presentation added

@@
 %% - \AND starts a new line, \And does not.
-\author{\\Research Report (Preparation for Research Master Thesis, Research Seminar) \\ Methodology and Statistics for the Behavioural, Biomedical and Social Sciences \AND Hanne Oberman\\Utrecht University}
+\author{Hanne Oberman\\Utrecht University \AND \\ Research Report (Preparation for Research Master Thesis, Research Seminar) \\ Methodology and Statistics for the Behavioural, Biomedical and Social Sciences}
    % \And Second Author\\Plus Affiliation}
@@
 We conclude that potential scale reduction factor $\widehat{R}$ and autocorrelation may be appropriate convergence diagnostics, but their performance differs from conventional applications to iterative algorithmic procedures.
-These results will be implemented into an interactive framework for the evaluation of multiple imputation methods that is under development as research Master thesis project.
-\textbf{TO DO still: check ref list, add legend to sim diags, check table and fig captions (abbrevations), references to tables, figs, appendix}
+These results will be implemented into an interactive framework for the evaluation of multiple imputation methods that is under development as Research Master's thesis.
+%\textbf{TO DO still: check worst lin pred ref, check ref list, add legend to sim diags, check table and fig captions (abbrevations), references to tables, figs, appendix}
 %%%%%%%%%%%%%
@@
 %% - Should be comma-separated and in sentence case.
-\Keywords{multiple imputation, convergence, \pkg{mice}, \proglang{R}}
+\Keywords{multiple imputation, convergence, mice, \proglang{R}}
 \Plainkeywords{multiple imputation, convergence, mice, R}
@@
 A popular answer to the ubiquitous problem of missing information is to use the framework of multiple imputation (MI), proposed by \cite{rubin87}.
-MI is an iterative algorithmic procedure in which each missing datapoint is `imputed' (i.e.
+MI is an iterative algorithmic procedure in which missing datapoints are `imputed' (i.e.
 filled in) several times.
@@
 If this variability is under-estimated, confidence intervals around estimates will be too narrow, which can yield spurious results.
-However, over-estimation of the variance between imputations results in unnecessarily wide confidence intervals, which can be costly because it lowers the statistical power.
+Over-estimation of the variance between imputations results in unnecessarily wide confidence intervals, which can be costly because it lowers the statistical power.
 Since both of these situations are undesirable, imputations and their variability should be evaluated.
@@
 multiple imputation algorithms. For reasons of brevity, we only focus on the MI algorithm implemented in \pkg{mice} \citep{mice}. 
-The convergence properties of the MI algorithm in \pkg{mice} are investigated through model-based simulation.%The results of this simulation study are guidelines for assessing convergence of MI algorithms.
+The convergence properties of this MI algorithm are investigated through model-based simulation\footnotemark.%The results of this simulation study are guidelines for assessing convergence of MI algorithms.
 These guidelines will be implemented in an interactive evaluation tool for \pkg{mice}, `ShinyMICE', which is currently under development.
-\footnote{All programming code used in this note is available from \href{https://github.com/gerkovink/shinyMice/simulation}{github.com/gerkovink/ShinyMICE/simulation}.} 
+\footnotetext{All programming code used in this note is available from \href{https://github.com/gerkovink/shinyMice/simulation}{github.com/gerkovink/ShinyMICE/simulation}.} 
 %I will replicate Lacerda et al.'s simulation study on $\widehat{R}$ \citep{lace07}, and develop novel guidelines for assessing convergence.
@@
 Basic familiarity with MI methodology is assumed.\footnote{For the theoretical foundation of MI, see \cite{rubin87}.
-For an accessible and comprehensive introduction to MI from an applied perspective, see \cite{buur18}.}
+For an accessible and comprehensive introduction to MI from an applied perspective, see e.g.
+\cite{buur18}.}
 
 Let $Y$ denote an $n \times p$ matrix containing the data values on $p$ variables for all $n$ units in a sample.
-The collection of observed data values in $Y$ is denoted by $Y_{obs}$, and will be referred to as `incomplete' or `observed' data interchangeably.
+The collection of observed data values in $Y$ is denoted by $Y_{obs}$, and will be referred to as `incomplete' or `observed' data.
 The missing part of $Y$ is denoted by $Y_{mis}$.
@@
 This note focuses on the algorithmic properties of the imputation step.
-The MI algorithm has an iterative nature.
+
+The MI algorithm in \pkg{mice} has an iterative nature.
 For each missing datapoint in $Y_{mis}$, $m$ `chains' of potential values are sampled.
@@
 Some default techniques in \pkg{mice} might not yield converged states at all \citep{murr18}.
-Algorithmic convergence should, therefore, be monitored carefully.
-In that respect is the current evaluation practice insufficient.
+Therefore, algorithmic convergence should be monitored carefully.
+The diagnostic evaluation of convergence is preferred over the current practice---visually inspecting imputation chains for signs of non-convergence---since the latter may be challenging to the untrained eye \citep[\S~6.5.2]{buur18}.
+%Diagnostic evaluation of algorithmic convergence does not share thisApplied researchers should be facilitated in evaluatingClear guidelines for the diagnostic evaluation of convergence would be preferred.
+
+%In that respect is the current evaluation practice insufficient.
 That is, the  recommendation is to visually inspect imputation chains for signs of non-convergence, which may be challenging to the untrained eye \citep[\S~6.5.2]{buur18}.
 Quantitative evaluation of the convergence of MI algorithms would be preferred.
+The application of convergence diagnostics to MI methods has not been systematically studied \citep{buur18}.
+The measures that are available for diagnosing convergence of iterative algorithmic procedures (Markov chain Monte Carlo methods; e.g., \pkg{mice} algorithms or Gibbs samplers) generally evaluate signs of non-convergence \citep{hoff09}.
+Non-convergence may be present as non-stationarity within chains (i.e., trending), or as slow mixing between chains (i.e., no intermingling).
+The mixing component of convergence may be evaluated with the potential scale reduction factor, $\widehat{R}$ \citep[a.k.a.
+`Gelman-Rubin statistic';][]{gelm92}\footnotemark.
+The stationarity component may be assessed through autocorrelations.
+Other convergence diagnostics are outside of the scope of this study.
+%(Several convergence diagnostics have been proposed for evaluating convergence of iterative algorithmic procedures
 
-The multiple imputation algorithm in \pkg{mice} is a special case of Markov chain Monte Carlo (MCMC) methods---a framework of iterative algorithmic procedures.
+\footnotetext{An adapted version of $\widehat{R}$ might also be used to diagnose non-stationarity.}
+
+% The multiple imputation algorithm in \pkg{mice} is a special case of Markov chain Monte Carlo (MCMC) methods---a framework of iterative algorithmic procedures.
 Within this framework, several convergence diagnostics have been proposed, but the application of these measures on MI algorithms has not been systematically studied \citep{buur18}.
-% For the framework of itrIt is not known whether convergence diagnostics that are available for the general framework that MICE is a special case of are appropriate for the evaluation of MI data.
+% % For the framework of itrIt is not known whether convergence diagnostics that are available for the general framework that MICE is a special case of are appropriate for the evaluation of MI data.
 This is the framework of Markoc chain Monte Carlo %Yet, reference manuals of promiment MI software do mention the importance of evaluating algorithmic convergence (e.g., \cite[p.XYZ]{mice}, \textbf{XYZ, XYZ}).
-%%%%%%%
-%This section provides a brief review of: \textbf{[INMIDDELS INCORRECT]}% briefly reviews?
-%1) how convergence is defined in the general context of iterative algorithmic procedures; 2) what diagnostic tools are available; and 3) which of these may apply to MI algorithms.
-
-Loosely speaking, MCMC algorithms aim to sample values from an unknown target distribution (e.g., the posterior distribution of a parameter in Gibbs samplers, or the distribution of missing values in MI algorithms).
+% %%%%%%%
+% %This section provides a brief review of: \textbf{[INMIDDELS INCORRECT]}% briefly reviews?
+% %1) how convergence is defined in the general context of iterative algorithmic procedures; 2) what diagnostic tools are available; and 3) which of these may apply to MI algorithms. 
+% 
+% Loosely speaking, MCMC algorithms aim to sample values from an unknown target distribution (e.g., the posterior distribution of a parameter in Gibbs samplers, or the distribution of missing values in MI algorithms).
 Convergence is reached once the algorithm samples exclusively from the target distribution \citep{cowl96}.
@@
 %\textbf{Figure XYZ shows a pathological case of non-convergence concerning both components.}
-%The aim of MCMC procedures is to sample from an unknown (theoretical) distribution.
+% %The aim of MCMC procedures is to sample from an unknown (theoretical) distribution.
 The stability of the solution denpends on two components of convergence: should have converged, two components should be satisfied.
-
-% Assessing convergence is difficult.
+% 
+% % Assessing convergence is difficult.
 We can never be certain of convergence, therefore convergence diagnostics evaluate signs of non-convergence \citep{hoff09}.
@@
 But sometimes you can know if your chain has not converged, so we at least check for this latter possibility" \cite[p.~101]{hoff09}
-% More importantly, in MCMC algorithms like MICE, "what is produced by the algorithm at convergence is not a single number or even a distribution, but rather a sample from a distribution" \citep[p.~883]{cowl96}.
+% % More importantly, in MCMC algorithms like MICE, "what is produced by the algorithm at convergence is not a single number or even a distribution, but rather a sample from a distribution" \citep[p.~883]{cowl96}.
 "automated convergence monitoring (as by a machine) is unsafe and should be avoided" \citep[p.~902]{cowl96}.
-%Convergence has two components: `mixing' and `stationarity' \citep[p.~284]{gelm13}.
+% %Convergence has two components: `mixing' and `stationarity' \citep[p.~284]{gelm13}.
 Mixing refers to the variation between chains of samples.
@@
 Ideally, both components will be satisfied: there are no trends within chains, and the chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm.
-
-\textbf{[COULD BE SHORTER!
+% 
+% \textbf{[COULD BE SHORTER!
 Maybe in footnote??]} Most convergence diagnostics target either of the two components.
@@
 Therefore, this study employs both $\widehat{R}$ and autocorrelation to investigate convergence, as recommended by \cite[p.~898]{cowl96}.
-%Therefore, the focus of this study is mainly on $\widehat{R}$. Stationarity will also be evaluated with autocorrelation as diagnostic, as recommended by \cite[p.~898]{cowl96}. 
-Other convergence diagnostics are outside the scope of this study.\footnotemark 
-\footnotetext{All of these methods evaluate the convergence of univariate scalar summaries (e.g., chain means or variances).
+% %Therefore, the focus of this study is mainly on $\widehat{R}$. Stationarity will also be evaluated with autocorrelation as diagnostic, as recommended by \cite[p.~898]{cowl96}. 
+% Other convergence diagnostics are outside the scope of this study.\footnotemark 
+% \footnotetext{All of these methods evaluate the convergence of univariate scalar summaries (e.g., chain means or variances).
 These convergence diagnostics cannot diagnose convergence of multivariable statistics (i.e., relations between scalar summaries).
@@
 Here, the focus is mainly on $\widehat{R}$ and autocorrelation.
-
-%Diagnose non-convergence in the mixing sense: $\widehat{R}$.
+% 
+% %Diagnose non-convergence in the mixing sense: $\widehat{R}$.
 Diagnose non-convergence in the stability sense: autocorrelation (AC).
@@
 Here, the focus is mainly on $\widehat{R}$ and autocorrelation.}
-
-% "In order to converge to a stationary distribution, a Markov chain needs to satisfy three important properties (Roberts 1996; Tierney 1996): irreducible, the chain must be able to reach all interesting parts of the state space; aperiodic, the chain should not oscillate between different states; recurrence, all interesting parts can be reached infinitely often, at least from almost all starting points.
+% 
+% % "In order to converge to a stationary distribution, a Markov chain needs to satisfy three important properties (Roberts 1996; Tierney 1996): irreducible, the chain must be able to reach all interesting parts of the state space; aperiodic, the chain should not oscillate between different states; recurrence, all interesting parts can be reached infinitely often, at least from almost all starting points.
 Do these properties hold for the MICE algorithm?
@@
 The development of practical tools that put these conditions to work for multiple imputation is still an ongoing research problem." \cite[par.~4.5]{buur18}
+%
 
-The convergence diagnostics under consideration are $\widehat{R}$ and autocorrelation.
+% The convergence diagnostics under consideration are $\widehat{R}$ and autocorrelation.
 Perfect convergence is reached when the variance between imputation chains is equivalent to the variance within chains ($\widehat{R} = 1$), and there is no dependency between subsequent iterations of imputation chains ($AC = 0$).
@@
 Over-dispersion implies that the initial values of the chains are `far away' from the target distribution and each other.
-The iteration at which chains are independent of their initial values is the point where mixing occurs.
+When all chains sample independent of their initial values, the mixing component of convergence is satisfied, and $\widehat{R}$-values will be close to one.
+High $\widehat{R}$-values thus indicate non-convergence.
+%The iteration at which chains are independent of their initial values is the point where mixing occurs.
 This is diagnosed with $\widehat{R}$-values close to one.
 %If there is no added value
-High $\widehat{R}$-values thus indicate slow mixing of chains, which is a sign of non-convergence.
+%High $\widehat{R}$-values thus indicate slow mixing of chains, which is a sign of non-convergence.
 %, e.g., because chains may depend on their initial value.
@@
 %If there is no added value compared to infinitely many iterations, the chains have converged. 
-The conventionally acceptable threshold for this convergence diagnostic was $\widehat{R} < 1.2$ \citep{gelm92}. Recently, \cite{veht19} proposed a more stringent threshold: $\widehat{R} < 1.01$.
+The conventionally acceptable threshold for convergence was $\widehat{R} < 1.2$ \citep{gelm92}. More recently, \cite{veht19} proposed a more stringent threshold of $\widehat{R} < 1.01$.
 
@@
 We can interpret %the sign and magnitude of ... Any non-zero $AC$-value ($-1 \leq AC \leq 1$) . $AC = 0$ indicates independence between $\theta$-values.
-$AC$-values as a measure of dependence within chains.
-Positive $AC$-values indicate recurrence (i.e., similarity between $\theta$-values of subsequent iterations), which may lead to trending.
+$AC$-values as a measure of stationarity.
+If $AC$-values are close to zero, there is no dependence between subsequent samples within imputation chains.
+Positive $AC$-values, however, indicate recurrence.
+If $\theta$-values of subsequent iterations are similar, trending may occur.
+Negative $AC$-values show no threat to the stationarity component of convergence.
+% (i.e., similarity between $\theta$-values of subsequent iterations), which may lead to trending.
 Negative $AC$-values occur when the $\theta$-values of subsequent iterations diverge from one-another.
@@
 %that we aim to evaluate with $AC$.
-On the contrary even---it may increase the variance of $\theta$.
-\textbf{speed up convergence???}  %have also indicate dependence, but do not threaten the stationarity component of convergence \textbf{[REMOVE INTERPRETATION HERE?]}.
+On the contrary even---negative $AC$-values indicate that $\theta$-values of subsequent iterations diverge from one-another, which may increase the variance of $\theta$ and speed up convergence.
+%have also indicate dependence, but do not threaten the stationarity component of convergence \textbf{[REMOVE INTERPRETATION HERE?]}.
 Instead of recurrence, there would be a `seesaw' effect within chains: subsequent $\theta$-values would deviate in opposite directions from the chain average $\theta^{(\cdot m)})$, increasing the variance of $\theta$.
-As convergence diagnostic, the interest is in positive $AC$-values.%\footnotemark 
-\footnote{The magnitude of $AC$-values can also be evaluated statistically, but that is outside of this note's scope.}
+As convergence diagnostic, the interest is therefore in positive $AC$-values.%\footnotemark 
+\footnote{Moreover, the magnitude of $AC$-values may be evaluated statistically, but that is outside of this note's scope.}
+
+In short, convergence is reached when there is no dependency between subsequent iterations of imputation chains ($AC = 0$), and chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm ($\widehat{R} = 1$).
 
@@
 This study evaluates whether $\widehat{R}$ and $AC$ could diagnose convergence of multiple imputation algorithms.
-There is, however, no baseline measure available to evaluate their performance against.
-\footnote{The current practice of visually inspecting imputation chains for signs of non-mixing or non-stationarity can only diagnose severely pathological cases of non-convergence.} 
-We will assess the appropriateness of the two convergence diagnostics in comparison to evaluation criteria recommended by \cite[\S~2.5.2]{buur18}.
-The performance diagnostics comprise of average bias, average confidence interval width, and empirical coverage rate across simulations.% %"We expect that in accordance with Neyman (1934), at least 95 percent of the confidence intervals should contain the true population value, although some room for simulation error should be taken into account to counteract the finite nature of our simulations"\cite{schou18}.
-\footnote{We could also look at distributional characteristics, and plausibility of imputed values, see \cite{vinknd} (n.d.).
+%There is, however, no baseline measure available to evaluate their performance against.
+%\footnote{The current practice of visually inspecting imputation chains for signs of non-mixing or non-stationarity can only diagnose severely pathological cases of non-convergence.} 
+We assess the performance of the two convergence diagnostics against the recommended evaluation criteria for MI methods \citep[i.e., average bias, average confidence interval width, and empirical coverage rate across simulations;][\S~2.5.2]{buur18}.
+%The performance diagnostics comprise of .% %"We expect that in accordance with Neyman (1934), at least 95 percent of the confidence intervals should contain the true population value, although some room for simulation error should be taken into account to counteract the finite nature of our simulations"\cite{schou18}.
+%\footnote{We could also look at distributional characteristics, and plausibility of imputed values, see \cite{vinknd} (n.d.).
 For now, this is outside of the scope of this study.} % Or use: Then evaluate the regular diagnostics of MI simulations, see Vink, n.d., and  convergence.
@@
 
-\textbf{[HIER NOG ZOOI]} We hypothesize that $\widehat{R}$ may not be an appropriate diagnostic, because it assumes over-dispersed initial values.
+%\textbf{[HIER NOG ZOOI]} We hypothesize that $\widehat{R}$ may not be an appropriate diagnostic, because it assumes over-dispersed initial values.
 In \pkg{mice}, initial values of the algorithm are chosen randomly from the observed data.
@@
 % Starting values of MICE chains: "For each j, fill in starting imputations $Y^0_j$ by random draws from $Y^{obs}_j$." \cite[\S~4.5.2]{buur18} before for loops over variables and imputations, see \url{https://stefvanbuuren.name/fimd/sec-FCS.html#def:mice}
-Without over-dispersed initial states, $\widehat{R}$ may falsely diagnose convergence \citep{broo98}.
+%Without over-dispersed initial states, $\widehat{R}$ may falsely diagnose convergence \citep{broo98}.
 %would not be sensitive enough to flag non-convergence of MI algorithms: "if over-dispersion does not hold, $\sigma_+^2$ can be too low, which can lead to falsely diagnosing convergence." \cite[p~437]{broo98}.
-This suggests that $\widehat{R}$ would not be sensitive enough to flag non-convergence of MI algorithms.
+%This suggests that $\widehat{R}$ would not be sensitive enough to flag non-convergence of MI algorithms.
 However, an empirical finding by \cite{lace07} shows that $\widehat{R}$ can indicate non-convergence at unconventionally long chain lengths.
@@
 
-\textbf{We hypothesize that $\widehat{R}$ will over-estimate non-convergence. No hypothesis was formulated about AC, results are exploratory.}
-%
+Based on an empirical finding \citep{lace07}, we hypothesize that $\widehat{R}$ will over-estimate non-convergence of MI algorithms.
+The threshold of $\widehat{R} < 1.01$ will then be too stringent for diagnosing convergence.
+This over-estimation may, however, be diminished because $\widehat{R}$ can falsely diagnose convergence if initial values of the algorithm are not appropriately over-dispersed \cite[p~437]{broo98}.
+In \pkg{mice}, initial values are chosen randomly from the observed data.
+Therefore, we cannot be certain that the initial values are over-dispersed.
+We expect this to have little effect on the hypothesized performance of $\widehat{R}$.
+No hypothesis was formulated about the performance of $AC$ as convergence diagnostic.
 %
@@
 We use model-based simulation in \proglang{R} to perform multiple imputation under 100 simulation conditions.
-In each condition, the MI algorithm comprises a different imputation chain length ($T = 1, 2, \dots, 100$).
+In each condition, the MI algorithm comprises of a different imputation chain length ($T = 1, 2, \dots, 100$).
 The number of simulation runs per condition is 1000.
 For each simulation condition (i.e., number of iterations) and each repetition (i.e., simulation run) we compute convergence diagnostics ($\widehat{R}$ and autocorrelation) and simulation diagnostics (bias, confidence interval width and coverage).
-These diagnostics are aggregated across repetitions to obtain averages per simulation condition.
+These diagnostics are aggregated across repetitions to obtain their average value per simulation condition.
 %We compute $\widehat{R}$ and autocorrelation (`convergence diagnostics') and performance measures (`simulation diagnostics') for each condition.
@@
 The quantity of scientific interest is regression coefficient $\beta_1$.
-The data generating model is a multivariate normal distribution with means structure $\mu$ and variance-covariance matrix $\Sigma$.
+The data generating model is a multivariate normal distribution with the following means structure and variance-covariance matrix:
 
@@
 \end{pmatrix}
-\end{bmatrix}\\[2\jot]
+\end{bmatrix}\\[2\jot].
 \end{align*}
 
-Outcome variable $Y$ is subsequently calculated as
+Data is simulated with the function \fct{mvtnorm::rmvnorm}, and outcome variable $Y$ is subsequently calculated as
 
@@
 \subsection{Amputation}
-The complete data is `amputed' once in each simulation repetition.
-That is, the \pkg{mice} function \fct{ampute} is used to impose a missingness mechanism upon the data.
-The missingness is univariate, and the probability to be missing is the same for all variables, namely 20\% (\texttt{prop = 0.8, mech = "MCAR"}).
+The complete data is `amputed' once for each simulation repetition.
+That is, the function \fct{mice::ampute} is used to impose a missingness mechanism upon the data.
+The missingness is univariate, and the probability to be missing is the same for all four variables, namely 20\% (\texttt{prop = 0.8, mech = "MCAR"}).
 This leaves 20\% of the rows completely observed.
@@
 \subsection{Imputation}
-Missing datapoints are imputed with the \pkg{mice} function \fct{mice}.
-All simulations are performed with Bayesian linear regression imputation (\texttt{method = "norm"}), and five imputation chains (\texttt{m = 5}).
+Missing datapoints are imputed with the function \fct{mice::mice}.
+All MI procedures are performed with Bayesian linear regression imputation (\texttt{method = "norm"}), and five imputation chains (\texttt{m = 5}).
 The number of iterations varies between simulation conditions (\texttt{maxit = 1, 2, \dots, 100}).
@@
 %, see \url{http://jsterman.scripts.mit.edu/docs/Rahmandad-Sterman%20Simulation%20Reporting%20Standards3.30%20update.pdf}.
-We extract not only the imputed values of each MI procedure (where $t = T$), but also information from the intermediate iterations ($t = 1, 2, \dots, T$).
+From the imputation step, we extract chain means and chain variances across iterations ($\theta$ at $t = 1,2,\dots,T$).
+%not only the imputed values of each MI procedure (where $t = T$), but also information from the intermediate iterations ($t = 1, 2, \dots, T$).
 The quantities of interest ($\theta$-values) are chain means and chain variances.
-We compute convergence diagnostics separately for each of the four variables in the dataset.
-\textbf{We report the maximum (absolute) value as the `worst linear predictor' of convergence [Rephrase or footnote??]}.
-$\widehat{R}$ is computed by implementing Vehtari et al.'s \citeyear{veht19} recommendations.
-$AC$ is computed with \pkg{stats} function \fct{acf}.
+We compute convergence diagnostics separately for both $\theta$s, and for each of the four variables in the dataset.
+We report the maximum (absolute) value as the `worst linear predictor' of convergence.
+$\widehat{R}$ is computed by implementing Vehtari et al.'s (\citeyear{veht19}) recommendations.
+$AC$ is computed with function \fct{stats::acf}.
 %%%%%%%%%%%%%%%%
@@
 
-To estimate the quantity of scientific interest, $Q$, we perform multiple linear regression on each completed dataset with the \pkg{stats} function \fct{lm}.
+To estimate the quantity of scientific interest, $Q$, we perform multiple linear regression on each completed dataset with the function \fct{stats::lm}.
 We obtain an estimated regression coefficient per imputation, which are pooled into a single estimate, $\bar{Q}$.
-We use the \pkg{mice} function \fct{pool} to get variance estimates according to Rubin's \citeyear{rubin87} rules, and utilize these to implement finite population pooling as per \citep{vink14}.
+We use the function \fct{mice::pool} to get variance estimates according to Rubin's (\citeyear{rubin87}) rules, and subsequently implement finite population pooling conform \cite{vink14}.
 
@@
 We compute bias as the difference between $\bar{Q}$ and $Q$.
-Confidence interval width (CIW) is defined as the difference between the lower and upper bound of the 95\% confidence interval around $\bar{Q}$ (CI95\%).
+Confidence interval width (CIW) is defined as the difference between the lower and upper bound of the 95\% confidence interval (CI95\%) around $\bar{Q}$.
 We compute the CI95\% bounds as
 
-$$\bar{Q} \pm t_{(m-1)} \times SE_{\bar{Q}}, \text{ where } t_{(m-1)} = 2.66 \text{ if } \alpha = .05. $$
+$$\bar{Q} \pm t_{(m-1)} \times SE_{\bar{Q}},$$
 
-$t_{(m-1)}$ is the quantile of a $t$-distribution with $m-1$ degrees of freedom, and $SE_{\bar{Q}}$ is the square root of the pooled variance estimate.
+where $t_{(m-1)}$ is the quantile of a $t$-distribution with $m-1$ degrees of freedom, and $SE_{\bar{Q}}$ is the square root of the pooled variance estimate.
 From bias and CIW, we calculate empirical coverage rates.
@@
 
-\begin{figure}[h]
-  \resizebox{\textwidth}{!}{ %notice the \resizebox{} command
-        \includegraphics{Figures/conv_diag_SEs.pdf}
-  }
-  \caption{Convergence diagnostics over 1000 MCMC simulations.}
-    \label{fig:conv2}
-\end{figure}
+% \begin{figure}[h]
+%   \resizebox{\textwidth}{!}{ %notice the \resizebox{} command
+%         \includegraphics{Figures/conv_diag_SEs.pdf}
+%   }
+%   \caption{Convergence diagnostics over 1000 MCMC simulations.}
+%     \label{fig:conv2}
+% \end{figure}
 
@@
   }
-  \caption{Simulation diagnostics over 1000 MCMC simulations.}
+  \caption{Simulation diagnostics over 1000 MCMC simulations. The dashed line depicts the average trend across simulation conditions (Loess line).}
     \label{fig:sim}
@@
 
-\begin{figure}[h]
-  \resizebox{\textwidth}{!}{ %notice the \resizebox{} command
-        \includegraphics{Figures/sim_diag_SEs.pdf}
-  }
-  \caption{Simulation diagnostics over 1000 MCMC simulations.}
-    \label{fig:sim2}
-\end{figure}
+% \begin{figure}[h]
+%   \resizebox{\textwidth}{!}{ %notice the \resizebox{} command
+%         \includegraphics{Figures/sim_diag_SEs.pdf}
+%   }
+%   \caption{Simulation diagnostics over 1000 MCMC simulations.}
+%     \label{fig:sim2}
+% \end{figure}
 
-% latex table generated in R 3.6.1 by xtable 1.8-4 package
-% Wed Dec 04 18:16:11 2019
+%latex table generated in R 3.6.1 by xtable 1.8-4 package
+%Wed Dec 04 18:16:11 2019
 \begin{table}[ht]
 \centering
-\caption{Simulation and convergence diagnostics over 1000 MCMC simulations.} 
+\caption{Simulation and convergence diagnostics over 1000 MCMC simulations.
+\footnotesize{No convergence diagnostics are reported for the condition where T=1, since these relative measures cannot be computed for a single value.}}
 \label{tab:results}
@@
 \subsection{Convergence diagnostics}
-It is apparent that there is a relation between convergence diagnostics $\widehat{R}$ and $AC$, and the number of iterations per simulation condition($T$).
-Generally speaking, we see a trend towards convergence as $T$ increases with each simulation condition ($\widehat{R}$-values approach one, and $AC$-values approach zero).
-The convergence diagnostics show more or less equivalent trends for chain means versus chain variances.
+It is apparent that there is a relation between the number of iterations per simulation condition ($T$) and the convergence diagnostics.
+Generally speaking, conditions with longer imputation chains (higher $T$) coincide with less signs of non-convergence ($\widehat{R}$-values approach one, and $AC$-values approach zero).
+We see more or less equivalent trends for chain means versus chain variances.
 We, therefore, only discuss the $\widehat{R}$ and $AC$-values of chain means.
@@
 In the intervening conditions ($3 \leq T \leq 5$), however, we observe a steep increase in $\widehat{R}$-values.
-This increase implies that convergence cannot be diagnosed at $T=3$, as would be the case based on the conventional threshold $\widehat{R} < 1.2$.
-\textbf{[This may be spurious??]} According to the widely used threshold $\widehat{R} < 1.1$, convergence can be diagnosed for conditions where $T>9$.
-If we use the recently recommended threshold $\widehat{R} < 1.01$, we would conclude that we cannot diagnose convergence in any simulation condition.
-\textbf{[Add something about $\widehat{R} <1$?]}
+This increase implies that non-convergence is under-estimated, and convergence should not be diagnosed at this point.
+Based on the conventional threshold $\widehat{R} < 1.2$, we would falsely diagnose convergence at $T=3$.
+According to the widely used threshold $\widehat{R} < 1.1$, convergence would be diagnosed for conditions where $T>9$.
+If we use the recently recommended threshold $\widehat{R} < 1.01$, we would conclude that convergence is reached in none of the simulation conditions.
+%\textbf{[Add something about $\widehat{R} <1$?]}
 
@@
 The lowest $AC$-value is obtained for $T=3$.
-At $T=5$, the $AC$-value reaches the level observed at $T=2$.
-Between $T=10$ and $T=30$ the increase plateaus off, and $AC$-values are stable around $T=70$.
-$AC$ is then indifferentiable from zero, indicating stationarity.
+%At $T=5$, the $AC$-value reaches the level observed at $T=2$.
+The gradual increase plateaus between $T=10$ and $T=30$.
+$AC$-values in conditions where $T>70$ are indifferentiable from zero, indicating stationarity.
 %Simulation conditions where $T>5$ have $AC$-values greater than $T=2$.
@@
 According to this diagnostic, none of the simulation conditions show signs of non-convergence, since we only observe negative or zero $AC$-values.
-
+The negative $AC$-values, however, indicate increasing divergence between subsequent samples in conditions where $T<4$.
+It may not be wise to terminating the algorithm at that point, but after `recovering' from this dip in $AC$-values.
+This would imply $T=6$ as the minimal number of iterations.
 % All observed autocorrelations  are negative or close to zero.
@@
 
-Taken together, we see that convergence may be diagnosed in simulation conditions where $T>10$.
+Taken together, we see that $T>3$ is the minimal requirement to diagnose convergence ($\widehat{R} < 1.2; AC \leq 0$).
+This threshold is, however, not sufficient, since we overlook the increase in $\widehat{R}$-values up-to conditions where $T>5$, and the convergence diagnostics only reach stability at $T>20$.
+%This implies that the current
 
@@
 We use average bias, average confidence interval width, and coverage rate as performance measures to evaluate $\widehat{R}$ and $AC$.
-We make the general observation that the simulation diagnostics behave as theorized for most simulation conditions (bias around zero, coverage rate around 95\%).
+We make the general observation that the simulation diagnostics behave as theorized for most simulation conditions (bias around zero, stable confidence interval widths, nominal coverage rate at 95\%).
 
@@
 The bias for $T=2$ is below average, but within the range of fluctuations.
-Conditions where $T>3$ \textbf{are OK, Loess line is flat for $T>20$}.
+Conditions where $T>3$ can be diagnosed as unbiased, but the average bias across iterations (as shown by a flat Loess line) only reaches stability at $T=20$.
 %the average bias between the regression coefficient of $X$ on $Y$ in the complete dataset, and $\bar{Q}$, \textbf{the pooled estimated regression coefficient and the ...
@@
 
-CIW is only clearly divergent for the simulation condition $T=1$.
-However, we want to be on the safe side, and not under-estimate the variance.
-Otherwise, we might end up with spurious inferences.
-Conditions where $T>3$ have similar CIWs, which seems to be stable across iterations.
-\textbf{Loess line is never completely flat.}%One might argue that there is some downward trending up-to thirthy to forty iterations.
+CIW is only clearly divergent in the simulation condition $T=1$, see Figure \ref{fig:sim}B.
+However, under-estimating the variance of $\bar{Q}$, which is the case for $T=3$, may yield spurious inferences.
+Only conditions where $T>3$ are therefore considered sufficient.
+These conditions have similar CIWs, but across iterations stability is not established (i.e., the Loess line is never completely flat within the 100 simulation conditions).%One might argue that there is some downward trending up-to thirthy to forty iterations.
 %After about twenty-five iterations the average CIW across all simulation conditions and repetitions is reached \Sexpr{round(mean(dat[, "CI width"]), 3)}.
 
-\textbf{Empirical coverage rate seems more or less stable for conditions where $T>1$.
-We see some over-coverage at $T=2$.
-We might over-estimate the variance, but better than under-estimating it.
+The empirical coverage rate across repetitions seems more or less stable for conditions where $T>1$, see Figure \ref{fig:sim}C.
+We see some over-coverage at $T=2$, but that is better than under-estimating the variance of $\bar{Q}$.
 On average, the coverage rate is somewhat higher than the expected nominal coverage of 95\% \citep{neym34}, namely \Sexpr{100*round(mean(dat[, "Cov.
 rate"]), 3)}\%.
-Loess line is never flat.}
+Similar to the CIWs, there is some trending across iterations (i.e., the Loess line is never flat).
 
-\textbf{From the simulation diagnostics, we see that simulation conditions where $T>3$ are generally OK.
+From the simulation diagnostics, we observe that unbiased estimates with nominal coverage rates were obtained in conditions where $T>3$.
+This suggests that as little as four iterations may be sufficient for the MI algorithm under the current circumstances.
+%  that simulation conditions where $T>3$ are generally OK.
 The performance measures indicate that for the current data, MI procedure, and quantity of interest, the algorithm should have maxit argument of four or more.
@@
 
-There is a discrepancy between what the convergence diagnostics and the performance measures indicate.
-While $T=4$ is OK with respect to simulation quantities, $T=3$ is the worst possible value according to the convergence diagnostics, and $T=4$ provides marginally better convergence diagnostics.
+In short, there is a discrepancy between what the convergence diagnostics and what the performance measures indicate.
+While $T=4$ seems sufficient with respect to simulation quantities, the condition where $T=4$ resulted in the `one-but-worst' values of both convergence diagnostics.
+Complete algorithmic convergence as indicated by $\widehat{R}$ and autocorrelation is not reached in conditions where $T<20$.
+%is the worst possible value according to the convergence diagnostics, and $T=4$ provides marginally better convergence diagnostics.
 
@@
 
-This note shows that the convergence of multiple imputation algorithms may be evaluated with convergence diagnostics $\widehat{R}$ and $AC$.
+This note shows that convergence diagnostics $\widehat{R}$ and $AC$ may diagnose convergence of multiple imputation algorithms, but their performance differs from conventional applications to iterative algorithmic procedures.
+%but the convergence of multiple imputation algorithms may be evaluated with convergence diagnostics .
 These measures, however, indicate non-convergence at imputation chain lengths that produce acceptable simulation diagnostics.
@@
 More `complicated' $Q$s (e.g., higher order effects or variance components) might show bias, under- or over-coverage at higher $T$.
+$\widehat{R}$ and autocorrelation indicate that algorithmic convergence may only be reached after twenty or even forty iterations, while unbiased, confidence valid estimates estimates may be obtained with as little as four iterations.
+These results are in agreement with the simulation hypothesis: $\widehat{R}$ over-estimates the severity of non-convergence when applied to MI procedures.
 
-At least $\widehat{R}$ does not under-estimate non-convergence.
-It's better to be too strict than too lenient.
-According to this simulation study, the recently proposed threshold $\widehat{R}$ is too stringent for MI algorithms, and the conventionally acceptable $\widehat{R} <1.2$ is too lenient.
+%performance diagnostics, it appears that as little as four iterations might be sufficient to obtain ..
+
+%At least $\widehat{R}$ does not under-estimate non-convergence. It's better to be too strict than too lenient. 
+According to this simulation study, the recently proposed threshold of $\widehat{R}<1.01$ may be too stringent for MI algorithms.
+Under the relatively easy missing data problem of the current study, the threshold was not reached.
+The other extreme of the $\widehat{R}$-thresholds, the conventionally acceptable $\widehat{R} <1.2$, may be too lenient for MI procedures.
+Applying this threshold to the current data, lead to falsely diagnosing convergence at $T = 3$.
+It appears that the widely used threshold of $\widehat{R} < 1.1$ suits MI algorithms the best.
+We might, however, also formulate a new threshold, specifically for the evaluation of MI algorithms.
+The current study suggests that $\widehat{R} < 1.05$ may be implemented, since that is the level at which the $\widehat{R}$ stabilize (around $T = 20$).
+
+The negative $AC$-values obtained in this study show no threat of non-stationarity.
+However, initial dip in $AC$-values may have implications for the default number of iterations in \pkg{mice} (\texttt{maxit = 5}).
+Terminating the algorithm at $T=5$ may not be the most appropriate, since this lead to the worst convergence, as indicated by $\widehat{R}$ and $AC$.
+Under the current specifications, $T>20$ would be more appropriate.
 
@@
 
-\textbf{Summary.}
-This note illustrates that conventional convergence diagnostics behave differently on MI data than other MCMC methods like Gibbs samplers in Bayesian analyses.
+% \textbf{Summary.}
+% This note illustrates that conventional convergence diagnostics behave differently on MI data than other MCMC methods like Gibbs samplers in Bayesian analyses.
 The most recent recommended threshold for $\widehat{R}$ ($\widehat{R} < 1.01$) may be too stringent for MI data.
@@
 With increased performance and storage capacities, doubling the default is fine I guess.}
-
-From the simulation diagnostics, it appears that as little as four iterations might be sufficient to obtain unbiased, confidence valid estimates.
-Convergence diagnostics $\widehat{R}$ and autocorrelation, however, indicate that convergence may only be reached after twenty or even forty iterations.
-
-\textbf{Does this mean that conv is not nec to obtain correct inferences?}
-
-\textbf{R hat below 1.}
-$\widehat{R}$ could theoretically not be smaller than one, yet it occurred several times in this study (see online appendix XYZ).
+% 
+% 
+% \textbf{Does this mean that conv is not nec to obtain correct inferences?}
+% 
+% \textbf{R hat below 1.}
+% $\widehat{R}$ could theoretically not be smaller than one, yet it occurred several times in this study (see online appendix XYZ).
 This can happen when the number of simulations is smaller than in `regular' MCMC processes [explain that fewer iterations is an advantage of MI, not a disadvantage compared to MCMC].
@@
 (Before convergence, we expect $\sigma^2$ to decrease with n, only increasing if the sequences explore a new area of parameter space, which would imply that the original sequences were not overdispersed for the particular scalar summary being monitored.)'' \cite[p~438]{broo98}.
-
-% \url{https://discourse.mc-stan.org/t/rhat-1-as-low-as-9-94e-01-why/9252/3}
-
-\textbf{Negative ACs.} autocorrelation is dangerous when positive.
+% 
+% % \url{https://discourse.mc-stan.org/t/rhat-1-as-low-as-9-94e-01-why/9252/3}
+% 
+% \textbf{Negative ACs.} autocorrelation is dangerous when positive.
 These autocorrelations are negative.
@@
 But why the dip??
-
-Assessing the stationarity component of convergence with $AC$ might be redundant, since high $AC$-values are implausible in MI procedures.
+% 
+% Assessing the stationarity component of convergence with $AC$ might be redundant, since high $AC$-values are implausible in MI procedures.
 That is, the randomness induced by the MI algorithm effectively mitigates the risk of dependency within chains.
 $AC$ would thus not be informative of the convergence of MI algorithms.
+% 
+% 
+% The observed dip in AC implies that default maxit value of five iterations is the worst possible number of iterations.
 
-The observed dip in AC implies that default maxit value of five iterations is the worst possible number of iterations.
-
-\textbf{Future research.} This study considered only an MCAR missingness mechanism.
-Necessary but not sufficient to demonstrate the performance of convergence diagnostics.
-This is just a proof of concept.
-Further research is needed to investigate the performance under violation of convergence, e.g.
-dependency between predictors (very high correlations).
+Since this study only considers only an MCAR missingness mechanism, results may not be extrapolated to other missing data problems.
+Proper performance of the convergence diagnostics under MCAR is necessary but not sufficient to demonstrate appropriateness of $\widehat{R}$ and $AC$ as convergence diagnostics.
+%This is just a proof of concept.
+Further research is needed to investigate their performance under clear violation of convergence, e.g.
+dependency between predictors (predictors with very high correlations).
 %We would like to have convergence measures for multivariable statistics (scalars?) of interest.
@@
 \textbf{[this paragraph is too negative, explain scope in intro instead.]}
-
-Also, conv diag for substantive model? wald test for AC = 0? 
+Until then, we have only shown that the convergence diagnostics can diagnose non-convergence of MI algorithms that trend towards a converged state.
+%Also, conv diag for substantive model? wald test for AC = 0? 
 %% -- Optional special unnumbered sections -------------------------------------

### 2019-12-15 Simulation/Writeup/NoteConvergence.tex
Presentation added

@@
 
-\begin{figure}[h]
-  \resizebox{\textwidth}{!}{ %notice the \resizebox{} command
-        \includegraphics{Figures/sim_diag_SEs.pdf}
-  }
-  \caption{Simulation diagnostics over 1000 MCMC simulations.}
-    \label{fig:sim2}
-\end{figure}
+% \begin{figure}[h]
+%   \resizebox{\textwidth}{!}{ %notice the \resizebox{} command
+%         \includegraphics{Figures/sim_diag_SEs.pdf}
+%   }
+%   \caption{Simulation diagnostics over 1000 MCMC simulations.}
+%     \label{fig:sim2}
+% \end{figure}
 
-% latex table generated in R 3.6.1 by xtable 1.8-4 package
-% Wed Dec 04 18:16:11 2019
+latex table generated in R 3.6.1 by xtable 1.8-4 package
+Wed Dec 04 18:16:11 2019
 \begin{table}[ht]
 \centering
-\caption{Simulation and convergence diagnostics over 1000 MCMC simulations.} 
+\caption{Simulation and convergence diagnostics over 1000 MCMC simulations.
+\footnotesize{No convergence diagnostics are reported for the condition where T=1, since these relative measures cannot be computed for a single value.}}
 \label{tab:results}
@@
 
-CIW is only clearly divergent for the simulation condition $T=1$.
-However, we want to be on the safe side, and not under-estimate the variance.
-Otherwise, we might end up with spurious inferences.
-Conditions where $T>3$ have similar CIWs, which seems to be stable across iterations.
-\textbf{Loess line is never completely flat.}%One might argue that there is some downward trending up-to thirthy to forty iterations.
+CIW is only clearly divergent in the simulation condition $T=1$, see Figure \ref{fig:sim}B.
+However, under-estimating the variance of $\bar{Q}$, which is the case for $T=3$, may yield spurious inferences.
+Only conditions where $T>3$ are therefore considered sufficient.
+These conditions have similar CIWs, but across iterations stability is not established (i.e., the Loess line is never completely flat within the 100 simulation conditions).%One might argue that there is some downward trending up-to thirthy to forty iterations.
 %After about twenty-five iterations the average CIW across all simulation conditions and repetitions is reached 0.928.
 
-\textbf{Empirical coverage rate seems more or less stable for conditions where $T>1$.
-We see some over-coverage at $T=2$.
-We might over-estimate the variance, but better than under-estimating it.
+The empirical coverage rate across repetitions seems more or less stable for conditions where $T>1$, see Figure \ref{fig:sim}C.
+We see some over-coverage at $T=2$, but that is better than under-estimating the variance of $\bar{Q}$.
 On average, the coverage rate is somewhat higher than the expected nominal coverage of 95\% \citep{neym34}, namely 95.3\%.
-Loess line is never flat.}
+Similar to the CIWs, there is some trending across iterations (i.e., the Loess line is never flat).
 
-\textbf{From the simulation diagnostics, we see that simulation conditions where $T>3$ are generally OK.
+From the simulation diagnostics, we observe that unbiased estimates with nominal coverage rates were obtained in conditions where $T>3$.
+This suggests that as little as four iterations may be sufficient for the MI algorithm under the current circumstances.
+%  that simulation conditions where $T>3$ are generally OK.
 The performance measures indicate that for the current data, MI procedure, and quantity of interest, the algorithm should have maxit argument of four or more.

### 2019-12-16 Presentation/Presentation.Rmd
update presentation after mice meeting

@@
 ## Algorithmic convergence
-<br><br>
-
 <center>
-<img src="figures/traceplot.png" alt="HTML5 Icon" width = 100%>
+<img src="figures/non-conv.png" alt="HTML5 Icon" width = 77%>
 <br>
@@
 
-## Mock code
+## Mock code | 
+Steps in the simulation study
 ```{r eval=FALSE}
@@
 
-## Results
-
+## Simulation diagnostics
 <center>
-<img src="figures/results.png" alt="HTML5 Icon" width = 100%>
+<img src="figures/results_sim.png" alt="HTML5 Icon" width = 100%>
 <br>
@@
 
-## Results | simulation diagnostics versus convergence diagnostics
+## Convergence diagnostics
+<center>
+<img src="figures/results_conv.png" alt="HTML5 Icon" width = 100%>
+<br>
 
+</center>
+
+## Results 
+<center>
+<img src="figures/results_sim_conv.png" alt="HTML5 Icon" width = 100%>
+<br>
+
+</center>
+
+## Results | Simulation diagnostics versus convergence diagnostics
 ```{r echo=FALSE}

### 2019-12-16 Presentation/plot_results.R

@@ -12,63 +12,6 @@ library("ggpubr")
 # load("C:/Users/User/Desktop/shinyMice/Simulation/Results/results17.Rdata.Rdata")
 # results <- dat
 
-# create
-R_plot <- results_with_CI %>% .[-1,] %>%
-  ggplot(aes(x = T, y = R_mean_X)) +
-  geom_line() +
-  geom_smooth(
-    method = "loess",
-    se = FALSE,
-    colour = "grey",
-    size = 0.5,
-    linetype = "dashed"
-  ) + 
-  xlab("") +
-  ylab(expression(paste(widehat(R)))) +
-  scale_x_continuous(breaks = seq(0, 100, by = 10)) +
-  theme_bw() +
-  theme(
-    panel.border = element_blank(),
-    panel.grid.major = element_blank(),
-    panel.grid.minor = element_blank(),
-    axis.line = element_line(colour = "black"),
-    legend.position =  ""
-  )
-
-AC_plot <- results_with_CI %>% .[-1,] %>%
-  ggplot(aes(x = T, y = AC_mean_X)) +
-  geom_line() +
-  geom_smooth(
-    method = "loess",
-    se = FALSE,
-    colour = "grey",
-    size = 0.5,
-    linetype = "dashed"
-  ) + 
-  xlab("") +
-  ylab("Auto-correlation") +
-  scale_x_continuous(breaks = seq(0, 100, by = 10)) +
-  theme_bw() +
-  theme(
-    panel.border = element_blank(),
-    panel.grid.major = element_blank(),
-    panel.grid.minor = element_blank(),
-    axis.line = element_line(colour = "black") ,
-    legend.position = ""
-  )
-
-figure1 <-
-  ggarrange(
-    R_plot,
-    AC_plot,
-    labels = c("A", "B"),
-    ncol = 1,
-    nrow = 2,
-    align = "v"
-  )
-
-figure1
-
 bias_plot <- ggplot(results_with_CI, aes(x = T, y = bias)) +
   geom_line() +
   geom_smooth(
@@ -134,7 +77,7 @@ figure2 <-
     bias_plot,
     CIW_plot,
     cov_plot,
-    labels = c("A", "B", "C"),
+    #labels = c("A", "B", "C"),
     ncol = 1,
     nrow = 3,
     align = "hv"
@@ -142,13 +85,91 @@ figure2 <-
 
 figure2
 
-figure3 <-
+# create
+R_plot <- results_with_CI %>% .[-1,] %>%
[…]

### 2019-12-16 Presentation/Presentation_Hanne.Rmd

--- 2019-12-16 Presentation/Presentation.Rmd
@@
 
-## This presentation | and the entire research archive is available on:
+<!-- ## This presentation | and the entire research archive is available on:  -->
 
-<center>
-[github.com/gerkovink/ShinyMICE/](https://github.com/gerkovink/ShinyMICE/)
-</center>
-<br><br><br>
-You can find all related materials, links and references there.
+<!-- <center> -->
+<!-- [github.com/gerkovink/ShinyMICE/](https://github.com/gerkovink/ShinyMICE/) -->
+<!-- </center> -->
+<!-- <br><br><br> -->
+<!-- You can find all related materials, links and references there.  -->
 
-## Content of the Presentation
+<!-- ## Content of the Presentation -->
 
-<div class="columns-2">
-<img src="figures/shinymice.png" alt="HTML5 Icon" width = 110%>
+<!-- <div class="columns-2"> -->
+<!-- <img src="figures/shinymice.png" alt="HTML5 Icon" width = 110%> -->
 
-<br><br><br>
+<!-- <br><br><br> -->
 
-- Introduction
-- Methodology
-- Simulation Results
-- Discussion
-</div>
+<!-- - Introduction -->
+<!-- - Methodology -->
+<!-- - Simulation Results -->
+<!-- - Discussion -->
+<!-- </div> -->
 
-# Introduction
+<!-- # Introduction -->
 
-## Multiple imputation
-- Missing data is ubiquitous [@alli02]. 
-- *Ad hoc* solutions may yield invalid inferences [@buur18].
-- @rubin87 proposed the framework of MI.
+<!-- ## Multiple imputation -->
+<!-- - Missing data is ubiquitous [@alli02].  -->
+<!-- - *Ad hoc* solutions may yield invalid inferences [@buur18]. -->
+<!-- - @rubin87 proposed the framework of MI. -->
 
@@
 
-# Methodology
+<!-- # Methodology -->
 
-## Mock code | 
-Steps in the simulation study
-```{r eval=FALSE}
-simulation <- function() {
-  mvtnorm(X,Z1,Z2) %>% 
-  mutate(Y~X+Z1+Z2) %>% 
-  for (max_iterations in 1:100) {
-    ampute() %>% 
-    impute() %T% 
-    convergence_diagnostics %>% 
-    lm(Y~X+Z1+Z2) %>% 
-    pool %>% 
-    simulation_diagnostics %>% 
-    c(., convergence_diagnostics)
-  }
-}
+<!-- ## Mock code -->
+<!-- ```{r eval=FALSE} -->
+<!-- simulation <- function() { -->
+<!--   mvtnorm(X,Z1,Z2) %>%  -->
+<!--   mutate(Y~X+Z1+Z2) %>%  -->
+<!--   for (max_iterations in 1:100) { -->
+<!--     ampute() %>%  -->
+<!--     impute() %T%  -->
+<!--     convergence_diagnostics %>%  -->
+<!--     lm(Y~X+Z1+Z2) %>%  -->
+<!--     pool %>%  -->
+<!--     simulation_diagnostics %>%  -->
+<!--     c(., convergence_diagnostics) -->
+<!--   } -->
+<!-- } -->
 
-replicate(simulation, n = 1000)
-```
+<!-- replicate(simulation, n = 1000) -->
+<!-- ``` -->
 
-# Simulation Results
+<!-- # Simulation Results -->
 
-## Simulation diagnostics
+## Results: simulation diagnostics
 <center>
@@
 
-## Convergence diagnostics
+## Results: convergence diagnostics
 <center>
@@
 
-## Results | Simulation diagnostics versus convergence diagnostics
-```{r echo=FALSE}
-dat %>% select("It.", "bias", "CIW", "cov", "R_mean_X", "AC_mean_X") %>% round(3) %>% datatable(options = list(pageLength = 5), rownames = FALSE)
-```
+<!-- ## Results | Simulation diagnostics versus convergence diagnostics -->
+<!-- ```{r echo=FALSE} -->
+<!-- dat %>% select("It.", "bias", "CIW", "cov", "R_mean_X", "AC_mean_X") %>% round(3) %>% datatable(options = list(pageLength = 5), rownames = FALSE) -->
+<!-- ``` -->
 
-## Inspect the autocorrelation dip
-```{r echo=FALSE}
-dat %>% round(3) %$%  plot_ly(x = It., y = AC_mean_X, 
-        text = paste("R hat: ", R_mean_X),
-        type = "scatter", mode = "lines") %>% 
-  layout(#title = "Auto-correlation across iterations",
-         xaxis = list(title = "Nr. of iterations"),
-         yaxis = list(title = "Auto-correlation (at lag 1)"))
-```
+<!-- ## Inspect the autocorrelation dip -->
+<!-- ```{r echo=FALSE} -->
+<!-- dat %>% round(3) %$%  plot_ly(x = It., y = AC_mean_X,  -->
+<!--         text = paste("R hat: ", R_mean_X), -->
+<!--         type = "scatter", mode = "lines") %>%  -->
+<!--   layout(#title = "Auto-correlation across iterations", -->
+<!--          xaxis = list(title = "Nr. of iterations"), -->
+<!--          yaxis = list(title = "Auto-correlation (at lag 1)")) -->
+<!-- ``` -->
 
-# Discussion
+<!-- # Discussion -->
 
@@
 
+## Results | Simulation diagnostics versus convergence diagnostics
+```{r echo=FALSE}
+dat %>% select("It.", "bias", "CIW", "cov", "R_mean_X", "AC_mean_X") %>% round(3) %>% datatable(options = list(pageLength = 5), rownames = FALSE)
+```
+
 ## Equations $\widehat{R}$

### 2020-01-13 2.ResearchReport/Presentation/Presentation.Rmd
Rename folders to make structure more clear, adjust readme file in 'main' repository.

--- 2019-12-16 Presentation/Presentation.Rmd
@@
 ## Multiple imputation
-- Missing data is ubiquitous [@alli02]. 
+- General statement:
+
+\[\text{Missing data is ubiquitous.}\] 
+<br>
+
 - *Ad hoc* solutions may yield invalid inferences [@buur18].
+<br>
+
 - @rubin87 proposed the framework of MI.
@@
 <br>
-
 <!-- Iteration number -->
 </center>
+
+## Equations $\widehat{R}$
+\[\begin{align}
+B&=\frac{T}{M-1} \sum_{m=1}^{M}\left(\bar{\theta}^{(\cdot m)}-\bar{\theta}^{(\cdot \cdot)}\right)^{2}, \quad \text{where} \quad \bar{\theta}^{(\cdot m)}=\frac{1}{T} \sum_{t=1}^{T} \theta^{(t m)}, \\
+&\quad \text{ and } \quad \bar{\theta}^{(\cdot \cdot)}=\frac{1}{M} \sum_{m=1}^{M} \bar{\theta}^{(\cdot m)} \\
+W&=\frac{1}{M} \sum_{m=1}^{M} s_{j}^{2},  \quad \text{where} \quad s_{m}^{2}=\frac{1}{T-1} \sum_{t=1}^{T}\left(\theta^{(t m)}-\bar{\theta}^{(\cdot m)}\right)^{2}  \\
+\widehat{R}&=\sqrt{\frac{\widehat{\operatorname{var}}^{+}(\theta | y)}{W}},
+\quad \text{where} \quad \widehat{\operatorname{var}}^{+}(\theta | y)=\frac{N-1}{N} W+\frac{1}{N} B.
+\end{align}\]
 
@@
 ```{r echo=FALSE}
-dat %>% select("It.", "bias", "CIW", "cov", "R_mean_X", "AC_mean_X") %>% round(3) %>% datatable(options = list(pageLength = 5), rownames = FALSE)
+dat %>% select("It.", "bias", "CIW", "cov", "R_mean_X", "AC_mean_X") %>% round(3) %>% datatable(options = list(pageLength = 10), rownames = FALSE)
 ```
@@
 
-## Equations $\widehat{R}$
-\[\begin{align}
-B&=\frac{T}{M-1} \sum_{m=1}^{M}\left(\bar{\theta}^{(\cdot m)}-\bar{\theta}^{(\cdot \cdot)}\right)^{2}, \quad \text{where} \quad \bar{\theta}^{(\cdot m)}=\frac{1}{T} \sum_{t=1}^{T} \theta^{(t m)}, \\
-&\quad \text{ and } \quad \bar{\theta}^{(\cdot \cdot)}=\frac{1}{M} \sum_{m=1}^{M} \bar{\theta}^{(\cdot m)} \\
-W&=\frac{1}{M} \sum_{m=1}^{M} s_{j}^{2},  \quad \text{where} \quad s_{m}^{2}=\frac{1}{T-1} \sum_{t=1}^{T}\left(\theta^{(t m)}-\bar{\theta}^{(\cdot m)}\right)^{2}  \\
-
-\widehat{R}&=\sqrt{\frac{\widehat{\operatorname{var}}^{+}(\theta | y)}{W}},
-\quad \text{where} \quad \widehat{\operatorname{var}}^{+}(\theta | y)=\frac{N-1}{N} W+\frac{1}{N} B.
-\end{align}\]
-
 <center>

### 2020-01-13 3.Thesis/article.Rnw
Rename folders to make structure more clear, adjust readme file in 'main' repository.

--- 2019-11-11 ResearchReport/article.Rnw
@@
 %% - \Shorttitle{} with LaTeX markup (if any), used as running title
-\title{\pkg{ShinyMICE}: an Evaluation Suite for Multiple Imputation}
-\Plaintitle{ShinyMICE: an Evaluation Suite for Multiple Imputation}
-\Shorttitle{\pkg{ShinyMICE}: an Evaluation Suite for Multiple Imputation}
+% \title{\pkg{ShinyMICE}: an Evaluation Suite for Multiple Imputation}
+% \Plaintitle{ShinyMICE: an Evaluation Suite for Multiple Imputation}
+% \Shorttitle{\pkg{ShinyMICE}: an Evaluation Suite for Multiple Imputation}
+\title{A Note on Convergence Diagnostics for Multiple Imputation using Chained Equations (MICE)}
+\Plaintitle{A Note on Convergence Diagnostics for Multiple Imputation using Chained Equations (MICE)}
+\Shorttitle{Convergence Diagnostics for MICE}
 
@@
 
-Subsequently, I will use\proglang{R}Shiny \citep{shiny} to implement the convergence indicator and existing evaluation measures in \pkg{ShinyMICE}, see Figure 1.
+Subsequently, I will use \proglang{R} Shiny \citep{shiny} to implement the convergence indicator and existing evaluation measures in \pkg{ShinyMICE}, see Figure 1.
 The application will at least contain methodology for: sensitivity analyses; data visualizations (e.g., scatter-plots, densities, cross-tabulations); and statistical evaluation of relations between variables pre- versus post-imputation (i.e., $\chi^2$~tests or $t$~tests).
@@
 
-\section{Simulation study}
+\section{Simulation study}% A Note on Convergence Diagnostics for Multiple Imputation using Chained Equations}
 
-\subsection{Simulation: Introduction}
+\subsection{Introduction}
 
@@
 
-R hat not working on MI data?
-"if over-dispersion does not hold, $\sigma_+^2$ can be too low, which can lead to falsely diagnosing convergence." \cite[p~437]{broo98}.
-Aim to replicate Lacerda et al.
-that R hat will not be smaller than 1.1 before iteration nr.
-50.
+Missing data is ubiquitous. MICE can solve stuff. [Copy Thesis Proposal Introduction.]
+
+Validity of inference MCMC algorithms threatned by non-convergence.
+Non-converegnce has two interpretations [look up source!!!]: mixing between chains and stability over iterations within chains.
+Diagnose non-convergence: $\widehat{R}$, autocorrelation, MC error, Geweke, the other one.
+However, existing methodology may not work on MI data.
+Focus on $\widehat{R}$.
+
+The potential scale reduction factor tells us how much variance could be shrunken down by running the chains infinitely long.
+That tells us something about how dependent the chains are on the starting values.
+If there is no dependence on the initial values anymore, the chains have converged (in the mixing sense of the word).
 
@@
 
+Why is R hat not applicable on MI data?
 R hat is not appropriate because it assumes overdispersed initial values, which means that the initial values of the m imputation chains should be 'far away' from the distribution that the chains are converging to (the modus of the distribution).
@@
 I should look this up.
-
-The potential scale reduction factor tells us how much variance could be shrunken down by running the chains infinitely long.
-That tells us something about how dependent the chains are on the starting values.
-If there is no dependence on the initial values anymore, the chains have converged (in the mixing sense of the word).
+And, "if over-dispersion does not hold, $\sigma_+^2$ can be too low, which can lead to falsely diagnosing convergence." \cite[p~437]{broo98}.
+Empirical finding suggests otherwise: that R hat will not be smaller than 1.1 before iteration nr.
+50.
+Therefore, the aim is to replicate empirical finding Lacerda et al.
 
@@
 
-Explain diagnostics here (see below).
-
-We want to evaluate whether the imputation procedure has converged.
-We can apply GR statistic Rhat to the first two moments of the variables of interest (mean and variance).
-The primary research interest is in determining whether Rhat is an appropriate convergence diagnostic, and if so, which level of stringency suits MI data.
+The aim is to evaluate whether the imputation procedure has converged.
+The primary research interest is in determining whether GR statistic Rhat is an appropriate convergence diagnostic, and if so, which level of stringency suits MI data.
+We can apply Rhat to the mean (or to the first two moments) of the variables of interest.
 The simulation diagnostics are as recommended by Stef van Buuren \cite{buur18}.
-That is, absolute bias confidence interval width, and empirical coverage rate (coverage probability?) across simulations.
-
-Hypothesis based on \cite{lace07} is that the conventional acceptable level of Rhat is too strict for MI data.
-We expect that the simulation diagnostics will indicate proper imputations before Rhat will.
-
-Evaluate: bias, variance, and cov rate to find problems that you might miss otherwise.
-There is no tradeoff between them.
-Also look at distributional char., and plausibility of imp.
-values, see \cite{vinknd}.
+That is, average bias, average confidence interval width, and empirical coverage rate (coverage probability) across simulations.
+Also look at distributional characteristics, and plausibility of imputed values, see \cite{vinknd}.
+% Or use: Then evaluate the regular diagnostics of MI simulations, see Vink, n.d., and  convergence.
+Diagnostics include absolute bias of the estimated regression coefficient, confidence interval width, and empirical coverage rate across simulations.
+Convergence is evaluated with $\hat{R}$, or potential scale reduction factor.
 
@@
 
+Hypothesis based on \cite{lace07} is that the conventional acceptable level of Rhat is too strict for MI data.
+We expect that the simulation diagnostics will indicate proper imputations before Rhat will.
+
 \subsection{Simulation: Simulation Approach}
 
-Set-up: simulate data, impute missing data points with varying number of iterations.
-Then evaluate the regular diagnostics of MI simulations, see Vink, n.d., and the convergence.
-Diagnostics include absolute bias of the estimated regression coefficient, confidence interval width, and empirical coverage rate (coverage probability?) across simulations.
-Convergence is evaluated with $\hat{R}$, or potential scale reduction factor.
-Proposed by Gelman and Rubin (1992), but since developed into ...
+In this study, 1000 MCMC simulations are performed. The simulation set-up consists of several steps, see pseudocode below. Simulation code is available in online appendix XYZ on Github.
+
+\begin{Code}
+# pseudocode of similation 
+simulate data with missingness
+for (number of simulation runs from 1 to 1000)
+  for (number of iterations from 1 to 100)
+    impute the missingness
+    compute convergence diagnostic
+    perform analysis
+    pool results
+    compute simulation diagnostics
+save convergence and simulation diagnostics
+\end{Code}
+
+The simulated data is a finite population of $N=1000$.
+The variables are dependent variable $Y$, independent variable $X$, and covariates $Z_1$ and $Z_2$.
+The quantity of scientific interest is the estimated regression coefficient of $X$ on $Y$ in linear regression model $Y \sim X+Z_1+Z_2$.
+The data generating model of the predictors is a multivariate normal distribution with means structure $\mu$, and variance-covariance matrix $\Sigma$.
+Outcome variable $Y$ is deduced from the predictor variables as follows:
+
+\begin{align*}
+\begin{pmatrix}X\\
+Z_{1}\\
+Z_{2}\\
+\epsilon
+\end{pmatrix} &\sim  N
+\begin{bmatrix}
+\begin{pmatrix}
+12\\
+3\\
+0.5\\
+0
+\end{pmatrix}\!\!,&
+\begin{pmatrix}
+4 & 4 & 1.8 & 0\\
+4 & 16 & 4.8 & 0\\
+1.8 & 4.8 & 9 & 0\\
+0 & 0 & 0 & 10
+\end{pmatrix}
+\end{bmatrix}\\[2\jot]
+Y &=  2 \times X + 0.5 \times Z_1 - 1 \times Z_2 + \epsilon
+\end{align*}
+
+After data generation, the complete data is 'amputed'.
+That is, the \pkg{mice} function \fct{ampute} is used to impose an MCAR missingness mechanism upon the data.
+The probability to be missing is the same for all cells, namely XYZ\%.
+Table 1 shows a summary of the generated complete data, and the amputed data.
+The amputed data is the starting point of each of the 1000 simulations.
+
+Missing data points are imputed with \pkg{mice} in \proglang{R}.
+All simulations are performed with imputation method 'norm' (Bayesian linear regression imputation), and five imputation chains.
+The number of iterations is varied over simulations ('maxit' argument between 1 and 100).
+For each number of iterations, simulation and convergence diagnostics are aggregated over 1000 MCMC simulations.
 % Missingness mechanism: "With MCAR missingness mechanisms, the probability to be missing is the same for all cases.
@@
 
-In this study, 1000 MCMC simulations are performed.
-
-\begin{Code}
-# pseudocode of similation 
-simulate data with missingness
-for (number of iterations from 1 to 100)
-  impute the missingness
-  compute convergence diagnostic
-  perform analysis
-  pool results
-  compute simulation diagnostics
-save diagnostics
-\end{Code}
-
-Simulation code is available in online appendix XYZ on Github.
-
 \subsection{Simulation: Results}
 
+Results over 1000 simulations are presented in Table XYZ.
 Rhat is not perfect for MI data.
@@
 The number of simulations is smaller than in 'regular' MCMC processes.
-Therefore, the '$n-1/n)$' correction factor can influence the estimated potential scale reduction facotr.
+Therefore, the '$(n-1/n)$' correction factor can influence the estimated potential scale reduction facotr.
 This downwards bias is in the opposite direction than expected.
+
+% latex table generated in R 3.6.1 by xtable 1.8-4 package
+% Tue Nov 12 17:43:11 2019
+\begin{table}[ht]
+\centering
+\begin{tabular}{lrrrrrr}
+  \hline
+It. & Bias & Emp. SE & CI width & Cov. rate & Rhat & Autocorr. \\ 
+  \hline
+  1 & -0.135 & 0.115 & 0.637 & 0.930 &  &   \\ 
+  2 & -0.088 & 0.125 & 0.691 & 0.990 & 1.499 & -0.500 \\ 
+  3 & -0.090 & 0.112 & 0.623 & 0.950 & 1.200 & -0.658 \\ 
+  4 & -0.093 & 0.116 & 0.645 & 0.980 & 1.146 & -0.738 \\ 
+  5 & -0.087 & 0.108 & 0.599 & 0.980 & 1.098 & -0.711 \\ 
+  6 & -0.092 & 0.116 & 0.644 & 0.950 & 1.090 & -0.674 \\ 
+  7 & -0.095 & 0.122 & 0.675 & 1.000 & 1.063 & -0.588 \\ 
+  8 & -0.099 & 0.113 & 0.626 & 0.970 & 1.058 & -0.523 \\ 
+  9 & -0.095 & 0.115 & 0.639 & 1.000 & 1.044 & -0.519 \\ 
+  10 & -0.090 & 0.111 & 0.618 & 0.990 & 1.048 & -0.414 \\ 
+  15 & -0.093 & 0.109 & 0.604 & 0.950 & 1.020 & -0.339 \\ 
+  20 & -0.098 & 0.109 & 0.606 & 0.980 & 1.023 & -0.234 \\ 
+  25 & -0.100 & 0.113 & 0.626 & 0.990 & 1.016 & -0.091 \\ 
+  30 & -0.091 & 0.118 & 0.654 & 0.960 & 1.014 & -0.153 \\ 
+  40 & -0.088 & 0.117 & 0.650 & 0.990 & 1.009 & -0.070 \\ 
+  50 & -0.087 & 0.119 & 0.662 & 0.990 & 1.009 & -0.029 \\ 
+  % 11 & -0.100 & 0.127 & 0.704 & 0.970 & 1.046 & -0.360 \\ 
+  % 12 & -0.096 & 0.115 & 0.641 & 0.980 & 1.032 & -0.354 \\ 
+  % 13 & -0.098 & 0.120 & 0.664 & 0.960 & 1.034 & -0.328 \\ 
+  % 14 & -0.094 & 0.113 & 0.630 & 0.990 & 1.030 & -0.353 \\ 
+  % 15 & -0.093 & 0.109 & 0.604 & 0.950 & 1.020 & -0.339 \\ 
+  % 16 & -0.094 & 0.114 & 0.630 & 0.980 & 1.025 & -0.272 \\ 
+  % 17 & -0.091 & 0.116 & 0.645 & 0.980 & 1.028 & -0.243 \\ 
+  % 18 & -0.091 & 0.116 & 0.645 & 0.990 & 1.023 & -0.179 \\ 
+  % 19 & -0.087 & 0.114 & 0.630 & 0.980 & 1.024 & -0.281 \\ 
+  % 20 & -0.098 & 0.109 & 0.606 & 0.980 & 1.023 & -0.234 \\ 
+  % 21 & -0.102 & 0.116 & 0.646 & 0.960 & 1.017 & -0.192 \\ 
+  % 22 & -0.095 & 0.114 & 0.633 & 0.980 & 1.016 & -0.119 \\ 
+  % 23 & -0.088 & 0.107 & 0.595 & 0.980 & 1.017 & -0.246 \\ 
+  % 24 & -0.097 & 0.112 & 0.621 & 0.980 & 1.016 & -0.148 \\ 
+  % 25 & -0.100 & 0.113 & 0.626 & 0.990 & 1.016 & -0.091 \\ 
+  % 26 & -0.096 & 0.113 & 0.628 & 0.980 & 1.017 & -0.177 \\ 
+  % 27 & -0.096 & 0.117 & 0.650 & 0.980 & 1.014 & -0.167 \\ 
+  % 28 & -0.102 & 0.116 & 0.645 & 0.960 & 1.015 & -0.038 \\ 
+  % 29 & -0.098 & 0.114 & 0.632 & 0.990 & 1.016 & -0.137 \\ 
+  % 30 & -0.091 & 0.118 & 0.654 & 0.960 & 1.014 & -0.153 \\ 
+  % 31 & -0.098 & 0.115 & 0.639 & 0.980 & 1.012 & -0.101 \\ 
+  % 32 & -0.096 & 0.113 & 0.626 & 0.940 & 1.011 & -0.095 \\ 
+  % 33 & -0.089 & 0.117 & 0.652 & 0.970 & 1.011 & -0.088 \\ 
+  % 34 & -0.093 & 0.115 & 0.639 & 0.960 & 1.013 & -0.061 \\ 
+  % 35 & -0.103 & 0.115 & 0.638 & 0.970 & 1.011 & -0.126 \\ 
+  % 36 & -0.090 & 0.113 & 0.630 & 0.980 & 1.010 & -0.161 \\ 
+  % 37 & -0.098 & 0.122 & 0.675 & 0.990 & 1.011 & -0.070 \\ 
+  % 38 & -0.095 & 0.118 & 0.656 & 0.970 & 1.011 & -0.119 \\ 
+  % 39 & -0.091 & 0.121 & 0.670 & 0.980 & 1.009 & -0.100 \\ 
+  % 40 & -0.088 & 0.117 & 0.650 & 0.990 & 1.009 & -0.070 \\ 
+  % 41 & -0.095 & 0.117 & 0.648 & 0.960 & 1.010 & -0.148 \\ 
+  % 42 & -0.095 & 0.118 & 0.657 & 0.990 & 1.009 & -0.117 \\ 
+  % 43 & -0.095 & 0.116 & 0.645 & 0.980 & 1.009 & -0.055 \\ 
+  % 44 & -0.101 & 0.125 & 0.692 & 0.980 & 1.009 & -0.068 \\ 
+  % 45 & -0.085 & 0.117 & 0.651 & 0.980 & 1.008 & -0.086 \\ 
+  % 46 & -0.092 & 0.117 & 0.649 & 0.970 & 1.008 & -0.048 \\ 
+  % 47 & -0.095 & 0.115 & 0.638 & 0.980 & 1.007 & -0.069 \\ 
+  % 48 & -0.096 & 0.112 & 0.620 & 0.970 & 1.009 & -0.076 \\ 
+  % 49 & -0.090 & 0.108 & 0.601 & 0.980 & 1.009 & -0.083 \\ 
+  % 50 & -0.087 & 0.119 & 0.662 & 0.990 & 1.009 & -0.029 \\ 
+   \hline
+\end{tabular}
+\end{table}
+
+And figure
+
+% \includegraphics[height=80mm,keepaspectratio]{res3.pdf}
+
+\begin{figure}[h]
+  \resizebox{\textwidth}{!}{ %notice the \resizebox{} command
+        \includegraphics{res3.pdf}
+  }      
+  \caption{Simulation and convergence diagnostics over 1000 MCMC simulations.}
+    \label{resultssim}
+\end{figure}
 

### 2020-01-13 3.Thesis/article.tex
Rename folders to make structure more clear, adjust readme file in 'main' repository.

--- 2020-01-13 3.Thesis/article.Rnw
@@
 %% - \Shorttitle{} with LaTeX markup (if any), used as running title
-% \title{\pkg{ShinyMICE}: an Evaluation Suite for Multiple Imputation}
-% \Plaintitle{ShinyMICE: an Evaluation Suite for Multiple Imputation}
-% \Shorttitle{\pkg{ShinyMICE}: an Evaluation Suite for Multiple Imputation}
-\title{A Note on Convergence Diagnostics for Multiple Imputation using Chained Equations (MICE)}
-\Plaintitle{A Note on Convergence Diagnostics for Multiple Imputation using Chained Equations (MICE)}
-\Shorttitle{Convergence Diagnostics for MICE}
+\title{\pkg{ShinyMICE}: an Evaluation Suite for Multiple Imputation}
+\Plaintitle{ShinyMICE: an Evaluation Suite for Multiple Imputation}
+\Shorttitle{\pkg{ShinyMICE}: an Evaluation Suite for Multiple Imputation}
 

### 2020-01-13 2.ResearchReport/3.CreateFigure.R

@@ -1,0 +1,214 @@
+# create figure to display results
+# requires the packages "dplyr", "ggplot2"
+# and the object 'results' created with '1.Excecute'
+
+# load package
+library("dplyr")
+library("ggplot2")
+library("ggpubr")
+
+# if necessary, load results
+# load("C:/Users/User/Desktop/shinyMice/Simulation/Results/results17.Rdata.Rdata")
+# results <- dat
+
+# create
+R_plot <- results_with_CI %>% .[-1,] %>%
+  ggplot(aes(x = T, y = R_var_X, color = "Chain variances")) +
+  geom_line(linetype = "dashed", size = 1) +
+  geom_line(aes(x = T, y = R_var_Z1, color = "Chain variances"),
+            linetype = "dashed",
+            size = 1) +
+  geom_line(aes(x = T, y = R_var_Z2, color = "Chain variances"),
+            linetype = "dashed",
+            size = 1) +
+  geom_line(aes(x = T, y = R_var_Y, color = "Chain variances"),
+            linetype = "dashed",
+            size = 1) +
+  geom_line(aes(x = T, y = R_mean_X, color = "Chain means")) +
+  geom_line(aes(x = T, y = R_mean_Z1, color = "Chain means")) +
+  geom_line(aes(x = T, y = R_mean_Z2, color = "Chain means")) +
+  geom_line(aes(x = T, y = R_mean_Y, color = "Chain means")) +
+  # geom_errorbar(
+  #   aes(ymin = R_mean_X_LL, ymax = R_mean_X_UL),
+  #   width = .2,
+  #   position = position_dodge(0.05)
+  # ) +
+  #geom_hline(yintercept = 1, color = "grey") +
+  #geom_hline(yintercept = 1.1) +
+  #geom_hline(yintercept = 1.01) +
+  xlab("") +
+  ylab(expression(paste(widehat(R)))) +
+  scale_x_continuous(breaks = seq(0, 100, by = 10)) +
+  scale_y_continuous(breaks = seq(.8, 2.2, by = .2)) +
+  scale_color_manual(
+    name = "Legend",
+    values = c("Chain means" = 1, "Chain variances" = 8),
+    guide = "legend"
+  ) +
+  guides(colour = guide_legend(override.aes = list(linetype = c(1, 8)))) +
+  theme_bw() +
+  theme(
+    panel.border = element_blank(),
[…]

### 2020-01-13 2.ResearchReport/Functions/Archive/Simulate.R

@@ -1,0 +1,85 @@
+# simulate imputations with varying niter
+# requires the package 'mice'
+
+# simulation function
+simulate.function <- function(data, n.iter, n.sim, true_effect = 2, m = 5) {
+  # set.seed(123)
+  pb <- txtProgressBar(min = 0, max = n.iter, style = 3)
+  z_value <- qnorm(.025,lower.tail = F)
+  m <- 5
+  #objects to store sims
+  temp <- out <- OUT <- list()
+  # objects to store converegnce diagnostics
+  r.hat <- matrix(NA, nrow = n.iter, ncol = n.sim)
+  # convergence <- matrix(NA, n.iter, ncol = 4)
+  # loop over all possible nrs of iterations
+  for (i in 1:n.iter) {
+    # repeat each analysis nsim times
+    for (j in 1:n.sim) {
+      # for each iterarion value, store each nsim 'repeatings' in lower case out
+      temp <-
+        mice(
+          data = data,
+          method = "norm",
+          maxit = i,
+          print = F
+        )
+      # extract r hat
+      if (i < 2) {r.hat[i, j] <- NA}
+      else {r.hat[i, j] <- max(as.numeric(Rhat.mice(temp)$Rhat.M.imp))}
+      
+      # apply analysis and extract est.
+      out[[i]] <- temp %>% 
+        pool(lm.mids(Y ~ X + Z1 + Z2)) %>% 
+               diagnostics.function(true_effect = true_effect, z_value = z_value)
+      
+      # extract diagnostics
+      # out[[i]] <- 
+    }
+    # for each iteration value, store the combined nsim simulations together
+    setTxtProgressBar(pb, i)
+    OUT[[j]] <- out
+  }
+  
+  # prepare for output
+  # convergence[, 1] <- rowMeans(r.hat, na.rm = T)
+  # convergence[, 2] <- rowMeans(bias)
+  # convergence[, 2] <- rowMeans(r.hat < 1.2, na.rm = T)
+  # convergence[, 3] <- rowMeans(r.hat < 1.1, na.rm = T)
+  # convergence[, 4] <- rowMeans(r.hat < 1.01, na.rm = T)
+  # colnames(convergence) <- c("Rhat", "Rhat_lenient", "Rhat_mid", "Rhat_stringent")
+  
+  # end function
+  close(pb)
+  return(OUT)
+}
+
+# # simulation function
+# simulate.function <- function(data, n.iter, n.sim) {
+#   # set.seed(123)
[…]

### 2020-01-13 Simulation/1.Execute.R

@@ -1,92 +1,0 @@
-# script to perform simulation study on convergence of mice
-# author: Hanne Oberman
-
-###
-
-# load required packages
-library(mice)
-library(mvtnorm)
-library(miceadds)
-library(data.table)
-library(tidyverse)
-# or if you don't want the entire tidyverse, just load these
-# library(dplyr)
-# library(ggplot2)
-# library(purrr)
-# library(magrittr)
-# library(tibble)
-
-# load simulation/evaluation functions
-source("Simulation/Functions/CreateData.R")
-source("Simulation/Functions/Convergence.R")
-source("Simulation/Functions/Convergence_supplement.R")
-source("Simulation/Functions/Autocorrelation.R")
-source("Simulation/Functions/Impute.R")
-source("Simulation/Functions/Evaluate.R")
-
-# simulation parameters
-populationsize <- 1000 #n of simulated dataset
-n.iter <- 100 #nr of iterations (varying 1:n.iter)
-n.sim <- 100 #nr of simulations per iteration value
-true_effect <- 2 #regression coefficient to be estimated
-
-# start simulation study
-set.seed(1111)
-
-###
-
-# simulate data once
-data <- data.simulation(n = populationsize, true_effect)
-# run ampute once to get patterns object (and )to be able to adjust it for multivariate missingness)
-amp_patterns <- ampute(data)$patterns
-# for multivariable missingness, uncomment this
-amp_patterns[1:4, 1] <- 0
-
-# combine separate functions into wrapper
-simulate <- function(data, n.iter, true_effect, patterns) {
-  pb <- txtProgressBar(min = 0, max = n.iter, style = 3)
-  
-  # remove values at random with 20 percent probability to be missing
-  ampdata <- ampute(data, patterns = amp_patterns, prop = 0.8, mech = "MCAR")$amp
-  
-  # object for output
-  res <- list()
-  # repeat mi procedure 'runs'  times for each nr of iterations
-  #for (run in 1:runs) {
-    for (i in 1:n.iter) {
-      res[[i]] <- test.impute(true_effect, data = ampdata, maxit = i)
-      setTxtProgressBar(pb, i)
-    }
-  #}
-  close(pb)
-  # output
-  names(res) <- 1:n.iter
-  res
-}
-
-# simulate
[…]

### 2020-01-13 2.ResearchReport/2.Presentation/Presentation_Hanne.Rmd
clean-up research report folder and edit readme

--- 2019-12-16 Presentation/Presentation_Hanne.Rmd
@@
 subtitle: "A Note on Convergence Diagnostics for Multiple Imputation Algorithms"
-author: "Hanne Oberman"
-date: "December 16, 2019"
+author: "Hanne Oberman, December 16, 2019"
+date: "Supervisors: Stef van Buuren and Gerko Vink | Mentor: Daniela Cianci"
 output: 
@@
 bibliography: ShinyMICE.bib
-
+nocite: | 
+  @buur18, @R, @mice, @shiny, @veht19
 ---
@@
 ```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = FALSE)
+
 library("mice")
@@
 
+<style>
+div.footnotes {
+  position: absolute;
+  bottom: 0;
+  margin-bottom: 10px;
+  width: 80%;
+  font-size: 0.6em;
+}
+</style>
+
+<script src="https://ajax.googleapis.com/ajax/libs/jquery/3.1.1/jquery.min.js"></script>
+
+<script>
+  $(document).ready(function() {
+    $('slide:not(.backdrop):not(.title-slide)').append('<div class=\"footnotes\">');
+
+    $('footnote').each(function(index) {
+      var text  = $(this).html();
+      var fnNum = (index+1).toString().sup();
+      $(this).html(text + fnNum);
+
+      var footnote   = fnNum + ': ' + $(this).attr('content') + '<br/>';
+      var oldContent = $(this).parents('slide').children('div.footnotes').html();
+      var newContent = oldContent + footnote;
+      $(this).parents('slide').children('div.footnotes').html(newContent);
+    });
+  });
+</script>
 <!-- ## Disclaimer -->
@@
 
-## Multiple imputation
+## Multiple imputation <footnote content="Figure adapted from van Buuren (2018)"></footnote>
+
 <br><br>
@@
 
-## Algorithmic convergence
+## Algorithmic convergence <footnote content="Figure adapted from van Buuren (2018)"></footnote>
 <center>
@@
 
-## Results: simulation diagnostics
+## Results: simulation diagnostics <footnote content="Results obtained using R (2019)"></footnote>
 <center>
@@
 
-## Results: convergence diagnostics
+## Results: convergence diagnostics <footnote content="Results obtained using R (2019)"></footnote>
 <center>
@@
 
-## Results 
+## Results <footnote content="Results obtained using R (2019)"></footnote>
 <center>
@@
 
-## Discussion
+## Discussion <footnote content="Figure adapted from mice (2014) and Shiny (2017)"></footnote>
 <center>
@@
 
-## Impression of ShinyMICE
+## Impression of ShinyMICE <footnote content="Figure adapted from mice (2014) and Shiny (2017)"></footnote>
 <center>
@@
 
-## Equations $\widehat{R}$
+## Equations $\widehat{R}$ <footnote content="Adapted from Vehtari et al. (2019)"></footnote>
 \[\begin{align}
@@
 W&=\frac{1}{M} \sum_{m=1}^{M} s_{j}^{2},  \quad \text{where} \quad s_{m}^{2}=\frac{1}{T-1} \sum_{t=1}^{T}\left(\theta^{(t m)}-\bar{\theta}^{(\cdot m)}\right)^{2}  \\
-
 \widehat{R}&=\sqrt{\frac{\widehat{\operatorname{var}}^{+}(\theta | y)}{W}},

### 2020-01-13 2.ResearchReport/1.SimulationStudy/Functions/Autocorrelation.R

@@ -1,0 +1,37 @@
+# Compute the maximum auto-correlation at lag 1
+# Used within imputation function
+
+autocorr_function <- function(imp, maxit, m = 5, n.var = 4, moment = "mean"){
+  
+  # object for for loop
+  ac_per_chain <- numeric(length = m)
+  ac_per_var <- numeric(length = n.var)
+  names(ac_per_var) <- attr(imp$chainMean, "dimnames")[[1]]
+  
+  # make suitable for convergence of mean and variance
+  if (moment == "mean") {
+    sims <- imp$chainMean
+  } else if (moment == "variance") {
+    sims <- imp$chainVar
+  }
+  
+  # compute convergence diagnostic per variable
+  for (v in 1:n.var) {
+    for (chain in 1:m) {
+      # ac[chain, v] <-
+      ac_per_chain[chain] <-
+        acf(sims[v, , chain],
+          lag.max = 1,
+          plot = F)$acf[-1] 
+    }
+    ac_per_var[v] <- ac_per_chain[which.max(abs(ac_per_chain))]
+}
+# output
+  return(ac_per_var)
+}
+
+# NB. This function can be adjusted to show the nr. of chains with significant auto-correlations,
+# output will then be the fraction of imputation chains with significant autocorrelation at lag 1.
+# To get this, add:
+# # critical value significance 
+# AC_lim <- qnorm((1 + .95) / 2) / sqrt(maxit)

### 2020-01-13 2.ResearchReport/Functions/Archive/Simulate.R

@@ -1,85 +1,0 @@
-# simulate imputations with varying niter
-# requires the package 'mice'
-
-# simulation function
-simulate.function <- function(data, n.iter, n.sim, true_effect = 2, m = 5) {
-  # set.seed(123)
-  pb <- txtProgressBar(min = 0, max = n.iter, style = 3)
-  z_value <- qnorm(.025,lower.tail = F)
-  m <- 5
-  #objects to store sims
-  temp <- out <- OUT <- list()
-  # objects to store converegnce diagnostics
-  r.hat <- matrix(NA, nrow = n.iter, ncol = n.sim)
-  # convergence <- matrix(NA, n.iter, ncol = 4)
-  # loop over all possible nrs of iterations
-  for (i in 1:n.iter) {
-    # repeat each analysis nsim times
-    for (j in 1:n.sim) {
-      # for each iterarion value, store each nsim 'repeatings' in lower case out
-      temp <-
-        mice(
-          data = data,
-          method = "norm",
-          maxit = i,
-          print = F
-        )
-      # extract r hat
-      if (i < 2) {r.hat[i, j] <- NA}
-      else {r.hat[i, j] <- max(as.numeric(Rhat.mice(temp)$Rhat.M.imp))}
-      
-      # apply analysis and extract est.
-      out[[i]] <- temp %>% 
-        pool(lm.mids(Y ~ X + Z1 + Z2)) %>% 
-               diagnostics.function(true_effect = true_effect, z_value = z_value)
-      
-      # extract diagnostics
-      # out[[i]] <- 
-    }
-    # for each iteration value, store the combined nsim simulations together
-    setTxtProgressBar(pb, i)
-    OUT[[j]] <- out
-  }
-  
-  # prepare for output
-  # convergence[, 1] <- rowMeans(r.hat, na.rm = T)
-  # convergence[, 2] <- rowMeans(bias)
-  # convergence[, 2] <- rowMeans(r.hat < 1.2, na.rm = T)
-  # convergence[, 3] <- rowMeans(r.hat < 1.1, na.rm = T)
-  # convergence[, 4] <- rowMeans(r.hat < 1.01, na.rm = T)
-  # colnames(convergence) <- c("Rhat", "Rhat_lenient", "Rhat_mid", "Rhat_stringent")
-  
-  # end function
-  close(pb)
-  return(OUT)
-}
-
-# # simulation function
-# simulate.function <- function(data, n.iter, n.sim) {
-#   # set.seed(123)
[…]

### 2020-01-13 2.ResearchReport/Presentation/plot_results.R

@@ -1,178 +1,0 @@
-# Figure for presentation
-# create figure to display results
-# requires the packages "dplyr", "ggplot2"
-# and the object 'results' created with '1.Excecute'
-
-# load package
-library("dplyr")
-library("ggplot2")
-library("ggpubr")
-
-# if necessary, load results
-# load("C:/Users/User/Desktop/shinyMice/Simulation/Results/results17.Rdata.Rdata")
-# results <- dat
-
-bias_plot <- ggplot(results_with_CI, aes(x = T, y = bias)) +
-  geom_line() +
-  geom_smooth(
-    method = "loess",
-    se = FALSE,
-    colour = "black",
-    size = 0.5,
-    linetype = "dashed"
-  ) +
-  xlab("") +
-  ylab("Bias") +
-  scale_x_continuous(breaks = seq(0, 100, by = 10)) +
-  theme_bw() +
-  theme(
-    panel.border = element_blank(),
-    panel.grid.major = element_blank(),
-    panel.grid.minor = element_blank(),
-    axis.line = element_line(colour = "black")
-  )
-
-CIW_plot <- ggplot(results_with_CI, aes(x = T, y = CIW)) +
-  geom_line() +
-  geom_smooth(
-    method = "loess",
-    se = FALSE,
-    colour = "black",
-    size = 0.5,
-    linetype = "dashed"
-  ) +
-  xlab("") +
-  ylab("CI width") +
-  scale_x_continuous(breaks = seq(0, 100, by = 10)) +
-  theme_bw() +
-  theme(
-    panel.border = element_blank(),
-    panel.grid.major = element_blank(),
-    panel.grid.minor = element_blank(),
-    axis.line = element_line(colour = "black")
-  )
-
-cov_plot <- ggplot(results_with_CI, aes(x = T, y = cov * 100)) +
-  geom_line() +
-  geom_smooth(
-    method = "loess",
-    se = FALSE,
-    colour = "black",
-    size = 0.5,
-    linetype = "dashed"
-  ) +
-  xlab("Number of iterations") +
-  ylab("Coverage (%)") +
-  scale_x_continuous(breaks = seq(0, 100, by = 10)) +
-  theme_bw() +
-  theme(
-    panel.border = element_blank(),
-    panel.grid.major = element_blank(),
-    panel.grid.minor = element_blank(),
[…]

### 2020-01-13 3.Thesis/Writeup/article.tex
clean-up thesis folder

--- 2019-11-11 ResearchReport/article.Rnw
@@
 %% - \Shorttitle{} with LaTeX markup (if any), used as running title
-\title{\pkg{ShinyMICE}: an Evaluation Suite for Multiple Imputation}
-\Plaintitle{ShinyMICE: an Evaluation Suite for Multiple Imputation}
-\Shorttitle{\pkg{ShinyMICE}: an Evaluation Suite for Multiple Imputation}
+% \title{\pkg{ShinyMICE}: an Evaluation Suite for Multiple Imputation}
+% \Plaintitle{ShinyMICE: an Evaluation Suite for Multiple Imputation}
+% \Shorttitle{\pkg{ShinyMICE}: an Evaluation Suite for Multiple Imputation}
+\title{\pkg{ShinyMICE}: An Evaluation Suite for Multiple Imputation}
+\Plaintitle{ShinyMICE: An Evaluation Suite for Multiple Imputation}
+\Shorttitle{Evaluation Suite for Multiple Imputation}
 
@@
 \Abstract{
-  This Research Report contains the Introduction and Methods section of the technical paper that will be submitted for publication in \emph{Journal of Statistical Software}.
-("There is no page limit, nor a limit on the number of figures or tables", see \url{https://www.jstatsoft.org/pages/view/authors}.) I have chosen to use the second format from the Research Report guidelines: "It is the first half of the thesis, i.e., there are no results included yet, but the report contains a full introduction including a literature review and a methods section that contains details about the data, instruments and or statistical procedures".
-The goal was to develop novel methodology and guidelines for evaluating multiple imputation methods, and implement these in an interactive evaluation framework for multiple imputation: \pkg{ShinyMICE}.
+  The goal of this Master's thesis was to develop novel methodology and guidelines for evaluating multiple imputation methods, and implement these in an interactive evaluation framework for multiple imputation: \pkg{ShinyMICE}.
   }
@@
 
-At some point, any scientist conducting statistical analyses will run into a missing data problem \citep{alli02}.
+\textbf{[From thesis proposal:]} At some point, any scientist conducting statistical analyses will run into a missing data problem \citep{alli02}.
 Missingness is problematic because statistical inference cannot be performed on incomplete data, and  ad hoc solutions can yield wildly invalid results \citep{buur18}.
@@
 
-With MI, many assumptions are made about the nature of the observed and missing parts of the data and their relation to the 'true' \emph{data generating model} \citep{buur18}.
+\textbf{[From thesis proposal:]} With MI, many assumptions are made about the nature of the observed and missing parts of the data and their relation to the 'true' \emph{data generating model} \citep{buur18}.
 Without proper evaluation of the imputations and the underlying assumptions, any drawn inference may erroneously be deemed valid.
@@
 
-\subsection{Theoretical Background from Thesis Proposal}
-
-The validity of the MI solution depends on numerous assumptions that cannot be verified from the observed data alone.
+\textbf{[From thesis proposal:]} The validity of the MI solution depends on numerous assumptions that cannot be verified from the observed data alone.
 So instead of statistical tests for assumptions, evaluation procedures have been developed.
@@
 \begin{enumerate}
-\item A missingness mechanism is said to be ignorable when the probability to be missing does not depend on the missing data itself.
+\item \textbf{[From thesis proposal:]} A missingness mechanism is said to be ignorable when the probability to be missing does not depend on the missing data itself.
 Violation of this assumption can gravely affect inferences.
@@
 %
-\item Congeneal imputation models capture all required relations between observed and missing parts of the data.
+\item \textbf{[From thesis proposal:]} Congeneal imputation models capture all required relations between observed and missing parts of the data.
 The extent to which this has been successful can be evaluated by plotting conditional distributions \citep{abay08}.
@@
 %
-\item The third assumption is met when the MI algorithm converges to a stable distribution.
+\item \textbf{[From thesis proposal:]} The third assumption is met when the MI algorithm converges to a stable distribution.
 However, conventional measures to diagnose convergence-- e.g., Gelman and Rubin's \citeyear{gelm92} statistic $\widehat{R}$ --are not applicable on multiply imputed data \citep{lace07}.
@@
 
-In short, the existing literature provides both possibilities and limitations to evaluating the validity of multiply imputed data.
+\textbf{[From thesis proposal:]} In short, the existing literature provides both possibilities and limitations to evaluating the validity of multiply imputed data.
 The goal of this research project is to develop novel methodology and guidelines for evaluating MI methods, and implement these in an interactive evaluation framework for multiple imputation.
@@
 
-\subsection{Approach from Thesis Proposal}
-
-Initially, the research project will consist of an investigation into algorithmic convergence of MI algorithms.
+\textbf{[From thesis proposal:]} Initially, the research project will consist of an investigation into algorithmic convergence of MI algorithms.
 I will replicate Lacerda et al.'s simulation study on $\widehat{R}$ \citep{lace07}, and develop novel guidelines for assessing convergence.
@@
 
-Subsequently, I will use\proglang{R}Shiny \citep{shiny} to implement the convergence indicator and existing evaluation measures in \pkg{ShinyMICE}, see Figure 1.
+\textbf{[From thesis proposal:]} Subsequently, I will use \proglang{R} Shiny \citep{shiny} to implement the convergence indicator and existing evaluation measures in \pkg{ShinyMICE}, see Figure 1.
 The application will at least contain methodology for: sensitivity analyses; data visualizations (e.g., scatter-plots, densities, cross-tabulations); and statistical evaluation of relations between variables pre- versus post-imputation (i.e., $\chi^2$~tests or $t$~tests).
 
-A working beta version of \pkg{ShinyMICE} will be considered a sufficient milestone to proceed with writing a technical paper on the methodology and the software.
+\textbf{[From thesis proposal:]} A working beta version of \pkg{ShinyMICE} will be considered a sufficient milestone to proceed with writing a technical paper on the methodology and the software.
 I will submit the paper for publication in \emph{Journal of Statistical Software}.
 Finally, \pkg{ShinyMICE} will be integrated into the existing MICE environment, and a vignette for applied researchers will be written.
-
 The \proglang{R} code and documentation of this project will be open source (available on Github).
 Since the study does not require the use of unpublished empirical data, I expect that the FETC will grant the label exempt.
-
-% \includegraphics[height=80mm,keepaspectratio]{impression} % this is not correct yet, it needs a figure label and stuff
 
@@
 
-\section{Simulation study}
-
-\subsection{Simulation: Introduction}
-
-\emph{It is written as a (mini) thesis, with an introduction, methods section, some results (i.e., preliminary analyses, or pilot simulations), and a discussion of results.
-The length of the research report should be maximally 2500 words of text (without references list and or tables and figures).
-Please do not include appendices, and no more than 6 tables or figures.
-Table and Figure captions do not count towards the word limit.
-An abstract may be included, but is not necessary.}
-
-R hat not working on MI data?
-"if over-dispersion does not hold, $\sigma_+^2$ can be too low, which can lead to falsely diagnosing convergence." \cite[p~437]{broo98}.
-Aim to replicate Lacerda et al.
-that R hat will not be smaller than 1.1 before iteration nr.
-50.
-
-As recommended by ... Rhat is computed as .... to be able to detect ... in the tails of the distribution.
-
-R hat is not appropriate because it assumes overdispersed initial values, which means that the initial values of the m imputation chains should be 'far away' from the distribution that the chains are converging to (the modus of the distribution).
-With MI procedures, initial values are chosen such that ...
-or maybe even estimated from the observed data.
-This means that at most one initial value is necessary, but probably none at al.
-I should look this up.
-
-The potential scale reduction factor tells us how much variance could be shrunken down by running the chains infinitely long.
-That tells us something about how dependent the chains are on the starting values.
-If there is no dependence on the initial values anymore, the chains have converged (in the mixing sense of the word).
-
-Convergence can also be interpreted as stable over iterations, or non-recurring.
-Non-recurrence can be evaluated with autocorrelation.
-Autocorrelation shows how dependent subsequent draws of an imputation chain are on the previous value.
-If there is a lot of dependence, draws at e.g.
-iteration five are significantly correlated with the value of the first draw.
-Ideally, we want the chains to intermingle nicely, without trends witihin chains.
-Autocorrelation above the confidence level indicate dependence within chains.
-
-\subsection{Simulation: Methodology}
-
-Explain diagnostics here (see below).
-
-We want to evaluate whether the imputation procedure has converged.
-We can apply GR statistic Rhat to the first two moments of the variables of interest (mean and variance).
-The primary research interest is in determining whether Rhat is an appropriate convergence diagnostic, and if so, which level of stringency suits MI data.
-The simulation diagnostics are as recommended by Stef van Buuren \cite{buur18}.
-That is, absolute bias confidence interval width, and empirical coverage rate (coverage probability?) across simulations.
-
-Hypothesis based on \cite{lace07} is that the conventional acceptable level of Rhat is too strict for MI data.
-We expect that the simulation diagnostics will indicate proper imputations before Rhat will.
-
-Evaluate: bias, variance, and cov rate to find problems that you might miss otherwise.
-There is no tradeoff between them.
-Also look at distributional char., and plausibility of imp.
-values, see \cite{vinknd}.
-
 % Distributional characteristics: In practice, the distribution of the incomplete data may differ greatly from the observed data.
@@
 When evaluating imputation routines, the evaluator should mention whether the routine is prone to deliver implausible value
-
-\subsection{Simulation: Simulation Approach}
-
-Set-up: simulate data, impute missing data points with varying number of iterations.
-Then evaluate the regular diagnostics of MI simulations, see Vink, n.d., and the convergence.
-Diagnostics include absolute bias of the estimated regression coefficient, confidence interval width, and empirical coverage rate (coverage probability?) across simulations.
-Convergence is evaluated with $\hat{R}$, or potential scale reduction factor.
-Proposed by Gelman and Rubin (1992), but since developed into ...
-% Missingness mechanism: "With MCAR missingness mechanisms, the probability to be missing is the same for all cases.
-This is a necessary simulation condition for evaluating the performance of imputation procedures.
-If an imputation method is not able to solve the problem (i.e.
-yield valid inference) under MCAR, the statistical properties of the procedure are not sound." (Vink, n.d., p.
-4)
-
-In this study, 1000 MCMC simulations are performed.
-
-\begin{Code}
-# pseudocode of similation 
-simulate data with missingness
-for (number of iterations from 1 to 100)
-  impute the missingness
-  compute convergence diagnostic
-  perform analysis
-  pool results
-  compute simulation diagnostics
-save diagnostics
-\end{Code}
-
-Simulation code is available in online appendix XYZ on Github.
-
-\subsection{Simulation: Results}
-
-Rhat is not perfect for MI data.
-The value shoudl theoretically not be smaller than one, yet it happened several times in this study.
-How could Rhat smaller than 1 occur?
-The number of simulations is smaller than in 'regular' MCMC processes.
-Therefore, the '$n-1/n)$' correction factor can influence the estimated potential scale reduction facotr.
-This downwards bias is in the opposite direction than expected.
-
-\subsection{Simulation Discussion}
-
-"The mixture-of-sequences variance, V, should stabilize as a function of n.
-(Before convergence, we expect $\sigma^2$ to decrease with n, only increasing if the sequences explore a new area of parameter space, which would imply that the original sequences were not overdispersed for the particular scalar summary being monitored.)" \cite[p~438]{broo98}.
-
-We would like to have convergence measures for multivariable statistics (scalars?) of interest.
-This is, however, dependent of the complete data model.
-The eigenvector decomposition methof proposed by McKay (?) should be implemented.
-I could not find any resources to apply this method and it is outside the scope of this thesis to investigate how this apprach could be implemented.
 
@@
 
-The results in this paper were obtained using \proglang{R}~\Sexpr{paste(R.Version()[6:7], collapse = ".")} with the \pkg{mice}~\Sexpr{paste(packageVersion("mice"), collapse = ".")} package.
+The results in this paper were obtained using \proglang{R}~3.6.2 with the \pkg{mice}~3.7.0 package.
 \proglang{R} itself and all packages used are available from the Comprehensive \proglang{R} Archive Network (CRAN) at \url{https://CRAN.R-project.org/}.

### 2020-01-13 3.Thesis/Writeup/Thesis.Rnw

--- 2020-01-13 3.Thesis/Writeup/article.tex
@@
 %% need no \usepackage{Sweave}
+\SweaveOpts{engine=R, eps=FALSE, keep.source = TRUE}
+<<preliminaries, echo=FALSE, results=hide>>=
+options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
+library("MASS")
+@
 
@@
 % data("quine", package = "MASS")
+@
 %
@@
 % box()
+@
 % \caption{\label{fig:quine} Frequency distribution for number of days absent
@@
 %   family = poisson)
+@
 %
@@
 % m_nbin <- glm.nb(Days ~ (Eth + Sex + Age + Lrn)^2, data = quine)
+@
 %
@@
 % BIC(m_pois, m_nbin)
+@
 % %
@@
 % summary(m_nbin)
+@
 
@@
 
-The results in this paper were obtained using \proglang{R}~3.6.2 with the \pkg{mice}~3.7.0 package.
+The results in this paper were obtained using \proglang{R}~\Sexpr{paste(R.Version()[6:7], collapse = ".")} with the \pkg{mice}~\Sexpr{paste(packageVersion("mice"), collapse = ".")} package.
 \proglang{R} itself and all packages used are available from the Comprehensive \proglang{R} Archive Network (CRAN) at \url{https://CRAN.R-project.org/}.

### 2020-01-15 2.ResearchReport/1.SimulationStudy/Writeup/NoteConvergence.Rnw

--- 2019-12-15 Simulation/Writeup/NoteConvergence.Rnw
@@
 \documentclass[article]{jss}
+
+%% -- Set-up document -------------------------------------------------------------
+%%
+%%
+%%
+%%
 
@@
 
-%% another package (only for this demo article)
-% \usepackage{framed}
-\usepackage{multirow, amsmath, tikz}
-\usetikzlibrary{fit, positioning}
+%% more packages 
+\usepackage{multirow, amsmath, tikz} % for table, equations, and flow chart figure (respectively)
+\usetikzlibrary{fit, positioning} % for flow chart figure
 
@@
 We conclude that potential scale reduction factor $\widehat{R}$ and autocorrelation may be appropriate convergence diagnostics, but their performance differs from conventional applications to iterative algorithmic procedures.
-These results will be implemented into an interactive framework for the evaluation of multiple imputation methods that is under development as Research Master's thesis.
-%\textbf{TO DO still: check worst lin pred ref, check ref list, add legend to sim diags, check table and fig captions (abbrevations), references to tables, figs, appendix}
-%%%%%%%%%%%%%
-%The ubiquitous problem of missing information may be circumvented by implementing the framework of multiple imputation (MI).
-With MI, missing datapoints are `filled in' (i.e.
-imputed) several times, and statistical inference is performed on each imputed data set.
-The validity of inferences may be threatened by non-convergence of this iterative algorithmic procedure, but diagnostics to evaluate algorithmic convergence have not been implemented in MI software.
-This note investigates the performance of two convergence diagnostics (\citep[$\widehat{R}$,][]{gelm92}; \citep[autocorrelation,][]{scha97}) through model-based simulation.
-To investigate convergence of the algorithm, the number of iterations is varied between one and one-hundred.
-Each of these 100 simulation conditions is replicated 1000 times.
-The convergence diagnostics are evaluated against three recommended simulation diagnostics (bias, confidence interval width, and coverage rate \citep{buur18}, since no benchmark diagnostic is available.
-This simulation study shows that three iterations may be sufficient to obtain ...
-\textbf{This Research Report contains a simulation study that serves as the basis of the technical paper that will be submitted for publication in \emph{Journal of Statistical Software}.
-I aim to publish a pre-print of this research report on \href{https://arxiv.org/}{arXiv}.
-That way, I can refer to the simulation study described here, without `bulking up' the technical paper that I want to submit for publication in JSS.
-Another option would be to attach this note as an online appendix to the technical paper on ShinyMICE.}
-%%%%%%%%%%%%
-% This Research Report contains a simulation study that serves as the basis of the technical paper that will be submitted for publication in \emph{Journal of Statistical Software}.
-I have chosen to use the first format from the Research Report guidelines: ``\emph{It is written as a (mini) thesis, with an introduction, methods section, some results (i.e., preliminary analyses, or pilot simulations), and a discussion of results.
-The length of the research report should be maximally 2500 words of text (without references list and or tables and figures).
-Please do not include appendices, and no more than 6 tables or figures.
-Table and Figure captions do not count towards the word limit.
-An abstract may be included, but is not necessary}''.
-I aim to publish a pre-print of this research report on \href{https://arxiv.org/}{ArXiv}.
-That way, I can refer to the simulation study described here, without 'bulking up' the technical paper that I want to submit for publication in JSS.
-Another option would be to attach this note as online appendix to the technical paper on ShinyMICE.
-  }
+These results will be implemented into an interactive framework for the evaluation of multiple imputation methods that is under development as Research Master's thesis.}
 
@@
 %% -- Introduction -------------------------------------------------------------
+%%
+%%
+%%
+%%
 
-\section{Introduction} \label{sec:intro} % curly brackets can handle typesetting
+\section{Introduction} \label{sec:intro}
 
@@
 Evaluation measures, however, are currently missing or under-developed in MI software, like the world-leading \pkg{mice} package \citep{mice} in \proglang{R} \citep{R}.
-%\textbf{ADD sentence to link eval measures to convergence!}
-%Therefore, I will answer the following question: 'Which measures are vital for evaluating the validity of multiply imputed data?'.
 The goal of this research project is to develop novel methodology and guidelines for evaluating MI methods.
@@
 \footnotetext{All programming code used in this note is available from \href{https://github.com/gerkovink/shinyMice/simulation}{github.com/gerkovink/ShinyMICE/simulation}.} 
-%I will replicate Lacerda et al.'s simulation study on $\widehat{R}$ \citep{lace07}, and develop novel guidelines for assessing convergence.
-Ideally, I will integrate several diagnostics (e.g., $\widehat{R}$, \emph{autocorrelation}, and \emph{simulation error}) into a single summary indicator to flag non-convergence.
 We will evaluate how convergence of the MI algorithm can be diagnosed.
 
-% for diagnostics for evalua one of these evaluation measures that is vital in evaluating MI procedures: a diagnostic to assess convergence of the MI algorithm.
-%The implementation limits itself to the MI algorithm
-%More specifically, this note focuses on the MI algorithm that is implemented in the \proglang{R} package \pkg{mice}: `Multiple Imputation using Chained Equations' (MICE) \textbf{[can be removed, can be found in the mice ref]}.
-
-%The research question that will be addressed is: %.
-a measure that is vital in evaluating one vital measure for evaluating MI procedures--an algorithmic convergence diagnostic.
-This note specifically investigates one vital evaluation measure: how to diagnose convergence of a multiple imputation algorithm.
-%This note focusses on one measure that is vital for evaluating the validity of multiply imputed data: a convergence diagnostic. of the MI algorithm. 
-%Therefore, the aim of this research project is to investigate which measures are vital for .
-how the convergence properties of the MI algorithm that is implemented in \pkg{mice}: `Multiple Imputation using Chained Equations' (MICE).
-This note addresses the question:
-%\emph{`How to diagnose convergence of the multiple imputation algorithm MICE?'.}
-
-%% -- Features ---------------------------------------------------------------
+%% -- Terminology ---------------------------------------------------------------
 \subsection{Terminology} \label{sec:terms}
 
-%The intended audience of this note consists of anyone who uses multiple imputation to solve missing data problems.
-
+%The intended audience of this note consists of anyone who uses multiple imputation to solve missing data problem. 
 This note follows notation and conventions of \pkg{mice} \citep{mice}.
-% to suit the \pkg{mice} environment.
+% to suit the \pkg{mice} environment, see \cite[p.~4]{mice}.
 %Deviations from the `original' notation by \cite{rubin87} are described in \citep[\S~2.2.3]{buur18}.
@@
 Which parts of $Y$ are missing is determined by the `missingness mechanism'.
-%The relation between $Y_{obs}$ and $Y_{mis}$ \textbf{REMOVE?
-Response indicator $R$ shows whether a data value in $Y$ is missing or observed.} The relation between $R$, $Y_{obs}$, and $Y_{mis}$ determines the missingness mechanism.
 This note only considers a `missing completely at random' (MCAR) mechanism, where the probability of being missing is equal for all $n \times p$ cells in $Y$ \citep{rubin87}.
-
-%In this note, the terms `unobserved' and `missing' data are used interchangeably to refer to $Y_{mis}$.
-The terms `incomplete' or `observed' data denote $Y_{obs}$.
-Incomplete data is the starting point of the multiple imputation procedure.
 
@@
 Since $Q$ is estimated on each completed dataset, $m$ separate $\hat{Q}$-values are obtained.
-These $m$ values are combined into a single pooled estimate $\bar{Q}$.}} %\footnotemark}
+These $m$ values are combined into a single pooled estimate $\bar{Q}$.}}
 \end{figure}
-
-%\footnotetext{Figure .}
-
-% "Let Q denote the quantity of scientific interest (e.g., a regression coefficient)." \cite[p.~4]{mice}.
-% "Let Y denote the n x p matrix containing the data values on p variables for all n units in the sample.
-We define the response indicator R as an n x p 0-1 matrix.
-The elements of Y and R are denoted by y i j and r i j , respectively, where i = 1 , ...
-, n and j = 1 , ...
-, p .
-If y i j is observed, then r i j = 1 , and if y i j is missing, then r i j = 0 ." ...
-The observed data are collectively denoted by Y o b s .
-The missing data are collectively denoted as Y m i s , and contain all elements y i j where r i j = 0 .
-When taken together Y = ( Y o b s , Y m i s ) contain the hypothetically complete data." \cite[par.~2.2]{buur18}
-% The data are said to be MCAR if Pr ( R = 0 | Y o b s , Y m i s , $\psi$ ) = Pr ( R = 0 | $\psi$ ) (2.1) so the probability of being missing depends only on some parameters $\psi$ , the overall probability of being missing.
-The data are said to be MAR if Pr ( R = 0 | Y o b s , Y m i s , $\psi$ ) = Pr ( R = 0 | Y o b s , $\psi$ ) (2.2) so the missingness probability may depend on observed information, including any design factors.
-Finally, the data are MNAR if Pr ( R = 0 | Y o b s , Y m i s , $\psi$ ) (2.3) does not simplify, so here the probability to be missing also depends on unobserved information, including Y m i s itself.
-\cite[par.~2.2]{buur18}
-
-% Also introduce: 
-% 
-% - Terminology (MCAR, MAR, MNAR)?
-% 
-% - Blue points are observed, the red points are imputed?
 
@@
 The diagnostic evaluation of convergence is preferred over the current practice---visually inspecting imputation chains for signs of non-convergence---since the latter may be challenging to the untrained eye \citep[\S~6.5.2]{buur18}.
-%Diagnostic evaluation of algorithmic convergence does not share thisApplied researchers should be facilitated in evaluatingClear guidelines for the diagnostic evaluation of convergence would be preferred.
 
-%In that respect is the current evaluation practice insufficient.
-That is, the  recommendation is to visually inspect imputation chains for signs of non-convergence, which may be challenging to the untrained eye \citep[\S~6.5.2]{buur18}.
-Quantitative evaluation of the convergence of MI algorithms would be preferred.
 The application of convergence diagnostics to MI methods has not been systematically studied \citep{buur18}.
@@
 Non-convergence may be present as non-stationarity within chains (i.e., trending), or as slow mixing between chains (i.e., no intermingling).
-The mixing component of convergence may be evaluated with the potential scale reduction factor, $\widehat{R}$ \citep[a.k.a.
-`Gelman-Rubin statistic';][]{gelm92}\footnotemark.
-The stationarity component may be assessed through autocorrelations.
-Other convergence diagnostics are outside of the scope of this study.
-%(Several convergence diagnostics have been proposed for evaluating convergence of iterative algorithmic procedures
-
-\footnotetext{An adapted version of $\widehat{R}$ might also be used to diagnose non-stationarity.}
-
-% The multiple imputation algorithm in \pkg{mice} is a special case of Markov chain Monte Carlo (MCMC) methods---a framework of iterative algorithmic procedures.
-Within this framework, several convergence diagnostics have been proposed, but the application of these measures on MI algorithms has not been systematically studied \citep{buur18}.
-% % For the framework of itrIt is not known whether convergence diagnostics that are available for the general framework that MICE is a special case of are appropriate for the evaluation of MI data.
-This is the framework of Markoc chain Monte Carlo %Yet, reference manuals of promiment MI software do mention the importance of evaluating algorithmic convergence (e.g., \cite[p.XYZ]{mice}, \textbf{XYZ, XYZ}).
-% %%%%%%%
-% %This section provides a brief review of: \textbf{[INMIDDELS INCORRECT]}% briefly reviews?
-% %1) how convergence is defined in the general context of iterative algorithmic procedures; 2) what diagnostic tools are available; and 3) which of these may apply to MI algorithms. 
-% 
-% Loosely speaking, MCMC algorithms aim to sample values from an unknown target distribution (e.g., the posterior distribution of a parameter in Gibbs samplers, or the distribution of missing values in MI algorithms).
-Convergence is reached once the algorithm samples exclusively from the target distribution \citep{cowl96}.
-Since this target distribution is unknown by definition, convergence can only be monitored by evaluating signs of non-convergence \citep{hoff09}.
-In practice, the definition of convergence consists of two components: `mixing' of chains together, and `stationarity' of individual chains \citep[p.~284]{gelm13}.
-The mixing component is satisfied when chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm.
-Stationarity refers to the absence of trending across iterations within chains.
-%\textbf{Figure XYZ shows a pathological case of non-convergence concerning both components.}
-% %The aim of MCMC procedures is to sample from an unknown (theoretical) distribution.
-The stability of the solution denpends on two components of convergence: should have converged, two components should be satisfied.
-% 
-% % Assessing convergence is difficult.
-We can never be certain of convergence, therefore convergence diagnostics evaluate signs of non-convergence \citep{hoff09}.
-% In general: "In general, you cannot know for sure if your chain has converged.
-But sometimes you can know if your chain has not converged, so we at least check for this latter possibility" \cite[p.~101]{hoff09}
-% % More importantly, in MCMC algorithms like MICE, "what is produced by the algorithm at convergence is not a single number or even a distribution, but rather a sample from a distribution" \citep[p.~883]{cowl96}.
-"automated convergence monitoring (as by a machine) is unsafe and should be avoided" \citep[p.~902]{cowl96}.
-% %Convergence has two components: `mixing' and `stationarity' \citep[p.~284]{gelm13}.
-Mixing refers to the variation between chains of samples.
-Stationarity has to do with stability of samples within chains.
-Ideally, both components will be satisfied: there are no trends within chains, and the chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm.
-% 
-% \textbf{[COULD BE SHORTER!
-Maybe in footnote??]} Most convergence diagnostics target either of the two components.
+% Most convergence diagnostics target either of the two components.
 The stationarity component of convergence may be evaluated with autocorrelation \citep[$AC$; ][]{scha97, gelm13}, numeric standard error \citep[or `MC error'; ][]{gewe92}, and Raftery and Lewis's (\citeyear{raft91}) procedure to determine the effect of trending within chains.
-The mixing component can be assessed with the potential scale reduction factor $\widehat{R}$ \citep[a.k.a.
+%\footnote{All of these methods evaluate the convergence of univariate scalar summaries (e.g., chain means or variances).
+These convergence diagnostics cannot diagnose convergence of multivariable statistics (i.e., relations between scalar summaries).
+\cite{buur18} proposed to implement multivariable evaluation through eigenvalue decomposition \cite{mack03}.
+This method is outside of the scope of the current study.}%  he eigenvector decomposition method proposed by McKay \citep{mcka03}.}
+% The mixing component can be assessed with the potential scale reduction factor $\widehat{R}$ \citep[a.k.a.
 `Gelman-Rubin statistic';][]{gelm92}.
@@
 Therefore, this study employs both $\widehat{R}$ and autocorrelation to investigate convergence, as recommended by \cite[p.~898]{cowl96}.
-% %Therefore, the focus of this study is mainly on $\widehat{R}$. Stationarity will also be evaluated with autocorrelation as diagnostic, as recommended by \cite[p.~898]{cowl96}. 
-% Other convergence diagnostics are outside the scope of this study.\footnotemark 
-% \footnotetext{All of these methods evaluate the convergence of univariate scalar summaries (e.g., chain means or variances).
-These convergence diagnostics cannot diagnose convergence of multivariable statistics (i.e., relations between scalar summaries).
-\cite{buur18} proposed to implement multivariable evaluation through eigenvalue decomposition \cite{mack03}.
-This method is outside of the scope of the current study.}%  he eigenvector decomposition method proposed by McKay \citep{XYZ}.} These four methods, however, are outside of the scope of this study.
-Here, the focus is mainly on $\widehat{R}$ and autocorrelation.
-% 
-% %Diagnose non-convergence in the mixing sense: $\widehat{R}$.
-Diagnose non-convergence in the stability sense: autocorrelation (AC).
-Also, it is possible to compute MC errors, the Geweke statistic \textbf{\citep{XYZ}}, or to determine the appropriate number of iterations for convergence by employing \textbf{Raftery and Lewis's (1992)} method.
-All of these methods evaluate the convergence of univariate scalars of interest (e.g., chain mean or chain variance).
-The convergence of multivariable statistics (i.e.
-relations between such scalars) may be evaluated via the eigenvector decomposition method proposed by McKay \citep{XYZ}.
-These four methods, however, are outside of the scope of this study.
-Here, the focus is mainly on $\widehat{R}$ and autocorrelation.}
-% 
-% % "In order to converge to a stationary distribution, a Markov chain needs to satisfy three important properties (Roberts 1996; Tierney 1996): irreducible, the chain must be able to reach all interesting parts of the state space; aperiodic, the chain should not oscillate between different states; recurrence, all interesting parts can be reached infinitely often, at least from almost all starting points.
-Do these properties hold for the MICE algorithm?
-Irreducibility is generally not a problem since the user has large control over the interesting parts of the state space.
-This flexibility is actually the main rationale for FCS instead of a joint model.
-Periodicity is a potential problem, and can arise in the situation where imputation models are clearly inconsistent.
-A rather artificial example of an oscillatory behavior occurs when Y 1 is imputed by Y 2 $\beta$ + $\epsilon$  1 and Y 2 is imputed by - Y 1 $\beta$ + $\epsilon$  2 for some fixed, nonzero $\beta$ .
-The sampler will oscillate between two qualitatively different states, so the correlation between Y 1 and Y 2 after imputing Y 1 will differ from that after imputing Y 2 .
-In general, we would like the statistical inferences to be independent of the stopping point.
-A way to diagnose the ping-pong problem, or order effect, is to stop the chain at different points.
-The stopping point should not affect the statistical inferences.
-The addition of noise to create imputations is a safeguard against periodicity, and allows the sampler to "break out" more easily.
-Non-recurrence may also be a potential difficulty, manifesting itself as explosive or non-stationary behavior.
-For example, if imputations are made by deterministic functions, the Markov chain may lock up.
-Such cases can sometimes be diagnosed from the trace lines of the sampler.
-See Section 6.5.2 for an example.
-As long as the parameters of imputation models are estimated from the data, non-recurrence is mild or absent.
-The required properties of the MCMC method can be translated into conditions on the eigenvalues of the matrix of transition probabilities (MacKay 2003, 372-73).
-The development of practical tools that put these conditions to work for multiple imputation is still an ongoing research problem." \cite[par.~4.5]{buur18}
-%
+The mixing component of convergence may be evaluated with the potential scale reduction factor, $\widehat{R}$ \citep[a.k.a.
+`Gelman-Rubin statistic';][]{gelm92}\footnotemark.
+The stationarity component may be assessed through autocorrelations.
+Other convergence diagnostics are outside of the scope of this study.
 
-% The convergence diagnostics under consideration are $\widehat{R}$ and autocorrelation.
-Perfect convergence is reached when the variance between imputation chains is equivalent to the variance within chains ($\widehat{R} = 1$), and there is no dependency between subsequent iterations of imputation chains ($AC = 0$).
-%
-%%%%%%%%%%%%%%%%%
-%As potential scale reduction factor, $\widehat{R}$ tells us by how much the variance of an estimate could be shrunken down if the chains were to be infinitely long \citep[$t \to \infty$;][]{gelm92}.
-If there is no added value compared to infinitely many iterations, the chains have converged.
-%%%%%%%%%%%%%%%%%%
-%
-%$\textbf{Explain the measures that I'm investigating here. R hat is ... autocorrelation is... How are they computed? What do they say? How are they computed on MI data? Looking at imputation chains. 
-%Explain that there is no baseline to compare the measures with. The only current convergence check is to visually inspect trace plots. (But in the trace-plots convergence seems immediate).}
-%
-%The potential scale reduction factor $\widehat{R}$ tells us by how much the variance of an estimate could be shrunken down if the chains were to be infinitely long \cite{gelm92}.
-That then informs us about how dependent the chains are on the starting values.
-If there is no dependence on the initial values anymore, the chains have converged (in the mixing sense of the word).
-%$\widehat{R}$ is then equal to one. 
-%
-%We can apply $\widehat{R}$ to the mean (or to the first two moments) of the variables of interest.
+\footnotetext{An adapted version of $\widehat{R}$ might also be used to diagnose non-stationarity.}
 
 To define $\widehat{R}$, we follow notation by \cite[p.~5]{veht19}.
-% The scalar summary of interest in this study is the average of the imputed values per variable.
-%As recommended by \cite{veht19} $\widehat{R}$ is computed as ....
-to be able to detect ...
-in the tails of the distribution.
-\cite[p.~5]{veth19}
 Let $M$ be the total number of chains, $T$ the number of iterations per chain, and $\theta$ the scalar summary of interest (e.g., chain mean or chain variance).
 For each chain ($m = 1, 2, \dots, M$), we estimate the variance of $\theta$, and average these to obtain within-chain variance $W$.
-
-%\textbf{REPHRASE: In the equations below, $N$ is the number of draws per chain, $M$ is the number of chains, and $S = MN$ is the total number of draws from all chains.
-For each scalar summary of interest $\theta$, we compute $B$ and $W$, the between- and within-chain variances:}
 
@@
 
-%The proportion of within chain variance $W$ against the weighted average of $B$ and $W$, $\widehat{\operatorname{var}}^{+}$ tells us how much variance of $\theta$ could be shrunken down if $N\to\infty$.
-$\widehat{\operatorname{var}}^{+}$ is computed as follows: %over-estimates the marginal posterior variance of the estimand, $\theta$.
-
-% \begin{equation*}
-% 
-% \end{equation*}
-
-%The potential scale reduction factor $\widehat{R}$ is obtained as:
 We can interpret $\widehat{R}$ as potential scale reduction factor since it indicates by how much the variance of $\theta$ could be shrunken down if an infinite number of iterations per chain would be run \citep{gelm92}.
@@
 High $\widehat{R}$-values thus indicate non-convergence.
-%The iteration at which chains are independent of their initial values is the point where mixing occurs.
-This is diagnosed with $\widehat{R}$-values close to one.
-%If there is no added value
-%High $\widehat{R}$-values thus indicate slow mixing of chains, which is a sign of non-convergence.
-%, e.g., because chains may depend on their initial value.
-%are of dependence across chains on initial values.
-Once there is no dependence on initial values anymore, there is no added value of continuing iterations.
-\textbf{[MOGELIJK OVERBODIGE ZIN HIER]} The measure then diagnoses algorithmic convergence ($\widehat{R} = 1$).
-%If there is no added value compared to infinitely many iterations, the chains have converged. 
-The conventionally acceptable threshold for convergence was $\widehat{R} < 1.2$ \citep{gelm92}. More recently, \cite{veht19} proposed a more stringent threshold of $\widehat{R} < 1.01$.
+The conventionally acceptable threshold for convergence was $\widehat{R} < 1.2$ \citep{gelm92}.
+More recently, \cite{veht19} proposed a more stringent threshold of $\widehat{R} < 1.01$.
 
@@
 
-We can interpret %the sign and magnitude of ... Any non-zero $AC$-value ($-1 \leq AC \leq 1$) . $AC = 0$ indicates independence between $\theta$-values.
-$AC$-values as a measure of stationarity.
+We can interpret $AC$-values as a measure of stationarity.
 If $AC$-values are close to zero, there is no dependence between subsequent samples within imputation chains.
@@
 Negative $AC$-values show no threat to the stationarity component of convergence.
-% (i.e., similarity between $\theta$-values of subsequent iterations), which may lead to trending.
-Negative $AC$-values occur when the $\theta$-values of subsequent iterations diverge from one-another.
-Dependence in this sense does not threaten the stationarity component of convergence.
-%that we aim to evaluate with $AC$.
 On the contrary even---negative $AC$-values indicate that $\theta$-values of subsequent iterations diverge from one-another, which may increase the variance of $\theta$ and speed up convergence.
-%have also indicate dependence, but do not threaten the stationarity component of convergence \textbf{[REMOVE INTERPRETATION HERE?]}.
-Instead of recurrence, there would be a `seesaw' effect within chains: subsequent $\theta$-values would deviate in opposite directions from the chain average $\theta^{(\cdot m)})$, increasing the variance of $\theta$.
-As convergence diagnostic, the interest is therefore in positive $AC$-values.%\footnotemark 
+As convergence diagnostic, the interest is therefore in positive $AC$-values.
 \footnote{Moreover, the magnitude of $AC$-values may be evaluated statistically, but that is outside of this note's scope.}
@@
 
-%, where $\theta$ is a scalar summary of interest (i.e., chain means and chain variances).
-%Non-recurrence can be evaluated with autocorrelation.
-autocorrelation shows how dependent subsequent draws of an imputation chain are on the previous value.
-If there is a lot of dependence, draws at e.g.
-iteration five are significantly correlated with the value of the first draw.
-%A high autocorrelation indicates dependence within chains.
-The magnitude of the AC can be interpreted qualitatively or quantitatively.
-Quantitative evaluation of the AC entails comparing the observed ACs to the critical values of a two-tailed 95\% confidence interval, divided by the square root of the number of iterations.
-%.
-critical value corresponding to the $z-$~value standard normal distribution with \textbf{XYZ} degrees of freedom.
-%  AC_lim <- qnorm(.975) / sqrt(maxit)
-
-%\textbf{HIER VALT TE SCHRAPPEN!} It is not known whether $\widehat{R}$ and $AC$ are appropriate measures to diagnose convergence of multiple imputation algorithms.
-
-% the conventional threshold $\widehat{R} < 1.1$ might be too stringent. \cite[pp.~49-50]{lace07} report
-% 
-% $\widehat{R}$ is sensitive enough to indicate non-convergence can  reported $\widehat{R}$-values  
-% 
-% An empirical finding, however, shows that the opposite may be true: \cite{lace07} report $\widehat{R}$-values above the threshold of $\widehat{R} < 1.1$ after fifty iterations.
-%that $\widehat{R}$ will not be smaller than 1.1 before iteration number 50.
-Therefore, the aim is to replicate empirical finding Lacerda et al.
-%
-
-%Conventional thresholds to diagnose non-convergence-- e.g., Gelman and Rubin's \citeyear{gelm92} statistic $\widehat{R} < 1.1$ --are not applicable on multiply imputed data \citep{lace07} \textbf{[Expand or remove Rhat here!
-Refer to future section?]}.
-Therefore, empirical researchers have to rely on visual inspection procedures that are theoretically equivalent to $\widehat{R}$ \citep{whit11} \textbf{[explain why this can still be appropriate, and numerical is not]}.
-Visually assessing convergence is not only challenging to the untrained eye, it might also be futile.
-
-% This research project consists of an investigation into algorithmic convergence of MI algorithms.
-I will replicate Lacerda et al.'s simulation study on $\widehat{R}$ \citep{lace07}, and develop novel guidelines for assessing convergence.
-Ideally, I will integrate several diagnostics (e.g., $\widehat{R}$, \emph{autocorrelation}, and \emph{simulation error}) into a single summary indicator to flag non-convergence.
-
+%% -- Hypothesis -------------------------------------------------------------
 \subsection{Simulation Hypothesis} \label{sec:hypothesis}
@@
 This study evaluates whether $\widehat{R}$ and $AC$ could diagnose convergence of multiple imputation algorithms.
-%There is, however, no baseline measure available to evaluate their performance against.
-%\footnote{The current practice of visually inspecting imputation chains for signs of non-mixing or non-stationarity can only diagnose severely pathological cases of non-convergence.} 
 We assess the performance of the two convergence diagnostics against the recommended evaluation criteria for MI methods \citep[i.e., average bias, average confidence interval width, and empirical coverage rate across simulations;][\S~2.5.2]{buur18}.
-%The performance diagnostics comprise of .% %"We expect that in accordance with Neyman (1934), at least 95 percent of the confidence intervals should contain the true population value, although some room for simulation error should be taken into account to counteract the finite nature of our simulations"\cite{schou18}.
+%That is, there is no baseline measure available to evaluate performance against.The current practice of visually inspecting imputation chains for signs of non-mixing or non-stationarity can only diagnose severely pathological cases of non-convergence.
 %\footnote{We could also look at distributional characteristics, and plausibility of imputed values, see \cite{vinknd} (n.d.).
-For now, this is outside of the scope of this study.} % Or use: Then evaluate the regular diagnostics of MI simulations, see Vink, n.d., and  convergence.
-Diagnostics include absolute bias of the estimated regression coefficient, confidence interval width, and empirical coverage rate across simulations.
-Convergence is evaluated with $\widehat{R}$, or potential scale reduction factor.
-
-%\textbf{[HIER NOG ZOOI]} We hypothesize that $\widehat{R}$ may not be an appropriate diagnostic, because it assumes over-dispersed initial values.
-In \pkg{mice}, initial values of the algorithm are chosen randomly from the observed data.
-Therefore, we cannot be certain that the initial values are over-dispersed.
-%The initial parameter value is noodzakelijk. Daarom is het lastig om in MI that starten vanuit overdisp state. 
-% Starting values of MICE chains: "For each j, fill in starting imputations $Y^0_j$ by random draws from $Y^{obs}_j$." \cite[\S~4.5.2]{buur18} before for loops over variables and imputations, see \url{https://stefvanbuuren.name/fimd/sec-FCS.html#def:mice}
-%Without over-dispersed initial states, $\widehat{R}$ may falsely diagnose convergence \citep{broo98}.
-%would not be sensitive enough to flag non-convergence of MI algorithms: "if over-dispersion does not hold, $\sigma_+^2$ can be too low, which can lead to falsely diagnosing convergence." \cite[p~437]{broo98}.
-%This suggests that $\widehat{R}$ would not be sensitive enough to flag non-convergence of MI algorithms.
-However, an empirical finding by \cite{lace07} shows that $\widehat{R}$ can indicate non-convergence at unconventionally long chain lengths.
-While the default number of iterations in \pkg{mice} is five, $\widehat{R} > 1.1$ was observed after 50 iterations, and $\widehat{R} > 1.01$ up-to 150 iterations.
-This suggests that $\widehat{R}$ would over-estimate non-convergence of MI algorithms.
-
+For now, this is outside of the scope of this study.}
 Based on an empirical finding \citep{lace07}, we hypothesize that $\widehat{R}$ will over-estimate non-convergence of MI algorithms.
@@
 No hypothesis was formulated about the performance of $AC$ as convergence diagnostic.
-%
-
-%Convergence diagnostics for MI methods have not been systematically studied \citep{buur18}.  Moreover, 
-%A fundamental assumption of MICE is convergence of the algorithm.
-MICE is a type of Markov chain Monte Carlo (MCMC) algorithm.
-In general, the validity of inference resulting from MCMC algorithms is threatened by non-convergence.
-Hence we use convergence diagnostics to flag non-convergence.
-But it is not known whether conventional convergence diagnostics for MCMC methods work on MI data.
-%the need for convergence diagnostics is extra pressing in the context of the MICE algorithm.
-
-%The primary research interest is in determining whether $\widehat{R}$ is an appropriate convergence diagnostic, and if so, which level of stringency suits MI data.
-\textbf{Or if there is a discrepancy between the visual and numeric inspection.
-There is no baseline to compare the convergence diagnostics with.
-Therefore, we will only evaluate performance measures.}
-
-%This simulation study hypothesizes that $\widehat{R}$ will over-estimate non-convergence.
-%\textbf{[Or use:] Hypothesis based on \cite{lace07} is that the conventional acceptable level of $\widehat{R}$ is too strict for MI data.
-We expect that the simulation diagnostics will indicate valid inference before $\widehat{R}$ will.}
-
-% Also introduce:
-% 
-% - missingness mechanims, ignorability?
-% % "The practical importance of the distinction between MCAR, MAR and MNAR is that it clarifies the conditions under which we can accurately estimate the scientifically interesting parameters without the need to know $\psi$" \cite[par.~2.2]{buur18}.
-% 
-% - Rubin's rules?
-% 
-% - FCS vs. JM?
-%
-
-% \subsubsection{What is already implemented?}
-% 
-% - Trace-plots
-% 
-% \subsubsection{What is not yet implemented, but exists?}
-% 
-% - $\widehat{R}$, but too stringent (new) threshold, and assumption of over-dispersed initial values of imputation chains not met.
-% % Potentially add something about updated (2019) version of $\widehat{R}$ and the new threshold of 1.01, see \citep{veht19}.
-% 
-% - autocorrelation.
-Schafer (1997, p.
-129) wrote on worst linear statistic.
-We could calculate the autocorrelation of that statistic to know that the algorithm converged elsewhere too.
-See autocorr function plot in SAS of worst linear function.
-% % Worst linear function in SAS see \url{https://support.sas.com/documentation/cdl/en/statug/63033/HTML/default/viewer.htm#statug_mi_sect027.htm}.
-% % Note: we're talking about missing data only, not the combined data (that autocorrelation is very high, as is the autocorrelation of deductively imputed values, like in the texp example). 
-% % "Applications of MICE with lowly correlated data therefore inject a lot of noise into the system.
-Hence, the autocorrelation over t will be low, and convergence will be rapid, and in fact immediate if all variables are independent.
-Thus, the incorporation of noise into the imputed data has pleasant side-effect of speeding up convergence" \citep{buur18}, par.
-4.5.
-% 
-% - Sensitivity analysis: Run algorithm several times and compare results. 
-% % "Imputers who do choose to use FCS should use flexible univariate models wherever possible and take care to assess apparent convergence of the algorithm, for example by computing traces of pooled estimates or other statistics and using standard MCMC diagnostics (Gelman et al., 2013, Chapter 11).
-It may also be helpful to examine the results of many independent runs of the algorithm with different initializations and to use random scans over the p variables to try to identify any convergence issues and mitigate possible order dependence" \cite[p.~19]{murr18}.
-% 
-% \subsubsection{What is not implemented, and NA?}
-% 
-% - $\widehat{R}$ threshold: Replicate simulation study and build a decision rule to solve the problem with $\widehat{R}$.
-% % "The monitoring statistic was computed for mean monthly earnings at each iteration in chains of length k = 200.
-Since calculation of the statistic requires M parallel sequences, m = 5 such chains were constructed.
-This value of m was informed by the preferred choice given in the literature on multiple imputation.
-The monitoring statistic computed at each iteration is presented in Figures 15 to 17 for each of the three missingness mechanisms.
-The red vertical line denotes ten iterations" \cite[p.~49]{lace07}.
-% 
-% - Stability of the solution: Possibly use the slope of means over iterations too to see whether there is trending.
-Or apply PCA on the imputed data and if that (the eigenvalues?) stays the same we know that the means and variances are stable as well, see McKay (?).
-% 
-% - MC error: MC error = SD/sqrt(number of iterations), where SD represents the variation across iterations.
-The MC error thus represents how much the means differ w.r.t.
-the iterations.
-MC error decreases as number of iterations increases.
-It should not be larger than 5\% of the sample standard deviation.
 
 %% -- Methods ---------------------------------------------------------------
-
 %%
@@
 
-%To investigate the convergence properties of the multiple imputation algorithm in \pkg{mice}, w
 We use model-based simulation in \proglang{R} to perform multiple imputation under 100 simulation conditions.
@@
 These diagnostics are aggregated across repetitions to obtain their average value per simulation condition.
-%We compute $\widehat{R}$ and autocorrelation (`convergence diagnostics') and performance measures (`simulation diagnostics') for each condition.
-%All simulation conditions are replicated 1000 times.
-%For each of the 100 conditions 1000 simulation runs are performed.
-for each of the 100 simulation conditions.
-\textbf{Estimands are convergence diagnostics.
-Performance measures are simulation diagnostics.
-where for one through one hundred the Rhat and ...
-are calculated.}
 The simulation set-up is summarized in the pseudo-code below.\footnote{The complete \proglang{R} script of the simulation study is available from \href{https://github.com/gerkovink/shinyMice/simulation}{github.com/gerkovink/ShinyMICE/simulation}.}
@@
 
+%% -- Data -------------------------------------------------------------
 \subsection{Data simulation}
+
 The point of origin for all simulation conditions is a finite population of $N=1000$.
@@
 Data is simulated with the function \fct{mvtnorm::rmvnorm}, and outcome variable $Y$ is subsequently calculated as 
-
 $$Y =  2X + .5Z_1 - Z_2 + \epsilon .$$
 
+%% -- Ampute -------------------------------------------------------------
 \subsection{Amputation}
+
 The complete data is `amputed' once for each simulation repetition.
@@
 %The resulting amputed data is equal for all simulation conditions in the same repetition.
-%"Furthermore, we generate three kinds of missingness proportions: 0.1, 0.5, and 0.9.
-Note that these values indicate the sampled proportion of incomplete cases in Y" \cite{schou18}.
 
+%% -- Impute -------------------------------------------------------------
 \subsection{Imputation}
+
 Missing datapoints are imputed with the function \fct{mice::mice}.
@@
 The number of iterations varies between simulation conditions (\texttt{maxit = 1, 2, \dots, 100}).
-% Missingness mechanism: "With MCAR missingness mechanisms, the probability to be missing is the same for all cases.
-This is a necessary simulation condition for evaluating the performance of imputation procedures.
-If an imputation method is not able to solve the problem (i.e.
-yield valid inference) under MCAR, the statistical properties of the procedure are not sound." (Vink, n.d., p.
-4)
-% "combine the m completed data inferences into a single inference following Rubin's (1987) rules (pp. 76-77)" \cite{schou18}.
 
+%% -- Convergence -------------------------------------------------------------
 \subsection{Convergence Diagnostics}
-% Add ``post-processing steps (e.g.
-aggregation computations, summary statistics, regressions on the simulation results) used to transform simulation outputs to reported results''.
-%, see \url{http://jsterman.scripts.mit.edu/docs/Rahmandad-Sterman%20Simulation%20Reporting%20Standards3.30%20update.pdf}.
+
 From the imputation step, we extract chain means and chain variances across iterations ($\theta$ at $t = 1,2,\dots,T$).  
-%not only the imputed values of each MI procedure (where $t = T$), but also information from the intermediate iterations ($t = 1, 2, \dots, T$).
-The quantities of interest ($\theta$-values) are chain means and chain variances.
 We compute convergence diagnostics separately for both $\theta$s, and for each of the four variables in the dataset.
@@
 $AC$ is computed with function \fct{stats::acf}.
-%%%%%%%%%%%%%%%%
-%From each imputation chain, we extract imputed values, chain means and chain variances ($\theta$-values).
-$\theta$-values are information from the intermediate iterations ($t = 1, 2, \dots, T$).
-The quantities of interest ($\theta$-values) are .
-We compute convergence diagnostics separately for each of the four variables in the dataset.
-\textbf{We report the maximum (absolute) value as the `worst linear predictor' of convergence [Rephrase or footnote??]}.
-$\widehat{R}$ is computed by implementing Vehtari et al.'s \citeyear{veht19} recommendations.
-$AC$ is computed with \pkg{stats} function \fct{acf}.
-%%%%%%%%%%%%%
-% 
-% 
-% extract chain means and chain variances per iteration $t$. From the MI algorithm Convergence diagnostics are computed for chain means and chain variances.   
-% $\widehat{R}$ is computed within imputation chains: . The maximum value across variables within the same simulation is reported. 
-% 
-% autocorrelation (AC) is computed within imputation chains,  The AC of the variable with the highest absolute AC value is reported per simulation.
 
+%% -- Analyze -------------------------------------------------------------
 \subsection{Analysis}
@@
 
+%% -- Performance -------------------------------------------------------------
 \subsection{Simulation diagnostics}
-% Distributional characteristics: In practice, the distribution of the incomplete data may differ greatly from the observed data.
-Under anything but the MCAR assumption, this can be expected.
-When evaluating imputations, the distributional shapes should be checked and diagnostic evaluations should be performed (see Abayomi et al., 2008, for an detailed overview of diagnostic evaluation for multivariate imputations).
-When anomalies are found, and if the imputation method is valid, there should be an explanation, especially in the controlled environment of a properly executed simulation study.
-% Plausibility of the imputed values: Plausible imputations - imputations that could be real values if they had been observed - are not a necessary condition for obtaining valid inference.
-However, in practice, especially when the imputer and the analyst are different persons, plausibility of imputations may be a desired property.
-When evaluating imputation routines, the evaluator should mention whether the routine is prone to deliver implausible value
 
@@
 %% -- Results ---------------------------------------------------------------
-
 %%
@@
 <<echo = FALSE, eval = TRUE>>=
-load("C:/Users/User/Desktop/shinyMice/Simulation/Results/results14.Rdata.Rdata")
+load("../Results/results.Rdata")
 names(dat) <- c("It.", "Bias", "CI width", "Cov. rate", "R chain mean", "R chain var", "AC chain mean", "AC chain var")
@@
 
+% % See "3.CreateFigure.R", I specifically chose to NOT generate the figures within this Rnw file because of journal guidelines.
 \begin{figure}[h]
@@
 
+% % With SEs
 % \begin{figure}[h]
@@
 
-%% --------------------------------
+%% -- Convergence -------------------------------------------------------------
 \subsection{Convergence diagnostics}
+
 It is apparent that there is a relation between the number of iterations per simulation condition ($T$) and the convergence diagnostics.
@@
 We, therefore, only discuss the $\widehat{R}$ and $AC$-values of chain means.
-%convergence diagnostics for the chain means.
-%do not differentiate in the interpretation of chain means .
 
@@
 If we use the recently recommended threshold $\widehat{R} < 1.01$, we would conclude that convergence is reached in none of the simulation conditions.
-%\textbf{[Add something about $\widehat{R} <1$?]}
-
-%The $\widehat{R}$-values for chain means and chain variances show equivalent trends, neither is there a clear difference between the four variables.
-Roughly speaking, we see that $\widehat{R}$ is higher for simulation conditions with a lower number of iterations, and approaches one for conditions with more iterations.
-
-%%%%
-%Figure \ref{fig:conv} shows an initial dip, a small bump upwards, and then a steep decline.
-This decline continues up-to iteration twenty, after which a more gradual decrease can be observed up-to iteration forty.
-The $\widehat{R}$-values are more or less stable after that.
-%\textbf{[Remove stuff here]} %The conventionally acceptable threshold for convergence, maximum 
-%$\widehat{R} < 1.2$ is reached at $T =3$, and conditions with six iterations or more.
-The widely used threshold $\widehat{R} < 1.1$ is met in conditions with ten or more iterations.
-The most recent recommended threshold is not reached within the 100 iterations considered in this simulation study.
-$\widehat{R} > 1.01$ for all simulation conditions (see online appendix for the full table of results).
-%%%%
 
@@
 $AC$-values in conditions where $T>70$ are indifferentiable from zero, indicating stationarity.
-%Simulation conditions where $T>5$ have $AC$-values greater than $T=2$.
 %We see an initial decrease between $T=2$ and $T=3$.
+Simulation conditions where $T>5$ have $AC$-values greater than $T=2$.
 According to this diagnostic, none of the simulation conditions show signs of non-convergence, since we only observe negative or zero $AC$-values.
@@
 This would imply $T=6$ as the minimal number of iterations.
-% All observed autocorrelations  are negative or close to zero.
-There are some differences between variables, and the ACs of chain variances seem lower than the ACs of chain means.
-The general trend in figure \ref{fig:conv} is that ACs decrease steeply as the number of iterations increases.
-\textbf{The lowest AC observed is %\Sexpr{round(min(dat[-1, c("AC chain mean", "AC chain var")]), 3)}
-% at four iterations.
-Simulation conditions with a higher number of iterations show a slow increase towards zero.
-autocorrelations do not plateau completely within the range of simulation conditions considered in this study.
-It seems that ACs might become positive if more iterations would be allowed.
-The average absolute maximum ACs are negative up to 77 iterations for chain means, and for chain variances all iterations considered here have negative ACs.}
-
-% \textbf{[Add simulation time (computational cost) to table?
-Add empirical SE of all diagnostics?
-Sample some random simulations and check distribution and plausibility of imputations?
-Use chain variance as scalar of interest for rhat and autocorrelation?]}
 
@@
 This threshold is, however, not sufficient, since we overlook the increase in $\widehat{R}$-values up-to conditions where $T>5$, and the convergence diagnostics only reach stability at $T>20$.
-%This implies that the current
 
-%% --------------------------------
+%% -- Performance -------------------------------------------------------------
 \subsection{Simulation diagnostics}
+
 We use average bias, average confidence interval width, and coverage rate as performance measures to evaluate $\widehat{R}$ and $AC$.
@@
 Conditions where $T>3$ can be diagnosed as unbiased, but the average bias across iterations (as shown by a flat Loess line) only reaches stability at $T=20$.
-%the average bias between the regression coefficient of $X$ on $Y$ in the complete dataset, and $\bar{Q}$, \textbf{the pooled estimated regression coefficient and the ...
-Average bias is close to zero for simulation conditions with two iterations or more.
-The simulation condition with one iteration has a negative bias.
-From two iterations upwards, the average bias across repetitions is stable.
-Across iterations two to one hundred, bias fluctuates within a narrow range around zero.
-Loess line is flat around it = 20.}
-% Average bias is not zero because there is only one incomplete dataset to be imputed. Bias is a sample effect.
-%Figure \ref{fig:sim}A shows that bias is fairly stable across simulation conditions.
-The condition $T=1$ clearly deviates from this trend with a negative bias.
-The bias for $T=2$ is below average, but within the range of fluctuations.
-Conditions where $T>3$ \textbf{are ok, Loess line is flat for $T>20$}.
 
@@
 Only conditions where $T>3$ are therefore considered sufficient.
-These conditions have similar CIWs, but across iterations stability is not established (i.e., the Loess line is never completely flat within the 100 simulation conditions).%One might argue that there is some downward trending up-to thirthy to forty iterations.
+These conditions have similar CIWs, but across iterations stability is not established (i.e., the Loess line is never completely flat within the 100 simulation conditions).
 %After about twenty-five iterations the average CIW across all simulation conditions and repetitions is reached \Sexpr{round(mean(dat[, "CI width"]), 3)}.
@@
 This suggests that as little as four iterations may be sufficient for the MI algorithm under the current circumstances.
-%  that simulation conditions where $T>3$ are generally OK.
-The performance measures indicate that for the current data, MI procedure, and quantity of interest, the algorithm should have maxit argument of four or more.
-This is, however, a simple simulation problem, with MCAR missingness mechanism, univariate missingness, and moderate fmi's [REMOVE??
-COMPUTE??].
-harder problems require more iterations.} %give some implications for these measures.
-connect convergence and simulation diagnostics.}
 
@@
 Complete algorithmic convergence as indicated by $\widehat{R}$ and autocorrelation is not reached in conditions where $T<20$.
-%is the worst possible value according to the convergence diagnostics, and $T=4$ provides marginally better convergence diagnostics.
 
@@
 This note shows that convergence diagnostics $\widehat{R}$ and $AC$ may diagnose convergence of multiple imputation algorithms, but their performance differs from conventional applications to iterative algorithmic procedures.
-%but the convergence of multiple imputation algorithms may be evaluated with convergence diagnostics .
-These measures, however, indicate non-convergence at imputation chain lengths that produce acceptable simulation diagnostics.
-This may be due to the quantity of scientific interest chosen.
-More `complicated' $Q$s (e.g., higher order effects or variance components) might show bias, under- or over-coverage at higher $T$.
 $\widehat{R}$ and autocorrelation indicate that algorithmic convergence may only be reached after twenty or even forty iterations, while unbiased, confidence valid estimates estimates may be obtained with as little as four iterations.
 These results are in agreement with the simulation hypothesis: $\widehat{R}$ over-estimates the severity of non-convergence when applied to MI procedures.
+%This may be due to the quantity of scientific interest chosen.
+More `complicated' $Q$s (e.g., higher order effects or variance components) might show bias, under- or over-coverage at higher $T$.
 
-%performance diagnostics, it appears that as little as four iterations might be sufficient to obtain ..
-
-%At least $\widehat{R}$ does not under-estimate non-convergence. It's better to be too strict than too lenient. 
 According to this simulation study, the recently proposed threshold of $\widehat{R}<1.01$ may be too stringent for MI algorithms.
@@
 Under the current specifications, $T>20$ would be more appropriate.
+% % The observed dip in AC implies that default maxit value of five iterations is the worst possible number of iterations.
+Moreover, the results of this study imply that assessing the stationarity component of convergence with $AC$ might be redundant, since high $AC$-values are implausible in MI procedures.
+That is, the randomness induced by the MI algorithm effectively mitigates the risk of dependency within chains.
+$AC$ would thus not be informative of the convergence of MI algorithms.
 
-%\textbf{Answer RQ.} How to diagnose non-convergence? It's complicated.
-%We will evaluate how convergence of the imputation algorithm can be diagnosed.
-
-% \textbf{Summary.}
-% This note illustrates that conventional convergence diagnostics behave differently on MI data than other MCMC methods like Gibbs samplers in Bayesian analyses.
-The most recent recommended threshold for $\widehat{R}$ ($\widehat{R} < 1.01$) may be too stringent for MI data.
-\textbf{However, using the 1.2 threshold may be too lenient because this was observed at three iterations and then again after six.
-If we would use 1.2, we might miss this increase after three iterations.
-The 1.1 threshold seems OK.
-Also, ten iterations are computationally not terrible.
-With increased performance and storage capacities, doubling the default is fine I guess.}
-% 
-% 
-% \textbf{Does this mean that conv is not nec to obtain correct inferences?}
-% 
-% \textbf{R hat below 1.}
-% $\widehat{R}$ could theoretically not be smaller than one, yet it occurred several times in this study (see online appendix XYZ).
+% textbf{[Add something about $\widehat{R} < 1$:]} $\widehat{R}$ could theoretically not be smaller than one, yet it occurred several times in this study.
 This can happen when the number of simulations is smaller than in `regular' MCMC processes [explain that fewer iterations is an advantage of MI, not a disadvantage compared to MCMC].
@@
 (Before convergence, we expect $\sigma^2$ to decrease with n, only increasing if the sequences explore a new area of parameter space, which would imply that the original sequences were not overdispersed for the particular scalar summary being monitored.)'' \cite[p~438]{broo98}.
+% % See \url{https://discourse.mc-stan.org/t/rhat-1-as-low-as-9-94e-01-why/9252/3}
 %
-% % \url{https://discourse.mc-stan.org/t/rhat-1-as-low-as-9-94e-01-why/9252/3}
-% 
-% \textbf{Negative ACs.} autocorrelation is dangerous when positive.
-These autocorrelations are negative.
-Still, we want the iterations to be stable and independent.
-Default maxit is five iterations now, should this be different?
-Are the 'waves' most pronounced at iteration 5?
-But why the dip??
-% 
-% Assessing the stationarity component of convergence with $AC$ might be redundant, since high $AC$-values are implausible in MI procedures.
-That is, the randomness induced by the MI algorithm effectively mitigates the risk of dependency within chains.
-$AC$ would thus not be informative of the convergence of MI algorithms.
-% 
-% 
-% The observed dip in AC implies that default maxit value of five iterations is the worst possible number of iterations.
 
@@
 %This is just a proof of concept.
-Further research is needed to investigate their performance under clear violation of convergence, e.g.
-dependency between predictors (predictors with very high correlations).
-%We would like to have convergence measures for multivariable statistics (scalars?) of interest.
-This is, however, dependent on the complete data model.
-The eigenvector decomposition method proposed by McKay (?) should be implemented.
-I could not find any resources to apply this method and it is outside the scope of this thesis to investigate how this approach could be implemented.
-\textbf{[this paragraph is too negative, explain scope in intro instead.]}
+Further research is needed to investigate their performance under clear violation of convergence, e.g. dependency between predictors (predictors with very high correlations). 
 Until then, we have only shown that the convergence diagnostics can diagnose non-convergence of MI algorithms that trend towards a converged state.
-%Also, conv diag for substantive model? wald test for AC = 0? 
+%Also for future research, implement Wernicke diagnostic, and look at developing a convergence diagnostic for substantive models, and implement a wald test for $AC = 0$.
+
 %% -- Optional special unnumbered sections -------------------------------------
-
+%%
+%%
+%%
+%%
 \section*{Computational details}
@@
 %% - With proper section titles and _not_ just "Appendix".
-
-\newpage
-
-\begin{appendix}
-
-\end{appendix}
+% 
+% \newpage
+% 
+% \begin{appendix}
+% 
+% \end{appendix}
 

### 2020-02-13 3.Thesis/inspect_incomplete_data.Rmd
developing evaluation of incomplete data before imputation (made a histogram and tests for 1 var in 1 dataset, now start expanding it).

---
title: "Investigate relations in incomplete datasets to inform imputation model"
output:
  html_document:
    df_print: paged
---

The goal is to provide plots and tests to inspect: 1) the relation between observed values of one variable and the missingness indicator of another; and 2) the relation between the observed values of two variables before imputation.

```{r echo=TRUE, message=FALSE, warning=FALSE}
# set-up environment
library(mice) # Data imputation
library(dplyr) # Data manipulation
library(magrittr)
library(broom)
library(ggplot2)
set.seed(123)
```

We will use the boys data as an example, specifically the variables age and gen. We will later apply it to all variables of this dataset, and then make it compatible for different datasets as well.

```{r}
# pre-processing
Rgen <- is.na(boys$gen)
levels(Rgen) <- c("Missing", "Observed")
data <- boys %>% as_tibble %>% cbind(Rgen)

# plot relation between observed values in age and missingness in gen
data %>%
  ggplot(aes(x = age, fill = is.na(gen))) +
  geom_histogram(aes(y = 100 * ..density..),
                 alpha = 0.5,
                 position = 'identity') +
  xlab("Age (years)") +
  ylab("Percentage observed") +
  scale_fill_manual(
    name = "Gen",
    breaks = c("FALSE", "TRUE"),
    labels = c("Observed", "Missing"),
    values = c("#006CC2", "#B61A51")
  )

# evaluate statistically
out_Rgen <- data %>% lm(age ~ is.na(gen), data = .)
out_int <- data %>% lm(age ~ 1, data = .)
anova(out_int, out_Rgen)
```

Now look at relations between the observed values of these variables. **Question:** Should I also visualize this relationship?

```{r}
out_gen <- data %>% lm(age ~ gen, data = .)
out_int_gen <- data %>% filter(!is.na(gen)) %>% lm(age ~ 1, data = .)
anova(out_int_gen, out_gen)
```

So both tests indicate that we should include age in the imputation model for gen.

### 2020-02-17 3.Thesis/inspect_incomplete_data.Rmd
used map to create density plots and tests for all missingness indicators at once.

@@
 
+# Part 1: Use 1 variable from 1 dataset
+
 ```{r echo=TRUE, message=FALSE, warning=FALSE}
@@
 library(ggplot2)
+library(purrr)
 set.seed(123)
@@
 ```{r}
-# pre-processing
-Rgen <- is.na(boys$gen)
-levels(Rgen) <- c("Missing", "Observed")
-data <- boys %>% as_tibble %>% cbind(Rgen)
+# # pre-processing
+# Rgen <- is.na(boys$gen)
+# levels(Rgen) <- c("Missing", "Observed")
+dat <- boys # %>% as_tibble # %>% cbind(Rgen)
 
 # plot relation between observed values in age and missingness in gen
-data %>%
+dat %>%
   ggplot(aes(x = age, fill = is.na(gen))) +
@@
 # evaluate statistically
-out_Rgen <- data %>% lm(age ~ is.na(gen), data = .)
-out_int <- data %>% lm(age ~ 1, data = .)
+out_Rgen <- dat %>% lm(age ~ is.na(gen), data = .)
+out_int <- dat %>% lm(age ~ 1, data = .)
 anova(out_int, out_Rgen)
 ```
+
+**To do:** Create a similar plot for categorical outcome variable.
 
@@
 ```{r}
-out_gen <- data %>% lm(age ~ gen, data = .)
-out_int_gen <- data %>% filter(!is.na(gen)) %>% lm(age ~ 1, data = .)
+out_gen <- dat %>% lm(age ~ gen, data = .)
+out_int_gen <- dat %>% filter(!is.na(gen)) %>% lm(age ~ 1, data = .)
 anova(out_int_gen, out_gen)
@@
 So both tests indicate that we should include age in the imputation model for gen.
+
+# Part 2: Use any variable from 1 dataset
+
+To apply these evaluations to all variables, we need a function. 
+```{r}
+# explore map functionalities
+test_miss <- function(.x){
+  return(is.na(.x))
+}
+map(c(1, NA, 4, 7), test_miss)
+
+# inline function
+map_df(c(1, NA, 4, 7), function(.x) {
+  return(data.frame(entry = .x, 
+                    is_NA = test_miss(.x)))
+})
+
+# shorter inline function
+map_df(c(1, NA, 4, 7), ~{
+  return(data.frame(entry = .x, 
+                    is_NA = test_miss(.x)))
+})
+
+# test map functionalities
+dat %>%  map_df(is.na)
+dat %>% map_df(~(data.frame(n_miss = sum(is.na(.x)),
+                                  class = class(.x))),
+                     .id = "variable")
+
+# apply map to our own problem
+interc_only <- dat %>% lm(age ~ 1, data = .)
+#age <- dat[,1]
+# output_for_table <- dat %>% 
+#   map(~{lm(dat$age~is.na(.x)) %>% 
+#       anova(interc_only, .)}, .id = "variable")
+dat %>% 
+  map(~{lm(age~is.na(.x), data = dat) %>% anova(interc_only, .)})
+
+# this works, but only for age as outcome. now apply to all variables.
+
+```
+
+**Question:** how do we evaluate these? Just look at p-values? Then, do we need some alpha inflation correction?
+
+Also create the density plots for all missingness indicators.
+
+```{r}
+# plot this for each missingness indicator
+dat %>%
+  map(~{ggplot(data=dat, aes(x = age, fill = is.na(.x))) +
+  geom_histogram(aes(y = 100 * ..density..),
+                 alpha = 0.5,
+                 position = 'identity') +
+  xlab("Age (years)") +
+  ylab("Percentage observed") +
+  scale_fill_manual(
+    name = names(.),
+    breaks = c("FALSE", "TRUE"),
+    labels = c("Observed", "Missing"),
+    values = c("#006CC2", "#B61A51")
+  )})
+
+# only the name of the legend does not work. maybe look into pmap?
+```
+
+# Part 3: Use any variable from any dataset

### 2020-02-18 3.Thesis/inspect_incomplete_data.Rmd
added functionalities to apply lm to all variables and missingness indicators

@@
 # plot this for each missingness indicator
-dat %>%
-  map(~{ggplot(data=dat, aes(x = age, fill = is.na(.x))) +
+vars <- names(dat)
+vars %>%
+  map(~{ggplot(data=dat, aes(x = age, fill = is.na(.data[[.x]]))) +
   geom_histogram(aes(y = 100 * ..density..),
@@
   scale_fill_manual(
-    name = names(.),
+    name = .x,
     breaks = c("FALSE", "TRUE"),
-    labels = c("Observed", "Missing"),
+    labels = c("observed", "missing"),
     values = c("#006CC2", "#B61A51")
@@
 
-# only the name of the legend does not work. maybe look into pmap?
+# now for all vars as outcome
+```
+# Part 2B: Apply to all variables as outcome
+
+This works for age now, with all missingness indicators. But not yet for other variables as outcome.
+
+```{r}
+# apply map to our own problem
+interc_only <- dat %>% lm(age ~ 1, data = .)
+vars <- names(dat)
+#dat %>% map(~class(.x)=="numeric")
+#dat[ ,map_lgl(dat, is.numeric)]
+#dat %>% map(~is.numeric(.x)) select_if(x, is.numeric)
+cont_vars <- dat %>% select_if(is.numeric) %>% names()
+#walk(cont_vars,~ walk(vars, ~ print(paste(.x, .y, sep = "-")),.y=.x))
+#test <- map(cont_vars, function(x) map(vars, function(y) c(x,y)))
+#test <- map(cont_vars, function(x) map(vars, function(y) c(dat[1,x],is.na(dat[1,y]))))
+test <- map(cont_vars, function(x) map(vars, function(y) lm(dat[,x]~is.na(dat[,y]))))
+
+# this works for the lm, but not yet for the intercep-only model
+#all_interc_only <- cont_vars %>% map(~lm(dat[,.x] ~ 1))
+# test2 <- map(cont_vars, function(x) map(vars, function(y) lm(dat[,x]~is.na(dat[,y])) %>% anova(interc_only[[x]], .)))
+# map(test, function(x) map(interc_only, function(y) anova(y,x)))
+test <- map(cont_vars, function(x) map(vars, function(y) {anova(lm(dat[,x]~1), lm(dat[,x]~is.na(dat[,y])))}))
+
 ```

### 2020-02-18 3.Thesis/inspect_incomplete_data.Rmd
created density plots for all continuous variables vs. the missingness indictors

@@
 # map(test, function(x) map(interc_only, function(y) anova(y,x)))
-test <- map(cont_vars, function(x) map(vars, function(y) {anova(lm(dat[,x]~1), lm(dat[,x]~is.na(dat[,y])))}))
+test2 <- map(cont_vars, function(x) map(vars, function(y) {anova(lm(dat[,x]~1), lm(dat[,x]~is.na(dat[,y])))}))
+
+```
+ Ok, I think this works. Now get it to work for the density plots.
+
+```{r}
+# create density plots for all continuous variables and all missingness indicators
+test <- map(cont_vars, function(x) map(vars, function(y) lm(dat[,x]~is.na(dat[,y]))))
+
+plots <- map(cont_vars, function(x) map(vars, function(z) {ggplot(data = dat, aes(x = dat[,x], fill = is.na(dat[,z]))) +
+  geom_histogram(alpha = 0.5, position = 'identity') +
+  xlab(paste0(x)) +
+  ylab("number of cases") +
+  scale_fill_manual(
+    name = z,
+    breaks = c("FALSE", "TRUE"),
+    labels = c("observed", "missing"),
+    values = c("#006CC2", "#B61A51")
+  )
+  }))
 

### 2020-02-19 3.Thesis/plot_imputation_densities.Rmd
created density plots of imputed values per case, but only for one variable and without map/rep to add iterations

---
title: "Plot density of imputations"
output: html_notebook
---

```{r echo=TRUE, message=FALSE, warning=FALSE}
# set-up environment
library(mice) # Data imputation
library(dplyr) # Data manipulation
# library(magrittr)
# library(broom)
library(ggplot2)
# library(purrr)
set.seed(123)
dat <- nhanes
mids <- list()

# impute missingness with one iteration
M <- 20
mids <- mice(dat, maxit = 1, m = M, print = FALSE)

####
# this should be repeated T times
####
# add an iteration
mids <- mice.mids(mids, maxit = 1, printFlag = FALSE)
# # traceplots
# plot(mids)

# plot density of imputed values
imputed_values <- mids[["imp"]]$bmi %>% mutate(case = row_number()) %>%  tidyr::gather("imp", "value", 1:M) %>% .[order(.$case, .$imp), ] #%>% group_by(case) %>% mutate(case_mean = mean(value)) %>% ungroup()
case_mean <- imputed_values %>% group_by(case) %>% summarise(mu = mean(value)) 
# # for one case
# imputed_values %>% filter(case==3) %>%  ggplot() +
#   geom_histogram(aes(x = value)) + 
#   geom_vline(xintercept = mean(imputed_values$value))

# # store names of cases
# missing_cases <- 1:max(imputed_values$case)#unique(imputed_values$case)
# # plot histograms
# missing_cases %>%
#   purrr::map(function(z) {imputed_values %>% filter(case==z) %>% ggplot() +
#   geom_histogram(aes(x = value)) 
#   })

# plot hist per case
imputed_values %>% ggplot(aes(x = value)) + 
  geom_histogram() +
  geom_vline(data = case_mean, mapping = aes(xintercept = mu)) +
  facet_wrap(~case)

```

### 2020-02-19 3.Thesis/plot_imputation_densities.Rmd
finalized plot with sd of imps per case of 1 var across it

@@
 set.seed(123)
-dat <- nhanes
-mids <- list()
+
+# parameters
+M <- 20
+T_max <- 100
 
 # impute missingness with one iteration
-M <- 20
-mids <- mice(dat, maxit = 1, m = M, print = FALSE)
+dat <- nhanes
 
@@
 ####
+plot <- list()
+distr <- list()
+
+for (it in 1:T_max){
+if (it == 1) {
+  mids <- mice(dat, maxit = 1, m = M, print = FALSE)
+} else {
 # add an iteration
-mids <- mice.mids(mids, maxit = 1, printFlag = FALSE)
-# # traceplots
-# plot(mids)
+mids <- mice.mids(mids, maxit = 1, printFlag = FALSE)}
 
@@
 imputed_values <- mids[["imp"]]$bmi %>% mutate(case = row_number()) %>%  tidyr::gather("imp", "value", 1:M) %>% .[order(.$case, .$imp), ] #%>% group_by(case) %>% mutate(case_mean = mean(value)) %>% ungroup()
-case_mean <- imputed_values %>% group_by(case) %>% summarise(mu = mean(value)) 
-# # for one case
-# imputed_values %>% filter(case==3) %>%  ggplot() +
-#   geom_histogram(aes(x = value)) + 
-#   geom_vline(xintercept = mean(imputed_values$value))
-
-# # store names of cases
-# missing_cases <- 1:max(imputed_values$case)#unique(imputed_values$case)
-# # plot histograms
-# missing_cases %>%
-#   purrr::map(function(z) {imputed_values %>% filter(case==z) %>% ggplot() +
-#   geom_histogram(aes(x = value)) 
-#   })
+distr[[it]] <- case_mean <- imputed_values %>% group_by(case) %>% summarise(mu = mean(value), sigma = sd(value))
 
 # plot hist per case
-imputed_values %>% ggplot(aes(x = value)) + 
+plot[[it]] <- imputed_values %>% ggplot(aes(x = value)) + 
   geom_histogram() +
@@
 
+# # plot density per case
+# imputed_values %>% ggplot(aes(x = value)) +
+#   geom_histogram(aes(y = ..density..)) +
+#   geom_density() +
+#   geom_vline(data = case_mean, mapping = aes(xintercept = mu)) +
+#   facet_wrap(~case)
+}
+
+# # this is just the between chain variance, right?
+# average_sd <- distr %>% purrr::map(~{mean(.$sigma)})
+# plot.ts(average_sd) #is this == plot(mids)?
+
+# now split by case
+purrr::map(1:9, function(x) purrr::map(distr, ~{.$sigma[.$case==x]})) %>% purrr::map(~plot.ts(.x))
+
+# or with ggplot
+sd_per_case_per_it <- purrr::map(1:9, function(x) purrr::map_dbl(distr, ~{.$sigma[.$case==x]})) %>% as.data.frame() %>% mutate(nr_it = row_number())
+
+colnames(sd_per_case_per_it) <- c(1:9, "nr_it")
+
+sd_per_case_per_it %>%  tidyr::gather("case", "sd", 1:9) %>% ggplot() + geom_line(aes(x = nr_it, y = sd)) + facet_wrap(~case)
+sd_per_case_per_it %>%  tidyr::gather("case", "sd", 1:9) %>% ggplot(aes(x = nr_it, y = sd, col = case)) + geom_line() + geom_smooth(se = F)
+
 ```

### 2020-02-20 3.Thesis/plot_imputation_densities.Rmd
re-wrote code to develop plots for all variables, not only bmi

--- 2020-02-19 3.Thesis/plot_imputation_densities.Rmd
@@
 title: "Plot density of imputations"
-output: html_notebook
+output:
+  html_document:
+    df_print: paged
 ---
@@
 library(ggplot2)
-# library(purrr)
+library(purrr)
 set.seed(123)
 dat <- nhanes
-mids <- list()
 
+# parameters
+M <- 20
+T_max <- 10
+
+```
+
+```{r}
+# for BMI only
+# objects
+plot <- list()
+distr <- list()
+
+# this should be repeated T times
+for (it in 1:T_max) {
+  if (it == 1) {
     # impute missingness with one iteration
-M <- 20
-mids <- mice(dat, maxit = 1, m = M, print = FALSE)
-
-####
-# this should be repeated T times
-####
+    mids <- mice(dat,
+                 maxit = 1,
+                 m = M,
+                 print = FALSE)
+  } else {
     # add an iteration
     mids <- mice.mids(mids, maxit = 1, printFlag = FALSE)
-# # traceplots
-# plot(mids)
+  }
 
   # plot density of imputed values
-imputed_values <- mids[["imp"]]$bmi %>% mutate(case = row_number()) %>%  tidyr::gather("imp", "value", 1:M) %>% .[order(.$case, .$imp), ] #%>% group_by(case) %>% mutate(case_mean = mean(value)) %>% ungroup()
-case_mean <- imputed_values %>% group_by(case) %>% summarise(mu = mean(value)) 
-# # for one case
-# imputed_values %>% filter(case==3) %>%  ggplot() +
-#   geom_histogram(aes(x = value)) + 
-#   geom_vline(xintercept = mean(imputed_values$value))
-
-# # store names of cases
-# missing_cases <- 1:max(imputed_values$case)#unique(imputed_values$case)
-# # plot histograms
-# missing_cases %>%
-#   purrr::map(function(z) {imputed_values %>% filter(case==z) %>% ggplot() +
-#   geom_histogram(aes(x = value)) 
-#   })
+  imputed_values <-
+    mids[["imp"]]$bmi %>% mutate(case = row_number()) %>%  tidyr::gather("imp", "value", 1:M) %>% .[order(.$case, .$imp),] #%>% group_by(case) %>% mutate(case_mean = mean(value)) %>% ungroup()
+  distr[[it]] <-
+    case_mean <-
+    imputed_values %>% group_by(case) %>% summarise(mu = mean(value), sigma = sd(value))
 
   # plot hist per case
-imputed_values %>% ggplot(aes(x = value)) + 
+  plot[[it]] <- imputed_values %>% ggplot(aes(x = value)) +
     geom_histogram() +
     geom_vline(data = case_mean, mapping = aes(xintercept = mu)) +
-  facet_wrap(~case)
+    facet_wrap( ~ case)
+  # would be nice to have these overlapping per case!
+
+  # # plot density per case
+  # imputed_values %>% ggplot(aes(x = value)) +
+  #   geom_histogram(aes(y = ..density..)) +
+  #   geom_density() +
+  #   geom_vline(data = case_mean, mapping = aes(xintercept = mu)) +
+  #   facet_wrap(~case)
+
+  ############################ make it work for all other vars
+  # vars <- map_dfr(dat, ~is.na(.x) %>% sum() > 0)
+  # imputed_values_list <- mids[["imp"]] %>% map(~{filter(.x %in% vars)})
+  #
+  # #filter(length(row.names(.x)))}) %>% map(~{mutate(case = row_number()) %>%  tidyr::gather("imp", "value", 1:M) %>% .[order(.$case, .$imp), ]})
+  ############################
+}
+
+# # this is just the between chain variance, right?
+# average_sd <- distr %>% purrr::map(~{mean(.$sigma)})
+# plot.ts(average_sd) #is this == plot(mids)?
+
+# # now split by case abd plot time series
+# purrr::map(1:9, function(x) purrr::map(distr, ~{.$sigma[.$case==x]})) %>% purrr::map(~plot.ts(.x))
+
+# or with ggplot to put them in 1 plot
+sd_per_case_per_it <-
+  purrr::map(1:9, function(x)
+    purrr::map_dbl(distr, ~ {
+      .$sigma[.$case == x]
+    })) %>% as.data.frame() %>% mutate(nr_it = row_number())
+
+colnames(sd_per_case_per_it) <- c(1:9, "nr_it")
+
+#sd_per_case_per_it %>%  tidyr::gather("case", "sd", 1:9) %>% ggplot() + geom_line(aes(x = nr_it, y = sd)) + facet_wrap(~case)
+sd_per_case_per_it %>%  tidyr::gather("case", "sd", 1:9) %>% ggplot(aes(x = nr_it, y = sd, col = case)) + geom_line() + geom_smooth(se = F)
 
 ```
+
+Now for any variable
+```{r}
+# objects
+imputed_values <- plot <- distr <- list()
+
+# which vars?
+nmis <- dat %>% map_dbl(~{is.na(.x) %>% sum()}) #%>% which(.>0) %>% names()
+#nmis <- apply(is.na(dat), 2, sum)
+vars <- names(nmis[nmis>0])
+
+# this should be repeated T times
+for (it in 1:T_max) {
+  if (it == 1) {
+    # impute missingness with one iteration
+    mids <- mice(dat,
+                 maxit = 1,
+                 m = M,
+                 print = FALSE)
+  } else {
+    # add an iteration
+    mids <- mice.mids(mids, maxit = 1, printFlag = FALSE)
+  }
+
+  # plot density of imputed values
+imputed_values[[it]] <-  mids[["imp"]][vars]
+}
+
+# #test3 <- map(1:T_max, function(x) map(imputed_values, function(y) y[vars]))
+# test4 <- map(imputed_values, function(x) x[vars])
+
+imputed_values[[1]]$bmi %>% mutate(case = row_number())
+#map(imputed_values, function(x) map(vars, function(y) {print(x)}))#mutate(case = row_number())})) #%>%  tidyr::gather("imp", "value", 1:M) %>% .[order(.$case, .$imp),]}) 
+test5 <- map(imputed_values, function(x) map(x, function(y) {as.data.frame(y) %>% mutate(case = row_number()) %>%  tidyr::gather("imp", "value", 1:M)}))  #%>% .[order(.$case, .$imp),]}) })) 
+# case_mean <- imputed_values %>% group_by(case) %>% summarise(mu = mean(value), sigma = sd(value))
+
+# # plot hist per case
+# test4[[1]] %>% map(~{ggplot(aes(x = value)) +
+#     geom_histogram() +
+#     geom_vline(data = case_mean, mapping = aes(xintercept = mu)) +
+#     facet_wrap( ~ case)})
+
+```

### 2020-02-22 3.Thesis/plot_imputation_densities.Rmd
update omdat laptop uitvalt

@@
 #map(imputed_values, function(x) map(vars, function(y) {print(x)}))#mutate(case = row_number())})) #%>%  tidyr::gather("imp", "value", 1:M) %>% .[order(.$case, .$imp),]}) 
-test5 <- map(imputed_values, function(x) map(x, function(y) {as.data.frame(y) %>% mutate(case = row_number()) %>%  tidyr::gather("imp", "value", 1:M)}))  #%>% .[order(.$case, .$imp),]}) })) 
+test5 <- map(imputed_values, function(.x) map(.x, function(.y) {as.data.frame(.y) %>% mutate(case = row_number()) %>%  tidyr::gather("imp", "value", 1:M)  %>% .[order(.$case),] %>% mutate(.x = .$value) }))#group_by(.$case) %>% summarise(mu = mean(.$value), sigma = sd(.$value))}) )
 # case_mean <- imputed_values %>% group_by(case) %>% summarise(mu = mean(value), sigma = sd(value))
+# test5[[1]]$bmi %>% group_by(case) %>% summarise(mu = mean(value), sigma = sd(value))  
+test5[[1]] %>% map(vars, ~{.x %>% rename(bmi = value)})
 

### 2020-02-26 3.Thesis/plot_imputation_densities.Rmd
same with different data

--- 2020-02-20 3.Thesis/plot_imputation_densities.Rmd
@@
 set.seed(123)
-dat <- nhanes
+dat <- boys
 
@@
 M <- 20
-T_max <- 10
+T_max <- 50
 
@@
                  maxit = 1,
+                 #method = "norm",
                  m = M,
@@
 #sd_per_case_per_it %>%  tidyr::gather("case", "sd", 1:9) %>% ggplot() + geom_line(aes(x = nr_it, y = sd)) + facet_wrap(~case)
-sd_per_case_per_it %>%  tidyr::gather("case", "sd", 1:9) %>% ggplot(aes(x = nr_it, y = sd, col = case)) + geom_line() + geom_smooth(se = F)
+sd_per_case_per_it %>%  tidyr::gather("case", "sd", 1:9) %>% ggplot(aes(x = nr_it, y = sd, col = case)) + geom_point() + geom_smooth(se = F)
 
@@
 
-Now for any variable
+Now for any variable, for one iteration nr. only.
 ```{r}
-# objects
-imputed_values <- plot <- distr <- list()
-
 # which vars?
+# only numeric
+num_vars <- dat %>% map(~{is.numeric(.x)})
+dat <- dat[,num_vars==T]
+# only with missings
 nmis <- dat %>% map_dbl(~{is.na(.x) %>% sum()}) #%>% which(.>0) %>% names()
@@
 vars <- names(nmis[nmis>0])
+
+# extract imputed values, reshape data
+imp_long <- mids[["imp"]][vars] %>% map( ~{
+  mutate(., case = rownames(.)) %>%  tidyr::gather("imp", "value", 1:M)
+  })
+
+# compute mean and sd for each case
+imp_descr <- imp_long %>% map( ~{
+  group_by(., case) %>% summarise(mu = mean(value), sigma = sd(value), sd1 = mu+sigma, sd2 = mu-sigma)
+})
+
+# plot each case with mean and sd per variable
+map2(imp_long, imp_descr, ~{
+  ggplot(.x, aes(x = value)) +
+    geom_histogram() +
+    geom_vline(.y, mapping = aes(xintercept = mu)) +
+    geom_vline(.y, mapping = aes(xintercept = sd1), linetype = "dotted") +
+    geom_vline(.y, mapping = aes(xintercept = sd2), linetype = "dotted") +
+    ggtitle("Distribution of imputed values per case (mean and 1 SD indicated)") +
+    #xlab(.y) +
+    facet_wrap( ~ case)
+})
+
+```
+
+Now try to automate this for all nrs of iterations
+```{r eval=FALSE, include=FALSE}
+# objects
+imputed_values <- plot <- distr <- list()
 
@@
 #map(imputed_values, function(x) map(vars, function(y) {print(x)}))#mutate(case = row_number())})) #%>%  tidyr::gather("imp", "value", 1:M) %>% .[order(.$case, .$imp),]}) 
-test5 <- map(imputed_values, function(x) map(x, function(y) {as.data.frame(y) %>% mutate(case = row_number()) %>%  tidyr::gather("imp", "value", 1:M)}))  #%>% .[order(.$case, .$imp),]}) })) 
+test5 <- map(imputed_values, function(.x) map(.x, function(.y) {as.data.frame(.y) %>% mutate(case = row_number()) %>%  tidyr::gather("imp", .y, 1:M)  %>% .[order(.$case),] %>% mutate(.x = .$value) }))#group_by(.$case) %>% summarise(mu = mean(.$value), sigma = sd(.$value))}) )
 # case_mean <- imputed_values %>% group_by(case) %>% summarise(mu = mean(value), sigma = sd(value))
+# test5[[1]]$bmi %>% group_by(case) %>% summarise(mu = mean(value), sigma = sd(value))
+
+#test5[[1]] %>% map(function(x) map(x, function(y) {rename(., y = .$value)}))
 

### 2020-02-27 3.Thesis/plot_imputation_densities.Rmd
Notes meeting with Gerko, and added Univariate bias plots (to compare with AC and R hat of chain means).

@@
 ```
+Bij it 1 is de random component gewoon een groter aandeel.
+Ookal is het numeriek niet geconv, maar de uitkomst klopt.
+Nu is het niet geconv, maar op de onzekerheid heeft het geen effect.
+Maak deze ook voor mu!
+Nu niet met 9 waarden, maar met de 21 van boys.
+En met hgt ipv bmi.
+
+Waarom itereren we dan nog? Bij it 2 zijn we al volledig onafh van de startwaarden.
+
+Mahal op complete set ipv missing data. wel de hele set nemen. misschien dat ie nodeloos stabiel wordt als je weinig missing hebt (maar dan conv ie ook heel snel).
 
@@
 
+map2(imp_long, imp_descr, ~{
+  ggplot(.x, aes(y = value, color = case)) +
+    geom_boxplot() 
+})
+
 ```
+
+Boxplot is makkelijker te zien!
+Gelijke var en means.
+Na imp heb je 1 plotje dat een boxpl is van alle geimp.
+x as cases, y as variatie.
+sommige cases heel veel spreiding en andere heel weinig.
+Dat is leuk!
+miss op y en kolommen op x as, dan boxplot met wel/niet een bepaalde var missing.
 

### 2020-02-27 2.ResearchReport/1.SimulationStudy/3.CreateFigure.R

@@ -8,9 +8,9 @@ library("ggplot2")
 library("ggpubr")
 
 # # if necessary, load results
-# load("C:/Users/User/Desktop/shinyMice/Simulation/Results/results.Rdata")
+load("C:/Users/User/Desktop/shinyMice/2.ResearchReport/1.SimulationStudy/Results/full_results.Rdata")
 # results <- dat
-
+results_with_CI <- dat #results
 # create plot for R hat
 R_plot <- results_with_CI %>% .[-1,] %>%
   ggplot(aes(x = T, y = R_var_X, color = "Chain variances")) +
@@ -108,7 +108,7 @@ figure1
 
 # create plot for bias
 bias_plot <- ggplot(results_with_CI, aes(x = T, y = bias)) +
-  geom_line() +
+  geom_point() +
   geom_smooth(
     method = "loess",
     se = FALSE,
@@ -116,12 +116,11 @@ bias_plot <- ggplot(results_with_CI, aes(x = T, y = bias)) +
     size = 0.5,
     linetype = "dashed"
   ) +
-  # # add error bar when `results_withSEs` is used
-  # geom_errorbar(
-  #   aes(ymin = bias_LL, ymax = bias_UL),
-  #   width = .2,
-  #   colour = "grey"
-  # ) +
+  geom_errorbar(
+    aes(ymin = bias_LL, ymax = bias_UL),
+    width = .2,
+    colour = "grey"
+  ) +
   xlab("") +
   ylab("Bias") +
   scale_x_continuous(breaks = seq(0, 100, by = 10)) +
@@ -132,6 +131,7 @@ bias_plot <- ggplot(results_with_CI, aes(x = T, y = bias)) +
     panel.grid.minor = element_blank(),
     axis.line = element_line(colour = "black")
   )
+
 
 # create plot for confidence interval width
 CIW_plot <- ggplot(results_with_CI, aes(x = T, y = CIW)) +
@@ -211,3 +211,85 @@ figure3 <-
   )
 # print figure
 figure3
+
+###################
+
+bias_univ_plot <- ggplot(results_with_CI, aes(x = T, y = bias_X, color = "X")) +
+  geom_point() +
+  geom_smooth(se = F) +
+  # geom_errorbar(
+  #   aes(ymin = bias_X_LL, ymax = bias_X_UL),
+  #   width = .2,
+  #   colour = "grey"
+  # ) +
[…]

### 2020-02-27 2.ResearchReport/1.SimulationStudy/4.SomeMorePlots.R

@@ -8,8 +8,8 @@ library("ggplot2")
 library("ggpubr")
 
 # # if necessary, load results
-# load("C:/Users/User/Desktop/shinyMice/2.ResearchReport/1.SimulationStudy/Results/full_results.Rdata")
-# results <- dat
+load("C:/Users/User/Desktop/shinyMice/2.ResearchReport/1.SimulationStudy/Results/full_results.Rdata")
+results <- dat
 
 # create plot for autocorrelations
 new_AC_plot <- results %>% .[-1,] %>%

### 2020-03-05 3.Thesis/quotes_to_include.Rmd
added file with references for manuscript, adjusted densities/mean/sd of imputations per iteration, per case

---
title: "Literature on non-convergence"
author: "Hanne Oberman"
date: "4-3-2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

"As MICE is an iterative procedure, it is important that convergence is achieved. This may be checked by computing, at
each cycle, the means of imputed values and/or the values of regression coefficients, and seeing if they are stable. For the
example in Section 9, these values appeared stable from the very first cycle. We have never found 10 cycles inadequate,
but larger numbers of cycles might in principle be required when incomplete variables are very strongly associated" (@whit11, p. 394).

# For ShinyMICE
"Determining How Many Imputations Are Needed
Working back from the computation of statistics and confidence intervals, we need
to ask how many imputations are required to generate useful statistical inferences. Surprisingly, the answer is often no more than five and sometimes as little as two or three.
The relative efficiency (RE) of the m imputations given a fraction of missing data
(Rubin, 1987, p. 114) is computed in standard error units as
...
Even when γ is .9, five imputations give a relative efficiency of .92. Consequently,
one rarely has to create more than five imputations, and 10 imputations are more than
suitable in almost any realistic application" (https://journals.sagepub.com/doi/pdf/10.1177/1094428103255532, p. 291).

### 2020-03-05 2.ResearchReport/1.SimulationStudy/4.SomeMorePlots.R

@@ -7,8 +7,8 @@ library("dplyr")
 library("ggplot2")
 library("ggpubr")
 
-# # if necessary, load results
-load("C:/Users/User/Desktop/shinyMice/2.ResearchReport/1.SimulationStudy/Results/full_results.Rdata")
+# # # if necessary, load results
+# load("C:/Users/User/Desktop/shinyMice/2.ResearchReport/1.SimulationStudy/Results/full_results.Rdata")
 results <- dat
 
 # create plot for autocorrelations
@@ -29,10 +29,10 @@ new_AC_plot <- results %>% .[-1,] %>%
   #           linetype = "dashed",
   #           size = 1) +
   ggplot() + 
-  geom_line(aes(x = T, y = AC_mean_X, color = "Chain means")) +
-  geom_line(aes(x = T, y = AC_mean_Z1, color = "Chain means")) +
-  geom_line(aes(x = T, y = AC_mean_Z2, color = "Chain means")) +
-  geom_line(aes(x = T, y = AC_mean_Y, color = "Chain means")) +
+  geom_line(aes(x = T, y = AC.mean.X, color = "Chain means")) +
+  geom_line(aes(x = T, y = AC.mean.Z1, color = "Chain means")) +
+  geom_line(aes(x = T, y = AC.mean.Z2, color = "Chain means")) +
+  geom_line(aes(x = T, y = AC.mean.Y, color = "Chain means")) +
   xlab(" ") +
   ylab("Auto-correlation") +
   scale_x_continuous(breaks = seq(0, 100, by = 10)) +
@@ -75,10 +75,10 @@ between_plot <- results %>% .[-1,] %>%
   #           linetype = "dashed",
   #           size = 1) +
   ggplot() +
-  geom_line(aes(x = T, y = between_mean_X, color = "Chain means")) +
-  geom_line(aes(x = T, y = between_mean_Z1, color = "Chain means")) +
-  geom_line(aes(x = T, y = between_mean_Z2, color = "Chain means")) +
-  geom_line(aes(x = T, y = between_mean_Y, color = "Chain means")) +
+  geom_line(aes(x = T, y = between.mean.X, color = "Chain means")) +
+  geom_line(aes(x = T, y = between.mean.Z1, color = "Chain means")) +
+  geom_line(aes(x = T, y = between.mean.Z2, color = "Chain means")) +
[…]

### 2020-03-05 2.ResearchReport/1.SimulationStudy/Functions/CreateData.R

@@ -27,3 +27,9 @@ data.simulation <- function(n = populationsize, true_effect = 2) {
   return(simdata)
   
 }
+
+###
+# ampute
+# 50% univar miss
+# sapply(A, function(x) sample(c(0, NA), length(x))
+# 

### 2020-03-05 4.ShinyMICE/plot_imputation_densities.Rmd
restructure non-conv branch to merge into master

--- 2020-02-27 3.Thesis/plot_imputation_densities.Rmd
@@
 dat <- boys
+
+#### hetzelfde met case nrs omgedraaid!
 
@@
   imputed_values <-
-    mids[["imp"]]$bmi %>% mutate(case = row_number()) %>%  tidyr::gather("imp", "value", 1:M) %>% .[order(.$case, .$imp),] #%>% group_by(case) %>% mutate(case_mean = mean(value)) %>% ungroup()
+    mids[["imp"]]$hgt %>% mutate(case = row_number()) %>%  tidyr::gather("imp", "value", 1:M) %>% .[order(.$case, .$imp),] #%>% group_by(case) %>% mutate(case_mean = mean(value)) %>% ungroup()
   distr[[it]] <-
@@
 
+n_cases <- max(distr[[1]]$case)
 # or with ggplot to put them in 1 plot
 sd_per_case_per_it <-
-  purrr::map(1:9, function(x)
+  purrr::map(1:n_cases, function(x)
     purrr::map_dbl(distr, ~ {
@@
 
-colnames(sd_per_case_per_it) <- c(1:9, "nr_it")
+colnames(sd_per_case_per_it) <- c(1:n_cases, "nr_it")
 
 #sd_per_case_per_it %>%  tidyr::gather("case", "sd", 1:9) %>% ggplot() + geom_line(aes(x = nr_it, y = sd)) + facet_wrap(~case)
-sd_per_case_per_it %>%  tidyr::gather("case", "sd", 1:9) %>% ggplot(aes(x = nr_it, y = sd, col = case)) + geom_point() + geom_smooth(se = F)
+sd_per_case_per_it %>%  tidyr::gather("case", "sd", 1:n_cases) %>% ggplot(aes(x = nr_it, y = sd, col = case)) + geom_point() + geom_smooth(se = F)
+
+############# means
+# or with ggplot to put them in 1 plot
+mean_per_case_per_it <-
+  purrr::map(1:n_cases, function(x)
+    purrr::map_dbl(distr, ~ {
+      .$mu[.$case == x]
+    })) %>% as.data.frame() %>% mutate(nr_it = row_number())
+
+colnames(mean_per_case_per_it) <- c(1:n_cases, "nr_it")
+
+#sd_per_case_per_it %>%  tidyr::gather("case", "sd", 1:9) %>% ggplot() + geom_line(aes(x = nr_it, y = sd)) + facet_wrap(~case)
+mean_per_case_per_it %>%  tidyr::gather("case", "mean", 1:n_cases) %>% ggplot(aes(x = nr_it, y = mean, col = case)) + geom_point() + geom_smooth(se = F)
 

### 2020-03-05 3.Thesis/1.SimulationStudy/2.CreateFigures.R

@@ -1,0 +1,456 @@
+# create figure to display results
+# requires the packages "dplyr", "ggplot2"
+# and the object 'results' created with '1.Excecute'
+
+# load package
+library("dplyr")
+library("ggplot2")
+library("ggpubr")
+
+# if necessary, load results
+# load("C:/Users/User/Desktop/shinyMice/2.ResearchReport/1.SimulationStudy/Results/full_results.Rdata")
+# results <- dat
+# results_with_CI <- dat #results
+
+# # create plot for R hat
+# R_plot <- results_with_CI %>% .[-1,] %>%
+#   ggplot(aes(x = T, y = R.var.X1, color = "Chain variances")) +
+#   geom_line(linetype = "dashed", size = 1) +
+#   geom_line(aes(x = T, y = R.var.X2, color = "Chain variances"),
+#             linetype = "dashed",
+#             size = 1) +
+#   geom_line(aes(x = T, y = R.var.X3, color = "Chain variances"),
+#             linetype = "dashed",
+#             size = 1) +
+#   geom_line(aes(x = T, y = R.var.Y, color = "Chain variances"),
+#             linetype = "dashed",
+#             size = 1) +
+#   geom_line(aes(x = T, y = R.mean.X1, color = "Chain means")) +
+#   geom_line(aes(x = T, y = R.mean.X2, color = "Chain means")) +
+#   geom_line(aes(x = T, y = R.mean.X3, color = "Chain means")) +
+#   geom_line(aes(x = T, y = R.mean.Y, color = "Chain means")) +
+#   # # add error bar when `results_withSEs` is used
+#   # geom_errorbar(
+#   #   aes(ymin = R.mean.X1_LL, ymax = R.mean.X1_UL),
+#   #   width = .2,
+#   #   position = position_dodge(0.05)
+#   # ) +
+#   xlab("") +
+#   ylab(expression(paste(widehat(R)))) +
+#   scale_x_continuous(breaks = seq(0, 100, by = 10)) +
+#   scale_y_continuous(breaks = seq(.8, 2.2, by = .2)) +
+#   scale_color_manual(name = "Legend", labels = c("Chain means", "Chain variances"), values = c("black", "grey"), guide = "legend") +
[…]

### 2020-03-05 3.Thesis/1.SimulationStudy/Functions/Evaluate.R

@@ -1,0 +1,77 @@
+# Extract aggregated simulation results by averaging over runs
+# requires the packages "purrr" and "data.table"
+
+evaluate.sim <-
+  function(sims,
+           n.iter,
+           mean_or_SE = "mean"
+           ) {
+    # organize output for evaluation
+    dt <- map(sims, as.data.table) %>% rbindlist(fill = TRUE)
+    # determine whether we want averages per simulation condition or MCMC SEs
+    if (mean_or_SE == "se") {
+      function_to_apply <- sd # or use: function(x) sqrt((var(x)))
+    } else if (mean_or_SE == "lower") {
+      function_to_apply <- function(x) quantile(x, .025, na.rm = TRUE)
+    } else if (mean_or_SE == "upper") {
+      function_to_apply <- function(x) quantile(x, .975, na.rm = TRUE)
+    } else {function_to_apply <- mean}
+
+    
+    # apply function to aggregate simulation runs per simulation condition
+    d <- dt[, lapply(.SD, function_to_apply)] %>% as.data.frame
+    # create object for output
+    output <- 1:n.iter %>% as.data.frame()
+    # attatch diagnostics
+    # vars <-
+    #   c(
+    #     # "bias",
+    #     # "CIW",
+    #     # "cov",
+    #     "bias_X",
+    #     "bias_Y",
+    #     "bias_Z1",
+    #     "bias_Z2",
+    #     "R_mean_X",
+    #     "R_mean_Z1",
+    #     "R_mean_Z2",
+    #     "R_mean_Y",
+    #     "R_var_X",
+    #     "R_var_Z1",
+    #     "R_var_Z2",
+    #     "R_var_Y",
+    #     "AC_mean_X",
+    #     "AC_mean_Z1",
+    #     "AC_mean_Z2",
+    #     "AC_mean_Y",
+    #     "AC_var_X",
+    #     "AC_var_Z1",
+    #     "AC_var_Z2",
+    #     "AC_var_Y",
+    #     "between_mean_X",
+    #     "between_mean_Z1",
+    #     "between_mean_Z2",
+    #     "between_mean_Y",
+    #     "between_var_X",
+    #     "between_var_Z1",
+    #     "between_var_Z2",
+    #     "between_var_Y",
+    #     "within_mean_X",
+    #     "within_mean_Z1",
+    #     "within_mean_Z2",
+    #     "within_mean_Y",
+    #     "within_var_X",
[…]

### 2020-03-12 2.ResearchReport/1.SimulationStudy/Writeup/NoteConvergence.tex
Copy simulation Research Report from master branch to merge branches

--- 2020-01-15 2.ResearchReport/1.SimulationStudy/Writeup/NoteConvergence.Rnw
@@
 
-<<echo = FALSE, eval = TRUE>>=
-load("../Results/results.Rdata")
-names(dat) <- c("It.", "Bias", "CI width", "Cov. rate", "R chain mean", "R chain var", "AC chain mean", "AC chain var")
-@
-
 Figures \ref{fig:conv} and \ref{fig:sim} display results per simulation condition ($T = 1,2,\dots,100$). A subset of simulation conditions is presented in Table \ref{tab:results}.
 
-% % See "3.CreateFigure.R", I specifically chose to NOT generate the figures within this Rnw file because of journal guidelines.
 \begin{figure}[h]
@@
 
-% % With SEs
 % \begin{figure}[h]
@@
 These conditions have similar CIWs, but across iterations stability is not established (i.e., the Loess line is never completely flat within the 100 simulation conditions).
-%After about twenty-five iterations the average CIW across all simulation conditions and repetitions is reached \Sexpr{round(mean(dat[, "CI width"]), 3)}.
+%After about twenty-five iterations the average CIW across all simulation conditions and repetitions is reached 0.927.
 
@@
 We see some over-coverage at $T=2$, but that is better than under-estimating the variance of $\bar{Q}$.
-On average, the coverage rate is somewhat higher than the expected nominal coverage of 95\% \citep{neym34}, namely \Sexpr{100*round(mean(dat[, "Cov.
-rate"]), 3)}\%.
+On average, the coverage rate is somewhat higher than the expected nominal coverage of 95\% \citep{neym34}, namely 95\%.
 Similar to the CIWs, there is some trending across iterations (i.e., the Loess line is never flat).
@@
 
-The results in this paper were obtained using \proglang{R}~\Sexpr{paste(R.Version()[6:7], collapse = ".")} \cite{R} with the \pkg{mice}~\Sexpr{paste(packageVersion("mice"), collapse = ".")} package \cite{mice}.
+The results in this paper were obtained using \proglang{R}~3.6.2 \cite{R} with the \pkg{mice}~3.7.0 package \cite{mice}.
 \proglang{R} itself and all packages used are available from the Comprehensive \proglang{R} Archive Network (CRAN) at \url{https://CRAN.R-project.org/}.

### 2020-03-12 2.ResearchReport/1.SimulationStudy/1.Execute.R

@@ -1,0 +1,94 @@
+# script to perform simulation study on convergence of mice
+# author: Hanne Oberman
+
+###
+
+# Creating a conflict because I want to keep the master files of this folder, does this work?
+
+# load required packages
+library(mice)
+library(mvtnorm)
+library(miceadds)
+library(data.table)
+library(tidyverse)
+# or if you don't want the entire tidyverse, just load these
+# library(dplyr)
+# library(ggplot2)
+# library(purrr)
+# library(magrittr)
+# library(tibble)
+
+# load simulation/evaluation functions
+source("Simulation/Functions/CreateData.R")
+source("Simulation/Functions/Convergence.R")
+source("Simulation/Functions/Convergence_supplement.R")
+source("Simulation/Functions/Autocorrelation.R")
+source("Simulation/Functions/Impute.R")
+source("Simulation/Functions/Evaluate.R")
+
+# simulation parameters
+populationsize <- 1000 #n of simulated dataset
+n.iter <- 100 #nr of iterations (varying 1:n.iter)
+n.sim <- 100 #nr of simulations per iteration value
+true_effect <- 2 #regression coefficient to be estimated
+
+# start simulation study
+set.seed(1111)
+
+###
+
+# simulate data once
+data <- data.simulation(n = populationsize, true_effect)
+# run ampute once to get patterns object (and )to be able to adjust it for multivariate missingness)
+amp_patterns <- ampute(data)$patterns
+# for multivariable missingness, uncomment this
+amp_patterns[1:4, 1] <- 0
+
+# combine separate functions into wrapper
+simulate <- function(data, n.iter, true_effect, patterns) {
+  pb <- txtProgressBar(min = 0, max = n.iter, style = 3)
+  
+  # remove values at random with 20 percent probability to be missing
+  ampdata <- ampute(data, patterns = amp_patterns, prop = 0.8, mech = "MCAR")$amp
+  
+  # object for output
+  res <- list()
+  # repeat mi procedure 'runs'  times for each nr of iterations
+  #for (run in 1:runs) {
+    for (i in 1:n.iter) {
+      res[[i]] <- test.impute(true_effect, data = ampdata, maxit = i)
+      setTxtProgressBar(pb, i)
+    }
+  #}
+  close(pb)
+  # output
+  names(res) <- 1:n.iter
+  res
+}
+
+# simulate
[…]

### 2020-03-12 2.ResearchReport/1.SimulationStudy/Functions/Convergence_supplement.R

@@ -1,0 +1,47 @@
+# Supplementary functions for Rhat function (see Convergence.R)
+# To get folded, split-half, rank-normalized Rhat
+
+# Creating a conflict because I want to keep the master files of this folder, does this work?
+
+# fold chains for rhat of tails
+fold_sims <- function(sims) {
+  # fold Markov chain for rhat of tails, adapted from rstan
+  abs(sims - median(sims))
+}
+
+# split chains with maxit > 4 to detect trending
+split_chains <- function(sims) {
+  # split Markov chains, adapted from rstan
+  niter <- dim(sims)[1]
+  # do not split if result will be chains of length 1
+  if (niter < 4)
+    return(sims)
+  # split each chain to get 2m chains
+  half <- niter / 2
+  cbind(sims[1:floor(half), ], sims[ceiling(half + 1):niter, ])
+}
+
+# rank-normalize chains because Gelman says so
+z_scale <- function(x) {
+  # rank-normalize Markov chain, copied from rstan
+  S <- length(x)
+  r <- rank(x, ties.method = 'average')
+  z <- qnorm((r - 1 / 2) / S)
+  if (!is.null(dim(x))) {
+    # output should have the input dimension
+    z <- array(z, dim = dim(x), dimnames = dimnames(x))
+  }
+  z
+}
+
+# compute rhat
+get.rhat <- function(sims, maxit = maxit) {
+  # compute potential scale reduction factor (rhat) for each variable in mids object
+  # equations adapted from Vehtari et al. (2019)
+  var_between <-
+    maxit * var(apply(sims, 2, mean))
+  var_within <- mean(apply(sims, 2, var))
+  rhat <-
+    sqrt((var_between / var_within + maxit - 1) / maxit)
+  return(rhat)
+}

### 2020-03-12 2.ResearchReport/1.SimulationStudy/Functions/Impute.R

@@ -1,0 +1,83 @@
+# Perform mi on mids object with varying maxit values
+# requires the package 'mice' and the function 'get.rhat'
+
+# Creating a conflict because I want to keep the master files of this folder, does this work?
+
+test.impute <- function(true_effect,
+                        data,
+                        m = 5,
+                        method = "norm",
+                        maxit,
+                        ...) {
+
+  # impute missing values
+  if (maxit == 1) {
+    impsim <<- mice(
+      data,
+      method = method,
+      m = m,
+      maxit = maxit,
+      print = FALSE,
+      ...
+    )
+  } else {
+    impsim <<- mice.mids(impsim, maxit = 1, printFlag = FALSE)
+  }
+  
+  # compute convergence diagnostics
+  if (maxit < 2) {
+    R_mean <- R_var <- AC_mean <- AC_var <- c(rep(NA, 4))
+  } else {
+    R_mean <-
+      rhat_function(impsim, maxit) #maximum Rhat across variables
+    R_var <-
+      rhat_function(impsim, maxit, moment = "variance") #maximum Rhat across variables
+    AC_mean <-
+      autocorr_function(impsim, maxit) #auto-correlation at lag 1
+    AC_var <-
+      autocorr_function(impsim, maxit, moment = "variance") #auto-correlation at lag 1
+  }
+  
+
+  # perform analysis
+  # mip <- unlist(pool(with(impsim, lm(Y ~ X + Z1 + Z2))))
+  mip <- impsim %>% with(lm(Y ~ X + Z1 + Z2)) %>% pool %>% .$pooled
+  
+  # compute simulation diagnostics
+  est <- mip$estimate[2] #estimated regression coefficient
+  bias <- est - true_effect #bias
+  SE <- sqrt(mip$b[2] + (mip$b[2] / m)) #pooled finite SE
+  CI.low <- est - qt(.975, df = m - 1) * SE #lower bound CI
+  CI.up <- est + qt(.975, df = m - 1) * SE #upper bound CI
+  CIW <- CI.up - CI.low #confidence interval width
+  cov <- CI.low < true_effect & true_effect < CI.up #coverage
+  
+  # output
+  return(
+    data.frame(
+      bias = bias,
+      CIW = CIW,
[…]

### 2020-03-12 4.ShinyMICE/inspect_incomplete_data.Rmd

--- 2020-02-18 3.Thesis/inspect_incomplete_data.Rmd
@@
 ---
-title: "Investigate relations in incomplete datasets to inform imputation model"
+title: "ShinyMICE: Investigate relations in incomplete datasets to inform imputation model"
 output:
@@
   map(~{ggplot(data=dat, aes(x = age, fill = is.na(.data[[.x]]))) +
-  geom_histogram(aes(y = 100 * ..density..),
-                 alpha = 0.5,
+  geom_histogram(alpha = 0.5,
                  position = 'identity') +
   xlab("Age (years)") +
-  ylab("Percentage observed") +
+  ylab("Count of cases") +
   scale_fill_manual(

### 2020-03-16 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
Added prediction performance measures (MAE, RMSE, etc.)

---
title: "Manuscript"
author: "Hanne Oberman"
date: "16-3-2020"
output: 
  pdf_document:
    keep_tex: true

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2020-03-16 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Added prediction performance measures (MAE, RMSE, etc.)

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Manuscript},
  pdfauthor={Hanne Oberman},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering

\title{Manuscript}
\author{Hanne Oberman}
\date{16-3-2020}

\begin{document}
\maketitle

\hypertarget{r-markdown}{%
\subsection{R Markdown}\label{r-markdown}}

This is an R Markdown document. Markdown is a simple formatting syntax
for authoring HTML, PDF, and MS Word documents. For more details on
using R Markdown see \url{http://rmarkdown.rstudio.com}.

When you click the \textbf{Knit} button a document will be generated
that includes both content as well as the output of any embedded R code
chunks within the document. You can embed an R code chunk like this:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{summary}\NormalTok{(cars)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##      speed           dist       
##  Min.   : 4.0   Min.   :  2.00  
##  1st Qu.:12.0   1st Qu.: 26.00  
##  Median :15.0   Median : 36.00  
##  Mean   :15.4   Mean   : 42.98  
##  3rd Qu.:19.0   3rd Qu.: 56.00  
##  Max.   :25.0   Max.   :120.00
\end{verbatim}

\hypertarget{including-plots}{%
\subsection{Including Plots}\label{including-plots}}

You can also embed plots, for example:

\includegraphics{manuscript_files/figure-latex/pressure-1.pdf}

Note that the \texttt{echo\ =\ FALSE} parameter was added to the code
chunk to prevent printing of the R code that generated the plot.

\end{document}

### 2020-03-16 3.Thesis/1.SimulationStudy/2.CreateFigures.R

@@ -443,9 +443,11 @@ bias_ests_plot <- results_with_CI %>% #.[-1,] %>%
     #legend.position = ""
   )
 
+
 ests_plot <-
   ggarrange(
     bias_ests_plot,
+    #R_sq_plot,
     PCA_plot,
     ncol = 1,
     nrow = 2,
@@ -454,3 +456,109 @@ ests_plot <-
 
 # print figure
 ests_plot
+
+####################################################################
+
+# prediction
+
+R_sq_plot <- results_with_CI %>% #.[-1,] %>% 
+  ggplot() +
+  geom_point(aes(x = T, y = R.sq, color = "R^2")) +
+  geom_smooth(aes(x = T, y = R.sq, color = "R^2"), se = FALSE) +
+  xlab("") +
+  ylab("Explained variance") +
+  #ggtitle("Multivariate estimates") +
+  theme_bw() +
+  theme(
+    panel.border = element_blank(),
+    panel.grid.major = element_blank(),
+    panel.grid.minor = element_blank(),
+    axis.line = element_line(colour = "black") #,
+    #legend.position = ""
+  )
+
+RMSE_plot <- results_with_CI %>% #.[-1,] %>% 
+  ggplot() +
+  geom_point(aes(x = T, y = RMSE.1, color = "m1")) +
+  geom_point(aes(x = T, y = RMSE.2, color = "m2")) +
+  geom_point(aes(x = T, y = RMSE.3, color = "m3")) +
+  geom_point(aes(x = T, y = RMSE.4, color = "m4")) +
+  geom_point(aes(x = T, y = RMSE.5, color = "m5")) +
+  geom_smooth(aes(x = T, y = RMSE.1, color = "m1"), se = FALSE) +
+  geom_smooth(aes(x = T, y = RMSE.2, color = "m2"), se = FALSE) +
+  geom_smooth(aes(x = T, y = RMSE.3, color = "m3"), se = FALSE) +
+  geom_smooth(aes(x = T, y = RMSE.4, color = "m4"), se = FALSE) +
+  geom_smooth(aes(x = T, y = RMSE.5, color = "m5"), se = FALSE) +
+  xlab("") +
+  ylab("Root mean squared error") +
+  #ggtitle("Multivariate estimates") +
+  theme_bw() +
+  theme(
+    panel.border = element_blank(),
+    panel.grid.major = element_blank(),
+    panel.grid.minor = element_blank(),
[…]

### 2020-03-16 3.Thesis/1.SimulationStudy/Functions/Impute.R

@@ -68,6 +68,24 @@ test.impute <- function(true_effect,
   
   # compute multivariate diagnostics
   bias_est <- true_effect - map(imputed, lm, formula = Y~X1+X2+X3) %>% pool() %>% .$pooled %>% .$estimate #%>% .[2]
+  #predicted_Y <- map(imputed, ~{true_effect[1] + true_effect["X1"] * .$X1 })
+  # bias_pred <- 1
+  # vars <- names(data)
+  R_sq <- lm.mids(Y~X1+X2+X3, impsim) %>% pool.r.squared() %>% .[1]
+  RMSE <- lm.mids(Y~X1+X2+X3, impsim) %>% .$analyses %>% map_dbl(., function(x){x$residuals %>% .^2 %>% mean() %>%  sqrt()})
+  MAE <- lm.mids(Y~X1+X2+X3, impsim) %>% .$analyses %>% map_dbl(., function(x){x$residuals %>% abs() %>% mean()})
+  error_var <- lm.mids(Y~X1+X2+X3, impsim) %>% .$analyses %>% map_dbl(., function(x){x$residuals %>% var()})
+  
+  # bias in pred is aggr measure
+  # bias variance of pred is R2 so bias in R2 summarises it quite nicely. so sq of correlation
+  # cell 1 and pred f cell 1: average absolute bias, mean abs error MEE (mean over errors)
+  # R2 is similarity in sign and effect 
+  # R2 is a nice measure with fisher trans first, then combine then combine, and transform again
+  # like RMSE R2 tells you how wrong you are and some info about spread.
+  # variance in pred vector 
+  # variance of the MAE and its variance is separate instead of seesaw
+  # MAE and variance is "I'm wrong on average, but if I have extremes I am far more wrong on"
+  # just compute all 4
   
   # bias_est <- pool(with(impsim, lm(Y ~ X + Z1 + Z2)))$pooled$estimate[2] #a <- pool(with(impsim, lm(Y ~ X + Z1 + Z2)))$pooled$estimate[2]
   # est <-  mip # mip$estimate[2] #estimated regression coefficient
@@ -89,7 +107,12 @@ test.impute <- function(true_effect,
     R.var = t(R_var),
     AC.var = t(AC_var),
     pca = t(pca),
-    bias.est = t(bias_est)
+    bias.est = t(bias_est),
+    R.sq = R_sq,
+    RMSE = t(RMSE),
+    MAE = t(MAE),
[…]

### 2020-03-16 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
Started improving Research Report manuscript for thesis

@@
 ---
-title: "Manuscript"
+title: "Missing the Point: Convergence of Multiple Imputation using Chained Equations Algorithms"
 author: "Hanne Oberman"
@@
     keep_tex: true
-
+bibliography: thesis.bib
 ---
@@
 
-## R Markdown
+# Introduction
 
-This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.
+Feedback:
 
-When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.
-You can embed an R code chunk like this:
+- Focus on convergence, not evaluation of MI in general.
 
-```{r cars}
-summary(cars)
-```
+- Combine Theoretical Background and Intro.
 
-## Including Plots
+At some point, any scientist conducting statistical analyses will run into a missing data problem [@alli02].
+Missingness is problematic because statistical inference cannot be performed on incomplete data without employing \emph{ad hoc} solutions (e.g., list-wise deletion), which may yield wildly invalid results [@buur18].
+A popular answer to the ubiquitous problem of missing information is to use the framework of multiple imputation (MI), proposed by @rubin87.
+MI is an iterative algorithmic procedure in which missing datapoints are `imputed' (i.e.
+filled in) several times.
+The variability between imputations is used to reflect how much uncertainty in the inference is introduced by the missingness.
+Therefore, MI can provide valid inferences despite missing information (**maybe add that this refers to CIs/p-values**).
 
-You can also embed plots, for example:
+To obtain valid inferences with MI, the variability between imputations should be properly represented [@rubin87; @buur18].
+If this variability is under-estimated, confidence intervals around estimates will be too narrow, which can yield spurious results.
+Over-estimation of the variance between imputations results in unnecessarily wide confidence intervals, which can be costly because it lowers the statistical power (**maybe add why this is costly**).
+Since both of these situations are undesirable, imputations and their variability should be evaluated.
+Evaluation measures, however, are currently missing or under-developed in MI software, like the world-leading `mice` package [@mice] in `R` [@R].
+The goal of this research project is to develop novel methodology and guidelines for evaluating MI methods.
+These tools will subsequently be implemented in an interactive evaluation framework for multiple imputation, which will aid applied researchers in drawing valid inference from incomplete datasets.
 
-```{r pressure, echo=FALSE}
-plot(pressure)
-```
+This note provides the theoretical foundation towards the diagnostic evaluation of multiple imputation algorithms.
+For reasons of brevity, we only focus on the MI algorithm implemented in `mice` [@mice].
+The convergence properties of this MI algorithm are investigated through model-based simulation [^1].
+The results of this simulation study are guidelines for assessing convergence of MI algorithms.
+<!-- These guidelines will be implemented in an interactive evaluation tool for `mice`, 'ShinyMICE', which is currently under development.
+We will evaluate how convergence of the MI algorithm can be diagnosed.
+-->
 
-Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
+[^1]: All programming code used in this note is available from [github.com/gerkovink/shinyMice/simulation]{https://github.com/gerkovink/ShinyMICE/simulation}.
+
+### Terminology
+
+<!-- The intended audience of this note consists of anyone who uses multiple imputation to solve missing data problem.  -->
+This note follows notation and conventions of `mice` [@mice]. 
+<!-- Deviations from the 'original' notation by @rubin87 are described in [@buur18, \S~2.2.3].  -->
+Basic familiarity with MI methodology is assumed.
+For the theoretical foundation of MI, see @rubin87.
+For an accessible and comprehensive introduction to MI from an applied perspective, see e.g.
+@buur18.
+
+Let $Y$ denote an $n \times p$ matrix containing the data values on $p$ variables for all $n$ units in a sample.
+The collection of observed data values in $Y$ is denoted by $Y_{obs}$, and will be referred to as 'incomplete' or 'observed' data.
+The missing part of $Y$ is denoted by $Y_{mis}$.
+Which parts of $Y$ are missing is determined by the `missingness mechanism'.
+This note only considers a 'missing completely at random' (MCAR) mechanism, where the probability of being missing is equal for all $n \times p$ cells in $Y$ [@rubin87].
+
+Figure 1 provides an overview of the steps involved with MI---from incomplete data, to $m$ completed datasets, to $m$ estimated quantities of interest ($\hat{Q}$s), to a single pooled estimate $\bar{Q}$.
+This note focuses on the algorithmic properties of the imputation step.
+
+The MI algorithm in `mice` has an iterative nature.
+For each missing datapoint in $Y_{mis}$, $m$ 'chains' of potential values are sampled.
+Only the ultimate sample that each chain lands on is imputed.
+The number of iterations per chain will be denoted with $T$, where $t$ varies over the integers $1, 2, \dots, T$.
+This is repeated $m$ times.
+Imputed values are the ultimate samples of a `chain' of For each imputed value ($t = T$), a s for each missing datapoint in $Y_{mis}$, .
+Each of the $m$ chains starts with an initial value, drawn randomly from $Y_{obs}$.
+The chains are terminated after a predefined number of iterations.
+, and subsequently used in the analysis and pooling steps.
+The collection of samples between the initial value (at $t=1$) and the imputed value (at $t=T$) will be referred to as an 'imputation chain'.
+
+# Methods
+
+# Results
+
+Feedback:
+
+- Remove the table.
+
+- Add more info about figure legends and axes.
+
+# Discussion
+
+# References

### 2020-03-16 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Started improving Research Report manuscript for thesis

@@
 \hypersetup{
-  pdftitle={Manuscript},
+  pdftitle={Missing the Point: Convergence of Multiple Imputation using Chained Equations Algorithms},
   pdfauthor={Hanne Oberman},
@@
 \usepackage[margin=1in]{geometry}
-\usepackage{color}
-\usepackage{fancyvrb}
-\newcommand{\VerbBar}{|}
-\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
-\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
-% Add ',fontsize=\small' for more characters per line
-\usepackage{framed}
-\definecolor{shadecolor}{RGB}{248,248,248}
-\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
-\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
-\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
-\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\BuiltInTok}[1]{#1}
-\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
-\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
-\newcommand{\ExtensionTok}[1]{#1}
-\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ImportTok}[1]{#1}
-\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\NormalTok}[1]{#1}
-\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
-\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
-\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\RegionMarkerTok}[1]{#1}
-\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
 \usepackage{graphicx,grffile}
@@
 
-\title{Manuscript}
+\title{Missing the Point: Convergence of Multiple Imputation using Chained
+Equations Algorithms}
 \author{Hanne Oberman}
@@
 
-\hypertarget{r-markdown}{%
-\subsection{R Markdown}\label{r-markdown}}
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
 
-This is an R Markdown document. Markdown is a simple formatting syntax
-for authoring HTML, PDF, and MS Word documents. For more details on
-using R Markdown see \url{http://rmarkdown.rstudio.com}.
+Feedback:
 
-When you click the \textbf{Knit} button a document will be generated
-that includes both content as well as the output of any embedded R code
-chunks within the document. You can embed an R code chunk like this:
+\begin{itemize}
+\item
+  Focus on convergence, not evaluation of MI in general.
+\item
+  Combine Theoretical Background and Intro.
+\end{itemize}
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{summary}\NormalTok{(cars)}
-\end{Highlighting}
-\end{Shaded}
+At some point, any scientist conducting statistical analyses will run
+into a missing data problem (Allison 2001). Missingness is problematic
+because statistical inference cannot be performed on incomplete data
+without employing \emph{ad hoc} solutions (e.g., list-wise deletion),
+which may yield wildly invalid results (Van Buuren 2018). A popular
+answer to the ubiquitous problem of missing information is to use the
+framework of multiple imputation (MI), proposed by Rubin (1987). MI is
+an iterative algorithmic procedure in which missing datapoints are
+`imputed' (i.e.~filled in) several times. The variability between
+imputations is used to reflect how much uncertainty in the inference is
+introduced by the missingness. Therefore, MI can provide valid
+inferences despite missing information (\textbf{maybe add that this
+refers to CIs/p-values}).
 
-\begin{verbatim}
-##      speed           dist       
-##  Min.   : 4.0   Min.   :  2.00  
-##  1st Qu.:12.0   1st Qu.: 26.00  
-##  Median :15.0   Median : 36.00  
-##  Mean   :15.4   Mean   : 42.98  
-##  3rd Qu.:19.0   3rd Qu.: 56.00  
-##  Max.   :25.0   Max.   :120.00
-\end{verbatim}
+To obtain valid inferences with MI, the variability between imputations
+should be properly represented (Rubin 1987; Van Buuren 2018). If this
+variability is under-estimated, confidence intervals around estimates
+will be too narrow, which can yield spurious results. Over-estimation of
+the variance between imputations results in unnecessarily wide
+confidence intervals, which can be costly because it lowers the
+statistical power (\textbf{maybe add why this is costly}). Since both of
+these situations are undesirable, imputations and their variability
+should be evaluated. Evaluation measures, however, are currently missing
+or under-developed in MI software, like the world-leading \texttt{mice}
+package (Van Buuren and Groothuis-Oudshoorn 2011) in \texttt{R} (R Core
+Team 2019). The goal of this research project is to develop novel
+methodology and guidelines for evaluating MI methods. These tools will
+subsequently be implemented in an interactive evaluation framework for
+multiple imputation, which will aid applied researchers in drawing valid
+inference from incomplete datasets.
 
-\hypertarget{including-plots}{%
-\subsection{Including Plots}\label{including-plots}}
+This note provides the theoretical foundation towards the diagnostic
+evaluation of multiple imputation algorithms. For reasons of brevity, we
+only focus on the MI algorithm implemented in \texttt{mice} (Van Buuren
+and Groothuis-Oudshoorn 2011). The convergence properties of this MI
+algorithm are investigated through model-based simulation.\footnote{All
+  programming code used in this note is available from
+  {[}github.com/gerkovink/shinyMice/simulation{]}\{\url{https://github.com/gerkovink/ShinyMICE/simulation}\}.}
+The results of this simulation study are guidelines for assessing
+convergence of MI algorithms.
 
-You can also embed plots, for example:
+\hypertarget{terminology}{%
+\subsubsection{Terminology}\label{terminology}}
 
-\includegraphics{manuscript_files/figure-latex/pressure-1.pdf}
+This note follows notation and conventions of \texttt{mice} (Van Buuren
+and Groothuis-Oudshoorn 2011). Basic familiarity with MI methodology is
+assumed. For the theoretical foundation of MI, see Rubin (1987). For an
+accessible and comprehensive introduction to MI from an applied
+perspective, see e.g.~Van Buuren (2018).
 
-Note that the \texttt{echo\ =\ FALSE} parameter was added to the code
-chunk to prevent printing of the R code that generated the plot.
+Let \(Y\) denote an \(n \times p\) matrix containing the data values on
+\(p\) variables for all \(n\) units in a sample. The collection of
+observed data values in \(Y\) is denoted by \(Y_{obs}\), and will be
+referred to as `incomplete' or `observed' data. The missing part of
+\(Y\) is denoted by \(Y_{mis}\). Which parts of \(Y\) are missing is
+determined by the `missingness mechanism'.\\
+This note only considers a `missing completely at random' (MCAR)
+mechanism, where the probability of being missing is equal for all
+\(n \times p\) cells in \(Y\) (Rubin 1987).
+
+Figure 1 provides an overview of the steps involved with MI---from
+incomplete data, to \(m\) completed datasets, to \(m\) estimated
+quantities of interest (\(\hat{Q}\)s), to a single pooled estimate
+\(\bar{Q}\). This note focuses on the algorithmic properties of the
+imputation step.
+
+The MI algorithm in \texttt{mice} has an iterative nature. For each
+missing datapoint in \(Y_{mis}\), \(m\) `chains' of potential values are
+sampled. Only the ultimate sample that each chain lands on is imputed.
+The number of iterations per chain will be denoted with \(T\), where
+\(t\) varies over the integers \(1, 2, \dots, T\). This is repeated
+\(m\) times. Imputed values are the ultimate samples of a `chain' of For
+each imputed value (\(t = T\)), a s for each missing datapoint in
+\(Y_{mis}\), . Each of the \(m\) chains starts with an initial value,
+drawn randomly from \(Y_{obs}\). The chains are terminated after a
+predefined number of iterations. , and subsequently used in the analysis
+and pooling steps. The collection of samples between the initial value
+(at \(t=1\)) and the imputed value (at \(t=T\)) will be referred to as
+an `imputation chain'.
+
+\hypertarget{methods}{%
+\section{Methods}\label{methods}}
+
+\hypertarget{results}{%
+\section{Results}\label{results}}
+
+Feedback:
+
+\begin{itemize}
+\item
+  Remove the table.
+\item
+  Add more info about figure legends and axes.
+\end{itemize}
+
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
+
+\hypertarget{references}{%
+\section*{References}\label{references}}
+\addcontentsline{toc}{section}{References}
+
+\hypertarget{refs}{}
+\leavevmode\hypertarget{ref-alli02}{}%
+Allison, Paul D. 2001. \emph{Missing Data}. Sage publications.
+
+\leavevmode\hypertarget{ref-R}{}%
+R Core Team. 2019. \emph{R: A Language and Environment for Statistical
+Computing}. Vienna, Austria. \url{https://www.R-project.org/}.
+
+\leavevmode\hypertarget{ref-rubin87}{}%
+Rubin, Donald B. 1987. \emph{Multiple Imputation for Nonresponse in
+Surveys}. Wiley Series in Probability and Mathematical Statistics
+Applied Probability and Statistics. New York, NY: Wiley.
+
+\leavevmode\hypertarget{ref-buur18}{}%
+Van Buuren, Stef. 2018. \emph{Flexible Imputation of Missing Data}.
+Chapman; Hall/CRC.
+
+\leavevmode\hypertarget{ref-mice}{}%
+Van Buuren, Stef, and Karin Groothuis-Oudshoorn. 2011. ``Mice:
+Multivariate Imputation by Chained Equations in R.'' \emph{Journal of
+Statistical Software} 45 (1): 1--67.
+\url{https://doi.org/10.18637/jss.v045.i03}.
 

### 2020-03-17 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
Adjust latex style to suit sociological methods and research journal

@@
 title: "Missing the Point: Convergence of Multiple Imputation using Chained Equations Algorithms"
-author: "Hanne Oberman"
+author: "Hanne I. Oberman"
 date: "16-3-2020"
@@
 bibliography: thesis.bib
+documentclass: sagej
+classoption: Royal, times, sageapa
 ---
@@
 
-# Introduction
+# Background
 
@@
 
-[^1]: All programming code used in this note is available from [github.com/gerkovink/shinyMice/simulation]{https://github.com/gerkovink/ShinyMICE/simulation}.
+[^1]: All programming code used in this note is available from [github.com/gerkovink/shinyMice](https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy).
 
@@
 
+### Theoretical Background
+
+There is no scientific consensus on the convergence properties of multiple imputation algorithms [@taka17].
+Some default techniques in `mice` might not yield converged states at all [@murr18].
+Therefore, algorithmic convergence should be monitored carefully.
+The diagnostic evaluation of convergence is preferred over the current practice---visually inspecting imputation chains for signs of non-convergence---since the latter may be challenging to the untrained eye [@buur18, \S~6.5.2].
+
+The application of convergence diagnostics to MI methods has not been systematically studied [@buur18].
+The measures that are available for diagnosing convergence of iterative algorithmic procedures (Markov chain Monte Carlo methods; e.g., `mice` algorithms or Gibbs samplers) generally evaluate signs of non-convergence [@hoff09].
+Non-convergence may be present as non-stationarity within chains (i.e., trending), or as slow mixing between chains (i.e., no intermingling).
+
+The stationarity component of convergence may be evaluated with autocorrelation [$AC$; @scha97; @gelm13], numeric standard error [or `MC error'; @gewe92], and Raftery and Lewis's [-@raft91] procedure to determine the effect of trending within chains.
+
+The mixing component can be assessed with the potential scale reduction factor $\widehat{R}$ [a.k.a.
+`Gelman-Rubin statistic'; @gelm92].
+With an adapted version of $\widehat{R}$, proposed by @veht19, we might also evaluate the stationarity component of convergence.
+This would make $\widehat{R}$ a general convergence diagnostic.
+The application of $\widehat{R}$ to assess stationarity has not been thoroughly investigated.
+Therefore, this study employs both $\widehat{R}$ and autocorrelation to investigate convergence, as recommended by [@cowl96, p.~898].
+
+The mixing component of convergence may be evaluated with the potential scale reduction factor, $\widehat{R}$ [a.k.a.
+`Gelman-Rubin statistic'; @gelm92].
+The stationarity component may be assessed through autocorrelations.
+An adapted version of $\widehat{R}$ might also be used to diagnose non-stationarity.
+Other convergence diagnostics are outside of the scope of this study.
+
+All of these methods evaluate the convergence of univariate scalar summaries (e.g., chain means or variances).
+These convergence diagnostics cannot diagnose convergence of multivariable statistics (i.e., relations between scalar summaries).
+@buur18 proposed to implement multivariable evaluation through eigenvalue decomposition [@mack03].
+This method is outside of the scope of the current study (**not anymore!**).
+
+To define $\widehat{R}$, we follow notation by [@veht19, p.~5].
+Let $M$ be the total number of chains, $T$ the number of iterations per chain, and $\theta$ the scalar summary of interest (e.g., chain mean or chain variance).
+For each chain ($m = 1, 2, \dots, M$), we estimate the variance of $\theta$, and average these to obtain within-chain variance $W$.
+
+\begin{align*}
+W&=\frac{1}{M} \sum_{m=1}^{M} s_{j}^{2},  \text { where } s_{m}^{2}=\frac{1}{T-1} \sum_{t=1}^{T}\left(\theta^{(t m)}-\bar{\theta}^{(\cdot m)}\right)^{2}. 
+<!-- %\text{ \cite[p.~5]{veht19}}  -->
+\end{align*}
+
+We then estimate between-chain variance $B$ as the variance of the collection of average $\theta$ per chain.
+
+\begin{align*}
+B&=\frac{T}{M-1} \sum_{m=1}^{M}\left(\bar{\theta}^{(\cdot m)}-\bar{\theta}^{(\cdot \cdot)}\right)^{2}, \text { where } \bar{\theta}^{(\cdot m)}=\frac{1}{T} \sum_{t=1}^{T} \theta^{(t m)}, \quad \bar{\theta}^{(\cdot \cdot)}=\frac{1}{M} \sum_{m=1}^{M} \bar{\theta}^{(\cdot m)}.
+<!-- %\text{ \cite[p.~5]{veht19}}  -->
+\end{align*}
+
+From the between- and within-chain variances we compute a weighted average, $\widehat{\operatorname{var}}^{+}$, which over-estimates the total variance of $\theta$.
+$\widehat{R}$ is then obtained as a ratio between the over-estimated total variance and the within-chain variance:
+
+\begin{equation*}
+\widehat{R}=\sqrt{\frac{\widehat{\operatorname{var}}^{+}(\theta | y)}{W}},
+\text{ where } \widehat{\operatorname{var}}^{+}(\theta | y)=\frac{N-1}{N} W+\frac{1}{N} B.
+\end{equation*}
+
+We can interpret $\widehat{R}$ as potential scale reduction factor since it indicates by how much the variance of $\theta$ could be shrunken down if an infinite number of iterations per chain would be run [@gelm92].
+This interpretation assumes that chains are 'over-dispersed' at $t=1$, and reach convergence as $T \to \infty$.
+Over-dispersion implies that the initial values of the chains are 'far away' from the target distribution and each other.
+When all chains sample independent of their initial values, the mixing component of convergence is satisfied, and $\widehat{R}$-values will be close to one.
+High $\widehat{R}$-values thus indicate non-convergence.
+The conventionally acceptable threshold for convergence was $\widehat{R} < 1.2$ [@gelm92].
+More recently, @veht19 proposed a more stringent threshold of $\widehat{R} < 1.01$.
+
+Following the same notation, we define autocorrelation as the correlation between two subsequent $\theta$-values within the same chain [@lync07, p.~147].
+In this study we only consider $AC$ at lag 1, i.e., the correlation between the $t^{th}$ and $t+1^{th}$ iteration of the same chain.
+
+\begin{equation*}
+AC = \left( \frac{T}{T-1} \right) \frac{\sum_{t=1}^{T-1}(\theta_t - \bar{\theta}^{(\cdot m)})(\theta_{t+1} - \bar{\theta}^{(\cdot m)})}{\sum_{t=1}^{T}(\theta_t - \bar{\theta}^{(\cdot m)})^2}.
+\end{equation*}
+
+We can interpret $AC$-values as a measure of stationarity.
+If $AC$-values are close to zero, there is no dependence between subsequent samples within imputation chains.
+Positive $AC$-values, however, indicate recurrence.
+If $\theta$-values of subsequent iterations are similar, trending may occur.
+Negative $AC$-values show no threat to the stationarity component of convergence.
+On the contrary even---negative $AC$-values indicate that $\theta$-values of subsequent iterations diverge from one-another, which may increase the variance of $\theta$ and speed up convergence.
+As convergence diagnostic, the interest is therefore in positive $AC$-values.
+Moreover, the magnitude of $AC$-values may be evaluated statistically, but that is outside of this note's scope.
+
+In short, convergence is reached when there is no dependency between subsequent iterations of imputation chains ($AC = 0$), and chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm ($\widehat{R} = 1$).
+
+### Simulation Hypothesis
+
+This study evaluates whether $\widehat{R}$ and $AC$ could diagnose convergence of multiple imputation algorithms.
+We assess the performance of the two convergence diagnostics against the recommended evaluation criteria for MI methods [i.e., average bias, average confidence interval width, and empirical coverage rate across simulations; @buur18, \S~2.5.2].
+That is, there is no baseline measure available to evaluate performance against.The current practice of visually inspecting imputation chains for signs of non-mixing or non-stationarity can only diagnose severely pathological cases of non-convergence.
+We could also look at distributional characteristics, and plausibility of imputed values, see @vinknd.
+For now, this is outside of the scope of this study.
+
+Based on an empirical finding [@lace07], we hypothesize that $\widehat{R}$ will over-estimate non-convergence of MI algorithms.
+The threshold of $\widehat{R} < 1.01$ will then be too stringent for diagnosing convergence.
+This over-estimation may, however, be diminished because $\widehat{R}$ can falsely diagnose convergence if initial values of the algorithm are not appropriately over-dispersed [@broo98, p~437].
+In `mice`, initial values are chosen randomly from the observed data.
+Therefore, we cannot be certain that the initial values are over-dispersed.
+We expect this to have little effect on the hypothesized performance of $\widehat{R}$.
+No hypothesis was formulated about the performance of $AC$ as convergence diagnostic.
+
 # Methods
+
+We use model-based simulation in `R` to perform multiple imputation under 100 simulation conditions.
+In each condition, the MI algorithm comprises of a different imputation chain length ($T = 1, 2, \dots, 100$).
+The number of simulation runs per condition is 1000.
+For each simulation condition (i.e., number of iterations) and each repetition (i.e., simulation run) we compute convergence diagnostics ($\widehat{R}$ and autocorrelation) and simulation diagnostics (bias, confidence interval width and coverage).
+These diagnostics are aggregated across repetitions to obtain their average value per simulation condition.
+The simulation set-up is summarized in the pseudo-code below.
+The complete `R` script of the simulation study is available from [github.com/gerkovink/shinyMice](https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy).
+
+```{r eval=FALSE}
+# pseudo-code of simulation 
+simulate data 
+for (number of simulation runs from 1 to 1000)
+  for (number of iterations from 1 to 100)
+    create missingness
+    impute the missingness
+    compute convergence diagnostics
+    perform analysis
+    pool results
+    compute simulation diagnostics
+aggregate convergence and simulation diagnostics
+```
+
+### Data Simulation
+
+The point of origin for all simulation conditions is a finite population of $N=1000$.
+The data are simulated to solve a multiple linear regression problem, where dependent variable $Y$ is regressed on independent variables $X_1$, $X_2$ and $X_3$:
+
+$$Y \sim \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3.$$
+
+The quantity of scientific interest is regression coefficient $\beta_1$.
+The data generating model is a multivariate normal distribution with the following means structure and variance-covariance matrix:
+
+\begin{align*}
+\begin{pmatrix}X_1\\
+X_2\\
+X_3\\
+\epsilon
+\end{pmatrix} \sim  N
+\begin{bmatrix}
+\begin{pmatrix}
+12\\
+3\\
+0.5\\
+0
+\end{pmatrix}\!\!,
+\begin{pmatrix}
+4 & 4 & 1.8 & 0\\
+4 & 16 & 4.8 & 0\\
+1.8 & 4.8 & 9 & 0\\
+0 & 0 & 0 & 100
+\end{pmatrix}
+\end{bmatrix}\\[2\jot].
+\end{align*}
+
+Data is simulated with the function `mvtnorm::rmvnorm()`, and outcome variable $Y$ is subsequently calculated as 
+$$Y =  2X_1 + .5X_2 - X_3 + \epsilon .$$
+The complete data is 'amputed' once for each simulation repetition.
+That is, the function `mice::ampute()` is used to impose a missingness mechanism upon the data.
+The missingness is univariate, and the probability to be missing is the same for all four variables, namely 20\% (`prop = 0.8, mech = "MCAR"`).
+This leaves 20\% of the rows completely observed.
+%The resulting amputed data is equal for all simulation conditions in the same repetition.
+
+Missing datapoints are imputed with the function `mice::mice()`.
+All MI procedures are performed with Bayesian linear regression imputation (`method = "norm"`), and five imputation chains (`m = 5`).
+The number of iterations varies between simulation conditions (`maxit = 1, 2, \dots, 100`).
+
+From the imputation step, we extract chain means and chain variances across iterations ($\theta$ at $t = 1,2,\dots,T$).  
+We compute convergence diagnostics separately for both $\theta$s, and for each of the four variables in the dataset.
+We report the maximum (absolute) value as the 'worst linear predictor' of convergence.
+$\widehat{R}$ is computed by implementing Vehtari et al.'s [-@veht19] recommendations.
+$AC$ is computed with function `stats::acf()`.
+
+To estimate the quantity of scientific interest, $Q$, we perform multiple linear regression on each completed dataset with the function `stats::lm()`.
+We obtain an estimated regression coefficient per imputation, which are pooled into a single estimate, $\bar{Q}$.
+We use the function `mice::pool()` to get variance estimates according to Rubin's [-@rubin87] rules, and subsequently implement finite population pooling conform @vink14.
+
+We compute bias as the difference between $\bar{Q}$ and $Q$.
+Confidence interval width (CIW) is defined as the difference between the lower and upper bound of the 95\% confidence interval (CI95\%) around $\bar{Q}$.
+We compute the CI95\% bounds as
+
+$$\bar{Q} \pm t_{(m-1)} \times SE_{\bar{Q}},$$
+
+where $t_{(m-1)}$ is the quantile of a $t$-distribution with $m-1$ degrees of freedom, and $SE_{\bar{Q}}$ is the square root of the pooled variance estimate.
+From bias and CIW, we calculate empirical coverage rates.
+Coverage rate is the proportion of simulations in which $Q$ is between the bounds of the CI95\% around $\bar{Q}$.
 
@@
 
+Figures 2 and 3 display results per simulation condition ($T = 1,2,\dots,100$).
+
+### Convergence Diagnostics
+
+It is apparent that there is a relation between the number of iterations per simulation condition ($T$) and the convergence diagnostics.
+Generally speaking, conditions with longer imputation chains (higher $T$) coincide with less signs of non-convergence ($\widehat{R}$-values approach one, and $AC$-values approach zero).
+We see more or less equivalent trends for chain means versus chain variances.
+We, therefore, only discuss the $\widehat{R}$ and $AC$-values of chain means.
+
+Figure 2A shows that $\widehat{R}$-values generally decrease with increasing imputation chain lengths.
+The decline stabilizes somewhere between the simulation conditions $T=30$ and $T=50$.
+The downward trend is most pronounced for $T=3$, and between $T = 5$ and $T = 10$.
+In the intervening conditions ($3 \leq T \leq 5$), however, we observe a steep increase in $\widehat{R}$-values.
+This increase implies that non-convergence is under-estimated, and convergence should not be diagnosed at this point.
+Based on the conventional threshold $\widehat{R} < 1.2$, we would falsely diagnose convergence at $T=3$.
+According to the widely used threshold $\widehat{R} < 1.1$, convergence would be diagnosed for conditions where $T>9$.
+If we use the recently recommended threshold $\widehat{R} < 1.01$, we would conclude that convergence is reached in none of the simulation conditions.
+
+The $AC$-values displayed in Figure 2B are almost uniformly increasing as a function of $T$.
+The only $AC$-value that deviates from this observation is for $T=2$.
+The lowest $AC$-value is obtained for $T=3$.
+%At $T=5$, the $AC$-value reaches the level observed at $T=2$.
+The gradual increase plateaus between $T=10$ and $T=30$.
+$AC$-values in conditions where $T>70$ are indifferentiable from zero, indicating stationarity.
+We see an initial decrease between $T=2$ and $T=3$.
+Simulation conditions where $T>5$ have $AC$-values greater than $T=2$.
+
+According to this diagnostic, none of the simulation conditions show signs of non-convergence, since we only observe negative or zero $AC$-values.
+The negative $AC$-values, however, indicate increasing divergence between subsequent samples in conditions where $T<4$.
+It may not be wise to terminating the algorithm at that point, but after `recovering' from this dip in $AC$-values.
+This would imply $T=6$ as the minimal number of iterations.
+
+Taken together, we see that $T>3$ is the minimal requirement to diagnose convergence ($\widehat{R} < 1.2; AC \leq 0$).
+This threshold is, however, not sufficient, since we overlook the increase in $\widehat{R}$-values up-to conditions where $T>5$, and the convergence diagnostics only reach stability at $T>20$.
+
+### Simulation Diagnostics
+
+We use average bias, average confidence interval width, and coverage rate as performance measures to evaluate $\widehat{R}$ and $AC$.
+We make the general observation that the simulation diagnostics behave as theorized for most simulation conditions (bias around zero, stable confidence interval widths, nominal coverage rate at 95\%).
+
+Figure 3A shows that bias is fairly stable across simulation conditions.
+The condition $T=1$ clearly deviates from this trend with a negative bias.
+The bias for $T=2$ is below average, but within the range of fluctuations.
+Conditions where $T>3$ can be diagnosed as unbiased, but the average bias across iterations (as shown by a flat Loess line) only reaches stability at $T=20$.
+
+CIW is only clearly divergent in the simulation condition $T=1$, see Figure 3B.
+However, under-estimating the variance of $\bar{Q}$, which is the case for $T=3$, may yield spurious inferences.
+Only conditions where $T>3$ are therefore considered sufficient.
+These conditions have similar CIWs, but across iterations stability is not established (i.e., the Loess line is never completely flat within the 100 simulation conditions).
+%After about twenty-five iterations the average CIW across all simulation conditions and repetitions is reached 0.927.
+
+The empirical coverage rate across repetitions seems more or less stable for conditions where $T>1$, see Figure 3C.
+We see some over-coverage at $T=2$, but that is better than under-estimating the variance of $\bar{Q}$.
+On average, the coverage rate is somewhat higher than the expected nominal coverage of 95\% [@neym34], namely 95\%.
+Similar to the CIWs, there is some trending across iterations (i.e., the Loess line is never flat).
+
+From the simulation diagnostics, we observe that unbiased estimates with nominal coverage rates were obtained in conditions where $T>3$.
+This suggests that as little as four iterations may be sufficient for the MI algorithm under the current circumstances.
+
+In short, there is a discrepancy between what the convergence diagnostics and what the performance measures indicate.
+While $T=4$ seems sufficient with respect to simulation quantities, the condition where $T=4$ resulted in the `one-but-worst' values of both convergence diagnostics.
+Complete algorithmic convergence as indicated by $\widehat{R}$ and autocorrelation is not reached in conditions where $T<20$.
+
 # Discussion
 
+This note shows that convergence diagnostics $\widehat{R}$ and $AC$ may diagnose convergence of multiple imputation algorithms, but their performance differs from conventional applications to iterative algorithmic procedures.
+$\widehat{R}$ and autocorrelation indicate that algorithmic convergence may only be reached after twenty or even forty iterations, while unbiased, confidence valid estimates estimates may be obtained with as little as four iterations.
+These results are in agreement with the simulation hypothesis: $\widehat{R}$ over-estimates the severity of non-convergence when applied to MI procedures.
+%This may be due to the quantity of scientific interest chosen.
+More `complicated' $Q$s (e.g., higher order effects or variance components) might show bias, under- or over-coverage at higher $T$.
+
+According to this simulation study, the recently proposed threshold of $\widehat{R}<1.01$ may be too stringent for MI algorithms.
+Under the relatively easy missing data problem of the current study, the threshold was not reached.
+The other extreme of the $\widehat{R}$-thresholds, the conventionally acceptable $\widehat{R} <1.2$, may be too lenient for MI procedures.
+Applying this threshold to the current data, lead to falsely diagnosing convergence at $T = 3$.
+It appears that the widely used threshold of $\widehat{R} < 1.1$ suits MI algorithms the best.
+We might, however, also formulate a new threshold, specifically for the evaluation of MI algorithms.
+The current study suggests that $\widehat{R} < 1.05$ may be implemented, since that is the level at which the $\widehat{R}$ stabilize (around $T = 20$).
+
+The negative $AC$-values obtained in this study show no threat of non-stationarity.
+However, initial dip in $AC$-values may have implications for the default number of iterations in `mice` (`maxit = 5`).
+Terminating the algorithm at $T=5$ may not be the most appropriate, since this lead to the worst convergence, as indicated by $\widehat{R}$ and $AC$.
+Under the current specifications, $T>20$ would be more appropriate.
+
+The observed dip in AC implies that default maxit value of five iterations is the worst possible number of iterations.
+Moreover, the results of this study imply that assessing the stationarity component of convergence with $AC$ might be redundant, since high $AC$-values are implausible in MI procedures.
+That is, the randomness induced by the MI algorithm effectively mitigates the risk of dependency within chains.
+$AC$ would thus not be informative of the convergence of MI algorithms.
+
+Since this study only considers only an MCAR missingness mechanism, results may not be extrapolated to other missing data problems.
+Proper performance of the convergence diagnostics under MCAR is necessary but not sufficient to demonstrate appropriateness of $\widehat{R}$ and $AC$ as convergence diagnostics.
+This is just a proof of concept.
+
+Further research is needed to investigate their performance under clear violation of convergence, e.g.
+dependency between predictors (predictors with very high correlations).
+Until then, we have only shown that the convergence diagnostics can diagnose non-convergence of MI algorithms that trend towards a converged state.
+Also for future research, implement Wernicke diagnostic, and look at developing a convergence diagnostic for substantive models, and implement a Wald test for $AC = 0$.
+
 # References

### 2020-03-17 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Adjust latex style to suit sociological methods and research journal

@@
 \documentclass[
-]{article}
+  Royal, times, sageapa]{sagej}
 \usepackage{lmodern}
@@
   pdftitle={Missing the Point: Convergence of Multiple Imputation using Chained Equations Algorithms},
-  pdfauthor={Hanne Oberman},
+  pdfauthor={Hanne I. Oberman},
   hidelinks,
@@
 \urlstyle{same} % disable monospaced font for URLs
-\usepackage[margin=1in]{geometry}
+\usepackage{color}
+\usepackage{fancyvrb}
+\newcommand{\VerbBar}{|}
+\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
+\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
+% Add ',fontsize=\small' for more characters per line
+\usepackage{framed}
+\definecolor{shadecolor}{RGB}{248,248,248}
+\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\BuiltInTok}[1]{#1}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
+\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
+\newcommand{\ExtensionTok}[1]{#1}
+\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ImportTok}[1]{#1}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\NormalTok}[1]{#1}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\RegionMarkerTok}[1]{#1}
+\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
 \usepackage{graphicx,grffile}
@@
 Equations Algorithms}
-\author{Hanne Oberman}
+\author{Hanne I. Oberman}
 \date{16-3-2020}
@@
 
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
+\hypertarget{background}{%
+\section{Background}\label{background}}
 
@@
   programming code used in this note is available from
-  {[}github.com/gerkovink/shinyMice/simulation{]}\{\url{https://github.com/gerkovink/ShinyMICE/simulation}\}.}
+  \href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy}{github.com/gerkovink/shinyMice}.}
 The results of this simulation study are guidelines for assessing
@@
 
+\hypertarget{theoretical-background}{%
+\subsubsection{Theoretical Background}\label{theoretical-background}}
+
+There is no scientific consensus on the convergence properties of
+multiple imputation algorithms (Takahashi 2017). Some default techniques
+in \texttt{mice} might not yield converged states at all (Murray 2018).
+Therefore, algorithmic convergence should be monitored carefully. The
+diagnostic evaluation of convergence is preferred over the current
+practice---visually inspecting imputation chains for signs of
+non-convergence---since the latter may be challenging to the untrained
+eye (Van Buuren 2018, \textasciitilde6.5.2).
+
+The application of convergence diagnostics to MI methods has not been
+systematically studied (Van Buuren 2018). The measures that are
+available for diagnosing convergence of iterative algorithmic procedures
+(Markov chain Monte Carlo methods; e.g., \texttt{mice} algorithms or
+Gibbs samplers) generally evaluate signs of non-convergence (Hoff 2009).
+Non-convergence may be present as non-stationarity within chains (i.e.,
+trending), or as slow mixing between chains (i.e., no intermingling).
+
+The stationarity component of convergence may be evaluated with
+autocorrelation (\(AC\); Schafer 1997; Gelman et al. 2013), numeric
+standard error (or `MC error'; Geweke 1992), and Raftery and Lewis's
+(1991) procedure to determine the effect of trending within chains.
+
+The mixing component can be assessed with the potential scale reduction
+factor \(\widehat{R}\) (a.k.a. `Gelman-Rubin statistic'; Gelman and
+Rubin 1992). With an adapted version of \(\widehat{R}\), proposed by
+Vehtari et al. (2019), we might also evaluate the stationarity component
+of convergence. This would make \(\widehat{R}\) a general convergence
+diagnostic. The application of \(\widehat{R}\) to assess stationarity
+has not been thoroughly investigated. Therefore, this study employs both
+\(\widehat{R}\) and autocorrelation to investigate convergence, as
+recommended by (Cowles and Carlin 1996, \textasciitilde898).
+
+The mixing component of convergence may be evaluated with the potential
+scale reduction factor, \(\widehat{R}\) (a.k.a. `Gelman-Rubin
+statistic'; Gelman and Rubin 1992). The stationarity component may be
+assessed through autocorrelations. An adapted version of \(\widehat{R}\)
+might also be used to diagnose non-stationarity. Other convergence
+diagnostics are outside of the scope of this study.
+
+All of these methods evaluate the convergence of univariate scalar
+summaries (e.g., chain means or variances). These convergence
+diagnostics cannot diagnose convergence of multivariable statistics
+(i.e., relations between scalar summaries). Van Buuren (2018) proposed
+to implement multivariable evaluation through eigenvalue decomposition
+(MacKay and Mac Kay 2003). This method is outside of the scope of the
+current study (\textbf{not anymore!}).
+
+To define \(\widehat{R}\), we follow notation by (Vehtari et al. 2019,
+\textasciitilde5). Let \(M\) be the total number of chains, \(T\) the
+number of iterations per chain, and \(\theta\) the scalar summary of
+interest (e.g., chain mean or chain variance). For each chain
+(\(m = 1, 2, \dots, M\)), we estimate the variance of \(\theta\), and
+average these to obtain within-chain variance \(W\).
+
+\begin{align*}
+W&=\frac{1}{M} \sum_{m=1}^{M} s_{j}^{2},  \text { where } s_{m}^{2}=\frac{1}{T-1} \sum_{t=1}^{T}\left(\theta^{(t m)}-\bar{\theta}^{(\cdot m)}\right)^{2}. 
+<!-- %\text{ \cite[p.~5]{veht19}}  -->
+\end{align*}
+
+We then estimate between-chain variance \(B\) as the variance of the
+collection of average \(\theta\) per chain.
+
+\begin{align*}
+B&=\frac{T}{M-1} \sum_{m=1}^{M}\left(\bar{\theta}^{(\cdot m)}-\bar{\theta}^{(\cdot \cdot)}\right)^{2}, \text { where } \bar{\theta}^{(\cdot m)}=\frac{1}{T} \sum_{t=1}^{T} \theta^{(t m)}, \quad \bar{\theta}^{(\cdot \cdot)}=\frac{1}{M} \sum_{m=1}^{M} \bar{\theta}^{(\cdot m)}.
+<!-- %\text{ \cite[p.~5]{veht19}}  -->
+\end{align*}
+
+From the between- and within-chain variances we compute a weighted
+average, \(\widehat{\operatorname{var}}^{+}\), which over-estimates the
+total variance of \(\theta\). \(\widehat{R}\) is then obtained as a
+ratio between the over-estimated total variance and the within-chain
+variance:
+
+\begin{equation*}
+\widehat{R}=\sqrt{\frac{\widehat{\operatorname{var}}^{+}(\theta | y)}{W}},
+\text{ where } \widehat{\operatorname{var}}^{+}(\theta | y)=\frac{N-1}{N} W+\frac{1}{N} B.
+\end{equation*}
+
+We can interpret \(\widehat{R}\) as potential scale reduction factor
+since it indicates by how much the variance of \(\theta\) could be
+shrunken down if an infinite number of iterations per chain would be run
+(Gelman and Rubin 1992). This interpretation assumes that chains are
+`over-dispersed' at \(t=1\), and reach convergence as \(T \to \infty\).
+Over-dispersion implies that the initial values of the chains are `far
+away' from the target distribution and each other. When all chains
+sample independent of their initial values, the mixing component of
+convergence is satisfied, and \(\widehat{R}\)-values will be close to
+one. High \(\widehat{R}\)-values thus indicate non-convergence. The
+conventionally acceptable threshold for convergence was
+\(\widehat{R} < 1.2\) (Gelman and Rubin 1992). More recently, Vehtari et
+al. (2019) proposed a more stringent threshold of
+\(\widehat{R} < 1.01\).
+
+Following the same notation, we define autocorrelation as the
+correlation between two subsequent \(\theta\)-values within the same
+chain (Lynch 2007, \textasciitilde147). In this study we only consider
+\(AC\) at lag 1, i.e., the correlation between the \(t^{th}\) and
+\(t+1^{th}\) iteration of the same chain.
+
+\begin{equation*}
+AC = \left( \frac{T}{T-1} \right) \frac{\sum_{t=1}^{T-1}(\theta_t - \bar{\theta}^{(\cdot m)})(\theta_{t+1} - \bar{\theta}^{(\cdot m)})}{\sum_{t=1}^{T}(\theta_t - \bar{\theta}^{(\cdot m)})^2}.
+\end{equation*}
+
+We can interpret \(AC\)-values as a measure of stationarity. If
+\(AC\)-values are close to zero, there is no dependence between
+subsequent samples within imputation chains. Positive \(AC\)-values,
+however, indicate recurrence. If \(\theta\)-values of subsequent
+iterations are similar, trending may occur. Negative \(AC\)-values show
+no threat to the stationarity component of convergence. On the contrary
+even---negative \(AC\)-values indicate that \(\theta\)-values of
+subsequent iterations diverge from one-another, which may increase the
+variance of \(\theta\) and speed up convergence. As convergence
+diagnostic, the interest is therefore in positive \(AC\)-values.
+Moreover, the magnitude of \(AC\)-values may be evaluated statistically,
+but that is outside of this note's scope.
+
+In short, convergence is reached when there is no dependency between
+subsequent iterations of imputation chains (\(AC = 0\)), and chains
+intermingle such that the only difference between the chains is caused
+by the randomness induced by the algorithm (\(\widehat{R} = 1\)).
+
+\hypertarget{simulation-hypothesis}{%
+\subsubsection{Simulation Hypothesis}\label{simulation-hypothesis}}
+
+This study evaluates whether \(\widehat{R}\) and \(AC\) could diagnose
+convergence of multiple imputation algorithms. We assess the performance
+of the two convergence diagnostics against the recommended evaluation
+criteria for MI methods (i.e., average bias, average confidence interval
+width, and empirical coverage rate across simulations; Van Buuren 2018,
+\textasciitilde2.5.2). That is, there is no baseline measure available
+to evaluate performance against.The current practice of visually
+inspecting imputation chains for signs of non-mixing or non-stationarity
+can only diagnose severely pathological cases of non-convergence. We
+could also look at distributional characteristics, and plausibility of
+imputed values, see Vink (n.d.). For now, this is outside of the scope
+of this study.
+
+Based on an empirical finding (Lacerda, Ardington, and Leibbrandt 2007),
+we hypothesize that \(\widehat{R}\) will over-estimate non-convergence
+of MI algorithms. The threshold of \(\widehat{R} < 1.01\) will then be
+too stringent for diagnosing convergence. This over-estimation may,
+however, be diminished because \(\widehat{R}\) can falsely diagnose
+convergence if initial values of the algorithm are not appropriately
+over-dispersed (Brooks and Gelman 1998, p\textasciitilde437). In
+\texttt{mice}, initial values are chosen randomly from the observed
+data. Therefore, we cannot be certain that the initial values are
+over-dispersed. We expect this to have little effect on the hypothesized
+performance of \(\widehat{R}\). No hypothesis was formulated about the
+performance of \(AC\) as convergence diagnostic.
+
 \hypertarget{methods}{%
 \section{Methods}\label{methods}}
+
+We use model-based simulation in \texttt{R} to perform multiple
+imputation under 100 simulation conditions. In each condition, the MI
+algorithm comprises of a different imputation chain length
+(\(T = 1, 2, \dots, 100\)). The number of simulation runs per condition
+is 1000. For each simulation condition (i.e., number of iterations) and
+each repetition (i.e., simulation run) we compute convergence
+diagnostics (\(\widehat{R}\) and autocorrelation) and simulation
+diagnostics (bias, confidence interval width and coverage). These
+diagnostics are aggregated across repetitions to obtain their average
+value per simulation condition. The simulation set-up is summarized in
+the pseudo-code below. The complete \texttt{R} script of the simulation
+study is available from
+\href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy}{github.com/gerkovink/shinyMice}.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{# pseudo-code of simulation }
+\NormalTok{simulate data }
+\ControlFlowTok{for}\NormalTok{ (number of simulation runs from }\DecValTok{1}\NormalTok{ to }\DecValTok{1000}\NormalTok{)}
+  \ControlFlowTok{for}\NormalTok{ (number of iterations from }\DecValTok{1}\NormalTok{ to }\DecValTok{100}\NormalTok{)}
+\NormalTok{    create missingness}
+\NormalTok{    impute the missingness}
+\NormalTok{    compute convergence diagnostics}
+\NormalTok{    perform analysis}
+\NormalTok{    pool results}
+\NormalTok{    compute simulation diagnostics}
+\NormalTok{aggregate convergence and simulation diagnostics}
+\end{Highlighting}
+\end{Shaded}
+
+\hypertarget{data-simulation}{%
+\subsubsection{Data Simulation}\label{data-simulation}}
+
+The point of origin for all simulation conditions is a finite population
+of \(N=1000\). The data are simulated to solve a multiple linear
+regression problem, where dependent variable \(Y\) is regressed on
+independent variables \(X_1\), \(X_2\) and \(X_3\):
+
+\[Y \sim \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3.\]
+
+The quantity of scientific interest is regression coefficient
+\(\beta_1\). The data generating model is a multivariate normal
+distribution with the following means structure and variance-covariance
+matrix:
+
+\begin{align*}
+\begin{pmatrix}X_1\\
+X_2\\
+X_3\\
+\epsilon
+\end{pmatrix} \sim  N
+\begin{bmatrix}
+\begin{pmatrix}
+12\\
+3\\
+0.5\\
+0
+\end{pmatrix}\!\!,
+\begin{pmatrix}
+4 & 4 & 1.8 & 0\\
+4 & 16 & 4.8 & 0\\
+1.8 & 4.8 & 9 & 0\\
+0 & 0 & 0 & 100
+\end{pmatrix}
+\end{bmatrix}\\[2\jot].
+\end{align*}
+
+Data is simulated with the function \texttt{mvtnorm::rmvnorm()}, and
+outcome variable \(Y\) is subsequently calculated as
+\[Y =  2X_1 + .5X_2 - X_3 + \epsilon .\] The complete data is `amputed'
+once for each simulation repetition. That is, the function
+\texttt{mice::ampute()} is used to impose a missingness mechanism upon
+the data. The missingness is univariate, and the probability to be
+missing is the same for all four variables, namely 20\%
+(\texttt{prop\ =\ 0.8,\ mech\ =\ "MCAR"}). This leaves 20\% of the rows
+completely observed. \%The resulting amputed data is equal for all
+simulation conditions in the same repetition.
+
+Missing datapoints are imputed with the function \texttt{mice::mice()}.
+All MI procedures are performed with Bayesian linear regression
+imputation (\texttt{method\ =\ "norm"}), and five imputation chains
+(\texttt{m\ =\ 5}). The number of iterations varies between simulation
+conditions (\texttt{maxit\ =\ 1,\ 2,\ \textbackslash{}dots,\ 100}).
+
+From the imputation step, we extract chain means and chain variances
+across iterations (\(\theta\) at \(t = 1,2,\dots,T\)).\\
+We compute convergence diagnostics separately for both \(\theta\)s, and
+for each of the four variables in the dataset. We report the maximum
+(absolute) value as the `worst linear predictor' of convergence.
+\(\widehat{R}\) is computed by implementing Vehtari et al.'s (2019)
+recommendations. \(AC\) is computed with function \texttt{stats::acf()}.
+
+To estimate the quantity of scientific interest, \(Q\), we perform
+multiple linear regression on each completed dataset with the function
+\texttt{stats::lm()}. We obtain an estimated regression coefficient per
+imputation, which are pooled into a single estimate, \(\bar{Q}\). We use
+the function \texttt{mice::pool()} to get variance estimates according
+to Rubin's (1987) rules, and subsequently implement finite population
+pooling conform Vink and Buuren (2014).
+
+We compute bias as the difference between \(\bar{Q}\) and \(Q\).
+Confidence interval width (CIW) is defined as the difference between the
+lower and upper bound of the 95\% confidence interval (CI95\%) around
+\(\bar{Q}\). We compute the CI95\% bounds as
+
+\[\bar{Q} \pm t_{(m-1)} \times SE_{\bar{Q}},\]
+
+where \(t_{(m-1)}\) is the quantile of a \(t\)-distribution with \(m-1\)
+degrees of freedom, and \(SE_{\bar{Q}}\) is the square root of the
+pooled variance estimate. From bias and CIW, we calculate empirical
+coverage rates. Coverage rate is the proportion of simulations in which
+\(Q\) is between the bounds of the CI95\% around \(\bar{Q}\).
 
@@
 
+Figures 2 and 3 display results per simulation condition
+(\(T = 1,2,\dots,100\)).
+
+\hypertarget{convergence-diagnostics}{%
+\subsubsection{Convergence Diagnostics}\label{convergence-diagnostics}}
+
+It is apparent that there is a relation between the number of iterations
+per simulation condition (\(T\)) and the convergence diagnostics.
+Generally speaking, conditions with longer imputation chains (higher
+\(T\)) coincide with less signs of non-convergence
+(\(\widehat{R}\)-values approach one, and \(AC\)-values approach zero).
+We see more or less equivalent trends for chain means versus chain
+variances. We, therefore, only discuss the \(\widehat{R}\) and
+\(AC\)-values of chain means.
+
+Figure 2A shows that \(\widehat{R}\)-values generally decrease with
+increasing imputation chain lengths. The decline stabilizes somewhere
+between the simulation conditions \(T=30\) and \(T=50\). The downward
+trend is most pronounced for \(T=3\), and between \(T = 5\) and
+\(T = 10\). In the intervening conditions (\(3 \leq T \leq 5\)),
+however, we observe a steep increase in \(\widehat{R}\)-values. This
+increase implies that non-convergence is under-estimated, and
+convergence should not be diagnosed at this point. Based on the
+conventional threshold \(\widehat{R} < 1.2\), we would falsely diagnose
+convergence at \(T=3\). According to the widely used threshold
+\(\widehat{R} < 1.1\), convergence would be diagnosed for conditions
+where \(T>9\). If we use the recently recommended threshold
+\(\widehat{R} < 1.01\), we would conclude that convergence is reached in
+none of the simulation conditions.
+
+The \(AC\)-values displayed in Figure 2B are almost uniformly increasing
+as a function of \(T\). The only \(AC\)-value that deviates from this
+observation is for \(T=2\). The lowest \(AC\)-value is obtained for
+\(T=3\). \%At \(T=5\), the \(AC\)-value reaches the level observed at
+\(T=2\). The gradual increase plateaus between \(T=10\) and \(T=30\).
+\(AC\)-values in conditions where \(T>70\) are indifferentiable from
+zero, indicating stationarity. We see an initial decrease between
+\(T=2\) and \(T=3\). Simulation conditions where \(T>5\) have
+\(AC\)-values greater than \(T=2\).
+
+According to this diagnostic, none of the simulation conditions show
+signs of non-convergence, since we only observe negative or zero
+\(AC\)-values. The negative \(AC\)-values, however, indicate increasing
+divergence between subsequent samples in conditions where \(T<4\). It
+may not be wise to terminating the algorithm at that point, but after
+`recovering' from this dip in \(AC\)-values. This would imply \(T=6\) as
+the minimal number of iterations.
+
+Taken together, we see that \(T>3\) is the minimal requirement to
+diagnose convergence (\(\widehat{R} < 1.2; AC \leq 0\)). This threshold
+is, however, not sufficient, since we overlook the increase in
+\(\widehat{R}\)-values up-to conditions where \(T>5\), and the
+convergence diagnostics only reach stability at \(T>20\).
+
+\hypertarget{simulation-diagnostics}{%
+\subsubsection{Simulation Diagnostics}\label{simulation-diagnostics}}
+
+We use average bias, average confidence interval width, and coverage
+rate as performance measures to evaluate \(\widehat{R}\) and \(AC\). We
+make the general observation that the simulation diagnostics behave as
+theorized for most simulation conditions (bias around zero, stable
+confidence interval widths, nominal coverage rate at 95\%).
+
+Figure 3A shows that bias is fairly stable across simulation conditions.
+The condition \(T=1\) clearly deviates from this trend with a negative
+bias. The bias for \(T=2\) is below average, but within the range of
+fluctuations. Conditions where \(T>3\) can be diagnosed as unbiased, but
+the average bias across iterations (as shown by a flat Loess line) only
+reaches stability at \(T=20\).
+
+CIW is only clearly divergent in the simulation condition \(T=1\), see
+Figure 3B. However, under-estimating the variance of \(\bar{Q}\), which
+is the case for \(T=3\), may yield spurious inferences. Only conditions
+where \(T>3\) are therefore considered sufficient. These conditions have
+similar CIWs, but across iterations stability is not established (i.e.,
+the Loess line is never completely flat within the 100 simulation
+conditions). \%After about twenty-five iterations the average CIW across
+all simulation conditions and repetitions is reached 0.927.
+
+The empirical coverage rate across repetitions seems more or less stable
+for conditions where \(T>1\), see Figure 3C. We see some over-coverage
+at \(T=2\), but that is better than under-estimating the variance of
+\(\bar{Q}\). On average, the coverage rate is somewhat higher than the
+expected nominal coverage of 95\% (Neyman 1934), namely 95\%. Similar to
+the CIWs, there is some trending across iterations (i.e., the Loess line
+is never flat).
+
+From the simulation diagnostics, we observe that unbiased estimates with
+nominal coverage rates were obtained in conditions where \(T>3\). This
+suggests that as little as four iterations may be sufficient for the MI
+algorithm under the current circumstances.
+
+In short, there is a discrepancy between what the convergence
+diagnostics and what the performance measures indicate. While \(T=4\)
+seems sufficient with respect to simulation quantities, the condition
+where \(T=4\) resulted in the `one-but-worst' values of both convergence
+diagnostics. Complete algorithmic convergence as indicated by
+\(\widehat{R}\) and autocorrelation is not reached in conditions where
+\(T<20\).
+
 \hypertarget{discussion}{%
 \section{Discussion}\label{discussion}}
+
+This note shows that convergence diagnostics \(\widehat{R}\) and \(AC\)
+may diagnose convergence of multiple imputation algorithms, but their
+performance differs from conventional applications to iterative
+algorithmic procedures.\\
+\(\widehat{R}\) and autocorrelation indicate that algorithmic
+convergence may only be reached after twenty or even forty iterations,
+while unbiased, confidence valid estimates estimates may be obtained
+with as little as four iterations. These results are in agreement with
+the simulation hypothesis: \(\widehat{R}\) over-estimates the severity
+of non-convergence when applied to MI procedures. \%This may be due to
+the quantity of scientific interest chosen. More `complicated' \(Q\)s
+(e.g., higher order effects or variance components) might show bias,
+under- or over-coverage at higher \(T\).
+
+According to this simulation study, the recently proposed threshold of
+\(\widehat{R}<1.01\) may be too stringent for MI algorithms. Under the
+relatively easy missing data problem of the current study, the threshold
+was not reached. The other extreme of the \(\widehat{R}\)-thresholds,
+the conventionally acceptable \(\widehat{R} <1.2\), may be too lenient
+for MI procedures. Applying this threshold to the current data, lead to
+falsely diagnosing convergence at \(T = 3\). It appears that the widely
+used threshold of \(\widehat{R} < 1.1\) suits MI algorithms the best. We
+might, however, also formulate a new threshold, specifically for the
+evaluation of MI algorithms. The current study suggests that
+\(\widehat{R} < 1.05\) may be implemented, since that is the level at
+which the \(\widehat{R}\) stabilize (around \(T = 20\)).
+
+The negative \(AC\)-values obtained in this study show no threat of
+non-stationarity. However, initial dip in \(AC\)-values may have
+implications for the default number of iterations in \texttt{mice}
+(\texttt{maxit\ =\ 5}). Terminating the algorithm at \(T=5\) may not be
+the most appropriate, since this lead to the worst convergence, as
+indicated by \(\widehat{R}\) and \(AC\). Under the current
+specifications, \(T>20\) would be more appropriate.
+
+The observed dip in AC implies that default maxit value of five
+iterations is the worst possible number of iterations. Moreover, the
+results of this study imply that assessing the stationarity component of
+convergence with \(AC\) might be redundant, since high \(AC\)-values are
+implausible in MI procedures. That is, the randomness induced by the MI
+algorithm effectively mitigates the risk of dependency within chains.
+\(AC\) would thus not be informative of the convergence of MI
+algorithms.
+
+Since this study only considers only an MCAR missingness mechanism,
+results may not be extrapolated to other missing data problems. Proper
+performance of the convergence diagnostics under MCAR is necessary but
+not sufficient to demonstrate appropriateness of \(\widehat{R}\) and
+\(AC\) as convergence diagnostics. This is just a proof of concept.
+
+Further research is needed to investigate their performance under clear
+violation of convergence, e.g.~dependency between predictors (predictors
+with very high correlations). Until then, we have only shown that the
+convergence diagnostics can diagnose non-convergence of MI algorithms
+that trend towards a converged state. Also for future research,
+implement Wernicke diagnostic, and look at developing a convergence
+diagnostic for substantive models, and implement a Wald test for
+\(AC = 0\).
 
@@
 
+\leavevmode\hypertarget{ref-broo98}{}%
+Brooks, Stephen P., and Andrew Gelman. 1998. ``General Methods for
+Monitoring Convergence of Iterative Simulations.'' \emph{Journal of
+Computational and Graphical Statistics} 7 (4): 434--55.
+\url{https://doi.org/10.1080/10618600.1998.10474787}.
+
+\leavevmode\hypertarget{ref-cowl96}{}%
+Cowles, Mary Kathryn, and Bradley P Carlin. 1996. ``Markov Chain Monte
+Carlo Convergence Diagnostics: A Comparative Review.'' \emph{Journal of
+the American Statistical Association} 91 (434): 883--904.
+
+\leavevmode\hypertarget{ref-gelm13}{}%
+Gelman, Andrew, John B. Carlin, Hal S. Stern, David B. Dunson, Aki
+Vehtari, and Donald B. Rubin. 2013. \emph{Bayesian Data Analysis}.
+Philadelphia, PA, United States: CRC Press LLC.
+
+\leavevmode\hypertarget{ref-gelm92}{}%
+Gelman, Andrew, and Donald B. Rubin. 1992. ``Inference from Iterative
+Simulation Using Multiple Sequences.'' \emph{Statistical Science} 7 (4):
+457--72. \url{https://doi.org/10.1214/ss/1177011136}.
+
+\leavevmode\hypertarget{ref-gewe92}{}%
+Geweke, John. 1992. ``Evaluating the Accuracy of Sampling-Based
+Approaches to the Calculations of Posterior Moments.'' \emph{Bayesian
+Statistics} 4: 641--49.
+
+\leavevmode\hypertarget{ref-hoff09}{}%
+Hoff, Peter D. 2009. \emph{A First Course in Bayesian Statistical
+Methods}. Springer Texts in Statistics. New York, NY: Springer New York.
+\url{https://doi.org/10.1007/978-0-387-92407-6}.
+
+\leavevmode\hypertarget{ref-lace07}{}%
+Lacerda, Miguel, Cally Ardington, and Murray Leibbrandt. 2007.
+``Sequential Regression Multiple Imputation for Incomplete Multivariate
+Data Using Markov Chain Monte Carlo.'' University of Cape Town, South
+Africa.
+
+\leavevmode\hypertarget{ref-lync07}{}%
+Lynch, Scott M. 2007. \emph{Introduction to Applied Bayesian Statistics
+and Estimation for Social Scientists}. Springer Science \& Business
+Media.
+
+\leavevmode\hypertarget{ref-mack03}{}%
+MacKay, David JC, and David JC Mac Kay. 2003. \emph{Information Theory,
+Inference and Learning Algorithms}. Cambridge university press.
+
+\leavevmode\hypertarget{ref-murr18}{}%
+Murray, Jared S. 2018. ``Multiple Imputation: A Review of Practical and
+Theoretical Findings.'' \emph{Statistical Science} 33 (2): 142--59.
+\url{https://doi.org/10.1214/18-STS644}.
+
+\leavevmode\hypertarget{ref-neym34}{}%
+Neyman, Jerzy. 1934. ``On the Two Different Aspects of the
+Representative Method: The Method of Stratified Sampling and the Method
+of Purposive Selection.'' \emph{Journal of the Royal Statistical
+Society} 97 (4): 558--625. \url{https://doi.org/10.2307/2342192}.
+
+\leavevmode\hypertarget{ref-raft91}{}%
+Raftery, Adrian E, and Steven Lewis. 1991. ``How Many Iterations in the
+Gibbs Sampler?'' Washington University Seatle, Department of Statistics,
+United States.
+
 \leavevmode\hypertarget{ref-R}{}%
@@
 Applied Probability and Statistics. New York, NY: Wiley.
+
+\leavevmode\hypertarget{ref-scha97}{}%
+Schafer, Joseph L. 1997. \emph{Analysis of Incomplete Multivariate
+Data}. Chapman; Hall/CRC.
+
+\leavevmode\hypertarget{ref-taka17}{}%
+Takahashi, Masayoshi. 2017. ``Statistical Inference in Missing Data by
+MCMC and Non-MCMC Multiple Imputation Algorithms: Assessing the Effects
+of Between-Imputation Iterations.'' \emph{Data Science Journal} 16
+(July): 37. \url{https://doi.org/10.5334/dsj-2017-037}.
 
@@
 
+\leavevmode\hypertarget{ref-veht19}{}%
+Vehtari, Aki, Andrew Gelman, Daniel Simpson, Bob Carpenter, and
+Paul-Christian Bürkner. 2019. ``Rank-Normalization, Folding, and
+Localization: An Improved \$\textbackslash widehat\{R\}\$ for Assessing
+Convergence of MCMC,'' March. \url{http://arxiv.org/abs/1903.08008}.
+
+\leavevmode\hypertarget{ref-vinknd}{}%
+Vink, Gerko. n.d. ``Towards a Standardized Evaluation of Multiple
+Imputation Routines.''
+
+\leavevmode\hypertarget{ref-vink14}{}%
+Vink, Gerko, and Stef van Buuren. 2014. ``Pooling Multiple Imputations
+When the Sample Happens to Be the Population.'' \emph{arXiv:1409.8542
+{[}Math, Stat{]}}, September. \url{http://arxiv.org/abs/1409.8542}.
+
 \end{document}

### 2020-03-20 3.Thesis/1.SimulationStudy/Writeup/Archive/quotes_to_include.Rmd
Removed stuff from Introduction

---
title: "Literature on non-convergence"
author: "Hanne Oberman"
date: "4-3-2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

"There are two general approaches for imputing the data multiply: joint modelling (JM) and fully conditional specification (FCS). Different techniques for MI using JM have been developed in the literature.1,10-13 This approach is useful when the data can be assumed to have a well-known multivariate distribution, such as the multivariate normal distribution. A variety of JM imputation algorithms for longitudinal data and multilevel missing data are also available in the literature.14-18 The FCS approach is much more flexible but potentially flawed by an incompatible distribution and does not require a joint multivariate distribution, but it specifies an imputation model for each missing variable using its fully conditional density. Fully means that conditioning is usually done on all other variables in the dataset. It is therefore also useful and can be implemented with ease for complex data structures. The idea of FCS has been implemented in the literature with different names: stochastic relaxation by Kennickell,19 variable-by-variable imputation by Brand,20 regression switching by van Buuren et al.,21 sequential regression by Raghunathan et al.,22 ordered pseudo-Gibbs sampler by Heckerman et al.,23 partially incompatible Markov chain Monte Carlo (MCMC) by Rubin,24 iterative univariate imputation by Gelman,25 MI by chained equations (MICE) by van Buuren and Oudshoorn26, van Buuren and Groothuis-Oudshoorn27 and White et al.7 and FCS by van Buuren.28 "

"As MICE is an iterative procedure, it is important that convergence is achieved. This may be checked by computing, at
each cycle, the means of imputed values and/or the values of regression coefficients, and seeing if they are stable. For the
example in Section 9, these values appeared stable from the very first cycle. We have never found 10 cycles inadequate,
but larger numbers of cycles might in principle be required when incomplete variables are very strongly associated" (@whit11, p. 394).

# For ShinyMICE
"Determining How Many Imputations Are Needed
Working back from the computation of statistics and confidence intervals, we need
to ask how many imputations are required to generate useful statistical inferences. Surprisingly, the answer is often no more than five and sometimes as little as two or three.
The relative efficiency (RE) of the m imputations given a fraction of missing data
(Rubin, 1987, p. 114) is computed in standard error units as
...
Even when γ is .9, five imputations give a relative efficiency of .92. Consequently,
one rarely has to create more than five imputations, and 10 imputations are more than
suitable in almost any realistic application" (https://journals.sagepub.com/doi/pdf/10.1177/1094428103255532, p. 291).

"Multiple imputation (MI) is the method of choice for many incomplete data problems. MI incorporates the uncertainty about the missing data by creating $m>2$ imputed data sets. Missing values are filled in under an imputation model. The imputed data that result from the imputation model is then analyzed by the analysis model. Separate analyses can be combined to get a single inference or set of estimates by making use of the combining rules derived by Rubin (1987)."

"Missing data problems are ubiquitous in observational data and common among social science applications. Statistical inference that does not adequately account for the missing data is widely known to lead to biased results and inflated (or deflated) variance estimates (King et al. 2001; Molenberghs et al. 2014; Rubin 1976; White and Carlin 2010). Even though most statistical software platforms provide methods that adequately handle missing data (the most popular of these is multiple imputations [MI]), they are often ignored by applied researchers."

"Statistical methods to deal with missing data have greatly increased in both number and sophistication over the past 20 years. Multiple imputation (MI) is now considered the leading method for dealing with item-missing data that are missing completely at random (MCAR) or missing at random (MAR; Rubin 1987, 1996; Schafer 1997). Because standard statistical software packages now have built-in routines for implementing MI, its usage is making its way into the analyses of ordinary (but statistically savvy) researchers. Research emphasizes the importance of including in the imputation model all variables from the intended analytic model as well as variables correlated with the potential mechanisms of missingness and variables correlated with the variables to be imputed (hereafter referred to as auxiliary variables) as they improve the capability of the model to predict the missing values (Collins, Schafer, and Kam 2001; Schafer 2003)".

"Multiple imputation (MI) is a popular method for repairing and analyzing data with missing values (Rubin 1987). Using MI, you fill in the missing values with imputed values that are drawn at random from a posterior predictive distribution that is conditioned on the observed values Yobs. You repeat the process of imputation M times to obtain M imputed data sets, which you analyze as though they were complete. You then combine the results of the M analyses to obtain a point estimate ...".

### 2020-03-20 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
Run simulation to evaluate predictions (explained variance and stuff)
Removed stuff from Introduction

@@
 ---
-title: "Missing the Point: Convergence of Multiple Imputation using Chained Equations Algorithms"
+title: "Missing the Point: Convergence of Multiple Imputation by Chained Equations Algorithms"
 author: "Hanne I. Oberman"
@@
     keep_tex: true
+    number_sections: true
 bibliography: thesis.bib
@@
 
-# Background
-
-Feedback:
-
-- Focus on convergence, not evaluation of MI in general.
-
-- Combine Theoretical Background and Intro.
+# Introduction
 
@@
 A popular answer to the ubiquitous problem of missing information is to use the framework of multiple imputation (MI), proposed by @rubin87.
-MI is an iterative algorithmic procedure in which missing datapoints are `imputed' (i.e.
+For the theoretical foundation of MI, see @rubin87.
+For an accessible and comprehensive introduction to MI from an applied perspective, see e.g.
+@buur18.
+With MI, missing datapoints are 'imputed' (i.e., filled in) via an *imputation model*.
+The imputation model can be a joint model for all variables, or several univariate conditional distributions (fully conditional specification).
+FCS is more flexible and "avoids the problem of specifying an appropriate joint imputation distribution and replaces this by the selection of appropriate univariate conditional distributions".
+With FCS, an iterative algorithmic procedure is used to ...
+multiple imputation by chained equations ...
+"The values for imputing the missing data are drawn from the posterior predictive distribution of the missing data conditional on the observed data."  There is no scientific consensus on the convergence properties of FCS algorithms [@taka17].
+Some default techniques in `mice` might not yield converged states at all [@murr18].
+Therefore, algorithmic convergence should be monitored carefully.
+
+Convergence cannot be determined for sure, because MICE is an MCMC method, and convergence of MCMC is not from scalar to a point, but from one distribution to another.
+Generated values will vary even after convergence.
+Therefore we can only diagnose non-convergence, not convergence [@hoff09].
+"a weak diagnostics is better than no diagnostic at all" [@cowl96].
+
+Current practice is visually inspecting imputation chains for signs of non-convergence, but may be challenging to the untrained eye [@buur18, \S~6.5.2].
+Therefore diagnostic evaluation of convergence would be preferred.
+However, not known whether MCMC diagnostics are appropriate for MICE ...
+The application of convergence diagnostics to MI methods has not been systematically studied [@buur18].
+How can non-convergence be diagnosed?
+How many iterations are sufficient/required?
+What are the effects of non-convergence on estimates, predictions and inference?
+What is an appropriate non-convergence diagnostic?
+Are the default number of iterations sufficient?
+Defaults in: SPSS (10; https://www.ibm.com/support/knowledgecenter/SSLVMB_24.0.0/spss/mva/syn_multiple_imputation_impute.html), Mplus (100??; https://pdfs.semanticscholar.org/e20e/29e008592cbfbaa567931f74cdfdb5451405.pdf?_ga=2.55354671.54033656.1584698748-527613517.1584698748), Stata (10; https://www.stata.com/manuals13/mi.pdf, p.
+139), Amelia (NA, because resampling, not convergence; https://cran.r-project.org/web/packages/Amelia/Amelia.pdf), en MI (30; https://cran.r-project.org/web/packages/mi/mi.pdf).
+
+For reasons of brevity, we only focus on the MI algorithm implemented in the world-leading MI software: the `R` [@R] package `mice` [@mice].
+The convergence properties of this MI algorithm are investigated through model-based simulation.
+The results of this simulation study are guidelines for assessing convergence of MI algorithms, which will aid applied researchers in drawing valid inference from incomplete datasets.
+
+<!-- MI is an iterative algorithmic procedure in which missing datapoints are `imputed' (i.e.
 filled in) several times.
@@
 Therefore, MI can provide valid inferences despite missing information (**maybe add that this refers to CIs/p-values**).
+-->
 
-To obtain valid inferences with MI, the variability between imputations should be properly represented [@rubin87; @buur18].
+<!-- To obtain valid inferences with MI, the variability between imputations should be properly represented [@rubin87; @buur18].
 If this variability is under-estimated, confidence intervals around estimates will be too narrow, which can yield spurious results.
@@
 Evaluation measures, however, are currently missing or under-developed in MI software, like the world-leading `mice` package [@mice] in `R` [@R].
-The goal of this research project is to develop novel methodology and guidelines for evaluating MI methods.
+-->
+<!-- The goal of this research project is to develop novel methodology and guidelines for evaluating MI methods.
 These tools will subsequently be implemented in an interactive evaluation framework for multiple imputation, which will aid applied researchers in drawing valid inference from incomplete datasets.
+-->
 
-This note provides the theoretical foundation towards the diagnostic evaluation of multiple imputation algorithms.
+<!-- This note provides the theoretical foundation towards the diagnostic evaluation of multiple imputation algorithms.
 For reasons of brevity, we only focus on the MI algorithm implemented in `mice` [@mice].
-The convergence properties of this MI algorithm are investigated through model-based simulation [^1].
+The convergence properties of this MI algorithm are investigated through model-based simulation.
 The results of this simulation study are guidelines for assessing convergence of MI algorithms.
+-->
 <!-- These guidelines will be implemented in an interactive evaluation tool for `mice`, 'ShinyMICE', which is currently under development.
@@
 
-[^1]: All programming code used in this note is available from [github.com/gerkovink/shinyMice](https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy).
+<!-- [^1]: All programming code used in this note is available from [github.com/gerkovink/shinyMice](https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy). -->
 
-### Terminology
+## Terminology
 
@@
 <!-- Deviations from the 'original' notation by @rubin87 are described in [@buur18, \S~2.2.3].  -->
-Basic familiarity with MI methodology is assumed.
-For the theoretical foundation of MI, see @rubin87.
-For an accessible and comprehensive introduction to MI from an applied perspective, see e.g.
-@buur18.
+<!-- Basic familiarity with MI methodology is assumed.   -->
 
@@
 The missing part of $Y$ is denoted by $Y_{mis}$.
-Which parts of $Y$ are missing is determined by the `missingness mechanism'.
-This note only considers a 'missing completely at random' (MCAR) mechanism, where the probability of being missing is equal for all $n \times p$ cells in $Y$ [@rubin87].
+Which parts of $Y$ are missing is determined by the 'missingness mechanism'.
 
@@
 
-### Theoretical Background
+# Convergence diagnostics
 
-There is no scientific consensus on the convergence properties of multiple imputation algorithms [@taka17].
-Some default techniques in `mice` might not yield converged states at all [@murr18].
-Therefore, algorithmic convergence should be monitored carefully.
-The diagnostic evaluation of convergence is preferred over the current practice---visually inspecting imputation chains for signs of non-convergence---since the latter may be challenging to the untrained eye [@buur18, \S~6.5.2].
-
-The application of convergence diagnostics to MI methods has not been systematically studied [@buur18].
-The measures that are available for diagnosing convergence of iterative algorithmic procedures (Markov chain Monte Carlo methods; e.g., `mice` algorithms or Gibbs samplers) generally evaluate signs of non-convergence [@hoff09].
-Non-convergence may be present as non-stationarity within chains (i.e., trending), or as slow mixing between chains (i.e., no intermingling).
+In iterative algorithmic procedures (Markov chain Monte Carlo methods; e.g., `mice` algorithms or Gibbs samplers) non-convergence may be present as non-stationarity within chains (i.e., trending), or as slow mixing between chains (i.e., no intermingling).
 
@@
 W&=\frac{1}{M} \sum_{m=1}^{M} s_{j}^{2},  \text { where } s_{m}^{2}=\frac{1}{T-1} \sum_{t=1}^{T}\left(\theta^{(t m)}-\bar{\theta}^{(\cdot m)}\right)^{2}. 
+\end{align*}
 <!-- %\text{ \cite[p.~5]{veht19}}  -->
-\end{align*}
 
@@
 B&=\frac{T}{M-1} \sum_{m=1}^{M}\left(\bar{\theta}^{(\cdot m)}-\bar{\theta}^{(\cdot \cdot)}\right)^{2}, \text { where } \bar{\theta}^{(\cdot m)}=\frac{1}{T} \sum_{t=1}^{T} \theta^{(t m)}, \quad \bar{\theta}^{(\cdot \cdot)}=\frac{1}{M} \sum_{m=1}^{M} \bar{\theta}^{(\cdot m)}.
+\end{align*}
 <!-- %\text{ \cite[p.~5]{veht19}}  -->
-\end{align*}
 
@@
 
-### Simulation Hypothesis
+## Simulation Hypothesis
 
@@
 
-# Methods
+# Simulation study
 
@@
 $$Y =  2X_1 + .5X_2 - X_3 + \epsilon .$$
+This note only considers a 'missing completely at random' (MCAR) mechanism, where the probability of being missing is equal for all $n \times p$ cells in $Y$ [@rubin87].
 The complete data is 'amputed' once for each simulation repetition.
@@
 
-- Remove the table.
-
 - Add more info about figure legends and axes.
 
-Figures 2 and 3 display results per simulation condition ($T = 1,2,\dots,100$).
+<!-- Figures 2 and 3 display results per simulation condition ($T = 1,2,\dots,100$).  -->
 

### 2020-03-20 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Run simulation to evaluate predictions (explained variance and stuff)
Removed stuff from Introduction

@@
   \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+\setcounter{secnumdepth}{5}
 
@@
 
-\hypertarget{background}{%
-\section{Background}\label{background}}
-
-Feedback:
-
-\begin{itemize}
-\item
-  Focus on convergence, not evaluation of MI in general.
-\item
-  Combine Theoretical Background and Intro.
-\end{itemize}
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
 
@@
 answer to the ubiquitous problem of missing information is to use the
-framework of multiple imputation (MI), proposed by Rubin (1987). MI is
-an iterative algorithmic procedure in which missing datapoints are
-`imputed' (i.e.~filled in) several times. The variability between
-imputations is used to reflect how much uncertainty in the inference is
-introduced by the missingness. Therefore, MI can provide valid
-inferences despite missing information (\textbf{maybe add that this
-refers to CIs/p-values}).
+framework of multiple imputation (MI), proposed by Rubin (1987). For the
+theoretical foundation of MI, see Rubin (1987). For an accessible and
+comprehensive introduction to MI from an applied perspective, see
+e.g.~Van Buuren (2018). With MI, missing datapoints are `imputed' (i.e.,
+filled in) via an \emph{imputation model}. The imputation model can be a
+joint model for all variables, or several univariate conditional
+distributions (fully conditional specification). FCS is more flexible
+and ``avoids the problem of specifying an appropriate joint imputation
+distribution and replaces this by the selection of appropriate
+univariate conditional distributions''. With FCS, an iterative
+algorithmic procedure is used to \ldots{} multiple imputation using
+chained equations \ldots{} ``The values for imputing the missing data
+are drawn from the posterior predictive distribution of the missing data
+conditional on the observed data.'' There is no scientific consensus on
+the convergence properties of FCS algorithms (Takahashi 2017). Some
+default techniques in \texttt{mice} might not yield converged states at
+all (Murray 2018). Therefore, algorithmic convergence should be
+monitored carefully.
 
-To obtain valid inferences with MI, the variability between imputations
-should be properly represented (Rubin 1987; Van Buuren 2018). If this
-variability is under-estimated, confidence intervals around estimates
-will be too narrow, which can yield spurious results. Over-estimation of
-the variance between imputations results in unnecessarily wide
-confidence intervals, which can be costly because it lowers the
-statistical power (\textbf{maybe add why this is costly}). Since both of
-these situations are undesirable, imputations and their variability
-should be evaluated. Evaluation measures, however, are currently missing
-or under-developed in MI software, like the world-leading \texttt{mice}
-package (Van Buuren and Groothuis-Oudshoorn 2011) in \texttt{R} (R Core
-Team 2019). The goal of this research project is to develop novel
-methodology and guidelines for evaluating MI methods. These tools will
-subsequently be implemented in an interactive evaluation framework for
-multiple imputation, which will aid applied researchers in drawing valid
-inference from incomplete datasets.
+Current practice is visually inspecting imputation chains for signs of
+non-convergence, but may be challenging to the untrained eye (Van Buuren
+2018, \textasciitilde6.5.2). Therefore diagnostic evaluation of
+convergence would be preferred. However, not known whether MCMC
+diagnostics are appropriate for MICE \ldots{} The application of
+convergence diagnostics to MI methods has not been systematically
+studied (Van Buuren 2018). How can non-convergence be diagnosed? How
+many iterations are sufficient/required? What are the effects of
+non-convergence on estimates, predictions and inference? What is an
+appropriate non-convergence diagnostic? Are the default number of
+iterations sufficient?
 
-This note provides the theoretical foundation towards the diagnostic
-evaluation of multiple imputation algorithms. For reasons of brevity, we
-only focus on the MI algorithm implemented in \texttt{mice} (Van Buuren
-and Groothuis-Oudshoorn 2011). The convergence properties of this MI
-algorithm are investigated through model-based simulation.\footnote{All
-  programming code used in this note is available from
-  \href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy}{github.com/gerkovink/shinyMice}.}
-The results of this simulation study are guidelines for assessing
-convergence of MI algorithms.
+Convergence cannot be determined for sure, because MICE is an MCMC
+method, and convergence of MCMC is not from scalar to a point, but from
+one distribution to another. Generated values will vary even after
+convergence. Therefore we can only diagnose non-convergence, not
+convergence (Hoff 2009). ``a weak diagnostics is better than no
+diagnostic at all'' (Cowles and Carlin 1996).
+
+Defaults is: SPSS (10;
+\url{https://www.ibm.com/support/knowledgecenter/SSLVMB_24.0.0/spss/mva/syn_multiple_imputation_impute.html}),
+Mplus (100??;
+\url{https://pdfs.semanticscholar.org/e20e/29e008592cbfbaa567931f74cdfdb5451405.pdf?_ga=2.55354671.54033656.1584698748-527613517.1584698748}),
+Stata (10; \url{https://www.stata.com/manuals13/mi.pdf}, p.~139), Amelia
+(NA, because resampling, not convergence;
+\url{https://cran.r-project.org/web/packages/Amelia/Amelia.pdf}), en MI
+(30; \url{https://cran.r-project.org/web/packages/mi/mi.pdf}).
+
+For reasons of brevity, we only focus on the MI algorithm implemented in
+the world-leading MI software: the \texttt{R} (R Core Team 2019) package
+\texttt{mice} (Van Buuren and Groothuis-Oudshoorn 2011). The convergence
+properties of this MI algorithm are investigated through model-based
+simulation. The results of this simulation study are guidelines for
+assessing convergence of MI algorithms, which will aid applied
+researchers in drawing valid inference from incomplete datasets.
 
 \hypertarget{terminology}{%
-\subsubsection{Terminology}\label{terminology}}
-
-This note follows notation and conventions of \texttt{mice} (Van Buuren
-and Groothuis-Oudshoorn 2011). Basic familiarity with MI methodology is
-assumed. For the theoretical foundation of MI, see Rubin (1987). For an
-accessible and comprehensive introduction to MI from an applied
-perspective, see e.g.~Van Buuren (2018).
+\subsection{Terminology}\label{terminology}}
 
@@
 
-\hypertarget{theoretical-background}{%
-\subsubsection{Theoretical Background}\label{theoretical-background}}
+\hypertarget{convergence-diagnostics}{%
+\section{Convergence diagnostics}\label{convergence-diagnostics}}
 
-There is no scientific consensus on the convergence properties of
-multiple imputation algorithms (Takahashi 2017). Some default techniques
-in \texttt{mice} might not yield converged states at all (Murray 2018).
-Therefore, algorithmic convergence should be monitored carefully. The
-diagnostic evaluation of convergence is preferred over the current
-practice---visually inspecting imputation chains for signs of
-non-convergence---since the latter may be challenging to the untrained
-eye (Van Buuren 2018, \textasciitilde6.5.2).
-
-The application of convergence diagnostics to MI methods has not been
-systematically studied (Van Buuren 2018). The measures that are
-available for diagnosing convergence of iterative algorithmic procedures
-(Markov chain Monte Carlo methods; e.g., \texttt{mice} algorithms or
-Gibbs samplers) generally evaluate signs of non-convergence (Hoff 2009).
-Non-convergence may be present as non-stationarity within chains (i.e.,
-trending), or as slow mixing between chains (i.e., no intermingling).
+In iterative algorithmic procedures (Markov chain Monte Carlo methods;
+e.g., \texttt{mice} algorithms or Gibbs samplers) non-convergence may be
+present as non-stationarity within chains (i.e., trending), or as slow
+mixing between chains (i.e., no intermingling).
 
@@
 W&=\frac{1}{M} \sum_{m=1}^{M} s_{j}^{2},  \text { where } s_{m}^{2}=\frac{1}{T-1} \sum_{t=1}^{T}\left(\theta^{(t m)}-\bar{\theta}^{(\cdot m)}\right)^{2}. 
-<!-- %\text{ \cite[p.~5]{veht19}}  -->
 \end{align*}
@@
 B&=\frac{T}{M-1} \sum_{m=1}^{M}\left(\bar{\theta}^{(\cdot m)}-\bar{\theta}^{(\cdot \cdot)}\right)^{2}, \text { where } \bar{\theta}^{(\cdot m)}=\frac{1}{T} \sum_{t=1}^{T} \theta^{(t m)}, \quad \bar{\theta}^{(\cdot \cdot)}=\frac{1}{M} \sum_{m=1}^{M} \bar{\theta}^{(\cdot m)}.
-<!-- %\text{ \cite[p.~5]{veht19}}  -->
 \end{align*}
@@
 \hypertarget{simulation-hypothesis}{%
-\subsubsection{Simulation Hypothesis}\label{simulation-hypothesis}}
+\subsection{Simulation Hypothesis}\label{simulation-hypothesis}}
 
@@
 
-\hypertarget{methods}{%
-\section{Methods}\label{methods}}
+\hypertarget{simulation-study}{%
+\section{Simulation study}\label{simulation-study}}
 
@@
 
-\hypertarget{convergence-diagnostics}{%
-\subsubsection{Convergence Diagnostics}\label{convergence-diagnostics}}
+\hypertarget{convergence-diagnostics-1}{%
+\subsubsection{Convergence
+Diagnostics}\label{convergence-diagnostics-1}}
 

### 2020-03-20 3.Thesis/1.SimulationStudy/1.Execute.R

@@ -29,7 +29,7 @@ source("3.Thesis/1.SimulationStudy/Functions/Evaluate.R")
 # simulation parameters
 populationsize <- 10000 #n of simulated dataset
 n.iter <- 30 #nr of iterations (varying 1:n.iter)
-n.sim <- 20 #nr of simulations per iteration value
+n.sim <- 2 #nr of simulations per iteration value
 true_effect <- 2 #regression coefficient to be estimated
 true_mean <- true_sd <- NA
 
@@ -40,12 +40,10 @@ set.seed(1111)
 
 # simulate data once
 data <- data.simulation(n = populationsize, true_effect)
-# run ampute once to get patterns object (and )to be able to adjust it for multivariate missingness)
-old_amp_patterns <- ampute(data)$patterns
-# for multivariable missingness, uncomment this
-# amp_patterns[1:4, 1] <- 0
-amp_patterns <- expand.grid(c(0,1),c(0,1), c(0,1), c(0,1)) %>% .[c(-1, -16),]# %>% unname()
-names(amp_patterns) <- names(old_amp_patterns)
+
+# create patterns object with multivariate missingness
+amp_patterns <- expand.grid(c(0,1),c(0,1), c(0,1), c(0,1)) %>% .[c(-1, -16),]
+names(amp_patterns) <- ampute(data)$patterns %>% names()
 
 # combine separate functions into wrapper
 simulate <- function(data, n.iter, true_effect, patterns) {
@@ -76,25 +74,19 @@ out <- replicate(n.sim, simulate(data = data, n.iter = n.iter, true_effect = tru
 
 # evaluate
 results <- evaluate.sim(sims = out, n.iter = n.iter)
-# uncomment for MCMC SEs
-MCMCSE <- evaluate.sim(sims = out, n.iter = n.iter, mean_or_SE = "se")
-# combine into one dataframe
-results_with_SE <- left_join(results, MCMCSE, by = "T", suffix = c("", ".SE"))
-# with empirical CI
+
+# # uncomment for MCMC SEs
+# MCMCSE <- evaluate.sim(sims = out, n.iter = n.iter, mean_or_SE = "se")
+# # combine into one dataframe
+# results_with_SE <- left_join(results, MCMCSE, by = "T", suffix = c("", ".SE"))
+
+# results with empirical CI
 CI_lower <- evaluate.sim(sims = out, n.iter = n.iter, mean_or_SE = "lower")
[…]

### 2020-03-22 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
Comments and changes to research report manuscript to fit new simulation (until methods section)

@@
 
-Current practice is visually inspecting imputation chains for signs of non-convergence, but may be challenging to the untrained eye [@buur18, \S~6.5.2].
+Current practice is visually inspecting imputation chains for signs of non-convergence, but may be challenging to the untrained eye [@buur18, $\S$ 6.5.2].
 Therefore diagnostic evaluation of convergence would be preferred.
@@
 Are the default number of iterations sufficient?
-Defaults in: SPSS (10; https://www.ibm.com/support/knowledgecenter/SSLVMB_24.0.0/spss/mva/syn_multiple_imputation_impute.html), Mplus (100??; https://pdfs.semanticscholar.org/e20e/29e008592cbfbaa567931f74cdfdb5451405.pdf?_ga=2.55354671.54033656.1584698748-527613517.1584698748), Stata (10; https://www.stata.com/manuals13/mi.pdf, p.
-139), Amelia (NA, because resampling, not convergence; https://cran.r-project.org/web/packages/Amelia/Amelia.pdf), en MI (30; https://cran.r-project.org/web/packages/mi/mi.pdf).
+Defaults in: SPSS = 10 [(link)](https://www.ibm.com/support/knowledgecenter/SSLVMB_24.0.0/spss/mva/syn_multiple_imputation_impute.html), Mplus = 100??
+[(link)](https://pdfs.semanticscholar.org/e20e/29e008592cbfbaa567931f74cdfdb5451405.pdf?_ga=2.55354671.54033656.1584698748-527613517.1584698748), Stata = 1; [(link)](https://www.stata.com/manuals13/mi.pdf, p.
+139), Amelia = NA, because resampling, not convergence [(link)](https://cran.r-project.org/web/packages/Amelia/Amelia.pdf), en MI = 30 [(link)](https://cran.r-project.org/web/packages/mi/mi.pdf).
 
@@
 
-## Terminology
+## Notation
 
 <!-- The intended audience of this note consists of anyone who uses multiple imputation to solve missing data problem.  -->
-This note follows notation and conventions of `mice` [@mice].
+<!-- This note follows notation and conventions of `mice` [@mice]. -->
 <!-- Deviations from the 'original' notation by @rubin87 are described in [@buur18, \S~2.2.3].  -->
@@
 
-Let $Y$ denote an $n \times p$ matrix containing the data values on $p$ variables for all $n$ units in a sample.
-The collection of observed data values in $Y$ is denoted by $Y_{obs}$, and will be referred to as 'incomplete' or 'observed' data.
-The missing part of $Y$ is denoted by $Y_{mis}$.
-Which parts of $Y$ are missing is determined by the 'missingness mechanism'.
+Let $y$ denote an $n \times p$ matrix containing the data values on $p$ variables for all $n$ units in a sample.
+The data value of unit $i$ ($i = 1, \dots, n$) on variable $j$ ($j = 1, \dots, p$) may be either observed or missing.
+The collection of observed data values in $y$ is denoted by $y_{obs}$; the missing part of $y$ is referred to as $y_{mis}$.
+For each datapoint in $y_{mis}$ $M \times T$ times plausible values are sampled, where $M$ is the number of imputations ($m = 1, \dots, M$) and $T$ is the number of iterations ($t = 1, \dots, T$).
+The collection of samples between the initial value (at $t=1$) and the imputed value (at $t=T$) will be referred to as an 'imputation chain'.
 
-Figure 1 provides an overview of the steps involved with MI---from incomplete data, to $m$ completed datasets, to $m$ estimated quantities of interest ($\hat{Q}$s), to a single pooled estimate $\bar{Q}$.
-This note focuses on the algorithmic properties of the imputation step.
+<!-- The number of imputations is denoted by $m$ (i.e., the number of times each datapoint in $y_{mis}$ is filled in).
+The aim of MI is to impute each datapoint in $y_{mis}$ $m$ times.
+Because of the iterative nature of the MICE algorithm -->
 
-The MI algorithm in `mice` has an iterative nature.
-For each missing datapoint in $Y_{mis}$, $m$ 'chains' of potential values are sampled.
-Only the ultimate sample that each chain lands on is imputed.
-The number of iterations per chain will be denoted with $T$, where $t$ varies over the integers $1, 2, \dots, T$.
-This is repeated $m$ times.
-Imputed values are the ultimate samples of a `chain' of For each imputed value ($t = T$), a s for each missing datapoint in $Y_{mis}$, .
-Each of the $m$ chains starts with an initial value, drawn randomly from $Y_{obs}$.
+<!-- The result is $m$ completed datasets, consisting of $y_{obs}$ and   The MI algorithm in `mice` has an iterative nature.
+For each missing datapoint in $y_{mis}$, $m$ 'chains' of potential values are sampled.
+-->
+<!-- Which parts of $y$ are missing is determined by the 'missingness mechanism'. -->
+
+<!-- Figure 1 provides an overview of the steps involved with MI---from incomplete data, to $m$ completed datasets, to $m$ estimated quantities of interest ($\hat{Q}$s), to a single pooled estimate $\bar{Q}$.
+-->
+<!-- This note focuses on the algorithmic properties of the imputation step.  -->
+
+<!-- Only the ultimate sample that each chain lands on is imputed. The number of iterations per chain will be denoted with $T$, where $t$ varies over the integers $1, 2, \dots, T$.  -->
+
+<!-- This is repeated $m$ times.
+Imputed values are the ultimate samples of a `chain' of For each imputed value ($t = T$), a s for each missing datapoint in $y_{mis}$, .
+Each of the $m$ chains starts with an initial value, drawn randomly from $y_{obs}$.
 The chains are terminated after a predefined number of iterations.
 , and subsequently used in the analysis and pooling steps.
-The collection of samples between the initial value (at $t=1$) and the imputed value (at $t=T$) will be referred to as an 'imputation chain'.
+-->
+<!-- The collection of samples between the initial value (at $t=1$) and the imputed value (at $t=T$) will be referred to as an 'imputation chain'.  -->
 
@@
 In iterative algorithmic procedures (Markov chain Monte Carlo methods; e.g., `mice` algorithms or Gibbs samplers) non-convergence may be present as non-stationarity within chains (i.e., trending), or as slow mixing between chains (i.e., no intermingling).
+(**add example/guidelines visual inspection here?**)
 
@@
 The application of $\widehat{R}$ to assess stationarity has not been thoroughly investigated.
-Therefore, this study employs both $\widehat{R}$ and autocorrelation to investigate convergence, as recommended by [@cowl96, p.~898].
+Therefore, this study employs both $\widehat{R}$ and autocorrelation to investigate convergence, as recommended by [@cowl96, p.
+898].
 
-The mixing component of convergence may be evaluated with the potential scale reduction factor, $\widehat{R}$ [a.k.a.
+<!-- The mixing component of convergence may be evaluated with the potential scale reduction factor, $\widehat{R}$ [a.k.a.
 `Gelman-Rubin statistic'; @gelm92].
@@
 Other convergence diagnostics are outside of the scope of this study.
+-->
 
-All of these methods evaluate the convergence of univariate scalar summaries (e.g., chain means or variances).
+Note that all of these methods evaluate the convergence of univariate scalar summaries (e.g., chain means or variances).
 These convergence diagnostics cannot diagnose convergence of multivariable statistics (i.e., relations between scalar summaries).
@@
 
-To define $\widehat{R}$, we follow notation by [@veht19, p.~5].
+## Potential scale reduction factor
+
+To define $\widehat{R}$, we follow notation by [@veht19, p.
+5].
 Let $M$ be the total number of chains, $T$ the number of iterations per chain, and $\theta$ the scalar summary of interest (e.g., chain mean or chain variance).
@@
 
-Following the same notation, we define autocorrelation as the correlation between two subsequent $\theta$-values within the same chain [@lync07, p.~147].
+## Autocorrelation
+
+Following the same notation, we define autocorrelation as the correlation between two subsequent $\theta$-values within the same chain [@lync07, p.
+147].
 In this study we only consider $AC$ at lag 1, i.e., the correlation between the $t^{th}$ and $t+1^{th}$ iteration of the same chain.
@@
 This study evaluates whether $\widehat{R}$ and $AC$ could diagnose convergence of multiple imputation algorithms.
-We assess the performance of the two convergence diagnostics against the recommended evaluation criteria for MI methods [i.e., average bias, average confidence interval width, and empirical coverage rate across simulations; @buur18, \S~2.5.2].
+We assess the performance of the two convergence diagnostics against the recommended evaluation criteria for MI methods [i.e., average bias, average confidence interval width, and empirical coverage rate across simulations; @buur18, $\S$ 2.5.2].
 That is, there is no baseline measure available to evaluate performance against.The current practice of visually inspecting imputation chains for signs of non-mixing or non-stationarity can only diagnose severely pathological cases of non-convergence.
@@
 The threshold of $\widehat{R} < 1.01$ will then be too stringent for diagnosing convergence.
-This over-estimation may, however, be diminished because $\widehat{R}$ can falsely diagnose convergence if initial values of the algorithm are not appropriately over-dispersed [@broo98, p~437].
+This over-estimation may, however, be diminished because $\widehat{R}$ can falsely diagnose convergence if initial values of the algorithm are not appropriately over-dispersed [@broo98, p.
+437].
 In `mice`, initial values are chosen randomly from the observed data.
@@
 
-We use model-based simulation in `R` to perform multiple imputation under 100 simulation conditions.
+The convergence of `mice` is investigated through model-based simulation in `R` [version `r paste(version$major,version$minor, sep = ".")`; @R].
+The simulation set-up is summarized in the pseudo-code below.
+The complete `R` script of the simulation study is available from [github.com/gerkovink/shinyMice](https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy).
+
+<!-- Simulation conditions are the number of iterations ($T = 1, 2, \dots, 100$) and the percentage missingness in the  -->
+
+<!-- to perform multiple imputation under 100 simulation conditions.
 In each condition, the MI algorithm comprises of a different imputation chain length ($T = 1, 2, \dots, 100$).
@@
 These diagnostics are aggregated across repetitions to obtain their average value per simulation condition.
-The simulation set-up is summarized in the pseudo-code below.
-The complete `R` script of the simulation study is available from [github.com/gerkovink/shinyMice](https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy).
+-->
 
@@
 
-### Data Simulation
-
-The point of origin for all simulation conditions is a finite population of $N=1000$.
-The data are simulated to solve a multiple linear regression problem, where dependent variable $Y$ is regressed on independent variables $X_1$, $X_2$ and $X_3$:
+A finite population of $N=1000$ is simulated to solve a multiple linear regression problem, where dependent variable $Y$ is regressed on independent variables $X_1$, $X_2$ and $X_3$:
 
@@
 
-The quantity of scientific interest is regression coefficient $\beta_1$.
-The data generating model is a multivariate normal distribution with the following means structure and variance-covariance matrix:
+The data generating model is a multivariate normal distribution
 
@@
 \end{pmatrix}
-\end{bmatrix}\\[2\jot].
+\end{bmatrix}\\[2\jot]
 \end{align*}
 
-Data is simulated with the function `mvtnorm::rmvnorm()`, and outcome variable $Y$ is subsequently calculated as 
+and simulated with the function `mvtnorm::rmvnorm()`. Outcome variable $Y$ is subsequently calculated as 
 $$Y =  2X_1 + .5X_2 - X_3 + \epsilon .$$
-This note only considers a 'missing completely at random' (MCAR) mechanism, where the probability of being missing is equal for all $n \times p$ cells in $Y$ [@rubin87].
+<!-- This note only considers a 'missing completely at random' (MCAR) mechanism, where the probability of being missing is equal for all $n \times p$ cells in $y$ [@rubin87].  -->
+
 The complete data is 'amputed' once for each simulation repetition.
 That is, the function `mice::ampute()` is used to impose a missingness mechanism upon the data.
-The missingness is univariate, and the probability to be missing is the same for all four variables, namely 20\% (`prop = 0.8, mech = "MCAR"`).
-This leaves 20\% of the rows completely observed.
-%The resulting amputed data is equal for all simulation conditions in the same repetition.
+**(change this: The missingness is univariate, and the probability to be missing is the same for all four variables, namely 20\% (`prop = 0.8, mech = "MCAR"`).
+This leaves 20\% of the rows completely observed).
+**
 
@@
 
-From the imputation step, we extract chain means and chain variances across iterations ($\theta$ at $t = 1,2,\dots,T$).  
-We compute convergence diagnostics separately for both $\theta$s, and for each of the four variables in the dataset.
+**(tot hier doorgelezen)**
+
+<!-- From the imputation step, we extract chain means and chain variances across iterations ($\theta$ at $t = 1,2,\dots,T$).   -->
+<!-- We compute convergence diagnostics separately for both $\theta$s, and for each of the four variables in the dataset.
 We report the maximum (absolute) value as the 'worst linear predictor' of convergence.
-$\widehat{R}$ is computed by implementing Vehtari et al.'s [-@veht19] recommendations.
-$AC$ is computed with function `stats::acf()`.
+-->
+$\widehat{R}$ is computed by implementing Vehtari et al.'s [-@veht19] recommendations. $AC$ is computed with function `stats::acf()`.
 
@@
 
-### Convergence Diagnostics
+```{r}
+# plot results
+load("../Results/results.Rdata")
+
+```
+
+## Convergence Diagnostics
 
@@
 
-### Simulation Diagnostics
+## Simulation Diagnostics
 

### 2020-03-22 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Comments and changes to research report manuscript to fit new simulation (until methods section)

@@
 \hypersetup{
-  pdftitle={Missing the Point: Convergence of Multiple Imputation using Chained Equations Algorithms},
+  pdftitle={Missing the Point: Convergence of Multiple Imputation by Chained Equations Algorithms},
   pdfauthor={Hanne I. Oberman},
@@
 
-\title{Missing the Point: Convergence of Multiple Imputation using Chained
+\title{Missing the Point: Convergence of Multiple Imputation by Chained
 Equations Algorithms}
@@
 univariate conditional distributions''. With FCS, an iterative
-algorithmic procedure is used to \ldots{} multiple imputation using
-chained equations \ldots{} ``The values for imputing the missing data
-are drawn from the posterior predictive distribution of the missing data
+algorithmic procedure is used to \ldots{} multiple imputation by chained
+equations \ldots{} ``The values for imputing the missing data are drawn
+from the posterior predictive distribution of the missing data
 conditional on the observed data.'' There is no scientific consensus on
@@
 monitored carefully.
-
-Current practice is visually inspecting imputation chains for signs of
-non-convergence, but may be challenging to the untrained eye (Van Buuren
-2018, \textasciitilde6.5.2). Therefore diagnostic evaluation of
-convergence would be preferred. However, not known whether MCMC
-diagnostics are appropriate for MICE \ldots{} The application of
-convergence diagnostics to MI methods has not been systematically
-studied (Van Buuren 2018). How can non-convergence be diagnosed? How
-many iterations are sufficient/required? What are the effects of
-non-convergence on estimates, predictions and inference? What is an
-appropriate non-convergence diagnostic? Are the default number of
-iterations sufficient?
 
@@
 
-Defaults is: SPSS (10;
-\url{https://www.ibm.com/support/knowledgecenter/SSLVMB_24.0.0/spss/mva/syn_multiple_imputation_impute.html}),
-Mplus (100??;
-\url{https://pdfs.semanticscholar.org/e20e/29e008592cbfbaa567931f74cdfdb5451405.pdf?_ga=2.55354671.54033656.1584698748-527613517.1584698748}),
-Stata (10; \url{https://www.stata.com/manuals13/mi.pdf}, p.~139), Amelia
-(NA, because resampling, not convergence;
-\url{https://cran.r-project.org/web/packages/Amelia/Amelia.pdf}), en MI
-(30; \url{https://cran.r-project.org/web/packages/mi/mi.pdf}).
+Current practice is visually inspecting imputation chains for signs of
+non-convergence, but may be challenging to the untrained eye (Van Buuren
+2018, \(\S\) 6.5.2). Therefore diagnostic evaluation of convergence
+would be preferred. However, not known whether MCMC diagnostics are
+appropriate for MICE \ldots{} The application of convergence diagnostics
+to MI methods has not been systematically studied (Van Buuren 2018). How
+can non-convergence be diagnosed? How many iterations are
+sufficient/required? What are the effects of non-convergence on
+estimates, predictions and inference? What is an appropriate
+non-convergence diagnostic? Are the default number of iterations
+sufficient? Defaults in: SPSS = 10
+\href{https://www.ibm.com/support/knowledgecenter/SSLVMB_24.0.0/spss/mva/syn_multiple_imputation_impute.html}{(link)},
+Mplus = 100??
+\href{https://pdfs.semanticscholar.org/e20e/29e008592cbfbaa567931f74cdfdb5451405.pdf?_ga=2.55354671.54033656.1584698748-527613517.1584698748}{(link)},
+Stata = 1;
+\href{https://www.stata.com/manuals13/mi.pdf,\%20p.\%20139}{(link)},
+Amelia = NA, because resampling, not convergence
+\href{https://cran.r-project.org/web/packages/Amelia/Amelia.pdf}{(link)},
+en MI = 30
+\href{https://cran.r-project.org/web/packages/mi/mi.pdf}{(link)}.
 
@@
 
-\hypertarget{terminology}{%
-\subsection{Terminology}\label{terminology}}
+\hypertarget{notation}{%
+\subsection{Notation}\label{notation}}
 
-Let \(Y\) denote an \(n \times p\) matrix containing the data values on
-\(p\) variables for all \(n\) units in a sample. The collection of
-observed data values in \(Y\) is denoted by \(Y_{obs}\), and will be
-referred to as `incomplete' or `observed' data. The missing part of
-\(Y\) is denoted by \(Y_{mis}\). Which parts of \(Y\) are missing is
-determined by the `missingness mechanism'.\\
-This note only considers a `missing completely at random' (MCAR)
-mechanism, where the probability of being missing is equal for all
-\(n \times p\) cells in \(Y\) (Rubin 1987).
-
-Figure 1 provides an overview of the steps involved with MI---from
-incomplete data, to \(m\) completed datasets, to \(m\) estimated
-quantities of interest (\(\hat{Q}\)s), to a single pooled estimate
-\(\bar{Q}\). This note focuses on the algorithmic properties of the
-imputation step.
-
-The MI algorithm in \texttt{mice} has an iterative nature. For each
-missing datapoint in \(Y_{mis}\), \(m\) `chains' of potential values are
-sampled. Only the ultimate sample that each chain lands on is imputed.
-The number of iterations per chain will be denoted with \(T\), where
-\(t\) varies over the integers \(1, 2, \dots, T\). This is repeated
-\(m\) times. Imputed values are the ultimate samples of a `chain' of For
-each imputed value (\(t = T\)), a s for each missing datapoint in
-\(Y_{mis}\), . Each of the \(m\) chains starts with an initial value,
-drawn randomly from \(Y_{obs}\). The chains are terminated after a
-predefined number of iterations. , and subsequently used in the analysis
-and pooling steps. The collection of samples between the initial value
-(at \(t=1\)) and the imputed value (at \(t=T\)) will be referred to as
-an `imputation chain'.
+Let \(y\) denote an \(n \times p\) matrix containing the data values on
+\(p\) variables for all \(n\) units in a sample. The data value of unit
+\(i\) (\(i = 1, \dots, n\)) on variable \(j\) (\(j = 1, \dots, p\)) may
+be either observed or missing. The collection of observed data values in
+\(y\) is denoted by \(y_{obs}\); the missing part of \(y\) is referred
+to as \(y_{mis}\). For each datapoint in \(y_{mis}\) \(M \times T\)
+times plausible values are sampled, where \(M\) is the number of
+imputations (\(m = 1, \dots, M\)) and \(T\) is the number of iterations
+(\(t = 1, \dots, T\)). The collection of samples between the initial
+value (at \(t=1\)) and the imputed value (at \(t=T\)) will be referred
+to as an `imputation chain'.
 
@@
 present as non-stationarity within chains (i.e., trending), or as slow
-mixing between chains (i.e., no intermingling).
+mixing between chains (i.e., no intermingling). (\textbf{add
+example/guidelines visual inspection here?})
 
@@
 \(\widehat{R}\) and autocorrelation to investigate convergence, as
-recommended by (Cowles and Carlin 1996, \textasciitilde898).
+recommended by (Cowles and Carlin 1996, 898).
 
-The mixing component of convergence may be evaluated with the potential
-scale reduction factor, \(\widehat{R}\) (a.k.a. `Gelman-Rubin
-statistic'; Gelman and Rubin 1992). The stationarity component may be
-assessed through autocorrelations. An adapted version of \(\widehat{R}\)
-might also be used to diagnose non-stationarity. Other convergence
-diagnostics are outside of the scope of this study.
-
-All of these methods evaluate the convergence of univariate scalar
-summaries (e.g., chain means or variances). These convergence
+Note that all of these methods evaluate the convergence of univariate
+scalar summaries (e.g., chain means or variances). These convergence
 diagnostics cannot diagnose convergence of multivariable statistics
@@
 
+\hypertarget{potential-scale-reduction-factor}{%
+\subsection{Potential scale reduction
+factor}\label{potential-scale-reduction-factor}}
+
 To define \(\widehat{R}\), we follow notation by (Vehtari et al. 2019,
-\textasciitilde5). Let \(M\) be the total number of chains, \(T\) the
-number of iterations per chain, and \(\theta\) the scalar summary of
-interest (e.g., chain mean or chain variance). For each chain
+5). Let \(M\) be the total number of chains, \(T\) the number of
+iterations per chain, and \(\theta\) the scalar summary of interest
+(e.g., chain mean or chain variance). For each chain
 (\(m = 1, 2, \dots, M\)), we estimate the variance of \(\theta\), and
@@
 
+\hypertarget{autocorrelation}{%
+\subsection{Autocorrelation}\label{autocorrelation}}
+
 Following the same notation, we define autocorrelation as the
 correlation between two subsequent \(\theta\)-values within the same
-chain (Lynch 2007, \textasciitilde147). In this study we only consider
-\(AC\) at lag 1, i.e., the correlation between the \(t^{th}\) and
-\(t+1^{th}\) iteration of the same chain.
+chain (Lynch 2007, 147). In this study we only consider \(AC\) at lag 1,
+i.e., the correlation between the \(t^{th}\) and \(t+1^{th}\) iteration
+of the same chain.
 
@@
 width, and empirical coverage rate across simulations; Van Buuren 2018,
-\textasciitilde2.5.2). That is, there is no baseline measure available
-to evaluate performance against.The current practice of visually
-inspecting imputation chains for signs of non-mixing or non-stationarity
-can only diagnose severely pathological cases of non-convergence. We
-could also look at distributional characteristics, and plausibility of
-imputed values, see Vink (n.d.). For now, this is outside of the scope
-of this study.
+\(\S\) 2.5.2). That is, there is no baseline measure available to
+evaluate performance against.The current practice of visually inspecting
+imputation chains for signs of non-mixing or non-stationarity can only
+diagnose severely pathological cases of non-convergence. We could also
+look at distributional characteristics, and plausibility of imputed
+values, see Vink (n.d.). For now, this is outside of the scope of this
+study.
 
@@
 convergence if initial values of the algorithm are not appropriately
-over-dispersed (Brooks and Gelman 1998, p\textasciitilde437). In
-\texttt{mice}, initial values are chosen randomly from the observed
-data. Therefore, we cannot be certain that the initial values are
-over-dispersed. We expect this to have little effect on the hypothesized
-performance of \(\widehat{R}\). No hypothesis was formulated about the
-performance of \(AC\) as convergence diagnostic.
+over-dispersed (Brooks and Gelman 1998, 437). In \texttt{mice}, initial
+values are chosen randomly from the observed data. Therefore, we cannot
+be certain that the initial values are over-dispersed. We expect this to
+have little effect on the hypothesized performance of \(\widehat{R}\).
+No hypothesis was formulated about the performance of \(AC\) as
+convergence diagnostic.
 
@@
 
-We use model-based simulation in \texttt{R} to perform multiple
-imputation under 100 simulation conditions. In each condition, the MI
-algorithm comprises of a different imputation chain length
-(\(T = 1, 2, \dots, 100\)). The number of simulation runs per condition
-is 1000. For each simulation condition (i.e., number of iterations) and
-each repetition (i.e., simulation run) we compute convergence
-diagnostics (\(\widehat{R}\) and autocorrelation) and simulation
-diagnostics (bias, confidence interval width and coverage). These
-diagnostics are aggregated across repetitions to obtain their average
-value per simulation condition. The simulation set-up is summarized in
-the pseudo-code below. The complete \texttt{R} script of the simulation
-study is available from
+The convergence of \texttt{mice} is investigated through model-based
+simulation in \texttt{R} (version 3.6.3; R Core Team 2019). The
+simulation set-up is summarized in the pseudo-code below. The complete
+\texttt{R} script of the simulation study is available from
 \href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy}{github.com/gerkovink/shinyMice}.
@@
 
-\hypertarget{data-simulation}{%
-\subsubsection{Data Simulation}\label{data-simulation}}
-
-The point of origin for all simulation conditions is a finite population
-of \(N=1000\). The data are simulated to solve a multiple linear
-regression problem, where dependent variable \(Y\) is regressed on
-independent variables \(X_1\), \(X_2\) and \(X_3\):
+A finite population of \(N=1000\) is simulated to solve a multiple
+linear regression problem, where dependent variable \(Y\) is regressed
+on independent variables \(X_1\), \(X_2\) and \(X_3\):
 
@@
 
-The quantity of scientific interest is regression coefficient
-\(\beta_1\). The data generating model is a multivariate normal
-distribution with the following means structure and variance-covariance
-matrix:
+The data generating model is a multivariate normal distribution
 
@@
 \end{pmatrix}
-\end{bmatrix}\\[2\jot].
+\end{bmatrix}\\[2\jot]
 \end{align*}
 
-Data is simulated with the function \texttt{mvtnorm::rmvnorm()}, and
-outcome variable \(Y\) is subsequently calculated as
-\[Y =  2X_1 + .5X_2 - X_3 + \epsilon .\] The complete data is `amputed'
-once for each simulation repetition. That is, the function
-\texttt{mice::ampute()} is used to impose a missingness mechanism upon
-the data. The missingness is univariate, and the probability to be
-missing is the same for all four variables, namely 20\%
-(\texttt{prop\ =\ 0.8,\ mech\ =\ "MCAR"}). This leaves 20\% of the rows
-completely observed. \%The resulting amputed data is equal for all
-simulation conditions in the same repetition.
+and simulated with the function \texttt{mvtnorm::rmvnorm()}. Outcome
+variable \(Y\) is subsequently calculated as
+\[Y =  2X_1 + .5X_2 - X_3 + \epsilon .\]
+
+The complete data is `amputed' once for each simulation repetition. That
+is, the function \texttt{mice::ampute()} is used to impose a missingness
+mechanism upon the data. \textbf{(change this: The missingness is
+univariate, and the probability to be missing is the same for all four
+variables, namely 20\% (\texttt{prop\ =\ 0.8,\ mech\ =\ "MCAR"}). This
+leaves 20\% of the rows completely observed). }
 
@@
 
-From the imputation step, we extract chain means and chain variances
-across iterations (\(\theta\) at \(t = 1,2,\dots,T\)).\\
-We compute convergence diagnostics separately for both \(\theta\)s, and
-for each of the four variables in the dataset. We report the maximum
-(absolute) value as the `worst linear predictor' of convergence.
+\textbf{(tot hier doorgelezen)}
+
 \(\widehat{R}\) is computed by implementing Vehtari et al.'s (2019)
@@
 \begin{itemize}
-\item
-  Remove the table.
+\tightlist
 \item
@@
 
-Figures 2 and 3 display results per simulation condition
-(\(T = 1,2,\dots,100\)).
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{# plot results}
+\KeywordTok{load}\NormalTok{(}\StringTok{"../Results/results.Rdata"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
 
 \hypertarget{convergence-diagnostics-1}{%
-\subsubsection{Convergence
-Diagnostics}\label{convergence-diagnostics-1}}
+\subsection{Convergence Diagnostics}\label{convergence-diagnostics-1}}
 
@@
 \hypertarget{simulation-diagnostics}{%
-\subsubsection{Simulation Diagnostics}\label{simulation-diagnostics}}
+\subsection{Simulation Diagnostics}\label{simulation-diagnostics}}
 

### 2020-03-22 3.Thesis/1.SimulationStudy/2.CreateFigures.R

@@ -569,3 +569,5 @@ pred_plot <-
 
 # print figure
 pred_plot
+
+

### 2020-03-23 3.Thesis/1.SimulationStudy/2.CreateFigures.R
Plots with patchwork package instead of ggarrange

@@ -5,224 +5,16 @@
 # load package
 library("dplyr")
 library("ggplot2")
-library("ggpubr")
+library("patchwork")
 
-# if necessary, load results
-# load("C:/Users/User/Desktop/shinyMice/2.ResearchReport/1.SimulationStudy/Results/full_results.Rdata")
-# results <- dat
-# results_with_CI <- dat #results
+# # if necessary, load results
+# load("C:/Users/User/Desktop/shinyMice/3.Thesis/1.SimulationStudy/Results/results.Rdata")
+# results_with_CI <- dat
 
-# # create plot for R hat
-# R_plot <- results_with_CI %>% .[-1,] %>%
-#   ggplot(aes(x = T, y = R.var.X1, color = "Chain variances")) +
-#   geom_line(linetype = "dashed", size = 1) +
-#   geom_line(aes(x = T, y = R.var.X2, color = "Chain variances"),
-#             linetype = "dashed",
-#             size = 1) +
-#   geom_line(aes(x = T, y = R.var.X3, color = "Chain variances"),
-#             linetype = "dashed",
-#             size = 1) +
-#   geom_line(aes(x = T, y = R.var.Y, color = "Chain variances"),
-#             linetype = "dashed",
-#             size = 1) +
-#   geom_line(aes(x = T, y = R.mean.X1, color = "Chain means")) +
-#   geom_line(aes(x = T, y = R.mean.X2, color = "Chain means")) +
-#   geom_line(aes(x = T, y = R.mean.X3, color = "Chain means")) +
-#   geom_line(aes(x = T, y = R.mean.Y, color = "Chain means")) +
-#   # # add error bar when `results_withSEs` is used
-#   # geom_errorbar(
-#   #   aes(ymin = R.mean.X1_LL, ymax = R.mean.X1_UL),
-#   #   width = .2,
-#   #   position = position_dodge(0.05)
-#   # ) +
-#   xlab("") +
-#   ylab(expression(paste(widehat(R)))) +
-#   scale_x_continuous(breaks = seq(0, 100, by = 10)) +
-#   scale_y_continuous(breaks = seq(.8, 2.2, by = .2)) +
-#   scale_color_manual(name = "Legend", labels = c("Chain means", "Chain variances"), values = c("black", "grey"), guide = "legend") +
[…]

### 2020-03-26 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Ran sim to show Gerko new plots (mean, var, est, pred).

@@
 \hypersetup{
-  pdftitle={Missing the Point: Convergence of Multiple Imputation by Chained Equations Algorithms},
+  pdftitle={Missing the Point: Detecting Non-Convergence in Multiple Imputation Procedures},
   pdfauthor={Hanne I. Oberman},
@@
 
-\title{Missing the Point: Convergence of Multiple Imputation by Chained
-Equations Algorithms}
+\title{Missing the Point: Detecting Non-Convergence in Multiple Imputation
+Procedures}
 \author{Hanne I. Oberman}
@@
 \textbf{(tot hier doorgelezen)}
+
+What to evaluate?
+
+\begin{itemize}
+\item
+  bias in regression coefficients
+\item
+  bias in \(R^2\) (= coefficient of determination)
+\item
+\end{itemize}
 

### 2020-03-30 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
Plots with patchwork package instead of ggarrange
Ran sim to show Gerko new plots (mean, var, est, pred).

@@
 knitr::opts_chunk$set(echo = TRUE)
+load("../Results/results.Rdata")
+source("../2.CreateFigures.R")
+
 ```
@@
 
+What to evaluate?
+
+- bias in regression coefficients
+
+- bias in $R^2$ (= coefficient of determination)
+
+-
+
 <!-- From the imputation step, we extract chain means and chain variances across iterations ($\theta$ at $t = 1,2,\dots,T$).   -->
@@
 
-```{r}
-# plot results
-load("../Results/results.Rdata")
+```{r, echo=FALSE, fig.cap="Convergence diagnostics (means)"}
+bias_means_plot + R_means_plot + AC_means_plot + plot_layout(nrow = 3, guides = "collect")
+```
 
+```{r, echo=FALSE, fig.cap="Convergence diagnostics (variances)"}
+bias_sds_plot + R_sds_plot + AC_sds_plot + plot_layout(nrow = 3, guides = "collect")
+```
+
+```{r, echo=FALSE, fig.cap="Convergence diagnostics (regression coefficients)"}
+bias_ests_plot + bias_R_sq_plot + cov_plot + plot_layout(nrow = 3, guides = "collect")
 ```

### 2020-03-30 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
Use rticles package instead of resolving issues in csl myself.

--- 2020-03-22 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
@@
 ---
-title: "Missing the Point: Convergence of Multiple Imputation by Chained Equations Algorithms"
-author: "Hanne I. Oberman"
-date: "16-3-2020"
+title: Missing the Point$\colon$ Non-Convergence in Multiple Imputation Procedures
+runninghead: Oberman
+author:
+- name: H. I. Oberman
+  num: 1
+address:
+- num: 1
+  org: Department of Methodology and Statistics, Utrecht University, Utrecht, The Netherlands
+corrauth: "Hanne Oberman, Sjoerd Groenman building, Utrecht Science Park, Utrecht, The Netherlands."
+email: h.i.oberman@uu.nl
+abstract: "Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Aenean ut elit odio.
+Donec fermentum tellus neque, vitae fringilla orci pretium vitae.
+Fusce maximus finibus facilisis.
+Donec ut ullamcorper turpis.
+Donec ut porta ipsum.
+Nullam cursus mauris a sapien ornare pulvinar.
+Aenean malesuada molestie erat quis mattis.
+Praesent scelerisque posuere faucibus.
+Praesent nunc nulla, ullamcorper ut ullamcorper sed, molestie ut est.
+Donec consequat libero nisi, non semper velit vulputate et.
+Quisque eleifend tincidunt ligula, bibendum finibus massa cursus eget."
+keywords: MICE; convergence; 
+classoption:
+  - Royal
+  - times
+bibliography: thesis.bib
+bibliographystyle: sageh
 output:
-  pdf_document:
-    keep_tex: true
-    number_sections: true
-bibliography: thesis.bib
-documentclass: sagej
-classoption: Royal, times, sageapa
+  rticles::sage_article:
+    keep_tex: yes
 ---
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+#knitr::opts_chunk$set(echo = TRUE)
+load("../Results/results.Rdata")
+source("../2.CreateFigures.R")
+
 ```
@@
 
-Feedback:
-
-- Add more info about figure legends and axes.
+**Add more info about figure legends and axes.**
 
@@
 
-```{r}
-# plot results
-load("../Results/results.Rdata")
+```{r, fig.cap="Convergence diagnostics (means)"}
+bias_means_plot + R_means_plot + AC_means_plot + plot_layout(nrow = 3, guides = "collect")
+```
 
+```{r, fig.cap="Convergence diagnostics (variances)"}
+bias_sds_plot + R_sds_plot + AC_sds_plot + plot_layout(nrow = 3, guides = "collect")
+```
+
+```{r, fig.cap="Convergence diagnostics (regression coefficients)"}
+bias_ests_plot + bias_R_sq_plot + cov_plot + plot_layout(nrow = 3, guides = "collect")
 ```
@@
 
-# References
+<!-- # References -->

### 2020-03-30 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Included plots with new results in manuscript while working in wrong branch. Going to fix it now.
Use rticles package instead of resolving issues in csl myself.

@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-%
-\documentclass[
-  Royal, times, sageapa]{sagej}
-\usepackage{lmodern}
-\usepackage{amssymb,amsmath}
-\usepackage{ifxetex,ifluatex}
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  pdftitle={Missing the Point: Detecting Non-Convergence in Multiple Imputation Procedures},
-  pdfauthor={Hanne I. Oberman},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
-\usepackage{color}
-\usepackage{fancyvrb}
-\newcommand{\VerbBar}{|}
-\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
-\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
-% Add ',fontsize=\small' for more characters per line
-\usepackage{framed}
-\definecolor{shadecolor}{RGB}{248,248,248}
-\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
-\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
-\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
-\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\BuiltInTok}[1]{#1}
-\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
-\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
-\newcommand{\ExtensionTok}[1]{#1}
-\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ImportTok}[1]{#1}
-\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\NormalTok}[1]{#1}
-\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
-\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
-\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\RegionMarkerTok}[1]{#1}
-\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\usepackage{graphicx,grffile}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{5}
+\documentclass[Royal,times,sageh]{sagej}
 
-\title{Missing the Point: Detecting Non-Convergence in Multiple Imputation
-Procedures}
-\author{Hanne I. Oberman}
-\date{16-3-2020}
+\usepackage{moreverb,url,natbib, multirow, tabularx}
+\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}
 
 \begin{document}
+
+\title{Missing the Point\(\colon\) Non-Convergence in Multiple Imputation
+Procedures}
+
+\runninghead{Oberman}
+
+\author{H. I. Oberman\affilnum{1}}
+
+\affiliation{\affilnum{1}{Department of Methodology and Statistics, Utrecht University, Utrecht,
+The Netherlands}}
+
+\corrauth{Hanne Oberman, Sjoerd Groenman building, Utrecht Science Park, Utrecht,
+The Netherlands.}
+
+\email{\href{mailto:h.i.oberman@uu.nl}{\nolinkurl{h.i.oberman@uu.nl}}}
+
+\begin{abstract}
+Lorem ipsum dolor sit amet, consectetur adipiscing elit. Aenean ut elit
+odio. Donec fermentum tellus neque, vitae fringilla orci pretium vitae.
+Fusce maximus finibus facilisis. Donec ut ullamcorper turpis. Donec ut
+porta ipsum. Nullam cursus mauris a sapien ornare pulvinar. Aenean
+malesuada molestie erat quis mattis. Praesent scelerisque posuere
+faucibus. Praesent nunc nulla, ullamcorper ut ullamcorper sed, molestie
+ut est. Donec consequat libero nisi, non semper velit vulputate et.
+Quisque eleifend tincidunt ligula, bibendum finibus massa cursus eget.
+Curabitur aliquet vehicula quam non pulvinar. Aliquam facilisis tortor
+nec purus finibus, sit amet elementum eros sodales. Ut porta porttitor
+vestibulum. Integer molestie, leo ut maximus aliquam, velit dui iaculis
+nibh, eget hendrerit purus risus sit amet dolor. Sed sed tincidunt ex.
+Curabitur imperdiet egestas tellus in iaculis. Maecenas ante neque,
+pretium vel nisl at, lobortis lacinia neque. In gravida elit vel
+volutpat imperdiet. Sed ut nulla arcu. Proin blandit interdum ex sit
+amet laoreet. Phasellus efficitur, sem hendrerit mattis dapibus, nunc
+tellus ornare nisi, nec eleifend enim nibh ac ipsum. Aenean tincidunt
+nisl sit amet facilisis faucibus. Donec odio erat, bibendum eu imperdiet
+sed, gravida luctus turpis.
+\end{abstract}
+
+\keywords{MICE; convergence;}
+
 \maketitle
@@
 At some point, any scientist conducting statistical analyses will run
-into a missing data problem (Allison 2001). Missingness is problematic
+into a missing data problem \citep{alli02}. Missingness is problematic
 because statistical inference cannot be performed on incomplete data
 without employing \emph{ad hoc} solutions (e.g., list-wise deletion),
-which may yield wildly invalid results (Van Buuren 2018). A popular
-answer to the ubiquitous problem of missing information is to use the
-framework of multiple imputation (MI), proposed by Rubin (1987). For the
-theoretical foundation of MI, see Rubin (1987). For an accessible and
+which may yield wildly invalid results \citep{buur18}. A popular answer
+to the ubiquitous problem of missing information is to use the framework
+of multiple imputation (MI), proposed by \citet{rubin87}. For the
+theoretical foundation of MI, see \citet{rubin87}. For an accessible and
 comprehensive introduction to MI from an applied perspective, see
-e.g.~Van Buuren (2018). With MI, missing datapoints are `imputed' (i.e.,
+e.g.~\citet{buur18}. With MI, missing datapoints are `imputed' (i.e.,
 filled in) via an \emph{imputation model}. The imputation model can be a
@@
 conditional on the observed data.'' There is no scientific consensus on
-the convergence properties of FCS algorithms (Takahashi 2017). Some
+the convergence properties of FCS algorithms \citep{taka17}. Some
 default techniques in \texttt{mice} might not yield converged states at
-all (Murray 2018). Therefore, algorithmic convergence should be
+all \citep{murr18}. Therefore, algorithmic convergence should be
 monitored carefully.
@@
 convergence. Therefore we can only diagnose non-convergence, not
-convergence (Hoff 2009). ``a weak diagnostics is better than no
-diagnostic at all'' (Cowles and Carlin 1996).
+convergence \citep{hoff09}. ``a weak diagnostics is better than no
+diagnostic at all'' \citep{cowl96}.
 
 Current practice is visually inspecting imputation chains for signs of
-non-convergence, but may be challenging to the untrained eye (Van Buuren
-2018, \(\S\) 6.5.2). Therefore diagnostic evaluation of convergence
-would be preferred. However, not known whether MCMC diagnostics are
-appropriate for MICE \ldots{} The application of convergence diagnostics
-to MI methods has not been systematically studied (Van Buuren 2018). How
-can non-convergence be diagnosed? How many iterations are
-sufficient/required? What are the effects of non-convergence on
-estimates, predictions and inference? What is an appropriate
-non-convergence diagnostic? Are the default number of iterations
-sufficient? Defaults in: SPSS = 10
+non-convergence, but may be challenging to the untrained eye
+\citep[\(\S\) 6.5.2]{buur18}. Therefore diagnostic evaluation of
+convergence would be preferred. However, not known whether MCMC
+diagnostics are appropriate for MICE \ldots{} The application of
+convergence diagnostics to MI methods has not been systematically
+studied \citep{buur18}. How can non-convergence be diagnosed? How many
+iterations are sufficient/required? What are the effects of
+non-convergence on estimates, predictions and inference? What is an
+appropriate non-convergence diagnostic? Are the default number of
+iterations sufficient? Defaults in: SPSS = 10
 \href{https://www.ibm.com/support/knowledgecenter/SSLVMB_24.0.0/spss/mva/syn_multiple_imputation_impute.html}{(link)},
@@
 For reasons of brevity, we only focus on the MI algorithm implemented in
-the world-leading MI software: the \texttt{R} (R Core Team 2019) package
-\texttt{mice} (Van Buuren and Groothuis-Oudshoorn 2011). The convergence
-properties of this MI algorithm are investigated through model-based
-simulation. The results of this simulation study are guidelines for
-assessing convergence of MI algorithms, which will aid applied
-researchers in drawing valid inference from incomplete datasets.
+the world-leading MI software: the \texttt{R} \citep{R} package
+\texttt{mice} \citep{mice}. The convergence properties of this MI
+algorithm are investigated through model-based simulation. The results
+of this simulation study are guidelines for assessing convergence of MI
+algorithms, which will aid applied researchers in drawing valid
+inference from incomplete datasets.
 
@@
 The stationarity component of convergence may be evaluated with
-autocorrelation (\(AC\); Schafer 1997; Gelman et al. 2013), numeric
-standard error (or `MC error'; Geweke 1992), and Raftery and Lewis's
-(1991) procedure to determine the effect of trending within chains.
+autocorrelation \citep[\(AC\);][]{scha97, gelm13}, numeric standard
+error \citep[or `MC error';][]{gewe92}, and Raftery and Lewis's
+\citeyearpar{raft91} procedure to determine the effect of trending
+within chains.
 
 The mixing component can be assessed with the potential scale reduction
-factor \(\widehat{R}\) (a.k.a. `Gelman-Rubin statistic'; Gelman and
-Rubin 1992). With an adapted version of \(\widehat{R}\), proposed by
-Vehtari et al. (2019), we might also evaluate the stationarity component
-of convergence. This would make \(\widehat{R}\) a general convergence
-diagnostic. The application of \(\widehat{R}\) to assess stationarity
-has not been thoroughly investigated. Therefore, this study employs both
-\(\widehat{R}\) and autocorrelation to investigate convergence, as
-recommended by (Cowles and Carlin 1996, 898).
+factor \(\widehat{R}\) \citep[a.k.a. `Gelman-Rubin
+statistic';][]{gelm92}. With an adapted version of \(\widehat{R}\),
+proposed by \citet{veht19}, we might also evaluate the stationarity
+component of convergence. This would make \(\widehat{R}\) a general
+convergence diagnostic. The application of \(\widehat{R}\) to assess
+stationarity has not been thoroughly investigated. Therefore, this study
+employs both \(\widehat{R}\) and autocorrelation to investigate
+convergence, as recommended by \citep[p.~898]{cowl96}.
 
@@
 diagnostics cannot diagnose convergence of multivariable statistics
-(i.e., relations between scalar summaries). Van Buuren (2018) proposed
-to implement multivariable evaluation through eigenvalue decomposition
-(MacKay and Mac Kay 2003). This method is outside of the scope of the
-current study (\textbf{not anymore!}).
+(i.e., relations between scalar summaries). \citet{buur18} proposed to
+implement multivariable evaluation through eigenvalue decomposition
+\citep{mack03}. This method is outside of the scope of the current study
+(\textbf{not anymore!}).
 
@@
 
-To define \(\widehat{R}\), we follow notation by (Vehtari et al. 2019,
-5). Let \(M\) be the total number of chains, \(T\) the number of
-iterations per chain, and \(\theta\) the scalar summary of interest
-(e.g., chain mean or chain variance). For each chain
-(\(m = 1, 2, \dots, M\)), we estimate the variance of \(\theta\), and
-average these to obtain within-chain variance \(W\).
+To define \(\widehat{R}\), we follow notation by \citep[p.~5]{veht19}.
+Let \(M\) be the total number of chains, \(T\) the number of iterations
+per chain, and \(\theta\) the scalar summary of interest (e.g., chain
+mean or chain variance). For each chain (\(m = 1, 2, \dots, M\)), we
+estimate the variance of \(\theta\), and average these to obtain
+within-chain variance \(W\).
 
@@
 shrunken down if an infinite number of iterations per chain would be run
-(Gelman and Rubin 1992). This interpretation assumes that chains are
+\citep{gelm92}. This interpretation assumes that chains are
 `over-dispersed' at \(t=1\), and reach convergence as \(T \to \infty\).
@@
 conventionally acceptable threshold for convergence was
-\(\widehat{R} < 1.2\) (Gelman and Rubin 1992). More recently, Vehtari et
-al. (2019) proposed a more stringent threshold of
-\(\widehat{R} < 1.01\).
+\(\widehat{R} < 1.2\) \citep{gelm92}. More recently, \citet{veht19}
+proposed a more stringent threshold of \(\widehat{R} < 1.01\).
 
@@
 correlation between two subsequent \(\theta\)-values within the same
-chain (Lynch 2007, 147). In this study we only consider \(AC\) at lag 1,
-i.e., the correlation between the \(t^{th}\) and \(t+1^{th}\) iteration
-of the same chain.
+chain \citep[p.~147]{lync07}. In this study we only consider \(AC\) at
+lag 1, i.e., the correlation between the \(t^{th}\) and \(t+1^{th}\)
+iteration of the same chain.
 
@@
 of the two convergence diagnostics against the recommended evaluation
-criteria for MI methods (i.e., average bias, average confidence interval
-width, and empirical coverage rate across simulations; Van Buuren 2018,
-\(\S\) 2.5.2). That is, there is no baseline measure available to
+criteria for MI methods \citep[i.e., average bias, average confidence
+interval width, and empirical coverage rate across simulations;][\(\S\)
+2.5.2]{buur18}. That is, there is no baseline measure available to
 evaluate performance against.The current practice of visually inspecting
@@
 look at distributional characteristics, and plausibility of imputed
-values, see Vink (n.d.). For now, this is outside of the scope of this
-study.
+values, see \citet{vinknd}. For now, this is outside of the scope of
+this study.
 
-Based on an empirical finding (Lacerda, Ardington, and Leibbrandt 2007),
-we hypothesize that \(\widehat{R}\) will over-estimate non-convergence
-of MI algorithms. The threshold of \(\widehat{R} < 1.01\) will then be
-too stringent for diagnosing convergence. This over-estimation may,
-however, be diminished because \(\widehat{R}\) can falsely diagnose
-convergence if initial values of the algorithm are not appropriately
-over-dispersed (Brooks and Gelman 1998, 437). In \texttt{mice}, initial
-values are chosen randomly from the observed data. Therefore, we cannot
-be certain that the initial values are over-dispersed. We expect this to
-have little effect on the hypothesized performance of \(\widehat{R}\).
-No hypothesis was formulated about the performance of \(AC\) as
-convergence diagnostic.
+Based on an empirical finding \citep{lace07}, we hypothesize that
+\(\widehat{R}\) will over-estimate non-convergence of MI algorithms. The
+threshold of \(\widehat{R} < 1.01\) will then be too stringent for
+diagnosing convergence. This over-estimation may, however, be diminished
+because \(\widehat{R}\) can falsely diagnose convergence if initial
+values of the algorithm are not appropriately over-dispersed
+\citep[p.~437]{broo98}. In \texttt{mice}, initial values are chosen
+randomly from the observed data. Therefore, we cannot be certain that
+the initial values are over-dispersed. We expect this to have little
+effect on the hypothesized performance of \(\widehat{R}\). No hypothesis
+was formulated about the performance of \(AC\) as convergence
+diagnostic.
 
@@
 The convergence of \texttt{mice} is investigated through model-based
-simulation in \texttt{R} (version 3.6.3; R Core Team 2019). The
-simulation set-up is summarized in the pseudo-code below. The complete
-\texttt{R} script of the simulation study is available from
+simulation in \texttt{R} \citep[version 3.6.3;][]{R}. The simulation
+set-up is summarized in the pseudo-code below. The complete \texttt{R}
+script of the simulation study is available from
 \href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy}{github.com/gerkovink/shinyMice}.
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{# pseudo-code of simulation }
-\NormalTok{simulate data }
-\ControlFlowTok{for}\NormalTok{ (number of simulation runs from }\DecValTok{1}\NormalTok{ to }\DecValTok{1000}\NormalTok{)}
-  \ControlFlowTok{for}\NormalTok{ (number of iterations from }\DecValTok{1}\NormalTok{ to }\DecValTok{100}\NormalTok{)}
-\NormalTok{    create missingness}
-\NormalTok{    impute the missingness}
-\NormalTok{    compute convergence diagnostics}
-\NormalTok{    perform analysis}
-\NormalTok{    pool results}
-\NormalTok{    compute simulation diagnostics}
-\NormalTok{aggregate convergence and simulation diagnostics}
-\end{Highlighting}
-\end{Shaded}
+\begin{verbatim}
+# pseudo-code of simulation 
+simulate data 
+for (number of simulation runs from 1 to 1000)
+  for (number of iterations from 1 to 100)
+    create missingness
+    impute the missingness
+    compute convergence diagnostics
+    perform analysis
+    pool results
+    compute simulation diagnostics
+aggregate convergence and simulation diagnostics
+\end{verbatim}
 
@@
 
-\(\widehat{R}\) is computed by implementing Vehtari et al.'s (2019)
-recommendations. \(AC\) is computed with function \texttt{stats::acf()}.
+\(\widehat{R}\) is computed by implementing Vehtari et al.'s
+\citeyearpar{veht19} recommendations. \(AC\) is computed with function
+\texttt{stats::acf()}.
 
@@
 the function \texttt{mice::pool()} to get variance estimates according
-to Rubin's (1987) rules, and subsequently implement finite population
-pooling conform Vink and Buuren (2014).
+to Rubin's \citeyearpar{rubin87} rules, and subsequently implement
+finite population pooling conform \citet{vink14}.
 
@@
 
-Feedback:
+\textbf{Add more info about figure legends and axes.}
 
-\begin{itemize}
-\tightlist
-\item
-  Add more info about figure legends and axes.
-\end{itemize}
+\begin{verbatim}
+bias_means_plot + R_means_plot + AC_means_plot + plot_layout(nrow = 3, guides = "collect")
+\end{verbatim}
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{# plot results}
-\KeywordTok{load}\NormalTok{(}\StringTok{"../Results/results.Rdata"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
+\begin{figure}
+\centering
+\includegraphics{manuscript_files/figure-latex/unnamed-chunk-2-1.pdf}
+\caption{Convergence diagnostics (means)}
+\end{figure}
+
+\begin{verbatim}
+bias_sds_plot + R_sds_plot + AC_sds_plot + plot_layout(nrow = 3, guides = "collect")
+\end{verbatim}
+
+\begin{figure}
+\centering
+\includegraphics{manuscript_files/figure-latex/unnamed-chunk-3-1.pdf}
+\caption{Convergence diagnostics (variances)}
+\end{figure}
+
+\begin{verbatim}
+bias_ests_plot + bias_R_sq_plot + cov_plot + plot_layout(nrow = 3, guides = "collect")
+\end{verbatim}
+
+\begin{figure}
+\centering
+\includegraphics{manuscript_files/figure-latex/unnamed-chunk-4-1.pdf}
+\caption{Convergence diagnostics (regression coefficients)}
+\end{figure}
 
@@
 \(\bar{Q}\). On average, the coverage rate is somewhat higher than the
-expected nominal coverage of 95\% (Neyman 1934), namely 95\%. Similar to
-the CIWs, there is some trending across iterations (i.e., the Loess line
-is never flat).
+expected nominal coverage of 95\% \citep{neym34}, namely 95\%. Similar
+to the CIWs, there is some trending across iterations (i.e., the Loess
+line is never flat).
 
@@
 
-\hypertarget{references}{%
-\section*{References}\label{references}}
-\addcontentsline{toc}{section}{References}
-
-\hypertarget{refs}{}
-\leavevmode\hypertarget{ref-alli02}{}%
-Allison, Paul D. 2001. \emph{Missing Data}. Sage publications.
-
-\leavevmode\hypertarget{ref-broo98}{}%
-Brooks, Stephen P., and Andrew Gelman. 1998. ``General Methods for
-Monitoring Convergence of Iterative Simulations.'' \emph{Journal of
-Computational and Graphical Statistics} 7 (4): 434--55.
-\url{https://doi.org/10.1080/10618600.1998.10474787}.
-
-\leavevmode\hypertarget{ref-cowl96}{}%
-Cowles, Mary Kathryn, and Bradley P Carlin. 1996. ``Markov Chain Monte
-Carlo Convergence Diagnostics: A Comparative Review.'' \emph{Journal of
-the American Statistical Association} 91 (434): 883--904.
-
-\leavevmode\hypertarget{ref-gelm13}{}%
-Gelman, Andrew, John B. Carlin, Hal S. Stern, David B. Dunson, Aki
-Vehtari, and Donald B. Rubin. 2013. \emph{Bayesian Data Analysis}.
-Philadelphia, PA, United States: CRC Press LLC.
-
-\leavevmode\hypertarget{ref-gelm92}{}%
-Gelman, Andrew, and Donald B. Rubin. 1992. ``Inference from Iterative
-Simulation Using Multiple Sequences.'' \emph{Statistical Science} 7 (4):
-457--72. \url{https://doi.org/10.1214/ss/1177011136}.
-
-\leavevmode\hypertarget{ref-gewe92}{}%
-Geweke, John. 1992. ``Evaluating the Accuracy of Sampling-Based
-Approaches to the Calculations of Posterior Moments.'' \emph{Bayesian
-Statistics} 4: 641--49.
-
-\leavevmode\hypertarget{ref-hoff09}{}%
-Hoff, Peter D. 2009. \emph{A First Course in Bayesian Statistical
-Methods}. Springer Texts in Statistics. New York, NY: Springer New York.
-\url{https://doi.org/10.1007/978-0-387-92407-6}.
-
-\leavevmode\hypertarget{ref-lace07}{}%
-Lacerda, Miguel, Cally Ardington, and Murray Leibbrandt. 2007.
-``Sequential Regression Multiple Imputation for Incomplete Multivariate
-Data Using Markov Chain Monte Carlo.'' University of Cape Town, South
-Africa.
-
-\leavevmode\hypertarget{ref-lync07}{}%
-Lynch, Scott M. 2007. \emph{Introduction to Applied Bayesian Statistics
-and Estimation for Social Scientists}. Springer Science \& Business
-Media.
-
-\leavevmode\hypertarget{ref-mack03}{}%
-MacKay, David JC, and David JC Mac Kay. 2003. \emph{Information Theory,
-Inference and Learning Algorithms}. Cambridge university press.
-
-\leavevmode\hypertarget{ref-murr18}{}%
-Murray, Jared S. 2018. ``Multiple Imputation: A Review of Practical and
-Theoretical Findings.'' \emph{Statistical Science} 33 (2): 142--59.
-\url{https://doi.org/10.1214/18-STS644}.
-
-\leavevmode\hypertarget{ref-neym34}{}%
-Neyman, Jerzy. 1934. ``On the Two Different Aspects of the
-Representative Method: The Method of Stratified Sampling and the Method
-of Purposive Selection.'' \emph{Journal of the Royal Statistical
-Society} 97 (4): 558--625. \url{https://doi.org/10.2307/2342192}.
-
-\leavevmode\hypertarget{ref-raft91}{}%
-Raftery, Adrian E, and Steven Lewis. 1991. ``How Many Iterations in the
-Gibbs Sampler?'' Washington University Seatle, Department of Statistics,
-United States.
-
-\leavevmode\hypertarget{ref-R}{}%
-R Core Team. 2019. \emph{R: A Language and Environment for Statistical
-Computing}. Vienna, Austria. \url{https://www.R-project.org/}.
-
-\leavevmode\hypertarget{ref-rubin87}{}%
-Rubin, Donald B. 1987. \emph{Multiple Imputation for Nonresponse in
-Surveys}. Wiley Series in Probability and Mathematical Statistics
-Applied Probability and Statistics. New York, NY: Wiley.
-
-\leavevmode\hypertarget{ref-scha97}{}%
-Schafer, Joseph L. 1997. \emph{Analysis of Incomplete Multivariate
-Data}. Chapman; Hall/CRC.
-
-\leavevmode\hypertarget{ref-taka17}{}%
-Takahashi, Masayoshi. 2017. ``Statistical Inference in Missing Data by
-MCMC and Non-MCMC Multiple Imputation Algorithms: Assessing the Effects
-of Between-Imputation Iterations.'' \emph{Data Science Journal} 16
-(July): 37. \url{https://doi.org/10.5334/dsj-2017-037}.
-
-\leavevmode\hypertarget{ref-buur18}{}%
-Van Buuren, Stef. 2018. \emph{Flexible Imputation of Missing Data}.
-Chapman; Hall/CRC.
-
-\leavevmode\hypertarget{ref-mice}{}%
-Van Buuren, Stef, and Karin Groothuis-Oudshoorn. 2011. ``Mice:
-Multivariate Imputation by Chained Equations in R.'' \emph{Journal of
-Statistical Software} 45 (1): 1--67.
-\url{https://doi.org/10.18637/jss.v045.i03}.
-
-\leavevmode\hypertarget{ref-veht19}{}%
-Vehtari, Aki, Andrew Gelman, Daniel Simpson, Bob Carpenter, and
-Paul-Christian Bürkner. 2019. ``Rank-Normalization, Folding, and
-Localization: An Improved \$\textbackslash widehat\{R\}\$ for Assessing
-Convergence of MCMC,'' March. \url{http://arxiv.org/abs/1903.08008}.
-
-\leavevmode\hypertarget{ref-vinknd}{}%
-Vink, Gerko. n.d. ``Towards a Standardized Evaluation of Multiple
-Imputation Routines.''
-
-\leavevmode\hypertarget{ref-vink14}{}%
-Vink, Gerko, and Stef van Buuren. 2014. ``Pooling Multiple Imputations
-When the Sample Happens to Be the Population.'' \emph{arXiv:1409.8542
-{[}Math, Stat{]}}, September. \url{http://arxiv.org/abs/1409.8542}.
+\bibliographystyle{sageh}
+\bibliography{thesis.bib}
 

### 2020-03-31 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Committing to switch branches. Not much changed.

@@
 Quisque eleifend tincidunt ligula, bibendum finibus massa cursus eget.
-Curabitur aliquet vehicula quam non pulvinar. Aliquam facilisis tortor
-nec purus finibus, sit amet elementum eros sodales. Ut porta porttitor
-vestibulum. Integer molestie, leo ut maximus aliquam, velit dui iaculis
-nibh, eget hendrerit purus risus sit amet dolor. Sed sed tincidunt ex.
-Curabitur imperdiet egestas tellus in iaculis. Maecenas ante neque,
-pretium vel nisl at, lobortis lacinia neque. In gravida elit vel
-volutpat imperdiet. Sed ut nulla arcu. Proin blandit interdum ex sit
-amet laoreet. Phasellus efficitur, sem hendrerit mattis dapibus, nunc
-tellus ornare nisi, nec eleifend enim nibh ac ipsum. Aenean tincidunt
-nisl sit amet facilisis faucibus. Donec odio erat, bibendum eu imperdiet
-sed, gravida luctus turpis.
 \end{abstract}

### 2020-03-31 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
Ran with 25 and 50% missingness

@@
 email: h.i.oberman@uu.nl
-abstract: "Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Aenean ut elit odio.
-Donec fermentum tellus neque, vitae fringilla orci pretium vitae.
-Fusce maximus finibus facilisis.
-Donec ut ullamcorper turpis.
-Donec ut porta ipsum.
-Nullam cursus mauris a sapien ornare pulvinar.
-Aenean malesuada molestie erat quis mattis.
-Praesent scelerisque posuere faucibus.
-Praesent nunc nulla, ullamcorper ut ullamcorper sed, molestie ut est.
-Donec consequat libero nisi, non semper velit vulputate et.
-Quisque eleifend tincidunt ligula, bibendum finibus massa cursus eget."
-keywords: MICE; convergence; 
+abstract: "Multple imputation by chained equations (mice) is a widely used tool to accommodate missing data.
+While empirical evidence supports the validity of inferences obtained using mice, there is no consensus on the convergence properties of the method.
+This paper provides insight into non-convergence of mice algorithms."
+keywords: MICE; convergence 
 classoption:
@@
 and simulated with the function `mvtnorm::rmvnorm()`. Outcome variable $Y$ is subsequently calculated as 
-$$Y =  2X_1 + .5X_2 - X_3 + \epsilon .$$
+$$Y =  1 + 2X_1 + .5X_2 - X_3 + \epsilon .$$
 <!-- This note only considers a 'missing completely at random' (MCAR) mechanism, where the probability of being missing is equal for all $n \times p$ cells in $y$ [@rubin87].  -->

### 2020-03-31 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Ran with 25 and 50% missingness

@@
 \begin{abstract}
-Lorem ipsum dolor sit amet, consectetur adipiscing elit. Aenean ut elit
-odio. Donec fermentum tellus neque, vitae fringilla orci pretium vitae.
-Fusce maximus finibus facilisis. Donec ut ullamcorper turpis. Donec ut
-porta ipsum. Nullam cursus mauris a sapien ornare pulvinar. Aenean
-malesuada molestie erat quis mattis. Praesent scelerisque posuere
-faucibus. Praesent nunc nulla, ullamcorper ut ullamcorper sed, molestie
-ut est. Donec consequat libero nisi, non semper velit vulputate et.
-Quisque eleifend tincidunt ligula, bibendum finibus massa cursus eget.
+Multple imputation by chained equations (mice) is a widely used tool to
+accommodate missing data. While empirical evidence supports the validity
+of inferences obtained using mice, there is no consensus on the
+convergence properties of the method. This paper provides insight into
+non-convergence of mice algorithms.
 \end{abstract}
 
-\keywords{MICE; convergence;}
+\keywords{MICE; convergence}
 
@@
 variable \(Y\) is subsequently calculated as
-\[Y =  2X_1 + .5X_2 - X_3 + \epsilon .\]
+\[Y =  1 + 2X_1 + .5X_2 - X_3 + \epsilon .\]
 

### 2020-04-01 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
Read entire research report and put comments in it for thesis manuscript. Ran 5,25,50,75,and 95% miss with 100 sims each.

@@
 #knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(fig.width = 5, echo = FALSE, fig.align = "center") 
 load("../Results/results.Rdata")
@@
 A popular answer to the ubiquitous problem of missing information is to use the framework of multiple imputation (MI), proposed by @rubin87.
-For the theoretical foundation of MI, see @rubin87.
+**(Call it mice already, then explain mi and ce sides of mice)** For the theoretical foundation of MI, see @rubin87.
 For an accessible and comprehensive introduction to MI from an applied perspective, see e.g.
 @buur18.
-With MI, missing datapoints are 'imputed' (i.e., filled in) via an *imputation model*.
-The imputation model can be a joint model for all variables, or several univariate conditional distributions (fully conditional specification).
-FCS is more flexible and "avoids the problem of specifying an appropriate joint imputation distribution and replaces this by the selection of appropriate univariate conditional distributions".
-With FCS, an iterative algorithmic procedure is used to ...
+With MI, missing datapoints are 'imputed' (i.e., filled in) via an *imputation model* (**term not required to understand this paper, so circumvent**) .
+(**Move this to mi vs.
+ce part in intro:** The imputation model can be a joint model for all variables, or several univariate conditional distributions (fully conditional specification).
+FCS is more flexible and "avoids the problem of specifying an appropriate joint imputation distribution and replaces this by the selection of appropriate univariate conditional distributions".)**
+
+**(Rewrite: With FCS, an iterative algorithmic procedure is used to ...
 multiple imputation by chained equations ...
-"The values for imputing the missing data are drawn from the posterior predictive distribution of the missing data conditional on the observed data."  There is no scientific consensus on the convergence properties of FCS algorithms [@taka17].
+"The values for imputing the missing data are drawn from the posterior predictive distribution of the missing data conditional on the observed data.")**
+
+There is no scientific consensus on the convergence properties of FCS algorithms [@taka17].
 Some default techniques in `mice` might not yield converged states at all [@murr18].
@@
 Defaults in: SPSS = 10 [(link)](https://www.ibm.com/support/knowledgecenter/SSLVMB_24.0.0/spss/mva/syn_multiple_imputation_impute.html), Mplus = 100??
-[(link)](https://pdfs.semanticscholar.org/e20e/29e008592cbfbaa567931f74cdfdb5451405.pdf?_ga=2.55354671.54033656.1584698748-527613517.1584698748), Stata = 1; [(link)](https://www.stata.com/manuals13/mi.pdf, p.
+[(link)](https://pdfs.semanticscholar.org/e20e/29e008592cbfbaa567931f74cdfdb5451405.pdf?_ga=2.55354671.54033656.1584698748-527613517.1584698748), Stata = 10 [(link)](https://www.stata.com/manuals13/mi.pdf, p.
 139), Amelia = NA, because resampling, not convergence [(link)](https://cran.r-project.org/web/packages/Amelia/Amelia.pdf), en MI = 30 [(link)](https://cran.r-project.org/web/packages/mi/mi.pdf).
@@
 The collection of observed data values in $y$ is denoted by $y_{obs}$; the missing part of $y$ is referred to as $y_{mis}$.
-For each datapoint in $y_{mis}$ $M \times T$ times plausible values are sampled, where $M$ is the number of imputations ($m = 1, \dots, M$) and $T$ is the number of iterations ($t = 1, \dots, T$).
-The collection of samples between the initial value (at $t=1$) and the imputed value (at $t=T$) will be referred to as an 'imputation chain'.
+**(This is not only notation, but theory too: For each datapoint in $y_{mis}$ $M \times T$ times plausible values are sampled, where $M$ is the number of imputations ($m = 1, \dots, M$) and $T$ is the number of iterations ($t = 1, \dots, T$).)** The collection of samples between the initial value (at $t=1$) and the imputed value (at $t=T$) will be referred to as an 'imputation chain'.
 
@@
 
-In iterative algorithmic procedures (Markov chain Monte Carlo methods; e.g., `mice` algorithms or Gibbs samplers) non-convergence may be present as non-stationarity within chains (i.e., trending), or as slow mixing between chains (i.e., no intermingling).
+**(Start with current guidelines, then continue with possible quant measures of the same things from MCMC lit.)** In iterative algorithmic procedures (Markov chain Monte Carlo methods; e.g., `mice` algorithms or Gibbs samplers) non-convergence may be present as non-stationarity within chains (i.e., trending), or as slow mixing between chains (i.e., no intermingling (**source??**)).
 (**add example/guidelines visual inspection here?**)
 
-The stationarity component of convergence may be evaluated with autocorrelation [$AC$; @scha97; @gelm13], numeric standard error [or `MC error'; @gewe92], and Raftery and Lewis's [-@raft91] procedure to determine the effect of trending within chains.
+The stationarity component of convergence may be evaluated with autocorrelation [$AC$; @scha97; @gelm13], numeric standard error [or `MC error'; @gewe92], and Raftery and Lewis's [-@raft91] procedure to determine (**the chain length required to cancel out trending**) the effect of trending within chains.
 
@@
 `Gelman-Rubin statistic'; @gelm92].
-With an adapted version of $\widehat{R}$, proposed by @veht19, we might also evaluate the stationarity component of convergence.
+With an adapted version of $\widehat{R}$, proposed by @veht19, the stationarity component of convergence might be evaluated as well.
 This would make $\widehat{R}$ a general convergence diagnostic.
@@
 \begin{align*}
-B&=\frac{T}{M-1} \sum_{m=1}^{M}\left(\bar{\theta}^{(\cdot m)}-\bar{\theta}^{(\cdot \cdot)}\right)^{2}, \text { where } \bar{\theta}^{(\cdot m)}=\frac{1}{T} \sum_{t=1}^{T} \theta^{(t m)}, \quad \bar{\theta}^{(\cdot \cdot)}=\frac{1}{M} \sum_{m=1}^{M} \bar{\theta}^{(\cdot m)}.
+B&=\frac{T}{M-1} \sum_{m=1}^{M}\left(\bar{\theta}^{(\cdot m)}-\bar{\theta}^{(\cdot \cdot)}\right)^{2}, \text { where } \bar{\theta}^{(\cdot m)}=\frac{1}{T} \sum_{t=1}^{T} \theta^{(t m)} \text{, } \bar{\theta}^{(\cdot \cdot)}=\frac{1}{M} \sum_{m=1}^{M} \bar{\theta}^{(\cdot m)}.
 \end{align*}
@@
 We assess the performance of the two convergence diagnostics against the recommended evaluation criteria for MI methods [i.e., average bias, average confidence interval width, and empirical coverage rate across simulations; @buur18, $\S$ 2.5.2].
-That is, there is no baseline measure available to evaluate performance against.The current practice of visually inspecting imputation chains for signs of non-mixing or non-stationarity can only diagnose severely pathological cases of non-convergence.
-We could also look at distributional characteristics, and plausibility of imputed values, see @vinknd.
-For now, this is outside of the scope of this study.
+That is, there is no baseline measure available to evaluate performance against.
+**(Move above: The current practice of visually inspecting imputation chains for signs of non-mixing or non-stationarity can only diagnose severely pathological cases of non-convergence.)**
+<!-- We could also look at distributional characteristics, and plausibility of imputed values, see @vinknd. For now, this is outside of the scope of this study. -->
 
@@
 We expect this to have little effect on the hypothesized performance of $\widehat{R}$.
-No hypothesis was formulated about the performance of $AC$ as convergence diagnostic.
+**(Add actual hypothesis.)** No hypothesis was formulated about the performance of $AC$ as convergence diagnostic.
 
@@
 
-A finite population of $N=1000$ is simulated to solve a multiple linear regression problem, where dependent variable $Y$ is regressed on independent variables $X_1$, $X_2$ and $X_3$:
+A finite population of $N=1000$ is simulated to solve a multiple linear regression problem, where dependent variable $Y$ is regressed on independent variables $X_1$, $X_2$ and $X_3$ (**check notation with betas here, and add what the quantity/-ies of scientific interest is/are!**):
 
@@
 -->
-$\widehat{R}$ is computed by implementing Vehtari et al.'s [-@veht19] recommendations. $AC$ is computed with function `stats::acf()`.
+$\widehat{R}$ (**of what??
+i.e., chain means and chain variances of each variable**) is computed by implementing Vehtari et al.'s [-@veht19] recommendations.
+$AC$ (**of what?
+and how is AC aggregated across imputations?
+i.e., mean in latest sim**) is computed with function `stats::acf()`.
 
@@
 
-We compute bias as the difference between $\bar{Q}$ and $Q$.
+We compute bias as the difference between $Q$ and $\bar{Q}$.
 Confidence interval width (CIW) is defined as the difference between the lower and upper bound of the 95\% confidence interval (CI95\%) around $\bar{Q}$.
@@
 
-**Add more info about figure legends and axes.**
+**(Add more info about figure legends and axes.)**
 
@@
 
-## Convergence Diagnostics
+## Convergence Diagnostics (pair convergnce diagnostics with respective estimates, not split between convergence and simulation diagnostics)
+
+**(Add different percentages missingness??)**
 
@@
 Generally speaking, conditions with longer imputation chains (higher $T$) coincide with less signs of non-convergence ($\widehat{R}$-values approach one, and $AC$-values approach zero).
-We see more or less equivalent trends for chain means versus chain variances.
-We, therefore, only discuss the $\widehat{R}$ and $AC$-values of chain means.
+<!-- We see more or less equivalent trends for chain means versus chain variances. We, therefore, only discuss the $\widehat{R}$ and $AC$-values of chain means.  -->
 
@@
 In the intervening conditions ($3 \leq T \leq 5$), however, we observe a steep increase in $\widehat{R}$-values.
-This increase implies that non-convergence is under-estimated, and convergence should not be diagnosed at this point.
+This increase implies that non-convergence is under-estimated, and convergence should not be diagnosed at this point (**is that true???**).
 Based on the conventional threshold $\widehat{R} < 1.2$, we would falsely diagnose convergence at $T=3$.
@@
 We see an initial decrease between $T=2$ and $T=3$.
-Simulation conditions where $T>5$ have $AC$-values greater than $T=2$.
+Simulation conditions where $T>5$ have $AC$-values greater than $T=2$ (**weird wording**).
 
 According to this diagnostic, none of the simulation conditions show signs of non-convergence, since we only observe negative or zero $AC$-values.
-The negative $AC$-values, however, indicate increasing divergence between subsequent samples in conditions where $T<4$.
-It may not be wise to terminating the algorithm at that point, but after `recovering' from this dip in $AC$-values.
+The negative $AC$-values, however, indicate increasing divergence (**move above!**) between subsequent samples in conditions where $T<4$.
+It may not be wise (**why?
+and how would we know?
+e.g., wider distrubutions?
+didn't seem like it...**) to terminating the algorithm at that point, but after `recovering' from this dip in $AC$-values.
 This would imply $T=6$ as the minimal number of iterations.
@@
 
-We use average bias, average confidence interval width, and coverage rate as performance measures to evaluate $\widehat{R}$ and $AC$.
+We use average bias, average confidence interval width, and coverage rate as performance measures to evaluate $\widehat{R}$ and $AC$ (**not the goal!
+we want to show the effects of non-conv, not just see how to diagnose it.**).
 We make the general observation that the simulation diagnostics behave as theorized for most simulation conditions (bias around zero, stable confidence interval widths, nominal coverage rate at 95\%).
@@
 The bias for $T=2$ is below average, but within the range of fluctuations.
-Conditions where $T>3$ can be diagnosed as unbiased, but the average bias across iterations (as shown by a flat Loess line) only reaches stability at $T=20$.
+Conditions where $T>3$ can be diagnosed as unbiased, but the average bias across iterations (as shown by a flat Loess line) only reaches stability at $T=20$ (**Loess might not be relevant, check and remove accordingly!**).
 
 CIW is only clearly divergent in the simulation condition $T=1$, see Figure 3B.
-However, under-estimating the variance of $\bar{Q}$, which is the case for $T=3$, may yield spurious inferences.
+However, under-estimating the variance of $\bar{Q}$, which is the case for $T=3$, may yield spurious inferences (**move above**).
 Only conditions where $T>3$ are therefore considered sufficient.
-These conditions have similar CIWs, but across iterations stability is not established (i.e., the Loess line is never completely flat within the 100 simulation conditions).
-%After about twenty-five iterations the average CIW across all simulation conditions and repetitions is reached 0.927.
+(**Possibly irrelevant Loess: These conditions have similar CIWs, but across iterations stability is not established (i.e., the Loess line is never completely flat within the 100 simulation conditions).**)
 
@@
 We see some over-coverage at $T=2$, but that is better than under-estimating the variance of $\bar{Q}$.
-On average, the coverage rate is somewhat higher than the expected nominal coverage of 95\% [@neym34], namely 95\%.
-Similar to the CIWs, there is some trending across iterations (i.e., the Loess line is never flat).
+On average, the coverage rate is somewhat higher than the expected nominal coverage of 95\% [@neym34] (**not anymore??**).
+(**Loess: Similar to the CIWs, there is some trending across iterations (i.e., the Loess line is never flat)**).
 
@@
 This note shows that convergence diagnostics $\widehat{R}$ and $AC$ may diagnose convergence of multiple imputation algorithms, but their performance differs from conventional applications to iterative algorithmic procedures.
+(**nope!
+it shows that MICE can lead to correct outcomes when they have not converged accroding to two common conv diags.
+This may be due to the measures (e.g., assumption of overdisp) or due to the Qs (lm reg coeff, not higher dimensional/more complex RQs).
+Add what %miss has to do with it.**)
 $\widehat{R}$ and autocorrelation indicate that algorithmic convergence may only be reached after twenty or even forty iterations, while unbiased, confidence valid estimates estimates may be obtained with as little as four iterations.
@@
 According to this simulation study, the recently proposed threshold of $\widehat{R}<1.01$ may be too stringent for MI algorithms.
-Under the relatively easy missing data problem of the current study, the threshold was not reached.
+(**This is only one of the goals: to give applied researchers a diagnostic to indicate that they should keep iterating.
+The other is the default in mice and other software packages, and yet another is ...
+i forgot**) Under the relatively easy missing data problem of the current study, the threshold was not reached.
 The other extreme of the $\widehat{R}$-thresholds, the conventionally acceptable $\widehat{R} <1.2$, may be too lenient for MI procedures.
-Applying this threshold to the current data, lead to falsely diagnosing convergence at $T = 3$.
+Applying this threshold to the current data, lead to falsely diagnosing convergence at $T = 3$ (**because it goes up after, not because it is not converged enough**).
 It appears that the widely used threshold of $\widehat{R} < 1.1$ suits MI algorithms the best.
 We might, however, also formulate a new threshold, specifically for the evaluation of MI algorithms.
-The current study suggests that $\widehat{R} < 1.05$ may be implemented, since that is the level at which the $\widehat{R}$ stabilize (around $T = 20$).
+The current study suggests that $\widehat{R} < 1.05$ may be implemented, since that is the level at which the $\widehat{R}$ stabilize (around $T = 20$) (**Not necessary for this Q, but maybe for more complicated Qs**).
 
@@
 However, initial dip in $AC$-values may have implications for the default number of iterations in `mice` (`maxit = 5`).
-Terminating the algorithm at $T=5$ may not be the most appropriate, since this lead to the worst convergence, as indicated by $\widehat{R}$ and $AC$.
+Terminating the algorithm at $T=5$ may not be the most appropriate, since this lead to the worst convergence (**nope, only for Rh, not AC**), as indicated by $\widehat{R}$ and $AC$.
 Under the current specifications, $T>20$ would be more appropriate.
@@
 Moreover, the results of this study imply that assessing the stationarity component of convergence with $AC$ might be redundant, since high $AC$-values are implausible in MI procedures.
-That is, the randomness induced by the MI algorithm effectively mitigates the risk of dependency within chains.
+That is, the randomness induced by the MI algorithm effectively mitigates the risk of dependency within chains (**add above!!**).
 $AC$ would thus not be informative of the convergence of MI algorithms.
 
-Since this study only considers only an MCAR missingness mechanism, results may not be extrapolated to other missing data problems.
+(**Add above, not here because it was a known limitation: Since this study only considers only an MCAR missingness mechanism, results may not be extrapolated to other missing data problems.
 Proper performance of the convergence diagnostics under MCAR is necessary but not sufficient to demonstrate appropriateness of $\widehat{R}$ and $AC$ as convergence diagnostics.
-This is just a proof of concept.
+This is just a proof of concept.**)
 
@@
 Until then, we have only shown that the convergence diagnostics can diagnose non-convergence of MI algorithms that trend towards a converged state.
-Also for future research, implement Wernicke diagnostic, and look at developing a convergence diagnostic for substantive models, and implement a Wald test for $AC = 0$.
+Also for future research, look at developing a convergence diagnostic for substantive models, and implement a Wald test for $AC = 0$.
 

### 2020-04-01 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Read entire research report and put comments in it for thesis manuscript. Ran 5,25,50,75,and 95% miss with 100 sims each.

@@
 to the ubiquitous problem of missing information is to use the framework
-of multiple imputation (MI), proposed by \citet{rubin87}. For the
+of multiple imputation (MI), proposed by \citet{rubin87}. \textbf{(Call
+it mice already, then explain mi and ce sides of mice)} For the
 theoretical foundation of MI, see \citet{rubin87}. For an accessible and
@@
 e.g.~\citet{buur18}. With MI, missing datapoints are `imputed' (i.e.,
-filled in) via an \emph{imputation model}. The imputation model can be a
-joint model for all variables, or several univariate conditional
-distributions (fully conditional specification). FCS is more flexible
-and ``avoids the problem of specifying an appropriate joint imputation
-distribution and replaces this by the selection of appropriate
-univariate conditional distributions''. With FCS, an iterative
-algorithmic procedure is used to \ldots{} multiple imputation by chained
-equations \ldots{} ``The values for imputing the missing data are drawn
-from the posterior predictive distribution of the missing data
-conditional on the observed data.'' There is no scientific consensus on
-the convergence properties of FCS algorithms \citep{taka17}. Some
-default techniques in \texttt{mice} might not yield converged states at
-all \citep{murr18}. Therefore, algorithmic convergence should be
-monitored carefully.
+filled in) via an \emph{imputation model} (\textbf{term not required to
+understand this paper, so circumvent}) . (\textbf{Move this to mi vs.~ce
+part in intro:} The imputation model can be a joint model for all
+variables, or several univariate conditional distributions (fully
+conditional specification). FCS is more flexible and ``avoids the
+problem of specifying an appropriate joint imputation distribution and
+replaces this by the selection of appropriate univariate conditional
+distributions''.)**
+
+\textbf{(Rewrite: With FCS, an iterative algorithmic procedure is used
+to \ldots{} multiple imputation by chained equations \ldots{} ``The
+values for imputing the missing data are drawn from the posterior
+predictive distribution of the missing data conditional on the observed
+data.'')}
+
+There is no scientific consensus on the convergence properties of FCS
+algorithms \citep{taka17}. Some default techniques in \texttt{mice}
+might not yield converged states at all \citep{murr18}. Therefore,
+algorithmic convergence should be monitored carefully.
 
@@
 \href{https://pdfs.semanticscholar.org/e20e/29e008592cbfbaa567931f74cdfdb5451405.pdf?_ga=2.55354671.54033656.1584698748-527613517.1584698748}{(link)},
-Stata = 1;
+Stata = 10
 \href{https://www.stata.com/manuals13/mi.pdf,\%20p.\%20139}{(link)},
@@
 \(y\) is denoted by \(y_{obs}\); the missing part of \(y\) is referred
-to as \(y_{mis}\). For each datapoint in \(y_{mis}\) \(M \times T\)
-times plausible values are sampled, where \(M\) is the number of
-imputations (\(m = 1, \dots, M\)) and \(T\) is the number of iterations
-(\(t = 1, \dots, T\)). The collection of samples between the initial
+to as \(y_{mis}\). \textbf{(This is not only notation, but theory too:
+For each datapoint in \(y_{mis}\) \(M \times T\) times plausible values
+are sampled, where \(M\) is the number of imputations
+(\(m = 1, \dots, M\)) and \(T\) is the number of iterations
+(\(t = 1, \dots, T\)).)} The collection of samples between the initial
 value (at \(t=1\)) and the imputed value (at \(t=T\)) will be referred
@@
 
-In iterative algorithmic procedures (Markov chain Monte Carlo methods;
-e.g., \texttt{mice} algorithms or Gibbs samplers) non-convergence may be
+\textbf{(Start with current guidelines, then continue with possible
+quant measures of the same things from MCMC lit.)} In iterative
+algorithmic procedures (Markov chain Monte Carlo methods; e.g.,
+\texttt{mice} algorithms or Gibbs samplers) non-convergence may be
 present as non-stationarity within chains (i.e., trending), or as slow
-mixing between chains (i.e., no intermingling). (\textbf{add
-example/guidelines visual inspection here?})
+mixing between chains (i.e., no intermingling (\textbf{source??})).
+(\textbf{add example/guidelines visual inspection here?})
 
@@
 error \citep[or `MC error';][]{gewe92}, and Raftery and Lewis's
-\citeyearpar{raft91} procedure to determine the effect of trending
-within chains.
+\citeyearpar{raft91} procedure to determine (\textbf{the chain length
+required to cancel out trending}) the effect of trending within chains.
 
@@
 statistic';][]{gelm92}. With an adapted version of \(\widehat{R}\),
-proposed by \citet{veht19}, we might also evaluate the stationarity
-component of convergence. This would make \(\widehat{R}\) a general
+proposed by \citet{veht19}, the stationarity component of convergence
+might be evaluated as well. This would make \(\widehat{R}\) a general
 convergence diagnostic. The application of \(\widehat{R}\) to assess
@@
 2.5.2]{buur18}. That is, there is no baseline measure available to
-evaluate performance against.The current practice of visually inspecting
-imputation chains for signs of non-mixing or non-stationarity can only
-diagnose severely pathological cases of non-convergence. We could also
-look at distributional characteristics, and plausibility of imputed
-values, see \citet{vinknd}. For now, this is outside of the scope of
-this study.
+evaluate performance against. \textbf{(Move above: The current practice
+of visually inspecting imputation chains for signs of non-mixing or
+non-stationarity can only diagnose severely pathological cases of
+non-convergence.)}
 
@@
 the initial values are over-dispersed. We expect this to have little
-effect on the hypothesized performance of \(\widehat{R}\). No hypothesis
-was formulated about the performance of \(AC\) as convergence
-diagnostic.
+effect on the hypothesized performance of \(\widehat{R}\). \textbf{(Add
+actual hypothesis.)} No hypothesis was formulated about the performance
+of \(AC\) as convergence diagnostic.
 
@@
 
-\begin{verbatim}
-# pseudo-code of simulation 
-simulate data 
-for (number of simulation runs from 1 to 1000)
-  for (number of iterations from 1 to 100)
-    create missingness
-    impute the missingness
-    compute convergence diagnostics
-    perform analysis
-    pool results
-    compute simulation diagnostics
-aggregate convergence and simulation diagnostics
-\end{verbatim}
-
 A finite population of \(N=1000\) is simulated to solve a multiple
 linear regression problem, where dependent variable \(Y\) is regressed
-on independent variables \(X_1\), \(X_2\) and \(X_3\):
+on independent variables \(X_1\), \(X_2\) and \(X_3\) (\textbf{check
+notation with betas here, and add what the quantity/-ies of scientific
+interest is/are!}):
 
@@
 
-\(\widehat{R}\) is computed by implementing Vehtari et al.'s
-\citeyearpar{veht19} recommendations. \(AC\) is computed with function
-\texttt{stats::acf()}.
+\(\widehat{R}\) (\textbf{of what?? i.e., chain means and chain variances
+of each variable}) is computed by implementing Vehtari et al.'s
+\citeyearpar{veht19} recommendations. \(AC\) (\textbf{of what? and how
+is AC aggregated across imputations? i.e., mean in latest sim}) is
+computed with function \texttt{stats::acf()}.
 
@@
 
-We compute bias as the difference between \(\bar{Q}\) and \(Q\).
+We compute bias as the difference between \(Q\) and \(\bar{Q}\).
 Confidence interval width (CIW) is defined as the difference between the
@@
 
-\textbf{Add more info about figure legends and axes.}
-
-\begin{verbatim}
-bias_means_plot + R_means_plot + AC_means_plot + plot_layout(nrow = 3, guides = "collect")
-\end{verbatim}
+\textbf{(Add more info about figure legends and axes.)}
 
 \begin{figure}
-\centering
-\includegraphics{manuscript_files/figure-latex/unnamed-chunk-2-1.pdf}
-\caption{Convergence diagnostics (means)}
+
+{\centering \includegraphics{manuscript_files/figure-latex/unnamed-chunk-2-1}
+
+}
+
+\caption{Convergence diagnostics (means)}\label{fig:unnamed-chunk-2}
 \end{figure}
 
-\begin{verbatim}
-bias_sds_plot + R_sds_plot + AC_sds_plot + plot_layout(nrow = 3, guides = "collect")
-\end{verbatim}
+\begin{figure}
+
+{\centering \includegraphics{manuscript_files/figure-latex/unnamed-chunk-3-1}
+
+}
+
+\caption{Convergence diagnostics (variances)}\label{fig:unnamed-chunk-3}
+\end{figure}
 
 \begin{figure}
-\centering
-\includegraphics{manuscript_files/figure-latex/unnamed-chunk-3-1.pdf}
-\caption{Convergence diagnostics (variances)}
+
+{\centering \includegraphics{manuscript_files/figure-latex/unnamed-chunk-4-1}
+
+}
+
+\caption{Convergence diagnostics (regression coefficients)}\label{fig:unnamed-chunk-4}
 \end{figure}
 
-\begin{verbatim}
-bias_ests_plot + bias_R_sq_plot + cov_plot + plot_layout(nrow = 3, guides = "collect")
-\end{verbatim}
+\hypertarget{convergence-diagnostics-pair-convergnce-diagnostics-with-respective-estimates-not-split-between-convergence-and-simulation-diagnostics}{%
+\subsection{Convergence Diagnostics (pair convergnce diagnostics with
+respective estimates, not split between convergence and simulation
+diagnostics)}\label{convergence-diagnostics-pair-convergnce-diagnostics-with-respective-estimates-not-split-between-convergence-and-simulation-diagnostics}}
 
-\begin{figure}
-\centering
-\includegraphics{manuscript_files/figure-latex/unnamed-chunk-4-1.pdf}
-\caption{Convergence diagnostics (regression coefficients)}
-\end{figure}
-
-\hypertarget{convergence-diagnostics-1}{%
-\subsection{Convergence Diagnostics}\label{convergence-diagnostics-1}}
+\textbf{(Add different percentages missingness??)}
 
@@
 (\(\widehat{R}\)-values approach one, and \(AC\)-values approach zero).
-We see more or less equivalent trends for chain means versus chain
-variances. We, therefore, only discuss the \(\widehat{R}\) and
-\(AC\)-values of chain means.
 
@@
 increase implies that non-convergence is under-estimated, and
-convergence should not be diagnosed at this point. Based on the
-conventional threshold \(\widehat{R} < 1.2\), we would falsely diagnose
-convergence at \(T=3\). According to the widely used threshold
-\(\widehat{R} < 1.1\), convergence would be diagnosed for conditions
-where \(T>9\). If we use the recently recommended threshold
+convergence should not be diagnosed at this point (\textbf{is that
+true???}). Based on the conventional threshold \(\widehat{R} < 1.2\), we
+would falsely diagnose convergence at \(T=3\). According to the widely
+used threshold \(\widehat{R} < 1.1\), convergence would be diagnosed for
+conditions where \(T>9\). If we use the recently recommended threshold
 \(\widehat{R} < 1.01\), we would conclude that convergence is reached in
@@
 \(T=2\) and \(T=3\). Simulation conditions where \(T>5\) have
-\(AC\)-values greater than \(T=2\).
+\(AC\)-values greater than \(T=2\) (\textbf{weird wording}).
 
@@
 \(AC\)-values. The negative \(AC\)-values, however, indicate increasing
-divergence between subsequent samples in conditions where \(T<4\). It
-may not be wise to terminating the algorithm at that point, but after
-`recovering' from this dip in \(AC\)-values. This would imply \(T=6\) as
-the minimal number of iterations.
+divergence (\textbf{move above!}) between subsequent samples in
+conditions where \(T<4\). It may not be wise (\textbf{why? and how would
+we know? e.g., wider distrubutions? didn't seem like it\ldots{}}) to
+terminating the algorithm at that point, but after `recovering' from
+this dip in \(AC\)-values. This would imply \(T=6\) as the minimal
+number of iterations.
 
@@
 We use average bias, average confidence interval width, and coverage
-rate as performance measures to evaluate \(\widehat{R}\) and \(AC\). We
-make the general observation that the simulation diagnostics behave as
-theorized for most simulation conditions (bias around zero, stable
-confidence interval widths, nominal coverage rate at 95\%).
+rate as performance measures to evaluate \(\widehat{R}\) and \(AC\)
+(\textbf{not the goal! we want to show the effects of non-conv, not just
+see how to diagnose it.}). We make the general observation that the
+simulation diagnostics behave as theorized for most simulation
+conditions (bias around zero, stable confidence interval widths, nominal
+coverage rate at 95\%).
 
@@
 the average bias across iterations (as shown by a flat Loess line) only
-reaches stability at \(T=20\).
+reaches stability at \(T=20\) (\textbf{Loess might not be relevant,
+check and remove accordingly!}).
 
@@
 Figure 3B. However, under-estimating the variance of \(\bar{Q}\), which
-is the case for \(T=3\), may yield spurious inferences. Only conditions
-where \(T>3\) are therefore considered sufficient. These conditions have
+is the case for \(T=3\), may yield spurious inferences (\textbf{move
+above}). Only conditions where \(T>3\) are therefore considered
+sufficient. (\textbf{Possibly irrelevant Loess: These conditions have
 similar CIWs, but across iterations stability is not established (i.e.,
 the Loess line is never completely flat within the 100 simulation
-conditions). \%After about twenty-five iterations the average CIW across
-all simulation conditions and repetitions is reached 0.927.
+conditions).})
 
@@
 \(\bar{Q}\). On average, the coverage rate is somewhat higher than the
-expected nominal coverage of 95\% \citep{neym34}, namely 95\%. Similar
-to the CIWs, there is some trending across iterations (i.e., the Loess
-line is never flat).
+expected nominal coverage of 95\% \citep{neym34} (\textbf{not
+anymore??}). (\textbf{Loess: Similar to the CIWs, there is some trending
+across iterations (i.e., the Loess line is never flat)}).
 
@@
 performance differs from conventional applications to iterative
-algorithmic procedures.\\
+algorithmic procedures. (\textbf{nope! it shows that MICE can lead to
+correct outcomes when they have not converged accroding to two common
+conv diags. This may be due to the measures (e.g., assumption of
+overdisp) or due to the Qs (lm reg coeff, not higher dimensional/more
+complex RQs). Add what \%miss has to do with it.})\\
 \(\widehat{R}\) and autocorrelation indicate that algorithmic
@@
 According to this simulation study, the recently proposed threshold of
-\(\widehat{R}<1.01\) may be too stringent for MI algorithms. Under the
-relatively easy missing data problem of the current study, the threshold
-was not reached. The other extreme of the \(\widehat{R}\)-thresholds,
-the conventionally acceptable \(\widehat{R} <1.2\), may be too lenient
-for MI procedures. Applying this threshold to the current data, lead to
-falsely diagnosing convergence at \(T = 3\). It appears that the widely
-used threshold of \(\widehat{R} < 1.1\) suits MI algorithms the best. We
-might, however, also formulate a new threshold, specifically for the
-evaluation of MI algorithms. The current study suggests that
-\(\widehat{R} < 1.05\) may be implemented, since that is the level at
-which the \(\widehat{R}\) stabilize (around \(T = 20\)).
+\(\widehat{R}<1.01\) may be too stringent for MI algorithms.
+(\textbf{This is only one of the goals: to give applied researchers a
+diagnostic to indicate that they should keep iterating. The other is the
+default in mice and other software packages, and yet another is \ldots{}
+i forgot}) Under the relatively easy missing data problem of the current
+study, the threshold was not reached. The other extreme of the
+\(\widehat{R}\)-thresholds, the conventionally acceptable
+\(\widehat{R} <1.2\), may be too lenient for MI procedures. Applying
+this threshold to the current data, lead to falsely diagnosing
+convergence at \(T = 3\) (\textbf{because it goes up after, not because
+it is not converged enough}). It appears that the widely used threshold
+of \(\widehat{R} < 1.1\) suits MI algorithms the best. We might,
+however, also formulate a new threshold, specifically for the evaluation
+of MI algorithms. The current study suggests that \(\widehat{R} < 1.05\)
+may be implemented, since that is the level at which the \(\widehat{R}\)
+stabilize (around \(T = 20\)) (\textbf{Not necessary for this Q, but
+maybe for more complicated Qs}).
 
@@
 (\texttt{maxit\ =\ 5}). Terminating the algorithm at \(T=5\) may not be
-the most appropriate, since this lead to the worst convergence, as
-indicated by \(\widehat{R}\) and \(AC\). Under the current
-specifications, \(T>20\) would be more appropriate.
+the most appropriate, since this lead to the worst convergence
+(\textbf{nope, only for Rh, not AC}), as indicated by \(\widehat{R}\)
+and \(AC\). Under the current specifications, \(T>20\) would be more
+appropriate.
 
@@
 implausible in MI procedures. That is, the randomness induced by the MI
-algorithm effectively mitigates the risk of dependency within chains.
-\(AC\) would thus not be informative of the convergence of MI
-algorithms.
+algorithm effectively mitigates the risk of dependency within chains
+(\textbf{add above!!}). \(AC\) would thus not be informative of the
+convergence of MI algorithms.
 
-Since this study only considers only an MCAR missingness mechanism,
-results may not be extrapolated to other missing data problems. Proper
+(\textbf{Add above, not here because it was a known limitation: Since
+this study only considers only an MCAR missingness mechanism, results
+may not be extrapolated to other missing data problems. Proper
 performance of the convergence diagnostics under MCAR is necessary but
 not sufficient to demonstrate appropriateness of \(\widehat{R}\) and
-\(AC\) as convergence diagnostics. This is just a proof of concept.
+\(AC\) as convergence diagnostics. This is just a proof of concept.})
 
@@
 convergence diagnostics can diagnose non-convergence of MI algorithms
-that trend towards a converged state. Also for future research,
-implement Wernicke diagnostic, and look at developing a convergence
-diagnostic for substantive models, and implement a Wald test for
-\(AC = 0\).
+that trend towards a converged state. Also for future research, look at
+developing a convergence diagnostic for substantive models, and
+implement a Wald test for \(AC = 0\).
 

### 2020-04-01 3.Thesis/1.SimulationStudy/1.Execute.R

@@ -29,7 +29,7 @@ source("3.Thesis/1.SimulationStudy/Functions/Evaluate.R")
 # simulation parameters
 populationsize <- 1000 #n of simulated dataset
 n.iter <- 50 #nr of iterations (varying 1:n.iter)
-n.sim <- 10 #nr of simulations per iteration value
+n.sim <- 100 #nr of simulations per iteration value
 true_effect <- 2 #regression coefficient to be estimated
 true_mean <- true_sd <- NA
 
@@ -54,7 +54,7 @@ simulate <- function(data, n.iter, true_effect, patterns) {
   ampdata <-
     ampute(data,
            patterns = amp_patterns,
-           prop = 0.95,
+           prop = 0.05,
            mech = "MCAR")$amp
   
   # object for output
@@ -111,4 +111,4 @@ dat <-
 ###
 
 # save for future reference
-# save.Rdata(results_with_CI, name = "newresults", path = "3.Thesis/1.SimulationStudy/Results")
+save.Rdata(dat, name = "5percentmiss", path = "3.Thesis/1.SimulationStudy/Results")

### 2020-04-02 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
Put different percentages missingness in the manuscript

@@
 email: h.i.oberman@uu.nl
-abstract: "Multple imputation by chained equations (mice) is a widely used tool to accommodate missing data.
+abstract: "(**Rewrite:**) Multple imputation by chained equations (mice) is a widely used tool to accommodate missing data.
 While empirical evidence supports the validity of inferences obtained using mice, there is no consensus on the convergence properties of the method.
@@
 ```{r setup, include=FALSE}
-#knitr::opts_chunk$set(echo = TRUE)
-knitr::opts_chunk$set(fig.width = 5, echo = FALSE, fig.align = "center") 
-load("../Results/results.Rdata")
-source("../2.CreateFigures.R")
-
+knitr::opts_chunk$set(fig.width = 4.8, fig.height = 3.6, echo = FALSE, fig.align = "left") 
 ```
@@
 
-At some point, any scientist conducting statistical analyses will run into a missing data problem [@alli02].
+At some point, any social scientist 
+<!-- conducting statistical analyses  -->
+will run into a missing data problem [@alli02].
 Missingness is problematic because statistical inference cannot be performed on incomplete data without employing \emph{ad hoc} solutions (e.g., list-wise deletion), which may yield wildly invalid results [@buur18].
-A popular answer to the ubiquitous problem of missing information is to use the framework of multiple imputation (MI), proposed by @rubin87.
-**(Call it mice already, then explain mi and ce sides of mice)** For the theoretical foundation of MI, see @rubin87.
+A popular method to obtain statistically valid results despite of missing data is 'multiple imputation' [MI; @rubin87].
+This paper investigates a potential threat to the validity of this method: algorithmic non-convergence.
+
+With multiple imputation, missing data are 'imputed' (i.e., filled in) several times by means of an iterative algorithmic procedure.
+The method ultimately results in an aggregated statistical inference.
+For the theoretical foundation of MI, see @rubin87.
 For an accessible and comprehensive introduction to MI from an applied perspective, see e.g.
 @buur18.
-With MI, missing datapoints are 'imputed' (i.e., filled in) via an *imputation model* (**term not required to understand this paper, so circumvent**) .
-(**Move this to mi vs.
-ce part in intro:** The imputation model can be a joint model for all variables, or several univariate conditional distributions (fully conditional specification).
-FCS is more flexible and "avoids the problem of specifying an appropriate joint imputation distribution and replaces this by the selection of appropriate univariate conditional distributions".)**
 
-**(Rewrite: With FCS, an iterative algorithmic procedure is used to ...
+based on a model of the observed databy means of an iterative algorithm 
+is the use of MICE: multiple imputation by chained equations.
+MICE is a type of multiple imputation procedure--a framework proposed by @rubin87--in which missing datapoints are 'imputed' (i.e., filled in) by means of an iterative algorithm (chained equations).
+Imputed values are obtained by modeling the (distribution of the) missing data conditional on the observed data.
+The technique is therefore also called 'fully conditional specification' (FCS).
+<!-- several times, and inferences are aggregated across imputed datasets.  -->
+<!-- and statistical inference is performed on each imputed dataset. Variation between the results refelect uncertainty due to missingness  -->
+<!-- CE: The imputation model can be a joint model for all variables, or several univariate conditional distributions (fully conditional specification).
+FCS is more flexible and "avoids the problem of specifying an appropriate joint imputation distribution and replaces this by the selection of appropriate univariate conditional distributions".
+-->
+
+<!-- (**Rewrite:** With FCS, an iterative algorithmic procedure is used to ...
 multiple imputation by chained equations ...
-"The values for imputing the missing data are drawn from the posterior predictive distribution of the missing data conditional on the observed data.")**
+"The values for imputing the missing data are drawn from the posterior predictive distribution of the missing data conditional on the observed data.")  -->
 
@@
 Current practice is visually inspecting imputation chains for signs of non-convergence, but may be challenging to the untrained eye [@buur18, $\S$ 6.5.2].
+Visually inspecting imputation chains (**for signs of non-mixing or non-stationarity**) can only diagnose severely pathological cases of non-convergence.
 Therefore diagnostic evaluation of convergence would be preferred.
@@
 The collection of observed data values in $y$ is denoted by $y_{obs}$; the missing part of $y$ is referred to as $y_{mis}$.
-**(This is not only notation, but theory too: For each datapoint in $y_{mis}$ $M \times T$ times plausible values are sampled, where $M$ is the number of imputations ($m = 1, \dots, M$) and $T$ is the number of iterations ($t = 1, \dots, T$).)** The collection of samples between the initial value (at $t=1$) and the imputed value (at $t=T$) will be referred to as an 'imputation chain'.
+(**This is not only notation, but theory too:** For each datapoint in $y_{mis}$ $M \times T$ times plausible values are sampled, where $M$ is the number of imputations ($m = 1, \dots, M$) and $T$ is the number of iterations ($t = 1, \dots, T$).) The collection of samples between the initial value (at $t=1$) and the imputed value (at $t=T$) will be referred to as an 'imputation chain'.
 
@@
 
+Negative $AC$-values indicate divergence between subsequent samples.
+It may not be wise (**why?
+and how would we know?
+e.g., wider distrubutions?
+didn't seem like it...**) to terminating the algorithm when AC is negative.
+High $AC$-values are implausible in MI procedures.
+That is, the randomness induced by the MI algorithm effectively mitigates the risk of dependency within chains.
+
 In short, convergence is reached when there is no dependency between subsequent iterations of imputation chains ($AC = 0$), and chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm ($\widehat{R} = 1$).
@@
 That is, there is no baseline measure available to evaluate performance against.
-**(Move above: The current practice of visually inspecting imputation chains for signs of non-mixing or non-stationarity can only diagnose severely pathological cases of non-convergence.)**
 <!-- We could also look at distributional characteristics, and plausibility of imputed values, see @vinknd. For now, this is outside of the scope of this study. -->
@@
 That is, the function `mice::ampute()` is used to impose a missingness mechanism upon the data.
-**(change this: The missingness is univariate, and the probability to be missing is the same for all four variables, namely 20\% (`prop = 0.8, mech = "MCAR"`).
+(**change this:** The missingness is univariate, and the probability to be missing is the same for all four variables, namely 20\% (`prop = 0.8, mech = "MCAR"`).
 This leaves 20\% of the rows completely observed).
-**
+**(Add different percentages missingness??)** This study only considers only an MCAR missingness mechanism.
+Therefore, results may not be extrapolated to other missing data problems.
+Proper performance of the convergence diagnostics under MCAR is necessary but not sufficient to demonstrate appropriateness of $\widehat{R}$ and $AC$ as convergence diagnostics.
+This is just a proof of concept.
 
@@
 The number of iterations varies between simulation conditions (`maxit = 1, 2, \dots, 100`).
-
-**(tot hier doorgelezen)**
 
@@
 where $t_{(m-1)}$ is the quantile of a $t$-distribution with $m-1$ degrees of freedom, and $SE_{\bar{Q}}$ is the square root of the pooled variance estimate.
-From bias and CIW, we calculate empirical coverage rates.
-Coverage rate is the proportion of simulations in which $Q$ is between the bounds of the CI95\% around $\bar{Q}$.
+**Why track CIW?** Under-estimating the variance of $\bar{Q}$ may yield spurious inferences.
+From bias and CIW, we calculate empirical coverage rates. Coverage rate is the proportion of simulations in which $Q$ is between the bounds of the CI95\% around $\bar{Q}$.
 
@@
 
-```{r, fig.cap="Convergence diagnostics (means)"}
-bias_means_plot + R_means_plot + AC_means_plot + plot_layout(nrow = 3, guides = "collect")
+```{r, include = FALSE}
+load("../Results/25percentmiss.Rdata")
+source("../2.CreateFigures.R")
 ```
 
-```{r, fig.cap="Convergence diagnostics (variances)"}
-bias_sds_plot + R_sds_plot + AC_sds_plot + plot_layout(nrow = 3, guides = "collect")
+```{r}
+bias_means_plot + R_means_plot + AC_means_plot + plot_layout(nrow = 3) + plot_annotation(title = "Convergence of means, 25% missingness")
 ```
 
-```{r, fig.cap="Convergence diagnostics (regression coefficients)"}
-bias_ests_plot + bias_R_sq_plot + cov_plot + plot_layout(nrow = 3, guides = "collect")
+```{r}
+bias_sds_plot + R_sds_plot + AC_sds_plot + plot_layout(nrow = 3) + plot_annotation(title = "Convergence of variances, 25% missingness")
 ```
 
-## Convergence Diagnostics (pair convergnce diagnostics with respective estimates, not split between convergence and simulation diagnostics)
+```{r}
+bias_ests_plot + bias_R_sq_plot + cov_plot + plot_layout(nrow = 3) + plot_annotation(title = "Convergence of effects, 25% missingness")
 
-**(Add different percentages missingness??)**
+```
+
+```{r, include = FALSE}
+load("../Results/50percentmiss.Rdata")
+source("../2.CreateFigures.R")
+```
+
+```{r}
+bias_means_plot + R_means_plot + AC_means_plot + plot_layout(nrow = 3) + plot_annotation(title = "Convergence of means, 50% missingness")
+```
+
+```{r}
+bias_sds_plot + R_sds_plot + AC_sds_plot + plot_layout(nrow = 3) + plot_annotation(title = "Convergence of variances, 50% missingness")
+```
+
+```{r}
+bias_ests_plot + bias_R_sq_plot + cov_plot + plot_layout(nrow = 3) + plot_annotation(title = "Convergence of effects, 50% missingness")
+
+```
+
+```{r, include = FALSE}
+load("../Results/75percentmiss.Rdata")
+source("../2.CreateFigures.R")
+```
+
+```{r}
+bias_means_plot + R_means_plot + AC_means_plot + plot_layout(nrow = 3) + plot_annotation(title = "Convergence of means, 75% missingness")
+```
+
+```{r}
+bias_sds_plot + R_sds_plot + AC_sds_plot + plot_layout(nrow = 3) + plot_annotation(title = "Convergence of variances, 75% missingness")
+```
+
+```{r}
+bias_ests_plot + bias_R_sq_plot + cov_plot + plot_layout(nrow = 3) + plot_annotation(title = "Convergence of effects, 75% missingness")
+
+```
+
+```{r, include = FALSE}
+load("../Results/95percentmiss.Rdata")
+source("../2.CreateFigures.R")
+```
+
+```{r}
+bias_means_plot + R_means_plot + AC_means_plot + plot_layout(nrow = 3) + plot_annotation(title = "Convergence of means, 95% missingness")
+```
+
+```{r}
+bias_sds_plot + R_sds_plot + AC_sds_plot + plot_layout(nrow = 3) + plot_annotation(title = "Convergence of variances, 95% missingness")
+```
+
+```{r}
+bias_ests_plot + bias_R_sq_plot + cov_plot + plot_layout(nrow = 3) + plot_annotation(title = "Convergence of effects, 95% missingness")
+
+```
+
+## Convergence Diagnostics
+
+**(Pair convergnce diagnostics with respective estimates, not split between convergence and simulation diagnostics)**
 
@@
 According to this diagnostic, none of the simulation conditions show signs of non-convergence, since we only observe negative or zero $AC$-values.
-The negative $AC$-values, however, indicate increasing divergence (**move above!**) between subsequent samples in conditions where $T<4$.
-It may not be wise (**why?
-and how would we know?
-e.g., wider distrubutions?
-didn't seem like it...**) to terminating the algorithm at that point, but after `recovering' from this dip in $AC$-values.
-This would imply $T=6$ as the minimal number of iterations.
 
@@
 CIW is only clearly divergent in the simulation condition $T=1$, see Figure 3B.
-However, under-estimating the variance of $\bar{Q}$, which is the case for $T=3$, may yield spurious inferences (**move above**).
 Only conditions where $T>3$ are therefore considered sufficient.
-(**Possibly irrelevant Loess: These conditions have similar CIWs, but across iterations stability is not established (i.e., the Loess line is never completely flat within the 100 simulation conditions).**)
+(**Possibly irrelevant Loess:** These conditions have similar CIWs, but across iterations stability is not established (i.e., the Loess line is never completely flat within the 100 simulation conditions).)
 
@@
 On average, the coverage rate is somewhat higher than the expected nominal coverage of 95\% [@neym34] (**not anymore??**).
-(**Loess: Similar to the CIWs, there is some trending across iterations (i.e., the Loess line is never flat)**).
+(**Loess:** Similar to the CIWs, there is some trending across iterations (i.e., the Loess line is never flat)).
 
@@
 The observed dip in AC implies that default maxit value of five iterations is the worst possible number of iterations.
-Moreover, the results of this study imply that assessing the stationarity component of convergence with $AC$ might be redundant, since high $AC$-values are implausible in MI procedures.
-That is, the randomness induced by the MI algorithm effectively mitigates the risk of dependency within chains (**add above!!**).
-$AC$ would thus not be informative of the convergence of MI algorithms.
-
-(**Add above, not here because it was a known limitation: Since this study only considers only an MCAR missingness mechanism, results may not be extrapolated to other missing data problems.
-Proper performance of the convergence diagnostics under MCAR is necessary but not sufficient to demonstrate appropriateness of $\widehat{R}$ and $AC$ as convergence diagnostics.
-This is just a proof of concept.**)
+Moreover, the results of this study imply that assessing the stationarity component of convergence with $AC$ might be redundant.
+<!-- $AC$ would thus not be informative of the convergence of MI algorithms.  -->
 

### 2020-04-02 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Put different percentages missingness in the manuscript

@@
 \begin{abstract}
-Multple imputation by chained equations (mice) is a widely used tool to
-accommodate missing data. While empirical evidence supports the validity
-of inferences obtained using mice, there is no consensus on the
-convergence properties of the method. This paper provides insight into
-non-convergence of mice algorithms.
+(\textbf{Rewrite:}) Multple imputation by chained equations (mice) is a
+widely used tool to accommodate missing data. While empirical evidence
+supports the validity of inferences obtained using mice, there is no
+consensus on the convergence properties of the method. This paper
+provides insight into non-convergence of mice algorithms.
 \end{abstract}
@@
 
-At some point, any scientist conducting statistical analyses will run
-into a missing data problem \citep{alli02}. Missingness is problematic
-because statistical inference cannot be performed on incomplete data
-without employing \emph{ad hoc} solutions (e.g., list-wise deletion),
-which may yield wildly invalid results \citep{buur18}. A popular answer
-to the ubiquitous problem of missing information is to use the framework
-of multiple imputation (MI), proposed by \citet{rubin87}. \textbf{(Call
-it mice already, then explain mi and ce sides of mice)} For the
+At some point, any social scientist will run into a missing data problem
+\citep{alli02}. Missingness is problematic because statistical inference
+cannot be performed on incomplete data without employing \emph{ad hoc}
+solutions (e.g., list-wise deletion), which may yield wildly invalid
+results \citep{buur18}. A popular method to obtain statistically valid
+results despite of missing data is `multiple imputation'
+\citep[MI;][]{rubin87}. This paper investigates a potential threat to
+the validity of this method: algorithmic non-convergence.
+
+With multiple imputation, missing data are `imputed' (i.e., filled in)
+several times by means of an iterative algorithmic procedure. The method
+ultimately results in an aggregated statistical inference. For the
 theoretical foundation of MI, see \citet{rubin87}. For an accessible and
 comprehensive introduction to MI from an applied perspective, see
-e.g.~\citet{buur18}. With MI, missing datapoints are `imputed' (i.e.,
-filled in) via an \emph{imputation model} (\textbf{term not required to
-understand this paper, so circumvent}) . (\textbf{Move this to mi vs.~ce
-part in intro:} The imputation model can be a joint model for all
-variables, or several univariate conditional distributions (fully
-conditional specification). FCS is more flexible and ``avoids the
-problem of specifying an appropriate joint imputation distribution and
-replaces this by the selection of appropriate univariate conditional
-distributions''.)**
+e.g.~\citet{buur18}.
 
-\textbf{(Rewrite: With FCS, an iterative algorithmic procedure is used
-to \ldots{} multiple imputation by chained equations \ldots{} ``The
-values for imputing the missing data are drawn from the posterior
-predictive distribution of the missing data conditional on the observed
-data.'')}
+based on a model of the observed databy means of an iterative algorithm
+is the use of MICE: multiple imputation by chained equations. MICE is a
+type of multiple imputation procedure--a framework proposed by
+\citet{rubin87}--in which missing datapoints are `imputed' (i.e., filled
+in) by means of an iterative algorithm (chained equations). Imputed
+values are obtained by modeling the (distribution of the) missing data
+conditional on the observed data. The technique is therefore also called
+`fully conditional specification' (FCS).
 
@@
 non-convergence, but may be challenging to the untrained eye
-\citep[\(\S\) 6.5.2]{buur18}. Therefore diagnostic evaluation of
-convergence would be preferred. However, not known whether MCMC
-diagnostics are appropriate for MICE \ldots{} The application of
+\citep[\(\S\) 6.5.2]{buur18}. Visually inspecting imputation chains
+(\textbf{for signs of non-mixing or non-stationarity}) can only diagnose
+severely pathological cases of non-convergence. Therefore diagnostic
+evaluation of convergence would be preferred. However, not known whether
+MCMC diagnostics are appropriate for MICE \ldots{} The application of
 convergence diagnostics to MI methods has not been systematically
@@
 \(y\) is denoted by \(y_{obs}\); the missing part of \(y\) is referred
-to as \(y_{mis}\). \textbf{(This is not only notation, but theory too:
+to as \(y_{mis}\). (\textbf{This is not only notation, but theory too:}
 For each datapoint in \(y_{mis}\) \(M \times T\) times plausible values
@@
 (\(m = 1, \dots, M\)) and \(T\) is the number of iterations
-(\(t = 1, \dots, T\)).)} The collection of samples between the initial
+(\(t = 1, \dots, T\)).) The collection of samples between the initial
 value (at \(t=1\)) and the imputed value (at \(t=T\)) will be referred
@@
 
+Negative \(AC\)-values indicate divergence between subsequent samples.
+It may not be wise (\textbf{why? and how would we know? e.g., wider
+distrubutions? didn't seem like it\ldots{}}) to terminating the
+algorithm when AC is negative. High \(AC\)-values are implausible in MI
+procedures. That is, the randomness induced by the MI algorithm
+effectively mitigates the risk of dependency within chains.
+
 In short, convergence is reached when there is no dependency between
@@
 2.5.2]{buur18}. That is, there is no baseline measure available to
-evaluate performance against. \textbf{(Move above: The current practice
-of visually inspecting imputation chains for signs of non-mixing or
-non-stationarity can only diagnose severely pathological cases of
-non-convergence.)}
+evaluate performance against.
 
@@
 is, the function \texttt{mice::ampute()} is used to impose a missingness
-mechanism upon the data. \textbf{(change this: The missingness is
+mechanism upon the data. (\textbf{change this:} The missingness is
 univariate, and the probability to be missing is the same for all four
 variables, namely 20\% (\texttt{prop\ =\ 0.8,\ mech\ =\ "MCAR"}). This
-leaves 20\% of the rows completely observed). }
+leaves 20\% of the rows completely observed). \textbf{(Add different
+percentages missingness??)} This study only considers only an MCAR
+missingness mechanism. Therefore, results may not be extrapolated to
+other missing data problems. Proper performance of the convergence
+diagnostics under MCAR is necessary but not sufficient to demonstrate
+appropriateness of \(\widehat{R}\) and \(AC\) as convergence
+diagnostics. This is just a proof of concept.
 
@@
 conditions (\texttt{maxit\ =\ 1,\ 2,\ \textbackslash{}dots,\ 100}).
-
-\textbf{(tot hier doorgelezen)}
 
@@
 degrees of freedom, and \(SE_{\bar{Q}}\) is the square root of the
-pooled variance estimate. From bias and CIW, we calculate empirical
-coverage rates. Coverage rate is the proportion of simulations in which
-\(Q\) is between the bounds of the CI95\% around \(\bar{Q}\).
+pooled variance estimate. \textbf{Why track CIW?} Under-estimating the
+variance of \(\bar{Q}\) may yield spurious inferences. From bias and
+CIW, we calculate empirical coverage rates. Coverage rate is the
+proportion of simulations in which \(Q\) is between the bounds of the
+CI95\% around \(\bar{Q}\).
 
@@
 
-\begin{figure}
+\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-3-1} \end{flushleft}
 
-{\centering \includegraphics{manuscript_files/figure-latex/unnamed-chunk-2-1}
+\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-4-1} \end{flushleft}
 
-}
+\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-5-1} \end{flushleft}
 
-\caption{Convergence diagnostics (means)}\label{fig:unnamed-chunk-2}
-\end{figure}
+\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-7-1} \end{flushleft}
 
-\begin{figure}
+\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-8-1} \end{flushleft}
 
-{\centering \includegraphics{manuscript_files/figure-latex/unnamed-chunk-3-1}
+\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-9-1} \end{flushleft}
 
-}
+\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-11-1} \end{flushleft}
 
-\caption{Convergence diagnostics (variances)}\label{fig:unnamed-chunk-3}
-\end{figure}
+\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-12-1} \end{flushleft}
 
-\begin{figure}
+\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-13-1} \end{flushleft}
 
-{\centering \includegraphics{manuscript_files/figure-latex/unnamed-chunk-4-1}
+\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-15-1} \end{flushleft}
 
-}
+\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-16-1} \end{flushleft}
 
-\caption{Convergence diagnostics (regression coefficients)}\label{fig:unnamed-chunk-4}
-\end{figure}
+\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-17-1} \end{flushleft}
 
-\hypertarget{convergence-diagnostics-pair-convergnce-diagnostics-with-respective-estimates-not-split-between-convergence-and-simulation-diagnostics}{%
-\subsection{Convergence Diagnostics (pair convergnce diagnostics with
-respective estimates, not split between convergence and simulation
-diagnostics)}\label{convergence-diagnostics-pair-convergnce-diagnostics-with-respective-estimates-not-split-between-convergence-and-simulation-diagnostics}}
+\hypertarget{convergence-diagnostics-1}{%
+\subsection{Convergence Diagnostics}\label{convergence-diagnostics-1}}
 
-\textbf{(Add different percentages missingness??)}
+\textbf{(Pair convergnce diagnostics with respective estimates, not
+split between convergence and simulation diagnostics)}
 
@@
 signs of non-convergence, since we only observe negative or zero
-\(AC\)-values. The negative \(AC\)-values, however, indicate increasing
-divergence (\textbf{move above!}) between subsequent samples in
-conditions where \(T<4\). It may not be wise (\textbf{why? and how would
-we know? e.g., wider distrubutions? didn't seem like it\ldots{}}) to
-terminating the algorithm at that point, but after `recovering' from
-this dip in \(AC\)-values. This would imply \(T=6\) as the minimal
-number of iterations.
+\(AC\)-values.
 
@@
 CIW is only clearly divergent in the simulation condition \(T=1\), see
-Figure 3B. However, under-estimating the variance of \(\bar{Q}\), which
-is the case for \(T=3\), may yield spurious inferences (\textbf{move
-above}). Only conditions where \(T>3\) are therefore considered
-sufficient. (\textbf{Possibly irrelevant Loess: These conditions have
+Figure 3B. Only conditions where \(T>3\) are therefore considered
+sufficient. (\textbf{Possibly irrelevant Loess:} These conditions have
 similar CIWs, but across iterations stability is not established (i.e.,
 the Loess line is never completely flat within the 100 simulation
-conditions).})
+conditions).)
 
@@
 expected nominal coverage of 95\% \citep{neym34} (\textbf{not
-anymore??}). (\textbf{Loess: Similar to the CIWs, there is some trending
-across iterations (i.e., the Loess line is never flat)}).
+anymore??}). (\textbf{Loess:} Similar to the CIWs, there is some
+trending across iterations (i.e., the Loess line is never flat)).
 
@@
 results of this study imply that assessing the stationarity component of
-convergence with \(AC\) might be redundant, since high \(AC\)-values are
-implausible in MI procedures. That is, the randomness induced by the MI
-algorithm effectively mitigates the risk of dependency within chains
-(\textbf{add above!!}). \(AC\) would thus not be informative of the
-convergence of MI algorithms.
-
-(\textbf{Add above, not here because it was a known limitation: Since
-this study only considers only an MCAR missingness mechanism, results
-may not be extrapolated to other missing data problems. Proper
-performance of the convergence diagnostics under MCAR is necessary but
-not sufficient to demonstrate appropriateness of \(\widehat{R}\) and
-\(AC\) as convergence diagnostics. This is just a proof of concept.})
+convergence with \(AC\) might be redundant.
 

### 2020-04-02 3.Thesis/1.SimulationStudy/2.CreateFigures.R

@@ -7,6 +7,16 @@ library("dplyr")
 library("ggplot2")
 library("patchwork")
 
+theme_update(plot.title = element_text(hjust = 0.5), 
+             panel.border = element_blank(),
+             panel.grid.major = element_blank(),
+             panel.grid.minor = element_blank(),
+             panel.background = element_blank(),
+             axis.line = element_line(colour = "black"),
+             legend.key = element_blank(),
+             legend.position = "bottom")
+
+
 # # if necessary, load results
 # load("C:/Users/User/Desktop/shinyMice/3.Thesis/1.SimulationStudy/Results/results.Rdata")
 # dat <- results_with_CI # <- dat
@@ -18,10 +28,10 @@ bias_means_plot <- dat %>% .[1:50,] %>% ggplot() +
   geom_hline(yintercept = 0,
              color = "grey",
              lwd = 2) +
-  geom_point(aes(x = T, y = bias.mean.Y, color = "Y")) +
-  geom_point(aes(x = T, y = bias.mean.X1, color = "X1")) +
-  geom_point(aes(x = T, y = bias.mean.X2, color = "X2")) +
-  geom_point(aes(x = T, y = bias.mean.X3, color = "X3")) +
+  geom_point(aes(x = T, y = bias.mean.Y, color = "Y"), size = .75) +
+  geom_point(aes(x = T, y = bias.mean.X1, color = "X1"), size = .75) +
+  geom_point(aes(x = T, y = bias.mean.X2, color = "X2"), size = .75) +
+  geom_point(aes(x = T, y = bias.mean.X3, color = "X3"), size = .75) +
   geom_line(aes(x = T, y = bias.mean.Y, color = "Y")) +
   geom_line(aes(x = T, y = bias.mean.X1, color = "X1")) +
   geom_line(aes(x = T, y = bias.mean.X2, color = "X2")) +
@@ -29,63 +39,44 @@ bias_means_plot <- dat %>% .[1:50,] %>% ggplot() +
   xlab("") +
   ylab(expression(paste("Bias in ", hat(mu)))) +
   labs(colour = "Legend") +
-  theme_bw() +
-  theme(
-    panel.border = element_blank(),
-    panel.grid.major = element_blank(),
-    panel.grid.minor = element_blank(),
[…]

### 2020-04-03 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
Before removing 9 of the 12 plots in the manuscript with the different missingness percentages.

@@
 
-With multiple imputation, missing data are 'imputed' (i.e., filled in) several times by means of an iterative algorithmic procedure.
-The method ultimately results in an aggregated statistical inference.
-For the theoretical foundation of MI, see @rubin87.
+The multiple imputation method can be roughly divided into 3 steps.
+First, each missing value in an incomplete dataset is 'imputed' (i.e., filled in) several times.
+Next, the analysis of scientific interest is performed on each of the completed datasets.
+And finally, the results of these analyses are pooled into a single aggregated statistic.
+The theoretical foundation of these streps was laid by Rubin [1972, -@rubin87].
 For an accessible and comprehensive introduction to MI from an applied perspective, see e.g.
 @buur18.
+In this paper, we focus on the first of these steps: the algorithmic procedure with which the missing values are imputed.
 
-based on a model of the observed databy means of an iterative algorithm 
-is the use of MICE: multiple imputation by chained equations.
+<!-- based on a model of the observed databy means of an iterative algorithm  -->
+<!-- is the use of MICE: multiple imputation by chained equations.
 MICE is a type of multiple imputation procedure--a framework proposed by @rubin87--in which missing datapoints are 'imputed' (i.e., filled in) by means of an iterative algorithm (chained equations).
@@
 The technique is therefore also called 'fully conditional specification' (FCS).
+-->
 <!-- several times, and inferences are aggregated across imputed datasets.  -->
@@
 
-There is no scientific consensus on the convergence properties of FCS algorithms [@taka17].
-Some default techniques in `mice` might not yield converged states at all [@murr18].
+There is no scientific consensus on the convergence properties of MI algorithms [@taka17].
+Some default techniques might not yield converged states at all [@murr18].
 Therefore, algorithmic convergence should be monitored carefully.
+The current practice is visually inspecting imputation chains for signs of non-convergence, but this may be challenging to the untrained eye [@buur18, $\S$ 6.5.2].
+Moreover, visually inspecting imputation chains can only diagnose severely pathological cases of non-convergence.
+Therefore, diagnostic evaluation of convergence would be preferred.
 
-Convergence cannot be determined for sure, because MICE is an MCMC method, and convergence of MCMC is not from scalar to a point, but from one distribution to another.
-Generated values will vary even after convergence.
-Therefore we can only diagnose non-convergence, not convergence [@hoff09].
-"a weak diagnostics is better than no diagnostic at all" [@cowl96].
+Convergence cannot be determined for sure, because MICE (**introduce MICE before!**) is a type of MCMC method, and convergence of MCMC algorithms is not from a scalar to a point, but from one distribution to another.
+Therefore, generated values will vary even after convergence.
+The aim of convergence diagnostics for MCMC methods is thus not to establish the point at which convergence is reached, but to diagnose severe non-convergence.
+<!-- Therefore we can only diagnose non-convergence, not convergence [@hoff09]. "a weak diagnostics is better than no diagnostic at all" [@cowl96]. -->
 
-Current practice is visually inspecting imputation chains for signs of non-convergence, but may be challenging to the untrained eye [@buur18, $\S$ 6.5.2].
-Visually inspecting imputation chains (**for signs of non-mixing or non-stationarity**) can only diagnose severely pathological cases of non-convergence.
-Therefore diagnostic evaluation of convergence would be preferred.
-However, not known whether MCMC diagnostics are appropriate for MICE ...
+Several MCMC convergence diagnostics exist, but it is not known whetherthese are appropriate for MICE.
 The application of convergence diagnostics to MI methods has not been systematically studied [@buur18].
-How can non-convergence be diagnosed?
+The open questions are, e.g.: How can non-convergence be diagnosed?
 How many iterations are sufficient/required?
@@
 Are the default number of iterations sufficient?
-Defaults in: SPSS = 10 [(link)](https://www.ibm.com/support/knowledgecenter/SSLVMB_24.0.0/spss/mva/syn_multiple_imputation_impute.html), Mplus = 100??
+The defaults of common MI software: SPSS = 10 [(link)](https://www.ibm.com/support/knowledgecenter/SSLVMB_24.0.0/spss/mva/syn_multiple_imputation_impute.html), Mplus = 100??
 [(link)](https://pdfs.semanticscholar.org/e20e/29e008592cbfbaa567931f74cdfdb5451405.pdf?_ga=2.55354671.54033656.1584698748-527613517.1584698748), Stata = 10 [(link)](https://www.stata.com/manuals13/mi.pdf, p.
@@
 
-The stationarity component of convergence may be evaluated with autocorrelation [$AC$; @scha97; @gelm13], numeric standard error [or `MC error'; @gewe92], and Raftery and Lewis's [-@raft91] procedure to determine (**the chain length required to cancel out trending**) the effect of trending within chains.
+The stationarity component of convergence may be evaluated with autocorrelation [$AC$; @scha97; @gelm13], numeric standard error [or `MC error'; @gewe92], and Raftery and Lewis's [-@raft91] procedure to determine  the effect of trending within chains (**or the chain length required to cancel out trending?**).
 
@@
 @buur18 proposed to implement multivariable evaluation through eigenvalue decomposition [@mack03].
-This method is outside of the scope of the current study (**not anymore!**).
+This method is outside of the scope of the current study (**not anymore?**).
 
@@
 
+## Univariate estimates and convergence diagnostics
+
+The bias in the estimates of the variable means show little to no difference between simulation conditions.
+It doesn't seem to matter how many iterations you use in the mice algorithm, the estimates are unbiased.
+Similarly, the bias in the estimated variances is more or less stable across simulation conditions.
+These univariate quantities appear to be unaffected by the number of iterations.
+
+When applied to the imputation chain means, $\widehat{R}$ indicates that the mice algorithm does not reach a converged state ($\widehat{R}$ = 1) in any of the simulation conditions.
+Neither is the most recent recommended threshold reached ($\widehat{R}$ < 1.01).
+The conventional $\widehat{R}$ threshold of 1.2 is reached in simulation conditions $T = 3$ and $T > 6$.
+(With the default number of iterations (maxit = 5), this dip in Rhat values would be spotted, so it is no problem.) The point at which an extra iteration does not seem to improve the Rhat value is around $T=30$.
+<!-- Rhat values are indistinguishable between simulation conditions when $T>30$.  -->
+<!-- The R hat values seemed to flatten off between 20 and 30 iterations.  -->
+
+Auto-correlations indicate no sign of trending within imputation chains.
+In most simulation conditions AC is smaller than or about equal to zero.
+Across simulation conditions, however, the autocorrelation curve does not trend towards 0.
+Autocorrelation values plateau off at a value of around .1 .
+This is a small positive autocorrelation, which would indicate some trending within chains.
+
+The $\widehat{R}$ and $AC$ values for the imputation chain variances show equal trends and are therefore not discussed separately.
+Taken together,  univariate estimates seem robust across simulation conditions.
+There is no clear effect of the number of iterations on the bias in these estimates, while the convergence diagnostics indicate that the algorithm did not reach a completely converged state (yet).
+
+## Multivariate estimates
+
+There is a clear bias in the regression estimates in simulation conditions where the number of iterations is smaller than four.
+In simulation conditions where $T > 5$ there is little to no bias in the estimated regression coefficients.
+In most simulation conditions nominal coverage is obtained (i.e., coverage rates of .95) for the confidence intervals around the regression coefficients.
+Conditions with only one or two iterations show some undercoverage.
+Since confidence interval width is stable across conditions, the undercoverage may be attributed to the bias in the estimated regression coefficients.
+
+If we look at the estimated proportion of explained variance in outcome variable Y we see that the coefficient of determination is underestimate estimated in conditions where the number of iterations is equal to two or less, and slightly overestimated in conditions where the number of iterations is equal to three or more.
+
+In short, we see that the minimum number of iterations required to obtain unbiased, confidence valid regression estimates is 5.
+This value, however, is dependent on the percentage of missing values.
+E.g., with 95% of cases having missing data we need at least seven iterations to obtain unbiased results.
+
+<!-- Bias in the estimated $R^2$ values does not improve  -->
+
 <!-- Figures 2 and 3 display results per simulation condition ($T = 1,2,\dots,100$).  -->
@@
 ```
+
+<!-- For 25% miss, no bias in estimated means, Rh flat from 30 it.
+(but <1.1 for it>11), AC never high but stable after 40 it.
+Bias in estimated variances only if it=1 or maybe 2?, while Rh of chain variances is flat around it 25 (smaller than 1.1 for it>9), and AC never positive, flat around it 30.
+Bias in estimated reg.
+coeff.
+only pronounced for the first iteration., while coverage rates are ok immediately on average, variance explained is marginally underestimated (bias of 0.5% on an effect of 19.25%).
+-->
+
+```{r include=FALSE}
+# chain means
+dat[-1, 6:9] %>% mutate(max_Rh = apply(., 1, max)) %>% .$max_Rh<1.2
+
+# chain variances
+dat[-1, 26:29] %>% mutate(max_Rh = apply(., 1, max)) %>% .$max_Rh<1.2
+
+```
+
+# Results [OLD, JUST FOR REFERENCE!]
 

### 2020-04-03 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Before removing 9 of the 12 plots in the manuscript with the different missingness percentages.

@@
 
-With multiple imputation, missing data are `imputed' (i.e., filled in)
-several times by means of an iterative algorithmic procedure. The method
-ultimately results in an aggregated statistical inference. For the
-theoretical foundation of MI, see \citet{rubin87}. For an accessible and
-comprehensive introduction to MI from an applied perspective, see
-e.g.~\citet{buur18}.
+The multiple imputation method can be roughly divided into 3 steps.
+First, each missing value in an incomplete dataset is `imputed' (i.e.,
+filled in) several times. Next, the analysis of scientific interest is
+performed on each of the completed datasets. And finally, the results of
+these analyses are pooled into a single aggregated statistic. The
+theoretical foundation of these streps was laid by Rubin
+\citeyearpar[1972,][]{rubin87}. For an accessible and comprehensive
+introduction to MI from an applied perspective, see e.g.~\citet{buur18}.
+In this paper, we focus on the first of these steps: the algorithmic
+procedure with which the missing values are imputed.
 
-based on a model of the observed databy means of an iterative algorithm
-is the use of MICE: multiple imputation by chained equations. MICE is a
-type of multiple imputation procedure--a framework proposed by
-\citet{rubin87}--in which missing datapoints are `imputed' (i.e., filled
-in) by means of an iterative algorithm (chained equations). Imputed
-values are obtained by modeling the (distribution of the) missing data
-conditional on the observed data. The technique is therefore also called
-`fully conditional specification' (FCS).
+There is no scientific consensus on the convergence properties of MI
+algorithms \citep{taka17}. Some default techniques might not yield
+converged states at all \citep{murr18}. Therefore, algorithmic
+convergence should be monitored carefully. The current practice is
+visually inspecting imputation chains for signs of non-convergence, but
+this may be challenging to the untrained eye \citep[\(\S\)
+6.5.2]{buur18}. Moreover, visually inspecting imputation chains can only
+diagnose severely pathological cases of non-convergence. Therefore,
+diagnostic evaluation of convergence would be preferred.
 
-There is no scientific consensus on the convergence properties of FCS
-algorithms \citep{taka17}. Some default techniques in \texttt{mice}
-might not yield converged states at all \citep{murr18}. Therefore,
-algorithmic convergence should be monitored carefully.
+Convergence cannot be determined for sure, because MICE
+(\textbf{introduce MICE before!}) is a type of MCMC method, and
+convergence of MCMC algorithms is not from a scalar to a point, but from
+one distribution to another. Therefore, generated values will vary even
+after convergence. The aim of convergence diagnostics for MCMC methods
+is thus not to establish the point at which convergence is reached, but
+to diagnose severe non-convergence.
 
-Convergence cannot be determined for sure, because MICE is an MCMC
-method, and convergence of MCMC is not from scalar to a point, but from
-one distribution to another. Generated values will vary even after
-convergence. Therefore we can only diagnose non-convergence, not
-convergence \citep{hoff09}. ``a weak diagnostics is better than no
-diagnostic at all'' \citep{cowl96}.
-
-Current practice is visually inspecting imputation chains for signs of
-non-convergence, but may be challenging to the untrained eye
-\citep[\(\S\) 6.5.2]{buur18}. Visually inspecting imputation chains
-(\textbf{for signs of non-mixing or non-stationarity}) can only diagnose
-severely pathological cases of non-convergence. Therefore diagnostic
-evaluation of convergence would be preferred. However, not known whether
-MCMC diagnostics are appropriate for MICE \ldots{} The application of
-convergence diagnostics to MI methods has not been systematically
-studied \citep{buur18}. How can non-convergence be diagnosed? How many
-iterations are sufficient/required? What are the effects of
-non-convergence on estimates, predictions and inference? What is an
-appropriate non-convergence diagnostic? Are the default number of
-iterations sufficient? Defaults in: SPSS = 10
+Several MCMC convergence diagnostics exist, but it is not known
+whetherthese are appropriate for MICE. The application of convergence
+diagnostics to MI methods has not been systematically studied
+\citep{buur18}. The open questions are, e.g.: How can non-convergence be
+diagnosed? How many iterations are sufficient/required? What are the
+effects of non-convergence on estimates, predictions and inference? What
+is an appropriate non-convergence diagnostic? Are the default number of
+iterations sufficient? The defaults of common MI software: SPSS = 10
 \href{https://www.ibm.com/support/knowledgecenter/SSLVMB_24.0.0/spss/mva/syn_multiple_imputation_impute.html}{(link)},
@@
 error \citep[or `MC error';][]{gewe92}, and Raftery and Lewis's
-\citeyearpar{raft91} procedure to determine (\textbf{the chain length
-required to cancel out trending}) the effect of trending within chains.
+\citeyearpar{raft91} procedure to determine the effect of trending
+within chains (\textbf{or the chain length required to cancel out
+trending?}).
 
@@
 \citep{mack03}. This method is outside of the scope of the current study
-(\textbf{not anymore!}).
+(\textbf{not anymore?}).
 
@@
 
+\hypertarget{univariate-estimates-and-convergence-diagnostics}{%
+\subsection{Univariate estimates and convergence
+diagnostics}\label{univariate-estimates-and-convergence-diagnostics}}
+
+The bias in the estimates of the variable means show little to no
+difference between simulation conditions. It doesn't seem to matter how
+many iterations you use in the mice algorithm, the estimates are
+unbiased. Similarly, the bias in the estimated variances is more or less
+stable across simulation conditions. These univariate quantities appear
+to be unaffected by the number of iterations.
+
+When applied to the imputation chain means, \(\widehat{R}\) indicates
+that the mice algorithm does not reach a converged state
+(\(\widehat{R}\) = 1) in any of the simulation conditions. Neither is
+the most recent recommended threshold reached (\(\widehat{R}\)
+\textless{} 1.01). The conventional \(\widehat{R}\) threshold of 1.2 is
+reached in simulation conditions \(T = 3\) and \(T > 6\). (With the
+default number of iterations (maxit = 5), this dip in Rhat values would
+be spotted, so it is no problem.) The point at which an extra iteration
+does not seem to improve the Rhat value is around \(T=30\).
+
+Auto-correlations indicate no sign of trending within imputation chains.
+In most simulation conditions AC is smaller than or about equal to zero.
+Across simulation conditions, however, the autocorrelation curve does
+not trend towards 0. Autocorrelation values plateau off at a value of
+around .1 . This is a small positive autocorrelation, which would
+indicate some trending within chains.
+
+The \(\widehat{R}\) and \(AC\) values for the imputation chain variances
+show equal trends and are therefore not discussed separately. Taken
+together, univariate estimates seem robust across simulation conditions.
+There is no clear effect of the number of iterations on the bias in
+these estimates, while the convergence diagnostics indicate that the
+algorithm did not reach a completely converged state (yet).
+
+\hypertarget{multivariate-estimates}{%
+\subsection{Multivariate estimates}\label{multivariate-estimates}}
+
+There is a clear bias in the regression estimates in simulation
+conditions where the number of iterations is smaller than four. In
+simulation conditions where \(T > 5\) there is little to no bias in the
+estimated regression coefficients. In most simulation conditions nominal
+coverage is obtained (i.e., coverage rates of .95) for the confidence
+intervals around the regression coefficients. Conditions with only one
+or two iterations show some undercoverage. Since confidence interval
+width is stable across conditions, the undercoverage may be attributed
+to the bias in the estimated regression coefficients.
+
+If we look at the estimated proportion of explained variance in outcome
+variable Y we see that the coefficient of determination is underestimate
+estimated in conditions where the number of iterations is equal to two
+or less, and slightly overestimated in conditions where the number of
+iterations is equal to three or more.
+
+In short, we see that the minimum number of iterations required to
+obtain unbiased, confidence valid regression estimates is 5. This value,
+however, is dependent on the percentage of missing values. E.g., with
+95\% of cases having missing data we need at least seven iterations to
+obtain unbiased results.
+
 \begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-3-1} \end{flushleft}
@@
 \begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-17-1} \end{flushleft}
+
+For 25\% miss, no bias in estimated means, Rh flat from 30 it. (but
+\textless1.1 for it\textgreater11), AC never high but stable after 40
+it. Bias in estimated variances only if it=1 or maybe 2?, while Rh of
+chain variances is flat around it 25 (smaller than 1.1 for
+it\textgreater9), and AC never positive, flat around it 30. Bias in
+estimated reg. coeff. only pronounced for the first iteration., while
+coverage rates are ok immediately on average, variance explained is
+marginally underestimated (bias of 0.5\% on an effect of 19.25\%).
+
+\hypertarget{results-old-just-for-reference}{%
+\section{Results {[}OLD, JUST FOR
+REFERENCE!{]}}\label{results-old-just-for-reference}}
 

### 2020-04-03 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
Removed excess plots from manuscript. Ready for feedback.

@@
 And finally, the results of these analyses are pooled into a single aggregated statistic.
-The theoretical foundation of these streps was laid by Rubin [1972, -@rubin87].
+The theoretical foundation of these streps was laid by Rubin [-@rubin76, -@rubin87].
 For an accessible and comprehensive introduction to MI from an applied perspective, see e.g.
@@
 ```{r, include = FALSE}
-load("../Results/25percentmiss.Rdata")
-source("../2.CreateFigures.R")
-```
-
-```{r}
-bias_means_plot + R_means_plot + AC_means_plot + plot_layout(nrow = 3) + plot_annotation(title = "Convergence of means, 25% missingness")
-```
-
-```{r}
-bias_sds_plot + R_sds_plot + AC_sds_plot + plot_layout(nrow = 3) + plot_annotation(title = "Convergence of variances, 25% missingness")
-```
-
-```{r}
-bias_ests_plot + bias_R_sq_plot + cov_plot + plot_layout(nrow = 3) + plot_annotation(title = "Convergence of effects, 25% missingness")
-
-```
-
-```{r, include = FALSE}
-load("../Results/50percentmiss.Rdata")
-source("../2.CreateFigures.R")
-```
-
-```{r}
-bias_means_plot + R_means_plot + AC_means_plot + plot_layout(nrow = 3) + plot_annotation(title = "Convergence of means, 50% missingness")
-```
-
-```{r}
-bias_sds_plot + R_sds_plot + AC_sds_plot + plot_layout(nrow = 3) + plot_annotation(title = "Convergence of variances, 50% missingness")
-```
-
-```{r}
-bias_ests_plot + bias_R_sq_plot + cov_plot + plot_layout(nrow = 3) + plot_annotation(title = "Convergence of effects, 50% missingness")
-
-```
-
-```{r, include = FALSE}
 load("../Results/75percentmiss.Rdata")
@@
 bias_ests_plot + bias_R_sq_plot + cov_plot + plot_layout(nrow = 3) + plot_annotation(title = "Convergence of effects, 75% missingness")
-
 ```
 
-```{r, include = FALSE}
-load("../Results/95percentmiss.Rdata")
-source("../2.CreateFigures.R")
-```
+```{r, echo = FALSE}
+load("../Results/combined.Rdata")
 
-```{r}
-bias_means_plot + R_means_plot + AC_means_plot + plot_layout(nrow = 3) + plot_annotation(title = "Convergence of means, 95% missingness")
-```
-
-```{r}
-bias_sds_plot + R_sds_plot + AC_sds_plot + plot_layout(nrow = 3) + plot_annotation(title = "Convergence of variances, 95% missingness")
-```
-
-```{r}
-bias_ests_plot + bias_R_sq_plot + cov_plot + plot_layout(nrow = 3) + plot_annotation(title = "Convergence of effects, 95% missingness")
+dat %>% ggplot(aes(x = T, y = bias.R.s, color = miss)) + 
+  geom_point() +
+  geom_line() +
+  ylab(expression(paste("Bias in ", hat(R ^ 2)))) +
+  labs(colour = "Legend: missingness percentage") +
+  plot_annotation(title = "The effect of missingness percentage, an example")
 

### 2020-04-03 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Removed excess plots from manuscript. Ready for feedback.

@@
 theoretical foundation of these streps was laid by Rubin
-\citeyearpar[1972,][]{rubin87}. For an accessible and comprehensive
-introduction to MI from an applied perspective, see e.g.~\citet{buur18}.
-In this paper, we focus on the first of these steps: the algorithmic
-procedure with which the missing values are imputed.
+\citeyearpar[\citeyearpar{rubin87}]{rubin76}. For an accessible and
+comprehensive introduction to MI from an applied perspective, see
+e.g.~\citet{buur18}. In this paper, we focus on the first of these
+steps: the algorithmic procedure with which the missing values are
+imputed.
 
@@
 
-\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-7-1} \end{flushleft}
-
-\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-8-1} \end{flushleft}
-
-\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-9-1} \end{flushleft}
-
-\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-11-1} \end{flushleft}
-
-\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-12-1} \end{flushleft}
-
-\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-13-1} \end{flushleft}
-
-\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-15-1} \end{flushleft}
-
-\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-16-1} \end{flushleft}
-
-\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-17-1} \end{flushleft}
-
-For 25\% miss, no bias in estimated means, Rh flat from 30 it. (but
-\textless1.1 for it\textgreater11), AC never high but stable after 40
-it. Bias in estimated variances only if it=1 or maybe 2?, while Rh of
-chain variances is flat around it 25 (smaller than 1.1 for
-it\textgreater9), and AC never positive, flat around it 30. Bias in
-estimated reg. coeff. only pronounced for the first iteration., while
-coverage rates are ok immediately on average, variance explained is
-marginally underestimated (bias of 0.5\% on an effect of 19.25\%).
+\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-6-1} \end{flushleft}
 

### 2020-04-06 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
Feedback Gerko

@@
 ---
-title: Missing the Point$\colon$ Non-Convergence in Multiple Imputation Procedures
+title: Missing the Point$\colon$ Non-Convergence in Iterative Imputation Procedures
 runninghead: Oberman
@@
 
-At some point, any social scientist 
-<!-- conducting statistical analyses  -->
-will run into a missing data problem [@alli02].
-Missingness is problematic because statistical inference cannot be performed on incomplete data without employing \emph{ad hoc} solutions (e.g., list-wise deletion), which may yield wildly invalid results [@buur18].
-A popular method to obtain statistically valid results despite of missing data is 'multiple imputation' [MI; @rubin87].
-This paper investigates a potential threat to the validity of this method: algorithmic non-convergence.
-
-The multiple imputation method can be roughly divided into 3 steps.
-First, each missing value in an incomplete dataset is 'imputed' (i.e., filled in) several times.
+Missing data problems are ubiquitous and pervasive. **or** Everybody who works with person data is likely to run into a missing data problem.  
+<!-- At some point, any social scientist  -->
+<!-- <!-- conducting statistical analyses  --> -->
+<!-- will run into a missing data problem [@alli02].  -->
+Missingness is problematic because statistical inference cannot be performed on the whole data without employing \emph{ad hoc} solutions (e.g., list-wise deletion), which may yield wildly invalid results [@buur18].
+A popular method to obtain statistically valid results despite of missing data is 'multiple imputation' [MI; @rubin76].
+The multiple imputation method can be roughly divided into 3 steps: First, each missing value in an incomplete dataset is 'imputed' (i.e., filled in) several times by some algorithmic procedure.
 Next, the analysis of scientific interest is performed on each of the completed datasets.
 And finally, the results of these analyses are pooled into a single aggregated statistic.
-The theoretical foundation of these streps was laid by Rubin [-@rubin76, -@rubin87].
+
+<!-- This paper investigates a potential threat to the validity of this method: algorithmic non-convergence.  -->
+
+<!-- The theoretical foundation of these streps was laid by Rubin [-@rubin76, -@rubin87].
 For an accessible and comprehensive introduction to MI from an applied perspective, see e.g.
@@
 In this paper, we focus on the first of these steps: the algorithmic procedure with which the missing values are imputed.
+-->
+
+Algorithms are often used to draw imputations from the posterior predictive distribution of the missing values.
+This introduces a potential threat to the validity of the imputations: what if the algorithm has not converged?
+Are the implications then to be trusted?
+And can we rely on the inference obtained on the completed data?
 
@@
 There is no scientific consensus on the convergence properties of MI algorithms [@taka17].
-Some default techniques might not yield converged states at all [@murr18].
+Some default techniques (e.g., 'predictive mean modeling') might not yield converged states at all [@murr18].
 Therefore, algorithmic convergence should be monitored carefully.
 The current practice is visually inspecting imputation chains for signs of non-convergence, but this may be challenging to the untrained eye [@buur18, $\S$ 6.5.2].
-Moreover, visually inspecting imputation chains can only diagnose severely pathological cases of non-convergence.
+Moreover, visually inspecting imputation chains may only diagnose severely pathological cases of non-convergence.
 Therefore, diagnostic evaluation of convergence would be preferred.
 
-Convergence cannot be determined for sure, because MICE (**introduce MICE before!**) is a type of MCMC method, and convergence of MCMC algorithms is not from a scalar to a point, but from one distribution to another.
+Iterative imputation algorithms such as mice, are MCMC methods, and convergence of MCMC algorithms is not from a scalar to a point but from one distribution to another.
 Therefore, generated values will vary even after convergence.
@@
 The open questions are, e.g.: How can non-convergence be diagnosed?
-How many iterations are sufficient/required?
+How many iterations are sufficient/needed?
 What are the effects of non-convergence on estimates, predictions and inference?
@@
 The collection of observed data values in $y$ is denoted by $y_{obs}$; the missing part of $y$ is referred to as $y_{mis}$.
-(**This is not only notation, but theory too:** For each datapoint in $y_{mis}$ $M \times T$ times plausible values are sampled, where $M$ is the number of imputations ($m = 1, \dots, M$) and $T$ is the number of iterations ($t = 1, \dots, T$).) The collection of samples between the initial value (at $t=1$) and the imputed value (at $t=T$) will be referred to as an 'imputation chain'.
+(**This is not only notation, but theory too:** For each datapoint in $y_{mis}$, we sample $M \times T$ times plausible values, where $M$ is the number of imputations ($m = 1, \dots, M$) and $T$ is the number of iterations ($t = 1, \dots, T$).) The collection of samples between the initial value (at $t=1$) and the final imputed value (at $t=T$) will be referred to as an 'imputation chain'.
 
@@
 
-Negative $AC$-values indicate divergence between subsequent samples.
+Negative $AC$-values indicate divergence within imputation chains.
+Subsequent sampled values within each imputation chain are less alike.
 It may not be wise (**why?
@@
 didn't seem like it...**) to terminating the algorithm when AC is negative.
-High $AC$-values are implausible in MI procedures.
-That is, the randomness induced by the MI algorithm effectively mitigates the risk of dependency within chains.
+**High $AC$-values are implausible in MI procedures.
+That is, the randomness induced by the MI algorithm effectively mitigates the risk of dependency within chains.**
 
@@
 
-and simulated with the function `mvtnorm::rmvnorm()`. Outcome variable $Y$ is subsequently calculated as 
+Outcome variable $Y$ is subsequently calculated as 
 $$Y =  1 + 2X_1 + .5X_2 - X_3 + \epsilon .$$
@@
 
-The complete data is 'amputed' once for each simulation repetition.
-That is, the function `mice::ampute()` is used to impose a missingness mechanism upon the data.
+**Add the function `mvtnorm::rmvnorm()` back in**
+
+The complete data is 'amputed' once for each simulation repetition with function `mice::ampute().
 (**change this:** The missingness is univariate, and the probability to be missing is the same for all four variables, namely 20\% (`prop = 0.8, mech = "MCAR"`).
@@
 All MI procedures are performed with Bayesian linear regression imputation (`method = "norm"`), and five imputation chains (`m = 5`).
-The number of iterations varies between simulation conditions (`maxit = 1, 2, \dots, 100`).
+The number of iterations varies between simulation conditions (`maxit = ` $$1, 2, \dots, 100$$).
 
@@
 We obtain an estimated regression coefficient per imputation, which are pooled into a single estimate, $\bar{Q}$.
-We use the function `mice::pool()` to get variance estimates according to Rubin's [-@rubin87] rules, and subsequently implement finite population pooling conform @vink14.
+We use the function `mice::pool()` to estimate u bar (**check wat dat nou precies is!!**) according to Rubin's [-@rubin87] rules, and subsequently implement finite population pooling conform @vink14.
 
@@
 
-## Univariate estimates and convergence diagnostics
+## Univariate estimates and convergence diagnostics (theta = mean and theta = variance)
 

### 2020-04-06 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Feedback Gerko

@@
 
-\title{Missing the Point\(\colon\) Non-Convergence in Multiple Imputation
+\title{Missing the Point\(\colon\) Non-Convergence in Iterative Imputation
 Procedures}
@@
 
-At some point, any social scientist will run into a missing data problem
-\citep{alli02}. Missingness is problematic because statistical inference
-cannot be performed on incomplete data without employing \emph{ad hoc}
-solutions (e.g., list-wise deletion), which may yield wildly invalid
-results \citep{buur18}. A popular method to obtain statistically valid
-results despite of missing data is `multiple imputation'
-\citep[MI;][]{rubin87}. This paper investigates a potential threat to
-the validity of this method: algorithmic non-convergence.
+Missing data problems are ubiquitous and pervasive. \textbf{or}
+Everybody who works with person data is likely to run into a missing
+data problem.\\
+--\textgreater{} Missingness is problematic because statistical
+inference cannot be performed on the whole data without employing
+\emph{ad hoc} solutions (e.g., list-wise deletion), which may yield
+wildly invalid results \citep{buur18}. A popular method to obtain
+statistically valid results despite of missing data is `multiple
+imputation' \citep[MI;][]{rubin76}. The multiple imputation method can
+be roughly divided into 3 steps: First, each missing value in an
+incomplete dataset is `imputed' (i.e., filled in) several times by some
+algorithmic procedure. Next, the analysis of scientific interest is
+performed on each of the completed datasets. And finally, the results of
+these analyses are pooled into a single aggregated statistic.
 
-The multiple imputation method can be roughly divided into 3 steps.
-First, each missing value in an incomplete dataset is `imputed' (i.e.,
-filled in) several times. Next, the analysis of scientific interest is
-performed on each of the completed datasets. And finally, the results of
-these analyses are pooled into a single aggregated statistic. The
-theoretical foundation of these streps was laid by Rubin
-\citeyearpar[\citeyearpar{rubin87}]{rubin76}. For an accessible and
-comprehensive introduction to MI from an applied perspective, see
-e.g.~\citet{buur18}. In this paper, we focus on the first of these
-steps: the algorithmic procedure with which the missing values are
-imputed.
+Algorithms are often used to draw imputations from the posterior
+predictive distribution of the missing values. This introduces a
+potential threat to the validity of the imputations: what if the
+algorithm has not converged? Are the implications then to be trusted?
+And can we rely on the inference obtained on the completed data?
 
 There is no scientific consensus on the convergence properties of MI
-algorithms \citep{taka17}. Some default techniques might not yield
-converged states at all \citep{murr18}. Therefore, algorithmic
-convergence should be monitored carefully. The current practice is
-visually inspecting imputation chains for signs of non-convergence, but
-this may be challenging to the untrained eye \citep[\(\S\)
-6.5.2]{buur18}. Moreover, visually inspecting imputation chains can only
-diagnose severely pathological cases of non-convergence. Therefore,
-diagnostic evaluation of convergence would be preferred.
+algorithms \citep{taka17}. Some default techniques (e.g., `predictive
+mean modeling') might not yield converged states at all \citep{murr18}.
+Therefore, algorithmic convergence should be monitored carefully. The
+current practice is visually inspecting imputation chains for signs of
+non-convergence, but this may be challenging to the untrained eye
+\citep[\(\S\) 6.5.2]{buur18}. Moreover, visually inspecting imputation
+chains may only diagnose severely pathological cases of non-convergence.
+Therefore, diagnostic evaluation of convergence would be preferred.
 
-Convergence cannot be determined for sure, because MICE
-(\textbf{introduce MICE before!}) is a type of MCMC method, and
-convergence of MCMC algorithms is not from a scalar to a point, but from
+Iterative imputation algorithms such as mice, are MCMC methods, and
+convergence of MCMC algorithms is not from a scalar to a point but from
 one distribution to another. Therefore, generated values will vary even
@@
 \citep{buur18}. The open questions are, e.g.: How can non-convergence be
-diagnosed? How many iterations are sufficient/required? What are the
+diagnosed? How many iterations are sufficient/needed? What are the
 effects of non-convergence on estimates, predictions and inference? What
@@
 to as \(y_{mis}\). (\textbf{This is not only notation, but theory too:}
-For each datapoint in \(y_{mis}\) \(M \times T\) times plausible values
-are sampled, where \(M\) is the number of imputations
+For each datapoint in \(y_{mis}\), we sample \(M \times T\) times
+plausible values, where \(M\) is the number of imputations
 (\(m = 1, \dots, M\)) and \(T\) is the number of iterations
 (\(t = 1, \dots, T\)).) The collection of samples between the initial
-value (at \(t=1\)) and the imputed value (at \(t=T\)) will be referred
-to as an `imputation chain'.
+value (at \(t=1\)) and the final imputed value (at \(t=T\)) will be
+referred to as an `imputation chain'.
 
@@
 
-Negative \(AC\)-values indicate divergence between subsequent samples.
+Negative \(AC\)-values indicate divergence within imputation chains.
+Subsequent sampled values within each imputation chain are less alike.
 It may not be wise (\textbf{why? and how would we know? e.g., wider
 distrubutions? didn't seem like it\ldots{}}) to terminating the
-algorithm when AC is negative. High \(AC\)-values are implausible in MI
-procedures. That is, the randomness induced by the MI algorithm
-effectively mitigates the risk of dependency within chains.
+algorithm when AC is negative. \textbf{High \(AC\)-values are
+implausible in MI procedures. That is, the randomness induced by the MI
+algorithm effectively mitigates the risk of dependency within chains.}
 
@@
 
-and simulated with the function \texttt{mvtnorm::rmvnorm()}. Outcome
-variable \(Y\) is subsequently calculated as
+Outcome variable \(Y\) is subsequently calculated as
 \[Y =  1 + 2X_1 + .5X_2 - X_3 + \epsilon .\]
 
-The complete data is `amputed' once for each simulation repetition. That
-is, the function \texttt{mice::ampute()} is used to impose a missingness
-mechanism upon the data. (\textbf{change this:} The missingness is
-univariate, and the probability to be missing is the same for all four
-variables, namely 20\% (\texttt{prop\ =\ 0.8,\ mech\ =\ "MCAR"}). This
-leaves 20\% of the rows completely observed). \textbf{(Add different
-percentages missingness??)} This study only considers only an MCAR
-missingness mechanism. Therefore, results may not be extrapolated to
-other missing data problems. Proper performance of the convergence
-diagnostics under MCAR is necessary but not sufficient to demonstrate
-appropriateness of \(\widehat{R}\) and \(AC\) as convergence
-diagnostics. This is just a proof of concept.
+\textbf{Add the function \texttt{mvtnorm::rmvnorm()} back in}
+
+The complete data is `amputed' once for each simulation repetition with
+function
+\texttt{mice::ampute().\ (**change\ this:**\ The\ missingness\ is\ univariate,\ and\ the\ probability\ to\ be\ missing\ is\ the\ same\ for\ all\ four\ variables,\ namely\ 20\textbackslash{}\%\ (}prop
+= 0.8, mech = ``MCAR''`). This leaves 20\% of the rows completely
+observed). \textbf{(Add different percentages missingness??)} This study
+only considers only an MCAR missingness mechanism. Therefore, results
+may not be extrapolated to other missing data problems. Proper
+performance of the convergence diagnostics under MCAR is necessary but
+not sufficient to demonstrate appropriateness of \(\widehat{R}\) and
+\(AC\) as convergence diagnostics. This is just a proof of concept.
 
@@
 (\texttt{m\ =\ 5}). The number of iterations varies between simulation
-conditions (\texttt{maxit\ =\ 1,\ 2,\ \textbackslash{}dots,\ 100}).
+conditions (\texttt{maxit\ =} \[1, 2, \dots, 100\]).
 
@@
 imputation, which are pooled into a single estimate, \(\bar{Q}\). We use
-the function \texttt{mice::pool()} to get variance estimates according
-to Rubin's \citeyearpar{rubin87} rules, and subsequently implement
-finite population pooling conform \citet{vink14}.
+the function \texttt{mice::pool()} to estimate u bar (\textbf{check wat
+dat nou precies is!!}) according to Rubin's \citeyearpar{rubin87} rules,
+and subsequently implement finite population pooling conform
+\citet{vink14}.
 
@@
 
-\hypertarget{univariate-estimates-and-convergence-diagnostics}{%
-\subsection{Univariate estimates and convergence
-diagnostics}\label{univariate-estimates-and-convergence-diagnostics}}
+\hypertarget{univariate-estimates-and-convergence-diagnostics-theta-mean-and-theta-variance}{%
+\subsection{Univariate estimates and convergence diagnostics (theta =
+mean and theta =
+variance)}\label{univariate-estimates-and-convergence-diagnostics-theta-mean-and-theta-variance}}
 

### 2020-04-09 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
Started working on convergence diagnostics for pca component. Only works for 1 simulation now.

@@
 <!-- At some point, any social scientist  -->
-<!-- <!-- conducting statistical analyses  --> -->
-<!-- will run into a missing data problem [@alli02].  -->
+<!-- conducting statistical analyses  --> <!-- will run into a missing data problem [@alli02].  -->
 Missingness is problematic because statistical inference cannot be performed on the whole data without employing \emph{ad hoc} solutions (e.g., list-wise deletion), which may yield wildly invalid results [@buur18].
@@
 
-**Add the function `mvtnorm::rmvnorm()` back in**
+**Add the function `mvtnorm::rmvnorm()` back in** use citation("mvtnorm")
 
-The complete data is 'amputed' once for each simulation repetition with function `mice::ampute().
+The complete data is 'amputed' once for each simulation repetition with function `mice::ampute()`.
 (**change this:** The missingness is univariate, and the probability to be missing is the same for all four variables, namely 20\% (`prop = 0.8, mech = "MCAR"`).
@@
 All MI procedures are performed with Bayesian linear regression imputation (`method = "norm"`), and five imputation chains (`m = 5`).
-The number of iterations varies between simulation conditions (`maxit = ` $$1, 2, \dots, 100$$).
+The number of iterations varies between simulation conditions (`maxit = ` $1, 2, \dots, 100$).
 
-What to evaluate?
+<!-- What to evaluate? -->
 
-- bias in regression coefficients
+<!-- - bias in regression coefficients -->
 
-- bias in $R^2$ (= coefficient of determination)
+<!-- - bias in $R^2$ (= coefficient of determination) -->
 
--
+<!-- -  -->
 
@@
 ```{r, include = FALSE}
-load("../Results/75percentmiss.Rdata")
+load("../Results/combined.Rdata")
 source("../2.CreateFigures.R")
@@
 ```{r, echo = FALSE}
-load("../Results/combined.Rdata")
+#load("../Results/combined.Rdata")
 

### 2020-04-09 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd

@@
 
+**Add: missingness percentage = proportion of cases with one or more missing values times 100 **
 <!-- The number of imputations is denoted by $m$ (i.e., the number of times each datapoint in $y_{mis}$ is filled in).

### 2020-04-09 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex

@@
 data problem.\\
---\textgreater{} Missingness is problematic because statistical
-inference cannot be performed on the whole data without employing
-\emph{ad hoc} solutions (e.g., list-wise deletion), which may yield
-wildly invalid results \citep{buur18}. A popular method to obtain
-statistically valid results despite of missing data is `multiple
-imputation' \citep[MI;][]{rubin76}. The multiple imputation method can
-be roughly divided into 3 steps: First, each missing value in an
-incomplete dataset is `imputed' (i.e., filled in) several times by some
-algorithmic procedure. Next, the analysis of scientific interest is
-performed on each of the completed datasets. And finally, the results of
-these analyses are pooled into a single aggregated statistic.
+Missingness is problematic because statistical inference cannot be
+performed on the whole data without employing \emph{ad hoc} solutions
+(e.g., list-wise deletion), which may yield wildly invalid results
+\citep{buur18}. A popular method to obtain statistically valid results
+despite of missing data is `multiple imputation' \citep[MI;][]{rubin76}.
+The multiple imputation method can be roughly divided into 3 steps:
+First, each missing value in an incomplete dataset is `imputed' (i.e.,
+filled in) several times by some algorithmic procedure. Next, the
+analysis of scientific interest is performed on each of the completed
+datasets. And finally, the results of these analyses are pooled into a
+single aggregated statistic.
 
@@
 The complete data is `amputed' once for each simulation repetition with
-function
-\texttt{mice::ampute().\ (**change\ this:**\ The\ missingness\ is\ univariate,\ and\ the\ probability\ to\ be\ missing\ is\ the\ same\ for\ all\ four\ variables,\ namely\ 20\textbackslash{}\%\ (}prop
-= 0.8, mech = ``MCAR''`). This leaves 20\% of the rows completely
-observed). \textbf{(Add different percentages missingness??)} This study
-only considers only an MCAR missingness mechanism. Therefore, results
-may not be extrapolated to other missing data problems. Proper
-performance of the convergence diagnostics under MCAR is necessary but
-not sufficient to demonstrate appropriateness of \(\widehat{R}\) and
-\(AC\) as convergence diagnostics. This is just a proof of concept.
+function \texttt{mice::ampute()}. (\textbf{change this:} The missingness
+is univariate, and the probability to be missing is the same for all
+four variables, namely 20\% (\texttt{prop\ =\ 0.8,\ mech\ =\ "MCAR"}).
+This leaves 20\% of the rows completely observed). \textbf{(Add
+different percentages missingness??)} This study only considers only an
+MCAR missingness mechanism. Therefore, results may not be extrapolated
+to other missing data problems. Proper performance of the convergence
+diagnostics under MCAR is necessary but not sufficient to demonstrate
+appropriateness of \(\widehat{R}\) and \(AC\) as convergence
+diagnostics. This is just a proof of concept.
 
@@
 (\texttt{m\ =\ 5}). The number of iterations varies between simulation
-conditions (\texttt{maxit\ =} \[1, 2, \dots, 100\]).
-
-What to evaluate?
-
-\begin{itemize}
-\item
-  bias in regression coefficients
-\item
-  bias in \(R^2\) (= coefficient of determination)
-\item
-\end{itemize}
+conditions (\texttt{maxit\ =} \(1, 2, \dots, 100\)).
 

### 2020-04-10 3.Thesis/1.SimulationStudy/Functions/PCA_convergence.R
Wrote new autocorrelation function instead :)

@@ -17,9 +17,11 @@ PCA_convergence <- function(sims, maxit = n.iter, reps = n.sim, m = 5) {
     
   # loop over nr of iterations
   for (it in 2:maxit){
+  
   # auto-correlation
   ac[it, k] <-
-   sim1[[k]][1:it,] %>% acf(lag.max = 1, plot = FALSE) %>% .$acf %>% .[2, , ] %>% diag() %>% "*"(-2 ) %>% mean()
+   sim1[[k]][1:it,] %>% ac_lag1()#acf(lag.max = 1, plot = FALSE) %>% .$acf %>% .[2, , ] %>% diag() %>% "*"(it/(it-3) ) %>% mean()
+  
   # r hat
   rhat_bulk <-
     sim1[[k]][1:it,] %>% as.matrix() %>% split_chains() %>% z_scale() %>% get.rhat(maxit = it)

### 2020-04-10 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
Wrote new autocorrelation function instead :)
Removed old plots, now all with missingness percentages

@@
 ```{r, include = FALSE}
-load("../Results/combined.Rdata")
+load("../Results/complete.Rdata")
 source("../2.CreateFigures.R")
@@
 
-```{r}
-bias_means_plot + R_means_plot + AC_means_plot + plot_layout(nrow = 3) + plot_annotation(title = "Convergence of means, 75% missingness")
-```
-
-```{r}
-bias_sds_plot + R_sds_plot + AC_sds_plot + plot_layout(nrow = 3) + plot_annotation(title = "Convergence of variances, 75% missingness")
-```
-
-```{r}
-bias_ests_plot + bias_R_sq_plot + cov_plot + plot_layout(nrow = 3) + plot_annotation(title = "Convergence of effects, 75% missingness")
-```
+<!-- ```{r} -->
+<!-- bias_means_plot + R_means_plot + AC_means_plot + plot_layout(nrow = 3) + plot_annotation(title = "Convergence of means, 75% missingness") -->
+<!-- ``` -->
 
@@
 
-dat %>% ggplot(aes(x = T, y = bias.R.s, color = miss)) + 
+dat %>% ggplot(aes(x = T, y = bias.R.s, color = as.factor(miss))) + 
   geom_point() +

### 2020-04-10 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Wrote new autocorrelation function instead :)
Removed old plots, now all with missingness percentages

@@
 referred to as an `imputation chain'.
+
+\textbf{Add: missingness percentage = proportion of cases with one or
+more missing values times 100 }
 
@@
 
-\textbf{Add the function \texttt{mvtnorm::rmvnorm()} back in}
+\textbf{Add the function \texttt{mvtnorm::rmvnorm()} back in} use
+citation(``mvtnorm'')
 
@@
 \begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-3-1} \end{flushleft}
-
-\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-4-1} \end{flushleft}
-
-\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-5-1} \end{flushleft}
-
-\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-6-1} \end{flushleft}
 

### 2020-04-10 3.Thesis/1.SimulationStudy/2.CreateFigures.R

@@ -2,317 +2,209 @@
 # requires the packages "dplyr", "ggplot2"
 # and the object 'results' created with '1.Excecute'
 
-# filter by missingness percentage
-miss_perc <- "75"
-
-# load package
+# load packages
 library("dplyr")
 library("ggplot2")
 library("patchwork")
 
-theme_update(plot.title = element_text(hjust = 0.5), 
-             panel.border = element_blank(),
-             panel.grid.major = element_blank(),
-             panel.grid.minor = element_blank(),
-             panel.background = element_blank(),
-             axis.line = element_line(colour = "black"),
-             legend.key = element_blank(),
-             legend.position = "bottom")
-
-# load data
+# # if necessary, load data
 # load("3.Thesis/1.SimulationStudy/Results/complete.Rdata")
 
-# R squared
-dat %>% ggplot(aes(x = T, y = bias.R.s, color = as.factor(miss))) + 
+# set default for plot layout
+theme_update(
+  plot.title = element_text(hjust = 0.5),
+  panel.border = element_blank(),
+  panel.grid.major = element_blank(),
+  panel.grid.minor = element_blank(),
+  panel.background = element_blank(),
+  axis.line = element_line(colour = "black"),
+  legend.key = element_blank(),
+  legend.position = "bottom"
+)
+
+
+####################
+#### UNIVARIATE ####
+####################
+
+# Univariate: mean estimate
+dat %>% ggplot(aes(x = T, y = bias.mean.X1, color = as.factor(miss))) +
   geom_hline(yintercept = 0,
              color = "grey",
              lwd = 2) +
-  geom_point() +
+  geom_point(size = .75) +
   geom_line() +
-  ylab(expression(paste("Bias in ", hat(R ^ 2)))) +
-  labs(colour = "Missingness (%)") +
-  plot_annotation(title = "The effect of missingness percentage, an example")
-
-# R hat for PCA
-dat %>% ggplot(aes(x = T, y = R.PCA, color = as.factor(miss))) + 
-  geom_hline(yintercept = 1,
-             color = "grey",
-             lwd = 2) +
-  geom_point() +
-  geom_line() +
[…]

### 2020-04-15 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
Include the acf into simulation again to show differences

@@
 
-Missing data problems are ubiquitous and pervasive. **or** Everybody who works with person data is likely to run into a missing data problem.  
-<!-- At some point, any social scientist  -->
-<!-- conducting statistical analyses  --> <!-- will run into a missing data problem [@alli02].  -->
-Missingness is problematic because statistical inference cannot be performed on the whole data without employing \emph{ad hoc} solutions (e.g., list-wise deletion), which may yield wildly invalid results [@buur18].
+<!-- (**or** Everybody who works with person data is likely to run into a missing data problem.) -->
+Missing data problems are ubiquitous and pervasive in the social and behavioral sciences.
+If a dataset contains just one missing value for a variable of interest, statistical inferences are undefined and will not produce any result.
+Incomplete observations are therefore ignored by default in many statistical packages (i.e., list-wise deletion is employed).
+Unfortunaltely, this *ad hoc* solution may yield wildly invalid results [@buur18].
+An alternative is to 'impute' (i.e., fill in) every missing value in an incomplete dataset.
+With imputation techniques, one or several completed datasets are created, on which statistical inferences can be performed.
+The case with several completed datasets is known as 'multiple imputation' [MI; @rubin76], and requires an additional step to pool the results accross imputations.
+By applying Rubin's rules, the pooled result will reflect the uncertainty in the data due to missingness.
+Under many circumstances, this yields an unbiased and confidence valid estimate of the inference on the true--but missing--data [@buur18].
+A popular method to obtain imputations is to use the 'Multiple Imputation by Chained Equations' algorithm, shorthand 'MICE'.
+MICE is an iterative algorithmic procedure to draw imputations from the posterior predictive distribution of the missing values.
+This introduces a potential threat to the validity of the imputations: what if the algorithm has not converged?
+Are the implications then to be trusted?
+And can we rely on the inference obtained on the completed data?
+These are all open questions, because the convergence properties of iterative imputation algortihms have not been systematically studied [@taka17; @buur18].
+In this paper, we investigate the convergence properties of iterative imputation procedures by means of model based simulation in `R` [@R].
+
+<!-- It has been shown that this pooled result MI will yield unbiased, confidence valid results under many missing data circumstances [@buur18]. To implement MI,  -->
+
+<!-- Rubin's rules [-@rubin76] may be used to reflect the uncertainty in the inferences due to missingness by pooling the results across 'imputations'.
+A popular method to obtain imputations   completions, the analysis of scientific interest can be performed.
+It has been , or to complete these observations by imputing the missing values.
+Imputation is the method of filling in  -->
+
+<!-- by which  That is, unless ad hoc solutions are employed (e.g., list-wise deletion), which can yield invalid results.
+contains missing values, statistical analyses cannot be performed without  only be performed a dataset contains missing values, statistical analyses can statistical inferences cannot be performed   -->
+<!-- <!-- At some point, any social scientist  --> -->
+<!-- <!-- conducting statistical analyses  --> <!-- will run into a missing data problem [@alli02].  --> -->
+<!-- Missingness is problematic because statistical inference cannot be performed on the whole data without employing \emph{ad hoc} solutions (e.g., list-wise deletion), which may yield wildly invalid results [@buur18].
 A popular method to obtain statistically valid results despite of missing data is 'multiple imputation' [MI; @rubin76].
@@
 And finally, the results of these analyses are pooled into a single aggregated statistic.
+-->
 
@@
 
-Algorithms are often used to draw imputations from the posterior predictive distribution of the missing values.
+<!-- Algorithms are often used to draw imputations from the posterior predictive distribution of the missing values.
 This introduces a potential threat to the validity of the imputations: what if the algorithm has not converged?
@@
 And can we rely on the inference obtained on the completed data?
+-->
 
@@
 
-There is no scientific consensus on the convergence properties of MI algorithms [@taka17].
-Some default techniques (e.g., 'predictive mean modeling') might not yield converged states at all [@murr18].
-Therefore, algorithmic convergence should be monitored carefully.
-The current practice is visually inspecting imputation chains for signs of non-convergence, but this may be challenging to the untrained eye [@buur18, $\S$ 6.5.2].
-Moreover, visually inspecting imputation chains may only diagnose severely pathological cases of non-convergence.
-Therefore, diagnostic evaluation of convergence would be preferred.
+**(There is no scientific consensus on the convergence properties of MI algorithms [@taka17].
+Some default MICE techniques (e.g., 'predictive mean modeling') might not yield converged states at all [@murr18].
+Therefore, algorithmic convergence should be monitored carefully.)**
 
-Iterative imputation algorithms such as mice, are MCMC methods, and convergence of MCMC algorithms is not from a scalar to a point but from one distribution to another.
+The recommended practice to monitor convergence is to visually inspect imputation chains for signs of non-convergence.
+This practice is insuffcient on two counts: 1) it may be challenging to the untrained eye [@buur18, $\S$ 6.5.2], and 2) only severely pathological cases of non-convergence may be diagnosed.
+Therefore, a quantitative, diagnostic evaluation of convergence would be preferred.
+
+Iterative imputation algorithms such as MICE are MCMC methods, and convergence of MCMC algorithms is not from a scalar to a point but from one distribution to another.
 Therefore, generated values will vary even after convergence.
-The aim of convergence diagnostics for MCMC methods is thus not to establish the point at which convergence is reached, but to diagnose severe non-convergence.
+The aim of convergence diagnostics for MCMC methods is thus not to establish the point at which convergence is reached, but to diagnose signs of  non-convergence.
 <!-- Therefore we can only diagnose non-convergence, not convergence [@hoff09]. "a weak diagnostics is better than no diagnostic at all" [@cowl96]. -->
 
-Several MCMC convergence diagnostics exist, but it is not known whetherthese are appropriate for MICE.
+Several MCMC convergence diagnostics exist, but it is not known whether these are appropriate for MICE.
 The application of convergence diagnostics to MI methods has not been systematically studied [@buur18].

### 2020-04-15 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Include the acf into simulation again to show differences

@@
 
-Missing data problems are ubiquitous and pervasive. \textbf{or}
-Everybody who works with person data is likely to run into a missing
-data problem.\\
+Missing data problems are ubiquitous and pervasive in the social and
+behavioral sciences. If a dataset contains just one missing value for a
+variable of interest, statistical inferences are undefined and will not
+yield results. Incomplete observations are therefore ignored by default
+in many statistical packages (i.e., list-wise deletion is employed).
+Unfortunaltely, this \emph{ad hoc} solution may yield wildly invalid
+results \citep{buur18}. An alternative is to `impute' (i.e., fill in)
+every missing value in the incomplete dataset. With imputation
+techniques, one or several completed datasets are created, on which
+statistical inferences can be performed. The case with several completed
+datasets is known as `multiple imputation' \citep[MI;][]{rubin76}, and
+requires and additional step to pool the results accross imputations.
+Because this pooled result will reflect the uncertainty in the data due
+to missingness, it is in many cases an unbiased and confidence valid
+estimate of the true (but missing) data inference \citep{buur18}. A
+popular method to obtain imputations is to use the `Multiple Imputation
+by Chained Equations' algorithm, shorthand `MICE'. MICE is an iterative
+algorithmic procedure to draw imputations from the posterior predictive
+distribution of the missing values. This introduces a potential threat
+to the validity of the imputations: what if the algorithm has not
+converged? Are the implications then to be trusted? And can we rely on
+the inference obtained on the completed data? These are open questions,
+because the convergence of iterative imputation algortihms has not been
+systematically studied \citep{buur18}. In this paper, we investigate the
+convergence properties of iterative imputation procedures by means of
+model based simulation in \texttt{R} list(title = ``R: A Language and
+Environment for Statistical Computing'', author = list(list(given = ``R
+Core Team'', family = NULL, role = NULL, email = NULL, comment = NULL)),
+organization = ``R Foundation for Statistical Computing'', address =
+``Vienna, Austria'', year = ``2020'', url =
+``\url{https://www.R-project.org/}'').
+
+Rubin's rules \citeyearpar{rubin76} may be used to reflect the
+uncertainty in the inferences due to missingness by pooling the results
+across `imputations'. A popular method to obtain imputations
+completions, the analysis of scientific interest can be performed. It
+has been , or to complete these observations by imputing the missing
+values. Imputation is the method of filling in
+
+by which That is, unless ad hoc solutions are employed (e.g., list-wise
+deletion), which can yield invalid results. contains missing values,
+statistical analyses cannot be performed without only be performed a
+dataset contains missing values, statistical analyses can statistical
+inferences cannot be performed\\
 Missingness is problematic because statistical inference cannot be

### 2020-04-15 3.Thesis/1.SimulationStudy/Functions/Autocorrelation.R

@@ -1,7 +1,7 @@
 # Compute the maximum auto-correlation at lag 1
 # Used within imputation function
 
-autocorr_function <- function(imp, maxit, m = 5, n.var = 4, moment = "mean"){
+autocorr_function <- function(imp, maxit, m = 5, n.var = 4, moment = "mean", ac_function = "ac"){
   
   # object for for loop
   ac_per_chain <- numeric(length = m)
@@ -18,15 +18,23 @@ autocorr_function <- function(imp, maxit, m = 5, n.var = 4, moment = "mean"){
   # compute convergence diagnostic per variable
   for (v in 1:n.var) {
     for (chain in 1:m) {
-      # ac[chain, v] <-
-      ac_per_chain[chain] <-
-        ac_lag1(sims[v, , chain])
+      if (ac_function == "acf"){
+        # use original afc
+        ac_per_chain[chain] <-	      
+          acf(sims[v, , chain],lag.max = 1, plot = F)$acf[2]
+      } else if (ac_function == "ac"){
+        # use my own ac
+        ac_per_chain[chain] <- ac_lag1(sims[v, , chain])
+       }
     }
     ac_per_var[v] <- max(ac_per_chain)
 }
 # output
   return(ac_per_var)
 }
+
+ 	          
+
 
 # NB. This function can be adjusted to show the nr. of chains with significant auto-correlations,
 # output will then be the fraction of imputation chains with significant autocorrelation at lag 1.

### 2020-04-15 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
Rewrote intro for peer feedback research seminar.

@@
 email: h.i.oberman@uu.nl
-abstract: "(**Rewrite:**) Multple imputation by chained equations (mice) is a widely used tool to accommodate missing data.
+abstract: "(**Rewrite:**) Multiple imputation by chained equations (mice) is a widely used tool to accommodate missing data.
 While empirical evidence supports the validity of inferences obtained using mice, there is no consensus on the convergence properties of the method.
@@
 
+**Note for peer review: things written in bold are reminders for myself or things that I am unsure off still.
+Also, I just decided to change things majorly in my results section.
+So right now, the figures do not correspond to the text.
+Please only skim that section, because I am definitely going to rewrite it.
+Thanks already!
+- Hanne**
+
 # Introduction
@@
 Incomplete observations are therefore ignored by default in many statistical packages (i.e., list-wise deletion is employed).
-Unfortunaltely, this *ad hoc* solution may yield wildly invalid results [@buur18].
+Unfortunately, this *ad hoc* solution may yield wildly invalid results [@buur18].
 An alternative is to 'impute' (i.e., fill in) every missing value in an incomplete dataset.
 With imputation techniques, one or several completed datasets are created, on which statistical inferences can be performed.
-The case with several completed datasets is known as 'multiple imputation' [MI; @rubin76], and requires an additional step to pool the results accross imputations.
+The case with several completed datasets is known as 'multiple imputation' [MI; @rubin76], and requires an additional step to pool the results across imputations.
 By applying Rubin's rules, the pooled result will reflect the uncertainty in the data due to missingness.
 Under many circumstances, this yields an unbiased and confidence valid estimate of the inference on the true--but missing--data [@buur18].
-A popular method to obtain imputations is to use the 'Multiple Imputation by Chained Equations' algorithm, shorthand 'MICE'.
+A popular method to obtain imputations is to use the 'Multiple Imputation by Chained Equations' algorithm, shorthand 'MICE'[@mice].
 MICE is an iterative algorithmic procedure to draw imputations from the posterior predictive distribution of the missing values.
@@
 And can we rely on the inference obtained on the completed data?
-These are all open questions, because the convergence properties of iterative imputation algortihms have not been systematically studied [@taka17; @buur18].
-In this paper, we investigate the convergence properties of iterative imputation procedures by means of model based simulation in `R` [@R].
+These are all open questions, because the convergence properties of iterative imputation algorithms have not been systematically studied [@buur18].
+Moreover, there is no scientific consensus on the convergence properties of MI algorithms [@taka17].
+Some default MICE techniques (e.g., 'predictive mean modeling') might not yield converged states at all [@murr18].
+Therefore, algorithmic convergence should be monitored carefully.
+
+<!-- In this paper, we investigate the convergence properties of iterative imputation procedures by means of model based simulation in `R` [@R].  -->
 
@@
 contains missing values, statistical analyses cannot be performed without  only be performed a dataset contains missing values, statistical analyses can statistical inferences cannot be performed   -->
-<!-- <!-- At some point, any social scientist  --> -->
-<!-- <!-- conducting statistical analyses  --> <!-- will run into a missing data problem [@alli02].  --> -->
+<!-- At some point, any social scientist  -->
+<!-- conducting statistical analyses  --> <!-- will run into a missing data problem [@alli02].  -->
 <!-- Missingness is problematic because statistical inference cannot be performed on the whole data without employing \emph{ad hoc} solutions (e.g., list-wise deletion), which may yield wildly invalid results [@buur18].
@@
 
-<!-- The theoretical foundation of these streps was laid by Rubin [-@rubin76, -@rubin87].
+<!-- The theoretical foundation of these steeps was laid by Rubin [-@rubin76, -@rubin87].
 For an accessible and comprehensive introduction to MI from an applied perspective, see e.g.
@@
 
-<!-- based on a model of the observed databy means of an iterative algorithm  -->
+<!-- based on a model of the observed data by means of an iterative algorithm  -->
 <!-- is the use of MICE: multiple imputation by chained equations.
@@
 <!-- several times, and inferences are aggregated across imputed datasets.  -->
-<!-- and statistical inference is performed on each imputed dataset. Variation between the results refelect uncertainty due to missingness  -->
+<!-- and statistical inference is performed on each imputed dataset. Variation between the results reflect uncertainty due to missingness  -->
 <!-- CE: The imputation model can be a joint model for all variables, or several univariate conditional distributions (fully conditional specification).
@@
 
-**(There is no scientific consensus on the convergence properties of MI algorithms [@taka17].
-Some default MICE techniques (e.g., 'predictive mean modeling') might not yield converged states at all [@murr18].
-Therefore, algorithmic convergence should be monitored carefully.)**
-
-The recommended practice to monitor convergence is to visually inspect imputation chains for signs of non-convergence.
-This practice is insuffcient on two counts: 1) it may be challenging to the untrained eye [@buur18, $\S$ 6.5.2], and 2) only severely pathological cases of non-convergence may be diagnosed.
+The recommended practice to monitor convergence is to visually inspect imputations for signs of non-convergence.
+This practice is insufficient on two counts: 1) it may be challenging to the untrained eye, and 2) only severely pathological cases of non-convergence may be diagnosed [@buur18, $\S$ 6.5.2].
 Therefore, a quantitative, diagnostic evaluation of convergence would be preferred.
 
-Iterative imputation algorithms such as MICE are MCMC methods, and convergence of MCMC algorithms is not from a scalar to a point but from one distribution to another.
-Therefore, generated values will vary even after convergence.
-The aim of convergence diagnostics for MCMC methods is thus not to establish the point at which convergence is reached, but to diagnose signs of  non-convergence.
+Iterative imputation algorithms such as MICE are Markov chain Monte Carlo (MCMC) methods, which means that convergence is not from a scalar to a point but from one distribution to another.
+The values generated by the algorithm (e.g., imputed values) will vary even after convergence.
+Therefore, the aim of convergence diagnostics for MCMC methods should not be to establish the point at which convergence is reached, but to monitor signs of  non-convergence.
+Several convergence diagnostics for MCMC methods exist, but it is not known whether these are appropriate for MICE.
+
 <!-- Therefore we can only diagnose non-convergence, not convergence [@hoff09]. "a weak diagnostics is better than no diagnostic at all" [@cowl96]. -->
 
-Several MCMC convergence diagnostics exist, but it is not known whether these are appropriate for MICE.
-The application of convergence diagnostics to MI methods has not been systematically studied [@buur18].
-The open questions are, e.g.: How can non-convergence be diagnosed?
-How many iterations are sufficient/needed?
-What are the effects of non-convergence on estimates, predictions and inference?
-What is an appropriate non-convergence diagnostic?
-Are the default number of iterations sufficient?
-The defaults of common MI software: SPSS = 10 [(link)](https://www.ibm.com/support/knowledgecenter/SSLVMB_24.0.0/spss/mva/syn_multiple_imputation_impute.html), Mplus = 100??
+Main question: How can non-convergence be diagnosed?
+Sub-questions: Are common MCMC non-convergence diagnostics appropriate for MICE?
+And if so, which threshold should be used?
+How many iterations are sufficient/needed to reach convergence?
+Are the default number of iterations sufficient (i.e., 5 in mice, 10 in SPSS and Stata, 30 in mi)?
+What are the effects of continued iterations on estimates, predictions and inferences?
+And finally, do the answers differ with varying missingness proportions?
+
+<!-- The defaults of common MI software: SPSS = 10 [(link)](https://www.ibm.com/support/knowledgecenter/SSLVMB_24.0.0/spss/mva/syn_multiple_imputation_impute.html), Mplus = 100??
 [(link)](https://pdfs.semanticscholar.org/e20e/29e008592cbfbaa567931f74cdfdb5451405.pdf?_ga=2.55354671.54033656.1584698748-527613517.1584698748), Stata = 10 [(link)](https://www.stata.com/manuals13/mi.pdf, p.
-139), Amelia = NA, because resampling, not convergence [(link)](https://cran.r-project.org/web/packages/Amelia/Amelia.pdf), en MI = 30 [(link)](https://cran.r-project.org/web/packages/mi/mi.pdf).
+139), Amelia = NA, because re-sampling, not convergence [(link)](https://cran.r-project.org/web/packages/Amelia/Amelia.pdf), en MI = 30 [(link)](https://cran.r-project.org/web/packages/mi/mi.pdf).
+-->
 
@@
 The collection of observed data values in $y$ is denoted by $y_{obs}$; the missing part of $y$ is referred to as $y_{mis}$.
-(**This is not only notation, but theory too:** For each datapoint in $y_{mis}$, we sample $M \times T$ times plausible values, where $M$ is the number of imputations ($m = 1, \dots, M$) and $T$ is the number of iterations ($t = 1, \dots, T$).) The collection of samples between the initial value (at $t=1$) and the final imputed value (at $t=T$) will be referred to as an 'imputation chain'.
-
-**Add: missingness percentage = proportion of cases with one or more missing values times 100 **
+For each datapoint in $y_{mis}$, we sample $M \times T$ times plausible values, where $M$ is the number of imputations ($m = 1, \dots, M$) and $T$ is the number of iterations ($t = 1, \dots, T$).
+The collection of samples between the initial value (at $t=1$) and the final imputed value (at $t=T$) will be referred to as an 'imputation chain'.
+**Add: missingness percentage = proportion of cases with one or more missing values **
 <!-- The number of imputations is denoted by $m$ (i.e., the number of times each datapoint in $y_{mis}$ is filled in).
@@
 
-**(Start with current guidelines, then continue with possible quant measures of the same things from MCMC lit.)** In iterative algorithmic procedures (Markov chain Monte Carlo methods; e.g., `mice` algorithms or Gibbs samplers) non-convergence may be present as non-stationarity within chains (i.e., trending), or as slow mixing between chains (i.e., no intermingling (**source??**)).
-(**add example/guidelines visual inspection here?**)
+The two challenges for convergence of iterative algorithms are mixing and stationarity [@gelm13], see Figure 1.
+Without mixing, imputation chains do not intermingle nicely.
+Without stationarity, there is trending within imputation chains.
+While the current practice is to inspect traceplots for signs of non-mixing and non-stationarity, diagnostic evaluation of these two challenges would be preferred.
+Non-stationarity within chains may be diagnosed with e.g., autocorrelation [$AC$; @scha97; @gelm13], numeric standard error ['MC error'; @gewe92], or Raftery and Lewis's [-@raft91] procedure to determine the effect of trending on the precision of estimates.
+A widely used diagnostic to monitor mixing between chains is the potential scale reduction factor $\widehat{R}$ ['Gelman-Rubin statistic'; @gelm92].
+With a recently proposed adaptation, $\widehat{R}$ might also serve to diagnose non-stationarity, but this has not yet been thoroughly investigated [@veht19].
+Therefore, mixing and stationarity will be evaluated separately in this study.
+As recommended [e.g., @cowl96, p.
+898], $AC$ and $\widehat{R}$ will be used.
 
-The stationarity component of convergence may be evaluated with autocorrelation [$AC$; @scha97; @gelm13], numeric standard error [or `MC error'; @gewe92], and Raftery and Lewis's [-@raft91] procedure to determine  the effect of trending within chains (**or the chain length required to cancel out trending?**).
+<!-- The current practice in diagnosing non-convergence is to inspect traceplots for signs of non-mixing and non-stationarity.   -->
 
-The mixing component can be assessed with the potential scale reduction factor $\widehat{R}$ [a.k.a.
+<!-- The mixing component can be assessed with the potential scale reduction factor $\widehat{R}$ [a.k.a.
 `Gelman-Rubin statistic'; @gelm92].
@@
 898].
+-->
 
@@
 
-Note that all of these methods evaluate the convergence of univariate scalar summaries (e.g., chain means or variances).
+<!-- Note that all of these methods evaluate the convergence of univariate scalar summaries (e.g., chain means or variances).
 These convergence diagnostics cannot diagnose convergence of multivariable statistics (i.e., relations between scalar summaries).
@@
 This method is outside of the scope of the current study (**not anymore?**).
+-->
 
@@
 
-We can interpret $AC$-values as a measure of stationarity.
-If $AC$-values are close to zero, there is no dependence between subsequent samples within imputation chains.
-Positive $AC$-values, however, indicate recurrence.
+We can interpret $AC$-values as a measure of stationarity. If $AC$-values are close to zero, there is no dependence between subsequent samples within imputation chains.
+
+Positive $AC$-values indicate recurrence.
 If $\theta$-values of subsequent iterations are similar, trending may occur.
@@
 Negative $AC$-values indicate divergence within imputation chains.
-Subsequent sampled values within each imputation chain are less alike.
-It may not be wise (**why?
-and how would we know?
-e.g., wider distrubutions?
-didn't seem like it...**) to terminating the algorithm when AC is negative.
-**High $AC$-values are implausible in MI procedures.
+**Subsequent sampled values within each imputation chain are less alike.
+High $AC$-values are implausible in MI procedures.
 That is, the randomness induced by the MI algorithm effectively mitigates the risk of dependency within chains.**
@@
 We expect this to have little effect on the hypothesized performance of $\widehat{R}$.
-**(Add actual hypothesis.)** No hypothesis was formulated about the performance of $AC$ as convergence diagnostic.
+**(Add actual hypothesis: we thought that mice would converge sooner than $\widehat{R}$ would indicate that it did)** No hypothesis was formulated about the performance of $AC$ as convergence diagnostic.
 
@@
 
-```{r eval=FALSE}
+```{r, eval=FALSE, echo=TRUE}
 # pseudo-code of simulation 
@@
 ```
+
+**(Add step for different missingness proportions.)**
 
@@
 
-**Add the function `mvtnorm::rmvnorm()` back in** use citation("mvtnorm")
+**(Add the function `mvtnorm::rmvnorm()` back in somewhere here? If yes, use citation("mvtnorm").)**
 
 The complete data is 'amputed' once for each simulation repetition with function `mice::ampute()`.
-(**change this:** The missingness is univariate, and the probability to be missing is the same for all four variables, namely 20\% (`prop = 0.8, mech = "MCAR"`).
+(**change this to reflect current simulation set-up with 5, 25, 50, 75, and 95% of cases having missing data:** The missingness is univariate, and the probability to be missing is the same for all four variables, namely 20\% (`prop = 0.8, mech = "MCAR"`).
 This leaves 20\% of the rows completely observed).
-**(Add different percentages missingness??)** This study only considers only an MCAR missingness mechanism.
+This study only considers only an MCAR missingness mechanism.
 Therefore, results may not be extrapolated to other missing data problems.
@@
 
-## Univariate estimates and convergence diagnostics (theta = mean and theta = variance)
+```{r, include = FALSE}
+load("../Results/test.Rdata")
+source("../2.CreateFigures.R")
+```
+
+```{r}
+mean_bias + plot_annotation(title = "Bias in mean estimate, convergence of chain means")
+mean_Rh 
+mean_ACF + plot_annotation(title = "Dotted lines are old AC computation, solid are new.")
+```
+
+## Univariate estimates and convergence diagnostics
 
@@
 In most simulation conditions nominal coverage is obtained (i.e., coverage rates of .95) for the confidence intervals around the regression coefficients.
-Conditions with only one or two iterations show some undercoverage.
-Since confidence interval width is stable across conditions, the undercoverage may be attributed to the bias in the estimated regression coefficients.
+Conditions with only one or two iterations show some under-coverage.
+Since confidence interval width is stable across conditions, the under-coverage may be attributed to the bias in the estimated regression coefficients.
 
 If we look at the estimated proportion of explained variance in outcome variable Y we see that the coefficient of determination is underestimate estimated in conditions where the number of iterations is equal to two or less, and slightly overestimated in conditions where the number of iterations is equal to three or more.
+
+```{r}
+est_bias + plot_annotation(title = "Bias in regression estimate, coverage rate, \n bias in variance explained")
+est_cov
+Rsq_bias 
+```
 
@@
 <!-- Figures 2 and 3 display results per simulation condition ($T = 1,2,\dots,100$).  -->
-
-```{r, include = FALSE}
-load("../Results/complete.Rdata")
-source("../2.CreateFigures.R")
-```
-
-<!-- ```{r} -->
-<!-- bias_means_plot + R_means_plot + AC_means_plot + plot_layout(nrow = 3) + plot_annotation(title = "Convergence of means, 75% missingness") -->
-<!-- ``` -->
-
-```{r, echo = FALSE}
-#load("../Results/combined.Rdata")
-
-dat %>% ggplot(aes(x = T, y = bias.R.s, color = as.factor(miss))) + 
-  geom_point() +
-  geom_line() +
-  ylab(expression(paste("Bias in ", hat(R ^ 2)))) +
-  labs(colour = "Legend: missingness percentage") +
-  plot_annotation(title = "The effect of missingness percentage, an example")
-
-```
 
@@
 
-# Results [OLD, JUST FOR REFERENCE!]
+<!-- # Results [OLD, JUST FOR REFERENCE!] -->
 
-## Convergence Diagnostics
+<!-- ## Convergence Diagnostics -->
 
-**(Pair convergnce diagnostics with respective estimates, not split between convergence and simulation diagnostics)**
+<!-- **(Pair convergence diagnostics with respective estimates, not split between convergence and simulation diagnostics)** -->
 
-It is apparent that there is a relation between the number of iterations per simulation condition ($T$) and the convergence diagnostics.
+<!-- It is apparent that there is a relation between the number of iterations per simulation condition ($T$) and the convergence diagnostics.
 Generally speaking, conditions with longer imputation chains (higher $T$) coincide with less signs of non-convergence ($\widehat{R}$-values approach one, and $AC$-values approach zero).
-<!-- We see more or less equivalent trends for chain means versus chain variances. We, therefore, only discuss the $\widehat{R}$ and $AC$-values of chain means.  -->
+-->
+<!-- <!-- We see more or less equivalent trends for chain means versus chain variances. We, therefore, only discuss the $\widehat{R}$ and $AC$-values of chain means.  --> -->
 
-Figure 2A shows that $\widehat{R}$-values generally decrease with increasing imputation chain lengths.
+<!-- Figure 2A shows that $\widehat{R}$-values generally decrease with increasing imputation chain lengths.
 The decline stabilizes somewhere between the simulation conditions $T=30$ and $T=50$.
@@
 If we use the recently recommended threshold $\widehat{R} < 1.01$, we would conclude that convergence is reached in none of the simulation conditions.
+-->
 
-The $AC$-values displayed in Figure 2B are almost uniformly increasing as a function of $T$.
+<!-- The $AC$-values displayed in Figure 2B are almost uniformly increasing as a function of $T$.
 The only $AC$-value that deviates from this observation is for $T=2$.
@@
 %At $T=5$, the $AC$-value reaches the level observed at $T=2$.
-The gradual increase plateaus between $T=10$ and $T=30$.
+-->
+<!-- The gradual increase plateaus between $T=10$ and $T=30$.
 $AC$-values in conditions where $T>70$ are indifferentiable from zero, indicating stationarity.
@@
 Simulation conditions where $T>5$ have $AC$-values greater than $T=2$ (**weird wording**).
+-->
 
-According to this diagnostic, none of the simulation conditions show signs of non-convergence, since we only observe negative or zero $AC$-values.
+<!-- According to this diagnostic, none of the simulation conditions show signs of non-convergence, since we only observe negative or zero $AC$-values.  -->
 
-Taken together, we see that $T>3$ is the minimal requirement to diagnose convergence ($\widehat{R} < 1.2; AC \leq 0$).
+<!-- Taken together, we see that $T>3$ is the minimal requirement to diagnose convergence ($\widehat{R} < 1.2; AC \leq 0$).
 This threshold is, however, not sufficient, since we overlook the increase in $\widehat{R}$-values up-to conditions where $T>5$, and the convergence diagnostics only reach stability at $T>20$.
+-->
 
-## Simulation Diagnostics
+<!-- ## Simulation Diagnostics -->
 
-We use average bias, average confidence interval width, and coverage rate as performance measures to evaluate $\widehat{R}$ and $AC$ (**not the goal!
+<!-- We use average bias, average confidence interval width, and coverage rate as performance measures to evaluate $\widehat{R}$ and $AC$ (**not the goal!
 we want to show the effects of non-conv, not just see how to diagnose it.**).
 We make the general observation that the simulation diagnostics behave as theorized for most simulation conditions (bias around zero, stable confidence interval widths, nominal coverage rate at 95\%).
+-->
 
-Figure 3A shows that bias is fairly stable across simulation conditions.
+<!-- Figure 3A shows that bias is fairly stable across simulation conditions.
 The condition $T=1$ clearly deviates from this trend with a negative bias.
@@
 Conditions where $T>3$ can be diagnosed as unbiased, but the average bias across iterations (as shown by a flat Loess line) only reaches stability at $T=20$ (**Loess might not be relevant, check and remove accordingly!**).
+-->
 
-CIW is only clearly divergent in the simulation condition $T=1$, see Figure 3B.
+<!-- CIW is only clearly divergent in the simulation condition $T=1$, see Figure 3B.
 Only conditions where $T>3$ are therefore considered sufficient.
-(**Possibly irrelevant Loess:** These conditions have similar CIWs, but across iterations stability is not established (i.e., the Loess line is never completely flat within the 100 simulation conditions).)
+(**Possibly irrelevant Loess:** These conditions have similar CIWs, but across iterations stability is not established (i.e., the Loess line is never completely flat within the 100 simulation conditions).)   -->
 
-The empirical coverage rate across repetitions seems more or less stable for conditions where $T>1$, see Figure 3C.
+<!-- The empirical coverage rate across repetitions seems more or less stable for conditions where $T>1$, see Figure 3C.
 We see some over-coverage at $T=2$, but that is better than under-estimating the variance of $\bar{Q}$.
@@
 (**Loess:** Similar to the CIWs, there is some trending across iterations (i.e., the Loess line is never flat)).
+-->
 
-From the simulation diagnostics, we observe that unbiased estimates with nominal coverage rates were obtained in conditions where $T>3$.
+<!-- From the simulation diagnostics, we observe that unbiased estimates with nominal coverage rates were obtained in conditions where $T>3$.
 This suggests that as little as four iterations may be sufficient for the MI algorithm under the current circumstances.
+-->
 
-In short, there is a discrepancy between what the convergence diagnostics and what the performance measures indicate.
+<!-- In short, there is a discrepancy between what the convergence diagnostics and what the performance measures indicate.
 While $T=4$ seems sufficient with respect to simulation quantities, the condition where $T=4$ resulted in the `one-but-worst' values of both convergence diagnostics.
 Complete algorithmic convergence as indicated by $\widehat{R}$ and autocorrelation is not reached in conditions where $T<20$.
+-->
 
@@
 (**nope!
-it shows that MICE can lead to correct outcomes when they have not converged accroding to two common conv diags.
+it shows that MICE can lead to correct outcomes when they have not converged according to two common conv diags.
 This may be due to the measures (e.g., assumption of overdisp) or due to the Qs (lm reg coeff, not higher dimensional/more complex RQs).

### 2020-04-15 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Rewrote intro for peer feedback research seminar.

--- 2020-04-09 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
@@
 \begin{abstract}
-(\textbf{Rewrite:}) Multple imputation by chained equations (mice) is a
+(\textbf{Rewrite:}) Multiple imputation by chained equations (mice) is a
 widely used tool to accommodate missing data. While empirical evidence
@@
 
+\textbf{Note for peer review: things written in bold are reminders for
+myself or things that I am unsure off still. Also, I just decided to
+change things majorly in my results section. So right now, the figures
+do not correspond to the text. Please only skim that section, because I
+am definitely going to rewrite it. Thanks already! - Hanne}
+
 \hypertarget{introduction}{%
@@
 
-Missing data problems are ubiquitous and pervasive. \textbf{or}
-Everybody who works with person data is likely to run into a missing
-data problem.\\
-Missingness is problematic because statistical inference cannot be
-performed on the whole data without employing \emph{ad hoc} solutions
-(e.g., list-wise deletion), which may yield wildly invalid results
-\citep{buur18}. A popular method to obtain statistically valid results
-despite of missing data is `multiple imputation' \citep[MI;][]{rubin76}.
-The multiple imputation method can be roughly divided into 3 steps:
-First, each missing value in an incomplete dataset is `imputed' (i.e.,
-filled in) several times by some algorithmic procedure. Next, the
-analysis of scientific interest is performed on each of the completed
-datasets. And finally, the results of these analyses are pooled into a
-single aggregated statistic.
+Missing data problems are ubiquitous and pervasive in the social and
+behavioral sciences. If a dataset contains just one missing value for a
+variable of interest, statistical inferences are undefined and will not
+produce any result. Incomplete observations are therefore ignored by
+default in many statistical packages (i.e., list-wise deletion is
+employed). Unfortunately, this \emph{ad hoc} solution may yield wildly
+invalid results \citep{buur18}. An alternative is to `impute' (i.e.,
+fill in) every missing value in an incomplete dataset. With imputation
+techniques, one or several completed datasets are created, on which
+statistical inferences can be performed. The case with several completed
+datasets is known as `multiple imputation' \citep[MI;][]{rubin76}, and
+requires an additional step to pool the results across imputations. By
+applying Rubin's rules, the pooled result will reflect the uncertainty
+in the data due to missingness. Under many circumstances, this yields an
+unbiased and confidence valid estimate of the inference on the true--but
+missing--data \citep{buur18}. A popular method to obtain imputations is
+to use the `Multiple Imputation by Chained Equations' algorithm,
+shorthand `MICE'\citep{mice}. MICE is an iterative algorithmic procedure
+to draw imputations from the posterior predictive distribution of the
+missing values. This introduces a potential threat to the validity of
+the imputations: what if the algorithm has not converged? Are the
+implications then to be trusted? And can we rely on the inference
+obtained on the completed data? These are all open questions, because
+the convergence properties of iterative imputation algorithms have not
+been systematically studied \citep{buur18}. Moreover, there is no
+scientific consensus on the convergence properties of MI algorithms
+\citep{taka17}. Some default MICE techniques (e.g., `predictive mean
+modeling') might not yield converged states at all \citep{murr18}.
+Therefore, algorithmic convergence should be monitored carefully.
 
-Algorithms are often used to draw imputations from the posterior
-predictive distribution of the missing values. This introduces a
-potential threat to the validity of the imputations: what if the
-algorithm has not converged? Are the implications then to be trusted?
-And can we rely on the inference obtained on the completed data?
+The recommended practice to monitor convergence is to visually inspect
+imputations for signs of non-convergence. This practice is insufficient
+on two counts: 1) it may be challenging to the untrained eye, and 2)
+only severely pathological cases of non-convergence may be diagnosed
+\citep[\(\S\) 6.5.2]{buur18}. Therefore, a quantitative, diagnostic
+evaluation of convergence would be preferred.
 
-There is no scientific consensus on the convergence properties of MI
-algorithms \citep{taka17}. Some default techniques (e.g., `predictive
-mean modeling') might not yield converged states at all \citep{murr18}.
-Therefore, algorithmic convergence should be monitored carefully. The
-current practice is visually inspecting imputation chains for signs of
-non-convergence, but this may be challenging to the untrained eye
-\citep[\(\S\) 6.5.2]{buur18}. Moreover, visually inspecting imputation
-chains may only diagnose severely pathological cases of non-convergence.
-Therefore, diagnostic evaluation of convergence would be preferred.
+Iterative imputation algorithms such as MICE are Markov chain Monte
+Carlo (MCMC) methods, which means that convergence is not from a scalar
+to a point but from one distribution to another. The values generated by
+the algorithm (e.g., imputed values) will vary even after convergence.
+Therefore, the aim of convergence diagnostics for MCMC methods should
+not be to establish the point at which convergence is reached, but to
+monitor signs of non-convergence. Several convergence diagnostics for
+MCMC methods exist, but it is not known whether these are appropriate
+for MICE.
 
-Iterative imputation algorithms such as mice, are MCMC methods, and
-convergence of MCMC algorithms is not from a scalar to a point but from
-one distribution to another. Therefore, generated values will vary even
-after convergence. The aim of convergence diagnostics for MCMC methods
-is thus not to establish the point at which convergence is reached, but
-to diagnose severe non-convergence.
-
-Several MCMC convergence diagnostics exist, but it is not known
-whetherthese are appropriate for MICE. The application of convergence
-diagnostics to MI methods has not been systematically studied
-\citep{buur18}. The open questions are, e.g.: How can non-convergence be
-diagnosed? How many iterations are sufficient/needed? What are the
-effects of non-convergence on estimates, predictions and inference? What
-is an appropriate non-convergence diagnostic? Are the default number of
-iterations sufficient? The defaults of common MI software: SPSS = 10
-\href{https://www.ibm.com/support/knowledgecenter/SSLVMB_24.0.0/spss/mva/syn_multiple_imputation_impute.html}{(link)},
-Mplus = 100??
-\href{https://pdfs.semanticscholar.org/e20e/29e008592cbfbaa567931f74cdfdb5451405.pdf?_ga=2.55354671.54033656.1584698748-527613517.1584698748}{(link)},
-Stata = 10
-\href{https://www.stata.com/manuals13/mi.pdf,\%20p.\%20139}{(link)},
-Amelia = NA, because resampling, not convergence
-\href{https://cran.r-project.org/web/packages/Amelia/Amelia.pdf}{(link)},
-en MI = 30
-\href{https://cran.r-project.org/web/packages/mi/mi.pdf}{(link)}.
+Main question: How can non-convergence be diagnosed? Sub-questions: Are
+common MCMC non-convergence diagnostics appropriate for MICE? And if so,
+which threshold should be used? How many iterations are
+sufficient/needed to reach convergence? Are the default number of
+iterations sufficient (i.e., 5 in mice, 10 in SPSS and Stata, 30 in mi)?
+What are the effects of continued iterations on estimates, predictions
+and inferences? And finally, do the answers differ with varying
+missingness proportions?
 
@@
 \(y\) is denoted by \(y_{obs}\); the missing part of \(y\) is referred
-to as \(y_{mis}\). (\textbf{This is not only notation, but theory too:}
-For each datapoint in \(y_{mis}\), we sample \(M \times T\) times
-plausible values, where \(M\) is the number of imputations
-(\(m = 1, \dots, M\)) and \(T\) is the number of iterations
-(\(t = 1, \dots, T\)).) The collection of samples between the initial
+to as \(y_{mis}\). For each datapoint in \(y_{mis}\), we sample
+\(M \times T\) times plausible values, where \(M\) is the number of
+imputations (\(m = 1, \dots, M\)) and \(T\) is the number of iterations
+(\(t = 1, \dots, T\)). The collection of samples between the initial
 value (at \(t=1\)) and the final imputed value (at \(t=T\)) will be
-referred to as an `imputation chain'.
+referred to as an `imputation chain'. \textbf{Add: missingness
+percentage = proportion of cases with one or more missing values }
 
@@
 
-\textbf{(Start with current guidelines, then continue with possible
-quant measures of the same things from MCMC lit.)} In iterative
-algorithmic procedures (Markov chain Monte Carlo methods; e.g.,
-\texttt{mice} algorithms or Gibbs samplers) non-convergence may be
-present as non-stationarity within chains (i.e., trending), or as slow
-mixing between chains (i.e., no intermingling (\textbf{source??})).
-(\textbf{add example/guidelines visual inspection here?})
-
-The stationarity component of convergence may be evaluated with
+The two challenges for convergence of iterative algorithms are mixing
+and stationarity \citep{gelm13}, see Figure 1. Without mixing,
+imputation chains do not intermingle nicely. Without stationarity, there
+is trending within imputation chains. While the current practice is to
+inspect traceplots for signs of non-mixing and non-stationarity,
+diagnostic evaluation of these two challenges would be preferred.
+Non-stationarity within chains may be diagnosed with e.g.,
 autocorrelation \citep[\(AC\);][]{scha97, gelm13}, numeric standard
-error \citep[or `MC error';][]{gewe92}, and Raftery and Lewis's
-\citeyearpar{raft91} procedure to determine the effect of trending
-within chains (\textbf{or the chain length required to cancel out
-trending?}).
-
-The mixing component can be assessed with the potential scale reduction
-factor \(\widehat{R}\) \citep[a.k.a. `Gelman-Rubin
-statistic';][]{gelm92}. With an adapted version of \(\widehat{R}\),
-proposed by \citet{veht19}, the stationarity component of convergence
-might be evaluated as well. This would make \(\widehat{R}\) a general
-convergence diagnostic. The application of \(\widehat{R}\) to assess
-stationarity has not been thoroughly investigated. Therefore, this study
-employs both \(\widehat{R}\) and autocorrelation to investigate
-convergence, as recommended by \citep[p.~898]{cowl96}.
-
-Note that all of these methods evaluate the convergence of univariate
-scalar summaries (e.g., chain means or variances). These convergence
-diagnostics cannot diagnose convergence of multivariable statistics
-(i.e., relations between scalar summaries). \citet{buur18} proposed to
-implement multivariable evaluation through eigenvalue decomposition
-\citep{mack03}. This method is outside of the scope of the current study
-(\textbf{not anymore?}).
+error \citep[`MC error';][]{gewe92}, or Raftery and Lewis's
+\citeyearpar{raft91} procedure to determine the effect of trending on
+the precision of estimates. A widely used diagnostic to monitor mixing
+between chains is the potential scale reduction factor \(\widehat{R}\)
+\citep[`Gelman-Rubin statistic';][]{gelm92}. With a recently proposed
+adaptation, \(\widehat{R}\) might also serve to diagnose
+non-stationarity, but this has not yet been thoroughly investigated
+\citep{veht19}. Therefore, mixing and stationarity will be evaluated
+separately in this study. As recommended \citep[e.g.,][p.~898]{cowl96},
+\(AC\) and \(\widehat{R}\) will be used.
 
@@
 \(AC\)-values are close to zero, there is no dependence between
-subsequent samples within imputation chains. Positive \(AC\)-values,
-however, indicate recurrence. If \(\theta\)-values of subsequent
-iterations are similar, trending may occur. Negative \(AC\)-values show
-no threat to the stationarity component of convergence. On the contrary
-even---negative \(AC\)-values indicate that \(\theta\)-values of
-subsequent iterations diverge from one-another, which may increase the
-variance of \(\theta\) and speed up convergence. As convergence
-diagnostic, the interest is therefore in positive \(AC\)-values.
-Moreover, the magnitude of \(AC\)-values may be evaluated statistically,
-but that is outside of this note's scope.
+subsequent samples within imputation chains.
+
+Positive \(AC\)-values indicate recurrence. If \(\theta\)-values of
+subsequent iterations are similar, trending may occur. Negative
+\(AC\)-values show no threat to the stationarity component of
+convergence. On the contrary even---negative \(AC\)-values indicate that
+\(\theta\)-values of subsequent iterations diverge from one-another,
+which may increase the variance of \(\theta\) and speed up convergence.
+As convergence diagnostic, the interest is therefore in positive
+\(AC\)-values. Moreover, the magnitude of \(AC\)-values may be evaluated
+statistically, but that is outside of this note's scope.
 
 Negative \(AC\)-values indicate divergence within imputation chains.
-Subsequent sampled values within each imputation chain are less alike.
-It may not be wise (\textbf{why? and how would we know? e.g., wider
-distrubutions? didn't seem like it\ldots{}}) to terminating the
-algorithm when AC is negative. \textbf{High \(AC\)-values are
-implausible in MI procedures. That is, the randomness induced by the MI
-algorithm effectively mitigates the risk of dependency within chains.}
+\textbf{Subsequent sampled values within each imputation chain are less
+alike. High \(AC\)-values are implausible in MI procedures. That is, the
+randomness induced by the MI algorithm effectively mitigates the risk of
+dependency within chains.}
 
@@
 effect on the hypothesized performance of \(\widehat{R}\). \textbf{(Add
-actual hypothesis.)} No hypothesis was formulated about the performance
-of \(AC\) as convergence diagnostic.
+actual hypothesis: we thought that mice would converge sooner than
+\(\widehat{R}\) would indicate that it did)} No hypothesis was
+formulated about the performance of \(AC\) as convergence diagnostic.
 
@@
 \href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy}{github.com/gerkovink/shinyMice}.
+
+\begin{verbatim}
+# pseudo-code of simulation 
+simulate data 
+for (number of simulation runs from 1 to 1000)
+  for (number of iterations from 1 to 100)
+    create missingness
+    impute the missingness
+    compute convergence diagnostics
+    perform analysis
+    pool results
+    compute simulation diagnostics
+aggregate convergence and simulation diagnostics
+\end{verbatim}
+
+\textbf{(Add step for different missingness proportions.)}
 
@@
 
-\textbf{Add the function \texttt{mvtnorm::rmvnorm()} back in}
+\textbf{(Add the function \texttt{mvtnorm::rmvnorm()} back in somewhere
+here? If yes, use citation(``mvtnorm'').)}
 
 The complete data is `amputed' once for each simulation repetition with
-function \texttt{mice::ampute()}. (\textbf{change this:} The missingness
-is univariate, and the probability to be missing is the same for all
-four variables, namely 20\% (\texttt{prop\ =\ 0.8,\ mech\ =\ "MCAR"}).
-This leaves 20\% of the rows completely observed). \textbf{(Add
-different percentages missingness??)} This study only considers only an
-MCAR missingness mechanism. Therefore, results may not be extrapolated
-to other missing data problems. Proper performance of the convergence
-diagnostics under MCAR is necessary but not sufficient to demonstrate
-appropriateness of \(\widehat{R}\) and \(AC\) as convergence
-diagnostics. This is just a proof of concept.
+function \texttt{mice::ampute()}. (\textbf{change this to reflect
+current simulation set-up with 5, 25, 50, 75, and 95\% of cases having
+missing data:} The missingness is univariate, and the probability to be
+missing is the same for all four variables, namely 20\%
+(\texttt{prop\ =\ 0.8,\ mech\ =\ "MCAR"}). This leaves 20\% of the rows
+completely observed). This study only considers only an MCAR missingness
+mechanism. Therefore, results may not be extrapolated to other missing
+data problems. Proper performance of the convergence diagnostics under
+MCAR is necessary but not sufficient to demonstrate appropriateness of
+\(\widehat{R}\) and \(AC\) as convergence diagnostics. This is just a
+proof of concept.
 
@@
 
-\hypertarget{univariate-estimates-and-convergence-diagnostics-theta-mean-and-theta-variance}{%
-\subsection{Univariate estimates and convergence diagnostics (theta =
-mean and theta =
-variance)}\label{univariate-estimates-and-convergence-diagnostics-theta-mean-and-theta-variance}}
+\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-3-1} \end{flushleft}
+
+\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-3-2} \end{flushleft}
+
+\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-3-3} \end{flushleft}
+
+\hypertarget{univariate-estimates-and-convergence-diagnostics}{%
+\subsection{Univariate estimates and convergence
+diagnostics}\label{univariate-estimates-and-convergence-diagnostics}}
 
@@
 intervals around the regression coefficients. Conditions with only one
-or two iterations show some undercoverage. Since confidence interval
-width is stable across conditions, the undercoverage may be attributed
+or two iterations show some under-coverage. Since confidence interval
+width is stable across conditions, the under-coverage may be attributed
 to the bias in the estimated regression coefficients.
@@
 
+\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-4-1} \end{flushleft}
+
+\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-4-2} \end{flushleft}
+
+\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-4-3} \end{flushleft}
+
 In short, we see that the minimum number of iterations required to
@@
 
-\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-3-1} \end{flushleft}
-
-\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-4-1} \end{flushleft}
-
-\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-5-1} \end{flushleft}
-
-\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-6-1} \end{flushleft}
-
-\hypertarget{results-old-just-for-reference}{%
-\section{Results {[}OLD, JUST FOR
-REFERENCE!{]}}\label{results-old-just-for-reference}}
-
-\hypertarget{convergence-diagnostics-1}{%
-\subsection{Convergence Diagnostics}\label{convergence-diagnostics-1}}
-
-\textbf{(Pair convergnce diagnostics with respective estimates, not
-split between convergence and simulation diagnostics)}
-
-It is apparent that there is a relation between the number of iterations
-per simulation condition (\(T\)) and the convergence diagnostics.
-Generally speaking, conditions with longer imputation chains (higher
-\(T\)) coincide with less signs of non-convergence
-(\(\widehat{R}\)-values approach one, and \(AC\)-values approach zero).
-
-Figure 2A shows that \(\widehat{R}\)-values generally decrease with
-increasing imputation chain lengths. The decline stabilizes somewhere
-between the simulation conditions \(T=30\) and \(T=50\). The downward
-trend is most pronounced for \(T=3\), and between \(T = 5\) and
-\(T = 10\). In the intervening conditions (\(3 \leq T \leq 5\)),
-however, we observe a steep increase in \(\widehat{R}\)-values. This
-increase implies that non-convergence is under-estimated, and
-convergence should not be diagnosed at this point (\textbf{is that
-true???}). Based on the conventional threshold \(\widehat{R} < 1.2\), we
-would falsely diagnose convergence at \(T=3\). According to the widely
-used threshold \(\widehat{R} < 1.1\), convergence would be diagnosed for
-conditions where \(T>9\). If we use the recently recommended threshold
-\(\widehat{R} < 1.01\), we would conclude that convergence is reached in
-none of the simulation conditions.
-
-The \(AC\)-values displayed in Figure 2B are almost uniformly increasing
-as a function of \(T\). The only \(AC\)-value that deviates from this
-observation is for \(T=2\). The lowest \(AC\)-value is obtained for
-\(T=3\). \%At \(T=5\), the \(AC\)-value reaches the level observed at
-\(T=2\). The gradual increase plateaus between \(T=10\) and \(T=30\).
-\(AC\)-values in conditions where \(T>70\) are indifferentiable from
-zero, indicating stationarity. We see an initial decrease between
-\(T=2\) and \(T=3\). Simulation conditions where \(T>5\) have
-\(AC\)-values greater than \(T=2\) (\textbf{weird wording}).
-
-According to this diagnostic, none of the simulation conditions show
-signs of non-convergence, since we only observe negative or zero
-\(AC\)-values.
-
-Taken together, we see that \(T>3\) is the minimal requirement to
-diagnose convergence (\(\widehat{R} < 1.2; AC \leq 0\)). This threshold
-is, however, not sufficient, since we overlook the increase in
-\(\widehat{R}\)-values up-to conditions where \(T>5\), and the
-convergence diagnostics only reach stability at \(T>20\).
-
-\hypertarget{simulation-diagnostics}{%
-\subsection{Simulation Diagnostics}\label{simulation-diagnostics}}
-
-We use average bias, average confidence interval width, and coverage
-rate as performance measures to evaluate \(\widehat{R}\) and \(AC\)
-(\textbf{not the goal! we want to show the effects of non-conv, not just
-see how to diagnose it.}). We make the general observation that the
-simulation diagnostics behave as theorized for most simulation
-conditions (bias around zero, stable confidence interval widths, nominal
-coverage rate at 95\%).
-
-Figure 3A shows that bias is fairly stable across simulation conditions.
-The condition \(T=1\) clearly deviates from this trend with a negative
-bias. The bias for \(T=2\) is below average, but within the range of
-fluctuations. Conditions where \(T>3\) can be diagnosed as unbiased, but
-the average bias across iterations (as shown by a flat Loess line) only
-reaches stability at \(T=20\) (\textbf{Loess might not be relevant,
-check and remove accordingly!}).
-
-CIW is only clearly divergent in the simulation condition \(T=1\), see
-Figure 3B. Only conditions where \(T>3\) are therefore considered
-sufficient. (\textbf{Possibly irrelevant Loess:} These conditions have
-similar CIWs, but across iterations stability is not established (i.e.,
-the Loess line is never completely flat within the 100 simulation
-conditions).)
-
-The empirical coverage rate across repetitions seems more or less stable
-for conditions where \(T>1\), see Figure 3C. We see some over-coverage
-at \(T=2\), but that is better than under-estimating the variance of
-\(\bar{Q}\). On average, the coverage rate is somewhat higher than the
-expected nominal coverage of 95\% \citep{neym34} (\textbf{not
-anymore??}). (\textbf{Loess:} Similar to the CIWs, there is some
-trending across iterations (i.e., the Loess line is never flat)).
-
-From the simulation diagnostics, we observe that unbiased estimates with
-nominal coverage rates were obtained in conditions where \(T>3\). This
-suggests that as little as four iterations may be sufficient for the MI
-algorithm under the current circumstances.
-
-In short, there is a discrepancy between what the convergence
-diagnostics and what the performance measures indicate. While \(T=4\)
-seems sufficient with respect to simulation quantities, the condition
-where \(T=4\) resulted in the `one-but-worst' values of both convergence
-diagnostics. Complete algorithmic convergence as indicated by
-\(\widehat{R}\) and autocorrelation is not reached in conditions where
-\(T<20\).
+--\textgreater{}
 
@@
 algorithmic procedures. (\textbf{nope! it shows that MICE can lead to
-correct outcomes when they have not converged accroding to two common
+correct outcomes when they have not converged according to two common
 conv diags. This may be due to the measures (e.g., assumption of

### 2020-04-16 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
Editing the methods section to be conform recommendations Morris et al.

@@
 
-Negative $AC$-values indicate divergence within imputation chains.
-**Subsequent sampled values within each imputation chain are less alike.
-High $AC$-values are implausible in MI procedures.
-That is, the randomness induced by the MI algorithm effectively mitigates the risk of dependency within chains.**
+Negative $AC$-values indicate divergence within imputation chains. **Subsequent sampled values within each imputation chain are less alike.**
 
@@
 **(Add actual hypothesis: we thought that mice would converge sooner than $\widehat{R}$ would indicate that it did)** No hypothesis was formulated about the performance of $AC$ as convergence diagnostic.
+**Added later: High $AC$-values are implausible in MI procedures.
+That is, the randomness induced by the MI algorithm effectively mitigates the risk of dependency within chains.**
 
@@
 
-The convergence of `mice` is investigated through model-based simulation in `R` [version `r paste(version$major,version$minor, sep = ".")`; @R].
+Convergence of the MICE algorithm is investigated through model-based simulation in `R` [version `r paste(version$major,version$minor, sep = ".")`; @R].
 The simulation set-up is summarized in the pseudo-code below.
 The complete `R` script of the simulation study is available from [github.com/gerkovink/shinyMice](https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy).
-
 <!-- Simulation conditions are the number of iterations ($T = 1, 2, \dots, 100$) and the percentage missingness in the  -->
@@
 
-```{r, eval=FALSE, echo=TRUE}
+```
 # pseudo-code of simulation 
-simulate data 
+1. simulate data 
+for (missingness proportions 5%, 25%, 50%, 75% and 95%)
   for (number of simulation runs from 1 to 1000)
+    2. create missingness
     for (number of iterations from 1 to 100)
-    create missingness
-    impute the missingness
-    compute convergence diagnostics
-    perform analysis
-    pool results
-    compute simulation diagnostics
-aggregate convergence and simulation diagnostics
+      3. impute missingness
+      4. compute convergence diagnostics
+      5. perform analysis of scientific interest
+      6. pool results across imputations
+      7. compute performance measures
+  8. aggregate outcomes 
+9. combine outcomes
 ```
+The aim of the simulation study is to evaluate the impact of: 1) terminating the MICE algorithm at different imputation chain lengths ($t = 1, 2, ..., 100$), and 2) varying the difficulty of the missingness problem (**different missingness proportions** $prop = .05, .25, .50, .75, .95$).
+For the purpose of this study, sampling variance is not of interest.
+Therefore, a single complete set may serve as comparative truth in all simulation runs [@vinknd].
 
-**(Add step for different missingness proportions.)**
-
-A finite population of $N=1000$ is simulated to solve a multiple linear regression problem, where dependent variable $Y$ is regressed on independent variables $X_1$, $X_2$ and $X_3$ (**check notation with betas here, and add what the quantity/-ies of scientific interest is/are!**):
-
-$$Y \sim \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3.$$
-
-The data generating model is a multivariate normal distribution
+The data-generating mechanism is a multivariate normal distribution, representing person data on three independent variables (from an unspecified social scientific field of study).
+A finite population of $N=1000$ is simulated using the function `mvtnorm::rmvnorm()`, where
 
@@
 \end{pmatrix}
-\end{bmatrix}\\[2\jot]
+\end{bmatrix}\!\!\text{, and}\\[2\jot]
+Y &=  1 + 2X_1 + .5X_2 - X_3 + \epsilon .
 \end{align*} 
+Subsequently, 
+We consider four quantities of scientific interest ['conceptual estimands'; @morr19], namely  The quantity of scientific interest is the estimated regression coeffiecent estimand is the A finite population of $N=1000$ is simulated to solve a multiple linear regression problem, where dependent variable $Y$ is regressed on independent variables $X_1$, $X_2$ and $X_3$ (**check notation with betas here, and add what the quantity/-ies of scientific interest is/are!**):
+
+$$Y \sim \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3.$$
+
+The data generating model is a multivariate normal distribution
 

### 2020-04-16 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Editing the methods section to be conform recommendations Morris et al.

@@
 \textbf{Subsequent sampled values within each imputation chain are less
-alike. High \(AC\)-values are implausible in MI procedures. That is, the
-randomness induced by the MI algorithm effectively mitigates the risk of
-dependency within chains.}
+alike.}
 
@@
 formulated about the performance of \(AC\) as convergence diagnostic.
+\textbf{Added later: High \(AC\)-values are implausible in MI
+procedures. That is, the randomness induced by the MI algorithm
+effectively mitigates the risk of dependency within chains.}
 
@@
 
-The convergence of \texttt{mice} is investigated through model-based
+Convergence of the MICE algorithm is investigated through model-based
 simulation in \texttt{R} \citep[version 3.6.3;][]{R}. The simulation
@@
 # pseudo-code of simulation 
-simulate data 
+1. simulate data 
+for (missingness proportions 5%, 25%, 50%, 75% and 95%)
   for (number of simulation runs from 1 to 1000)
+    2. create missingness
     for (number of iterations from 1 to 100)
-    create missingness
-    impute the missingness
-    compute convergence diagnostics
-    perform analysis
-    pool results
-    compute simulation diagnostics
-aggregate convergence and simulation diagnostics
+      3. impute missingness
+      4. compute convergence diagnostics
+      5. perform analysis of scientific interest
+      6. pool results across imputations
+      7. compute performance measures
+  8. aggregate outcomes 
+9. combine outcomes
 \end{verbatim}
 
-\textbf{(Add step for different missingness proportions.)}
+The aim of the simulation study is to evaluate the impact of: 1)
+terminating the MICE algorithm at different imputation chain lengths
+(\(t = 1, 2, ..., 100\)), and 2) varying the difficulty of the
+missingness problem (\textbf{different missingness proportions}
+\(prop = .05, .25, .50, .75, .95\)). For the purpose of this study,
+sampling variance is not of interest. Therefore, a single complete set
+may serve as comparative truth in all simulation runs \citep{vinknd}.
 
-A finite population of \(N=1000\) is simulated to solve a multiple
-linear regression problem, where dependent variable \(Y\) is regressed
-on independent variables \(X_1\), \(X_2\) and \(X_3\) (\textbf{check
-notation with betas here, and add what the quantity/-ies of scientific
-interest is/are!}):
-
-\[Y \sim \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3.\]
-
-The data generating model is a multivariate normal distribution
+The data-generating mechanism is a multivariate normal distribution. We
+simulate a finite population of \(N=1000\) to represent person data for
+an unspecified social scientific field of study.
 
@@
 \epsilon
-\end{pmatrix} \sim  N
+\end{pmatrix} &\sim  N
 \begin{bmatrix}
@@
 \end{pmatrix}
-\end{bmatrix}\\[2\jot]
+\end{bmatrix}\!\!\text{, and}\\[2\jot]
+Y &=  1 + 2X_1 + .5X_2 - X_3 + \epsilon .
 \end{align*}
+
+A finite population of \(N=1000\) is simulated to solve a multiple
+linear regression problem, where dependent variable \(Y\) is regressed
+on independent variables \(X_1\), \(X_2\) and \(X_3\) (\textbf{check
+notation with betas here, and add what the quantity/-ies of scientific
+interest is/are!}):
+
+\[Y \sim \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3.\]
+
+The data generating model is a multivariate normal distribution
 
@@
 
-\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-3-1} \end{flushleft}
+\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-2-1} \end{flushleft}
 
-\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-3-2} \end{flushleft}
+\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-2-2} \end{flushleft}
 
-\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-3-3} \end{flushleft}
+\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-2-3} \end{flushleft}
 
@@
 
-\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-4-1} \end{flushleft}
+\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-3-1} \end{flushleft}
 
-\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-4-2} \end{flushleft}
+\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-3-2} \end{flushleft}
 
-\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-4-3} \end{flushleft}
+\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-3-3} \end{flushleft}
 

### 2020-04-17 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
Minor changes in methods section

@@
 Sub-questions: Are common MCMC non-convergence diagnostics appropriate for MICE?
-And if so, which threshold should be used?
-How many iterations are sufficient/needed to reach convergence?
+And if so, which threshold should be used to diagnose non-convergence?
+How many iterations are sufficient/needed to be able to diagnose non-convergence?
 Are the default number of iterations sufficient (i.e., 5 in mice, 10 in SPSS and Stata, 30 in mi)?
-What are the effects of continued iterations on estimates, predictions and inferences?
+**(These are part of the methods, not RQ per se: What are the effects of continued iterations on estimates, predictions and inferences?
 And finally, do the answers differ with varying missingness proportions?
+That is, we vary the nr of iterarions and the missingness proportion because we assume that the algorithm has not reached convergence at $t=1$, and performs worse with moderate to high missingness (not so much with little or a LOT of missingness).)**
 
@@
       7. compute performance measures
-  8. aggregate outcomes 
-9. combine outcomes
+  8. aggregate outcomes across simulation runs 
+9. combine outcomes of all missingness proportions
 ```
-The aim of the simulation study is to evaluate the impact of: 1) terminating the MICE algorithm at different imputation chain lengths ($t = 1, 2, ..., 100$), and 2) varying the difficulty of the missingness problem (**different missingness proportions** $prop = .05, .25, .50, .75, .95$).
-For the purpose of this study, sampling variance is not of interest.
-Therefore, a single complete set may serve as comparative truth in all simulation runs [@vinknd].
+The aim of the simulation study is to evaluate the impact of inducing non-convergence by: 1) terminating the MICE algorithm at different imputation chain lengths ($t = 1, 2, ..., 100$), and 2) varying the missingness proportions ($p_{miss} = .05, .25, .50, .75, .95$).
+The assumption underlying the different number of iterations is that the algorithm generally does reach convergence at $t=1$, because the initial values are sampled randomly from the set of observed datapoints.
+**As the number of iterations goes up, the imputation chains will become independent of the inital values until the point at which an extra added iteration does not lead to a more converged state.** The second set of experimental conditions under consideration--the missingness proportions--are chosen to reflect the difficulty of the missingness problem.
+The inherent assumption is that low missingness proportions will lead to quick algorithmic convergence, since there is a lot of information in the observed data.
+Higher missingness proportions then cause slower convergence.
+However, if the fraction of missing information is very high, there is so little information in the data that the random component in the algorithm will take the overhand and a stable but very uncertain (high variance) point will be reached.
 
-The data-generating mechanism is a multivariate normal distribution, representing person data on three independent variables (from an unspecified social scientific field of study).
-A finite population of $N=1000$ is simulated using the function `mvtnorm::rmvnorm()`, where
+The data-generating mechanism is a multivariate normal distribution, representing person data on three independent variables (from an unspecified social scientific field of study). Let
 
@@
 X_2\\
-X_3\\
-\epsilon
-\end{pmatrix} \sim  N
+X_3
+\end{pmatrix} \sim  \mathcal{N}
 \begin{bmatrix}
@@
 3\\
-0.5\\
-0
+0.5
 \end{pmatrix}\!\!,
@@
 4 & 16 & 4.8 & 0\\
-1.8 & 4.8 & 9 & 0\\
-0 & 0 & 0 & 100
+1.8 & 4.8 & 9 & 0
 \end{pmatrix}
-\end{bmatrix}\!\!\text{, and}\\[2\jot]
-Y &=  1 + 2X_1 + .5X_2 - X_3 + \epsilon .
+\end{bmatrix}\!\!\text{.}\\[2\jot]
 \end{align*} 
-Subsequently, 
-We consider four quantities of scientific interest ['conceptual estimands'; @morr19], namely  The quantity of scientific interest is the estimated regression coeffiecent estimand is the A finite population of $N=1000$ is simulated to solve a multiple linear regression problem, where dependent variable $Y$ is regressed on independent variables $X_1$, $X_2$ and $X_3$ (**check notation with betas here, and add what the quantity/-ies of scientific interest is/are!**):
+For the purpose of this study, sampling variance is not of interest.
+Therefore, a single complete set may serve as comparative truth in all simulation runs [@vinknd].
+A finite population of $N=1000$ is simulated using the function `mvtnorm::rmvnorm()`.
+Subsequently, a fourth variable is constructed to serve as dependent variable in a multiple linear regression problem.
+Let
+$$
+Y_i =  1 + 2X_{1i} + .5X_{2i} - X_{3i} + \epsilon_i ,
+$$
+where $i = 1, 2, ..., N$ and $\epsilon \sim \mathcal{N}(0, 100)$.
+
+We consider four quantities of scientific interest ['conceptual estimands'; @morr19], namely  ...
+solve a multiple linear regression problem, where dependent variable $Y$ is regressed on independent variables $X_1$, $X_2$ and $X_3$ (**check notation with betas here, and add what the quantity/-ies of scientific interest is/are!**):
 
 $$Y \sim \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3.$$
-
-The data generating model is a multivariate normal distribution
-
-Outcome variable $Y$ is subsequently calculated as 
-$$Y =  1 + 2X_1 + .5X_2 - X_3 + \epsilon .$$
 <!-- This note only considers a 'missing completely at random' (MCAR) mechanism, where the probability of being missing is equal for all $n \times p$ cells in $y$ [@rubin87].  -->
-
-**(Add the function `mvtnorm::rmvnorm()` back in somewhere here? If yes, use citation("mvtnorm").)**
 

### 2020-04-17 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Minor changes in methods section

@@
 common MCMC non-convergence diagnostics appropriate for MICE? And if so,
-which threshold should be used? How many iterations are
-sufficient/needed to reach convergence? Are the default number of
-iterations sufficient (i.e., 5 in mice, 10 in SPSS and Stata, 30 in mi)?
-What are the effects of continued iterations on estimates, predictions
-and inferences? And finally, do the answers differ with varying
-missingness proportions?
+which threshold should be used to diagnose non-convergence? How many
+iterations are sufficient/needed to be able to diagnose non-convergence?
+Are the default number of iterations sufficient (i.e., 5 in mice, 10 in
+SPSS and Stata, 30 in mi)? \textbf{(These are part of the methods, not
+RQ per se: What are the effects of continued iterations on estimates,
+predictions and inferences? And finally, do the answers differ with
+varying missingness proportions? That is, we vary the nr of iterarions
+and the missingness proportion because we assume that the algorithm has
+not reached convergence at \(t=1\), and performs worse with moderate to
+high missingness (not so much with little or a LOT of missingness).)}
 
@@
       7. compute performance measures
-  8. aggregate outcomes 
-9. combine outcomes
+  8. aggregate outcomes across simulation runs 
+9. combine outcomes of all missingness proportions
 \end{verbatim}
 
-The aim of the simulation study is to evaluate the impact of: 1)
-terminating the MICE algorithm at different imputation chain lengths
-(\(t = 1, 2, ..., 100\)), and 2) varying the difficulty of the
-missingness problem (\textbf{different missingness proportions}
-\(prop = .05, .25, .50, .75, .95\)). For the purpose of this study,
-sampling variance is not of interest. Therefore, a single complete set
-may serve as comparative truth in all simulation runs \citep{vinknd}.
+The aim of the simulation study is to evaluate the impact of inducing
+non-convergence by: 1) terminating the MICE algorithm at different
+imputation chain lengths (\(t = 1, 2, ..., 100\)), and 2) varying the
+missingness proportions (\(p_{miss} = .05, .25, .50, .75, .95\)). The
+assumption underlying the different number of iterations is that the
+algorithm generally does reach convergence at \(t=1\), because the
+initial values are sampled randomly from the set of observed datapoints.
+\textbf{As the number of iterations goes up, the imputation chains will
+become independent of the inital values until the point at which an
+extra added iteration does not lead to a more converged state.} The
+second set of experimental conditions under consideration--the
+missingness proportions--are chosen to reflect the difficulty of the
+missingness problem. The inherent assumption is that low missingness
+proportions will lead to quick algorithmic convergence, since there is a
+lot of information in the observed data. Higher missingness proportions
+then cause slower convergence. However, if the fraction of missing
+information is very high, there is so little information in the data
+that the random component in the algorithm will take the overhand and a
+stable but very uncertain (high variance) point will be reached.
 
-The data-generating mechanism is a multivariate normal distribution. We
-simulate a finite population of \(N=1000\) to represent person data for
-an unspecified social scientific field of study.
+The data-generating mechanism is a multivariate normal distribution,
+representing person data on three independent variables (from an
+unspecified social scientific field of study). Let
 
@@
 X_2\\
-X_3\\
-\epsilon
-\end{pmatrix} &\sim  N
+X_3
+\end{pmatrix} \sim  \mathcal{N}
 \begin{bmatrix}
@@
 4 & 16 & 4.8 & 0\\
-1.8 & 4.8 & 9 & 0\\
-0 & 0 & 0 & 100
+1.8 & 4.8 & 9 & 0
 \end{pmatrix}
-\end{bmatrix}\!\!\text{, and}\\[2\jot]
-Y &=  1 + 2X_1 + .5X_2 - X_3 + \epsilon .
-\end{align*}
+\end{bmatrix}\!\!\text{.}\\[2\jot]
+\end{align*} For the purpose of this study, sampling variance is not of
+interest. Therefore, a single complete set may serve as comparative
+truth in all simulation runs \citep{vinknd}. A finite population of
+\(N=1000\) is simulated using the function \texttt{mvtnorm::rmvnorm()}.
+Subsequently, a fourth variable is constructed to serve as dependent
+variable in a multiple linear regression problem. Let \[
+Y_i =  1 + 2X_{1i} + .5X_{2i} - X_{3i} + \epsilon_i ,
+\] where \(i = 1, 2, ..., N\) and \(\epsilon \sim \mathcal{N}(0, 100)\).
 
-A finite population of \(N=1000\) is simulated to solve a multiple
-linear regression problem, where dependent variable \(Y\) is regressed
-on independent variables \(X_1\), \(X_2\) and \(X_3\) (\textbf{check
+We consider four quantities of scientific interest \citep[`conceptual
+estimands';][]{morr19}, namely \ldots{} solve a multiple linear
+regression problem, where dependent variable \(Y\) is regressed on
+independent variables \(X_1\), \(X_2\) and \(X_3\) (\textbf{check
 notation with betas here, and add what the quantity/-ies of scientific
@@
 \[Y \sim \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3.\]
-
-The data generating model is a multivariate normal distribution
-
-Outcome variable \(Y\) is subsequently calculated as
-\[Y =  1 + 2X_1 + .5X_2 - X_3 + \epsilon .\]
-
-\textbf{(Add the function \texttt{mvtnorm::rmvnorm()} back in somewhere
-here? If yes, use citation(``mvtnorm'').)}
 

### 2020-04-20 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
removed for loops for efficiency, but did not fix pca convergence again.

@@
 email: h.i.oberman@uu.nl
-abstract: "(**Rewrite:**) Multiple imputation by chained equations (mice) is a widely used tool to accommodate missing data.
+abstract: "(**Rewrite:**) Multiple imputation by chained equations (MICE) is a widely used tool to accommodate missing data.
 While empirical evidence supports the validity of inferences obtained using mice, there is no consensus on the convergence properties of the method.
@@
 With imputation techniques, one or several completed datasets are created, on which statistical inferences can be performed.
-The case with several completed datasets is known as 'multiple imputation' [MI; @rubin76], and requires an additional step to pool the results across imputations.
+The case with several completed datasets is known as 'multiple imputation' [MI; @rubin76], and requires an additional step to pool the results across imputations **see figure 1??**.
 By applying Rubin's rules, the pooled result will reflect the uncertainty in the data due to missingness.
@@
 The collection of observed data values in $y$ is denoted by $y_{obs}$; the missing part of $y$ is referred to as $y_{mis}$.
-For each datapoint in $y_{mis}$, we sample $M \times T$ times plausible values, where $M$ is the number of imputations ($m = 1, \dots, M$) and $T$ is the number of iterations ($t = 1, \dots, T$).
+For each datapoint in $y_{mis}$, we sample $M \times T$ times plausible values, where $M$ is the number of imputations ($m = 1, 2, \dots, M$) and $T$ is the number of iterations ($t = 1, 2, \dots, T$).
 The collection of samples between the initial value (at $t=1$) and the final imputed value (at $t=T$) will be referred to as an 'imputation chain'.
-**Add: missingness percentage = proportion of cases with one or more missing values **
+**Add: missingness percentage = proportion of cases with one or more missing values, Q is quantity of scientific interest: "A scientific estimand $Q$ is a quantity of scientific interest that we can calculate if we would observe the entire population" [@buur18, par 2.3.1].
+**
 <!-- The number of imputations is denoted by $m$ (i.e., the number of times each datapoint in $y_{mis}$ is filled in).
@@
 
-From the between- and within-chain variances we compute a weighted average, $\widehat{\operatorname{var}}^{+}$, which over-estimates the total variance of $\theta$.
+From the between- and within-chain variances we compute a weighted average, $\widehat{\operatorname{var}}^{+}$, which over-estimates the total variance of $\theta$ **remove or explain why**.
 $\widehat{R}$ is then obtained as a ratio between the over-estimated total variance and the within-chain variance:
@@
 
-Based on an empirical finding [@lace07], we hypothesize that $\widehat{R}$ will over-estimate non-convergence of MI algorithms.
+Based on an empirical finding [@lace07], we hypothesize that $\widehat{R}$ will over-estimate non-convergence of MI algorithms **explain why?**.
 The threshold of $\widehat{R} < 1.01$ will then be too stringent for diagnosing convergence.
@@
 Therefore, we cannot be certain that the initial values are over-dispersed.
-We expect this to have little effect on the hypothesized performance of $\widehat{R}$.
+We expect this to have little effect on the hypothesized performance of $\widehat{R}$ **explain that the randomness induced by the MICE algorithm would take care of this??**.
 **(Add actual hypothesis: we thought that mice would converge sooner than $\widehat{R}$ would indicate that it did)** No hypothesis was formulated about the performance of $AC$ as convergence diagnostic.
@@
 
-Convergence of the MICE algorithm is investigated through model-based simulation in `R` [version `r paste(version$major,version$minor, sep = ".")`; @R].
+**Add to this section: 1) explicit mention of simulation conditions; 2) emphasize that the plots are averages across repetitions, not within MICE; 3) sample effects due to single complete dataset** Convergence of the MICE algorithm is investigated through model-based simulation in `R` [version `r paste(version$major,version$minor, sep = ".")`; @R].
 The simulation set-up is summarized in the pseudo-code below.

### 2020-04-20 3.Thesis/1.SimulationStudy/Functions/Autocorrelation_supplement.R

@@ -11,7 +11,7 @@ ac_lag1 <- function(x){
     # compute the correlation between the chain and itself, one iteration delayed
     ac[m] <- cor(x[-dim(x)[1], m], x[-1, m])
   }
-  return(max(ac))
+  return(ac)
 }
 
 # chain <- 1:4

### 2020-04-20 3.Thesis/1.SimulationStudy/Functions/Evaluate.R

@@ -6,30 +6,36 @@ evaluate.sim <-
            n.iter,
            mean_or_SE = "mean"
            ) {
-    # organize output for evaluation
-    dt <- map(sims, as.data.table) %>% rbindlist(fill = TRUE)
+    
     # determine whether we want averages per simulation condition or MCMC SEs
     if (mean_or_SE == "se") {
-      function_to_apply <- sd # or use: function(x) sqrt((var(x)))
+      function_to_apply <- function(x) sd(x, na.rm = TRUE) # or use: function(x) sqrt((var(x)))
     } else if (mean_or_SE == "lower") {
       function_to_apply <- function(x) quantile(x, .025, na.rm = TRUE)
     } else if (mean_or_SE == "upper") {
       function_to_apply <- function(x) quantile(x, .975, na.rm = TRUE)
-    } else {function_to_apply <- mean}
+    } else {function_to_apply <- function(x) mean(x, na.rm = TRUE)}
 
     
-    # apply function to aggregate simulation runs per simulation condition
-    d <- dt[, lapply(.SD, function_to_apply)] %>% as.data.frame
-    # create object for output
-    output <- 1:n.iter %>% as.data.frame()
-    # add names
-    vars <- names(sims[[1]][[1]])
-    for (var in vars) {
-      output <- cbind(output, d[, grep(var, names(d))] %>% t)
-    }
-    rownames(output) <- NULL
-    colnames(output) <- c("T", vars)
+    # organize output for evaluation
+    dt <- map(sims, as.data.table) %>% rbindlist(fill = TRUE) #%>% group_by(t,p)
+    # averages
+    res <- aggregate(.~t+p, dt, function_to_apply, na.action = na.pass) %>% dplyr::na_if(., "NaN")
+
+    # # apply function to aggregate simulation runs per simulation condition
+    # d <- dt[, lapply(.SD, function_to_apply)] %>% as.data.frame
+    
+    # # create object for output
+    # output <- 1:n.iter %>% as.data.frame()
+    
+    # # add names
+    # vars <- names(sims[[1]][[1]])
+    # for (var in vars) {
+    #   output <- cbind(output, d[, grep(var, names(d))] %>% t)
+    # }
+    # rownames(output) <- NULL
+    # colnames(output) <- c("T", vars)
     
     #output
-    return(output)
+    return(res)
   }

### 2020-04-21 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
New results section (draft)

@@
 
+Research question when and how to diagnose non convergence?
+
+QUESTION: How?
+
+The two challenges of convergence are stationarity and mixing point.
+Mixing and stationarity can be evaluated visually.
+Common practice is to evaluate stationarity and mixing in trace plots for scalar summaries of interest.
+In mice these scalar summaries of interest are chain means and chain variances.
+As Van Buren describes users can also specify a model specific scalar summary (e.g., a regression coefficient).
+search a user specified scalar summary however is not universal to all complete data problems.
+Therefore as inspired by someone 2002 we will use introduce a scalar summary that summarizes the multivariate state space of the algorithm.
+In this study we use the first eigenvalue of the variance covariance matrix of the completed data set per imputation.
+The methods under evaluation are Rhat and AC.
+These will be computed for the three scalar summaries of interest: chain means, chain variances, and the first eigenvalue of the variance-covariance matrix across imputations.
+
+The two methods will be evaluated with an estimand per scalar summary of interest.
+It is assumed that when the estimates are unbiased (and confidence valid), the algorithm is sufficiently converged.
+The R hat and AC values with chain means as theta are evaluated against the bias in univariate mean estimates (**just for Y??**).
+Similarly, the performance measure for Rhat and AC applied to chain variances is the bias in estimated standard deviations.
+To evaluate the performance of Rhat and AC on the eigenvalues, we will use bias in R squared and in the estimated regression coefficient, the coverage rate of the CI95% of the regression estimate, and the CI length.
+**remove the term performance measure above and add that performance measures are bias in all estimands, and cov and CIL of regression coeff.**
+
+ANSWER
+
+In general there is more bias in the estimate in conditions with a higher proportion of missingness and or a lower number of iterations.
+As required the values of or hat R lower for conditions with a higher number off iterations.
+End somewhat higher 4 conditions with a higher percentage of missing data.
+However autocorrelation did not show a decrease with increase number of iterations.
+After replacing the ACF function with a manual calculation for autocorrelation the autocorrelation values were indeed decreasing with a higher number of iterations.
+Evaluation with the performance measures shows that are had an autocorrelation are conservative: signs of non-convergence in conditions where estimands are unbiased and or confidence valid.
+As expected complete convergence (as indicated by autocorrelation equals 0; R hat equals 1)  is not observed.
+
+QUESTION: When?
+
+Usually, convergence is diagnosed when Rhat < 1.2 or 1.1 or 1.01. And a t-test is performed for AC values == 0. Performance measures are the same as above: unbiased, confidence valid estimates.
+
+ANSWER
+
+Our head is equal to 1.2 is not stringent enough because the performance measures R not unbiased and confidence fellas fell it 4 every condition in which our hat is smaller than one point 2.
+At least, not when our hat is smaller than one point two fur first occurs.
+The threshold R hat is equal to one point one is somewhat conservative with respect to the performance measures.
+It seems OK.
+Our hat is equal to 1.01 is too stringent compared to the performance measures (**but may be necessary in more complex complete data models???**).
+
+RECOMMENDATIONS
+
+For empirical researchers: 1) Check trace plots for pathological non convergence and adjust imputation model if necessary.
+2) use our hat wait 1.1 ash threshold and autocorrelation with **?** as threshold .
+Keep iterating until these thresholds are reached.
+3) Do not use the R function ACF.
+Instead compute autocorrelations manually , see for example GitHub link.
+4) Track your own scalar summary of interest.
+This is advanced but explained in Van Buren 2018.
+Compute R hat and autocorrelation values for this scalar summary.
+
+LIMITATIONS/FUTURE RESEARCH
+
+Pmm,
+M(N)AR,
+Empirical data,
+Vignette,
+ShinyMICE,
+AC across imputations, not iterations (unique for imp data??)
+
+If I don't include it in my study, also: Significance of AC values; Estimate (PCA or reg. coeff.) versus Rhat and AC on the estimate
+
+## Start methods section
+
 **Add to this section: 1) explicit mention of simulation conditions; 2) emphasize that the plots are averages across repetitions, not within MICE; 3) sample effects due to single complete dataset** Convergence of the MICE algorithm is investigated through model-based simulation in `R` [version `r paste(version$major,version$minor, sep = ".")`; @R].
@@
 ```{r, include = FALSE}
-load("../Results/test.Rdata")
-source("../2.CreateFigures.R")
+# load("../Results/test.Rdata")
+# source("../2.CreateFigures.R")
 ```
@@
 ```{r}
-mean_bias + plot_annotation(title = "Bias in mean estimate, convergence of chain means")
-mean_Rh 
-mean_ACF + plot_annotation(title = "Dotted lines are old AC computation, solid are new.")
+# mean_bias + plot_annotation(title = "Bias in mean estimate, convergence of chain means")
+# mean_Rh 
+# mean_ACF + plot_annotation(title = "Dotted lines are old AC computation, solid are new.")
 ```
@@
 ```{r}
-est_bias + plot_annotation(title = "Bias in regression estimate, coverage rate, \n bias in variance explained")
-est_cov
-Rsq_bias 
+# est_bias + plot_annotation(title = "Bias in regression estimate, coverage rate, \n bias in variance explained")
+# est_cov
+# Rsq_bias 
 ```
@@
 ```{r include=FALSE}
-# chain means
-dat[-1, 6:9] %>% mutate(max_Rh = apply(., 1, max)) %>% .$max_Rh<1.2
-
-# chain variances
-dat[-1, 26:29] %>% mutate(max_Rh = apply(., 1, max)) %>% .$max_Rh<1.2
+# # chain means
+# dat[-1, 6:9] %>% mutate(max_Rh = apply(., 1, max)) %>% .$max_Rh<1.2
+# 
+# # chain variances
+# dat[-1, 26:29] %>% mutate(max_Rh = apply(., 1, max)) %>% .$max_Rh<1.2
 

### 2020-04-21 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
New results section (draft)

@@
 \begin{abstract}
-(\textbf{Rewrite:}) Multiple imputation by chained equations (mice) is a
+(\textbf{Rewrite:}) Multiple imputation by chained equations (MICE) is a
 widely used tool to accommodate missing data. While empirical evidence
@@
 datasets is known as `multiple imputation' \citep[MI;][]{rubin76}, and
-requires an additional step to pool the results across imputations. By
-applying Rubin's rules, the pooled result will reflect the uncertainty
-in the data due to missingness. Under many circumstances, this yields an
-unbiased and confidence valid estimate of the inference on the true--but
-missing--data \citep{buur18}. A popular method to obtain imputations is
-to use the `Multiple Imputation by Chained Equations' algorithm,
-shorthand `MICE'\citep{mice}. MICE is an iterative algorithmic procedure
-to draw imputations from the posterior predictive distribution of the
-missing values. This introduces a potential threat to the validity of
-the imputations: what if the algorithm has not converged? Are the
-implications then to be trusted? And can we rely on the inference
-obtained on the completed data? These are all open questions, because
-the convergence properties of iterative imputation algorithms have not
-been systematically studied \citep{buur18}. Moreover, there is no
-scientific consensus on the convergence properties of MI algorithms
-\citep{taka17}. Some default MICE techniques (e.g., `predictive mean
-modeling') might not yield converged states at all \citep{murr18}.
-Therefore, algorithmic convergence should be monitored carefully.
+requires an additional step to pool the results across imputations
+\textbf{see figure 1??}. By applying Rubin's rules, the pooled result
+will reflect the uncertainty in the data due to missingness. Under many
+circumstances, this yields an unbiased and confidence valid estimate of
+the inference on the true--but missing--data \citep{buur18}. A popular
+method to obtain imputations is to use the `Multiple Imputation by
+Chained Equations' algorithm, shorthand `MICE'\citep{mice}. MICE is an
+iterative algorithmic procedure to draw imputations from the posterior
+predictive distribution of the missing values. This introduces a
+potential threat to the validity of the imputations: what if the
+algorithm has not converged? Are the implications then to be trusted?
+And can we rely on the inference obtained on the completed data? These
+are all open questions, because the convergence properties of iterative
+imputation algorithms have not been systematically studied
+\citep{buur18}. Moreover, there is no scientific consensus on the
+convergence properties of MI algorithms \citep{taka17}. Some default
+MICE techniques (e.g., `predictive mean modeling') might not yield
+converged states at all \citep{murr18}. Therefore, algorithmic
+convergence should be monitored carefully.
 
@@
 \(M \times T\) times plausible values, where \(M\) is the number of
-imputations (\(m = 1, \dots, M\)) and \(T\) is the number of iterations
-(\(t = 1, \dots, T\)). The collection of samples between the initial
-value (at \(t=1\)) and the final imputed value (at \(t=T\)) will be
-referred to as an `imputation chain'. \textbf{Add: missingness
-percentage = proportion of cases with one or more missing values }
+imputations (\(m = 1, 2, \dots, M\)) and \(T\) is the number of
+iterations (\(t = 1, 2, \dots, T\)). The collection of samples between
+the initial value (at \(t=1\)) and the final imputed value (at \(t=T\))
+will be referred to as an `imputation chain'. \textbf{Add: missingness
+percentage = proportion of cases with one or more missing values, Q is
+quantity of scientific interest: ``A scientific estimand \(Q\) is a
+quantity of scientific interest that we can calculate if we would
+observe the entire population'' \citep[par 2.3.1]{buur18}. }
 
@@
 average, \(\widehat{\operatorname{var}}^{+}\), which over-estimates the
-total variance of \(\theta\). \(\widehat{R}\) is then obtained as a
-ratio between the over-estimated total variance and the within-chain
-variance:
+total variance of \(\theta\) \textbf{remove or explain why}.
+\(\widehat{R}\) is then obtained as a ratio between the over-estimated
+total variance and the within-chain variance:
 
@@
 Based on an empirical finding \citep{lace07}, we hypothesize that
-\(\widehat{R}\) will over-estimate non-convergence of MI algorithms. The
-threshold of \(\widehat{R} < 1.01\) will then be too stringent for
-diagnosing convergence. This over-estimation may, however, be diminished
-because \(\widehat{R}\) can falsely diagnose convergence if initial
-values of the algorithm are not appropriately over-dispersed
-\citep[p.~437]{broo98}. In \texttt{mice}, initial values are chosen
-randomly from the observed data. Therefore, we cannot be certain that
-the initial values are over-dispersed. We expect this to have little
-effect on the hypothesized performance of \(\widehat{R}\). \textbf{(Add
-actual hypothesis: we thought that mice would converge sooner than
-\(\widehat{R}\) would indicate that it did)} No hypothesis was
-formulated about the performance of \(AC\) as convergence diagnostic.
-\textbf{Added later: High \(AC\)-values are implausible in MI
-procedures. That is, the randomness induced by the MI algorithm
-effectively mitigates the risk of dependency within chains.}
+\(\widehat{R}\) will over-estimate non-convergence of MI algorithms
+\textbf{explain why?}. The threshold of \(\widehat{R} < 1.01\) will then
+be too stringent for diagnosing convergence. This over-estimation may,
+however, be diminished because \(\widehat{R}\) can falsely diagnose
+convergence if initial values of the algorithm are not appropriately
+over-dispersed \citep[p.~437]{broo98}. In \texttt{mice}, initial values
+are chosen randomly from the observed data. Therefore, we cannot be
+certain that the initial values are over-dispersed. We expect this to
+have little effect on the hypothesized performance of \(\widehat{R}\)
+\textbf{explain that the randomness induced by the MICE algorithm would
+take care of this??}. \textbf{(Add actual hypothesis: we thought that
+mice would converge sooner than \(\widehat{R}\) would indicate that it
+did)} No hypothesis was formulated about the performance of \(AC\) as
+convergence diagnostic. \textbf{Added later: High \(AC\)-values are
+implausible in MI procedures. That is, the randomness induced by the MI
+algorithm effectively mitigates the risk of dependency within chains.}
 
@@
 
+Research question when and how to diagnose non convergence?
+
+QUESTION: How?
+
+The two challenges of convergence are stationarity and mixing point.
+Mixing and stationarity can be evaluated visually. Common practice is to
+evaluate stationarity and mixing in trace plots for scalar summaries of
+interest. In mice these scalar summaries of interest are chain means and
+chain variances. As Van Buren describes users can also specify a model
+specific scalar summary (e.g., a regression coefficient). search a user
+specified scalar summary however is not universal to all complete data
+problems. Therefore as inspired by someone 2002 we will use introduce a
+scalar summary that summarizes the multivariate state space of the
+algorithm. In this study we use the first eigenvalue of the variance
+covariance matrix of the completed data set per imputation. The methods
+under evaluation are Rhat and AC. These will be computed for the three
+scalar summaries of interest: chain means, chain variances, and the
+first eigenvalue of the variance-covariance matrix across imputations.
+
+The two methods will be evaluated with an estimand per scalar summary of
+interest. It is assumed that when the estimates are unbiased (and
+confidence valid), the algorithm is sufficiently converged. The R hat
+and AC values with chain means as theta are evaluated against the bias
+in univariate mean estimates (\textbf{just for Y??}). Similarly, the
+performance measure for Rhat and AC applied to chain variances is the
+bias in estimated standard deviations. To evaluate the performance of
+Rhat and AC on the eigenvalues, we will use bias in R squared and in the
+estimated regression coefficient, the coverage rate of the CI95\% of the
+regression estimate, and the CI length. \textbf{remove the term
+performance measure above and add that performance measures are bias in
+all estimands, and cov and CIL of regression coeff.}
+
+ANSWER
+
+In general there is more bias in the estimate in conditions with a
+higher proportion of missingness and or a lower number of iterations. As
+required the values of or hat R lower for conditions with a higher
+number off iterations. End somewhat higher 4 conditions with a higher
+percentage of missing data. However autocorrelation did not show a
+decrease with increase number of iterations. After replacing the ACF
+function with a manual calculation for autocorrelation the
+autocorrelation values were indeed decreasing with a higher number of
+iterations. Evaluation with the performance measures shows that are had
+an autocorrelation are conservative: signs of non-convergence in
+conditions where estimands are unbiased and or confidence valid. As
+expected complete convergence (as indicated by autocorrelation equals 0;
+R hat equals 1) is not observed.
+
+QUESTION: When?
+
+Usually, convergence is diagnosed when Rhat \textless{} 1.2 or 1.1 or
+1.01. And a t-test is performed for AC values == 0. Performance measures
+are the same as above: unbiased, confidence valid estimates.
+
+ANSWER
+
+Our head is equal to 1.2 is not stringent enough because the performance
+measures R not unbiased and confidence fellas fell it 4 every condition
+in which our hat is smaller than one point 2. At least, not when our hat
+is smaller than one point two fur first occurs. The threshold R hat is
+equal to one point one is somewhat conservative with respect to the
+performance measures. It seems OK. Our hat is equal to 1.01 is too
+stringent compared to the performance measures (\textbf{but may be
+necessary in more complex complete data models???}).
+
+RECOMMENDATIONS
+
+For empirical researchers: 1) Check trace plots for pathological non
+convergence and adjust imputation model if necessary. 2) use our hat
+wait 1.1 ash threshold and autocorrelation with \textbf{?} as threshold
+. Keep iterating until these thresholds are reached. 3) Do not use the R
+function ACF. Instead compute autocorrelations manually , see for
+example GitHub link. 4) Track your own scalar summary of interest. This
+is advanced but explained in Van Buren 2018. Compute R hat and
+autocorrelation values for this scalar summary.
+
+LIMITATIONS/FUTURE RESEARCH
+
+Pmm, M(N)AR, Empirical data, Vignette, ShinyMICE, AC across imputations,
+not iterations (unique for imp data??)
+
+If I don't include it in my study, also: Significance of AC values;
+Estimate (PCA or reg. coeff.) versus Rhat and AC on the estimate
+
+\hypertarget{start-methods-section}{%
+\subsection{Start methods section}\label{start-methods-section}}
+
+\textbf{Add to this section: 1) explicit mention of simulation
+conditions; 2) emphasize that the plots are averages across repetitions,
+not within MICE; 3) sample effects due to single complete dataset}
 Convergence of the MICE algorithm is investigated through model-based
@@
 
-\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-2-1} \end{flushleft}
-
-\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-2-2} \end{flushleft}
-
-\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-2-3} \end{flushleft}
-
 \hypertarget{univariate-estimates-and-convergence-diagnostics}{%
@@
 iterations is equal to three or more.
-
-\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-3-1} \end{flushleft}
-
-\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-3-2} \end{flushleft}
-
-\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-3-3} \end{flushleft}
 

### 2020-04-21 3.Thesis/1.SimulationStudy/1.Execute.R

@@ -31,7 +31,7 @@ source("3.Thesis/1.SimulationStudy/Functions/PCA_convergence.R")
 # simulation parameters
 populationsize <- 1000 #n of simulated dataset
 n.iter <- 100 #nr of iterations (varying 1:n.iter)
-n.sim <- 20 #nr of simulations per iteration value
+n.sim <- 100 #nr of simulations per iteration value
 p.miss <- c(.05, .25, .5, .75, .95)
 true_effect <- 2 #regression coefficient to be estimated
 true_mean <- true_sd <- NA

### 2020-04-22 3.Thesis/1.SimulationStudy/Writeup/diagram.Rnw
Add diagram of steps in MI back in

\documentclass{article}

%% recommended packages

%% more packages 
\usepackage{multirow, amsmath, tikz} % for table, equations, and flow chart figure (respectively)
\usetikzlibrary{fit, positioning} % for flow chart figure

\begin{document}

\begin{figure}[ht]
%\label{fig:steps2}
\centering
	\large{\begin{tikzpicture}
	\tikzstyle{main}=[circle, minimum size = 25mm, thick, draw =black!80, node distance = 20mm]
\tikzstyle{connect}=[-latex, thick]
\node[main, fill = white!100] (data) [label=below:Incomplete data] {$\{y_{obs}, y_{mis}\}$};
\node[main] (mids) [right=of data] { $\{y_{obs}, y_{imp}\}$};
\node[main] (mids2) [above=10mm of mids] {$\{y_{obs}, y_{imp}\}$ };
\node[main] (mids3) [below=10mm of mids,label=below: Completed data (per $m$)] { $\{y_{obs}, y_{imp}\}$};
\node[main] (mira) [right=of mids] {$\hat{Q}$};
\node[main] (mira2) [above=10mm of mira] {$\hat{Q}$ };
\node[main] (mira3) [below=10mm of mira,label=below: Results (per $m$)] {$\hat{Q}$ };
\node[main, fill = white!100] (mipo) [right=of mira,label=below:Pooled results] {$\bar{Q}$ };
\path (data) edge [connect] (mids)
      (data) edge [connect] (mids2)
      (data) edge [connect] (mids3)
      (mids) edge [connect] (mira)
      (mids2) edge [connect] (mira2)
      (mids3) edge [connect] (mira3)
		  (mira) edge [connect] (mipo)
		  (mira2) edge [connect] (mipo)
		  (mira3) edge [connect] (mipo);
\end{tikzpicture}}
%\caption{Scheme of the main steps in multiple imputation \citep[$m = 3$; adapted from][\S~1.4.1]{buur18}. {\footnotesize Missing data in dataset $Y$ is `imputed' (i.e., filled in) $m$ times. The imputed data is combined with the observed data ($Y_{obs}$) to create $m$ completed datasets. On each completed dataset, the analysis of scientific interest (or `substantive model') is performed. The quantity of scientific interest (e.g., a regression coefficient) is denoted with $Q$. Since $Q$ is estimated on each completed dataset, $m$ separate $\hat{Q}$-values are obtained. These $m$ values are combined into a single pooled estimate $\bar{Q}$. $\{y_{obs}, y_{mis}\}$ }} 
\end{figure}

\end{document}

### 2020-04-22 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
Add diagram of steps in MI back in

@@
 "The values for imputing the missing data are drawn from the posterior predictive distribution of the missing data conditional on the observed data.") -->
+
+```{r}
+# names <- c("A", "B", "C", "D", "E")
+# M <- matrix(nrow = 6, ncol = 6, byrow = FALSE, data = 0)
+# M[2, 3] <- M[3, 1] <- M[4, 2] <- M[4, 3] <- "flow"
+# plotmat(M, curve = 0, name = names, lwd = 1, box.lwd = 2, cex.txt = 0.8, box.type = "circle", box.prop = 1.0)
+# 
+# M <- matrix(nrow = 6, ncol = 6, byrow = TRUE, data = c(
+# # p n z d b f
+# 0, 1, 0, 0, 0, 0, #p
+# 0, 0, 4, 10, 11, 0, #n
+# 2, 0, 0, 0, 0, 0, #z
+# 8, 0, 13, 0, 0, 12, #d
+# 9, 0, 0, 7, 0, 0, #b
+# 0, 0, 5, 0, 0, 0 #f
+# ))
+# 
+# 
+# plotmat(M, pos = c(1, 2, 1, 2), curve = 0, name = names,
+# lwd = 1, box.lwd = 2, cex.txt = 0.8,
+# box.type = "square", box.prop = 0.5, arr.type = "triangle",
+# arr.pos = 0.4, shadow.size = 0.01, prefix = "f",
+# main = "NPZZDD model")
+# #
+# 
+# library(diagram)
+# data <- c(0, "'.47*'", 0,
+#           0, 0, 0, 
+#           "'.36*'", "'.33*'", 0)
+# M<- matrix (nrow=3, ncol=4, byrow = TRUE, data=0)
+# plot<- plotmat (M, 
+#                 name= c( "Math self-efficacy","Math ability", "Interest in the math major"), 
+#                 box.type = "rect", box.size = 0.12, box.prop=0.5,  curve=0)
+# 
+# library(semPlot)
+# # A silly dataset:
+# A <- B <- C <- D <- E <- F <- rnorm(10)
+# DF <- data.frame(A, B, C, D, E, F)
+# 
+# # Two regressions:
+# res1 <- lm(C ~ B, data = DF)
+# res11 <- lm(D ~ B, data = DF)
+# res12 <- lm(E ~ B, data = DF)
+# res2 <- lm(A ~ B + C, data = DF)
+# res3 <- lm(A ~ D + E + F, data = DF)
+# 
+# # Plot both in the same path diagram in two ways:
+# semPaths(res1 + res2 + res3 + res11 + res12, "model", "est", intercepts = FALSE)
+# semPaths(list(res1, res2), "model", "est", intercepts = FALSE)
+```
 

### 2020-04-22 3.Thesis/1.SimulationStudy/Writeup/images/diagram.tex
Add example of non-convergence

--- 2020-04-22 3.Thesis/1.SimulationStudy/Writeup/diagram.Rnw
@@
-\documentclass{article}
-
+\documentclass{standalone}
 %% recommended packages
@@
 %% more packages 
-\usepackage{multirow, amsmath, tikz} % for table, equations, and flow chart figure (respectively)
-\usetikzlibrary{fit, positioning} % for flow chart figure
+\usepackage{amsmath, tikz} % for table, equations, and flow chart figure (respectively)
+\usetikzlibrary{fit, positioning, shapes} % for flow chart figure
 
+\usepackage{Sweave}
 \begin{document}
 
-\begin{figure}[ht]
-%\label{fig:steps2}
-\centering
-	\large{\begin{tikzpicture}
-	\tikzstyle{main}=[circle, minimum size = 25mm, thick, draw =black!80, node distance = 20mm]
-\tikzstyle{connect}=[-latex, thick]
+\begin{tikzpicture}
+\tikzstyle{main}=[rectangle, minimum width = 21mm, minimum height = 12mm, thick, draw =black!80, node distance = 17mm]
+\tikzstyle{connect}=[draw, -latex, thick]
 \node[main, fill = white!100] (data) [label=below:Incomplete data] {$\{y_{obs}, y_{mis}\}$};
@@
 \node[main] (mids3) [below=10mm of mids,label=below: Completed data (per $m$)] { $\{y_{obs}, y_{imp}\}$};
-\node[main] (mira) [right=of mids] {$\hat{Q}$};
-\node[main] (mira2) [above=10mm of mira] {$\hat{Q}$ };
-\node[main] (mira3) [below=10mm of mira,label=below: Results (per $m$)] {$\hat{Q}$ };
-\node[main, fill = white!100] (mipo) [right=of mira,label=below:Pooled results] {$\bar{Q}$ };
+\node[main] (mira) [right=of mids] {\large $\hat{Q}$};
+\node[main] (mira2) [above=10mm of mira] {\large $\hat{Q}$ };
+\node[main] (mira3) [below=10mm of mira,label=below: Results (per $m$)] {\large $\hat{Q}$ };
+\node[main, fill = white!100] (mipo) [right=of mira,label=below:Pooled results] {\large $\bar{Q}$ };
 \path (data) edge [connect] (mids)
@@
 		  (mira3) edge [connect] (mipo);
-\end{tikzpicture}}
+\end{tikzpicture}
 %\caption{Scheme of the main steps in multiple imputation \citep[$m = 3$; adapted from][\S~1.4.1]{buur18}.
@@
 $\{y_{obs}, y_{mis}\}$ }}
-\end{figure}
 

### 2020-04-22 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
Add example of non-convergence

--- 2020-04-21 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(fig.width = 4.8, fig.height = 3.6, echo = FALSE, fig.align = "left") 
+knitr::opts_chunk$set(fig.width = 4.8, fig.height = 3.6, echo = FALSE, fig.align = "center") 
 ```
@@
 The case with several completed datasets is known as 'multiple imputation' [MI; @rubin76], and requires an additional step to pool the results across imputations **see figure 1??**.
+
+**Move this after notation??** Figure 1 provides an overview of the steps involved with MI---from incomplete data, to $M$ completed datasets, to $M$ estimated quantities of interest ($\hat{Q}$s), to a single pooled estimate $\bar{Q}$.
+Missing data in $y$ is 'imputed' (i.e., filled in) $M$ times.
+The imputed data ($y_{imp}$) is combined with the observed data ($y_{obs}$) to create $M$ completed datasets.
+On each completed dataset, the analysis of scientific interest (the 'substantive model', or 'complete data model') is performed.
+The quantity of scientific interest (e.g., a regression coefficient) is denoted with $Q$.
+Since $Q$ is estimated on each completed dataset, $M$ separate $\hat{Q}$-values are obtained.
+These $M$ values are combined into a single pooled estimate $\bar{Q}$.
+**ALSO: make clear that iteration happens within the first step.**
+
+```{r, out.width="\\linewidth", include=TRUE, fig.align="center", fig.cap="Scheme of the main steps in multiple imputation."}
+knitr::include_graphics("./images/diagram.pdf") #adapted from [@buur18] par. 1.4.1.
+```
+
 By applying Rubin's rules, the pooled result will reflect the uncertainty in the data due to missingness.
@@
 And finally, do the answers differ with varying missingness proportions?
-That is, we vary the nr of iterarions and the missingness proportion because we assume that the algorithm has not reached convergence at $t=1$, and performs worse with moderate to high missingness (not so much with little or a LOT of missingness).)**
+That is, we vary the nr of iterations and the missingness proportion because we assume that the algorithm has not reached convergence at $t=1$, and performs worse with moderate to high missingness (not so much with little or a LOT of missingness).)**
 
@@
 
-The two challenges for convergence of iterative algorithms are mixing and stationarity [@gelm13], see Figure 1.
+The two challenges for convergence of iterative algorithms are mixing and stationarity [@gelm13], see Figure XYZ.
 Without mixing, imputation chains do not intermingle nicely.
@@
 898], $AC$ and $\widehat{R}$ will be used.
+
+```{r, out.width='.49\\linewidth', fig.width=3.5, fig.height=3,fig.show='hold',fig.cap="Pathological non-convergence versus typical convergence."}
+load("./images/nonc.Rdata")
+load("./images/conv.Rdata")
+par(mar = c(2, 2, 0.1, 0.1))
+plot(nonc)
+plot(conv)
+
+```
 
@@
 In this study we use the first eigenvalue of the variance covariance matrix of the completed data set per imputation.
-The methods under evaluation are Rhat and AC.
+The methods under evaluation are $\widehat{R}$ and $AC$.
 These will be computed for the three scalar summaries of interest: chain means, chain variances, and the first eigenvalue of the variance-covariance matrix across imputations.
@@
 It is assumed that when the estimates are unbiased (and confidence valid), the algorithm is sufficiently converged.
-The R hat and AC values with chain means as theta are evaluated against the bias in univariate mean estimates (**just for Y??**).
-Similarly, the performance measure for Rhat and AC applied to chain variances is the bias in estimated standard deviations.
-To evaluate the performance of Rhat and AC on the eigenvalues, we will use bias in R squared and in the estimated regression coefficient, the coverage rate of the CI95% of the regression estimate, and the CI length.
+The $\widehat{R}$ and $AC$ values with chain means as theta are evaluated against the bias in univariate mean estimates (**just for Y??**).
+Similarly, the performance measure for Rhat and $AC$ applied to chain variances is the bias in estimated standard deviations.
+To evaluate the performance of Rhat and $AC$ on the eigenvalues, we will use bias in R squared and in the estimated regression coefficient, the coverage rate of the CI95% of the regression estimate, and the CI length.
 **remove the term performance measure above and add that performance measures are bias in all estimands, and cov and CIL of regression coeff.**
@@
 In general there is more bias in the estimate in conditions with a higher proportion of missingness and or a lower number of iterations.
-As required the values of or hat R lower for conditions with a higher number off iterations.
+As required the values of o$\widehat{R}$ R lower for conditions with a higher number off iterations.
 End somewhat higher 4 conditions with a higher percentage of missing data.
@@
 Evaluation with the performance measures shows that are had an autocorrelation are conservative: signs of non-convergence in conditions where estimands are unbiased and or confidence valid.
-As expected complete convergence (as indicated by autocorrelation equals 0; R hat equals 1)  is not observed.
+As expected complete convergence (as indicated by autocorrelation equals 0; $\widehat{R}$ equals 1)  is not observed.
 
@@
 
-Usually, convergence is diagnosed when Rhat < 1.2 or 1.1 or 1.01. And a t-test is performed for AC values == 0. Performance measures are the same as above: unbiased, confidence valid estimates.
+Usually, convergence is diagnosed when Rhat < 1.2 or 1.1 or 1.01. And a t-test is performed for $AC$ values == 0. Performance measures are the same as above: unbiased, confidence valid estimates.
 
@@
 
-Our head is equal to 1.2 is not stringent enough because the performance measures R not unbiased and confidence fellas fell it 4 every condition in which our hat is smaller than one point 2.
-At least, not when our hat is smaller than one point two fur first occurs.
-The threshold R hat is equal to one point one is somewhat conservative with respect to the performance measures.
+Our head is equal to 1.2 is not stringent enough because the performance measures R not unbiased and confidence fellas fell it 4 every condition in which $\widehat{R}$ is smaller than one point 2.
+At least, not when $\widehat{R}$ is smaller than one point two fur first occurs.
+The threshold $\widehat{R}$ is equal to one point one is somewhat conservative with respect to the performance measures.
 It seems OK.
-Our hat is equal to 1.01 is too stringent compared to the performance measures (**but may be necessary in more complex complete data models???**).
+$\widehat{R}$ is equal to 1.01 is too stringent compared to the performance measures (**but may be necessary in more complex complete data models???**).
 
@@
 For empirical researchers: 1) Check trace plots for pathological non convergence and adjust imputation model if necessary.
-2) use our hat wait 1.1 ash threshold and autocorrelation with **?** as threshold .
+2) use $\widehat{R}$ wait 1.1 ash threshold and autocorrelation with **?** as threshold .
 Keep iterating until these thresholds are reached.
@@
 This is advanced but explained in Van Buren 2018.
-Compute R hat and autocorrelation values for this scalar summary.
+Compute $\widehat{R}$ and autocorrelation values for this scalar summary.
 
@@
 
-If I don't include it in my study, also: Significance of AC values; Estimate (PCA or reg. coeff.) versus Rhat and AC on the estimate
+If I don't include it in my study, also: Significance of $AC$ values; Estimate (PCA or reg. coeff.) versus Rhat and $AC$ on the estimate
 
@@
 The assumption underlying the different number of iterations is that the algorithm generally does reach convergence at $t=1$, because the initial values are sampled randomly from the set of observed datapoints.
-**As the number of iterations goes up, the imputation chains will become independent of the inital values until the point at which an extra added iteration does not lead to a more converged state.** The second set of experimental conditions under consideration--the missingness proportions--are chosen to reflect the difficulty of the missingness problem.
+**As the number of iterations goes up, the imputation chains will become independent of the initial values until the point at which an extra added iteration does not lead to a more converged state.** The second set of experimental conditions under consideration--the missingness proportions--are chosen to reflect the difficulty of the missingness problem.
 The inherent assumption is that low missingness proportions will lead to quick algorithmic convergence, since there is a lot of information in the observed data.
@@
 $AC$ (**of what?
-and how is AC aggregated across imputations?
+and how is $AC$ aggregated across imputations?
 i.e., mean in latest sim**) is computed with function `stats::acf()`.
@@
 We obtain an estimated regression coefficient per imputation, which are pooled into a single estimate, $\bar{Q}$.
-We use the function `mice::pool()` to estimate u bar (**check wat dat nou precies is!!**) according to Rubin's [-@rubin87] rules, and subsequently implement finite population pooling conform @vink14.
+We use the function `mice::pool()` to estimate u bar (**check definition!**) according to Rubin's [-@rubin87] rules, and subsequently implement finite population pooling conform @vink14.
 
@@
 # mean_Rh 
-# mean_ACF + plot_annotation(title = "Dotted lines are old AC computation, solid are new.")
+# mean_ACF + plot_annotation(title = "Dotted lines are old $AC$ computation, solid are new.")
 ```
@@
 <!-- Rhat values are indistinguishable between simulation conditions when $T>30$.  -->
-<!-- The R hat values seemed to flatten off between 20 and 30 iterations.  -->
+<!-- The $\widehat{R}$ values seemed to flatten off between 20 and 30 iterations.  -->
 
 Auto-correlations indicate no sign of trending within imputation chains.
-In most simulation conditions AC is smaller than or about equal to zero.
+In most simulation conditions $AC$ is smaller than or about equal to zero.
 Across simulation conditions, however, the autocorrelation curve does not trend towards 0.
@@
 <!-- For 25% miss, no bias in estimated means, Rh flat from 30 it.
-(but <1.1 for it>11), AC never high but stable after 40 it.
-Bias in estimated variances only if it=1 or maybe 2?, while Rh of chain variances is flat around it 25 (smaller than 1.1 for it>9), and AC never positive, flat around it 30.
+(but <1.1 for it>11), $AC$ never high but stable after 40 it.
+Bias in estimated variances only if it=1 or maybe 2?, while Rh of chain variances is flat around it 25 (smaller than 1.1 for it>9), and $AC$ never positive, flat around it 30.
 Bias in estimated reg.
@@
 
-<!-- Taken together, we see that $T>3$ is the minimal requirement to diagnose convergence ($\widehat{R} < 1.2; AC \leq 0$).
+<!-- Taken together, we see that $T>3$ is the minimal requirement to diagnose convergence ($\widehat{R} < 1.2; $AC$ \leq 0$).
 This threshold is, however, not sufficient, since we overlook the increase in $\widehat{R}$-values up-to conditions where $T>5$, and the convergence diagnostics only reach stability at $T>20$.
@@
 However, initial dip in $AC$-values may have implications for the default number of iterations in `mice` (`maxit = 5`).
-Terminating the algorithm at $T=5$ may not be the most appropriate, since this lead to the worst convergence (**nope, only for Rh, not AC**), as indicated by $\widehat{R}$ and $AC$.
+Terminating the algorithm at $T=5$ may not be the most appropriate, since this lead to the worst convergence (**nope, only for Rh, not $AC$**), as indicated by $\widehat{R}$ and $AC$.
 Under the current specifications, $T>20$ would be more appropriate.
 
-The observed dip in AC implies that default maxit value of five iterations is the worst possible number of iterations.
+The observed dip in $AC$ implies that default maxit value of five iterations is the worst possible number of iterations.
 Moreover, the results of this study imply that assessing the stationarity component of convergence with $AC$ might be redundant.

### 2020-04-22 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Add example of non-convergence

@@
 requires an additional step to pool the results across imputations
-\textbf{see figure 1??}. By applying Rubin's rules, the pooled result
-will reflect the uncertainty in the data due to missingness. Under many
-circumstances, this yields an unbiased and confidence valid estimate of
-the inference on the true--but missing--data \citep{buur18}. A popular
-method to obtain imputations is to use the `Multiple Imputation by
-Chained Equations' algorithm, shorthand `MICE'\citep{mice}. MICE is an
-iterative algorithmic procedure to draw imputations from the posterior
-predictive distribution of the missing values. This introduces a
-potential threat to the validity of the imputations: what if the
-algorithm has not converged? Are the implications then to be trusted?
-And can we rely on the inference obtained on the completed data? These
-are all open questions, because the convergence properties of iterative
-imputation algorithms have not been systematically studied
-\citep{buur18}. Moreover, there is no scientific consensus on the
-convergence properties of MI algorithms \citep{taka17}. Some default
-MICE techniques (e.g., `predictive mean modeling') might not yield
-converged states at all \citep{murr18}. Therefore, algorithmic
-convergence should be monitored carefully.
+\textbf{see figure 1??}.
+
+\textbf{Move this after notation??} Figure 1 provides an overview of the
+steps involved with MI---from incomplete data, to \(M\) completed
+datasets, to \(M\) estimated quantities of interest (\(\hat{Q}\)s), to a
+single pooled estimate \(\bar{Q}\). Missing data in \(y\) is `imputed'
+(i.e., filled in) \(M\) times. The imputed data (\(y_{imp}\)) is
+combined with the observed data (\(y_{obs}\)) to create \(M\) completed
+datasets. On each completed dataset, the analysis of scientific interest
+(the `substantive model', or `complete data model') is performed. The
+quantity of scientific interest (e.g., a regression coefficient) is
+denoted with \(Q\). Since \(Q\) is estimated on each completed dataset,
+\(M\) separate \(\hat{Q}\)-values are obtained. These \(M\) values are
+combined into a single pooled estimate \(\bar{Q}\). \textbf{ALSO: make
+clear that iteration happens within the first step.}
+
+\begin{figure}
+
+{\centering \includegraphics[width=\linewidth]{./images/diagram}
+
+}
+
+\caption{Scheme of the main steps in multiple imputation.}\label{fig:unnamed-chunk-1}
+\end{figure}
+
+By applying Rubin's rules, the pooled result will reflect the
+uncertainty in the data due to missingness. Under many circumstances,
+this yields an unbiased and confidence valid estimate of the inference
+on the true--but missing--data \citep{buur18}. A popular method to
+obtain imputations is to use the `Multiple Imputation by Chained
+Equations' algorithm, shorthand `MICE'\citep{mice}. MICE is an iterative
+algorithmic procedure to draw imputations from the posterior predictive
+distribution of the missing values. This introduces a potential threat
+to the validity of the imputations: what if the algorithm has not
+converged? Are the implications then to be trusted? And can we rely on
+the inference obtained on the completed data? These are all open
+questions, because the convergence properties of iterative imputation
+algorithms have not been systematically studied \citep{buur18}.
+Moreover, there is no scientific consensus on the convergence properties
+of MI algorithms \citep{taka17}. Some default MICE techniques (e.g.,
+`predictive mean modeling') might not yield converged states at all
+\citep{murr18}. Therefore, algorithmic convergence should be monitored
+carefully.
 
@@
 predictions and inferences? And finally, do the answers differ with
-varying missingness proportions? That is, we vary the nr of iterarions
+varying missingness proportions? That is, we vary the nr of iterations
 and the missingness proportion because we assume that the algorithm has
@@
 The two challenges for convergence of iterative algorithms are mixing
-and stationarity \citep{gelm13}, see Figure 1. Without mixing,
+and stationarity \citep{gelm13}, see Figure XYZ. Without mixing,
 imputation chains do not intermingle nicely. Without stationarity, there
@@
 \(AC\) and \(\widehat{R}\) will be used.
+
+\begin{figure}
+
+{\centering \includegraphics[width=.49\linewidth]{manuscript_files/figure-latex/unnamed-chunk-2-1} \includegraphics[width=.49\linewidth]{manuscript_files/figure-latex/unnamed-chunk-2-2}
+
+}
+
+\caption{Pathological non-convergence versus typical convergence.}\label{fig:unnamed-chunk-2}
+\end{figure}
 
@@
 covariance matrix of the completed data set per imputation. The methods
-under evaluation are Rhat and AC. These will be computed for the three
-scalar summaries of interest: chain means, chain variances, and the
-first eigenvalue of the variance-covariance matrix across imputations.
+under evaluation are \(\widehat{R}\) and \(AC\). These will be computed
+for the three scalar summaries of interest: chain means, chain
+variances, and the first eigenvalue of the variance-covariance matrix
+across imputations.
 
@@
 interest. It is assumed that when the estimates are unbiased (and
-confidence valid), the algorithm is sufficiently converged. The R hat
-and AC values with chain means as theta are evaluated against the bias
-in univariate mean estimates (\textbf{just for Y??}). Similarly, the
-performance measure for Rhat and AC applied to chain variances is the
-bias in estimated standard deviations. To evaluate the performance of
-Rhat and AC on the eigenvalues, we will use bias in R squared and in the
-estimated regression coefficient, the coverage rate of the CI95\% of the
-regression estimate, and the CI length. \textbf{remove the term
-performance measure above and add that performance measures are bias in
-all estimands, and cov and CIL of regression coeff.}
+confidence valid), the algorithm is sufficiently converged. The
+\(\widehat{R}\) and \(AC\) values with chain means as theta are
+evaluated against the bias in univariate mean estimates (\textbf{just
+for Y??}). Similarly, the performance measure for Rhat and \(AC\)
+applied to chain variances is the bias in estimated standard deviations.
+To evaluate the performance of Rhat and \(AC\) on the eigenvalues, we
+will use bias in R squared and in the estimated regression coefficient,
+the coverage rate of the CI95\% of the regression estimate, and the CI
+length. \textbf{remove the term performance measure above and add that
+performance measures are bias in all estimands, and cov and CIL of
+regression coeff.}
 
@@
 higher proportion of missingness and or a lower number of iterations. As
-required the values of or hat R lower for conditions with a higher
-number off iterations. End somewhat higher 4 conditions with a higher
-percentage of missing data. However autocorrelation did not show a
-decrease with increase number of iterations. After replacing the ACF
+required the values of o\(\widehat{R}\) R lower for conditions with a
+higher number off iterations. End somewhat higher 4 conditions with a
+higher percentage of missing data. However autocorrelation did not show
+a decrease with increase number of iterations. After replacing the ACF
 function with a manual calculation for autocorrelation the
@@
 expected complete convergence (as indicated by autocorrelation equals 0;
-R hat equals 1) is not observed.
+\(\widehat{R}\) equals 1) is not observed.
 
@@
 Usually, convergence is diagnosed when Rhat \textless{} 1.2 or 1.1 or
-1.01. And a t-test is performed for AC values == 0. Performance measures
-are the same as above: unbiased, confidence valid estimates.
+1.01. And a t-test is performed for \(AC\) values == 0. Performance
+measures are the same as above: unbiased, confidence valid estimates.
 
@@
 measures R not unbiased and confidence fellas fell it 4 every condition
-in which our hat is smaller than one point 2. At least, not when our hat
-is smaller than one point two fur first occurs. The threshold R hat is
-equal to one point one is somewhat conservative with respect to the
-performance measures. It seems OK. Our hat is equal to 1.01 is too
-stringent compared to the performance measures (\textbf{but may be
-necessary in more complex complete data models???}).
+in which \(\widehat{R}\) is smaller than one point 2. At least, not when
+\(\widehat{R}\) is smaller than one point two fur first occurs. The
+threshold \(\widehat{R}\) is equal to one point one is somewhat
+conservative with respect to the performance measures. It seems OK.
+\(\widehat{R}\) is equal to 1.01 is too stringent compared to the
+performance measures (\textbf{but may be necessary in more complex
+complete data models???}).
 
@@
 For empirical researchers: 1) Check trace plots for pathological non
-convergence and adjust imputation model if necessary. 2) use our hat
-wait 1.1 ash threshold and autocorrelation with \textbf{?} as threshold
-. Keep iterating until these thresholds are reached. 3) Do not use the R
-function ACF. Instead compute autocorrelations manually , see for
-example GitHub link. 4) Track your own scalar summary of interest. This
-is advanced but explained in Van Buren 2018. Compute R hat and
-autocorrelation values for this scalar summary.
+convergence and adjust imputation model if necessary. 2) use
+\(\widehat{R}\) wait 1.1 ash threshold and autocorrelation with
+\textbf{?} as threshold . Keep iterating until these thresholds are
+reached. 3) Do not use the R function ACF. Instead compute
+autocorrelations manually , see for example GitHub link. 4) Track your
+own scalar summary of interest. This is advanced but explained in Van
+Buren 2018. Compute \(\widehat{R}\) and autocorrelation values for this
+scalar summary.
 
@@
 
-If I don't include it in my study, also: Significance of AC values;
-Estimate (PCA or reg. coeff.) versus Rhat and AC on the estimate
+If I don't include it in my study, also: Significance of \(AC\) values;
+Estimate (PCA or reg. coeff.) versus Rhat and \(AC\) on the estimate
 
@@
 \textbf{As the number of iterations goes up, the imputation chains will
-become independent of the inital values until the point at which an
+become independent of the initial values until the point at which an
 extra added iteration does not lead to a more converged state.} The
@@
 \citeyearpar{veht19} recommendations. \(AC\) (\textbf{of what? and how
-is AC aggregated across imputations? i.e., mean in latest sim}) is
+is \(AC\) aggregated across imputations? i.e., mean in latest sim}) is
 computed with function \texttt{stats::acf()}.
@@
 imputation, which are pooled into a single estimate, \(\bar{Q}\). We use
-the function \texttt{mice::pool()} to estimate u bar (\textbf{check wat
-dat nou precies is!!}) according to Rubin's \citeyearpar{rubin87} rules,
-and subsequently implement finite population pooling conform
-\citet{vink14}.
+the function \texttt{mice::pool()} to estimate u bar (\textbf{check
+definition!}) according to Rubin's \citeyearpar{rubin87} rules, and
+subsequently implement finite population pooling conform \citet{vink14}.
 
@@
 Auto-correlations indicate no sign of trending within imputation chains.
-In most simulation conditions AC is smaller than or about equal to zero.
-Across simulation conditions, however, the autocorrelation curve does
-not trend towards 0. Autocorrelation values plateau off at a value of
-around .1 . This is a small positive autocorrelation, which would
+In most simulation conditions \(AC\) is smaller than or about equal to
+zero. Across simulation conditions, however, the autocorrelation curve
+does not trend towards 0. Autocorrelation values plateau off at a value
+of around .1 . This is a small positive autocorrelation, which would
 indicate some trending within chains.
@@
 the most appropriate, since this lead to the worst convergence
-(\textbf{nope, only for Rh, not AC}), as indicated by \(\widehat{R}\)
-and \(AC\). Under the current specifications, \(T>20\) would be more
-appropriate.
+(\textbf{nope, only for Rh, not \(AC\)}), as indicated by
+\(\widehat{R}\) and \(AC\). Under the current specifications, \(T>20\)
+would be more appropriate.
 
-The observed dip in AC implies that default maxit value of five
+The observed dip in \(AC\) implies that default maxit value of five
 iterations is the worst possible number of iterations. Moreover, the

### 2020-04-22 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
Compute rhat and ac for pathological non-convergence case. Adjust functions by adding wrapper for general convergence function

@@
 
-**Move this after notation??** Figure 1 provides an overview of the steps involved with MI---from incomplete data, to $M$ completed datasets, to $M$ estimated quantities of interest ($\hat{Q}$s), to a single pooled estimate $\bar{Q}$.
+**Move this after notation??** Figure \ref{fig:diagram} provides an overview of the steps involved with MI---from incomplete data, to $M$ completed datasets, to $M$ estimated quantities of interest ($\hat{Q}$s), to a single pooled estimate $\bar{Q}$.
 Missing data in $y$ is 'imputed' (i.e., filled in) $M$ times.
@@
 
-```{r, out.width="\\linewidth", include=TRUE, fig.align="center", fig.cap="Scheme of the main steps in multiple imputation."}
+```{r diagram, out.width="\\linewidth", include=TRUE, fig.align="center", fig.cap="Scheme of the main steps in multiple imputation."}
 knitr::include_graphics("./images/diagram.pdf") #adapted from [@buur18] par. 1.4.1.
@@
 
-Main question: How can non-convergence be diagnosed?
-Sub-questions: Are common MCMC non-convergence diagnostics appropriate for MICE?
+### Research question: how and when to diagnose non convergence?
+
+Sub-questions: How can non-convergence be diagnosed?
+Are common MCMC non-convergence diagnostics appropriate for MICE?
 And if so, which threshold should be used to diagnose non-convergence?
@@
 
-The two challenges for convergence of iterative algorithms are mixing and stationarity [@gelm13], see Figure XYZ.
+The two challenges for convergence of iterative algorithms are mixing and stationarity [@gelm13], see Figure \ref{fig:non-conv}.
 Without mixing, imputation chains do not intermingle nicely.
@@
 
-```{r, out.width='.49\\linewidth', fig.width=3.5, fig.height=3,fig.show='hold',fig.cap="Pathological non-convergence versus typical convergence."}
+```{r non-conv, out.width='.49\\linewidth', fig.width=3.5, fig.height=3,fig.show='hold',fig.cap="Pathological non-convergence versus typical convergence."}
 load("./images/nonc.Rdata")
@@
 
-Research question when and how to diagnose non convergence?
-
-QUESTION: How?
+## How to diagnose non-convergence?
 
@@
 In mice these scalar summaries of interest are chain means and chain variances.
-As Van Buren describes users can also specify a model specific scalar summary (e.g., a regression coefficient).
-search a user specified scalar summary however is not universal to all complete data problems.
-Therefore as inspired by someone 2002 we will use introduce a scalar summary that summarizes the multivariate state space of the algorithm.
+As @buur18 describes, users can also specify a model-specific scalar summary (e.g., a regression coefficient).
+Such a user-specified scalar summary, however, is not universal to all complete data problems.
+Therefore, as inspired by **someone** (2002), we will use introduce a scalar summary that summarizes the multivariate state space of the algorithm.
 In this study we use the first eigenvalue of the variance covariance matrix of the completed data set per imputation.
@@
 To evaluate the performance of Rhat and $AC$ on the eigenvalues, we will use bias in R squared and in the estimated regression coefficient, the coverage rate of the CI95% of the regression estimate, and the CI length.
-**remove the term performance measure above and add that performance measures are bias in all estimands, and cov and CIL of regression coeff.**
+**remove the term performance measure above and add that performance measures are bias in all estimands, and cov and CI length of regression coeff.**
 
-ANSWER
-
-In general there is more bias in the estimate in conditions with a higher proportion of missingness and or a lower number of iterations.
-As required the values of o$\widehat{R}$ R lower for conditions with a higher number off iterations.
-End somewhat higher 4 conditions with a higher percentage of missing data.
+**Result:** In general there is more bias in the estimate in conditions with a higher proportion of missingness and or a lower number of iterations.
+As required the values of $\widehat{R}$ are lower for conditions with a higher number off iterations.
+End somewhat higher in conditions with a higher percentage of missing data.
 However autocorrelation did not show a decrease with increase number of iterations.
@@
 
-QUESTION: When?
+## When to diagnose non-convergence?
 
@@
 
-ANSWER
-
-Our head is equal to 1.2 is not stringent enough because the performance measures R not unbiased and confidence fellas fell it 4 every condition in which $\widehat{R}$ is smaller than one point 2.
+**Result:** $\widehat{R}$ is equal to 1.2 is not stringent enough because the performance measures R not unbiased and confidence fellas fell it 4 every condition in which $\widehat{R}$ is smaller than one point 2.
 At least, not when $\widehat{R}$ is smaller than one point two fur first occurs.
@@
 
-RECOMMENDATIONS
-
-For empirical researchers: 1) Check trace plots for pathological non convergence and adjust imputation model if necessary.
-2) use $\widehat{R}$ wait 1.1 ash threshold and autocorrelation with **?** as threshold .
-Keep iterating until these thresholds are reached.
-3) Do not use the R function ACF.
-Instead compute autocorrelations manually , see for example GitHub link.
-4) Track your own scalar summary of interest.
-This is advanced but explained in Van Buren 2018.
-Compute $\widehat{R}$ and autocorrelation values for this scalar summary.
-
-LIMITATIONS/FUTURE RESEARCH
-
-Pmm,
-M(N)AR,
-Empirical data,
-Vignette,
-ShinyMICE,
-AC across imputations, not iterations (unique for imp data??)
-
-If I don't include it in my study, also: Significance of $AC$ values; Estimate (PCA or reg. coeff.) versus Rhat and $AC$ on the estimate
-
-## Start methods section
+## Start old methods section
 
@@
 
-```{r, include = FALSE}
+```{r load-res, include = FALSE}
 # load("../Results/test.Rdata")
@@
 
-```{r}
+```{r res-univ}
 # mean_bias + plot_annotation(title = "Bias in mean estimate, convergence of chain means")
@@
 
-```{r}
+```{r res-multi}
 # est_bias + plot_annotation(title = "Bias in regression estimate, coverage rate, \n bias in variance explained")
@@
 
-```{r include=FALSE}
+```{r thresh-Rh, include=FALSE}
 # # chain means
@@
 
+## Recommendations for empirical researchers
+
+For empirical researchers: 1) Check trace plots for pathological non convergence and adjust imputation model if necessary.
+2) use $\widehat{R}$ wait 1.1 ash threshold and autocorrelation with **?** as threshold .
+Keep iterating until these thresholds are reached.
+3) Do not use the R function ACF.
+Instead compute autocorrelations manually , see for example GitHub link.
+4) Track your own scalar summary of interest.
+This is advanced but explained in Van Buren 2018.
+Compute $\widehat{R}$ and autocorrelation values for this scalar summary.
+
+## Recommendations for future research
+
+Pmm,
+M(N)AR,
+Empirical data,
+Vignette,
+ShinyMICE,
+AC across imputations, not iterations (unique for imp data??)
+
+If I don't include it in my study, also: Significance of $AC$ values; Estimate (PCA or reg. coeff.) versus Rhat and $AC$ on the estimate
+
 <!-- # References -->

### 2020-04-22 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Compute rhat and ac for pathological non-convergence case. Adjust functions by adding wrapper for general convergence function

@@
 
-\textbf{Move this after notation??} Figure 1 provides an overview of the
-steps involved with MI---from incomplete data, to \(M\) completed
-datasets, to \(M\) estimated quantities of interest (\(\hat{Q}\)s), to a
-single pooled estimate \(\bar{Q}\). Missing data in \(y\) is `imputed'
-(i.e., filled in) \(M\) times. The imputed data (\(y_{imp}\)) is
-combined with the observed data (\(y_{obs}\)) to create \(M\) completed
-datasets. On each completed dataset, the analysis of scientific interest
-(the `substantive model', or `complete data model') is performed. The
-quantity of scientific interest (e.g., a regression coefficient) is
-denoted with \(Q\). Since \(Q\) is estimated on each completed dataset,
-\(M\) separate \(\hat{Q}\)-values are obtained. These \(M\) values are
-combined into a single pooled estimate \(\bar{Q}\). \textbf{ALSO: make
-clear that iteration happens within the first step.}
+\textbf{Move this after notation??} Figure \ref{fig:diagram} provides an
+overview of the steps involved with MI---from incomplete data, to \(M\)
+completed datasets, to \(M\) estimated quantities of interest
+(\(\hat{Q}\)s), to a single pooled estimate \(\bar{Q}\). Missing data in
+\(y\) is `imputed' (i.e., filled in) \(M\) times. The imputed data
+(\(y_{imp}\)) is combined with the observed data (\(y_{obs}\)) to create
+\(M\) completed datasets. On each completed dataset, the analysis of
+scientific interest (the `substantive model', or `complete data model')
+is performed. The quantity of scientific interest (e.g., a regression
+coefficient) is denoted with \(Q\). Since \(Q\) is estimated on each
+completed dataset, \(M\) separate \(\hat{Q}\)-values are obtained. These
+\(M\) values are combined into a single pooled estimate \(\bar{Q}\).
+\textbf{ALSO: make clear that iteration happens within the first step.}
 
@@
 
-\caption{Scheme of the main steps in multiple imputation.}\label{fig:unnamed-chunk-1}
+\caption{Scheme of the main steps in multiple imputation.}\label{fig:diagram}
 \end{figure}
@@
 
-Main question: How can non-convergence be diagnosed? Sub-questions: Are
-common MCMC non-convergence diagnostics appropriate for MICE? And if so,
-which threshold should be used to diagnose non-convergence? How many
+\hypertarget{research-question-how-and-when-to-diagnose-non-convergence}{%
+\subsubsection{Research question: how and when to diagnose non
+convergence?}\label{research-question-how-and-when-to-diagnose-non-convergence}}
+
+Sub-questions: How can non-convergence be diagnosed? Are common MCMC
+non-convergence diagnostics appropriate for MICE? And if so, which
+threshold should be used to diagnose non-convergence? How many
 iterations are sufficient/needed to be able to diagnose non-convergence?
@@
 The two challenges for convergence of iterative algorithms are mixing
-and stationarity \citep{gelm13}, see Figure XYZ. Without mixing,
-imputation chains do not intermingle nicely. Without stationarity, there
-is trending within imputation chains. While the current practice is to
-inspect traceplots for signs of non-mixing and non-stationarity,
-diagnostic evaluation of these two challenges would be preferred.
-Non-stationarity within chains may be diagnosed with e.g.,
+and stationarity \citep{gelm13}, see Figure \ref{fig:non-conv}. Without
+mixing, imputation chains do not intermingle nicely. Without
+stationarity, there is trending within imputation chains. While the
+current practice is to inspect traceplots for signs of non-mixing and
+non-stationarity, diagnostic evaluation of these two challenges would be
+preferred. Non-stationarity within chains may be diagnosed with e.g.,
 autocorrelation \citep[\(AC\);][]{scha97, gelm13}, numeric standard
@@
 
-{\centering \includegraphics[width=.49\linewidth]{manuscript_files/figure-latex/unnamed-chunk-2-1} \includegraphics[width=.49\linewidth]{manuscript_files/figure-latex/unnamed-chunk-2-2}
+{\centering \includegraphics[width=.49\linewidth]{manuscript_files/figure-latex/non-conv-1} \includegraphics[width=.49\linewidth]{manuscript_files/figure-latex/non-conv-2}
 
@@
 
-\caption{Pathological non-convergence versus typical convergence.}\label{fig:unnamed-chunk-2}
+\caption{Pathological non-convergence versus typical convergence.}\label{fig:non-conv}
 \end{figure}
@@
 
-Research question when and how to diagnose non convergence?
-
-QUESTION: How?
+\hypertarget{how-to-diagnose-non-convergence}{%
+\subsection{How to diagnose
+non-convergence?}\label{how-to-diagnose-non-convergence}}
 
@@
 interest. In mice these scalar summaries of interest are chain means and
-chain variances. As Van Buren describes users can also specify a model
-specific scalar summary (e.g., a regression coefficient). search a user
-specified scalar summary however is not universal to all complete data
-problems. Therefore as inspired by someone 2002 we will use introduce a
-scalar summary that summarizes the multivariate state space of the
-algorithm. In this study we use the first eigenvalue of the variance
-covariance matrix of the completed data set per imputation. The methods
-under evaluation are \(\widehat{R}\) and \(AC\). These will be computed
-for the three scalar summaries of interest: chain means, chain
-variances, and the first eigenvalue of the variance-covariance matrix
-across imputations.
+chain variances. As \citet{buur18} describes, users can also specify a
+model-specific scalar summary (e.g., a regression coefficient). Such a
+user-specified scalar summary, however, is not universal to all complete
+data problems. Therefore, as inspired by \textbf{someone} (2002), we
+will use introduce a scalar summary that summarizes the multivariate
+state space of the algorithm. In this study we use the first eigenvalue
+of the variance covariance matrix of the completed data set per
+imputation. The methods under evaluation are \(\widehat{R}\) and \(AC\).
+These will be computed for the three scalar summaries of interest: chain
+means, chain variances, and the first eigenvalue of the
+variance-covariance matrix across imputations.
 
@@
 length. \textbf{remove the term performance measure above and add that
-performance measures are bias in all estimands, and cov and CIL of
+performance measures are bias in all estimands, and cov and CI length of
 regression coeff.}
 
-ANSWER
+\textbf{Result:} In general there is more bias in the estimate in
+conditions with a higher proportion of missingness and or a lower number
+of iterations. As required the values of \(\widehat{R}\) are lower for
+conditions with a higher number off iterations. End somewhat higher in
+conditions with a higher percentage of missing data. However
+autocorrelation did not show a decrease with increase number of
+iterations. After replacing the ACF function with a manual calculation
+for autocorrelation the autocorrelation values were indeed decreasing
+with a higher number of iterations. Evaluation with the performance
+measures shows that are had an autocorrelation are conservative: signs
+of non-convergence in conditions where estimands are unbiased and or
+confidence valid. As expected complete convergence (as indicated by
+autocorrelation equals 0; \(\widehat{R}\) equals 1) is not observed.
 
-In general there is more bias in the estimate in conditions with a
-higher proportion of missingness and or a lower number of iterations. As
-required the values of o\(\widehat{R}\) R lower for conditions with a
-higher number off iterations. End somewhat higher 4 conditions with a
-higher percentage of missing data. However autocorrelation did not show
-a decrease with increase number of iterations. After replacing the ACF
-function with a manual calculation for autocorrelation the
-autocorrelation values were indeed decreasing with a higher number of
-iterations. Evaluation with the performance measures shows that are had
-an autocorrelation are conservative: signs of non-convergence in
-conditions where estimands are unbiased and or confidence valid. As
-expected complete convergence (as indicated by autocorrelation equals 0;
-\(\widehat{R}\) equals 1) is not observed.
-
-QUESTION: When?
+\hypertarget{when-to-diagnose-non-convergence}{%
+\subsection{When to diagnose
+non-convergence?}\label{when-to-diagnose-non-convergence}}
 
@@
 
-ANSWER
-
-Our head is equal to 1.2 is not stringent enough because the performance
-measures R not unbiased and confidence fellas fell it 4 every condition
-in which \(\widehat{R}\) is smaller than one point 2. At least, not when
-\(\widehat{R}\) is smaller than one point two fur first occurs. The
-threshold \(\widehat{R}\) is equal to one point one is somewhat
-conservative with respect to the performance measures. It seems OK.
-\(\widehat{R}\) is equal to 1.01 is too stringent compared to the
-performance measures (\textbf{but may be necessary in more complex
-complete data models???}).
-
-RECOMMENDATIONS
-
-For empirical researchers: 1) Check trace plots for pathological non
-convergence and adjust imputation model if necessary. 2) use
-\(\widehat{R}\) wait 1.1 ash threshold and autocorrelation with
-\textbf{?} as threshold . Keep iterating until these thresholds are
-reached. 3) Do not use the R function ACF. Instead compute
-autocorrelations manually , see for example GitHub link. 4) Track your
-own scalar summary of interest. This is advanced but explained in Van
-Buren 2018. Compute \(\widehat{R}\) and autocorrelation values for this
-scalar summary.
-
-LIMITATIONS/FUTURE RESEARCH
-
-Pmm, M(N)AR, Empirical data, Vignette, ShinyMICE, AC across imputations,
-not iterations (unique for imp data??)
-
-If I don't include it in my study, also: Significance of \(AC\) values;
-Estimate (PCA or reg. coeff.) versus Rhat and \(AC\) on the estimate
+\textbf{Result:} \(\widehat{R}\) is equal to 1.2 is not stringent enough
+because the performance measures R not unbiased and confidence fellas
+fell it 4 every condition in which \(\widehat{R}\) is smaller than one
+point 2. At least, not when \(\widehat{R}\) is smaller than one point
+two fur first occurs. The threshold \(\widehat{R}\) is equal to one
+point one is somewhat conservative with respect to the performance
+measures. It seems OK. \(\widehat{R}\) is equal to 1.01 is too stringent
+compared to the performance measures (\textbf{but may be necessary in
+more complex complete data models???}).
 
@@
 
+\hypertarget{recommendations-for-empirical-researchers}{%
+\subsection{Recommendations for empirical
+researchers}\label{recommendations-for-empirical-researchers}}
+
+For empirical researchers: 1) Check trace plots for pathological non
+convergence and adjust imputation model if necessary. 2) use
+\(\widehat{R}\) wait 1.1 ash threshold and autocorrelation with
+\textbf{?} as threshold . Keep iterating until these thresholds are
+reached. 3) Do not use the R function ACF. Instead compute
+autocorrelations manually , see for example GitHub link. 4) Track your
+own scalar summary of interest. This is advanced but explained in Van
+Buren 2018. Compute \(\widehat{R}\) and autocorrelation values for this
+scalar summary.
+
+\hypertarget{recommendations-for-future-research}{%
+\subsection{Recommendations for future
+research}\label{recommendations-for-future-research}}
+
+Pmm, M(N)AR, Empirical data, Vignette, ShinyMICE, AC across imputations,
+not iterations (unique for imp data??)
+
+If I don't include it in my study, also: Significance of \(AC\) values;
+Estimate (PCA or reg. coeff.) versus Rhat and \(AC\) on the estimate
+
 \bibliographystyle{sageh}

### 2020-04-22 3.Thesis/1.SimulationStudy/Functions/Autocorrelation_supplement.R

@@ -1,33 +1,59 @@
 # autocorrelation supplement
+# requires library(purrr) and dplyr/magrittr
 
-# function for lag 1 AC for one or more imputation chains 
-ac_lag1 <- function(x){
-  # # make input the same object type
-  # x <- as.matrix(x)
-  # # create output object with length equal to the number of imputation chains of the input
-  # ac <- matrix(NA, 1, dim(x)[2])
-  # # for each imputation
-  # for (m in 1:dim(x)[2]) {
-  #   # compute the correlation between the chain and itself, one iteration delayed
-  #   ac[m] <- cor(x[-dim(x)[1], m], x[-1, m])
-  # }
-  ac <- map_dbl(1:dim(x)[2], function(m){
-    cor(x[-dim(x)[1], m], x[-1, m]) 
-  })
-  names(ac) <- paste0("ac.", names(x))
-  #names(ac) <- names(x)
-  if(!is.na(ac)[1]){
+# function for lag 1 AC for one or more imputation chains
+ac_lag1 <- function(x) {
+  # input: object with theta values (rows are iterations, columns are imputations)
+  # output: autocorrelation per imputation, with max and mean across imputations
+  
+  # make input ready for mapping
+  if (is.numeric(x)) {
+    x <- as.data.frame(x)
+  }
+  
+  # compute ac
+  ac <- map_dbl(1:dim(x)[2], function(m) {
+    cor(x[-dim(x)[1], m], x[-1, m])
+  }) %>% set_names(paste0("ac.", names(x)))
+  
+  # add mean and max for iterations > 2
+  if (!is.na(ac)[1]) {
     ac <- c(ac, mean.ac = mean(ac), max.ac = max(ac))
-    }
+  }
+  
+  # output
   return(data.frame(t(ac)))
 }
 
-acf_lag1 <- function(x){acf(x, lag.max = 1, plot = FALSE) %>% .$acf %>% .[2, , ] %>% diag() %>% set_names(paste0("acf.", names(x)))}
+# adjusted stats::acf() function for lag 1 AC for one or more imputation chains
+acf_lag1 <- function(x) {
+  # input: object with theta values (rows are iterations, columns are imputations)
+  # output: autocorrelation per imputation, with max and mean across imputations
+  
[…]

### 2020-04-23 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Both Rhat and AC work properly now for all iterations. Last version with old functions. Cleaning up Functions folder after this.

@@
 
-\textbf{Note for peer review: things written in bold are reminders for
-myself or things that I am unsure off still. Also, I just decided to
-change things majorly in my results section. So right now, the figures
-do not correspond to the text. Please only skim that section, because I
-am definitely going to rewrite it. Thanks already! - Hanne}
+\textbf{Dear Daniela, I have just implemented some major changes in my
+Methods and Results sections. Therefore, these are kind of a mess.
+Please do not waste any time on spelling/wording there, I will continue
+improving it. On the other hand, if you have general advice, please tell
+me because I'm happy to move thngs around still! Figures in the Results
+are currently missing, but will help in understanding the paper. Sorry
+that you won't have that now! Also, things written in bold are reminders
+for myself or things that I am unsure off still. Thanks already! -
+Hanne}
 
@@
 \(p\) variables for all \(n\) units in a sample. The data value of unit
-\(i\) (\(i = 1, \dots, n\)) on variable \(j\) (\(j = 1, \dots, p\)) may
-be either observed or missing. The collection of observed data values in
-\(y\) is denoted by \(y_{obs}\); the missing part of \(y\) is referred
-to as \(y_{mis}\). For each datapoint in \(y_{mis}\), we sample
-\(M \times T\) times plausible values, where \(M\) is the number of
-imputations (\(m = 1, 2, \dots, M\)) and \(T\) is the number of
-iterations (\(t = 1, 2, \dots, T\)). The collection of samples between
-the initial value (at \(t=1\)) and the final imputed value (at \(t=T\))
-will be referred to as an `imputation chain'. \textbf{Add: missingness
+\(i\) (\(i = 1, 2, \dots, n\)) on variable \(j\)
+(\(j = 1, 2, \dots, p\)) may be either observed or missing. The
+collection of observed data values in \(y\) is denoted by \(y_{obs}\);
+the missing part of \(y\) is referred to as \(y_{mis}\). For each
+datapoint in \(y_{mis}\), we sample \(M \times T\) times plausible
+values, where \(M\) is the number of imputations
+(\(m = 1, 2, \dots, M\)) and \(T\) is the number of iterations
+(\(t = 1, 2, \dots, T\)). The collection of samples between the initial
+value (at \(t=1\)) and the final imputed value (at \(t=T\)) will be
+referred to as an `imputation chain'. \textbf{Add: missingness
 percentage = proportion of cases with one or more missing values, Q is
@@
 quantity of scientific interest that we can calculate if we would
-observe the entire population'' \citep[par 2.3.1]{buur18}. }
+observe the entire population'' \citep[par 2.3.1]{buur18}, and
+\(\theta\)s are scalar summaries of interest in the iterative algorithm
+(e.g., chain means; the average of the imputed values in each imputation
+chain). }
 
@@
 
-The two challenges of convergence are stationarity and mixing point.
-Mixing and stationarity can be evaluated visually. Common practice is to
-evaluate stationarity and mixing in trace plots for scalar summaries of
-interest. In mice these scalar summaries of interest are chain means and
-chain variances. As \citet{buur18} describes, users can also specify a
-model-specific scalar summary (e.g., a regression coefficient). Such a
+The two challenges of convergence are stationarity and mixing. Mixing
+and stationarity can be inspected visually, by evaluating traceplots of
+scalar summaries of interest (\(\theta\)s; e.g., chain means and chain
+variances). As \citet{buur18} describes, users can also specify a
+model-specific scalar summary (e.g., a regression coefficient). A
 user-specified scalar summary, however, is not universal to all complete
 data problems. Therefore, as inspired by \textbf{someone} (2002), we
-will use introduce a scalar summary that summarizes the multivariate
-state space of the algorithm. In this study we use the first eigenvalue
-of the variance covariance matrix of the completed data set per
-imputation. The methods under evaluation are \(\widehat{R}\) and \(AC\).
-These will be computed for the three scalar summaries of interest: chain
-means, chain variances, and the first eigenvalue of the
-variance-covariance matrix across imputations.
+will use introduce a \(\theta\) that summarizes the multivariate state
+space of the algorithm. Namely, the first eigenvalue of the
+variance-covariance matrix of the \(M\) completed datasets.
 
-The two methods will be evaluated with an estimand per scalar summary of
-interest. It is assumed that when the estimates are unbiased (and
-confidence valid), the algorithm is sufficiently converged. The
-\(\widehat{R}\) and \(AC\) values with chain means as theta are
-evaluated against the bias in univariate mean estimates (\textbf{just
-for Y??}). Similarly, the performance measure for Rhat and \(AC\)
-applied to chain variances is the bias in estimated standard deviations.
-To evaluate the performance of Rhat and \(AC\) on the eigenvalues, we
-will use bias in R squared and in the estimated regression coefficient,
-the coverage rate of the CI95\% of the regression estimate, and the CI
-length. \textbf{remove the term performance measure above and add that
-performance measures are bias in all estimands, and cov and CI length of
-regression coeff.}
+The methods under evaluation are \(\widehat{R}\) and \(AC\). These will
+be computed for the three scalar summaries of interest: chain means
+(i.e., per imputation, per variable: on \(y_{imp, m, j}\)), chain
+variances (i.e., per imputation, per variable: on \(y_{imp, m, j}\)),
+and the first eigenvalue of the variance-covariance matrix across
+imputations (i.e., per imputation: on \{\(y_{obs}, y_{imp, m}\)\})
+\textbf{add user-specified \(\theta\): estimated regression coefficient
+per imputation}.
 
-\textbf{Result:} In general there is more bias in the estimate in
-conditions with a higher proportion of missingness and or a lower number
-of iterations. As required the values of \(\widehat{R}\) are lower for
-conditions with a higher number off iterations. End somewhat higher in
-conditions with a higher percentage of missing data. However
-autocorrelation did not show a decrease with increase number of
-iterations. After replacing the ACF function with a manual calculation
-for autocorrelation the autocorrelation values were indeed decreasing
-with a higher number of iterations. Evaluation with the performance
-measures shows that are had an autocorrelation are conservative: signs
-of non-convergence in conditions where estimands are unbiased and or
-confidence valid. As expected complete convergence (as indicated by
-autocorrelation equals 0; \(\widehat{R}\) equals 1) is not observed.
+The performance of two methods (\(\widehat{R}\) and \(AC\)) on each of
+these scalar summaries of interest will be evaluated with several
+estimands (quantities of scientific interest; \(Q\)s). It is assumed
+that when \(\bar{Q}\) (the pooled result across imputations) is an
+unbiased (and confidence valid) estimate of \(Q\), the algorithm is
+sufficiently converged. For each scalar \(\theta\), one estimand \(Q\)
+is defined that may be of interest in empirical research. The
+\(\widehat{R}\) and \(AC\) values with chain means as \(\theta\)s are
+evaluated against the bias in univariate mean estimates (\textbf{all
+variables or just for Y??}). Similarly, the performance measure for
+\(\widehat{R}\) and \(AC\) applied to chain variances is the bias in
+estimated standard deviations. To evaluate the performance of
+\(\widehat{R}\) and \(AC\) on the eigenvalues, we will use bias in the
+coefficient of determination, \(R^2\). Additionally, we will evaluate
+the estimated regression coefficients, the coverage rate of the CI95\%
+of the regression estimates, and the CI length (\textbf{see definitions
+in old Methods}). \textbf{remove the term performance measure here and
+add an extra alinea to define performance measures as bias in all
+estimands, and coverage rate and CI length of regression coefficients.}
+
+\textbf{Move this to Results:} As expected, conditions with a higher
+proportion of missingness and/or a lower number of iterations show more
+signs of non-convergence, as indicated by more extreme bias in the
+estimated \(Q\)s. Roughly speaking this means that MICE is indeed not
+converged as \(t=1\), and converges gradually as \(t\) increases. The
+point at wich an additional iteration does not lead to an improvement of
+the estimates depends on the difficulty of the missingness
+problem---between XYZ and XYZ. This means that the algorithm has
+converged sufficiently (\textbf{under the current specifications}).
+
+As required for a method to diagnose non-convergence, the values of
+\(\widehat{R}\) follow a trend similar to the performance measures.
+\(\widehat{R}\) values are generally lower in conditions with a higher
+number of iterations, and somewhat higher in conditions with a higher
+percentage of missing data. \textbf{Decide where to specify the two
+calculations of AC:} Initially, autocorrelation did not show a decrease
+with an increasing number of iterations. But, after replacing the ACF
+function with a manual calculation for autocorrelation, the
+autocorrelation values were indeed decreasing with a higher number of
+iterations.
+
+Evaluation with the performance measures shows that are had an
+autocorrelation are conservative: they indicate signs of non-convergence
+in conditions where \(\bar{Q}\)s are unbiased and or confidence valid
+estimates of \(Q\).
 
@@
 
-Usually, convergence is diagnosed when Rhat \textless{} 1.2 or 1.1 or
-1.01. And a t-test is performed for \(AC\) values == 0. Performance
-measures are the same as above: unbiased, confidence valid estimates.
+Upon convergence, \(\widehat{R}=1\) and \(AC=0\), but in MCMC algorithms
+these are very unlikely (because it's a distribution, not a point to
+converge to). In practice thresholds are used to diagnose
+non-convergence. Usually, non-convergence is diagnosed when
+\(\widehat{R}\) \textgreater{} 1.2 or 1.1 or even 1.01. And a t-test is
+performed to assess whether \(AC\) is significantly different from zero.
+Performance measures are the same as above: unbiased, confidence valid
+estimates.
 
-\textbf{Result:} \(\widehat{R}\) is equal to 1.2 is not stringent enough
-because the performance measures R not unbiased and confidence fellas
-fell it 4 every condition in which \(\widehat{R}\) is smaller than one
-point 2. At least, not when \(\widehat{R}\) is smaller than one point
-two fur first occurs. The threshold \(\widehat{R}\) is equal to one
-point one is somewhat conservative with respect to the performance
-measures. It seems OK. \(\widehat{R}\) is equal to 1.01 is too stringent
-compared to the performance measures (\textbf{but may be necessary in
-more complex complete data models???}).
+\textbf{Move this to Results:} As expected, complete convergence
+(\(\widehat{R}=1\) and \(AC=0\)) is not observed. But what about the
+thresholds for practice? \(\widehat{R}\) \textless1.2 is not stringent
+enough, because conditions in which \(\widehat{R}\) is smaller than 1.2
+do not all yield unbiased estimates of the \(Q\)s. Moreover, there is a
+dip in the \(\widehat{R}\) values, after which \(\widehat{R}\)
+\textgreater{} 1.2 again. If the algorithms would be terminated at the
+iteration where \(\widehat{R}\) \textless{} 1.2 occurs for the first
+time, the increase in \(\widehat{R}\) values might be missed. The
+threshold \(\widehat{R}\) \textless{} 1.1 is somewhat conservative with
+respect to the performance measures. But it seems OK. \(\widehat{R}\)
+\textless{} 1.01 is too stringent compared to the performance measures.
+And it is not obtained for any number of iterations (\textbf{but may be
+necessary in more complex complete data models???}).
 
-\hypertarget{start-methods-section}{%
-\subsection{Start methods section}\label{start-methods-section}}
+\hypertarget{start-old-methods-section}{%
+\subsection{Start old methods section}\label{start-old-methods-section}}
 
@@
 
-\hypertarget{results}{%
-\section{Results}\label{results}}
+\hypertarget{old-results-section}{%
+\section{Old Results section}\label{old-results-section}}
 
@@
 reached in simulation conditions \(T = 3\) and \(T > 6\). (With the
-default number of iterations (maxit = 5), this dip in Rhat values would
-be spotted, so it is no problem.) The point at which an extra iteration
-does not seem to improve the Rhat value is around \(T=30\).
+default number of iterations (maxit = 5), this dip in \(\widehat{R}\)
+values would be spotted, so it is no problem.) The point at which an
+extra iteration does not seem to improve the \(\widehat{R}\) value is
+around \(T=30\).
 
@@
 If I don't include it in my study, also: Significance of \(AC\) values;
-Estimate (PCA or reg. coeff.) versus Rhat and \(AC\) on the estimate
+Estimate (PCA or reg. coeff.) versus \(\widehat{R}\) and \(AC\) on the
+estimate
 

### 2020-04-23 3.Thesis/1.SimulationStudy/Functions/Rhat_supplement.R

@@ -14,18 +14,19 @@ rhat_adapted <- function(x) {
   # quit function if there are not enough iterations
   t <- dim(x)[1]
   if (is.null(t)) {
-    r.hat <- data.frame(iteration = 1, r.hat = NA)
+    r.hat <- data.frame(iteration = 1, max.r.hat = NA, r.hat = NA)
   }
   else {
-  r.hat <- map_dfr(2:t, function(it){
-  # compute r hat in all ways described by Vehtari et al. (2019)
-  rhat_original <- x[1:it,] %>% get.rhat()
-  rhat_bulk <- x[1:it,] %>% split_chains() %>% z_scale() %>% get.rhat()
-  rhat_tail <- x[1:it,] %>% fold_sims() %>% split_chains() %>% z_scale() %>% get.rhat()
-  max.rhat <- max(rhat_original, rhat_bulk, rhat_tail) %>% data.frame() %>% set_names("r.hat") 
-  }) %>% rbind(NA,.) %>% cbind(iteration = 1:t, .)
+    r.hat <- map_dfr(2:t, function(it) {
+      # compute r hat in all ways described by Vehtari et al. (2019)
+      rhat_original <- x[1:it, ] %>% get.rhat()
+      rhat_bulk <-
+        x[1:it, ] %>% split_chains() %>% z_scale() %>% get.rhat()
+      rhat_tail <-
+        x[1:it, ] %>% fold_sims() %>% split_chains() %>% z_scale() %>% get.rhat()
+      max(rhat_bulk, rhat_tail) %>% data.frame(max.r.hat = ., r.hat = rhat_original) #%>% set_names("r.hat")
+    }) %>% rbind(NA, .) %>% cbind(iteration = 1:t, .)
   }
   # output
   return(r.hat)
 }
-

### 2020-04-23 2.ResearchReport/1.SimulationStudy/Writeup/NoteConvergence.tex

@@
 These $m$ values are combined into a single pooled estimate $\bar{Q}$.}}
+\end{figure}
+
+\begin{figure}
+%\label{fig:steps2}
+\centering
+	\large{\begin{tikzpicture}
+	\tikzstyle{main}=[circle, minimum size = 25mm, thick, draw =black!80, node distance = 20mm]
+\tikzstyle{connect}=[-latex, thick]
+\node[main, fill = white!100] (data) [label=below:Incomplete data] {$\{y_{obs}, y_{mis}\}$};
+\node[main] (mids) [right=of data] { $\{y_{obs}, y_{imp}\}$};
+\node[main] (mids2) [above=10mm of mids] {$\{y_{obs}, y_{imp}\}$ };
+\node[main] (mids3) [below=10mm of mids,label=below: Completed data (per $m$)] { $\{y_{obs}, y_{imp}\}$};
+\node[main] (mira) [right=of mids] {$\hat{Q}$};
+\node[main] (mira2) [above=10mm of mira] {$\hat{Q}$ };
+\node[main] (mira3) [below=10mm of mira,label=below: Results (per $m$)] {$\hat{Q}$ };
+\node[main, fill = white!100] (mipo) [right=of mira,label=below:Pooled results] {$\bar{Q}$ };
+\path (data) edge [connect] (mids)
+      (data) edge [connect] (mids2)
+      (data) edge [connect] (mids3)
+      (mids) edge [connect] (mira)
+      (mids2) edge [connect] (mira2)
+      (mids3) edge [connect] (mira3)
+		  (mira) edge [connect] (mipo)
+		  (mira2) edge [connect] (mipo)
+		  (mira3) edge [connect] (mipo);
+\end{tikzpicture}}
+%\caption{Scheme of the main steps in multiple imputation \citep[$m = 3$; adapted from][\S~1.4.1]{buur18}.
+{\footnotesize Missing data in dataset $Y$ is `imputed' (i.e., filled in) $m$ times.
+The imputed data is combined with the observed data ($Y_{obs}$) to create $m$ completed datasets.
+On each completed dataset, the analysis of scientific interest (or `substantive model') is performed.
+The quantity of scientific interest (e.g., a regression coefficient) is denoted with $Q$.
+Since $Q$ is estimated on each completed dataset, $m$ separate $\hat{Q}$-values are obtained.
+These $m$ values are combined into a single pooled estimate $\bar{Q}$.
+$\{y_{obs}, y_{mis}\}$ }}
 \end{figure}

### 2020-04-23 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd

@@
 
-**Note for peer review: things written in bold are reminders for myself or things that I am unsure off still.
-Also, I just decided to change things majorly in my results section.
-So right now, the figures do not correspond to the text.
-Please only skim that section, because I am definitely going to rewrite it.
+**Dear Daniela, I have just implemented some major changes in my Methods and Results sections.
+Therefore, these are kind of a mess.
+Please do not waste any time on spelling/wording there, I will continue improving it.
+On the other hand, if you have general advice, please tell me because I'm happy to move thngs around still!
+Figures in the Results are currently missing, but will help in understanding the paper.
+Sorry that you won't have that now!
+Also, things written in bold are reminders for myself or things that I am unsure off still.
 Thanks already!
@@
 Let $y$ denote an $n \times p$ matrix containing the data values on $p$ variables for all $n$ units in a sample.
-The data value of unit $i$ ($i = 1, \dots, n$) on variable $j$ ($j = 1, \dots, p$) may be either observed or missing.
+The data value of unit $i$ ($i = 1, 2, \dots, n$) on variable $j$ ($j = 1, 2, \dots, p$) may be either observed or missing.
 The collection of observed data values in $y$ is denoted by $y_{obs}$; the missing part of $y$ is referred to as $y_{mis}$.
@@
 The collection of samples between the initial value (at $t=1$) and the final imputed value (at $t=T$) will be referred to as an 'imputation chain'.
-**Add: missingness percentage = proportion of cases with one or more missing values, Q is quantity of scientific interest: "A scientific estimand $Q$ is a quantity of scientific interest that we can calculate if we would observe the entire population" [@buur18, par 2.3.1].
+**Add: missingness percentage = proportion of cases with one or more missing values, Q is quantity of scientific interest: "A scientific estimand $Q$ is a quantity of scientific interest that we can calculate if we would observe the entire population" [@buur18, par 2.3.1], and $\theta$s are scalar summaries of interest in the iterative algorithm (e.g., chain means; the average of the imputed values in each imputation chain).
 **
@@
 
-The two challenges of convergence are stationarity and mixing point.
-Mixing and stationarity can be evaluated visually.
-Common practice is to evaluate stationarity and mixing in trace plots for scalar summaries of interest.
-In mice these scalar summaries of interest are chain means and chain variances.
+The two challenges of convergence are stationarity and mixing.
+Mixing and stationarity can be inspected visually, by evaluating traceplots of scalar summaries of interest ($\theta$s; e.g., chain means and chain variances).
 As @buur18 describes, users can also specify a model-specific scalar summary (e.g., a regression coefficient).
-Such a user-specified scalar summary, however, is not universal to all complete data problems.
-Therefore, as inspired by **someone** (2002), we will use introduce a scalar summary that summarizes the multivariate state space of the algorithm.
-In this study we use the first eigenvalue of the variance covariance matrix of the completed data set per imputation.
+A user-specified scalar summary, however, is not universal to all complete data problems.
+Therefore, as inspired by **someone** (2002), we will use introduce a $\theta$ that summarizes the multivariate state space of the algorithm.
+Namely, the first eigenvalue of the variance-covariance matrix of the $M$ completed datasets.
+
 The methods under evaluation are $\widehat{R}$ and $AC$.
-These will be computed for the three scalar summaries of interest: chain means, chain variances, and the first eigenvalue of the variance-covariance matrix across imputations.
+These will be computed for the three scalar summaries of interest: chain means (i.e., per imputation, per variable: on $y_{imp, m, j}$), chain variances (i.e., per imputation, per variable: on $y_{imp, m, j}$), and the first eigenvalue of the variance-covariance matrix across imputations (i.e., per imputation: on {$y_{obs}, y_{imp, m}$}) **add user-specified $\theta$: estimated regression coefficient per imputation**.
 
-The two methods will be evaluated with an estimand per scalar summary of interest.
-It is assumed that when the estimates are unbiased (and confidence valid), the algorithm is sufficiently converged.
-The $\widehat{R}$ and $AC$ values with chain means as theta are evaluated against the bias in univariate mean estimates (**just for Y??**).
-Similarly, the performance measure for Rhat and $AC$ applied to chain variances is the bias in estimated standard deviations.
-To evaluate the performance of Rhat and $AC$ on the eigenvalues, we will use bias in R squared and in the estimated regression coefficient, the coverage rate of the CI95% of the regression estimate, and the CI length.
-**remove the term performance measure above and add that performance measures are bias in all estimands, and cov and CI length of regression coeff.**
+The performance of two methods ($\widehat{R}$ and $AC$) on each of these scalar summaries of interest will be evaluated with several estimands (quantities of scientific interest; $Q$s).
+It is assumed that when $\bar{Q}$ (the pooled result across imputations) is an unbiased (and confidence valid) estimate of $Q$, the algorithm is sufficiently converged.
+For each scalar $\theta$, one estimand $Q$ is defined that may be of interest in empirical research.
+The $\widehat{R}$ and $AC$ values with chain means as $\theta$s are evaluated against the bias in univariate mean estimates (**all variables or just for Y??**).
+Similarly, the performance measure for $\widehat{R}$ and $AC$ applied to chain variances is the bias in estimated standard deviations.
+To evaluate the performance of $\widehat{R}$ and $AC$ on the eigenvalues, we will use bias in the coefficient of determination, $R^2$.
+Additionally, we will evaluate the estimated regression coefficients, the coverage rate of the CI95% of the regression estimates, and the CI length (**see definitions in old Methods**).
+**remove the term performance measure here and add an extra alinea to define performance measures as  bias in all estimands, and coverage rate and CI length of regression coefficients.**
 
-**Result:** In general there is more bias in the estimate in conditions with a higher proportion of missingness and or a lower number of iterations.
-As required the values of $\widehat{R}$ are lower for conditions with a higher number off iterations.
-End somewhat higher in conditions with a higher percentage of missing data.
-However autocorrelation did not show a decrease with increase number of iterations.
-After replacing the ACF function with a manual calculation for autocorrelation the autocorrelation values were indeed decreasing with a higher number of iterations.
-Evaluation with the performance measures shows that are had an autocorrelation are conservative: signs of non-convergence in conditions where estimands are unbiased and or confidence valid.
-As expected complete convergence (as indicated by autocorrelation equals 0; $\widehat{R}$ equals 1)  is not observed.
+**Move this to Results:** As expected, conditions with a higher proportion of missingness and/or a lower number of iterations show more signs of non-convergence, as indicated by more extreme bias in the estimated $Q$s.
+Roughly speaking this means that MICE is indeed not converged as $t=1$, and converges gradually as $t$ increases.
+The point at wich an additional iteration does not lead to an improvement of the estimates depends on the difficulty of the missingness problem---between XYZ and XYZ.
+This means that the algorithm has converged sufficiently (**under the current specifications**).
+
+As required for a method to diagnose non-convergence, the values of $\widehat{R}$ follow a trend similar to the performance measures.
+$\widehat{R}$ values are generally lower in conditions with a higher number of iterations, and somewhat higher in conditions with a higher percentage of missing data.
+**Decide where to specify the two calculations of AC:** Initially, autocorrelation did not show a decrease with an increasing number of iterations.
+But, after replacing the ACF function with a manual calculation for autocorrelation, the autocorrelation values were indeed decreasing with a higher number of iterations.
+
+Evaluation with the performance measures shows that are had an autocorrelation are conservative: they indicate signs of non-convergence in conditions where $\bar{Q}$s are unbiased and or confidence valid estimates of $Q$.
 
@@
 
-Usually, convergence is diagnosed when Rhat < 1.2 or 1.1 or 1.01. And a t-test is performed for $AC$ values == 0. Performance measures are the same as above: unbiased, confidence valid estimates.
+Upon convergence, $\widehat{R}=1$ and $AC=0$, but in MCMC algorithms these are very unlikely (because it's a distribution, not a point to converge to).
+In practice thresholds are used to diagnose non-convergence.
+Usually, non-convergence is diagnosed when $\widehat{R}$  > 1.2 or 1.1 or even 1.01.
+And a t-test is performed to assess whether $AC$ is significantly different from zero.
+Performance measures are the same as above: unbiased, confidence valid estimates.
 
-**Result:** $\widehat{R}$ is equal to 1.2 is not stringent enough because the performance measures R not unbiased and confidence fellas fell it 4 every condition in which $\widehat{R}$ is smaller than one point 2.
-At least, not when $\widehat{R}$ is smaller than one point two fur first occurs.
-The threshold $\widehat{R}$ is equal to one point one is somewhat conservative with respect to the performance measures.
-It seems OK.
-$\widehat{R}$ is equal to 1.01 is too stringent compared to the performance measures (**but may be necessary in more complex complete data models???**).
+**Move this to Results:** As expected, complete convergence ($\widehat{R}=1$ and $AC=0$) is not observed.
+But what about the thresholds for practice?
+$\widehat{R}$ <1.2 is not stringent enough, because conditions in which $\widehat{R}$ is smaller than 1.2 do not all yield unbiased estimates of the $Q$s.
+Moreover, there is a dip in the $\widehat{R}$ values, after which $\widehat{R}$ > 1.2 again.
+If the algorithms would be terminated at the iteration where $\widehat{R}$ < 1.2 occurs for the first time, the increase in $\widehat{R}$ values might be missed.
+The threshold $\widehat{R}$ < 1.1 is somewhat conservative with respect to the performance measures.
+But it seems OK.
+$\widehat{R}$ < 1.01 is too stringent compared to the performance measures.
+And it is not obtained for any number of iterations (**but may be necessary in more complex complete data models???**).
 
@@
 
-# Results
+# Old Results section
 
@@
 The conventional $\widehat{R}$ threshold of 1.2 is reached in simulation conditions $T = 3$ and $T > 6$.
-(With the default number of iterations (maxit = 5), this dip in Rhat values would be spotted, so it is no problem.) The point at which an extra iteration does not seem to improve the Rhat value is around $T=30$.
-<!-- Rhat values are indistinguishable between simulation conditions when $T>30$.  -->
+(With the default number of iterations (maxit = 5), this dip in $\widehat{R}$ values would be spotted, so it is no problem.) The point at which an extra iteration does not seem to improve the $\widehat{R}$ value is around $T=30$.
+<!-- $\widehat{R}$ values are indistinguishable between simulation conditions when $T>30$.  -->
 <!-- The $\widehat{R}$ values seemed to flatten off between 20 and 30 iterations.  -->
@@
 
-If I don't include it in my study, also: Significance of $AC$ values; Estimate (PCA or reg. coeff.) versus Rhat and $AC$ on the estimate
+If I don't include it in my study, also: Significance of $AC$ values; Estimate (PCA or reg. coeff.) versus $\widehat{R}$ and $AC$ on the estimate
 

### 2020-04-24 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
Add plot with rhat and ac to show pathological non-conv

@@
 
-**Dear Daniela, I have just implemented some major changes in my Methods and Results sections.
-Therefore, these are kind of a mess.
-Please do not waste any time on spelling/wording there, I will continue improving it.
-On the other hand, if you have general advice, please tell me because I'm happy to move thngs around still!
-Figures in the Results are currently missing, but will help in understanding the paper.
-Sorry that you won't have that now!
-Also, things written in bold are reminders for myself or things that I am unsure off still.
-Thanks already!
-- Hanne**
-
 # Introduction
@@
 If a dataset contains just one missing value for a variable of interest, statistical inferences are undefined and will not produce any result.
-Incomplete observations are therefore ignored by default in many statistical packages (i.e., list-wise deletion is employed).
+To circumvent this, incomplete observations are ignored by default in many statistical packages (i.e., list-wise deletion is employed).
 Unfortunately, this *ad hoc* solution may yield wildly invalid results [@buur18].
@@
 These $M$ values are combined into a single pooled estimate $\bar{Q}$.
-**ALSO: make clear that iteration happens within the first step.**
+**ALSO: make clear that iteration happens within the first step, between incomplete and completed data.**
 

### 2020-04-24 3.Thesis/1.SimulationStudy/Functions/Convergence.R

@@ -1,6 +1,9 @@
 # Convergence function to combine Rhat and AC
 # Requires 'dplyr', and the functions in 'AC.R', 'AC_supplement.R', 'Rhat.R', and 'Rhat_supplement.R'
 
-convergence <- function(x, ac_function = "ac") {
-  left_join(rhat_adapted(x), ac_adapted(x, ac_function), by = "iteration")
+convergence <- function(x, include_acf = FALSE) {
+  out <-  left_join(rhat_adapted(x), ac_adapted(x), by = "iteration") 
+  if(include_acf){ out <- left_join(out, ac_adapted(x, "acf"), by = "iteration")
+  }
+  return(out)
 }

### 2020-04-24 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
Placed new figures in manuscript

@@
 
-```{r non-conv, out.width='.49\\linewidth', fig.width=3.5, fig.height=3,fig.show='hold',fig.cap="Pathological non-convergence versus typical convergence."}
-load("./images/nonc.Rdata")
-load("./images/conv.Rdata")
+```{r non-conv, out.width='.49\\linewidth', fig.width=3.5, fig.height=3,fig.show='hold',fig.cap="Typical convergence versus pathological non-convergence."}
+load("./images/nonc_plot.Rdata")
+load("./images/conv_plot.Rdata")
 par(mar = c(2, 2, 0.1, 0.1))
+plot(conv)
 plot(nonc)
-plot(conv)
 
+```
+
+```{r diagnostics, fig.cap="Convergence diagnostics pathological non-convergence versus typical convergence."}
+library(ggplot2)
+library(patchwork)
+theme_update(
+  plot.title = element_text(hjust = 0.5),
+  panel.border = element_blank(),
+  panel.grid.major = element_blank(),
+  panel.grid.minor = element_blank(),
+  panel.background = element_blank(),
+  axis.line = element_line(colour = "black"),
+  legend.key = element_blank(),
+  legend.position = "bottom",
+  legend.margin = margin(0,0,0,0),
+  # adjust text size
+  text = element_text(size = 8),
+  title = element_text(size = 8)
+  # axis.title = element_text(size = 8),
+  # axis.text = element_text(size = 7),
+  # legend.text = element_text(size = 7)
+)
+
+load("./images/diagnostics_plot.Rdata")
+diagnostics_plot
 ```

### 2020-04-24 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Placed new figures in manuscript

@@
 
-\textbf{Dear Daniela, I have just implemented some major changes in my
-Methods and Results sections. Therefore, these are kind of a mess.
-Please do not waste any time on spelling/wording there, I will continue
-improving it. On the other hand, if you have general advice, please tell
-me because I'm happy to move thngs around still! Figures in the Results
-are currently missing, but will help in understanding the paper. Sorry
-that you won't have that now! Also, things written in bold are reminders
-for myself or things that I am unsure off still. Thanks already! -
-Hanne}
-
 \hypertarget{introduction}{%
@@
 variable of interest, statistical inferences are undefined and will not
-produce any result. Incomplete observations are therefore ignored by
-default in many statistical packages (i.e., list-wise deletion is
-employed). Unfortunately, this \emph{ad hoc} solution may yield wildly
-invalid results \citep{buur18}. An alternative is to `impute' (i.e.,
-fill in) every missing value in an incomplete dataset. With imputation
-techniques, one or several completed datasets are created, on which
-statistical inferences can be performed. The case with several completed
-datasets is known as `multiple imputation' \citep[MI;][]{rubin76}, and
-requires an additional step to pool the results across imputations
-\textbf{see figure 1??}.
+produce any result. To circumvent this, incomplete observations are
+ignored by default in many statistical packages (i.e., list-wise
+deletion is employed). Unfortunately, this \emph{ad hoc} solution may
+yield wildly invalid results \citep{buur18}. An alternative is to
+`impute' (i.e., fill in) every missing value in an incomplete dataset.
+With imputation techniques, one or several completed datasets are
+created, on which statistical inferences can be performed. The case with
+several completed datasets is known as `multiple imputation'
+\citep[MI;][]{rubin76}, and requires an additional step to pool the
+results across imputations \textbf{see figure 1??}.
 
@@
 \(M\) values are combined into a single pooled estimate \(\bar{Q}\).
-\textbf{ALSO: make clear that iteration happens within the first step.}
+\textbf{ALSO: make clear that iteration happens within the first step,
+between incomplete and completed data.}
 
@@
 
-\caption{Pathological non-convergence versus typical convergence.}\label{fig:non-conv}
+\caption{Typical convergence versus pathological non-convergence.}\label{fig:non-conv}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics{manuscript_files/figure-latex/diagnostics-1}
+
+}
+
+\caption{Convergence diagnostics pathological non-convergence versus typical convergence.}\label{fig:diagnostics}
 \end{figure}

### 2020-04-24 3.Thesis/1.SimulationStudy/1.Execute.R

@@ -14,7 +14,6 @@ library(tidyverse)
 # library(ggplot2)
 # library(purrr)
 # library(magrittr)
-# library(tibble)
 
 # load simulation/evaluation functions
 source("3.Thesis/1.SimulationStudy/Functions/CreateData.R")
@@ -28,8 +27,8 @@ source("3.Thesis/1.SimulationStudy/Functions/AC_supplement.R")
 
 # simulation parameters
 populationsize <- 1000 #n of simulated dataset
-n.iter <- 20 #nr of iterations (varying 1:n.iter)
-n.sim <- 2 #nr of simulations per iteration value
+n.iter <- 100 #nr of iterations (varying 1:n.iter)
+n.sim <- 5 #nr of simulations per iteration value
 p.miss <- c(.05, .25, .5, .75, .95)
 true.effect <- 2 #regression coefficient to be estimated
 
@@ -107,6 +106,6 @@ results <-
                                                                         by = c("t", "p"),
                                                                         suffix = c("", ".UL"))
 
-# # save results
-# save(out, file = "3.Thesis/1.SimulationStudy/Results/raw.Rdata")
-# save(results, file = "3.Thesis/1.SimulationStudy/Results/complete.Rdata")
+# save results
+save(out, file = "3.Thesis/1.SimulationStudy/Results/raw.Rdata")
+save(results, file = "3.Thesis/1.SimulationStudy/Results/complete.Rdata")

### 2020-04-24 3.Thesis/1.SimulationStudy/2.CreateFigures.R

@@ -7,8 +7,8 @@ library(dplyr)
 library(ggplot2)
 library(patchwork)
 
-# if necessary, load data
-load("3.Thesis/1.SimulationStudy/Results/results.Rdata")
+# # if necessary, load data
+# load("3.Thesis/1.SimulationStudy/Results/complete.Rdata")
 
 # set default for plot layout
 theme_update(
@@ -28,18 +28,26 @@ theme_update(
 ####################
 
 # Univariate: mean estimate
-mean_bias <- results %>% ggplot(aes(x = t, y = bias.mean.X1, color = as.factor(p*100))) +
+mean_bias <- results %>% ggplot(aes(x = t, y = bias.mean.Y, color = as.factor(p*100))) +
   geom_hline(yintercept = 0,
              color = "grey",
              lwd = 2) +
   geom_point(size = .75) +
+  # geom_errorbar(
+  #       aes(x = t, ymin = bias.mean.Y.LL, ymax = bias.mean.Y.UL),
+  #       width = .2,
+  #       color = "grey"
+  #     ) +
   geom_line() +
   xlab("Number of iterations") +
-  ylab(expression(paste("Bias in ", hat(mu)))) +
-  labs(colour = "Missingness (%)")
+  # ylab(expression(paste("Bias in ", hat(mu)))) +
+  ylab(expression(paste(bar(Q) - Q))) +
+  ggtitle("Bias in \bar{Q}; Q = \mu_Y = 25.81") +
+  labs(colour = "Missingness (%)") 
+
 
 # Univariate: convergence chain mean
-mean_Rh <- dat %>% ggplot(aes(x = t, y = R.mean.X1, color = as.factor(p*100))) +
+mean_Rh <- results %>% ggplot(aes(x = t, y = max.r.hat.chain.mean.Y, color = as.factor(p*100))) +
   geom_hline(yintercept = 1,
              color = "grey",
              lwd = 2) +
@@ -49,7 +57,7 @@ mean_Rh <- dat %>% ggplot(aes(x = t, y = R.mean.X1, color = as.factor(p*100))) +
   ylab(expression(paste(widehat(R)))) +
   labs(colour = "Missingness (%)")
 
-mean_AC <- dat %>% ggplot(aes(x = t, y = AC.mean.X1, color = as.factor(p*100))) +
+mean_AC <- results %>% ggplot(aes(x = t, y = ac.max.chain.mean.Y, color = as.factor(p*100))) +
   geom_hline(yintercept = 0,
[…]

### 2020-04-24 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
Edit figures in Results section manuscript.

--- 2020-04-24 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(fig.width = 4.8, fig.height = 3.6, echo = FALSE, fig.align = "center") 
+knitr::opts_chunk$set(fig.width = 4.8, echo = FALSE, fig.align = "center") 
+# knitr::opts_chunk$set(fig.width = 4.8, fig.height = 3.6, echo = FALSE, fig.align = "center")
+
+# packages
+library(ggplot2)
+library(patchwork)
+
+# results and functions
+load("../Results/complete.Rdata")
+source("../2.CreateFigures.R")
+
+# default graphing behavior
+theme_update(
+  plot.title = element_text(hjust = 0.5),
+  plot.subtitle = element_text(hjust = 0.5),
+  panel.border = element_blank(),
+  panel.grid.major = element_blank(),
+  panel.grid.minor = element_blank(),
+  panel.background = element_blank(),
+  axis.line = element_line(colour = "black"),
+  legend.key = element_blank(),
+  legend.position = "bottom",
+  legend.margin = margin(0,0,0,0),
+  # adjust text size
+  text = element_text(size = 8),
+  title = element_text(size = 8)
+  # axis.title = element_text(size = 8),
+  # axis.text = element_text(size = 7),
+  # legend.text = element_text(size = 7)
+)
 ```
@@
 
-```{r diagram, out.width="\\linewidth", include=TRUE, fig.align="center", fig.cap="Scheme of the main steps in multiple imputation."}
+```{r diagram, out.width="\\linewidth", include=TRUE, fig.cap="Scheme of the main steps in multiple imputation."}
 knitr::include_graphics("./images/diagram.pdf") #adapted from [@buur18] par. 1.4.1.
@@
 
-```{r non-conv, out.width='.49\\linewidth', fig.width=3.5, fig.height=3,fig.show='hold',fig.cap="Pathological non-convergence versus typical convergence."}
-load("./images/nonc.Rdata")
-load("./images/conv.Rdata")
+```{r non-conv, out.width='.49\\linewidth', fig.width=3.5, fig.height=3,fig.show='hold',fig.cap="Typical convergence versus pathological non-convergence."}
+load("./images/nonc_plot.Rdata")
+load("./images/conv_plot.Rdata")
 par(mar = c(2, 2, 0.1, 0.1))
+plot(conv)
 plot(nonc)
-plot(conv)
 
+```
+
+```{r diagnostics, fig.cap="Convergence diagnostics pathological non-convergence versus typical convergence."}
+load("./images/diagnostics_plot.Rdata")
+diagnostics_plot
 ```
@@
 
-```{r load-res, include = FALSE}
-# load("../Results/test.Rdata")
-# source("../2.CreateFigures.R")
+```{r mean, fig.height = 6.8, fig.cap="Convergence diagnostics chain mean."}
+mean_Rh + mean_AC + mean_bias + plot_layout(guides = "collect", ncol = 1) + plot_annotation(title = "Performance of convergence diagnostics", subtitle = bquote(theta ~ "= chain mean of Y in" ~ y[imp][",m"] ~ "; Q = " ~ mu[Y] ~ "= 25.81"))
 ```
 
-```{r res-univ}
-# mean_bias + plot_annotation(title = "Bias in mean estimate, convergence of chain means")
-# mean_Rh 
-# mean_ACF + plot_annotation(title = "Dotted lines are old $AC$ computation, solid are new.")
+```{r sd, fig.height = 6.8, fig.cap="Convergence diagnostics chain variance."}
+var_Rh + var_AC + sd_bias + plot_layout(guides = "collect", ncol = 1) + plot_annotation(title = "Performance of convergence diagnostics", subtitle = bquote(theta ~ "= chain variance of Y in" ~ y[imp][",m"] ~ "; Q = " ~ sigma[Y] ~ "= 11.32"))
+```
+```{r est, fig.height = 6.8, fig.cap="Convergence diagnostics regression coefficient."}
+est_Rh + est_AC + est_bias + plot_layout(guides = "collect", ncol = 1) + plot_annotation(title = "Performance of convergence diagnostics", subtitle = bquote(theta ~ "= " ~ hat(beta)[2] ~ "in {" ~ y[obs] ~ "," ~ y[imp][",m"] ~ "}" ~ "; Q = " ~ beta[2] ~ "= 2.06"))
+```
+
+```{r pred, fig.height = 6.8, fig.cap="Convergence diagnostics regression coefficient."}
+PCA_Rh + PCA_AC + Rsq_bias + plot_layout(guides = "collect", ncol = 1) + plot_annotation(title = "Performance of convergence diagnostics", subtitle = bquote(theta ~ "= first eigenvalue of" ~ Sigma ~ "in {" ~ y[obs] ~ "," ~ y[imp][",m"] ~ "}" ~ "; Q = explained variance = 19.25"))
+
 ```

### 2020-04-24 3.Thesis/1.SimulationStudy/2.CreateFigures.R

@@ -33,28 +33,23 @@ mean_bias <- results %>% ggplot(aes(x = t, y = bias.mean.Y, color = as.factor(p*
              color = "grey",
              lwd = 2) +
   geom_point(size = .75) +
-  # geom_errorbar(
-  #       aes(x = t, ymin = bias.mean.Y.LL, ymax = bias.mean.Y.UL),
-  #       width = .2,
-  #       color = "grey"
-  #     ) +
   geom_line() +
   xlab("Number of iterations") +
-  # ylab(expression(paste("Bias in ", hat(mu)))) +
-  ylab(expression(paste(bar(Q) - Q))) +
-  ggtitle("Bias in \bar{Q}; Q = \mu_Y = 25.81") +
+  ylab(bquote("Bias ("~bar(Q)-Q~")")) +
+  #ggtitle(bquote("A) Bias in " ~ bar(mu[Y]) ~ "; " ~ mu[Y] ~ "= 25.81")) +
   labs(colour = "Missingness (%)") 
 
 
 # Univariate: convergence chain mean
-mean_Rh <- results %>% ggplot(aes(x = t, y = max.r.hat.chain.mean.Y, color = as.factor(p*100))) +
+mean_Rh <- results %>% ggplot(aes(x = t, y = r.hat.max.chain.mean.Y, color = as.factor(p*100))) +
   geom_hline(yintercept = 1,
              color = "grey",
              lwd = 2) +
   geom_point(size = .75, na.rm = TRUE) +
   geom_line(na.rm = TRUE) +
-  xlab("Number of iterations") +
-  ylab(expression(paste(widehat(R)))) +
+  xlab("") +
+  ylab(bquote(widehat(R)~" ("~theta~")")) +
+  #ggtitle(bquote("B) " ~ widehat(R) ~ "of chain means" ~ y[imp][",Y"])) +
   labs(colour = "Missingness (%)")
 
 mean_AC <- results %>% ggplot(aes(x = t, y = ac.max.chain.mean.Y, color = as.factor(p*100))) +
@@ -63,9 +58,13 @@ mean_AC <- results %>% ggplot(aes(x = t, y = ac.max.chain.mean.Y, color = as.fac
              lwd = 2) +
   geom_point(size = .75, na.rm = TRUE) +
   geom_line(na.rm = TRUE) +
-  xlab("Number of iterations") +
-  ylab("Autocorrelation") +
+  xlab("") +
+  ylab(bquote("AC ("~theta~")")) +
[…]

### 2020-04-24 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
Manuscript versie voor feedback Gerko

@@
 ---
-title: Missing the Point$\colon$ Non-Convergence in Iterative Imputation Procedures
+title: "Missing the Point: Non-Convergence in Iterative Imputation Algorithms"
 runninghead: Oberman
@@
 <!-- (**or** Everybody who works with person data is likely to run into a missing data problem.) -->
-Missing data problems are ubiquitous and pervasive in the social and behavioral sciences.
-If a dataset contains just one missing value for a variable of interest, statistical inferences are undefined and will not produce any result.
-To circumvent this, incomplete observations are ignored by default in many statistical packages (i.e., list-wise deletion is employed).
+Anyone who analyzes person-data may run into a missing data problem.
+Missing data is not only ubiquitous, but also tedious.
+If a dataset contains just one incomplete observation, statistical inferences are undefined and will not produce any results.
+To circumvent this, many statistical packages employ list-wise deletion by default (i.e., ignoring incomplete observations).
 Unfortunately, this *ad hoc* solution may yield wildly invalid results [@buur18].
-An alternative is to 'impute' (i.e., fill in) every missing value in an incomplete dataset.
-With imputation techniques, one or several completed datasets are created, on which statistical inferences can be performed.
-The case with several completed datasets is known as 'multiple imputation' [MI; @rubin76], and requires an additional step to pool the results across imputations **see figure 1??**.
+An alternative is to 'impute' (i.e., fill in) the missing values in the incomplete observations.
+Subsequently, statistical inferences can be performed on the completed dataset.
+By repeating this process several times, a distribution of plausible results may be obtained, which reflects the uncertainty in the data due to missingness.
+This technique is known as 'multiple imputation' [MI; @rubin76].
+MI yields unbiased and confidence valid estimates of the true---but missing---data inference under many circumstances [@buur18].
 
-**Move this after notation??** Figure \ref{fig:diagram} provides an overview of the steps involved with MI---from incomplete data, to $M$ completed datasets, to $M$ estimated quantities of interest ($\hat{Q}$s), to a single pooled estimate $\bar{Q}$.
+**Move this after terminology??** Figure \ref{fig:diagram} provides an overview of the steps involved with MI---from incomplete data, to $M$ completed datasets, to $M$ estimated quantities of interest ($\hat{Q}$s), to a single pooled estimate $\bar{Q}$.
 Missing data in $y$ is 'imputed' (i.e., filled in) $M$ times.
@@
 
-By applying Rubin's rules, the pooled result will reflect the uncertainty in the data due to missingness.
-Under many circumstances, this yields an unbiased and confidence valid estimate of the inference on the true--but missing--data [@buur18].
 A popular method to obtain imputations is to use the 'Multiple Imputation by Chained Equations' algorithm, shorthand 'MICE'[@mice].
 MICE is an iterative algorithmic procedure to draw imputations from the posterior predictive distribution of the missing values.
-This introduces a potential threat to the validity of the imputations: what if the algorithm has not converged?
+This introduces a potential threat to the validity of the imputations: What if the algorithm has not converged?
 Are the implications then to be trusted?
@@
 
-The recommended practice to monitor convergence is to visually inspect imputations for signs of non-convergence.
-This practice is insufficient on two counts: 1) it may be challenging to the untrained eye, and 2) only severely pathological cases of non-convergence may be diagnosed [@buur18, $\S$ 6.5.2].
+Currently, the recommended practice for evaluating convergence is to visually inspect imputations for signs of non-convergence.
+This method is insufficient on two counts: 1) it may be challenging to the untrained eye, and 2) only severely pathological cases of non-convergence may be diagnosed [@buur18, $\S$ 6.5.2].
 Therefore, a quantitative, diagnostic evaluation of convergence would be preferred.
 
-Iterative imputation algorithms such as MICE are Markov chain Monte Carlo (MCMC) methods, which means that convergence is not from a scalar to a point but from one distribution to another.
+Monitoring convergence of iterative imputation algorithms diagnostically is challenging.
+Iterative imputation algorithms such as MICE are Markov chain Monte Carlo (MCMC) methods.
+In MCMC methods, convergence is not from a scalar to a point but from one distribution to another.
 The values generated by the algorithm (e.g., imputed values) will vary even after convergence.
-Therefore, the aim of convergence diagnostics for MCMC methods should not be to establish the point at which convergence is reached, but to monitor signs of  non-convergence.
-Several convergence diagnostics for MCMC methods exist, but it is not known whether these are appropriate for MICE.
-
+Therefore, the aim of convergence diagnostics for MCMC methods is not to establish the point at which convergence is reached, but to monitor signs of non-convergence.
+Several of such diagnostics exist for MCMC methods, but it is not known whether these are appropriate for MICE.
 <!-- Therefore we can only diagnose non-convergence, not convergence [@hoff09]. "a weak diagnostics is better than no diagnostic at all" [@cowl96]. -->
 
-### Research question: how and when to diagnose non convergence?
+In this paper we investigate how non-convergence in iterative imputation algorithms may be diagnosed, how well these methods perform, and at which point convergence may safely be assumed.
+For reasons of brevity, we only focus on the MI algorithm implemented in the world-leading MI software: the `R` [@R] package `mice` [@mice].
+The convergence properties of this MI algorithm are investigated through model-based simulation.
+The results of this simulation study are guidelines for assessing convergence of MI algorithms, which will aid applied researchers in drawing valid inference from incomplete datasets.
 
-Sub-questions: How can non-convergence be diagnosed?
+**Include sub-questions?:** How can non-convergence be diagnosed?
 Are common MCMC non-convergence diagnostics appropriate for MICE?
@@
 Are the default number of iterations sufficient (i.e., 5 in mice, 10 in SPSS and Stata, 30 in mi)?
-**(These are part of the methods, not RQ per se: What are the effects of continued iterations on estimates, predictions and inferences?
+
+<!-- **(These are part of the methods, not RQ per se: What are the effects of continued iterations on estimates, predictions and inferences?
 And finally, do the answers differ with varying missingness proportions?
-That is, we vary the nr of iterations and the missingness proportion because we assume that the algorithm has not reached convergence at $t=1$, and performs worse with moderate to high missingness (not so much with little or a LOT of missingness).)**
+That is, we vary the nr of iterations and the missingness proportion because we assume that the algorithm has not reached convergence at $t=1$, and performs worse with moderate to high missingness (not so much with little or a LOT of missingness).)** -->
 
@@
 -->
-
-For reasons of brevity, we only focus on the MI algorithm implemented in the world-leading MI software: the `R` [@R] package `mice` [@mice].
-The convergence properties of this MI algorithm are investigated through model-based simulation.
-The results of this simulation study are guidelines for assessing convergence of MI algorithms, which will aid applied researchers in drawing valid inference from incomplete datasets.
 
@@
 
-## Notation
+## Terminology
 
@@
 
-# Convergence diagnostics
+# How may non-convergence be diagnosed?
 
-The two challenges for convergence of iterative algorithms are mixing and stationarity [@gelm13], see Figure \ref{fig:non-conv}.
+The two challenges for convergence of iterative algorithms are mixing and stationarity [@gelm13].
 Without mixing, imputation chains do not intermingle nicely.
-Without stationarity, there is trending within imputation chains.
-While the current practice is to inspect traceplots for signs of non-mixing and non-stationarity, diagnostic evaluation of these two challenges would be preferred.
+Without stationarity, there is trending within imputation chains, see Figure \ref{fig:non-conv}.
+**Explain what we see, namely example by @buur18 reproduced, showing the traceplots of chain means for some variable.
+The first plot is typical convergence of MICE, the second is pathological non-convergence because of a mis-specified imputation model.
+Each line is an imputation.
+In the first plot, the lines intermingle nicely and there is little trending.
+In the second, there is a lot of trending and the lines do not cross much.
+Importantly, the chain means at the last iteration (the imputed value per $m$) are very different between the two plots.
+The algorithm with the mis-specified model yields imputed values that are on average a magnitude two larger than those of the typically converged algorithm.
+This shows the importance of reaching convergend states in iterative imputation algorithms.**
+
 Non-stationarity within chains may be diagnosed with e.g., autocorrelation [$AC$; @scha97; @gelm13], numeric standard error ['MC error'; @gewe92], or Raftery and Lewis's [-@raft91] procedure to determine the effect of trending on the precision of estimates.
@@
 
-```
-
-```{r diagnostics, fig.cap="Convergence diagnostics pathological non-convergence versus typical convergence."}
-load("./images/diagnostics_plot.Rdata")
-diagnostics_plot
 ```
@@
 
-We can interpret $AC$-values as a measure of stationarity. If $AC$-values are close to zero, there is no dependence between subsequent samples within imputation chains.
-
-Positive $AC$-values indicate recurrence.
+We can interpret $AC$-values as a measure of stationarity.
+If $AC$-values are close to zero, there is no dependence between subsequent samples within imputation chains.
+Negative $AC$-values indicate divergence within imputation chains.
+**Subsequent sampled values within each imputation chain are less alike.** Positive $AC$-values indicate recurrence.
 If $\theta$-values of subsequent iterations are similar, trending may occur.
@@
 As convergence diagnostic, the interest is therefore in positive $AC$-values.
-Moreover, the magnitude of $AC$-values may be evaluated statistically, but that is outside of this note's scope.
+**Maybe remove:** Moreover, the magnitude of $AC$-values may be evaluated statistically, but that is outside of this note's scope.
 
-Negative $AC$-values indicate divergence within imputation chains. **Subsequent sampled values within each imputation chain are less alike.**
+## In practice/Answer
+**To assess whether $\widehat{R}$ and $AC$ may be appropriate non-convergence diagnostics for MICE algorithms, the following condtion must hold when applied on the two algorithms plotted above: the methods indicate worse performance for the mis-specified model with pathological non-convergence (i.e., higher $\widehat{R}$- and $AC$-values than the typical performance).
+And additionally the methods should reflect the increasing convergence in the typical convergence situation as the number of iterations goes up.**
 
-In short, convergence is reached when there is no dependency between subsequent iterations of imputation chains ($AC = 0$), and chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm ($\widehat{R} = 1$).
+**In figure \ref{fig:diagnostics}A (**add panel labels**), the chain means are plotted again, now together.
+So are the non-convergence diagnostics.
+Panel B shows $\widehat{R}$ as computed by implementing @veht19 's recommendations.
+As required, $\widehat{R}$ indicates less signs of non-convergence as the number of iterations goes up in the typical convergence situation.
+The superior performance of the typical convergence over the pathological non-convergnce is less prominent, even flipped for $t<4$.
+Panel C displays the $AC$  as computed with the R function stats::acf().
+When we look at this panel, we must conclude that something weird is going on.
+The $AC$-values indicate equal performance (upto $t=5$) for the typical convergence and the pathological non-convergnce, while there is obvious trending in the latter.
+Moreover, the best convergence (as indicated by the lowest $AC$-value) is observed at $t=2$, but looking at the chain means in panel A, there should be some signs of trending upto iteration number seven.
+After consulting the documentation on stats::acf(), we conclude that this implementation of $AC$ is not suitable for iterative imputation algorithms.
+There is a correction factor for a mathematical shortcut that works in the limit.
+According to ...
+the function works when the number of iterations > 50.
+The default number of iterations in mice, however,  is 5.
+Therefore, we compute $AC$ manually, see panel D.
+The AC values in this plot do meet the requirements (...) and will therefore be used in the simulation study.**
+
+```{r diagnostics, fig.cap="Convergence diagnostics pathological non-convergence versus typical convergence."}
+load("./images/diagnostics_plot.Rdata")
+diagnostics_plot
+```
 
 ## Simulation Hypothesis
+
+Convergence is reached when there is no dependency between subsequent iterations of imputation chains ($AC = 0$), and chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm ($\widehat{R} = 1$).
+We expect that a completely convergenced state will not be reached.
 
@@
 **Added later: High $AC$-values are implausible in MI procedures.
-That is, the randomness induced by the MI algorithm effectively mitigates the risk of dependency within chains.**
+That is, the randomness induced by the MI algorithm effectively mitigates the risk of dependency within chains.
+But this doesn't seem the case in practice??**
 
@@
 A user-specified scalar summary, however, is not universal to all complete data problems.
-Therefore, as inspired by **someone** (2002), we will use introduce a $\theta$ that summarizes the multivariate state space of the algorithm.
+Therefore, as inspired by [@mack03], we will use introduce a $\theta$ that summarizes the multivariate state space of the algorithm.
 Namely, the first eigenvalue of the variance-covariance matrix of the $M$ completed datasets.
@@
 
-For empirical researchers: 1) Check trace plots for pathological non convergence and adjust imputation model if necessary.
+**For empirical researchers: 1) Check trace plots for pathological non convergence and adjust imputation model if necessary.
 2) use $\widehat{R}$ wait 1.1 ash threshold and autocorrelation with **?** as threshold .
@@
 Compute $\widehat{R}$ and autocorrelation values for this scalar summary.
+**
 
@@
 
-Pmm,
-M(N)AR,
-Empirical data,
-Vignette,
-ShinyMICE,
-AC across imputations, not iterations (unique for imp data??)
+**Pmm, M(N)AR, Empirical data, Vignette, ShinyMICE, AC across imputations, not iterations (unique for imp data??)**
 
-If I don't include it in my study, also: Significance of $AC$ values; Estimate (PCA or reg. coeff.) versus $\widehat{R}$ and $AC$ on the estimate
+**If I don't include it in my study, also: Significance of $AC$ values**
 

### 2020-04-24 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Edit figures in Results section manuscript.
Manuscript versie voor feedback Gerko

@@
 
-\title{Missing the Point\(\colon\) Non-Convergence in Iterative Imputation
-Procedures}
+\title{Missing the Point: Non-Convergence in Iterative Imputation Algorithms}
 
@@
 
-Missing data problems are ubiquitous and pervasive in the social and
-behavioral sciences. If a dataset contains just one missing value for a
-variable of interest, statistical inferences are undefined and will not
-produce any result. To circumvent this, incomplete observations are
-ignored by default in many statistical packages (i.e., list-wise
-deletion is employed). Unfortunately, this \emph{ad hoc} solution may
-yield wildly invalid results \citep{buur18}. An alternative is to
-`impute' (i.e., fill in) every missing value in an incomplete dataset.
-With imputation techniques, one or several completed datasets are
-created, on which statistical inferences can be performed. The case with
-several completed datasets is known as `multiple imputation'
-\citep[MI;][]{rubin76}, and requires an additional step to pool the
-results across imputations \textbf{see figure 1??}.
+Anyone who analyzes person-data may run into a missing data problem.
+Missing data is not only ubiquitous, but also tedious. If a dataset
+contains just one incomplete observation, statistical inferences are
+undefined and will not produce any results. To circumvent this, many
+statistical packages employ list-wise deletion by default (i.e.,
+ignoring incomplete observations). Unfortunately, this \emph{ad hoc}
+solution may yield wildly invalid results \citep{buur18}. An alternative
+is to `impute' (i.e., fill in) the missing values in the incomplete
+observations. Subsequently, statistical inferences can be performed on
+the completed dataset. By repeating this process several times, a
+distribution of plausible results may be obtained, which reflects the
+uncertainty in the data due to missingness. This technique is known as
+`multiple imputation' \citep[MI;][]{rubin76}. MI yields unbiased and
+confidence valid estimates of the true---but missing---data inference
+under many circumstances \citep{buur18}.
 
-\textbf{Move this after notation??} Figure \ref{fig:diagram} provides an
-overview of the steps involved with MI---from incomplete data, to \(M\)
-completed datasets, to \(M\) estimated quantities of interest
+\textbf{Move this after terminology??} Figure \ref{fig:diagram} provides
+an overview of the steps involved with MI---from incomplete data, to
+\(M\) completed datasets, to \(M\) estimated quantities of interest
 (\(\hat{Q}\)s), to a single pooled estimate \(\bar{Q}\). Missing data in
@@
 
-By applying Rubin's rules, the pooled result will reflect the
-uncertainty in the data due to missingness. Under many circumstances,
-this yields an unbiased and confidence valid estimate of the inference
-on the true--but missing--data \citep{buur18}. A popular method to
-obtain imputations is to use the `Multiple Imputation by Chained
-Equations' algorithm, shorthand `MICE'\citep{mice}. MICE is an iterative
-algorithmic procedure to draw imputations from the posterior predictive
-distribution of the missing values. This introduces a potential threat
-to the validity of the imputations: what if the algorithm has not
-converged? Are the implications then to be trusted? And can we rely on
-the inference obtained on the completed data? These are all open
-questions, because the convergence properties of iterative imputation
-algorithms have not been systematically studied \citep{buur18}.
-Moreover, there is no scientific consensus on the convergence properties
-of MI algorithms \citep{taka17}. Some default MICE techniques (e.g.,
-`predictive mean modeling') might not yield converged states at all
-\citep{murr18}. Therefore, algorithmic convergence should be monitored
-carefully.
+A popular method to obtain imputations is to use the `Multiple
+Imputation by Chained Equations' algorithm, shorthand
+`MICE'\citep{mice}. MICE is an iterative algorithmic procedure to draw
+imputations from the posterior predictive distribution of the missing
+values. This introduces a potential threat to the validity of the
+imputations: What if the algorithm has not converged? Are the
+implications then to be trusted? And can we rely on the inference
+obtained on the completed data? These are all open questions, because
+the convergence properties of iterative imputation algorithms have not
+been systematically studied \citep{buur18}. Moreover, there is no
+scientific consensus on the convergence properties of MI algorithms
+\citep{taka17}. Some default MICE techniques (e.g., `predictive mean
+modeling') might not yield converged states at all \citep{murr18}.
+Therefore, algorithmic convergence should be monitored carefully.
 
-The recommended practice to monitor convergence is to visually inspect
-imputations for signs of non-convergence. This practice is insufficient
-on two counts: 1) it may be challenging to the untrained eye, and 2)
-only severely pathological cases of non-convergence may be diagnosed
-\citep[\(\S\) 6.5.2]{buur18}. Therefore, a quantitative, diagnostic
-evaluation of convergence would be preferred.
+Currently, the recommended practice for evaluating convergence is to
+visually inspect imputations for signs of non-convergence. This method
+is insufficient on two counts: 1) it may be challenging to the untrained
+eye, and 2) only severely pathological cases of non-convergence may be
+diagnosed \citep[\(\S\) 6.5.2]{buur18}. Therefore, a quantitative,
+diagnostic evaluation of convergence would be preferred.
 
-Iterative imputation algorithms such as MICE are Markov chain Monte
-Carlo (MCMC) methods, which means that convergence is not from a scalar
-to a point but from one distribution to another. The values generated by
-the algorithm (e.g., imputed values) will vary even after convergence.
-Therefore, the aim of convergence diagnostics for MCMC methods should
-not be to establish the point at which convergence is reached, but to
-monitor signs of non-convergence. Several convergence diagnostics for
-MCMC methods exist, but it is not known whether these are appropriate
-for MICE.
+Monitoring convergence of iterative imputation algorithms diagnostically
+is challenging. Iterative imputation algorithms such as MICE are Markov
+chain Monte Carlo (MCMC) methods. In MCMC methods, convergence is not
+from a scalar to a point but from one distribution to another. The
+values generated by the algorithm (e.g., imputed values) will vary even
+after convergence. Therefore, the aim of convergence diagnostics for
+MCMC methods is not to establish the point at which convergence is
+reached, but to monitor signs of non-convergence. Several of such
+diagnostics exist for MCMC methods, but it is not known whether these
+are appropriate for MICE.
 
-\hypertarget{research-question-how-and-when-to-diagnose-non-convergence}{%
-\subsubsection{Research question: how and when to diagnose non
-convergence?}\label{research-question-how-and-when-to-diagnose-non-convergence}}
+In this paper we investigate how non-convergence in iterative imputation
+algorithms may be diagnosed, how well these methods perform, and at
+which point convergence may safely be assumed. For reasons of brevity,
+we only focus on the MI algorithm implemented in the world-leading MI
+software: the \texttt{R} \citep{R} package \texttt{mice} \citep{mice}.
+The convergence properties of this MI algorithm are investigated through
+model-based simulation. The results of this simulation study are
+guidelines for assessing convergence of MI algorithms, which will aid
+applied researchers in drawing valid inference from incomplete datasets.
 
-Sub-questions: How can non-convergence be diagnosed? Are common MCMC
-non-convergence diagnostics appropriate for MICE? And if so, which
-threshold should be used to diagnose non-convergence? How many
+\textbf{Include sub-questions?:} How can non-convergence be diagnosed?
+Are common MCMC non-convergence diagnostics appropriate for MICE? And if
+so, which threshold should be used to diagnose non-convergence? How many
 iterations are sufficient/needed to be able to diagnose non-convergence?
 Are the default number of iterations sufficient (i.e., 5 in mice, 10 in
-SPSS and Stata, 30 in mi)? \textbf{(These are part of the methods, not
-RQ per se: What are the effects of continued iterations on estimates,
-predictions and inferences? And finally, do the answers differ with
-varying missingness proportions? That is, we vary the nr of iterations
-and the missingness proportion because we assume that the algorithm has
-not reached convergence at \(t=1\), and performs worse with moderate to
-high missingness (not so much with little or a LOT of missingness).)}
+SPSS and Stata, 30 in mi)?
 
-For reasons of brevity, we only focus on the MI algorithm implemented in
-the world-leading MI software: the \texttt{R} \citep{R} package
-\texttt{mice} \citep{mice}. The convergence properties of this MI
-algorithm are investigated through model-based simulation. The results
-of this simulation study are guidelines for assessing convergence of MI
-algorithms, which will aid applied researchers in drawing valid
-inference from incomplete datasets.
-
-\hypertarget{notation}{%
-\subsection{Notation}\label{notation}}
+\hypertarget{terminology}{%
+\subsection{Terminology}\label{terminology}}
 
@@
 
-\hypertarget{convergence-diagnostics}{%
-\section{Convergence diagnostics}\label{convergence-diagnostics}}
+\hypertarget{how-may-non-convergence-be-diagnosed}{%
+\section{How may non-convergence be
+diagnosed?}\label{how-may-non-convergence-be-diagnosed}}
 
 The two challenges for convergence of iterative algorithms are mixing
-and stationarity \citep{gelm13}, see Figure \ref{fig:non-conv}. Without
-mixing, imputation chains do not intermingle nicely. Without
-stationarity, there is trending within imputation chains. While the
-current practice is to inspect traceplots for signs of non-mixing and
-non-stationarity, diagnostic evaluation of these two challenges would be
-preferred. Non-stationarity within chains may be diagnosed with e.g.,
+and stationarity \citep{gelm13}. Without mixing, imputation chains do
+not intermingle nicely. Without stationarity, there is trending within
+imputation chains, see Figure \ref{fig:non-conv}. \textbf{Explain what
+we see, namely example by \citet{buur18} reproduced, showing the
+traceplots of chain means for some variable. The first plot is typical
+convergence of MICE, the second is pathological non-convergence because
+of a mis-specified imputation model. Each line is an imputation. In the
+first plot, the lines intermingle nicely and there is little trending.
+In the second, there is a lot of trending and the lines do not cross
+much. Importantly, the chain means at the last iteration (the imputed
+value per \(m\)) are very different between the two plots. The algorithm
+with the mis-specified model yields imputed values that are on average a
+magnitude two larger than those of the typically converged algorithm.
+This shows the importance of reaching convergend states in iterative
+imputation algorithms.}
+
+Non-stationarity within chains may be diagnosed with e.g.,
 autocorrelation \citep[\(AC\);][]{scha97, gelm13}, numeric standard
@@
 \caption{Typical convergence versus pathological non-convergence.}\label{fig:non-conv}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics{manuscript_files/figure-latex/diagnostics-1}
-
-}
-
-\caption{Convergence diagnostics pathological non-convergence versus typical convergence.}\label{fig:diagnostics}
 \end{figure}
@@
 \(AC\)-values are close to zero, there is no dependence between
-subsequent samples within imputation chains.
+subsequent samples within imputation chains. Negative \(AC\)-values
+indicate divergence within imputation chains. \textbf{Subsequent sampled
+values within each imputation chain are less alike.} Positive
+\(AC\)-values indicate recurrence. If \(\theta\)-values of subsequent
+iterations are similar, trending may occur. Negative \(AC\)-values show
+no threat to the stationarity component of convergence. On the contrary
+even---negative \(AC\)-values indicate that \(\theta\)-values of
+subsequent iterations diverge from one-another, which may increase the
+variance of \(\theta\) and speed up convergence. As convergence
+diagnostic, the interest is therefore in positive \(AC\)-values.
+\textbf{Maybe remove:} Moreover, the magnitude of \(AC\)-values may be
+evaluated statistically, but that is outside of this note's scope.
 
-Positive \(AC\)-values indicate recurrence. If \(\theta\)-values of
-subsequent iterations are similar, trending may occur. Negative
-\(AC\)-values show no threat to the stationarity component of
-convergence. On the contrary even---negative \(AC\)-values indicate that
-\(\theta\)-values of subsequent iterations diverge from one-another,
-which may increase the variance of \(\theta\) and speed up convergence.
-As convergence diagnostic, the interest is therefore in positive
-\(AC\)-values. Moreover, the magnitude of \(AC\)-values may be evaluated
-statistically, but that is outside of this note's scope.
+\hypertarget{in-practiceanswer}{%
+\subsection{In practice/Answer}\label{in-practiceanswer}}
 
-Negative \(AC\)-values indicate divergence within imputation chains.
-\textbf{Subsequent sampled values within each imputation chain are less
-alike.}
+\textbf{To assess whether \(\widehat{R}\) and \(AC\) may be appropriate
+non-convergence diagnostics for MICE algorithms, the following condtion
+must hold when applied on the two algorithms plotted above: the methods
+indicate worse performance for the mis-specified model with pathological
+non-convergence (i.e., higher \(\widehat{R}\)- and \(AC\)-values than
+the typical performance). And additionally the methods should reflect
+the increasing convergence in the typical convergence situation as the
+number of iterations goes up.}
 
-In short, convergence is reached when there is no dependency between
-subsequent iterations of imputation chains (\(AC = 0\)), and chains
-intermingle such that the only difference between the chains is caused
-by the randomness induced by the algorithm (\(\widehat{R} = 1\)).
+\textbf{In figure \ref{fig:diagnostics}A (}add panel labels\textbf{),
+the chain means are plotted again, now together. So are the
+non-convergence diagnostics. Panel B shows \(\widehat{R}\) as computed
+by implementing \citet{veht19} 's recommendations. As required,
+\(\widehat{R}\) indicates less signs of non-convergence as the number of
+iterations goes up in the typical convergence situation. The superior
+performance of the typical convergence over the pathological
+non-convergnce is less prominent, even flipped for \(t<4\). Panel C
+displays the \(AC\) as computed with the R function stats::acf(). When
+we look at this panel, we must conclude that something weird is going
+on. The \(AC\)-values indicate equal performance (upto \(t=5\)) for the
+typical convergence and the pathological non-convergnce, while there is
+obvious trending in the latter. Moreover, the best convergence (as
+indicated by the lowest \(AC\)-value) is observed at \(t=2\), but
+looking at the chain means in panel A, there should be some signs of
+trending upto iteration number seven. After consulting the documentation
+on stats::acf(), we conclude that this implementation of \(AC\) is not
+suitable for iterative imputation algorithms. There is a correction
+factor for a mathematical shortcut that works in the limit. According to
+\ldots{} the function works when the number of iterations \textgreater{}
+50. The default number of iterations in mice, however, is 5. Therefore,
+we compute \(AC\) manually, see panel D. The AC values in this plot do
+meet the requirements (\ldots) and will therefore be used in the
+simulation study.}
+
+\begin{figure}
+
+{\centering \includegraphics{manuscript_files/figure-latex/diagnostics-1}
+
+}
+
+\caption{Convergence diagnostics pathological non-convergence versus typical convergence.}\label{fig:diagnostics}
+\end{figure}
 
@@
 \subsection{Simulation Hypothesis}\label{simulation-hypothesis}}
+
+Convergence is reached when there is no dependency between subsequent
+iterations of imputation chains (\(AC = 0\)), and chains intermingle
+such that the only difference between the chains is caused by the
+randomness induced by the algorithm (\(\widehat{R} = 1\)). We expect
+that a completely convergenced state will not be reached.
 
@@
 implausible in MI procedures. That is, the randomness induced by the MI
-algorithm effectively mitigates the risk of dependency within chains.}
+algorithm effectively mitigates the risk of dependency within chains.
+But this doesn't seem the case in practice??}
 
@@
 user-specified scalar summary, however, is not universal to all complete
-data problems. Therefore, as inspired by \textbf{someone} (2002), we
-will use introduce a \(\theta\) that summarizes the multivariate state
-space of the algorithm. Namely, the first eigenvalue of the
-variance-covariance matrix of the \(M\) completed datasets.
+data problems. Therefore, as inspired by \citep{mack03}, we will use
+introduce a \(\theta\) that summarizes the multivariate state space of
+the algorithm. Namely, the first eigenvalue of the variance-covariance
+matrix of the \(M\) completed datasets.
 
@@
 
+\begin{figure}
+
+{\centering \includegraphics{manuscript_files/figure-latex/mean-1}
+
+}
+
+\caption{Convergence diagnostics chain mean.}\label{fig:mean}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics{manuscript_files/figure-latex/sd-1}
+
+}
+
+\caption{Convergence diagnostics chain variance.}\label{fig:sd}
+\end{figure}
+\begin{figure}
+
+{\centering \includegraphics{manuscript_files/figure-latex/est-1}
+
+}
+
+\caption{Convergence diagnostics regression coefficient.}\label{fig:est}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics{manuscript_files/figure-latex/pred-1}
+
+}
+
+\caption{Convergence diagnostics regression coefficient.}\label{fig:pred}
+\end{figure}
+
 \hypertarget{univariate-estimates-and-convergence-diagnostics}{%
@@
 
-For empirical researchers: 1) Check trace plots for pathological non
-convergence and adjust imputation model if necessary. 2) use
-\(\widehat{R}\) wait 1.1 ash threshold and autocorrelation with
-\textbf{?} as threshold . Keep iterating until these thresholds are
-reached. 3) Do not use the R function ACF. Instead compute
-autocorrelations manually , see for example GitHub link. 4) Track your
-own scalar summary of interest. This is advanced but explained in Van
-Buren 2018. Compute \(\widehat{R}\) and autocorrelation values for this
-scalar summary.
+\textbf{For empirical researchers: 1) Check trace plots for pathological
+non convergence and adjust imputation model if necessary. 2) use
+\(\widehat{R}\) wait 1.1 ash threshold and autocorrelation with }?** as
+threshold . Keep iterating until these thresholds are reached. 3) Do not
+use the R function ACF. Instead compute autocorrelations manually , see
+for example GitHub link. 4) Track your own scalar summary of interest.
+This is advanced but explained in Van Buren 2018. Compute
+\(\widehat{R}\) and autocorrelation values for this scalar summary. **
 
@@
 
-Pmm, M(N)AR, Empirical data, Vignette, ShinyMICE, AC across imputations,
-not iterations (unique for imp data??)
+\textbf{Pmm, M(N)AR, Empirical data, Vignette, ShinyMICE, AC across
+imputations, not iterations (unique for imp data??)}
 
-If I don't include it in my study, also: Significance of \(AC\) values;
-Estimate (PCA or reg. coeff.) versus \(\widehat{R}\) and \(AC\) on the
-estimate
+\textbf{If I don't include it in my study, also: Significance of \(AC\)
+values}
 

### 2020-04-24 3.Thesis/1.SimulationStudy/Writeup/images/non-convergence.R

@@ -36,7 +36,7 @@ save(nonc, file = "3.Thesis/1.SimulationStudy/Writeup/images/nonc_plot.Rdata")
 
 # diagnostics on chain means of wgt (rows are iterations, columns are imputations)
 diagnostics1 <-
-  imp.bmi1$chainMean[3, ,] %>% cbind(patho = "Typical", convergence(., include_acf = TRUE), .)
+  imp.bmi1$chainMean[3, ,] %>% cbind(patho = "Pathological", convergence(., include_acf = TRUE), .)
 
 #####################
 # Regular convergence
@@ -63,7 +63,7 @@ save(conv, file = "3.Thesis/1.SimulationStudy/Writeup/images/conv_plot.Rdata")
 
 # diagnostics on chain means of wgt (rows are iterations, columns are imputations)
 diagnostics2 <-
-  imp.bmi2$chainMean[3, ,] %>% cbind(patho = "Pathological", convergence(., include_acf = TRUE), .)
+  imp.bmi2$chainMean[3, ,] %>% cbind(patho = "Typical", convergence(., include_acf = TRUE), .)
 
 #####################
 # Combine diagnostics

### 2020-04-27 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
After spellcheck

@@
 author:
-- name: H. I. Oberman
+  name: H. I. Oberman
   num: 1
 address:
-- num: 1
+  num: 1
   org: Department of Methodology and Statistics, Utrecht University, Utrecht, The Netherlands
@@
 Anyone who analyzes person-data may run into a missing data problem.
-Missing data is not only ubiquitous, but also tedious.
+Missing data is not only ubiquitous but also tedious.
 If a dataset contains just one incomplete observation, statistical inferences are undefined and will not produce any results.
@@
 
-<!-- (**Rewrite:** With FCS, an iterative algorithmic procedure is used to ...
-multiple imputation by chained equations ...
+<!-- (**Rewrite:** With FCS, an iterative algorithmic procedure is used to...
+multiple imputation by chained equations...
 "The values for imputing the missing data are drawn from the posterior predictive distribution of the missing data conditional on the observed data.") -->
@@
 
-In this paper we investigate how non-convergence in iterative imputation algorithms may be diagnosed, how well these methods perform, and at which point convergence may safely be assumed.
+In this paper, we investigate how non-convergence in iterative imputation algorithms may be diagnosed, how well these methods perform, and at which point convergence may safely be assumed.
 For reasons of brevity, we only focus on the MI algorithm implemented in the world-leading MI software: the `R` [@R] package `mice` [@mice].
@@
 <!-- This is repeated $m$ times.
-Imputed values are the ultimate samples of a `chain' of For each imputed value ($t = T$), a s for each missing datapoint in $y_{mis}$, .
+Imputed values are the ultimate samples of a `chain' of For each imputed value ($t = T$), a s for each missing datapoint in $y_{mis}$,.
 Each of the $m$ chains starts with an initial value, drawn randomly from $y_{obs}$.
@@
 The algorithm with the mis-specified model yields imputed values that are on average a magnitude two larger than those of the typically converged algorithm.
-This shows the importance of reaching convergend states in iterative imputation algorithms.**
+This shows the importance of reaching converged states in iterative imputation algorithms.**
 
@@
 147].
-In this study we only consider $AC$ at lag 1, i.e., the correlation between the $t^{th}$ and $t+1^{th}$ iteration of the same chain.
+In this study, we only consider $AC$ at lag 1, i.e., the correlation between the $t^{th}$ and $t+1^{th}$ iteration of the same chain.
 
@@
 Negative $AC$-values show no threat to the stationarity component of convergence.
-On the contrary even---negative $AC$-values indicate that $\theta$-values of subsequent iterations diverge from one-another, which may increase the variance of $\theta$ and speed up convergence.
+On the contrary even---negative $AC$-values indicate that $\theta$-values of subsequent iterations diverge from one another, which may increase the variance of $\theta$ and speed up convergence.
 As convergence diagnostic, the interest is therefore in positive $AC$-values.
@@
 ## In practice/Answer
-**To assess whether $\widehat{R}$ and $AC$ may be appropriate non-convergence diagnostics for MICE algorithms, the following condtion must hold when applied on the two algorithms plotted above: the methods indicate worse performance for the mis-specified model with pathological non-convergence (i.e., higher $\widehat{R}$- and $AC$-values than the typical performance).
+**To assess whether $\widehat{R}$ and $AC$ may be appropriate non-convergence diagnostics for MICE algorithms, the following condition must hold when applied on the two algorithms plotted above: the methods indicate worse performance for the mis-specified model with pathological non-convergence (i.e., higher $\widehat{R}$- and $AC$-values than the typical performance).
 And additionally the methods should reflect the increasing convergence in the typical convergence situation as the number of iterations goes up.**
@@
 As required, $\widehat{R}$ indicates less signs of non-convergence as the number of iterations goes up in the typical convergence situation.
-The superior performance of the typical convergence over the pathological non-convergnce is less prominent, even flipped for $t<4$.
+The superior performance of the typical convergence over the pathological non-convergence is less prominent, even flipped for $t<4$.
 Panel C displays the $AC$ as computed with the R function stats::acf().
 When we look at this panel, we must conclude that something weird is going on.
-The $AC$-values indicate equal performance (upto $t=5$) for the typical convergence and the pathological non-convergnce, while there is obvious trending in the latter.
-Moreover, the best convergence (as indicated by the lowest $AC$-value) is observed at $t=2$, but looking at the chain means in panel A, there should be some signs of trending upto iteration number seven.
+The $AC$-values indicate equal performance (up-to $t=5$) for the typical convergence and the pathological non-convergence, while there is obvious trending in the latter.
+Moreover, the best convergence (as indicated by the lowest $AC$-value) is observed at $t=2$, but looking at the chain means in panel A, there should be some signs of trending up-to iteration number seven.
 After consulting the documentation on stats::acf(), we conclude that this implementation of $AC$ is not suitable for iterative imputation algorithms.
 There is a correction factor for a mathematical shortcut that works in the limit.
-According to ...
+According to...
 the function works when the number of iterations > 50.
@@
 Convergence is reached when there is no dependency between subsequent iterations of imputation chains ($AC = 0$), and chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm ($\widehat{R} = 1$).
-We expect that a completely convergenced state will not be reached.
+We expect that a completely converged state will not be reached.
 
@@
 A user-specified scalar summary, however, is not universal to all complete data problems.
-Therefore, as inspired by [@mack03], we will use introduce a $\theta$ that summarizes the multivariate state space of the algorithm.
+Therefore, as inspired by [@mack03], we will propose a $\theta$ that summarizes the multivariate state space of the algorithm.
 Namely, the first eigenvalue of the variance-covariance matrix of the $M$ completed datasets.
@@
 Additionally, we will evaluate the estimated regression coefficients, the coverage rate of the CI95% of the regression estimates, and the CI length (**see definitions in old Methods**).
-**remove the term performance measure here and add an extra alinea to define performance measures as  bias in all estimands, and coverage rate and CI length of regression coefficients.**
+**remove the term performance measure here and add an extra paragraph to define performance measures as bias in all estimands, and coverage rate and CI length of regression coefficients.**
 
@@
 Roughly speaking this means that MICE is indeed not converged as $t=1$, and converges gradually as $t$ increases.
-The point at wich an additional iteration does not lead to an improvement of the estimates depends on the difficulty of the missingness problem---between XYZ and XYZ.
+The point at which an additional iteration does not lead to an improvement of the estimates depends on the difficulty of the missingness problem---between XYZ and XYZ.
 This means that the algorithm has converged sufficiently (**under the current specifications**).
@@
 
-Evaluation with the performance measures shows that are had an autocorrelation are conservative: they indicate signs of non-convergence in conditions where $\bar{Q}$s are unbiased and or confidence valid estimates of $Q$.
+Evaluation with the performance measures shows that $\widehat{R}$ and autocorrelation are conservative: they indicate signs of non-convergence in conditions where $\bar{Q}$s are unbiased and or confidence valid estimates of $Q$.
 
@@
 Upon convergence, $\widehat{R}=1$ and $AC=0$, but in MCMC algorithms these are very unlikely (because it's a distribution, not a point to converge to).
-In practice thresholds are used to diagnose non-convergence.
+In practice, thresholds are used to diagnose non-convergence.
 Usually, non-convergence is diagnosed when $\widehat{R}$ > 1.2 or 1.1 or even 1.01.
@@
 If the algorithms would be terminated at the iteration where $\widehat{R}$ < 1.2 occurs for the first time, the increase in $\widehat{R}$ values might be missed.
-The threshold $\widehat{R}$ < 1.1 is somewhat conservative with respect to the performance measures.
+The threshold $\widehat{R}$ < 1.1 is somewhat conservative in comparison with the performance measures.
 But it seems OK.
@@
 ```
-The aim of the simulation study is to evaluate the impact of inducing non-convergence by: 1) terminating the MICE algorithm at different imputation chain lengths ($t = 1, 2, ..., 100$), and 2) varying the missingness proportions ($p_{miss} = .05, .25, .50, .75, .95$).
+The aim of the simulation study is to evaluate the impact of inducing non-convergence by: 1) terminating the MICE algorithm at different imputation chain lengths ($t = 1, 2,..., 100$), and 2) varying the missingness proportions ($p_{miss} =.05,.25,.50,.75,.95$).
 The assumption underlying the different number of iterations is that the algorithm generally does reach convergence at $t=1$, because the initial values are sampled randomly from the set of observed datapoints.
@@
 \end{align*} 
-For the purpose of this study, sampling variance is not of interest.
+In this study, sampling variance is not of interest.
 Therefore, a single complete set may serve as comparative truth in all simulation runs [@vinknd].
@@
 $$
-Y_i =  1 + 2X_{1i} + .5X_{2i} - X_{3i} + \epsilon_i ,
+Y_i = 1 + 2X_{1i} +.5X_{2i} - X_{3i} + \epsilon_i ,
 $$
-where $i = 1, 2, ..., N$ and $\epsilon \sim \mathcal{N}(0, 100)$.
+where $i = 1, 2,..., N$ and $\epsilon \sim \mathcal{N}(0, 100)$.
 
-We consider four quantities of scientific interest ['conceptual estimands'; @morr19], namely  ...
-solve a multiple linear regression problem, where dependent variable $Y$ is regressed on independent variables $X_1$, $X_2$ and $X_3$ (**check notation with betas here, and add what the quantity/-ies of scientific interest is/are!**):
+We consider four quantities of scientific interest ['conceptual estimands'; @morr19], namely...
+solve a multiple linear regression problem, where dependent variable $Y$ is regressed on independent variables $X_1$, $X_2$ and $X_3$ (**check notation with betas here, and add what the quantity/-ies of scientific interest is/are!
+Namely, the descriptive statistics of all variables (mean and standard deviation), $\beta_2$, and the percentage of variance in $Y$ explained (coefficient of determination; $R^2$)**):
 
@@
 Therefore, results may not be extrapolated to other missing data problems.
-Proper performance of the convergence diagnostics under MCAR is necessary but not sufficient to demonstrate appropriateness of $\widehat{R}$ and $AC$ as convergence diagnostics.
-This is just a proof of concept.
+Proper performance of the convergence diagnostics under MCAR is necessary but not sufficient to demonstrate the appropriateness of $\widehat{R}$ and $AC$ as convergence diagnostics.
+Convergence diagnostics should therefore at least apply to the MCAR situation, before more complex missingness should be explored.
 
-Missing datapoints are imputed with the function `mice::mice()`.
+Missing datapoints in $y$ are imputed with the function `mice::mice()`.
 All MI procedures are performed with Bayesian linear regression imputation (`method = "norm"`), and five imputation chains (`m = 5`).
 The number of iterations varies between simulation conditions (`maxit = ` $1, 2, \dots, 100$).
+Each repetition of the simulation thus starts from 1 complete dataset, then splits into 5 missingness conditions ($p =.05,.25,.5,.75,.95$), each of the amputed datasets is imputed 5 times ($m = 1, 2,..., 5$) for each iteration condition ($t = 1, 2,..., 100$).
+From these $5\times5\times100$ imputations, we extract the $\theta$s to apply $\widehat{R}$ and $AC$ on (**i.e., chain means, chain variances, $\beta$s per imputation, and the first eigenvalue of the variance-covariance matrix per imputation**).
+We calculate $\widehat{R}$ by implementing Vehtari et al.'s [-@veht19] recommendations, and $AC$ as the correlation between the $t^{th}$ and the $(t+1)^{th}$ iteration.
+For each $\theta$ we estimate the corresponding scientific estimand $Q$.
 
-<!-- What to evaluate? -->
-
-<!-- - bias in regression coefficients -->
-
-<!-- - bias in $R^2$ (= coefficient of determination) -->
-
-<!-- -  -->
-
-<!-- From the imputation step, we extract chain means and chain variances across iterations ($\theta$ at $t = 1,2,\dots,T$).   -->
-<!-- We compute convergence diagnostics separately for both $\theta$s, and for each of the four variables in the dataset.
-We report the maximum (absolute) value as the 'worst linear predictor' of convergence.
--->
-$\widehat{R}$ (**of what??
-i.e., chain means and chain variances of each variable**) is computed by implementing Vehtari et al.'s [-@veht19] recommendations.
-$AC$ (**of what?
-and how is $AC$ aggregated across imputations?
-i.e., mean in latest sim**) is computed with function `stats::acf()`.
-
-To estimate the quantity of scientific interest, $Q$, we perform multiple linear regression on each completed dataset with the function `stats::lm()`.
+The estimator for each $Q$ is $\bar{Q}$---the pooled aggregate of the ($\hat{Q}$)s across imputations.
+To calculate the $\bar{Q}$s, we combine the observed data ($y_{obs}$) and the imputed data for each imputation ($y_{imp,m}$) with the function `mice::complete()`.
+Descriptive statistics are computed as the average across imputations for each variable ($\mu_j$ and $\sigma_j$, where $j = Y, X_1, X_2, X_3$).
+Estimated regression coefficients are obtained with the function `stats::lm()` for each imputation, and then pooled conform @vink14.
+The coefficient of determination is estimated for each imputation, and pooled using `mice::pool.r.squared()`.
+<!-- perform multiple linear regression on each completed dataset.
 We obtain an estimated regression coefficient per imputation, which are pooled into a single estimate, $\bar{Q}$.
 We use the function `mice::pool()` to estimate u bar (**check definition!**) according to Rubin's [-@rubin87] rules, and subsequently implement finite population pooling conform @vink14.
+-->
 
-We compute bias as the difference between $Q$ and $\bar{Q}$.
-Confidence interval width (CIW) is defined as the difference between the lower and upper bound of the 95\% confidence interval (CI95\%) around $\bar{Q}$.
-We compute the CI95\% bounds as
+The performance of $\widehat{R}$ and $AC$ is assessed by comparing $\bar{Q}$s with $Q$s.
+For each $Q$, we compute bias as $\bar{Q} - Q$.
+For $Q=\beta_j$, we also compute the empirical coverage rate (CR).
+CR is defined as the percentage of simulation repetitions in which the 95\% confidence interval (CI) around $\bar{Q}$ covers the true estimand $Q$.
+Let
 
-$$\bar{Q} \pm t_{(m-1)} \times SE_{\bar{Q}},$$
+$$\text{CI} = \bar{Q} \pm t_{(M-1)} \times SE_{\bar{Q}},$$
 
-where $t_{(m-1)}$ is the quantile of a $t$-distribution with $m-1$ degrees of freedom, and $SE_{\bar{Q}}$ is the square root of the pooled variance estimate.
-**Why track CIW?** Under-estimating the variance of $\bar{Q}$ may yield spurious inferences.
-From bias and CIW, we calculate empirical coverage rates. Coverage rate is the proportion of simulations in which $Q$ is between the bounds of the CI95\% around $\bar{Q}$.
+where $t_{(M-1)}$ is the quantile of a $t$-distribution with $M-1$ degrees of freedom, and $SE_{\bar{Q}}$ is the square root of the pooled variance estimate.
+**Remove this?
+** Under-estimating the variance of $\bar{Q}$ may yield spurious inferences.
+Confidence interval width (CIW) is defined as the difference between the lower and upper bound of the 95\% confidence interval (CI) around $\bar{Q}$ From bias and CIW, we calculate empirical coverage rates.
+Coverage rate is the proportion of simulations in which $Q$ is between the bounds of the CI95\% around $\bar{Q}$.
 
 # Old Results section
+
+For reasons of brevity, we only discuss the convergence diagnostics for the $Q$ with the worst performance in terms of bias.
+For the descriptve statistics, the magnitude of the bias was the largest in $Y$ (i.e, $j = Y$ in $\mu_j$ and $\sigma_j$).
+For the $Q=\beta_j$, $j=2$ showed the worst perfomance (i.e, the effect of $X_1$ on $Y$.
 
@@
 
-The bias in the estimates of the variable means show little to no difference between simulation conditions.
+The bias in the estimates of the variable means shows little to no difference between simulation conditions.
 It doesn't seem to matter how many iterations you use in the mice algorithm, the estimates are unbiased.
@@
 Neither is the most recent recommended threshold reached ($\widehat{R}$ < 1.01).
-The conventional $\widehat{R}$ threshold of 1.2 is reached in simulation conditions $T = 3$ and $T > 6$.
+**This is only for 75% missingness:** The conventional $\widehat{R}$ threshold of 1.2 is reached in simulation conditions $T = 3$ and $T > 6$.
 (With the default number of iterations (maxit = 5), this dip in $\widehat{R}$ values would be spotted, so it is no problem.) The point at which an extra iteration does not seem to improve the $\widehat{R}$ value is around $T=30$.
@@
 
-Auto-correlations indicate no sign of trending within imputation chains.
+Autocorrelations indicate no sign of trending **(NOT correct anymore, fluke of the stats::acf() function!!)** within imputation chains.
 In most simulation conditions $AC$ is smaller than or about equal to zero.
 Across simulation conditions, however, the autocorrelation curve does not trend towards 0.
-Autocorrelation values plateau off at a value of around .1 .
+Autocorrelation values plateau off at a value of around.1.
 This is a small positive autocorrelation, which would indicate some trending within chains.
 
-The $\widehat{R}$ and $AC$ values for the imputation chain variances show equal trends and are therefore not discussed separately.
+The $\widehat{R}$ and $AC$ values for the imputation chain variances show equal trends to the chain means and are therefore not discussed separately.
 Taken together, univariate estimates seem robust across simulation conditions.
@@
 
-There is a clear bias in the regression estimates in simulation conditions where the number of iterations is smaller than four.
+**This is only true for 75% missingness:** There is a clear bias in the regression estimates in simulation conditions where the number of iterations is smaller than four.
 In simulation conditions where $T > 5$ there is little to no bias in the estimated regression coefficients.
-In most simulation conditions nominal coverage is obtained (i.e., coverage rates of .95) for the confidence intervals around the regression coefficients.
+In most simulation conditions nominal coverage is obtained (i.e., coverage rates of.95) for the confidence intervals around the regression coefficients.
 Conditions with only one or two iterations show some under-coverage.
@@
 
-If we look at the estimated proportion of explained variance in outcome variable Y we see that the coefficient of determination is underestimate estimated in conditions where the number of iterations is equal to two or less, and slightly overestimated in conditions where the number of iterations is equal to three or more.
-
-```{r res-multi}
-# est_bias + plot_annotation(title = "Bias in regression estimate, coverage rate, \n bias in variance explained")
-# est_cov
-# Rsq_bias 
-```
+**This is only true for 75% missingness:** If we look at the estimated proportion of explained variance in outcome variable Y we see that the coefficient of determination is underestimate estimated in conditions where the number of iterations is equal to two or less, and slightly overestimated in conditions where the number of iterations is equal to three or more.
 
@@
 E.g., with 95% of cases having missing data we need at least seven iterations to obtain unbiased results.
-
-<!-- Bias in the estimated $R^2$ values does not improve  -->
-
-<!-- Figures 2 and 3 display results per simulation condition ($T = 1,2,\dots,100$).  -->
 
@@
 # # chain means
-# dat[-1, 6:9] %>% mutate(max_Rh = apply(., 1, max)) %>% .$max_Rh<1.2
+# dat[-1, 6:9] %>% mutate(max_Rh = apply(., 1, max)) %>%.$max_Rh<1.2
 # 
 # # chain variances
-# dat[-1, 26:29] %>% mutate(max_Rh = apply(., 1, max)) %>% .$max_Rh<1.2
+# dat[-1, 26:29] %>% mutate(max_Rh = apply(., 1, max)) %>%.$max_Rh<1.2
 
 ```
-
-<!-- # Results [OLD, JUST FOR REFERENCE!] -->
-
-<!-- ## Convergence Diagnostics -->
-
-<!-- **(Pair convergence diagnostics with respective estimates, not split between convergence and simulation diagnostics)** -->
-
-<!-- It is apparent that there is a relation between the number of iterations per simulation condition ($T$) and the convergence diagnostics.
-Generally speaking, conditions with longer imputation chains (higher $T$) coincide with less signs of non-convergence ($\widehat{R}$-values approach one, and $AC$-values approach zero).
--->
-<!-- <!-- We see more or less equivalent trends for chain means versus chain variances. We, therefore, only discuss the $\widehat{R}$ and $AC$-values of chain means.  --> -->
-
-<!-- Figure 2A shows that $\widehat{R}$-values generally decrease with increasing imputation chain lengths.
-The decline stabilizes somewhere between the simulation conditions $T=30$ and $T=50$.
-The downward trend is most pronounced for $T=3$, and between $T = 5$ and $T = 10$.
-In the intervening conditions ($3 \leq T \leq 5$), however, we observe a steep increase in $\widehat{R}$-values.
-This increase implies that non-convergence is under-estimated, and convergence should not be diagnosed at this point (**is that true???**).
-Based on the conventional threshold $\widehat{R} < 1.2$, we would falsely diagnose convergence at $T=3$.
-According to the widely used threshold $\widehat{R} < 1.1$, convergence would be diagnosed for conditions where $T>9$.
-If we use the recently recommended threshold $\widehat{R} < 1.01$, we would conclude that convergence is reached in none of the simulation conditions.
--->
-
-<!-- The $AC$-values displayed in Figure 2B are almost uniformly increasing as a function of $T$.
-The only $AC$-value that deviates from this observation is for $T=2$.
-The lowest $AC$-value is obtained for $T=3$.
-%At $T=5$, the $AC$-value reaches the level observed at $T=2$.
--->
-<!-- The gradual increase plateaus between $T=10$ and $T=30$.
-$AC$-values in conditions where $T>70$ are indifferentiable from zero, indicating stationarity.
-We see an initial decrease between $T=2$ and $T=3$.
-Simulation conditions where $T>5$ have $AC$-values greater than $T=2$ (**weird wording**).
--->
-
-<!-- According to this diagnostic, none of the simulation conditions show signs of non-convergence, since we only observe negative or zero $AC$-values.  -->
-
-<!-- Taken together, we see that $T>3$ is the minimal requirement to diagnose convergence ($\widehat{R} < 1.2; $AC$ \leq 0$).
-This threshold is, however, not sufficient, since we overlook the increase in $\widehat{R}$-values up-to conditions where $T>5$, and the convergence diagnostics only reach stability at $T>20$.
--->
-
-<!-- ## Simulation Diagnostics -->
-
-<!-- We use average bias, average confidence interval width, and coverage rate as performance measures to evaluate $\widehat{R}$ and $AC$ (**not the goal!
-we want to show the effects of non-conv, not just see how to diagnose it.**).
-We make the general observation that the simulation diagnostics behave as theorized for most simulation conditions (bias around zero, stable confidence interval widths, nominal coverage rate at 95\%).
--->
-
-<!-- Figure 3A shows that bias is fairly stable across simulation conditions.
-The condition $T=1$ clearly deviates from this trend with a negative bias.
-The bias for $T=2$ is below average, but within the range of fluctuations.
-Conditions where $T>3$ can be diagnosed as unbiased, but the average bias across iterations (as shown by a flat Loess line) only reaches stability at $T=20$ (**Loess might not be relevant, check and remove accordingly!**).
--->
-
-<!-- CIW is only clearly divergent in the simulation condition $T=1$, see Figure 3B.
-Only conditions where $T>3$ are therefore considered sufficient.
-(**Possibly irrelevant Loess:** These conditions have similar CIWs, but across iterations stability is not established (i.e., the Loess line is never completely flat within the 100 simulation conditions).)   -->
-
-<!-- The empirical coverage rate across repetitions seems more or less stable for conditions where $T>1$, see Figure 3C.
-We see some over-coverage at $T=2$, but that is better than under-estimating the variance of $\bar{Q}$.
-On average, the coverage rate is somewhat higher than the expected nominal coverage of 95\% [@neym34] (**not anymore??**).
-(**Loess:** Similar to the CIWs, there is some trending across iterations (i.e., the Loess line is never flat)).
--->
-
-<!-- From the simulation diagnostics, we observe that unbiased estimates with nominal coverage rates were obtained in conditions where $T>3$.
-This suggests that as little as four iterations may be sufficient for the MI algorithm under the current circumstances.
--->
-
-<!-- In short, there is a discrepancy between what the convergence diagnostics and what the performance measures indicate.
-While $T=4$ seems sufficient with respect to simulation quantities, the condition where $T=4$ resulted in the `one-but-worst' values of both convergence diagnostics.
-Complete algorithmic convergence as indicated by $\widehat{R}$ and autocorrelation is not reached in conditions where $T<20$.
--->
 
@@
 Add what %miss has to do with it.**)
-$\widehat{R}$ and autocorrelation indicate that algorithmic convergence may only be reached after twenty or even forty iterations, while unbiased, confidence valid estimates estimates may be obtained with as little as four iterations.
+
+$\widehat{R}$ and autocorrelation indicate that algorithmic convergence may only be reached after twenty or even forty iterations, while unbiased, confidence valid estimates may be obtained with as little as four iterations.
 These results are in agreement with the simulation hypothesis: $\widehat{R}$ over-estimates the severity of non-convergence when applied to MI procedures.
 %This may be due to the quantity of scientific interest chosen.
-More `complicated' $Q$s (e.g., higher order effects or variance components) might show bias, under- or over-coverage at higher $T$.
+More `complicated' $Q$s (e.g., higher-order effects or variance components) might show bias, under- or over-coverage at higher $T$.
 
@@
 (**This is only one of the goals: to give applied researchers a diagnostic to indicate that they should keep iterating.
-The other is the default in mice and other software packages, and yet another is ...
+The other is the default in mice and other software packages, and yet another is...
 i forgot**) Under the relatively easy missing data problem of the current study, the threshold was not reached.
@@
 The negative $AC$-values obtained in this study show no threat of non-stationarity.
-However, initial dip in $AC$-values may have implications for the default number of iterations in `mice` (`maxit = 5`).
+However, the initial dip in $AC$-values **(which disappeared!)** may have implications for the default number of iterations in `mice` (`maxit = 5`).
 Terminating the algorithm at $T=5$ may not be the most appropriate, since this lead to the worst convergence (**nope, only for Rh, not $AC$**), as indicated by $\widehat{R}$ and $AC$.
 Under the current specifications, $T>20$ would be more appropriate.
-
-The observed dip in $AC$ implies that default maxit value of five iterations is the worst possible number of iterations.
-Moreover, the results of this study imply that assessing the stationarity component of convergence with $AC$ might be redundant.
-<!-- $AC$ would thus not be informative of the convergence of MI algorithms.  -->
 
@@
 
-**For empirical researchers: 1) Check trace plots for pathological non convergence and adjust imputation model if necessary.
-2) use $\widehat{R}$ wait 1.1 ash threshold and autocorrelation with **?** as threshold .
+For empirical researchers: 1) Check trace plots for pathological non convergence and adjust imputation model if necessary.
+2) use $\widehat{R}$ wait 1.1 ash threshold and autocorrelation with **?** as threshold.
 Keep iterating until these thresholds are reached.
 3) Do not use the R function ACF.
-Instead compute autocorrelations manually , see for example GitHub link.
+Instead, compute autocorrelations manually **(see e.g., [GitHub link])**.
 4) Track your own scalar summary of interest.
-This is advanced but explained in Van Buren 2018.
+This is somewhat advanced but explained in Van Buren 2018.
 Compute $\widehat{R}$ and autocorrelation values for this scalar summary.
-**
+5) **Something** about the novel $\theta$ that is 'substantive model-independent'.
 
@@
 
-**Pmm, M(N)AR, Empirical data, Vignette, ShinyMICE, AC across imputations, not iterations (unique for imp data??)**
+**Pmm, M(N)AR, Empirical data, Vignette, ShinyMICE, AC across imputations, not iterations (new convergence diagnostic unique for iterative imputation??).**
 
-**If I don't include it in my study, also: Significance of $AC$ values**
+**If I don't include it in my study, also: Significance of $AC$ values.**
 

### 2020-04-27 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
After spellcheck

@@
 Anyone who analyzes person-data may run into a missing data problem.
-Missing data is not only ubiquitous, but also tedious. If a dataset
+Missing data is not only ubiquitous but also tedious. If a dataset
 contains just one incomplete observation, statistical inferences are
@@
 
-In this paper we investigate how non-convergence in iterative imputation
-algorithms may be diagnosed, how well these methods perform, and at
-which point convergence may safely be assumed. For reasons of brevity,
-we only focus on the MI algorithm implemented in the world-leading MI
-software: the \texttt{R} \citep{R} package \texttt{mice} \citep{mice}.
-The convergence properties of this MI algorithm are investigated through
-model-based simulation. The results of this simulation study are
-guidelines for assessing convergence of MI algorithms, which will aid
-applied researchers in drawing valid inference from incomplete datasets.
+In this paper, we investigate how non-convergence in iterative
+imputation algorithms may be diagnosed, how well these methods perform,
+and at which point convergence may safely be assumed. For reasons of
+brevity, we only focus on the MI algorithm implemented in the
+world-leading MI software: the \texttt{R} \citep{R} package
+\texttt{mice} \citep{mice}. The convergence properties of this MI
+algorithm are investigated through model-based simulation. The results
+of this simulation study are guidelines for assessing convergence of MI
+algorithms, which will aid applied researchers in drawing valid
+inference from incomplete datasets.
 
@@
 magnitude two larger than those of the typically converged algorithm.
-This shows the importance of reaching convergend states in iterative
+This shows the importance of reaching converged states in iterative
 imputation algorithms.}
@@
 correlation between two subsequent \(\theta\)-values within the same
-chain \citep[p.~147]{lync07}. In this study we only consider \(AC\) at
+chain \citep[p.~147]{lync07}. In this study, we only consider \(AC\) at
 lag 1, i.e., the correlation between the \(t^{th}\) and \(t+1^{th}\)
@@
 even---negative \(AC\)-values indicate that \(\theta\)-values of
-subsequent iterations diverge from one-another, which may increase the
+subsequent iterations diverge from one another, which may increase the
 variance of \(\theta\) and speed up convergence. As convergence
@@
 \textbf{To assess whether \(\widehat{R}\) and \(AC\) may be appropriate
-non-convergence diagnostics for MICE algorithms, the following condtion
+non-convergence diagnostics for MICE algorithms, the following condition
 must hold when applied on the two algorithms plotted above: the methods
@@
 performance of the typical convergence over the pathological
-non-convergnce is less prominent, even flipped for \(t<4\). Panel C
+non-convergence is less prominent, even flipped for \(t<4\). Panel C
 displays the \(AC\) as computed with the R function stats::acf(). When
 we look at this panel, we must conclude that something weird is going
-on. The \(AC\)-values indicate equal performance (upto \(t=5\)) for the
-typical convergence and the pathological non-convergnce, while there is
+on. The \(AC\)-values indicate equal performance (up-to \(t=5\)) for the
+typical convergence and the pathological non-convergence, while there is
 obvious trending in the latter. Moreover, the best convergence (as
@@
 looking at the chain means in panel A, there should be some signs of
-trending upto iteration number seven. After consulting the documentation
-on stats::acf(), we conclude that this implementation of \(AC\) is not
-suitable for iterative imputation algorithms. There is a correction
-factor for a mathematical shortcut that works in the limit. According to
-\ldots{} the function works when the number of iterations \textgreater{}
-50. The default number of iterations in mice, however, is 5. Therefore,
-we compute \(AC\) manually, see panel D. The AC values in this plot do
-meet the requirements (\ldots) and will therefore be used in the
-simulation study.}
+trending up-to iteration number seven. After consulting the
+documentation on stats::acf(), we conclude that this implementation of
+\(AC\) is not suitable for iterative imputation algorithms. There is a
+correction factor for a mathematical shortcut that works in the limit.
+According to\ldots{} the function works when the number of iterations
+\textgreater{} 50. The default number of iterations in mice, however, is
+5. Therefore, we compute \(AC\) manually, see panel D. The AC values in
+this plot do meet the requirements (\ldots) and will therefore be used
+in the simulation study.}
 
@@
 randomness induced by the algorithm (\(\widehat{R} = 1\)). We expect
-that a completely convergenced state will not be reached.
+that a completely converged state will not be reached.
 
@@
 user-specified scalar summary, however, is not universal to all complete
-data problems. Therefore, as inspired by \citep{mack03}, we will use
-introduce a \(\theta\) that summarizes the multivariate state space of
-the algorithm. Namely, the first eigenvalue of the variance-covariance
+data problems. Therefore, as inspired by \citep{mack03}, we will propose
+a \(\theta\) that summarizes the multivariate state space of the
+algorithm. Namely, the first eigenvalue of the variance-covariance
 matrix of the \(M\) completed datasets.
@@
 in old Methods}). \textbf{remove the term performance measure here and
-add an extra alinea to define performance measures as bias in all
+add an extra paragraph to define performance measures as bias in all
 estimands, and coverage rate and CI length of regression coefficients.}
@@
 converged as \(t=1\), and converges gradually as \(t\) increases. The
-point at wich an additional iteration does not lead to an improvement of
-the estimates depends on the difficulty of the missingness
+point at which an additional iteration does not lead to an improvement
+of the estimates depends on the difficulty of the missingness
 problem---between XYZ and XYZ. This means that the algorithm has
@@
 
-Evaluation with the performance measures shows that are had an
+Evaluation with the performance measures shows that \(\widehat{R}\) and
 autocorrelation are conservative: they indicate signs of non-convergence
@@
 these are very unlikely (because it's a distribution, not a point to
-converge to). In practice thresholds are used to diagnose
+converge to). In practice, thresholds are used to diagnose
 non-convergence. Usually, non-convergence is diagnosed when
@@
 time, the increase in \(\widehat{R}\) values might be missed. The
-threshold \(\widehat{R}\) \textless{} 1.1 is somewhat conservative with
-respect to the performance measures. But it seems OK. \(\widehat{R}\)
-\textless{} 1.01 is too stringent compared to the performance measures.
-And it is not obtained for any number of iterations (\textbf{but may be
-necessary in more complex complete data models???}).
+threshold \(\widehat{R}\) \textless{} 1.1 is somewhat conservative in
+comparison with the performance measures. But it seems OK.
+\(\widehat{R}\) \textless{} 1.01 is too stringent compared to the
+performance measures. And it is not obtained for any number of
+iterations (\textbf{but may be necessary in more complex complete data
+models???}).
 
@@
 non-convergence by: 1) terminating the MICE algorithm at different
-imputation chain lengths (\(t = 1, 2, ..., 100\)), and 2) varying the
-missingness proportions (\(p_{miss} = .05, .25, .50, .75, .95\)). The
+imputation chain lengths (\(t = 1, 2,..., 100\)), and 2) varying the
+missingness proportions (\(p_{miss} =.05,.25,.50,.75,.95\)). The
 assumption underlying the different number of iterations is that the
@@
 \end{bmatrix}\!\!\text{.}\\[2\jot]
-\end{align*} For the purpose of this study, sampling variance is not of
-interest. Therefore, a single complete set may serve as comparative
-truth in all simulation runs \citep{vinknd}. A finite population of
-\(N=1000\) is simulated using the function \texttt{mvtnorm::rmvnorm()}.
-Subsequently, a fourth variable is constructed to serve as dependent
-variable in a multiple linear regression problem. Let \[
-Y_i =  1 + 2X_{1i} + .5X_{2i} - X_{3i} + \epsilon_i ,
-\] where \(i = 1, 2, ..., N\) and \(\epsilon \sim \mathcal{N}(0, 100)\).
+\end{align*} In this study, sampling variance is not of interest.
+Therefore, a single complete set may serve as comparative truth in all
+simulation runs \citep{vinknd}. A finite population of \(N=1000\) is
+simulated using the function \texttt{mvtnorm::rmvnorm()}. Subsequently,
+a fourth variable is constructed to serve as dependent variable in a
+multiple linear regression problem. Let \[
+Y_i = 1 + 2X_{1i} +.5X_{2i} - X_{3i} + \epsilon_i ,
+\] where \(i = 1, 2,..., N\) and \(\epsilon \sim \mathcal{N}(0, 100)\).
 
 We consider four quantities of scientific interest \citep[`conceptual
-estimands';][]{morr19}, namely \ldots{} solve a multiple linear
+estimands';][]{morr19}, namely\ldots{} solve a multiple linear
 regression problem, where dependent variable \(Y\) is regressed on
@@
 notation with betas here, and add what the quantity/-ies of scientific
-interest is/are!}):
+interest is/are! Namely, the descriptive statistics of all variables
+(mean and standard deviation), \(\beta_2\), and the percentage of
+variance in \(Y\) explained (coefficient of determination; \(R^2\))}):
 
@@
 data problems. Proper performance of the convergence diagnostics under
-MCAR is necessary but not sufficient to demonstrate appropriateness of
-\(\widehat{R}\) and \(AC\) as convergence diagnostics. This is just a
-proof of concept.
+MCAR is necessary but not sufficient to demonstrate the appropriateness
+of \(\widehat{R}\) and \(AC\) as convergence diagnostics. Convergence
+diagnostics should therefore at least apply to the MCAR situation,
+before more complex missingness should be explored.
 
-Missing datapoints are imputed with the function \texttt{mice::mice()}.
-All MI procedures are performed with Bayesian linear regression
-imputation (\texttt{method\ =\ "norm"}), and five imputation chains
-(\texttt{m\ =\ 5}). The number of iterations varies between simulation
-conditions (\texttt{maxit\ =} \(1, 2, \dots, 100\)).
+Missing datapoints in \(y\) are imputed with the function
+\texttt{mice::mice()}. All MI procedures are performed with Bayesian
+linear regression imputation (\texttt{method\ =\ "norm"}), and five
+imputation chains (\texttt{m\ =\ 5}). The number of iterations varies
+between simulation conditions (\texttt{maxit\ =} \(1, 2, \dots, 100\)).
+Each repetition of the simulation thus starts from 1 complete dataset,
+then splits into 5 missingness conditions (\(p =.05,.25,.5,.75,.95\)),
+each of the amputed datasets is imputed 5 times (\(m = 1, 2,..., 5\))
+for each iteration condition (\(t = 1, 2,..., 100\)). From these
+\(5\times5\times100\) imputations, we extract the \(\theta\)s to apply
+\(\widehat{R}\) and \(AC\) on (\textbf{i.e., chain means, chain
+variances, \(\beta\)s per imputation, and the first eigenvalue of the
+variance-covariance matrix per imputation}). We calculate
+\(\widehat{R}\) by implementing Vehtari et al.'s \citeyearpar{veht19}
+recommendations, and \(AC\) as the correlation between the \(t^{th}\)
+and the \((t+1)^{th}\) iteration. For each \(\theta\) we estimate the
+corresponding scientific estimand \(Q\).
 
-\(\widehat{R}\) (\textbf{of what?? i.e., chain means and chain variances
-of each variable}) is computed by implementing Vehtari et al.'s
-\citeyearpar{veht19} recommendations. \(AC\) (\textbf{of what? and how
-is \(AC\) aggregated across imputations? i.e., mean in latest sim}) is
-computed with function \texttt{stats::acf()}.
+The estimator for each \(Q\) is \(\bar{Q}\)---the pooled aggregate of
+the (\(\hat{Q}\))s across imputations. To calculate the \(\bar{Q}\)s, we
+combine the observed data (\(y_{obs}\)) and the imputed data for each
+imputation (\(y_{imp,m}\)) with the function \texttt{mice::complete()}.
+Descriptive statistics are computed as the average across imputations
+for each variable (\(\mu_j\) and \(\sigma_j\), where
+\(j = Y, X_1, X_2, X_3\)). Estimated regression coefficients are
+obtained with the function \texttt{stats::lm()} for each imputation, and
+then pooled conform \citet{vink14}. The coefficient of determination is
+estimated for each imputation, and pooled using
+\texttt{mice::pool.r.squared()}.
 
-To estimate the quantity of scientific interest, \(Q\), we perform
-multiple linear regression on each completed dataset with the function
-\texttt{stats::lm()}. We obtain an estimated regression coefficient per
-imputation, which are pooled into a single estimate, \(\bar{Q}\). We use
-the function \texttt{mice::pool()} to estimate u bar (\textbf{check
-definition!}) according to Rubin's \citeyearpar{rubin87} rules, and
-subsequently implement finite population pooling conform \citet{vink14}.
+The performance of \(\widehat{R}\) and \(AC\) is assessed by comparing
+\(\bar{Q}\)s with \(Q\)s. For each \(Q\), we compute bias as
+\(\bar{Q} - Q\). For \(Q=\beta_j\), we also compute the empirical
+coverage rate (CR). CR is defined as the percentage of simulation
+repetitions in which the 95\% confidence interval (CI) around
+\(\bar{Q}\) covers the true estimand \(Q\). Let
 
-We compute bias as the difference between \(Q\) and \(\bar{Q}\).
-Confidence interval width (CIW) is defined as the difference between the
-lower and upper bound of the 95\% confidence interval (CI95\%) around
-\(\bar{Q}\). We compute the CI95\% bounds as
+\[\text{CI} = \bar{Q} \pm t_{(M-1)} \times SE_{\bar{Q}},\]
 
-\[\bar{Q} \pm t_{(m-1)} \times SE_{\bar{Q}},\]
-
-where \(t_{(m-1)}\) is the quantile of a \(t\)-distribution with \(m-1\)
+where \(t_{(M-1)}\) is the quantile of a \(t\)-distribution with \(M-1\)
 degrees of freedom, and \(SE_{\bar{Q}}\) is the square root of the
-pooled variance estimate. \textbf{Why track CIW?} Under-estimating the
-variance of \(\bar{Q}\) may yield spurious inferences. From bias and
-CIW, we calculate empirical coverage rates. Coverage rate is the
-proportion of simulations in which \(Q\) is between the bounds of the
-CI95\% around \(\bar{Q}\).
+pooled variance estimate. \textbf{Remove this? } Under-estimating the
+variance of \(\bar{Q}\) may yield spurious inferences. Confidence
+interval width (CIW) is defined as the difference between the lower and
+upper bound of the 95\% confidence interval (CI) around \(\bar{Q}\) From
+bias and CIW, we calculate empirical coverage rates. Coverage rate is
+the proportion of simulations in which \(Q\) is between the bounds of
+the CI95\% around \(\bar{Q}\).
 
@@
 \section{Old Results section}\label{old-results-section}}
+
+For reasons of brevity, we only discuss the convergence diagnostics for
+the \(Q\) with the worst performance in terms of bias. For the
+descriptve statistics, the magnitude of the bias was the largest in
+\(Y\) (i.e, \(j = Y\) in \(\mu_j\) and \(\sigma_j\)). For the
+\(Q=\beta_j\), \(j=2\) showed the worst perfomance (i.e, the effect of
+\(X_1\) on \(Y\).
 
@@
 
-The bias in the estimates of the variable means show little to no
+The bias in the estimates of the variable means shows little to no
 difference between simulation conditions. It doesn't seem to matter how
@@
 the most recent recommended threshold reached (\(\widehat{R}\)
-\textless{} 1.01). The conventional \(\widehat{R}\) threshold of 1.2 is
-reached in simulation conditions \(T = 3\) and \(T > 6\). (With the
-default number of iterations (maxit = 5), this dip in \(\widehat{R}\)
-values would be spotted, so it is no problem.) The point at which an
-extra iteration does not seem to improve the \(\widehat{R}\) value is
-around \(T=30\).
+\textless{} 1.01). \textbf{This is only for 75\% missingness:} The
+conventional \(\widehat{R}\) threshold of 1.2 is reached in simulation
+conditions \(T = 3\) and \(T > 6\). (With the default number of
+iterations (maxit = 5), this dip in \(\widehat{R}\) values would be
+spotted, so it is no problem.) The point at which an extra iteration
+does not seem to improve the \(\widehat{R}\) value is around \(T=30\).
 
-Auto-correlations indicate no sign of trending within imputation chains.
-In most simulation conditions \(AC\) is smaller than or about equal to
-zero. Across simulation conditions, however, the autocorrelation curve
-does not trend towards 0. Autocorrelation values plateau off at a value
-of around .1 . This is a small positive autocorrelation, which would
-indicate some trending within chains.
+Autocorrelations indicate no sign of trending \textbf{(NOT correct
+anymore, fluke of the stats::acf() function!!)} within imputation
+chains. In most simulation conditions \(AC\) is smaller than or about
+equal to zero. Across simulation conditions, however, the
+autocorrelation curve does not trend towards 0. Autocorrelation values
+plateau off at a value of around.1. This is a small positive
+autocorrelation, which would indicate some trending within chains.
 
 The \(\widehat{R}\) and \(AC\) values for the imputation chain variances
-show equal trends and are therefore not discussed separately. Taken
-together, univariate estimates seem robust across simulation conditions.
-There is no clear effect of the number of iterations on the bias in
-these estimates, while the convergence diagnostics indicate that the
-algorithm did not reach a completely converged state (yet).
+show equal trends to the chain means and are therefore not discussed
+separately. Taken together, univariate estimates seem robust across
+simulation conditions. There is no clear effect of the number of
+iterations on the bias in these estimates, while the convergence
+diagnostics indicate that the algorithm did not reach a completely
+converged state (yet).
 
@@
 
-There is a clear bias in the regression estimates in simulation
-conditions where the number of iterations is smaller than four. In
-simulation conditions where \(T > 5\) there is little to no bias in the
-estimated regression coefficients. In most simulation conditions nominal
-coverage is obtained (i.e., coverage rates of .95) for the confidence
-intervals around the regression coefficients. Conditions with only one
-or two iterations show some under-coverage. Since confidence interval
-width is stable across conditions, the under-coverage may be attributed
-to the bias in the estimated regression coefficients.
+\textbf{This is only true for 75\% missingness:} There is a clear bias
+in the regression estimates in simulation conditions where the number of
+iterations is smaller than four. In simulation conditions where
+\(T > 5\) there is little to no bias in the estimated regression
+coefficients. In most simulation conditions nominal coverage is obtained
+(i.e., coverage rates of.95) for the confidence intervals around the
+regression coefficients. Conditions with only one or two iterations show
+some under-coverage. Since confidence interval width is stable across
+conditions, the under-coverage may be attributed to the bias in the
+estimated regression coefficients.
 
-If we look at the estimated proportion of explained variance in outcome
-variable Y we see that the coefficient of determination is underestimate
-estimated in conditions where the number of iterations is equal to two
-or less, and slightly overestimated in conditions where the number of
-iterations is equal to three or more.
+\textbf{This is only true for 75\% missingness:} If we look at the
+estimated proportion of explained variance in outcome variable Y we see
+that the coefficient of determination is underestimate estimated in
+conditions where the number of iterations is equal to two or less, and
+slightly overestimated in conditions where the number of iterations is
+equal to three or more.
 
@@
 obtain unbiased results.
-
---\textgreater{}
 
@@
 overdisp) or due to the Qs (lm reg coeff, not higher dimensional/more
-complex RQs). Add what \%miss has to do with it.})\\
+complex RQs). Add what \%miss has to do with it.})
+
 \(\widehat{R}\) and autocorrelation indicate that algorithmic
 convergence may only be reached after twenty or even forty iterations,
-while unbiased, confidence valid estimates estimates may be obtained
-with as little as four iterations. These results are in agreement with
-the simulation hypothesis: \(\widehat{R}\) over-estimates the severity
-of non-convergence when applied to MI procedures. \%This may be due to
-the quantity of scientific interest chosen. More `complicated' \(Q\)s
-(e.g., higher order effects or variance components) might show bias,
-under- or over-coverage at higher \(T\).
+while unbiased, confidence valid estimates may be obtained with as
+little as four iterations. These results are in agreement with the
+simulation hypothesis: \(\widehat{R}\) over-estimates the severity of
+non-convergence when applied to MI procedures. \%This may be due to the
+quantity of scientific interest chosen. More `complicated' \(Q\)s (e.g.,
+higher-order effects or variance components) might show bias, under- or
+over-coverage at higher \(T\).
 
@@
 diagnostic to indicate that they should keep iterating. The other is the
-default in mice and other software packages, and yet another is \ldots{}
+default in mice and other software packages, and yet another is\ldots{}
 i forgot}) Under the relatively easy missing data problem of the current
@@
 The negative \(AC\)-values obtained in this study show no threat of
-non-stationarity. However, initial dip in \(AC\)-values may have
-implications for the default number of iterations in \texttt{mice}
-(\texttt{maxit\ =\ 5}). Terminating the algorithm at \(T=5\) may not be
-the most appropriate, since this lead to the worst convergence
-(\textbf{nope, only for Rh, not \(AC\)}), as indicated by
-\(\widehat{R}\) and \(AC\). Under the current specifications, \(T>20\)
-would be more appropriate.
-
-The observed dip in \(AC\) implies that default maxit value of five
-iterations is the worst possible number of iterations. Moreover, the
-results of this study imply that assessing the stationarity component of
-convergence with \(AC\) might be redundant.
+non-stationarity. However, the initial dip in \(AC\)-values
+\textbf{(which disappeared!)} may have implications for the default
+number of iterations in \texttt{mice} (\texttt{maxit\ =\ 5}).
+Terminating the algorithm at \(T=5\) may not be the most appropriate,
+since this lead to the worst convergence (\textbf{nope, only for Rh, not
+\(AC\)}), as indicated by \(\widehat{R}\) and \(AC\). Under the current
+specifications, \(T>20\) would be more appropriate.
 
@@
 
-\textbf{For empirical researchers: 1) Check trace plots for pathological
-non convergence and adjust imputation model if necessary. 2) use
-\(\widehat{R}\) wait 1.1 ash threshold and autocorrelation with }?** as
-threshold . Keep iterating until these thresholds are reached. 3) Do not
-use the R function ACF. Instead compute autocorrelations manually , see
-for example GitHub link. 4) Track your own scalar summary of interest.
-This is advanced but explained in Van Buren 2018. Compute
-\(\widehat{R}\) and autocorrelation values for this scalar summary. **
+For empirical researchers: 1) Check trace plots for pathological non
+convergence and adjust imputation model if necessary. 2) use
+\(\widehat{R}\) wait 1.1 ash threshold and autocorrelation with
+\textbf{?} as threshold. Keep iterating until these thresholds are
+reached. 3) Do not use the R function ACF. Instead, compute
+autocorrelations manually \textbf{(see e.g., {[}GitHub link{]})}. 4)
+Track your own scalar summary of interest. This is somewhat advanced but
+explained in Van Buren 2018. Compute \(\widehat{R}\) and autocorrelation
+values for this scalar summary. 5) \textbf{Something} about the novel
+\(\theta\) that is `substantive model-independent'.
 
@@
 \textbf{Pmm, M(N)AR, Empirical data, Vignette, ShinyMICE, AC across
-imputations, not iterations (unique for imp data??)}
+imputations, not iterations (new convergence diagnostic unique for
+iterative imputation??).}
 
 \textbf{If I don't include it in my study, also: Significance of \(AC\)
-values}
+values.}
 

### 2020-04-27 3.Thesis/1.SimulationStudy/Writeup/images/diagram.Rnw
After textual feedback Gerko

--- 2020-04-22 3.Thesis/1.SimulationStudy/Writeup/images/diagram.tex
@@
 
-\usepackage{Sweave}
 \begin{document}
@@
 \node[main] (mids2) [above=10mm of mids] {$\{y_{obs}, y_{imp}\}$ };
-\node[main] (mids3) [below=10mm of mids,label=below: Completed data (per $m$)] { $\{y_{obs}, y_{imp}\}$};
+\node[main] (mids3) [below=10mm of mids,label=below: $M$ completed datasets] { $\{y_{obs}, y_{imp}\}$};
 \node[main] (mira) [right=of mids] {\large $\hat{Q}$};
 \node[main] (mira2) [above=10mm of mira] {\large $\hat{Q}$ };
-\node[main] (mira3) [below=10mm of mira,label=below: Results (per $m$)] {\large $\hat{Q}$ };
+\node[main] (mira3) [below=10mm of mira,label=below: $M$ results] {\large $\hat{Q}$ };
 \node[main, fill = white!100] (mipo) [right=of mira,label=below:Pooled results] {\large $\bar{Q}$ };

### 2020-04-27 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
After textual feedback Gerko

@@
 email: h.i.oberman@uu.nl
-abstract: "(**Rewrite:**) Multiple imputation by chained equations (MICE) is a widely used tool to accommodate missing data.
-While empirical evidence supports the validity of inferences obtained using mice, there is no consensus on the convergence properties of the method.
-This paper provides insight into non-convergence of mice algorithms."
+abstract: "(**Rewrite:**) Iterative imputation by chained equations (MICE) is a widely used tool to accommodate missing data.
+While it is widely accepted that valid inferences can be obtained with this technique, these inferences all rely on algorithmic convergence.
+There is no consensus on how to evaluate the convergence properties of the method.
+This paper provides insight into identifying non-convergence of iterative impuation algorithms."
 keywords: MICE; convergence 
@@
 Anyone who analyzes person-data may run into a missing data problem.
-Missing data is not only ubiquitous but also tedious.
+Missing data is not only ubiquitous but also **tedious**.
 If a dataset contains just one incomplete observation, statistical inferences are undefined and will not produce any results.
@@
 This technique is known as 'multiple imputation' [MI; @rubin76].
-MI yields unbiased and confidence valid estimates of the true---but missing---data inference under many circumstances [@buur18].
+MI has proven to be a powerful technique to yield unbiased and confidence valid estimates of the true---but missing---data inference under many circumstances [@buur18].
 
@@
 The imputed data ($y_{imp}$) is combined with the observed data ($y_{obs}$) to create $M$ completed datasets.
-On each completed dataset, the analysis of scientific interest (the 'substantive model', or 'complete data model') is performed.
+On each completed dataset, the analysis of scientific interest is performed.
 The quantity of scientific interest (e.g., a regression coefficient) is denoted with $Q$.
@@
 These $M$ values are combined into a single pooled estimate $\bar{Q}$.
-**ALSO: make clear that iteration happens within the first step, between incomplete and completed data.**
 
@@
 These are all open questions, because the convergence properties of iterative imputation algorithms have not been systematically studied [@buur18].
-Moreover, there is no scientific consensus on the convergence properties of MI algorithms [@taka17].
+Moreover, there is no scientific consensus on how to evaluate convergence of MI algorithms [@taka17].
 Some default MICE techniques (e.g., 'predictive mean modeling') might not yield converged states at all [@murr18].
@@
 In this paper, we investigate how non-convergence in iterative imputation algorithms may be diagnosed, how well these methods perform, and at which point convergence may safely be assumed.
-For reasons of brevity, we only focus on the MI algorithm implemented in the world-leading MI software: the `R` [@R] package `mice` [@mice].
-The convergence properties of this MI algorithm are investigated through model-based simulation.
+For reasons of brevity, we only focus on the MI algorithm implemented in the popular `mice` package [@mice] in `R` [@R].
+The convergence properties of the MICE algorithm are investigated through model-based simulation.
 The results of this simulation study are guidelines for assessing convergence of MI algorithms, which will aid applied researchers in drawing valid inference from incomplete datasets.
@@
 Are the default number of iterations sufficient (i.e., 5 in mice, 10 in SPSS and Stata, 30 in mi)?
+**How severe is it when the algorithm has not converged?
+And what are guidelines for practice?**
 
@@
 
-## Terminology
+## Terminology/Notation
 
@@
 
-# How may non-convergence be diagnosed?
+# Identifying non-convergence
 
-The two challenges for convergence of iterative algorithms are mixing and stationarity [@gelm13].
-Without mixing, imputation chains do not intermingle nicely.
-Without stationarity, there is trending within imputation chains, see Figure \ref{fig:non-conv}.
+There are two requirements for convergence of iterative algorithms: mixing and stationarity [@gelm13].
+Without mixing, imputation chains do not intermingle nicely, **indicating that ...**.
+Without stationarity, there is trending within imputation chains, which implies that further iterations would yiled a different set of imputations.
+In Figure \ref{fig:non-conv}...
 **Explain what we see, namely example by @buur18 reproduced, showing the traceplots of chain means for some variable.
@@
 Each line is an imputation.
-In the first plot, the lines intermingle nicely and there is little trending.
-In the second, there is a lot of trending and the lines do not cross much.
+In the first plot, the chains intermingle nicely and there is little to no trending.
+In the second plot, there is a lot of trending and some chains do not intermingle.
 Importantly, the chain means at the last iteration (the imputed value per $m$) are very different between the two plots.
@@
 As recommended [e.g., @cowl96, p.
-898], $AC$ and $\widehat{R}$ will be used.
+898], $AC$ and $\widehat{R}$ will be used **to diagnose non-convergence**.
 
@@
 
-From the between- and within-chain variances we compute a weighted average, $\widehat{\operatorname{var}}^{+}$, which over-estimates the total variance of $\theta$ **remove or explain why**.
+From the between- and within-chain variances we compute a weighted average, $\widehat{\operatorname{var}}^{+}$, which over-estimates the total variance of $\theta$ **remove or explain why, or leave in**.
 $\widehat{R}$ is then obtained as a ratio between the over-estimated total variance and the within-chain variance:
@@
 As convergence diagnostic, the interest is therefore in positive $AC$-values.
-**Maybe remove:** Moreover, the magnitude of $AC$-values may be evaluated statistically, but that is outside of this note's scope.
+**Maybe remove:** Moreover, the magnitude of $AC$-values may be evaluated statistically, but that is outside of our scope.
 
 ## In practice/Answer
-**To assess whether $\widehat{R}$ and $AC$ may be appropriate non-convergence diagnostics for MICE algorithms, the following condition must hold when applied on the two algorithms plotted above: the methods indicate worse performance for the mis-specified model with pathological non-convergence (i.e., higher $\widehat{R}$- and $AC$-values than the typical performance).
+**To assess whether $\widehat{R}$ and $AC$ may be appropriate non-convergence identifiers for iterative imputation algorithms, the following condition must hold when applied on the two algorithms plotted above: the methods indicate worse performance for the mis-specified model with pathological non-convergence (i.e., higher $\widehat{R}$- and $AC$-values than the typical performance).
 And additionally the methods should reflect the increasing convergence in the typical convergence situation as the number of iterations goes up.**
 
-**In figure \ref{fig:diagnostics}A (**add panel labels**), the chain means are plotted again, now together.
-So are the non-convergence diagnostics.
+**In figure \ref{fig:diagnostics}A (**add panel labels**), the chain means are plotted again, now together---as are the non-convergence diagnostics.
 Panel B shows $\widehat{R}$ as computed by implementing @veht19 's recommendations.
-As required, $\widehat{R}$ indicates less signs of non-convergence as the number of iterations goes up in the typical convergence situation.
+As required, $\widehat{R}$ indicates less signs of non-convergence as the number of iterations goes up **in the typical convergence situation**.
 The superior performance of the typical convergence over the pathological non-convergence is less prominent, even flipped for $t<4$.
-Panel C displays the $AC$ as computed with the R function stats::acf().
-When we look at this panel, we must conclude that something weird is going on.
+Panel C displays the $AC$ as computed with the R function `stats::acf()`.
+When we look at this panel, we conclude something weird.
 The $AC$-values indicate equal performance (up-to $t=5$) for the typical convergence and the pathological non-convergence, while there is obvious trending in the latter.
 Moreover, the best convergence (as indicated by the lowest $AC$-value) is observed at $t=2$, but looking at the chain means in panel A, there should be some signs of trending up-to iteration number seven.
-After consulting the documentation on stats::acf(), we conclude that this implementation of $AC$ is not suitable for iterative imputation algorithms.
+After consulting the documentation on stats::acf() [@R], we conclude that this implementation of $AC$ is not suitable for iterative imputation algorithms.
 There is a correction factor for a mathematical shortcut that works in the limit.
-According to...
-the function works when the number of iterations > 50.
-The default number of iterations in mice, however, is 5.
+According to @box15, the function works when the number of iterations > 50.
+The default number of iterations in iterative imputation, however, is often lower.
 Therefore, we compute $AC$ manually, see panel D.
@@
 
-```{r diagnostics, fig.cap="Convergence diagnostics pathological non-convergence versus typical convergence."}
+```{r diagnostics, fig.cap="Convergence diagnostics pathological non-convergence versus typical convergence. **ADD PANEL LABELS**"}
 load("./images/diagnostics_plot.Rdata")
@@
 
-Convergence is reached when there is no dependency between subsequent iterations of imputation chains ($AC = 0$), and chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm ($\widehat{R} = 1$).
-We expect that a completely converged state will not be reached.
+For multiple imputation algorithms, it holds that convergence is reached when there is no dependency between subsequent iterations of imputation chains ($AC = 0$), and chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm ($\widehat{R} = 1$).
+**We expect that a completely converged state will not be reached.**
 
@@
 We assess the performance of the two convergence diagnostics against the recommended evaluation criteria for MI methods [i.e., average bias, average confidence interval width, and empirical coverage rate across simulations; @buur18, $\S$ 2.5.2].
-That is, there is no baseline measure available to evaluate performance against.
+**That is, there is no baseline measure available to evaluate performance against.**
 <!-- We could also look at distributional characteristics, and plausibility of imputed values, see @vinknd. For now, this is outside of the scope of this study. -->
@@
 437].
-In `mice`, initial values are chosen randomly from the observed data.
+In e.g.
+`mice`, initial values are chosen randomly from the observed data.
 Therefore, we cannot be certain that the initial values are over-dispersed.
@@
 Namely, the first eigenvalue of the variance-covariance matrix of the $M$ completed datasets.
+**Explain model-dependent...**
 
@@
 
-The performance of two methods ($\widehat{R}$ and $AC$) on each of these scalar summaries of interest will be evaluated with several estimands (quantities of scientific interest; $Q$s).
-It is assumed that when $\bar{Q}$ (the pooled result across imputations) is an unbiased (and confidence valid) estimate of $Q$, the algorithm is sufficiently converged.
+The performance of two methods ($\widehat{R}$ and $AC$) on each of these scalar summaries of interest will be evaluated with several estimands, i.e.
+quantities of scientific interest $Q$.
+It is assumed that when $\bar{Q}$ (the pooled result across imputations) is an unbiased and confidence valid estimate of $Q$, the algorithm is sufficiently converged.
 For each scalar $\theta$, one estimand $Q$ is defined that may be of interest in empirical research.
@@
 
-Upon convergence, $\widehat{R}=1$ and $AC=0$, but in MCMC algorithms these are very unlikely (because it's a distribution, not a point to converge to).
-In practice, thresholds are used to diagnose non-convergence.
-Usually, non-convergence is diagnosed when $\widehat{R}$ > 1.2 or 1.1 or even 1.01.
+Upon convergence, $\widehat{R}=1$ and $AC=0$, which are unlikely thresholds for MCMC algorithms, because **it's** convergence to a distribution.
+In practice,  non-convergence is usually diagnosed when $\widehat{R}$ > 1.2 or 1.1 or even 1.01.
 And a t-test is performed to assess whether $AC$ is significantly different from zero.
@@
 
-**Move this to Results:** As expected, complete convergence ($\widehat{R}=1$ and $AC=0$) is not observed.
+**Move this to Results:** As expected, complete convergence ($\widehat{R}=1$ and $AC=0$) is never observed.
 But what about the thresholds for practice?
@@
 1. simulate data 
+for (number of simulation runs from 1 to 1000)
  for (missingness proportions 5%, 25%, 50%, 75% and 95%)
- for (number of simulation runs from 1 to 1000)
   2. create missingness
@@
    3. impute missingness
-   4. compute convergence diagnostics
-   5. perform analysis of scientific interest
+   4. perform analysis of scientific interest
+   5. compute convergence diagnostics 
    6. pool results across imputations
    7. compute performance measures
- 8. aggregate outcomes across simulation runs 
-9. combine outcomes of all missingness proportions
+ 8. combine outcomes of all missingness proportions
+9. aggregate outcomes across simulation runs
+
 ```
@@
 
-The data-generating mechanism is a multivariate normal distribution, representing person data on three independent variables (from an unspecified social scientific field of study). Let
+The data-generating mechanism is a multivariate normal distribution, representing person data on three independent variables (from an unspecified social scientific field of study). Let **what?**
 
@@
 In this study, sampling variance is not of interest.
-Therefore, a single complete set may serve as comparative truth in all simulation runs [@vinknd].
-A finite population of $N=1000$ is simulated using the function `mvtnorm::rmvnorm()`.
+Therefore, a single complete set may serve as comparative truth in all simulation runs [@vink14].
+A finite population of $N=1000$ is simulated using the `mvtnorm` package [@mvtnorm].
 Subsequently, a fourth variable is constructed to serve as dependent variable in a multiple linear regression problem.
@@
 
-We consider four quantities of scientific interest ['conceptual estimands'; @morr19], namely...
-solve a multiple linear regression problem, where dependent variable $Y$ is regressed on independent variables $X_1$, $X_2$ and $X_3$ (**check notation with betas here, and add what the quantity/-ies of scientific interest is/are!
-Namely, the descriptive statistics of all variables (mean and standard deviation), $\beta_2$, and the percentage of variance in $Y$ explained (coefficient of determination; $R^2$)**):
+**We consider four quantities of scientific interest ['conceptual estimands'; @morr19], namely the descriptive statistics of all variables (mean and standard deviation), $\beta_2$, and the percentage of variance in $Y$ explained (coefficient of determination; $R^2$).
+We solve a multiple linear regression problem, where dependent variable $Y$ is regressed on independent variables $X_1$, $X_2$ and $X_3$**
 
@@
 
-The complete data is 'amputed' once for each simulation repetition with function `mice::ampute()`.
-(**change this to reflect current simulation set-up with 5, 25, 50, 75, and 95% of cases having missing data:** The missingness is univariate, and the probability to be missing is the same for all four variables, namely 20\% (`prop = 0.8, mech = "MCAR"`).
+The complete data is *amputed* once for each simulation repetition with function `mice::ampute()`.
+
+(**change this to reflect current simulation set-up with 5, 25, 50, 75, and 95% of cases having missing data: The missingness is univariate, and the probability to be missing is the same for all four variables, namely 20\% (`prop = 0.8, mech = "MCAR"`).
 This leaves 20\% of the rows completely observed).
-This study only considers only an MCAR missingness mechanism.
-Therefore, results may not be extrapolated to other missing data problems.
-Proper performance of the convergence diagnostics under MCAR is necessary but not sufficient to demonstrate the appropriateness of $\widehat{R}$ and $AC$ as convergence diagnostics.
-Convergence diagnostics should therefore at least apply to the MCAR situation, before more complex missingness should be explored.
+**
 
-Missing datapoints in $y$ are imputed with the function `mice::mice()`.
+Proper performance of the convergence diagnostics under at least missing completely at random (MCAR) missingness mechanism is necessary to demonstrate the appropriateness of $\widehat{R}$ and $AC$ as convergence diagnostics.
+However, results may not be extrapolated to other missingness mechanisms.
+**Convergence diagnostics should therefore at least apply to the MCAR situation, before more complex missingness should be explored.**
+
+Missing datapoints in $y$ are imputed with the `mice` package [@mice].
 All MI procedures are performed with Bayesian linear regression imputation (`method = "norm"`), and five imputation chains (`m = 5`).

### 2020-04-27 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
After textual feedback Gerko

@@
 \begin{abstract}
-(\textbf{Rewrite:}) Multiple imputation by chained equations (MICE) is a
-widely used tool to accommodate missing data. While empirical evidence
-supports the validity of inferences obtained using mice, there is no
-consensus on the convergence properties of the method. This paper
-provides insight into non-convergence of mice algorithms.
+(\textbf{Rewrite:}) Iterative imputation by chained equations (MICE) is
+a widely used tool to accommodate missing data. While it is widely
+accepted that valid inferences can be obtained with this technique,
+these inferences all rely on algorithmic convergence. There is no
+consensus on how to evaluate the convergence properties of the method.
+This paper provides insight into identifying non-convergence of
+iterative impuation algorithms.
 \end{abstract}
@@
 Anyone who analyzes person-data may run into a missing data problem.
-Missing data is not only ubiquitous but also tedious. If a dataset
-contains just one incomplete observation, statistical inferences are
-undefined and will not produce any results. To circumvent this, many
+Missing data is not only ubiquitous but also \textbf{tedious}. If a
+dataset contains just one incomplete observation, statistical inferences
+are undefined and will not produce any results. To circumvent this, many
 statistical packages employ list-wise deletion by default (i.e.,
@@
 uncertainty in the data due to missingness. This technique is known as
-`multiple imputation' \citep[MI;][]{rubin76}. MI yields unbiased and
-confidence valid estimates of the true---but missing---data inference
-under many circumstances \citep{buur18}.
+`multiple imputation' \citep[MI;][]{rubin76}. MI has proven to be a
+powerful technique to yield unbiased and confidence valid estimates of
+the true---but missing---data inference under many circumstances
+\citep{buur18}.
 
@@
 \(M\) completed datasets. On each completed dataset, the analysis of
-scientific interest (the `substantive model', or `complete data model')
-is performed. The quantity of scientific interest (e.g., a regression
-coefficient) is denoted with \(Q\). Since \(Q\) is estimated on each
-completed dataset, \(M\) separate \(\hat{Q}\)-values are obtained. These
-\(M\) values are combined into a single pooled estimate \(\bar{Q}\).
-\textbf{ALSO: make clear that iteration happens within the first step,
-between incomplete and completed data.}
+scientific interest is performed. The quantity of scientific interest
+(e.g., a regression coefficient) is denoted with \(Q\). Since \(Q\) is
+estimated on each completed dataset, \(M\) separate \(\hat{Q}\)-values
+are obtained. These \(M\) values are combined into a single pooled
+estimate \(\bar{Q}\).
 
@@
 been systematically studied \citep{buur18}. Moreover, there is no
-scientific consensus on the convergence properties of MI algorithms
+scientific consensus on how to evaluate convergence of MI algorithms
 \citep{taka17}. Some default MICE techniques (e.g., `predictive mean
@@
 and at which point convergence may safely be assumed. For reasons of
-brevity, we only focus on the MI algorithm implemented in the
-world-leading MI software: the \texttt{R} \citep{R} package
-\texttt{mice} \citep{mice}. The convergence properties of this MI
-algorithm are investigated through model-based simulation. The results
-of this simulation study are guidelines for assessing convergence of MI
-algorithms, which will aid applied researchers in drawing valid
-inference from incomplete datasets.
+brevity, we only focus on the MI algorithm implemented in the popular
+\texttt{mice} package \citep{mice} in \texttt{R} \citep{R}. The
+convergence properties of the MICE algorithm are investigated through
+model-based simulation. The results of this simulation study are
+guidelines for assessing convergence of MI algorithms, which will aid
+applied researchers in drawing valid inference from incomplete datasets.
 
@@
 Are the default number of iterations sufficient (i.e., 5 in mice, 10 in
-SPSS and Stata, 30 in mi)?
+SPSS and Stata, 30 in mi)? \textbf{How severe is it when the algorithm
+has not converged? And what are guidelines for practice?}
 
-\hypertarget{terminology}{%
-\subsection{Terminology}\label{terminology}}
+\hypertarget{terminologynotation}{%
+\subsection{Terminology/Notation}\label{terminologynotation}}
 
@@
 
-\hypertarget{how-may-non-convergence-be-diagnosed}{%
-\section{How may non-convergence be
-diagnosed?}\label{how-may-non-convergence-be-diagnosed}}
+\hypertarget{identifying-non-convergence}{%
+\section{Identifying
+non-convergence}\label{identifying-non-convergence}}
 
-The two challenges for convergence of iterative algorithms are mixing
-and stationarity \citep{gelm13}. Without mixing, imputation chains do
-not intermingle nicely. Without stationarity, there is trending within
-imputation chains, see Figure \ref{fig:non-conv}. \textbf{Explain what
+There are two requirements for convergence of iterative algorithms:
+mixing and stationarity \citep{gelm13}. Without mixing, imputation
+chains do not intermingle nicely, \textbf{indicating that \ldots{}}.
+Without stationarity, there is trending within imputation chains, which
+implies that further iterations would yiled a different set of
+imputations. In Figure \ref{fig:non-conv}\ldots{} \textbf{Explain what
 we see, namely example by \citet{buur18} reproduced, showing the
@@
 of a mis-specified imputation model. Each line is an imputation. In the
-first plot, the lines intermingle nicely and there is little trending.
-In the second, there is a lot of trending and the lines do not cross
-much. Importantly, the chain means at the last iteration (the imputed
-value per \(m\)) are very different between the two plots. The algorithm
-with the mis-specified model yields imputed values that are on average a
-magnitude two larger than those of the typically converged algorithm.
-This shows the importance of reaching converged states in iterative
-imputation algorithms.}
+first plot, the chains intermingle nicely and there is little to no
+trending. In the second plot, there is a lot of trending and some chains
+do not intermingle. Importantly, the chain means at the last iteration
+(the imputed value per \(m\)) are very different between the two plots.
+The algorithm with the mis-specified model yields imputed values that
+are on average a magnitude two larger than those of the typically
+converged algorithm. This shows the importance of reaching converged
+states in iterative imputation algorithms.}
 
@@
 separately in this study. As recommended \citep[e.g.,][p.~898]{cowl96},
-\(AC\) and \(\widehat{R}\) will be used.
+\(AC\) and \(\widehat{R}\) will be used \textbf{to diagnose
+non-convergence}.
 
@@
 average, \(\widehat{\operatorname{var}}^{+}\), which over-estimates the
-total variance of \(\theta\) \textbf{remove or explain why}.
-\(\widehat{R}\) is then obtained as a ratio between the over-estimated
-total variance and the within-chain variance:
+total variance of \(\theta\) \textbf{remove or explain why, or leave
+in}. \(\widehat{R}\) is then obtained as a ratio between the
+over-estimated total variance and the within-chain variance:
 
@@
 \textbf{Maybe remove:} Moreover, the magnitude of \(AC\)-values may be
-evaluated statistically, but that is outside of this note's scope.
+evaluated statistically, but that is outside of our scope.
 
@@
 \textbf{To assess whether \(\widehat{R}\) and \(AC\) may be appropriate
-non-convergence diagnostics for MICE algorithms, the following condition
-must hold when applied on the two algorithms plotted above: the methods
-indicate worse performance for the mis-specified model with pathological
-non-convergence (i.e., higher \(\widehat{R}\)- and \(AC\)-values than
-the typical performance). And additionally the methods should reflect
-the increasing convergence in the typical convergence situation as the
-number of iterations goes up.}
+non-convergence identifiers for iterative imputation algorithms, the
+following condition must hold when applied on the two algorithms plotted
+above: the methods indicate worse performance for the mis-specified
+model with pathological non-convergence (i.e., higher \(\widehat{R}\)-
+and \(AC\)-values than the typical performance). And additionally the
+methods should reflect the increasing convergence in the typical
+convergence situation as the number of iterations goes up.}
 
 \textbf{In figure \ref{fig:diagnostics}A (}add panel labels\textbf{),
-the chain means are plotted again, now together. So are the
+the chain means are plotted again, now together---as are the
 non-convergence diagnostics. Panel B shows \(\widehat{R}\) as computed
@@
 \(\widehat{R}\) indicates less signs of non-convergence as the number of
-iterations goes up in the typical convergence situation. The superior
-performance of the typical convergence over the pathological
+iterations goes up }in the typical convergence situation\textbf{. The
+superior performance of the typical convergence over the pathological
 non-convergence is less prominent, even flipped for \(t<4\). Panel C
-displays the \(AC\) as computed with the R function stats::acf(). When
-we look at this panel, we must conclude that something weird is going
-on. The \(AC\)-values indicate equal performance (up-to \(t=5\)) for the
-typical convergence and the pathological non-convergence, while there is
-obvious trending in the latter. Moreover, the best convergence (as
-indicated by the lowest \(AC\)-value) is observed at \(t=2\), but
+displays the \(AC\) as computed with the R function
+\texttt{stats::acf()}. When we look at this panel, we conclude something
+weird. The \(AC\)-values indicate equal performance (up-to \(t=5\)) for
+the typical convergence and the pathological non-convergence, while
+there is obvious trending in the latter. Moreover, the best convergence
+(as indicated by the lowest \(AC\)-value) is observed at \(t=2\), but
 looking at the chain means in panel A, there should be some signs of
 trending up-to iteration number seven. After consulting the
-documentation on stats::acf(), we conclude that this implementation of
-\(AC\) is not suitable for iterative imputation algorithms. There is a
-correction factor for a mathematical shortcut that works in the limit.
-According to\ldots{} the function works when the number of iterations
-\textgreater{} 50. The default number of iterations in mice, however, is
-5. Therefore, we compute \(AC\) manually, see panel D. The AC values in
-this plot do meet the requirements (\ldots) and will therefore be used
-in the simulation study.}
+documentation on stats::acf() \citep{R}, we conclude that this
+implementation of \(AC\) is not suitable for iterative imputation
+algorithms. There is a correction factor for a mathematical shortcut
+that works in the limit. According to \citet{box15}, the function works
+when the number of iterations \textgreater{} 50. The default number of
+iterations in iterative imputation, however, is often lower. Therefore,
+we compute \(AC\) manually, see panel D. The AC values in this plot do
+meet the requirements (\ldots) and will therefore be used in the
+simulation study.}
 
@@
 
-\caption{Convergence diagnostics pathological non-convergence versus typical convergence.}\label{fig:diagnostics}
+\caption{Convergence diagnostics pathological non-convergence versus typical convergence. **ADD PANEL LABELS**}\label{fig:diagnostics}
 \end{figure}
@@
 
-Convergence is reached when there is no dependency between subsequent
-iterations of imputation chains (\(AC = 0\)), and chains intermingle
-such that the only difference between the chains is caused by the
-randomness induced by the algorithm (\(\widehat{R} = 1\)). We expect
-that a completely converged state will not be reached.
+For multiple imputation algorithms, it holds that convergence is reached
+when there is no dependency between subsequent iterations of imputation
+chains (\(AC = 0\)), and chains intermingle such that the only
+difference between the chains is caused by the randomness induced by the
+algorithm (\(\widehat{R} = 1\)). \textbf{We expect that a completely
+converged state will not be reached.}
 
@@
 interval width, and empirical coverage rate across simulations;][\(\S\)
-2.5.2]{buur18}. That is, there is no baseline measure available to
-evaluate performance against.
+2.5.2]{buur18}. \textbf{That is, there is no baseline measure available
+to evaluate performance against.}
 
@@
 convergence if initial values of the algorithm are not appropriately
-over-dispersed \citep[p.~437]{broo98}. In \texttt{mice}, initial values
-are chosen randomly from the observed data. Therefore, we cannot be
-certain that the initial values are over-dispersed. We expect this to
+over-dispersed \citep[p.~437]{broo98}. In e.g.~\texttt{mice}, initial
+values are chosen randomly from the observed data. Therefore, we cannot
+be certain that the initial values are over-dispersed. We expect this to
 have little effect on the hypothesized performance of \(\widehat{R}\)
@@
 algorithm. Namely, the first eigenvalue of the variance-covariance
-matrix of the \(M\) completed datasets.
+matrix of the \(M\) completed datasets. \textbf{Explain
+model-dependent\ldots{}}
 
@@
 these scalar summaries of interest will be evaluated with several
-estimands (quantities of scientific interest; \(Q\)s). It is assumed
+estimands, i.e.~quantities of scientific interest \(Q\). It is assumed
 that when \(\bar{Q}\) (the pooled result across imputations) is an
-unbiased (and confidence valid) estimate of \(Q\), the algorithm is
+unbiased and confidence valid estimate of \(Q\), the algorithm is
 sufficiently converged. For each scalar \(\theta\), one estimand \(Q\)
@@
 
-Upon convergence, \(\widehat{R}=1\) and \(AC=0\), but in MCMC algorithms
-these are very unlikely (because it's a distribution, not a point to
-converge to). In practice, thresholds are used to diagnose
-non-convergence. Usually, non-convergence is diagnosed when
+Upon convergence, \(\widehat{R}=1\) and \(AC=0\), which are unlikely
+thresholds for MCMC algorithms, because \textbf{it's} convergence to a
+distribution. In practice, non-convergence is usually diagnosed when
 \(\widehat{R}\) \textgreater{} 1.2 or 1.1 or even 1.01. And a t-test is
@@
 \textbf{Move this to Results:} As expected, complete convergence
-(\(\widehat{R}=1\) and \(AC=0\)) is not observed. But what about the
+(\(\widehat{R}=1\) and \(AC=0\)) is never observed. But what about the
 thresholds for practice? \(\widehat{R}\) \textless1.2 is not stringent
@@
    3. impute missingness
-   4. compute convergence diagnostics
-   5. perform analysis of scientific interest
+   4. perform analysis of scientific interest
+   5. compute convergence diagnostics 
    6. pool results across imputations
    7. compute performance measures
- 8. aggregate outcomes across simulation runs 
-9. combine outcomes of all missingness proportions
+ 8. combine outcomes of all missingness proportions
+9. aggregate outcomes across simulation runs 
 \end{verbatim}
@@
 representing person data on three independent variables (from an
-unspecified social scientific field of study). Let
+unspecified social scientific field of study). Let \textbf{what?}
 
@@
 Therefore, a single complete set may serve as comparative truth in all
-simulation runs \citep{vinknd}. A finite population of \(N=1000\) is
-simulated using the function \texttt{mvtnorm::rmvnorm()}. Subsequently,
-a fourth variable is constructed to serve as dependent variable in a
-multiple linear regression problem. Let \[
+simulation runs \citep{vink14}. A finite population of \(N=1000\) is
+simulated using the \texttt{mvtnorm} package \citep{mvtnorm}.
+Subsequently, a fourth variable is constructed to serve as dependent
+variable in a multiple linear regression problem. Let \[
 Y_i = 1 + 2X_{1i} +.5X_{2i} - X_{3i} + \epsilon_i ,
@@
 
-We consider four quantities of scientific interest \citep[`conceptual
-estimands';][]{morr19}, namely\ldots{} solve a multiple linear
-regression problem, where dependent variable \(Y\) is regressed on
-independent variables \(X_1\), \(X_2\) and \(X_3\) (\textbf{check
-notation with betas here, and add what the quantity/-ies of scientific
-interest is/are! Namely, the descriptive statistics of all variables
-(mean and standard deviation), \(\beta_2\), and the percentage of
-variance in \(Y\) explained (coefficient of determination; \(R^2\))}):
+\textbf{We consider four quantities of scientific interest
+\citep[`conceptual estimands';][]{morr19}, namely the descriptive
+statistics of all variables (mean and standard deviation), \(\beta_2\),
+and the percentage of variance in \(Y\) explained (coefficient of
+determination; \(R^2\)). We solve a multiple linear regression problem,
+where dependent variable \(Y\) is regressed on independent variables
+\(X_1\), \(X_2\) and \(X_3\)}
 
@@
 
-The complete data is `amputed' once for each simulation repetition with
-function \texttt{mice::ampute()}. (\textbf{change this to reflect
-current simulation set-up with 5, 25, 50, 75, and 95\% of cases having
-missing data:} The missingness is univariate, and the probability to be
-missing is the same for all four variables, namely 20\%
-(\texttt{prop\ =\ 0.8,\ mech\ =\ "MCAR"}). This leaves 20\% of the rows
-completely observed). This study only considers only an MCAR missingness
-mechanism. Therefore, results may not be extrapolated to other missing
-data problems. Proper performance of the convergence diagnostics under
-MCAR is necessary but not sufficient to demonstrate the appropriateness
-of \(\widehat{R}\) and \(AC\) as convergence diagnostics. Convergence
-diagnostics should therefore at least apply to the MCAR situation,
-before more complex missingness should be explored.
+The complete data is \emph{amputed} once for each simulation repetition
+with function \texttt{mice::ampute()}.
 
-Missing datapoints in \(y\) are imputed with the function
-\texttt{mice::mice()}. All MI procedures are performed with Bayesian
-linear regression imputation (\texttt{method\ =\ "norm"}), and five
-imputation chains (\texttt{m\ =\ 5}). The number of iterations varies
-between simulation conditions (\texttt{maxit\ =} \(1, 2, \dots, 100\)).
-Each repetition of the simulation thus starts from 1 complete dataset,
-then splits into 5 missingness conditions (\(p =.05,.25,.5,.75,.95\)),
-each of the amputed datasets is imputed 5 times (\(m = 1, 2,..., 5\))
-for each iteration condition (\(t = 1, 2,..., 100\)). From these
+(\textbf{change this to reflect current simulation set-up with 5, 25,
+50, 75, and 95\% of cases having missing data: The missingness is
+univariate, and the probability to be missing is the same for all four
+variables, namely 20\% (\texttt{prop\ =\ 0.8,\ mech\ =\ "MCAR"}). This
+leaves 20\% of the rows completely observed). }
+
+Proper performance of the convergence diagnostics under at least missing
+completely at random (MCAR) missingness mechanism is necessary to
+demonstrate the appropriateness of \(\widehat{R}\) and \(AC\) as
+convergence diagnostics. However, results may not be extrapolated to
+other missingness mechanisms. \textbf{Convergence diagnostics should
+therefore at least apply to the MCAR situation, before more complex
+missingness should be explored.}
+
+Missing datapoints in \(y\) are imputed with the \texttt{mice} package
+\citep{mice}. All MI procedures are performed with Bayesian linear
+regression imputation (\texttt{method\ =\ "norm"}), and five imputation
+chains (\texttt{m\ =\ 5}). The number of iterations varies between
+simulation conditions (\texttt{maxit\ =} \(1, 2, \dots, 100\)). Each
+repetition of the simulation thus starts from 1 complete dataset, then
+splits into 5 missingness conditions (\(p =.05,.25,.5,.75,.95\)), each
+of the amputed datasets is imputed 5 times (\(m = 1, 2,..., 5\)) for
+each iteration condition (\(t = 1, 2,..., 100\)). From these
 \(5\times5\times100\) imputations, we extract the \(\theta\)s to apply

### 2020-04-27 3.Thesis/1.SimulationStudy/Writeup/images/non-convergence.R

@@ -5,6 +5,7 @@
 library(mice)
 library(purrr)
 library(dplyr)
+library(ggplot2)
 library(patchwork)
 source('./3.Thesis/1.SimulationStudy/Functions/Convergence.R')
 source('./3.Thesis/1.SimulationStudy/Functions/AC.R')
@@ -30,7 +31,7 @@ nonc <- plot(imp.bmi1, "wgt")[1]
 nonc$layout <- c(1, 1)
 nonc[["condlevels"]][[".ms"]] <- c("", "sd")
 nonc$ylab <- "Chain mean"
-nonc$main <- list(label = "Pathological non-convergence", fontsize=12, fontface = 1)
+nonc$main <- list(label = "Non-convergence", fontsize=12, fontface = 1)
 nonc$x.scales$tick.number <-10
 save(nonc, file = "3.Thesis/1.SimulationStudy/Writeup/images/nonc_plot.Rdata")
 
@@ -149,7 +150,7 @@ ac <-
   geom_line(na.rm = TRUE) +
   scale_x_continuous(breaks = 1:10) +
   xlab("Iteration") +
-  ylab("AC (manual)") +
+  ylab("AC") +
   ggtitle(expression(paste("AC of ", theta, " (manual calculation)"))) + 
   theme(legend.position = "")
 

### 2020-04-28 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
Including feedback meeting Gerko

@@
 email: h.i.oberman@uu.nl
-abstract: "(**Rewrite:**) Iterative imputation by chained equations (MICE) is a widely used tool to accommodate missing data.
+abstract: "Iterative imputation is a popular tool to accommodate missing data.
 While it is widely accepted that valid inferences can be obtained with this technique, these inferences all rely on algorithmic convergence.
@@
 This paper provides insight into identifying non-convergence of iterative impuation algorithms."
-keywords: MICE; convergence 
+keywords: iterative imputation; non-convergence; mice
 classoption:
@@
 Anyone who analyzes person-data may run into a missing data problem.
-Missing data is not only ubiquitous but also **tedious**.
+Missing data is not only ubiquitous, but treating it can also be tedious.
 If a dataset contains just one incomplete observation, statistical inferences are undefined and will not produce any results.
@@
 Unfortunately, this *ad hoc* solution may yield wildly invalid results [@buur18].
-An alternative is to 'impute' (i.e., fill in) the missing values in the incomplete observations.
+An alternative is to *impute* (i.e., fill in) the missing values in the incomplete observations.
 Subsequently, statistical inferences can be performed on the completed dataset.
@@
 
-**Move this after terminology??** Figure \ref{fig:diagram} provides an overview of the steps involved with MI---from incomplete data, to $M$ completed datasets, to $M$ estimated quantities of interest ($\hat{Q}$s), to a single pooled estimate $\bar{Q}$.
-Missing data in $y$ is 'imputed' (i.e., filled in) $M$ times.
-The imputed data ($y_{imp}$) is combined with the observed data ($y_{obs}$) to create $M$ completed datasets.
+Figure \ref{fig:diagram} provides an overview of the steps involved with MI---from incomplete data, to $M$ multiply imputed datasets, to $M$ estimated quantities of interest $\hat{Q}$s, to a single pooled estimate $\bar{Q}$.
+Missing data in dataset $y$ is imputed $M$ times.
+The imputed data $y_{imp}$ is combined with the observed data $y_{obs}$ to create $M$ completed datasets.
 On each completed dataset, the analysis of scientific interest is performed.
@@
 These $M$ values are combined into a single pooled estimate $\bar{Q}$.
+
+<!-- Q is quantity of scientific interest: "A scientific estimand $Q$ is a quantity of scientific interest that we can calculate if we would observe the entire population" [@buur18, par 2.3.1] -->
 
@@
 
-Currently, the recommended practice for evaluating convergence is to visually inspect imputations for signs of non-convergence.
+Currently, the recommended practice for evaluating convergence in iterative imputation algorithms is to visually inspect imputations for signs of non-convergence.
+**(Add chain means and variances here already?
+Because it's insufficient that these are univariate?)**.
 This method is insufficient on two counts: 1) it may be challenging to the untrained eye, and 2) only severely pathological cases of non-convergence may be diagnosed [@buur18, $\S$ 6.5.2].
 Therefore, a quantitative, diagnostic evaluation of convergence would be preferred.
-
-Monitoring convergence of iterative imputation algorithms diagnostically is challenging.
+Yet, monitoring convergence of iterative imputation algorithms diagnostically is challenging.
 Iterative imputation algorithms such as MICE are Markov chain Monte Carlo (MCMC) methods.
@@
 Therefore, the aim of convergence diagnostics for MCMC methods is not to establish the point at which convergence is reached, but to monitor signs of non-convergence.
-Several of such diagnostics exist for MCMC methods, but it is not known whether these are appropriate for MICE.
+Several of such diagnostics exist for MCMC methods, but it is not known whether these are appropriate for iterative imputation algorithms.
 <!-- Therefore we can only diagnose non-convergence, not convergence [@hoff09]. "a weak diagnostics is better than no diagnostic at all" [@cowl96]. -->
@@
 In this paper, we investigate how non-convergence in iterative imputation algorithms may be diagnosed, how well these methods perform, and at which point convergence may safely be assumed.
-For reasons of brevity, we only focus on the MI algorithm implemented in the popular `mice` package [@mice] in `R` [@R].
+For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular `mice` package [@mice] in `R` [@R].
 The convergence properties of the MICE algorithm are investigated through model-based simulation.
@@
 **How severe is it when the algorithm has not converged?
-And what are guidelines for practice?**
+And what are guidelines for practice?
+Can the parameter of interest, estimand $Q$, be correct when the algorithm is not (yet) converged, and vice versa?**
 
@@
 
-## Terminology/Notation
+## Some notation
 
@@
 
-Let $y$ denote an $n \times p$ matrix containing the data values on $p$ variables for all $n$ units in a sample.
-The data value of unit $i$ ($i = 1, 2, \dots, n$) on variable $j$ ($j = 1, 2, \dots, p$) may be either observed or missing.
-The collection of observed data values in $y$ is denoted by $y_{obs}$; the missing part of $y$ is referred to as $y_{mis}$.
+Let $y$ denote an $n \times k$ matrix containing the data values on $k$ variables for all $n$ units in a sample.
+The data value of unit $i$ ($i = 1, 2, \dots, n$) on variable $j$ ($j = 1, 2, \dots, k$) may be either observed or missing.
+**The number of units $i$ with at least one missing data value can be divided by the total number of units $n$ to obtain the 'missingness proportion'  $p_{mis}$.** The collection of observed data values in $y$ is denoted by $y_{obs}$; the missing part of $y$ is referred to as $y_{mis}$.
 For each datapoint in $y_{mis}$, we sample $M \times T$ times plausible values, where $M$ is the number of imputations ($m = 1, 2, \dots, M$) and $T$ is the number of iterations ($t = 1, 2, \dots, T$).
 The collection of samples between the initial value (at $t=1$) and the final imputed value (at $t=T$) will be referred to as an 'imputation chain'.
-**Add: missingness percentage = proportion of cases with one or more missing values, Q is quantity of scientific interest: "A scientific estimand $Q$ is a quantity of scientific interest that we can calculate if we would observe the entire population" [@buur18, par 2.3.1], and $\theta$s are scalar summaries of interest in the iterative algorithm (e.g., chain means; the average of the imputed values in each imputation chain).
+**Add: $\theta$s are scalar summaries of interest in the iterative algorithm (e.g., chain means; the average of the imputed values in each imputation chain).
 **
@@
 With a recently proposed adaptation, $\widehat{R}$ might also serve to diagnose non-stationarity, but this has not yet been thoroughly investigated [@veht19].
-Therefore, mixing and stationarity will be evaluated separately in this study.
-As recommended [e.g., @cowl96, p.
-898], $AC$ and $\widehat{R}$ will be used **to diagnose non-convergence**.
+Therefore, use we $AC$ and $\widehat{R}$ to evaluate stationarity and mixing separately, as recommended by e.g., @cowl96, p.
+898.
 
@@
 
-To define $\widehat{R}$, we follow notation by [@veht19, p.
-5].
-Let $M$ be the total number of chains, $T$ the number of iterations per chain, and $\theta$ the scalar summary of interest (e.g., chain mean or chain variance).
+An updated version of $\widehat{R}$ has been proposed by @veht19, (p.
+5).
+**This version may be suitable for iterative imputation.** Let $M$ be the total number of chains, $T$ the number of iterations per chain, and $\theta$ the scalar summary of interest (e.g., chain mean or chain variance).
 For each chain ($m = 1, 2, \dots, M$), we estimate the variance of $\theta$, and average these to obtain within-chain variance $W$.
@@
 According to @box15, the function works when the number of iterations > 50.
-The default number of iterations in iterative imputation, however, is often lower.
+The default number of iterations in iterative imputation, however, is often much lower.
 Therefore, we compute $AC$ manually, see panel D.
 The AC values in this plot do meet the requirements (...) and will therefore be used in the simulation study.**
+
+**acf is niet bedoeld om weinig iteraties te... niet een fout in acf, alleen hier niet van toepassing**
 
@@
 
-## Simulation Hypothesis
+## Simulation Hypothesis (als bullets!!)
 
@@
 Based on an empirical finding [@lace07], we hypothesize that $\widehat{R}$ will over-estimate non-convergence of MI algorithms **explain why?**.
-The threshold of $\widehat{R} < 1.01$ will then be too stringent for diagnosing convergence.
+Lacerda concluded that rhat dropped below 1.2 after at least 50 iterations.
+Therefore...
+Vehtari's proposed threshold of $\widehat{R} < 1.01$ will then be too stringent for diagnosing convergence.
 This over-estimation may, however, be diminished because $\widehat{R}$ can falsely diagnose convergence if initial values of the algorithm are not appropriately over-dispersed [@broo98, p.
@@
 Therefore, we cannot be certain that the initial values are over-dispersed.
-We expect this to have little effect on the hypothesized performance of $\widehat{R}$ **explain that the randomness induced by the MICE algorithm would take care of this??**.
-**(Add actual hypothesis: we thought that mice would converge sooner than $\widehat{R}$ would indicate that it did)** No hypothesis was formulated about the performance of $AC$ as convergence diagnostic.
+We expect this to have little effect on the hypothesized performance of $\widehat{R}$.
+**(Add actual hypothesis: we thought that mice would converge sooner than $\widehat{R}$ would indicate that it did)** We hypothesize that high AC values are implausible in converged MI algorithms.
 **Added later: High $AC$-values are implausible in MI procedures.
@@
 
-The two challenges of convergence are stationarity and mixing.
-Mixing and stationarity can be inspected visually, by evaluating traceplots of scalar summaries of interest ($\theta$s; e.g., chain means and chain variances).
+**Move to section above** Mixing and stationarity have histroically been inspected visually, by evaluating traceplots of scalar summaries of interest ($\theta$s; e.g., chain means and chain variances).
 As @buur18 describes, users can also specify a model-specific scalar summary (e.g., a regression coefficient).
 A user-specified scalar summary, however, is not universal to all complete data problems.
-Therefore, as inspired by [@mack03], we will propose a $\theta$ that summarizes the multivariate state space of the algorithm.
+Therefore, as inspired by [@mack03], we propose a $\theta$ that summarizes the multivariate state space of the algorithm.
 Namely, the first eigenvalue of the variance-covariance matrix of the $M$ completed datasets.
-**Explain model-dependent...**
+**The first eigenvalue has the appealing property that is not dependent on the substantive model of interest.**
 
 The methods under evaluation are $\widehat{R}$ and $AC$.
-These will be computed for the three scalar summaries of interest: chain means (i.e., per imputation, per variable: on $y_{imp, m, j}$), chain variances (i.e., per imputation, per variable: on $y_{imp, m, j}$), and the first eigenvalue of the variance-covariance matrix across imputations (i.e., per imputation: on {$y_{obs}, y_{imp, m}$}) **add user-specified $\theta$: estimated regression coefficient per imputation**.
+These will be computed for the **four** scalar summaries of interest: chain means (i.e., per imputation, per variable: on $y_{imp, m, j}$), chain variances (i.e., per imputation, per variable: on $y_{imp, m, j}$), and the first eigenvalue of the variance-covariance matrix across imputations (i.e., per imputation: on {$y_{obs}, y_{imp, m}$}) **add user-specified $\theta$: estimated regression coefficient per imputation**.
 
@@
 
-Upon convergence, $\widehat{R}=1$ and $AC=0$, which are unlikely thresholds for MCMC algorithms, because **it's** convergence to a distribution.
+Upon convergence, $\widehat{R}=1$ and $AC=0$, which are unlikely thresholds for MCMC algorithms, because of its convergence to a distribution.
 In practice,  non-convergence is usually diagnosed when $\widehat{R}$ > 1.2 or 1.1 or even 1.01.
@@
 ```
-The aim of the simulation study is to evaluate the impact of inducing non-convergence by: 1) terminating the MICE algorithm at different imputation chain lengths ($t = 1, 2,..., 100$), and 2) varying the missingness proportions ($p_{miss} =.05,.25,.50,.75,.95$).
+The aim of the simulation study is to evaluate the impact of inducing non-convergence by: 1) terminating the MICE algorithm at different imputation chain lengths ($t = 1, 2,..., 100$), and 2) varying the missingness proportions ($p_{mis} =.05,.25,.50,.75,.95$).
 The assumption underlying the different number of iterations is that the algorithm generally does reach convergence at $t=1$, because the initial values are sampled randomly from the set of observed datapoints.
@@
 
-The data-generating mechanism is a multivariate normal distribution, representing person data on three independent variables (from an unspecified social scientific field of study). Let **what?**
+The data-generating mechanism is a multivariate normal distribution, representing person data on three predictor variables (from an unspecified social scientific field of study).
+Let our predictor space be defined by the following a multivariate normal distribution
 
@@
 A finite population of $N=1000$ is simulated using the `mvtnorm` package [@mvtnorm].
-Subsequently, a fourth variable is constructed to serve as dependent variable in a multiple linear regression problem.
+Subsequently, a fourth variable is constructed to serve as outcome variable in a multiple linear regression problem.
 Let
@@
 **We consider four quantities of scientific interest ['conceptual estimands'; @morr19], namely the descriptive statistics of all variables (mean and standard deviation), $\beta_2$, and the percentage of variance in $Y$ explained (coefficient of determination; $R^2$).
-We solve a multiple linear regression problem, where dependent variable $Y$ is regressed on independent variables $X_1$, $X_2$ and $X_3$**
+We solve a multiple linear regression problem, where outcome $Y$ is regressed on predictors $X_1$, $X_2$ and $X_3$**
 
@@
 The number of iterations varies between simulation conditions (`maxit = ` $1, 2, \dots, 100$).
-Each repetition of the simulation thus starts from 1 complete dataset, then splits into 5 missingness conditions ($p =.05,.25,.5,.75,.95$), each of the amputed datasets is imputed 5 times ($m = 1, 2,..., 5$) for each iteration condition ($t = 1, 2,..., 100$).
+Each repetition of the simulation thus starts from 1 complete dataset, then splits into 5 missingness conditions ($p_{mis} =.05,.25,.5,.75,.95$), each of the amputed datasets is imputed 5 times ($m = 1, 2,..., 5$) for each iteration condition ($t = 1, 2,..., 100$).
 From these $5\times5\times100$ imputations, we extract the $\theta$s to apply $\widehat{R}$ and $AC$ on (**i.e., chain means, chain variances, $\beta$s per imputation, and the first eigenvalue of the variance-covariance matrix per imputation**).
@@
 
-The estimator for each $Q$ is $\bar{Q}$---the pooled aggregate of the ($\hat{Q}$)s across imputations.
-To calculate the $\bar{Q}$s, we combine the observed data ($y_{obs}$) and the imputed data for each imputation ($y_{imp,m}$) with the function `mice::complete()`.
+The estimator for each $Q$ is $\bar{Q}$---the pooled aggregate of the $\hat{Q}$s across imputations.
+To calculate the $\bar{Q}$s, we combine the observed data $y_{obs}$ and the imputed data for each imputation $y_{imp,m}$ with the function `mice::complete()`.
 Descriptive statistics are computed as the average across imputations for each variable ($\mu_j$ and $\sigma_j$, where $j = Y, X_1, X_2, X_3$).

### 2020-04-28 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Including feedback meeting Gerko

@@
 \begin{abstract}
-(\textbf{Rewrite:}) Iterative imputation by chained equations (MICE) is
-a widely used tool to accommodate missing data. While it is widely
-accepted that valid inferences can be obtained with this technique,
-these inferences all rely on algorithmic convergence. There is no
-consensus on how to evaluate the convergence properties of the method.
-This paper provides insight into identifying non-convergence of
-iterative impuation algorithms.
+Iterative imputation is a popular tool to accommodate missing data.
+While it is widely accepted that valid inferences can be obtained with
+this technique, these inferences all rely on algorithmic convergence.
+There is no consensus on how to evaluate the convergence properties of
+the method. This paper provides insight into identifying non-convergence
+of iterative impuation algorithms.
 \end{abstract}
 
-\keywords{MICE; convergence}
+\keywords{iterative imputation; non-convergence; mice}
 
@@
 Anyone who analyzes person-data may run into a missing data problem.
-Missing data is not only ubiquitous but also \textbf{tedious}. If a
-dataset contains just one incomplete observation, statistical inferences
-are undefined and will not produce any results. To circumvent this, many
-statistical packages employ list-wise deletion by default (i.e.,
-ignoring incomplete observations). Unfortunately, this \emph{ad hoc}
-solution may yield wildly invalid results \citep{buur18}. An alternative
-is to `impute' (i.e., fill in) the missing values in the incomplete
-observations. Subsequently, statistical inferences can be performed on
-the completed dataset. By repeating this process several times, a
-distribution of plausible results may be obtained, which reflects the
-uncertainty in the data due to missingness. This technique is known as
-`multiple imputation' \citep[MI;][]{rubin76}. MI has proven to be a
-powerful technique to yield unbiased and confidence valid estimates of
-the true---but missing---data inference under many circumstances
-\citep{buur18}.
+Missing data is not only ubiquitous, but treating it can also be
+tedious. If a dataset contains just one incomplete observation,
+statistical inferences are undefined and will not produce any results.
+To circumvent this, many statistical packages employ list-wise deletion
+by default (i.e., ignoring incomplete observations). Unfortunately, this
+\emph{ad hoc} solution may yield wildly invalid results \citep{buur18}.
+An alternative is to \emph{impute} (i.e., fill in) the missing values in
+the incomplete observations. Subsequently, statistical inferences can be
+performed on the completed dataset. By repeating this process several
+times, a distribution of plausible results may be obtained, which
+reflects the uncertainty in the data due to missingness. This technique
+is known as `multiple imputation' \citep[MI;][]{rubin76}. MI has proven
+to be a powerful technique to yield unbiased and confidence valid
+estimates of the true---but missing---data inference under many
+circumstances \citep{buur18}.
 
-\textbf{Move this after terminology??} Figure \ref{fig:diagram} provides
-an overview of the steps involved with MI---from incomplete data, to
-\(M\) completed datasets, to \(M\) estimated quantities of interest
-(\(\hat{Q}\)s), to a single pooled estimate \(\bar{Q}\). Missing data in
-\(y\) is `imputed' (i.e., filled in) \(M\) times. The imputed data
-(\(y_{imp}\)) is combined with the observed data (\(y_{obs}\)) to create
-\(M\) completed datasets. On each completed dataset, the analysis of
-scientific interest is performed. The quantity of scientific interest
-(e.g., a regression coefficient) is denoted with \(Q\). Since \(Q\) is
-estimated on each completed dataset, \(M\) separate \(\hat{Q}\)-values
-are obtained. These \(M\) values are combined into a single pooled
-estimate \(\bar{Q}\).
+Figure \ref{fig:diagram} provides an overview of the steps involved with
+MI---from incomplete data, to \(M\) multiply imputed datasets, to \(M\)
+estimated quantities of interest \(\hat{Q}\)s, to a single pooled
+estimate \(\bar{Q}\). Missing data in dataset \(y\) is imputed \(M\)
+times. The imputed data \(y_{imp}\) is combined with the observed data
+\(y_{obs}\) to create \(M\) completed datasets. On each completed
+dataset, the analysis of scientific interest is performed. The quantity
+of scientific interest (e.g., a regression coefficient) is denoted with
+\(Q\). Since \(Q\) is estimated on each completed dataset, \(M\)
+separate \(\hat{Q}\)-values are obtained. These \(M\) values are
+combined into a single pooled estimate \(\bar{Q}\).
 
@@
 
-Currently, the recommended practice for evaluating convergence is to
-visually inspect imputations for signs of non-convergence. This method
-is insufficient on two counts: 1) it may be challenging to the untrained
-eye, and 2) only severely pathological cases of non-convergence may be
-diagnosed \citep[\(\S\) 6.5.2]{buur18}. Therefore, a quantitative,
-diagnostic evaluation of convergence would be preferred.
-
-Monitoring convergence of iterative imputation algorithms diagnostically
-is challenging. Iterative imputation algorithms such as MICE are Markov
-chain Monte Carlo (MCMC) methods. In MCMC methods, convergence is not
-from a scalar to a point but from one distribution to another. The
-values generated by the algorithm (e.g., imputed values) will vary even
-after convergence. Therefore, the aim of convergence diagnostics for
-MCMC methods is not to establish the point at which convergence is
-reached, but to monitor signs of non-convergence. Several of such
-diagnostics exist for MCMC methods, but it is not known whether these
-are appropriate for MICE.
+Currently, the recommended practice for evaluating convergence in
+iterative imputation algorithms is to visually inspect imputations for
+signs of non-convergence. \textbf{(Add chain means and variances here
+already? Because it's insufficient that these are univariate?)}. This
+method is insufficient on two counts: 1) it may be challenging to the
+untrained eye, and 2) only severely pathological cases of
+non-convergence may be diagnosed \citep[\(\S\) 6.5.2]{buur18}.
+Therefore, a quantitative, diagnostic evaluation of convergence would be
+preferred. Yet, monitoring convergence of iterative imputation
+algorithms diagnostically is challenging. Iterative imputation
+algorithms such as MICE are Markov chain Monte Carlo (MCMC) methods. In
+MCMC methods, convergence is not from a scalar to a point but from one
+distribution to another. The values generated by the algorithm (e.g.,
+imputed values) will vary even after convergence. Therefore, the aim of
+convergence diagnostics for MCMC methods is not to establish the point
+at which convergence is reached, but to monitor signs of
+non-convergence. Several of such diagnostics exist for MCMC methods, but
+it is not known whether these are appropriate for iterative imputation
+algorithms.
 
@@
 and at which point convergence may safely be assumed. For reasons of
-brevity, we only focus on the MI algorithm implemented in the popular
-\texttt{mice} package \citep{mice} in \texttt{R} \citep{R}. The
-convergence properties of the MICE algorithm are investigated through
-model-based simulation. The results of this simulation study are
-guidelines for assessing convergence of MI algorithms, which will aid
-applied researchers in drawing valid inference from incomplete datasets.
+brevity, we only focus on the iterative imputation algorithm implemented
+in the popular \texttt{mice} package \citep{mice} in \texttt{R}
+\citep{R}. The convergence properties of the MICE algorithm are
+investigated through model-based simulation. The results of this
+simulation study are guidelines for assessing convergence of MI
+algorithms, which will aid applied researchers in drawing valid
+inference from incomplete datasets.
 
@@
 SPSS and Stata, 30 in mi)? \textbf{How severe is it when the algorithm
-has not converged? And what are guidelines for practice?}
+has not converged? And what are guidelines for practice? Can the
+parameter of interest, estimand \(Q\), be correct when the algorithm is
+not (yet) converged, and vice versa?}
 
-\hypertarget{terminologynotation}{%
-\subsection{Terminology/Notation}\label{terminologynotation}}
+\hypertarget{some-notation}{%
+\subsection{Some notation}\label{some-notation}}
 
-Let \(y\) denote an \(n \times p\) matrix containing the data values on
-\(p\) variables for all \(n\) units in a sample. The data value of unit
+Let \(y\) denote an \(n \times k\) matrix containing the data values on
+\(k\) variables for all \(n\) units in a sample. The data value of unit
 \(i\) (\(i = 1, 2, \dots, n\)) on variable \(j\)
-(\(j = 1, 2, \dots, p\)) may be either observed or missing. The
-collection of observed data values in \(y\) is denoted by \(y_{obs}\);
-the missing part of \(y\) is referred to as \(y_{mis}\). For each
-datapoint in \(y_{mis}\), we sample \(M \times T\) times plausible
-values, where \(M\) is the number of imputations
-(\(m = 1, 2, \dots, M\)) and \(T\) is the number of iterations
-(\(t = 1, 2, \dots, T\)). The collection of samples between the initial
-value (at \(t=1\)) and the final imputed value (at \(t=T\)) will be
-referred to as an `imputation chain'. \textbf{Add: missingness
-percentage = proportion of cases with one or more missing values, Q is
-quantity of scientific interest: ``A scientific estimand \(Q\) is a
-quantity of scientific interest that we can calculate if we would
-observe the entire population'' \citep[par 2.3.1]{buur18}, and
-\(\theta\)s are scalar summaries of interest in the iterative algorithm
-(e.g., chain means; the average of the imputed values in each imputation
-chain). }
+(\(j = 1, 2, \dots, k\)) may be either observed or missing. \textbf{The
+number of units \(i\) with at least one missing data value can be
+divided by the total number of units \(n\) to obtain the `missingness
+proportion' \(p_{mis}\).} The collection of observed data values in
+\(y\) is denoted by \(y_{obs}\); the missing part of \(y\) is referred
+to as \(y_{mis}\). For each datapoint in \(y_{mis}\), we sample
+\(M \times T\) times plausible values, where \(M\) is the number of
+imputations (\(m = 1, 2, \dots, M\)) and \(T\) is the number of
+iterations (\(t = 1, 2, \dots, T\)). The collection of samples between
+the initial value (at \(t=1\)) and the final imputed value (at \(t=T\))
+will be referred to as an `imputation chain'. \textbf{Add: \(\theta\)s
+are scalar summaries of interest in the iterative algorithm (e.g., chain
+means; the average of the imputed values in each imputation chain). }
 
@@
 non-stationarity, but this has not yet been thoroughly investigated
-\citep{veht19}. Therefore, mixing and stationarity will be evaluated
-separately in this study. As recommended \citep[e.g.,][p.~898]{cowl96},
-\(AC\) and \(\widehat{R}\) will be used \textbf{to diagnose
-non-convergence}.
+\citep{veht19}. Therefore, use we \(AC\) and \(\widehat{R}\) to evaluate
+stationarity and mixing separately, as recommended by e.g.,
+\citet{cowl96}, p.~898.
 
@@
 
-To define \(\widehat{R}\), we follow notation by \citep[p.~5]{veht19}.
-Let \(M\) be the total number of chains, \(T\) the number of iterations
-per chain, and \(\theta\) the scalar summary of interest (e.g., chain
-mean or chain variance). For each chain (\(m = 1, 2, \dots, M\)), we
-estimate the variance of \(\theta\), and average these to obtain
-within-chain variance \(W\).
+An updated version of \(\widehat{R}\) has been proposed by
+\citet{veht19}, (p.~5). \textbf{This version may be suitable for
+iterative imputation.} Let \(M\) be the total number of chains, \(T\)
+the number of iterations per chain, and \(\theta\) the scalar summary of
+interest (e.g., chain mean or chain variance). For each chain
+(\(m = 1, 2, \dots, M\)), we estimate the variance of \(\theta\), and
+average these to obtain within-chain variance \(W\).
 
@@
 when the number of iterations \textgreater{} 50. The default number of
-iterations in iterative imputation, however, is often lower. Therefore,
-we compute \(AC\) manually, see panel D. The AC values in this plot do
-meet the requirements (\ldots) and will therefore be used in the
-simulation study.}
+iterations in iterative imputation, however, is often much lower.
+Therefore, we compute \(AC\) manually, see panel D. The AC values in
+this plot do meet the requirements (\ldots) and will therefore be used
+in the simulation study.}
+
+\textbf{acf is niet bedoeld om weinig iteraties te\ldots{} niet een fout
+in acf, alleen hier niet van toepassing}
 
@@
 
-\hypertarget{simulation-hypothesis}{%
-\subsection{Simulation Hypothesis}\label{simulation-hypothesis}}
+\hypertarget{simulation-hypothesis-als-bullets}{%
+\subsection{Simulation Hypothesis (als
+bullets!!)}\label{simulation-hypothesis-als-bullets}}
 
@@
 \(\widehat{R}\) will over-estimate non-convergence of MI algorithms
-\textbf{explain why?}. The threshold of \(\widehat{R} < 1.01\) will then
-be too stringent for diagnosing convergence. This over-estimation may,
-however, be diminished because \(\widehat{R}\) can falsely diagnose
-convergence if initial values of the algorithm are not appropriately
-over-dispersed \citep[p.~437]{broo98}. In e.g.~\texttt{mice}, initial
-values are chosen randomly from the observed data. Therefore, we cannot
-be certain that the initial values are over-dispersed. We expect this to
-have little effect on the hypothesized performance of \(\widehat{R}\)
-\textbf{explain that the randomness induced by the MICE algorithm would
-take care of this??}. \textbf{(Add actual hypothesis: we thought that
-mice would converge sooner than \(\widehat{R}\) would indicate that it
-did)} No hypothesis was formulated about the performance of \(AC\) as
-convergence diagnostic. \textbf{Added later: High \(AC\)-values are
-implausible in MI procedures. That is, the randomness induced by the MI
-algorithm effectively mitigates the risk of dependency within chains.
-But this doesn't seem the case in practice??}
+\textbf{explain why?}. Lacerda concluded that rhat dropped below 1.2
+after at least 50 iterations. Therefore\ldots{} Vehtari's proposed
+threshold of \(\widehat{R} < 1.01\) will then be too stringent for
+diagnosing convergence. This over-estimation may, however, be diminished
+because \(\widehat{R}\) can falsely diagnose convergence if initial
+values of the algorithm are not appropriately over-dispersed
+\citep[p.~437]{broo98}. In e.g.~\texttt{mice}, initial values are chosen
+randomly from the observed data. Therefore, we cannot be certain that
+the initial values are over-dispersed. We expect this to have little
+effect on the hypothesized performance of \(\widehat{R}\). \textbf{(Add
+actual hypothesis: we thought that mice would converge sooner than
+\(\widehat{R}\) would indicate that it did)} We hypothesize that high AC
+values are implausible in converged MI algorithms. \textbf{Added later:
+High \(AC\)-values are implausible in MI procedures. That is, the
+randomness induced by the MI algorithm effectively mitigates the risk of
+dependency within chains. But this doesn't seem the case in practice??}
 
@@
 
-The two challenges of convergence are stationarity and mixing. Mixing
-and stationarity can be inspected visually, by evaluating traceplots of
-scalar summaries of interest (\(\theta\)s; e.g., chain means and chain
-variances). As \citet{buur18} describes, users can also specify a
-model-specific scalar summary (e.g., a regression coefficient). A
-user-specified scalar summary, however, is not universal to all complete
-data problems. Therefore, as inspired by \citep{mack03}, we will propose
-a \(\theta\) that summarizes the multivariate state space of the
-algorithm. Namely, the first eigenvalue of the variance-covariance
-matrix of the \(M\) completed datasets. \textbf{Explain
-model-dependent\ldots{}}
+\textbf{Move to section above} Mixing and stationarity have histroically
+been inspected visually, by evaluating traceplots of scalar summaries of
+interest (\(\theta\)s; e.g., chain means and chain variances). As
+\citet{buur18} describes, users can also specify a model-specific scalar
+summary (e.g., a regression coefficient). A user-specified scalar
+summary, however, is not universal to all complete data problems.
+Therefore, as inspired by \citep{mack03}, we propose a \(\theta\) that
+summarizes the multivariate state space of the algorithm. Namely, the
+first eigenvalue of the variance-covariance matrix of the \(M\)
+completed datasets. \textbf{The first eigenvalue has the appealing
+property that is not dependent on the substantive model of interest.}
 
 The methods under evaluation are \(\widehat{R}\) and \(AC\). These will
-be computed for the three scalar summaries of interest: chain means
-(i.e., per imputation, per variable: on \(y_{imp, m, j}\)), chain
+be computed for the \textbf{four} scalar summaries of interest: chain
+means (i.e., per imputation, per variable: on \(y_{imp, m, j}\)), chain
 variances (i.e., per imputation, per variable: on \(y_{imp, m, j}\)),
@@
 Upon convergence, \(\widehat{R}=1\) and \(AC=0\), which are unlikely
-thresholds for MCMC algorithms, because \textbf{it's} convergence to a
+thresholds for MCMC algorithms, because of its convergence to a
 distribution. In practice, non-convergence is usually diagnosed when
@@
 imputation chain lengths (\(t = 1, 2,..., 100\)), and 2) varying the
-missingness proportions (\(p_{miss} =.05,.25,.50,.75,.95\)). The
+missingness proportions (\(p_{mis} =.05,.25,.50,.75,.95\)). The
 assumption underlying the different number of iterations is that the
@@
 The data-generating mechanism is a multivariate normal distribution,
-representing person data on three independent variables (from an
-unspecified social scientific field of study). Let \textbf{what?}
+representing person data on three predictor variables (from an
+unspecified social scientific field of study). Let our predictor space
+be defined by the following a multivariate normal distribution
 
@@
 simulated using the \texttt{mvtnorm} package \citep{mvtnorm}.
-Subsequently, a fourth variable is constructed to serve as dependent
+Subsequently, a fourth variable is constructed to serve as outcome
 variable in a multiple linear regression problem. Let \[
@@
 determination; \(R^2\)). We solve a multiple linear regression problem,
-where dependent variable \(Y\) is regressed on independent variables
-\(X_1\), \(X_2\) and \(X_3\)}
+where outcome \(Y\) is regressed on predictors \(X_1\), \(X_2\) and
+\(X_3\)}
 
@@
 repetition of the simulation thus starts from 1 complete dataset, then
-splits into 5 missingness conditions (\(p =.05,.25,.5,.75,.95\)), each
-of the amputed datasets is imputed 5 times (\(m = 1, 2,..., 5\)) for
-each iteration condition (\(t = 1, 2,..., 100\)). From these
+splits into 5 missingness conditions (\(p_{mis} =.05,.25,.5,.75,.95\)),
+each of the amputed datasets is imputed 5 times (\(m = 1, 2,..., 5\))
+for each iteration condition (\(t = 1, 2,..., 100\)). From these
 \(5\times5\times100\) imputations, we extract the \(\theta\)s to apply
@@
 The estimator for each \(Q\) is \(\bar{Q}\)---the pooled aggregate of
-the (\(\hat{Q}\))s across imputations. To calculate the \(\bar{Q}\)s, we
-combine the observed data (\(y_{obs}\)) and the imputed data for each
-imputation (\(y_{imp,m}\)) with the function \texttt{mice::complete()}.
+the \(\hat{Q}\)s across imputations. To calculate the \(\bar{Q}\)s, we
+combine the observed data \(y_{obs}\) and the imputed data for each
+imputation \(y_{imp,m}\) with the function \texttt{mice::complete()}.
 Descriptive statistics are computed as the average across imputations

### 2020-04-28 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
With feedback Research Seminar

@@
 
-**Include sub-questions?:** How can non-convergence be diagnosed?
+**Include sub-questions?:** How can non-convergence be identified diagnostically?
 Are common MCMC non-convergence diagnostics appropriate for MICE?
@@
 With a recently proposed adaptation, $\widehat{R}$ might also serve to diagnose non-stationarity, but this has not yet been thoroughly investigated [@veht19].
-Therefore, use we $AC$ and $\widehat{R}$ to evaluate stationarity and mixing separately, as recommended by e.g., @cowl96, p.
-898.
+Therefore, use we $\widehat{R}$ and $AC$ to evaluate mixing and stationarity separately, as recommended by e.g., @cowl96.
+<!-- @cowl96, p. 898. -->
 
@@
 
-## Simulation Hypothesis (als bullets!!)
+## Simulation Hypothesis (in bullets!! maybe move to simulation section?)
 
@@
 
-The data-generating mechanism is a multivariate normal distribution, representing person data on three predictor variables (from an unspecified social scientific field of study).
+The data-generating mechanism is a multivariate normal distribution, representing person-data on three predictor variables (from an unspecified social scientific field of study).
 Let our predictor space be defined by the following a multivariate normal distribution
@@
 
-Proper performance of the convergence diagnostics under at least missing completely at random (MCAR) missingness mechanism is necessary to demonstrate the appropriateness of $\widehat{R}$ and $AC$ as convergence diagnostics.
+**Maybe move to discussion:**Proper performance of the convergence diagnostics under at least missing completely at random (MCAR) missingness mechanism is necessary to demonstrate the appropriateness of $\widehat{R}$ and $AC$ as convergence diagnostics.
 However, results may not be extrapolated to other missingness mechanisms.

### 2020-04-28 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
With feedback Research Seminar

@@
 
-\textbf{Include sub-questions?:} How can non-convergence be diagnosed?
-Are common MCMC non-convergence diagnostics appropriate for MICE? And if
-so, which threshold should be used to diagnose non-convergence? How many
-iterations are sufficient/needed to be able to diagnose non-convergence?
-Are the default number of iterations sufficient (i.e., 5 in mice, 10 in
-SPSS and Stata, 30 in mi)? \textbf{How severe is it when the algorithm
-has not converged? And what are guidelines for practice? Can the
-parameter of interest, estimand \(Q\), be correct when the algorithm is
-not (yet) converged, and vice versa?}
+\textbf{Include sub-questions?:} How can non-convergence be identified
+diagnostically? Are common MCMC non-convergence diagnostics appropriate
+for MICE? And if so, which threshold should be used to diagnose
+non-convergence? How many iterations are sufficient/needed to be able to
+diagnose non-convergence? Are the default number of iterations
+sufficient (i.e., 5 in mice, 10 in SPSS and Stata, 30 in mi)?
+\textbf{How severe is it when the algorithm has not converged? And what
+are guidelines for practice? Can the parameter of interest, estimand
+\(Q\), be correct when the algorithm is not (yet) converged, and vice
+versa?}
 
@@
 non-stationarity, but this has not yet been thoroughly investigated
-\citep{veht19}. Therefore, use we \(AC\) and \(\widehat{R}\) to evaluate
-stationarity and mixing separately, as recommended by e.g.,
-\citet{cowl96}, p.~898.
+\citep{veht19}. Therefore, use we \(\widehat{R}\) and \(AC\) to evaluate
+mixing and stationarity separately, as recommended by e.g.,
+\citet{cowl96}.
 
@@
 
-\hypertarget{simulation-hypothesis-als-bullets}{%
-\subsection{Simulation Hypothesis (als
-bullets!!)}\label{simulation-hypothesis-als-bullets}}
+\hypertarget{simulation-hypothesis-in-bullets-maybe-move-to-simulation-section}{%
+\subsection{Simulation Hypothesis (in bullets!! maybe move to simulation
+section?)}\label{simulation-hypothesis-in-bullets-maybe-move-to-simulation-section}}
 
@@
 The data-generating mechanism is a multivariate normal distribution,
-representing person data on three predictor variables (from an
+representing person-data on three predictor variables (from an
 unspecified social scientific field of study). Let our predictor space
@@
 
-Proper performance of the convergence diagnostics under at least missing
-completely at random (MCAR) missingness mechanism is necessary to
-demonstrate the appropriateness of \(\widehat{R}\) and \(AC\) as
-convergence diagnostics. However, results may not be extrapolated to
-other missingness mechanisms. \textbf{Convergence diagnostics should
-therefore at least apply to the MCAR situation, before more complex
-missingness should be explored.}
+\textbf{Maybe move to discussion:}Proper performance of the convergence
+diagnostics under at least missing completely at random (MCAR)
+missingness mechanism is necessary to demonstrate the appropriateness of
+\(\widehat{R}\) and \(AC\) as convergence diagnostics. However, results
+may not be extrapolated to other missingness mechanisms.
+\textbf{Convergence diagnostics should therefore at least apply to the
+MCAR situation, before more complex missingness should be explored.}
 

### 2020-04-28 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
Plot labels

@@
 This paper provides insight into identifying non-convergence of iterative impuation algorithms."
-keywords: iterative imputation; non-convergence; mice
+keywords: Iterative imputation; non-convergence; mice
 classoption:
@@
 
+# Grading
+
+Scientific contribution
+(clear research question, convincing argumentation that the study adds to the scientific literature)
+
+Theory
+(use of theory, clear hypotheses, relevant literature)
+
+Method
+(correct data collection/handling, measures, choice and description of method of analysis)
+
+Results
+(correct interpretation of findings, clear presentation)
+Discussion and conclusions
+(summary, broader implications, limitations, future research)
+
+Written presentation
+(writing, structure, tables and figures, references)
+
 <!-- # References -->

### 2020-04-28 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Plot labels

@@
 
-\keywords{iterative imputation; non-convergence; mice}
+\keywords{Iterative imputation; non-convergence; mice}
 
@@
 
+\hypertarget{grading}{%
+\section{Grading}\label{grading}}
+
+Scientific contribution (clear research question, convincing
+argumentation that the study adds to the scientific literature)
+
+Theory (use of theory, clear hypotheses, relevant literature)
+
+Method (correct data collection/handling, measures, choice and
+description of method of analysis)
+
+Results (correct interpretation of findings, clear presentation)
+Discussion and conclusions (summary, broader implications, limitations,
+future research)
+
+Written presentation (writing, structure, tables and figures,
+references)
+
 \bibliographystyle{sageh}

### 2020-04-28 3.Thesis/1.SimulationStudy/Writeup/images/non-convergence.R

@@ -106,10 +106,10 @@ theta <- ggplot(diagnostics) +
   geom_line(aes(x = iteration, y = `Chain 4`, color = patho), na.rm = TRUE) +
   geom_line(aes(x = iteration, y = `Chain 5`, color = patho), na.rm = TRUE) +
   scale_x_continuous(breaks = 1:10) +
-  xlab("Iteration") +
-  ylab("Chain mean") +
+  xlab("") +
+  ylab(bquote(theta)) +
   # ggtitle(expression(paste("Chain means (", theta, ")"))) +
-  ggtitle(expression(paste("Scalar summary (", theta, ")"))) +
+  #ggtitle(expression(paste("Scalar summary (", theta, ")"))) +
   theme(legend.position = "")
 
 # plot rhat
@@ -121,9 +121,9 @@ rhat <-
   geom_point(size = .75, na.rm = TRUE) +
   geom_line(na.rm = TRUE) +
   scale_x_continuous(breaks = 1:10) +
-  xlab("Iteration") +
-  ylab(expression(paste(widehat(R)))) +
-  ggtitle(expression(paste(widehat(R), " of ", theta))) + 
+  xlab("") +
+  ylab(bquote(widehat(R)~" of "~theta)) +
+  #ggtitle(expression(paste(widehat(R), " of ", theta))) + 
   labs(color = "Convergence")
 
 # plot default ac (with acf())
@@ -136,8 +136,8 @@ acf <-
   geom_line(na.rm = TRUE) +
   scale_x_continuous(breaks = 1:10) +
   xlab("Iteration") +
-  ylab("AC") +
-  ggtitle(expression(paste("AC of ", theta, " (default calculation)"))) + 
+  ylab(bquote("AC of "~theta~"(default)")) +
+  #ggtitle(expression(paste("AC of ", theta, " (default calculation)"))) + 
   theme(legend.position = "")
 
 # plot manual ac
@@ -150,12 +150,12 @@ ac <-
   geom_line(na.rm = TRUE) +
   scale_x_continuous(breaks = 1:10) +
   xlab("Iteration") +
-  ylab("AC") +
-  ggtitle(expression(paste("AC of ", theta, " (manual calculation)"))) + 
+  ylab(bquote("AC of "~theta~"(manual)")) +
+  #ggtitle(expression(paste("AC of ", theta, " (manual calculation)"))) + 
   theme(legend.position = "")
 
 # combine
[…]

### 2020-04-29 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
Merge old and new methods sections

@@
 This paper provides insight into identifying non-convergence of iterative impuation algorithms."
-keywords: Iterative imputation; non-convergence; mice
+keywords: missing data, iterative imputation, non-convergence, mice
 classoption:
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(fig.width = 4.8, echo = FALSE, fig.align = "center") 
+knitr::opts_chunk$set(fig.width = 4.9, echo = FALSE, fig.align = "center") 
 # knitr::opts_chunk$set(fig.width = 4.8, fig.height = 3.6, echo = FALSE, fig.align = "center")
@@
 
-<!-- (**or** Everybody who works with person data is likely to run into a missing data problem.) -->
 Anyone who analyzes person-data may run into a missing data problem.
@@
 
-<!-- In this paper, we investigate the convergence properties of iterative imputation procedures by means of model based simulation in `R` [@R]. -->
+Currently, the recommended practice for evaluating convergence in iterative imputation algorithms is to visually inspect imputations for signs of non-convergence.
+**Add that this refers to chain means and variances already?
+It's insufficient that these are univariate, so multivariate state space of the algorithm may not be converged when univariates are ok.
+And @buur18's suggestion to track user-defined statistics (e.g., regression coefficient) may be somewhat advanced for empirical researchers.
+Moreover, this is not model-independent, i.e., only apply to one substantive model, while one of the advantages of MI is that missing data problem and scientific problem are split.
+Then we can say 'we propose a novel...' and integrate this: Note that usually we only evaluate the convergence of univariate scalar summaries (e.g., chain means or variances).
+With these we cannot diagnose convergence of multivariable statistics (i.e., relations between scalar summaries).
+Van Buuren [-@buur18, $\S$ 4.5.2] proposed to implement multivariable evaluation of the MICE algorithm through eigenvalue decomposition building on the work of @mack03.
+Eigenvalues of a covariance matrix are a measure of the data's covariance.**.
 
-<!-- It has been shown that this pooled result MI will yield unbiased, confidence valid results under many missing data circumstances [@buur18]. To implement MI, -->
-
-<!-- Rubin's rules [-@rubin76] may be used to reflect the uncertainty in the inferences due to missingness by pooling the results across 'imputations'.
-A popular method to obtain imputations  completions, the analysis of scientific interest can be performed.
-It has been , or to complete these observations by imputing the missing values.
-Imputation is the method of filling in -->
-
-<!-- by which That is, unless ad hoc solutions are employed (e.g., list-wise deletion), which can yield invalid results.
-contains missing values, statistical analyses cannot be performed without only be performed a dataset contains missing values, statistical analyses can statistical inferences cannot be performed  -->
-<!-- At some point, any social scientist -->
-<!-- conducting statistical analyses --> <!-- will run into a missing data problem [@alli02]. -->
-<!-- Missingness is problematic because statistical inference cannot be performed on the whole data without employing \emph{ad hoc} solutions (e.g., list-wise deletion), which may yield wildly invalid results [@buur18].
-A popular method to obtain statistically valid results despite of missing data is 'multiple imputation' [MI; @rubin76].
-The multiple imputation method can be roughly divided into 3 steps: First, each missing value in an incomplete dataset is 'imputed' (i.e., filled in) several times by some algorithmic procedure.
-Next, the analysis of scientific interest is performed on each of the completed datasets.
-And finally, the results of these analyses are pooled into a single aggregated statistic.
--->
-
-<!-- This paper investigates a potential threat to the validity of this method: algorithmic non-convergence. -->
-
-<!-- The theoretical foundation of these steeps was laid by Rubin [-@rubin76, -@rubin87].
-For an accessible and comprehensive introduction to MI from an applied perspective, see e.g.
-@buur18.
-In this paper, we focus on the first of these steps: the algorithmic procedure with which the missing values are imputed.
--->
-
-<!-- Algorithms are often used to draw imputations from the posterior predictive distribution of the missing values.
-This introduces a potential threat to the validity of the imputations: what if the algorithm has not converged?
-Are the implications then to be trusted?
-And can we rely on the inference obtained on the completed data?
--->
-
-<!-- based on a model of the observed data by means of an iterative algorithm -->
-<!-- is the use of MICE: multiple imputation by chained equations.
-MICE is a type of multiple imputation procedure--a framework proposed by @rubin87--in which missing datapoints are 'imputed' (i.e., filled in) by means of an iterative algorithm (chained equations).
-Imputed values are obtained by modeling the (distribution of the) missing data conditional on the observed data.
-The technique is therefore also called 'fully conditional specification' (FCS).
--->
-<!-- several times, and inferences are aggregated across imputed datasets. -->
-<!-- and statistical inference is performed on each imputed dataset. Variation between the results reflect uncertainty due to missingness -->
-<!-- CE: The imputation model can be a joint model for all variables, or several univariate conditional distributions (fully conditional specification).
-FCS is more flexible and "avoids the problem of specifying an appropriate joint imputation distribution and replaces this by the selection of appropriate univariate conditional distributions".
--->
-
-<!-- (**Rewrite:** With FCS, an iterative algorithmic procedure is used to...
-multiple imputation by chained equations...
-"The values for imputing the missing data are drawn from the posterior predictive distribution of the missing data conditional on the observed data.") -->
-
-Currently, the recommended practice for evaluating convergence in iterative imputation algorithms is to visually inspect imputations for signs of non-convergence.
-**(Add chain means and variances here already?
-Because it's insufficient that these are univariate?)**.
 This method is insufficient on two counts: 1) it may be challenging to the untrained eye, and 2) only severely pathological cases of non-convergence may be diagnosed [@buur18, $\S$ 6.5.2].
@@
 The values generated by the algorithm (e.g., imputed values) will vary even after convergence.
-Therefore, the aim of convergence diagnostics for MCMC methods is not to establish the point at which convergence is reached, but to monitor signs of non-convergence.
+Therefore, the aim of convergence diagnostics for MCMC methods is not to establish the point at which convergence is reached, but to monitor signs of non-convergence [@hoff09].
 Several of such diagnostics exist for MCMC methods, but it is not known whether these are appropriate for iterative imputation algorithms.
-<!-- Therefore we can only diagnose non-convergence, not convergence [@hoff09]. "a weak diagnostics is better than no diagnostic at all" [@cowl96]. -->
+<!-- "a weak diagnostics is better than no diagnostic at all" [@cowl96]. -->
 
@@
 
-**Include sub-questions?:** How can non-convergence be identified diagnostically?
+**Include sub-questions!** How can non-convergence be identified diagnostically?
 Are common MCMC non-convergence diagnostics appropriate for MICE?
@@
 And what are guidelines for practice?
-Can the parameter of interest, estimand $Q$, be correct when the algorithm is not (yet) converged, and vice versa?**
-
-<!-- **(These are part of the methods, not RQ per se: What are the effects of continued iterations on estimates, predictions and inferences?
-And finally, do the answers differ with varying missingness proportions?
-That is, we vary the nr of iterations and the missingness proportion because we assume that the algorithm has not reached convergence at $t=1$, and performs worse with moderate to high missingness (not so much with little or a LOT of missingness).)** -->
+Can the parameter of interest, estimand $Q$, be correct when the algorithm is not (yet) converged, and vice versa?** *(Maybe add these too?
+What are the effects of continued iterations on estimates, predictions and inferences?
+Do the answers differ with varying missingness proportions?
+That is, we vary the nr of iterations and the missingness proportion because we assume that the algorithm has not reached convergence at $t=1$, and performs worse with moderate to high missingness (not so much with little or a LOT of missingness).)*
 
@@
 
-<!-- MI is an iterative algorithmic procedure in which missing datapoints are `imputed' (i.e.
-filled in) several times.
-The variability between imputations is used to reflect how much uncertainty in the inference is introduced by the missingness.
-Therefore, MI can provide valid inferences despite missing information (**maybe add that this refers to CIs/p-values**).
--->
-
-<!-- To obtain valid inferences with MI, the variability between imputations should be properly represented [@rubin87; @buur18].
-If this variability is under-estimated, confidence intervals around estimates will be too narrow, which can yield spurious results.
-Over-estimation of the variance between imputations results in unnecessarily wide confidence intervals, which can be costly because it lowers the statistical power (**maybe add why this is costly**).
-Since both of these situations are undesirable, imputations and their variability should be evaluated.
-Evaluation measures, however, are currently missing or under-developed in MI software, like the world-leading `mice` package [@mice] in `R` [@R].
--->
 <!-- The goal of this research project is to develop novel methodology and guidelines for evaluating MI methods.
@@
 
-<!-- This note provides the theoretical foundation towards the diagnostic evaluation of multiple imputation algorithms.
-For reasons of brevity, we only focus on the MI algorithm implemented in `mice` [@mice].
-The convergence properties of this MI algorithm are investigated through model-based simulation.
-The results of this simulation study are guidelines for assessing convergence of MI algorithms.
--->
-<!-- These guidelines will be implemented in an interactive evaluation tool for `mice`, 'ShinyMICE', which is currently under development.
-We will evaluate how convergence of the MI algorithm can be diagnosed.
--->
-
-<!-- [^1]: All programming code used in this note is available from [github.com/gerkovink/shinyMice](https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy). -->
-
 ## Some notation
-
-<!-- The intended audience of this note consists of anyone who uses multiple imputation to solve missing data problem. -->
-<!-- This note follows notation and conventions of `mice` [@mice]. -->
-<!-- Deviations from the 'original' notation by @rubin87 are described in [@buur18, \S~2.2.3]. -->
-<!-- Basic familiarity with MI methodology is assumed.  -->
 
@@
 **
-<!-- The number of imputations is denoted by $m$ (i.e., the number of times each datapoint in $y_{mis}$ is filled in).
-The aim of MI is to impute each datapoint in $y_{mis}$ $m$ times.
-Because of the iterative nature of the MICE algorithm -->
-
-<!-- The result is $m$ completed datasets, consisting of $y_{obs}$ and  The MI algorithm in `mice` has an iterative nature.
-For each missing datapoint in $y_{mis}$, $m$ 'chains' of potential values are sampled.
--->
-<!-- Which parts of $y$ are missing is determined by the 'missingness mechanism'. -->
-
-<!-- Figure 1 provides an overview of the steps involved with MI---from incomplete data, to $m$ completed datasets, to $m$ estimated quantities of interest ($\hat{Q}$s), to a single pooled estimate $\bar{Q}$.
--->
-<!-- This note focuses on the algorithmic properties of the imputation step. -->
-
-<!-- Only the ultimate sample that each chain lands on is imputed. The number of iterations per chain will be denoted with $T$, where $t$ varies over the integers $1, 2, \dots, T$. -->
-
-<!-- This is repeated $m$ times.
-Imputed values are the ultimate samples of a `chain' of For each imputed value ($t = T$), a s for each missing datapoint in $y_{mis}$,.
-Each of the $m$ chains starts with an initial value, drawn randomly from $y_{obs}$.
-The chains are terminated after a predefined number of iterations.
-, and subsequently used in the analysis and pooling steps.
--->
-<!-- The collection of samples between the initial value (at $t=1$) and the imputed value (at $t=T$) will be referred to as an 'imputation chain'. -->
 
 # Identifying non-convergence
+
+Mixing and stationarity have histroically been inspected visually, by evaluating traceplots of scalar summaries of interest ($\theta$s; e.g., chain means and chain variances).
+As @buur18 describes, users can also specify a model-specific scalar summary (e.g., a regression coefficient).
+A user-specified scalar summary, however, is not universal to all complete data problems.
+Therefore, as inspired by [@mack03], we propose a $\theta$ that summarizes the multivariate state space of the algorithm.
+Namely, the first eigenvalue of the variance-covariance matrix of the $M$ completed datasets.
+**The first eigenvalue has the appealing property that is not dependent on the substantive model of interest.**
+
+**Since convergence of iterative imputation algorithms is in distribution, there is not a unique point at which the algorithm reaches a converged state.
+We can only evaluate scalar summaries of the multivariate state space of the algorithm, $\theta$s.
+The recommended scalar summaries to evaluate are chain means and chain variances.
+Additionally, researchers may "monitor some statistic of scientific interest" [@buur18, $\S$ 6.5.2].
+Because of the issues mentioned above, we propose a novel $\theta$ to monitor: the first eigenvalue of the variance-covariance matrix of the completed data.
+Let $\lambda_1 \geq \lambda_2 \geq ...
+\geq \lambda_j$ be the eigenvalues of $\Sigma$ in each of the $M$ completed datasets ${y_{obs}, y_{imp}}$.
+$\lambda_1$ is measure that summarises the covariances in the completed datasets.**
 
@@
 Without stationarity, there is trending within imputation chains, which implies that further iterations would yiled a different set of imputations.
-In Figure \ref{fig:non-conv}...
+
+To illustrate what non-convergence looks like in MI, we reproduce the example from van Buuren [-@buur18, $\S$ 6.5.2].
+Figure \ref{fig:non-conv} shows the traceplot for one of the variables in the example.
+we see the average of the imputed values for a variable $j$ in $y_{imp}$.
+The average imputed values in the left hand plot are about a magnitude 2 lower than the right hand plot.
+So non-convergence has a impact on the average imputed value for $j$ in $y_{imp, m}$.
+This difference (presumably) translates to bias in the pooled estimate $\bar{\mu}_j$ as well.
+And maybe also to higher order statistics, e.g.
+variances, covariances, etc.
+Therefore, it's important to converge.
 **Explain what we see, namely example by @buur18 reproduced, showing the traceplots of chain means for some variable.
@@
 This shows the importance of reaching converged states in iterative imputation algorithms.**
+
+## Methods under evaluation
+
+<!-- Requirements for diagnostics to identify non-convergence in iterative imputation are: 1) work for a relatively low number of iterations (defaults in popular iterative imputation software between 5 and 30); 2) may be applied to different $\theta$s, including model-independent $\theta$s (e.g., chain mean, or eigenvalue).
+-->
 
@@
 
-<!-- The current practice in diagnosing non-convergence is to inspect traceplots for signs of non-mixing and non-stationarity.  -->
+### Potential scale reduction factor
 
-<!-- The mixing component can be assessed with the potential scale reduction factor $\widehat{R}$ [a.k.a.
-`Gelman-Rubin statistic'; @gelm92].
-With an adapted version of $\widehat{R}$, proposed by @veht19, the stationarity component of convergence might be evaluated as well.
-This would make $\widehat{R}$ a general convergence diagnostic.
-The application of $\widehat{R}$ to assess stationarity has not been thoroughly investigated.
-Therefore, this study employs both $\widehat{R}$ and autocorrelation to investigate convergence, as recommended by [@cowl96, p.
-898].
--->
-
-<!-- The mixing component of convergence may be evaluated with the potential scale reduction factor, $\widehat{R}$ [a.k.a.
-`Gelman-Rubin statistic'; @gelm92].
-The stationarity component may be assessed through autocorrelations.
-An adapted version of $\widehat{R}$ might also be used to diagnose non-stationarity.
-Other convergence diagnostics are outside of the scope of this study.
--->
-
-<!-- Note that all of these methods evaluate the convergence of univariate scalar summaries (e.g., chain means or variances).
-These convergence diagnostics cannot diagnose convergence of multivariable statistics (i.e., relations between scalar summaries).
-@buur18 proposed to implement multivariable evaluation through eigenvalue decomposition [@mack03].
-This method is outside of the scope of the current study (**not anymore?**).
--->
-
-## Potential scale reduction factor
-
-An updated version of $\widehat{R}$ has been proposed by @veht19, (p.
+An updated version of $\widehat{R}$ has been proposed by @veht19 (p.
 5).
@@
 
-## Autocorrelation
+### Autocorrelation
 
@@
 
-## In practice/Answer
-**To assess whether $\widehat{R}$ and $AC$ may be appropriate non-convergence identifiers for iterative imputation algorithms, the following condition must hold when applied on the two algorithms plotted above: the methods indicate worse performance for the mis-specified model with pathological non-convergence (i.e., higher $\widehat{R}$- and $AC$-values than the typical performance).
+## In practice
+
+**Before using them in the simulation, the two methods show at least be able to distinguish between the two algorithms plotted in Figure 2.
+From visual inspection we know that the typical convergence has some signs of non-mixing around iteration 2, and little trending.
+In the non-convergence situation, there is a lot of trending upto iteration 6, after which the chains reach a somewhat more stationary state.
+Mixing in this situation gets worse from the first iteration onward, and gradually gets better around iteration 7.
+To assess whether $\widehat{R}$ and $AC$ may be appropriate non-convergence identifiers for iterative imputation algorithms, the following condition must hold: the methods indicate worse performance for the mis-specified model with pathological non-convergence (i.e., higher $\widehat{R}$- and $AC$-values than the typical performance).
 And additionally the methods should reflect the increasing convergence in the typical convergence situation as the number of iterations goes up.**
 
-**In figure \ref{fig:diagnostics}A (**add panel labels**), the chain means are plotted again, now together---as are the non-convergence diagnostics.
+**In Figure \ref{fig:diagnostics}A, the chain means from Figure \ref{fig:non-conv} are plotted again, now together---as are the non-convergence diagnostics.
 Panel B shows $\widehat{R}$ as computed by implementing @veht19 's recommendations.
@@
 
-```{r diagnostics, fig.cap="Convergence diagnostics pathological non-convergence versus typical convergence. **ADD PANEL LABELS**"}
+```{r diagnostics, fig.cap="Convergence diagnostics pathological non-convergence versus typical convergence."}
 load("./images/diagnostics_plot.Rdata")
@@
 
-## Simulation Hypothesis (in bullets!! maybe move to simulation section?)
+# Simulation study
 
-For multiple imputation algorithms, it holds that convergence is reached when there is no dependency between subsequent iterations of imputation chains ($AC = 0$), and chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm ($\widehat{R} = 1$).
+**Add to this section: 1) explicit mention of simulation conditions; 2) emphasize that the plots are averages across repetitions, not within MICE; 3) sample effects due to single complete dataset**
+
+The aim of the simulation study is to evaluate the impact of inducing non-convergence on several quantities of scientific interest $Q$, and to subsequently evaluate the performance of $\widehat{R}$ and $AC$ in identifying the non-convergence.
+
+Non-convergence will be induced by: 1) varying the missingness proportions in $y$ ($p_{mis} =.05,.25,.50,.75,.95$), and 2) terminating the MICE algorithm at different imputation chain lengths ($t = 1, 2,..., 100$).
+The assumption underlying the different number of iterations is that the algorithm generally does reach convergence at $t=1$.
+The initial values in the first iteration are sampled randomly from the set of observed datapoints.
+**As the number of iterations goes up, the imputation chains become independent of the initial values until the point at which an extra added iteration does not lead to a more converged state.** The second set of experimental conditions under consideration--the missingness proportions--are chosen to reflect the difficulty of the missingness problem.
+The inherent assumption is that low missingness proportions will lead to quick algorithmic convergence, since there is a lot of information in the observed data.
+Higher missingness proportions then cause slower convergence.
+However, if the fraction of missing information is very high, there is so little information in the data that the random component in the algorithm will take the overhand and a stable but very uncertain (high variance) state will be reached.
+
+## Hypothesis
+
+We hypothesize that simulation conditions with a higer $p_miss$ and lower $t$ have inferior convergence.
+We assume that when $\bar{Q}$ is an unbiased and confidence valid estimate of $Q$, the algorithm is *sufficiently* converged.
+
+**(bruggetje hier)** For multiple imputation algorithms, it holds that convergence is reached when there is no dependency between subsequent iterations of imputation chains ($AC = 0$), and chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm ($\widehat{R} = 1$).
 **We expect that a completely converged state will not be reached.**
 
-This study evaluates whether $\widehat{R}$ and $AC$ could diagnose convergence of multiple imputation algorithms.
+<!-- This study evaluates whether $\widehat{R}$ and $AC$ could diagnose convergence of multiple imputation algorithms.
 We assess the performance of the two convergence diagnostics against the recommended evaluation criteria for MI methods [i.e., average bias, average confidence interval width, and empirical coverage rate across simulations; @buur18, $\S$ 2.5.2].
-**That is, there is no baseline measure available to evaluate performance against.**
+**That is, there is no baseline measure available to evaluate performance against.** -->
+
 <!-- We could also look at distributional characteristics, and plausibility of imputed values, see @vinknd. For now, this is outside of the scope of this study. -->
@@
 Lacerda concluded that rhat dropped below 1.2 after at least 50 iterations.
-Therefore...
-Vehtari's proposed threshold of $\widehat{R} < 1.01$ will then be too stringent for diagnosing convergence.
+Therefore we hypothesize that Vehtari's proposed threshold of $\widehat{R} < 1.01$ will be too stringent for diagnosing convergence in iterative imputation.
 This over-estimation may, however, be diminished because $\widehat{R}$ can falsely diagnose convergence if initial values of the algorithm are not appropriately over-dispersed [@broo98, p.
@@
 
-# Simulation study
+## Set-up
 
-## How to diagnose non-convergence?
-
-**Move to section above** Mixing and stationarity have histroically been inspected visually, by evaluating traceplots of scalar summaries of interest ($\theta$s; e.g., chain means and chain variances).
-As @buur18 describes, users can also specify a model-specific scalar summary (e.g., a regression coefficient).
-A user-specified scalar summary, however, is not universal to all complete data problems.
-Therefore, as inspired by [@mack03], we propose a $\theta$ that summarizes the multivariate state space of the algorithm.
-Namely, the first eigenvalue of the variance-covariance matrix of the $M$ completed datasets.
-**The first eigenvalue has the appealing property that is not dependent on the substantive model of interest.**
-
-The methods under evaluation are $\widehat{R}$ and $AC$.
-These will be computed for the **four** scalar summaries of interest: chain means (i.e., per imputation, per variable: on $y_{imp, m, j}$), chain variances (i.e., per imputation, per variable: on $y_{imp, m, j}$), and the first eigenvalue of the variance-covariance matrix across imputations (i.e., per imputation: on {$y_{obs}, y_{imp, m}$}) **add user-specified $\theta$: estimated regression coefficient per imputation**.
-
-The performance of two methods ($\widehat{R}$ and $AC$) on each of these scalar summaries of interest will be evaluated with several estimands, i.e.
-quantities of scientific interest $Q$.
-It is assumed that when $\bar{Q}$ (the pooled result across imputations) is an unbiased and confidence valid estimate of $Q$, the algorithm is sufficiently converged.
-For each scalar $\theta$, one estimand $Q$ is defined that may be of interest in empirical research.
-The $\widehat{R}$ and $AC$ values with chain means as $\theta$s are evaluated against the bias in univariate mean estimates (**all variables or just for Y??**).
-Similarly, the performance measure for $\widehat{R}$ and $AC$ applied to chain variances is the bias in estimated standard deviations.
-To evaluate the performance of $\widehat{R}$ and $AC$ on the eigenvalues, we will use bias in the coefficient of determination, $R^2$.
-Additionally, we will evaluate the estimated regression coefficients, the coverage rate of the CI95% of the regression estimates, and the CI length (**see definitions in old Methods**).
-**remove the term performance measure here and add an extra paragraph to define performance measures as bias in all estimands, and coverage rate and CI length of regression coefficients.**
-
-**Move this to Results:** As expected, conditions with a higher proportion of missingness and/or a lower number of iterations show more signs of non-convergence, as indicated by more extreme bias in the estimated $Q$s.
-Roughly speaking this means that MICE is indeed not converged as $t=1$, and converges gradually as $t$ increases.
-The point at which an additional iteration does not lead to an improvement of the estimates depends on the difficulty of the missingness problem---between XYZ and XYZ.
-This means that the algorithm has converged sufficiently (**under the current specifications**).
-
-As required for a method to diagnose non-convergence, the values of $\widehat{R}$ follow a trend similar to the performance measures.
-$\widehat{R}$ values are generally lower in conditions with a higher number of iterations, and somewhat higher in conditions with a higher percentage of missing data.
-**Decide where to specify the two calculations of AC:** Initially, autocorrelation did not show a decrease with an increasing number of iterations.
-But, after replacing the ACF function with a manual calculation for autocorrelation, the autocorrelation values were indeed decreasing with a higher number of iterations.
-
-Evaluation with the performance measures shows that $\widehat{R}$ and autocorrelation are conservative: they indicate signs of non-convergence in conditions where $\bar{Q}$s are unbiased and or confidence valid estimates of $Q$.
-
-## When to diagnose non-convergence?
-
-Upon convergence, $\widehat{R}=1$ and $AC=0$, which are unlikely thresholds for MCMC algorithms, because of its convergence to a distribution.
-In practice,  non-convergence is usually diagnosed when $\widehat{R}$ > 1.2 or 1.1 or even 1.01.
-And a t-test is performed to assess whether $AC$ is significantly different from zero.
-Performance measures are the same as above: unbiased, confidence valid estimates.
-
-**Move this to Results:** As expected, complete convergence ($\widehat{R}=1$ and $AC=0$) is never observed.
-But what about the thresholds for practice?
-$\widehat{R}$ <1.2 is not stringent enough, because conditions in which $\widehat{R}$ is smaller than 1.2 do not all yield unbiased estimates of the $Q$s.
-Moreover, there is a dip in the $\widehat{R}$ values, after which $\widehat{R}$ > 1.2 again.
-If the algorithms would be terminated at the iteration where $\widehat{R}$ < 1.2 occurs for the first time, the increase in $\widehat{R}$ values might be missed.
-The threshold $\widehat{R}$ < 1.1 is somewhat conservative in comparison with the performance measures.
-But it seems OK.
-$\widehat{R}$ < 1.01 is too stringent compared to the performance measures.
-And it is not obtained for any number of iterations (**but may be necessary in more complex complete data models???**).
-
-## Start old methods section
-
-**Add to this section: 1) explicit mention of simulation conditions; 2) emphasize that the plots are averages across repetitions, not within MICE; 3) sample effects due to single complete dataset** Convergence of the MICE algorithm is investigated through model-based simulation in `R` [version `r paste(version$major,version$minor, sep = ".")`; @R].
+Convergence of the MICE algorithm is investigated through model-based simulation in `R` [version `r paste(version$major,version$minor, sep = ".")`; @R].
 The simulation set-up is summarized in the pseudo-code below.
 The complete `R` script of the simulation study is available from [github.com/gerkovink/shinyMice](https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy).
-<!-- Simulation conditions are the number of iterations ($T = 1, 2, \dots, 100$) and the percentage missingness in the -->
-
-<!-- to perform multiple imputation under 100 simulation conditions.
-In each condition, the MI algorithm comprises of a different imputation chain length ($T = 1, 2, \dots, 100$).
-The number of simulation runs per condition is 1000.
-For each simulation condition (i.e., number of iterations) and each repetition (i.e., simulation run) we compute convergence diagnostics ($\widehat{R}$ and autocorrelation) and simulation diagnostics (bias, confidence interval width and coverage).
-These diagnostics are aggregated across repetitions to obtain their average value per simulation condition.
--->
 
@@
 ```
-The aim of the simulation study is to evaluate the impact of inducing non-convergence by: 1) terminating the MICE algorithm at different imputation chain lengths ($t = 1, 2,..., 100$), and 2) varying the missingness proportions ($p_{mis} =.05,.25,.50,.75,.95$).
-The assumption underlying the different number of iterations is that the algorithm generally does reach convergence at $t=1$, because the initial values are sampled randomly from the set of observed datapoints.
-**As the number of iterations goes up, the imputation chains will become independent of the initial values until the point at which an extra added iteration does not lead to a more converged state.** The second set of experimental conditions under consideration--the missingness proportions--are chosen to reflect the difficulty of the missingness problem.
-The inherent assumption is that low missingness proportions will lead to quick algorithmic convergence, since there is a lot of information in the observed data.
-Higher missingness proportions then cause slower convergence.
-However, if the fraction of missing information is very high, there is so little information in the data that the random component in the algorithm will take the overhand and a stable but very uncertain (high variance) point will be reached.
 
 The data-generating mechanism is a multivariate normal distribution, representing person-data on three predictor variables (from an unspecified social scientific field of study).
-Let our predictor space be defined by the following a multivariate normal distribution
-
+Let our predictor space be defined as
 \begin{align*}
@@
 \end{align*}
+
 In this study, sampling variance is not of interest.
@@
 A finite population of $N=1000$ is simulated using the `mvtnorm` package [@mvtnorm].
-Subsequently, a fourth variable is constructed to serve as outcome variable in a multiple linear regression problem.
+Subsequently, a fourth variable is constructed  as outcome variable in a multiple linear regression problem.
 Let
@@
 
-**We consider four quantities of scientific interest ['conceptual estimands'; @morr19], namely the descriptive statistics of all variables (mean and standard deviation), $\beta_2$, and the percentage of variance in $Y$ explained (coefficient of determination; $R^2$).
-We solve a multiple linear regression problem, where outcome $Y$ is regressed on predictors $X_1$, $X_2$ and $X_3$**
-
-$$Y \sim \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3.$$
-<!-- This note only considers a 'missing completely at random' (MCAR) mechanism, where the probability of being missing is equal for all $n \times p$ cells in $y$ [@rubin87]. -->
-
 The complete data is *amputed* once for each simulation repetition with function `mice::ampute()`.
-
-(**change this to reflect current simulation set-up with 5, 25, 50, 75, and 95% of cases having missing data: The missingness is univariate, and the probability to be missing is the same for all four variables, namely 20\% (`prop = 0.8, mech = "MCAR"`).
-This leaves 20\% of the rows completely observed).
-**
-
-**Maybe move to discussion:**Proper performance of the convergence diagnostics under at least missing completely at random (MCAR) missingness mechanism is necessary to demonstrate the appropriateness of $\widehat{R}$ and $AC$ as convergence diagnostics.
+The missingness is multivariate, and the probability to be missing is the same for all $n \times k$ cells in $y$ ['MCAR'; @rubin87].
+Proper performance of the convergence diagnostics under at least 'missing completely at random' missingness mechanism is necessary to demonstrate the appropriateness of $\widehat{R}$ and $AC$ as convergence diagnostics.
 However, results may not be extrapolated to other missingness mechanisms.
-**Convergence diagnostics should therefore at least apply to the MCAR situation, before more complex missingness should be explored.**
+The missinness proportion varies between conditions ($p_{mis} =.05,.25,.5,.75,.95$).
+<!-- This note only considers a 'missing completely at random' (MCAR) mechanism, where the probability of being missing is equal for all $n \times k$ cells in $y$ [@rubin87]. -->
 
@@
 The number of iterations varies between simulation conditions (`maxit = ` $1, 2, \dots, 100$).
-Each repetition of the simulation thus starts from 1 complete dataset, then splits into 5 missingness conditions ($p_{mis} =.05,.25,.5,.75,.95$), each of the amputed datasets is imputed 5 times ($m = 1, 2,..., 5$) for each iteration condition ($t = 1, 2,..., 100$).
+<!-- Each repetition of the simulation thus starts from 1 complete dataset, then splits into 5 missingness conditions ($p_{mis} =.05,.25,.5,.75,.95$), each of the amputed datasets is imputed 5 times ($m = 1, 2,..., 5$) for each iteration condition ($t = 1, 2,..., 100$).
 From these $5\times5\times100$ imputations, we extract the $\theta$s to apply $\widehat{R}$ and $AC$ on (**i.e., chain means, chain variances, $\beta$s per imputation, and the first eigenvalue of the variance-covariance matrix per imputation**).
+-->
+
+The methods under evaluation, $\widehat{R}$ and $AC$, are computed for four scalar summaries $\theta$: chain means (i.e., the mean in each $y_{imp, m, j}$), chain variances (i.e., the variance in each $y_{imp, m, j}$), the statistic of scientific interest $\hat{Q}$ in each {$y_{obs}, y_{imp, m}$}, and the first eigenvalue of the variance-covariance matrix in each {$y_{obs}, y_{imp, m}$}, $\lambda_1$.
 We calculate $\widehat{R}$ by implementing Vehtari et al.'s [-@veht19] recommendations, and $AC$ as the correlation between the $t^{th}$ and the $(t+1)^{th}$ iteration.
 For each $\theta$ we estimate the corresponding scientific estimand $Q$.
+
+The performance of two methods on each of these scalar summaries will be evaluated with several estimands, i.e.
+quantities of scientific interest $Q$.
+For each $\theta$, one estimand $Q$ is defined that may be of interest in empirical research.
+The $\widehat{R}$ and $AC$ values with chain means as $\theta$s are evaluated against the bias in univariate mean estimates.
+Similarly, the performance measure for $\widehat{R}$ and $AC$ applied to chain variances is the bias in estimated standard deviations.
+To evaluate the performance of $\widehat{R}$ and $AC$ on the eigenvalues, we will use bias in the coefficient of determination, $R^2$.
+Additionally, we will evaluate the estimated regression coefficients, the coverage rate of the CI95% of the regression estimates, and the CI length (**see definitions in old Methods**).
+**remove the term performance measure here and add an extra paragraph to define performance measures as bias in all estimands, and coverage rate and CI length of regression coefficients.**
+
+We consider four quantities of scientific interest, namely the descriptive statistics of all variables (mean and standard deviation), $\beta_2$, and the percentage of variance in $Y$ explained (coefficient of determination $R^2 \times 100$).
+We solve a multiple linear regression problem, where outcome $Y$ is regressed on predictors $X_1$, $X_2$ and $X_3$
+
+$$Y \sim \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3.$$
 
@@
 The coefficient of determination is estimated for each imputation, and pooled using `mice::pool.r.squared()`.
-<!-- perform multiple linear regression on each completed dataset.
-We obtain an estimated regression coefficient per imputation, which are pooled into a single estimate, $\bar{Q}$.
-We use the function `mice::pool()` to estimate u bar (**check definition!**) according to Rubin's [-@rubin87] rules, and subsequently implement finite population pooling conform @vink14.
--->
 
@@
 
-# Old Results section
+# Results
 
@@
 
+As expected, conditions with a higher proportion of missingness and/or a lower number of iterations show more signs of non-convergence, as indicated by more extreme bias in the estimated $Q$s.
+Roughly speaking this means that MICE is indeed not converged as $t=1$, and converges gradually as $t$ increases.
+The point at which an additional iteration does not lead to an improvement of the estimates depends on the difficulty of the missingness problem---between XYZ and XYZ.
+This means that the algorithm has converged sufficiently (**under the current specifications**).
+
+As required for a method to diagnose non-convergence, the values of $\widehat{R}$ follow a trend similar to the performance measures.
+$\widehat{R}$ values are generally lower in conditions with a higher number of iterations, and somewhat higher in conditions with a higher percentage of missing data.
+Autocorrelation values were indeed decreasing with a higher number of iterations.
+
+Evaluation with the performance measures shows that $\widehat{R}$ and autocorrelation are conservative: they indicate signs of non-convergence in conditions where $\bar{Q}$s are unbiased and or confidence valid estimates of $Q$.
+
+## Old Results section
+
 **(Add more info about figure legends and axes.)**
@@
 ```{r mean, fig.height = 6.8, fig.cap="Convergence diagnostics chain mean."}
-mean_Rh + mean_AC + mean_bias + plot_layout(guides = "collect", ncol = 1) + plot_annotation(title = "Performance of convergence diagnostics", subtitle = bquote(theta ~ "= chain mean of Y in" ~ y[imp][",m"] ~ "; Q = " ~ mu[Y] ~ "= 25.81"))
+mean_Rh + mean_AC + mean_bias + plot_layout(guides = "collect", ncol = 1) + plot_annotation(tag_levels = "A", tag_suffix = ".", title = "Performance of convergence diagnostics", subtitle = bquote(theta ~ "= chain mean of Y in" ~ y[imp][",m"] ~ "; Q = " ~ mu[Y] ~ "= 25.81"))
 ```
@@
 ```{r sd, fig.height = 6.8, fig.cap="Convergence diagnostics chain variance."}
-var_Rh + var_AC + sd_bias + plot_layout(guides = "collect", ncol = 1) + plot_annotation(title = "Performance of convergence diagnostics", subtitle = bquote(theta ~ "= chain variance of Y in" ~ y[imp][",m"] ~ "; Q = " ~ sigma[Y] ~ "= 11.32"))
+var_Rh + var_AC + sd_bias + plot_layout(guides = "collect", ncol = 1) + plot_annotation(tag_levels = "A", tag_suffix = ".", title = "Performance of convergence diagnostics", subtitle = bquote(theta ~ "= chain variance of Y in" ~ y[imp][",m"] ~ "; Q = " ~ sigma[Y] ~ "= 11.32"))
 ```
 ```{r est, fig.height = 6.8, fig.cap="Convergence diagnostics regression coefficient."}
-est_Rh + est_AC + est_bias + plot_layout(guides = "collect", ncol = 1) + plot_annotation(title = "Performance of convergence diagnostics", subtitle = bquote(theta ~ "= " ~ hat(beta)[2] ~ "in {" ~ y[obs] ~ "," ~ y[imp][",m"] ~ "}" ~ "; Q = " ~ beta[2] ~ "= 2.06"))
+est_Rh + est_AC + est_bias + plot_layout(guides = "collect", ncol = 1) + plot_annotation(tag_levels = "A", tag_suffix = ".", title = "Performance of convergence diagnostics", subtitle = bquote(theta ~ "= " ~ hat(beta)[2] ~ "in {" ~ y[obs] ~ "," ~ y[imp][",m"] ~ "}" ~ "; Q = " ~ beta[2] ~ "= 2.06"))
 ```
@@
 ```{r pred, fig.height = 6.8, fig.cap="Convergence diagnostics regression coefficient."}
-PCA_Rh + PCA_AC + Rsq_bias + plot_layout(guides = "collect", ncol = 1) + plot_annotation(title = "Performance of convergence diagnostics", subtitle = bquote(theta ~ "= first eigenvalue of" ~ Sigma ~ "in {" ~ y[obs] ~ "," ~ y[imp][",m"] ~ "}" ~ "; Q = explained variance = 19.25"))
+PCA_Rh + PCA_AC + Rsq_bias + plot_layout(guides = "collect", ncol = 1) + plot_annotation(tag_levels = "A", tag_suffix = ".", title = "Performance of convergence diagnostics", subtitle = bquote(theta ~ "= first eigenvalue of" ~ Sigma ~ "in {" ~ y[obs] ~ "," ~ y[imp][",m"] ~ "}" ~ "; Q = explained variance = 19.25"))
 
@@
 (With the default number of iterations (maxit = 5), this dip in $\widehat{R}$ values would be spotted, so it is no problem.) The point at which an extra iteration does not seem to improve the $\widehat{R}$ value is around $T=30$.
-<!-- $\widehat{R}$ values are indistinguishable between simulation conditions when $T>30$. -->
-<!-- The $\widehat{R}$ values seemed to flatten off between 20 and 30 iterations. -->
 
@@
 
-<!-- For 25% miss, no bias in estimated means, Rh flat from 30 it.
-(but <1.1 for it>11), $AC$ never high but stable after 40 it.
-Bias in estimated variances only if it=1 or maybe 2?, while Rh of chain variances is flat around it 25 (smaller than 1.1 for it>9), and $AC$ never positive, flat around it 30.
-Bias in estimated reg.
-coeff.
-only pronounced for the first iteration., while coverage rates are ok immediately on average, variance explained is marginally underestimated (bias of 0.5% on an effect of 19.25%).
--->
-
 ```{r thresh-Rh, include=FALSE}
@@
 # Discussion
+
+Upon convergence, $\widehat{R}=1$ and $AC=0$, which are unlikely thresholds for MCMC algorithms, because of its convergence to a distribution.
+In practice, non-convergence is usually diagnosed when $\widehat{R}$ > 1.2 or 1.1 or even 1.01.
+And a t-test is performed to assess whether $AC$ is significantly different from zero.
+
+As expected, complete convergence ($\widehat{R}=1$ and $AC=0$) is never observed.
+But what about the thresholds for practice?
+$\widehat{R}$ <1.2 is not stringent enough, because conditions in which $\widehat{R}$ is smaller than 1.2 do not all yield unbiased estimates of the $Q$s.
+Moreover, there is a dip in the $\widehat{R}$ values, after which $\widehat{R}$ > 1.2 again.
+If the algorithms would be terminated at the iteration where $\widehat{R}$ < 1.2 occurs for the first time, the increase in $\widehat{R}$ values might be missed.
+The threshold $\widehat{R}$ < 1.1 is somewhat conservative in comparison with the performance measures.
+Overall it seems OK.
+$\widehat{R}$ < 1.01 is too stringent compared to the performance measures.
+And it is not obtained for any number of iterations (**but may be necessary in more complex complete data models???**).
 

### 2020-04-29 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Merge old and new methods sections

@@
 
-\keywords{Iterative imputation; non-convergence; mice}
+\keywords{missing data, iterative imputation, non-convergence, mice}
 
@@
 iterative imputation algorithms is to visually inspect imputations for
-signs of non-convergence. \textbf{(Add chain means and variances here
-already? Because it's insufficient that these are univariate?)}. This
-method is insufficient on two counts: 1) it may be challenging to the
-untrained eye, and 2) only severely pathological cases of
+signs of non-convergence. \textbf{Add that this refers to chain means
+and variances already? It's insufficient that these are univariate, so
+multivariate state space of the algorithm may not be converged when
+univariates are ok. And \citet{buur18}`s suggestion to track
+user-defined statistics (e.g., regression coefficient) may be somewhat
+advanced for empirical researchers. Moreover, this is not
+model-independent, i.e., only apply to one substantive model, while one
+of the advantages of MI is that missing data problem and scientific
+problem are split. Then we can say 'we propose a novel\ldots{}' and
+integrate this: Note that usually we only evaluate the convergence of
+univariate scalar summaries (e.g., chain means or variances). With these
+we cannot diagnose convergence of multivariable statistics (i.e.,
+relations between scalar summaries). Van Buuren \citeyearpar[\(\S\)
+4.5.2]{buur18} proposed to implement multivariable evaluation of the
+MICE algorithm through eigenvalue decomposition building on the work of
+\citet{mack03}. Eigenvalues of a covariance matrix are a measure of the
+data's covariance.}.
+
+This method is insufficient on two counts: 1) it may be challenging to
+the untrained eye, and 2) only severely pathological cases of
 non-convergence may be diagnosed \citep[\(\S\) 6.5.2]{buur18}.
@@
 convergence diagnostics for MCMC methods is not to establish the point
-at which convergence is reached, but to monitor signs of
-non-convergence. Several of such diagnostics exist for MCMC methods, but
+at which convergence is reached, but to monitor signs of non-convergence
+\citep{hoff09}. Several of such diagnostics exist for MCMC methods, but
 it is not known whether these are appropriate for iterative imputation
@@
 
-\textbf{Include sub-questions?:} How can non-convergence be identified
+\textbf{Include sub-questions!} How can non-convergence be identified
 diagnostically? Are common MCMC non-convergence diagnostics appropriate
@@
 \(Q\), be correct when the algorithm is not (yet) converged, and vice
-versa?}
+versa?} \emph{(Maybe add these too? What are the effects of continued
+iterations on estimates, predictions and inferences? Do the answers
+differ with varying missingness proportions? That is, we vary the nr of
+iterations and the missingness proportion because we assume that the
+algorithm has not reached convergence at \(t=1\), and performs worse
+with moderate to high missingness (not so much with little or a LOT of
+missingness).)}
 
@@
 
+Mixing and stationarity have histroically been inspected visually, by
+evaluating traceplots of scalar summaries of interest (\(\theta\)s;
+e.g., chain means and chain variances). As \citet{buur18} describes,
+users can also specify a model-specific scalar summary (e.g., a
+regression coefficient). A user-specified scalar summary, however, is
+not universal to all complete data problems. Therefore, as inspired by
+\citep{mack03}, we propose a \(\theta\) that summarizes the multivariate
+state space of the algorithm. Namely, the first eigenvalue of the
+variance-covariance matrix of the \(M\) completed datasets. \textbf{The
+first eigenvalue has the appealing property that is not dependent on the
+substantive model of interest.}
+
+\textbf{Since convergence of iterative imputation algorithms is in
+distribution, there is not a unique point at which the algorithm reaches
+a converged state. We can only evaluate scalar summaries of the
+multivariate state space of the algorithm, \(\theta\)s. The recommended
+scalar summaries to evaluate are chain means and chain variances.
+Additionally, researchers may ``monitor some statistic of scientific
+interest'' \citep[\(\S\) 6.5.2]{buur18}. Because of the issues mentioned
+above, we propose a novel \(\theta\) to monitor: the first eigenvalue of
+the variance-covariance matrix of the completed data. Let
+\(\lambda_1 \geq \lambda_2 \geq ... \geq \lambda_j\) be the eigenvalues
+of \(\Sigma\) in each of the \(M\) completed datasets
+\({y_{obs}, y_{imp}}\). \(\lambda_1\) is measure that summarises the
+covariances in the completed datasets.}
+
 There are two requirements for convergence of iterative algorithms:
@@
 implies that further iterations would yiled a different set of
-imputations. In Figure \ref{fig:non-conv}\ldots{} \textbf{Explain what
-we see, namely example by \citet{buur18} reproduced, showing the
-traceplots of chain means for some variable. The first plot is typical
-convergence of MICE, the second is pathological non-convergence because
-of a mis-specified imputation model. Each line is an imputation. In the
-first plot, the chains intermingle nicely and there is little to no
-trending. In the second plot, there is a lot of trending and some chains
-do not intermingle. Importantly, the chain means at the last iteration
-(the imputed value per \(m\)) are very different between the two plots.
-The algorithm with the mis-specified model yields imputed values that
-are on average a magnitude two larger than those of the typically
-converged algorithm. This shows the importance of reaching converged
-states in iterative imputation algorithms.}
+imputations.
+
+To illustrate what non-convergence looks like in MI, we reproduce the
+example from van Buuren \citeyearpar[\(\S\) 6.5.2]{buur18}. Figure
+\ref{fig:non-conv} shows the traceplot for one of the variables in the
+example. we see the average of the imputed values for a variable \(j\)
+in \(y_{imp}\). The average imputed values in the left hand plot are
+about a magnitude 2 lower than the right hand plot. So non-convergence
+has a impact on the average imputed value for \(j\) in \(y_{imp, m}\).
+This difference (presumably) translates to bias in the pooled estimate
+\(\bar{\mu}_j\) as well. And maybe also to higher order statistics,
+e.g.~variances, covariances, etc. Therefore, it's important to converge.
+\textbf{Explain what we see, namely example by \citet{buur18}
+reproduced, showing the traceplots of chain means for some variable. The
+first plot is typical convergence of MICE, the second is pathological
+non-convergence because of a mis-specified imputation model. Each line
+is an imputation. In the first plot, the chains intermingle nicely and
+there is little to no trending. In the second plot, there is a lot of
+trending and some chains do not intermingle. Importantly, the chain
+means at the last iteration (the imputed value per \(m\)) are very
+different between the two plots. The algorithm with the mis-specified
+model yields imputed values that are on average a magnitude two larger
+than those of the typically converged algorithm. This shows the
+importance of reaching converged states in iterative imputation
+algorithms.}
+
+\hypertarget{methods-under-evaluation}{%
+\subsection{Methods under evaluation}\label{methods-under-evaluation}}
 
@@
 \hypertarget{potential-scale-reduction-factor}{%
-\subsection{Potential scale reduction
+\subsubsection{Potential scale reduction
 factor}\label{potential-scale-reduction-factor}}
@@
 An updated version of \(\widehat{R}\) has been proposed by
-\citet{veht19}, (p.~5). \textbf{This version may be suitable for
+\citet{veht19} (p.~5). \textbf{This version may be suitable for
 iterative imputation.} Let \(M\) be the total number of chains, \(T\)
@@
 \hypertarget{autocorrelation}{%
-\subsection{Autocorrelation}\label{autocorrelation}}
+\subsubsection{Autocorrelation}\label{autocorrelation}}
 
@@
 
-\hypertarget{in-practiceanswer}{%
-\subsection{In practice/Answer}\label{in-practiceanswer}}
+\hypertarget{in-practice}{%
+\subsection{In practice}\label{in-practice}}
 
-\textbf{To assess whether \(\widehat{R}\) and \(AC\) may be appropriate
-non-convergence identifiers for iterative imputation algorithms, the
-following condition must hold when applied on the two algorithms plotted
-above: the methods indicate worse performance for the mis-specified
+\textbf{Before using them in the simulation, the two methods show at
+least be able to distinguish between the two algorithms plotted in
+Figure 2. From visual inspection we know that the typical convergence
+has some signs of non-mixing around iteration 2, and little trending. In
+the non-convergence situation, there is a lot of trending upto iteration
+6, after which the chains reach a somewhat more stationary state. Mixing
+in this situation gets worse from the first iteration onward, and
+gradually gets better around iteration 7. To assess whether
+\(\widehat{R}\) and \(AC\) may be appropriate non-convergence
+identifiers for iterative imputation algorithms, the following condition
+must hold: the methods indicate worse performance for the mis-specified
 model with pathological non-convergence (i.e., higher \(\widehat{R}\)-
@@
 
-\textbf{In figure \ref{fig:diagnostics}A (}add panel labels\textbf{),
-the chain means are plotted again, now together---as are the
+\textbf{In Figure \ref{fig:diagnostics}A, the chain means from Figure
+\ref{fig:non-conv} are plotted again, now together---as are the
 non-convergence diagnostics. Panel B shows \(\widehat{R}\) as computed
@@
 
-\caption{Convergence diagnostics pathological non-convergence versus typical convergence. **ADD PANEL LABELS**}\label{fig:diagnostics}
+\caption{Convergence diagnostics pathological non-convergence versus typical convergence.}\label{fig:diagnostics}
 \end{figure}
 
-\hypertarget{simulation-hypothesis-in-bullets-maybe-move-to-simulation-section}{%
-\subsection{Simulation Hypothesis (in bullets!! maybe move to simulation
-section?)}\label{simulation-hypothesis-in-bullets-maybe-move-to-simulation-section}}
+\hypertarget{simulation-study}{%
+\section{Simulation study}\label{simulation-study}}
 
-For multiple imputation algorithms, it holds that convergence is reached
-when there is no dependency between subsequent iterations of imputation
-chains (\(AC = 0\)), and chains intermingle such that the only
-difference between the chains is caused by the randomness induced by the
-algorithm (\(\widehat{R} = 1\)). \textbf{We expect that a completely
-converged state will not be reached.}
+\textbf{Add to this section: 1) explicit mention of simulation
+conditions; 2) emphasize that the plots are averages across repetitions,
+not within MICE; 3) sample effects due to single complete dataset}
 
-This study evaluates whether \(\widehat{R}\) and \(AC\) could diagnose
-convergence of multiple imputation algorithms. We assess the performance
-of the two convergence diagnostics against the recommended evaluation
-criteria for MI methods \citep[i.e., average bias, average confidence
-interval width, and empirical coverage rate across simulations;][\(\S\)
-2.5.2]{buur18}. \textbf{That is, there is no baseline measure available
-to evaluate performance against.}
+The aim of the simulation study is to evaluate the impact of inducing
+non-convergence on several quantities of scientific interest \(Q\), and
+to subsequently evaluate the performance of \(\widehat{R}\) and \(AC\)
+in identifying the non-convergence.
+
+Non-convergence will be induced by: 1) varying the missingness
+proportions in \(y\) (\(p_{mis} =.05,.25,.50,.75,.95\)), and 2)
+terminating the MICE algorithm at different imputation chain lengths
+(\(t = 1, 2,..., 100\)). The assumption underlying the different number
+of iterations is that the algorithm generally does reach convergence at
+\(t=1\). The initial values in the first iteration are sampled randomly
+from the set of observed datapoints. \textbf{As the number of iterations
+goes up, the imputation chains become independent of the initial values
+until the point at which an extra added iteration does not lead to a
+more converged state.} The second set of experimental conditions under
+consideration--the missingness proportions--are chosen to reflect the
+difficulty of the missingness problem. The inherent assumption is that
+low missingness proportions will lead to quick algorithmic convergence,
+since there is a lot of information in the observed data. Higher
+missingness proportions then cause slower convergence. However, if the
+fraction of missing information is very high, there is so little
+information in the data that the random component in the algorithm will
+take the overhand and a stable but very uncertain (high variance) state
+will be reached.
+
+\hypertarget{hypothesis}{%
+\subsection{Hypothesis}\label{hypothesis}}
+
+We hypothesize that simulation conditions with a higer \(p_miss\) and
+lower \(t\) have inferior convergence. We assume that when \(\bar{Q}\)
+is an unbiased and confidence valid estimate of \(Q\), the algorithm is
+\emph{sufficiently} converged.
+
+\textbf{(bruggetje hier)} For multiple imputation algorithms, it holds
+that convergence is reached when there is no dependency between
+subsequent iterations of imputation chains (\(AC = 0\)), and chains
+intermingle such that the only difference between the chains is caused
+by the randomness induced by the algorithm (\(\widehat{R} = 1\)).
+\textbf{We expect that a completely converged state will not be
+reached.}
 
@@
 \textbf{explain why?}. Lacerda concluded that rhat dropped below 1.2
-after at least 50 iterations. Therefore\ldots{} Vehtari's proposed
-threshold of \(\widehat{R} < 1.01\) will then be too stringent for
-diagnosing convergence. This over-estimation may, however, be diminished
-because \(\widehat{R}\) can falsely diagnose convergence if initial
-values of the algorithm are not appropriately over-dispersed
-\citep[p.~437]{broo98}. In e.g.~\texttt{mice}, initial values are chosen
-randomly from the observed data. Therefore, we cannot be certain that
-the initial values are over-dispersed. We expect this to have little
-effect on the hypothesized performance of \(\widehat{R}\). \textbf{(Add
-actual hypothesis: we thought that mice would converge sooner than
-\(\widehat{R}\) would indicate that it did)} We hypothesize that high AC
-values are implausible in converged MI algorithms. \textbf{Added later:
-High \(AC\)-values are implausible in MI procedures. That is, the
-randomness induced by the MI algorithm effectively mitigates the risk of
-dependency within chains. But this doesn't seem the case in practice??}
+after at least 50 iterations. Therefore we hypothesize that Vehtari's
+proposed threshold of \(\widehat{R} < 1.01\) will be too stringent for
+diagnosing convergence in iterative imputation. This over-estimation
+may, however, be diminished because \(\widehat{R}\) can falsely diagnose
+convergence if initial values of the algorithm are not appropriately
+over-dispersed \citep[p.~437]{broo98}. In e.g.~\texttt{mice}, initial
+values are chosen randomly from the observed data. Therefore, we cannot
+be certain that the initial values are over-dispersed. We expect this to
+have little effect on the hypothesized performance of \(\widehat{R}\).
+\textbf{(Add actual hypothesis: we thought that mice would converge
+sooner than \(\widehat{R}\) would indicate that it did)} We hypothesize
+that high AC values are implausible in converged MI algorithms.
+\textbf{Added later: High \(AC\)-values are implausible in MI
+procedures. That is, the randomness induced by the MI algorithm
+effectively mitigates the risk of dependency within chains. But this
+doesn't seem the case in practice??}
 
-\hypertarget{simulation-study}{%
-\section{Simulation study}\label{simulation-study}}
+\hypertarget{set-up}{%
+\subsection{Set-up}\label{set-up}}
 
-\hypertarget{how-to-diagnose-non-convergence}{%
-\subsection{How to diagnose
-non-convergence?}\label{how-to-diagnose-non-convergence}}
-
-\textbf{Move to section above} Mixing and stationarity have histroically
-been inspected visually, by evaluating traceplots of scalar summaries of
-interest (\(\theta\)s; e.g., chain means and chain variances). As
-\citet{buur18} describes, users can also specify a model-specific scalar
-summary (e.g., a regression coefficient). A user-specified scalar
-summary, however, is not universal to all complete data problems.
-Therefore, as inspired by \citep{mack03}, we propose a \(\theta\) that
-summarizes the multivariate state space of the algorithm. Namely, the
-first eigenvalue of the variance-covariance matrix of the \(M\)
-completed datasets. \textbf{The first eigenvalue has the appealing
-property that is not dependent on the substantive model of interest.}
-
-The methods under evaluation are \(\widehat{R}\) and \(AC\). These will
-be computed for the \textbf{four} scalar summaries of interest: chain
-means (i.e., per imputation, per variable: on \(y_{imp, m, j}\)), chain
-variances (i.e., per imputation, per variable: on \(y_{imp, m, j}\)),
-and the first eigenvalue of the variance-covariance matrix across
-imputations (i.e., per imputation: on \{\(y_{obs}, y_{imp, m}\)\})
-\textbf{add user-specified \(\theta\): estimated regression coefficient
-per imputation}.
-
-The performance of two methods (\(\widehat{R}\) and \(AC\)) on each of
-these scalar summaries of interest will be evaluated with several
-estimands, i.e.~quantities of scientific interest \(Q\). It is assumed
-that when \(\bar{Q}\) (the pooled result across imputations) is an
-unbiased and confidence valid estimate of \(Q\), the algorithm is
-sufficiently converged. For each scalar \(\theta\), one estimand \(Q\)
-is defined that may be of interest in empirical research. The
-\(\widehat{R}\) and \(AC\) values with chain means as \(\theta\)s are
-evaluated against the bias in univariate mean estimates (\textbf{all
-variables or just for Y??}). Similarly, the performance measure for
-\(\widehat{R}\) and \(AC\) applied to chain variances is the bias in
-estimated standard deviations. To evaluate the performance of
-\(\widehat{R}\) and \(AC\) on the eigenvalues, we will use bias in the
-coefficient of determination, \(R^2\). Additionally, we will evaluate
-the estimated regression coefficients, the coverage rate of the CI95\%
-of the regression estimates, and the CI length (\textbf{see definitions
-in old Methods}). \textbf{remove the term performance measure here and
-add an extra paragraph to define performance measures as bias in all
-estimands, and coverage rate and CI length of regression coefficients.}
-
-\textbf{Move this to Results:} As expected, conditions with a higher
-proportion of missingness and/or a lower number of iterations show more
-signs of non-convergence, as indicated by more extreme bias in the
-estimated \(Q\)s. Roughly speaking this means that MICE is indeed not
-converged as \(t=1\), and converges gradually as \(t\) increases. The
-point at which an additional iteration does not lead to an improvement
-of the estimates depends on the difficulty of the missingness
-problem---between XYZ and XYZ. This means that the algorithm has
-converged sufficiently (\textbf{under the current specifications}).
-
-As required for a method to diagnose non-convergence, the values of
-\(\widehat{R}\) follow a trend similar to the performance measures.
-\(\widehat{R}\) values are generally lower in conditions with a higher
-number of iterations, and somewhat higher in conditions with a higher
-percentage of missing data. \textbf{Decide where to specify the two
-calculations of AC:} Initially, autocorrelation did not show a decrease
-with an increasing number of iterations. But, after replacing the ACF
-function with a manual calculation for autocorrelation, the
-autocorrelation values were indeed decreasing with a higher number of
-iterations.
-
-Evaluation with the performance measures shows that \(\widehat{R}\) and
-autocorrelation are conservative: they indicate signs of non-convergence
-in conditions where \(\bar{Q}\)s are unbiased and or confidence valid
-estimates of \(Q\).
-
-\hypertarget{when-to-diagnose-non-convergence}{%
-\subsection{When to diagnose
-non-convergence?}\label{when-to-diagnose-non-convergence}}
-
-Upon convergence, \(\widehat{R}=1\) and \(AC=0\), which are unlikely
-thresholds for MCMC algorithms, because of its convergence to a
-distribution. In practice, non-convergence is usually diagnosed when
-\(\widehat{R}\) \textgreater{} 1.2 or 1.1 or even 1.01. And a t-test is
-performed to assess whether \(AC\) is significantly different from zero.
-Performance measures are the same as above: unbiased, confidence valid
-estimates.
-
-\textbf{Move this to Results:} As expected, complete convergence
-(\(\widehat{R}=1\) and \(AC=0\)) is never observed. But what about the
-thresholds for practice? \(\widehat{R}\) \textless1.2 is not stringent
-enough, because conditions in which \(\widehat{R}\) is smaller than 1.2
-do not all yield unbiased estimates of the \(Q\)s. Moreover, there is a
-dip in the \(\widehat{R}\) values, after which \(\widehat{R}\)
-\textgreater{} 1.2 again. If the algorithms would be terminated at the
-iteration where \(\widehat{R}\) \textless{} 1.2 occurs for the first
-time, the increase in \(\widehat{R}\) values might be missed. The
-threshold \(\widehat{R}\) \textless{} 1.1 is somewhat conservative in
-comparison with the performance measures. But it seems OK.
-\(\widehat{R}\) \textless{} 1.01 is too stringent compared to the
-performance measures. And it is not obtained for any number of
-iterations (\textbf{but may be necessary in more complex complete data
-models???}).
-
-\hypertarget{start-old-methods-section}{%
-\subsection{Start old methods section}\label{start-old-methods-section}}
-
-\textbf{Add to this section: 1) explicit mention of simulation
-conditions; 2) emphasize that the plots are averages across repetitions,
-not within MICE; 3) sample effects due to single complete dataset}
 Convergence of the MICE algorithm is investigated through model-based
@@
 
-The aim of the simulation study is to evaluate the impact of inducing
-non-convergence by: 1) terminating the MICE algorithm at different
-imputation chain lengths (\(t = 1, 2,..., 100\)), and 2) varying the
-missingness proportions (\(p_{mis} =.05,.25,.50,.75,.95\)). The
-assumption underlying the different number of iterations is that the
-algorithm generally does reach convergence at \(t=1\), because the
-initial values are sampled randomly from the set of observed datapoints.
-\textbf{As the number of iterations goes up, the imputation chains will
-become independent of the initial values until the point at which an
-extra added iteration does not lead to a more converged state.} The
-second set of experimental conditions under consideration--the
-missingness proportions--are chosen to reflect the difficulty of the
-missingness problem. The inherent assumption is that low missingness
-proportions will lead to quick algorithmic convergence, since there is a
-lot of information in the observed data. Higher missingness proportions
-then cause slower convergence. However, if the fraction of missing
-information is very high, there is so little information in the data
-that the random component in the algorithm will take the overhand and a
-stable but very uncertain (high variance) point will be reached.
-
 The data-generating mechanism is a multivariate normal distribution,
@@
 unspecified social scientific field of study). Let our predictor space
-be defined by the following a multivariate normal distribution
-
-\begin{align*}
+be defined as \begin{align*}
 \begin{pmatrix}X_1\\
@@
 \end{bmatrix}\!\!\text{.}\\[2\jot]
-\end{align*} In this study, sampling variance is not of interest.
-Therefore, a single complete set may serve as comparative truth in all
-simulation runs \citep{vink14}. A finite population of \(N=1000\) is
-simulated using the \texttt{mvtnorm} package \citep{mvtnorm}.
-Subsequently, a fourth variable is constructed to serve as outcome
-variable in a multiple linear regression problem. Let \[
+\end{align*}
+
+In this study, sampling variance is not of interest. Therefore, a single
+complete set may serve as comparative truth in all simulation runs
+\citep{vink14}. A finite population of \(N=1000\) is simulated using the
+\texttt{mvtnorm} package \citep{mvtnorm}. Subsequently, a fourth
+variable is constructed as outcome variable in a multiple linear
+regression problem. Let \[
 Y_i = 1 + 2X_{1i} +.5X_{2i} - X_{3i} + \epsilon_i ,
@@
 
-\textbf{We consider four quantities of scientific interest
-\citep[`conceptual estimands';][]{morr19}, namely the descriptive
-statistics of all variables (mean and standard deviation), \(\beta_2\),
-and the percentage of variance in \(Y\) explained (coefficient of
-determination; \(R^2\)). We solve a multiple linear regression problem,
-where outcome \(Y\) is regressed on predictors \(X_1\), \(X_2\) and
-\(X_3\)}
-
-\[Y \sim \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3.\]
-
 The complete data is \emph{amputed} once for each simulation repetition
-with function \texttt{mice::ampute()}.
-
-(\textbf{change this to reflect current simulation set-up with 5, 25,
-50, 75, and 95\% of cases having missing data: The missingness is
-univariate, and the probability to be missing is the same for all four
-variables, namely 20\% (\texttt{prop\ =\ 0.8,\ mech\ =\ "MCAR"}). This
-leaves 20\% of the rows completely observed). }
-
-\textbf{Maybe move to discussion:}Proper performance of the convergence
-diagnostics under at least missing completely at random (MCAR)
+with function \texttt{mice::ampute()}. The missingness is multivariate,
+and the probability to be missing is the same for all \(n \times k\)
+cells in \(y\) \citep[`MCAR';][]{rubin87}. Proper performance of the
+convergence diagnostics under at least `missing completely at random'
 missingness mechanism is necessary to demonstrate the appropriateness of
 \(\widehat{R}\) and \(AC\) as convergence diagnostics. However, results
-may not be extrapolated to other missingness mechanisms.
-\textbf{Convergence diagnostics should therefore at least apply to the
-MCAR situation, before more complex missingness should be explored.}
+may not be extrapolated to other missingness mechanisms. The missinness
+proportion varies between conditions (\(p_{mis} =.05,.25,.5,.75,.95\)).
 
@@
 chains (\texttt{m\ =\ 5}). The number of iterations varies between
-simulation conditions (\texttt{maxit\ =} \(1, 2, \dots, 100\)). Each
-repetition of the simulation thus starts from 1 complete dataset, then
-splits into 5 missingness conditions (\(p_{mis} =.05,.25,.5,.75,.95\)),
-each of the amputed datasets is imputed 5 times (\(m = 1, 2,..., 5\))
-for each iteration condition (\(t = 1, 2,..., 100\)). From these
-\(5\times5\times100\) imputations, we extract the \(\theta\)s to apply
-\(\widehat{R}\) and \(AC\) on (\textbf{i.e., chain means, chain
-variances, \(\beta\)s per imputation, and the first eigenvalue of the
-variance-covariance matrix per imputation}). We calculate
-\(\widehat{R}\) by implementing Vehtari et al.'s \citeyearpar{veht19}
-recommendations, and \(AC\) as the correlation between the \(t^{th}\)
-and the \((t+1)^{th}\) iteration. For each \(\theta\) we estimate the
-corresponding scientific estimand \(Q\).
+simulation conditions (\texttt{maxit\ =} \(1, 2, \dots, 100\)).
+
+The methods under evaluation, \(\widehat{R}\) and \(AC\), are computed
+for four scalar summaries \(\theta\): chain means (i.e., the mean in
+each \(y_{imp, m, j}\)), chain variances (i.e., the variance in each
+\(y_{imp, m, j}\)), the statistic of scientific interest \(\hat{Q}\) in
+each \{\(y_{obs}, y_{imp, m}\)\}, and the first eigenvalue of the
+variance-covariance matrix in each \{\(y_{obs}, y_{imp, m}\)\},
+\(\lambda_1\). We calculate \(\widehat{R}\) by implementing Vehtari et
+al.'s \citeyearpar{veht19} recommendations, and \(AC\) as the
+correlation between the \(t^{th}\) and the \((t+1)^{th}\) iteration. For
+each \(\theta\) we estimate the corresponding scientific estimand \(Q\).
+
+The performance of two methods on each of these scalar summaries will be
+evaluated with several estimands, i.e.~quantities of scientific interest
+\(Q\). For each \(\theta\), one estimand \(Q\) is defined that may be of
+interest in empirical research. The \(\widehat{R}\) and \(AC\) values
+with chain means as \(\theta\)s are evaluated against the bias in
+univariate mean estimates. Similarly, the performance measure for
+\(\widehat{R}\) and \(AC\) applied to chain variances is the bias in
+estimated standard deviations. To evaluate the performance of
+\(\widehat{R}\) and \(AC\) on the eigenvalues, we will use bias in the
+coefficient of determination, \(R^2\). Additionally, we will evaluate
+the estimated regression coefficients, the coverage rate of the CI95\%
+of the regression estimates, and the CI length (\textbf{see definitions
+in old Methods}). \textbf{remove the term performance measure here and
+add an extra paragraph to define performance measures as bias in all
+estimands, and coverage rate and CI length of regression coefficients.}
+
+We consider four quantities of scientific interest, namely the
+descriptive statistics of all variables (mean and standard deviation),
+\(\beta_2\), and the percentage of variance in \(Y\) explained
+(coefficient of determination \(R^2 \times 100\)). We solve a multiple
+linear regression problem, where outcome \(Y\) is regressed on
+predictors \(X_1\), \(X_2\) and \(X_3\)
+
+\[Y \sim \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3.\]
 
@@
 
-\hypertarget{old-results-section}{%
-\section{Old Results section}\label{old-results-section}}
+\hypertarget{results}{%
+\section{Results}\label{results}}
 
@@
 \(X_1\) on \(Y\).
+
+As expected, conditions with a higher proportion of missingness and/or a
+lower number of iterations show more signs of non-convergence, as
+indicated by more extreme bias in the estimated \(Q\)s. Roughly speaking
+this means that MICE is indeed not converged as \(t=1\), and converges
+gradually as \(t\) increases. The point at which an additional iteration
+does not lead to an improvement of the estimates depends on the
+difficulty of the missingness problem---between XYZ and XYZ. This means
+that the algorithm has converged sufficiently (\textbf{under the current
+specifications}).
+
+As required for a method to diagnose non-convergence, the values of
+\(\widehat{R}\) follow a trend similar to the performance measures.
+\(\widehat{R}\) values are generally lower in conditions with a higher
+number of iterations, and somewhat higher in conditions with a higher
+percentage of missing data. Autocorrelation values were indeed
+decreasing with a higher number of iterations.
+
+Evaluation with the performance measures shows that \(\widehat{R}\) and
+autocorrelation are conservative: they indicate signs of non-convergence
+in conditions where \(\bar{Q}\)s are unbiased and or confidence valid
+estimates of \(Q\).
+
+\hypertarget{old-results-section}{%
+\subsection{Old Results section}\label{old-results-section}}
 
@@
 
+Upon convergence, \(\widehat{R}=1\) and \(AC=0\), which are unlikely
+thresholds for MCMC algorithms, because of its convergence to a
+distribution. In practice, non-convergence is usually diagnosed when
+\(\widehat{R}\) \textgreater{} 1.2 or 1.1 or even 1.01. And a t-test is
+performed to assess whether \(AC\) is significantly different from zero.
+
+As expected, complete convergence (\(\widehat{R}=1\) and \(AC=0\)) is
+never observed. But what about the thresholds for practice?
+\(\widehat{R}\) \textless1.2 is not stringent enough, because conditions
+in which \(\widehat{R}\) is smaller than 1.2 do not all yield unbiased
+estimates of the \(Q\)s. Moreover, there is a dip in the \(\widehat{R}\)
+values, after which \(\widehat{R}\) \textgreater{} 1.2 again. If the
+algorithms would be terminated at the iteration where \(\widehat{R}\)
+\textless{} 1.2 occurs for the first time, the increase in
+\(\widehat{R}\) values might be missed. The threshold \(\widehat{R}\)
+\textless{} 1.1 is somewhat conservative in comparison with the
+performance measures. Overall it seems OK. \(\widehat{R}\) \textless{}
+1.01 is too stringent compared to the performance measures. And it is
+not obtained for any number of iterations (\textbf{but may be necessary
+in more complex complete data models???}).
+
 This note shows that convergence diagnostics \(\widehat{R}\) and \(AC\)

### 2020-04-30 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
add regular rhat in figure 3, adjust methods section to include hypotheses and better structure with qs and thetas.

@@
  text = element_text(size = 8),
- title = element_text(size = 8)
+ title = element_text(size = 8),
+ legend.title = element_text(size = 7)
+
  # axis.title = element_text(size = 8),
@@
 
+The goal of imputation through e.g.
+FCS/MICE is to pbtain a multivariate distribution by iterating over the sequence of univariate imputations.
+Traditionally, the inspection of the multivariate convergence is done by visually inspecting the univariate chain means and variances over the iterations, or by tracking the Rhat and autocorrelation.
+The downside to these approaches is that they either focus on the univariate state space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest.
+Focusing on the convergence of these outcome parameters may influence this procedure in the sense that the model of  evaluation favors the model of interest.
+Ideally, one would like to evaluate a model-independent parameter that summarizes the multivariate nature of the data.
+We propose eigenvalue as such a parameter, because it....
+
 Mixing and stationarity have histroically been inspected visually, by evaluating traceplots of scalar summaries of interest ($\theta$s; e.g., chain means and chain variances).
@@
 
-```{r non-conv, out.width='.49\\linewidth', fig.width=3.5, fig.height=3,fig.show='hold',fig.cap="Typical convergence versus pathological non-convergence."}
+```{r non-conv, out.width='.49\\linewidth', fig.width=3.5, fig.height=3,fig.show='hold',fig.cap="Typical convergence versus pathological non-convergence.
+Please note that this is the same data with a different imputation model, leading to different imputations (see the y axis!).
+Wel min of meer dezelfde origin op iteration 1."}
 load("./images/nonc_plot.Rdata")
@@
 
-```{r diagnostics, fig.cap="Convergence diagnostics pathological non-convergence versus typical convergence."}
-load("./images/diagnostics_plot.Rdata")
+We assume that complete convergence, defined as $\widehat{R} = 1$ and $AC = 0$, will not be observed.
+Because even in the most converged state, the algorithm will show some signs of non-mixing and non-stationarity.
+Upon *sufficient* convergence, the imputation chains will intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm ($\widehat{R} > 1$), and there may be some dependency between subsequent iterations of imputation chains ($AC > 0$).
+
+```{r diagnostics, fig.cap="Convergence diagnostics applied on the imputation algorithms of Figure 2."}
+load("./images/diagnostics_plot.Rdata") #bquote(theta ~ "= chain mean in "~ y[imp][",m"])
 diagnostics_plot
@@
 
-**Add to this section: 1) explicit mention of simulation conditions; 2) emphasize that the plots are averages across repetitions, not within MICE; 3) sample effects due to single complete dataset**
+The aim of the simulation study is to evaluate the impact of inducing non-convergence in the MICE algorithm on several quantities of scientific interest $Q$.
+Subsequently we will evaluate how well $\widehat{R}$ and $AC$ perform in identifying the effects of non-convergence.
+And finally, we will formulate an informed advice on the requirements to safely assume *sufficient* convergence in practice.
+That is, we assume that the algorithm is *sufficiently* converged when each estimate $\bar{Q}$ is an unbiased and confidence valid estimate of the corresponding estimand $Q$.
 
-The aim of the simulation study is to evaluate the impact of inducing non-convergence on several quantities of scientific interest $Q$, and to subsequently evaluate the performance of $\widehat{R}$ and $AC$ in identifying the non-convergence.
+Non-convergence will be induced by: 1) increasing the missingness proportion $p_{mis}$ in dataset $y$ incrementally, and 2) terminating the imputation algorithm at a varying number of iterations $T$.
+The first set of simulation conditions---the missingness proportions---is chosen to reflect the difficulty of the missingness problem.
+The underlying assumption is that low missingness proportions lead to quick algorithmic convergence, since there is a lot of information in the observed data.
+Higher missingness proportions should yield slower convergence.
+Unless, however, the fraction of missing information is so high that the random component in the imputation algorithm outweighs the information in the observed data.
+Then, convergence to a stable but highly variable state may be reached instanteneously.
+We assume that the incremental missingness proportions in our study will result in a corresponding increase in signs of non-convergence.
 
-Non-convergence will be induced by: 1) varying the missingness proportions in $y$ ($p_{mis} =.05,.25,.50,.75,.95$), and 2) terminating the MICE algorithm at different imputation chain lengths ($t = 1, 2,..., 100$).
-The assumption underlying the different number of iterations is that the algorithm generally does reach convergence at $t=1$.
-The initial values in the first iteration are sampled randomly from the set of observed datapoints.
-**As the number of iterations goes up, the imputation chains become independent of the initial values until the point at which an extra added iteration does not lead to a more converged state.** The second set of experimental conditions under consideration--the missingness proportions--are chosen to reflect the difficulty of the missingness problem.
-The inherent assumption is that low missingness proportions will lead to quick algorithmic convergence, since there is a lot of information in the observed data.
-Higher missingness proportions then cause slower convergence.
-However, if the fraction of missing information is very high, there is so little information in the data that the random component in the algorithm will take the overhand and a stable but very uncertain (high variance) state will be reached.
+The assumption inherent to the second set of simulation conditions---the number of iterations---is that terminating the imputation algorithm too early causes non-convergence.
+Generally, the algorithm will not reach convergence if $T=1$, because the imputed values in the first iteration (at $t=1$) depend on the initial values of the algorithm (which are sampled randomly from the set of observed datapoints).
+As the number of iterations increases, the imputation chains should become independent of the initial values, until the point at which adding an extra iteration does not lead to a more converged state.
+We assume that we can induce non-convergence at least in conditions where $T$ is smaller than the default number of iterations in `mice`, $T=5$.
 
-## Hypothesis
+## Hypotheses
 
-We hypothesize that simulation conditions with a higer $p_miss$ and lower $t$ have inferior convergence.
-We assume that when $\bar{Q}$ is an unbiased and confidence valid estimate of $Q$, the algorithm is *sufficiently* converged.
+1.
+We expect that simulation conditions with a high missingness proportion $p_{mis}$ and a low number of iterations $T$ will result in biased, invalid estimates of the quantities of scientific interest, $Q$s.
 
-**(bruggetje hier)** For multiple imputation algorithms, it holds that convergence is reached when there is no dependency between subsequent iterations of imputation chains ($AC = 0$), and chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm ($\widehat{R} = 1$).
-**We expect that a completely converged state will not be reached.**
+2.
+We hypothesize that $\widehat{R}$ and $AC$ will correctly identify signs of non-convergence in those simulation conditions where the $\bar{Q}$s are *not* unbiased and confidence valid estimates of the $Q$s.
+
+3.
+We hypothesize that the recommended thresholds to diagnose non-convergence with $\widehat{R}$ ($\widehat{R} > 1.2$, $\widehat{R} > 1.1$, and $\widehat{R} > 1.01$) may be too stringent for iterative imputation applications.
+In an empirical study, where $\widehat{R}$ was used to inform the required imputation chain length, it took as many as 50 iterations to overcome the conventional non-convergence threshold $\widehat{R}>1.2$.
+Yet, scientific estimates were insensitive to continued iteration after $t>5$ [@lace07].
+We therefore suspect that $\widehat{R}$ may over-estimate signs of non-convergence in iterative imputation algorithms.
+In contrast to this, it may occur that the signs of non-convergence are under-estimated by $\widehat{R}$, in exceptional cases where the initial values of the algorithm are not appropriately over-dispersed [@broo98, p.
+437].
+In e.g.
+`mice`, initial values are chosen randomly from the observed data, hence we cannot be certain of over-dispersion in the initial values.
+In practice, we do not expect this to cause problems for identifying non-convergence with $\widehat{R}$.
+
+4.
+We expect that high $AC$ values are implausible in iterative imputation algorithms with typical convergence.
+That is, after only a few iterations, the randomness induced by the algorithm will effectively mitigate the risk of dependency within chains.
+
+<!-- **(bruggetje hier)** For multiple imputation algorithms, it holds that convergence is reached when there is no dependency between subsequent iterations of imputation chains ($AC = 0$), and chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm ($\widehat{R} = 1$).
+**We expect that a completely converged state will not be reached.**  -->
 
@@
 
-Based on an empirical finding [@lace07], we hypothesize that $\widehat{R}$ will over-estimate non-convergence of MI algorithms **explain why?**.
+<!-- Based on an empirical finding [@lace07], we hypothesize that $\widehat{R}$ will over-estimate non-convergence of MI algorithms **explain why?**.
 Lacerda concluded that rhat dropped below 1.2 after at least 50 iterations.
@@
 We expect this to have little effect on the hypothesized performance of $\widehat{R}$.
-**(Add actual hypothesis: we thought that mice would converge sooner than $\widehat{R}$ would indicate that it did)** We hypothesize that high AC values are implausible in converged MI algorithms.
-**Added later: High $AC$-values are implausible in MI procedures.
-That is, the randomness induced by the MI algorithm effectively mitigates the risk of dependency within chains.
-But this doesn't seem the case in practice??**
+**(Add actual hypothesis: we thought that mice would converge sooner than $\widehat{R}$ would indicate that it did).
+On the one hand, $\widehat{R}$ is known to under-estimate signs of non-convergence when the initial values of the algorithm are not appropriately over-dispersed [@broo98, p.
+437], which may be the case for iterative imputation algorithms.
+In e.g.
+`mice`, initial values are chosen randomly from the observed data.
+Therefore, we cannot be certain that the initial values are sufficiently over-dispersed.
+**  -->
 
@@
    4. perform analysis of scientific interest
-   5. compute convergence diagnostics 
+   5. compute non-convergence diagnostics 
    6. pool results across imputations
@@
 ```
+### Data-generating mechanism.
 
-The data-generating mechanism is a multivariate normal distribution, representing person-data on three predictor variables (from an unspecified social scientific field of study).
-Let our predictor space be defined as
-\begin{align*}
+In this study, sampling variance is not of interest.
+Therefore, a single complete dataset may serve as comparative truth in all simulation repetitions [@vink14].
+The data-generating mechanism is a multivariate normal distribution, representing person-data on three predictor variables in a multiple linear regression problem **(from an unspecified social scientific field of study)**.
+Let the predictor space be defined as
+$$
+\begin{aligned}
 \begin{pmatrix}X_1\\
@@
 \end{bmatrix}\!\!\text{.}\\[2\jot]
-\end{align*}
-
-In this study, sampling variance is not of interest.
-Therefore, a single complete set may serve as comparative truth in all simulation runs [@vink14].
-A finite population of $N=1000$ is simulated using the `mvtnorm` package [@mvtnorm].
-Subsequently, a fourth variable is constructed  as outcome variable in a multiple linear regression problem.
-Let
+\end{aligned}
+$$
+A finite population of $N=1000$ is simulated using the `mvtnorm` package [@mvtnorm]. Subsequently, a fourth variable is constructed as outcome $Y$. For each unit $i = 1, 2,..., N$, let 
 $$
@@
 $$
-where $i = 1, 2,..., N$ and $\epsilon \sim \mathcal{N}(0, 100)$.
+where $\epsilon \sim \mathcal{N}(0, 100)$. From the complete set, we obtain the true values of each quantity of scientific interest $Q$.
 
-The complete data is *amputed* once for each simulation repetition with function `mice::ampute()`.
-The missingness is multivariate, and the probability to be missing is the same for all $n \times k$ cells in $y$ ['MCAR'; @rubin87].
-Proper performance of the convergence diagnostics under at least 'missing completely at random' missingness mechanism is necessary to demonstrate the appropriateness of $\widehat{R}$ and $AC$ as convergence diagnostics.
-However, results may not be extrapolated to other missingness mechanisms.
-The missinness proportion varies between conditions ($p_{mis} =.05,.25,.5,.75,.95$).
-<!-- This note only considers a 'missing completely at random' (MCAR) mechanism, where the probability of being missing is equal for all $n \times k$ cells in $y$ [@rubin87]. -->
+### Scientific estimands.
 
-Missing datapoints in $y$ are imputed with the `mice` package [@mice].
-All MI procedures are performed with Bayesian linear regression imputation (`method = "norm"`), and five imputation chains (`m = 5`).
-The number of iterations varies between simulation conditions (`maxit = ` $1, 2, \dots, 100$).
+We consider four types of $Q$s that are often of interest in empirical research.
+Namely, two descriptive statistics: the mean $\mu_j$ and standard deviation $\sigma_j$ of each variable $j = Y, X_1, X_2, X_3$.
+We also consider the regression coefficients of the predictors: $\beta_1$, $\beta_2$, and $\beta_3$ in regression equation
+$$Y' = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3,$$
+where $Y'$ is the expected value of the outcome.
+And finally, the proportion of variance explained by the set of predictors: coefficient of determination $r^2 \times 100$ **(note that lower case $r$ is used to avoid confusion with non-convergence diagnostic $\widehat{R}$)**.
+The true values of these scientific estimands are calculated with the R functions `base::mean()`, `stats::sd()`, and `stats::lm()` [@R].
+Each of these $Q$s is estimated with a corresponding $\bar{Q}$, after the complete dataset is first *amputed* (with a varying missingness proportion $p_{mis}$) and subsequently imputed (with a varying number of iterations $T$).
+
+### Methods.
+
+The complete dataset is amputed with the `mice` package [function `mice::ampute()`; @mice].
+All missingness is multivariate, and conform a 'missing completely at random' missingness mechanism [@rubin87].
+I.e., the probability to be missing is the same for all $N \times k$ cells in $y$, conditional on the missingness proportion ($p_{mis} =.05,.25,.5,.75,.95$).
+<!-- This note only considers a 'missing completely at random' (MCAR) mechanism, where the probability of being missing is equal for all $n \times k$ cells in $y$ [@rubin87].
+The probability to be missing is the same for all $n \times k$ cells in $y$ [a 'missing completely at random' missingness mechanism; @rubin87].
+-->
+
+Missing datapoints in $y$ are imputed with the `mice()` function [@mice].
+All imputation procedures are performed with Bayesian linear regression imputation, and five imputation chains ($M=5$).
+The number of iterations varies between simulation conditions ($T=1,2,...,100$).
 <!-- Each repetition of the simulation thus starts from 1 complete dataset, then splits into 5 missingness conditions ($p_{mis} =.05,.25,.5,.75,.95$), each of the amputed datasets is imputed 5 times ($m = 1, 2,..., 5$) for each iteration condition ($t = 1, 2,..., 100$).
@@
 
-The methods under evaluation, $\widehat{R}$ and $AC$, are computed for four scalar summaries $\theta$: chain means (i.e., the mean in each $y_{imp, m, j}$), chain variances (i.e., the variance in each $y_{imp, m, j}$), the statistic of scientific interest $\hat{Q}$ in each {$y_{obs}, y_{imp, m}$}, and the first eigenvalue of the variance-covariance matrix in each {$y_{obs}, y_{imp, m}$}, $\lambda_1$.
-We calculate $\widehat{R}$ by implementing Vehtari et al.'s [-@veht19] recommendations, and $AC$ as the correlation between the $t^{th}$ and the $(t+1)^{th}$ iteration.
-For each $\theta$ we estimate the corresponding scientific estimand $Q$.
+After each run of the `mice` algorithm, the function `mice::complete()` is used to combine the imputed data $y_{imp,m}$ with the observed data $y_{obs}$, resulting in five completed datasets {$y_{obs}, y_{imp, m}$}, where $m = 1, 2, ..., 5$.
 
-The performance of two methods on each of these scalar summaries will be evaluated with several estimands, i.e.
-quantities of scientific interest $Q$.
-For each $\theta$, one estimand $Q$ is defined that may be of interest in empirical research.
-The $\widehat{R}$ and $AC$ values with chain means as $\theta$s are evaluated against the bias in univariate mean estimates.
-Similarly, the performance measure for $\widehat{R}$ and $AC$ applied to chain variances is the bias in estimated standard deviations.
-To evaluate the performance of $\widehat{R}$ and $AC$ on the eigenvalues, we will use bias in the coefficient of determination, $R^2$.
-Additionally, we will evaluate the estimated regression coefficients, the coverage rate of the CI95% of the regression estimates, and the CI length (**see definitions in old Methods**).
-**remove the term performance measure here and add an extra paragraph to define performance measures as bias in all estimands, and coverage rate and CI length of regression coefficients.**
+- on the 5 imputations $y_{imp,m}$: get mean and var within imp chains as theta to compute conv diag --> then complete with `mice::complete()`
 
-We consider four quantities of scientific interest, namely the descriptive statistics of all variables (mean and standard deviation), $\beta_2$, and the percentage of variance in $Y$ explained (coefficient of determination $R^2 \times 100$).
-We solve a multiple linear regression problem, where outcome $Y$ is regressed on predictors $X_1$, $X_2$ and $X_3$
-
-$$Y \sim \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3.$$
+- on the 5 completed datasets {$y_{obs}, y_{imp, m}$}: 1) estimate Qhats for mu and sigma --> pool as mean() --> get Qbars of mu and sigma; 2) compute lambda as theta for conv diag; 3) perform lm to get Qhats for betas and r^2 --> use Qhats of beta as theta to compute conv diag --> pool Qhats of beta and r^2 to get Qbars
 
@@
 
-The performance of $\widehat{R}$ and $AC$ is assessed by comparing $\bar{Q}$s with $Q$s.
+For each run of the imputation algorithm, we compute non-convergence diagnostics $\widehat{R}$ and $AC$.
+We calculate $\widehat{R}$ by implementing Vehtari et al.'s [-@veht19] recommendations, and $AC$ as the correlation between the $t^{th}$ and the $(t+1)^{th}$ iteration.
+The two methods are applied to four scalar summaries $\theta$: chain means (i.e., the mean in each $y_{imp, m, j}$), chain variances (i.e., the variance in each $y_{imp, m, j}$), a statistic of scientific interest $\hat{Q}$ in each {$y_{obs}, y_{imp, m}$}, and the first eigenvalue of the variance-covariance matrix in each {$y_{obs}, y_{imp, m}$}, $\lambda_1$.
+
+The $\widehat{R}$ and $AC$ values with chain means as $\theta$s are evaluated against the bias in univariate mean estimates.
+Similarly, the performance measure for $\widehat{R}$ and $AC$ applied to chain variances is the bias in estimated standard deviations.
+To evaluate the performance of $\widehat{R}$ and $AC$ on the eigenvalues, we will use bias in the coefficient of determination, $R^2$.
+Additionally, we will evaluate the estimated regression coefficients, the coverage rate of the CI95% of the regression estimates, and the CI length (**see definitions in old Methods**).
+**remove the term performance measure here and add an extra paragraph to define performance measures as bias in all estimands, and coverage rate and CI length of regression coefficients.**
+
+### Performance measures.
+
+The impact of inducing non-convergence in the iterative imputation algorithm is assessed by comparing $\bar{Q}$s with $Q$s.
 For each $Q$, we compute bias as $\bar{Q} - Q$.
-For $Q=\beta_j$, we also compute the empirical coverage rate (CR).
+For the regression coefficients $Q=\beta_{1,2,3}$, we also compute the empirical coverage rate (CR).
 CR is defined as the percentage of simulation repetitions in which the 95\% confidence interval (CI) around $\bar{Q}$ covers the true estimand $Q$.
 Let
+$$\text{CI} = \bar{Q} \pm t_{(M-1)} \times SE_{\bar{Q}},$$
+where $t_{(M-1)}$ is the quantile of a $t$-distribution with $M-1$ degrees of freedom, and $SE_{\bar{Q}}$ is the square root of the pooled variance estimate.
+Since under-estimating the variance of $\bar{Q}$ may yield spurious inferences, we also inspect CI width (CIW): the difference between the lower and upper bound of the 95\% confidence interval around $\bar{Q}$.
 
-$$\text{CI} = \bar{Q} \pm t_{(M-1)} \times SE_{\bar{Q}},$$
-
-where $t_{(M-1)}$ is the quantile of a $t$-distribution with $M-1$ degrees of freedom, and $SE_{\bar{Q}}$ is the square root of the pooled variance estimate.
-**Remove this?
-** Under-estimating the variance of $\bar{Q}$ may yield spurious inferences.
-Confidence interval width (CIW) is defined as the difference between the lower and upper bound of the 95\% confidence interval (CI) around $\bar{Q}$ From bias and CIW, we calculate empirical coverage rates.
-Coverage rate is the proportion of simulations in which $Q$ is between the bounds of the CI95\% around $\bar{Q}$.
+The performance of two methods is evaluated against the bias and confidence validity of the estimated $Q$s.
 
 # Results
+
+**Add to this section: 1) emphasize that the plots are averages across repetitions, not within MICE; 2) sample effects due to single complete dataset.**
 
@@
 
+# qnorm((1 + .95) / 2) / sqrt(t)
+
 ```
@@
 # Discussion
+
+*Proper performance under a 'missing completely at random' missingness mechanism is necessary to demonstrate the appropriateness of $\widehat{R}$ and $AC$ as non-convergence diagnostics.
+However, results may not be extrapolated to other missingness mechanisms.*
 

### 2020-04-30 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
add regular rhat in figure 3, adjust methods section to include hypotheses and better structure with qs and thetas.

@@
 
+The goal of imputation through e.g.~FCS/MICE is to pbtain a multivariate
+distribution by iterating over the sequence of univariate imputations.
+Traditionally, the inspection of the multivariate convergence is done by
+visually inspecting the univariate chain means and variances over the
+iterations, or by tracking the Rhat and autocorrelation. The downside to
+these approaches is that they either focus on the univariate state
+space, or primarily track the change over the iterations of a
+multivariate outcome conform the scientific model of interest. Focusing
+on the convergence of these outcome parameters may influence this
+procedure in the sense that the model of evaluation favors the model of
+interest. Ideally, one would like to evaluate a model-independent
+parameter that summarizes the multivariate nature of the data. We
+propose eigenvalue as such a parameter, because it\ldots.
+
 Mixing and stationarity have histroically been inspected visually, by
@@
 
-\caption{Typical convergence versus pathological non-convergence.}\label{fig:non-conv}
+\caption{Typical convergence versus pathological non-convergence.
+Please note that this is the same data with a different imputation model, leading to different imputations (see the y axis!).
+Wel min of meer dezelfde origin op iteration 1.}\label{fig:non-conv}
 \end{figure}
@@
 
+We assume that complete convergence, defined as \(\widehat{R} = 1\) and
+\(AC = 0\), will not be observed. Because even in the most converged
+state, the algorithm will show some signs of non-mixing and
+non-stationarity. Upon \emph{sufficient} convergence, the imputation
+chains will intermingle such that the only difference between the chains
+is caused by the randomness induced by the algorithm
+(\(\widehat{R} > 1\)), and there may be some dependency between
+subsequent iterations of imputation chains (\(AC > 0\)).
+
 \begin{figure}
@@
 
-\caption{Convergence diagnostics pathological non-convergence versus typical convergence.}\label{fig:diagnostics}
+\caption{Convergence diagnostics applied on the imputation algorithms of Figure 2.}\label{fig:diagnostics}
 \end{figure}
@@
 
-\textbf{Add to this section: 1) explicit mention of simulation
-conditions; 2) emphasize that the plots are averages across repetitions,
-not within MICE; 3) sample effects due to single complete dataset}
+The aim of the simulation study is to evaluate the impact of inducing
+non-convergence in the MICE algorithm on several quantities of
+scientific interest \(Q\). Subsequently we will evaluate how well
+\(\widehat{R}\) and \(AC\) perform in identifying the effects of
+non-convergence. And finally, we will formulate an informed advice on
+the requirements to safely assume \emph{sufficient} convergence in
+practice. That is, we assume that the algorithm is \emph{sufficiently}
+converged when each estimate \(\bar{Q}\) is an unbiased and confidence
+valid estimate of the corresponding estimand \(Q\).
 
-The aim of the simulation study is to evaluate the impact of inducing
-non-convergence on several quantities of scientific interest \(Q\), and
-to subsequently evaluate the performance of \(\widehat{R}\) and \(AC\)
-in identifying the non-convergence.
+Non-convergence will be induced by: 1) increasing the missingness
+proportion \(p_{mis}\) in dataset \(y\) incrementally, and 2)
+terminating the imputation algorithm at a varying number of iterations
+\(T\). The first set of simulation conditions---the missingness
+proportions---is chosen to reflect the difficulty of the missingness
+problem. The underlying assumption is that low missingness proportions
+lead to quick algorithmic convergence, since there is a lot of
+information in the observed data. Higher missingness proportions should
+yield slower convergence. Unless, however, the fraction of missing
+information is so high that the random component in the imputation
+algorithm outweighs the information in the observed data. Then,
+convergence to a stable but highly variable state may be reached
+instanteneously. We assume that the incremental missingness proportions
+in our study will result in a corresponding increase in signs of
+non-convergence.
 
-Non-convergence will be induced by: 1) varying the missingness
-proportions in \(y\) (\(p_{mis} =.05,.25,.50,.75,.95\)), and 2)
-terminating the MICE algorithm at different imputation chain lengths
-(\(t = 1, 2,..., 100\)). The assumption underlying the different number
-of iterations is that the algorithm generally does reach convergence at
-\(t=1\). The initial values in the first iteration are sampled randomly
-from the set of observed datapoints. \textbf{As the number of iterations
-goes up, the imputation chains become independent of the initial values
-until the point at which an extra added iteration does not lead to a
-more converged state.} The second set of experimental conditions under
-consideration--the missingness proportions--are chosen to reflect the
-difficulty of the missingness problem. The inherent assumption is that
-low missingness proportions will lead to quick algorithmic convergence,
-since there is a lot of information in the observed data. Higher
-missingness proportions then cause slower convergence. However, if the
-fraction of missing information is very high, there is so little
-information in the data that the random component in the algorithm will
-take the overhand and a stable but very uncertain (high variance) state
-will be reached.
+The assumption inherent to the second set of simulation conditions---the
+number of iterations---is that terminating the imputation algorithm too
+early causes non-convergence. Generally, the algorithm will not reach
+convergence if \(T=1\), because the imputed values in the first
+iteration (at \(t=1\)) depend on the initial values of the algorithm
+(which are sampled randomly from the set of observed datapoints). As the
+number of iterations increases, the imputation chains should become
+independent of the initial values, until the point at which adding an
+extra iteration does not lead to a more converged state. We assume that
+we can induce non-convergence at least in conditions where \(T\) is
+smaller than the default number of iterations in \texttt{mice}, \(T=5\).
 
-\hypertarget{hypothesis}{%
-\subsection{Hypothesis}\label{hypothesis}}
+\hypertarget{hypotheses}{%
+\subsection{Hypotheses}\label{hypotheses}}
 
-We hypothesize that simulation conditions with a higer \(p_miss\) and
-lower \(t\) have inferior convergence. We assume that when \(\bar{Q}\)
-is an unbiased and confidence valid estimate of \(Q\), the algorithm is
-\emph{sufficiently} converged.
-
-\textbf{(bruggetje hier)} For multiple imputation algorithms, it holds
-that convergence is reached when there is no dependency between
-subsequent iterations of imputation chains (\(AC = 0\)), and chains
-intermingle such that the only difference between the chains is caused
-by the randomness induced by the algorithm (\(\widehat{R} = 1\)).
-\textbf{We expect that a completely converged state will not be
-reached.}
-
-Based on an empirical finding \citep{lace07}, we hypothesize that
-\(\widehat{R}\) will over-estimate non-convergence of MI algorithms
-\textbf{explain why?}. Lacerda concluded that rhat dropped below 1.2
-after at least 50 iterations. Therefore we hypothesize that Vehtari's
-proposed threshold of \(\widehat{R} < 1.01\) will be too stringent for
-diagnosing convergence in iterative imputation. This over-estimation
-may, however, be diminished because \(\widehat{R}\) can falsely diagnose
-convergence if initial values of the algorithm are not appropriately
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\item
+  We expect that simulation conditions with a high missingness
+  proportion \(p_{mis}\) and a low number of iterations \(T\) will
+  result in biased, invalid estimates of the quantities of scientific
+  interest, \(Q\)s.
+\item
+  We hypothesize that \(\widehat{R}\) and \(AC\) will correctly identify
+  signs of non-convergence in those simulation conditions where the
+  \(\bar{Q}\)s are \emph{not} unbiased and confidence valid estimates of
+  the \(Q\)s.
+\item
+  We hypothesize that the recommended thresholds to diagnose
+  non-convergence with \(\widehat{R}\) (\(\widehat{R} > 1.2\),
+  \(\widehat{R} > 1.1\), and \(\widehat{R} > 1.01\)) may be too
+  stringent for iterative imputation applications. In an empirical
+  study, where \(\widehat{R}\) was used to inform the required
+  imputation chain length, it took as many as 50 iterations to overcome
+  the conventional non-convergence threshold \(\widehat{R}>1.2\). Yet,
+  scientific estimates were insensitive to continued iteration after
+  \(t>5\) \citep{lace07}. We therefore suspect that \(\widehat{R}\) may
+  over-estimate signs of non-convergence in iterative imputation
+  algorithms. In contrast to this, it may occur that the signs of
+  non-convergence are under-estimated by \(\widehat{R}\), in exceptional
+  cases where the initial values of the algorithm are not appropriately
   over-dispersed \citep[p.~437]{broo98}. In e.g.~\texttt{mice}, initial
-values are chosen randomly from the observed data. Therefore, we cannot
-be certain that the initial values are over-dispersed. We expect this to
-have little effect on the hypothesized performance of \(\widehat{R}\).
-\textbf{(Add actual hypothesis: we thought that mice would converge
-sooner than \(\widehat{R}\) would indicate that it did)} We hypothesize
-that high AC values are implausible in converged MI algorithms.
-\textbf{Added later: High \(AC\)-values are implausible in MI
-procedures. That is, the randomness induced by the MI algorithm
-effectively mitigates the risk of dependency within chains. But this
-doesn't seem the case in practice??}
+  values are chosen randomly from the observed data, hence we cannot be
+  certain of over-dispersion in the initial values. In practice, we do
+  not expect this to cause problems for identifying non-convergence with
+  \(\widehat{R}\).
+\item
+  We expect that high \(AC\) values are implausible in iterative
+  imputation algorithms with typical convergence. That is, after only a
+  few iterations, the randomness induced by the algorithm will
+  effectively mitigate the risk of dependency within chains.
+\end{enumerate}
 
@@
 
-The data-generating mechanism is a multivariate normal distribution,
-representing person-data on three predictor variables (from an
-unspecified social scientific field of study). Let our predictor space
-be defined as \begin{align*}
+\hypertarget{data-generating-mechanism.}{%
+\subsubsection{Data-generating
+mechanism.}\label{data-generating-mechanism.}}
+
+In this study, sampling variance is not of interest. Therefore, a single
+complete dataset may serve as comparative truth in all simulation
+repetitions \citep{vink14}. The data-generating mechanism is a
+multivariate normal distribution, representing person-data on three
+predictor variables in a multiple linear regression problem
+\textbf{(from an unspecified social scientific field of study)}. Let the
+predictor space be defined as \[
+\begin{aligned}
 \begin{pmatrix}X_1\\
@@
 \end{bmatrix}\!\!\text{.}\\[2\jot]
-\end{align*}
+\end{aligned}
+\] A finite population of \(N=1000\) is simulated using the
+\texttt{mvtnorm} package \citep{mvtnorm}. Subsequently, a fourth
+variable is constructed as outcome \(Y\). For each unit
+\(i = 1, 2,..., N\), let \[
+Y_i = 1 + 2X_{1i} +.5X_{2i} - X_{3i} + \epsilon_i ,
+\] where \(\epsilon \sim \mathcal{N}(0, 100)\). From the complete set,
+we obtain the true values of each quantity of scientific interest \(Q\).
 
-In this study, sampling variance is not of interest. Therefore, a single
-complete set may serve as comparative truth in all simulation runs
-\citep{vink14}. A finite population of \(N=1000\) is simulated using the
-\texttt{mvtnorm} package \citep{mvtnorm}. Subsequently, a fourth
-variable is constructed as outcome variable in a multiple linear
-regression problem. Let \[
-Y_i = 1 + 2X_{1i} +.5X_{2i} - X_{3i} + \epsilon_i ,
-\] where \(i = 1, 2,..., N\) and \(\epsilon \sim \mathcal{N}(0, 100)\).
+\hypertarget{scientific-estimands.}{%
+\subsubsection{Scientific estimands.}\label{scientific-estimands.}}
 
-The complete data is \emph{amputed} once for each simulation repetition
-with function \texttt{mice::ampute()}. The missingness is multivariate,
-and the probability to be missing is the same for all \(n \times k\)
-cells in \(y\) \citep[`MCAR';][]{rubin87}. Proper performance of the
-convergence diagnostics under at least `missing completely at random'
-missingness mechanism is necessary to demonstrate the appropriateness of
-\(\widehat{R}\) and \(AC\) as convergence diagnostics. However, results
-may not be extrapolated to other missingness mechanisms. The missinness
-proportion varies between conditions (\(p_{mis} =.05,.25,.5,.75,.95\)).
+We consider four types of \(Q\)s that are often of interest in empirical
+research. Namely, two descriptive statistics: the mean \(\mu_j\) and
+standard deviation \(\sigma_j\) of each variable
+\(j = Y, X_1, X_2, X_3\). We also consider the regression coefficients
+of the predictors: \(\beta_1\), \(\beta_2\), and \(\beta_3\) in
+regression equation
+\[Y' = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3,\] where \(Y'\)
+is the expected value of the outcome. And finally, the proportion of
+variance explained by the set of predictors: coefficient of
+determination \(r^2 \times 100\) \textbf{(note that lower case \(r\) is
+used to avoid confusion with non-convergence diagnostic
+\(\widehat{R}\))}. The true values of these scientific estimands are
+calculated with the R functions \texttt{base::mean()},
+\texttt{stats::sd()}, and \texttt{stats::lm()} \citep{R}. Each of these
+\(Q\)s is estimated with a corresponding \(\bar{Q}\), after the complete
+dataset is first \emph{amputed} (with a varying missingness proportion
+\(p_{mis}\)) and subsequently imputed (with a varying number of
+iterations \(T\)).
 
-Missing datapoints in \(y\) are imputed with the \texttt{mice} package
-\citep{mice}. All MI procedures are performed with Bayesian linear
-regression imputation (\texttt{method\ =\ "norm"}), and five imputation
-chains (\texttt{m\ =\ 5}). The number of iterations varies between
-simulation conditions (\texttt{maxit\ =} \(1, 2, \dots, 100\)).
+\hypertarget{methods.}{%
+\subsubsection{Methods.}\label{methods.}}
 
-The methods under evaluation, \(\widehat{R}\) and \(AC\), are computed
-for four scalar summaries \(\theta\): chain means (i.e., the mean in
-each \(y_{imp, m, j}\)), chain variances (i.e., the variance in each
-\(y_{imp, m, j}\)), the statistic of scientific interest \(\hat{Q}\) in
-each \{\(y_{obs}, y_{imp, m}\)\}, and the first eigenvalue of the
-variance-covariance matrix in each \{\(y_{obs}, y_{imp, m}\)\},
-\(\lambda_1\). We calculate \(\widehat{R}\) by implementing Vehtari et
-al.'s \citeyearpar{veht19} recommendations, and \(AC\) as the
-correlation between the \(t^{th}\) and the \((t+1)^{th}\) iteration. For
-each \(\theta\) we estimate the corresponding scientific estimand \(Q\).
+The complete dataset is amputed with the \texttt{mice} package
+\citep[function \texttt{mice::ampute()};][]{mice}. All missingness is
+multivariate, and conform a `missing completely at random' missingness
+mechanism \citep{rubin87}. I.e., the probability to be missing is the
+same for all \(N \times k\) cells in \(y\), conditional on the
+missingness proportion (\(p_{mis} =.05,.25,.5,.75,.95\)).
 
-The performance of two methods on each of these scalar summaries will be
-evaluated with several estimands, i.e.~quantities of scientific interest
-\(Q\). For each \(\theta\), one estimand \(Q\) is defined that may be of
-interest in empirical research. The \(\widehat{R}\) and \(AC\) values
-with chain means as \(\theta\)s are evaluated against the bias in
-univariate mean estimates. Similarly, the performance measure for
-\(\widehat{R}\) and \(AC\) applied to chain variances is the bias in
-estimated standard deviations. To evaluate the performance of
-\(\widehat{R}\) and \(AC\) on the eigenvalues, we will use bias in the
-coefficient of determination, \(R^2\). Additionally, we will evaluate
-the estimated regression coefficients, the coverage rate of the CI95\%
-of the regression estimates, and the CI length (\textbf{see definitions
-in old Methods}). \textbf{remove the term performance measure here and
-add an extra paragraph to define performance measures as bias in all
-estimands, and coverage rate and CI length of regression coefficients.}
+Missing datapoints in \(y\) are imputed with the \texttt{mice()}
+function \citep{mice}. All imputation procedures are performed with
+Bayesian linear regression imputation, and five imputation chains
+(\(M=5\)). The number of iterations varies between simulation conditions
+(\(T=1,2,...,100\)).
 
-We consider four quantities of scientific interest, namely the
-descriptive statistics of all variables (mean and standard deviation),
-\(\beta_2\), and the percentage of variance in \(Y\) explained
-(coefficient of determination \(R^2 \times 100\)). We solve a multiple
-linear regression problem, where outcome \(Y\) is regressed on
-predictors \(X_1\), \(X_2\) and \(X_3\)
+After each run of the \texttt{mice} algorithm, the function
+\texttt{mice::complete()} is used to combine the imputed data
+\(y_{imp,m}\) with the observed data \(y_{obs}\), resulting in five
+completed datasets \{\(y_{obs}, y_{imp, m}\)\}, where
+\(m = 1, 2, ..., 5\).
 
-\[Y \sim \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3.\]
+\begin{itemize}
+\item
+  on the 5 imputations \(y_{imp,m}\): get mean and var within imp chains
+  as theta to compute conv diag --\textgreater{} then complete with
+  \texttt{mice::complete()}
+\item
+  on the 5 completed datasets \{\(y_{obs}, y_{imp, m}\)\}: 1) estimate
+  Qhats for mu and sigma --\textgreater{} pool as mean()
+  --\textgreater{} get Qbars of mu and sigma; 2) compute lambda as theta
+  for conv diag; 3) perform lm to get Qhats for betas and r\^{}2
+  --\textgreater{} use Qhats of beta as theta to compute conv diag
+  --\textgreater{} pool Qhats of beta and r\^{}2 to get Qbars
+\end{itemize}
 
@@
 
-The performance of \(\widehat{R}\) and \(AC\) is assessed by comparing
-\(\bar{Q}\)s with \(Q\)s. For each \(Q\), we compute bias as
-\(\bar{Q} - Q\). For \(Q=\beta_j\), we also compute the empirical
-coverage rate (CR). CR is defined as the percentage of simulation
-repetitions in which the 95\% confidence interval (CI) around
-\(\bar{Q}\) covers the true estimand \(Q\). Let
+For each run of the imputation algorithm, we compute non-convergence
+diagnostics \(\widehat{R}\) and \(AC\). We calculate \(\widehat{R}\) by
+implementing Vehtari et al.'s \citeyearpar{veht19} recommendations, and
+\(AC\) as the correlation between the \(t^{th}\) and the \((t+1)^{th}\)
+iteration. The two methods are applied to four scalar summaries
+\(\theta\): chain means (i.e., the mean in each \(y_{imp, m, j}\)),
+chain variances (i.e., the variance in each \(y_{imp, m, j}\)), a
+statistic of scientific interest \(\hat{Q}\) in each
+\{\(y_{obs}, y_{imp, m}\)\}, and the first eigenvalue of the
+variance-covariance matrix in each \{\(y_{obs}, y_{imp, m}\)\},
+\(\lambda_1\).
 
-\[\text{CI} = \bar{Q} \pm t_{(M-1)} \times SE_{\bar{Q}},\]
+The \(\widehat{R}\) and \(AC\) values with chain means as \(\theta\)s
+are evaluated against the bias in univariate mean estimates. Similarly,
+the performance measure for \(\widehat{R}\) and \(AC\) applied to chain
+variances is the bias in estimated standard deviations. To evaluate the
+performance of \(\widehat{R}\) and \(AC\) on the eigenvalues, we will
+use bias in the coefficient of determination, \(R^2\). Additionally, we
+will evaluate the estimated regression coefficients, the coverage rate
+of the CI95\% of the regression estimates, and the CI length
+(\textbf{see definitions in old Methods}). \textbf{remove the term
+performance measure here and add an extra paragraph to define
+performance measures as bias in all estimands, and coverage rate and CI
+length of regression coefficients.}
 
-where \(t_{(M-1)}\) is the quantile of a \(t\)-distribution with \(M-1\)
+\hypertarget{performance-measures.}{%
+\subsubsection{Performance measures.}\label{performance-measures.}}
+
+The impact of inducing non-convergence in the iterative imputation
+algorithm is assessed by comparing \(\bar{Q}\)s with \(Q\)s. For each
+\(Q\), we compute bias as \(\bar{Q} - Q\). For the regression
+coefficients \(Q=\beta_{1,2,3}\), we also compute the empirical coverage
+rate (CR). CR is defined as the percentage of simulation repetitions in
+which the 95\% confidence interval (CI) around \(\bar{Q}\) covers the
+true estimand \(Q\). Let
+\[\text{CI} = \bar{Q} \pm t_{(M-1)} \times SE_{\bar{Q}},\] where
+\(t_{(M-1)}\) is the quantile of a \(t\)-distribution with \(M-1\)
 degrees of freedom, and \(SE_{\bar{Q}}\) is the square root of the
-pooled variance estimate. \textbf{Remove this? } Under-estimating the
-variance of \(\bar{Q}\) may yield spurious inferences. Confidence
-interval width (CIW) is defined as the difference between the lower and
-upper bound of the 95\% confidence interval (CI) around \(\bar{Q}\) From
-bias and CIW, we calculate empirical coverage rates. Coverage rate is
-the proportion of simulations in which \(Q\) is between the bounds of
-the CI95\% around \(\bar{Q}\).
+pooled variance estimate. Since under-estimating the variance of
+\(\bar{Q}\) may yield spurious inferences, we also inspect CI width
+(CIW): the difference between the lower and upper bound of the 95\%
+confidence interval around \(\bar{Q}\).
+
+The performance of two methods is evaluated against the bias and
+confidence validity of the estimated \(Q\)s.
 
@@
 \section{Results}\label{results}}
+
+\textbf{Add to this section: 1) emphasize that the plots are averages
+across repetitions, not within MICE; 2) sample effects due to single
+complete dataset.}
 
@@
 
+\emph{Proper performance under a `missing completely at random'
+missingness mechanism is necessary to demonstrate the appropriateness of
+\(\widehat{R}\) and \(AC\) as non-convergence diagnostics. However,
+results may not be extrapolated to other missingness mechanisms.}
+
 Upon convergence, \(\widehat{R}=1\) and \(AC=0\), which are unlikely

### 2020-04-30 3.Thesis/1.SimulationStudy/Writeup/images/non-convergence.R

@@ -37,7 +37,7 @@ save(nonc, file = "3.Thesis/1.SimulationStudy/Writeup/images/nonc_plot.Rdata")
 
 # diagnostics on chain means of wgt (rows are iterations, columns are imputations)
 diagnostics1 <-
-  imp.bmi1$chainMean[3, ,] %>% cbind(patho = "Pathological", convergence(., include_acf = TRUE), .)
+  imp.bmi1$chainMean[3, ,] %>% cbind(patho = "Non-", convergence(., include_acf = TRUE), .)
 
 #####################
 # Regular convergence
@@ -97,109 +97,130 @@ theme_update(
   legend.margin = margin(0,0,0,0),
   # adjust text size
   text = element_text(size = 8),
-  title = element_text(size = 8)
+  title = element_text(size = 8),
   # axis.title = element_text(size = 8),
   # axis.text = element_text(size = 7),
-  # legend.text = element_text(size = 7)
+  legend.title = element_text(size = 7)
 )
 
 # plot chain means
 theta <- ggplot(diagnostics) +
   geom_point(aes(x = iteration, y = `Chain 1`, color = patho),
-             size = .75,
+             size = .5,
              na.rm = TRUE) +
   geom_point(aes(x = iteration, y = `Chain 2`, color = patho),
-             size = .75,
+             size = .5,
              na.rm = TRUE) +
   geom_point(aes(x = iteration, y = `Chain 3`, color = patho),
-             size = .75,
+             size = .5,
              na.rm = TRUE) +
   geom_point(aes(x = iteration, y = `Chain 4`, color = patho),
-             size = .75,
+             size = .5,
              na.rm = TRUE) +
   geom_point(aes(x = iteration, y = `Chain 5`, color = patho),
-             size = .75,
+             size = .5,
              na.rm = TRUE) +
-  geom_line(aes(x = iteration, y = `Chain 1`, color = patho), na.rm = TRUE) +
-  geom_line(aes(x = iteration, y = `Chain 2`, color = patho), na.rm = TRUE) +
-  geom_line(aes(x = iteration, y = `Chain 3`, color = patho), na.rm = TRUE) +
[…]

### 2020-05-01 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
Add critical value to plot for AC

@@
 
-```{r diagnostics, fig.cap="Convergence diagnostics applied on the imputation algorithms of Figure 2."}
+```{r diagnostics, fig.cap="Convergence diagnostics applied on the imputation algorithms of Figure 2. $\\theta$ = chain mean in $y_{imp, m}$."}
 load("./images/diagnostics_plot.Rdata") #bquote(theta ~ "= chain mean in "~ y[imp][",m"])
@@
 
+We want to know what the effect is of inducing non-conv of the bias in all Qs and the confidence validity of the betas.
+So we plot the bias for each Q against the number of iterations in the condition and the missingness proportion in the condition.
+We look at averages across simulation repetitions.
+Then, we want to know if conditions in which the Qs are affected by non-convergence (on average) are also identified by the conv diag.
+We do this by looking at the bias in Q against the
+
 **Add to this section: 1) emphasize that the plots are averages across repetitions, not within MICE; 2) sample effects due to single complete dataset.**
@@
 ```{r mean, fig.height = 6.8, fig.cap="Convergence diagnostics chain mean."}
-mean_Rh + mean_AC + mean_bias + plot_layout(guides = "collect", ncol = 1) + plot_annotation(tag_levels = "A", tag_suffix = ".", title = "Performance of convergence diagnostics", subtitle = bquote(theta ~ "= chain mean of Y in" ~ y[imp][",m"] ~ "; Q = " ~ mu[Y] ~ "= 25.81"))
+mean_bias + mean_Rh + mean_AC + plot_layout(guides = "collect", ncol = 1) + plot_annotation(tag_levels = "A", tag_suffix = ".", title = bquote("Q = " ~ mu[Y] ~ "= 25.81; " ~ theta ~ "= chain mean of Y in" ~ y[imp][",m"])) #& theme(legend.position = "bottom")
 ```
@@
 ```{r sd, fig.height = 6.8, fig.cap="Convergence diagnostics chain variance."}
-var_Rh + var_AC + sd_bias + plot_layout(guides = "collect", ncol = 1) + plot_annotation(tag_levels = "A", tag_suffix = ".", title = "Performance of convergence diagnostics", subtitle = bquote(theta ~ "= chain variance of Y in" ~ y[imp][",m"] ~ "; Q = " ~ sigma[Y] ~ "= 11.32"))
+sd_bias + var_Rh + var_AC + plot_layout(guides = "collect", ncol = 1) + plot_annotation(tag_levels = "A", tag_suffix = ".", title = bquote("Q = " ~ sigma[Y] ~ "= 11.32; " ~ theta ~ "= chain variance of Y in" ~ y[imp][",m"]))
 ```
+
 ```{r est, fig.height = 6.8, fig.cap="Convergence diagnostics regression coefficient."}
-est_Rh + est_AC + est_bias + plot_layout(guides = "collect", ncol = 1) + plot_annotation(tag_levels = "A", tag_suffix = ".", title = "Performance of convergence diagnostics", subtitle = bquote(theta ~ "= " ~ hat(beta)[2] ~ "in {" ~ y[obs] ~ "," ~ y[imp][",m"] ~ "}" ~ "; Q = " ~ beta[2] ~ "= 2.06"))
+est_bias + est_Rh + est_AC + plot_layout(guides = "collect", ncol = 1) + plot_annotation(tag_levels = "A", tag_suffix = ".", title = bquote("Q = " ~ beta[2] ~ "= 2.06; " ~ theta ~ "= " ~ hat(beta)[2] ~ "in {" ~ y[obs] ~ "," ~ y[imp][",m"] ~ "}"))
 ```
@@
 ```{r pred, fig.height = 6.8, fig.cap="Convergence diagnostics regression coefficient."}
-PCA_Rh + PCA_AC + Rsq_bias + plot_layout(guides = "collect", ncol = 1) + plot_annotation(tag_levels = "A", tag_suffix = ".", title = "Performance of convergence diagnostics", subtitle = bquote(theta ~ "= first eigenvalue of" ~ Sigma ~ "in {" ~ y[obs] ~ "," ~ y[imp][",m"] ~ "}" ~ "; Q = explained variance = 19.25"))
-
+Rsq_bias + PCA_Rh + PCA_AC + plot_layout(guides = "collect", ncol = 1) + plot_annotation(tag_levels = "A", tag_suffix = ".", title = bquote("Q = " ~ r^2 ~ "\u00D7 100 = 19.25; " ~ theta ~ "= " ~ lambda[1] ~ "in {" ~ y[obs] ~ "," ~ y[imp][",m"] ~ "}")) #first eigenvalue of" ~ Sigma ~
 ```

### 2020-05-01 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Add critical value to plot for AC

@@
 
-\caption{Convergence diagnostics applied on the imputation algorithms of Figure 2.}\label{fig:diagnostics}
+\caption{Convergence diagnostics applied on the imputation algorithms of Figure 2. $\theta$ = chain mean in $y_{imp, m}$.}\label{fig:diagnostics}
 \end{figure}
@@
 
+We want to know what the effect is of inducing non-conv of the bias in
+all Qs and the confidence validity of the betas. So we plot the bias for
+each Q against the number of iterations in the condition and the
+missingness proportion in the condition. We look at averages across
+simulation repetitions. Then, we want to know if conditions in which the
+Qs are affected by non-convergence (on average) are also identified by
+the conv diag. We do this by looking at the bias in Q against the
+
 \textbf{Add to this section: 1) emphasize that the plots are averages
@@
 \end{figure}
+
 \begin{figure}

### 2020-05-01 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
New results section and removed bullshit from the discussion

@@
 # packages
+library(dplyr)
 library(ggplot2)
@@
 load("../Results/complete.Rdata") 
+results <- results %>% filter(t<50)
 source("../2.CreateFigures.R")
@@
 
+Upon convergence, $\widehat{R}=1$ and $AC=0$, which are unlikely thresholds for MCMC algorithms, because of its convergence to a distribution.
+In practice, non-convergence is usually diagnosed when $\widehat{R}$ > 1.2 or 1.1 or even 1.01.
+And a t-test is performed to assess whether $AC$ is significantly different from zero.
+
+We assume that complete convergence, defined as $\widehat{R} = 1$ and $AC = 0$, will not be observed.
+Because even in the most converged state, the algorithm will show some signs of non-mixing and non-stationarity.
+Upon *sufficient* convergence, the imputation chains will intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm ($\widehat{R} > 1$), and there may be some dependency between subsequent iterations of imputation chains ($AC > 0$).
+
 **Before using them in the simulation, the two methods show at least be able to distinguish between the two algorithms plotted in Figure 2.
@@
 **acf is niet bedoeld om weinig iteraties te... niet een fout in acf, alleen hier niet van toepassing**
-
-We assume that complete convergence, defined as $\widehat{R} = 1$ and $AC = 0$, will not be observed.
-Because even in the most converged state, the algorithm will show some signs of non-mixing and non-stationarity.
-Upon *sufficient* convergence, the imputation chains will intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm ($\widehat{R} > 1$), and there may be some dependency between subsequent iterations of imputation chains ($AC > 0$).
 
@@
 Yet, scientific estimates were insensitive to continued iteration after $t>5$ [@lace07].
-We therefore suspect that $\widehat{R}$ may over-estimate signs of non-convergence in iterative imputation algorithms.
-In contrast to this, it may occur that the signs of non-convergence are under-estimated by $\widehat{R}$, in exceptional cases where the initial values of the algorithm are not appropriately over-dispersed [@broo98, p.
+We therefore suspect that $\widehat{R}$ may over-estimate signs of non-convergence in iterative imputation algorithms, compared to the bias the estimates.
+In contrast to this, it may occur that signs of non-convergence are under-estimated by $\widehat{R}$, in exceptional cases where the initial values of the algorithm are not appropriately over-dispersed [@broo98, p.
 437].
@@
 We expect that high $AC$ values are implausible in iterative imputation algorithms with typical convergence.
-That is, after only a few iterations, the randomness induced by the algorithm will effectively mitigate the risk of dependency within chains.
+That is, after only a few iterations, the randomness induced by the algorithm should effectively mitigate the risk of dependency within chains.
 
@@
 We want to know what the effect is of inducing non-conv of the bias in all Qs and the confidence validity of the betas.
-So we plot the bias for each Q against the number of iterations in the condition and the missingness proportion in the condition.
+So we plot the bias for each Q against the number of iterations and the missingness proportion in the condition.
 We look at averages across simulation repetitions.
+Is there more bias in conditions where non-conv was induced?
+(NOPE for univariates, YES for multivar.).
+
 Then, we want to know if conditions in which the Qs are affected by non-convergence (on average) are also identified by the conv diag.
-We do this by looking at the bias in Q against the
+We do this by looking at the bias in Q against the conv diag applied to related thetas.
+Is the point at which there is no more bias the same point at which the conv diag indicate minimal signs of non-conv?
+(NOPE, they are too conservative).
 
-**Add to this section: 1) emphasize that the plots are averages across repetitions, not within MICE; 2) sample effects due to single complete dataset.**
+Is one of the thetas better? (NAH, doesn't really matter).
+
+What should the thresholds be? (LOWER than recommended) And the default nr of iterarions? (MAYBE increase to 10?)
+
+**Add to this section: 1) emphasize that the plots are averages across repetitions, not within MICE; 2) sample effects due to single complete dataset.
+3) Add more info about figure legends and axes.)**
 
@@
 For the descriptve statistics, the magnitude of the bias was the largest in $Y$ (i.e, $j = Y$ in $\mu_j$ and $\sigma_j$).
-For the $Q=\beta_j$, $j=2$ showed the worst perfomance (i.e, the effect of $X_1$ on $Y$.
+For the $Q=\beta_{1,2,3}$, $\beta_1$ showed the worst perfomance (i.e, the effect of $X_1$ on $Y$.
+Moreover, we only show conditions where $T\leq50$ in the figures.
+Full results are available from **[link github repo]**.
+
+## In general
 
@@
 
-Evaluation with the performance measures shows that $\widehat{R}$ and autocorrelation are conservative: they indicate signs of non-convergence in conditions where $\bar{Q}$s are unbiased and or confidence valid estimates of $Q$.
+Evaluation with the performance measures shows that $\widehat{R}$ is conservative: it identifies signs of non-convergence in conditions where the $\bar{Q}$s are unbiased and/or confidence valid estimates of the $Q$s.
 
-## Old Results section
+## Means
 
-**(Add more info about figure legends and axes.)**
+**Message: after one iteration (T=2), the algorithm is as well converged as the missingness proportion allows.
+Sooner even with lower missingness proportions (T=1 might be ok for p_mis=5 and 25).** Number of it has little influence on bias in hat{Q}, only in first iteration some bias in missingness conditions of 25 and up.
+There is a clear different between the missingness conditions.
+The higher the missingness prop, the larger the magnitude of the bias.
+For this Q, t=2 would be sufficient.
+And only for p_mis = 5 the bar{Q}s are unbiased.
+All other p_mis conditions have bias irrespective of the number of iterations.
+
+Rhat differs a lot across iteration conditions.
+But not structurally for missingness proportions.
+Highest Rhats for the condition t=2, while the most bias is for (t=1), but there is no Rhat for that by definition.
+But for missingness, 25% seems the most signs of non-conv, while it has one of the least bias.
+Threshold of 1.2 reached in condition it = 3, p_mis = 5 and 75, and then for it>6, p_mis= 5 and 50, and it>10 for all p_mis.
+Threshold 1.1 reached in in it=12, p_mis = all but 75, 1.01 only reached in it = 93, p_mis = 75.
+While for this Q, t=2 would be sufficient.
+
+AC mainly shows effects of the set of iteration conditions.
+AC indicates less signs of non-conv in conditions with higher number of iterations.
+For iteration conditions where t<10, there are also some effects of the missingness proportions.
+The p_mis condtions 25 and 95 % indicate the most signs of non-convergence.
+For p_mis = 95% this corresponds to the bias in Qbar.
+For 25% miss it doesn't.
+The thresholds defined by the critical value is only met in the conditions it=3, p_mis = 5,75, and 95%.
 
@@
 
+## Standard deviations
+
+**Message: even with possible effect of initial values (T=1), the algorithm is as well converged as the missingness proportion allows.** For $Q=\sigma_Y$ there is no effect visible of the iteration conditions, but clearly for the p_mis conditions.
+Most bias for p_mis = 95, followed by p_mis=50 and 75.
+For the least possible bias we may use any number of iterations $T\geq1$, but none of the p_mis results in unbiased bar{Q}s.
+
+There is a clear effect of the number of iterations on Rhat.
+Conditions with lower Ts have more signs of non-conv, as identified by Rhat.
+There is little difference between missingness conditions.
+Only for conditions where $T<5$ the missingness proportions seem to affect whether Rhat indicates non-conv.
+The bias in bar{Q} seems unrelated to the magnitude of Rhat.
+Conditions with more bias (p_mis = 50,75,95) are not identified to have more signs of non-convergence.
+The thresholds are therefore useless (because depend mostly on t, not so much p_mis).
+
+AC is affected by both t and p_mis conditions.
+But conditions with worst bias (p_mis = 50, 75, 95) are not identified as worst converging according to AC.
+Moreover, only for T=3 the AC is significantly different from zero.
+
 ```{r sd, fig.height = 6.8, fig.cap="Convergence diagnostics chain variance."}
@@
 
+## Regression coefficients
+
+**Message: after four iterations (T=5), the algorithm is as well converged as the missingness proportion allows.
+Sooner even with lower missingness proportions (T=1 might be ok for p_mis=5 and 25).** The figure for $Q=\beta_1$ shows that the $\bar{Q}$s are affected by both the number of iterations and the missingness proportion.
+For T=1, the magnitude of the bias (almost) follows the order of the missingness proportions.
+More missingness causes more bias in the $\bar{Q}$.
+In general, a lower nr of iterations results in more bias, but this depends on the missingness proportion.
+For p_mis=95, there is more bias in conditions where T<6.
+When p_mis = 50, it's T<3.
+For p_mis = 75 and 25, it's only for T<2.
+And conditions where p_mis = 5 are unaffected on average.
+However, none of the conditions result in completely unbiased estimates of $Q$.
+
+We should see this in the conv diag: higher Rhat and AC for p_mis=75 and 95, and for condtions where $T<2-6$, depending on p_mis.
+
+Rhat depends somewhat on the missingness proportion.
+For conditions where T<10, the Rhat values follow the p_mis more or less.
+But what we actually want to know is whether they are indicators of bias.
+Conditions with the worst bias (high p_mis, low T) have the largest Rhat values.
+The thresholds are not appropriate to diagnose sufficient convergence.
+All thresholds over-estimate the signs of non-convergence compared to the bias in Qbar.
+However, a higher threshold than 1.2 may result in terminating the algorithm after 3 iterations, which is insufficient for the p_mis = 95 condition.
+Threshold of 1.25 or 1.3 could work for the other p_mis conditions.
+
+AC is very affected by the p_mis.
+The two conditions with the most bias (p_mis = 75 and 95) are the only ones identified as non-converging, as they should.
+However, the p_mis = 75 condition is only identified in the T=3 condtion, while the bias is the least pronounced there.
+And the AC only identifies non-convergence in the p_mis = 95 conditions where T>10.
+In the default situation, T=5, this would be overlooked.
+Also, the AC shows the opposite of what you would expect from a theta that is based on the observed data: shouldn't there be higher autocorrelations in conditions with comparatively more observed data in the completed datasets?
+I don't understand this.
+
+**Add how interesting it is because we literally track Q and are still not ok.**
+
 ```{r est, fig.height = 6.8, fig.cap="Convergence diagnostics regression coefficient."}
-est_bias + est_Rh + est_AC + plot_layout(guides = "collect", ncol = 1) + plot_annotation(tag_levels = "A", tag_suffix = ".", title = bquote("Q = " ~ beta[2] ~ "= 2.06; " ~ theta ~ "= " ~ hat(beta)[2] ~ "in {" ~ y[obs] ~ "," ~ y[imp][",m"] ~ "}"))
+est_bias + est_Rh + est_AC + plot_layout(guides = "collect", ncol = 1) + plot_annotation(tag_levels = "A", tag_suffix = ".", title = bquote("Q = " ~ beta[1] ~ "= 2.06; " ~ theta ~ "= " ~ hat(Q) ~ "in {" ~ y[obs] ~ "," ~ y[imp][",m"] ~ "}"))
 ```
+
+## Variance explained
+
+**Message: after two iterations (T=3), the algorithm is as well converged as the missingness proportion allows.
+Sooner even with lower missingness proportions (T=1 might be ok for p_mis=5 and 25).** Bias is the worst for conditions where p_mis = 95, T<3.
+Followed by p_mis = 75, T=1; and p_mis = 50, T=1.
+Conditions where p_mis = 25 or 5 seem unaffected by the number of iterations.
+Minimal bias is obtained in all iteration conditions.
+We should see this in Rhat and AC: worst convergence in p_mis = 95 and 50; some non-conv in conditions where p_mis = 75 and T=1, and no non-conv in conditions where p_mis = 25 or 5.
+The sign of the bias is logical: when there is less info in the data, the relations between the predictors and the outcome is under-estimated.
+
+Rhat behaves oppositely to what we wanted.
+The lowest Rhat values for conditions where T<10 are observed for the two missingness conditions with the worst bias.
+However, the highest Rhat observed is indeed for the missingness condition with the worst bias, p_mis =95.
+But only at T=5 (while bias is no worse than anywhere else in this p_mis condition, conditional on iterations condition T>2).
+So neither of the thresholds works perfectly.
+
+AC shows a similar trend to the biases: more signs of non-conv for higher p_mis and lower T.
+The threshold, however, performs less well.
+It identifies signs of non-convergence for all conditions where $T=3$, irrespective of p_mis.
+Then nothing, and then for p_mis =95, T>7, and p_mis=50, T >14, and p_mis = 75, T>20.
+This does not corresond to the bias in $\bar{Q}$.
+Maybe use AC>.5 at T=5 as threshold?
 
@@
 
-## Univariate estimates and convergence diagnostics
+<!-- ## Univariate estimates and convergence diagnostics  -->
 
-The bias in the estimates of the variable means shows little to no difference between simulation conditions.
+<!-- The bias in the estimates of the variable means shows little to no difference between simulation conditions.
 It doesn't seem to matter how many iterations you use in the mice algorithm, the estimates are unbiased.
@@
 These univariate quantities appear to be unaffected by the number of iterations.
+-->
 
-When applied to the imputation chain means, $\widehat{R}$ indicates that the mice algorithm does not reach a converged state ($\widehat{R}$ = 1) in any of the simulation conditions.
+<!-- When applied to the imputation chain means, $\widehat{R}$ indicates that the mice algorithm does not reach a converged state ($\widehat{R}$ = 1) in any of the simulation conditions.
 Neither is the most recent recommended threshold reached ($\widehat{R}$ < 1.01).
@@
 (With the default number of iterations (maxit = 5), this dip in $\widehat{R}$ values would be spotted, so it is no problem.) The point at which an extra iteration does not seem to improve the $\widehat{R}$ value is around $T=30$.
+-->
 
-Autocorrelations indicate no sign of trending **(NOT correct anymore, fluke of the stats::acf() function!!)** within imputation chains.
+<!-- Autocorrelations indicate no sign of trending **(NOT correct anymore, fluke of the stats::acf() function!!)** within imputation chains.
 In most simulation conditions $AC$ is smaller than or about equal to zero.
@@
 This is a small positive autocorrelation, which would indicate some trending within chains.
+-->
 
-The $\widehat{R}$ and $AC$ values for the imputation chain variances show equal trends to the chain means and are therefore not discussed separately.
+<!-- The $\widehat{R}$ and $AC$ values for the imputation chain variances show equal trends to the chain means and are therefore not discussed separately.
 Taken together, univariate estimates seem robust across simulation conditions.
 There is no clear effect of the number of iterations on the bias in these estimates, while the convergence diagnostics indicate that the algorithm did not reach a completely converged state (yet).
+-->
 
-## Multivariate estimates
+<!-- ## Multivariate estimates  -->
 
-**This is only true for 75% missingness:** There is a clear bias in the regression estimates in simulation conditions where the number of iterations is smaller than four.
+<!-- **This is only true for 75% missingness:** There is a clear bias in the regression estimates in simulation conditions where the number of iterations is smaller than four.
 In simulation conditions where $T > 5$ there is little to no bias in the estimated regression coefficients.
@@
 Since confidence interval width is stable across conditions, the under-coverage may be attributed to the bias in the estimated regression coefficients.
+-->
 
-**This is only true for 75% missingness:** If we look at the estimated proportion of explained variance in outcome variable Y we see that the coefficient of determination is underestimate estimated in conditions where the number of iterations is equal to two or less, and slightly overestimated in conditions where the number of iterations is equal to three or more.
-
-In short, we see that the minimum number of iterations required to obtain unbiased, confidence valid regression estimates is 5.
-This value, however, is dependent on the percentage of missing values.
-E.g., with 95% of cases having missing data we need at least seven iterations to obtain unbiased results.
+<!-- **This is only true for 75% missingness:** If we look at the estimated proportion of explained variance in outcome variable Y we see that the coefficient of determination is underestimate estimated in conditions where the number of iterations is equal to two or less, and slightly overestimated in conditions where the number of iterations is equal to three or more.
+-->
 
@@
 
-*Proper performance under a 'missing completely at random' missingness mechanism is necessary to demonstrate the appropriateness of $\widehat{R}$ and $AC$ as non-convergence diagnostics.
-However, results may not be extrapolated to other missingness mechanisms.*
+In short, univariate estimates seem robust against terminating the algorithm early: There is no clear effect of the number of iterations on the bias in these estimates.
+The bias in the Qs only depends on the p_mis.
+Yet, the convergence diagnostics indicate that the algorithm did not reach a completely converged state (yet).
 
-Upon convergence, $\widehat{R}=1$ and $AC=0$, which are unlikely thresholds for MCMC algorithms, because of its convergence to a distribution.
-In practice, non-convergence is usually diagnosed when $\widehat{R}$ > 1.2 or 1.1 or even 1.01.
-And a t-test is performed to assess whether $AC$ is significantly different from zero.
+$\widehat{R}$ and autocorrelation indicate that algorithmic convergence may only be reached after twenty or even forty iterations, while unbiased, confidence valid estimates may be obtained with as little as one iteration.
+These results are in agreement with the simulation hypothesis: $\widehat{R}$ over-estimates the severity of non-convergence when applied to MI procedures.
 
-As expected, complete convergence ($\widehat{R}=1$ and $AC=0$) is never observed.
+With that, we show that MICE can lead to correct outcomes when they have not converged according to two common conv diags.
+This may be due to the methods (and their thresholds) or due to the Qs (descriptives and multivariate linear regression, not higher dimensional/more complex Qs).
+More `complicated' $Q$s (e.g., higher-order effects or variance components) might show bias, under- or over-coverage at higher $T$, as indiated by Rhat and AC.
+Add what %miss has to do with it.
+
+<!-- As expected, complete convergence ($\widehat{R}=1$ and $AC=0$) is never observed.
 But what about the thresholds for practice?
@@
 And it is not obtained for any number of iterations (**but may be necessary in more complex complete data models???**).
+-->
 
-This note shows that convergence diagnostics $\widehat{R}$ and $AC$ may diagnose convergence of multiple imputation algorithms, but their performance differs from conventional applications to iterative algorithmic procedures.
-(**nope!
-it shows that MICE can lead to correct outcomes when they have not converged according to two common conv diags.
-This may be due to the measures (e.g., assumption of overdisp) or due to the Qs (lm reg coeff, not higher dimensional/more complex RQs).
-Add what %miss has to do with it.**)
-
-$\widehat{R}$ and autocorrelation indicate that algorithmic convergence may only be reached after twenty or even forty iterations, while unbiased, confidence valid estimates may be obtained with as little as four iterations.
-These results are in agreement with the simulation hypothesis: $\widehat{R}$ over-estimates the severity of non-convergence when applied to MI procedures.
-%This may be due to the quantity of scientific interest chosen.
-More `complicated' $Q$s (e.g., higher-order effects or variance components) might show bias, under- or over-coverage at higher $T$.
-
-According to this simulation study, the recently proposed threshold of $\widehat{R}<1.01$ may be too stringent for MI algorithms.
-(**This is only one of the goals: to give applied researchers a diagnostic to indicate that they should keep iterating.
-The other is the default in mice and other software packages, and yet another is...
-i forgot**) Under the relatively easy missing data problem of the current study, the threshold was not reached.
+<!-- According to this simulation study, the recently proposed threshold of $\widehat{R}<1.01$ may be too stringent for MI algorithms.
+Under the relatively easy missing data problem of the current study, the threshold was not reached.
 The other extreme of the $\widehat{R}$-thresholds, the conventionally acceptable $\widehat{R} <1.2$, may be too lenient for MI procedures.
@@
 The current study suggests that $\widehat{R} < 1.05$ may be implemented, since that is the level at which the $\widehat{R}$ stabilize (around $T = 20$) (**Not necessary for this Q, but maybe for more complicated Qs**).
+-->
 
-The negative $AC$-values obtained in this study show no threat of non-stationarity.
+<!-- The negative $AC$-values obtained in this study show no threat of non-stationarity.
 However, the initial dip in $AC$-values **(which disappeared!)** may have implications for the default number of iterations in `mice` (`maxit = 5`).
@@
 Under the current specifications, $T>20$ would be more appropriate.
-
-Further research is needed to investigate their performance under clear violation of convergence, e.g.
-dependency between predictors (predictors with very high correlations).
-Until then, we have only shown that the convergence diagnostics can diagnose non-convergence of MI algorithms that trend towards a converged state.
-Also for future research, look at developing a convergence diagnostic for substantive models, and implement a Wald test for $AC = 0$.
+-->
 
@@
 
-For empirical researchers: 1) Check trace plots for pathological non convergence and adjust imputation model if necessary.
-2) use $\widehat{R}$ wait 1.1 ash threshold and autocorrelation with **?** as threshold.
-Keep iterating until these thresholds are reached.
-3) Do not use the R function ACF.
-Instead, compute autocorrelations manually **(see e.g., [GitHub link])**.
-4) Track your own scalar summary of interest.
-This is somewhat advanced but explained in Van Buren 2018.
-Compute $\widehat{R}$ and autocorrelation values for this scalar summary.
-5) **Something** about the novel $\theta$ that is 'substantive model-independent'.
+For empirical researchers:
+
+1. Check trace plots for pathological non convergence and adjust imputation model if necessary.
+
+2. use $\widehat{R}$ wait 1.1 ash threshold and autocorrelation with **?** as threshold. Keep iterating until these thresholds are reached.
+
+3. Do not use the R function ACF. Instead, compute autocorrelations manually **(see e.g., [GitHub link])**.
+
+4. Track your own scalar summary of interest. This is somewhat advanced but explained in Van Buren 2018. Compute $\widehat{R}$ and autocorrelation values for this scalar summary.
+
+5. **Something** about the novel $\theta$ that is 'substantive model-independent'.
 
@@
 
-**Pmm, M(N)AR, Empirical data, Vignette, ShinyMICE, AC across imputations, not iterations (new convergence diagnostic unique for iterative imputation??).**
+Further research is needed to investigate their performance under clear violation of convergence, e.g. dependency between predictors (predictors with very high correlations). Also:
 
-**If I don't include it in my study, also: Significance of $AC$ values.**
+- Induce non-convergence with mis-specified imputation model.
+
+- More difficult missingness problems, i.e., M(N)AR instead of MCAR.
+Proper performance under a 'missing completely at random' missingness mechanism is necessary to demonstrate the appropriateness of $\widehat{R}$ and $AC$ as non-convergence diagnostics.
+However, results may not be extrapolated to other missingness mechanisms.
+
+- Imputation technique with questionable convergence properties, e.g., PMM.
+
+- Apply recommendations to empirical data and write a vignette for applied researchers.
+
+- Implement recommendations in ShinyMICE.
+
+- Develop an additional convergence diagnostic unique for iterative imputation: AC across imputations, instead of iterations.
 
@@
 
-Scientific contribution
+- Scientific contribution 
 (clear research question, convincing argumentation that the study adds to the scientific literature)
 
-Theory
+- Theory
 (use of theory, clear hypotheses, relevant literature)
 
-Method
+- Method
 (correct data collection/handling, measures, choice and description of method of analysis)
 
-Results
+- Results
 (correct interpretation of findings, clear presentation)
-Discussion and conclusions
+
+- Discussion and conclusions
 (summary, broader implications, limitations, future research)
 
-Written presentation
+- Written presentation
 (writing, structure, tables and figures, references)

### 2020-05-01 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
New results section and removed bullshit from the discussion

@@
 
+Upon convergence, \(\widehat{R}=1\) and \(AC=0\), which are unlikely
+thresholds for MCMC algorithms, because of its convergence to a
+distribution. In practice, non-convergence is usually diagnosed when
+\(\widehat{R}\) \textgreater{} 1.2 or 1.1 or even 1.01. And a t-test is
+performed to assess whether \(AC\) is significantly different from zero.
+
+We assume that complete convergence, defined as \(\widehat{R} = 1\) and
+\(AC = 0\), will not be observed. Because even in the most converged
+state, the algorithm will show some signs of non-mixing and
+non-stationarity. Upon \emph{sufficient} convergence, the imputation
+chains will intermingle such that the only difference between the chains
+is caused by the randomness induced by the algorithm
+(\(\widehat{R} > 1\)), and there may be some dependency between
+subsequent iterations of imputation chains (\(AC > 0\)).
+
 \textbf{Before using them in the simulation, the two methods show at
@@
 in acf, alleen hier niet van toepassing}
-
-We assume that complete convergence, defined as \(\widehat{R} = 1\) and
-\(AC = 0\), will not be observed. Because even in the most converged
-state, the algorithm will show some signs of non-mixing and
-non-stationarity. Upon \emph{sufficient} convergence, the imputation
-chains will intermingle such that the only difference between the chains
-is caused by the randomness induced by the algorithm
-(\(\widehat{R} > 1\)), and there may be some dependency between
-subsequent iterations of imputation chains (\(AC > 0\)).
 
@@
 all Qs and the confidence validity of the betas. So we plot the bias for
-each Q against the number of iterations in the condition and the
-missingness proportion in the condition. We look at averages across
-simulation repetitions. Then, we want to know if conditions in which the
-Qs are affected by non-convergence (on average) are also identified by
-the conv diag. We do this by looking at the bias in Q against the
+each Q against the number of iterations and the missingness proportion
+in the condition. We look at averages across simulation repetitions. Is
+there more bias in conditions where non-conv was induced? (NOPE for
+univariates, YES for multivar.).
+
+Then, we want to know if conditions in which the Qs are affected by
+non-convergence (on average) are also identified by the conv diag. We do
+this by looking at the bias in Q against the conv diag applied to
+related thetas. Is the point at which there is no more bias the same
+point at which the conv diag indicate minimal signs of non-conv? (NOPE,
+they are too conservative).
+
+Is one of the thetas better? (NAH, doesn't really matter).
+
+What should the thresholds be? (LOWER than recommended) And the default
+nr of iterarions? (MAYBE increase to 10?)
 
@@
 across repetitions, not within MICE; 2) sample effects due to single
-complete dataset.}
+complete dataset. 3) Add more info about figure legends and axes.)}
 
@@
 \(Y\) (i.e, \(j = Y\) in \(\mu_j\) and \(\sigma_j\)). For the
-\(Q=\beta_j\), \(j=2\) showed the worst perfomance (i.e, the effect of
-\(X_1\) on \(Y\).
+\(Q=\beta_{1,2,3}\), \(\beta_1\) showed the worst perfomance (i.e, the
+effect of \(X_1\) on \(Y\). Moreover, we only show conditions where
+\(T\leq50\) in the figures. Full results are available from
+\textbf{{[}link github repo{]}}.
+
+\hypertarget{in-general}{%
+\subsection{In general}\label{in-general}}
 
@@
 
-Evaluation with the performance measures shows that \(\widehat{R}\) and
-autocorrelation are conservative: they indicate signs of non-convergence
-in conditions where \(\bar{Q}\)s are unbiased and or confidence valid
-estimates of \(Q\).
+Evaluation with the performance measures shows that \(\widehat{R}\) is
+conservative: it identifies signs of non-convergence in conditions where
+the \(\bar{Q}\)s are unbiased and/or confidence valid estimates of the
+\(Q\)s.
 
-\hypertarget{old-results-section}{%
-\subsection{Old Results section}\label{old-results-section}}
+\hypertarget{means}{%
+\subsection{Means}\label{means}}
 
-\textbf{(Add more info about figure legends and axes.)}
+\textbf{Message: after one iteration (T=2), the algorithm is as well
+converged as the missingness proportion allows. Sooner even with lower
+missingness proportions (T=1 might be ok for p\_mis=5 and 25).} Number
+of it has little influence on bias in hat\{Q\}, only in first iteration
+some bias in missingness conditions of 25 and up. There is a clear
+different between the missingness conditions. The higher the missingness
+prop, the larger the magnitude of the bias. For this Q, t=2 would be
+sufficient. And only for p\_mis = 5 the bar\{Q\}s are unbiased. All
+other p\_mis conditions have bias irrespective of the number of
+iterations.
+
+Rhat differs a lot across iteration conditions. But not structurally for
+missingness proportions. Highest Rhats for the condition t=2, while the
+most bias is for (t=1), but there is no Rhat for that by definition. But
+for missingness, 25\% seems the most signs of non-conv, while it has one
+of the least bias. Threshold of 1.2 reached in condition it = 3, p\_mis
+= 5 and 75, and then for it\textgreater6, p\_mis= 5 and 50, and
+it\textgreater10 for all p\_mis. Threshold 1.1 reached in in it=12,
+p\_mis = all but 75, 1.01 only reached in it = 93, p\_mis = 75. While
+for this Q, t=2 would be sufficient.
+
+AC mainly shows effects of the set of iteration conditions. AC indicates
+less signs of non-conv in conditions with higher number of iterations.
+For iteration conditions where t\textless10, there are also some effects
+of the missingness proportions. The p\_mis condtions 25 and 95 \%
+indicate the most signs of non-convergence. For p\_mis = 95\% this
+corresponds to the bias in Qbar. For 25\% miss it doesn't. The
+thresholds defined by the critical value is only met in the conditions
+it=3, p\_mis = 5,75, and 95\%.
 
@@
 
+\hypertarget{standard-deviations}{%
+\subsection{Standard deviations}\label{standard-deviations}}
+
+\textbf{Message: even with possible effect of initial values (T=1), the
+algorithm is as well converged as the missingness proportion allows.}
+For \(Q=\sigma_Y\) there is no effect visible of the iteration
+conditions, but clearly for the p\_mis conditions. Most bias for p\_mis
+= 95, followed by p\_mis=50 and 75. For the least possible bias we may
+use any number of iterations \(T\geq1\), but none of the p\_mis results
+in unbiased bar\{Q\}s.
+
+There is a clear effect of the number of iterations on Rhat. Conditions
+with lower Ts have more signs of non-conv, as identified by Rhat. There
+is little difference between missingness conditions. Only for conditions
+where \(T<5\) the missingness proportions seem to affect whether Rhat
+indicates non-conv. The bias in bar\{Q\} seems unrelated to the
+magnitude of Rhat. Conditions with more bias (p\_mis = 50,75,95) are not
+identified to have more signs of non-convergence. The thresholds are
+therefore useless (because depend mostly on t, not so much p\_mis).
+
+AC is affected by both t and p\_mis conditions. But conditions with
+worst bias (p\_mis = 50, 75, 95) are not identified as worst converging
+according to AC. Moreover, only for T=3 the AC is significantly
+different from zero.
+
 \begin{figure}
@@
 \end{figure}
+
+\hypertarget{regression-coefficients}{%
+\subsection{Regression coefficients}\label{regression-coefficients}}
+
+\textbf{Message: after four iterations (T=5), the algorithm is as well
+converged as the missingness proportion allows. Sooner even with lower
+missingness proportions (T=1 might be ok for p\_mis=5 and 25).} The
+figure for \(Q=\beta_1\) shows that the \(\bar{Q}\)s are affected by
+both the number of iterations and the missingness proportion. For T=1,
+the magnitude of the bias (almost) follows the order of the missingness
+proportions. More missingness causes more bias in the \(\bar{Q}\). In
+general, a lower nr of iterations results in more bias, but this depends
+on the missingness proportion. For p\_mis=95, there is more bias in
+conditions where T\textless6. When p\_mis = 50, it's T\textless3. For
+p\_mis = 75 and 25, it's only for T\textless2. And conditions where
+p\_mis = 5 are unaffected on average. However, none of the conditions
+result in completely unbiased estimates of \(Q\).
+
+We should see this in the conv diag: higher Rhat and AC for p\_mis=75
+and 95, and for condtions where \(T<2-6\), depending on p\_mis.
+
+Rhat depends somewhat on the missingness proportion. For conditions
+where T\textless10, the Rhat values follow the p\_mis more or less. But
+what we actually want to know is whether they are indicators of bias.
+Conditions with the worst bias (high p\_mis, low T) have the largest
+Rhat values. The thresholds are not appropriate to diagnose sufficient
+convergence. All thresholds over-estimate the signs of non-convergence
+compared to the bias in Qbar. However, a higher threshold than 1.2 may
+result in terminating the algorithm after 3 iterations, which is
+insufficient for the p\_mis = 95 condition. Threshold of 1.25 or 1.3
+could work for the other p\_mis conditions.
+
+AC is very affected by the p\_mis. The two conditions with the most bias
+(p\_mis = 75 and 95) are the only ones identified as non-converging, as
+they should. However, the p\_mis = 75 condition is only identified in
+the T=3 condtion, while the bias is the least pronounced there. And the
+AC only identifies non-convergence in the p\_mis = 95 conditions where
+T\textgreater10. In the default situation, T=5, this would be
+overlooked. Also, the AC shows the opposite of what you would expect
+from a theta that is based on the observed data: shouldn't there be
+higher autocorrelations in conditions with comparatively more observed
+data in the completed datasets? I don't understand this.
+
+\textbf{Add how interesting it is because we literally track Q and are
+still not ok.}
 
@@
 
+\hypertarget{variance-explained}{%
+\subsection{Variance explained}\label{variance-explained}}
+
+\textbf{Message: after two iterations (T=3), the algorithm is as well
+converged as the missingness proportion allows. Sooner even with lower
+missingness proportions (T=1 might be ok for p\_mis=5 and 25).} Bias is
+the worst for conditions where p\_mis = 95, T\textless3. Followed by
+p\_mis = 75, T=1; and p\_mis = 50, T=1. Conditions where p\_mis = 25 or
+5 seem unaffected by the number of iterations. Minimal bias is obtained
+in all iteration conditions. We should see this in Rhat and AC: worst
+convergence in p\_mis = 95 and 50; some non-conv in conditions where
+p\_mis = 75 and T=1, and no non-conv in conditions where p\_mis = 25 or
+5. The sign of the bias is logical: when there is less info in the data,
+the relations between the predictors and the outcome is under-estimated.
+
+Rhat behaves oppositely to what we wanted. The lowest Rhat values for
+conditions where T\textless10 are observed for the two missingness
+conditions with the worst bias. However, the highest Rhat observed is
+indeed for the missingness condition with the worst bias, p\_mis =95.
+But only at T=5 (while bias is no worse than anywhere else in this
+p\_mis condition, conditional on iterations condition T\textgreater2).
+So neither of the thresholds works perfectly.
+
+AC shows a similar trend to the biases: more signs of non-conv for
+higher p\_mis and lower T. The threshold, however, performs less well.
+It identifies signs of non-convergence for all conditions where \(T=3\),
+irrespective of p\_mis. Then nothing, and then for p\_mis =95,
+T\textgreater7, and p\_mis=50, T \textgreater14, and p\_mis = 75,
+T\textgreater20. This does not corresond to the bias in \(\bar{Q}\).
+Maybe use AC\textgreater.5 at T=5 as threshold?
+
 \begin{figure}
@@
 
-\hypertarget{univariate-estimates-and-convergence-diagnostics}{%
-\subsection{Univariate estimates and convergence
-diagnostics}\label{univariate-estimates-and-convergence-diagnostics}}
-
-The bias in the estimates of the variable means shows little to no
-difference between simulation conditions. It doesn't seem to matter how
-many iterations you use in the mice algorithm, the estimates are
-unbiased. Similarly, the bias in the estimated variances is more or less
-stable across simulation conditions. These univariate quantities appear
-to be unaffected by the number of iterations.
-
-When applied to the imputation chain means, \(\widehat{R}\) indicates
-that the mice algorithm does not reach a converged state
-(\(\widehat{R}\) = 1) in any of the simulation conditions. Neither is
-the most recent recommended threshold reached (\(\widehat{R}\)
-\textless{} 1.01). \textbf{This is only for 75\% missingness:} The
-conventional \(\widehat{R}\) threshold of 1.2 is reached in simulation
-conditions \(T = 3\) and \(T > 6\). (With the default number of
-iterations (maxit = 5), this dip in \(\widehat{R}\) values would be
-spotted, so it is no problem.) The point at which an extra iteration
-does not seem to improve the \(\widehat{R}\) value is around \(T=30\).
-
-Autocorrelations indicate no sign of trending \textbf{(NOT correct
-anymore, fluke of the stats::acf() function!!)} within imputation
-chains. In most simulation conditions \(AC\) is smaller than or about
-equal to zero. Across simulation conditions, however, the
-autocorrelation curve does not trend towards 0. Autocorrelation values
-plateau off at a value of around.1. This is a small positive
-autocorrelation, which would indicate some trending within chains.
-
-The \(\widehat{R}\) and \(AC\) values for the imputation chain variances
-show equal trends to the chain means and are therefore not discussed
-separately. Taken together, univariate estimates seem robust across
-simulation conditions. There is no clear effect of the number of
-iterations on the bias in these estimates, while the convergence
-diagnostics indicate that the algorithm did not reach a completely
-converged state (yet).
-
-\hypertarget{multivariate-estimates}{%
-\subsection{Multivariate estimates}\label{multivariate-estimates}}
-
-\textbf{This is only true for 75\% missingness:} There is a clear bias
-in the regression estimates in simulation conditions where the number of
-iterations is smaller than four. In simulation conditions where
-\(T > 5\) there is little to no bias in the estimated regression
-coefficients. In most simulation conditions nominal coverage is obtained
-(i.e., coverage rates of.95) for the confidence intervals around the
-regression coefficients. Conditions with only one or two iterations show
-some under-coverage. Since confidence interval width is stable across
-conditions, the under-coverage may be attributed to the bias in the
-estimated regression coefficients.
-
-\textbf{This is only true for 75\% missingness:} If we look at the
-estimated proportion of explained variance in outcome variable Y we see
-that the coefficient of determination is underestimate estimated in
-conditions where the number of iterations is equal to two or less, and
-slightly overestimated in conditions where the number of iterations is
-equal to three or more.
-
-In short, we see that the minimum number of iterations required to
-obtain unbiased, confidence valid regression estimates is 5. This value,
-however, is dependent on the percentage of missing values. E.g., with
-95\% of cases having missing data we need at least seven iterations to
-obtain unbiased results.
-
 \hypertarget{discussion}{%
@@
 
-\emph{Proper performance under a `missing completely at random'
-missingness mechanism is necessary to demonstrate the appropriateness of
-\(\widehat{R}\) and \(AC\) as non-convergence diagnostics. However,
-results may not be extrapolated to other missingness mechanisms.}
-
-Upon convergence, \(\widehat{R}=1\) and \(AC=0\), which are unlikely
-thresholds for MCMC algorithms, because of its convergence to a
-distribution. In practice, non-convergence is usually diagnosed when
-\(\widehat{R}\) \textgreater{} 1.2 or 1.1 or even 1.01. And a t-test is
-performed to assess whether \(AC\) is significantly different from zero.
-
-As expected, complete convergence (\(\widehat{R}=1\) and \(AC=0\)) is
-never observed. But what about the thresholds for practice?
-\(\widehat{R}\) \textless1.2 is not stringent enough, because conditions
-in which \(\widehat{R}\) is smaller than 1.2 do not all yield unbiased
-estimates of the \(Q\)s. Moreover, there is a dip in the \(\widehat{R}\)
-values, after which \(\widehat{R}\) \textgreater{} 1.2 again. If the
-algorithms would be terminated at the iteration where \(\widehat{R}\)
-\textless{} 1.2 occurs for the first time, the increase in
-\(\widehat{R}\) values might be missed. The threshold \(\widehat{R}\)
-\textless{} 1.1 is somewhat conservative in comparison with the
-performance measures. Overall it seems OK. \(\widehat{R}\) \textless{}
-1.01 is too stringent compared to the performance measures. And it is
-not obtained for any number of iterations (\textbf{but may be necessary
-in more complex complete data models???}).
-
-This note shows that convergence diagnostics \(\widehat{R}\) and \(AC\)
-may diagnose convergence of multiple imputation algorithms, but their
-performance differs from conventional applications to iterative
-algorithmic procedures. (\textbf{nope! it shows that MICE can lead to
-correct outcomes when they have not converged according to two common
-conv diags. This may be due to the measures (e.g., assumption of
-overdisp) or due to the Qs (lm reg coeff, not higher dimensional/more
-complex RQs). Add what \%miss has to do with it.})
+In short, univariate estimates seem robust against terminating the
+algorithm early: There is no clear effect of the number of iterations on
+the bias in these estimates. The bias in the Qs only depends on the
+p\_mis. Yet, the convergence diagnostics indicate that the algorithm did
+not reach a completely converged state (yet).
 
@@
 while unbiased, confidence valid estimates may be obtained with as
-little as four iterations. These results are in agreement with the
+little as one iteration. These results are in agreement with the
 simulation hypothesis: \(\widehat{R}\) over-estimates the severity of
-non-convergence when applied to MI procedures. \%This may be due to the
-quantity of scientific interest chosen. More `complicated' \(Q\)s (e.g.,
-higher-order effects or variance components) might show bias, under- or
-over-coverage at higher \(T\).
+non-convergence when applied to MI procedures.
 
-According to this simulation study, the recently proposed threshold of
-\(\widehat{R}<1.01\) may be too stringent for MI algorithms.
-(\textbf{This is only one of the goals: to give applied researchers a
-diagnostic to indicate that they should keep iterating. The other is the
-default in mice and other software packages, and yet another is\ldots{}
-i forgot}) Under the relatively easy missing data problem of the current
-study, the threshold was not reached. The other extreme of the
-\(\widehat{R}\)-thresholds, the conventionally acceptable
-\(\widehat{R} <1.2\), may be too lenient for MI procedures. Applying
-this threshold to the current data, lead to falsely diagnosing
-convergence at \(T = 3\) (\textbf{because it goes up after, not because
-it is not converged enough}). It appears that the widely used threshold
-of \(\widehat{R} < 1.1\) suits MI algorithms the best. We might,
-however, also formulate a new threshold, specifically for the evaluation
-of MI algorithms. The current study suggests that \(\widehat{R} < 1.05\)
-may be implemented, since that is the level at which the \(\widehat{R}\)
-stabilize (around \(T = 20\)) (\textbf{Not necessary for this Q, but
-maybe for more complicated Qs}).
-
-The negative \(AC\)-values obtained in this study show no threat of
-non-stationarity. However, the initial dip in \(AC\)-values
-\textbf{(which disappeared!)} may have implications for the default
-number of iterations in \texttt{mice} (\texttt{maxit\ =\ 5}).
-Terminating the algorithm at \(T=5\) may not be the most appropriate,
-since this lead to the worst convergence (\textbf{nope, only for Rh, not
-\(AC\)}), as indicated by \(\widehat{R}\) and \(AC\). Under the current
-specifications, \(T>20\) would be more appropriate.
-
-Further research is needed to investigate their performance under clear
-violation of convergence, e.g.~dependency between predictors (predictors
-with very high correlations). Until then, we have only shown that the
-convergence diagnostics can diagnose non-convergence of MI algorithms
-that trend towards a converged state. Also for future research, look at
-developing a convergence diagnostic for substantive models, and
-implement a Wald test for \(AC = 0\).
+With that, we show that MICE can lead to correct outcomes when they have
+not converged according to two common conv diags. This may be due to the
+methods (and their thresholds) or due to the Qs (descriptives and
+multivariate linear regression, not higher dimensional/more complex Qs).
+More `complicated' \(Q\)s (e.g., higher-order effects or variance
+components) might show bias, under- or over-coverage at higher \(T\), as
+indiated by Rhat and AC. Add what \%miss has to do with it.
 
@@
 
-For empirical researchers: 1) Check trace plots for pathological non
-convergence and adjust imputation model if necessary. 2) use
-\(\widehat{R}\) wait 1.1 ash threshold and autocorrelation with
+For empirical researchers:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\item
+  Check trace plots for pathological non convergence and adjust
+  imputation model if necessary.
+\item
+  use \(\widehat{R}\) wait 1.1 ash threshold and autocorrelation with
   \textbf{?} as threshold. Keep iterating until these thresholds are
-reached. 3) Do not use the R function ACF. Instead, compute
-autocorrelations manually \textbf{(see e.g., {[}GitHub link{]})}. 4)
-Track your own scalar summary of interest. This is somewhat advanced but
-explained in Van Buren 2018. Compute \(\widehat{R}\) and autocorrelation
-values for this scalar summary. 5) \textbf{Something} about the novel
-\(\theta\) that is `substantive model-independent'.
+  reached.
+\item
+  Do not use the R function ACF. Instead, compute autocorrelations
+  manually \textbf{(see e.g., {[}GitHub link{]})}.
+\item
+  Track your own scalar summary of interest. This is somewhat advanced
+  but explained in Van Buren 2018. Compute \(\widehat{R}\) and
+  autocorrelation values for this scalar summary.
+\item
+  \textbf{Something} about the novel \(\theta\) that is `substantive
+  model-independent'.
+\end{enumerate}
 
@@
 
-\textbf{Pmm, M(N)AR, Empirical data, Vignette, ShinyMICE, AC across
-imputations, not iterations (new convergence diagnostic unique for
-iterative imputation??).}
+Further research is needed to investigate their performance under clear
+violation of convergence, e.g.~dependency between predictors (predictors
+with very high correlations). Also:
 
-\textbf{If I don't include it in my study, also: Significance of \(AC\)
-values.}
+\begin{itemize}
+\item
+  Induce non-convergence with mis-specified imputation model.
+\item
+  More difficult missingness problems, i.e., M(N)AR instead of MCAR.
+  Proper performance under a `missing completely at random' missingness
+  mechanism is necessary to demonstrate the appropriateness of
+  \(\widehat{R}\) and \(AC\) as non-convergence diagnostics. However,
+  results may not be extrapolated to other missingness mechanisms.
+\item
+  Imputation technique with questionable convergence properties, e.g.,
+  PMM.
+\item
+  Apply recommendations to empirical data and write a vignette for
+  applied researchers.
+\item
+  Implement recommendations in ShinyMICE.
+\item
+  Develop an additional convergence diagnostic unique for iterative
+  imputation: AC across imputations, instead of iterations.
+\end{itemize}
 
@@
 
+\begin{itemize}
+\item
   Scientific contribution (clear research question, convincing
   argumentation that the study adds to the scientific literature)
-
+\item
   Theory (use of theory, clear hypotheses, relevant literature)
-
+\item
   Method (correct data collection/handling, measures, choice and
   description of method of analysis)
-
+\item
   Results (correct interpretation of findings, clear presentation)
-Discussion and conclusions (summary, broader implications, limitations,
-future research)
-
+\item
+  Discussion and conclusions (summary, broader implications,
+  limitations, future research)
+\item
   Written presentation (writing, structure, tables and figures,
   references)
+\end{itemize}
 

### 2020-05-02 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
Remove comments from Methods section, rewrite first part of Results section to correspond to the newly formulated simulation objectives.

@@
 
-<!-- Q is quantity of scientific interest: "A scientific estimand $Q$ is a quantity of scientific interest that we can calculate if we would observe the entire population" [@buur18, par 2.3.1] -->
+<!-- $Q$ is quantity of scientific interest: "A scientific estimand $Q$ is a quantity of scientific interest that we can calculate if we would observe the entire population" [@buur18, par 2.3.1] -->
 
@@
 This method is insufficient on two counts: 1) it may be challenging to the untrained eye, and 2) only severely pathological cases of non-convergence may be diagnosed [@buur18, $\S$ 6.5.2].
-Therefore, a quantitative, diagnostic evaluation of convergence would be preferred.
-Yet, monitoring convergence of iterative imputation algorithms diagnostically is challenging.
+Therefore, a quantitative, diagnostic evaluation of convergence would be preferred---yet not straightforward.
 Iterative imputation algorithms such as MICE are Markov chain Monte Carlo (MCMC) methods.
-In MCMC methods, convergence is not from a scalar to a point but from one distribution to another.
+In MCMC methods, convergence is not from a scalar to a point, but from one distribution to another.
 The values generated by the algorithm (e.g., imputed values) will vary even after convergence.
-Therefore, the aim of convergence diagnostics for MCMC methods is not to establish the point at which convergence is reached, but to monitor signs of non-convergence [@hoff09].
-Several of such diagnostics exist for MCMC methods, but it is not known whether these are appropriate for iterative imputation algorithms.
+Since MCMC algorithms do not reach a unique point at which convergence is established, diagnostics we may only identify signs of non-convergence [@hoff09].
+Several of such diagnostics exist, but it is not known whether these are appropriate for iterative imputation algorithms.
 <!-- "a weak diagnostics is better than no diagnostic at all" [@cowl96]. -->
@@
 The results of this simulation study are guidelines for assessing convergence of MI algorithms, which will aid applied researchers in drawing valid inference from incomplete datasets.
+**add societal relevance: what is the consequence of non-convergence?**
 
-**Include sub-questions!** How can non-convergence be identified diagnostically?
+**Include sub-questions and add structure of the paper!** How can non-convergence be identified diagnostically?
 Are common MCMC non-convergence diagnostics appropriate for MICE?
@@
 The data value of unit $i$ ($i = 1, 2, \dots, n$) on variable $j$ ($j = 1, 2, \dots, k$) may be either observed or missing.
-**The number of units $i$ with at least one missing data value can be divided by the total number of units $n$ to obtain the 'missingness proportion'  $p_{mis}$.** The collection of observed data values in $y$ is denoted by $y_{obs}$; the missing part of $y$ is referred to as $y_{mis}$.
+**The number of units $i$ in dataset $y$ with at least one missing data value, divided by the total number of units $n$, is called the missingness proportion in $p_{mis}$.** The collection of observed data values in $y$ is denoted by $y_{obs}$; the missing part of $y$ is referred to as $y_{mis}$.
 For each datapoint in $y_{mis}$, we sample $M \times T$ times plausible values, where $M$ is the number of imputations ($m = 1, 2, \dots, M$) and $T$ is the number of iterations ($t = 1, 2, \dots, T$).
@@
 
-The goal of imputation through e.g.
-FCS/MICE is to pbtain a multivariate distribution by iterating over the sequence of univariate imputations.
-Traditionally, the inspection of the multivariate convergence is done by visually inspecting the univariate chain means and variances over the iterations, or by tracking the Rhat and autocorrelation.
+**The goal of imputation through e.g.
+MICE is to obtain a multivariate distribution by iterating over the sequence of univariate imputations.
+Traditionally, the inspection of the multivariate convergence is done by visually inspecting the univariate chain means and variances over the iterations, or by tracking the $\widehat{R}$ and autocorrelation.
 The downside to these approaches is that they either focus on the univariate state space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest.
@@
 We propose eigenvalue as such a parameter, because it....
+**
 
@@
 Therefore, we compute $AC$ manually, see panel D.
-The AC values in this plot do meet the requirements (...) and will therefore be used in the simulation study.**
+The $AC$ values in this plot do meet the requirements (...) and will therefore be used in the simulation study.**
 
@@
 The aim of the simulation study is to evaluate the impact of inducing non-convergence in the MICE algorithm on several quantities of scientific interest $Q$.
-Subsequently we will evaluate how well $\widehat{R}$ and $AC$ perform in identifying the effects of non-convergence.
+Subsequently we will evaluate how well $\widehat{R}$ and $AC$ perform in identifying the effects of non-convergence (**add thetas**).
 And finally, we will formulate an informed advice on the requirements to safely assume *sufficient* convergence in practice.
 That is, we assume that the algorithm is *sufficiently* converged when each estimate $\bar{Q}$ is an unbiased and confidence valid estimate of the corresponding estimand $Q$.
+(**add to formulate advice on thetas?**)
 
@@
 That is, after only a few iterations, the randomness induced by the algorithm should effectively mitigate the risk of dependency within chains.
+
+5. **Something about thetas**
 
@@
  8. combine outcomes of all missingness proportions
-9. aggregate outcomes across simulation runs
+9. aggregate outcomes of all simulation runs
 
@@
 $$
-where $\epsilon \sim \mathcal{N}(0, 100)$. From the complete set, we obtain the true values of each quantity of scientific interest $Q$.
+where $\epsilon \sim \mathcal{N}(0, 100)$.
+From the complete set, we obtain the true values of several quantities of scientific interest, $Q$s.
+**MOVE THIS NEXT PART TO TOP OF SIMULATION SECTION?** $Q$s are the targets that will be estimated in each simulation repetition, after first *amputing* the complete data with varying missingness proportions, then imputing the missingness with a varying number of iterations, then estimating the $Q$s on the $M$ completed datasets, and finally pooling the $M$ $\hat{Q}$s to obtain a $\bar{Q}$ for each $Q$.
+**Add that we try to predict the effect of non-converegence (i.e., the difference between $Q$ and $\bar{Q}$) with the convergence diagnostics $\widehat{R}$ and $AC$, by appliying these methods to several scalar summaries of the state space of the algorithm, $\theta$s.**
 
@@
 And finally, the proportion of variance explained by the set of predictors: coefficient of determination $r^2 \times 100$ **(note that lower case $r$ is used to avoid confusion with non-convergence diagnostic $\widehat{R}$)**.
-The true values of these scientific estimands are calculated with the R functions `base::mean()`, `stats::sd()`, and `stats::lm()` [@R].
-Each of these $Q$s is estimated with a corresponding $\bar{Q}$, after the complete dataset is first *amputed* (with a varying missingness proportion $p_{mis}$) and subsequently imputed (with a varying number of iterations $T$).
+The true values of these scientific estimands are calculated on the complete data with the R functions `base::mean()`, `stats::sd()`, and `stats::lm()` [@R].
+<!-- Each of these $Q$s is estimated with a corresponding $\bar{Q}$, after the complete dataset is first *amputed* (with a varying missingness proportion $p_{mis}$) and subsequently imputed (with a varying number of iterations $T$).
+-->
 
@@
 
-The complete dataset is amputed with the `mice` package [function `mice::ampute()`; @mice].
+In each simulation repetition, we ampute the complete dataset five times to obtain five different incomplete datasets, $y$s.
+The ratio between $y_{obs}$ and $y_{mis}$ in each $y$ depends on the missingness proportion: $p_{mis} =.05,.25,.5,.75,.95$.
+We ampute the complete data with the `mice` package [function `mice::ampute()`; @mice].
 All missingness is multivariate, and conform a 'missing completely at random' missingness mechanism [@rubin87].
-I.e., the probability to be missing is the same for all $N \times k$ cells in $y$, conditional on the missingness proportion ($p_{mis} =.05,.25,.5,.75,.95$).
+I.e., the probability to be missing is the same for all $N \times k$ cells in $y$, conditional on the missingness proportion $p_{mis}$.
 <!-- This note only considers a 'missing completely at random' (MCAR) mechanism, where the probability of being missing is equal for all $n \times k$ cells in $y$ [@rubin87].
@@
 
-Missing datapoints in $y$ are imputed with the `mice()` function [@mice].
+Missing datapoints in the incomplete datasets, $y$s, are imputed with the `mice()` function [@mice].
 All imputation procedures are performed with Bayesian linear regression imputation, and five imputation chains ($M=5$).
-The number of iterations varies between simulation conditions ($T=1,2,...,100$).
+Because $M=5$, each run of the `mice()` function results in five sets of imputations, $y_{imp,m}$s, where $m = 1, 2, ..., 5$.
+The number of iterations in each run of the algorithm varies between simulation conditions ($T=1,2,...,100$).
 <!-- Each repetition of the simulation thus starts from 1 complete dataset, then splits into 5 missingness conditions ($p_{mis} =.05,.25,.5,.75,.95$), each of the amputed datasets is imputed 5 times ($m = 1, 2,..., 5$) for each iteration condition ($t = 1, 2,..., 100$).
@@
 
-After each run of the `mice` algorithm, the function `mice::complete()` is used to combine the imputed data $y_{imp,m}$ with the observed data $y_{obs}$, resulting in five completed datasets {$y_{obs}, y_{imp, m}$}, where $m = 1, 2, ..., 5$.
+From the 5 imputations $y_{imp,m}$: get chain means and chain variances from imputation object to use as scalar summary $\theta$ in convergence diagnostics.
+Then use `mice::complete()` to combine the imputed data $y_{imp,m}$ with the observed data $y_{obs}$, resulting in five completed datasets {$y_{obs}, y_{imp, m}$}.
 
-- on the 5 imputations $y_{imp,m}$: get mean and var within imp chains as theta to compute conv diag --> then complete with `mice::complete()`
+On the 5 completed datasets {$y_{obs}, y_{imp, m}$}, we:
 
-- on the 5 completed datasets {$y_{obs}, y_{imp, m}$}: 1) estimate Qhats for mu and sigma --> pool as mean() --> get Qbars of mu and sigma; 2) compute lambda as theta for conv diag; 3) perform lm to get Qhats for betas and r^2 --> use Qhats of beta as theta to compute conv diag --> pool Qhats of beta and r^2 to get Qbars
+- Estimate $\hat{Q}$s for $Q= \mu$ and $Q=\sigma$ with `base::mean()`, and `stats::sd()` [@R]. Then pool the $\hat{Q}$s with `base::mean()` to get $\bar{Q}$s.
 
-The estimator for each $Q$ is $\bar{Q}$---the pooled aggregate of the $\hat{Q}$s across imputations.
+- Compute $\lambda_1$, to use as $\theta$ in convergence diagnostics.
+By definition, the first eigenvalue of the variance-covariance matrix, $\lambda_1$, is equal to the variance of the principal component solution of each {$y_{obs}, y_{imp, m}$}, which we calculate with `stats::princomp()` [@R].
+
+- Perform multiple linear regression with `stats::lm()` [@R] to get the $\hat{Q}$s for $Q=\beta$ and $Q=r^2$.
+Pool the $\hat{Q}$s conform @vink14 to get $\bar{Q}$s for the $\beta$s and $r^2$ (**check  pooling of  $r^2$, instead of `mice::pool.r.squared()`**).
+Additionally, use the $\hat{Q}$s of $Q=\beta$ as $\theta$ to compute convergence diagnostics.
+
+<!-- After each run of the `mice` algorithm, the function `mice::complete()` is used to combine the imputed data $y_{imp,m}$ with the observed data $y_{obs}$, resulting in five completed datasets {$y_{obs}, y_{imp, m}$}.
+-->
+
+<!-- The estimator for each $Q$ is $\bar{Q}$---the pooled aggregate of the $\hat{Q}$s across imputations.
 To calculate the $\bar{Q}$s, we combine the observed data $y_{obs}$ and the imputed data for each imputation $y_{imp,m}$ with the function `mice::complete()`.
@@
 The coefficient of determination is estimated for each imputation, and pooled using `mice::pool.r.squared()`.
+-->
 
-For each run of the imputation algorithm, we compute non-convergence diagnostics $\widehat{R}$ and $AC$.
+We apply non-convergence diagnostics $\widehat{R}$ and $AC$ to each scalar summary $\theta$.
 We calculate $\widehat{R}$ by implementing Vehtari et al.'s [-@veht19] recommendations, and $AC$ as the correlation between the $t^{th}$ and the $(t+1)^{th}$ iteration.
-The two methods are applied to four scalar summaries $\theta$: chain means (i.e., the mean in each $y_{imp, m, j}$), chain variances (i.e., the variance in each $y_{imp, m, j}$), a statistic of scientific interest $\hat{Q}$ in each {$y_{obs}, y_{imp, m}$}, and the first eigenvalue of the variance-covariance matrix in each {$y_{obs}, y_{imp, m}$}, $\lambda_1$.
-
-The $\widehat{R}$ and $AC$ values with chain means as $\theta$s are evaluated against the bias in univariate mean estimates.
-Similarly, the performance measure for $\widehat{R}$ and $AC$ applied to chain variances is the bias in estimated standard deviations.
-To evaluate the performance of $\widehat{R}$ and $AC$ on the eigenvalues, we will use bias in the coefficient of determination, $R^2$.
-Additionally, we will evaluate the estimated regression coefficients, the coverage rate of the CI95% of the regression estimates, and the CI length (**see definitions in old Methods**).
-**remove the term performance measure here and add an extra paragraph to define performance measures as bias in all estimands, and coverage rate and CI length of regression coefficients.**
 
@@
 
-The impact of inducing non-convergence in the iterative imputation algorithm is assessed by comparing $\bar{Q}$s with $Q$s.
-For each $Q$, we compute bias as $\bar{Q} - Q$.
+We assess the impact of inducing non-convergence in the iterative imputation algorithm by comparing $\bar{Q}$s with $Q$s.
+For each $Q$, we calculate bias as $\bar{Q} - Q$.
 For the regression coefficients $Q=\beta_{1,2,3}$, we also compute the empirical coverage rate (CR).
@@
 where $t_{(M-1)}$ is the quantile of a $t$-distribution with $M-1$ degrees of freedom, and $SE_{\bar{Q}}$ is the square root of the pooled variance estimate.
-Since under-estimating the variance of $\bar{Q}$ may yield spurious inferences, we also inspect CI width (CIW): the difference between the lower and upper bound of the 95\% confidence interval around $\bar{Q}$.
+If we obtain nominal coverage (CR = 95%), we can conclude that the $\bar{Q}$s are conficence valid estimates of the $Q$s.
+Finally, we inspect CI width (CIW): the difference between the lower and upper bound of the 95\% confidence interval around $\bar{Q}$.
+CIW is of interest because too short CIs indicate under-estimation of the variance in $\bar{Q}$, which may yield spurious inferences.
 
-The performance of two methods is evaluated against the bias and confidence validity of the estimated $Q$s.
+With these performance measures, we also evaluate how well the non-convergence diagnostics $\widehat{R}$ and $AC$ can identify simulation conditions in which the $\bar{Q}$s are *not* unbiased, confidence valid estimates of the $Q$s.
+**Rephrase and add corresponding $Q$s:** The two methods are applied to four scalar summaries $\theta$: chain means (i.e., the mean in each $y_{imp, m, j}$), chain variances (i.e., the variance in each $y_{imp, m, j}$), the estimated regression coefficients $\beta_{1,2,3}$ in each {$y_{obs}, y_{imp, m}$} (i.e., a user-defined statistic of scientific interest $\hat{Q}$), and the first eigenvalue of the variance-covariance matrix in each {$y_{obs}, y_{imp, m}$}, $\lambda_1$.
 
@@
 
-We want to know what the effect is of inducing non-conv of the bias in all Qs and the confidence validity of the betas.
-So we plot the bias for each Q against the number of iterations and the missingness proportion in the condition.
-We look at averages across simulation repetitions.
-Is there more bias in conditions where non-conv was induced?
-(NOPE for univariates, YES for multivar.).
-
-Then, we want to know if conditions in which the Qs are affected by non-convergence (on average) are also identified by the conv diag.
-We do this by looking at the bias in Q against the conv diag applied to related thetas.
-Is the point at which there is no more bias the same point at which the conv diag indicate minimal signs of non-conv?
-(NOPE, they are too conservative).
-
-Is one of the thetas better? (NAH, doesn't really matter).
-
-What should the thresholds be? (LOWER than recommended) And the default nr of iterarions? (MAYBE increase to 10?)
-
-**Add to this section: 1) emphasize that the plots are averages across repetitions, not within MICE; 2) sample effects due to single complete dataset.
+**Add to this section: 1) emphasize that the plots are averages across repetitions, not within MICE; 2) sample effects due to single complete dataset??
 3) Add more info about figure legends and axes.)**
 
-For reasons of brevity, we only discuss the convergence diagnostics for the $Q$ with the worst performance in terms of bias.
-For the descriptve statistics, the magnitude of the bias was the largest in $Y$ (i.e, $j = Y$ in $\mu_j$ and $\sigma_j$).
-For the $Q=\beta_{1,2,3}$, $\beta_1$ showed the worst perfomance (i.e, the effect of $X_1$ on $Y$.
-Moreover, we only show conditions where $T\leq50$ in the figures.
+Primarily, we want to know what the effect is of inducing non-convergence on the estimated $Q$s.
+Specifically, we are interested in the bias of $\bar{Q}$ for all $Q$s and the confidence validity of $\bar{Q}$ for $Q=\beta$.
+Therefore, we plot these performance measures against the simulation conditions (the number of iterations $T$ and the missingness proportion $p_{mis}$).
+We look at averages across simulation repetitions.
+
+Secondly, we want to know if conditions in which the $Q$s are affected by non-convergence are also identified by the non-convergance diagnostics $\widehat{R}$ and $AC$.
+We do this, again, by looking at the performance measures for the $Q$s, but now we evaluate these against what $\widehat{R}$ and $AC$ indicate about signs of non-the convergence in the $\theta$s.
+
+- Is there more bias in conditions where non-convergence was induced? (NOPE for univariates, YES for multivar.).
+
+- Is the point of *sufficient* convergence (at which the $\bar{Q}$s are unbiased and confidence valid estimates of the $Q$s) the same point as identified by the non-convergence diagnostics?
+(NOPE, they are too conservative).**
+
+- Is one of the thetas better? (NAH, doesn't really matter).
+
+- What should the thresholds be? (LOWER than recommended).
+
+- And the default nr of iterarions? (MAYBE increase to 10?).
+
+For reasons of brevity, we only discuss the convergence diagnostics for the $Q$s with the worst performance in terms of bias.
+Bias in the  descriptve statistics ($Q=\mu$ and $Q=\sigma$) was most pronounced in the outcome variable $Y$.
+Of the regression coefficients, $Q=\beta$, the largest bias was observed in the effect of $X_1$ on $Y$, $\beta_1$.
+Results for $Q=r^2$ are reported as-is.
+In the figures, however, we only present simulation conditions where $T\leq50$.
 Full results are available from **[link github repo]**.
@@
 
-As expected, conditions with a higher proportion of missingness and/or a lower number of iterations show more signs of non-convergence, as indicated by more extreme bias in the estimated $Q$s.
-Roughly speaking this means that MICE is indeed not converged as $t=1$, and converges gradually as $t$ increases.
+As expected, conditions with a higher proportion of missingness and/or a lower number of iterations show more signs of non-convergence.
+I.e., on average these conditions portet more extreme bias in the estimated $Q$s, and non-nominal coverages.
+Roughly speaking this means that MICE is indeed not converged at $t=1$, and converges gradually as $t$ increases to $T$.
 The point at which an additional iteration does not lead to an improvement of the estimates depends on the difficulty of the missingness problem---between XYZ and XYZ.
@@
 As required for a method to diagnose non-convergence, the values of $\widehat{R}$ follow a trend similar to the performance measures.
-$\widehat{R}$ values are generally lower in conditions with a higher number of iterations, and somewhat higher in conditions with a higher percentage of missing data.
-Autocorrelation values were indeed decreasing with a higher number of iterations.
+**Add split between convergence diagnostics versus performance measures, and convergence diagnostics versus simulation conditions.** $\widehat{R}$ values are generally lower in conditions with a higher number of iterations, and somewhat higher in conditions with a higher percentage of missing data.
+Autocorrelation values were indeed decreasing with a higher number of iterations, and lower in conditions with less missingness.
 
@@
 
-**Message: after one iteration (T=2), the algorithm is as well converged as the missingness proportion allows.
-Sooner even with lower missingness proportions (T=1 might be ok for p_mis=5 and 25).** Number of it has little influence on bias in hat{Q}, only in first iteration some bias in missingness conditions of 25 and up.
+**Message: after one iteration ($T$=2), the algorithm is as well converged as the missingness proportion allows.
+Sooner even with lower missingness proportions ($T$=1 might be ok for $p_{mis}$ = 5 and 25).** Number of it has little influence on bias in $\hat{Q}$, only in first iteration some bias in missingness conditions of 25 and up.
 There is a clear different between the missingness conditions.
 The higher the missingness prop, the larger the magnitude of the bias.
-For this Q, t=2 would be sufficient.
-And only for p_mis = 5 the bar{Q}s are unbiased.
-All other p_mis conditions have bias irrespective of the number of iterations.
+For this $Q$, t=2 would be sufficient.
+And only for $p_{mis}$ = 5 the $\bar{Q}$s are unbiased.
+All other $p_{mis}$ conditions have bias irrespective of the number of iterations.
 
-Rhat differs a lot across iteration conditions.
+$\widehat{R}$ differs a lot across iteration conditions.
 But not structurally for missingness proportions.
-Highest Rhats for the condition t=2, while the most bias is for (t=1), but there is no Rhat for that by definition.
+Highest $\widehat{R}$s for the condition $T$=2, while the most bias is for ($T$=1), but there is no $\widehat{R}$ for that by definition.
 But for missingness, 25% seems the most signs of non-conv, while it has one of the least bias.
-Threshold of 1.2 reached in condition it = 3, p_mis = 5 and 75, and then for it>6, p_mis= 5 and 50, and it>10 for all p_mis.
-Threshold 1.1 reached in in it=12, p_mis = all but 75, 1.01 only reached in it = 93, p_mis = 75.
-While for this Q, t=2 would be sufficient.
+Threshold of 1.2 reached in condition it = 3, $p_{mis}$ = 5 and 75, and then for it>6, $p_{mis}$ =  5 and 50, and $T$>10 for all $p_{mis}$.
+Threshold 1.1 reached in in $T$=12, $p_{mis}$ = all but 75, 1.01 only reached in $T$ = 93, $p_{mis}$ = 75.
+While for this $Q$, $T$=2 would be sufficient.
 
-AC mainly shows effects of the set of iteration conditions.
-AC indicates less signs of non-conv in conditions with higher number of iterations.
-For iteration conditions where t<10, there are also some effects of the missingness proportions.
-The p_mis condtions 25 and 95 % indicate the most signs of non-convergence.
-For p_mis = 95% this corresponds to the bias in Qbar.
+$AC$ mainly shows effects of the set of iteration conditions.
+$AC$ indicates less signs of non-convergencein conditions with higher number of iterations.
+For iteration conditions where $T$<10, there are also some effects of the missingness proportions.
+The $p_{mis}$ condtions 25 and 95 % indicate the most signs of non-convergence.
+For $p_{mis}$ = 95% this corresponds to the bias in Qbar.
 For 25% miss it doesn't.
-The thresholds defined by the critical value is only met in the conditions it=3, p_mis = 5,75, and 95%.
+The thresholds defined by the critical value is only met in the conditions it=3, $p_{mis}$ = 5,75, and 95%.
 
@@
 
-**Message: even with possible effect of initial values (T=1), the algorithm is as well converged as the missingness proportion allows.** For $Q=\sigma_Y$ there is no effect visible of the iteration conditions, but clearly for the p_mis conditions.
-Most bias for p_mis = 95, followed by p_mis=50 and 75.
-For the least possible bias we may use any number of iterations $T\geq1$, but none of the p_mis results in unbiased bar{Q}s.
+**Message: even with possible effect of initial values ($T$=1), the algorithm is as well converged as the missingness proportion allows.** For $Q=\sigma_Y$ there is no effect visible of the iteration conditions, but clearly for the $p_{mis}$ conditions.
+Most bias for $p_{mis}$ = 95, followed by $p_{mis}$ = 50 and 75.
+For the least possible bias we may use any number of iterations $T\geq1$, but none of the $p_{mis}$ results in unbiased $\bar{Q}$s.
 
-There is a clear effect of the number of iterations on Rhat.
-Conditions with lower Ts have more signs of non-conv, as identified by Rhat.
+There is a clear effect of the number of iterations on $\widehat{R}$.
+Conditions with lower Ts have more signs of non-conv, as identified by $\widehat{R}$.
 There is little difference between missingness conditions.
-Only for conditions where $T<5$ the missingness proportions seem to affect whether Rhat indicates non-conv.
-The bias in bar{Q} seems unrelated to the magnitude of Rhat.
-Conditions with more bias (p_mis = 50,75,95) are not identified to have more signs of non-convergence.
-The thresholds are therefore useless (because depend mostly on t, not so much p_mis).
+Only for conditions where $T<5$ the missingness proportions seem to affect whether $\widehat{R}$ indicates non-convergence The bias in $\bar{Q}$ seems unrelated to the magnitude of $\widehat{R}$.
+Conditions with more bias ($p_{mis}$ = 50,75,95) are not identified to have more signs of non-convergence.
+The thresholds are therefore useless (because depend mostly on $T$, not so much $p_{mis}$).
 
-AC is affected by both t and p_mis conditions.
-But conditions with worst bias (p_mis = 50, 75, 95) are not identified as worst converging according to AC.
-Moreover, only for T=3 the AC is significantly different from zero.
+$AC$ is affected by both t and $p_{mis}$ conditions.
+But conditions with worst bias ($p_{mis}$ = 50, 75, 95) are not identified as worst converging according to AC.
+Moreover, only for $T$ =3 the $AC$ is significantly different from zero.
 
@@
 
-**Message: after four iterations (T=5), the algorithm is as well converged as the missingness proportion allows.
-Sooner even with lower missingness proportions (T=1 might be ok for p_mis=5 and 25).** The figure for $Q=\beta_1$ shows that the $\bar{Q}$s are affected by both the number of iterations and the missingness proportion.
-For T=1, the magnitude of the bias (almost) follows the order of the missingness proportions.
+**Message: after four iterations ($T$=5), the algorithm is as well converged as the missingness proportion allows.
+Sooner even with lower missingness proportions ($T$=1 might be ok for $p_{mis}$ = 5 and 25).** The figure for $Q=\beta_1$ shows that the $\bar{Q}$s are affected by both the number of iterations and the missingness proportion.
+For $T$ =1, the magnitude of the bias (almost) follows the order of the missingness proportions.
 More missingness causes more bias in the $\bar{Q}$.
 In general, a lower nr of iterations results in more bias, but this depends on the missingness proportion.
-For p_mis=95, there is more bias in conditions where T<6.
-When p_mis = 50, it's T<3.
-For p_mis = 75 and 25, it's only for T<2.
-And conditions where p_mis = 5 are unaffected on average.
+For $p_{mis}$ = 95, there is more bias in conditions where $T$<6.
+When $p_{mis}$ = 50, it's $T$<3.
+For $p_{mis}$ = 75 and 25, it's only for $T$<2.
+And conditions where $p_{mis}$ = 5 are unaffected on average.
 However, none of the conditions result in completely unbiased estimates of $Q$.
 
-We should see this in the conv diag: higher Rhat and AC for p_mis=75 and 95, and for condtions where $T<2-6$, depending on p_mis.
+We should see this in the conv diag: higher $\widehat{R}$ and $AC$ for $p_{mis}$ = 75 and 95, and for condtions where $T<2-6$, depending on $p_{mis}$.
 
-Rhat depends somewhat on the missingness proportion.
-For conditions where T<10, the Rhat values follow the p_mis more or less.
+$\widehat{R}$ depends somewhat on the missingness proportion.
+For conditions where $T$<10, the $\widehat{R}$ values follow the $p_{mis}$ more or less.
 But what we actually want to know is whether they are indicators of bias.
-Conditions with the worst bias (high p_mis, low T) have the largest Rhat values.
+Conditions with the worst bias (high $p_{mis}$, low T) have the largest $\widehat{R}$ values.
 The thresholds are not appropriate to diagnose sufficient convergence.
 All thresholds over-estimate the signs of non-convergence compared to the bias in Qbar.
-However, a higher threshold than 1.2 may result in terminating the algorithm after 3 iterations, which is insufficient for the p_mis = 95 condition.
-Threshold of 1.25 or 1.3 could work for the other p_mis conditions.
+However, a higher threshold than 1.2 may result in terminating the algorithm after 3 iterations, which is insufficient for the $p_{mis}$ = 95 condition.
+Threshold of 1.25 or 1.3 could work for the other $p_{mis}$ conditions.
 
-AC is very affected by the p_mis.
-The two conditions with the most bias (p_mis = 75 and 95) are the only ones identified as non-converging, as they should.
-However, the p_mis = 75 condition is only identified in the T=3 condtion, while the bias is the least pronounced there.
-And the AC only identifies non-convergence in the p_mis = 95 conditions where T>10.
-In the default situation, T=5, this would be overlooked.
-Also, the AC shows the opposite of what you would expect from a theta that is based on the observed data: shouldn't there be higher autocorrelations in conditions with comparatively more observed data in the completed datasets?
+$AC$ is very affected by the p_mis.
+The two conditions with the most bias ($p_{mis}$ = 75 and 95) are the only ones identified as non-converging, as they should.
+However, the $p_{mis}$ = 75 condition is only identified in the $T$ =3 condtion, while the bias is the least pronounced there.
+And the $AC$ only identifies non-convergence in the $p_{mis}$ = 95 conditions where $T$>10.
+In the default situation, $T$ =5, this would be overlooked.
+Also, the $AC$ shows the opposite of what you would expect from a theta that is based on the observed data: shouldn't there be higher autocorrelations in conditions with comparatively more observed data in the completed datasets?
 I don't understand this.
 
-**Add how interesting it is because we literally track Q and are still not ok.**
+**Add how interesting it is because we literally track $Q$ and are still not ok.**
 
@@
 
-**Message: after two iterations (T=3), the algorithm is as well converged as the missingness proportion allows.
-Sooner even with lower missingness proportions (T=1 might be ok for p_mis=5 and 25).** Bias is the worst for conditions where p_mis = 95, T<3.
-Followed by p_mis = 75, T=1; and p_mis = 50, T=1.
-Conditions where p_mis = 25 or 5 seem unaffected by the number of iterations.
+**Message: after two iterations ($T$=3), the algorithm is as well converged as the missingness proportion allows.
+Sooner even with lower missingness proportions ($T$=1 might be ok for $p_{mis}$ = 5 and 25).** Bias is the worst for conditions where $p_{mis}$ = 95, $T$ <3.
+Followed by $p_{mis}$ = 75, $T$ =1; and $p_{mis}$ = 50, $T$ = 1.
+Conditions where $p_{mis}$ = 25 or 5 seem unaffected by the number of iterations.
 Minimal bias is obtained in all iteration conditions.
-We should see this in Rhat and AC: worst convergence in p_mis = 95 and 50; some non-conv in conditions where p_mis = 75 and T=1, and no non-conv in conditions where p_mis = 25 or 5.
+We should see this in $\widehat{R}$ and AC: worst convergence in $p_{mis}$ = 95 and 50; some non-convergencein conditions where $p_{mis}$ = 75 and $T$ =1, and no non-convergencein conditions where $p_{mis}$ = 25 or 5.
 The sign of the bias is logical: when there is less info in the data, the relations between the predictors and the outcome is under-estimated.
 
-Rhat behaves oppositely to what we wanted.
-The lowest Rhat values for conditions where T<10 are observed for the two missingness conditions with the worst bias.
-However, the highest Rhat observed is indeed for the missingness condition with the worst bias, p_mis =95.
-But only at T=5 (while bias is no worse than anywhere else in this p_mis condition, conditional on iterations condition T>2).
+$\widehat{R}$ behaves oppositely to what we wanted.
+The lowest $\widehat{R}$ values for conditions where $T$<10 are observed for the two missingness conditions with the worst bias.
+However, the highest $\widehat{R}$ observed is indeed for the missingness condition with the worst bias, $p_{mis}$ =95.
+But only at $T$ =5 (while bias is no worse than anywhere else in this $p_{mis}$ condition, conditional on iterations condition $T$>2).
 So neither of the thresholds works perfectly.
 
-AC shows a similar trend to the biases: more signs of non-conv for higher p_mis and lower T.
+$AC$ shows a similar trend to the biases: more signs of non-convergencefor higher $p_{mis}$ and lower T.
 The threshold, however, performs less well.
 It identifies signs of non-convergence for all conditions where $T=3$, irrespective of p_mis.
-Then nothing, and then for p_mis =95, T>7, and p_mis=50, T >14, and p_mis = 75, T>20.
+Then nothing, and then for $p_{mis}$ =95, $T$>7, and $p_{mis}$ = 50, $T$ >14, and $p_{mis}$ = 75, $T$>20.
 This does not corresond to the bias in $\bar{Q}$.
-Maybe use AC>.5 at T=5 as threshold?
+Maybe use AC>.5 at $T$ =5 as threshold?
 
@@
 
-In short, univariate estimates seem robust against terminating the algorithm early: There is no clear effect of the number of iterations on the bias in these estimates.
-The bias in the Qs only depends on the p_mis.
+In short, univariate estimates seem robust against terminating the algorithm early: There is no clear effect of the number of iterations on the bias in the estimated means and standard deviations.
+The bias in these $Q$s only depends on the $p_{mis}$, with higher $p_{mis}$ resulting in more bias.
 Yet, the convergence diagnostics indicate that the algorithm did not reach a completely converged state (yet).
@@
 With that, we show that MICE can lead to correct outcomes when they have not converged according to two common conv diags.
-This may be due to the methods (and their thresholds) or due to the Qs (descriptives and multivariate linear regression, not higher dimensional/more complex Qs).
-More `complicated' $Q$s (e.g., higher-order effects or variance components) might show bias, under- or over-coverage at higher $T$, as indiated by Rhat and AC.
+This may be due to the methods (and their thresholds) or due to the $Q$s (descriptives and multivariate linear regression, not higher dimensional/more complex $Q$s).
+More `complicated' $Q$s (e.g., higher-order effects or variance components) might show bias, under- or over-coverage at higher $T$, as indiated by $\widehat{R}$ and AC.
 Add what %miss has to do with it.
@@
 We might, however, also formulate a new threshold, specifically for the evaluation of MI algorithms.
-The current study suggests that $\widehat{R} < 1.05$ may be implemented, since that is the level at which the $\widehat{R}$ stabilize (around $T = 20$) (**Not necessary for this Q, but maybe for more complicated Qs**).
+The current study suggests that $\widehat{R} < 1.05$ may be implemented, since that is the level at which the $\widehat{R}$ stabilize (around $T = 20$) (**Not necessary for this Q, but maybe for more complicated $Q$s**).
 -->
@@
 
-- Develop an additional convergence diagnostic unique for iterative imputation: AC across imputations, instead of iterations.
+- Develop an additional convergence diagnostic unique for iterative imputation: $AC$ across imputations, instead of iterations.
 

### 2020-05-02 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Remove comments from Methods section, rewrite first part of Results section to correspond to the newly formulated simulation objectives.

@@
 Therefore, a quantitative, diagnostic evaluation of convergence would be
-preferred. Yet, monitoring convergence of iterative imputation
-algorithms diagnostically is challenging. Iterative imputation
-algorithms such as MICE are Markov chain Monte Carlo (MCMC) methods. In
-MCMC methods, convergence is not from a scalar to a point but from one
+preferred---yet not straightforward. Iterative imputation algorithms
+such as MICE are Markov chain Monte Carlo (MCMC) methods. In MCMC
+methods, convergence is not from a scalar to a point, but from one
 distribution to another. The values generated by the algorithm (e.g.,
-imputed values) will vary even after convergence. Therefore, the aim of
-convergence diagnostics for MCMC methods is not to establish the point
-at which convergence is reached, but to monitor signs of non-convergence
-\citep{hoff09}. Several of such diagnostics exist for MCMC methods, but
-it is not known whether these are appropriate for iterative imputation
-algorithms.
+imputed values) will vary even after convergence. Since MCMC algorithms
+do not reach a unique point at which convergence is established,
+diagnostics we may only identify signs of non-convergence
+\citep{hoff09}. Several of such diagnostics exist, but it is not known
+whether these are appropriate for iterative imputation algorithms.
 
@@
 algorithms, which will aid applied researchers in drawing valid
-inference from incomplete datasets.
-
-\textbf{Include sub-questions!} How can non-convergence be identified
-diagnostically? Are common MCMC non-convergence diagnostics appropriate
-for MICE? And if so, which threshold should be used to diagnose
-non-convergence? How many iterations are sufficient/needed to be able to
-diagnose non-convergence? Are the default number of iterations
-sufficient (i.e., 5 in mice, 10 in SPSS and Stata, 30 in mi)?
-\textbf{How severe is it when the algorithm has not converged? And what
-are guidelines for practice? Can the parameter of interest, estimand
-\(Q\), be correct when the algorithm is not (yet) converged, and vice
-versa?} \emph{(Maybe add these too? What are the effects of continued
-iterations on estimates, predictions and inferences? Do the answers
-differ with varying missingness proportions? That is, we vary the nr of
-iterations and the missingness proportion because we assume that the
-algorithm has not reached convergence at \(t=1\), and performs worse
-with moderate to high missingness (not so much with little or a LOT of
-missingness).)}
+inference from incomplete datasets. \textbf{add societal relevance: what
+is the consequence of non-convergence?}
+
+\textbf{Include sub-questions and add structure of the paper!} How can
+non-convergence be identified diagnostically? Are common MCMC
+non-convergence diagnostics appropriate for MICE? And if so, which
+threshold should be used to diagnose non-convergence? How many
+iterations are sufficient/needed to be able to diagnose non-convergence?
+Are the default number of iterations sufficient (i.e., 5 in mice, 10 in
+SPSS and Stata, 30 in mi)? \textbf{How severe is it when the algorithm
+has not converged? And what are guidelines for practice? Can the
+parameter of interest, estimand \(Q\), be correct when the algorithm is
+not (yet) converged, and vice versa?} \emph{(Maybe add these too? What
+are the effects of continued iterations on estimates, predictions and
+inferences? Do the answers differ with varying missingness proportions?
+That is, we vary the nr of iterations and the missingness proportion
+because we assume that the algorithm has not reached convergence at
+\(t=1\), and performs worse with moderate to high missingness (not so
+much with little or a LOT of missingness).)}
 
@@
 (\(j = 1, 2, \dots, k\)) may be either observed or missing. \textbf{The
-number of units \(i\) with at least one missing data value can be
-divided by the total number of units \(n\) to obtain the `missingness
-proportion' \(p_{mis}\).} The collection of observed data values in
-\(y\) is denoted by \(y_{obs}\); the missing part of \(y\) is referred
-to as \(y_{mis}\). For each datapoint in \(y_{mis}\), we sample
+number of units \(i\) in dataset \(y\) with at least one missing data
+value, divided by the total number of units \(n\), is called the
+missingness proportion in \(p_{mis}\).} The collection of observed data
+values in \(y\) is denoted by \(y_{obs}\); the missing part of \(y\) is
+referred to as \(y_{mis}\). For each datapoint in \(y_{mis}\), we sample
 \(M \times T\) times plausible values, where \(M\) is the number of
@@
 
-The goal of imputation through e.g.~FCS/MICE is to pbtain a multivariate
-distribution by iterating over the sequence of univariate imputations.
-Traditionally, the inspection of the multivariate convergence is done by
-visually inspecting the univariate chain means and variances over the
-iterations, or by tracking the Rhat and autocorrelation. The downside to
-these approaches is that they either focus on the univariate state
-space, or primarily track the change over the iterations of a
-multivariate outcome conform the scientific model of interest. Focusing
-on the convergence of these outcome parameters may influence this
-procedure in the sense that the model of evaluation favors the model of
-interest. Ideally, one would like to evaluate a model-independent
-parameter that summarizes the multivariate nature of the data. We
-propose eigenvalue as such a parameter, because it\ldots.
+\textbf{The goal of imputation through e.g.~MICE is to obtain a
+multivariate distribution by iterating over the sequence of univariate
+imputations. Traditionally, the inspection of the multivariate
+convergence is done by visually inspecting the univariate chain means
+and variances over the iterations, or by tracking the \(\widehat{R}\)
+and autocorrelation. The downside to these approaches is that they
+either focus on the univariate state space, or primarily track the
+change over the iterations of a multivariate outcome conform the
+scientific model of interest. Focusing on the convergence of these
+outcome parameters may influence this procedure in the sense that the
+model of evaluation favors the model of interest. Ideally, one would
+like to evaluate a model-independent parameter that summarizes the
+multivariate nature of the data. We propose eigenvalue as such a
+parameter, because it\ldots. }
 
@@
 iterations in iterative imputation, however, is often much lower.
-Therefore, we compute \(AC\) manually, see panel D. The AC values in
+Therefore, we compute \(AC\) manually, see panel D. The \(AC\) values in
 this plot do meet the requirements (\ldots) and will therefore be used
@@
 \(\widehat{R}\) and \(AC\) perform in identifying the effects of
-non-convergence. And finally, we will formulate an informed advice on
-the requirements to safely assume \emph{sufficient} convergence in
-practice. That is, we assume that the algorithm is \emph{sufficiently}
-converged when each estimate \(\bar{Q}\) is an unbiased and confidence
-valid estimate of the corresponding estimand \(Q\).
+non-convergence (\textbf{add thetas}). And finally, we will formulate an
+informed advice on the requirements to safely assume \emph{sufficient}
+convergence in practice. That is, we assume that the algorithm is
+\emph{sufficiently} converged when each estimate \(\bar{Q}\) is an
+unbiased and confidence valid estimate of the corresponding estimand
+\(Q\). (\textbf{add to formulate advice on thetas?})
 
@@
   over-estimate signs of non-convergence in iterative imputation
-  algorithms. In contrast to this, it may occur that the signs of
-  non-convergence are under-estimated by \(\widehat{R}\), in exceptional
-  cases where the initial values of the algorithm are not appropriately
-  over-dispersed \citep[p.~437]{broo98}. In e.g.~\texttt{mice}, initial
-  values are chosen randomly from the observed data, hence we cannot be
-  certain of over-dispersion in the initial values. In practice, we do
-  not expect this to cause problems for identifying non-convergence with
-  \(\widehat{R}\).
+  algorithms, compared to the bias the estimates. In contrast to this,
+  it may occur that signs of non-convergence are under-estimated by
+  \(\widehat{R}\), in exceptional cases where the initial values of the
+  algorithm are not appropriately over-dispersed \citep[p.~437]{broo98}.
+  In e.g.~\texttt{mice}, initial values are chosen randomly from the
+  observed data, hence we cannot be certain of over-dispersion in the
+  initial values. In practice, we do not expect this to cause problems
+  for identifying non-convergence with \(\widehat{R}\).
 \item
@@
   imputation algorithms with typical convergence. That is, after only a
-  few iterations, the randomness induced by the algorithm will
+  few iterations, the randomness induced by the algorithm should
   effectively mitigate the risk of dependency within chains.
+\item
+  \textbf{Something about thetas}
 \end{enumerate}
@@
  8. combine outcomes of all missingness proportions
-9. aggregate outcomes across simulation runs 
+9. aggregate outcomes of all simulation runs 
 \end{verbatim}
@@
 \] where \(\epsilon \sim \mathcal{N}(0, 100)\). From the complete set,
-we obtain the true values of each quantity of scientific interest \(Q\).
+we obtain the true values of several quantities of scientific interest,
+\(Q\)s. \textbf{MOVE THIS NEXT PART TO TOP OF SIMULATION SECTION?}
+\(Q\)s are the targets that will be estimated in each simulation
+repetition, after first \emph{amputing} the complete data with varying
+missingness proportions, then imputing the missingness with a varying
+number of iterations, then estimating the \(Q\)s on the \(M\) completed
+datasets, and finally pooling the \(M\) \(\hat{Q}\)s to obtain a
+\(\bar{Q}\) for each \(Q\). \textbf{Add that we try to predict the
+effect of non-converegence (i.e., the difference between \(Q\) and
+\(\bar{Q}\)) with the convergence diagnostics \(\widehat{R}\) and
+\(AC\), by appliying these methods to several scalar summaries of the
+state space of the algorithm, \(\theta\)s.}
 
@@
 \(\widehat{R}\))}. The true values of these scientific estimands are
-calculated with the R functions \texttt{base::mean()},
-\texttt{stats::sd()}, and \texttt{stats::lm()} \citep{R}. Each of these
-\(Q\)s is estimated with a corresponding \(\bar{Q}\), after the complete
-dataset is first \emph{amputed} (with a varying missingness proportion
-\(p_{mis}\)) and subsequently imputed (with a varying number of
-iterations \(T\)).
+calculated on the complete data with the R functions
+\texttt{base::mean()}, \texttt{stats::sd()}, and \texttt{stats::lm()}
+\citep{R}.
 
@@
 
-The complete dataset is amputed with the \texttt{mice} package
-\citep[function \texttt{mice::ampute()};][]{mice}. All missingness is
-multivariate, and conform a `missing completely at random' missingness
-mechanism \citep{rubin87}. I.e., the probability to be missing is the
-same for all \(N \times k\) cells in \(y\), conditional on the
-missingness proportion (\(p_{mis} =.05,.25,.5,.75,.95\)).
-
-Missing datapoints in \(y\) are imputed with the \texttt{mice()}
-function \citep{mice}. All imputation procedures are performed with
-Bayesian linear regression imputation, and five imputation chains
-(\(M=5\)). The number of iterations varies between simulation conditions
+In each simulation repetition, we ampute the complete dataset five times
+to obtain five different incomplete datasets, \(y\)s. The ratio between
+\(y_{obs}\) and \(y_{mis}\) in each \(y\) depends on the missingness
+proportion: \(p_{mis} =.05,.25,.5,.75,.95\). We ampute the complete data
+with the \texttt{mice} package \citep[function
+\texttt{mice::ampute()};][]{mice}. All missingness is multivariate, and
+conform a `missing completely at random' missingness mechanism
+\citep{rubin87}. I.e., the probability to be missing is the same for all
+\(N \times k\) cells in \(y\), conditional on the missingness proportion
+\(p_{mis}\).
+
+Missing datapoints in the incomplete datasets, \(y\)s, are imputed with
+the \texttt{mice()} function \citep{mice}. All imputation procedures are
+performed with Bayesian linear regression imputation, and five
+imputation chains (\(M=5\)). Because \(M=5\), each run of the
+\texttt{mice()} function results in five sets of imputations,
+\(y_{imp,m}\)s, where \(m = 1, 2, ..., 5\). The number of iterations in
+each run of the algorithm varies between simulation conditions
 (\(T=1,2,...,100\)).
 
-After each run of the \texttt{mice} algorithm, the function
-\texttt{mice::complete()} is used to combine the imputed data
-\(y_{imp,m}\) with the observed data \(y_{obs}\), resulting in five
-completed datasets \{\(y_{obs}, y_{imp, m}\)\}, where
-\(m = 1, 2, ..., 5\).
+From the 5 imputations \(y_{imp,m}\): get chain means and chain
+variances from imputation object to use as scalar summary \(\theta\) in
+convergence diagnostics. Then use \texttt{mice::complete()} to combine
+the imputed data \(y_{imp,m}\) with the observed data \(y_{obs}\),
+resulting in five completed datasets \{\(y_{obs}, y_{imp, m}\)\}.
+
+On the 5 completed datasets \{\(y_{obs}, y_{imp, m}\)\}, we:
 
@@
 \item
-  on the 5 imputations \(y_{imp,m}\): get mean and var within imp chains
-  as theta to compute conv diag --\textgreater{} then complete with
-  \texttt{mice::complete()}
-\item
-  on the 5 completed datasets \{\(y_{obs}, y_{imp, m}\)\}: 1) estimate
-  Qhats for mu and sigma --\textgreater{} pool as mean()
-  --\textgreater{} get Qbars of mu and sigma; 2) compute lambda as theta
-  for conv diag; 3) perform lm to get Qhats for betas and r\^{}2
-  --\textgreater{} use Qhats of beta as theta to compute conv diag
-  --\textgreater{} pool Qhats of beta and r\^{}2 to get Qbars
+  Estimate \(\hat{Q}\)s for \(Q= \mu\) and \(Q=\sigma\) with
+  \texttt{base::mean()}, and \texttt{stats::sd()} \citep{R}. Then pool
+  the \(\hat{Q}\)s with \texttt{base::mean()} to get \(\bar{Q}\)s.
+\item
+  Compute \(\lambda_1\), to use as \(\theta\) in convergence
+  diagnostics. By definition, the first eigenvalue of the
+  variance-covariance matrix, \(\lambda_1\), is equal to the variance of
+  the principal component solution of each \{\(y_{obs}, y_{imp, m}\)\},
+  which we calculate with \texttt{stats::princomp()} \citep{R}.
+\item
+  Perform multiple linear regression with \texttt{stats::lm()} \citep{R}
+  to get the \(\hat{Q}\)s for \(Q=\beta\) and \(Q=r^2\). Pool the
+  \(\hat{Q}\)s conform \citet{vink14} to get \(\bar{Q}\)s for the
+  \(\beta\)s and \(r^2\) (\textbf{check pooling of \(r^2\), instead of
+  \texttt{mice::pool.r.squared()}}). Additionally, use the \(\hat{Q}\)s
+  of \(Q=\beta\) as \(\theta\) to compute convergence diagnostics.
 \end{itemize}
 
-The estimator for each \(Q\) is \(\bar{Q}\)---the pooled aggregate of
-the \(\hat{Q}\)s across imputations. To calculate the \(\bar{Q}\)s, we
-combine the observed data \(y_{obs}\) and the imputed data for each
-imputation \(y_{imp,m}\) with the function \texttt{mice::complete()}.
-Descriptive statistics are computed as the average across imputations
-for each variable (\(\mu_j\) and \(\sigma_j\), where
-\(j = Y, X_1, X_2, X_3\)). Estimated regression coefficients are
-obtained with the function \texttt{stats::lm()} for each imputation, and
-then pooled conform \citet{vink14}. The coefficient of determination is
-estimated for each imputation, and pooled using
-\texttt{mice::pool.r.squared()}.
-
-For each run of the imputation algorithm, we compute non-convergence
-diagnostics \(\widehat{R}\) and \(AC\). We calculate \(\widehat{R}\) by
-implementing Vehtari et al.'s \citeyearpar{veht19} recommendations, and
-\(AC\) as the correlation between the \(t^{th}\) and the \((t+1)^{th}\)
-iteration. The two methods are applied to four scalar summaries
-\(\theta\): chain means (i.e., the mean in each \(y_{imp, m, j}\)),
-chain variances (i.e., the variance in each \(y_{imp, m, j}\)), a
-statistic of scientific interest \(\hat{Q}\) in each
-\{\(y_{obs}, y_{imp, m}\)\}, and the first eigenvalue of the
-variance-covariance matrix in each \{\(y_{obs}, y_{imp, m}\)\},
-\(\lambda_1\).
-
-The \(\widehat{R}\) and \(AC\) values with chain means as \(\theta\)s
-are evaluated against the bias in univariate mean estimates. Similarly,
-the performance measure for \(\widehat{R}\) and \(AC\) applied to chain
-variances is the bias in estimated standard deviations. To evaluate the
-performance of \(\widehat{R}\) and \(AC\) on the eigenvalues, we will
-use bias in the coefficient of determination, \(R^2\). Additionally, we
-will evaluate the estimated regression coefficients, the coverage rate
-of the CI95\% of the regression estimates, and the CI length
-(\textbf{see definitions in old Methods}). \textbf{remove the term
-performance measure here and add an extra paragraph to define
-performance measures as bias in all estimands, and coverage rate and CI
-length of regression coefficients.}
+We apply non-convergence diagnostics \(\widehat{R}\) and \(AC\) to each
+scalar summary \(\theta\). We calculate \(\widehat{R}\) by implementing
+Vehtari et al.'s \citeyearpar{veht19} recommendations, and \(AC\) as the
+correlation between the \(t^{th}\) and the \((t+1)^{th}\) iteration.
 
@@
 
-The impact of inducing non-convergence in the iterative imputation
-algorithm is assessed by comparing \(\bar{Q}\)s with \(Q\)s. For each
-\(Q\), we compute bias as \(\bar{Q} - Q\). For the regression
+We assess the impact of inducing non-convergence in the iterative
+imputation algorithm by comparing \(\bar{Q}\)s with \(Q\)s. For each
+\(Q\), we calculate bias as \(\bar{Q} - Q\). For the regression
 coefficients \(Q=\beta_{1,2,3}\), we also compute the empirical coverage
@@
 degrees of freedom, and \(SE_{\bar{Q}}\) is the square root of the
-pooled variance estimate. Since under-estimating the variance of
-\(\bar{Q}\) may yield spurious inferences, we also inspect CI width
-(CIW): the difference between the lower and upper bound of the 95\%
-confidence interval around \(\bar{Q}\).
-
-The performance of two methods is evaluated against the bias and
-confidence validity of the estimated \(Q\)s.
+pooled variance estimate. If we obtain nominal coverage (CR = 95\%), we
+can conclude that the \(\bar{Q}\)s are conficence valid estimates of the
+\(Q\)s. Finally, we inspect CI width (CIW): the difference between the
+lower and upper bound of the 95\% confidence interval around
+\(\bar{Q}\). CIW is of interest because too short CIs indicate
+under-estimation of the variance in \(\bar{Q}\), which may yield
+spurious inferences.
+
+With these performance measures, we also evaluate how well the
+non-convergence diagnostics \(\widehat{R}\) and \(AC\) can identify
+simulation conditions in which the \(\bar{Q}\)s are \emph{not} unbiased,
+confidence valid estimates of the \(Q\)s. \textbf{Rephrase and add
+corresponding \(Q\)s:} The two methods are applied to four scalar
+summaries \(\theta\): chain means (i.e., the mean in each
+\(y_{imp, m, j}\)), chain variances (i.e., the variance in each
+\(y_{imp, m, j}\)), the estimated regression coefficients
+\(\beta_{1,2,3}\) in each \{\(y_{obs}, y_{imp, m}\)\} (i.e., a
+user-defined statistic of scientific interest \(\hat{Q}\)), and the
+first eigenvalue of the variance-covariance matrix in each
+\{\(y_{obs}, y_{imp, m}\)\}, \(\lambda_1\).
 
@@
 
-We want to know what the effect is of inducing non-conv of the bias in
-all Qs and the confidence validity of the betas. So we plot the bias for
-each Q against the number of iterations and the missingness proportion
-in the condition. We look at averages across simulation repetitions. Is
-there more bias in conditions where non-conv was induced? (NOPE for
-univariates, YES for multivar.).
-
-Then, we want to know if conditions in which the Qs are affected by
-non-convergence (on average) are also identified by the conv diag. We do
-this by looking at the bias in Q against the conv diag applied to
-related thetas. Is the point at which there is no more bias the same
-point at which the conv diag indicate minimal signs of non-conv? (NOPE,
-they are too conservative).
-
-Is one of the thetas better? (NAH, doesn't really matter).
-
-What should the thresholds be? (LOWER than recommended) And the default
-nr of iterarions? (MAYBE increase to 10?)
-
 \textbf{Add to this section: 1) emphasize that the plots are averages
 across repetitions, not within MICE; 2) sample effects due to single
-complete dataset. 3) Add more info about figure legends and axes.)}
+complete dataset?? 3) Add more info about figure legends and axes.)}
+
+Primarily, we want to know what the effect is of inducing
+non-convergence on the estimated \(Q\)s. Specifically, we are interested
+in the bias of \(\bar{Q}\) for all \(Q\)s and the confidence validity of
+\(\bar{Q}\) for \(Q=\beta\). Therefore, we plot these performance
+measures against the simulation conditions (the number of iterations
+\(T\) and the missingness proportion \(p_{mis}\)). We look at averages
+across simulation repetitions.
+
+Secondly, we want to know if conditions in which the \(Q\)s are affected
+by non-convergence are also identified by the non-convergance
+diagnostics \(\widehat{R}\) and \(AC\). We do this, again, by looking at
+the performance measures for the \(Q\)s, but now we evaluate these
+against what \(\widehat{R}\) and \(AC\) indicate about signs of non-the
+convergence in the \(\theta\)s.
+
+\begin{itemize}
+\item
+  Is there more bias in conditions where non-convergence was induced?
+  (NOPE for univariates, YES for multivar.).
+\item
+  Is the point of \emph{sufficient} convergence (at which the
+  \(\bar{Q}\)s are unbiased and confidence valid estimates of the
+  \(Q\)s) the same point as identified by the non-convergence
+  diagnostics? (NOPE, they are too conservative).**
+\item
+  Is one of the thetas better? (NAH, doesn't really matter).
+\item
+  What should the thresholds be? (LOWER than recommended).
+\item
+  And the default nr of iterarions? (MAYBE increase to 10?).
+\end{itemize}
 
 For reasons of brevity, we only discuss the convergence diagnostics for
-the \(Q\) with the worst performance in terms of bias. For the
-descriptve statistics, the magnitude of the bias was the largest in
-\(Y\) (i.e, \(j = Y\) in \(\mu_j\) and \(\sigma_j\)). For the
-\(Q=\beta_{1,2,3}\), \(\beta_1\) showed the worst perfomance (i.e, the
-effect of \(X_1\) on \(Y\). Moreover, we only show conditions where
-\(T\leq50\) in the figures. Full results are available from
-\textbf{{[}link github repo{]}}.
+the \(Q\)s with the worst performance in terms of bias. Bias in the
+descriptve statistics (\(Q=\mu\) and \(Q=\sigma\)) was most pronounced
+in the outcome variable \(Y\). Of the regression coefficients,
+\(Q=\beta\), the largest bias was observed in the effect of \(X_1\) on
+\(Y\), \(\beta_1\). Results for \(Q=r^2\) are reported as-is. In the
+figures, however, we only present simulation conditions where
+\(T\leq50\). Full results are available from \textbf{{[}link github
+repo{]}}.
 
@@
 As expected, conditions with a higher proportion of missingness and/or a
-lower number of iterations show more signs of non-convergence, as
-indicated by more extreme bias in the estimated \(Q\)s. Roughly speaking
-this means that MICE is indeed not converged as \(t=1\), and converges
-gradually as \(t\) increases. The point at which an additional iteration
-does not lead to an improvement of the estimates depends on the
-difficulty of the missingness problem---between XYZ and XYZ. This means
-that the algorithm has converged sufficiently (\textbf{under the current
-specifications}).
+lower number of iterations show more signs of non-convergence. I.e., on
+average these conditions portet more extreme bias in the estimated
+\(Q\)s, and non-nominal coverages. Roughly speaking this means that MICE
+is indeed not converged at \(t=1\), and converges gradually as \(t\)
+increases to \(T\). The point at which an additional iteration does not
+lead to an improvement of the estimates depends on the difficulty of the
+missingness problem---between XYZ and XYZ. This means that the algorithm
+has converged sufficiently (\textbf{under the current specifications}).
 
@@
 \(\widehat{R}\) follow a trend similar to the performance measures.
+\textbf{Add split between convergence diagnostics versus performance
+measures, and convergence diagnostics versus simulation conditions.}
 \(\widehat{R}\) values are generally lower in conditions with a higher
@@
 percentage of missing data. Autocorrelation values were indeed
-decreasing with a higher number of iterations.
+decreasing with a higher number of iterations, and lower in conditions
+with less missingness.
 
@@
 
-\textbf{Message: after one iteration (T=2), the algorithm is as well
+\textbf{Message: after one iteration (\(T\)=2), the algorithm is as well
 converged as the missingness proportion allows. Sooner even with lower
-missingness proportions (T=1 might be ok for p\_mis=5 and 25).} Number
-of it has little influence on bias in hat\{Q\}, only in first iteration
-some bias in missingness conditions of 25 and up. There is a clear
-different between the missingness conditions. The higher the missingness
-prop, the larger the magnitude of the bias. For this Q, t=2 would be
-sufficient. And only for p\_mis = 5 the bar\{Q\}s are unbiased. All
-other p\_mis conditions have bias irrespective of the number of
-iterations.
-
-Rhat differs a lot across iteration conditions. But not structurally for
-missingness proportions. Highest Rhats for the condition t=2, while the
-most bias is for (t=1), but there is no Rhat for that by definition. But
-for missingness, 25\% seems the most signs of non-conv, while it has one
-of the least bias. Threshold of 1.2 reached in condition it = 3, p\_mis
-= 5 and 75, and then for it\textgreater6, p\_mis= 5 and 50, and
-it\textgreater10 for all p\_mis. Threshold 1.1 reached in in it=12,
-p\_mis = all but 75, 1.01 only reached in it = 93, p\_mis = 75. While
-for this Q, t=2 would be sufficient.
-
-AC mainly shows effects of the set of iteration conditions. AC indicates
-less signs of non-conv in conditions with higher number of iterations.
-For iteration conditions where t\textless10, there are also some effects
-of the missingness proportions. The p\_mis condtions 25 and 95 \%
-indicate the most signs of non-convergence. For p\_mis = 95\% this
-corresponds to the bias in Qbar. For 25\% miss it doesn't. The
-thresholds defined by the critical value is only met in the conditions
-it=3, p\_mis = 5,75, and 95\%.
+missingness proportions (\(T\)=1 might be ok for \(p_{mis}\) = 5 and
+25).} Number of it has little influence on bias in \(\hat{Q}\), only in
+first iteration some bias in missingness conditions of 25 and up. There
+is a clear different between the missingness conditions. The higher the
+missingness prop, the larger the magnitude of the bias. For this \(Q\),
+t=2 would be sufficient. And only for \(p_{mis}\) = 5 the \(\bar{Q}\)s
+are unbiased. All other \(p_{mis}\) conditions have bias irrespective of
+the number of iterations.
+
+\(\widehat{R}\) differs a lot across iteration conditions. But not
+structurally for missingness proportions. Highest \(\widehat{R}\)s for
+the condition \(T\)=2, while the most bias is for (\(T\)=1), but there
+is no \(\widehat{R}\) for that by definition. But for missingness, 25\%
+seems the most signs of non-conv, while it has one of the least bias.
+Threshold of 1.2 reached in condition it = 3, \(p_{mis}\) = 5 and 75,
+and then for it\textgreater6, \(p_{mis}\) = 5 and 50, and
+\(T\)\textgreater10 for all \(p_{mis}\). Threshold 1.1 reached in in
+\(T\)=12, \(p_{mis}\) = all but 75, 1.01 only reached in \(T\) = 93,
+\(p_{mis}\) = 75. While for this \(Q\), \(T\)=2 would be sufficient.
+
+\(AC\) mainly shows effects of the set of iteration conditions. \(AC\)
+indicates less signs of non-convergencein conditions with higher number
+of iterations. For iteration conditions where \(T\)\textless10, there
+are also some effects of the missingness proportions. The \(p_{mis}\)
+condtions 25 and 95 \% indicate the most signs of non-convergence. For
+\(p_{mis}\) = 95\% this corresponds to the bias in Qbar. For 25\% miss
+it doesn't. The thresholds defined by the critical value is only met in
+the conditions it=3, \(p_{mis}\) = 5,75, and 95\%.
 
@@
 
-\textbf{Message: even with possible effect of initial values (T=1), the
-algorithm is as well converged as the missingness proportion allows.}
-For \(Q=\sigma_Y\) there is no effect visible of the iteration
-conditions, but clearly for the p\_mis conditions. Most bias for p\_mis
-= 95, followed by p\_mis=50 and 75. For the least possible bias we may
-use any number of iterations \(T\geq1\), but none of the p\_mis results
-in unbiased bar\{Q\}s.
-
-There is a clear effect of the number of iterations on Rhat. Conditions
-with lower Ts have more signs of non-conv, as identified by Rhat. There
-is little difference between missingness conditions. Only for conditions
-where \(T<5\) the missingness proportions seem to affect whether Rhat
-indicates non-conv. The bias in bar\{Q\} seems unrelated to the
-magnitude of Rhat. Conditions with more bias (p\_mis = 50,75,95) are not
-identified to have more signs of non-convergence. The thresholds are
-therefore useless (because depend mostly on t, not so much p\_mis).
-
-AC is affected by both t and p\_mis conditions. But conditions with
-worst bias (p\_mis = 50, 75, 95) are not identified as worst converging
-according to AC. Moreover, only for T=3 the AC is significantly
-different from zero.
+\textbf{Message: even with possible effect of initial values (\(T\)=1),
+the algorithm is as well converged as the missingness proportion
+allows.} For \(Q=\sigma_Y\) there is no effect visible of the iteration
+conditions, but clearly for the \(p_{mis}\) conditions. Most bias for
+\(p_{mis}\) = 95, followed by \(p_{mis}\) = 50 and 75. For the least
+possible bias we may use any number of iterations \(T\geq1\), but none
+of the \(p_{mis}\) results in unbiased \(\bar{Q}\)s.
+
+There is a clear effect of the number of iterations on \(\widehat{R}\).
+Conditions with lower Ts have more signs of non-conv, as identified by
+\(\widehat{R}\). There is little difference between missingness
+conditions. Only for conditions where \(T<5\) the missingness
+proportions seem to affect whether \(\widehat{R}\) indicates
+non-convergence The bias in \(\bar{Q}\) seems unrelated to the magnitude
+of \(\widehat{R}\). Conditions with more bias (\(p_{mis}\) = 50,75,95)
+are not identified to have more signs of non-convergence. The thresholds
+are therefore useless (because depend mostly on \(T\), not so much
+\(p_{mis}\)).
+
+\(AC\) is affected by both t and \(p_{mis}\) conditions. But conditions
+with worst bias (\(p_{mis}\) = 50, 75, 95) are not identified as worst
+converging according to AC. Moreover, only for \(T\) =3 the \(AC\) is
+significantly different from zero.
 
@@
 
-\textbf{Message: after four iterations (T=5), the algorithm is as well
-converged as the missingness proportion allows. Sooner even with lower
-missingness proportions (T=1 might be ok for p\_mis=5 and 25).} The
-figure for \(Q=\beta_1\) shows that the \(\bar{Q}\)s are affected by
-both the number of iterations and the missingness proportion. For T=1,
-the magnitude of the bias (almost) follows the order of the missingness
-proportions. More missingness causes more bias in the \(\bar{Q}\). In
-general, a lower nr of iterations results in more bias, but this depends
-on the missingness proportion. For p\_mis=95, there is more bias in
-conditions where T\textless6. When p\_mis = 50, it's T\textless3. For
-p\_mis = 75 and 25, it's only for T\textless2. And conditions where
-p\_mis = 5 are unaffected on average. However, none of the conditions
-result in completely unbiased estimates of \(Q\).
-
-We should see this in the conv diag: higher Rhat and AC for p\_mis=75
-and 95, and for condtions where \(T<2-6\), depending on p\_mis.
-
-Rhat depends somewhat on the missingness proportion. For conditions
-where T\textless10, the Rhat values follow the p\_mis more or less. But
-what we actually want to know is whether they are indicators of bias.
-Conditions with the worst bias (high p\_mis, low T) have the largest
-Rhat values. The thresholds are not appropriate to diagnose sufficient
-convergence. All thresholds over-estimate the signs of non-convergence
-compared to the bias in Qbar. However, a higher threshold than 1.2 may
-result in terminating the algorithm after 3 iterations, which is
-insufficient for the p\_mis = 95 condition. Threshold of 1.25 or 1.3
-could work for the other p\_mis conditions.
-
-AC is very affected by the p\_mis. The two conditions with the most bias
-(p\_mis = 75 and 95) are the only ones identified as non-converging, as
-they should. However, the p\_mis = 75 condition is only identified in
-the T=3 condtion, while the bias is the least pronounced there. And the
-AC only identifies non-convergence in the p\_mis = 95 conditions where
-T\textgreater10. In the default situation, T=5, this would be
-overlooked. Also, the AC shows the opposite of what you would expect
-from a theta that is based on the observed data: shouldn't there be
-higher autocorrelations in conditions with comparatively more observed
-data in the completed datasets? I don't understand this.
-
-\textbf{Add how interesting it is because we literally track Q and are
-still not ok.}
+\textbf{Message: after four iterations (\(T\)=5), the algorithm is as
+well converged as the missingness proportion allows. Sooner even with
+lower missingness proportions (\(T\)=1 might be ok for \(p_{mis}\) = 5
+and 25).} The figure for \(Q=\beta_1\) shows that the \(\bar{Q}\)s are
+affected by both the number of iterations and the missingness
+proportion. For \(T\) =1, the magnitude of the bias (almost) follows the
+order of the missingness proportions. More missingness causes more bias
+in the \(\bar{Q}\). In general, a lower nr of iterations results in more
+bias, but this depends on the missingness proportion. For \(p_{mis}\) =
+95, there is more bias in conditions where \(T\)\textless6. When
+\(p_{mis}\) = 50, it's \(T\)\textless3. For \(p_{mis}\) = 75 and 25,
+it's only for \(T\)\textless2. And conditions where \(p_{mis}\) = 5 are
+unaffected on average. However, none of the conditions result in
+completely unbiased estimates of \(Q\).
+
+We should see this in the conv diag: higher \(\widehat{R}\) and \(AC\)
+for \(p_{mis}\) = 75 and 95, and for condtions where \(T<2-6\),
+depending on \(p_{mis}\).
+
+\(\widehat{R}\) depends somewhat on the missingness proportion. For
+conditions where \(T\)\textless10, the \(\widehat{R}\) values follow the
+\(p_{mis}\) more or less. But what we actually want to know is whether
+they are indicators of bias. Conditions with the worst bias (high
+\(p_{mis}\), low T) have the largest \(\widehat{R}\) values. The
+thresholds are not appropriate to diagnose sufficient convergence. All
+thresholds over-estimate the signs of non-convergence compared to the
+bias in Qbar. However, a higher threshold than 1.2 may result in
+terminating the algorithm after 3 iterations, which is insufficient for
+the \(p_{mis}\) = 95 condition. Threshold of 1.25 or 1.3 could work for
+the other \(p_{mis}\) conditions.
+
+\(AC\) is very affected by the p\_mis. The two conditions with the most
+bias (\(p_{mis}\) = 75 and 95) are the only ones identified as
+non-converging, as they should. However, the \(p_{mis}\) = 75 condition
+is only identified in the \(T\) =3 condtion, while the bias is the least
+pronounced there. And the \(AC\) only identifies non-convergence in the
+\(p_{mis}\) = 95 conditions where \(T\)\textgreater10. In the default
+situation, \(T\) =5, this would be overlooked. Also, the \(AC\) shows
+the opposite of what you would expect from a theta that is based on the
+observed data: shouldn't there be higher autocorrelations in conditions
+with comparatively more observed data in the completed datasets? I don't
+understand this.
+
+\textbf{Add how interesting it is because we literally track \(Q\) and
+are still not ok.}
 
@@
 
-\textbf{Message: after two iterations (T=3), the algorithm is as well
-converged as the missingness proportion allows. Sooner even with lower
-missingness proportions (T=1 might be ok for p\_mis=5 and 25).} Bias is
-the worst for conditions where p\_mis = 95, T\textless3. Followed by
-p\_mis = 75, T=1; and p\_mis = 50, T=1. Conditions where p\_mis = 25 or
-5 seem unaffected by the number of iterations. Minimal bias is obtained
-in all iteration conditions. We should see this in Rhat and AC: worst
-convergence in p\_mis = 95 and 50; some non-conv in conditions where
-p\_mis = 75 and T=1, and no non-conv in conditions where p\_mis = 25 or
-5. The sign of the bias is logical: when there is less info in the data,
-the relations between the predictors and the outcome is under-estimated.
-
-Rhat behaves oppositely to what we wanted. The lowest Rhat values for
-conditions where T\textless10 are observed for the two missingness
-conditions with the worst bias. However, the highest Rhat observed is
-indeed for the missingness condition with the worst bias, p\_mis =95.
-But only at T=5 (while bias is no worse than anywhere else in this
-p\_mis condition, conditional on iterations condition T\textgreater2).
-So neither of the thresholds works perfectly.
-
-AC shows a similar trend to the biases: more signs of non-conv for
-higher p\_mis and lower T. The threshold, however, performs less well.
-It identifies signs of non-convergence for all conditions where \(T=3\),
-irrespective of p\_mis. Then nothing, and then for p\_mis =95,
-T\textgreater7, and p\_mis=50, T \textgreater14, and p\_mis = 75,
-T\textgreater20. This does not corresond to the bias in \(\bar{Q}\).
-Maybe use AC\textgreater.5 at T=5 as threshold?
+\textbf{Message: after two iterations (\(T\)=3), the algorithm is as
+well converged as the missingness proportion allows. Sooner even with
+lower missingness proportions (\(T\)=1 might be ok for \(p_{mis}\) = 5
+and 25).} Bias is the worst for conditions where \(p_{mis}\) = 95, \(T\)
+\textless3. Followed by \(p_{mis}\) = 75, \(T\) =1; and \(p_{mis}\) =
+50, \(T\) = 1. Conditions where \(p_{mis}\) = 25 or 5 seem unaffected by
+the number of iterations. Minimal bias is obtained in all iteration
+conditions. We should see this in \(\widehat{R}\) and AC: worst
+convergence in \(p_{mis}\) = 95 and 50; some non-convergencein
+conditions where \(p_{mis}\) = 75 and \(T\) =1, and no non-convergencein
+conditions where \(p_{mis}\) = 25 or 5. The sign of the bias is logical:
+when there is less info in the data, the relations between the
+predictors and the outcome is under-estimated.
+
+\(\widehat{R}\) behaves oppositely to what we wanted. The lowest
+\(\widehat{R}\) values for conditions where \(T\)\textless10 are
+observed for the two missingness conditions with the worst bias.
+However, the highest \(\widehat{R}\) observed is indeed for the
+missingness condition with the worst bias, \(p_{mis}\) =95. But only at
+\(T\) =5 (while bias is no worse than anywhere else in this \(p_{mis}\)
+condition, conditional on iterations condition \(T\)\textgreater2). So
+neither of the thresholds works perfectly.
+
+\(AC\) shows a similar trend to the biases: more signs of
+non-convergencefor higher \(p_{mis}\) and lower T. The threshold,
+however, performs less well. It identifies signs of non-convergence for
+all conditions where \(T=3\), irrespective of p\_mis. Then nothing, and
+then for \(p_{mis}\) =95, \(T\)\textgreater7, and \(p_{mis}\) = 50,
+\(T\) \textgreater14, and \(p_{mis}\) = 75, \(T\)\textgreater20. This
+does not corresond to the bias in \(\bar{Q}\). Maybe use
+AC\textgreater.5 at \(T\) =5 as threshold?
 
@@
 algorithm early: There is no clear effect of the number of iterations on
-the bias in these estimates. The bias in the Qs only depends on the
-p\_mis. Yet, the convergence diagnostics indicate that the algorithm did
-not reach a completely converged state (yet).
+the bias in the estimated means and standard deviations. The bias in
+these \(Q\)s only depends on the \(p_{mis}\), with higher \(p_{mis}\)
+resulting in more bias. Yet, the convergence diagnostics indicate that
+the algorithm did not reach a completely converged state (yet).
 
@@
 not converged according to two common conv diags. This may be due to the
-methods (and their thresholds) or due to the Qs (descriptives and
-multivariate linear regression, not higher dimensional/more complex Qs).
-More `complicated' \(Q\)s (e.g., higher-order effects or variance
-components) might show bias, under- or over-coverage at higher \(T\), as
-indiated by Rhat and AC. Add what \%miss has to do with it.
+methods (and their thresholds) or due to the \(Q\)s (descriptives and
+multivariate linear regression, not higher dimensional/more complex
+\(Q\)s). More `complicated' \(Q\)s (e.g., higher-order effects or
+variance components) might show bias, under- or over-coverage at higher
+\(T\), as indiated by \(\widehat{R}\) and AC. Add what \%miss has to do
+with it.
 
@@
   Develop an additional convergence diagnostic unique for iterative
-  imputation: AC across imputations, instead of iterations.
+  imputation: \(AC\) across imputations, instead of iterations.
 \end{itemize}

### 2020-05-03 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
Remove or rewrite parts in bold face until results

@@
 This technique is known as 'multiple imputation' [MI; @rubin76].
-MI has proven to be a powerful technique to yield unbiased and confidence valid estimates of the true---but missing---data inference under many circumstances [@buur18].
+MI has proven to be a powerful technique to draw valid inferences from incomplete data under many circumstances [@buur18].
 
-Figure \ref{fig:diagram} provides an overview of the steps involved with MI---from incomplete data, to $M$ multiply imputed datasets, to $M$ estimated quantities of interest $\hat{Q}$s, to a single pooled estimate $\bar{Q}$.
-Missing data in dataset $y$ is imputed $M$ times.
+Figure \ref{fig:diagram} provides an overview of the steps involved with MI.
+Missing data $y_{mis}$ in an incomplete dataset is imputed $M$ times.
 The imputed data $y_{imp}$ is combined with the observed data $y_{obs}$ to create $M$ completed datasets.
-On each completed dataset, the analysis of scientific interest is performed.
+On each completed dataset, an analysis of scientific interest is performed.
 The quantity of scientific interest (e.g., a regression coefficient) is denoted with $Q$.
@@
 These $M$ values are combined into a single pooled estimate $\bar{Q}$.
+The premise of multiple imputation is that $\bar{Q}$ is an unbiased and confidence valid estimate of the true---but missing---data inference.
 
@@
 
-```{r diagram, out.width="\\linewidth", include=TRUE, fig.cap="Scheme of the main steps in multiple imputation."}
+```{r diagram, out.width="\\linewidth", include=TRUE, fig.cap="Scheme of the main steps in multiple imputation (where $M=3$)---from an incomplete dataset, to $M$ multiply imputed datasets, to $M$ estimated quantities of scientific interest $\\hat{Q}$s, to a single pooled estimate $\\bar{Q}$."}
 knitr::include_graphics("./images/diagram.pdf") #adapted from [@buur18] par. 1.4.1.
@@
 A popular method to obtain imputations is to use the 'Multiple Imputation by Chained Equations' algorithm, shorthand 'MICE'[@mice].
-MICE is an iterative algorithmic procedure to draw imputations from the posterior predictive distribution of the missing values.
-This introduces a potential threat to the validity of the imputations: What if the algorithm has not converged?
+With MICE, imputed values $y_{imp}$ are drawn from the posterior predictive distribution of the missing values $y_{mis}$.
+The algorithm is named after the iterative algorithmic procedure by which imputed values are generated: a multivariate distribution is obtained by iterating over a sequence of univariate imputations.
+The iterative nature of algorithms like MICE introduces a potential threat to the validity of the imputations: What if the algorithm has not converged?
 Are the implications then to be trusted?
@@
 These are all open questions, because the convergence properties of iterative imputation algorithms have not been systematically studied [@buur18].
-Moreover, there is no scientific consensus on how to evaluate convergence of MI algorithms [@taka17].
+Moreover, there is no scientific consensus on how to evaluate convergence of imputation algorithms [@taka17].
 Some default MICE techniques (e.g., 'predictive mean modeling') might not yield converged states at all [@murr18].
@@
 
-Currently, the recommended practice for evaluating convergence in iterative imputation algorithms is to visually inspect imputations for signs of non-convergence.
-**Add that this refers to chain means and variances already?
-It's insufficient that these are univariate, so multivariate state space of the algorithm may not be converged when univariates are ok.
-And @buur18's suggestion to track user-defined statistics (e.g., regression coefficient) may be somewhat advanced for empirical researchers.
-Moreover, this is not model-independent, i.e., only apply to one substantive model, while one of the advantages of MI is that missing data problem and scientific problem are split.
-Then we can say 'we propose a novel...' and integrate this: Note that usually we only evaluate the convergence of univariate scalar summaries (e.g., chain means or variances).
-With these we cannot diagnose convergence of multivariable statistics (i.e., relations between scalar summaries).
-Van Buuren [-@buur18, $\S$ 4.5.2] proposed to implement multivariable evaluation of the MICE algorithm through eigenvalue decomposition building on the work of @mack03.
-Eigenvalues of a covariance matrix are a measure of the data's covariance.**.
-
-This method is insufficient on two counts: 1) it may be challenging to the untrained eye, and 2) only severely pathological cases of non-convergence may be diagnosed [@buur18, $\S$ 6.5.2].
-Therefore, a quantitative, diagnostic evaluation of convergence would be preferred---yet not straightforward.
+Currently, the recommended practice for evaulating the convergence of iterative imputation algorithms is through visual inspection.
+That is, the state space of the algorithm is monitored by plotting a scalar summary $\theta$ against the iteration number.
+**[Segway to WHAT TO MONITOR?
+Also, this paragraph is more or less the same as the third paragraph in the second section.
+Merge or remove one of the two?]** Typically, the $\theta$s under evaluation are chain means and variances (**define!**).
+These $\theta$s may be insufficient because they are univariate summaries of the state space of the algorithm, while MICE is not only concerned with univariate estimates, but the entire multivariate distribution of $y_{imp}$.
+A suggestion by @buur18 for a multivariate $\theta$ to monitor is the quantity of scientific interest $Q$ (e.g., a regression coefficient).
+Implementing this, however, might be somewhat advanced for empirical researchers.
+Moreover, this scalar summary is not model-independent, i.e., it only applies to one model of scientific interest.
+Yet, one of the advantages of MI is that the missing data problem and scientific problem are solved independently.
+Therefore, this $\theta$ is not sufficient either.
+Van Buuren [-@buur18, $\S$ 4.5.2] also proposed multivariable evaluation of the MICE algorithm through eigenvalue decomposition, building on the work of @mack03.
+Eigenvalues of a variance-covariance matrix are a measure of the data's total covariance.
+This would be a model-independent, multivariate scalar summary to monitor, but it has not yet been implemented.
+<!-- Recommendation: "the different streams should be freely intermingled with one another, without showing any definite trends". -->
+**[Segway to HOW TO MONITOR?]** Monitoring convergence visually is insufficient on two counts: 1) it may be challenging to the untrained eye, and 2) only severely pathological cases of non-convergence may be diagnosed [@buur18, $\S$ 6.5.2].
+Therefore, a quantitative, diagnostic evaluation of convergence would be preferred---although not straightforward.
 Iterative imputation algorithms such as MICE are Markov chain Monte Carlo (MCMC) methods.
@@
 The values generated by the algorithm (e.g., imputed values) will vary even after convergence.
-Since MCMC algorithms do not reach a unique point at which convergence is established, diagnostics we may only identify signs of non-convergence [@hoff09].
-Several of such diagnostics exist, but it is not known whether these are appropriate for iterative imputation algorithms.
+Since MCMC algorithms do not reach a unique point at which convergence is established, diagnostics  may only identify signs of *non*-convergence [@hoff09].
+Several of such non-convergence diagnostics exist, but it is not known whether these are appropriate for iterative imputation algorithms.
 <!-- "a weak diagnostics is better than no diagnostic at all" [@cowl96]. -->
 
-In this paper, we investigate how non-convergence in iterative imputation algorithms may be diagnosed, how well these methods perform, and at which point convergence may safely be assumed.
-For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular `mice` package [@mice] in `R` [@R].
-The convergence properties of the MICE algorithm are investigated through model-based simulation.
-The results of this simulation study are guidelines for assessing convergence of MI algorithms, which will aid applied researchers in drawing valid inference from incomplete datasets.
-**add societal relevance: what is the consequence of non-convergence?**
+In this paper, we investigate how non-convergence in iterative imputation algorithms may be diagnosed.
+First, we define which non-convergence diagnostics will be considered, and evaluate how these diagnostics could be appropriate for iterative imputation applications.
+In the second part of this paper, we will investigate the impact of inducing non-convergence in iterative imputation algorithms by means of model-based simulation in `R` [@R].
+The aim of the simulation is to determine whether unbiased, confidence valid inferences may be obtained under different levels of non-convergence (**and at which point convergence may safely be assumed**).
+Additionally, we will assess how well different non-convergence diagnostics perform in identifying signs of non-convergence.
+The results of this study are guidelines for identifying non-convergence in iterative imptation algorithms, which will aid applied researchers in drawing valid inference from incomplete datasets.
+**[add more societal relevance: what is the consequence of non-convergence?
+e.g., biased estimates, invalid inference, etc.]** For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular `mice` package [@mice] in `R` [@R].
 
-**Include sub-questions and add structure of the paper!** How can non-convergence be identified diagnostically?
+**[Include sub-questions and make structure of the paper clear!]** How can non-convergence be identified diagnostically?
 Are common MCMC non-convergence diagnostics appropriate for MICE?
@@
 Are the default number of iterations sufficient (i.e., 5 in mice, 10 in SPSS and Stata, 30 in mi)?
-**How severe is it when the algorithm has not converged?
+How severe is it when the algorithm has not converged?
 And what are guidelines for practice?
-Can the parameter of interest, estimand $Q$, be correct when the algorithm is not (yet) converged, and vice versa?** *(Maybe add these too?
-What are the effects of continued iterations on estimates, predictions and inferences?
-Do the answers differ with varying missingness proportions?
-That is, we vary the nr of iterations and the missingness proportion because we assume that the algorithm has not reached convergence at $t=1$, and performs worse with moderate to high missingness (not so much with little or a LOT of missingness).)*
+Can the parameter of interest $Q$ be correct when the algorithm is not (yet) converged, and vice versa?
 
@@
 The data value of unit $i$ ($i = 1, 2, \dots, n$) on variable $j$ ($j = 1, 2, \dots, k$) may be either observed or missing.
-**The number of units $i$ in dataset $y$ with at least one missing data value, divided by the total number of units $n$, is called the missingness proportion in $p_{mis}$.** The collection of observed data values in $y$ is denoted by $y_{obs}$; the missing part of $y$ is referred to as $y_{mis}$.
-For each datapoint in $y_{mis}$, we sample $M \times T$ times plausible values, where $M$ is the number of imputations ($m = 1, 2, \dots, M$) and $T$ is the number of iterations ($t = 1, 2, \dots, T$).
+The number of units $i$ in dataset $y$ with at least one missing value, divided by the total number of units $n$, is called the missingness proportion $p_{mis}$.
+The collection of observed data values in $y$ is denoted by $y_{obs}$; the missing part of $y$ is referred to as $y_{mis}$.
+For each datapoint in $y_{mis}$, we sample $M \times T$ plausible values, where $M$ is the number of imputations ($m = 1, 2, \dots, M$) and $T$ is the number of iterations ($t = 1, 2, \dots, T$) in the imputation algorithm.
 The collection of samples between the initial value (at $t=1$) and the final imputed value (at $t=T$) will be referred to as an 'imputation chain'.
-**Add: $\theta$s are scalar summaries of interest in the iterative algorithm (e.g., chain means; the average of the imputed values in each imputation chain).
+**$\theta$s are scalar summaries of the algorithm at a certain iteration (e.g., chain means; the average of the imputed values in each imputation chain).
 **
@@
 
-**The goal of imputation through e.g.
-MICE is to obtain a multivariate distribution by iterating over the sequence of univariate imputations.
-Traditionally, the inspection of the multivariate convergence is done by visually inspecting the univariate chain means and variances over the iterations, or by tracking the $\widehat{R}$ and autocorrelation.
+There are two requirements for convergence of iterative algorithms: mixing and stationarity [@gelm13].
+Without mixing, imputation chains do not intermingle nicely, indicating that the distribution of imputed values differs across imputations.
+Chains may be 'stuck' at a local optimum, instead of sampling imputed values from the entire predictive posterior distribution of the missing values.
+This may cause under-estimation of the variance within chains, which results in spurious, invalid inferences.
+Without stationarity, there is trending within imputation chains.
+Trending implies that further iterations would yield a different set of imputations.
+Iterative imputation algorithms that have not yet reached stationarity, may thus yield biased estimates.
+
+<!-- **The goal of imputation through e.g.
+MICE is to obtain a multivariate distribution by iterating over a sequence of univariate imputations.
+Traditionally, the inspection of the multivariate convergence is done by visually inspecting the univariate chain means and variances over the iterations.
 The downside to these approaches is that they either focus on the univariate state space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest.
@@
 We propose eigenvalue as such a parameter, because it....
-**
+** -->
 
-Mixing and stationarity have histroically been inspected visually, by evaluating traceplots of scalar summaries of interest ($\theta$s; e.g., chain means and chain variances).
-As @buur18 describes, users can also specify a model-specific scalar summary (e.g., a regression coefficient).
-A user-specified scalar summary, however, is not universal to all complete data problems.
-Therefore, as inspired by [@mack03], we propose a $\theta$ that summarizes the multivariate state space of the algorithm.
-Namely, the first eigenvalue of the variance-covariance matrix of the $M$ completed datasets.
-**The first eigenvalue has the appealing property that is not dependent on the substantive model of interest.**
+**[Move to INTRO???]** To illustrate what non-convergence looks like in iterative imputation algorithms, we reproduce an example from van Buuren [-@buur18, $\S$ 6.5.2].
+Figure \ref{fig:non-conv} displays a subset of the example: a variable that we will call $j$.
+We see two traceplots of the chain means for $j$---i.e., the average of the imputed values for $j$ in each imputation $y_{imp,m}$, plotted against the iteration number.
+Each line portrays one imputation.
+The plot on the lefthand side of the figure shows typical convergence of an iterative imputation algorithm.
+The righthand side displays pathological non-convergence, induced by purposefully mis-specifying the imputation model.
+**[Leave this in if it's after mixing and stationarity are defined, otherwise remove?] In the typical convergence situation, the imputation chains intermingle nicely and there is little to no trending.
+In the non-convergence plot, there is a lot of trending and some chains do not intermingle.** Importantly, the chain means at the last iteration (the imputed value per $m$) are very different between the two plots.
+The algorithm with the mis-specified model yields imputed values that are on average a factor two larger than those of the typically converged algorithm.
+That means that non-convergence has a impact on the mean of $j$ in $y_{imp, m}$.
+This difference (presumably) translates to bias in the $M$ completed data estimates $\hat{\mu}_{j,m}$, and consequently also to the pooled estimate $\bar{\mu}_j$.
+Non-convergence thus leads to biased estimates.
+This shows the importance of reaching converged states in iterative imputation algorithms.
 
-**Since convergence of iterative imputation algorithms is in distribution, there is not a unique point at which the algorithm reaches a converged state.
+```{r non-conv, out.width='.49\\linewidth', fig.width=3.5, fig.height=3,fig.show='hold',fig.cap="Typical convergence versus pathological non-convergence.
+Please note that this is the same data with a different imputation model, leading to different imputations (e.g., see the units on the y-axis)."}
+load("./images/nonc_plot.Rdata")
+load("./images/conv_plot.Rdata")
+par(mar = c(2, 2, 0.1, 0.1))
+plot(conv)
+plot(nonc)
+```
+
+**[Move to INTRO???]** Convergence has historically been inspected visually, by monitoring the imputation algorithm over iterations.
+This is typically done through traceplots.
+In a traceplot, the iteraton number is plotted against a certain $\theta$.
+$\theta$s are scalar summaries of the state space of the algorithm at a specific iteration.
+The recommended scalar summaries to evaluate are chain means and chain variances.
+Additionally, researchers may "monitor some statistic of scientific interest" [@buur18, $\S$ 6.5.2].
+As @buur18 describes, researchers can specify their quantity of scientific interest $Q$ (e.g., a regression coefficient) as scalar summary $\theta$.
+Such a user-defined scalar summary, however, may be somewhat advanced for empirical researchers.
+And moreover, it is not universal to all complete data problems.
+Focusing on the convergence of outcome parameters may influence the iterative imputation procedure in the sense that the model of evaluation favors the model of interest.
+The downside to these two approaches is that they either focus on the univariate state space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest.
+Ideally, one would like to evaluate a model-independent parameter that summarizes the multivariate nature of the data.
+Therefore, we propose a $\theta$ that summarizes the multivariate state space of the algorithm, but is independent from the model of scientific interest.
+We propose $\lambda_1$ as such a scalar summary.
+We define $\lambda_1$ as the first eigenvalue of the variance-covariance matrix of the completed data.
+Let $\lambda_1 \geq \lambda_2 \geq ...
+\geq \lambda_j$ be the eigenvalues of $\Sigma$ in each of the $M$ completed datasets ${y_{obs}, y_{imp}}$.
+$\lambda_1$ is measure that summarises the covariances in the completed datasets.
+**The first eigenvalue has the appealing property that is not dependent on the substantive model of interest.
+Eigenvalue decomposition is inspired by @mack03.**
+
+<!-- **Since convergence of iterative imputation algorithms is in distribution, there is not a unique point at which the algorithm reaches a converged state.
 We can only evaluate scalar summaries of the multivariate state space of the algorithm, $\theta$s.
@@
 \geq \lambda_j$ be the eigenvalues of $\Sigma$ in each of the $M$ completed datasets ${y_{obs}, y_{imp}}$.
-$\lambda_1$ is measure that summarises the covariances in the completed datasets.**
+$\lambda_1$ is measure that summarises the covariances in the completed datasets.**  -->
 
-There are two requirements for convergence of iterative algorithms: mixing and stationarity [@gelm13].
-Without mixing, imputation chains do not intermingle nicely, **indicating that ...**.
-Without stationarity, there is trending within imputation chains, which implies that further iterations would yiled a different set of imputations.
-
-To illustrate what non-convergence looks like in MI, we reproduce the example from van Buuren [-@buur18, $\S$ 6.5.2].
-Figure \ref{fig:non-conv} shows the traceplot for one of the variables in the example.
-we see the average of the imputed values for a variable $j$ in $y_{imp}$.
-The average imputed values in the left hand plot are about a magnitude 2 lower than the right hand plot.
-So non-convergence has a impact on the average imputed value for $j$ in $y_{imp, m}$.
-This difference (presumably) translates to bias in the pooled estimate $\bar{\mu}_j$ as well.
-And maybe also to higher order statistics, e.g.
-variances, covariances, etc.
-Therefore, it's important to converge.
-**Explain what we see, namely example by @buur18 reproduced, showing the traceplots of chain means for some variable.
-The first plot is typical convergence of MICE, the second is pathological non-convergence because of a mis-specified imputation model.
-Each line is an imputation.
-In the first plot, the chains intermingle nicely and there is little to no trending.
-In the second plot, there is a lot of trending and some chains do not intermingle.
-Importantly, the chain means at the last iteration (the imputed value per $m$) are very different between the two plots.
-The algorithm with the mis-specified model yields imputed values that are on average a magnitude two larger than those of the typically converged algorithm.
-This shows the importance of reaching converged states in iterative imputation algorithms.**
-
-## Methods under evaluation
+## Diagnostics under evaluation
 
@@
 
+Non-convergence diagnostics for MCMC algorithms typically identify problems in either one of the two requirements for converegnce: mixing or stationarity.
 Non-stationarity within chains may be diagnosed with e.g., autocorrelation [$AC$; @scha97; @gelm13], numeric standard error ['MC error'; @gewe92], or Raftery and Lewis's [-@raft91] procedure to determine the effect of trending on the precision of estimates.
@@
 With a recently proposed adaptation, $\widehat{R}$ might also serve to diagnose non-stationarity, but this has not yet been thoroughly investigated [@veht19].
-Therefore, use we $\widehat{R}$ and $AC$ to evaluate mixing and stationarity separately, as recommended by e.g., @cowl96.
+Therefore, we use $\widehat{R}$ and $AC$ to evaluate mixing and stationarity separately, as recommended by e.g., @cowl96.
 <!-- @cowl96, p. 898. -->
-
-```{r non-conv, out.width='.49\\linewidth', fig.width=3.5, fig.height=3,fig.show='hold',fig.cap="Typical convergence versus pathological non-convergence.
-Please note that this is the same data with a different imputation model, leading to different imputations (see the y axis!).
-Wel min of meer dezelfde origin op iteration 1."}
-load("./images/nonc_plot.Rdata")
-load("./images/conv_plot.Rdata")
-par(mar = c(2, 2, 0.1, 0.1))
-plot(conv)
-plot(nonc)
-
-```
 
@@
 
-An updated version of $\widehat{R}$ has been proposed by @veht19 (p.
-5).
-**This version may be suitable for iterative imputation.** Let $M$ be the total number of chains, $T$ the number of iterations per chain, and $\theta$ the scalar summary of interest (e.g., chain mean or chain variance).
+An updated version of $\widehat{R}$ has been proposed by Vehtari et al.
+[-@veht19, p.
+5].
+**This version may be suitable for iterative imputation.** Let $M$ be the total number of chains, $T$ the number of iterations per chain, and $\theta$ the scalar summary of interest (e.g., chain means).
 For each chain ($m = 1, 2, \dots, M$), we estimate the variance of $\theta$, and average these to obtain within-chain variance $W$.
@@
 As convergence diagnostic, the interest is therefore in positive $AC$-values.
-**Maybe remove:** Moreover, the magnitude of $AC$-values may be evaluated statistically, but that is outside of our scope.
+The magnitude of $AC$-values may be evaluated statistically [**add reference**].
+
+### Thresholds
+
+<!-- We assume that complete convergence, defined as $\widehat{R} = 1$ and $AC = 0$, will not be observed.  -->
+Upon convergence, $\widehat{R}=1$ and $AC=0$, which are unlikely thresholds for MCMC algorithms, because of its convergence to a distribution.
+Even in the most converged state, the algorithm will show some signs of non-mixing and non-stationarity.
+Upon *sufficient* convergence, the imputation chains will intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm ($\widehat{R} > 1$), and there may be some dependency between subsequent iterations of imputation chains ($AC > 0$).
+In practice, non-convergence is usually diagnosed when $\widehat{R}$ > 1.2 or 1.1 or even 1.01.
+And a t-test is performed to assess whether $AC$ is significantly different from zero.
 
@@
 
-Upon convergence, $\widehat{R}=1$ and $AC=0$, which are unlikely thresholds for MCMC algorithms, because of its convergence to a distribution.
-In practice, non-convergence is usually diagnosed when $\widehat{R}$ > 1.2 or 1.1 or even 1.01.
-And a t-test is performed to assess whether $AC$ is significantly different from zero.
+Before we evaluate the performance of the non-convergence diagnostics $\widehat{R}=1$ and $AC=0$ quantitatively through simulation, we apply them to the example of pathological non-convergence by @buur18 that we reproduced above.
+We want to be able to draw the same conclusions as we did from visually inspecting the two algorithms.
+But $\widehat{R}=1$ and $AC=0$ should at least be able to distinguish between the two algorithms plotted in Figure \ref{fig:non-conv}.
+From visual inspection, we know that the typically converged algorithm initially portrayed some signs of non-mixing (around $t=2$), but intermigled nicely overall.
+Additionally, there was very little trending.
+The algorithm with pathological non-convergence showed severe non-mixing, although this gradually improved (from $t=7$).
+Moreover, there was a lot of trending as well initially (up-to $t=6$), after which the chains reached a somewhat more stationary state.
+To assess whether $\widehat{R}$ and $AC$ may be appropriate non-convergence identifiers for iterative imputation algorithms, they should identify the same trends as the visual inspection.
+But at least, the following condition must hold: the methods should indicate worse performance for the mis-specified model with pathological non-convergence than the typically converged algorithm (i.e., higher $\widehat{R}$- and $AC$-values).
 
-We assume that complete convergence, defined as $\widehat{R} = 1$ and $AC = 0$, will not be observed.
-Because even in the most converged state, the algorithm will show some signs of non-mixing and non-stationarity.
-Upon *sufficient* convergence, the imputation chains will intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm ($\widehat{R} > 1$), and there may be some dependency between subsequent iterations of imputation chains ($AC > 0$).
-
-**Before using them in the simulation, the two methods show at least be able to distinguish between the two algorithms plotted in Figure 2.
-From visual inspection we know that the typical convergence has some signs of non-mixing around iteration 2, and little trending.
-In the non-convergence situation, there is a lot of trending upto iteration 6, after which the chains reach a somewhat more stationary state.
-Mixing in this situation gets worse from the first iteration onward, and gradually gets better around iteration 7.
-To assess whether $\widehat{R}$ and $AC$ may be appropriate non-convergence identifiers for iterative imputation algorithms, the following condition must hold: the methods indicate worse performance for the mis-specified model with pathological non-convergence (i.e., higher $\widehat{R}$- and $AC$-values than the typical performance).
-And additionally the methods should reflect the increasing convergence in the typical convergence situation as the number of iterations goes up.**
-
-**In Figure \ref{fig:diagnostics}A, the chain means from Figure \ref{fig:non-conv} are plotted again, now together---as are the non-convergence diagnostics.
-Panel B shows $\widehat{R}$ as computed by implementing @veht19 's recommendations.
-As required, $\widehat{R}$ indicates less signs of non-convergence as the number of iterations goes up **in the typical convergence situation**.
-The superior performance of the typical convergence over the pathological non-convergence is less prominent, even flipped for $t<4$.
-Panel C displays the $AC$ as computed with the R function `stats::acf()`.
-When we look at this panel, we conclude something weird.
-The $AC$-values indicate equal performance (up-to $t=5$) for the typical convergence and the pathological non-convergence, while there is obvious trending in the latter.
-Moreover, the best convergence (as indicated by the lowest $AC$-value) is observed at $t=2$, but looking at the chain means in panel A, there should be some signs of trending up-to iteration number seven.
-After consulting the documentation on stats::acf() [@R], we conclude that this implementation of $AC$ is not suitable for iterative imputation algorithms.
-There is a correction factor for a mathematical shortcut that works in the limit.
-According to @box15, the function works when the number of iterations > 50.
-The default number of iterations in iterative imputation, however, is often much lower.
-Therefore, we compute $AC$ manually, see panel D.
-The $AC$ values in this plot do meet the requirements (...) and will therefore be used in the simulation study.**
-
-**acf is niet bedoeld om weinig iteraties te... niet een fout in acf, alleen hier niet van toepassing**
+In Figure \ref{fig:diagnostics}A, the chain means from Figure \ref{fig:non-conv} are plotted again---now together.
+Panel B shows two versions of calculating $AC$s applied to the $\theta$s of panel A: the default calculation with `R` function `stats::acf()` [@R], and manual calculation as the correlation between $\theta$ in iteration $t$ and $\theta$ in iteration $t+1$.
+Panel C shows the traditional computation of $\widehat{R}$, and panel D shows $\widehat{R}$ as computed by implementing @veht19 's recommendations.
 
@@
 
+When we look at panel B, we conclude something weird.
+The $AC$-values calculated with the default function indicate equal performance for the typical convergence and the pathological non-convergence (up-to $t=5$), while there is obvious trending in the $\theta$s of the latter.
+Moreover, the best convergence (as indicated by the lowest $AC$-value) is observed at $t=2$, but looking at the chain means in panel A, there should be some signs of trending up-to iteration number seven.
+After consulting the documentation on `stats::acf()` [@R], we conclude that this $AC$ function is not suited for iterative imputation algorithms.
+The function is optimized for performance when $t\geq50$ [@box15], while the default number of iterations in iterative imputation is often much lower.
+Therefore, we compute $AC$ manually, see the solid line.
+These $AC$ values indicate non-stationarity as expected.
+We therefore calculate $AC$ manually.
+
+**Describe panel C here and why it is not optimal]** The $\widehat{R}$-values in panel D do meet the requirements specified above.
+$\widehat{R}$ as computed conform @veht19 indeed indicate less signs of non-convergence as the number of iterations goes up.
+<!-- However, the superior performance of the typical convergence over the pathological non-convergence is less prominent, even flipped for $t<4$.  -->
+
 # Simulation study
@@
 The aim of the simulation study is to evaluate the impact of inducing non-convergence in the MICE algorithm on several quantities of scientific interest $Q$.
-Subsequently we will evaluate how well $\widehat{R}$ and $AC$ perform in identifying the effects of non-convergence (**add thetas**).
+Subsequently we will evaluate how well $\widehat{R}$ and $AC$ perform in identifying the effects of non-convergence **[and performance of different thetas]**.
 And finally, we will formulate an informed advice on the requirements to safely assume *sufficient* convergence in practice.
 That is, we assume that the algorithm is *sufficiently* converged when each estimate $\bar{Q}$ is an unbiased and confidence valid estimate of the corresponding estimand $Q$.
-(**add to formulate advice on thetas?**)
+**[and formulate advice on thetas?]**
 
@@
 In an empirical study, where $\widehat{R}$ was used to inform the required imputation chain length, it took as many as 50 iterations to overcome the conventional non-convergence threshold $\widehat{R}>1.2$.
-Yet, scientific estimates were insensitive to continued iteration after $t>5$ [@lace07].
+Yet, scientific estimates were insensitive to continued iteration from $t=5$ onwards [@lace07].
 We therefore suspect that $\widehat{R}$ may over-estimate signs of non-convergence in iterative imputation algorithms, compared to the bias the estimates.
@@
 From the complete set, we obtain the true values of several quantities of scientific interest, $Q$s.
-**MOVE THIS NEXT PART TO TOP OF SIMULATION SECTION?** $Q$s are the targets that will be estimated in each simulation repetition, after first *amputing* the complete data with varying missingness proportions, then imputing the missingness with a varying number of iterations, then estimating the $Q$s on the $M$ completed datasets, and finally pooling the $M$ $\hat{Q}$s to obtain a $\bar{Q}$ for each $Q$.
-**Add that we try to predict the effect of non-converegence (i.e., the difference between $Q$ and $\bar{Q}$) with the convergence diagnostics $\widehat{R}$ and $AC$, by appliying these methods to several scalar summaries of the state space of the algorithm, $\theta$s.**
+**[MOVE THIS NEXT PART TO TOP OF SIMULATION SECTION?]** $Q$s are the targets that will be estimated in each simulation repetition, after first *amputing* the complete data with varying missingness proportions, then imputing the missingness with a varying number of iterations, then estimating the $Q$s on the $M$ completed datasets, and finally pooling the $M$ estimates $\hat{Q}$ to obtain a single $\bar{Q}$ for each $Q$.
+**We try to predict the effect of non-converegence (i.e., the difference between $Q$ and $\bar{Q}$) with the convergence diagnostics $\widehat{R}$ and $AC$, by appliying these methods to several scalar summaries of the state space of the algorithm, $\theta$s.**
 
@@
 And finally, the proportion of variance explained by the set of predictors: coefficient of determination $r^2 \times 100$ **(note that lower case $r$ is used to avoid confusion with non-convergence diagnostic $\widehat{R}$)**.
-The true values of these scientific estimands are calculated on the complete data with the R functions `base::mean()`, `stats::sd()`, and `stats::lm()` [@R].
+The true values of these scientific estimands are calculated on the complete data with the `R` functions `base::mean()`, `stats::sd()`, and `stats::lm()` [@R].
 <!-- Each of these $Q$s is estimated with a corresponding $\bar{Q}$, after the complete dataset is first *amputed* (with a varying missingness proportion $p_{mis}$) and subsequently imputed (with a varying number of iterations $T$).
@@
 
-In each simulation repetition, we ampute the complete dataset five times to obtain five different incomplete datasets, $y$s.
+In each simulation repetition, we *ampute* the complete dataset five times to obtain five different incomplete datasets, $y$s.
 The ratio between $y_{obs}$ and $y_{mis}$ in each $y$ depends on the missingness proportion: $p_{mis} =.05,.25,.5,.75,.95$.
 We ampute the complete data with the `mice` package [function `mice::ampute()`; @mice].
-All missingness is multivariate, and conform a 'missing completely at random' missingness mechanism [@rubin87].
+We consider all possible univariate and multivariate patterns of missingness, conform a 'missing completely at random' missingness mechanism [@rubin87].
 I.e., the probability to be missing is the same for all $N \times k$ cells in $y$, conditional on the missingness proportion $p_{mis}$.
@@
 
-Missing datapoints in the incomplete datasets, $y$s, are imputed with the `mice()` function [@mice].
+Missing datapoints in each incomplete dataset $y$ are imputed with the `mice()` function [@mice].
 All imputation procedures are performed with Bayesian linear regression imputation, and five imputation chains ($M=5$).
-Because $M=5$, each run of the `mice()` function results in five sets of imputations, $y_{imp,m}$s, where $m = 1, 2, ..., 5$.
+Because $M=5$, each run of the `mice()` function results in five sets of imputations: $y_{imp,m}$, where $m = 1, 2, ..., 5$.
 The number of iterations in each run of the algorithm varies between simulation conditions ($T=1,2,...,100$).
@@
 
-From the 5 imputations $y_{imp,m}$: get chain means and chain variances from imputation object to use as scalar summary $\theta$ in convergence diagnostics.
-Then use `mice::complete()` to combine the imputed data $y_{imp,m}$ with the observed data $y_{obs}$, resulting in five completed datasets {$y_{obs}, y_{imp, m}$}.
+From the 5 imputations $y_{imp,m}$ we obtain the chain means and chain variances for each variable $j$.
+These will serve as scalar summary $\theta$ to apply non-convergence diagnostics $\widehat{R}$ and $AC$ on.
+We then use `mice::complete()` to combine the imputed data $y_{imp,m}$ with the observed data $y_{obs}$, resulting in five completed datasets {$y_{obs}, y_{imp, m}$}.
 
@@
 
-- Estimate $\hat{Q}$s for $Q= \mu$ and $Q=\sigma$ with `base::mean()`, and `stats::sd()` [@R]. Then pool the $\hat{Q}$s with `base::mean()` to get $\bar{Q}$s.
+- Estimate $\hat{Q}$s for $Q= \mu$ and $Q=\sigma$ with `base::mean()`, and `stats::sd()` [@R]. We pool the $\hat{Q}$s with `base::mean()` to get $\bar{Q}$s.
 
-- Compute $\lambda_1$, to use as $\theta$ in convergence diagnostics.
+- Compute $\lambda_1$, to use as $\theta$ in the two convergence diagnostics.
 By definition, the first eigenvalue of the variance-covariance matrix, $\lambda_1$, is equal to the variance of the principal component solution of each {$y_{obs}, y_{imp, m}$}, which we calculate with `stats::princomp()` [@R].
@@
 - Perform multiple linear regression with `stats::lm()` [@R] to get the $\hat{Q}$s for $Q=\beta$ and $Q=r^2$.
-Pool the $\hat{Q}$s conform @vink14 to get $\bar{Q}$s for the $\beta$s and $r^2$ (**check  pooling of  $r^2$, instead of `mice::pool.r.squared()`**).
-Additionally, use the $\hat{Q}$s of $Q=\beta$ as $\theta$ to compute convergence diagnostics.
+We pool the $\hat{Q}$s conform @vink14 to get $\bar{Q}$s for the $\beta$s and $r^2$ **[check  pooling of  $r^2$, instead of `mice::pool.r.squared()`]**.
+Additionally, we use the $\hat{Q}$s of $Q=\beta$ as $\theta$ to apply the convergence diagnostics to.
 
@@
 With these performance measures, we also evaluate how well the non-convergence diagnostics $\widehat{R}$ and $AC$ can identify simulation conditions in which the $\bar{Q}$s are *not* unbiased, confidence valid estimates of the $Q$s.
-**Rephrase and add corresponding $Q$s:** The two methods are applied to four scalar summaries $\theta$: chain means (i.e., the mean in each $y_{imp, m, j}$), chain variances (i.e., the variance in each $y_{imp, m, j}$), the estimated regression coefficients $\beta_{1,2,3}$ in each {$y_{obs}, y_{imp, m}$} (i.e., a user-defined statistic of scientific interest $\hat{Q}$), and the first eigenvalue of the variance-covariance matrix in each {$y_{obs}, y_{imp, m}$}, $\lambda_1$.
+**Rephrase and add corresponding $Q$s:** We measure the performance of $\widehat{R}$ and $AC$ by the four scalar summaries $\theta$: chain means (i.e., the mean in each $y_{imp, m, j}$) versus $Q=\mu$, chain variances (i.e., the variance in each $y_{imp, m, j}$) versus $Q=\sigma$, the estimated regression coefficients $\beta$ in each {$y_{obs}, y_{imp, m}$} (i.e., a user-defined statistic of scientific interest $\hat{Q}$) against $Q=\beta$, and the first eigenvalue of the variance-covariance matrix in each {$y_{obs}, y_{imp, m}$}, $\lambda_1$ versus $Q=r^2$.
 
@@
 
-**Add to this section: 1) emphasize that the plots are averages across repetitions, not within MICE; 2) sample effects due to single complete dataset??
-3) Add more info about figure legends and axes.)**
-
+**[Add to this section: 1) emphasize that the iterations in the plots are averages across repetitions, not within one run of MICE; 2) sample effects due to single complete dataset??
+3) Add more info about figure legends and axes.]**
 Primarily, we want to know what the effect is of inducing non-convergence on the estimated $Q$s.
@@
 Secondly, we want to know if conditions in which the $Q$s are affected by non-convergence are also identified by the non-convergance diagnostics $\widehat{R}$ and $AC$.
-We do this, again, by looking at the performance measures for the $Q$s, but now we evaluate these against what $\widehat{R}$ and $AC$ indicate about signs of non-the convergence in the $\theta$s.
+We do this, again, by looking at the performance measures for the $Q$s, but now we evaluate these against what $\widehat{R}$ and $AC$ indicate about signs of non-convergence in the $\theta$s.
 
+**[In short:]**
 - Is there more bias in conditions where non-convergence was induced? (NOPE for univariates, YES for multivar.).
@@
 - Is the point of *sufficient* convergence (at which the $\bar{Q}$s are unbiased and confidence valid estimates of the $Q$s) the same point as identified by the non-convergence diagnostics?
-(NOPE, they are too conservative).**
+(NOPE, they are too conservative).
 
-- Is one of the thetas better? (NAH, doesn't really matter).
+- Is one of the thetas better? (NAH, doesn't really matter, probably multivariate is better).
 
@@
 As expected, conditions with a higher proportion of missingness and/or a lower number of iterations show more signs of non-convergence.
-I.e., on average these conditions portet more extreme bias in the estimated $Q$s, and non-nominal coverages.
+I.e., on average these conditions portray more extreme bias in the estimated $Q$s, and non-nominal coverages.
 Roughly speaking this means that MICE is indeed not converged at $t=1$, and converges gradually as $t$ increases to $T$.
@@
 
+Uitsmijter: To conclude, in the case of drawing inference from incomplete data, convergence of the iterative imputation algorithm is often a convenience but not a necessity.
+
 # Grading

### 2020-05-03 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Remove or rewrite parts in bold face until results

@@
 is known as `multiple imputation' \citep[MI;][]{rubin76}. MI has proven
-to be a powerful technique to yield unbiased and confidence valid
-estimates of the true---but missing---data inference under many
-circumstances \citep{buur18}.
+to be a powerful technique to draw valid inferences from incomplete data
+under many circumstances \citep{buur18}.
 
 Figure \ref{fig:diagram} provides an overview of the steps involved with
-MI---from incomplete data, to \(M\) multiply imputed datasets, to \(M\)
-estimated quantities of interest \(\hat{Q}\)s, to a single pooled
-estimate \(\bar{Q}\). Missing data in dataset \(y\) is imputed \(M\)
+MI. Missing data \(y_{mis}\) in an incomplete dataset is imputed \(M\)
 times. The imputed data \(y_{imp}\) is combined with the observed data
 \(y_{obs}\) to create \(M\) completed datasets. On each completed
-dataset, the analysis of scientific interest is performed. The quantity
+dataset, an analysis of scientific interest is performed. The quantity
 of scientific interest (e.g., a regression coefficient) is denoted with
@@
 separate \(\hat{Q}\)-values are obtained. These \(M\) values are
-combined into a single pooled estimate \(\bar{Q}\).
+combined into a single pooled estimate \(\bar{Q}\). The premise of
+multiple imputation is that \(\bar{Q}\) is an unbiased and confidence
+valid estimate of the true---but missing---data inference.
 
@@
 
-\caption{Scheme of the main steps in multiple imputation.}\label{fig:diagram}
+\caption{Scheme of the main steps in multiple imputation (where $M=3$)---from an incomplete dataset, to $M$ multiply imputed datasets, to $M$ estimated quantities of scientific interest $\hat{Q}$s, to a single pooled estimate $\bar{Q}$.}\label{fig:diagram}
 \end{figure}
@@
 Imputation by Chained Equations' algorithm, shorthand
-`MICE'\citep{mice}. MICE is an iterative algorithmic procedure to draw
-imputations from the posterior predictive distribution of the missing
-values. This introduces a potential threat to the validity of the
-imputations: What if the algorithm has not converged? Are the
-implications then to be trusted? And can we rely on the inference
-obtained on the completed data? These are all open questions, because
-the convergence properties of iterative imputation algorithms have not
-been systematically studied \citep{buur18}. Moreover, there is no
-scientific consensus on how to evaluate convergence of MI algorithms
-\citep{taka17}. Some default MICE techniques (e.g., `predictive mean
-modeling') might not yield converged states at all \citep{murr18}.
-Therefore, algorithmic convergence should be monitored carefully.
-
-Currently, the recommended practice for evaluating convergence in
-iterative imputation algorithms is to visually inspect imputations for
-signs of non-convergence. \textbf{Add that this refers to chain means
-and variances already? It's insufficient that these are univariate, so
-multivariate state space of the algorithm may not be converged when
-univariates are ok. And \citet{buur18}`s suggestion to track
-user-defined statistics (e.g., regression coefficient) may be somewhat
-advanced for empirical researchers. Moreover, this is not
-model-independent, i.e., only apply to one substantive model, while one
-of the advantages of MI is that missing data problem and scientific
-problem are split. Then we can say 'we propose a novel\ldots{}' and
-integrate this: Note that usually we only evaluate the convergence of
-univariate scalar summaries (e.g., chain means or variances). With these
-we cannot diagnose convergence of multivariable statistics (i.e.,
-relations between scalar summaries). Van Buuren \citeyearpar[\(\S\)
-4.5.2]{buur18} proposed to implement multivariable evaluation of the
-MICE algorithm through eigenvalue decomposition building on the work of
-\citet{mack03}. Eigenvalues of a covariance matrix are a measure of the
-data's covariance.}.
-
-This method is insufficient on two counts: 1) it may be challenging to
-the untrained eye, and 2) only severely pathological cases of
-non-convergence may be diagnosed \citep[\(\S\) 6.5.2]{buur18}.
-Therefore, a quantitative, diagnostic evaluation of convergence would be
-preferred---yet not straightforward. Iterative imputation algorithms
-such as MICE are Markov chain Monte Carlo (MCMC) methods. In MCMC
-methods, convergence is not from a scalar to a point, but from one
-distribution to another. The values generated by the algorithm (e.g.,
-imputed values) will vary even after convergence. Since MCMC algorithms
-do not reach a unique point at which convergence is established,
-diagnostics we may only identify signs of non-convergence
-\citep{hoff09}. Several of such diagnostics exist, but it is not known
-whether these are appropriate for iterative imputation algorithms.
+`MICE'\citep{mice}. With MICE, imputed values \(y_{imp}\) are drawn from
+the posterior predictive distribution of the missing values \(y_{mis}\).
+The algorithm is named after the iterative algorithmic procedure by
+which imputed values are generated: a multivariate distribution is
+obtained by iterating over a sequence of univariate imputations. The
+iterative nature of algorithms like MICE introduces a potential threat
+to the validity of the imputations: What if the algorithm has not
+converged? Are the implications then to be trusted? And can we rely on
+the inference obtained on the completed data? These are all open
+questions, because the convergence properties of iterative imputation
+algorithms have not been systematically studied \citep{buur18}.
+Moreover, there is no scientific consensus on how to evaluate
+convergence of imputation algorithms \citep{taka17}. Some default MICE
+techniques (e.g., `predictive mean modeling') might not yield converged
+states at all \citep{murr18}. Therefore, algorithmic convergence should
+be monitored carefully.
+
+Currently, the recommended practice for evaulating the convergence of
+iterative imputation algorithms is through visual inspection. That is,
+the state space of the algorithm is monitored by plotting a scalar
+summary \(\theta\) against the iteration number. \textbf{{[}Segway to
+WHAT TO MONITOR? Also, this paragraph is more or less the same as the
+third paragraph in the second section. Merge or remove one of the
+two?{]}} Typically, the \(\theta\)s under evaluation are chain means and
+variances (\textbf{define!}). These \(\theta\)s may be insufficient
+because they are univariate summaries of the state space of the
+algorithm, while MICE is not only concerned with univariate estimates,
+but the entire multivariate distribution of \(y_{imp}\). A suggestion by
+\citet{buur18} for a multivariate \(\theta\) to monitor is the quantity
+of scientific interest \(Q\) (e.g., a regression coefficient).
+Implementing this, however, might be somewhat advanced for empirical
+researchers. Moreover, this scalar summary is not model-independent,
+i.e., it only applies to one model of scientific interest. Yet, one of
+the advantages of MI is that the missing data problem and scientific
+problem are solved independently. Therefore, this \(\theta\) is not
+sufficient either. Van Buuren \citeyearpar[\(\S\) 4.5.2]{buur18} also
+proposed multivariable evaluation of the MICE algorithm through
+eigenvalue decomposition, building on the work of \citet{mack03}.
+Eigenvalues of a variance-covariance matrix are a measure of the data's
+total covariance. This would be a model-independent, multivariate scalar
+summary to monitor, but it has not yet been implemented.
+\textbf{{[}Segway to HOW TO MONITOR?{]}} Monitoring convergence visually
+is insufficient on two counts: 1) it may be challenging to the untrained
+eye, and 2) only severely pathological cases of non-convergence may be
+diagnosed \citep[\(\S\) 6.5.2]{buur18}. Therefore, a quantitative,
+diagnostic evaluation of convergence would be preferred---although not
+straightforward. Iterative imputation algorithms such as MICE are Markov
+chain Monte Carlo (MCMC) methods. In MCMC methods, convergence is not
+from a scalar to a point, but from one distribution to another. The
+values generated by the algorithm (e.g., imputed values) will vary even
+after convergence. Since MCMC algorithms do not reach a unique point at
+which convergence is established, diagnostics may only identify signs of
+\emph{non}-convergence \citep{hoff09}. Several of such non-convergence
+diagnostics exist, but it is not known whether these are appropriate for
+iterative imputation algorithms.
 
 In this paper, we investigate how non-convergence in iterative
-imputation algorithms may be diagnosed, how well these methods perform,
-and at which point convergence may safely be assumed. For reasons of
-brevity, we only focus on the iterative imputation algorithm implemented
-in the popular \texttt{mice} package \citep{mice} in \texttt{R}
-\citep{R}. The convergence properties of the MICE algorithm are
-investigated through model-based simulation. The results of this
-simulation study are guidelines for assessing convergence of MI
+imputation algorithms may be diagnosed. First, we define which
+non-convergence diagnostics will be considered, and evaluate how these
+diagnostics could be appropriate for iterative imputation applications.
+In the second part of this paper, we will investigate the impact of
+inducing non-convergence in iterative imputation algorithms by means of
+model-based simulation in \texttt{R} \citep{R}. The aim of the
+simulation is to determine whether unbiased, confidence valid inferences
+may be obtained under different levels of non-convergence (\textbf{and
+at which point convergence may safely be assumed}). Additionally, we
+will assess how well different non-convergence diagnostics perform in
+identifying signs of non-convergence. The results of this study are
+guidelines for identifying non-convergence in iterative imptation
 algorithms, which will aid applied researchers in drawing valid
-inference from incomplete datasets. \textbf{add societal relevance: what
-is the consequence of non-convergence?}
-
-\textbf{Include sub-questions and add structure of the paper!} How can
-non-convergence be identified diagnostically? Are common MCMC
-non-convergence diagnostics appropriate for MICE? And if so, which
-threshold should be used to diagnose non-convergence? How many
+inference from incomplete datasets. \textbf{{[}add more societal
+relevance: what is the consequence of non-convergence? e.g., biased
+estimates, invalid inference, etc.{]}} For reasons of brevity, we only
+focus on the iterative imputation algorithm implemented in the popular
+\texttt{mice} package \citep{mice} in \texttt{R} \citep{R}.
+
+\textbf{{[}Include sub-questions and make structure of the paper
+clear!{]}} How can non-convergence be identified diagnostically? Are
+common MCMC non-convergence diagnostics appropriate for MICE? And if so,
+which threshold should be used to diagnose non-convergence? How many
 iterations are sufficient/needed to be able to diagnose non-convergence?
 Are the default number of iterations sufficient (i.e., 5 in mice, 10 in
-SPSS and Stata, 30 in mi)? \textbf{How severe is it when the algorithm
-has not converged? And what are guidelines for practice? Can the
-parameter of interest, estimand \(Q\), be correct when the algorithm is
-not (yet) converged, and vice versa?} \emph{(Maybe add these too? What
-are the effects of continued iterations on estimates, predictions and
-inferences? Do the answers differ with varying missingness proportions?
-That is, we vary the nr of iterations and the missingness proportion
-because we assume that the algorithm has not reached convergence at
-\(t=1\), and performs worse with moderate to high missingness (not so
-much with little or a LOT of missingness).)}
+SPSS and Stata, 30 in mi)? How severe is it when the algorithm has not
+converged? And what are guidelines for practice? Can the parameter of
+interest \(Q\) be correct when the algorithm is not (yet) converged, and
+vice versa?
 
@@
 \(i\) (\(i = 1, 2, \dots, n\)) on variable \(j\)
-(\(j = 1, 2, \dots, k\)) may be either observed or missing. \textbf{The
-number of units \(i\) in dataset \(y\) with at least one missing data
-value, divided by the total number of units \(n\), is called the
-missingness proportion in \(p_{mis}\).} The collection of observed data
-values in \(y\) is denoted by \(y_{obs}\); the missing part of \(y\) is
-referred to as \(y_{mis}\). For each datapoint in \(y_{mis}\), we sample
-\(M \times T\) times plausible values, where \(M\) is the number of
-imputations (\(m = 1, 2, \dots, M\)) and \(T\) is the number of
-iterations (\(t = 1, 2, \dots, T\)). The collection of samples between
-the initial value (at \(t=1\)) and the final imputed value (at \(t=T\))
-will be referred to as an `imputation chain'. \textbf{Add: \(\theta\)s
-are scalar summaries of interest in the iterative algorithm (e.g., chain
-means; the average of the imputed values in each imputation chain). }
+(\(j = 1, 2, \dots, k\)) may be either observed or missing. The number
+of units \(i\) in dataset \(y\) with at least one missing value, divided
+by the total number of units \(n\), is called the missingness proportion
+\(p_{mis}\). The collection of observed data values in \(y\) is denoted
+by \(y_{obs}\); the missing part of \(y\) is referred to as \(y_{mis}\).
+For each datapoint in \(y_{mis}\), we sample \(M \times T\) plausible
+values, where \(M\) is the number of imputations
+(\(m = 1, 2, \dots, M\)) and \(T\) is the number of iterations
+(\(t = 1, 2, \dots, T\)) in the imputation algorithm. The collection of
+samples between the initial value (at \(t=1\)) and the final imputed
+value (at \(t=T\)) will be referred to as an `imputation chain'.
+\textbf{\(\theta\)s are scalar summaries of the algorithm at a certain
+iteration (e.g., chain means; the average of the imputed values in each
+imputation chain). }
 
@@
 
-\textbf{The goal of imputation through e.g.~MICE is to obtain a
-multivariate distribution by iterating over the sequence of univariate
-imputations. Traditionally, the inspection of the multivariate
-convergence is done by visually inspecting the univariate chain means
-and variances over the iterations, or by tracking the \(\widehat{R}\)
-and autocorrelation. The downside to these approaches is that they
-either focus on the univariate state space, or primarily track the
-change over the iterations of a multivariate outcome conform the
-scientific model of interest. Focusing on the convergence of these
-outcome parameters may influence this procedure in the sense that the
-model of evaluation favors the model of interest. Ideally, one would
-like to evaluate a model-independent parameter that summarizes the
-multivariate nature of the data. We propose eigenvalue as such a
-parameter, because it\ldots. }
-
-Mixing and stationarity have histroically been inspected visually, by
-evaluating traceplots of scalar summaries of interest (\(\theta\)s;
-e.g., chain means and chain variances). As \citet{buur18} describes,
-users can also specify a model-specific scalar summary (e.g., a
-regression coefficient). A user-specified scalar summary, however, is
-not universal to all complete data problems. Therefore, as inspired by
-\citep{mack03}, we propose a \(\theta\) that summarizes the multivariate
-state space of the algorithm. Namely, the first eigenvalue of the
-variance-covariance matrix of the \(M\) completed datasets. \textbf{The
-first eigenvalue has the appealing property that is not dependent on the
-substantive model of interest.}
-
-\textbf{Since convergence of iterative imputation algorithms is in
-distribution, there is not a unique point at which the algorithm reaches
-a converged state. We can only evaluate scalar summaries of the
-multivariate state space of the algorithm, \(\theta\)s. The recommended
-scalar summaries to evaluate are chain means and chain variances.
-Additionally, researchers may ``monitor some statistic of scientific
-interest'' \citep[\(\S\) 6.5.2]{buur18}. Because of the issues mentioned
-above, we propose a novel \(\theta\) to monitor: the first eigenvalue of
-the variance-covariance matrix of the completed data. Let
+There are two requirements for convergence of iterative algorithms:
+mixing and stationarity \citep{gelm13}. Without mixing, imputation
+chains do not intermingle nicely, indicating that the distribution of
+imputed values differs across imputations. Chains may be `stuck' at a
+local optimum, instead of sampling imputed values from the entire
+predictive posterior distribution of the missing values. This may cause
+under-estimation of the variance within chains, which results in
+spurious, invalid inferences. Without stationarity, there is trending
+within imputation chains. Trending implies that further iterations would
+yield a different set of imputations. Iterative imputation algorithms
+that have not yet reached stationarity, may thus yield biased estimates.
+
+\textbf{{[}Move to INTRO???{]}} To illustrate what non-convergence looks
+like in iterative imputation algorithms, we reproduce an example from
+van Buuren \citeyearpar[\(\S\) 6.5.2]{buur18}. Figure \ref{fig:non-conv}
+displays a subset of the example: a variable that we will call \(j\). We
+see two traceplots of the chain means for \(j\)---i.e., the average of
+the imputed values for \(j\) in each imputation \(y_{imp,m}\), plotted
+against the iteration number. Each line portrays one imputation. The
+plot on the lefthand side of the figure shows typical convergence of an
+iterative imputation algorithm. The righthand side displays pathological
+non-convergence, induced by purposefully mis-specifying the imputation
+model. \textbf{{[}Leave this in if it's after mixing and stationarity
+are defined, otherwise remove?{]} In the typical convergence situation,
+the imputation chains intermingle nicely and there is little to no
+trending. In the non-convergence plot, there is a lot of trending and
+some chains do not intermingle.} Importantly, the chain means at the
+last iteration (the imputed value per \(m\)) are very different between
+the two plots. The algorithm with the mis-specified model yields imputed
+values that are on average a factor two larger than those of the
+typically converged algorithm. That means that non-convergence has a
+impact on the mean of \(j\) in \(y_{imp, m}\). This difference
+(presumably) translates to bias in the \(M\) completed data estimates
+\(\hat{\mu}_{j,m}\), and consequently also to the pooled estimate
+\(\bar{\mu}_j\). Non-convergence thus leads to biased estimates. This
+shows the importance of reaching converged states in iterative
+imputation algorithms.
+
+\begin{figure}
+
+{\centering \includegraphics[width=.49\linewidth]{manuscript_files/figure-latex/non-conv-1} \includegraphics[width=.49\linewidth]{manuscript_files/figure-latex/non-conv-2}
+
+}
+
+\caption{Typical convergence versus pathological non-convergence.
+Please note that this is the same data with a different imputation model, leading to different imputations (e.g., see the units on the y-axis).}\label{fig:non-conv}
+\end{figure}
+
+\textbf{{[}Move to INTRO???{]}} Convergence has historically been
+inspected visually, by monitoring the imputation algorithm over
+iterations. This is typically done through traceplots. In a traceplot,
+the iteraton number is plotted against a certain \(\theta\). \(\theta\)s
+are scalar summaries of the state space of the algorithm at a specific
+iteration. The recommended scalar summaries to evaluate are chain means
+and chain variances. Additionally, researchers may ``monitor some
+statistic of scientific interest'' \citep[\(\S\) 6.5.2]{buur18}. As
+\citet{buur18} describes, researchers can specify their quantity of
+scientific interest \(Q\) (e.g., a regression coefficient) as scalar
+summary \(\theta\). Such a user-defined scalar summary, however, may be
+somewhat advanced for empirical researchers. And moreover, it is not
+universal to all complete data problems. Focusing on the convergence of
+outcome parameters may influence the iterative imputation procedure in
+the sense that the model of evaluation favors the model of interest. The
+downside to these two approaches is that they either focus on the
+univariate state space, or primarily track the change over the
+iterations of a multivariate outcome conform the scientific model of
+interest. Ideally, one would like to evaluate a model-independent
+parameter that summarizes the multivariate nature of the data.
+Therefore, we propose a \(\theta\) that summarizes the multivariate
+state space of the algorithm, but is independent from the model of
+scientific interest. We propose \(\lambda_1\) as such a scalar summary.
+We define \(\lambda_1\) as the first eigenvalue of the
+variance-covariance matrix of the completed data. Let
 \(\lambda_1 \geq \lambda_2 \geq ... \geq \lambda_j\) be the eigenvalues
@@
 \({y_{obs}, y_{imp}}\). \(\lambda_1\) is measure that summarises the
-covariances in the completed datasets.}
-
-There are two requirements for convergence of iterative algorithms:
-mixing and stationarity \citep{gelm13}. Without mixing, imputation
-chains do not intermingle nicely, \textbf{indicating that \ldots{}}.
-Without stationarity, there is trending within imputation chains, which
-implies that further iterations would yiled a different set of
-imputations.
-
-To illustrate what non-convergence looks like in MI, we reproduce the
-example from van Buuren \citeyearpar[\(\S\) 6.5.2]{buur18}. Figure
-\ref{fig:non-conv} shows the traceplot for one of the variables in the
-example. we see the average of the imputed values for a variable \(j\)
-in \(y_{imp}\). The average imputed values in the left hand plot are
-about a magnitude 2 lower than the right hand plot. So non-convergence
-has a impact on the average imputed value for \(j\) in \(y_{imp, m}\).
-This difference (presumably) translates to bias in the pooled estimate
-\(\bar{\mu}_j\) as well. And maybe also to higher order statistics,
-e.g.~variances, covariances, etc. Therefore, it's important to converge.
-\textbf{Explain what we see, namely example by \citet{buur18}
-reproduced, showing the traceplots of chain means for some variable. The
-first plot is typical convergence of MICE, the second is pathological
-non-convergence because of a mis-specified imputation model. Each line
-is an imputation. In the first plot, the chains intermingle nicely and
-there is little to no trending. In the second plot, there is a lot of
-trending and some chains do not intermingle. Importantly, the chain
-means at the last iteration (the imputed value per \(m\)) are very
-different between the two plots. The algorithm with the mis-specified
-model yields imputed values that are on average a magnitude two larger
-than those of the typically converged algorithm. This shows the
-importance of reaching converged states in iterative imputation
-algorithms.}
-
-\hypertarget{methods-under-evaluation}{%
-\subsection{Methods under evaluation}\label{methods-under-evaluation}}
-
-Non-stationarity within chains may be diagnosed with e.g.,
-autocorrelation \citep[\(AC\);][]{scha97, gelm13}, numeric standard
-error \citep[`MC error';][]{gewe92}, or Raftery and Lewis's
+covariances in the completed datasets. \textbf{The first eigenvalue has
+the appealing property that is not dependent on the substantive model of
+interest. Eigenvalue decomposition is inspired by \citet{mack03}.}
+
+\hypertarget{diagnostics-under-evaluation}{%
+\subsection{Diagnostics under
+evaluation}\label{diagnostics-under-evaluation}}
+
+Non-convergence diagnostics for MCMC algorithms typically identify
+problems in either one of the two requirements for converegnce: mixing
+or stationarity. Non-stationarity within chains may be diagnosed with
+e.g., autocorrelation \citep[\(AC\);][]{scha97, gelm13}, numeric
+standard error \citep[`MC error';][]{gewe92}, or Raftery and Lewis's
 \citeyearpar{raft91} procedure to determine the effect of trending on
@@
 non-stationarity, but this has not yet been thoroughly investigated
-\citep{veht19}. Therefore, use we \(\widehat{R}\) and \(AC\) to evaluate
+\citep{veht19}. Therefore, we use \(\widehat{R}\) and \(AC\) to evaluate
 mixing and stationarity separately, as recommended by e.g.,
 \citet{cowl96}.
-
-\begin{figure}
-
-{\centering \includegraphics[width=.49\linewidth]{manuscript_files/figure-latex/non-conv-1} \includegraphics[width=.49\linewidth]{manuscript_files/figure-latex/non-conv-2}
-
-}
-
-\caption{Typical convergence versus pathological non-convergence.
-Please note that this is the same data with a different imputation model, leading to different imputations (see the y axis!).
-Wel min of meer dezelfde origin op iteration 1.}\label{fig:non-conv}
-\end{figure}
 
@@
 
-An updated version of \(\widehat{R}\) has been proposed by
-\citet{veht19} (p.~5). \textbf{This version may be suitable for
+An updated version of \(\widehat{R}\) has been proposed by Vehtari et
+al.~\citeyearpar[p.~5]{veht19}. \textbf{This version may be suitable for
 iterative imputation.} Let \(M\) be the total number of chains, \(T\)
 the number of iterations per chain, and \(\theta\) the scalar summary of
-interest (e.g., chain mean or chain variance). For each chain
-(\(m = 1, 2, \dots, M\)), we estimate the variance of \(\theta\), and
-average these to obtain within-chain variance \(W\).
+interest (e.g., chain means). For each chain (\(m = 1, 2, \dots, M\)),
+we estimate the variance of \(\theta\), and average these to obtain
+within-chain variance \(W\).
 
@@
 variance of \(\theta\) and speed up convergence. As convergence
-diagnostic, the interest is therefore in positive \(AC\)-values.
-\textbf{Maybe remove:} Moreover, the magnitude of \(AC\)-values may be
-evaluated statistically, but that is outside of our scope.
+diagnostic, the interest is therefore in positive \(AC\)-values. The
+magnitude of \(AC\)-values may be evaluated statistically {[}\textbf{add
+reference}{]}.
+
+\hypertarget{thresholds}{%
+\subsubsection{Thresholds}\label{thresholds}}
+
+Upon convergence, \(\widehat{R}=1\) and \(AC=0\), which are unlikely
+thresholds for MCMC algorithms, because of its convergence to a
+distribution. Even in the most converged state, the algorithm will show
+some signs of non-mixing and non-stationarity. Upon \emph{sufficient}
+convergence, the imputation chains will intermingle such that the only
+difference between the chains is caused by the randomness induced by the
+algorithm (\(\widehat{R} > 1\)), and there may be some dependency
+between subsequent iterations of imputation chains (\(AC > 0\)). In
+practice, non-convergence is usually diagnosed when \(\widehat{R}\)
+\textgreater{} 1.2 or 1.1 or even 1.01. And a t-test is performed to
+assess whether \(AC\) is significantly different from zero.
 
@@
 
-Upon convergence, \(\widehat{R}=1\) and \(AC=0\), which are unlikely
-thresholds for MCMC algorithms, because of its convergence to a
-distribution. In practice, non-convergence is usually diagnosed when
-\(\widehat{R}\) \textgreater{} 1.2 or 1.1 or even 1.01. And a t-test is
-performed to assess whether \(AC\) is significantly different from zero.
-
-We assume that complete convergence, defined as \(\widehat{R} = 1\) and
-\(AC = 0\), will not be observed. Because even in the most converged
-state, the algorithm will show some signs of non-mixing and
-non-stationarity. Upon \emph{sufficient} convergence, the imputation
-chains will intermingle such that the only difference between the chains
-is caused by the randomness induced by the algorithm
-(\(\widehat{R} > 1\)), and there may be some dependency between
-subsequent iterations of imputation chains (\(AC > 0\)).
-
-\textbf{Before using them in the simulation, the two methods show at
-least be able to distinguish between the two algorithms plotted in
-Figure 2. From visual inspection we know that the typical convergence
-has some signs of non-mixing around iteration 2, and little trending. In
-the non-convergence situation, there is a lot of trending upto iteration
-6, after which the chains reach a somewhat more stationary state. Mixing
-in this situation gets worse from the first iteration onward, and
-gradually gets better around iteration 7. To assess whether
-\(\widehat{R}\) and \(AC\) may be appropriate non-convergence
-identifiers for iterative imputation algorithms, the following condition
-must hold: the methods indicate worse performance for the mis-specified
-model with pathological non-convergence (i.e., higher \(\widehat{R}\)-
-and \(AC\)-values than the typical performance). And additionally the
-methods should reflect the increasing convergence in the typical
-convergence situation as the number of iterations goes up.}
-
-\textbf{In Figure \ref{fig:diagnostics}A, the chain means from Figure
-\ref{fig:non-conv} are plotted again, now together---as are the
-non-convergence diagnostics. Panel B shows \(\widehat{R}\) as computed
-by implementing \citet{veht19} 's recommendations. As required,
-\(\widehat{R}\) indicates less signs of non-convergence as the number of
-iterations goes up }in the typical convergence situation\textbf{. The
-superior performance of the typical convergence over the pathological
-non-convergence is less prominent, even flipped for \(t<4\). Panel C
-displays the \(AC\) as computed with the R function
-\texttt{stats::acf()}. When we look at this panel, we conclude something
-weird. The \(AC\)-values indicate equal performance (up-to \(t=5\)) for
-the typical convergence and the pathological non-convergence, while
-there is obvious trending in the latter. Moreover, the best convergence
-(as indicated by the lowest \(AC\)-value) is observed at \(t=2\), but
-looking at the chain means in panel A, there should be some signs of
-trending up-to iteration number seven. After consulting the
-documentation on stats::acf() \citep{R}, we conclude that this
-implementation of \(AC\) is not suitable for iterative imputation
-algorithms. There is a correction factor for a mathematical shortcut
-that works in the limit. According to \citet{box15}, the function works
-when the number of iterations \textgreater{} 50. The default number of
-iterations in iterative imputation, however, is often much lower.
-Therefore, we compute \(AC\) manually, see panel D. The \(AC\) values in
-this plot do meet the requirements (\ldots) and will therefore be used
-in the simulation study.}
-
-\textbf{acf is niet bedoeld om weinig iteraties te\ldots{} niet een fout
-in acf, alleen hier niet van toepassing}
+Before we evaluate the performance of the non-convergence diagnostics
+\(\widehat{R}=1\) and \(AC=0\) quantitatively through simulation, we
+apply them to the example of pathological non-convergence by
+\citet{buur18} that we reproduced above. We want to be able to draw the
+same conclusions as we did from visually inspecting the two algorithms.
+But \(\widehat{R}=1\) and \(AC=0\) should at least be able to
+distinguish between the two algorithms plotted in Figure
+\ref{fig:non-conv}. From visual inspection, we know that the typically
+converged algorithm initially portrayed some signs of non-mixing (around
+\(t=2\)), but intermigled nicely overall. Additionally, there was very
+little trending. The algorithm with pathological non-convergence showed
+severe non-mixing, although this gradually improved (from \(t=7\)).
+Moreover, there was a lot of trending as well initially (up-to \(t=6\)),
+after which the chains reached a somewhat more stationary state. To
+assess whether \(\widehat{R}\) and \(AC\) may be appropriate
+non-convergence identifiers for iterative imputation algorithms, they
+should identify the same trends as the visual inspection. But at least,
+the following condition must hold: the methods should indicate worse
+performance for the mis-specified model with pathological
+non-convergence than the typically converged algorithm (i.e., higher
+\(\widehat{R}\)- and \(AC\)-values).
+
+In Figure \ref{fig:diagnostics}A, the chain means from Figure
+\ref{fig:non-conv} are plotted again---now together. Panel B shows two
+versions of calculating \(AC\)s applied to the \(\theta\)s of panel A:
+the default calculation with \texttt{R} function \texttt{stats::acf()}
+\citep{R}, and manual calculation as the correlation between \(\theta\)
+in iteration \(t\) and \(\theta\) in iteration \(t+1\). Panel C shows
+the traditional computation of \(\widehat{R}\), and panel D shows
+\(\widehat{R}\) as computed by implementing \citet{veht19} 's
+recommendations.
 
@@
 \end{figure}
+
+When we look at panel B, we conclude something weird. The \(AC\)-values
+calculated with the default function indicate equal performance for the
+typical convergence and the pathological non-convergence (up-to
+\(t=5\)), while there is obvious trending in the \(\theta\)s of the
+latter. Moreover, the best convergence (as indicated by the lowest
+\(AC\)-value) is observed at \(t=2\), but looking at the chain means in
+panel A, there should be some signs of trending up-to iteration number
+seven. After consulting the documentation on \texttt{stats::acf()}
+\citep{R}, we conclude that this \(AC\) function is not suited for
+iterative imputation algorithms. The function is optimized for
+performance when \(t\geq50\) \citep{box15}, while the default number of
+iterations in iterative imputation is often much lower. Therefore, we
+compute \(AC\) manually, see the solid line. These \(AC\) values
+indicate non-stationarity as expected. We therefore calculate \(AC\)
+manually.
+
+\textbf{Describe panel C here and why it is not optimal{]}} The
+\(\widehat{R}\)-values in panel D do meet the requirements specified
+above. \(\widehat{R}\) as computed conform \citet{veht19} indeed
+indicate less signs of non-convergence as the number of iterations goes
+up.
 
@@
 \(\widehat{R}\) and \(AC\) perform in identifying the effects of
-non-convergence (\textbf{add thetas}). And finally, we will formulate an
-informed advice on the requirements to safely assume \emph{sufficient}
-convergence in practice. That is, we assume that the algorithm is
-\emph{sufficiently} converged when each estimate \(\bar{Q}\) is an
-unbiased and confidence valid estimate of the corresponding estimand
-\(Q\). (\textbf{add to formulate advice on thetas?})
+non-convergence \textbf{{[}and performance of different thetas{]}}. And
+finally, we will formulate an informed advice on the requirements to
+safely assume \emph{sufficient} convergence in practice. That is, we
+assume that the algorithm is \emph{sufficiently} converged when each
+estimate \(\bar{Q}\) is an unbiased and confidence valid estimate of the
+corresponding estimand \(Q\). \textbf{{[}and formulate advice on
+thetas?{]}}
 
@@
   the conventional non-convergence threshold \(\widehat{R}>1.2\). Yet,
-  scientific estimates were insensitive to continued iteration after
-  \(t>5\) \citep{lace07}. We therefore suspect that \(\widehat{R}\) may
-  over-estimate signs of non-convergence in iterative imputation
-  algorithms, compared to the bias the estimates. In contrast to this,
-  it may occur that signs of non-convergence are under-estimated by
-  \(\widehat{R}\), in exceptional cases where the initial values of the
-  algorithm are not appropriately over-dispersed \citep[p.~437]{broo98}.
-  In e.g.~\texttt{mice}, initial values are chosen randomly from the
-  observed data, hence we cannot be certain of over-dispersion in the
-  initial values. In practice, we do not expect this to cause problems
-  for identifying non-convergence with \(\widehat{R}\).
+  scientific estimates were insensitive to continued iteration from
+  \(t=5\) onwards \citep{lace07}. We therefore suspect that
+  \(\widehat{R}\) may over-estimate signs of non-convergence in
+  iterative imputation algorithms, compared to the bias the estimates.
+  In contrast to this, it may occur that signs of non-convergence are
+  under-estimated by \(\widehat{R}\), in exceptional cases where the
+  initial values of the algorithm are not appropriately over-dispersed
+  \citep[p.~437]{broo98}. In e.g.~\texttt{mice}, initial values are
+  chosen randomly from the observed data, hence we cannot be certain of
+  over-dispersion in the initial values. In practice, we do not expect
+  this to cause problems for identifying non-convergence with
+  \(\widehat{R}\).
 \item
@@
 we obtain the true values of several quantities of scientific interest,
-\(Q\)s. \textbf{MOVE THIS NEXT PART TO TOP OF SIMULATION SECTION?}
+\(Q\)s. \textbf{{[}MOVE THIS NEXT PART TO TOP OF SIMULATION SECTION?{]}}
 \(Q\)s are the targets that will be estimated in each simulation
@@
 number of iterations, then estimating the \(Q\)s on the \(M\) completed
-datasets, and finally pooling the \(M\) \(\hat{Q}\)s to obtain a
-\(\bar{Q}\) for each \(Q\). \textbf{Add that we try to predict the
+datasets, and finally pooling the \(M\) estimates \(\hat{Q}\) to obtain
+a single \(\bar{Q}\) for each \(Q\). \textbf{We try to predict the
 effect of non-converegence (i.e., the difference between \(Q\) and
@@
 \(\widehat{R}\))}. The true values of these scientific estimands are
-calculated on the complete data with the R functions
+calculated on the complete data with the \texttt{R} functions
 \texttt{base::mean()}, \texttt{stats::sd()}, and \texttt{stats::lm()}
@@
 
-In each simulation repetition, we ampute the complete dataset five times
-to obtain five different incomplete datasets, \(y\)s. The ratio between
-\(y_{obs}\) and \(y_{mis}\) in each \(y\) depends on the missingness
-proportion: \(p_{mis} =.05,.25,.5,.75,.95\). We ampute the complete data
-with the \texttt{mice} package \citep[function
-\texttt{mice::ampute()};][]{mice}. All missingness is multivariate, and
-conform a `missing completely at random' missingness mechanism
-\citep{rubin87}. I.e., the probability to be missing is the same for all
-\(N \times k\) cells in \(y\), conditional on the missingness proportion
-\(p_{mis}\).
-
-Missing datapoints in the incomplete datasets, \(y\)s, are imputed with
-the \texttt{mice()} function \citep{mice}. All imputation procedures are
+In each simulation repetition, we \emph{ampute} the complete dataset
+five times to obtain five different incomplete datasets, \(y\)s. The
+ratio between \(y_{obs}\) and \(y_{mis}\) in each \(y\) depends on the
+missingness proportion: \(p_{mis} =.05,.25,.5,.75,.95\). We ampute the
+complete data with the \texttt{mice} package \citep[function
+\texttt{mice::ampute()};][]{mice}. We consider all possible univariate
+and multivariate patterns of missingness, conform a `missing completely
+at random' missingness mechanism \citep{rubin87}. I.e., the probability
+to be missing is the same for all \(N \times k\) cells in \(y\),
+conditional on the missingness proportion \(p_{mis}\).
+
+Missing datapoints in each incomplete dataset \(y\) are imputed with the
+\texttt{mice()} function \citep{mice}. All imputation procedures are
 performed with Bayesian linear regression imputation, and five
 imputation chains (\(M=5\)). Because \(M=5\), each run of the
-\texttt{mice()} function results in five sets of imputations,
-\(y_{imp,m}\)s, where \(m = 1, 2, ..., 5\). The number of iterations in
+\texttt{mice()} function results in five sets of imputations:
+\(y_{imp,m}\), where \(m = 1, 2, ..., 5\). The number of iterations in
 each run of the algorithm varies between simulation conditions
@@
 
-From the 5 imputations \(y_{imp,m}\): get chain means and chain
-variances from imputation object to use as scalar summary \(\theta\) in
-convergence diagnostics. Then use \texttt{mice::complete()} to combine
-the imputed data \(y_{imp,m}\) with the observed data \(y_{obs}\),
-resulting in five completed datasets \{\(y_{obs}, y_{imp, m}\)\}.
+From the 5 imputations \(y_{imp,m}\) we obtain the chain means and chain
+variances for each variable \(j\). These will serve as scalar summary
+\(\theta\) to apply non-convergence diagnostics \(\widehat{R}\) and
+\(AC\) on. We then use \texttt{mice::complete()} to combine the imputed
+data \(y_{imp,m}\) with the observed data \(y_{obs}\), resulting in five
+completed datasets \{\(y_{obs}, y_{imp, m}\)\}.
 
@@
   Estimate \(\hat{Q}\)s for \(Q= \mu\) and \(Q=\sigma\) with
-  \texttt{base::mean()}, and \texttt{stats::sd()} \citep{R}. Then pool
-  the \(\hat{Q}\)s with \texttt{base::mean()} to get \(\bar{Q}\)s.
-\item
-  Compute \(\lambda_1\), to use as \(\theta\) in convergence
+  \texttt{base::mean()}, and \texttt{stats::sd()} \citep{R}. We pool the
+  \(\hat{Q}\)s with \texttt{base::mean()} to get \(\bar{Q}\)s.
+\item
+  Compute \(\lambda_1\), to use as \(\theta\) in the two convergence
   diagnostics. By definition, the first eigenvalue of the
@@
   Perform multiple linear regression with \texttt{stats::lm()} \citep{R}
-  to get the \(\hat{Q}\)s for \(Q=\beta\) and \(Q=r^2\). Pool the
+  to get the \(\hat{Q}\)s for \(Q=\beta\) and \(Q=r^2\). We pool the
   \(\hat{Q}\)s conform \citet{vink14} to get \(\bar{Q}\)s for the
-  \(\beta\)s and \(r^2\) (\textbf{check pooling of \(r^2\), instead of
-  \texttt{mice::pool.r.squared()}}). Additionally, use the \(\hat{Q}\)s
-  of \(Q=\beta\) as \(\theta\) to compute convergence diagnostics.
+  \(\beta\)s and \(r^2\) \textbf{{[}check pooling of \(r^2\), instead of
+  \texttt{mice::pool.r.squared()}{]}}. Additionally, we use the
+  \(\hat{Q}\)s of \(Q=\beta\) as \(\theta\) to apply the convergence
+  diagnostics to.
 \end{itemize}
@@
 confidence valid estimates of the \(Q\)s. \textbf{Rephrase and add
-corresponding \(Q\)s:} The two methods are applied to four scalar
-summaries \(\theta\): chain means (i.e., the mean in each
-\(y_{imp, m, j}\)), chain variances (i.e., the variance in each
-\(y_{imp, m, j}\)), the estimated regression coefficients
-\(\beta_{1,2,3}\) in each \{\(y_{obs}, y_{imp, m}\)\} (i.e., a
-user-defined statistic of scientific interest \(\hat{Q}\)), and the
-first eigenvalue of the variance-covariance matrix in each
-\{\(y_{obs}, y_{imp, m}\)\}, \(\lambda_1\).
+corresponding \(Q\)s:} We measure the performance of \(\widehat{R}\) and
+\(AC\) by the four scalar summaries \(\theta\): chain means (i.e., the
+mean in each \(y_{imp, m, j}\)) versus \(Q=\mu\), chain variances (i.e.,
+the variance in each \(y_{imp, m, j}\)) versus \(Q=\sigma\), the
+estimated regression coefficients \(\beta\) in each
+\{\(y_{obs}, y_{imp, m}\)\} (i.e., a user-defined statistic of
+scientific interest \(\hat{Q}\)) against \(Q=\beta\), and the first
+eigenvalue of the variance-covariance matrix in each
+\{\(y_{obs}, y_{imp, m}\)\}, \(\lambda_1\) versus \(Q=r^2\).
 
@@
 
-\textbf{Add to this section: 1) emphasize that the plots are averages
-across repetitions, not within MICE; 2) sample effects due to single
-complete dataset?? 3) Add more info about figure legends and axes.)}
-
-Primarily, we want to know what the effect is of inducing
-non-convergence on the estimated \(Q\)s. Specifically, we are interested
-in the bias of \(\bar{Q}\) for all \(Q\)s and the confidence validity of
-\(\bar{Q}\) for \(Q=\beta\). Therefore, we plot these performance
-measures against the simulation conditions (the number of iterations
-\(T\) and the missingness proportion \(p_{mis}\)). We look at averages
-across simulation repetitions.
+\textbf{{[}Add to this section: 1) emphasize that the iterations in the
+plots are averages across repetitions, not within one run of MICE; 2)
+sample effects due to single complete dataset?? 3) Add more info about
+figure legends and axes.{]}} Primarily, we want to know what the effect
+is of inducing non-convergence on the estimated \(Q\)s. Specifically, we
+are interested in the bias of \(\bar{Q}\) for all \(Q\)s and the
+confidence validity of \(\bar{Q}\) for \(Q=\beta\). Therefore, we plot
+these performance measures against the simulation conditions (the number
+of iterations \(T\) and the missingness proportion \(p_{mis}\)). We look
+at averages across simulation repetitions.
 
@@
 the performance measures for the \(Q\)s, but now we evaluate these
-against what \(\widehat{R}\) and \(AC\) indicate about signs of non-the
-convergence in the \(\theta\)s.
+against what \(\widehat{R}\) and \(AC\) indicate about signs of
+non-convergence in the \(\theta\)s.
+
+\textbf{{[}In short:{]}} - Is there more bias in conditions where
+non-convergence was induced? (NOPE for univariates, YES for multivar.).
 
 \begin{itemize}
-\item
-  Is there more bias in conditions where non-convergence was induced?
-  (NOPE for univariates, YES for multivar.).
 \item
@@
   \(Q\)s) the same point as identified by the non-convergence
-  diagnostics? (NOPE, they are too conservative).**
-\item
-  Is one of the thetas better? (NAH, doesn't really matter).
+  diagnostics? (NOPE, they are too conservative).
+\item
+  Is one of the thetas better? (NAH, doesn't really matter, probably
+  multivariate is better).
 \item
@@
 lower number of iterations show more signs of non-convergence. I.e., on
-average these conditions portet more extreme bias in the estimated
+average these conditions portray more extreme bias in the estimated
 \(Q\)s, and non-nominal coverages. Roughly speaking this means that MICE
@@
 
+Uitsmijter: To conclude, in the case of drawing inference from
+incomplete data, convergence of the iterative imputation algorithm is
+often a convenience but not a necessity.
+
 \hypertarget{grading}{%

### 2020-05-04 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
Add requirements thesis submission

@@
 output:
- rticles::sage_article:
+  bookdown::pdf_book:
+    # includes:
+    #   in_header: preamble.tex
     keep_tex: yes
+    base_format: rticles::sage_article
 ---
@@
 **[Add to this section: 1) emphasize that the iterations in the plots are averages across repetitions, not within one run of MICE; 2) sample effects due to single complete dataset??
-3) Add more info about figure legends and axes.]**
+3) Add more info about figure legends and axes 4) the ac and rhat are independent of missingness, not the thetas.]**
 Primarily, we want to know what the effect is of inducing non-convergence on the estimated $Q$s.

### 2020-05-04 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Add requirements thesis submission

@@
 
-\affiliation{\affilnum{1}{Department of Methodology and Statistics, Utrecht University, Utrecht,
-The Netherlands}}
+\affiliation{\affilnum{1}{Department of Methodology and Statistics, Utrecht University, Utrecht, The Netherlands}}
 
-\corrauth{Hanne Oberman, Sjoerd Groenman building, Utrecht Science Park, Utrecht,
-The Netherlands.}
+\corrauth{Hanne Oberman, Sjoerd Groenman building, Utrecht Science Park, Utrecht, The Netherlands.}
 
@@
 Iterative imputation is a popular tool to accommodate missing data.
-While it is widely accepted that valid inferences can be obtained with
-this technique, these inferences all rely on algorithmic convergence.
-There is no consensus on how to evaluate the convergence properties of
-the method. This paper provides insight into identifying non-convergence
-of iterative impuation algorithms.
+While it is widely accepted that valid inferences can be obtained with this technique, these inferences all rely on algorithmic convergence.
+There is no consensus on how to evaluate the convergence properties of the method.
+This paper provides insight into identifying non-convergence of iterative impuation algorithms.
 \end{abstract}
@@
 Anyone who analyzes person-data may run into a missing data problem.
-Missing data is not only ubiquitous, but treating it can also be
-tedious. If a dataset contains just one incomplete observation,
-statistical inferences are undefined and will not produce any results.
-To circumvent this, many statistical packages employ list-wise deletion
-by default (i.e., ignoring incomplete observations). Unfortunately, this
-\emph{ad hoc} solution may yield wildly invalid results \citep{buur18}.
-An alternative is to \emph{impute} (i.e., fill in) the missing values in
-the incomplete observations. Subsequently, statistical inferences can be
-performed on the completed dataset. By repeating this process several
-times, a distribution of plausible results may be obtained, which
-reflects the uncertainty in the data due to missingness. This technique
-is known as `multiple imputation' \citep[MI;][]{rubin76}. MI has proven
-to be a powerful technique to draw valid inferences from incomplete data
-under many circumstances \citep{buur18}.
+Missing data is not only ubiquitous, but treating it can also be tedious.
+If a dataset contains just one incomplete observation, statistical inferences are undefined and will not produce any results.
+To circumvent this, many statistical packages employ list-wise deletion by default (i.e., ignoring incomplete observations).
+Unfortunately, this \emph{ad hoc} solution may yield wildly invalid results \citep{buur18}.
+An alternative is to \emph{impute} (i.e., fill in) the missing values in the incomplete observations.
+Subsequently, statistical inferences can be performed on the completed dataset.
+By repeating this process several times, a distribution of plausible results may be obtained, which reflects the uncertainty in the data due to missingness.
+This technique is known as `multiple imputation' \citep[MI;][]{rubin76}.
+MI has proven to be a powerful technique to draw valid inferences from incomplete data under many circumstances \citep{buur18}.
 
-Figure \ref{fig:diagram} provides an overview of the steps involved with
-MI. Missing data \(y_{mis}\) in an incomplete dataset is imputed \(M\)
-times. The imputed data \(y_{imp}\) is combined with the observed data
-\(y_{obs}\) to create \(M\) completed datasets. On each completed
-dataset, an analysis of scientific interest is performed. The quantity
-of scientific interest (e.g., a regression coefficient) is denoted with
-\(Q\). Since \(Q\) is estimated on each completed dataset, \(M\)
-separate \(\hat{Q}\)-values are obtained. These \(M\) values are
-combined into a single pooled estimate \(\bar{Q}\). The premise of
-multiple imputation is that \(\bar{Q}\) is an unbiased and confidence
-valid estimate of the true---but missing---data inference.
+Figure \ref{fig:diagram} provides an overview of the steps involved with MI.
+Missing data \(y_{mis}\) in an incomplete dataset is imputed \(M\) times.
+The imputed data \(y_{imp}\) is combined with the observed data \(y_{obs}\) to create \(M\) completed datasets.
+On each completed dataset, an analysis of scientific interest is performed.
+The quantity of scientific interest (e.g., a regression coefficient) is denoted with \(Q\).
+Since \(Q\) is estimated on each completed dataset, \(M\) separate \(\hat{Q}\)-values are obtained.
+These \(M\) values are combined into a single pooled estimate \(\bar{Q}\).
+The premise of multiple imputation is that \(\bar{Q}\) is an unbiased and confidence valid estimate of the true---but missing---data inference.
 
@@
 
-A popular method to obtain imputations is to use the `Multiple
-Imputation by Chained Equations' algorithm, shorthand
-`MICE'\citep{mice}. With MICE, imputed values \(y_{imp}\) are drawn from
-the posterior predictive distribution of the missing values \(y_{mis}\).
-The algorithm is named after the iterative algorithmic procedure by
-which imputed values are generated: a multivariate distribution is
-obtained by iterating over a sequence of univariate imputations. The
-iterative nature of algorithms like MICE introduces a potential threat
-to the validity of the imputations: What if the algorithm has not
-converged? Are the implications then to be trusted? And can we rely on
-the inference obtained on the completed data? These are all open
-questions, because the convergence properties of iterative imputation
-algorithms have not been systematically studied \citep{buur18}.
-Moreover, there is no scientific consensus on how to evaluate
-convergence of imputation algorithms \citep{taka17}. Some default MICE
-techniques (e.g., `predictive mean modeling') might not yield converged
-states at all \citep{murr18}. Therefore, algorithmic convergence should
-be monitored carefully.
+A popular method to obtain imputations is to use the `Multiple Imputation by Chained Equations' algorithm, shorthand `MICE'\citep{mice}.
+With MICE, imputed values \(y_{imp}\) are drawn from the posterior predictive distribution of the missing values \(y_{mis}\).
+The algorithm is named after the iterative algorithmic procedure by which imputed values are generated: a multivariate distribution is obtained by iterating over a sequence of univariate imputations.
+The iterative nature of algorithms like MICE introduces a potential threat to the validity of the imputations: What if the algorithm has not converged?
+Are the implications then to be trusted?
+And can we rely on the inference obtained on the completed data?
+These are all open questions, because the convergence properties of iterative imputation algorithms have not been systematically studied \citep{buur18}.
+Moreover, there is no scientific consensus on how to evaluate convergence of imputation algorithms \citep{taka17}.
+Some default MICE techniques (e.g., `predictive mean modeling') might not yield converged states at all \citep{murr18}.
+Therefore, algorithmic convergence should be monitored carefully.
 
-Currently, the recommended practice for evaulating the convergence of
-iterative imputation algorithms is through visual inspection. That is,
-the state space of the algorithm is monitored by plotting a scalar
-summary \(\theta\) against the iteration number. \textbf{{[}Segway to
-WHAT TO MONITOR? Also, this paragraph is more or less the same as the
-third paragraph in the second section. Merge or remove one of the
-two?{]}} Typically, the \(\theta\)s under evaluation are chain means and
-variances (\textbf{define!}). These \(\theta\)s may be insufficient
-because they are univariate summaries of the state space of the
-algorithm, while MICE is not only concerned with univariate estimates,
-but the entire multivariate distribution of \(y_{imp}\). A suggestion by
-\citet{buur18} for a multivariate \(\theta\) to monitor is the quantity
-of scientific interest \(Q\) (e.g., a regression coefficient).
-Implementing this, however, might be somewhat advanced for empirical
-researchers. Moreover, this scalar summary is not model-independent,
-i.e., it only applies to one model of scientific interest. Yet, one of
-the advantages of MI is that the missing data problem and scientific
-problem are solved independently. Therefore, this \(\theta\) is not
-sufficient either. Van Buuren \citeyearpar[\(\S\) 4.5.2]{buur18} also
-proposed multivariable evaluation of the MICE algorithm through
-eigenvalue decomposition, building on the work of \citet{mack03}.
-Eigenvalues of a variance-covariance matrix are a measure of the data's
-total covariance. This would be a model-independent, multivariate scalar
-summary to monitor, but it has not yet been implemented.
-\textbf{{[}Segway to HOW TO MONITOR?{]}} Monitoring convergence visually
-is insufficient on two counts: 1) it may be challenging to the untrained
-eye, and 2) only severely pathological cases of non-convergence may be
-diagnosed \citep[\(\S\) 6.5.2]{buur18}. Therefore, a quantitative,
-diagnostic evaluation of convergence would be preferred---although not
-straightforward. Iterative imputation algorithms such as MICE are Markov
-chain Monte Carlo (MCMC) methods. In MCMC methods, convergence is not
-from a scalar to a point, but from one distribution to another. The
-values generated by the algorithm (e.g., imputed values) will vary even
-after convergence. Since MCMC algorithms do not reach a unique point at
-which convergence is established, diagnostics may only identify signs of
-\emph{non}-convergence \citep{hoff09}. Several of such non-convergence
-diagnostics exist, but it is not known whether these are appropriate for
-iterative imputation algorithms.
+Currently, the recommended practice for evaulating the convergence of iterative imputation algorithms is through visual inspection.
+That is, the state space of the algorithm is monitored by plotting a scalar summary \(\theta\) against the iteration number.
+\textbf{{[}Segway to WHAT TO MONITOR?
+Also, this paragraph is more or less the same as the third paragraph in the second section.
+Merge or remove one of the two?{]}} Typically, the \(\theta\)s under evaluation are chain means and variances (\textbf{define!}).
+These \(\theta\)s may be insufficient because they are univariate summaries of the state space of the algorithm, while MICE is not only concerned with univariate estimates, but the entire multivariate distribution of \(y_{imp}\).
+A suggestion by \citet{buur18} for a multivariate \(\theta\) to monitor is the quantity of scientific interest \(Q\) (e.g., a regression coefficient).
+Implementing this, however, might be somewhat advanced for empirical researchers.
+Moreover, this scalar summary is not model-independent, i.e., it only applies to one model of scientific interest.
+Yet, one of the advantages of MI is that the missing data problem and scientific problem are solved independently.
+Therefore, this \(\theta\) is not sufficient either.
+Van Buuren \citeyearpar[\(\S\) 4.5.2]{buur18} also proposed multivariable evaluation of the MICE algorithm through eigenvalue decomposition, building on the work of \citet{mack03}.
+Eigenvalues of a variance-covariance matrix are a measure of the data's total covariance.
+This would be a model-independent, multivariate scalar summary to monitor, but it has not yet been implemented.
+\textbf{{[}Segway to HOW TO MONITOR?{]}} Monitoring convergence visually is insufficient on two counts: 1) it may be challenging to the untrained eye, and 2) only severely pathological cases of non-convergence may be diagnosed \citep[\(\S\) 6.5.2]{buur18}.
+Therefore, a quantitative, diagnostic evaluation of convergence would be preferred---although not straightforward.
+Iterative imputation algorithms such as MICE are Markov chain Monte Carlo (MCMC) methods.
+In MCMC methods, convergence is not from a scalar to a point, but from one distribution to another.
+The values generated by the algorithm (e.g., imputed values) will vary even after convergence.
+Since MCMC algorithms do not reach a unique point at which convergence is established, diagnostics may only identify signs of \emph{non}-convergence \citep{hoff09}.
+Several of such non-convergence diagnostics exist, but it is not known whether these are appropriate for iterative imputation algorithms.
 
-In this paper, we investigate how non-convergence in iterative
-imputation algorithms may be diagnosed. First, we define which
-non-convergence diagnostics will be considered, and evaluate how these
-diagnostics could be appropriate for iterative imputation applications.
-In the second part of this paper, we will investigate the impact of
-inducing non-convergence in iterative imputation algorithms by means of
-model-based simulation in \texttt{R} \citep{R}. The aim of the
-simulation is to determine whether unbiased, confidence valid inferences
-may be obtained under different levels of non-convergence (\textbf{and
-at which point convergence may safely be assumed}). Additionally, we
-will assess how well different non-convergence diagnostics perform in
-identifying signs of non-convergence. The results of this study are
-guidelines for identifying non-convergence in iterative imptation
-algorithms, which will aid applied researchers in drawing valid
-inference from incomplete datasets. \textbf{{[}add more societal
-relevance: what is the consequence of non-convergence? e.g., biased
-estimates, invalid inference, etc.{]}} For reasons of brevity, we only
-focus on the iterative imputation algorithm implemented in the popular
-\texttt{mice} package \citep{mice} in \texttt{R} \citep{R}.
+In this paper, we investigate how non-convergence in iterative imputation algorithms may be diagnosed.
+First, we define which non-convergence diagnostics will be considered, and evaluate how these diagnostics could be appropriate for iterative imputation applications.
+In the second part of this paper, we will investigate the impact of inducing non-convergence in iterative imputation algorithms by means of model-based simulation in \texttt{R} \citep{R}.
+The aim of the simulation is to determine whether unbiased, confidence valid inferences may be obtained under different levels of non-convergence (\textbf{and at which point convergence may safely be assumed}).
+Additionally, we will assess how well different non-convergence diagnostics perform in identifying signs of non-convergence.
+The results of this study are guidelines for identifying non-convergence in iterative imptation algorithms, which will aid applied researchers in drawing valid inference from incomplete datasets.
+\textbf{{[}add more societal relevance: what is the consequence of non-convergence?
+e.g., biased estimates, invalid inference, etc.{]}} For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular \texttt{mice} package \citep{mice} in \texttt{R} \citep{R}.
 
-\textbf{{[}Include sub-questions and make structure of the paper
-clear!{]}} How can non-convergence be identified diagnostically? Are
-common MCMC non-convergence diagnostics appropriate for MICE? And if so,
-which threshold should be used to diagnose non-convergence? How many
-iterations are sufficient/needed to be able to diagnose non-convergence?
-Are the default number of iterations sufficient (i.e., 5 in mice, 10 in
-SPSS and Stata, 30 in mi)? How severe is it when the algorithm has not
-converged? And what are guidelines for practice? Can the parameter of
-interest \(Q\) be correct when the algorithm is not (yet) converged, and
-vice versa?
+\textbf{{[}Include sub-questions and make structure of the paper clear!{]}} How can non-convergence be identified diagnostically?
+Are common MCMC non-convergence diagnostics appropriate for MICE?
+And if so, which threshold should be used to diagnose non-convergence?
+How many iterations are sufficient/needed to be able to diagnose non-convergence?
+Are the default number of iterations sufficient (i.e., 5 in mice, 10 in SPSS and Stata, 30 in mi)?
+How severe is it when the algorithm has not converged?
+And what are guidelines for practice?
+Can the parameter of interest \(Q\) be correct when the algorithm is not (yet) converged, and vice versa?
 
@@
 
-Let \(y\) denote an \(n \times k\) matrix containing the data values on
-\(k\) variables for all \(n\) units in a sample. The data value of unit
-\(i\) (\(i = 1, 2, \dots, n\)) on variable \(j\)
-(\(j = 1, 2, \dots, k\)) may be either observed or missing. The number
-of units \(i\) in dataset \(y\) with at least one missing value, divided
-by the total number of units \(n\), is called the missingness proportion
-\(p_{mis}\). The collection of observed data values in \(y\) is denoted
-by \(y_{obs}\); the missing part of \(y\) is referred to as \(y_{mis}\).
-For each datapoint in \(y_{mis}\), we sample \(M \times T\) plausible
-values, where \(M\) is the number of imputations
-(\(m = 1, 2, \dots, M\)) and \(T\) is the number of iterations
-(\(t = 1, 2, \dots, T\)) in the imputation algorithm. The collection of
-samples between the initial value (at \(t=1\)) and the final imputed
-value (at \(t=T\)) will be referred to as an `imputation chain'.
-\textbf{\(\theta\)s are scalar summaries of the algorithm at a certain
-iteration (e.g., chain means; the average of the imputed values in each
-imputation chain). }
+Let \(y\) denote an \(n \times k\) matrix containing the data values on \(k\) variables for all \(n\) units in a sample.
+The data value of unit \(i\) (\(i = 1, 2, \dots, n\)) on variable \(j\) (\(j = 1, 2, \dots, k\)) may be either observed or missing.
+The number of units \(i\) in dataset \(y\) with at least one missing value, divided by the total number of units \(n\), is called the missingness proportion \(p_{mis}\).
+The collection of observed data values in \(y\) is denoted by \(y_{obs}\); the missing part of \(y\) is referred to as \(y_{mis}\).
+For each datapoint in \(y_{mis}\), we sample \(M \times T\) plausible values, where \(M\) is the number of imputations (\(m = 1, 2, \dots, M\)) and \(T\) is the number of iterations (\(t = 1, 2, \dots, T\)) in the imputation algorithm.
+The collection of samples between the initial value (at \(t=1\)) and the final imputed value (at \(t=T\)) will be referred to as an `imputation chain'.
+\textbf{\(\theta\)s are scalar summaries of the algorithm at a certain iteration (e.g., chain means; the average of the imputed values in each imputation chain).
+}
 
 \hypertarget{identifying-non-convergence}{%
-\section{Identifying
-non-convergence}\label{identifying-non-convergence}}
+\section{Identifying non-convergence}\label{identifying-non-convergence}}
 
-There are two requirements for convergence of iterative algorithms:
-mixing and stationarity \citep{gelm13}. Without mixing, imputation
-chains do not intermingle nicely, indicating that the distribution of
-imputed values differs across imputations. Chains may be `stuck' at a
-local optimum, instead of sampling imputed values from the entire
-predictive posterior distribution of the missing values. This may cause
-under-estimation of the variance within chains, which results in
-spurious, invalid inferences. Without stationarity, there is trending
-within imputation chains. Trending implies that further iterations would
-yield a different set of imputations. Iterative imputation algorithms
-that have not yet reached stationarity, may thus yield biased estimates.
+There are two requirements for convergence of iterative algorithms: mixing and stationarity \citep{gelm13}.
+Without mixing, imputation chains do not intermingle nicely, indicating that the distribution of imputed values differs across imputations.
+Chains may be `stuck' at a local optimum, instead of sampling imputed values from the entire predictive posterior distribution of the missing values.
+This may cause under-estimation of the variance within chains, which results in spurious, invalid inferences.
+Without stationarity, there is trending within imputation chains.
+Trending implies that further iterations would yield a different set of imputations.
+Iterative imputation algorithms that have not yet reached stationarity, may thus yield biased estimates.
 
-\textbf{{[}Move to INTRO???{]}} To illustrate what non-convergence looks
-like in iterative imputation algorithms, we reproduce an example from
-van Buuren \citeyearpar[\(\S\) 6.5.2]{buur18}. Figure \ref{fig:non-conv}
-displays a subset of the example: a variable that we will call \(j\). We
-see two traceplots of the chain means for \(j\)---i.e., the average of
-the imputed values for \(j\) in each imputation \(y_{imp,m}\), plotted
-against the iteration number. Each line portrays one imputation. The
-plot on the lefthand side of the figure shows typical convergence of an
-iterative imputation algorithm. The righthand side displays pathological
-non-convergence, induced by purposefully mis-specifying the imputation
-model. \textbf{{[}Leave this in if it's after mixing and stationarity
-are defined, otherwise remove?{]} In the typical convergence situation,
-the imputation chains intermingle nicely and there is little to no
-trending. In the non-convergence plot, there is a lot of trending and
-some chains do not intermingle.} Importantly, the chain means at the
-last iteration (the imputed value per \(m\)) are very different between
-the two plots. The algorithm with the mis-specified model yields imputed
-values that are on average a factor two larger than those of the
-typically converged algorithm. That means that non-convergence has a
-impact on the mean of \(j\) in \(y_{imp, m}\). This difference
-(presumably) translates to bias in the \(M\) completed data estimates
-\(\hat{\mu}_{j,m}\), and consequently also to the pooled estimate
-\(\bar{\mu}_j\). Non-convergence thus leads to biased estimates. This
-shows the importance of reaching converged states in iterative
-imputation algorithms.
+\textbf{{[}Move to INTRO???{]}} To illustrate what non-convergence looks like in iterative imputation algorithms, we reproduce an example from van Buuren \citeyearpar[\(\S\) 6.5.2]{buur18}.
+Figure \ref{fig:non-conv} displays a subset of the example: a variable that we will call \(j\).
+We see two traceplots of the chain means for \(j\)---i.e., the average of the imputed values for \(j\) in each imputation \(y_{imp,m}\), plotted against the iteration number.
+Each line portrays one imputation.
+The plot on the lefthand side of the figure shows typical convergence of an iterative imputation algorithm.
+The righthand side displays pathological non-convergence, induced by purposefully mis-specifying the imputation model.
+\textbf{{[}Leave this in if it's after mixing and stationarity are defined, otherwise remove?{]} In the typical convergence situation, the imputation chains intermingle nicely and there is little to no trending.
+In the non-convergence plot, there is a lot of trending and some chains do not intermingle.} Importantly, the chain means at the last iteration (the imputed value per \(m\)) are very different between the two plots.
+The algorithm with the mis-specified model yields imputed values that are on average a factor two larger than those of the typically converged algorithm.
+That means that non-convergence has a impact on the mean of \(j\) in \(y_{imp, m}\).
+This difference (presumably) translates to bias in the \(M\) completed data estimates \(\hat{\mu}_{j,m}\), and consequently also to the pooled estimate \(\bar{\mu}_j\).
+Non-convergence thus leads to biased estimates.
+This shows the importance of reaching converged states in iterative imputation algorithms.
 
@@
 
-\textbf{{[}Move to INTRO???{]}} Convergence has historically been
-inspected visually, by monitoring the imputation algorithm over
-iterations. This is typically done through traceplots. In a traceplot,
-the iteraton number is plotted against a certain \(\theta\). \(\theta\)s
-are scalar summaries of the state space of the algorithm at a specific
-iteration. The recommended scalar summaries to evaluate are chain means
-and chain variances. Additionally, researchers may ``monitor some
-statistic of scientific interest'' \citep[\(\S\) 6.5.2]{buur18}. As
-\citet{buur18} describes, researchers can specify their quantity of
-scientific interest \(Q\) (e.g., a regression coefficient) as scalar
-summary \(\theta\). Such a user-defined scalar summary, however, may be
-somewhat advanced for empirical researchers. And moreover, it is not
-universal to all complete data problems. Focusing on the convergence of
-outcome parameters may influence the iterative imputation procedure in
-the sense that the model of evaluation favors the model of interest. The
-downside to these two approaches is that they either focus on the
-univariate state space, or primarily track the change over the
-iterations of a multivariate outcome conform the scientific model of
-interest. Ideally, one would like to evaluate a model-independent
-parameter that summarizes the multivariate nature of the data.
-Therefore, we propose a \(\theta\) that summarizes the multivariate
-state space of the algorithm, but is independent from the model of
-scientific interest. We propose \(\lambda_1\) as such a scalar summary.
-We define \(\lambda_1\) as the first eigenvalue of the
-variance-covariance matrix of the completed data. Let
-\(\lambda_1 \geq \lambda_2 \geq ... \geq \lambda_j\) be the eigenvalues
-of \(\Sigma\) in each of the \(M\) completed datasets
-\({y_{obs}, y_{imp}}\). \(\lambda_1\) is measure that summarises the
-covariances in the completed datasets. \textbf{The first eigenvalue has
-the appealing property that is not dependent on the substantive model of
-interest. Eigenvalue decomposition is inspired by \citet{mack03}.}
+\textbf{{[}Move to INTRO???{]}} Convergence has historically been inspected visually, by monitoring the imputation algorithm over iterations.
+This is typically done through traceplots.
+In a traceplot, the iteraton number is plotted against a certain \(\theta\).
+\(\theta\)s are scalar summaries of the state space of the algorithm at a specific iteration.
+The recommended scalar summaries to evaluate are chain means and chain variances.
+Additionally, researchers may ``monitor some statistic of scientific interest'' \citep[\(\S\) 6.5.2]{buur18}.
+As \citet{buur18} describes, researchers can specify their quantity of scientific interest \(Q\) (e.g., a regression coefficient) as scalar summary \(\theta\).
+Such a user-defined scalar summary, however, may be somewhat advanced for empirical researchers.
+And moreover, it is not universal to all complete data problems.
+Focusing on the convergence of outcome parameters may influence the iterative imputation procedure in the sense that the model of evaluation favors the model of interest.
+The downside to these two approaches is that they either focus on the univariate state space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest.
+Ideally, one would like to evaluate a model-independent parameter that summarizes the multivariate nature of the data.
+Therefore, we propose a \(\theta\) that summarizes the multivariate state space of the algorithm, but is independent from the model of scientific interest.
+We propose \(\lambda_1\) as such a scalar summary.
+We define \(\lambda_1\) as the first eigenvalue of the variance-covariance matrix of the completed data.
+Let \(\lambda_1 \geq \lambda_2 \geq ...
+\geq \lambda_j\) be the eigenvalues of \(\Sigma\) in each of the \(M\) completed datasets \({y_{obs}, y_{imp}}\).
+\(\lambda_1\) is measure that summarises the covariances in the completed datasets.
+\textbf{The first eigenvalue has the appealing property that is not dependent on the substantive model of interest.
+Eigenvalue decomposition is inspired by \citet{mack03}.}
 
 \hypertarget{diagnostics-under-evaluation}{%
-\subsection{Diagnostics under
-evaluation}\label{diagnostics-under-evaluation}}
+\subsection{Diagnostics under evaluation}\label{diagnostics-under-evaluation}}
 
-Non-convergence diagnostics for MCMC algorithms typically identify
-problems in either one of the two requirements for converegnce: mixing
-or stationarity. Non-stationarity within chains may be diagnosed with
-e.g., autocorrelation \citep[\(AC\);][]{scha97, gelm13}, numeric
-standard error \citep[`MC error';][]{gewe92}, or Raftery and Lewis's
-\citeyearpar{raft91} procedure to determine the effect of trending on
-the precision of estimates. A widely used diagnostic to monitor mixing
-between chains is the potential scale reduction factor \(\widehat{R}\)
-\citep[`Gelman-Rubin statistic';][]{gelm92}. With a recently proposed
-adaptation, \(\widehat{R}\) might also serve to diagnose
-non-stationarity, but this has not yet been thoroughly investigated
-\citep{veht19}. Therefore, we use \(\widehat{R}\) and \(AC\) to evaluate
-mixing and stationarity separately, as recommended by e.g.,
-\citet{cowl96}.
+Non-convergence diagnostics for MCMC algorithms typically identify problems in either one of the two requirements for converegnce: mixing or stationarity.
+Non-stationarity within chains may be diagnosed with e.g., autocorrelation \citep[\(AC\);][]{scha97, gelm13}, numeric standard error \citep[`MC error';][]{gewe92}, or Raftery and Lewis's \citeyearpar{raft91} procedure to determine the effect of trending on the precision of estimates.
+A widely used diagnostic to monitor mixing between chains is the potential scale reduction factor \(\widehat{R}\) \citep[`Gelman-Rubin statistic';][]{gelm92}.
+With a recently proposed adaptation, \(\widehat{R}\) might also serve to diagnose non-stationarity, but this has not yet been thoroughly investigated \citep{veht19}.
+Therefore, we use \(\widehat{R}\) and \(AC\) to evaluate mixing and stationarity separately, as recommended by e.g., \citet{cowl96}.
 
 \hypertarget{potential-scale-reduction-factor}{%
-\subsubsection{Potential scale reduction
-factor}\label{potential-scale-reduction-factor}}
+\subsubsection{Potential scale reduction factor}\label{potential-scale-reduction-factor}}
 
-An updated version of \(\widehat{R}\) has been proposed by Vehtari et
-al.~\citeyearpar[p.~5]{veht19}. \textbf{This version may be suitable for
-iterative imputation.} Let \(M\) be the total number of chains, \(T\)
-the number of iterations per chain, and \(\theta\) the scalar summary of
-interest (e.g., chain means). For each chain (\(m = 1, 2, \dots, M\)),
-we estimate the variance of \(\theta\), and average these to obtain
-within-chain variance \(W\).
+An updated version of \(\widehat{R}\) has been proposed by Vehtari et al.~\citeyearpar[p.~5]{veht19}.
+\textbf{This version may be suitable for iterative imputation.} Let \(M\) be the total number of chains, \(T\) the number of iterations per chain, and \(\theta\) the scalar summary of interest (e.g., chain means).
+For each chain (\(m = 1, 2, \dots, M\)), we estimate the variance of \(\theta\), and average these to obtain within-chain variance \(W\).
 
@@
 
-We then estimate between-chain variance \(B\) as the variance of the
-collection of average \(\theta\) per chain.
+We then estimate between-chain variance \(B\) as the variance of the collection of average \(\theta\) per chain.
 
@@
 
-From the between- and within-chain variances we compute a weighted
-average, \(\widehat{\operatorname{var}}^{+}\), which over-estimates the
-total variance of \(\theta\) \textbf{remove or explain why, or leave
-in}. \(\widehat{R}\) is then obtained as a ratio between the
-over-estimated total variance and the within-chain variance:
+From the between- and within-chain variances we compute a weighted average, \(\widehat{\operatorname{var}}^{+}\), which over-estimates the total variance of \(\theta\) \textbf{remove or explain why, or leave in}.
+\(\widehat{R}\) is then obtained as a ratio between the over-estimated total variance and the within-chain variance:
 
@@
 
-We can interpret \(\widehat{R}\) as potential scale reduction factor
-since it indicates by how much the variance of \(\theta\) could be
-shrunken down if an infinite number of iterations per chain would be run
-\citep{gelm92}. This interpretation assumes that chains are
-`over-dispersed' at \(t=1\), and reach convergence as \(T \to \infty\).
-Over-dispersion implies that the initial values of the chains are `far
-away' from the target distribution and each other. When all chains
-sample independent of their initial values, the mixing component of
-convergence is satisfied, and \(\widehat{R}\)-values will be close to
-one. High \(\widehat{R}\)-values thus indicate non-convergence. The
-conventionally acceptable threshold for convergence was
-\(\widehat{R} < 1.2\) \citep{gelm92}. More recently, \citet{veht19}
-proposed a more stringent threshold of \(\widehat{R} < 1.01\).
+We can interpret \(\widehat{R}\) as potential scale reduction factor since it indicates by how much the variance of \(\theta\) could be shrunken down if an infinite number of iterations per chain would be run \citep{gelm92}.
+This interpretation assumes that chains are `over-dispersed' at \(t=1\), and reach convergence as \(T \to \infty\).
+Over-dispersion implies that the initial values of the chains are `far away' from the target distribution and each other.
+When all chains sample independent of their initial values, the mixing component of convergence is satisfied, and \(\widehat{R}\)-values will be close to one.
+High \(\widehat{R}\)-values thus indicate non-convergence.
+The conventionally acceptable threshold for convergence was \(\widehat{R} < 1.2\) \citep{gelm92}.
+More recently, \citet{veht19} proposed a more stringent threshold of \(\widehat{R} < 1.01\).
 
@@
 
-Following the same notation, we define autocorrelation as the
-correlation between two subsequent \(\theta\)-values within the same
-chain \citep[p.~147]{lync07}. In this study, we only consider \(AC\) at
-lag 1, i.e., the correlation between the \(t^{th}\) and \(t+1^{th}\)
-iteration of the same chain.
+Following the same notation, we define autocorrelation as the correlation between two subsequent \(\theta\)-values within the same chain \citep[p.~147]{lync07}.
+In this study, we only consider \(AC\) at lag 1, i.e., the correlation between the \(t^{th}\) and \(t+1^{th}\) iteration of the same chain.
 
@@
 
-We can interpret \(AC\)-values as a measure of stationarity. If
-\(AC\)-values are close to zero, there is no dependence between
-subsequent samples within imputation chains. Negative \(AC\)-values
-indicate divergence within imputation chains. \textbf{Subsequent sampled
-values within each imputation chain are less alike.} Positive
-\(AC\)-values indicate recurrence. If \(\theta\)-values of subsequent
-iterations are similar, trending may occur. Negative \(AC\)-values show
-no threat to the stationarity component of convergence. On the contrary
-even---negative \(AC\)-values indicate that \(\theta\)-values of
-subsequent iterations diverge from one another, which may increase the
-variance of \(\theta\) and speed up convergence. As convergence
-diagnostic, the interest is therefore in positive \(AC\)-values. The
-magnitude of \(AC\)-values may be evaluated statistically {[}\textbf{add
-reference}{]}.
+We can interpret \(AC\)-values as a measure of stationarity.
+If \(AC\)-values are close to zero, there is no dependence between subsequent samples within imputation chains.
+Negative \(AC\)-values indicate divergence within imputation chains.
+\textbf{Subsequent sampled values within each imputation chain are less alike.} Positive \(AC\)-values indicate recurrence.
+If \(\theta\)-values of subsequent iterations are similar, trending may occur.
+Negative \(AC\)-values show no threat to the stationarity component of convergence.
+On the contrary even---negative \(AC\)-values indicate that \(\theta\)-values of subsequent iterations diverge from one another, which may increase the variance of \(\theta\) and speed up convergence.
+As convergence diagnostic, the interest is therefore in positive \(AC\)-values.
+The magnitude of \(AC\)-values may be evaluated statistically {[}\textbf{add reference}{]}.
 
@@
 
-Upon convergence, \(\widehat{R}=1\) and \(AC=0\), which are unlikely
-thresholds for MCMC algorithms, because of its convergence to a
-distribution. Even in the most converged state, the algorithm will show
-some signs of non-mixing and non-stationarity. Upon \emph{sufficient}
-convergence, the imputation chains will intermingle such that the only
-difference between the chains is caused by the randomness induced by the
-algorithm (\(\widehat{R} > 1\)), and there may be some dependency
-between subsequent iterations of imputation chains (\(AC > 0\)). In
-practice, non-convergence is usually diagnosed when \(\widehat{R}\)
-\textgreater{} 1.2 or 1.1 or even 1.01. And a t-test is performed to
-assess whether \(AC\) is significantly different from zero.
+Upon convergence, \(\widehat{R}=1\) and \(AC=0\), which are unlikely thresholds for MCMC algorithms, because of its convergence to a distribution.
+Even in the most converged state, the algorithm will show some signs of non-mixing and non-stationarity.
+Upon \emph{sufficient} convergence, the imputation chains will intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm (\(\widehat{R} > 1\)), and there may be some dependency between subsequent iterations of imputation chains (\(AC > 0\)).
+In practice, non-convergence is usually diagnosed when \(\widehat{R}\) \textgreater{} 1.2 or 1.1 or even 1.01.
+And a t-test is performed to assess whether \(AC\) is significantly different from zero.
 
@@
 
-Before we evaluate the performance of the non-convergence diagnostics
-\(\widehat{R}=1\) and \(AC=0\) quantitatively through simulation, we
-apply them to the example of pathological non-convergence by
-\citet{buur18} that we reproduced above. We want to be able to draw the
-same conclusions as we did from visually inspecting the two algorithms.
-But \(\widehat{R}=1\) and \(AC=0\) should at least be able to
-distinguish between the two algorithms plotted in Figure
-\ref{fig:non-conv}. From visual inspection, we know that the typically
-converged algorithm initially portrayed some signs of non-mixing (around
-\(t=2\)), but intermigled nicely overall. Additionally, there was very
-little trending. The algorithm with pathological non-convergence showed
-severe non-mixing, although this gradually improved (from \(t=7\)).
-Moreover, there was a lot of trending as well initially (up-to \(t=6\)),
-after which the chains reached a somewhat more stationary state. To
-assess whether \(\widehat{R}\) and \(AC\) may be appropriate
-non-convergence identifiers for iterative imputation algorithms, they
-should identify the same trends as the visual inspection. But at least,
-the following condition must hold: the methods should indicate worse
-performance for the mis-specified model with pathological
-non-convergence than the typically converged algorithm (i.e., higher
-\(\widehat{R}\)- and \(AC\)-values).
+Before we evaluate the performance of the non-convergence diagnostics \(\widehat{R}=1\) and \(AC=0\) quantitatively through simulation, we apply them to the example of pathological non-convergence by \citet{buur18} that we reproduced above.
+We want to be able to draw the same conclusions as we did from visually inspecting the two algorithms.
+But \(\widehat{R}=1\) and \(AC=0\) should at least be able to distinguish between the two algorithms plotted in Figure \ref{fig:non-conv}.
+From visual inspection, we know that the typically converged algorithm initially portrayed some signs of non-mixing (around \(t=2\)), but intermigled nicely overall.
+Additionally, there was very little trending.
+The algorithm with pathological non-convergence showed severe non-mixing, although this gradually improved (from \(t=7\)).
+Moreover, there was a lot of trending as well initially (up-to \(t=6\)), after which the chains reached a somewhat more stationary state.
+To assess whether \(\widehat{R}\) and \(AC\) may be appropriate non-convergence identifiers for iterative imputation algorithms, they should identify the same trends as the visual inspection.
+But at least, the following condition must hold: the methods should indicate worse performance for the mis-specified model with pathological non-convergence than the typically converged algorithm (i.e., higher \(\widehat{R}\)- and \(AC\)-values).
 
-In Figure \ref{fig:diagnostics}A, the chain means from Figure
-\ref{fig:non-conv} are plotted again---now together. Panel B shows two
-versions of calculating \(AC\)s applied to the \(\theta\)s of panel A:
-the default calculation with \texttt{R} function \texttt{stats::acf()}
-\citep{R}, and manual calculation as the correlation between \(\theta\)
-in iteration \(t\) and \(\theta\) in iteration \(t+1\). Panel C shows
-the traditional computation of \(\widehat{R}\), and panel D shows
-\(\widehat{R}\) as computed by implementing \citet{veht19} 's
-recommendations.
+In Figure \ref{fig:diagnostics}A, the chain means from Figure \ref{fig:non-conv} are plotted again---now together.
+Panel B shows two versions of calculating \(AC\)s applied to the \(\theta\)s of panel A: the default calculation with \texttt{R} function \texttt{stats::acf()} \citep{R}, and manual calculation as the correlation between \(\theta\) in iteration \(t\) and \(\theta\) in iteration \(t+1\).
+Panel C shows the traditional computation of \(\widehat{R}\), and panel D shows \(\widehat{R}\) as computed by implementing \citet{veht19} 's recommendations.
 
@@
 
-When we look at panel B, we conclude something weird. The \(AC\)-values
-calculated with the default function indicate equal performance for the
-typical convergence and the pathological non-convergence (up-to
-\(t=5\)), while there is obvious trending in the \(\theta\)s of the
-latter. Moreover, the best convergence (as indicated by the lowest
-\(AC\)-value) is observed at \(t=2\), but looking at the chain means in
-panel A, there should be some signs of trending up-to iteration number
-seven. After consulting the documentation on \texttt{stats::acf()}
-\citep{R}, we conclude that this \(AC\) function is not suited for
-iterative imputation algorithms. The function is optimized for
-performance when \(t\geq50\) \citep{box15}, while the default number of
-iterations in iterative imputation is often much lower. Therefore, we
-compute \(AC\) manually, see the solid line. These \(AC\) values
-indicate non-stationarity as expected. We therefore calculate \(AC\)
-manually.
+When we look at panel B, we conclude something weird.
+The \(AC\)-values calculated with the default function indicate equal performance for the typical convergence and the pathological non-convergence (up-to \(t=5\)), while there is obvious trending in the \(\theta\)s of the latter.
+Moreover, the best convergence (as indicated by the lowest \(AC\)-value) is observed at \(t=2\), but looking at the chain means in panel A, there should be some signs of trending up-to iteration number seven.
+After consulting the documentation on \texttt{stats::acf()} \citep{R}, we conclude that this \(AC\) function is not suited for iterative imputation algorithms.
+The function is optimized for performance when \(t\geq50\) \citep{box15}, while the default number of iterations in iterative imputation is often much lower.
+Therefore, we compute \(AC\) manually, see the solid line.
+These \(AC\) values indicate non-stationarity as expected.
+We therefore calculate \(AC\) manually.
 
-\textbf{Describe panel C here and why it is not optimal{]}} The
-\(\widehat{R}\)-values in panel D do meet the requirements specified
-above. \(\widehat{R}\) as computed conform \citet{veht19} indeed
-indicate less signs of non-convergence as the number of iterations goes
-up.
+\textbf{Describe panel C here and why it is not optimal{]}} The \(\widehat{R}\)-values in panel D do meet the requirements specified above.
+\(\widehat{R}\) as computed conform \citet{veht19} indeed indicate less signs of non-convergence as the number of iterations goes up.
 
@@
 
-The aim of the simulation study is to evaluate the impact of inducing
-non-convergence in the MICE algorithm on several quantities of
-scientific interest \(Q\). Subsequently we will evaluate how well
-\(\widehat{R}\) and \(AC\) perform in identifying the effects of
-non-convergence \textbf{{[}and performance of different thetas{]}}. And
-finally, we will formulate an informed advice on the requirements to
-safely assume \emph{sufficient} convergence in practice. That is, we
-assume that the algorithm is \emph{sufficiently} converged when each
-estimate \(\bar{Q}\) is an unbiased and confidence valid estimate of the
-corresponding estimand \(Q\). \textbf{{[}and formulate advice on
-thetas?{]}}
+The aim of the simulation study is to evaluate the impact of inducing non-convergence in the MICE algorithm on several quantities of scientific interest \(Q\).
+Subsequently we will evaluate how well \(\widehat{R}\) and \(AC\) perform in identifying the effects of non-convergence \textbf{{[}and performance of different thetas{]}}.
+And finally, we will formulate an informed advice on the requirements to safely assume \emph{sufficient} convergence in practice.
+That is, we assume that the algorithm is \emph{sufficiently} converged when each estimate \(\bar{Q}\) is an unbiased and confidence valid estimate of the corresponding estimand \(Q\).
+\textbf{{[}and formulate advice on thetas?{]}}
 
-Non-convergence will be induced by: 1) increasing the missingness
-proportion \(p_{mis}\) in dataset \(y\) incrementally, and 2)
-terminating the imputation algorithm at a varying number of iterations
-\(T\). The first set of simulation conditions---the missingness
-proportions---is chosen to reflect the difficulty of the missingness
-problem. The underlying assumption is that low missingness proportions
-lead to quick algorithmic convergence, since there is a lot of
-information in the observed data. Higher missingness proportions should
-yield slower convergence. Unless, however, the fraction of missing
-information is so high that the random component in the imputation
-algorithm outweighs the information in the observed data. Then,
-convergence to a stable but highly variable state may be reached
-instanteneously. We assume that the incremental missingness proportions
-in our study will result in a corresponding increase in signs of
-non-convergence.
+Non-convergence will be induced by: 1) increasing the missingness proportion \(p_{mis}\) in dataset \(y\) incrementally, and 2) terminating the imputation algorithm at a varying number of iterations \(T\).
+The first set of simulation conditions---the missingness proportions---is chosen to reflect the difficulty of the missingness problem.
+The underlying assumption is that low missingness proportions lead to quick algorithmic convergence, since there is a lot of information in the observed data.
+Higher missingness proportions should yield slower convergence.
+Unless, however, the fraction of missing information is so high that the random component in the imputation algorithm outweighs the information in the observed data.
+Then, convergence to a stable but highly variable state may be reached instanteneously.
+We assume that the incremental missingness proportions in our study will result in a corresponding increase in signs of non-convergence.
 
-The assumption inherent to the second set of simulation conditions---the
-number of iterations---is that terminating the imputation algorithm too
-early causes non-convergence. Generally, the algorithm will not reach
-convergence if \(T=1\), because the imputed values in the first
-iteration (at \(t=1\)) depend on the initial values of the algorithm
-(which are sampled randomly from the set of observed datapoints). As the
-number of iterations increases, the imputation chains should become
-independent of the initial values, until the point at which adding an
-extra iteration does not lead to a more converged state. We assume that
-we can induce non-convergence at least in conditions where \(T\) is
-smaller than the default number of iterations in \texttt{mice}, \(T=5\).
+The assumption inherent to the second set of simulation conditions---the number of iterations---is that terminating the imputation algorithm too early causes non-convergence.
+Generally, the algorithm will not reach convergence if \(T=1\), because the imputed values in the first iteration (at \(t=1\)) depend on the initial values of the algorithm (which are sampled randomly from the set of observed datapoints).
+As the number of iterations increases, the imputation chains should become independent of the initial values, until the point at which adding an extra iteration does not lead to a more converged state.
+We assume that we can induce non-convergence at least in conditions where \(T\) is smaller than the default number of iterations in \texttt{mice}, \(T=5\).
 
@@
 \item
-  We expect that simulation conditions with a high missingness
-  proportion \(p_{mis}\) and a low number of iterations \(T\) will
-  result in biased, invalid estimates of the quantities of scientific
-  interest, \(Q\)s.
+  We expect that simulation conditions with a high missingness proportion \(p_{mis}\) and a low number of iterations \(T\) will result in biased, invalid estimates of the quantities of scientific interest, \(Q\)s.
 \item
-  We hypothesize that \(\widehat{R}\) and \(AC\) will correctly identify
-  signs of non-convergence in those simulation conditions where the
-  \(\bar{Q}\)s are \emph{not} unbiased and confidence valid estimates of
-  the \(Q\)s.
+  We hypothesize that \(\widehat{R}\) and \(AC\) will correctly identify signs of non-convergence in those simulation conditions where the \(\bar{Q}\)s are \emph{not} unbiased and confidence valid estimates of the \(Q\)s.
 \item
-  We hypothesize that the recommended thresholds to diagnose
-  non-convergence with \(\widehat{R}\) (\(\widehat{R} > 1.2\),
-  \(\widehat{R} > 1.1\), and \(\widehat{R} > 1.01\)) may be too
-  stringent for iterative imputation applications. In an empirical
-  study, where \(\widehat{R}\) was used to inform the required
-  imputation chain length, it took as many as 50 iterations to overcome
-  the conventional non-convergence threshold \(\widehat{R}>1.2\). Yet,
-  scientific estimates were insensitive to continued iteration from
-  \(t=5\) onwards \citep{lace07}. We therefore suspect that
-  \(\widehat{R}\) may over-estimate signs of non-convergence in
-  iterative imputation algorithms, compared to the bias the estimates.
-  In contrast to this, it may occur that signs of non-convergence are
-  under-estimated by \(\widehat{R}\), in exceptional cases where the
-  initial values of the algorithm are not appropriately over-dispersed
-  \citep[p.~437]{broo98}. In e.g.~\texttt{mice}, initial values are
-  chosen randomly from the observed data, hence we cannot be certain of
-  over-dispersion in the initial values. In practice, we do not expect
-  this to cause problems for identifying non-convergence with
-  \(\widehat{R}\).
+  We hypothesize that the recommended thresholds to diagnose non-convergence with \(\widehat{R}\) (\(\widehat{R} > 1.2\), \(\widehat{R} > 1.1\), and \(\widehat{R} > 1.01\)) may be too stringent for iterative imputation applications.
+In an empirical study, where \(\widehat{R}\) was used to inform the required imputation chain length, it took as many as 50 iterations to overcome the conventional non-convergence threshold \(\widehat{R}>1.2\).
+Yet, scientific estimates were insensitive to continued iteration from \(t=5\) onwards \citep{lace07}.
+We therefore suspect that \(\widehat{R}\) may over-estimate signs of non-convergence in iterative imputation algorithms, compared to the bias the estimates.
+In contrast to this, it may occur that signs of non-convergence are under-estimated by \(\widehat{R}\), in exceptional cases where the initial values of the algorithm are not appropriately over-dispersed \citep[p.~437]{broo98}.
+In e.g.~\texttt{mice}, initial values are chosen randomly from the observed data, hence we cannot be certain of over-dispersion in the initial values.
+In practice, we do not expect this to cause problems for identifying non-convergence with \(\widehat{R}\).
 \item
-  We expect that high \(AC\) values are implausible in iterative
-  imputation algorithms with typical convergence. That is, after only a
-  few iterations, the randomness induced by the algorithm should
-  effectively mitigate the risk of dependency within chains.
+  We expect that high \(AC\) values are implausible in iterative imputation algorithms with typical convergence.
+That is, after only a few iterations, the randomness induced by the algorithm should effectively mitigate the risk of dependency within chains.
 \item
@@
 
-Convergence of the MICE algorithm is investigated through model-based
-simulation in \texttt{R} \citep[version 3.6.3;][]{R}. The simulation
-set-up is summarized in the pseudo-code below. The complete \texttt{R}
-script of the simulation study is available from
-\href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy}{github.com/gerkovink/shinyMice}.
+Convergence of the MICE algorithm is investigated through model-based simulation in \texttt{R} \citep[version 3.6.3;][]{R}.
+The simulation set-up is summarized in the pseudo-code below.
+The complete \texttt{R} script of the simulation study is available from \href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy}{github.com/gerkovink/shinyMice}.
 
@@
 \hypertarget{data-generating-mechanism.}{%
-\subsubsection{Data-generating
-mechanism.}\label{data-generating-mechanism.}}
+\subsubsection{Data-generating mechanism.}\label{data-generating-mechanism.}}
 
-In this study, sampling variance is not of interest. Therefore, a single
-complete dataset may serve as comparative truth in all simulation
-repetitions \citep{vink14}. The data-generating mechanism is a
-multivariate normal distribution, representing person-data on three
-predictor variables in a multiple linear regression problem
-\textbf{(from an unspecified social scientific field of study)}. Let the
-predictor space be defined as \[
+In this study, sampling variance is not of interest.
+Therefore, a single complete dataset may serve as comparative truth in all simulation repetitions \citep{vink14}.
+The data-generating mechanism is a multivariate normal distribution, representing person-data on three predictor variables in a multiple linear regression problem \textbf{(from an unspecified social scientific field of study)}.
+Let the predictor space be defined as
+\[
 \begin{aligned}
@@
 \end{aligned}
-\] A finite population of \(N=1000\) is simulated using the
-\texttt{mvtnorm} package \citep{mvtnorm}. Subsequently, a fourth
-variable is constructed as outcome \(Y\). For each unit
-\(i = 1, 2,..., N\), let \[
+\]
+A finite population of \(N=1000\) is simulated using the \texttt{mvtnorm} package \citep{mvtnorm}.
+Subsequently, a fourth variable is constructed as outcome \(Y\).
+For each unit \(i = 1, 2,..., N\), let
+\[
 Y_i = 1 + 2X_{1i} +.5X_{2i} - X_{3i} + \epsilon_i ,
-\] where \(\epsilon \sim \mathcal{N}(0, 100)\). From the complete set,
-we obtain the true values of several quantities of scientific interest,
-\(Q\)s. \textbf{{[}MOVE THIS NEXT PART TO TOP OF SIMULATION SECTION?{]}}
-\(Q\)s are the targets that will be estimated in each simulation
-repetition, after first \emph{amputing} the complete data with varying
-missingness proportions, then imputing the missingness with a varying
-number of iterations, then estimating the \(Q\)s on the \(M\) completed
-datasets, and finally pooling the \(M\) estimates \(\hat{Q}\) to obtain
-a single \(\bar{Q}\) for each \(Q\). \textbf{We try to predict the
-effect of non-converegence (i.e., the difference between \(Q\) and
-\(\bar{Q}\)) with the convergence diagnostics \(\widehat{R}\) and
-\(AC\), by appliying these methods to several scalar summaries of the
-state space of the algorithm, \(\theta\)s.}
+\]
+where \(\epsilon \sim \mathcal{N}(0, 100)\).
+From the complete set, we obtain the true values of several quantities of scientific interest, \(Q\)s.
+\textbf{{[}MOVE THIS NEXT PART TO TOP OF SIMULATION SECTION?{]}} \(Q\)s are the targets that will be estimated in each simulation repetition, after first \emph{amputing} the complete data with varying missingness proportions, then imputing the missingness with a varying number of iterations, then estimating the \(Q\)s on the \(M\) completed datasets, and finally pooling the \(M\) estimates \(\hat{Q}\) to obtain a single \(\bar{Q}\) for each \(Q\).
+\textbf{We try to predict the effect of non-converegence (i.e., the difference between \(Q\) and \(\bar{Q}\)) with the convergence diagnostics \(\widehat{R}\) and \(AC\), by appliying these methods to several scalar summaries of the state space of the algorithm, \(\theta\)s.}
 
@@
 
-We consider four types of \(Q\)s that are often of interest in empirical
-research. Namely, two descriptive statistics: the mean \(\mu_j\) and
-standard deviation \(\sigma_j\) of each variable
-\(j = Y, X_1, X_2, X_3\). We also consider the regression coefficients
-of the predictors: \(\beta_1\), \(\beta_2\), and \(\beta_3\) in
-regression equation
-\[Y' = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3,\] where \(Y'\)
-is the expected value of the outcome. And finally, the proportion of
-variance explained by the set of predictors: coefficient of
-determination \(r^2 \times 100\) \textbf{(note that lower case \(r\) is
-used to avoid confusion with non-convergence diagnostic
-\(\widehat{R}\))}. The true values of these scientific estimands are
-calculated on the complete data with the \texttt{R} functions
-\texttt{base::mean()}, \texttt{stats::sd()}, and \texttt{stats::lm()}
-\citep{R}.
+We consider four types of \(Q\)s that are often of interest in empirical research.
+Namely, two descriptive statistics: the mean \(\mu_j\) and standard deviation \(\sigma_j\) of each variable \(j = Y, X_1, X_2, X_3\).
+We also consider the regression coefficients of the predictors: \(\beta_1\), \(\beta_2\), and \(\beta_3\) in regression equation
+\[Y' = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3,\]
+where \(Y'\) is the expected value of the outcome.
+And finally, the proportion of variance explained by the set of predictors: coefficient of determination \(r^2 \times 100\) \textbf{(note that lower case \(r\) is used to avoid confusion with non-convergence diagnostic \(\widehat{R}\))}.
+The true values of these scientific estimands are calculated on the complete data with the \texttt{R} functions \texttt{base::mean()}, \texttt{stats::sd()}, and \texttt{stats::lm()} \citep{R}.
 
@@
 
-In each simulation repetition, we \emph{ampute} the complete dataset
-five times to obtain five different incomplete datasets, \(y\)s. The
-ratio between \(y_{obs}\) and \(y_{mis}\) in each \(y\) depends on the
-missingness proportion: \(p_{mis} =.05,.25,.5,.75,.95\). We ampute the
-complete data with the \texttt{mice} package \citep[function
-\texttt{mice::ampute()};][]{mice}. We consider all possible univariate
-and multivariate patterns of missingness, conform a `missing completely
-at random' missingness mechanism \citep{rubin87}. I.e., the probability
-to be missing is the same for all \(N \times k\) cells in \(y\),
-conditional on the missingness proportion \(p_{mis}\).
+In each simulation repetition, we \emph{ampute} the complete dataset five times to obtain five different incomplete datasets, \(y\)s.
+The ratio between \(y_{obs}\) and \(y_{mis}\) in each \(y\) depends on the missingness proportion: \(p_{mis} =.05,.25,.5,.75,.95\).
+We ampute the complete data with the \texttt{mice} package \citep[function \texttt{mice::ampute()};][]{mice}.
+We consider all possible univariate and multivariate patterns of missingness, conform a `missing completely at random' missingness mechanism \citep{rubin87}.
+I.e., the probability to be missing is the same for all \(N \times k\) cells in \(y\), conditional on the missingness proportion \(p_{mis}\).
 
-Missing datapoints in each incomplete dataset \(y\) are imputed with the
-\texttt{mice()} function \citep{mice}. All imputation procedures are
-performed with Bayesian linear regression imputation, and five
-imputation chains (\(M=5\)). Because \(M=5\), each run of the
-\texttt{mice()} function results in five sets of imputations:
-\(y_{imp,m}\), where \(m = 1, 2, ..., 5\). The number of iterations in
-each run of the algorithm varies between simulation conditions
-(\(T=1,2,...,100\)).
+Missing datapoints in each incomplete dataset \(y\) are imputed with the \texttt{mice()} function \citep{mice}.
+All imputation procedures are performed with Bayesian linear regression imputation, and five imputation chains (\(M=5\)).
+Because \(M=5\), each run of the \texttt{mice()} function results in five sets of imputations: \(y_{imp,m}\), where \(m = 1, 2, ..., 5\).
+The number of iterations in each run of the algorithm varies between simulation conditions (\(T=1,2,...,100\)).
 
-From the 5 imputations \(y_{imp,m}\) we obtain the chain means and chain
-variances for each variable \(j\). These will serve as scalar summary
-\(\theta\) to apply non-convergence diagnostics \(\widehat{R}\) and
-\(AC\) on. We then use \texttt{mice::complete()} to combine the imputed
-data \(y_{imp,m}\) with the observed data \(y_{obs}\), resulting in five
-completed datasets \{\(y_{obs}, y_{imp, m}\)\}.
+From the 5 imputations \(y_{imp,m}\) we obtain the chain means and chain variances for each variable \(j\).
+These will serve as scalar summary \(\theta\) to apply non-convergence diagnostics \(\widehat{R}\) and \(AC\) on.
+We then use \texttt{mice::complete()} to combine the imputed data \(y_{imp,m}\) with the observed data \(y_{obs}\), resulting in five completed datasets \{\(y_{obs}, y_{imp, m}\)\}.
 
@@
 \item
-  Estimate \(\hat{Q}\)s for \(Q= \mu\) and \(Q=\sigma\) with
-  \texttt{base::mean()}, and \texttt{stats::sd()} \citep{R}. We pool the
-  \(\hat{Q}\)s with \texttt{base::mean()} to get \(\bar{Q}\)s.
+  Estimate \(\hat{Q}\)s for \(Q= \mu\) and \(Q=\sigma\) with \texttt{base::mean()}, and \texttt{stats::sd()} \citep{R}. We pool the \(\hat{Q}\)s with \texttt{base::mean()} to get \(\bar{Q}\)s.
 \item
-  Compute \(\lambda_1\), to use as \(\theta\) in the two convergence
-  diagnostics. By definition, the first eigenvalue of the
-  variance-covariance matrix, \(\lambda_1\), is equal to the variance of
-  the principal component solution of each \{\(y_{obs}, y_{imp, m}\)\},
-  which we calculate with \texttt{stats::princomp()} \citep{R}.
+  Compute \(\lambda_1\), to use as \(\theta\) in the two convergence diagnostics.
+By definition, the first eigenvalue of the variance-covariance matrix, \(\lambda_1\), is equal to the variance of the principal component solution of each \{\(y_{obs}, y_{imp, m}\)\}, which we calculate with \texttt{stats::princomp()} \citep{R}.
 \item
-  Perform multiple linear regression with \texttt{stats::lm()} \citep{R}
-  to get the \(\hat{Q}\)s for \(Q=\beta\) and \(Q=r^2\). We pool the
-  \(\hat{Q}\)s conform \citet{vink14} to get \(\bar{Q}\)s for the
-  \(\beta\)s and \(r^2\) \textbf{{[}check pooling of \(r^2\), instead of
-  \texttt{mice::pool.r.squared()}{]}}. Additionally, we use the
-  \(\hat{Q}\)s of \(Q=\beta\) as \(\theta\) to apply the convergence
-  diagnostics to.
+  Perform multiple linear regression with \texttt{stats::lm()} \citep{R} to get the \(\hat{Q}\)s for \(Q=\beta\) and \(Q=r^2\).
+We pool the \(\hat{Q}\)s conform \citet{vink14} to get \(\bar{Q}\)s for the \(\beta\)s and \(r^2\) \textbf{{[}check pooling of \(r^2\), instead of \texttt{mice::pool.r.squared()}{]}}.
+Additionally, we use the \(\hat{Q}\)s of \(Q=\beta\) as \(\theta\) to apply the convergence diagnostics to.
 \end{itemize}
 
-We apply non-convergence diagnostics \(\widehat{R}\) and \(AC\) to each
-scalar summary \(\theta\). We calculate \(\widehat{R}\) by implementing
-Vehtari et al.'s \citeyearpar{veht19} recommendations, and \(AC\) as the
-correlation between the \(t^{th}\) and the \((t+1)^{th}\) iteration.
+We apply non-convergence diagnostics \(\widehat{R}\) and \(AC\) to each scalar summary \(\theta\).
+We calculate \(\widehat{R}\) by implementing Vehtari et al.'s \citeyearpar{veht19} recommendations, and \(AC\) as the correlation between the \(t^{th}\) and the \((t+1)^{th}\) iteration.
 
@@
 
-We assess the impact of inducing non-convergence in the iterative
-imputation algorithm by comparing \(\bar{Q}\)s with \(Q\)s. For each
-\(Q\), we calculate bias as \(\bar{Q} - Q\). For the regression
-coefficients \(Q=\beta_{1,2,3}\), we also compute the empirical coverage
-rate (CR). CR is defined as the percentage of simulation repetitions in
-which the 95\% confidence interval (CI) around \(\bar{Q}\) covers the
-true estimand \(Q\). Let
-\[\text{CI} = \bar{Q} \pm t_{(M-1)} \times SE_{\bar{Q}},\] where
-\(t_{(M-1)}\) is the quantile of a \(t\)-distribution with \(M-1\)
-degrees of freedom, and \(SE_{\bar{Q}}\) is the square root of the
-pooled variance estimate. If we obtain nominal coverage (CR = 95\%), we
-can conclude that the \(\bar{Q}\)s are conficence valid estimates of the
-\(Q\)s. Finally, we inspect CI width (CIW): the difference between the
-lower and upper bound of the 95\% confidence interval around
-\(\bar{Q}\). CIW is of interest because too short CIs indicate
-under-estimation of the variance in \(\bar{Q}\), which may yield
-spurious inferences.
+We assess the impact of inducing non-convergence in the iterative imputation algorithm by comparing \(\bar{Q}\)s with \(Q\)s.
+For each \(Q\), we calculate bias as \(\bar{Q} - Q\).
+For the regression coefficients \(Q=\beta_{1,2,3}\), we also compute the empirical coverage rate (CR).
+CR is defined as the percentage of simulation repetitions in which the 95\% confidence interval (CI) around \(\bar{Q}\) covers the true estimand \(Q\).
+Let
+\[\text{CI} = \bar{Q} \pm t_{(M-1)} \times SE_{\bar{Q}},\]
+where \(t_{(M-1)}\) is the quantile of a \(t\)-distribution with \(M-1\) degrees of freedom, and \(SE_{\bar{Q}}\) is the square root of the pooled variance estimate.
+If we obtain nominal coverage (CR = 95\%), we can conclude that the \(\bar{Q}\)s are conficence valid estimates of the \(Q\)s.
+Finally, we inspect CI width (CIW): the difference between the lower and upper bound of the 95\% confidence interval around \(\bar{Q}\).
+CIW is of interest because too short CIs indicate under-estimation of the variance in \(\bar{Q}\), which may yield spurious inferences.
 
-With these performance measures, we also evaluate how well the
-non-convergence diagnostics \(\widehat{R}\) and \(AC\) can identify
-simulation conditions in which the \(\bar{Q}\)s are \emph{not} unbiased,
-confidence valid estimates of the \(Q\)s. \textbf{Rephrase and add
-corresponding \(Q\)s:} We measure the performance of \(\widehat{R}\) and
-\(AC\) by the four scalar summaries \(\theta\): chain means (i.e., the
-mean in each \(y_{imp, m, j}\)) versus \(Q=\mu\), chain variances (i.e.,
-the variance in each \(y_{imp, m, j}\)) versus \(Q=\sigma\), the
-estimated regression coefficients \(\beta\) in each
-\{\(y_{obs}, y_{imp, m}\)\} (i.e., a user-defined statistic of
-scientific interest \(\hat{Q}\)) against \(Q=\beta\), and the first
-eigenvalue of the variance-covariance matrix in each
-\{\(y_{obs}, y_{imp, m}\)\}, \(\lambda_1\) versus \(Q=r^2\).
+With these performance measures, we also evaluate how well the non-convergence diagnostics \(\widehat{R}\) and \(AC\) can identify simulation conditions in which the \(\bar{Q}\)s are \emph{not} unbiased, confidence valid estimates of the \(Q\)s.
+\textbf{Rephrase and add corresponding \(Q\)s:} We measure the performance of \(\widehat{R}\) and \(AC\) by the four scalar summaries \(\theta\): chain means (i.e., the mean in each \(y_{imp, m, j}\)) versus \(Q=\mu\), chain variances (i.e., the variance in each \(y_{imp, m, j}\)) versus \(Q=\sigma\), the estimated regression coefficients \(\beta\) in each \{\(y_{obs}, y_{imp, m}\)\} (i.e., a user-defined statistic of scientific interest \(\hat{Q}\)) against \(Q=\beta\), and the first eigenvalue of the variance-covariance matrix in each \{\(y_{obs}, y_{imp, m}\)\}, \(\lambda_1\) versus \(Q=r^2\).
 
@@
 
-\textbf{{[}Add to this section: 1) emphasize that the iterations in the
-plots are averages across repetitions, not within one run of MICE; 2)
-sample effects due to single complete dataset?? 3) Add more info about
-figure legends and axes.{]}} Primarily, we want to know what the effect
-is of inducing non-convergence on the estimated \(Q\)s. Specifically, we
-are interested in the bias of \(\bar{Q}\) for all \(Q\)s and the
-confidence validity of \(\bar{Q}\) for \(Q=\beta\). Therefore, we plot
-these performance measures against the simulation conditions (the number
-of iterations \(T\) and the missingness proportion \(p_{mis}\)). We look
-at averages across simulation repetitions.
+\textbf{{[}Add to this section: 1) emphasize that the iterations in the plots are averages across repetitions, not within one run of MICE; 2) sample effects due to single complete dataset??
+3) Add more info about figure legends and axes 4) the ac and rhat are independent of missingness, not the thetas.{]}}
+Primarily, we want to know what the effect is of inducing non-convergence on the estimated \(Q\)s.
+Specifically, we are interested in the bias of \(\bar{Q}\) for all \(Q\)s and the confidence validity of \(\bar{Q}\) for \(Q=\beta\).
+Therefore, we plot these performance measures against the simulation conditions (the number of iterations \(T\) and the missingness proportion \(p_{mis}\)).
+We look at averages across simulation repetitions.
 
-Secondly, we want to know if conditions in which the \(Q\)s are affected
-by non-convergence are also identified by the non-convergance
-diagnostics \(\widehat{R}\) and \(AC\). We do this, again, by looking at
-the performance measures for the \(Q\)s, but now we evaluate these
-against what \(\widehat{R}\) and \(AC\) indicate about signs of
-non-convergence in the \(\theta\)s.
+Secondly, we want to know if conditions in which the \(Q\)s are affected by non-convergence are also identified by the non-convergance diagnostics \(\widehat{R}\) and \(AC\).
+We do this, again, by looking at the performance measures for the \(Q\)s, but now we evaluate these against what \(\widehat{R}\) and \(AC\) indicate about signs of non-convergence in the \(\theta\)s.
 
-\textbf{{[}In short:{]}} - Is there more bias in conditions where
-non-convergence was induced? (NOPE for univariates, YES for multivar.).
+\textbf{{[}In short:{]}}
+- Is there more bias in conditions where non-convergence was induced? (NOPE for univariates, YES for multivar.).
 
@@
 \item
-  Is the point of \emph{sufficient} convergence (at which the
-  \(\bar{Q}\)s are unbiased and confidence valid estimates of the
-  \(Q\)s) the same point as identified by the non-convergence
-  diagnostics? (NOPE, they are too conservative).
+  Is the point of \emph{sufficient} convergence (at which the \(\bar{Q}\)s are unbiased and confidence valid estimates of the \(Q\)s) the same point as identified by the non-convergence diagnostics?
+(NOPE, they are too conservative).
 \item
-  Is one of the thetas better? (NAH, doesn't really matter, probably
-  multivariate is better).
+  Is one of the thetas better? (NAH, doesn't really matter, probably multivariate is better).
 \item
@@
 
-For reasons of brevity, we only discuss the convergence diagnostics for
-the \(Q\)s with the worst performance in terms of bias. Bias in the
-descriptve statistics (\(Q=\mu\) and \(Q=\sigma\)) was most pronounced
-in the outcome variable \(Y\). Of the regression coefficients,
-\(Q=\beta\), the largest bias was observed in the effect of \(X_1\) on
-\(Y\), \(\beta_1\). Results for \(Q=r^2\) are reported as-is. In the
-figures, however, we only present simulation conditions where
-\(T\leq50\). Full results are available from \textbf{{[}link github
-repo{]}}.
+For reasons of brevity, we only discuss the convergence diagnostics for the \(Q\)s with the worst performance in terms of bias.
+Bias in the descriptve statistics (\(Q=\mu\) and \(Q=\sigma\)) was most pronounced in the outcome variable \(Y\).
+Of the regression coefficients, \(Q=\beta\), the largest bias was observed in the effect of \(X_1\) on \(Y\), \(\beta_1\).
+Results for \(Q=r^2\) are reported as-is.
+In the figures, however, we only present simulation conditions where \(T\leq50\).
+Full results are available from \textbf{{[}link github repo{]}}.
 
@@
 
-As expected, conditions with a higher proportion of missingness and/or a
-lower number of iterations show more signs of non-convergence. I.e., on
-average these conditions portray more extreme bias in the estimated
-\(Q\)s, and non-nominal coverages. Roughly speaking this means that MICE
-is indeed not converged at \(t=1\), and converges gradually as \(t\)
-increases to \(T\). The point at which an additional iteration does not
-lead to an improvement of the estimates depends on the difficulty of the
-missingness problem---between XYZ and XYZ. This means that the algorithm
-has converged sufficiently (\textbf{under the current specifications}).
+As expected, conditions with a higher proportion of missingness and/or a lower number of iterations show more signs of non-convergence.
+I.e., on average these conditions portray more extreme bias in the estimated \(Q\)s, and non-nominal coverages.
+Roughly speaking this means that MICE is indeed not converged at \(t=1\), and converges gradually as \(t\) increases to \(T\).
+The point at which an additional iteration does not lead to an improvement of the estimates depends on the difficulty of the missingness problem---between XYZ and XYZ.
+This means that the algorithm has converged sufficiently (\textbf{under the current specifications}).
 
-As required for a method to diagnose non-convergence, the values of
-\(\widehat{R}\) follow a trend similar to the performance measures.
-\textbf{Add split between convergence diagnostics versus performance
-measures, and convergence diagnostics versus simulation conditions.}
-\(\widehat{R}\) values are generally lower in conditions with a higher
-number of iterations, and somewhat higher in conditions with a higher
-percentage of missing data. Autocorrelation values were indeed
-decreasing with a higher number of iterations, and lower in conditions
-with less missingness.
+As required for a method to diagnose non-convergence, the values of \(\widehat{R}\) follow a trend similar to the performance measures.
+\textbf{Add split between convergence diagnostics versus performance measures, and convergence diagnostics versus simulation conditions.} \(\widehat{R}\) values are generally lower in conditions with a higher number of iterations, and somewhat higher in conditions with a higher percentage of missing data.
+Autocorrelation values were indeed decreasing with a higher number of iterations, and lower in conditions with less missingness.
 
-Evaluation with the performance measures shows that \(\widehat{R}\) is
-conservative: it identifies signs of non-convergence in conditions where
-the \(\bar{Q}\)s are unbiased and/or confidence valid estimates of the
-\(Q\)s.
+Evaluation with the performance measures shows that \(\widehat{R}\) is conservative: it identifies signs of non-convergence in conditions where the \(\bar{Q}\)s are unbiased and/or confidence valid estimates of the \(Q\)s.
 
@@
 
-\textbf{Message: after one iteration (\(T\)=2), the algorithm is as well
-converged as the missingness proportion allows. Sooner even with lower
-missingness proportions (\(T\)=1 might be ok for \(p_{mis}\) = 5 and
-25).} Number of it has little influence on bias in \(\hat{Q}\), only in
-first iteration some bias in missingness conditions of 25 and up. There
-is a clear different between the missingness conditions. The higher the
-missingness prop, the larger the magnitude of the bias. For this \(Q\),
-t=2 would be sufficient. And only for \(p_{mis}\) = 5 the \(\bar{Q}\)s
-are unbiased. All other \(p_{mis}\) conditions have bias irrespective of
-the number of iterations.
+\textbf{Message: after one iteration (\(T\)=2), the algorithm is as well converged as the missingness proportion allows.
+Sooner even with lower missingness proportions (\(T\)=1 might be ok for \(p_{mis}\) = 5 and 25).} Number of it has little influence on bias in \(\hat{Q}\), only in first iteration some bias in missingness conditions of 25 and up.
+There is a clear different between the missingness conditions.
+The higher the missingness prop, the larger the magnitude of the bias.
+For this \(Q\), t=2 would be sufficient.
+And only for \(p_{mis}\) = 5 the \(\bar{Q}\)s are unbiased.
+All other \(p_{mis}\) conditions have bias irrespective of the number of iterations.
 
-\(\widehat{R}\) differs a lot across iteration conditions. But not
-structurally for missingness proportions. Highest \(\widehat{R}\)s for
-the condition \(T\)=2, while the most bias is for (\(T\)=1), but there
-is no \(\widehat{R}\) for that by definition. But for missingness, 25\%
-seems the most signs of non-conv, while it has one of the least bias.
-Threshold of 1.2 reached in condition it = 3, \(p_{mis}\) = 5 and 75,
-and then for it\textgreater6, \(p_{mis}\) = 5 and 50, and
-\(T\)\textgreater10 for all \(p_{mis}\). Threshold 1.1 reached in in
-\(T\)=12, \(p_{mis}\) = all but 75, 1.01 only reached in \(T\) = 93,
-\(p_{mis}\) = 75. While for this \(Q\), \(T\)=2 would be sufficient.
+\(\widehat{R}\) differs a lot across iteration conditions.
+But not structurally for missingness proportions.
+Highest \(\widehat{R}\)s for the condition \(T\)=2, while the most bias is for (\(T\)=1), but there is no \(\widehat{R}\) for that by definition.
+But for missingness, 25\% seems the most signs of non-conv, while it has one of the least bias.
+Threshold of 1.2 reached in condition it = 3, \(p_{mis}\) = 5 and 75, and then for it\textgreater6, \(p_{mis}\) = 5 and 50, and \(T\)\textgreater10 for all \(p_{mis}\).
+Threshold 1.1 reached in in \(T\)=12, \(p_{mis}\) = all but 75, 1.01 only reached in \(T\) = 93, \(p_{mis}\) = 75.
+While for this \(Q\), \(T\)=2 would be sufficient.
 
-\(AC\) mainly shows effects of the set of iteration conditions. \(AC\)
-indicates less signs of non-convergencein conditions with higher number
-of iterations. For iteration conditions where \(T\)\textless10, there
-are also some effects of the missingness proportions. The \(p_{mis}\)
-condtions 25 and 95 \% indicate the most signs of non-convergence. For
-\(p_{mis}\) = 95\% this corresponds to the bias in Qbar. For 25\% miss
-it doesn't. The thresholds defined by the critical value is only met in
-the conditions it=3, \(p_{mis}\) = 5,75, and 95\%.
+\(AC\) mainly shows effects of the set of iteration conditions.
+\(AC\) indicates less signs of non-convergencein conditions with higher number of iterations.
+For iteration conditions where \(T\)\textless10, there are also some effects of the missingness proportions.
+The \(p_{mis}\) condtions 25 and 95 \% indicate the most signs of non-convergence.
+For \(p_{mis}\) = 95\% this corresponds to the bias in Qbar.
+For 25\% miss it doesn't.
+The thresholds defined by the critical value is only met in the conditions it=3, \(p_{mis}\) = 5,75, and 95\%.
 
@@
 
-\textbf{Message: even with possible effect of initial values (\(T\)=1),
-the algorithm is as well converged as the missingness proportion
-allows.} For \(Q=\sigma_Y\) there is no effect visible of the iteration
-conditions, but clearly for the \(p_{mis}\) conditions. Most bias for
-\(p_{mis}\) = 95, followed by \(p_{mis}\) = 50 and 75. For the least
-possible bias we may use any number of iterations \(T\geq1\), but none
-of the \(p_{mis}\) results in unbiased \(\bar{Q}\)s.
+\textbf{Message: even with possible effect of initial values (\(T\)=1), the algorithm is as well converged as the missingness proportion allows.} For \(Q=\sigma_Y\) there is no effect visible of the iteration conditions, but clearly for the \(p_{mis}\) conditions.
+Most bias for \(p_{mis}\) = 95, followed by \(p_{mis}\) = 50 and 75.
+For the least possible bias we may use any number of iterations \(T\geq1\), but none of the \(p_{mis}\) results in unbiased \(\bar{Q}\)s.
 
 There is a clear effect of the number of iterations on \(\widehat{R}\).
-Conditions with lower Ts have more signs of non-conv, as identified by
-\(\widehat{R}\). There is little difference between missingness
-conditions. Only for conditions where \(T<5\) the missingness
-proportions seem to affect whether \(\widehat{R}\) indicates
-non-convergence The bias in \(\bar{Q}\) seems unrelated to the magnitude
-of \(\widehat{R}\). Conditions with more bias (\(p_{mis}\) = 50,75,95)
-are not identified to have more signs of non-convergence. The thresholds
-are therefore useless (because depend mostly on \(T\), not so much
-\(p_{mis}\)).
+Conditions with lower Ts have more signs of non-conv, as identified by \(\widehat{R}\).
+There is little difference between missingness conditions.
+Only for conditions where \(T<5\) the missingness proportions seem to affect whether \(\widehat{R}\) indicates non-convergence The bias in \(\bar{Q}\) seems unrelated to the magnitude of \(\widehat{R}\).
+Conditions with more bias (\(p_{mis}\) = 50,75,95) are not identified to have more signs of non-convergence.
+The thresholds are therefore useless (because depend mostly on \(T\), not so much \(p_{mis}\)).
 
-\(AC\) is affected by both t and \(p_{mis}\) conditions. But conditions
-with worst bias (\(p_{mis}\) = 50, 75, 95) are not identified as worst
-converging according to AC. Moreover, only for \(T\) =3 the \(AC\) is
-significantly different from zero.
+\(AC\) is affected by both t and \(p_{mis}\) conditions.
+But conditions with worst bias (\(p_{mis}\) = 50, 75, 95) are not identified as worst converging according to AC.
+Moreover, only for \(T\) =3 the \(AC\) is significantly different from zero.
 
@@
 
-\textbf{Message: after four iterations (\(T\)=5), the algorithm is as
-well converged as the missingness proportion allows. Sooner even with
-lower missingness proportions (\(T\)=1 might be ok for \(p_{mis}\) = 5
-and 25).} The figure for \(Q=\beta_1\) shows that the \(\bar{Q}\)s are
-affected by both the number of iterations and the missingness
-proportion. For \(T\) =1, the magnitude of the bias (almost) follows the
-order of the missingness proportions. More missingness causes more bias
-in the \(\bar{Q}\). In general, a lower nr of iterations results in more
-bias, but this depends on the missingness proportion. For \(p_{mis}\) =
-95, there is more bias in conditions where \(T\)\textless6. When
-\(p_{mis}\) = 50, it's \(T\)\textless3. For \(p_{mis}\) = 75 and 25,
-it's only for \(T\)\textless2. And conditions where \(p_{mis}\) = 5 are
-unaffected on average. However, none of the conditions result in
-completely unbiased estimates of \(Q\).
+\textbf{Message: after four iterations (\(T\)=5), the algorithm is as well converged as the missingness proportion allows.
+Sooner even with lower missingness proportions (\(T\)=1 might be ok for \(p_{mis}\) = 5 and 25).} The figure for \(Q=\beta_1\) shows that the \(\bar{Q}\)s are affected by both the number of iterations and the missingness proportion.
+For \(T\) =1, the magnitude of the bias (almost) follows the order of the missingness proportions.
+More missingness causes more bias in the \(\bar{Q}\).
+In general, a lower nr of iterations results in more bias, but this depends on the missingness proportion.
+For \(p_{mis}\) = 95, there is more bias in conditions where \(T\)\textless6.
+When \(p_{mis}\) = 50, it's \(T\)\textless3.
+For \(p_{mis}\) = 75 and 25, it's only for \(T\)\textless2.
+And conditions where \(p_{mis}\) = 5 are unaffected on average.
+However, none of the conditions result in completely unbiased estimates of \(Q\).
 
-We should see this in the conv diag: higher \(\widehat{R}\) and \(AC\)
-for \(p_{mis}\) = 75 and 95, and for condtions where \(T<2-6\),
-depending on \(p_{mis}\).
+We should see this in the conv diag: higher \(\widehat{R}\) and \(AC\) for \(p_{mis}\) = 75 and 95, and for condtions where \(T<2-6\), depending on \(p_{mis}\).
 
-\(\widehat{R}\) depends somewhat on the missingness proportion. For
-conditions where \(T\)\textless10, the \(\widehat{R}\) values follow the
-\(p_{mis}\) more or less. But what we actually want to know is whether
-they are indicators of bias. Conditions with the worst bias (high
-\(p_{mis}\), low T) have the largest \(\widehat{R}\) values. The
-thresholds are not appropriate to diagnose sufficient convergence. All
-thresholds over-estimate the signs of non-convergence compared to the
-bias in Qbar. However, a higher threshold than 1.2 may result in
-terminating the algorithm after 3 iterations, which is insufficient for
-the \(p_{mis}\) = 95 condition. Threshold of 1.25 or 1.3 could work for
-the other \(p_{mis}\) conditions.
+\(\widehat{R}\) depends somewhat on the missingness proportion.
+For conditions where \(T\)\textless10, the \(\widehat{R}\) values follow the \(p_{mis}\) more or less.
+But what we actually want to know is whether they are indicators of bias.
+Conditions with the worst bias (high \(p_{mis}\), low T) have the largest \(\widehat{R}\) values.
+The thresholds are not appropriate to diagnose sufficient convergence.
+All thresholds over-estimate the signs of non-convergence compared to the bias in Qbar.
+However, a higher threshold than 1.2 may result in terminating the algorithm after 3 iterations, which is insufficient for the \(p_{mis}\) = 95 condition.
+Threshold of 1.25 or 1.3 could work for the other \(p_{mis}\) conditions.
 
-\(AC\) is very affected by the p\_mis. The two conditions with the most
-bias (\(p_{mis}\) = 75 and 95) are the only ones identified as
-non-converging, as they should. However, the \(p_{mis}\) = 75 condition
-is only identified in the \(T\) =3 condtion, while the bias is the least
-pronounced there. And the \(AC\) only identifies non-convergence in the
-\(p_{mis}\) = 95 conditions where \(T\)\textgreater10. In the default
-situation, \(T\) =5, this would be overlooked. Also, the \(AC\) shows
-the opposite of what you would expect from a theta that is based on the
-observed data: shouldn't there be higher autocorrelations in conditions
-with comparatively more observed data in the completed datasets? I don't
-understand this.
+\(AC\) is very affected by the p\_mis.
+The two conditions with the most bias (\(p_{mis}\) = 75 and 95) are the only ones identified as non-converging, as they should.
+However, the \(p_{mis}\) = 75 condition is only identified in the \(T\) =3 condtion, while the bias is the least pronounced there.
+And the \(AC\) only identifies non-convergence in the \(p_{mis}\) = 95 conditions where \(T\)\textgreater10.
+In the default situation, \(T\) =5, this would be overlooked.
+Also, the \(AC\) shows the opposite of what you would expect from a theta that is based on the observed data: shouldn't there be higher autocorrelations in conditions with comparatively more observed data in the completed datasets?
+I don't understand this.
 
-\textbf{Add how interesting it is because we literally track \(Q\) and
-are still not ok.}
+\textbf{Add how interesting it is because we literally track \(Q\) and are still not ok.}
 
@@
 
-\textbf{Message: after two iterations (\(T\)=3), the algorithm is as
-well converged as the missingness proportion allows. Sooner even with
-lower missingness proportions (\(T\)=1 might be ok for \(p_{mis}\) = 5
-and 25).} Bias is the worst for conditions where \(p_{mis}\) = 95, \(T\)
-\textless3. Followed by \(p_{mis}\) = 75, \(T\) =1; and \(p_{mis}\) =
-50, \(T\) = 1. Conditions where \(p_{mis}\) = 25 or 5 seem unaffected by
-the number of iterations. Minimal bias is obtained in all iteration
-conditions. We should see this in \(\widehat{R}\) and AC: worst
-convergence in \(p_{mis}\) = 95 and 50; some non-convergencein
-conditions where \(p_{mis}\) = 75 and \(T\) =1, and no non-convergencein
-conditions where \(p_{mis}\) = 25 or 5. The sign of the bias is logical:
-when there is less info in the data, the relations between the
-predictors and the outcome is under-estimated.
+\textbf{Message: after two iterations (\(T\)=3), the algorithm is as well converged as the missingness proportion allows.
+Sooner even with lower missingness proportions (\(T\)=1 might be ok for \(p_{mis}\) = 5 and 25).} Bias is the worst for conditions where \(p_{mis}\) = 95, \(T\) \textless3.
+Followed by \(p_{mis}\) = 75, \(T\) =1; and \(p_{mis}\) = 50, \(T\) = 1.
+Conditions where \(p_{mis}\) = 25 or 5 seem unaffected by the number of iterations.
+Minimal bias is obtained in all iteration conditions.
+We should see this in \(\widehat{R}\) and AC: worst convergence in \(p_{mis}\) = 95 and 50; some non-convergencein conditions where \(p_{mis}\) = 75 and \(T\) =1, and no non-convergencein conditions where \(p_{mis}\) = 25 or 5.
+The sign of the bias is logical: when there is less info in the data, the relations between the predictors and the outcome is under-estimated.
 
-\(\widehat{R}\) behaves oppositely to what we wanted. The lowest
-\(\widehat{R}\) values for conditions where \(T\)\textless10 are
-observed for the two missingness conditions with the worst bias.
-However, the highest \(\widehat{R}\) observed is indeed for the
-missingness condition with the worst bias, \(p_{mis}\) =95. But only at
-\(T\) =5 (while bias is no worse than anywhere else in this \(p_{mis}\)
-condition, conditional on iterations condition \(T\)\textgreater2). So
-neither of the thresholds works perfectly.
+\(\widehat{R}\) behaves oppositely to what we wanted.
+The lowest \(\widehat{R}\) values for conditions where \(T\)\textless10 are observed for the two missingness conditions with the worst bias.
+However, the highest \(\widehat{R}\) observed is indeed for the missingness condition with the worst bias, \(p_{mis}\) =95.
+But only at \(T\) =5 (while bias is no worse than anywhere else in this \(p_{mis}\) condition, conditional on iterations condition \(T\)\textgreater2).
+So neither of the thresholds works perfectly.
 
-\(AC\) shows a similar trend to the biases: more signs of
-non-convergencefor higher \(p_{mis}\) and lower T. The threshold,
-however, performs less well. It identifies signs of non-convergence for
-all conditions where \(T=3\), irrespective of p\_mis. Then nothing, and
-then for \(p_{mis}\) =95, \(T\)\textgreater7, and \(p_{mis}\) = 50,
-\(T\) \textgreater14, and \(p_{mis}\) = 75, \(T\)\textgreater20. This
-does not corresond to the bias in \(\bar{Q}\). Maybe use
-AC\textgreater.5 at \(T\) =5 as threshold?
+\(AC\) shows a similar trend to the biases: more signs of non-convergencefor higher \(p_{mis}\) and lower T.
+The threshold, however, performs less well.
+It identifies signs of non-convergence for all conditions where \(T=3\), irrespective of p\_mis.
+Then nothing, and then for \(p_{mis}\) =95, \(T\)\textgreater7, and \(p_{mis}\) = 50, \(T\) \textgreater14, and \(p_{mis}\) = 75, \(T\)\textgreater20.
+This does not corresond to the bias in \(\bar{Q}\).
+Maybe use AC\textgreater.5 at \(T\) =5 as threshold?
 
@@
 
-In short, univariate estimates seem robust against terminating the
-algorithm early: There is no clear effect of the number of iterations on
-the bias in the estimated means and standard deviations. The bias in
-these \(Q\)s only depends on the \(p_{mis}\), with higher \(p_{mis}\)
-resulting in more bias. Yet, the convergence diagnostics indicate that
-the algorithm did not reach a completely converged state (yet).
+In short, univariate estimates seem robust against terminating the algorithm early: There is no clear effect of the number of iterations on the bias in the estimated means and standard deviations.
+The bias in these \(Q\)s only depends on the \(p_{mis}\), with higher \(p_{mis}\) resulting in more bias.
+Yet, the convergence diagnostics indicate that the algorithm did not reach a completely converged state (yet).
 
-\(\widehat{R}\) and autocorrelation indicate that algorithmic
-convergence may only be reached after twenty or even forty iterations,
-while unbiased, confidence valid estimates may be obtained with as
-little as one iteration. These results are in agreement with the
-simulation hypothesis: \(\widehat{R}\) over-estimates the severity of
-non-convergence when applied to MI procedures.
+\(\widehat{R}\) and autocorrelation indicate that algorithmic convergence may only be reached after twenty or even forty iterations, while unbiased, confidence valid estimates may be obtained with as little as one iteration.
+These results are in agreement with the simulation hypothesis: \(\widehat{R}\) over-estimates the severity of non-convergence when applied to MI procedures.
 
-With that, we show that MICE can lead to correct outcomes when they have
-not converged according to two common conv diags. This may be due to the
-methods (and their thresholds) or due to the \(Q\)s (descriptives and
-multivariate linear regression, not higher dimensional/more complex
-\(Q\)s). More `complicated' \(Q\)s (e.g., higher-order effects or
-variance components) might show bias, under- or over-coverage at higher
-\(T\), as indiated by \(\widehat{R}\) and AC. Add what \%miss has to do
-with it.
+With that, we show that MICE can lead to correct outcomes when they have not converged according to two common conv diags.
+This may be due to the methods (and their thresholds) or due to the \(Q\)s (descriptives and multivariate linear regression, not higher dimensional/more complex \(Q\)s).
+More `complicated' \(Q\)s (e.g., higher-order effects or variance components) might show bias, under- or over-coverage at higher \(T\), as indiated by \(\widehat{R}\) and AC.
+Add what \%miss has to do with it.
 
 \hypertarget{recommendations-for-empirical-researchers}{%
-\subsection{Recommendations for empirical
-researchers}\label{recommendations-for-empirical-researchers}}
+\subsection{Recommendations for empirical researchers}\label{recommendations-for-empirical-researchers}}
 
@@
 \item
-  Check trace plots for pathological non convergence and adjust
-  imputation model if necessary.
+  Check trace plots for pathological non convergence and adjust imputation model if necessary.
 \item
-  use \(\widehat{R}\) wait 1.1 ash threshold and autocorrelation with
-  \textbf{?} as threshold. Keep iterating until these thresholds are
-  reached.
+  use \(\widehat{R}\) wait 1.1 ash threshold and autocorrelation with \textbf{?} as threshold. Keep iterating until these thresholds are reached.
 \item
-  Do not use the R function ACF. Instead, compute autocorrelations
-  manually \textbf{(see e.g., {[}GitHub link{]})}.
+  Do not use the R function ACF. Instead, compute autocorrelations manually \textbf{(see e.g., {[}GitHub link{]})}.
 \item
-  Track your own scalar summary of interest. This is somewhat advanced
-  but explained in Van Buren 2018. Compute \(\widehat{R}\) and
-  autocorrelation values for this scalar summary.
+  Track your own scalar summary of interest. This is somewhat advanced but explained in Van Buren 2018. Compute \(\widehat{R}\) and autocorrelation values for this scalar summary.
 \item
-  \textbf{Something} about the novel \(\theta\) that is `substantive
-  model-independent'.
+  \textbf{Something} about the novel \(\theta\) that is `substantive model-independent'.
 \end{enumerate}
@@
 \hypertarget{recommendations-for-future-research}{%
-\subsection{Recommendations for future
-research}\label{recommendations-for-future-research}}
+\subsection{Recommendations for future research}\label{recommendations-for-future-research}}
 
-Further research is needed to investigate their performance under clear
-violation of convergence, e.g.~dependency between predictors (predictors
-with very high correlations). Also:
+Further research is needed to investigate their performance under clear violation of convergence, e.g.~dependency between predictors (predictors with very high correlations). Also:
 
@@
   More difficult missingness problems, i.e., M(N)AR instead of MCAR.
-  Proper performance under a `missing completely at random' missingness
-  mechanism is necessary to demonstrate the appropriateness of
-  \(\widehat{R}\) and \(AC\) as non-convergence diagnostics. However,
-  results may not be extrapolated to other missingness mechanisms.
+Proper performance under a `missing completely at random' missingness mechanism is necessary to demonstrate the appropriateness of \(\widehat{R}\) and \(AC\) as non-convergence diagnostics.
+However, results may not be extrapolated to other missingness mechanisms.
 \item
-  Imputation technique with questionable convergence properties, e.g.,
-  PMM.
+  Imputation technique with questionable convergence properties, e.g., PMM.
 \item
-  Apply recommendations to empirical data and write a vignette for
-  applied researchers.
+  Apply recommendations to empirical data and write a vignette for applied researchers.
 \item
@@
 \item
-  Develop an additional convergence diagnostic unique for iterative
-  imputation: \(AC\) across imputations, instead of iterations.
+  Develop an additional convergence diagnostic unique for iterative imputation: \(AC\) across imputations, instead of iterations.
 \end{itemize}
 
-Uitsmijter: To conclude, in the case of drawing inference from
-incomplete data, convergence of the iterative imputation algorithm is
-often a convenience but not a necessity.
+Uitsmijter: To conclude, in the case of drawing inference from incomplete data, convergence of the iterative imputation algorithm is often a convenience but not a necessity.
 
@@
 \item
-  Scientific contribution (clear research question, convincing
-  argumentation that the study adds to the scientific literature)
+  Scientific contribution
+  (clear research question, convincing argumentation that the study adds to the scientific literature)
 \item
-  Theory (use of theory, clear hypotheses, relevant literature)
+  Theory
+  (use of theory, clear hypotheses, relevant literature)
 \item
-  Method (correct data collection/handling, measures, choice and
-  description of method of analysis)
+  Method
+  (correct data collection/handling, measures, choice and description of method of analysis)
 \item
-  Results (correct interpretation of findings, clear presentation)
+  Results
+  (correct interpretation of findings, clear presentation)
 \item
-  Discussion and conclusions (summary, broader implications,
-  limitations, future research)
+  Discussion and conclusions
+  (summary, broader implications, limitations, future research)
 \item
-  Written presentation (writing, structure, tables and figures,
-  references)
+  Written presentation
+  (writing, structure, tables and figures, references)
 \end{itemize}

### 2020-05-04 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
Draft for final feedback

--- 2020-05-03 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
@@
 There is no consensus on how to evaluate the convergence properties of the method.
-This paper provides insight into identifying non-convergence of iterative impuation algorithms."
+This paper provides insight into identifying non-convergence of iterative imputation algorithms."
 keywords: missing data, iterative imputation, non-convergence, mice
@@
 
-Currently, the recommended practice for evaulating the convergence of iterative imputation algorithms is through visual inspection.
+Currently, the recommended practice for evaluating the convergence of iterative imputation algorithms is through visual inspection.
 That is, the state space of the algorithm is monitored by plotting a scalar summary $\theta$ against the iteration number.
@@
 The aim of the simulation is to determine whether unbiased, confidence valid inferences may be obtained under different levels of non-convergence (**and at which point convergence may safely be assumed**).
-Additionally, we will assess how well different non-convergence diagnostics perform in identifying signs of non-convergence.
-The results of this study are guidelines for identifying non-convergence in iterative imptation algorithms, which will aid applied researchers in drawing valid inference from incomplete datasets.
+Additionally, we will assess how well different non-convergence diagnostics perform in identifying signs of non-convergence (**and on which scalar summary of the state space of the algorithm these diagnostics should be applied**).
+The results of this study are guidelines for identifying non-convergence in iterative imputation algorithms, which will aid applied researchers in drawing valid inference from incomplete datasets.
 **[add more societal relevance: what is the consequence of non-convergence?
-e.g., biased estimates, invalid inference, etc.]** For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular `mice` package [@mice] in `R` [@R].
+see example of non-convergence in Figure 2 (e.g., biased estimates, invalid inference, etc.)]** For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular `mice` package [@mice] in `R` [@R].
 
@@
 The collection of samples between the initial value (at $t=1$) and the final imputed value (at $t=T$) will be referred to as an 'imputation chain'.
-**$\theta$s are scalar summaries of the algorithm at a certain iteration (e.g., chain means; the average of the imputed values in each imputation chain).
-**
+**[If not defined elsewhere, add: $\theta$s are scalar summaries of the algorithm at a certain iteration (e.g., chain means; the average of the imputed values in each imputation chain).] **
 
@@
 Each line portrays one imputation.
-The plot on the lefthand side of the figure shows typical convergence of an iterative imputation algorithm.
-The righthand side displays pathological non-convergence, induced by purposefully mis-specifying the imputation model.
+The plot on the left-hand side of the figure shows typical convergence of an iterative imputation algorithm.
+The right-hand side displays pathological non-convergence, induced by purposefully mis-specifying the imputation model.
 **[Leave this in if it's after mixing and stationarity are defined, otherwise remove?] In the typical convergence situation, the imputation chains intermingle nicely and there is little to no trending.
@@
 This is typically done through traceplots.
-In a traceplot, the iteraton number is plotted against a certain $\theta$.
+In a traceplot, the iteration number is plotted against a certain $\theta$.
 $\theta$s are scalar summaries of the state space of the algorithm at a specific iteration.
@@
 \geq \lambda_j$ be the eigenvalues of $\Sigma$ in each of the $M$ completed datasets ${y_{obs}, y_{imp}}$.
-$\lambda_1$ is measure that summarises the covariances in the completed datasets.
+$\lambda_1$ is measure that summarizes the covariances in the completed datasets.
 **The first eigenvalue has the appealing property that is not dependent on the substantive model of interest.
@@
 \geq \lambda_j$ be the eigenvalues of $\Sigma$ in each of the $M$ completed datasets ${y_{obs}, y_{imp}}$.
-$\lambda_1$ is measure that summarises the covariances in the completed datasets.**  -->
+$\lambda_1$ is measure that summarizes the covariances in the completed datasets.**  -->
 
@@
 
-Non-convergence diagnostics for MCMC algorithms typically identify problems in either one of the two requirements for converegnce: mixing or stationarity.
+Non-convergence diagnostics for MCMC algorithms typically identify problems in either one of the two requirements for convergence: mixing or stationarity.
 Non-stationarity within chains may be diagnosed with e.g., autocorrelation [$AC$; @scha97; @gelm13], numeric standard error ['MC error'; @gewe92], or Raftery and Lewis's [-@raft91] procedure to determine the effect of trending on the precision of estimates.
@@
 
+The potential scale reduction factor $\widehat{R}$ was coined by @gelm92.
 An updated version of $\widehat{R}$ has been proposed by Vehtari et al.
@@
 5].
-**This version may be suitable for iterative imputation.** Let $M$ be the total number of chains, $T$ the number of iterations per chain, and $\theta$ the scalar summary of interest (e.g., chain means).
+**This version is better suited to detect non-convergence in the tails of distributions, by using three transformations on the scalar summary that it is applied to.
+Namely, folded, rank-normalized, split-chain $\theta$s.
+The Vehtari et al.
+version of $\widehat{R}$ may be suitable for iterative imputation.
+We therefore use their definition of the diagnostic.** Let $M$ be the total number of chains, $T$ the number of iterations per chain, and $\theta$ the scalar summary of interest (e.g., chain means).
 For each chain ($m = 1, 2, \dots, M$), we estimate the variance of $\theta$, and average these to obtain within-chain variance $W$.
@@
 
-We then estimate between-chain variance $B$ as the variance of the collection of average $\theta$ per chain.
+\noindent We then estimate between-chain variance $B$ as the variance of the collection of average $\theta$ per chain.
 
@@
 
-From the between- and within-chain variances we compute a weighted average, $\widehat{\operatorname{var}}^{+}$, which over-estimates the total variance of $\theta$ **remove or explain why, or leave in**.
-$\widehat{R}$ is then obtained as a ratio between the over-estimated total variance and the within-chain variance:
+\noindent From the between- and within-chain variances we compute a weighted average, $\widehat{\operatorname{var}}^{+}$, which approximates the total variance of $\theta$.
+$\widehat{R}$ is then obtained as a ratio between the total variance and the within-chain variance:
 
@@
 Over-dispersion implies that the initial values of the chains are 'far away' from the target distribution and each other.
-When all chains sample independent of their initial values, the mixing component of convergence is satisfied, and $\widehat{R}$-values will be close to one.
+When the sampled values in each chain are independent of the chain's initial value, the mixing component of convergence is satisfied.
+The variance between chains is then equivalent to the variance within chains, and $\widehat{R}$-values will be close to one.
 High $\widehat{R}$-values thus indicate non-convergence.
@@
 But $\widehat{R}=1$ and $AC=0$ should at least be able to distinguish between the two algorithms plotted in Figure \ref{fig:non-conv}.
-From visual inspection, we know that the typically converged algorithm initially portrayed some signs of non-mixing (around $t=2$), but intermigled nicely overall.
+From visual inspection, we know that the typically converged algorithm initially portrayed some signs of non-mixing (around $t=2$), but intermingled nicely overall.
 Additionally, there was very little trending.
@@
 Panel C shows the traditional computation of $\widehat{R}$, and panel D shows $\widehat{R}$ as computed by implementing @veht19 's recommendations.
+**[Add labels for the two types of Rhat to the figure!]**
 
@@
 **Describe panel C here and why it is not optimal]** The $\widehat{R}$-values in panel D do meet the requirements specified above.
-$\widehat{R}$ as computed conform @veht19 indeed indicate less signs of non-convergence as the number of iterations goes up.
+$\widehat{R}$ as computed conform @veht19 does indeed indicate less signs of non-convergence as the number of iterations goes up.
+**[explain the dip in Rhat values at t=2.
+Namely, because we can only use 2 of the 3 tricks by @veht19 if the number of iterations is very low (t<4).
+That's why the $\widehat{R}$s are more similar to the traditional GR.]**
 <!-- However, the superior performance of the typical convergence over the pathological non-convergence is less prominent, even flipped for $t<4$.  -->
@@
 Unless, however, the fraction of missing information is so high that the random component in the imputation algorithm outweighs the information in the observed data.
-Then, convergence to a stable but highly variable state may be reached instanteneously.
+Then, convergence to a stable but highly variable state may be reached instantaneously.
 We assume that the incremental missingness proportions in our study will result in a corresponding increase in signs of non-convergence.
@@
 In an empirical study, where $\widehat{R}$ was used to inform the required imputation chain length, it took as many as 50 iterations to overcome the conventional non-convergence threshold $\widehat{R}>1.2$.
-Yet, scientific estimates were insensitive to continued iteration from $t=5$ onwards [@lace07].
+Yet, scientific estimates were insensitive to continued iteration from $t=5$ onward [@lace07].
 We therefore suspect that $\widehat{R}$ may over-estimate signs of non-convergence in iterative imputation algorithms, compared to the bias the estimates.
@@
 
-5. **Something about thetas**
+5.
+**[Something about thetas, that multivariate thetas are better at detecting non-convergence than univariate thetas.
+The non-convergence diagnostics applied on the chain means and chain variances will fail to identify signs of non-convergence when the multivariate Qs are affected.]**
 
@@
 From the complete set, we obtain the true values of several quantities of scientific interest, $Q$s.
-**[MOVE THIS NEXT PART TO TOP OF SIMULATION SECTION?]** $Q$s are the targets that will be estimated in each simulation repetition, after first *amputing* the complete data with varying missingness proportions, then imputing the missingness with a varying number of iterations, then estimating the $Q$s on the $M$ completed datasets, and finally pooling the $M$ estimates $\hat{Q}$ to obtain a single $\bar{Q}$ for each $Q$.
-**We try to predict the effect of non-converegence (i.e., the difference between $Q$ and $\bar{Q}$) with the convergence diagnostics $\widehat{R}$ and $AC$, by appliying these methods to several scalar summaries of the state space of the algorithm, $\theta$s.**
+**[MOVE this next part to the start of the Simulation Study section?] $Q$s are the targets that will be estimated in each simulation repetition, after first *amputing* the complete data with varying missingness proportions, then imputing the missingness with a varying number of iterations, then estimating the $Q$s on the $M$ completed datasets, and finally pooling the $M$ estimates $\hat{Q}$ to obtain a single $\bar{Q}$ for each $Q$.
+We try to predict the effect of non-convergence (i.e., the difference between $Q$ and $\bar{Q}$) with the convergence diagnostics $\widehat{R}$ and $AC$, by applying these methods to several scalar summaries of the state space of the algorithm, $\theta$s.**
 
@@
 - Perform multiple linear regression with `stats::lm()` [@R] to get the $\hat{Q}$s for $Q=\beta$ and $Q=r^2$.
-We pool the $\hat{Q}$s conform @vink14 to get $\bar{Q}$s for the $\beta$s and $r^2$ **[check  pooling of  $r^2$, instead of `mice::pool.r.squared()`]**.
+We pool the $\hat{Q}$s conform @vink14 to get $\bar{Q}$s for the $\beta$s and $r^2$ **[check  finite pooling of  $r^2$, instead of `mice::pool.r.squared()`]**.
 Additionally, we use the $\hat{Q}$s of $Q=\beta$ as $\theta$ to apply the convergence diagnostics to.
@@
 where $t_{(M-1)}$ is the quantile of a $t$-distribution with $M-1$ degrees of freedom, and $SE_{\bar{Q}}$ is the square root of the pooled variance estimate.
-If we obtain nominal coverage (CR = 95%), we can conclude that the $\bar{Q}$s are conficence valid estimates of the $Q$s.
+If we obtain nominal coverage (CR = 95%), we can conclude that the $\bar{Q}$s are confidence valid estimates of the $Q$s.
 Finally, we inspect CI width (CIW): the difference between the lower and upper bound of the 95\% confidence interval around $\bar{Q}$.
@@
 With these performance measures, we also evaluate how well the non-convergence diagnostics $\widehat{R}$ and $AC$ can identify simulation conditions in which the $\bar{Q}$s are *not* unbiased, confidence valid estimates of the $Q$s.
-**Rephrase and add corresponding $Q$s:** We measure the performance of $\widehat{R}$ and $AC$ by the four scalar summaries $\theta$: chain means (i.e., the mean in each $y_{imp, m, j}$) versus $Q=\mu$, chain variances (i.e., the variance in each $y_{imp, m, j}$) versus $Q=\sigma$, the estimated regression coefficients $\beta$ in each {$y_{obs}, y_{imp, m}$} (i.e., a user-defined statistic of scientific interest $\hat{Q}$) against $Q=\beta$, and the first eigenvalue of the variance-covariance matrix in each {$y_{obs}, y_{imp, m}$}, $\lambda_1$ versus $Q=r^2$.
+**[Rephrase this:] We measure the performance of $\widehat{R}$ and $AC$ by the four scalar summaries $\theta$: chain means (i.e., the mean in each $y_{imp, m, j}$) versus $Q=\mu$, chain variances (i.e., the variance in each $y_{imp, m, j}$) versus $Q=\sigma$, the estimated regression coefficients $\beta$ in each {$y_{obs}, y_{imp, m}$} (i.e., a user-defined statistic of scientific interest $\hat{Q}$) against $Q=\beta$, and the first eigenvalue of the variance-covariance matrix in each {$y_{obs}, y_{imp, m}$}, $\lambda_1$ versus $Q=r^2$.**
+
+**[Remove this?] Primarily, we want to know what the effect is of inducing non-convergence on the estimated $Q$s.
+More specifically, we are interested in the bias of $\bar{Q}$ for all $Q$s and the confidence validity of $\bar{Q}$ for $Q=\beta$.
+Therefore, we evaluate these performance measures against the simulation conditions (the number of iterations $T$ and the missingness proportion $p_{mis}$).
+We look at averages across simulation repetitions.
+Secondly, we want to know if conditions in which the $Q$s are affected by non-convergence are also identified by the non-convergence diagnostics $\widehat{R}$ and $AC$.
+We do this, again, by looking at the performance measures for the $Q$s, but now we evaluate these against what $\widehat{R}$ and $AC$ indicate about signs of non-convergence in the $\theta$s.**
 
@@
 
-**[Add to this section: 1) emphasize that the iterations in the plots are averages across repetitions, not within one run of MICE; 2) sample effects due to single complete dataset??
-3) Add more info about figure legends and axes.]**
-Primarily, we want to know what the effect is of inducing non-convergence on the estimated $Q$s.
-Specifically, we are interested in the bias of $\bar{Q}$ for all $Q$s and the confidence validity of $\bar{Q}$ for $Q=\beta$.
-Therefore, we plot these performance measures against the simulation conditions (the number of iterations $T$ and the missingness proportion $p_{mis}$).
-We look at averages across simulation repetitions.
+**[Add to this section: 1) emphasize that the iterations in the plots are averages across repetitions, not within one run of MICE; 2) Add more info about figure legends and axes]**
 
-Secondly, we want to know if conditions in which the $Q$s are affected by non-convergence are also identified by the non-convergance diagnostics $\widehat{R}$ and $AC$.
-We do this, again, by looking at the performance measures for the $Q$s, but now we evaluate these against what $\widehat{R}$ and $AC$ indicate about signs of non-convergence in the $\theta$s.
+Our results show the impact of inducing non-convergence in an iterative imputation algorithm on several scientific estimates, and whether the non-convergence diagnostics $\widehat{R}$ and $AC$ identify the signs of non-convergence in the algorithm.
+For reasons of brevity, we only discuss the non-convergence diagnostics for estimates with the worst performance in terms of bias.
+Bias in the descriptive statistics ($Q=\mu$ and $Q=\sigma$) is most pronounced for the outcome variable $Y$.
+Of the regression coefficients ($Q=\beta$) we observe the largest bias for $\beta_1$ (the effect of $X_1$ on $Y$).
+There is only one estimate for $Q=r^2$ to consider.
+In the figures, we present simulation conditions where the number of iterations is $1 \leq T\leq50$, since the results are more or less stable for conditions where $T \leq 30$.
+Full results are available from [github.com/gerkovink/shinyMice](https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy).
 
-**[In short:]**
-- Is there more bias in conditions where non-convergence was induced? (NOPE for univariates, YES for multivar.).
+## Quantities of scientific interest
 
-- Is the point of *sufficient* convergence (at which the $\bar{Q}$s are unbiased and confidence valid estimates of the $Q$s) the same point as identified by the non-convergence diagnostics?
-(NOPE, they are too conservative).
+```{r Qs, fig.height = 6.8, fig.cap="Impact of non-convergence on statistical inferences."}
+mean_bias + sd_bias + est_bias + est_ciw + est_cov + Rsq_bias + plot_layout(ncol = 2, guides = "collect")
+```
 
-- Is one of the thetas better? (NAH, doesn't really matter, probably multivariate is better).
+Figure \ref{fig:Qs} displays the influence of inducing non-convergence on the estimated quantities of scientific interest, $Q$s.
+**[Add panel labels, describe panels]** As expected, conditions with a higher proportion of missingness and/or a lower number of iterations are impacted more severely by non-convergence.
+I.e., on average these conditions portray more extreme bias in the estimated $Q$s, and non-nominal coverages.
+Roughly speaking, this means that the algorithm is indeed not converged at $t=1$, and converges gradually as $t$ increases to $T$.
+The point at which an additional iteration does not lead to an improvement in the estimates depends on quantity of scientific interest $Q$ and the missingness proportion $p_{mis}$.
+The highest number of iterations that is necessary to reach this point across all simulation conditions is $T=7$.
+This implies that, under the current specifications, no more than seven iterations are necessary to obtain optimal**/the best possible** performance in terms of bias and confidence validity.
 
-- What should the thresholds be? (LOWER than recommended).
+### $Q=\mu$.
+The estimated univariate means seem unaffected by the number of iterations in the algorithm.
+The magnitude of the bias in $\bar{Q}$ depends exclusively on the missingness proportion $p_{mis}$, with higher missingness proportions causing more extreme bias.
+Note however, that the simulation condition with the largest overall bias resulted in a $\bar{Q}$ under-estimated the true value of $Q$ by only 0.02 units (where $Q=\mu_Y=25.81; \sigma_Y = 11.32$).
+We conclude that---conditional on the missingness proportion---all simulation conditions ($T \geq 1$) have equal performance.
+According to this scientific estimand $Q$, non-convergence does not impact the validity of the inferences.
+**[Add: up-to 50% missingness everything is relatively unbiased, while 75% and 95% missingness show very different results.
+Why is that?
+Add sd of Y in the figure to show that the bias is relative: a bias of 0.02 for a variable with a mean of 25.81, and sd of 11.32 is nothing.]**
 
-- And the default nr of iterarions? (MAYBE increase to 10?).
+### $Q=\sigma$.
+The estimated standard deviations show impact of non-convergence similar to the means.
+The magnitude of the bias $\bar{Q}$ follows the incremental order of the missingness proportions.
+The number of iterations in the algorithm has little effect on the bias.
+Only the estimates in simulation conditions with just one iteration ($T=1$) differ from other conditions that have the same missingness proportion but a higher number of iterations.
+Interestingly, the bias in conditions where $T=1$ is *less* severe than after continued iteration.
+For the least possible bias, we may use any number of iterations $T\geq1$, but only $p_{mis} = 0.05$ results in completely unbiased $\bar{Q}$s.
 
-For reasons of brevity, we only discuss the convergence diagnostics for the $Q$s with the worst performance in terms of bias.
-Bias in the  descriptve statistics ($Q=\mu$ and $Q=\sigma$) was most pronounced in the outcome variable $Y$.
-Of the regression coefficients, $Q=\beta$, the largest bias was observed in the effect of $X_1$ on $Y$, $\beta_1$.
-Results for $Q=r^2$ are reported as-is.
-In the figures, however, we only present simulation conditions where $T\leq50$.
-Full results are available from **[link github repo]**.
+### $Q=\beta$.
+For the estimated regression coefficients, we consider bias, confidence interval width (CIW), and coverage rate (CR) as performance measures.
+The magnitude of the bias, again, depends on the missingness proportion.
+But for this $Q$ we also observe a clear effect of the number of iterations.
+Conditions with a lower number of iterations yield more bias in $\bar{Q}$.
+This is especially pronounced in conditions with high missingness proportions: with 95% missingness we need at least seven iterations before an additional iteration does not lead to improved estimates.
+Conditions with a lower missingness proportion (e.g., $p_{mis} = {0.5}$), produce *optimal* estimates with as many as two iterations.
+Approximately unbiased estimates are obtained for all conditions where $p_{mis}\leq0.75$ and $T\leq6$.
+**[Add CIW and CR]**
 
-## In general
+### $Q=r^2$.
+The bias in the estimated explained variance perfectly follows both sets of simulation conditions: The worst performance is observed for the highest missingness proportion and the lowest number of iterations, while the best performance occurs with less missingness and more iterations.
+Only conditions with the lowest missingness proportion ($p_{mis}=0.05$) results in unbiased estimates for every number of iterations.
+In other conditions there is at least some bias irrespective of $T$.
+The inferences do not improve in any condition where $T \geq 6$.
+This suggests that the algorithm is as converged as necessary to obtain valid inferences.
 
-As expected, conditions with a higher proportion of missingness and/or a lower number of iterations show more signs of non-convergence.
-I.e., on average these conditions portray more extreme bias in the estimated $Q$s, and non-nominal coverages.
-Roughly speaking this means that MICE is indeed not converged at $t=1$, and converges gradually as $t$ increases to $T$.
-The point at which an additional iteration does not lead to an improvement of the estimates depends on the difficulty of the missingness problem---between XYZ and XYZ.
-This means that the algorithm has converged sufficiently (**under the current specifications**).
+<!-- The sign of the bias is logical: when there is less info in the data, the relations between the predictors and the outcome is under-estimated. -->
 
-As required for a method to diagnose non-convergence, the values of $\widehat{R}$ follow a trend similar to the performance measures.
-**Add split between convergence diagnostics versus performance measures, and convergence diagnostics versus simulation conditions.** $\widehat{R}$ values are generally lower in conditions with a higher number of iterations, and somewhat higher in conditions with a higher percentage of missing data.
-Autocorrelation values were indeed decreasing with a higher number of iterations, and lower in conditions with less missingness.
+## Non-convergence diagnostics
 
-Evaluation with the performance measures shows that $\widehat{R}$ is conservative: it identifies signs of non-convergence in conditions where the $\bar{Q}$s are unbiased and/or confidence valid estimates of the $Q$s.
+In Figure \ref{fig:rhats-acs} we see the results of the non-convergence diagnostics $\widehat{R}$ and $AC$.
+**[Add panel labels and explain how to read the plots]** As expected, the diagnostics indicate more signs of non-convergence in simulation conditions with a lower number of iterations.
+The missingness proportions, however, only seem to impact multivariate convergence.
+When $\widehat{R}$ and $AC$ are applied to  univariate $\theta$s, they identify the same amount of non-convergence, irrespective of the missingness proportion.
+When the diagnostics are applied to multivariate $\theta$s, they generally identify more signs of non-convergence for higher missingness proportions.
 
-## Means
+None of the simulation conditions that we consider resulted in complete convergence of the algorithm (defined as $\widehat{R}=1$ and $AC=0$).
+The thresholds for diagnosing non-convergence that are used in practice seem too conservative compared to the performance measures: $\widehat{R}$ and $AC$ identify signs of non-convergence in conditions where the $\bar{Q}$s are unbiased and confidence valid estimates of the $Q$s.
+The most recent recommended threshold for diagnosing non-convergence with $\widehat{R}$ is never overcome in this study (all $\widehat{R}$>1.01).
+Apparently, this threshold is too stringent for iterative imputation algorithms, and will not be discussed further.
 
-**Message: after one iteration ($T$=2), the algorithm is as well converged as the missingness proportion allows.
-Sooner even with lower missingness proportions ($T$=1 might be ok for $p_{mis}$ = 5 and 25).** Number of it has little influence on bias in $\hat{Q}$, only in first iteration some bias in missingness conditions of 25 and up.
+```{r rhats-acs, fig.height = 6.8, fig.cap="Non-convergence diagnostics. [Split up into 2 or 4 figures??]"}
+mean_Rh + mean_AC + var_Rh + var_AC + est_Rh + est_AC + PCA_Rh + PCA_AC + plot_layout(guides = "collect", ncol = 2)
+```
+
+<!-- **Message: after one iteration ($T$=2), the algorithm is as well converged as the missingness proportion allows.
+Sooner even with lower missingness proportions---$T$=1 might be sufficient for $p_{mis}$ = 5, 25, and 50.** Number of iterations has little influence on bias in $\hat{Q}$, only in first iteration there is some bias in missingness conditions of 25 and up.
 There is a clear different between the missingness conditions.
@@
 All other $p_{mis}$ conditions have bias irrespective of the number of iterations.
+-->
 
-$\widehat{R}$ differs a lot across iteration conditions.
-But not structurally for missingness proportions.
-Highest $\widehat{R}$s for the condition $T$=2, while the most bias is for ($T$=1), but there is no $\widehat{R}$ for that by definition.
-But for missingness, 25% seems the most signs of non-conv, while it has one of the least bias.
-Threshold of 1.2 reached in condition it = 3, $p_{mis}$ = 5 and 75, and then for it>6, $p_{mis}$ =  5 and 50, and $T$>10 for all $p_{mis}$.
-Threshold 1.1 reached in in $T$=12, $p_{mis}$ = all but 75, 1.01 only reached in $T$ = 93, $p_{mis}$ = 75.
-While for this $Q$, $T$=2 would be sufficient.
+### $\theta=$ chain mean.
+When $\widehat{R}$ and $AC$ are applied to the chain means, we notice no impact of the missingness proportions. The results only depend on the number of iterations per simulation condition.
 
-$AC$ mainly shows effects of the set of iteration conditions.
-$AC$ indicates less signs of non-convergencein conditions with higher number of iterations.
-For iteration conditions where $T$<10, there are also some effects of the missingness proportions.
-The $p_{mis}$ condtions 25 and 95 % indicate the most signs of non-convergence.
-For $p_{mis}$ = 95% this corresponds to the bias in Qbar.
-For 25% miss it doesn't.
-The thresholds defined by the critical value is only met in the conditions it=3, $p_{mis}$ = 5,75, and 95%.
+**[MAYBE remove all this because it has no relation to the bias in Q] The highest $\widehat{R}$-values are obtained for the condition $T=2$.
+We observe the same 'dip' in $\widehat{R}$ as we did in Figure \ref{fig:diagnostics} for conditions with three iterations.
+$\widehat{R}$ gradually decreases as the number of iterations increases across conditions.
+By the traditional threshold  $\widehat{R}>1.2$ we should conclude that there is non-convergence in conditions where $T<8$.
+The common threshold $\widehat{R}>1.1$ is overcome in conditions where $T>14$.
+**
+
+**Autocorrelations indicate some signs of non-stationarity in all simulation conditions, but primarily in conditions where $T<6$.
+The threshold for diagnosing non-convergence is not reached in any condition.
+We would thus conclude that the algorithm is sufficiently converged regardless of the number of iterations or missingness proportion.
+**
+
+We evaluate the diagnostics with chain means as $\theta$ against the bias in $Q=\mu$.
+According to the bias in the univariate means, we should conclude that two iterations is sufficient, and we should see that $p_{mis}=0.75$ and $p_{mis}=0.95$ have worse convergence than the other missingness proportions.
+This is not the case.
 
 ```{r mean, fig.height = 6.8, fig.cap="Convergence diagnostics chain mean."}
-mean_bias + mean_Rh + mean_AC + plot_layout(guides = "collect", ncol = 1) + plot_annotation(tag_levels = "A", tag_suffix = ".", title = bquote("Q = " ~ mu[Y] ~ "= 25.81; " ~ theta ~ "= chain mean of Y in" ~ y[imp][",m"])) #& theme(legend.position = "bottom")
+# mean_bias + mean_Rh + mean_AC + plot_layout(guides = "collect", ncol = 1) + plot_annotation(tag_levels = "A", tag_suffix = ".", title = bquote("Q = " ~ mu[Y] ~ "= 25.81; " ~ theta ~ "= chain mean of Y in" ~ y[imp][",m"])) #& theme(legend.position = "bottom")
 ```
 
-## Standard deviations
-
-**Message: even with possible effect of initial values ($T$=1), the algorithm is as well converged as the missingness proportion allows.** For $Q=\sigma_Y$ there is no effect visible of the iteration conditions, but clearly for the $p_{mis}$ conditions.
-Most bias for $p_{mis}$ = 95, followed by $p_{mis}$ = 50 and 75.
-For the least possible bias we may use any number of iterations $T\geq1$, but none of the $p_{mis}$ results in unbiased $\bar{Q}$s.
-
-There is a clear effect of the number of iterations on $\widehat{R}$.
-Conditions with lower Ts have more signs of non-conv, as identified by $\widehat{R}$.
-There is little difference between missingness conditions.
-Only for conditions where $T<5$ the missingness proportions seem to affect whether $\widehat{R}$ indicates non-convergence The bias in $\bar{Q}$ seems unrelated to the magnitude of $\widehat{R}$.
-Conditions with more bias ($p_{mis}$ = 50,75,95) are not identified to have more signs of non-convergence.
-The thresholds are therefore useless (because depend mostly on $T$, not so much $p_{mis}$).
-
-$AC$ is affected by both t and $p_{mis}$ conditions.
-But conditions with worst bias ($p_{mis}$ = 50, 75, 95) are not identified as worst converging according to AC.
-Moreover, only for $T$ =3 the $AC$ is significantly different from zero.
+### $\theta=$ chain variance.
+The non-convergence diagnostics applied to the chain variances as $\theta$ yield highly similar results to the chain means.
+We therefore evaluate these results immediately against the corresponding $Q$, without describing the $\widehat{R}$ and $AC$ values.
+While the bias in the estimated standard deviations  mainly depends on the missingness proportion of each condition, $\widehat{R}$ and $AC$ are influenced mostly by the number of iterations.
+We thus conclude that there is no added value of evaluating non-convergence with chain variances as $\theta$ in the diagnostics.
 
 ```{r sd, fig.height = 6.8, fig.cap="Convergence diagnostics chain variance."}
-sd_bias + var_Rh + var_AC + plot_layout(guides = "collect", ncol = 1) + plot_annotation(tag_levels = "A", tag_suffix = ".", title = bquote("Q = " ~ sigma[Y] ~ "= 11.32; " ~ theta ~ "= chain variance of Y in" ~ y[imp][",m"]))
+# sd_bias + var_Rh + var_AC + plot_layout(guides = "collect", ncol = 1) + plot_annotation(tag_levels = "A", tag_suffix = ".", title = bquote("Q = " ~ sigma[Y] ~ "= 11.32; " ~ theta ~ "= chain variance of Y in" ~ y[imp][",m"]))
 ```
 
-## Regression coefficients
+### $\theta=\hat{Q}$.
+The non-convergence diagnostics are applied to the estimated regression coefficients in each imputation, and evaluated against the bias in the corresponding pooled estimate.
 
-**Message: after four iterations ($T$=5), the algorithm is as well converged as the missingness proportion allows.
-Sooner even with lower missingness proportions ($T$=1 might be ok for $p_{mis}$ = 5 and 25).** The figure for $Q=\beta_1$ shows that the $\bar{Q}$s are affected by both the number of iterations and the missingness proportion.
-For $T$ =1, the magnitude of the bias (almost) follows the order of the missingness proportions.
-More missingness causes more bias in the $\bar{Q}$.
-In general, a lower nr of iterations results in more bias, but this depends on the missingness proportion.
-For $p_{mis}$ = 95, there is more bias in conditions where $T$<6.
-When $p_{mis}$ = 50, it's $T$<3.
-For $p_{mis}$ = 75 and 25, it's only for $T$<2.
-And conditions where $p_{mis}$ = 5 are unaffected on average.
-However, none of the conditions result in completely unbiased estimates of $Q$.
+$\widehat{R}$ indicates signs of non-convergence across all missingness proportions and each number of iterations, in the expected order: simulation conditions with higher missingness proportions and less iterations are identified as the worst converged.
+The thresholds to diagnose non-convergence are overcome in different conditions, depending on the number of iterations and the missingness proportion.
+The most lenient threshold ($\widehat{R}>1.2$) is overcome at $T\geq9$ for $p_{mis} = 0.05$, and $T\geq12$ for $p_{mis} = 0.95$.
+The conventional $\widehat{R}>1.1$ is surpassed at $T\geq12$ and $T\geq24$, respectively.
 
-We should see this in the conv diag: higher $\widehat{R}$ and $AC$ for $p_{mis}$ = 75 and 95, and for condtions where $T<2-6$, depending on $p_{mis}$.
+For the autocorrelations applied to the $\hat{Q}$s, we see an even stronger effect of the missingness proportions.
+Especially the conditions where $p_{mis} = 0.75$ and $p_{mis} = 0.95$ are identified as non-stationary.
+The threshold used to evaluate whether $AC=0$, however, is only reached in conditions with at least ten iterations.
+This suggests that the default number of iterations may be too low to detect non-stationarity with $AC$.
+However, the critical value for diagnosing non-convergence with $AC$ decreases as a function of the number of iterations.
+The conclusion to identify non-stationarity in conditions where $T\geq10$ and $p_{mis} = 0.95$ might therefore be spurious.
+**[Also, the $AC$ across missingness proportions shows the opposite of what you would expect: shouldn't there be higher autocorrelations in conditions with comparatively more observed data in the completed datasets?
+I don't understand this.]**
 
-$\widehat{R}$ depends somewhat on the missingness proportion.
-For conditions where $T$<10, the $\widehat{R}$ values follow the $p_{mis}$ more or less.
-But what we actually want to know is whether they are indicators of bias.
-Conditions with the worst bias (high $p_{mis}$, low T) have the largest $\widehat{R}$ values.
-The thresholds are not appropriate to diagnose sufficient convergence.
-All thresholds over-estimate the signs of non-convergence compared to the bias in Qbar.
-However, a higher threshold than 1.2 may result in terminating the algorithm after 3 iterations, which is insufficient for the $p_{mis}$ = 95 condition.
-Threshold of 1.25 or 1.3 could work for the other $p_{mis}$ conditions.
-
-$AC$ is very affected by the p_mis.
-The two conditions with the most bias ($p_{mis}$ = 75 and 95) are the only ones identified as non-converging, as they should.
-However, the $p_{mis}$ = 75 condition is only identified in the $T$ =3 condtion, while the bias is the least pronounced there.
-And the $AC$ only identifies non-convergence in the $p_{mis}$ = 95 conditions where $T$>10.
-In the default situation, $T$ =5, this would be overlooked.
-Also, the $AC$ shows the opposite of what you would expect from a theta that is based on the observed data: shouldn't there be higher autocorrelations in conditions with comparatively more observed data in the completed datasets?
-I don't understand this.
-
-**Add how interesting it is because we literally track $Q$ and are still not ok.**
+Compared to the bias in $\bar{Q}$, the diagnostics applied to the $\hat{Q}$s over-estimate the severity of the non-convergence. This is odd, since the $\theta$ and $Q$ are directly related.
 
 ```{r est, fig.height = 6.8, fig.cap="Convergence diagnostics regression coefficient."}
-est_bias + est_Rh + est_AC + plot_layout(guides = "collect", ncol = 1) + plot_annotation(tag_levels = "A", tag_suffix = ".", title = bquote("Q = " ~ beta[1] ~ "= 2.06; " ~ theta ~ "= " ~ hat(Q) ~ "in {" ~ y[obs] ~ "," ~ y[imp][",m"] ~ "}"))
+# est_bias + est_Rh + est_AC + plot_layout(guides = "collect", ncol = 1) + plot_annotation(tag_levels = "A", tag_suffix = ".", title = bquote("Q = " ~ beta[1] ~ "= 2.06; " ~ theta ~ "= " ~ hat(Q) ~ "in {" ~ y[obs] ~ "," ~ y[imp][",m"] ~ "}"))
 ```
 
-## Variance explained
+### $\theta=\lambda_1$.
+Finally, we evaluate the non-convergence diagnostics applied to the first eigenvalue of the variance-covariance matrices of the completed data.
+Similar to the results the the regression coefficients as $\theta$s, we see effects of both the missingness proportion and number of iterations in each simulation condition.
+The most signs of non-convergence are identified in conditions where $p_{mis}$ is large and $T$ is small.
 
-**Message: after two iterations ($T$=3), the algorithm is as well converged as the missingness proportion allows.
-Sooner even with lower missingness proportions ($T$=1 might be ok for $p_{mis}$ = 5 and 25).** Bias is the worst for conditions where $p_{mis}$ = 95, $T$ <3.
-Followed by $p_{mis}$ = 75, $T$ =1; and $p_{mis}$ = 50, $T$ = 1.
-Conditions where $p_{mis}$ = 25 or 5 seem unaffected by the number of iterations.
-Minimal bias is obtained in all iteration conditions.
-We should see this in $\widehat{R}$ and AC: worst convergence in $p_{mis}$ = 95 and 50; some non-convergencein conditions where $p_{mis}$ = 75 and $T$ =1, and no non-convergencein conditions where $p_{mis}$ = 25 or 5.
-The sign of the bias is logical: when there is less info in the data, the relations between the predictors and the outcome is under-estimated.
+Notwithstanding the similarities with $\theta=\hat{Q}$, there is a big difference in $\widehat{R}$-values for conditions where $T<4$.
+In these conditions, $\widehat{R}$ behaves oppositely of our expectation: conditions with the lowest missingness proportions yield the highest $\widehat{R}$-values.
+In conditions with a higher number of iterations, this artifact disappears.
+This suggests that the combination of $\widehat{R}$ and $\lambda_1$ is only appropriate for imputation algorithms where the number of iterations at least four.
+Interpreting the $\widehat{R}$ values of conditions where $T<3$ may lead to the incorrect conclusion that only conditions with *low* missingness proportions show signs of non-convergence.
+For applications as a model-independent multivariate $\theta$, we should therefore only interpret the $\widehat{R}$-values when $T\geq4$.
+Using this cut-off we conveniently ignore the initial 'dip' at $T=3$ as well.
+The traditional threshold $\widehat{R} > 1.2$ is overcome in conditions where $p_{mis} = .95$ and $T\geq10$, and for all other missingness proportions in conditions where $T\geq8$.
+With the $\widehat{R} > 1.1$ threshold, the number of iterations are $T\geq20$ and $T\geq16$, respectively.
 
-$\widehat{R}$ behaves oppositely to what we wanted.
-The lowest $\widehat{R}$ values for conditions where $T$<10 are observed for the two missingness conditions with the worst bias.
-However, the highest $\widehat{R}$ observed is indeed for the missingness condition with the worst bias, $p_{mis}$ =95.
-But only at $T$ =5 (while bias is no worse than anywhere else in this $p_{mis}$ condition, conditional on iterations condition $T$>2).
-So neither of the thresholds works perfectly.
+The $AC$-values largely depend on the missingness proportion in each condition.
+Conditions where $p_{mis}=.95$ show statistically significant autocorrelations for any number of iterations.
+The other extreme, conditions where $p_{mis}=.05$, does not yield $AC$ values above the threshold.
+The $AC$ values with this missingness proportion are equivalent for all conditions where $T\geq6$, which suggests that the algorithm is as stationary as it can be.
+A similar point is only observed after twenty iterations in conditions with higher missingness proportions.
 
-$AC$ shows a similar trend to the biases: more signs of non-convergencefor higher $p_{mis}$ and lower T.
-The threshold, however, performs less well.
-It identifies signs of non-convergence for all conditions where $T=3$, irrespective of p_mis.
-Then nothing, and then for $p_{mis}$ =95, $T$>7, and $p_{mis}$ = 50, $T$ >14, and $p_{mis}$ = 75, $T$>20.
-This does not corresond to the bias in $\bar{Q}$.
-Maybe use AC>.5 at $T$ =5 as threshold?
+In comparison to the $Q$s, the diagnostics again over-estimate the signs of non-convergence.
+Conditions with low missingness proportions (e.g., 5% or 25% of the complete data) yield unbiased, confidence valid estimates with as little as two iterations.
+The convergence diagnostics, however, indicate improved convergence up-to eight iterations according to $AC$ and thirty to forty iterations according to $\widehat{R}$.
+The model-independent $\theta$ $\lambda_1$ seems as well-equipped to diagnose signs of non-convergence as $\theta=\hat{Q}$.
+The univariate $\theta$s do not differentiate between bias induced by the missingness proportions.
 
 ```{r pred, fig.height = 6.8, fig.cap="Convergence diagnostics regression coefficient."}
-Rsq_bias + PCA_Rh + PCA_AC + plot_layout(guides = "collect", ncol = 1) + plot_annotation(tag_levels = "A", tag_suffix = ".", title = bquote("Q = " ~ r^2 ~ "\u00D7 100 = 19.25; " ~ theta ~ "= " ~ lambda[1] ~ "in {" ~ y[obs] ~ "," ~ y[imp][",m"] ~ "}")) #first eigenvalue of" ~ Sigma ~
+# Rsq_bias + PCA_Rh + PCA_AC + plot_layout(guides = "collect", ncol = 1) + plot_annotation(tag_levels = "A", tag_suffix = ".", title = bquote("Q = " ~ r^2 ~ "\u00D7 100 = 19.25; " ~ theta ~ "= " ~ lambda[1] ~ "in {" ~ y[obs] ~ "," ~ y[imp][",m"] ~ "}")) #first eigenvalue of" ~ Sigma ~
 ```
@@
 
+## In short [remove later!]
+
+- Is there more bias in conditions where non-convergence was induced? (NOPE for univariates, YES for multivar.).
+
+- Is the point of *sufficient* convergence (at which the $\bar{Q}$s are unbiased and confidence valid estimates of the $Q$s) the same point as identified by the non-convergence diagnostics?
+(NOPE, they are too conservative).
+
+- Is one of the thetas better? (YES, while it doesn't really matter as long as it's multivariate, $\lambda_1$ is model-independent and therefore preferred).
+
+- What should the thresholds be? (much HIGHER than recommended for $\widehat{R}$, difficult to say for $AC$).
+
+- And the default nr of iterations? (MAYBE increase to 10 to detect significant $AC$ and be sure not to miss the dip in $\widehat{R}$?).
+
 # Discussion
 
-In short, univariate estimates seem robust against terminating the algorithm early: There is no clear effect of the number of iterations on the bias in the estimated means and standard deviations.
-The bias in these $Q$s only depends on the $p_{mis}$, with higher $p_{mis}$ resulting in more bias.
-Yet, the convergence diagnostics indicate that the algorithm did not reach a completely converged state (yet).
+In short, iterative imputation algorithms may yield approximately unbiased, confidence valid estimates under different levels of induced non-convergence.
+Univariate estimates seem robust against terminating the algorithm early: There is no clear effect of the number of iterations on the bias in estimated means and standard deviations.
+The bias in these $Q$s was only impacted by the missingness proporion $p_{mis}$, with higher $p_{mis}$ resulting in more bias.
+Bias in multivariate estimates depended on both $p_{mis}$ and the number of iterations.
+However, the only effect of the number of iterations was observed when the algorithm was terminated at seven iterations or earlier.
+This suggests that continued iteration beyond seven iterations does not yield better estimates.
+Yet, the non-convergence diagnostics $\widehat{R}$ and $AC$ indicated that the iterative imputation algorithm did not reach a stable state before twenty to thirty iterations.
+**[MERGE or remove] $\widehat{R}$ and autocorrelation indicate that there are signs of non-convergence in the algorithm up-to at least twenty iterations, while unbiased, confidence valid estimates may be obtained with as little as one iteration.
+This observation is in agreement with **one of [still add the others!]** the simulation hypotheses: $\widehat{R}$ over-estimates the severity of non-convergence when applied to MI procedures.**
 
-$\widehat{R}$ and autocorrelation indicate that algorithmic convergence may only be reached after twenty or even forty iterations, while unbiased, confidence valid estimates may be obtained with as little as one iteration.
-These results are in agreement with the simulation hypothesis: $\widehat{R}$ over-estimates the severity of non-convergence when applied to MI procedures.
-
-With that, we show that MICE can lead to correct outcomes when they have not converged according to two common conv diags.
-This may be due to the methods (and their thresholds) or due to the $Q$s (descriptives and multivariate linear regression, not higher dimensional/more complex $Q$s).
-More `complicated' $Q$s (e.g., higher-order effects or variance components) might show bias, under- or over-coverage at higher $T$, as indiated by $\widehat{R}$ and AC.
-Add what %miss has to do with it.
+Over-estimation of the signs of non-convergence may be due to the methods (and their thresholds) or due to the $Q$s (descriptive and multivariate linear regression, not higher dimensional/more complex $Q$s).
+More complicated $Q$s (e.g., higher-order effects or variance components) might show bias, under- or over-coverage at higher $T$.
+Which might be why $\widehat{R}$ and $AC$ identify non-convergence.
 
@@
 
-1. Check trace plots for pathological non convergence and adjust imputation model if necessary.
+1. Check traceplots for signs of pathological non-convergence. Adjust imputation model if necessary.
 
-2. use $\widehat{R}$ wait 1.1 ash threshold and autocorrelation with **?** as threshold. Keep iterating until these thresholds are reached.
+2. Monitor non-convergence diagnostics $\widehat{R}$ and $AC$. Use the thresholds for $\widehat{R}>1.2$ and autocorrelation  **???**. Keep iterating until these thresholds are reached.
 
-3. Do not use the R function ACF. Instead, compute autocorrelations manually **(see e.g., [GitHub link])**.
+3.
+Do not use the traditional calculation for $\widehat{R}$, or the `R` function `stats::acf()` to compute $AC$.
+Instead, calculate $\widehat{R}$ conform @veht19 and compute autocorrelations manually (see e.g., [github.com/gerkovink/shinyMice](https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy)).
 
-4. Track your own scalar summary of interest. This is somewhat advanced but explained in Van Buren 2018. Compute $\widehat{R}$ and autocorrelation values for this scalar summary.
+4. Track your own scalar summary of interest (see e.g., @buur18). Compute $\widehat{R}$ and autocorrelation values for this scalar summary.
 
@@
 
-Further research is needed to investigate their performance under clear violation of convergence, e.g. dependency between predictors (predictors with very high correlations). Also:
+Further research is needed to investigate the performance of iterative imputation algorithms under clear violation of convergence, e.g.
+dependency between predictors (predictors with very high correlations).
+For example:
 
@@
 
-- More difficult missingness problems, i.e., M(N)AR instead of MCAR.
-Proper performance under a 'missing completely at random' missingness mechanism is necessary to demonstrate the appropriateness of $\widehat{R}$ and $AC$ as non-convergence diagnostics.
+- Introduce more difficult missingness problems, i.e., M(N)AR instead of MCAR.
+That is, proper performance under a 'missing completely at random' missingness mechanism is necessary to demonstrate the appropriateness of $\widehat{R}$ and $AC$ as non-convergence diagnostics.
 However, results may not be extrapolated to other missingness mechanisms.
@@
 
-Uitsmijter: To conclude, in the case of drawing inference from incomplete data, convergence of the iterative imputation algorithm is often a convenience but not a necessity.
+In short, we have shown that an iterative imputation algorithm can yield correct outcomes, even when a converged state is not reached according to two common non-convergence diagnostics.
+Continued iteration just burns a lot of computing time, without improving the statistical inferences.
+To conclude, in the case of drawing inference from incomplete data, convergence of the iterative imputation algorithm is often a convenience but not a necessity.
 
-# Grading
+<!-- # Grading -->
 
-- Scientific contribution 
-(clear research question, convincing argumentation that the study adds to the scientific literature)
+<!-- - Scientific contribution  -->
+<!-- (clear research question, convincing argumentation that the study adds to the scientific literature) -->
 
-- Theory
-(use of theory, clear hypotheses, relevant literature)
+<!-- - Theory -->
+<!-- (use of theory, clear hypotheses, relevant literature) -->
 
-- Method
-(correct data collection/handling, measures, choice and description of method of analysis)
+<!-- - Method -->
+<!-- (correct data collection/handling, measures, choice and description of method of analysis) -->
 
-- Results
-(correct interpretation of findings, clear presentation)
+<!-- - Results -->
+<!-- (correct interpretation of findings, clear presentation) -->
 
-- Discussion and conclusions
-(summary, broader implications, limitations, future research)
+<!-- - Discussion and conclusions -->
+<!-- (summary, broader implications, limitations, future research) -->
 
-- Written presentation
-(writing, structure, tables and figures, references)
+<!-- - Written presentation -->
+<!-- (writing, structure, tables and figures, references) -->
 

### 2020-05-04 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Draft for final feedback

@@
 There is no consensus on how to evaluate the convergence properties of the method.
-This paper provides insight into identifying non-convergence of iterative impuation algorithms.
+This paper provides insight into identifying non-convergence of iterative imputation algorithms.
 \end{abstract}
@@
 
-Currently, the recommended practice for evaulating the convergence of iterative imputation algorithms is through visual inspection.
+Currently, the recommended practice for evaluating the convergence of iterative imputation algorithms is through visual inspection.
 That is, the state space of the algorithm is monitored by plotting a scalar summary \(\theta\) against the iteration number.
@@
 The aim of the simulation is to determine whether unbiased, confidence valid inferences may be obtained under different levels of non-convergence (\textbf{and at which point convergence may safely be assumed}).
-Additionally, we will assess how well different non-convergence diagnostics perform in identifying signs of non-convergence.
-The results of this study are guidelines for identifying non-convergence in iterative imptation algorithms, which will aid applied researchers in drawing valid inference from incomplete datasets.
+Additionally, we will assess how well different non-convergence diagnostics perform in identifying signs of non-convergence (\textbf{and on which scalar summary of the state space of the algorithm these diagnostics should be applied}).
+The results of this study are guidelines for identifying non-convergence in iterative imputation algorithms, which will aid applied researchers in drawing valid inference from incomplete datasets.
 \textbf{{[}add more societal relevance: what is the consequence of non-convergence?
-e.g., biased estimates, invalid inference, etc.{]}} For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular \texttt{mice} package \citep{mice} in \texttt{R} \citep{R}.
+see example of non-convergence in Figure 2 (e.g., biased estimates, invalid inference, etc.){]}} For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular \texttt{mice} package \citep{mice} in \texttt{R} \citep{R}.
 
@@
 The collection of samples between the initial value (at \(t=1\)) and the final imputed value (at \(t=T\)) will be referred to as an `imputation chain'.
-\textbf{\(\theta\)s are scalar summaries of the algorithm at a certain iteration (e.g., chain means; the average of the imputed values in each imputation chain).
-}
+\textbf{{[}If not defined elsewhere, add: \(\theta\)s are scalar summaries of the algorithm at a certain iteration (e.g., chain means; the average of the imputed values in each imputation chain).{]} }
 
@@
 Each line portrays one imputation.
-The plot on the lefthand side of the figure shows typical convergence of an iterative imputation algorithm.
-The righthand side displays pathological non-convergence, induced by purposefully mis-specifying the imputation model.
+The plot on the left-hand side of the figure shows typical convergence of an iterative imputation algorithm.
+The right-hand side displays pathological non-convergence, induced by purposefully mis-specifying the imputation model.
 \textbf{{[}Leave this in if it's after mixing and stationarity are defined, otherwise remove?{]} In the typical convergence situation, the imputation chains intermingle nicely and there is little to no trending.
@@
 This is typically done through traceplots.
-In a traceplot, the iteraton number is plotted against a certain \(\theta\).
+In a traceplot, the iteration number is plotted against a certain \(\theta\).
 \(\theta\)s are scalar summaries of the state space of the algorithm at a specific iteration.
@@
 \geq \lambda_j\) be the eigenvalues of \(\Sigma\) in each of the \(M\) completed datasets \({y_{obs}, y_{imp}}\).
-\(\lambda_1\) is measure that summarises the covariances in the completed datasets.
+\(\lambda_1\) is measure that summarizes the covariances in the completed datasets.
 \textbf{The first eigenvalue has the appealing property that is not dependent on the substantive model of interest.
@@
 
-Non-convergence diagnostics for MCMC algorithms typically identify problems in either one of the two requirements for converegnce: mixing or stationarity.
+Non-convergence diagnostics for MCMC algorithms typically identify problems in either one of the two requirements for convergence: mixing or stationarity.
 Non-stationarity within chains may be diagnosed with e.g., autocorrelation \citep[\(AC\);][]{scha97, gelm13}, numeric standard error \citep[`MC error';][]{gewe92}, or Raftery and Lewis's \citeyearpar{raft91} procedure to determine the effect of trending on the precision of estimates.
@@
 
+The potential scale reduction factor \(\widehat{R}\) was coined by \citet{gelm92}.
 An updated version of \(\widehat{R}\) has been proposed by Vehtari et al.~\citeyearpar[p.~5]{veht19}.
-\textbf{This version may be suitable for iterative imputation.} Let \(M\) be the total number of chains, \(T\) the number of iterations per chain, and \(\theta\) the scalar summary of interest (e.g., chain means).
+\textbf{This version is better suited to detect non-convergence in the tails of distributions, by using three transformations on the scalar summary that it is applied to.
+Namely, folded, rank-normalized, split-chain \(\theta\)s.
+The Vehtari et al.~version of \(\widehat{R}\) may be suitable for iterative imputation.
+We therefore use their definition of the diagnostic.} Let \(M\) be the total number of chains, \(T\) the number of iterations per chain, and \(\theta\) the scalar summary of interest (e.g., chain means).
 For each chain (\(m = 1, 2, \dots, M\)), we estimate the variance of \(\theta\), and average these to obtain within-chain variance \(W\).
@@
 
-We then estimate between-chain variance \(B\) as the variance of the collection of average \(\theta\) per chain.
+\noindent We then estimate between-chain variance \(B\) as the variance of the collection of average \(\theta\) per chain.
 
@@
 
-From the between- and within-chain variances we compute a weighted average, \(\widehat{\operatorname{var}}^{+}\), which over-estimates the total variance of \(\theta\) \textbf{remove or explain why, or leave in}.
-\(\widehat{R}\) is then obtained as a ratio between the over-estimated total variance and the within-chain variance:
+\noindent From the between- and within-chain variances we compute a weighted average, \(\widehat{\operatorname{var}}^{+}\), which approximates the total variance of \(\theta\).
+\(\widehat{R}\) is then obtained as a ratio between the total variance and the within-chain variance:
 
@@
 Over-dispersion implies that the initial values of the chains are `far away' from the target distribution and each other.
-When all chains sample independent of their initial values, the mixing component of convergence is satisfied, and \(\widehat{R}\)-values will be close to one.
+When the sampled values in each chain are independent of the chain's initial value, the mixing component of convergence is satisfied.
+The variance between chains is then equivalent to the variance within chains, and \(\widehat{R}\)-values will be close to one.
 High \(\widehat{R}\)-values thus indicate non-convergence.
@@
 But \(\widehat{R}=1\) and \(AC=0\) should at least be able to distinguish between the two algorithms plotted in Figure \ref{fig:non-conv}.
-From visual inspection, we know that the typically converged algorithm initially portrayed some signs of non-mixing (around \(t=2\)), but intermigled nicely overall.
+From visual inspection, we know that the typically converged algorithm initially portrayed some signs of non-mixing (around \(t=2\)), but intermingled nicely overall.
 Additionally, there was very little trending.
@@
 Panel C shows the traditional computation of \(\widehat{R}\), and panel D shows \(\widehat{R}\) as computed by implementing \citet{veht19} 's recommendations.
+\textbf{{[}Add labels for the two types of Rhat to the figure!{]}}
 
@@
 \textbf{Describe panel C here and why it is not optimal{]}} The \(\widehat{R}\)-values in panel D do meet the requirements specified above.
-\(\widehat{R}\) as computed conform \citet{veht19} indeed indicate less signs of non-convergence as the number of iterations goes up.
+\(\widehat{R}\) as computed conform \citet{veht19} does indeed indicate less signs of non-convergence as the number of iterations goes up.
+\textbf{\citep[explain the dip in Rhat values at t=2.
+Namely, because we can only use 2 of the 3 tricks by][ if the number of iterations is very low (t\textless4).
+That's why the \(\widehat{R}\)s are more similar to the traditional GR.]{veht19}}
 
@@
 In an empirical study, where \(\widehat{R}\) was used to inform the required imputation chain length, it took as many as 50 iterations to overcome the conventional non-convergence threshold \(\widehat{R}>1.2\).
-Yet, scientific estimates were insensitive to continued iteration from \(t=5\) onwards \citep{lace07}.
+Yet, scientific estimates were insensitive to continued iteration from \(t=5\) onward \citep{lace07}.
 We therefore suspect that \(\widehat{R}\) may over-estimate signs of non-convergence in iterative imputation algorithms, compared to the bias the estimates.
@@
 \item
-  \textbf{Something about thetas}
+  \textbf{{[}Something about thetas, that multivariate thetas are better at detecting non-convergence than univariate thetas.
+The non-convergence diagnostics applied on the chain means and chain variances will fail to identify signs of non-convergence when the multivariate Qs are affected.{]}}
 \end{enumerate}
@@
 From the complete set, we obtain the true values of several quantities of scientific interest, \(Q\)s.
-\textbf{{[}MOVE THIS NEXT PART TO TOP OF SIMULATION SECTION?{]}} \(Q\)s are the targets that will be estimated in each simulation repetition, after first \emph{amputing} the complete data with varying missingness proportions, then imputing the missingness with a varying number of iterations, then estimating the \(Q\)s on the \(M\) completed datasets, and finally pooling the \(M\) estimates \(\hat{Q}\) to obtain a single \(\bar{Q}\) for each \(Q\).
-\textbf{We try to predict the effect of non-converegence (i.e., the difference between \(Q\) and \(\bar{Q}\)) with the convergence diagnostics \(\widehat{R}\) and \(AC\), by appliying these methods to several scalar summaries of the state space of the algorithm, \(\theta\)s.}
+\textbf{{[}MOVE this next part to the start of the Simulation Study section?{]} \(Q\)s are the targets that will be estimated in each simulation repetition, after first \emph{amputing} the complete data with varying missingness proportions, then imputing the missingness with a varying number of iterations, then estimating the \(Q\)s on the \(M\) completed datasets, and finally pooling the \(M\) estimates \(\hat{Q}\) to obtain a single \(\bar{Q}\) for each \(Q\).
+We try to predict the effect of non-convergence (i.e., the difference between \(Q\) and \(\bar{Q}\)) with the convergence diagnostics \(\widehat{R}\) and \(AC\), by applying these methods to several scalar summaries of the state space of the algorithm, \(\theta\)s.}
 
@@
   Perform multiple linear regression with \texttt{stats::lm()} \citep{R} to get the \(\hat{Q}\)s for \(Q=\beta\) and \(Q=r^2\).
-We pool the \(\hat{Q}\)s conform \citet{vink14} to get \(\bar{Q}\)s for the \(\beta\)s and \(r^2\) \textbf{{[}check pooling of \(r^2\), instead of \texttt{mice::pool.r.squared()}{]}}.
+We pool the \(\hat{Q}\)s conform \citet{vink14} to get \(\bar{Q}\)s for the \(\beta\)s and \(r^2\) \textbf{{[}check finite pooling of \(r^2\), instead of \texttt{mice::pool.r.squared()}{]}}.
 Additionally, we use the \(\hat{Q}\)s of \(Q=\beta\) as \(\theta\) to apply the convergence diagnostics to.
@@
 where \(t_{(M-1)}\) is the quantile of a \(t\)-distribution with \(M-1\) degrees of freedom, and \(SE_{\bar{Q}}\) is the square root of the pooled variance estimate.
-If we obtain nominal coverage (CR = 95\%), we can conclude that the \(\bar{Q}\)s are conficence valid estimates of the \(Q\)s.
+If we obtain nominal coverage (CR = 95\%), we can conclude that the \(\bar{Q}\)s are confidence valid estimates of the \(Q\)s.
 Finally, we inspect CI width (CIW): the difference between the lower and upper bound of the 95\% confidence interval around \(\bar{Q}\).
@@
 With these performance measures, we also evaluate how well the non-convergence diagnostics \(\widehat{R}\) and \(AC\) can identify simulation conditions in which the \(\bar{Q}\)s are \emph{not} unbiased, confidence valid estimates of the \(Q\)s.
-\textbf{Rephrase and add corresponding \(Q\)s:} We measure the performance of \(\widehat{R}\) and \(AC\) by the four scalar summaries \(\theta\): chain means (i.e., the mean in each \(y_{imp, m, j}\)) versus \(Q=\mu\), chain variances (i.e., the variance in each \(y_{imp, m, j}\)) versus \(Q=\sigma\), the estimated regression coefficients \(\beta\) in each \{\(y_{obs}, y_{imp, m}\)\} (i.e., a user-defined statistic of scientific interest \(\hat{Q}\)) against \(Q=\beta\), and the first eigenvalue of the variance-covariance matrix in each \{\(y_{obs}, y_{imp, m}\)\}, \(\lambda_1\) versus \(Q=r^2\).
+\textbf{{[}Rephrase this:{]} We measure the performance of \(\widehat{R}\) and \(AC\) by the four scalar summaries \(\theta\): chain means (i.e., the mean in each \(y_{imp, m, j}\)) versus \(Q=\mu\), chain variances (i.e., the variance in each \(y_{imp, m, j}\)) versus \(Q=\sigma\), the estimated regression coefficients \(\beta\) in each \{\(y_{obs}, y_{imp, m}\)\} (i.e., a user-defined statistic of scientific interest \(\hat{Q}\)) against \(Q=\beta\), and the first eigenvalue of the variance-covariance matrix in each \{\(y_{obs}, y_{imp, m}\)\}, \(\lambda_1\) versus \(Q=r^2\).}
+
+\textbf{{[}Remove this?{]} Primarily, we want to know what the effect is of inducing non-convergence on the estimated \(Q\)s.
+More specifically, we are interested in the bias of \(\bar{Q}\) for all \(Q\)s and the confidence validity of \(\bar{Q}\) for \(Q=\beta\).
+Therefore, we evaluate these performance measures against the simulation conditions (the number of iterations \(T\) and the missingness proportion \(p_{mis}\)).
+We look at averages across simulation repetitions.
+Secondly, we want to know if conditions in which the \(Q\)s are affected by non-convergence are also identified by the non-convergence diagnostics \(\widehat{R}\) and \(AC\).
+We do this, again, by looking at the performance measures for the \(Q\)s, but now we evaluate these against what \(\widehat{R}\) and \(AC\) indicate about signs of non-convergence in the \(\theta\)s.}
 
@@
 
-\textbf{{[}Add to this section: 1) emphasize that the iterations in the plots are averages across repetitions, not within one run of MICE; 2) sample effects due to single complete dataset??
-3) Add more info about figure legends and axes 4) the ac and rhat are independent of missingness, not the thetas.{]}}
-Primarily, we want to know what the effect is of inducing non-convergence on the estimated \(Q\)s.
-Specifically, we are interested in the bias of \(\bar{Q}\) for all \(Q\)s and the confidence validity of \(\bar{Q}\) for \(Q=\beta\).
-Therefore, we plot these performance measures against the simulation conditions (the number of iterations \(T\) and the missingness proportion \(p_{mis}\)).
-We look at averages across simulation repetitions.
+\textbf{{[}Add to this section: 1) emphasize that the iterations in the plots are averages across repetitions, not within one run of MICE; 2) Add more info about figure legends and axes{]}}
 
-Secondly, we want to know if conditions in which the \(Q\)s are affected by non-convergence are also identified by the non-convergance diagnostics \(\widehat{R}\) and \(AC\).
-We do this, again, by looking at the performance measures for the \(Q\)s, but now we evaluate these against what \(\widehat{R}\) and \(AC\) indicate about signs of non-convergence in the \(\theta\)s.
+Our results show the impact of inducing non-convergence in an iterative imputation algorithm on several scientific estimates, and whether the non-convergence diagnostics \(\widehat{R}\) and \(AC\) identify the signs of non-convergence in the algorithm.
+For reasons of brevity, we only discuss the non-convergence diagnostics for estimates with the worst performance in terms of bias.
+Bias in the descriptive statistics (\(Q=\mu\) and \(Q=\sigma\)) is most pronounced for the outcome variable \(Y\).
+Of the regression coefficients (\(Q=\beta\)) we observe the largest bias for \(\beta_1\) (the effect of \(X_1\) on \(Y\)).
+There is only one estimate for \(Q=r^2\) to consider.
+In the figures, we present simulation conditions where the number of iterations is \(1 \leq T\leq50\), since the results are more or less stable for conditions where \(T \leq 30\).
+Full results are available from \href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy}{github.com/gerkovink/shinyMice}.
 
-\textbf{{[}In short:{]}}
-- Is there more bias in conditions where non-convergence was induced? (NOPE for univariates, YES for multivar.).
+\hypertarget{quantities-of-scientific-interest}{%
+\subsection{Quantities of scientific interest}\label{quantities-of-scientific-interest}}
+
+\begin{figure}
+
+{\centering \includegraphics{manuscript_files/figure-latex/Qs-1}
+
+}
+
+\caption{Impact of non-convergence on statistical inferences.}\label{fig:Qs}
+\end{figure}
+
+Figure \ref{fig:Qs} displays the influence of inducing non-convergence on the estimated quantities of scientific interest, \(Q\)s.
+\textbf{{[}Add panel labels, describe panels{]}} As expected, conditions with a higher proportion of missingness and/or a lower number of iterations are impacted more severely by non-convergence.
+I.e., on average these conditions portray more extreme bias in the estimated \(Q\)s, and non-nominal coverages.
+Roughly speaking, this means that the algorithm is indeed not converged at \(t=1\), and converges gradually as \(t\) increases to \(T\).
+The point at which an additional iteration does not lead to an improvement in the estimates depends on quantity of scientific interest \(Q\) and the missingness proportion \(p_{mis}\).
+The highest number of iterations that is necessary to reach this point across all simulation conditions is \(T=7\).
+This implies that, under the current specifications, no more than seven iterations are necessary to obtain optimal\textbf{/the best possible} performance in terms of bias and confidence validity.
+
+\hypertarget{qmu.}{%
+\subsubsection{\texorpdfstring{\(Q=\mu\).}{Q=\textbackslash mu.}}\label{qmu.}}
+
+The estimated univariate means seem unaffected by the number of iterations in the algorithm.
+The magnitude of the bias in \(\bar{Q}\) depends exclusively on the missingness proportion \(p_{mis}\), with higher missingness proportions causing more extreme bias.
+Note however, that the simulation condition with the largest overall bias resulted in a \(\bar{Q}\) under-estimated the true value of \(Q\) by only 0.02 units (where \(Q=\mu_Y=25.81; \sigma_Y = 11.32\)).
+We conclude that---conditional on the missingness proportion---all simulation conditions (\(T \geq 1\)) have equal performance.
+According to this scientific estimand \(Q\), non-convergence does not impact the validity of the inferences.
+\textbf{{[}Add: up-to 50\% missingness everything is relatively unbiased, while 75\% and 95\% missingness show very different results.
+Why is that?
+Add sd of Y in the figure to show that the bias is relative: a bias of 0.02 for a variable with a mean of 25.81, and sd of 11.32 is nothing.{]}}
+
+\hypertarget{qsigma.}{%
+\subsubsection{\texorpdfstring{\(Q=\sigma\).}{Q=\textbackslash sigma.}}\label{qsigma.}}
+
+The estimated standard deviations show impact of non-convergence similar to the means.
+The magnitude of the bias \(\bar{Q}\) follows the incremental order of the missingness proportions.
+The number of iterations in the algorithm has little effect on the bias.
+Only the estimates in simulation conditions with just one iteration (\(T=1\)) differ from other conditions that have the same missingness proportion but a higher number of iterations.
+Interestingly, the bias in conditions where \(T=1\) is \emph{less} severe than after continued iteration.
+For the least possible bias, we may use any number of iterations \(T\geq1\), but only \(p_{mis} = 0.05\) results in completely unbiased \(\bar{Q}\)s.
+
+\hypertarget{qbeta.}{%
+\subsubsection{\texorpdfstring{\(Q=\beta\).}{Q=\textbackslash beta.}}\label{qbeta.}}
+
+For the estimated regression coefficients, we consider bias, confidence interval width (CIW), and coverage rate (CR) as performance measures.
+The magnitude of the bias, again, depends on the missingness proportion.
+But for this \(Q\) we also observe a clear effect of the number of iterations.
+Conditions with a lower number of iterations yield more bias in \(\bar{Q}\).
+This is especially pronounced in conditions with high missingness proportions: with 95\% missingness we need at least seven iterations before an additional iteration does not lead to improved estimates.
+Conditions with a lower missingness proportion (e.g., \(p_{mis} = {0.5}\)), produce \emph{optimal} estimates with as many as two iterations.
+Approximately unbiased estimates are obtained for all conditions where \(p_{mis}\leq0.75\) and \(T\leq6\).
+\textbf{{[}Add CIW and CR{]}}
+
+\hypertarget{qr2.}{%
+\subsubsection{\texorpdfstring{\(Q=r^2\).}{Q=r\^{}2.}}\label{qr2.}}
+
+The bias in the estimated explained variance perfectly follows both sets of simulation conditions: The worst performance is observed for the highest missingness proportion and the lowest number of iterations, while the best performance occurs with less missingness and more iterations.
+Only conditions with the lowest missingness proportion (\(p_{mis}=0.05\)) results in unbiased estimates for every number of iterations.
+In other conditions there is at least some bias irrespective of \(T\).
+The inferences do not improve in any condition where \(T \geq 6\).
+This suggests that the algorithm is as converged as necessary to obtain valid inferences.
+
+\hypertarget{non-convergence-diagnostics}{%
+\subsection{Non-convergence diagnostics}\label{non-convergence-diagnostics}}
+
+In Figure \ref{fig:rhats-acs} we see the results of the non-convergence diagnostics \(\widehat{R}\) and \(AC\).
+\textbf{{[}Add panel labels and explain how to read the plots{]}} As expected, the diagnostics indicate more signs of non-convergence in simulation conditions with a lower number of iterations.
+The missingness proportions, however, only seem to impact multivariate convergence.
+When \(\widehat{R}\) and \(AC\) are applied to univariate \(\theta\)s, they identify the same amount of non-convergence, irrespective of the missingness proportion.
+When the diagnostics are applied to multivariate \(\theta\)s, they generally identify more signs of non-convergence for higher missingness proportions.
+
+None of the simulation conditions that we consider resulted in complete convergence of the algorithm (defined as \(\widehat{R}=1\) and \(AC=0\)).
+The thresholds for diagnosing non-convergence that are used in practice seem too conservative compared to the performance measures: \(\widehat{R}\) and \(AC\) identify signs of non-convergence in conditions where the \(\bar{Q}\)s are unbiased and confidence valid estimates of the \(Q\)s.
+The most recent recommended threshold for diagnosing non-convergence with \(\widehat{R}\) is never overcome in this study (all \(\widehat{R}\)\textgreater1.01).
+Apparently, this threshold is too stringent for iterative imputation algorithms, and will not be discussed further.
+
+\begin{figure}
+
+{\centering \includegraphics{manuscript_files/figure-latex/rhats-acs-1}
+
+}
+
+\caption{Non-convergence diagnostics. [Split up into 2 or 4 figures??]}\label{fig:rhats-acs}
+\end{figure}
+
+\hypertarget{theta-chain-mean.}{%
+\subsubsection{\texorpdfstring{\(\theta=\) chain mean.}{\textbackslash theta= chain mean.}}\label{theta-chain-mean.}}
+
+When \(\widehat{R}\) and \(AC\) are applied to the chain means, we notice no impact of the missingness proportions. The results only depend on the number of iterations per simulation condition.
+
+\textbf{{[}MAYBE remove all this because it has no relation to the bias in Q{]} The highest \(\widehat{R}\)-values are obtained for the condition \(T=2\).
+We observe the same `dip' in \(\widehat{R}\) as we did in Figure \ref{fig:diagnostics} for conditions with three iterations.
+\(\widehat{R}\) gradually decreases as the number of iterations increases across conditions.
+By the traditional threshold \(\widehat{R}>1.2\) we should conclude that there is non-convergence in conditions where \(T<8\).
+The common threshold \(\widehat{R}>1.1\) is overcome in conditions where \(T>14\).
+}
+
+\textbf{Autocorrelations indicate some signs of non-stationarity in all simulation conditions, but primarily in conditions where \(T<6\).
+The threshold for diagnosing non-convergence is not reached in any condition.
+We would thus conclude that the algorithm is sufficiently converged regardless of the number of iterations or missingness proportion.
+}
+
+We evaluate the diagnostics with chain means as \(\theta\) against the bias in \(Q=\mu\).
+According to the bias in the univariate means, we should conclude that two iterations is sufficient, and we should see that \(p_{mis}=0.75\) and \(p_{mis}=0.95\) have worse convergence than the other missingness proportions.
+This is not the case.
+
+\hypertarget{theta-chain-variance.}{%
+\subsubsection{\texorpdfstring{\(\theta=\) chain variance.}{\textbackslash theta= chain variance.}}\label{theta-chain-variance.}}
+
+The non-convergence diagnostics applied to the chain variances as \(\theta\) yield highly similar results to the chain means.
+We therefore evaluate these results immediately against the corresponding \(Q\), without describing the \(\widehat{R}\) and \(AC\) values.
+While the bias in the estimated standard deviations mainly depends on the missingness proportion of each condition, \(\widehat{R}\) and \(AC\) are influenced mostly by the number of iterations.
+We thus conclude that there is no added value of evaluating non-convergence with chain variances as \(\theta\) in the diagnostics.
+
+\hypertarget{thetahatq.}{%
+\subsubsection{\texorpdfstring{\(\theta=\hat{Q}\).}{\textbackslash theta=\textbackslash hat\{Q\}.}}\label{thetahatq.}}
+
+The non-convergence diagnostics are applied to the estimated regression coefficients in each imputation, and evaluated against the bias in the corresponding pooled estimate.
+
+\(\widehat{R}\) indicates signs of non-convergence across all missingness proportions and each number of iterations, in the expected order: simulation conditions with higher missingness proportions and less iterations are identified as the worst converged.
+The thresholds to diagnose non-convergence are overcome in different conditions, depending on the number of iterations and the missingness proportion.
+The most lenient threshold (\(\widehat{R}>1.2\)) is overcome at \(T\geq9\) for \(p_{mis} = 0.05\), and \(T\geq12\) for \(p_{mis} = 0.95\).
+The conventional \(\widehat{R}>1.1\) is surpassed at \(T\geq12\) and \(T\geq24\), respectively.
+
+For the autocorrelations applied to the \(\hat{Q}\)s, we see an even stronger effect of the missingness proportions.
+Especially the conditions where \(p_{mis} = 0.75\) and \(p_{mis} = 0.95\) are identified as non-stationary.
+The threshold used to evaluate whether \(AC=0\), however, is only reached in conditions with at least ten iterations.
+This suggests that the default number of iterations may be too low to detect non-stationarity with \(AC\).
+However, the critical value for diagnosing non-convergence with \(AC\) decreases as a function of the number of iterations.
+The conclusion to identify non-stationarity in conditions where \(T\geq10\) and \(p_{mis} = 0.95\) might therefore be spurious.
+\textbf{{[}Also, the \(AC\) across missingness proportions shows the opposite of what you would expect: shouldn't there be higher autocorrelations in conditions with comparatively more observed data in the completed datasets?
+I don't understand this.{]}}
+
+Compared to the bias in \(\bar{Q}\), the diagnostics applied to the \(\hat{Q}\)s over-estimate the severity of the non-convergence. This is odd, since the \(\theta\) and \(Q\) are directly related.
+
+\hypertarget{thetalambda_1.}{%
+\subsubsection{\texorpdfstring{\(\theta=\lambda_1\).}{\textbackslash theta=\textbackslash lambda\_1.}}\label{thetalambda_1.}}
+
+Finally, we evaluate the non-convergence diagnostics applied to the first eigenvalue of the variance-covariance matrices of the completed data.
+Similar to the results the the regression coefficients as \(\theta\)s, we see effects of both the missingness proportion and number of iterations in each simulation condition.
+The most signs of non-convergence are identified in conditions where \(p_{mis}\) is large and \(T\) is small.
+
+Notwithstanding the similarities with \(\theta=\hat{Q}\), there is a big difference in \(\widehat{R}\)-values for conditions where \(T<4\).
+In these conditions, \(\widehat{R}\) behaves oppositely of our expectation: conditions with the lowest missingness proportions yield the highest \(\widehat{R}\)-values.
+In conditions with a higher number of iterations, this artifact disappears.
+This suggests that the combination of \(\widehat{R}\) and \(\lambda_1\) is only appropriate for imputation algorithms where the number of iterations at least four.
+Interpreting the \(\widehat{R}\) values of conditions where \(T<3\) may lead to the incorrect conclusion that only conditions with \emph{low} missingness proportions show signs of non-convergence.
+For applications as a model-independent multivariate \(\theta\), we should therefore only interpret the \(\widehat{R}\)-values when \(T\geq4\).
+Using this cut-off we conveniently ignore the initial `dip' at \(T=3\) as well.
+The traditional threshold \(\widehat{R} > 1.2\) is overcome in conditions where \(p_{mis} = .95\) and \(T\geq10\), and for all other missingness proportions in conditions where \(T\geq8\).
+With the \(\widehat{R} > 1.1\) threshold, the number of iterations are \(T\geq20\) and \(T\geq16\), respectively.
+
+The \(AC\)-values largely depend on the missingness proportion in each condition.
+Conditions where \(p_{mis}=.95\) show statistically significant autocorrelations for any number of iterations.
+The other extreme, conditions where \(p_{mis}=.05\), does not yield \(AC\) values above the threshold.
+The \(AC\) values with this missingness proportion are equivalent for all conditions where \(T\geq6\), which suggests that the algorithm is as stationary as it can be.
+A similar point is only observed after twenty iterations in conditions with higher missingness proportions.
+
+In comparison to the \(Q\)s, the diagnostics again over-estimate the signs of non-convergence.
+Conditions with low missingness proportions (e.g., 5\% or 25\% of the complete data) yield unbiased, confidence valid estimates with as little as two iterations.
+The convergence diagnostics, however, indicate improved convergence up-to eight iterations according to \(AC\) and thirty to forty iterations according to \(\widehat{R}\).
+The model-independent \(\theta\) \(\lambda_1\) seems as well-equipped to diagnose signs of non-convergence as \(\theta=\hat{Q}\).
+The univariate \(\theta\)s do not differentiate between bias induced by the missingness proportions.
+
+\hypertarget{in-short-remove-later}{%
+\subsection{In short {[}remove later!{]}}\label{in-short-remove-later}}
 
 \begin{itemize}
+\item
+  Is there more bias in conditions where non-convergence was induced? (NOPE for univariates, YES for multivar.).
 \item
@@
 \item
-  Is one of the thetas better? (NAH, doesn't really matter, probably multivariate is better).
+  Is one of the thetas better? (YES, while it doesn't really matter as long as it's multivariate, \(\lambda_1\) is model-independent and therefore preferred).
 \item
-  What should the thresholds be? (LOWER than recommended).
+  What should the thresholds be? (much HIGHER than recommended for \(\widehat{R}\), difficult to say for \(AC\)).
 \item
-  And the default nr of iterarions? (MAYBE increase to 10?).
+  And the default nr of iterations? (MAYBE increase to 10 to detect significant \(AC\) and be sure not to miss the dip in \(\widehat{R}\)?).
 \end{itemize}
-
-For reasons of brevity, we only discuss the convergence diagnostics for the \(Q\)s with the worst performance in terms of bias.
-Bias in the descriptve statistics (\(Q=\mu\) and \(Q=\sigma\)) was most pronounced in the outcome variable \(Y\).
-Of the regression coefficients, \(Q=\beta\), the largest bias was observed in the effect of \(X_1\) on \(Y\), \(\beta_1\).
-Results for \(Q=r^2\) are reported as-is.
-In the figures, however, we only present simulation conditions where \(T\leq50\).
-Full results are available from \textbf{{[}link github repo{]}}.
-
-\hypertarget{in-general}{%
-\subsection{In general}\label{in-general}}
-
-As expected, conditions with a higher proportion of missingness and/or a lower number of iterations show more signs of non-convergence.
-I.e., on average these conditions portray more extreme bias in the estimated \(Q\)s, and non-nominal coverages.
-Roughly speaking this means that MICE is indeed not converged at \(t=1\), and converges gradually as \(t\) increases to \(T\).
-The point at which an additional iteration does not lead to an improvement of the estimates depends on the difficulty of the missingness problem---between XYZ and XYZ.
-This means that the algorithm has converged sufficiently (\textbf{under the current specifications}).
-
-As required for a method to diagnose non-convergence, the values of \(\widehat{R}\) follow a trend similar to the performance measures.
-\textbf{Add split between convergence diagnostics versus performance measures, and convergence diagnostics versus simulation conditions.} \(\widehat{R}\) values are generally lower in conditions with a higher number of iterations, and somewhat higher in conditions with a higher percentage of missing data.
-Autocorrelation values were indeed decreasing with a higher number of iterations, and lower in conditions with less missingness.
-
-Evaluation with the performance measures shows that \(\widehat{R}\) is conservative: it identifies signs of non-convergence in conditions where the \(\bar{Q}\)s are unbiased and/or confidence valid estimates of the \(Q\)s.
-
-\hypertarget{means}{%
-\subsection{Means}\label{means}}
-
-\textbf{Message: after one iteration (\(T\)=2), the algorithm is as well converged as the missingness proportion allows.
-Sooner even with lower missingness proportions (\(T\)=1 might be ok for \(p_{mis}\) = 5 and 25).} Number of it has little influence on bias in \(\hat{Q}\), only in first iteration some bias in missingness conditions of 25 and up.
-There is a clear different between the missingness conditions.
-The higher the missingness prop, the larger the magnitude of the bias.
-For this \(Q\), t=2 would be sufficient.
-And only for \(p_{mis}\) = 5 the \(\bar{Q}\)s are unbiased.
-All other \(p_{mis}\) conditions have bias irrespective of the number of iterations.
-
-\(\widehat{R}\) differs a lot across iteration conditions.
-But not structurally for missingness proportions.
-Highest \(\widehat{R}\)s for the condition \(T\)=2, while the most bias is for (\(T\)=1), but there is no \(\widehat{R}\) for that by definition.
-But for missingness, 25\% seems the most signs of non-conv, while it has one of the least bias.
-Threshold of 1.2 reached in condition it = 3, \(p_{mis}\) = 5 and 75, and then for it\textgreater6, \(p_{mis}\) = 5 and 50, and \(T\)\textgreater10 for all \(p_{mis}\).
-Threshold 1.1 reached in in \(T\)=12, \(p_{mis}\) = all but 75, 1.01 only reached in \(T\) = 93, \(p_{mis}\) = 75.
-While for this \(Q\), \(T\)=2 would be sufficient.
-
-\(AC\) mainly shows effects of the set of iteration conditions.
-\(AC\) indicates less signs of non-convergencein conditions with higher number of iterations.
-For iteration conditions where \(T\)\textless10, there are also some effects of the missingness proportions.
-The \(p_{mis}\) condtions 25 and 95 \% indicate the most signs of non-convergence.
-For \(p_{mis}\) = 95\% this corresponds to the bias in Qbar.
-For 25\% miss it doesn't.
-The thresholds defined by the critical value is only met in the conditions it=3, \(p_{mis}\) = 5,75, and 95\%.
-
-\begin{figure}
-
-{\centering \includegraphics{manuscript_files/figure-latex/mean-1}
-
-}
-
-\caption{Convergence diagnostics chain mean.}\label{fig:mean}
-\end{figure}
-
-\hypertarget{standard-deviations}{%
-\subsection{Standard deviations}\label{standard-deviations}}
-
-\textbf{Message: even with possible effect of initial values (\(T\)=1), the algorithm is as well converged as the missingness proportion allows.} For \(Q=\sigma_Y\) there is no effect visible of the iteration conditions, but clearly for the \(p_{mis}\) conditions.
-Most bias for \(p_{mis}\) = 95, followed by \(p_{mis}\) = 50 and 75.
-For the least possible bias we may use any number of iterations \(T\geq1\), but none of the \(p_{mis}\) results in unbiased \(\bar{Q}\)s.
-
-There is a clear effect of the number of iterations on \(\widehat{R}\).
-Conditions with lower Ts have more signs of non-conv, as identified by \(\widehat{R}\).
-There is little difference between missingness conditions.
-Only for conditions where \(T<5\) the missingness proportions seem to affect whether \(\widehat{R}\) indicates non-convergence The bias in \(\bar{Q}\) seems unrelated to the magnitude of \(\widehat{R}\).
-Conditions with more bias (\(p_{mis}\) = 50,75,95) are not identified to have more signs of non-convergence.
-The thresholds are therefore useless (because depend mostly on \(T\), not so much \(p_{mis}\)).
-
-\(AC\) is affected by both t and \(p_{mis}\) conditions.
-But conditions with worst bias (\(p_{mis}\) = 50, 75, 95) are not identified as worst converging according to AC.
-Moreover, only for \(T\) =3 the \(AC\) is significantly different from zero.
-
-\begin{figure}
-
-{\centering \includegraphics{manuscript_files/figure-latex/sd-1}
-
-}
-
-\caption{Convergence diagnostics chain variance.}\label{fig:sd}
-\end{figure}
-
-\hypertarget{regression-coefficients}{%
-\subsection{Regression coefficients}\label{regression-coefficients}}
-
-\textbf{Message: after four iterations (\(T\)=5), the algorithm is as well converged as the missingness proportion allows.
-Sooner even with lower missingness proportions (\(T\)=1 might be ok for \(p_{mis}\) = 5 and 25).} The figure for \(Q=\beta_1\) shows that the \(\bar{Q}\)s are affected by both the number of iterations and the missingness proportion.
-For \(T\) =1, the magnitude of the bias (almost) follows the order of the missingness proportions.
-More missingness causes more bias in the \(\bar{Q}\).
-In general, a lower nr of iterations results in more bias, but this depends on the missingness proportion.
-For \(p_{mis}\) = 95, there is more bias in conditions where \(T\)\textless6.
-When \(p_{mis}\) = 50, it's \(T\)\textless3.
-For \(p_{mis}\) = 75 and 25, it's only for \(T\)\textless2.
-And conditions where \(p_{mis}\) = 5 are unaffected on average.
-However, none of the conditions result in completely unbiased estimates of \(Q\).
-
-We should see this in the conv diag: higher \(\widehat{R}\) and \(AC\) for \(p_{mis}\) = 75 and 95, and for condtions where \(T<2-6\), depending on \(p_{mis}\).
-
-\(\widehat{R}\) depends somewhat on the missingness proportion.
-For conditions where \(T\)\textless10, the \(\widehat{R}\) values follow the \(p_{mis}\) more or less.
-But what we actually want to know is whether they are indicators of bias.
-Conditions with the worst bias (high \(p_{mis}\), low T) have the largest \(\widehat{R}\) values.
-The thresholds are not appropriate to diagnose sufficient convergence.
-All thresholds over-estimate the signs of non-convergence compared to the bias in Qbar.
-However, a higher threshold than 1.2 may result in terminating the algorithm after 3 iterations, which is insufficient for the \(p_{mis}\) = 95 condition.
-Threshold of 1.25 or 1.3 could work for the other \(p_{mis}\) conditions.
-
-\(AC\) is very affected by the p\_mis.
-The two conditions with the most bias (\(p_{mis}\) = 75 and 95) are the only ones identified as non-converging, as they should.
-However, the \(p_{mis}\) = 75 condition is only identified in the \(T\) =3 condtion, while the bias is the least pronounced there.
-And the \(AC\) only identifies non-convergence in the \(p_{mis}\) = 95 conditions where \(T\)\textgreater10.
-In the default situation, \(T\) =5, this would be overlooked.
-Also, the \(AC\) shows the opposite of what you would expect from a theta that is based on the observed data: shouldn't there be higher autocorrelations in conditions with comparatively more observed data in the completed datasets?
-I don't understand this.
-
-\textbf{Add how interesting it is because we literally track \(Q\) and are still not ok.}
-
-\begin{figure}
-
-{\centering \includegraphics{manuscript_files/figure-latex/est-1}
-
-}
-
-\caption{Convergence diagnostics regression coefficient.}\label{fig:est}
-\end{figure}
-
-\hypertarget{variance-explained}{%
-\subsection{Variance explained}\label{variance-explained}}
-
-\textbf{Message: after two iterations (\(T\)=3), the algorithm is as well converged as the missingness proportion allows.
-Sooner even with lower missingness proportions (\(T\)=1 might be ok for \(p_{mis}\) = 5 and 25).} Bias is the worst for conditions where \(p_{mis}\) = 95, \(T\) \textless3.
-Followed by \(p_{mis}\) = 75, \(T\) =1; and \(p_{mis}\) = 50, \(T\) = 1.
-Conditions where \(p_{mis}\) = 25 or 5 seem unaffected by the number of iterations.
-Minimal bias is obtained in all iteration conditions.
-We should see this in \(\widehat{R}\) and AC: worst convergence in \(p_{mis}\) = 95 and 50; some non-convergencein conditions where \(p_{mis}\) = 75 and \(T\) =1, and no non-convergencein conditions where \(p_{mis}\) = 25 or 5.
-The sign of the bias is logical: when there is less info in the data, the relations between the predictors and the outcome is under-estimated.
-
-\(\widehat{R}\) behaves oppositely to what we wanted.
-The lowest \(\widehat{R}\) values for conditions where \(T\)\textless10 are observed for the two missingness conditions with the worst bias.
-However, the highest \(\widehat{R}\) observed is indeed for the missingness condition with the worst bias, \(p_{mis}\) =95.
-But only at \(T\) =5 (while bias is no worse than anywhere else in this \(p_{mis}\) condition, conditional on iterations condition \(T\)\textgreater2).
-So neither of the thresholds works perfectly.
-
-\(AC\) shows a similar trend to the biases: more signs of non-convergencefor higher \(p_{mis}\) and lower T.
-The threshold, however, performs less well.
-It identifies signs of non-convergence for all conditions where \(T=3\), irrespective of p\_mis.
-Then nothing, and then for \(p_{mis}\) =95, \(T\)\textgreater7, and \(p_{mis}\) = 50, \(T\) \textgreater14, and \(p_{mis}\) = 75, \(T\)\textgreater20.
-This does not corresond to the bias in \(\bar{Q}\).
-Maybe use AC\textgreater.5 at \(T\) =5 as threshold?
-
-\begin{figure}
-
-{\centering \includegraphics{manuscript_files/figure-latex/pred-1}
-
-}
-
-\caption{Convergence diagnostics regression coefficient.}\label{fig:pred}
-\end{figure}
 
@@
 
-In short, univariate estimates seem robust against terminating the algorithm early: There is no clear effect of the number of iterations on the bias in the estimated means and standard deviations.
-The bias in these \(Q\)s only depends on the \(p_{mis}\), with higher \(p_{mis}\) resulting in more bias.
-Yet, the convergence diagnostics indicate that the algorithm did not reach a completely converged state (yet).
+In short, iterative imputation algorithms may yield approximately unbiased, confidence valid estimates under different levels of induced non-convergence.
+Univariate estimates seem robust against terminating the algorithm early: There is no clear effect of the number of iterations on the bias in estimated means and standard deviations.
+The bias in these \(Q\)s was only impacted by the missingness proporion \(p_{mis}\), with higher \(p_{mis}\) resulting in more bias.
+Bias in multivariate estimates depended on both \(p_{mis}\) and the number of iterations.
+However, the only effect of the number of iterations was observed when the algorithm was terminated at seven iterations or earlier.
+This suggests that continued iteration beyond seven iterations does not yield better estimates.
+Yet, the non-convergence diagnostics \(\widehat{R}\) and \(AC\) indicated that the iterative imputation algorithm did not reach a stable state before twenty to thirty iterations.
+\textbf{{[}MERGE or remove{]} \(\widehat{R}\) and autocorrelation indicate that there are signs of non-convergence in the algorithm up-to at least twenty iterations, while unbiased, confidence valid estimates may be obtained with as little as one iteration.
+This observation is in agreement with }one of {[}still add the others!{]}** the simulation hypotheses: \(\widehat{R}\) over-estimates the severity of non-convergence when applied to MI procedures.**
 
-\(\widehat{R}\) and autocorrelation indicate that algorithmic convergence may only be reached after twenty or even forty iterations, while unbiased, confidence valid estimates may be obtained with as little as one iteration.
-These results are in agreement with the simulation hypothesis: \(\widehat{R}\) over-estimates the severity of non-convergence when applied to MI procedures.
-
-With that, we show that MICE can lead to correct outcomes when they have not converged according to two common conv diags.
-This may be due to the methods (and their thresholds) or due to the \(Q\)s (descriptives and multivariate linear regression, not higher dimensional/more complex \(Q\)s).
-More `complicated' \(Q\)s (e.g., higher-order effects or variance components) might show bias, under- or over-coverage at higher \(T\), as indiated by \(\widehat{R}\) and AC.
-Add what \%miss has to do with it.
+Over-estimation of the signs of non-convergence may be due to the methods (and their thresholds) or due to the \(Q\)s (descriptive and multivariate linear regression, not higher dimensional/more complex \(Q\)s).
+More complicated \(Q\)s (e.g., higher-order effects or variance components) might show bias, under- or over-coverage at higher \(T\).
+Which might be why \(\widehat{R}\) and \(AC\) identify non-convergence.
 
@@
 \item
-  Check trace plots for pathological non convergence and adjust imputation model if necessary.
+  Check traceplots for signs of pathological non-convergence. Adjust imputation model if necessary.
 \item
-  use \(\widehat{R}\) wait 1.1 ash threshold and autocorrelation with \textbf{?} as threshold. Keep iterating until these thresholds are reached.
+  Monitor non-convergence diagnostics \(\widehat{R}\) and \(AC\). Use the thresholds for \(\widehat{R}>1.2\) and autocorrelation \textbf{???}. Keep iterating until these thresholds are reached.
 \item
-  Do not use the R function ACF. Instead, compute autocorrelations manually \textbf{(see e.g., {[}GitHub link{]})}.
+  Do not use the traditional calculation for \(\widehat{R}\), or the \texttt{R} function \texttt{stats::acf()} to compute \(AC\).
+Instead, calculate \(\widehat{R}\) conform \citet{veht19} and compute autocorrelations manually (see e.g., \href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy}{github.com/gerkovink/shinyMice}).
 \item
-  Track your own scalar summary of interest. This is somewhat advanced but explained in Van Buren 2018. Compute \(\widehat{R}\) and autocorrelation values for this scalar summary.
+  Track your own scalar summary of interest (see e.g., \citet{buur18}). Compute \(\widehat{R}\) and autocorrelation values for this scalar summary.
 \item
@@
 
-Further research is needed to investigate their performance under clear violation of convergence, e.g.~dependency between predictors (predictors with very high correlations). Also:
+Further research is needed to investigate the performance of iterative imputation algorithms under clear violation of convergence, e.g.~dependency between predictors (predictors with very high correlations).
+For example:
 
@@
 \item
-  More difficult missingness problems, i.e., M(N)AR instead of MCAR.
-Proper performance under a `missing completely at random' missingness mechanism is necessary to demonstrate the appropriateness of \(\widehat{R}\) and \(AC\) as non-convergence diagnostics.
+  Introduce more difficult missingness problems, i.e., M(N)AR instead of MCAR.
+That is, proper performance under a `missing completely at random' missingness mechanism is necessary to demonstrate the appropriateness of \(\widehat{R}\) and \(AC\) as non-convergence diagnostics.
 However, results may not be extrapolated to other missingness mechanisms.
@@
 
-Uitsmijter: To conclude, in the case of drawing inference from incomplete data, convergence of the iterative imputation algorithm is often a convenience but not a necessity.
-
-\hypertarget{grading}{%
-\section{Grading}\label{grading}}
-
-\begin{itemize}
-\item
-  Scientific contribution
-  (clear research question, convincing argumentation that the study adds to the scientific literature)
-\item
-  Theory
-  (use of theory, clear hypotheses, relevant literature)
-\item
-  Method
-  (correct data collection/handling, measures, choice and description of method of analysis)
-\item
-  Results
-  (correct interpretation of findings, clear presentation)
-\item
-  Discussion and conclusions
-  (summary, broader implications, limitations, future research)
-\item
-  Written presentation
-  (writing, structure, tables and figures, references)
-\end{itemize}
+In short, we have shown that an iterative imputation algorithm can yield correct outcomes, even when a converged state is not reached according to two common non-convergence diagnostics.
+Continued iteration just burns a lot of computing time, without improving the statistical inferences.
+To conclude, in the case of drawing inference from incomplete data, convergence of the iterative imputation algorithm is often a convenience but not a necessity.
 

### 2020-05-05 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
Add title page for hand in

@@
   bookdown::pdf_book:
-    # includes:
-    #   in_header: preamble.tex
+    base_format: rticles::sage_article
     keep_tex: yes
-    base_format: rticles::sage_article
 ---

### 2020-05-05 3.Thesis/1.SimulationStudy/Writeup/1.CombineTitlePageWithThesis.R

@@ -1,0 +1,3 @@
+library(pdftools)
+pdf_combine(c("preamble.pdf", "manuscript.pdf"), 
+            output = "Oberman,H.I.-MSBBSS.pdf")

### 2020-05-06 3.Thesis/Writeup/2.Manuscript.Rmd
Re-structure repo to make clear how Figures are created.

--- 2020-05-05 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
@@
 # packages
+library(mice)
 library(dplyr)
@@
 
-# results and functions
+# load example non-convergence and figures
+load("../Results/example_nonconv.Rdata")
+load("../Results/example_conv.Rdata")
+load("../Results/example_diagnostics.Rdata")
+source("../3.PlotExampleNon-convergence.R")
+
+# load results and figures
 load("../Results/complete.Rdata") 
-results <- results %>% filter(t<50)
-source("../2.CreateFigures.R")
+results <- results %>% filter(t<51)
+source("../4.PlotResults.R")
 
@@
  legend.title = element_text(size = 7)
-
- # axis.title = element_text(size = 8),
- # axis.text = element_text(size = 7),
- # legend.text = element_text(size = 7)
 )
@@
 ```{r diagram, out.width="\\linewidth", include=TRUE, fig.cap="Scheme of the main steps in multiple imputation (where $M=3$)---from an incomplete dataset, to $M$ multiply imputed datasets, to $M$ estimated quantities of scientific interest $\\hat{Q}$s, to a single pooled estimate $\\bar{Q}$."}
-knitr::include_graphics("./images/diagram.pdf") #adapted from [@buur18] par. 1.4.1.
+knitr::include_graphics("../Figures/diagram.pdf") #adapted from [@buur18] par. 1.4.1.
 ```
@@
 Please note that this is the same data with a different imputation model, leading to different imputations (e.g., see the units on the y-axis)."}
-load("./images/nonc_plot.Rdata")
-load("./images/conv_plot.Rdata")
 par(mar = c(2, 2, 0.1, 0.1))
-plot(conv)
-plot(nonc)
+con_plot
+non_plot
 ```
@@
 ```{r diagnostics, fig.cap="Convergence diagnostics applied on the imputation algorithms of Figure 2. $\\theta$ = chain mean in $y_{imp, m}$."}
-load("./images/diagnostics_plot.Rdata") #bquote(theta ~ "= chain mean in "~ y[imp][",m"])
-diagnostics_plot 
+# load("./images/diagnostics_plot.Rdata") #bquote(theta ~ "= chain mean in "~ y[imp][",m"])
+(theta + plot_layout(guides = "collect", ncol = 2)) + (ac_both + plot_layout(guides = "keep")) + old_rhat + new_rhat +  plot_annotation(tag_levels = "A", tag_suffix = ".") # add 'ncol=1' for plots under each other
 ```
@@
 
-Convergence of the MICE algorithm is investigated through model-based simulation in `R` [version `r paste(version$major,version$minor, sep = ".")`; @R].
+Convergence of the MICE algorithm is investigated through model-based simulation in `R` [version 3.6.3; @R].
 The simulation set-up is summarized in the pseudo-code below.

### 2020-05-06 3.Thesis/Writeup/2.Manuscript.tex
Re-structure repo to make clear how Figures are created.

--- 2020-05-04 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
@@
 
-{\centering \includegraphics[width=\linewidth]{./images/diagram}
+{\centering \includegraphics[width=\linewidth]{../Figures/diagram}
 
@@
 
-{\centering \includegraphics[width=.49\linewidth]{manuscript_files/figure-latex/non-conv-1} \includegraphics[width=.49\linewidth]{manuscript_files/figure-latex/non-conv-2}
+{\centering \includegraphics[width=.49\linewidth]{2.Manuscript_files/figure-latex/non-conv-1} \includegraphics[width=.49\linewidth]{2.Manuscript_files/figure-latex/non-conv-2}
 
@@
 
-{\centering \includegraphics{manuscript_files/figure-latex/diagnostics-1}
+{\centering \includegraphics{2.Manuscript_files/figure-latex/diagnostics-1}
 
@@
 
-{\centering \includegraphics{manuscript_files/figure-latex/Qs-1}
+{\centering \includegraphics{2.Manuscript_files/figure-latex/Qs-1}
 
@@
 
-{\centering \includegraphics{manuscript_files/figure-latex/rhats-acs-1}
+{\centering \includegraphics{2.Manuscript_files/figure-latex/rhats-acs-1}
 

### 2020-05-06 3.Thesis/1.SimulationStudy/Writeup/images/non-convergence.R

@@ -1,234 +1,0 @@
-# create figure to display pathological non-convergence
-# requires the packages 'mice', 'dplyr', 'ggplot2'
-
-# load packages and functions
-library(mice)
-library(purrr)
-library(dplyr)
-library(ggplot2)
-library(patchwork)
-source('./3.Thesis/1.SimulationStudy/Functions/Convergence.R')
-source('./3.Thesis/1.SimulationStudy/Functions/AC.R')
-source('./3.Thesis/1.SimulationStudy/Functions/AC_supplement.R')
-source('./3.Thesis/1.SimulationStudy/Functions/Rhat.R')
-source('./3.Thesis/1.SimulationStudy/Functions/Rhat_supplement.R')
-
-#####################
-# Pathological non-c.
-#####################
-meth <- make.method(boys)
-meth["bmi"] <- "~I(wgt / (hgt / 100)^2)"
-imp.bmi1 <- mice(
-  boys,
-  meth = meth,
-  maxit = 10,
-  print = FALSE,
-  seed = 60109
-)
-
-# plot chain means of wgt (rows are iterations, columns are imputations)
-nonc <- plot(imp.bmi1, "wgt")[1]
-nonc$layout <- c(1, 1)
-nonc[["condlevels"]][[".ms"]] <- c("", "sd")
-nonc$ylab <- "Chain mean"
-nonc$main <- list(label = "Non-convergence", fontsize=12, fontface = 1)
-nonc$x.scales$tick.number <-10
-save(nonc, file = "3.Thesis/1.SimulationStudy/Writeup/images/nonc_plot.Rdata")
-
-# diagnostics on chain means of wgt (rows are iterations, columns are imputations)
-diagnostics1 <-
-  imp.bmi1$chainMean[3, ,] %>% cbind(patho = "Non-convergence", convergence(., include_acf = TRUE), .) %>% mutate(signif = c(NA, NA, 1, qnorm((1 + .95) / 2) / sqrt(iteration[-(1:3)])))
-
-
-#####################
-# Regular convergence
-#####################
-pred <- make.predictorMatrix(boys)
-pred[c("hgt", "wgt"), "bmi"] <- 0
-imp.bmi2 <- mice(
-  boys,
-  meth = meth,
-  pred = pred,
-  maxit = 10,
-  print = FALSE,
-  seed = 60109
-)
-
-# plot chain means of wgt (rows are iterations, columns are imputations)
-conv <- plot(imp.bmi2, "wgt")[1]
-conv$layout <- c(1, 1)
[…]

### 2020-05-06 3.Thesis/Writeup/2.Manuscript.Rmd
Update results about Qs, not yet about thetas

@@
 Merge or remove one of the two?]** Typically, the $\theta$s under evaluation are chain means and variances (**define!**).
-These $\theta$s may be insufficient because they are univariate summaries of the state space of the algorithm, while MICE is not only concerned with univariate estimates, but the entire multivariate distribution of $y_{imp}$.
+These $\theta$s may be insufficient because they are summaries of the univariate state space of the algorithm, while MICE is not only concerned with univariate estimates, but the entire multivariate distribution of $y_{imp}$.
 A suggestion by @buur18 for a multivariate $\theta$ to monitor is the quantity of scientific interest $Q$ (e.g., a regression coefficient).
@@
 In MCMC methods, convergence is not from a scalar to a point, but from one distribution to another.
-The values generated by the algorithm (e.g., imputed values) will vary even after convergence.
+The values generated by the algorithm (e.g., imputed values) will vary even after convergence [@gelm13].
 Since MCMC algorithms do not reach a unique point at which convergence is established, diagnostics  may only identify signs of *non*-convergence [@hoff09].
@@
 Iterative imputation algorithms that have not yet reached stationarity, may thus yield biased estimates.
+
+<!-- When we declare approximate convergence, we are actually concluding that each individual sequence appears stationary and that the observed sequences -->
+<!-- have mixed well with each other. @gelm13, p. 288-->
 
@@
 
+**[Add: effective sample size is used if we want independent samples *within* chains.
+In MI, we only use the final estimate.
+Also, computing n_eff assumes infinitely many iterations, which is fine in Bayesian analyses where T is often at least one thousand.
+But the default in iterative imp is often substantially lower.]**
+
 ### Potential scale reduction factor
@@
 **This version is better suited to detect non-convergence in the tails of distributions, by using three transformations on the scalar summary that it is applied to.
-Namely, folded, rank-normalized, split-chain $\theta$s.
+Namely, rank-normalization, folding, and localization.
 The Vehtari et al.
@@
 As convergence diagnostic, the interest is therefore in positive $AC$-values.
-The magnitude of $AC$-values may be evaluated statistically [**add reference**].
+The magnitude of $AC$-values may be evaluated statistically, using a Wald test with $AC=0$ as null hypothesis [@box15].
 
@@
 
+<!-- We assess discrepancy from convergence via practical significance (or some conventional version thereof, such as Rh > 1.1) @gelm13, p. 288 -->
+
+<!-- We can use effective sample size ^neff to give us a sense of the precision obtained from our simulations.
+As we have discussed in Section 10.5, for many purposes it should suffice to have 100 or even 10 independent simulation draws.
+(If neff = 10, the simulation standard error is increased by p 1 + 1/10 = 1.05).
+As a default rule, we suggest running the simulation until ^neff is at least 5m, that is, until there are the equivalent of at least 10 independent draws per sequence (recall that m is twice the number of sequences, as we have split each sequence into two parts so that Rb can assess stationarity as well as mixing).
+Having an effective sample size of 10 per sequence should typically correspond to stability of all the simulated sequences.
+@gelm13, p.
+287 -->
+
 ## In practice
 
-Before we evaluate the performance of the non-convergence diagnostics $\widehat{R}=1$ and $AC=0$ quantitatively through simulation, we apply them to the example of pathological non-convergence by @buur18 that we reproduced above.
+Before we evaluate the performance of the non-convergence diagnostics $\widehat{R}$ and $AC$ quantitatively through simulation, we apply them to the example of pathological non-convergence by @buur18 that we reproduced above.
 We want to be able to draw the same conclusions as we did from visually inspecting the two algorithms.
-But $\widehat{R}=1$ and $AC=0$ should at least be able to distinguish between the two algorithms plotted in Figure \ref{fig:non-conv}.
+But $\widehat{R}$ and $AC$ should at least be able to distinguish between the two algorithms plotted in Figure \ref{fig:non-conv}.
 From visual inspection, we know that the typically converged algorithm initially portrayed some signs of non-mixing (around $t=2$), but intermingled nicely overall.
@@
 <!-- However, the superior performance of the typical convergence over the pathological non-convergence is less prominent, even flipped for $t<4$.  -->
+
+**In conclusion, we will use $\widehat{R}$ conform @veht19 and $AC$ computed manually as non-convergence diagnostics in the simulation study.
+The diagnostics will be applied to four $\theta$s of interest: chain means, chain variances, a quantity of scientific interest, and the first eigenvalue of the variance-covariance matrix.**
 
@@
 Yet, scientific estimates were insensitive to continued iteration from $t=5$ onward [@lace07].
-We therefore suspect that $\widehat{R}$ may over-estimate signs of non-convergence in iterative imputation algorithms, compared to the bias the estimates.
+We therefore suspect that $\widehat{R}$ may over-estimate signs of non-convergence in iterative imputation algorithms, compared to the validity of estimates.
 In contrast to this, it may occur that signs of non-convergence are under-estimated by $\widehat{R}$, in exceptional cases where the initial values of the algorithm are not appropriately over-dispersed [@broo98, p.
@@
 
-Convergence of the MICE algorithm is investigated through model-based simulation in `R` [version 3.6.3; @R].
+We investigate non-convergence of the MICE algorithm through model-based simulation in `R` [version 3.6.3; @R].
+The number of simulation repetitions is 1000.
 The simulation set-up is summarized in the pseudo-code below.
@@
 There is only one estimate for $Q=r^2$ to consider.
-In the figures, we present simulation conditions where the number of iterations is $1 \leq T\leq50$, since the results are more or less stable for conditions where $T \leq 30$.
+In the figures, we present simulation conditions where the number of iterations is $1 \leq T\leq50$, since the results are more or less stable for conditions where $T \geq 30$.
 Full results are available from [github.com/gerkovink/shinyMice](https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy).
@@
 ```{r Qs, fig.height = 6.8, fig.cap="Impact of non-convergence on statistical inferences."}
-mean_bias + sd_bias + est_bias + est_ciw + est_cov + Rsq_bias + plot_layout(ncol = 2, guides = "collect")
+mean_bias + sd_bias + Rsq_bias + est_bias + est_cov + est_ciw + plot_layout(ncol = 2, guides = "collect")
 ```
@@
 Figure \ref{fig:Qs} displays the influence of inducing non-convergence on the estimated quantities of scientific interest, $Q$s.
-**[Add panel labels, describe panels]** As expected, conditions with a higher proportion of missingness and/or a lower number of iterations are impacted more severely by non-convergence.
+**[Add panel labels, describe panels]** Overall, conditions with a higher proportion of missingness and/or a lower number of iterations are impacted more severely by non-convergence.
 I.e., on average these conditions portray more extreme bias in the estimated $Q$s, and non-nominal coverages.
@@
 The highest number of iterations that is necessary to reach this point across all simulation conditions is $T=7$.
-This implies that, under the current specifications, no more than seven iterations are necessary to obtain optimal**/the best possible** performance in terms of bias and confidence validity.
+This implies that, under the current specifications, no more than seven iterations are necessary to obtain *sufficient* performance in terms of bias and confidence validity.
 
@@
 The estimated univariate means seem unaffected by the number of iterations in the algorithm.
-The magnitude of the bias in $\bar{Q}$ depends exclusively on the missingness proportion $p_{mis}$, with higher missingness proportions causing more extreme bias.
-Note however, that the simulation condition with the largest overall bias resulted in a $\bar{Q}$ under-estimated the true value of $Q$ by only 0.02 units (where $Q=\mu_Y=25.81; \sigma_Y = 11.32$).
-We conclude that---conditional on the missingness proportion---all simulation conditions ($T \geq 1$) have equal performance.
-According to this scientific estimand $Q$, non-convergence does not impact the validity of the inferences.
+This implies that *suffient* inferences may be obtained with as little as one iteration ($T \geq 1$).
+The magnitude of the bias in $\bar{Q}$ depends exclusively on the missingness proportion $p_{mis}$, with missingness proportions of 75% and 95% causing more extreme bias than other conditions.
+Unbiased estimates are obtained in all conditions where $p_{mis}\leq.50$.
+Note however, that even in the conditions where $p_{mis}\geq.75$, the bias is relatively small: $\mu_Y$ is maximally under-estimated by 0.02 units, while the true value of $\mu_Y$ is 25.81 ($\sigma_Y = 11.32$).
+We conclude that non-convergence does not substantially impact the validity of the inferences for $Q=\mu$.
 **[Add: up-to 50% missingness everything is relatively unbiased, while 75% and 95% missingness show very different results.
@@
 ### $Q=\sigma$.
-The estimated standard deviations show impact of non-convergence similar to the means.
+The estimated standard deviations are largely affected by the missingness proportions, but not to much by the number of iterations.
+*Sufficient* inferences are obtained in conditions with just two iterations.
+Interestingly, the bias in first iteration is actually *less*  extreme than for any other number of iterations.
+We may therefore conclude that no iteration at all is needed to obtain *sufficient* estimates for $Q=\sigma$.
+Completely unbiased estimates are only obtained in conditions where $p_{mis}=.05$.
+Conditions where $p_{mis}\geq.25$ are impacted by non-convergence in order of increasing missingness proportion.
+<!-- in the first iteration.
+how impact of non-convergence similar to the means.
 The magnitude of the bias $\bar{Q}$ follows the incremental order of the missingness proportions.
@@
 For the least possible bias, we may use any number of iterations $T\geq1$, but only $p_{mis} = 0.05$ results in completely unbiased $\bar{Q}$s.
+-->
+
+### $Q=r^2$.
+Bias in the estimated variance explained depends both on missingness proportions and the number of iterations.
+The magnitude of the bias clearly follows the incremental order of the missingness proportions.
+We, therefore, observe *sufficient* estimates at a different number of iterations for each missingness proportion.
+For $p_{mis}\leq.25$, the minimum number of iterations is two ($T\geq2$), while $p_{mis}=.50$ requires $T\geq4$, and $p_{mis}=.75$ requires $T\geq5$.
+The highest number of iterations needed to obtain *sufficient* estimates is seven ($p_{mis}=.95$; $T\geq7$).
+Unbiased estimates, however, may only be obtained if the missingness proportion is at most 25%.
+**[interpret sign of bias]**
+
+<!-- while the best performance occurs with less missingness and more iterations.
+Only conditions with the lowest missingness proportion ($p_{mis}=0.05$) results in unbiased estimates for every number of iterations.
+In other conditions there is at least some bias irrespective of $T$.
+The inferences do not improve in any condition where $T \geq 6$.
+This suggests that the algorithm is as converged as necessary to obtain valid inferences.
+-->
 
@@
 For the estimated regression coefficients, we consider bias, confidence interval width (CIW), and coverage rate (CR) as performance measures.
-The magnitude of the bias, again, depends on the missingness proportion.
-But for this $Q$ we also observe a clear effect of the number of iterations.
+Unbiased estimates are obtained in conditions where $p_{mis}\leq.50$ and $T\geq3$.
+Approximately confidence valid estimates are obtained in any condition where $T\geq3$.
+
 Conditions with a lower number of iterations yield more bias in $\bar{Q}$.
@@
 
-### $Q=r^2$.
-The bias in the estimated explained variance perfectly follows both sets of simulation conditions: The worst performance is observed for the highest missingness proportion and the lowest number of iterations, while the best performance occurs with less missingness and more iterations.
-Only conditions with the lowest missingness proportion ($p_{mis}=0.05$) results in unbiased estimates for every number of iterations.
-In other conditions there is at least some bias irrespective of $T$.
-The inferences do not improve in any condition where $T \geq 6$.
-This suggests that the algorithm is as converged as necessary to obtain valid inferences.
-
 <!-- The sign of the bias is logical: when there is less info in the data, the relations between the predictors and the outcome is under-estimated. -->
@@
 In Figure \ref{fig:rhats-acs} we see the results of the non-convergence diagnostics $\widehat{R}$ and $AC$.
-**[Add panel labels and explain how to read the plots]** As expected, the diagnostics indicate more signs of non-convergence in simulation conditions with a lower number of iterations.
+**[Add panel labels and explain how to read the plots]** As expected, the diagnostics identify more signs of non-convergence in simulation conditions with a lower number of iterations.
 The missingness proportions, however, only seem to impact multivariate convergence.

### 2020-05-06 3.Thesis/Writeup/2.Manuscript.tex
Update results about Qs, not yet about thetas

@@
 Merge or remove one of the two?{]}} Typically, the \(\theta\)s under evaluation are chain means and variances (\textbf{define!}).
-These \(\theta\)s may be insufficient because they are univariate summaries of the state space of the algorithm, while MICE is not only concerned with univariate estimates, but the entire multivariate distribution of \(y_{imp}\).
+These \(\theta\)s may be insufficient because they are summaries of the univariate state space of the algorithm, while MICE is not only concerned with univariate estimates, but the entire multivariate distribution of \(y_{imp}\).
 A suggestion by \citet{buur18} for a multivariate \(\theta\) to monitor is the quantity of scientific interest \(Q\) (e.g., a regression coefficient).
@@
 In MCMC methods, convergence is not from a scalar to a point, but from one distribution to another.
-The values generated by the algorithm (e.g., imputed values) will vary even after convergence.
+The values generated by the algorithm (e.g., imputed values) will vary even after convergence \citep{gelm13}.
 Since MCMC algorithms do not reach a unique point at which convergence is established, diagnostics may only identify signs of \emph{non}-convergence \citep{hoff09}.
@@
 
+\textbf{{[}Add: effective sample size is used if we want independent samples \emph{within} chains.
+In MI, we only use the final estimate.
+Also, computing n\_eff assumes infinitely many iterations, which is fine in Bayesian analyses where T is often at least one thousand.
+But the default in iterative imp is often substantially lower.{]}}
+
 \hypertarget{potential-scale-reduction-factor}{%
@@
 \textbf{This version is better suited to detect non-convergence in the tails of distributions, by using three transformations on the scalar summary that it is applied to.
-Namely, folded, rank-normalized, split-chain \(\theta\)s.
+Namely, rank-normalization, folding, and localization.
 The Vehtari et al.~version of \(\widehat{R}\) may be suitable for iterative imputation.
@@
 As convergence diagnostic, the interest is therefore in positive \(AC\)-values.
-The magnitude of \(AC\)-values may be evaluated statistically {[}\textbf{add reference}{]}.
+The magnitude of \(AC\)-values may be evaluated statistically, using a Wald test with \(AC=0\) as null hypothesis \citep{box15}.
 
@@
 
-Before we evaluate the performance of the non-convergence diagnostics \(\widehat{R}=1\) and \(AC=0\) quantitatively through simulation, we apply them to the example of pathological non-convergence by \citet{buur18} that we reproduced above.
+Before we evaluate the performance of the non-convergence diagnostics \(\widehat{R}\) and \(AC\) quantitatively through simulation, we apply them to the example of pathological non-convergence by \citet{buur18} that we reproduced above.
 We want to be able to draw the same conclusions as we did from visually inspecting the two algorithms.
-But \(\widehat{R}=1\) and \(AC=0\) should at least be able to distinguish between the two algorithms plotted in Figure \ref{fig:non-conv}.
+But \(\widehat{R}\) and \(AC\) should at least be able to distinguish between the two algorithms plotted in Figure \ref{fig:non-conv}.
 From visual inspection, we know that the typically converged algorithm initially portrayed some signs of non-mixing (around \(t=2\)), but intermingled nicely overall.
@@
 
+\textbf{In conclusion, we will use \(\widehat{R}\) conform \citet{veht19} and \(AC\) computed manually as non-convergence diagnostics in the simulation study.
+The diagnostics will be applied to four \(\theta\)s of interest: chain means, chain variances, a quantity of scientific interest, and the first eigenvalue of the variance-covariance matrix.}
+
 \hypertarget{simulation-study}{%
@@
 Yet, scientific estimates were insensitive to continued iteration from \(t=5\) onward \citep{lace07}.
-We therefore suspect that \(\widehat{R}\) may over-estimate signs of non-convergence in iterative imputation algorithms, compared to the bias the estimates.
+We therefore suspect that \(\widehat{R}\) may over-estimate signs of non-convergence in iterative imputation algorithms, compared to the validity of estimates.
 In contrast to this, it may occur that signs of non-convergence are under-estimated by \(\widehat{R}\), in exceptional cases where the initial values of the algorithm are not appropriately over-dispersed \citep[p.~437]{broo98}.
@@
 
-Convergence of the MICE algorithm is investigated through model-based simulation in \texttt{R} \citep[version 3.6.3;][]{R}.
+We investigate non-convergence of the MICE algorithm through model-based simulation in \texttt{R} \citep[version 3.6.3;][]{R}.
+The number of simulation repetitions is 1000.
 The simulation set-up is summarized in the pseudo-code below.
@@
 There is only one estimate for \(Q=r^2\) to consider.
-In the figures, we present simulation conditions where the number of iterations is \(1 \leq T\leq50\), since the results are more or less stable for conditions where \(T \leq 30\).
+In the figures, we present simulation conditions where the number of iterations is \(1 \leq T\leq50\), since the results are more or less stable for conditions where \(T \geq 30\).
 Full results are available from \href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy}{github.com/gerkovink/shinyMice}.
@@
 The highest number of iterations that is necessary to reach this point across all simulation conditions is \(T=7\).
-This implies that, under the current specifications, no more than seven iterations are necessary to obtain optimal\textbf{/the best possible} performance in terms of bias and confidence validity.
+This implies that, under the current specifications, no more than seven iterations are necessary to obtain \emph{sufficient} performance in terms of bias and confidence validity.
 
@@
 In Figure \ref{fig:rhats-acs} we see the results of the non-convergence diagnostics \(\widehat{R}\) and \(AC\).
-\textbf{{[}Add panel labels and explain how to read the plots{]}} As expected, the diagnostics indicate more signs of non-convergence in simulation conditions with a lower number of iterations.
+\textbf{{[}Add panel labels and explain how to read the plots{]}} As expected, the diagnostics identify more signs of non-convergence in simulation conditions with a lower number of iterations.
 The missingness proportions, however, only seem to impact multivariate convergence.

### 2020-05-06 3.Thesis/Writeup/2.Manuscript.Rmd
Reread Results until thetas

@@
 Unbiased estimates are obtained in conditions where $p_{mis}\leq.50$ and $T\geq3$.
-Approximately confidence valid estimates are obtained in any condition where $T\geq3$.
+Approximately confidence valid estimates are obtained in any condition where $T\geq3$---despite of persistant bias in conditions where $p_{mis}\geq.75$.
+These approximately nominal coverages are explained by the CIWs: conditions with higher missingness proportions result in more uncertainty in the estimates (i.e., wider CIs), which diminish the effect of the bias.
+*Sufficient* inferences in terms of confidence validity may be obtained when $T\geq3$, while we need at least seven iterations to reach this point in terms of bias.
+Continued iteration after $T\geq7$ seems futile.
+<!-- One exception is the CIW of the most extreme condition ($p_{mis}=.95$; $T=1$). -->
 
-Conditions with a lower number of iterations yield more bias in $\bar{Q}$.
+<!-- Conditions with a lower number of iterations yield more bias in $\bar{Q}$.
 This is especially pronounced in conditions with high missingness proportions: with 95% missingness we need at least seven iterations before an additional iteration does not lead to improved estimates.
@@
 Approximately unbiased estimates are obtained for all conditions where $p_{mis}\leq0.75$ and $T\leq6$.
-**[Add CIW and CR]**
+**[Add CIW and CR]** -->
 
@@
 
-In Figure \ref{fig:rhats-acs} we see the results of the non-convergence diagnostics $\widehat{R}$ and $AC$.
-**[Add panel labels and explain how to read the plots]** As expected, the diagnostics identify more signs of non-convergence in simulation conditions with a lower number of iterations.
+We evaluate the non-convergence diagnostics $\widehat{R}$ and $AC$ against the performance measures described above (i.e., bias and confidence validity of the quantities of scientific interest $Q$).
+The diagnostics should identify those simulation conditions that are negatively impacted by non-convergence.
+We differentiate between approximately unbiased and confidence valid estimates, and *sufficiently* valid estimates.
+The former is defined by the interplay between $Q$, $p_{mis}$, and $T$.
+The latter is observed in all simulation conditions where $T\geq7$, irrespective of the $Q$ and $p_{mis}$.
+We thus expect the most signs of non-convergence in conditions where $T<7$.
+
+<!-- As we have observed, the affected conditions are $p_{mis}=.95$ for all all $Q$s are impacted by the missingness proportions, but only the multivariate quantities of scientific interest are affected by the number of iterations.
+-->
+
+Figure \ref{fig:rhats-acs} displays the non-convergence diagnostics $\widehat{R}$ and $AC$ for each of the four $\theta$s under consideration.
+**[Add panel labels and explain how to read the plots]** In general, the diagnostics identify more signs of non-convergence in simulation conditions with a lower number of iterations.
 The missingness proportions, however, only seem to impact multivariate convergence.
-When $\widehat{R}$ and $AC$ are applied to  univariate $\theta$s, they identify the same amount of non-convergence, irrespective of the missingness proportion.
+When $\widehat{R}$ and $AC$ are applied to  univariate $\theta$s, they identify the same amount of non-convergence for any missingness proportion.
 When the diagnostics are applied to multivariate $\theta$s, they generally identify more signs of non-convergence for higher missingness proportions.
 
-None of the simulation conditions that we consider resulted in complete convergence of the algorithm (defined as $\widehat{R}=1$ and $AC=0$).
-The thresholds for diagnosing non-convergence that are used in practice seem too conservative compared to the performance measures: $\widehat{R}$ and $AC$ identify signs of non-convergence in conditions where the $\bar{Q}$s are unbiased and confidence valid estimates of the $Q$s.
+As expected, none of the simulation conditions that we consider resulted in complete convergence of the algorithm (defined as $\widehat{R}=1$ and $AC=0$).
+Yet, the thresholds used in practice to diagnose non-convergence seem too conservative as well.
+$\widehat{R}$ and $AC$ identify signs of non-convergence in conditions where the $\bar{Q}$s are unbiased and confidence valid estimates of the $Q$s.
 The most recent recommended threshold for diagnosing non-convergence with $\widehat{R}$ is never overcome in this study (all $\widehat{R}$>1.01).
-Apparently, this threshold is too stringent for iterative imputation algorithms, and will not be discussed further.
+Apparently, this threshold is too stringent for iterative imputation algorithms, and will not be discussed any further.
 

### 2020-05-06 3.Thesis/Writeup/2.Manuscript.tex
Reread Results until thetas

@@
 Figure \ref{fig:Qs} displays the influence of inducing non-convergence on the estimated quantities of scientific interest, \(Q\)s.
-\textbf{{[}Add panel labels, describe panels{]}} As expected, conditions with a higher proportion of missingness and/or a lower number of iterations are impacted more severely by non-convergence.
+\textbf{{[}Add panel labels, describe panels{]}} Overall, conditions with a higher proportion of missingness and/or a lower number of iterations are impacted more severely by non-convergence.
 I.e., on average these conditions portray more extreme bias in the estimated \(Q\)s, and non-nominal coverages.
@@
 The estimated univariate means seem unaffected by the number of iterations in the algorithm.
-The magnitude of the bias in \(\bar{Q}\) depends exclusively on the missingness proportion \(p_{mis}\), with higher missingness proportions causing more extreme bias.
-Note however, that the simulation condition with the largest overall bias resulted in a \(\bar{Q}\) under-estimated the true value of \(Q\) by only 0.02 units (where \(Q=\mu_Y=25.81; \sigma_Y = 11.32\)).
-We conclude that---conditional on the missingness proportion---all simulation conditions (\(T \geq 1\)) have equal performance.
-According to this scientific estimand \(Q\), non-convergence does not impact the validity of the inferences.
+This implies that \emph{suffient} inferences may be obtained with as little as one iteration (\(T \geq 1\)).
+The magnitude of the bias in \(\bar{Q}\) depends exclusively on the missingness proportion \(p_{mis}\), with missingness proportions of 75\% and 95\% causing more extreme bias than other conditions.
+Unbiased estimates are obtained in all conditions where \(p_{mis}\leq.50\).
+Note however, that even in the conditions where \(p_{mis}\geq.75\), the bias is relatively small: \(\mu_Y\) is maximally under-estimated by 0.02 units, while the true value of \(\mu_Y\) is 25.81 (\(\sigma_Y = 11.32\)).
+We conclude that non-convergence does not substantially impact the validity of the inferences for \(Q=\mu\).
 \textbf{{[}Add: up-to 50\% missingness everything is relatively unbiased, while 75\% and 95\% missingness show very different results.
@@
 
-The estimated standard deviations show impact of non-convergence similar to the means.
-The magnitude of the bias \(\bar{Q}\) follows the incremental order of the missingness proportions.
-The number of iterations in the algorithm has little effect on the bias.
-Only the estimates in simulation conditions with just one iteration (\(T=1\)) differ from other conditions that have the same missingness proportion but a higher number of iterations.
-Interestingly, the bias in conditions where \(T=1\) is \emph{less} severe than after continued iteration.
-For the least possible bias, we may use any number of iterations \(T\geq1\), but only \(p_{mis} = 0.05\) results in completely unbiased \(\bar{Q}\)s.
+The estimated standard deviations are largely affected by the missingness proportions, but not to much by the number of iterations.
+\emph{Sufficient} inferences are obtained in conditions with just two iterations.
+Interestingly, the bias in first iteration is actually \emph{less} extreme than for any other number of iterations.
+We may therefore conclude that no iteration at all is needed to obtain \emph{sufficient} estimates for \(Q=\sigma\).
+Completely unbiased estimates are only obtained in conditions where \(p_{mis}=.05\).
+Conditions where \(p_{mis}\geq.25\) are impacted by non-convergence in order of increasing missingness proportion.
+
+\hypertarget{qr2.}{%
+\subsubsection{\texorpdfstring{\(Q=r^2\).}{Q=r\^{}2.}}\label{qr2.}}
+
+Bias in the estimated variance explained depends both on missingness proportions and the number of iterations.
+The magnitude of the bias clearly follows the incremental order of the missingness proportions.
+We, therefore, observe \emph{sufficient} estimates at a different number of iterations for each missingness proportion.
+For \(p_{mis}\leq.25\), the minimum number of iterations is two (\(T\geq2\)), while \(p_{mis}=.50\) requires \(T\geq4\), and \(p_{mis}=.75\) requires \(T\geq5\).
+The highest number of iterations needed to obtain \emph{sufficient} estimates is seven (\(p_{mis}=.95\); \(T\geq7\)).
+Unbiased estimates, however, may only be obtained if the missingness proportion is at most 25\%.
+\textbf{{[}interpret sign of bias{]}}
 
@@
 For the estimated regression coefficients, we consider bias, confidence interval width (CIW), and coverage rate (CR) as performance measures.
-The magnitude of the bias, again, depends on the missingness proportion.
-But for this \(Q\) we also observe a clear effect of the number of iterations.
-Conditions with a lower number of iterations yield more bias in \(\bar{Q}\).
-This is especially pronounced in conditions with high missingness proportions: with 95\% missingness we need at least seven iterations before an additional iteration does not lead to improved estimates.
-Conditions with a lower missingness proportion (e.g., \(p_{mis} = {0.5}\)), produce \emph{optimal} estimates with as many as two iterations.
-Approximately unbiased estimates are obtained for all conditions where \(p_{mis}\leq0.75\) and \(T\leq6\).
-\textbf{{[}Add CIW and CR{]}}
-
-\hypertarget{qr2.}{%
-\subsubsection{\texorpdfstring{\(Q=r^2\).}{Q=r\^{}2.}}\label{qr2.}}
-
-The bias in the estimated explained variance perfectly follows both sets of simulation conditions: The worst performance is observed for the highest missingness proportion and the lowest number of iterations, while the best performance occurs with less missingness and more iterations.
-Only conditions with the lowest missingness proportion (\(p_{mis}=0.05\)) results in unbiased estimates for every number of iterations.
-In other conditions there is at least some bias irrespective of \(T\).
-The inferences do not improve in any condition where \(T \geq 6\).
-This suggests that the algorithm is as converged as necessary to obtain valid inferences.
+Unbiased estimates are obtained in conditions where \(p_{mis}\leq.50\) and \(T\geq3\).
+Approximately confidence valid estimates are obtained in any condition where \(T\geq3\)---despite of persistant bias in conditions where \(p_{mis}\geq.75\).
+These approximately nominal coverages are explained by the CIWs: conditions with higher missingness proportions result in more uncertainty in the estimates (i.e., wider CIs), which diminish the effect of the bias.
+\emph{Sufficient} inferences in terms of confidence validity may be obtained when \(T\geq3\), while we need at least seven iterations to reach this point in terms of bias.
+Continued iteration after \(T\geq7\) seems futile.
 
@@
 
-In Figure \ref{fig:rhats-acs} we see the results of the non-convergence diagnostics \(\widehat{R}\) and \(AC\).
-\textbf{{[}Add panel labels and explain how to read the plots{]}} As expected, the diagnostics identify more signs of non-convergence in simulation conditions with a lower number of iterations.
+We evaluate the non-convergence diagnostics \(\widehat{R}\) and \(AC\) against the performance measures described above (i.e., bias and confidence validity of the quantities of scientific interest \(Q\)).
+The diagnostics should identify those simulation conditions that are negatively impacted by non-convergence.
+We differentiate between approximately unbiased and confidence valid estimates, and \emph{sufficiently} valid estimates.
+The former is defined by the interplay between \(Q\), \(p_{mis}\), and \(T\).
+The latter is observed in all simulation conditions where \(T\geq7\), irrespective of the \(Q\) and \(p_{mis}\).
+We thus expect the most signs of non-convergence in conditions where \(T<7\).
+
+Figure \ref{fig:rhats-acs} displays the non-convergence diagnostics \(\widehat{R}\) and \(AC\) for each of the four \(\theta\)s under consideration.
+\textbf{{[}Add panel labels and explain how to read the plots{]}} In general, the diagnostics identify more signs of non-convergence in simulation conditions with a lower number of iterations.
 The missingness proportions, however, only seem to impact multivariate convergence.
-When \(\widehat{R}\) and \(AC\) are applied to univariate \(\theta\)s, they identify the same amount of non-convergence, irrespective of the missingness proportion.
+When \(\widehat{R}\) and \(AC\) are applied to univariate \(\theta\)s, they identify the same amount of non-convergence for any missingness proportion.
 When the diagnostics are applied to multivariate \(\theta\)s, they generally identify more signs of non-convergence for higher missingness proportions.
 
-None of the simulation conditions that we consider resulted in complete convergence of the algorithm (defined as \(\widehat{R}=1\) and \(AC=0\)).
-The thresholds for diagnosing non-convergence that are used in practice seem too conservative compared to the performance measures: \(\widehat{R}\) and \(AC\) identify signs of non-convergence in conditions where the \(\bar{Q}\)s are unbiased and confidence valid estimates of the \(Q\)s.
+As expected, none of the simulation conditions that we consider resulted in complete convergence of the algorithm (defined as \(\widehat{R}=1\) and \(AC=0\)).
+Yet, the thresholds used in practice to diagnose non-convergence seem too conservative as well.
+\(\widehat{R}\) and \(AC\) identify signs of non-convergence in conditions where the \(\bar{Q}\)s are unbiased and confidence valid estimates of the \(Q\)s.
 The most recent recommended threshold for diagnosing non-convergence with \(\widehat{R}\) is never overcome in this study (all \(\widehat{R}\)\textgreater1.01).
-Apparently, this threshold is too stringent for iterative imputation algorithms, and will not be discussed further.
+Apparently, this threshold is too stringent for iterative imputation algorithms, and will not be discussed any further.
 

### 2020-05-06 3.Thesis/Writeup/2.Manuscript.Rmd
Emphasize unaffected univariate Qs

@@
 This implies that, under the current specifications, no more than seven iterations are necessary to obtain *sufficient* performance in terms of bias and confidence validity.
+**[only substantial effect of non-conv in multivariate $Q$s]**
 
@@
 We conclude that non-convergence does not substantially impact the validity of the inferences for $Q=\mu$.
-**[Add: up-to 50% missingness everything is relatively unbiased, while 75% and 95% missingness show very different results.
-Why is that?
-Add sd of Y in the figure to show that the bias is relative: a bias of 0.02 for a variable with a mean of 25.81, and sd of 11.32 is nothing.]**
 
@@
 Conditions where $p_{mis}\geq.25$ are impacted by non-convergence in order of increasing missingness proportion.
+Similar to the bias for $Q=\mu$ the magnitude of the bias in $Q=\sigma$ is not substantial in any condition.
+We conclude that both of the univariate $Q$s are practically unaffected by non-convergence.
 <!-- in the first iteration.
@@
 The diagnostics should identify those simulation conditions that are negatively impacted by non-convergence.
-We differentiate between approximately unbiased and confidence valid estimates, and *sufficiently* valid estimates.
-The former is defined by the interplay between $Q$, $p_{mis}$, and $T$.
-The latter is observed in all simulation conditions where $T\geq7$, irrespective of the $Q$ and $p_{mis}$.
-We thus expect the most signs of non-convergence in conditions where $T<7$.
+We expect the most signs of non-convergence in conditions where *sufficient* estimates are not obtained (high missingness proportions, $T<7$), and at least some signs of non-convergence in conditions where estimates are not approximately unbiased ($p_{mis}=.95$, any number of iterations).
 
@@
 ### $\theta=$ chain mean.
-When $\widehat{R}$ and $AC$ are applied to the chain means, we notice no impact of the missingness proportions. The results only depend on the number of iterations per simulation condition.
+When $\widehat{R}$ and $AC$ are applied to the chain means, **we fail to identify the non-convergence in conditions where $p_{mis}=.95$**.
+We notice no impact of the missingness proportions.
+The results only depend on the number of iterations per simulation condition.
 

### 2020-05-07 3.Thesis/Writeup/2.Manuscript.Rmd
After feedback Gerko (not yet with notes meeting)

@@
 There is no consensus on how to evaluate the convergence properties of the method.
-This paper provides insight into identifying non-convergence of iterative imputation algorithms."
+This paper provides insight into identifying non-convergence of iterative imputation algorithms.
+We conclude that in the case of drawing inference from incomplete data, convergence of the iterative imputation algorithm is often a convenience but not a necessity."
 keywords: missing data, iterative imputation, non-convergence, mice
@@
 
+# Overview of sections
+
+Intro
+
+- This is missingness
+
+- This is how we solve missingness
+
+- For that we need an algorithm, but what if it has not converged?
+
+- That is what we solve in this paper
+
+Algorithmic convergence
+
+- What is algorithmic convergence?
+
+- When is it converged? Stationarity versus mixing.
+
+- How is convergence currently investigated? And how should it be investigated? Multivariate versus univariate.
+
+Methods
+
+- Which methods are there? And how do they relate to mixing and stationarity?
+
+- $\widehat{R}$
+
+- AC
+
+- We can use these on any $\theta$
+
+The Simulation study, Results, Discussion.
+
 # Introduction
@@
 Anyone who analyzes person-data may run into a missing data problem.
-Missing data is not only ubiquitous, but treating it can also be tedious.
+Missing data is not only ubiquitous across science, but treating it can also be a tedious task.
 If a dataset contains just one incomplete observation, statistical inferences are undefined and will not produce any results.
@@
 Figure \ref{fig:diagram} provides an overview of the steps involved with MI.
-Missing data $y_{mis}$ in an incomplete dataset is imputed $M$ times.
+The missing part $y_{mis}$ of an incomplete dataset is imputed $M$ times.
 The imputed data $y_{imp}$ is combined with the observed data $y_{obs}$ to create $M$ completed datasets.
@@
 These $M$ values are combined into a single pooled estimate $\bar{Q}$.
-The premise of multiple imputation is that $\bar{Q}$ is an unbiased and confidence valid estimate of the true---but missing---data inference.
+The premise of multiple imputation is that $\bar{Q}$ is an unbiased and confidence valid estimate of the true---but unobserved---inference.
 
@@
 And can we rely on the inference obtained on the completed data?
-These are all open questions, because the convergence properties of iterative imputation algorithms have not been systematically studied [@buur18].
+
+These remain open questions, because the convergence properties of iterative imputation algorithms have not been systematically studied [@buur18].
 Moreover, there is no scientific consensus on how to evaluate convergence of imputation algorithms [@taka17].
-Some default MICE techniques (e.g., 'predictive mean modeling') might not yield converged states at all [@murr18].
+Some default MICE techniques (e.g., 'predictive mean matching') might not yield converged states at all [@murr18].
 Therefore, algorithmic convergence should be monitored carefully.
@@
 Currently, the recommended practice for evaluating the convergence of iterative imputation algorithms is through visual inspection.
-That is, the state space of the algorithm is monitored by plotting a scalar summary $\theta$ against the iteration number.
-**[Segway to WHAT TO MONITOR?
-Also, this paragraph is more or less the same as the third paragraph in the second section.
-Merge or remove one of the two?]** Typically, the $\theta$s under evaluation are chain means and variances (**define!**).
-These $\theta$s may be insufficient because they are summaries of the univariate state space of the algorithm, while MICE is not only concerned with univariate estimates, but the entire multivariate distribution of $y_{imp}$.
-A suggestion by @buur18 for a multivariate $\theta$ to monitor is the quantity of scientific interest $Q$ (e.g., a regression coefficient).
-Implementing this, however, might be somewhat advanced for empirical researchers.
-Moreover, this scalar summary is not model-independent, i.e., it only applies to one model of scientific interest.
-Yet, one of the advantages of MI is that the missing data problem and scientific problem are solved independently.
-Therefore, this $\theta$ is not sufficient either.
-Van Buuren [-@buur18, $\S$ 4.5.2] also proposed multivariable evaluation of the MICE algorithm through eigenvalue decomposition, building on the work of @mack03.
-Eigenvalues of a variance-covariance matrix are a measure of the data's total covariance.
-This would be a model-independent, multivariate scalar summary to monitor, but it has not yet been implemented.
-<!-- Recommendation: "the different streams should be freely intermingled with one another, without showing any definite trends". -->
-**[Segway to HOW TO MONITOR?]** Monitoring convergence visually is insufficient on two counts: 1) it may be challenging to the untrained eye, and 2) only severely pathological cases of non-convergence may be diagnosed [@buur18, $\S$ 6.5.2].
+After running the imputation algorithm for a certain number of iterations, its convergence is monitiored by plotting a scalar summary of the state space of the algorithm $\theta$ against the iteration number.
+Typically, the $\theta$s under evaluation are chain means and variances---the average and the variance of the imputed values per imputation.
+Monitoring convergence by inspecting traceplots of these $\theta$s is insufficient on two counts: 1) it may be challenging to the untrained eye, and 2) only severely pathological cases of non-convergence may be diagnosed [@buur18, $\S$ 6.5.2] **[Add that convergence of univariate thetas cannot guarantee multivariate convergence?]**.
 Therefore, a quantitative, diagnostic evaluation of convergence would be preferred---although not straightforward.
@@
 The collection of samples between the initial value (at $t=1$) and the final imputed value (at $t=T$) will be referred to as an 'imputation chain'.
-**[If not defined elsewhere, add: $\theta$s are scalar summaries of the algorithm at a certain iteration (e.g., chain means; the average of the imputed values in each imputation chain).] **
+<!-- **[If not defined elsewhere, add: $\theta$s are scalar summaries of the algorithm at a certain iteration (e.g., chain means; the average of the imputed values in each imputation chain).] ** -->
 
-# Identifying non-convergence
+# Algorithmic non-convergence
 
@@
 
-**[Move to INTRO???]** To illustrate what non-convergence looks like in iterative imputation algorithms, we reproduce an example from van Buuren [-@buur18, $\S$ 6.5.2].
+**[SEGWAY to this section]** To illustrate what non-convergence looks like in iterative imputation algorithms, we reproduce an example from van Buuren [-@buur18, $\S$ 6.5.2].
 Figure \ref{fig:non-conv} displays a subset of the example: a variable that we will call $j$.
@@
 The right-hand side displays pathological non-convergence, induced by purposefully mis-specifying the imputation model.
-**[Leave this in if it's after mixing and stationarity are defined, otherwise remove?] In the typical convergence situation, the imputation chains intermingle nicely and there is little to no trending.
+**In the typical convergence situation, the imputation chains intermingle nicely and there is little to no trending.
 In the non-convergence plot, there is a lot of trending and some chains do not intermingle.** Importantly, the chain means at the last iteration (the imputed value per $m$) are very different between the two plots.
@@
 
-**[Move to INTRO???]** Convergence has historically been inspected visually, by monitoring the imputation algorithm over iterations.
+**[MERGE with section below]**.
+Inspecting non-convergence through traceplots of chain means and chain variances may be insufficient because they are summaries of the univariate state space of the algorithm, while MICE is not only concerned with a sigle column, but the entire multivariate distribution of $y_{imp}$.
+Ideally, a multivariate $\theta$ should be monitored.
+A suggestion by @buur18 for a multivariate $\theta$ to monitor is the quantity of scientific interest $Q$ (e.g., a regression coefficient), which usually is multivariate in nature.
+Implementing this, however, might be somewhat advanced for empirical researchers.
+Moreover, this scalar summary is not model-independent, i.e., it only applies to one model of scientific interest.
+Yet, one of the advantages of MI is that the missing data problem and scientific problem are solved independently.
+On these grounds, such a $\theta$ can be considered insufficient too.
+Van Buuren [-@buur18, $\S$ 4.5.2] also proposed multivariable evaluation of the MICE algorithm through eigenvalue decomposition, building on the work of @mack03.
+Eigenvalues of a variance-covariance matrix are a measure of the data's total covariance.
+Such a $\theta$ would be a model-independent, multivariate scalar summary to monitor, but it has not yet been implemented.
+<!-- Recommendation: "the different streams should be freely intermingled with one another, without showing any definite trends". -->
+
+**[MERGE with section above]** Convergence has historically been inspected visually, by monitoring the imputation algorithm over iterations.
 This is typically done through traceplots.
@@
 
-## Diagnostics under evaluation
+# Non-convergence diagnostics
 
@@
 
-**[Add: effective sample size is used if we want independent samples *within* chains.
+**[Add: we don't use effective sample size and MC error because they assume independent samples *within* chains.
 In MI, we only use the final estimate.
@@
 
-### Potential scale reduction factor
+## Potential scale reduction factor
 
@@
 
-### Autocorrelation
+## Autocorrelation
 
@@
 
-### Thresholds
+## Thresholds
 
@@
 Subsequently we will evaluate how well $\widehat{R}$ and $AC$ perform in identifying the effects of non-convergence **[and performance of different thetas]**.
-And finally, we will formulate an informed advice on the requirements to safely assume *sufficient* convergence in practice.
-That is, we assume that the algorithm is *sufficiently* converged when each estimate $\bar{Q}$ is an unbiased and confidence valid estimate of the corresponding estimand $Q$.
+And finally, we will formulate an informed advice on the requirements to safely assume sufficient convergence in practice.
+That is, we assume that the algorithm is sufficiently converged when each estimate $\bar{Q}$ is an unbiased and confidence valid estimate of the corresponding estimand $Q$.
 **[and formulate advice on thetas?]**
@@
 We assume that we can induce non-convergence at least in conditions where $T$ is smaller than the default number of iterations in `mice`, $T=5$.
+
+**[SEGWAY to this section] $Q$s are the targets that will be estimated in each simulation repetition, after first *amputing* the complete data with varying missingness proportions, then imputing the missingness with a varying number of iterations, then estimating the $Q$s on the $M$ completed datasets, and finally pooling the $M$ estimates $\hat{Q}$ to obtain a single $\bar{Q}$ for each $Q$.
+We try to predict the effect of non-convergence (i.e., the difference between $Q$ and $\bar{Q}$) with the convergence diagnostics $\widehat{R}$ and $AC$, by applying these methods to several scalar summaries of the state space of the algorithm, $\theta$s.**
 
@@
 
-5.
-**[Something about thetas, that multivariate thetas are better at detecting non-convergence than univariate thetas.
-The non-convergence diagnostics applied on the chain means and chain variances will fail to identify signs of non-convergence when the multivariate Qs are affected.]**
+5. We further hypothesize that multivariate $\theta$s are better at detecting non-convergence than univariate $\theta$s.
 
@@
 Therefore, a single complete dataset may serve as comparative truth in all simulation repetitions [@vink14].
-The data-generating mechanism is a multivariate normal distribution, representing person-data on three predictor variables in a multiple linear regression problem **(from an unspecified social scientific field of study)**.
+The data-generating mechanism is a multivariate normal distribution, representing person-data on three predictor variables in a multiple linear regression problem.
 Let the predictor space be defined as
@@
 $$
-where $\epsilon \sim \mathcal{N}(0, 100)$.
-From the complete set, we obtain the true values of several quantities of scientific interest, $Q$s.
-**[MOVE this next part to the start of the Simulation Study section?] $Q$s are the targets that will be estimated in each simulation repetition, after first *amputing* the complete data with varying missingness proportions, then imputing the missingness with a varying number of iterations, then estimating the $Q$s on the $M$ completed datasets, and finally pooling the $M$ estimates $\hat{Q}$ to obtain a single $\bar{Q}$ for each $Q$.
-We try to predict the effect of non-convergence (i.e., the difference between $Q$ and $\bar{Q}$) with the convergence diagnostics $\widehat{R}$ and $AC$, by applying these methods to several scalar summaries of the state space of the algorithm, $\theta$s.**
+where $\epsilon \sim \mathcal{N}(0, 100)$. From the complete set, we obtain the true values of several quantities of scientific interest, $Q$s.
 
@@
 where $Y'$ is the expected value of the outcome.
-And finally, the proportion of variance explained by the set of predictors: coefficient of determination $r^2 \times 100$ **(note that lower case $r$ is used to avoid confusion with non-convergence diagnostic $\widehat{R}$)**.
+And finally, the proportion of variance explained by the set of predictors: coefficient of determination $r^2$ **(note that lower case $r$ is used to avoid confusion with non-convergence diagnostic $\widehat{R}$)**.
 The true values of these scientific estimands are calculated on the complete data with the `R` functions `base::mean()`, `stats::sd()`, and `stats::lm()` [@R].
@@
 
-With these performance measures, we also evaluate how well the non-convergence diagnostics $\widehat{R}$ and $AC$ can identify simulation conditions in which the $\bar{Q}$s are *not* unbiased, confidence valid estimates of the $Q$s.
-**[Rephrase this:] We measure the performance of $\widehat{R}$ and $AC$ by the four scalar summaries $\theta$: chain means (i.e., the mean in each $y_{imp, m, j}$) versus $Q=\mu$, chain variances (i.e., the variance in each $y_{imp, m, j}$) versus $Q=\sigma$, the estimated regression coefficients $\beta$ in each {$y_{obs}, y_{imp, m}$} (i.e., a user-defined statistic of scientific interest $\hat{Q}$) against $Q=\beta$, and the first eigenvalue of the variance-covariance matrix in each {$y_{obs}, y_{imp, m}$}, $\lambda_1$ versus $Q=r^2$.**
+With these performance measures, we also evaluate how well the non-convergence diagnostics $\widehat{R}$ and $AC$ can identify simulation conditions in which the inferences are affected by non-convergence, i.e., where $\bar{Q}$ is *not* an unbiased, confidence valid estimate of $Q$.
+**[REPHRASE this:] For performance of univariate $Q$s, we apply the non-convergence diagnostics to univariate $\theta$s.
+When $Q$ is multivariate, we look at signs of non-convergence identified in multivariate $\theta$s.**
 
-**[Remove this?] Primarily, we want to know what the effect is of inducing non-convergence on the estimated $Q$s.
+- For $Q=\mu$, $\theta$s are chain means.
+
+- For $Q=\sigma$, $\theta$s are chain variances.
+
+- For $Q=\beta$, $\theta$s are estimated $\beta$s per imputation.
+
+- For $Q=r^2$, $\theta$s are the first eigenvalue of the variance-covariance matrix.
+
+<!-- We measure the performance of $\widehat{R}$ and $AC$ by the four scalar summaries $\theta$: chain means (i.e., the mean in each $y_{imp, m, j}$) versus $Q=\mu$, chain variances (i.e., the variance in each $y_{imp, m, j}$) versus $Q=\sigma$, the estimated regression coefficients $\beta$ in each {$y_{obs}, y_{imp, m}$} (i.e., a user-defined statistic of scientific interest $\hat{Q}$) against $Q=\beta$, and the first eigenvalue of the variance-covariance matrix in each {$y_{obs}, y_{imp, m}$}, $\lambda_1$ versus $Q=r^2$.** -->
+
+<!-- **[Remove this?] Primarily, we want to know what the effect is of inducing non-convergence on the estimated $Q$s.
 More specifically, we are interested in the bias of $\bar{Q}$ for all $Q$s and the confidence validity of $\bar{Q}$ for $Q=\beta$.
@@
 Secondly, we want to know if conditions in which the $Q$s are affected by non-convergence are also identified by the non-convergence diagnostics $\widehat{R}$ and $AC$.
-We do this, again, by looking at the performance measures for the $Q$s, but now we evaluate these against what $\widehat{R}$ and $AC$ indicate about signs of non-convergence in the $\theta$s.**
+We do this, again, by looking at the performance measures for the $Q$s, but now we evaluate these against what $\widehat{R}$ and $AC$ indicate about signs of non-convergence in the $\theta$s.** -->
 
@@
 
-**[Add to this section: 1) emphasize that the iterations in the plots are averages across repetitions, not within one run of MICE; 2) Add more info about figure legends and axes]**
+**[Add more info about figure legends and axes]**
 
@@
 Figure \ref{fig:Qs} displays the influence of inducing non-convergence on the estimated quantities of scientific interest, $Q$s.
-**[Add panel labels, describe panels]** Overall, conditions with a higher proportion of missingness and/or a lower number of iterations are impacted more severely by non-convergence.
-I.e., on average these conditions portray more extreme bias in the estimated $Q$s, and non-nominal coverages.
-Roughly speaking, this means that the algorithm is indeed not converged at $t=1$, and converges gradually as $t$ increases to $T$.
+**[Add panel labels, describe panels]** Simulation conditions that are impacted by non-convergence portray more extreme bias in the estimated $Q$s, and non-nominal coverages.
+Overall, we see that conditions with a higher proportion of missingness are affected more severely by non-convergence.
+Multivariate $Q$s are also substantially influenced by the number of iterations.
+This means that the algorithm has indeed not converged at $t=1$, and converges gradually as $t$ moves to $T$.
 The point at which an additional iteration does not lead to an improvement in the estimates depends on quantity of scientific interest $Q$ and the missingness proportion $p_{mis}$.
 The highest number of iterations that is necessary to reach this point across all simulation conditions is $T=7$.
-This implies that, under the current specifications, no more than seven iterations are necessary to obtain *sufficient* performance in terms of bias and confidence validity.
-**[only substantial effect of non-conv in multivariate $Q$s]**
+This implies that, under the current specifications, no more than seven iterations are necessary to obtain sufficient performance in terms of bias and confidence validity.
 
@@
 The estimated univariate means seem unaffected by the number of iterations in the algorithm.
-This implies that *suffient* inferences may be obtained with as little as one iteration ($T \geq 1$).
-The magnitude of the bias in $\bar{Q}$ depends exclusively on the missingness proportion $p_{mis}$, with missingness proportions of 75% and 95% causing more extreme bias than other conditions.
+This implies that valid inferences may be obtained with as little as one iteration ($T \geq 1$).
+The magnitude of the bias in $\bar{Q}$ depends solely on the missingness proportion $p_{mis}$, with missingness proportions of 75% and 95% leading to more extreme biases than other missingness proportions.
 Unbiased estimates are obtained in all conditions where $p_{mis}\leq.50$.
-Note however, that even in the conditions where $p_{mis}\geq.75$, the bias is relatively small: $\mu_Y$ is maximally under-estimated by 0.02 units, while the true value of $\mu_Y$ is 25.81 ($\sigma_Y = 11.32$).
+Note however, that even in the conditions where $p_{mis}\geq.75$, the bias is small: $\mu_Y$ is maximally under-estimated by 0.02 units, while the true value of $\mu_Y$ is 25.81 ($\sigma_Y = 11.32$).
 We conclude that non-convergence does not substantially impact the validity of the inferences for $Q=\mu$.
@@
 The estimated standard deviations are largely affected by the missingness proportions, but not to much by the number of iterations.
-*Sufficient* inferences are obtained in conditions with just two iterations.
+Valid inferences are obtained in conditions with just two iterations.
 Interestingly, the bias in first iteration is actually *less* extreme than for any other number of iterations.
-We may therefore conclude that no iteration at all is needed to obtain *sufficient* estimates for $Q=\sigma$.
+We may therefore conclude that no iteration at all is needed to obtain approximately unbiased estimates for $Q=\sigma$.
 Completely unbiased estimates are only obtained in conditions where $p_{mis}=.05$.
@@
 The magnitude of the bias clearly follows the incremental order of the missingness proportions.
-We, therefore, observe *sufficient* estimates at a different number of iterations for each missingness proportion.
+We, therefore, observe approximately unbiased estimates at a different number of iterations for each missingness proportion.
 For $p_{mis}\leq.25$, the minimum number of iterations is two ($T\geq2$), while $p_{mis}=.50$ requires $T\geq4$, and $p_{mis}=.75$ requires $T\geq5$.
-The highest number of iterations needed to obtain *sufficient* estimates is seven ($p_{mis}=.95$; $T\geq7$).
+The highest number of iterations needed to obtain valid inferences is seven ($p_{mis}=.95$; $T\geq7$).
 Unbiased estimates, however, may only be obtained if the missingness proportion is at most 25%.
@@
 These approximately nominal coverages are explained by the CIWs: conditions with higher missingness proportions result in more uncertainty in the estimates (i.e., wider CIs), which diminish the effect of the bias.
-*Sufficient* inferences in terms of confidence validity may be obtained when $T\geq3$, while we need at least seven iterations to reach this point in terms of bias.
+Confidence valid inferences may be obtained when $T\geq3$, while we need at least seven iterations to reach valid inferences in terms of bias.
 Continued iteration after $T\geq7$ seems futile.
@@
 As expected, none of the simulation conditions that we consider resulted in complete convergence of the algorithm (defined as $\widehat{R}=1$ and $AC=0$).
-Yet, the thresholds used in practice to diagnose non-convergence seem too conservative as well.
+However, the thresholds used in practice to diagnose non-convergence seem too conservative for imputation purposes.
 $\widehat{R}$ and $AC$ identify signs of non-convergence in conditions where the $\bar{Q}$s are unbiased and confidence valid estimates of the $Q$s.
@@
 ### $\theta=$ chain mean.
-When $\widehat{R}$ and $AC$ are applied to the chain means, **we fail to identify the non-convergence in conditions where $p_{mis}=.95$**.
-We notice no impact of the missingness proportions.
-The results only depend on the number of iterations per simulation condition.
-
-**[MAYBE remove all this because it has no relation to the bias in Q] The highest $\widehat{R}$-values are obtained for the condition $T=2$.
-We observe the same 'dip' in $\widehat{R}$ as we did in Figure \ref{fig:diagnostics} for conditions with three iterations.
-$\widehat{R}$ gradually decreases as the number of iterations increases across conditions.
-By the traditional threshold  $\widehat{R}>1.2$ we should conclude that there is non-convergence in conditions where $T<8$.
-The common threshold $\widehat{R}>1.1$ is overcome in conditions where $T>14$.
-**
-
-**Autocorrelations indicate some signs of non-stationarity in all simulation conditions, but primarily in conditions where $T<6$.
-The threshold for diagnosing non-convergence is not reached in any condition.
-We would thus conclude that the algorithm is sufficiently converged regardless of the number of iterations or missingness proportion.
-**
-
-We evaluate the diagnostics with chain means as $\theta$ against the bias in $Q=\mu$.
-According to the bias in the univariate means, we should conclude that two iterations is sufficient, and we should see that $p_{mis}=0.75$ and $p_{mis}=0.95$ have worse convergence than the other missingness proportions.
-This is not the case.
+We evaluate the performance of the non-convergence diagnostics applied to chain means against the bias in $Q=\mu$.
+Based on the bias in $Q$, we concluded that one iteration is sufficient to obtain valid inferences, irrespective of the missingness proportion.
+The non-convergence diagnostics, however, identify some signs of non-convergence in conditions with as many as thirty iterations.
+This does not correspond to the impact of non-convergence in the univariate $Q$s.
+**Compared to the multivariate $Q$s [add before that we evaluate the thetas against all Qs?], $\widehat{R}$ and $AC$ applied to the chain means fail to identify the non-convergence in conditions where $p_{mis}=.95$.**
 
@@
 The non-convergence diagnostics applied to the chain variances as $\theta$ yield highly similar results to the chain means.
-We therefore evaluate these results immediately against the corresponding $Q$, without describing the $\widehat{R}$ and $AC$ values.
-While the bias in the estimated standard deviations  mainly depends on the missingness proportion of each condition, $\widehat{R}$ and $AC$ are influenced mostly by the number of iterations.
-We thus conclude that there is no added value of evaluating non-convergence with chain variances as $\theta$ in the diagnostics.
+**[Segway to performance of $Q=\sigma$] While the bias in the estimated standard deviations mainly depends on the missingness proportion of each condition, $\widehat{R}$ and $AC$ are influenced mostly by the number of iterations.
+We thus conclude that there is no added value of evaluating non-convergence with chain variances as $\theta$ in the diagnostics.**
 
@@
 The conclusion to identify non-stationarity in conditions where $T\geq10$ and $p_{mis} = 0.95$ might therefore be spurious.
-**[Also, the $AC$ across missingness proportions shows the opposite of what you would expect: shouldn't there be higher autocorrelations in conditions with comparatively more observed data in the completed datasets?
-I don't understand this.]**
 
-Compared to the bias in $\bar{Q}$, the diagnostics applied to the $\hat{Q}$s over-estimate the severity of the non-convergence. This is odd, since the $\theta$ and $Q$ are directly related.
+<!-- **[Also, the $AC$ across missingness proportions shows the opposite of what you would expect: shouldn't there be higher autocorrelations in conditions with comparatively more observed data in the completed datasets?
+I don't understand this.]** -->
+
+Compared to the bias in $\bar{Q}$, the diagnostics applied to the $\hat{Q}$s over-estimate the severity of the non-convergence. 
+<!-- This is odd, since the $\theta$ and $Q$ are directly related.   -->
 
@@
 
-## In short [remove later!]
-
-- Is there more bias in conditions where non-convergence was induced? (NOPE for univariates, YES for multivar.).
-
-- Is the point of *sufficient* convergence (at which the $\bar{Q}$s are unbiased and confidence valid estimates of the $Q$s) the same point as identified by the non-convergence diagnostics?
-(NOPE, they are too conservative).
-
-- Is one of the thetas better? (YES, while it doesn't really matter as long as it's multivariate, $\lambda_1$ is model-independent and therefore preferred).
-
-- What should the thresholds be? (much HIGHER than recommended for $\widehat{R}$, difficult to say for $AC$).
-
-- And the default nr of iterations? (MAYBE increase to 10 to detect significant $AC$ and be sure not to miss the dip in $\widehat{R}$?).
-
 # Discussion
@@
 Which might be why $\widehat{R}$ and $AC$ identify non-convergence.
+
+```{r eval=FALSE, include=FALSE}
+# **[INCLUDE this] The highest $\widehat{R}$-values are obtained for the condition $T=2$.
+We observe the same 'dip' in $\widehat{R}$ as we did in Figure \ref{fig:diagnostics} for conditions with three iterations.
+$\widehat{R}$ gradually decreases as the number of iterations increases across conditions.
+By the traditional threshold  $\widehat{R}>1.2$ we should conclude that there is non-convergence in conditions where $T<8$.
+The common threshold $\widehat{R}>1.1$ is overcome in conditions where $T>14$.
+**
+#   
+#   **Autocorrelations indicate some signs of non-stationarity in all simulation conditions, but primarily in conditions where $T<6$.
+The threshold for diagnosing non-convergence is not reached in any condition.
+We would thus conclude that the algorithm is sufficiently converged regardless of the number of iterations or missingness proportion.
+**
+```
+
+Some other interesting findings are **[REPHRASE]**
+
+- Is there more bias in conditions where non-convergence was induced? (YES for multivar., NOPE for univariates).
+
+- Is the point of *sufficient* convergence (at which the $\bar{Q}$s are unbiased and confidence valid estimates of the $Q$s) the same point as identified by the non-convergence diagnostics?
+(NOPE, they are too conservative).
+
+- Is one of the thetas better? (YES, while it doesn't really matter as long as it's multivariate, $\lambda_1$ is model-independent and therefore preferred).
+
+- What should the thresholds be? (much HIGHER than recommended for $\widehat{R}$, difficult to say for $AC$).
+
+- And the default nr of iterations? (MAYBE increase to 10 to detect significant $AC$ and be sure not to miss the dip in $\widehat{R}$?).
 
@@
 
-For empirical researchers:
+After imputation it is always wise to take the following steps:
 
-1. Check traceplots for signs of pathological non-convergence. Adjust imputation model if necessary.
+1. First, check traceplots for signs of pathological non-convergence. Adjust imputation model if necessary.
 
@@
 
-3.
+3. Track $\theta$s over iterations, e.g., the novel $\theta$ that is 'substantive model-independent'.
+
+4. Track your own scalar summary of interest (see e.g., @buur18). Compute $\widehat{R}$ and autocorrelation values for this scalar summary.
+
+5.
 Do not use the traditional calculation for $\widehat{R}$, or the `R` function `stats::acf()` to compute $AC$.
@@
 
-4. Track your own scalar summary of interest (see e.g., @buur18). Compute $\widehat{R}$ and autocorrelation values for this scalar summary.
-
-5. **Something** about the novel $\theta$ that is 'substantive model-independent'.
-
 ## Recommendations for future research
 
+Much remains unknown.
+about non-convergence in iterative imputation algorithms.
 Further research is needed to investigate the performance of iterative imputation algorithms under clear violation of convergence, e.g.
@@
 
-- Implement recommendations in ShinyMICE.
+- Implement recommendations in imputation software, e.g., evaluation environment for applied researchers.
 
@@
 
-In short, we have shown that an iterative imputation algorithm can yield correct outcomes, even when a converged state is not reached according to two common non-convergence diagnostics.
-Continued iteration just burns a lot of computing time, without improving the statistical inferences.
+In short, we have shown that an iterative imputation algorithm can yield correct outcomes, even when a converged state has not yet formally been reached.
+Any further iterations would then burn computational resources without improving the statistical inferences.
 To conclude, in the case of drawing inference from incomplete data, convergence of the iterative imputation algorithm is often a convenience but not a necessity.

### 2020-05-07 3.Thesis/Writeup/2.Manuscript.tex
After feedback Gerko (not yet with notes meeting)

@@
 This paper provides insight into identifying non-convergence of iterative imputation algorithms.
+We conclude that in the case of drawing inference from incomplete data, convergence of the iterative imputation algorithm is often a convenience but not a necessity.
 \end{abstract}
@@
 
+\hypertarget{overview-of-sections}{%
+\section{Overview of sections}\label{overview-of-sections}}
+
+Intro
+
+\begin{itemize}
+\item
+  This is missingness
+\item
+  This is how we solve missingness
+\item
+  For that we need an algorithm, but what if it has not converged?
+\item
+  That is what we solve in this paper
+\end{itemize}
+
+Algorithmic convergence
+
+\begin{itemize}
+\item
+  What is algorithmic convergence?
+\item
+  When is it converged? Stationarity versus mixing.
+\item
+  How is convergence currently investigated? And how should it be investigated? Multivariate versus univariate.
+\end{itemize}
+
+Methods
+
+\begin{itemize}
+\item
+  Which methods are there? And how do they relate to mixing and stationarity?
+\item
+  \(\widehat{R}\)
+\item
+  AC
+\item
+  We can use these on any \(\theta\)
+\end{itemize}
+
+The Simulation study, Results, Discussion.
+
 \hypertarget{introduction}{%
@@
 Figure \ref{fig:diagram} provides an overview of the steps involved with MI.
-Missing data \(y_{mis}\) in an incomplete dataset is imputed \(M\) times.
+The missing part \(y_{mis}\) of an incomplete dataset is imputed \(M\) times.
 The imputed data \(y_{imp}\) is combined with the observed data \(y_{obs}\) to create \(M\) completed datasets.
@@
 These \(M\) values are combined into a single pooled estimate \(\bar{Q}\).
-The premise of multiple imputation is that \(\bar{Q}\) is an unbiased and confidence valid estimate of the true---but missing---data inference.
+The premise of multiple imputation is that \(\bar{Q}\) is an unbiased and confidence valid estimate of the true---but unobserved---inference.
 
@@
 And can we rely on the inference obtained on the completed data?
-These are all open questions, because the convergence properties of iterative imputation algorithms have not been systematically studied \citep{buur18}.
+
+These remain open questions, because the convergence properties of iterative imputation algorithms have not been systematically studied \citep{buur18}.
 Moreover, there is no scientific consensus on how to evaluate convergence of imputation algorithms \citep{taka17}.
-Some default MICE techniques (e.g., `predictive mean modeling') might not yield converged states at all \citep{murr18}.
+Some default MICE techniques (e.g., `predictive mean matching') might not yield converged states at all \citep{murr18}.
 Therefore, algorithmic convergence should be monitored carefully.
@@
 Currently, the recommended practice for evaluating the convergence of iterative imputation algorithms is through visual inspection.
-That is, the state space of the algorithm is monitored by plotting a scalar summary \(\theta\) against the iteration number.
-\textbf{{[}Segway to WHAT TO MONITOR?
-Also, this paragraph is more or less the same as the third paragraph in the second section.
-Merge or remove one of the two?{]}} Typically, the \(\theta\)s under evaluation are chain means and variances (\textbf{define!}).
-These \(\theta\)s may be insufficient because they are summaries of the univariate state space of the algorithm, while MICE is not only concerned with univariate estimates, but the entire multivariate distribution of \(y_{imp}\).
-A suggestion by \citet{buur18} for a multivariate \(\theta\) to monitor is the quantity of scientific interest \(Q\) (e.g., a regression coefficient).
-Implementing this, however, might be somewhat advanced for empirical researchers.
-Moreover, this scalar summary is not model-independent, i.e., it only applies to one model of scientific interest.
-Yet, one of the advantages of MI is that the missing data problem and scientific problem are solved independently.
-Therefore, this \(\theta\) is not sufficient either.
-Van Buuren \citeyearpar[\(\S\) 4.5.2]{buur18} also proposed multivariable evaluation of the MICE algorithm through eigenvalue decomposition, building on the work of \citet{mack03}.
-Eigenvalues of a variance-covariance matrix are a measure of the data's total covariance.
-This would be a model-independent, multivariate scalar summary to monitor, but it has not yet been implemented.
-\textbf{{[}Segway to HOW TO MONITOR?{]}} Monitoring convergence visually is insufficient on two counts: 1) it may be challenging to the untrained eye, and 2) only severely pathological cases of non-convergence may be diagnosed \citep[\(\S\) 6.5.2]{buur18}.
+After running the imputation algorithm for a certain number of iterations, its convergence is monitiored by plotting a scalar summary of the state space of the algorithm \(\theta\) against the iteration number.
+Typically, the \(\theta\)s under evaluation are chain means and variances---the average and the variance of the imputed values per imputation.
+Monitoring convergence by inspecting traceplots of these \(\theta\)s is insufficient on two counts: 1) it may be challenging to the untrained eye, and 2) only severely pathological cases of non-convergence may be diagnosed \citep[\(\S\) 6.5.2]{buur18} \textbf{{[}Add that convergence of univariate thetas cannot guarantee multivariate convergence?{]}}.
 Therefore, a quantitative, diagnostic evaluation of convergence would be preferred---although not straightforward.
@@
 The collection of samples between the initial value (at \(t=1\)) and the final imputed value (at \(t=T\)) will be referred to as an `imputation chain'.
-\textbf{{[}If not defined elsewhere, add: \(\theta\)s are scalar summaries of the algorithm at a certain iteration (e.g., chain means; the average of the imputed values in each imputation chain).{]} }
 
-\hypertarget{identifying-non-convergence}{%
-\section{Identifying non-convergence}\label{identifying-non-convergence}}
+\hypertarget{algorithmic-non-convergence}{%
+\section{Algorithmic non-convergence}\label{algorithmic-non-convergence}}
 
@@
 
-\textbf{{[}Move to INTRO???{]}} To illustrate what non-convergence looks like in iterative imputation algorithms, we reproduce an example from van Buuren \citeyearpar[\(\S\) 6.5.2]{buur18}.
+\textbf{{[}SEGWAY to this section{]}} To illustrate what non-convergence looks like in iterative imputation algorithms, we reproduce an example from van Buuren \citeyearpar[\(\S\) 6.5.2]{buur18}.
 Figure \ref{fig:non-conv} displays a subset of the example: a variable that we will call \(j\).
@@
 The right-hand side displays pathological non-convergence, induced by purposefully mis-specifying the imputation model.
-\textbf{{[}Leave this in if it's after mixing and stationarity are defined, otherwise remove?{]} In the typical convergence situation, the imputation chains intermingle nicely and there is little to no trending.
+\textbf{In the typical convergence situation, the imputation chains intermingle nicely and there is little to no trending.
 In the non-convergence plot, there is a lot of trending and some chains do not intermingle.} Importantly, the chain means at the last iteration (the imputed value per \(m\)) are very different between the two plots.
@@
 
-\textbf{{[}Move to INTRO???{]}} Convergence has historically been inspected visually, by monitoring the imputation algorithm over iterations.
+\textbf{{[}MERGE with section below{]}}.
+Inspecting non-convergence through traceplots of chain means and chain variances may be insufficient because they are summaries of the univariate state space of the algorithm, while MICE is not only concerned with a sigle column, but the entire multivariate distribution of \(y_{imp}\).
+Ideally, a multivariate \(\theta\) should be monitored.
+A suggestion by \citet{buur18} for a multivariate \(\theta\) to monitor is the quantity of scientific interest \(Q\) (e.g., a regression coefficient), which usually is multivariate in nature.
+Implementing this, however, might be somewhat advanced for empirical researchers.
+Moreover, this scalar summary is not model-independent, i.e., it only applies to one model of scientific interest.
+Yet, one of the advantages of MI is that the missing data problem and scientific problem are solved independently.
+On these grounds, such a \(\theta\) can be considered insufficient too.
+Van Buuren \citeyearpar[\(\S\) 4.5.2]{buur18} also proposed multivariable evaluation of the MICE algorithm through eigenvalue decomposition, building on the work of \citet{mack03}.
+Eigenvalues of a variance-covariance matrix are a measure of the data's total covariance.
+Such a \(\theta\) would be a model-independent, multivariate scalar summary to monitor, but it has not yet been implemented.
+
+\textbf{{[}MERGE with section above{]}} Convergence has historically been inspected visually, by monitoring the imputation algorithm over iterations.
 This is typically done through traceplots.
@@
 
-\hypertarget{diagnostics-under-evaluation}{%
-\subsection{Diagnostics under evaluation}\label{diagnostics-under-evaluation}}
+\hypertarget{non-convergence-diagnostics}{%
+\section{Non-convergence diagnostics}\label{non-convergence-diagnostics}}
 
@@
 
-\textbf{{[}Add: effective sample size is used if we want independent samples \emph{within} chains.
+\textbf{{[}Add: we don't use effective sample size and MC error because they assume independent samples \emph{within} chains.
 In MI, we only use the final estimate.
@@
 \hypertarget{potential-scale-reduction-factor}{%
-\subsubsection{Potential scale reduction factor}\label{potential-scale-reduction-factor}}
+\subsection{Potential scale reduction factor}\label{potential-scale-reduction-factor}}
 
@@
 \hypertarget{autocorrelation}{%
-\subsubsection{Autocorrelation}\label{autocorrelation}}
+\subsection{Autocorrelation}\label{autocorrelation}}
 
@@
 \hypertarget{thresholds}{%
-\subsubsection{Thresholds}\label{thresholds}}
+\subsection{Thresholds}\label{thresholds}}
 
@@
 Subsequently we will evaluate how well \(\widehat{R}\) and \(AC\) perform in identifying the effects of non-convergence \textbf{{[}and performance of different thetas{]}}.
-And finally, we will formulate an informed advice on the requirements to safely assume \emph{sufficient} convergence in practice.
-That is, we assume that the algorithm is \emph{sufficiently} converged when each estimate \(\bar{Q}\) is an unbiased and confidence valid estimate of the corresponding estimand \(Q\).
+And finally, we will formulate an informed advice on the requirements to safely assume sufficient convergence in practice.
+That is, we assume that the algorithm is sufficiently converged when each estimate \(\bar{Q}\) is an unbiased and confidence valid estimate of the corresponding estimand \(Q\).
 \textbf{{[}and formulate advice on thetas?{]}}
@@
 We assume that we can induce non-convergence at least in conditions where \(T\) is smaller than the default number of iterations in \texttt{mice}, \(T=5\).
+
+\textbf{{[}SEGWAY to this section{]} \(Q\)s are the targets that will be estimated in each simulation repetition, after first \emph{amputing} the complete data with varying missingness proportions, then imputing the missingness with a varying number of iterations, then estimating the \(Q\)s on the \(M\) completed datasets, and finally pooling the \(M\) estimates \(\hat{Q}\) to obtain a single \(\bar{Q}\) for each \(Q\).
+We try to predict the effect of non-convergence (i.e., the difference between \(Q\) and \(\bar{Q}\)) with the convergence diagnostics \(\widehat{R}\) and \(AC\), by applying these methods to several scalar summaries of the state space of the algorithm, \(\theta\)s.}
 
@@
 \item
-  \textbf{{[}Something about thetas, that multivariate thetas are better at detecting non-convergence than univariate thetas.
-The non-convergence diagnostics applied on the chain means and chain variances will fail to identify signs of non-convergence when the multivariate Qs are affected.{]}}
+  We further hypothesize that multivariate \(\theta\)s are better at detecting non-convergence than univariate \(\theta\)s.
 \end{enumerate}
@@
 Therefore, a single complete dataset may serve as comparative truth in all simulation repetitions \citep{vink14}.
-The data-generating mechanism is a multivariate normal distribution, representing person-data on three predictor variables in a multiple linear regression problem \textbf{(from an unspecified social scientific field of study)}.
+The data-generating mechanism is a multivariate normal distribution, representing person-data on three predictor variables in a multiple linear regression problem.
 Let the predictor space be defined as
@@
 \]
-where \(\epsilon \sim \mathcal{N}(0, 100)\).
-From the complete set, we obtain the true values of several quantities of scientific interest, \(Q\)s.
-\textbf{{[}MOVE this next part to the start of the Simulation Study section?{]} \(Q\)s are the targets that will be estimated in each simulation repetition, after first \emph{amputing} the complete data with varying missingness proportions, then imputing the missingness with a varying number of iterations, then estimating the \(Q\)s on the \(M\) completed datasets, and finally pooling the \(M\) estimates \(\hat{Q}\) to obtain a single \(\bar{Q}\) for each \(Q\).
-We try to predict the effect of non-convergence (i.e., the difference between \(Q\) and \(\bar{Q}\)) with the convergence diagnostics \(\widehat{R}\) and \(AC\), by applying these methods to several scalar summaries of the state space of the algorithm, \(\theta\)s.}
+where \(\epsilon \sim \mathcal{N}(0, 100)\). From the complete set, we obtain the true values of several quantities of scientific interest, \(Q\)s.
 
@@
 where \(Y'\) is the expected value of the outcome.
-And finally, the proportion of variance explained by the set of predictors: coefficient of determination \(r^2 \times 100\) \textbf{(note that lower case \(r\) is used to avoid confusion with non-convergence diagnostic \(\widehat{R}\))}.
+And finally, the proportion of variance explained by the set of predictors: coefficient of determination \(r^2\) \textbf{(note that lower case \(r\) is used to avoid confusion with non-convergence diagnostic \(\widehat{R}\))}.
 The true values of these scientific estimands are calculated on the complete data with the \texttt{R} functions \texttt{base::mean()}, \texttt{stats::sd()}, and \texttt{stats::lm()} \citep{R}.
@@
 
-With these performance measures, we also evaluate how well the non-convergence diagnostics \(\widehat{R}\) and \(AC\) can identify simulation conditions in which the \(\bar{Q}\)s are \emph{not} unbiased, confidence valid estimates of the \(Q\)s.
-\textbf{{[}Rephrase this:{]} We measure the performance of \(\widehat{R}\) and \(AC\) by the four scalar summaries \(\theta\): chain means (i.e., the mean in each \(y_{imp, m, j}\)) versus \(Q=\mu\), chain variances (i.e., the variance in each \(y_{imp, m, j}\)) versus \(Q=\sigma\), the estimated regression coefficients \(\beta\) in each \{\(y_{obs}, y_{imp, m}\)\} (i.e., a user-defined statistic of scientific interest \(\hat{Q}\)) against \(Q=\beta\), and the first eigenvalue of the variance-covariance matrix in each \{\(y_{obs}, y_{imp, m}\)\}, \(\lambda_1\) versus \(Q=r^2\).}
+With these performance measures, we also evaluate how well the non-convergence diagnostics \(\widehat{R}\) and \(AC\) can identify simulation conditions in which the inferences are affected by non-convergence, i.e., where \(\bar{Q}\) is \emph{not} an unbiased, confidence valid estimate of \(Q\).
+\textbf{{[}REPHRASE this:{]} For performance of univariate \(Q\)s, we apply the non-convergence diagnostics to univariate \(\theta\)s.
+When \(Q\) is multivariate, we look at signs of non-convergence identified in multivariate \(\theta\)s.}
 
-\textbf{{[}Remove this?{]} Primarily, we want to know what the effect is of inducing non-convergence on the estimated \(Q\)s.
-More specifically, we are interested in the bias of \(\bar{Q}\) for all \(Q\)s and the confidence validity of \(\bar{Q}\) for \(Q=\beta\).
-Therefore, we evaluate these performance measures against the simulation conditions (the number of iterations \(T\) and the missingness proportion \(p_{mis}\)).
-We look at averages across simulation repetitions.
-Secondly, we want to know if conditions in which the \(Q\)s are affected by non-convergence are also identified by the non-convergence diagnostics \(\widehat{R}\) and \(AC\).
-We do this, again, by looking at the performance measures for the \(Q\)s, but now we evaluate these against what \(\widehat{R}\) and \(AC\) indicate about signs of non-convergence in the \(\theta\)s.}
+\begin{itemize}
+\item
+  For \(Q=\mu\), \(\theta\)s are chain means.
+\item
+  For \(Q=\sigma\), \(\theta\)s are chain variances.
+\item
+  For \(Q=\beta\), \(\theta\)s are estimated \(\beta\)s per imputation.
+\item
+  For \(Q=r^2\), \(\theta\)s are the first eigenvalue of the variance-covariance matrix.
+\end{itemize}
 
@@
 
-\textbf{{[}Add to this section: 1) emphasize that the iterations in the plots are averages across repetitions, not within one run of MICE; 2) Add more info about figure legends and axes{]}}
+\textbf{{[}Add more info about figure legends and axes{]}}
 
@@
 Figure \ref{fig:Qs} displays the influence of inducing non-convergence on the estimated quantities of scientific interest, \(Q\)s.
-\textbf{{[}Add panel labels, describe panels{]}} Overall, conditions with a higher proportion of missingness and/or a lower number of iterations are impacted more severely by non-convergence.
-I.e., on average these conditions portray more extreme bias in the estimated \(Q\)s, and non-nominal coverages.
-Roughly speaking, this means that the algorithm is indeed not converged at \(t=1\), and converges gradually as \(t\) increases to \(T\).
+\textbf{{[}Add panel labels, describe panels{]}} Simulation conditions that are impacted by non-convergence portray more extreme bias in the estimated \(Q\)s, and non-nominal coverages.
+Overall, we see that conditions with a higher proportion of missingness are affected more severely by non-convergence.
+Multivariate \(Q\)s are also substantially influenced by the number of iterations.
+This means that the algorithm has indeed not converged at \(t=1\), and converges gradually as \(t\) moves to \(T\).
 The point at which an additional iteration does not lead to an improvement in the estimates depends on quantity of scientific interest \(Q\) and the missingness proportion \(p_{mis}\).
 The highest number of iterations that is necessary to reach this point across all simulation conditions is \(T=7\).
-This implies that, under the current specifications, no more than seven iterations are necessary to obtain \emph{sufficient} performance in terms of bias and confidence validity.
+This implies that, under the current specifications, no more than seven iterations are necessary to obtain sufficient performance in terms of bias and confidence validity.
 
@@
 The estimated univariate means seem unaffected by the number of iterations in the algorithm.
-This implies that \emph{suffient} inferences may be obtained with as little as one iteration (\(T \geq 1\)).
-The magnitude of the bias in \(\bar{Q}\) depends exclusively on the missingness proportion \(p_{mis}\), with missingness proportions of 75\% and 95\% causing more extreme bias than other conditions.
+This implies that valid inferences may be obtained with as little as one iteration (\(T \geq 1\)).
+The magnitude of the bias in \(\bar{Q}\) depends solely on the missingness proportion \(p_{mis}\), with missingness proportions of 75\% and 95\% leading to more extreme biases than other missingness proportions.
 Unbiased estimates are obtained in all conditions where \(p_{mis}\leq.50\).
-Note however, that even in the conditions where \(p_{mis}\geq.75\), the bias is relatively small: \(\mu_Y\) is maximally under-estimated by 0.02 units, while the true value of \(\mu_Y\) is 25.81 (\(\sigma_Y = 11.32\)).
+Note however, that even in the conditions where \(p_{mis}\geq.75\), the bias is small: \(\mu_Y\) is maximally under-estimated by 0.02 units, while the true value of \(\mu_Y\) is 25.81 (\(\sigma_Y = 11.32\)).
 We conclude that non-convergence does not substantially impact the validity of the inferences for \(Q=\mu\).
-\textbf{{[}Add: up-to 50\% missingness everything is relatively unbiased, while 75\% and 95\% missingness show very different results.
-Why is that?
-Add sd of Y in the figure to show that the bias is relative: a bias of 0.02 for a variable with a mean of 25.81, and sd of 11.32 is nothing.{]}}
 
@@
 The estimated standard deviations are largely affected by the missingness proportions, but not to much by the number of iterations.
-\emph{Sufficient} inferences are obtained in conditions with just two iterations.
+Valid inferences are obtained in conditions with just two iterations.
 Interestingly, the bias in first iteration is actually \emph{less} extreme than for any other number of iterations.
-We may therefore conclude that no iteration at all is needed to obtain \emph{sufficient} estimates for \(Q=\sigma\).
+We may therefore conclude that no iteration at all is needed to obtain approximately unbiased estimates for \(Q=\sigma\).
 Completely unbiased estimates are only obtained in conditions where \(p_{mis}=.05\).
 Conditions where \(p_{mis}\geq.25\) are impacted by non-convergence in order of increasing missingness proportion.
+Similar to the bias for \(Q=\mu\) the magnitude of the bias in \(Q=\sigma\) is not substantial in any condition.
+We conclude that both of the univariate \(Q\)s are practically unaffected by non-convergence.
 
@@
 The magnitude of the bias clearly follows the incremental order of the missingness proportions.
-We, therefore, observe \emph{sufficient} estimates at a different number of iterations for each missingness proportion.
+We, therefore, observe approximately unbiased estimates at a different number of iterations for each missingness proportion.
 For \(p_{mis}\leq.25\), the minimum number of iterations is two (\(T\geq2\)), while \(p_{mis}=.50\) requires \(T\geq4\), and \(p_{mis}=.75\) requires \(T\geq5\).
-The highest number of iterations needed to obtain \emph{sufficient} estimates is seven (\(p_{mis}=.95\); \(T\geq7\)).
+The highest number of iterations needed to obtain valid inferences is seven (\(p_{mis}=.95\); \(T\geq7\)).
 Unbiased estimates, however, may only be obtained if the missingness proportion is at most 25\%.
@@
 These approximately nominal coverages are explained by the CIWs: conditions with higher missingness proportions result in more uncertainty in the estimates (i.e., wider CIs), which diminish the effect of the bias.
-\emph{Sufficient} inferences in terms of confidence validity may be obtained when \(T\geq3\), while we need at least seven iterations to reach this point in terms of bias.
+Confidence valid inferences may be obtained when \(T\geq3\), while we need at least seven iterations to reach valid inferences in terms of bias.
 Continued iteration after \(T\geq7\) seems futile.
 
-\hypertarget{non-convergence-diagnostics}{%
-\subsection{Non-convergence diagnostics}\label{non-convergence-diagnostics}}
+\hypertarget{non-convergence-diagnostics-1}{%
+\subsection{Non-convergence diagnostics}\label{non-convergence-diagnostics-1}}
 
@@
 The diagnostics should identify those simulation conditions that are negatively impacted by non-convergence.
-We differentiate between approximately unbiased and confidence valid estimates, and \emph{sufficiently} valid estimates.
-The former is defined by the interplay between \(Q\), \(p_{mis}\), and \(T\).
-The latter is observed in all simulation conditions where \(T\geq7\), irrespective of the \(Q\) and \(p_{mis}\).
-We thus expect the most signs of non-convergence in conditions where \(T<7\).
+We expect the most signs of non-convergence in conditions where \emph{sufficient} estimates are not obtained (high missingness proportions, \(T<7\)), and at least some signs of non-convergence in conditions where estimates are not approximately unbiased (\(p_{mis}=.95\), any number of iterations).
 
@@
 As expected, none of the simulation conditions that we consider resulted in complete convergence of the algorithm (defined as \(\widehat{R}=1\) and \(AC=0\)).
-Yet, the thresholds used in practice to diagnose non-convergence seem too conservative as well.
+However, the thresholds used in practice to diagnose non-convergence seem too conservative for imputation purposes.
 \(\widehat{R}\) and \(AC\) identify signs of non-convergence in conditions where the \(\bar{Q}\)s are unbiased and confidence valid estimates of the \(Q\)s.
@@
 
-When \(\widehat{R}\) and \(AC\) are applied to the chain means, we notice no impact of the missingness proportions. The results only depend on the number of iterations per simulation condition.
-
-\textbf{{[}MAYBE remove all this because it has no relation to the bias in Q{]} The highest \(\widehat{R}\)-values are obtained for the condition \(T=2\).
-We observe the same `dip' in \(\widehat{R}\) as we did in Figure \ref{fig:diagnostics} for conditions with three iterations.
-\(\widehat{R}\) gradually decreases as the number of iterations increases across conditions.
-By the traditional threshold \(\widehat{R}>1.2\) we should conclude that there is non-convergence in conditions where \(T<8\).
-The common threshold \(\widehat{R}>1.1\) is overcome in conditions where \(T>14\).
-}
-
-\textbf{Autocorrelations indicate some signs of non-stationarity in all simulation conditions, but primarily in conditions where \(T<6\).
-The threshold for diagnosing non-convergence is not reached in any condition.
-We would thus conclude that the algorithm is sufficiently converged regardless of the number of iterations or missingness proportion.
-}
-
-We evaluate the diagnostics with chain means as \(\theta\) against the bias in \(Q=\mu\).
-According to the bias in the univariate means, we should conclude that two iterations is sufficient, and we should see that \(p_{mis}=0.75\) and \(p_{mis}=0.95\) have worse convergence than the other missingness proportions.
-This is not the case.
+We evaluate the performance of the non-convergence diagnostics applied to chain means against the bias in \(Q=\mu\).
+Based on the bias in \(Q\), we concluded that one iteration is sufficient to obtain valid inferences, irrespective of the missingness proportion.
+The non-convergence diagnostics, however, identify some signs of non-convergence in conditions with as many as thirty iterations.
+This does not correspond to the impact of non-convergence in the univariate \(Q\)s.
+\textbf{Compared to the multivariate \(Q\)s {[}add before that we evaluate the thetas against all Qs?{]}, \(\widehat{R}\) and \(AC\) applied to the chain means fail to identify the non-convergence in conditions where \(p_{mis}=.95\).}
 
@@
 The non-convergence diagnostics applied to the chain variances as \(\theta\) yield highly similar results to the chain means.
-We therefore evaluate these results immediately against the corresponding \(Q\), without describing the \(\widehat{R}\) and \(AC\) values.
-While the bias in the estimated standard deviations mainly depends on the missingness proportion of each condition, \(\widehat{R}\) and \(AC\) are influenced mostly by the number of iterations.
-We thus conclude that there is no added value of evaluating non-convergence with chain variances as \(\theta\) in the diagnostics.
+\textbf{{[}Segway to performance of \(Q=\sigma\){]} While the bias in the estimated standard deviations mainly depends on the missingness proportion of each condition, \(\widehat{R}\) and \(AC\) are influenced mostly by the number of iterations.
+We thus conclude that there is no added value of evaluating non-convergence with chain variances as \(\theta\) in the diagnostics.}
 
@@
 The conclusion to identify non-stationarity in conditions where \(T\geq10\) and \(p_{mis} = 0.95\) might therefore be spurious.
-\textbf{{[}Also, the \(AC\) across missingness proportions shows the opposite of what you would expect: shouldn't there be higher autocorrelations in conditions with comparatively more observed data in the completed datasets?
-I don't understand this.{]}}
 
-Compared to the bias in \(\bar{Q}\), the diagnostics applied to the \(\hat{Q}\)s over-estimate the severity of the non-convergence. This is odd, since the \(\theta\) and \(Q\) are directly related.
+Compared to the bias in \(\bar{Q}\), the diagnostics applied to the \(\hat{Q}\)s over-estimate the severity of the non-convergence.
 
@@
 
-\hypertarget{in-short-remove-later}{%
-\subsection{In short {[}remove later!{]}}\label{in-short-remove-later}}
-
-\begin{itemize}
-\item
-  Is there more bias in conditions where non-convergence was induced? (NOPE for univariates, YES for multivar.).
-\item
-  Is the point of \emph{sufficient} convergence (at which the \(\bar{Q}\)s are unbiased and confidence valid estimates of the \(Q\)s) the same point as identified by the non-convergence diagnostics?
-(NOPE, they are too conservative).
-\item
-  Is one of the thetas better? (YES, while it doesn't really matter as long as it's multivariate, \(\lambda_1\) is model-independent and therefore preferred).
-\item
-  What should the thresholds be? (much HIGHER than recommended for \(\widehat{R}\), difficult to say for \(AC\)).
-\item
-  And the default nr of iterations? (MAYBE increase to 10 to detect significant \(AC\) and be sure not to miss the dip in \(\widehat{R}\)?).
-\end{itemize}
-
 \hypertarget{discussion}{%
@@
 
+Some other interesting findings are \textbf{{[}REPHRASE{]}}
+
+\begin{itemize}
+\item
+  Is there more bias in conditions where non-convergence was induced? (YES for multivar., NOPE for univariates).
+\item
+  Is the point of \emph{sufficient} convergence (at which the \(\bar{Q}\)s are unbiased and confidence valid estimates of the \(Q\)s) the same point as identified by the non-convergence diagnostics?
+(NOPE, they are too conservative).
+\item
+  Is one of the thetas better? (YES, while it doesn't really matter as long as it's multivariate, \(\lambda_1\) is model-independent and therefore preferred).
+\item
+  What should the thresholds be? (much HIGHER than recommended for \(\widehat{R}\), difficult to say for \(AC\)).
+\item
+  And the default nr of iterations? (MAYBE increase to 10 to detect significant \(AC\) and be sure not to miss the dip in \(\widehat{R}\)?).
+\end{itemize}
+
 \hypertarget{recommendations-for-empirical-researchers}{%
@@
 
-For empirical researchers:
+After imputation it is always wise to take the following steps:
 
@@
 \item
-  Check traceplots for signs of pathological non-convergence. Adjust imputation model if necessary.
+  First, check traceplots for signs of pathological non-convergence. Adjust imputation model if necessary.
 \item
@@
 \item
-  Do not use the traditional calculation for \(\widehat{R}\), or the \texttt{R} function \texttt{stats::acf()} to compute \(AC\).
-Instead, calculate \(\widehat{R}\) conform \citet{veht19} and compute autocorrelations manually (see e.g., \href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy}{github.com/gerkovink/shinyMice}).
+  Track \(\theta\)s over iterations, e.g., the novel \(\theta\) that is `substantive model-independent'.
 \item
@@
 \item
-  \textbf{Something} about the novel \(\theta\) that is `substantive model-independent'.
+  Do not use the traditional calculation for \(\widehat{R}\), or the \texttt{R} function \texttt{stats::acf()} to compute \(AC\).
+Instead, calculate \(\widehat{R}\) conform \citet{veht19} and compute autocorrelations manually (see e.g., \href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy}{github.com/gerkovink/shinyMice}).
 \end{enumerate}
@@
 
+Much remains unknown.
+about non-convergence in iterative imputation algorithms.
 Further research is needed to investigate the performance of iterative imputation algorithms under clear violation of convergence, e.g.~dependency between predictors (predictors with very high correlations).
@@
 \item
-  Implement recommendations in ShinyMICE.
+  Implement recommendations in imputation software, e.g., evaluation environment for applied researchers.
 \item

### 2020-05-07 3.Thesis/Figures/DiagramMI.Rnw
Feedback Stef

--- 2020-04-27 3.Thesis/1.SimulationStudy/Writeup/images/diagram.Rnw
@@
 \begin{tikzpicture}
-\tikzstyle{main}=[rectangle, minimum width = 21mm, minimum height = 12mm, thick, draw =black!80, node distance = 17mm]
-\tikzstyle{connect}=[draw, -latex, thick]
-\node[main, fill = white!100] (data) [label=below:Incomplete data] {$\{y_{obs}, y_{mis}\}$};
-\node[main] (mids) [right=of data] { $\{y_{obs}, y_{imp}\}$};
-\node[main] (mids2) [above=10mm of mids] {$\{y_{obs}, y_{imp}\}$ };
-\node[main] (mids3) [below=10mm of mids,label=below: $M$ completed datasets] { $\{y_{obs}, y_{imp}\}$};
-\node[main] (mira) [right=of mids] {\large $\hat{Q}$};
-\node[main] (mira2) [above=10mm of mira] {\large $\hat{Q}$ };
-\node[main] (mira3) [below=10mm of mira,label=below: $M$ results] {\large $\hat{Q}$ };
-\node[main, fill = white!100] (mipo) [right=of mira,label=below:Pooled results] {\large $\bar{Q}$ };
-\path (data) edge [connect] (mids)
-      (data) edge [connect] (mids2)
-      (data) edge [connect] (mids3)
-      (mids) edge [connect] (mira)
-      (mids2) edge [connect] (mira2)
-      (mids3) edge [connect] (mira3)
-		  (mira) edge [connect] (mipo)
-		  (mira2) edge [connect] (mipo)
-		  (mira3) edge [connect] (mipo);
+\tikzstyle{main}=[rectangle, minimum width = 22mm, minimum height = 9mm, thick, draw =black!80, node distance = 14mm]
+\tikzstyle{connect}=[draw, -latex, thick, out=180,in=270, looseness=0]
+\node[main, fill = white!100] (data) [label=below:Incomplete data] {$\{y_{\rm obs}, y_{\rm mis}\}$};
+\node[main] (mids) [right=of data] { $\{y_{\rm obs}, \dot y_{\rm imp, \ell}\}$};
+\node[main] (mids2) [above=10mm of mids] {$\{y_{\rm obs}, \dot y_{\rm imp, \ell}\}$ };
+\node[main] (mids3) [below=10mm of mids,label=below: $m$ completed datasets] { $\{y_{\rm obs}, \dot y_{\rm imp, \ell}\}$};
+\node[main] (mira) [right=of mids] {$\hat{Q}_\ell$};
+\node[main] (mira2) [above=10mm of mira] { $\hat{Q}_\ell$ };
+\node[main] (mira3) [below=10mm of mira,label=below: $m$ results] { $\hat{Q}_\ell$ };
+\node[main, fill = white!100] (mipo) [right=of mira,label=below:Pooled result] {$\bar{Q}$ };
+\path      
+    (data.east) edge [connect] node [right] {} (mids.west)
+(data.east) edge [connect] node [right] {} (mids2.west)
+(data.east) edge [connect] node [right] {} (mids3.west)
+      (mids.east) edge [connect] node [right] {} (mira.west)
+      (mids2.east) edge [connect] node [right] {} (mira2.west)
+      (mids3.east) edge [connect] node [right] {} (mira3.west)
+		(mira.east) edge [connect] node [right] {} (mipo.west)
+		(mira2.east) edge [connect] node [right] {} (mipo.west)
+		(mira3.east) edge [connect] node [right] {} (mipo.west)
+;
 \end{tikzpicture}

### 2020-05-07 3.Thesis/Writeup/2.Manuscript.Rmd
Feedback Stef

@@
 Missing data is not only ubiquitous across science, but treating it can also be a tedious task.
-If a dataset contains just one incomplete observation, statistical inferences are undefined and will not produce any results.
+If a dataset contains just one incomplete observation, calculations are not defined, **hence** statistical models cannot do fitted to the data.
 To circumvent this, many statistical packages employ list-wise deletion by default (i.e., ignoring incomplete observations).
-Unfortunately, this *ad hoc* solution may yield wildly invalid results [@buur18].
-An alternative is to *impute* (i.e., fill in) the missing values in the incomplete observations.
-Subsequently, statistical inferences can be performed on the completed dataset.
+Unfortunately, this *ad hoc* solution may yield invalid results [@buur18].
+An alternative is to apply imputation.
+With imputation, we 'fill in' the missing values in an incomplete dataset.
+Subsequently, the model of scientific interest can be fitted to the completed dataset.
 By repeating this process several times, a distribution of plausible results may be obtained, which reflects the uncertainty in the data due to missingness.
@@
 Figure \ref{fig:diagram} provides an overview of the steps involved with MI.
-The missing part $y_{mis}$ of an incomplete dataset is imputed $M$ times.
-The imputed data $y_{imp}$ is combined with the observed data $y_{obs}$ to create $M$ completed datasets.
-On each completed dataset, an analysis of scientific interest is performed.
-The quantity of scientific interest (e.g., a regression coefficient) is denoted with $Q$.
-Since $Q$ is estimated on each completed dataset, $M$ separate $\hat{Q}$-values are obtained.
-These $M$ values are combined into a single pooled estimate $\bar{Q}$.
-The premise of multiple imputation is that $\bar{Q}$ is an unbiased and confidence valid estimate of the true---but unobserved---inference.
+The missing part $y_{\rm mis}$ of an incomplete dataset is imputed $m$ times.
+This creates $m$ sets of imputed data $\dot y_{\rm imp, \ell}$, where $\ell = 1, 2, ..., m$.
+The imputed data is then combined with the observed data $y_{\rm obs}$ to create $m$ completed datasets.
+On each of these datasets the analysis of scientific interest is performed to estimate $Q$: the quantity of scientific interest (e.g., a regression coefficient).
+Since $Q$ is estimated on each completed dataset, $m$ separate $\hat{Q}_\ell$-values are obtained.
+Finally, the $m$ $\hat{Q}_\ell$-values are combined into a single pooled estimate $\bar{Q}$.
+The premise of multiple imputation is that $\bar{Q}$ is an unbiased and confidence-valid estimate of the true---but unobserved---inference [@rubin96].
 
@@
 
-```{r diagram, out.width="\\linewidth", include=TRUE, fig.cap="Scheme of the main steps in multiple imputation (where $M=3$)---from an incomplete dataset, to $M$ multiply imputed datasets, to $M$ estimated quantities of scientific interest $\\hat{Q}$s, to a single pooled estimate $\\bar{Q}$."}
-knitr::include_graphics("../Figures/diagram.pdf") #adapted from [@buur18] par. 1.4.1.
+```{r diagram, out.width="\\linewidth", include=TRUE, fig.cap="Scheme of the main steps in multiple imputation (where $m=3$)---from an incomplete dataset, to $m$ multiply imputed datasets, to $m$ estimated quantities of scientific interest $\\hat{Q}$s, to a single pooled estimate $\\bar{Q}$."}
+knitr::include_graphics("../Figures/DiagramMI.pdf") #adapted from [@buur18] par. 1.4.1.
 ```
@@
 A popular method to obtain imputations is to use the 'Multiple Imputation by Chained Equations' algorithm, shorthand 'MICE'[@mice].
-With MICE, imputed values $y_{imp}$ are drawn from the posterior predictive distribution of the missing values $y_{mis}$.
+With MICE, imputed values $\dot y_{\rm imp, \ell}$ are drawn from the posterior predictive distribution of the missing values $y_{\rm mis}$.
 The algorithm is named after the iterative algorithmic procedure by which imputed values are generated: a multivariate distribution is obtained by iterating over a sequence of univariate imputations.
 The iterative nature of algorithms like MICE introduces a potential threat to the validity of the imputations: What if the algorithm has not converged?
-Are the implications then to be trusted?
+Are the imputations then to be trusted?
 And can we rely on the inference obtained on the completed data?
 
-These remain open questions, because the convergence properties of iterative imputation algorithms have not been systematically studied [@buur18].
-Moreover, there is no scientific consensus on how to evaluate convergence of imputation algorithms [@taka17].
-Some default MICE techniques (e.g., 'predictive mean matching') might not yield converged states at all [@murr18].
-Therefore, algorithmic convergence should be monitored carefully.
+These remain open questions, since the convergence properties of iterative imputation algorithms have not been systematically studied [@buur18].
+There is no scientific consensus on how to evaluate convergence of imputation algorithms [@taka17].
+**Moreover, the convergence of chained equations algorithms such as MICE under certain default imputation models (e.g., 'predictive mean matching') is an entirely open question [@murr18].** Therefore, algorithmic convergence should be monitored carefully.
 
@@
 Typically, the $\theta$s under evaluation are chain means and variances---the average and the variance of the imputed values per imputation.
-Monitoring convergence by inspecting traceplots of these $\theta$s is insufficient on two counts: 1) it may be challenging to the untrained eye, and 2) only severely pathological cases of non-convergence may be diagnosed [@buur18, $\S$ 6.5.2] **[Add that convergence of univariate thetas cannot guarantee multivariate convergence?]**.
+Monitoring convergence by inspecting traceplots of these $\theta$s may be undesirable for several reasons on two counts: 1) it may be challenging to the untrained eye, and 2) only severely pathological cases of non-convergence may be diagnosed [@buur18, $\S$ 6.5.2] and, 3) there is not an objective measure that quantifies convergence.
 Therefore, a quantitative, diagnostic evaluation of convergence would be preferred---although not straightforward.
-Iterative imputation algorithms such as MICE are Markov chain Monte Carlo (MCMC) methods.
+Iterative imputation algorithms such as MICE are special cases of Markov chain Monte Carlo (MCMC) methods.
 In MCMC methods, convergence is not from a scalar to a point, but from one distribution to another.
@@
 
-In this paper, we investigate how non-convergence in iterative imputation algorithms may be diagnosed.
-First, we define which non-convergence diagnostics will be considered, and evaluate how these diagnostics could be appropriate for iterative imputation applications.
-In the second part of this paper, we will investigate the impact of inducing non-convergence in iterative imputation algorithms by means of model-based simulation in `R` [@R].
-The aim of the simulation is to determine whether unbiased, confidence valid inferences may be obtained under different levels of non-convergence (**and at which point convergence may safely be assumed**).
+In this paper, we study different methods for assessing non-convergence in iterative imputation algorithms.
+We define several diagnostics and evaluate how these diagnostics could be appropriate for iterative imputation applications.
+We then adress the impact of inducing non-convergence in iterative imputation algorithms by means of model-based simulation in `R` [@R].
+The aim of the simulation is to determine whether unbiased, confidence-valid inferences may be obtained under different levels of non-convergence (**and at which point convergence may safely be assumed**).
 Additionally, we will assess how well different non-convergence diagnostics perform in identifying signs of non-convergence (**and on which scalar summary of the state space of the algorithm these diagnostics should be applied**).
-The results of this study are guidelines for identifying non-convergence in iterative imputation algorithms, which will aid applied researchers in drawing valid inference from incomplete datasets.
+We translate the results of the study into guidelines for applied researchers.
 **[add more societal relevance: what is the consequence of non-convergence?
 see example of non-convergence in Figure 2 (e.g., biased estimates, invalid inference, etc.)]** For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular `mice` package [@mice] in `R` [@R].
+
+<!-- The results of this study are guidelines for identifying non-convergence in iterative imputation algorithms, which will aid applied researchers in drawing valid inference from incomplete datasets.
+-->
 
@@
 The data value of unit $i$ ($i = 1, 2, \dots, n$) on variable $j$ ($j = 1, 2, \dots, k$) may be either observed or missing.
-The number of units $i$ in dataset $y$ with at least one missing value, divided by the total number of units $n$, is called the missingness proportion $p_{mis}$.
-The collection of observed data values in $y$ is denoted by $y_{obs}$; the missing part of $y$ is referred to as $y_{mis}$.
-For each datapoint in $y_{mis}$, we sample $M \times T$ plausible values, where $M$ is the number of imputations ($m = 1, 2, \dots, M$) and $T$ is the number of iterations ($t = 1, 2, \dots, T$) in the imputation algorithm.
+The number of units $i$ in dataset $y$ with at least one missing value, divided by the total number of units $n$, is called the missingness proportion $p_{\rm mis}$.
+The collection of observed data values in $y$ is denoted by $y_{\rm obs}$; the missing part of $y$ is referred to as $y_{\rm mis}$.
+For each datapoint in $y_{\rm mis}$, we sample $m \times T$ plausible values, where $m$ is the number of imputations ($\ell = 1, 2, \dots, m$) and $T$ is the number of iterations ($t = 1, 2, \dots, T$) in the imputation algorithm.
 The collection of samples between the initial value (at $t=1$) and the final imputed value (at $t=T$) will be referred to as an 'imputation chain'.
@@
 Without stationarity, there is trending within imputation chains.
-Trending implies that further iterations would yield a different set of imputations.
+Trending implies that further iterations would yield a systematically lower or higher set of imputations.
 Iterative imputation algorithms that have not yet reached stationarity, may thus yield biased estimates.
@@
 Figure \ref{fig:non-conv} displays a subset of the example: a variable that we will call $j$.
-We see two traceplots of the chain means for $j$---i.e., the average of the imputed values for $j$ in each imputation $y_{imp,m}$, plotted against the iteration number.
+We see two traceplots of the chain means for $j$---i.e., the average of the imputed values for $j$ in each imputation $\dot y_{\rm imp, \ell}$, plotted against the iteration number.
 Each line portrays one imputation.
@@
 **In the typical convergence situation, the imputation chains intermingle nicely and there is little to no trending.
-In the non-convergence plot, there is a lot of trending and some chains do not intermingle.** Importantly, the chain means at the last iteration (the imputed value per $m$) are very different between the two plots.
+In the non-convergence plot, there is a lot of trending and some chains do not intermingle.** Importantly, the chain means at the last iteration (the imputed value per imputation $\ell$) are very different between the two plots.
 The algorithm with the mis-specified model yields imputed values that are on average a factor two larger than those of the typically converged algorithm.
-That means that non-convergence has a impact on the mean of $j$ in $y_{imp, m}$.
-This difference (presumably) translates to bias in the $M$ completed data estimates $\hat{\mu}_{j,m}$, and consequently also to the pooled estimate $\bar{\mu}_j$.
+That means that non-convergence has a impact on the mean of $j$ in $\dot y_{\rm imp, \ell}$.
+This difference (presumably) translates to bias in the $m$ completed data estimates $\hat{\mu}_{j,\ell}$, and consequently also to the pooled estimate $\bar{\mu}_j$.
 Non-convergence thus leads to biased estimates.
@@
 **[MERGE with section below]**.
-Inspecting non-convergence through traceplots of chain means and chain variances may be insufficient because they are summaries of the univariate state space of the algorithm, while MICE is not only concerned with a sigle column, but the entire multivariate distribution of $y_{imp}$.
+Inspecting non-convergence through traceplots of chain means and chain variances may be insufficient because they are univariate summaries of the state space of the algorithm, while MICE is not only concerned with a single column, but the entire multivariate distribution of $y_{\rm imp}$.
 Ideally, a multivariate $\theta$ should be monitored.
 A suggestion by @buur18 for a multivariate $\theta$ to monitor is the quantity of scientific interest $Q$ (e.g., a regression coefficient), which usually is multivariate in nature.
-Implementing this, however, might be somewhat advanced for empirical researchers.
+Implementing this, however, might be somewhat too technical for empirical researchers.
 Moreover, this scalar summary is not model-independent, i.e., it only applies to one model of scientific interest.
@@
 Eigenvalues of a variance-covariance matrix are a measure of the data's total covariance.
-Such a $\theta$ would be a model-independent, multivariate scalar summary to monitor, but it has not yet been implemented.
+Such a $\theta$ would be a model-independent, multivariate scalar summary to monitor, but it is not implemented in `mice`.
 <!-- Recommendation: "the different streams should be freely intermingled with one another, without showing any definite trends". -->
@@
 $\theta$s are scalar summaries of the state space of the algorithm at a specific iteration.
-The recommended scalar summaries to evaluate are chain means and chain variances.
+The default scalar summaries in `mice` are chain means and chain variances.
 Additionally, researchers may "monitor some statistic of scientific interest" [@buur18, $\S$ 6.5.2].
@@
 Therefore, we propose a $\theta$ that summarizes the multivariate state space of the algorithm, but is independent from the model of scientific interest.
-We propose $\lambda_1$ as such a scalar summary.
-We define $\lambda_1$ as the first eigenvalue of the variance-covariance matrix of the completed data.
-Let $\lambda_1 \geq \lambda_2 \geq ...
-\geq \lambda_j$ be the eigenvalues of $\Sigma$ in each of the $M$ completed datasets ${y_{obs}, y_{imp}}$.
-$\lambda_1$ is measure that summarizes the covariances in the completed datasets.
+We propose $\lambda_{1, \ell}$ as such a scalar summary.
+We define $\lambda_{1, \ell}$ as the first eigenvalue of the variance-covariance matrix of the completed data.
+Let $\lambda_{1, \ell} \geq \lambda_{2, \ell} \geq ...
+\geq \lambda_{j, \ell}$ be the eigenvalues of $\Sigma_\ell$ in each of the $m$ completed datasets {$y_{\rm obs}, \dot y_{\rm imp, \ell}$}.
+$\lambda_{1, \ell}$ is measure that summarizes the covariances in the completed datasets.
 **The first eigenvalue has the appealing property that is not dependent on the substantive model of interest.
@@
 Because of the issues mentioned above, we propose a novel $\theta$ to monitor: the first eigenvalue of the variance-covariance matrix of the completed data.
-Let $\lambda_1 \geq \lambda_2 \geq ...
-\geq \lambda_j$ be the eigenvalues of $\Sigma$ in each of the $M$ completed datasets ${y_{obs}, y_{imp}}$.
-$\lambda_1$ is measure that summarizes the covariances in the completed datasets.**  -->
+Let $\lambda_{1, \ell} \geq \lambda_2 \geq ...
+\geq \lambda_j$ be the eigenvalues of $\Sigma$ in each of the $m$ completed datasets ${y_{\rm obs}, y_{\rm imp}}$.
+$\lambda_{1, \ell}$ is measure that summarizes the covariances in the completed datasets.**  -->
 
@@
 With a recently proposed adaptation, $\widehat{R}$ might also serve to diagnose non-stationarity, but this has not yet been thoroughly investigated [@veht19].
-Therefore, we use $\widehat{R}$ and $AC$ to evaluate mixing and stationarity separately, as recommended by e.g., @cowl96.
+Therefore, we use $\widehat{R}$ **[WHICH?
+Explain that we'll use both]** and $AC$ to evaluate mixing and stationarity separately, as recommended by e.g., @cowl96.
 <!-- @cowl96, p. 898. -->
@@
 version of $\widehat{R}$ may be suitable for iterative imputation.
-We therefore use their definition of the diagnostic.** Let $M$ be the total number of chains, $T$ the number of iterations per chain, and $\theta$ the scalar summary of interest (e.g., chain means).
-For each chain ($m = 1, 2, \dots, M$), we estimate the variance of $\theta$, and average these to obtain within-chain variance $W$.
+We therefore use their definition of the diagnostic.** Let $m$ be the total number of chains, $T$ the number of iterations per chain ($T\geq2$), and $\theta$ the scalar summary of interest (e.g., chain means).
+For each chain ($\ell = 1, 2, \dots, m$), we estimate the variance of $\theta$, and average these to obtain within-chain variance $W$.
 
 \begin{align*}
-W&=\frac{1}{M} \sum_{m=1}^{M} s_{j}^{2}, \text { where } s_{m}^{2}=\frac{1}{T-1} \sum_{t=1}^{T}\left(\theta^{(t m)}-\bar{\theta}^{(\cdot m)}\right)^{2}. 
+W&=\frac{1}{m} \sum_{\ell=1}^{m} s_{j}^{2}, \text { where } s_{\ell}^{2}=\frac{1}{T-1} \sum_{t=1}^{T}\left(\theta^{(t \ell)}-\bar{\theta}^{(\cdot \ell)}\right)^{2}. 
 \end{align*}
@@
 
-\noindent We then estimate between-chain variance $B$ as the variance of the collection of average $\theta$ per chain.
+\noindent We then estimate between-chain variance $B$ as the variance of the collection of average $\theta$ per chain. **[Note that this is not the usual B in MI]**
 
 \begin{align*}
-B&=\frac{T}{M-1} \sum_{m=1}^{M}\left(\bar{\theta}^{(\cdot m)}-\bar{\theta}^{(\cdot \cdot)}\right)^{2}, \text { where } \bar{\theta}^{(\cdot m)}=\frac{1}{T} \sum_{t=1}^{T} \theta^{(t m)} \text{, } \bar{\theta}^{(\cdot \cdot)}=\frac{1}{M} \sum_{m=1}^{M} \bar{\theta}^{(\cdot m)}.
+B&=\frac{T}{m-1} \sum_{\ell=1}^{m}\left(\bar{\theta}^{(\cdot \ell)}-\bar{\theta}^{(\cdot \cdot)}\right)^{2}, \text { where } \bar{\theta}^{(\cdot \ell)}=\frac{1}{T} \sum_{t=1}^{T} \theta^{(t \ell)} \text{, } \bar{\theta}^{(\cdot \cdot)}=\frac{1}{m} \sum_{\ell=1}^{m} \bar{\theta}^{(\cdot \ell)}.
 \end{align*}
@@
 High $\widehat{R}$-values thus indicate non-convergence.
-The conventionally acceptable threshold for convergence was $\widehat{R} < 1.2$ [@gelm92].
-More recently, @veht19 proposed a more stringent threshold of $\widehat{R} < 1.01$.
 
@@
 We can interpret $AC$-values as a measure of stationarity.
-If $AC$-values are close to zero, there is no dependence between subsequent samples within imputation chains.
-Negative $AC$-values indicate divergence within imputation chains.
-**Subsequent sampled values within each imputation chain are less alike.** Positive $AC$-values indicate recurrence.
-If $\theta$-values of subsequent iterations are similar, trending may occur.
+$AC$-values close to zero indicate  independence between subsequent $\theta$s in imputation chains.
 Negative $AC$-values show no threat to the stationarity component of convergence.
@@
 As convergence diagnostic, the interest is therefore in positive $AC$-values.
+Positive $AC$-values occur when $\theta$-values of subsequent iterations are similar.
+When high $\theta$-values are  followed by high $\theta$-values, and low $\theta$-values are followed by low $\theta$-values, we speak of recurrence.
+Recurrence within imputation chains indicates trending.
+Ergo, non-stationarity may be diagnosed when $AC$ values are larger than zero.
+
+<!-- Negative $AC$-values indicate divergence within imputation chains.
+**Subsequent sampled values within each imputation chain are less alike.** Positive $AC$-values indicate recurrence.
+If $\theta$-values of subsequent iterations are similar, trending may occur.
+-->
+
+## Thresholds
+
+<!-- We assume that complete convergence, defined as $\widehat{R} = 1$ and $AC = 0$, will not be observed.  -->
+It is unlikely that an MCMC algorithm will achieve the ideal values of $\widehat{R}=1$ and $AC=0$.
+Because of its convergence to a distribution, the algorithm will show some signs of non-mixing and non-stationarity even in the most converged state.
+Upon approximate convergence, the imputation chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm ($\widehat{R} > 1$), and there is some dependency between subsequent iterations of imputation chains ($AC > 0$).
+In practice, non-convergence is therefore diagnosed when non-convergence diagnostics exeed a certain threshold.
+The conventional threshold to diagnose non-mixing is $\widehat{R} > 1.2$ [@gelm92].
+Another generally accepted threshold is $\widehat{R} > 1.1$ [@gelm13].
+And more recently, @veht19 proposed an even more stringent threshold of $\widehat{R} > 1.01$.
 The magnitude of $AC$-values may be evaluated statistically, using a Wald test with $AC=0$ as null hypothesis [@box15].
-
-## Thresholds
-
-<!-- We assume that complete convergence, defined as $\widehat{R} = 1$ and $AC = 0$, will not be observed.  -->
-Upon convergence, $\widehat{R}=1$ and $AC=0$, which are unlikely thresholds for MCMC algorithms, because of its convergence to a distribution.
-Even in the most converged state, the algorithm will show some signs of non-mixing and non-stationarity.
-Upon *sufficient* convergence, the imputation chains will intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm ($\widehat{R} > 1$), and there may be some dependency between subsequent iterations of imputation chains ($AC > 0$).
-In practice, non-convergence is usually diagnosed when $\widehat{R}$ > 1.2 or 1.1 or even 1.01.
-And a t-test is performed to assess whether $AC$ is significantly different from zero.
+$AC$-values that are significantly higher than zero indicate non-stationarity.
+**[think about one-sided test]**
 
@@
 
-```{r diagnostics, fig.cap="Convergence diagnostics applied on the imputation algorithms of Figure 2. $\\theta$ = chain mean in $y_{imp, m}$."}
+```{r diagnostics, fig.cap="Convergence diagnostics applied on the imputation algorithms of Figure 2. $\\theta$ = chain mean in $ \\dot y_{\\rm imp, \\ell} $."}
 # load("./images/diagnostics_plot.Rdata") #bquote(theta ~ "= chain mean in "~ y[imp][",m"])
@@
 
-**Describe panel C here and why it is not optimal]** The $\widehat{R}$-values in panel D do meet the requirements specified above.
+**Describe panel C here and why it is not optimal: NB.
+In panel C: De grotere variantie overall zorgt er kennelijk voor dat $\widehat{R}$ groeit.
+De $\widehat{R}$ geeft de foutieve boodschap dat convergentie steeds slechter gaat.
+Verklaring: de overdispersie assumptie klopt niet als er trend is.]** The $\widehat{R}$-values in panel D do meet the requirements specified above.
 $\widehat{R}$ as computed conform @veht19 does indeed indicate less signs of non-convergence as the number of iterations goes up.
@@
 
-The aim of the simulation study is to evaluate the impact of inducing non-convergence in the MICE algorithm on several quantities of scientific interest $Q$.
+The aim of the simulation study is to evaluate the **impact of inducing non-convergence ["early stopping" en "proportion of complete cases"]** in the MICE algorithm on several quantities of scientific interest $Q$.
 Subsequently we will evaluate how well $\widehat{R}$ and $AC$ perform in identifying the effects of non-convergence **[and performance of different thetas]**.
-And finally, we will formulate an informed advice on the requirements to safely assume sufficient convergence in practice.
-That is, we assume that the algorithm is sufficiently converged when each estimate $\bar{Q}$ is an unbiased and confidence valid estimate of the corresponding estimand $Q$.
+And finally, we will formulate an informed advice on the requirements to safely conclude sufficient convergence in practice.
+That is, we conclude that  convergence is sufficient when each estimate $\bar{Q}$ is an unbiased and confidence-valid estimate of the corresponding estimand $Q$.
 **[and formulate advice on thetas?]**
 
-Non-convergence will be induced by: 1) increasing the missingness proportion $p_{mis}$ in dataset $y$ incrementally, and 2) terminating the imputation algorithm at a varying number of iterations $T$.
+Non-convergence will be induced by: 1) increasing the missingness proportion $p_{\rm mis}$ in dataset $y$ incrementally, and 2) terminating the imputation algorithm at a varying number of iterations $T$.
 The first set of simulation conditions---the missingness proportions---is chosen to reflect the difficulty of the missingness problem.
@@
 Then, convergence to a stable but highly variable state may be reached instantaneously.
-We assume that the incremental missingness proportions in our study will result in a corresponding increase in signs of non-convergence.
+We expect that the incremental missingness proportions in our study will result in a corresponding increase in signs of non-convergence.
 
@@
 
-**[SEGWAY to this section] $Q$s are the targets that will be estimated in each simulation repetition, after first *amputing* the complete data with varying missingness proportions, then imputing the missingness with a varying number of iterations, then estimating the $Q$s on the $M$ completed datasets, and finally pooling the $M$ estimates $\hat{Q}$ to obtain a single $\bar{Q}$ for each $Q$.
+**[SEGWAY to this section] $Q$s are the targets that will be estimated in each simulation repetition, after first *amputing* the complete data with varying missingness proportions, then imputing the missingness with a varying number of iterations, then estimating the $Q$s on the $m$ completed datasets, and finally pooling the $m$ estimates $\hat{Q}_\ell$ to obtain a single $\bar{Q}$ for each $Q$.
 We try to predict the effect of non-convergence (i.e., the difference between $Q$ and $\bar{Q}$) with the convergence diagnostics $\widehat{R}$ and $AC$, by applying these methods to several scalar summaries of the state space of the algorithm, $\theta$s.**
@@
 1.
-We expect that simulation conditions with a high missingness proportion $p_{mis}$ and a low number of iterations $T$ will result in biased, invalid estimates of the quantities of scientific interest, $Q$s.
+We expect that simulation conditions with a high missingness proportion $p_{\rm mis}$ and a low number of iterations $T$ will more often result in biased, invalid estimates of the quantities of scientific interest, $Q$s.
 
 2.
-We hypothesize that $\widehat{R}$ and $AC$ will correctly identify signs of non-convergence in those simulation conditions where the $\bar{Q}$s are *not* unbiased and confidence valid estimates of the $Q$s.
+We hypothesize that $\widehat{R}$ and $AC$ will correctly identify signs of non-convergence in those simulation conditions where the $\bar{Q}$s are *not* unbiased and confidence-valid estimates of the $Q$s.
 
@@
 The simulation set-up is summarized in the pseudo-code below.
-The complete `R` script of the simulation study is available from [github.com/gerkovink/shinyMice](https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy).
+The complete `R` script of the simulation study is available from [github.com/gerkovink/shinyMice](https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/).
 
@@
 \begin{pmatrix}
-4 & 4 & 1.8 & 0\\
-4 & 16 & 4.8 & 0\\
-1.8 & 4.8 & 9 & 0
+4 & 4 & 1.8 \\
+4 & 16 & 4.8\\
+1.8 & 4.8 & 9
 \end{pmatrix}
@@
 $$
-A finite population of $N=1000$ is simulated using the `mvtnorm` package [@mvtnorm]. Subsequently, a fourth variable is constructed as outcome $Y$. For each unit $i = 1, 2,..., N$, let 
+A finite population of $N=1000$ is simulated using the `mvtnorm` package [@mvtnorm]. Subsequently, a fourth variable is constructed as outcome variable $Y$. For each unit $i = 1, 2,..., N$, let 
 $$
@@
 $$
-where $\epsilon \sim \mathcal{N}(0, 100)$. From the complete set, we obtain the true values of several quantities of scientific interest, $Q$s.
+where $\epsilon \sim \mathcal{N}(0, 100)$.
+This results in a complete set $y$, with observed values for all units $i$ on all variables $j$ ($j =  Y, X_1, X_2, X_3$).
+From the complete set we obtain the true values of the scientific estimands $Q$.
 
@@
 And finally, the proportion of variance explained by the set of predictors: coefficient of determination $r^2$ **(note that lower case $r$ is used to avoid confusion with non-convergence diagnostic $\widehat{R}$)**.
-The true values of these scientific estimands are calculated on the complete data with the `R` functions `base::mean()`, `stats::sd()`, and `stats::lm()` [@R].
-<!-- Each of these $Q$s is estimated with a corresponding $\bar{Q}$, after the complete dataset is first *amputed* (with a varying missingness proportion $p_{mis}$) and subsequently imputed (with a varying number of iterations $T$).
+The true values of these scientific estimands are calculated on the complete data.
+<!-- Each of these $Q$s is estimated with a corresponding $\bar{Q}$, after the complete dataset is first *amputed* (with a varying missingness proportion $p_{\rm mis}$) and subsequently imputed (with a varying number of iterations $T$).
 -->
@@
 In each simulation repetition, we *ampute* the complete dataset five times to obtain five different incomplete datasets, $y$s.
-The ratio between $y_{obs}$ and $y_{mis}$ in each $y$ depends on the missingness proportion: $p_{mis} =.05,.25,.5,.75,.95$.
+The ratio between $y_{\rm obs}$ and $y_{\rm mis}$ in each $y$ depends on the missingness proportion: $p_{\rm mis} =.05,.25,.5,.75,.95$.
 We ampute the complete data with the `mice` package [function `mice::ampute()`; @mice].
 We consider all possible univariate and multivariate patterns of missingness, conform a 'missing completely at random' missingness mechanism [@rubin87].
-I.e., the probability to be missing is the same for all $N \times k$ cells in $y$, conditional on the missingness proportion $p_{mis}$.
+I.e., the probability to be missing is the same for all $N \times k$ cells in $y$, conditional on the missingness proportion $p_{\rm mis}$.
 <!-- This note only considers a 'missing completely at random' (MCAR) mechanism, where the probability of being missing is equal for all $n \times k$ cells in $y$ [@rubin87].
@@
 Missing datapoints in each incomplete dataset $y$ are imputed with the `mice()` function [@mice].
-All imputation procedures are performed with Bayesian linear regression imputation, and five imputation chains ($M=5$).
-Because $M=5$, each run of the `mice()` function results in five sets of imputations: $y_{imp,m}$, where $m = 1, 2, ..., 5$.
+All imputation procedures are performed with Bayesian linear regression imputation, and five imputation chains ($m=5$).
+Because $m=5$, each run of the `mice()` function results in five sets of imputations: $\dot y_{\rm imp, \ell}$, where $m = 1, 2, ..., 5$.
 The number of iterations in each run of the algorithm varies between simulation conditions ($T=1,2,...,100$).
-<!-- Each repetition of the simulation thus starts from 1 complete dataset, then splits into 5 missingness conditions ($p_{mis} =.05,.25,.5,.75,.95$), each of the amputed datasets is imputed 5 times ($m = 1, 2,..., 5$) for each iteration condition ($t = 1, 2,..., 100$).
+<!-- Each repetition of the simulation thus starts from 1 complete dataset, then splits into 5 missingness conditions ($p_{\rm mis} =.05,.25,.5,.75,.95$), each of the amputed datasets is imputed 5 times ($m = 1, 2,..., 5$) for each iteration condition ($t = 1, 2,..., 100$).
 From these $5\times5\times100$ imputations, we extract the $\theta$s to apply $\widehat{R}$ and $AC$ on (**i.e., chain means, chain variances, $\beta$s per imputation, and the first eigenvalue of the variance-covariance matrix per imputation**).
@@
 
-From the 5 imputations $y_{imp,m}$ we obtain the chain means and chain variances for each variable $j$.
+From the 5 imputations $\dot y_{\rm imp, \ell}$ we obtain the chain means and chain variances for each variable $j$.
 These will serve as scalar summary $\theta$ to apply non-convergence diagnostics $\widehat{R}$ and $AC$ on.
-We then use `mice::complete()` to combine the imputed data $y_{imp,m}$ with the observed data $y_{obs}$, resulting in five completed datasets {$y_{obs}, y_{imp, m}$}.
-
-On the 5 completed datasets {$y_{obs}, y_{imp, m}$}, we:
-
-- Estimate $\hat{Q}$s for $Q= \mu$ and $Q=\sigma$ with `base::mean()`, and `stats::sd()` [@R]. We pool the $\hat{Q}$s with `base::mean()` to get $\bar{Q}$s.
-
-- Compute $\lambda_1$, to use as $\theta$ in the two convergence diagnostics.
-By definition, the first eigenvalue of the variance-covariance matrix, $\lambda_1$, is equal to the variance of the principal component solution of each {$y_{obs}, y_{imp, m}$}, which we calculate with `stats::princomp()` [@R].
-
-- Perform multiple linear regression with `stats::lm()` [@R] to get the $\hat{Q}$s for $Q=\beta$ and $Q=r^2$.
-We pool the $\hat{Q}$s conform @vink14 to get $\bar{Q}$s for the $\beta$s and $r^2$ **[check  finite pooling of  $r^2$, instead of `mice::pool.r.squared()`]**.
-Additionally, we use the $\hat{Q}$s of $Q=\beta$ as $\theta$ to apply the convergence diagnostics to.
-
-<!-- After each run of the `mice` algorithm, the function `mice::complete()` is used to combine the imputed data $y_{imp,m}$ with the observed data $y_{obs}$, resulting in five completed datasets {$y_{obs}, y_{imp, m}$}.
--->
-
-<!-- The estimator for each $Q$ is $\bar{Q}$---the pooled aggregate of the $\hat{Q}$s across imputations.
-To calculate the $\bar{Q}$s, we combine the observed data $y_{obs}$ and the imputed data for each imputation $y_{imp,m}$ with the function `mice::complete()`.
+We then use `mice::complete()` to combine the imputed data $\dot y_{\rm imp, \ell}$ with the observed data $y_{\rm obs}$, resulting in five completed datasets {$y_{\rm obs}, y_{imp, m}$}.
+
+On the 5 completed datasets {$y_{\rm obs}, y_{imp, m}$}, we:
+
+- Estimate $\hat{Q}_\ell$s for $Q= \mu$ and $Q=\sigma$ with `base::mean()`, and `stats::sd()` [@R]. We pool the $\hat{Q}_\ell$s with `base::mean()` to get $\bar{Q}$s.
+
+- Compute $\lambda_{1, \ell}$, to use as $\theta$ in the two convergence diagnostics.
+By definition, the first eigenvalue of the variance-covariance matrix, $\lambda_{1, \ell}$, is equal to the variance of the principal component solution of each {$y_{\rm obs}, y_{imp, m}$}, which we calculate with `stats::princomp()` [@R].
+
+- Perform multiple linear regression with `stats::lm()` [@R] to get the $\hat{Q}_\ell$s for $Q=\beta$ and $Q=r^2$.
+We pool the $\hat{Q}_\ell$s conform @vink14 to get $\bar{Q}$s for the $\beta$s and $r^2$ **[check  finite pooling of  $r^2$, instead of `mice::pool.r.squared()`]**.
+Additionally, we use the $\hat{Q}_\ell$s of $Q=\beta$ as $\theta$ to apply the convergence diagnostics to.
+
+<!-- After each run of the `mice` algorithm, the function `mice::complete()` is used to combine the imputed data $\dot y_{\rm imp, \ell}$ with the observed data $y_{\rm obs}$, resulting in five completed datasets {$y_{\rm obs}, y_{imp, m}$}.
+-->
+
+<!-- The estimator for each $Q$ is $\bar{Q}$---the pooled aggregate of the $\hat{Q}_\ell$s across imputations.
+To calculate the $\bar{Q}$s, we combine the observed data $y_{\rm obs}$ and the imputed data for each imputation $\dot y_{\rm imp, \ell}$ with the function `mice::complete()`.
 Descriptive statistics are computed as the average across imputations for each variable ($\mu_j$ and $\sigma_j$, where $j = Y, X_1, X_2, X_3$).
@@
 
-We assess the impact of inducing non-convergence in the iterative imputation algorithm by comparing $\bar{Q}$s with $Q$s.
+We assess the performance of the iterative imputation algorithm in each simulation condition by comparing $\bar{Q}$s with $Q$s.
 For each $Q$, we calculate bias as $\bar{Q} - Q$.
@@
 $$\text{CI} = \bar{Q} \pm t_{(M-1)} \times SE_{\bar{Q}},$$
-where $t_{(M-1)}$ is the quantile of a $t$-distribution with $M-1$ degrees of freedom, and $SE_{\bar{Q}}$ is the square root of the pooled variance estimate.
-If we obtain nominal coverage (CR = 95%), we can conclude that the $\bar{Q}$s are confidence valid estimates of the $Q$s.
+where $t_{(M-1)}$ is the quantile of a $t$-distribution with $m-1$ degrees of freedom, and $SE_{\bar{Q}}$ is the square root of the pooled variance estimate.
+If we obtain nominal coverage (CR = 95%), we can conclude that the $\bar{Q}$s are confidence-valid estimates of the $Q$s.
 Finally, we inspect CI width (CIW): the difference between the lower and upper bound of the 95\% confidence interval around $\bar{Q}$.
-CIW is of interest because too short CIs indicate under-estimation of the variance in $\bar{Q}$, which may yield spurious inferences.
-
-With these performance measures, we also evaluate how well the non-convergence diagnostics $\widehat{R}$ and $AC$ can identify simulation conditions in which the inferences are affected by non-convergence, i.e., where $\bar{Q}$ is *not* an unbiased, confidence valid estimate of $Q$.
+CIW is of interest because it is a measure of efficiency.
+Under nominal coverage, short CIs are preferred.
+Too short CIs, however, indicate under-estimation of the variance in $\bar{Q}$, **which may yield inferences with low statistical power if CIW is large**.
+
+With these performance measures, we also evaluate how well the non-convergence diagnostics $\widehat{R}$ and $AC$ can identify simulation conditions in which the inferences are affected by non-convergence, i.e., where $\bar{Q}$ is *not* an unbiased, confidence-valid estimate of $Q$.
 **[REPHRASE this:] For performance of univariate $Q$s, we apply the non-convergence diagnostics to univariate $\theta$s.
@@
 
-- For $Q=\mu$, $\theta$s are chain means.
+- For $Q=\mu$, $\theta$s are chain means **[ADD that these should be unbiased because MCAR!]**.
 
@@
 
-- For $Q=r^2$, $\theta$s are the first eigenvalue of the variance-covariance matrix.
-
-<!-- We measure the performance of $\widehat{R}$ and $AC$ by the four scalar summaries $\theta$: chain means (i.e., the mean in each $y_{imp, m, j}$) versus $Q=\mu$, chain variances (i.e., the variance in each $y_{imp, m, j}$) versus $Q=\sigma$, the estimated regression coefficients $\beta$ in each {$y_{obs}, y_{imp, m}$} (i.e., a user-defined statistic of scientific interest $\hat{Q}$) against $Q=\beta$, and the first eigenvalue of the variance-covariance matrix in each {$y_{obs}, y_{imp, m}$}, $\lambda_1$ versus $Q=r^2$.** -->
+- For $Q=r^2$, $\theta$s are the first eigenvalue of the variance-covariance matrix, $\lambda_{1,\ell}$.
+
+<!-- We measure the performance of $\widehat{R}$ and $AC$ by the four scalar summaries $\theta$: chain means (i.e., the mean in each $y_{imp, m, j}$) versus $Q=\mu$, chain variances (i.e., the variance in each $y_{imp, m, j}$) versus $Q=\sigma$, the estimated regression coefficients $\beta$ in each {$y_{\rm obs}, y_{imp, m}$} (i.e., a user-defined statistic of scientific interest $\hat{Q}_\ell$) against $Q=\beta$, and the first eigenvalue of the variance-covariance matrix in each {$y_{\rm obs}, y_{imp, m}$}, $\lambda_{1, \ell}$ versus $Q=r^2$.** -->
 
 <!-- **[Remove this?] Primarily, we want to know what the effect is of inducing non-convergence on the estimated $Q$s.
-More specifically, we are interested in the bias of $\bar{Q}$ for all $Q$s and the confidence validity of $\bar{Q}$ for $Q=\beta$.
-Therefore, we evaluate these performance measures against the simulation conditions (the number of iterations $T$ and the missingness proportion $p_{mis}$).
+More specifically, we are interested in the bias of $\bar{Q}$ for all $Q$s and the confidence-validity of $\bar{Q}$ for $Q=\beta$.
+Therefore, we evaluate these performance measures against the simulation conditions (the number of iterations $T$ and the missingness proportion $p_{\rm mis}$).
 We look at averages across simulation repetitions.
@@
 
-Our results show the impact of inducing non-convergence in an iterative imputation algorithm on several scientific estimates, and whether the non-convergence diagnostics $\widehat{R}$ and $AC$ identify the signs of non-convergence in the algorithm.
+Our results show the impact of **inducing non-convergence ["early stopping" en "proportion of complete cases" ]** in an iterative imputation algorithm on several scientific estimates, and whether the non-convergence diagnostics $\widehat{R}$ and $AC$ identify the signs of non-convergence in the algorithm.
 For reasons of brevity, we only discuss the non-convergence diagnostics for estimates with the worst performance in terms of bias.
@@
 
-```{r Qs, fig.height = 6.8, fig.cap="Impact of non-convergence on statistical inferences."}
+```{r Qs, fig.height = 6.8, fig.cap="Impact of non-convergence on statistical inferences.
+Depicted are the bias, coverage rate (CR) and confidence interval width (CIW) of the estimates $\\bar{Q}$.
+We present the subset of $Q$s with the worst performance in terms of bias: the mean and standard deviation of $Y$, and the regression coefficient of $X_1$."}
 mean_bias + sd_bias + Rsq_bias + est_bias + est_cov + est_ciw + plot_layout(ncol = 2, guides = "collect")
@@
 
-Figure \ref{fig:Qs} displays the influence of inducing non-convergence on the estimated quantities of scientific interest, $Q$s.
+Figure \ref{fig:Qs} displays the influence of **inducing non-convergence ["early stopping" en "proportion of complete cases"]** on the estimated quantities of scientific interest, $Q$s.
 **[Add panel labels, describe panels]** Simulation conditions that are impacted by non-convergence portray more extreme bias in the estimated $Q$s, and non-nominal coverages.
-Overall, we see that conditions with a higher proportion of missingness are affected more severely by non-convergence.
-Multivariate $Q$s are also substantially influenced by the number of iterations.
-This means that the algorithm has indeed not converged at $t=1$, and converges gradually as $t$ moves to $T$.
-The point at which an additional iteration does not lead to an improvement in the estimates depends on quantity of scientific interest $Q$ and the missingness proportion $p_{mis}$.
-The highest number of iterations that is necessary to reach this point across all simulation conditions is $T=7$.
-This implies that, under the current specifications, no more than seven iterations are necessary to obtain sufficient performance in terms of bias and confidence validity.
 
@@
 This implies that valid inferences may be obtained with as little as one iteration ($T \geq 1$).
-The magnitude of the bias in $\bar{Q}$ depends solely on the missingness proportion $p_{mis}$, with missingness proportions of 75% and 95% leading to more extreme biases than other missingness proportions.
-Unbiased estimates are obtained in all conditions where $p_{mis}\leq.50$.
-Note however, that even in the conditions where $p_{mis}\geq.75$, the bias is small: $\mu_Y$ is maximally under-estimated by 0.02 units, while the true value of $\mu_Y$ is 25.81 ($\sigma_Y = 11.32$).
+The magnitude of the bias in $\bar{Q}$ depends solely on the missingness proportion $p_{\rm mis}$, with missingness proportions of 75% and 95% leading to more extreme biases than other missingness proportions **[note that these are curious]**.
+Unbiased estimates are obtained in all conditions where $p_{\rm mis}\leq.50$.
+Note however, that even in the conditions where $p_{\rm mis}\geq.75$, the bias is small: $\mu_Y$ is maximally under-estimated by 0.02 units, while the true value of $\mu_Y$ is 25.81 ($\sigma_Y = 11.32$).
 We conclude that non-convergence does not substantially impact the validity of the inferences for $Q=\mu$.
+**[To be expected, because MCAR]**
 
@@
 Interestingly, the bias in first iteration is actually *less* extreme than for any other number of iterations.
-We may therefore conclude that no iteration at all is needed to obtain approximately unbiased estimates for $Q=\sigma$.
-Completely unbiased estimates are only obtained in conditions where $p_{mis}=.05$.
-Conditions where $p_{mis}\geq.25$ are impacted by non-convergence in order of increasing missingness proportion.
-Similar to the bias for $Q=\mu$ the magnitude of the bias in $Q=\sigma$ is not substantial in any condition.
-We conclude that both of the univariate $Q$s are practically unaffected by non-convergence.
+**[Due to MCAR and randomly drawn initial values?]** We may therefore conclude that no iteration at all is needed to obtain approximately unbiased estimates for $Q=\sigma$.
+Completely unbiased estimates are only obtained in conditions where $p_{\rm mis}=.05$.
+Conditions where $p_{\rm mis}\geq.25$ are impacted by non-convergence in order of increasing missingness proportion.
+Similar to the bias for $Q=\mu$ the magnitude of the bias in $Q=\sigma$ is negligible.
+We therefore conclude that neither of the univariate $Q$s is affected substantially by non-convergence.
 <!-- in the first iteration.
@@
 Interestingly, the bias in conditions where $T=1$ is *less* severe than after continued iteration.
-For the least possible bias, we may use any number of iterations $T\geq1$, but only $p_{mis} = 0.05$ results in completely unbiased $\bar{Q}$s.
+For the least possible bias, we may use any number of iterations $T\geq1$, but only $p_{\rm mis} = 0.05$ results in completely unbiased $\bar{Q}$s.
 -->
@@
 The magnitude of the bias clearly follows the incremental order of the missingness proportions.
-We, therefore, observe approximately unbiased estimates at a different number of iterations for each missingness proportion.
-For $p_{mis}\leq.25$, the minimum number of iterations is two ($T\geq2$), while $p_{mis}=.50$ requires $T\geq4$, and $p_{mis}=.75$ requires $T\geq5$.
-The highest number of iterations needed to obtain valid inferences is seven ($p_{mis}=.95$; $T\geq7$).
-Unbiased estimates, however, may only be obtained if the missingness proportion is at most 25%.
-**[interpret sign of bias]**
+Completely unbiased estimates may only be obtained if the missingness proportion is at most 25%.
+Approximately unbiased estimates are observed for any missingness proportion.
+For $p_{\rm mis}\leq.25$, the minimum number of iterations is two ($T\geq2$),  for $p_{\rm mis}=.50$ it is $T\geq4$, and $p_{\rm mis}=.75$ requires $T\geq5$.
+The highest number of iterations needed to obtain valid inferences is seven ($p_{\rm mis}=.95$; $T\geq7$).
+**[interpret sign of bias?
+ADD what this implies: dat - zelfs bij een gebiasede methode - je maar 7 iteraties nodig hebt om het eindresultaat te bereiken.]**
 
 <!-- while the best performance occurs with less missingness and more iterations.
-Only conditions with the lowest missingness proportion ($p_{mis}=0.05$) results in unbiased estimates for every number of iterations.
+Only conditions with the lowest missingness proportion ($p_{\rm mis}=0.05$) results in unbiased estimates for every number of iterations.
 In other conditions there is at least some bias irrespective of $T$.
@@
 For the estimated regression coefficients, we consider bias, confidence interval width (CIW), and coverage rate (CR) as performance measures.
-Unbiased estimates are obtained in conditions where $p_{mis}\leq.50$ and $T\geq3$.
-Approximately confidence valid estimates are obtained in any condition where $T\geq3$---despite of persistant bias in conditions where $p_{mis}\geq.75$.
-These approximately nominal coverages are explained by the CIWs: conditions with higher missingness proportions result in more uncertainty in the estimates (i.e., wider CIs), which diminish the effect of the bias.
-Confidence valid inferences may be obtained when $T\geq3$, while we need at least seven iterations to reach valid inferences in terms of bias.
+**[Split panels]** Unbiased estimates are obtained in conditions where $p_{\rm mis}\leq.50$ and $T\geq3$.
+Approximately confidence-valid estimates are obtained in any condition where $T\geq3$---despite of persistant bias in conditions where $p_{\rm mis}\geq.75$.
+These approximately nominal coverages are explained by the CIWs: as expected, conditions with higher missingness proportions result in more uncertainty in the estimates (i.e., wider CIs), which diminish the effect of the bias.
+confidence-valid inferences may be obtained when $T\geq3$, while we need at least seven iterations to reach valid inferences in terms of bias.
 Continued iteration after $T\geq7$ seems futile.
-<!-- One exception is the CIW of the most extreme condition ($p_{mis}=.95$; $T=1$). -->
+**[add max T for other betas?]**
+<!-- One exception is the CIW of the most extreme condition ($p_{\rm mis}=.95$; $T=1$). -->
+
+### Summary
+
+Overall, we see that conditions with a higher proportion of missingness are affected more severely by non-convergence.
+Multivariate $Q$s are also substantially influenced by the number of iterations.
+This means that the algorithm has indeed not converged at $t=1$, and converges gradually as $t$ moves to $T$.
+The point at which an additional iteration does not lead to an improvement in the estimates depends on quantity of scientific interest $Q$ and the missingness proportion $p_{\rm mis}$.
+The highest number of iterations that is necessary to reach this point across all simulation conditions is $T=7$.
+This implies that, under the current specifications, no more than seven iterations are necessary to obtain sufficient performance in terms of bias and confidence-validity.
 
@@
 This is especially pronounced in conditions with high missingness proportions: with 95% missingness we need at least seven iterations before an additional iteration does not lead to improved estimates.
-Conditions with a lower missingness proportion (e.g., $p_{mis} = {0.5}$), produce *optimal* estimates with as many as two iterations.
-Approximately unbiased estimates are obtained for all conditions where $p_{mis}\leq0.75$ and $T\leq6$.
+Conditions with a lower missingness proportion (e.g., $p_{\rm mis} = {0.5}$), produce *optimal* estimates with as many as two iterations.
+Approximately unbiased estimates are obtained for all conditions where $p_{\rm mis}\leq0.75$ and $T\leq6$.
 **[Add CIW and CR]** -->
@@
 
-We evaluate the non-convergence diagnostics $\widehat{R}$ and $AC$ against the performance measures described above (i.e., bias and confidence validity of the quantities of scientific interest $Q$).
+We evaluate the non-convergence diagnostics $\widehat{R}$ and $AC$ against the performance measures described above (i.e., bias and confidence-validity of the quantities of scientific interest $Q$).
 The diagnostics should identify those simulation conditions that are negatively impacted by non-convergence.
-We expect the most signs of non-convergence in conditions where *sufficient* estimates are not obtained (high missingness proportions, $T<7$), and at least some signs of non-convergence in conditions where estimates are not approximately unbiased ($p_{mis}=.95$, any number of iterations).
-
-<!-- As we have observed, the affected conditions are $p_{mis}=.95$ for all all $Q$s are impacted by the missingness proportions, but only the multivariate quantities of scientific interest are affected by the number of iterations.
--->
-
-Figure \ref{fig:rhats-acs} displays the non-convergence diagnostics $\widehat{R}$ and $AC$ for each of the four $\theta$s under consideration.
-**[Add panel labels and explain how to read the plots]** In general, the diagnostics identify more signs of non-convergence in simulation conditions with a lower number of iterations.
-The missingness proportions, however, only seem to impact multivariate convergence.
-When $\widehat{R}$ and $AC$ are applied to  univariate $\theta$s, they identify the same amount of non-convergence for any missingness proportion.
-When the diagnostics are applied to multivariate $\theta$s, they generally identify more signs of non-convergence for higher missingness proportions.
-
-As expected, none of the simulation conditions that we consider resulted in complete convergence of the algorithm (defined as $\widehat{R}=1$ and $AC=0$).
-However, the thresholds used in practice to diagnose non-convergence seem too conservative for imputation purposes.
-$\widehat{R}$ and $AC$ identify signs of non-convergence in conditions where the $\bar{Q}$s are unbiased and confidence valid estimates of the $Q$s.
-The most recent recommended threshold for diagnosing non-convergence with $\widehat{R}$ is never overcome in this study (all $\widehat{R}$>1.01).
-Apparently, this threshold is too stringent for iterative imputation algorithms, and will not be discussed any further.
+We expect the most signs of non-convergence in conditions where valid estimates are not obtained (high missingness proportions, $T<7$), and at least some signs of non-convergence in conditions where estimates are not approximately unbiased ($p_{\rm mis}=.95$, any number of iterations).
+
+<!-- As we have observed, the affected conditions are $p_{\rm mis}=.95$ for all all $Q$s are impacted by the missingness proportions, but only the multivariate quantities of scientific interest are affected by the number of iterations.
+-->
+
+Figure \ref{fig:rhats-acs} displays the non-convergence diagnostics $\widehat{R}$ and $AC$ for each of the four $\theta$s under consideration. **[Add panel labels and explain how to read the plots]**
 
@@
 <!-- **Message: after one iteration ($T$=2), the algorithm is as well converged as the missingness proportion allows.
-Sooner even with lower missingness proportions---$T$=1 might be sufficient for $p_{mis}$ = 5, 25, and 50.** Number of iterations has little influence on bias in $\hat{Q}$, only in first iteration there is some bias in missingness conditions of 25 and up.
+Sooner even with lower missingness proportions---$T$=1 might be sufficient for $p_{\rm mis}$ = 5, 25, and 50.** Number of iterations has little influence on bias in $\hat{Q}_\ell$, only in first iteration there is some bias in missingness conditions of 25 and up.
 There is a clear different between the missingness conditions.
@@
 For this $Q$, t=2 would be sufficient.
-And only for $p_{mis}$ = 5 the $\bar{Q}$s are unbiased.
-All other $p_{mis}$ conditions have bias irrespective of the number of iterations.
+And only for $p_{\rm mis}$ = 5 the $\bar{Q}$s are unbiased.
+All other $p_{\rm mis}$ conditions have bias irrespective of the number of iterations.
 -->
@@
 ### $\theta=$ chain mean.
+**[As expected, $\widehat{R}$ lowers with the number of iterations.
+At a cut-point of 1.2 , we have convergence at about 8 iteration, whereas we need 12 and 25 iteration for more conservative cut-off 1.1 and 1.05.
+We do not reach the very strict 1.01.]**
+
 We evaluate the performance of the non-convergence diagnostics applied to chain means against the bias in $Q=\mu$.
@@
 This does not correspond to the impact of non-convergence in the univariate $Q$s.
-**Compared to the multivariate $Q$s [add before that we evaluate the thetas against all Qs?], $\widehat{R}$ and $AC$ applied to the chain means fail to identify the non-convergence in conditions where $p_{mis}=.95$.**
+**Compared to the multivariate $Q$s [add before that we evaluate the thetas against all Qs?], $\widehat{R}$ and $AC$ applied to the chain means fail to identify the non-convergence in conditions where $p_{\rm mis}=.95$.**
 
@@
 
-### $\theta=\hat{Q}$.
+### $\theta=\hat{Q}_\ell$.
+**[We zien weer hetzelfde voor laag-medium, beetje meer voor extreme misisnh data.]**
+
 The non-convergence diagnostics are applied to the estimated regression coefficients in each imputation, and evaluated against the bias in the corresponding pooled estimate.
@@
 The thresholds to diagnose non-convergence are overcome in different conditions, depending on the number of iterations and the missingness proportion.
-The most lenient threshold ($\widehat{R}>1.2$) is overcome at $T\geq9$ for $p_{mis} = 0.05$, and $T\geq12$ for $p_{mis} = 0.95$.
+The most lenient threshold ($\widehat{R}>1.2$) is overcome at $T\geq9$ for $p_{\rm mis} = 0.05$, and $T\geq12$ for $p_{\rm mis} = 0.95$.
 The conventional $\widehat{R}>1.1$ is surpassed at $T\geq12$ and $T\geq24$, respectively.
 
-For the autocorrelations applied to the $\hat{Q}$s, we see an even stronger effect of the missingness proportions.
-Especially the conditions where $p_{mis} = 0.75$ and $p_{mis} = 0.95$ are identified as non-stationary.
+For the autocorrelations applied to the $\hat{Q}_\ell$s, we see an even stronger effect of the missingness proportions.
+Especially the conditions where $p_{\rm mis} = 0.75$ and $p_{\rm mis} = 0.95$ are identified as non-stationary.
 The threshold used to evaluate whether $AC=0$, however, is only reached in conditions with at least ten iterations.
@@
 However, the critical value for diagnosing non-convergence with $AC$ decreases as a function of the number of iterations.
-The conclusion to identify non-stationarity in conditions where $T\geq10$ and $p_{mis} = 0.95$ might therefore be spurious.
+The conclusion to identify non-stationarity in conditions where $T\geq10$ and $p_{\rm mis} = 0.95$ might therefore be spurious.
 
@@
 
-Compared to the bias in $\bar{Q}$, the diagnostics applied to the $\hat{Q}$s over-estimate the severity of the non-convergence. 
+Compared to the bias in $\bar{Q}$, the diagnostics applied to the $\hat{Q}_\ell$s over-estimate the severity of the non-convergence.
+
+### Summary
+
+In general, the diagnostics identify more signs of non-convergence in simulation conditions with a lower number of iterations.
+The missingness proportions, however, only seem to impact multivariate convergence.
+When $\widehat{R}$ and $AC$ are applied to  univariate $\theta$s, they identify the same amount of non-convergence for any missingness proportion.
+When the diagnostics are applied to multivariate $\theta$s, they generally identify more signs of non-convergence for higher missingness proportions.
+
+As expected, none of the simulation conditions that we consider resulted in complete convergence of the algorithm (defined as $\widehat{R}=1$ and $AC=0$).
+However, the thresholds used in practice to diagnose non-convergence seem too conservative for imputation purposes.
+$\widehat{R}$ and $AC$ identify signs of non-convergence in conditions where the $\bar{Q}$s are unbiased and confidence-valid estimates of the $Q$s.
+The most recent recommended threshold for diagnosing non-convergence with $\widehat{R}$ is never overcome in this study (all $\widehat{R}$>1.01).
+Apparently, this threshold is too stringent for iterative imputation algorithms, and will not be discussed any further.
+
 <!-- This is odd, since the $\theta$ and $Q$ are directly related.   -->
@@
 
-### $\theta=\lambda_1$.
+### $\theta=\lambda_{1, \ell}$.
 Finally, we evaluate the non-convergence diagnostics applied to the first eigenvalue of the variance-covariance matrices of the completed data.
 Similar to the results the the regression coefficients as $\theta$s, we see effects of both the missingness proportion and number of iterations in each simulation condition.
-The most signs of non-convergence are identified in conditions where $p_{mis}$ is large and $T$ is small.
-
-Notwithstanding the similarities with $\theta=\hat{Q}$, there is a big difference in $\widehat{R}$-values for conditions where $T<4$.
+The most signs of non-convergence are identified in conditions where $p_{\rm mis}$ is large and $T$ is small.
+
+Notwithstanding the similarities with $\theta=\hat{Q}_\ell$, there is a big difference in $\widehat{R}$-values for conditions where $T<4$.
 In these conditions, $\widehat{R}$ behaves oppositely of our expectation: conditions with the lowest missingness proportions yield the highest $\widehat{R}$-values.
 In conditions with a higher number of iterations, this artifact disappears.
-This suggests that the combination of $\widehat{R}$ and $\lambda_1$ is only appropriate for imputation algorithms where the number of iterations at least four.
+This suggests that the combination of $\widehat{R}$ and $\lambda_{1, \ell}$ is only appropriate for imputation algorithms where the number of iterations at least four.
 Interpreting the $\widehat{R}$ values of conditions where $T<3$ may lead to the incorrect conclusion that only conditions with *low* missingness proportions show signs of non-convergence.
@@
 Using this cut-off we conveniently ignore the initial 'dip' at $T=3$ as well.
-The traditional threshold $\widehat{R} > 1.2$ is overcome in conditions where $p_{mis} = .95$ and $T\geq10$, and for all other missingness proportions in conditions where $T\geq8$.
+The traditional threshold $\widehat{R} > 1.2$ is overcome in conditions where $p_{\rm mis} = .95$ and $T\geq10$, and for all other missingness proportions in conditions where $T\geq8$.
 With the $\widehat{R} > 1.1$ threshold, the number of iterations are $T\geq20$ and $T\geq16$, respectively.
@@
 The $AC$-values largely depend on the missingness proportion in each condition.
-Conditions where $p_{mis}=.95$ show statistically significant autocorrelations for any number of iterations.
-The other extreme, conditions where $p_{mis}=.05$, does not yield $AC$ values above the threshold.
+Conditions where $p_{\rm mis}=.95$ show statistically significant autocorrelations for any number of iterations.
+The other extreme, conditions where $p_{\rm mis}=.05$, does not yield $AC$ values above the threshold.
 The $AC$ values with this missingness proportion are equivalent for all conditions where $T\geq6$, which suggests that the algorithm is as stationary as it can be.
@@
 In comparison to the $Q$s, the diagnostics again over-estimate the signs of non-convergence.
-Conditions with low missingness proportions (e.g., 5% or 25% of the complete data) yield unbiased, confidence valid estimates with as little as two iterations.
+Conditions with low missingness proportions (e.g., 5% or 25% of the complete data) yield unbiased, confidence-valid estimates with as little as two iterations.
 The convergence diagnostics, however, indicate improved convergence up-to eight iterations according to $AC$ and thirty to forty iterations according to $\widehat{R}$.
-The model-independent $\theta$ $\lambda_1$ seems as well-equipped to diagnose signs of non-convergence as $\theta=\hat{Q}$.
+The model-independent $\theta$ $\lambda_{1, \ell}$ seems as well-equipped to diagnose signs of non-convergence as $\theta=\hat{Q}_\ell$.
 The univariate $\theta$s do not differentiate between bias induced by the missingness proportions.
@@
 
-In short, iterative imputation algorithms may yield approximately unbiased, confidence valid estimates under different levels of induced non-convergence.
+We arrive back at the original questions that we posed.
+
+- Non-convergence can be identified with non-convergence diagnostics $\widehat{R}$ and $AC$.
+But common thresholds are too conservative.
+Why?
+Part of the distribution is already determined by the observed data.
+in regular MCMC the entire distribution is unknown (**priors??**).
+In iterative imputation algorithms we have the same characteristics as MCMC in the limit, but because we combine the known distribution with the unknown distribution, convergence is reached much sooner.
+
+- Inferences are unbiased much quicker than the algorithm has converged.
+All things univariate are converged almost intantly after initiating the iterative algorithm with a valid imputation model.
+But less quick with multivariate parameters.
+
+In short, iterative imputation algorithms may yield approximately unbiased, confidence-valid estimates under different levels of induced non-convergence.
 Univariate estimates seem robust against terminating the algorithm early: There is no clear effect of the number of iterations on the bias in estimated means and standard deviations.
-The bias in these $Q$s was only impacted by the missingness proporion $p_{mis}$, with higher $p_{mis}$ resulting in more bias.
-Bias in multivariate estimates depended on both $p_{mis}$ and the number of iterations.
+The bias in these $Q$s was only impacted by the missingness proporion $p_{\rm mis}$, with higher $p_{\rm mis}$ resulting in more bias.
+Bias in multivariate estimates depended on both $p_{\rm mis}$ and the number of iterations.
 However, the only effect of the number of iterations was observed when the algorithm was terminated at seven iterations or earlier.
@@
 Yet, the non-convergence diagnostics $\widehat{R}$ and $AC$ indicated that the iterative imputation algorithm did not reach a stable state before twenty to thirty iterations.
-**[MERGE or remove] $\widehat{R}$ and autocorrelation indicate that there are signs of non-convergence in the algorithm up-to at least twenty iterations, while unbiased, confidence valid estimates may be obtained with as little as one iteration.
+**[MERGE or remove] $\widehat{R}$ and autocorrelation indicate that there are signs of non-convergence in the algorithm up-to at least twenty iterations, while unbiased, confidence-valid estimates may be obtained with as little as one iteration.
 This observation is in agreement with **one of [still add the others!]** the simulation hypotheses: $\widehat{R}$ over-estimates the severity of non-convergence when applied to MI procedures.**
@@
 
-- Is the point of *sufficient* convergence (at which the $\bar{Q}$s are unbiased and confidence valid estimates of the $Q$s) the same point as identified by the non-convergence diagnostics?
+- Is the point of *sufficient* convergence (at which the $\bar{Q}$s are unbiased and confidence-valid estimates of the $Q$s) the same point as identified by the non-convergence diagnostics?
 (NOPE, they are too conservative).
 
-- Is one of the thetas better? (YES, while it doesn't really matter as long as it's multivariate, $\lambda_1$ is model-independent and therefore preferred).
+- Is one of the thetas better? (YES, while it doesn't really matter as long as it's multivariate, $\lambda_{1, \ell}$ is model-independent and therefore preferred).
 
@@
 
-2. Monitor non-convergence diagnostics $\widehat{R}$ and $AC$. Use the thresholds for $\widehat{R}>1.2$ and autocorrelation  **???**. Keep iterating until these thresholds are reached.
-
-3. Track $\theta$s over iterations, e.g., the novel $\theta$ that is 'substantive model-independent'.
-
-4. Track your own scalar summary of interest (see e.g., @buur18). Compute $\widehat{R}$ and autocorrelation values for this scalar summary.
-
-5.
+2.
+Monitor non-convergence diagnostics $\widehat{R}$ and $AC$.
 Do not use the traditional calculation for $\widehat{R}$, or the `R` function `stats::acf()` to compute $AC$.
 Instead, calculate $\widehat{R}$ conform @veht19 and compute autocorrelations manually (see e.g., [github.com/gerkovink/shinyMice](https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy)).
+
+3. Track $\theta$s over iterations, e.g., the novel $\theta$ that is 'substantive model-independent'.
+
+4. Track your own scalar summary of interest (see e.g., @buur18). Compute $\widehat{R}$ and autocorrelation values for this scalar summary.
+
+5.
+Use the threshold $\widehat{R}>1.2$ for mixing, and assess stationarity by plotting the autocorrelation over iterations.
+Keep iterating until the threshold for mixing is reached, and until a reasonably decreasing asympototic $AC$ value is obtained.
+At that point, inferences are unlikely to improve by continued iteration.
 
@@
 To conclude, in the case of drawing inference from incomplete data, convergence of the iterative imputation algorithm is often a convenience but not a necessity.
+**[Our study found that - in the cases considered - inferential validity was achieved after 5-10 iterations, much earlier than indicated by the $\widehat{R}$ and AC diagnostics.
+Of course, it never hurts to iterate longer, but such calculations hardly bring added value.]**
 

### 2020-05-07 3.Thesis/Writeup/2.Manuscript.tex
Feedback Stef

@@
 Missing data is not only ubiquitous across science, but treating it can also be a tedious task.
-If a dataset contains just one incomplete observation, statistical inferences are undefined and will not produce any results.
+If a dataset contains just one incomplete observation, calculations are not defined, \textbf{hence} statistical models cannot do fitted to the data.
 To circumvent this, many statistical packages employ list-wise deletion by default (i.e., ignoring incomplete observations).
-Unfortunately, this \emph{ad hoc} solution may yield wildly invalid results \citep{buur18}.
-An alternative is to \emph{impute} (i.e., fill in) the missing values in the incomplete observations.
-Subsequently, statistical inferences can be performed on the completed dataset.
+Unfortunately, this \emph{ad hoc} solution may yield invalid results \citep{buur18}.
+An alternative is to apply imputation.
+With imputation, we `fill in' the missing values in an incomplete dataset.
+Subsequently, the model of scientific interest can be fitted to the completed dataset.
 By repeating this process several times, a distribution of plausible results may be obtained, which reflects the uncertainty in the data due to missingness.
@@
 Figure \ref{fig:diagram} provides an overview of the steps involved with MI.
-The missing part \(y_{mis}\) of an incomplete dataset is imputed \(M\) times.
-The imputed data \(y_{imp}\) is combined with the observed data \(y_{obs}\) to create \(M\) completed datasets.
-On each completed dataset, an analysis of scientific interest is performed.
-The quantity of scientific interest (e.g., a regression coefficient) is denoted with \(Q\).
-Since \(Q\) is estimated on each completed dataset, \(M\) separate \(\hat{Q}\)-values are obtained.
-These \(M\) values are combined into a single pooled estimate \(\bar{Q}\).
-The premise of multiple imputation is that \(\bar{Q}\) is an unbiased and confidence valid estimate of the true---but unobserved---inference.
+The missing part \(y_{\rm mis}\) of an incomplete dataset is imputed \(m\) times.
+This creates \(m\) sets of imputed data \(\dot y_{\rm imp, \ell}\), where \(\ell = 1, 2, ..., m\).
+The imputed data is then combined with the observed data \(y_{\rm obs}\) to create \(m\) completed datasets.
+On each of these datasets the analysis of scientific interest is performed to estimate \(Q\): the quantity of scientific interest (e.g., a regression coefficient).
+Since \(Q\) is estimated on each completed dataset, \(m\) separate \(\hat{Q}_\ell\)-values are obtained.
+Finally, the \(m\) \(\hat{Q}_\ell\)-values are combined into a single pooled estimate \(\bar{Q}\).
+The premise of multiple imputation is that \(\bar{Q}\) is an unbiased and confidence-valid estimate of the true---but unobserved---inference \citep{rubin96}.
 
@@
 
-{\centering \includegraphics[width=\linewidth]{../Figures/diagram}
+{\centering \includegraphics[width=\linewidth]{../Figures/DiagramMI}
 
@@
 
-\caption{Scheme of the main steps in multiple imputation (where $M=3$)---from an incomplete dataset, to $M$ multiply imputed datasets, to $M$ estimated quantities of scientific interest $\hat{Q}$s, to a single pooled estimate $\bar{Q}$.}\label{fig:diagram}
+\caption{Scheme of the main steps in multiple imputation (where $m=3$)---from an incomplete dataset, to $m$ multiply imputed datasets, to $m$ estimated quantities of scientific interest $\hat{Q}$s, to a single pooled estimate $\bar{Q}$.}\label{fig:diagram}
 \end{figure}
@@
 A popular method to obtain imputations is to use the `Multiple Imputation by Chained Equations' algorithm, shorthand `MICE'\citep{mice}.
-With MICE, imputed values \(y_{imp}\) are drawn from the posterior predictive distribution of the missing values \(y_{mis}\).
+With MICE, imputed values \(\dot y_{\rm imp, \ell}\) are drawn from the posterior predictive distribution of the missing values \(y_{\rm mis}\).
 The algorithm is named after the iterative algorithmic procedure by which imputed values are generated: a multivariate distribution is obtained by iterating over a sequence of univariate imputations.
 The iterative nature of algorithms like MICE introduces a potential threat to the validity of the imputations: What if the algorithm has not converged?
-Are the implications then to be trusted?
+Are the imputations then to be trusted?
 And can we rely on the inference obtained on the completed data?
 
-These remain open questions, because the convergence properties of iterative imputation algorithms have not been systematically studied \citep{buur18}.
-Moreover, there is no scientific consensus on how to evaluate convergence of imputation algorithms \citep{taka17}.
-Some default MICE techniques (e.g., `predictive mean matching') might not yield converged states at all \citep{murr18}.
-Therefore, algorithmic convergence should be monitored carefully.
+These remain open questions, since the convergence properties of iterative imputation algorithms have not been systematically studied \citep{buur18}.
+There is no scientific consensus on how to evaluate convergence of imputation algorithms \citep{taka17}.
+\textbf{Moreover, the convergence of chained equations algorithms such as MICE under certain default imputation models (e.g., `predictive mean matching') is an entirely open question \citep{murr18}.} Therefore, algorithmic convergence should be monitored carefully.
 
@@
 Typically, the \(\theta\)s under evaluation are chain means and variances---the average and the variance of the imputed values per imputation.
-Monitoring convergence by inspecting traceplots of these \(\theta\)s is insufficient on two counts: 1) it may be challenging to the untrained eye, and 2) only severely pathological cases of non-convergence may be diagnosed \citep[\(\S\) 6.5.2]{buur18} \textbf{{[}Add that convergence of univariate thetas cannot guarantee multivariate convergence?{]}}.
+Monitoring convergence by inspecting traceplots of these \(\theta\)s may be undesirable for several reasons on two counts: 1) it may be challenging to the untrained eye, and 2) only severely pathological cases of non-convergence may be diagnosed \citep[\(\S\) 6.5.2]{buur18} and, 3) there is not an objective measure that quantifies convergence.
 Therefore, a quantitative, diagnostic evaluation of convergence would be preferred---although not straightforward.
-Iterative imputation algorithms such as MICE are Markov chain Monte Carlo (MCMC) methods.
+Iterative imputation algorithms such as MICE are special cases of Markov chain Monte Carlo (MCMC) methods.
 In MCMC methods, convergence is not from a scalar to a point, but from one distribution to another.
@@
 
-In this paper, we investigate how non-convergence in iterative imputation algorithms may be diagnosed.
-First, we define which non-convergence diagnostics will be considered, and evaluate how these diagnostics could be appropriate for iterative imputation applications.
-In the second part of this paper, we will investigate the impact of inducing non-convergence in iterative imputation algorithms by means of model-based simulation in \texttt{R} \citep{R}.
-The aim of the simulation is to determine whether unbiased, confidence valid inferences may be obtained under different levels of non-convergence (\textbf{and at which point convergence may safely be assumed}).
+In this paper, we study different methods for assessing non-convergence in iterative imputation algorithms.
+We define several diagnostics and evaluate how these diagnostics could be appropriate for iterative imputation applications.
+We then adress the impact of inducing non-convergence in iterative imputation algorithms by means of model-based simulation in \texttt{R} \citep{R}.
+The aim of the simulation is to determine whether unbiased, confidence-valid inferences may be obtained under different levels of non-convergence (\textbf{and at which point convergence may safely be assumed}).
 Additionally, we will assess how well different non-convergence diagnostics perform in identifying signs of non-convergence (\textbf{and on which scalar summary of the state space of the algorithm these diagnostics should be applied}).
-The results of this study are guidelines for identifying non-convergence in iterative imputation algorithms, which will aid applied researchers in drawing valid inference from incomplete datasets.
+We translate the results of the study into guidelines for applied researchers.
 \textbf{{[}add more societal relevance: what is the consequence of non-convergence?
@@
 The data value of unit \(i\) (\(i = 1, 2, \dots, n\)) on variable \(j\) (\(j = 1, 2, \dots, k\)) may be either observed or missing.
-The number of units \(i\) in dataset \(y\) with at least one missing value, divided by the total number of units \(n\), is called the missingness proportion \(p_{mis}\).
-The collection of observed data values in \(y\) is denoted by \(y_{obs}\); the missing part of \(y\) is referred to as \(y_{mis}\).
-For each datapoint in \(y_{mis}\), we sample \(M \times T\) plausible values, where \(M\) is the number of imputations (\(m = 1, 2, \dots, M\)) and \(T\) is the number of iterations (\(t = 1, 2, \dots, T\)) in the imputation algorithm.
+The number of units \(i\) in dataset \(y\) with at least one missing value, divided by the total number of units \(n\), is called the missingness proportion \(p_{\rm mis}\).
+The collection of observed data values in \(y\) is denoted by \(y_{\rm obs}\); the missing part of \(y\) is referred to as \(y_{\rm mis}\).
+For each datapoint in \(y_{\rm mis}\), we sample \(m \times T\) plausible values, where \(m\) is the number of imputations (\(\ell = 1, 2, \dots, m\)) and \(T\) is the number of iterations (\(t = 1, 2, \dots, T\)) in the imputation algorithm.
 The collection of samples between the initial value (at \(t=1\)) and the final imputed value (at \(t=T\)) will be referred to as an `imputation chain'.
@@
 Figure \ref{fig:non-conv} displays a subset of the example: a variable that we will call \(j\).
-We see two traceplots of the chain means for \(j\)---i.e., the average of the imputed values for \(j\) in each imputation \(y_{imp,m}\), plotted against the iteration number.
+We see two traceplots of the chain means for \(j\)---i.e., the average of the imputed values for \(j\) in each imputation \(\dot y_{\rm imp, \ell}\), plotted against the iteration number.
 Each line portrays one imputation.
@@
 \textbf{In the typical convergence situation, the imputation chains intermingle nicely and there is little to no trending.
-In the non-convergence plot, there is a lot of trending and some chains do not intermingle.} Importantly, the chain means at the last iteration (the imputed value per \(m\)) are very different between the two plots.
+In the non-convergence plot, there is a lot of trending and some chains do not intermingle.} Importantly, the chain means at the last iteration (the imputed value per imputation \(\ell\)) are very different between the two plots.
 The algorithm with the mis-specified model yields imputed values that are on average a factor two larger than those of the typically converged algorithm.
-That means that non-convergence has a impact on the mean of \(j\) in \(y_{imp, m}\).
-This difference (presumably) translates to bias in the \(M\) completed data estimates \(\hat{\mu}_{j,m}\), and consequently also to the pooled estimate \(\bar{\mu}_j\).
+That means that non-convergence has a impact on the mean of \(j\) in \(\dot y_{\rm imp, \ell}\).
+This difference (presumably) translates to bias in the \(m\) completed data estimates \(\hat{\mu}_{j,\ell}\), and consequently also to the pooled estimate \(\bar{\mu}_j\).
 Non-convergence thus leads to biased estimates.
@@
 \textbf{{[}MERGE with section below{]}}.
-Inspecting non-convergence through traceplots of chain means and chain variances may be insufficient because they are summaries of the univariate state space of the algorithm, while MICE is not only concerned with a sigle column, but the entire multivariate distribution of \(y_{imp}\).
+Inspecting non-convergence through traceplots of chain means and chain variances may be insufficient because they are univariate summaries of the state space of the algorithm, while MICE is not only concerned with a single column, but the entire multivariate distribution of \(y_{\rm imp}\).
 Ideally, a multivariate \(\theta\) should be monitored.
 A suggestion by \citet{buur18} for a multivariate \(\theta\) to monitor is the quantity of scientific interest \(Q\) (e.g., a regression coefficient), which usually is multivariate in nature.
-Implementing this, however, might be somewhat advanced for empirical researchers.
+Implementing this, however, might be somewhat too technical for empirical researchers.
 Moreover, this scalar summary is not model-independent, i.e., it only applies to one model of scientific interest.
@@
 Eigenvalues of a variance-covariance matrix are a measure of the data's total covariance.
-Such a \(\theta\) would be a model-independent, multivariate scalar summary to monitor, but it has not yet been implemented.
+Such a \(\theta\) would be a model-independent, multivariate scalar summary to monitor, but it is not implemented in \texttt{mice}.
 
@@
 \(\theta\)s are scalar summaries of the state space of the algorithm at a specific iteration.
-The recommended scalar summaries to evaluate are chain means and chain variances.
+The default scalar summaries in \texttt{mice} are chain means and chain variances.
 Additionally, researchers may ``monitor some statistic of scientific interest'' \citep[\(\S\) 6.5.2]{buur18}.
@@
 Therefore, we propose a \(\theta\) that summarizes the multivariate state space of the algorithm, but is independent from the model of scientific interest.
-We propose \(\lambda_1\) as such a scalar summary.
-We define \(\lambda_1\) as the first eigenvalue of the variance-covariance matrix of the completed data.
-Let \(\lambda_1 \geq \lambda_2 \geq ...
-\geq \lambda_j\) be the eigenvalues of \(\Sigma\) in each of the \(M\) completed datasets \({y_{obs}, y_{imp}}\).
-\(\lambda_1\) is measure that summarizes the covariances in the completed datasets.
+We propose \(\lambda_{1, \ell}\) as such a scalar summary.
+We define \(\lambda_{1, \ell}\) as the first eigenvalue of the variance-covariance matrix of the completed data.
+Let \(\lambda_{1, \ell} \geq \lambda_{2, \ell} \geq ...
+\geq \lambda_{j, \ell}\) be the eigenvalues of \(\Sigma_\ell\) in each of the \(m\) completed datasets \{\(y_{\rm obs}, \dot y_{\rm imp, \ell}\)\}.
+\(\lambda_{1, \ell}\) is measure that summarizes the covariances in the completed datasets.
 \textbf{The first eigenvalue has the appealing property that is not dependent on the substantive model of interest.
@@
 With a recently proposed adaptation, \(\widehat{R}\) might also serve to diagnose non-stationarity, but this has not yet been thoroughly investigated \citep{veht19}.
-Therefore, we use \(\widehat{R}\) and \(AC\) to evaluate mixing and stationarity separately, as recommended by e.g., \citet{cowl96}.
+Therefore, we use \(\widehat{R}\) \textbf{{[}WHICH?
+Explain that we'll use both{]}} and \(AC\) to evaluate mixing and stationarity separately, as recommended by e.g., \citet{cowl96}.
 
@@
 The Vehtari et al.~version of \(\widehat{R}\) may be suitable for iterative imputation.
-We therefore use their definition of the diagnostic.} Let \(M\) be the total number of chains, \(T\) the number of iterations per chain, and \(\theta\) the scalar summary of interest (e.g., chain means).
-For each chain (\(m = 1, 2, \dots, M\)), we estimate the variance of \(\theta\), and average these to obtain within-chain variance \(W\).
+We therefore use their definition of the diagnostic.} Let \(m\) be the total number of chains, \(T\) the number of iterations per chain (\(T\geq2\)), and \(\theta\) the scalar summary of interest (e.g., chain means).
+For each chain (\(\ell = 1, 2, \dots, m\)), we estimate the variance of \(\theta\), and average these to obtain within-chain variance \(W\).
 
 \begin{align*}
-W&=\frac{1}{M} \sum_{m=1}^{M} s_{j}^{2}, \text { where } s_{m}^{2}=\frac{1}{T-1} \sum_{t=1}^{T}\left(\theta^{(t m)}-\bar{\theta}^{(\cdot m)}\right)^{2}. 
+W&=\frac{1}{m} \sum_{\ell=1}^{m} s_{j}^{2}, \text { where } s_{\ell}^{2}=\frac{1}{T-1} \sum_{t=1}^{T}\left(\theta^{(t \ell)}-\bar{\theta}^{(\cdot \ell)}\right)^{2}. 
 \end{align*}
 
-\noindent We then estimate between-chain variance \(B\) as the variance of the collection of average \(\theta\) per chain.
+\noindent We then estimate between-chain variance \(B\) as the variance of the collection of average \(\theta\) per chain. \textbf{{[}Note that this is not the usual B in MI{]}}
 
 \begin{align*}
-B&=\frac{T}{M-1} \sum_{m=1}^{M}\left(\bar{\theta}^{(\cdot m)}-\bar{\theta}^{(\cdot \cdot)}\right)^{2}, \text { where } \bar{\theta}^{(\cdot m)}=\frac{1}{T} \sum_{t=1}^{T} \theta^{(t m)} \text{, } \bar{\theta}^{(\cdot \cdot)}=\frac{1}{M} \sum_{m=1}^{M} \bar{\theta}^{(\cdot m)}.
+B&=\frac{T}{m-1} \sum_{\ell=1}^{m}\left(\bar{\theta}^{(\cdot \ell)}-\bar{\theta}^{(\cdot \cdot)}\right)^{2}, \text { where } \bar{\theta}^{(\cdot \ell)}=\frac{1}{T} \sum_{t=1}^{T} \theta^{(t \ell)} \text{, } \bar{\theta}^{(\cdot \cdot)}=\frac{1}{m} \sum_{\ell=1}^{m} \bar{\theta}^{(\cdot \ell)}.
 \end{align*}
@@
 High \(\widehat{R}\)-values thus indicate non-convergence.
-The conventionally acceptable threshold for convergence was \(\widehat{R} < 1.2\) \citep{gelm92}.
-More recently, \citet{veht19} proposed a more stringent threshold of \(\widehat{R} < 1.01\).
 
@@
 We can interpret \(AC\)-values as a measure of stationarity.
-If \(AC\)-values are close to zero, there is no dependence between subsequent samples within imputation chains.
-Negative \(AC\)-values indicate divergence within imputation chains.
-\textbf{Subsequent sampled values within each imputation chain are less alike.} Positive \(AC\)-values indicate recurrence.
-If \(\theta\)-values of subsequent iterations are similar, trending may occur.
+\(AC\)-values close to zero indicate independence between subsequent \(\theta\)s in imputation chains.
 Negative \(AC\)-values show no threat to the stationarity component of convergence.
@@
 As convergence diagnostic, the interest is therefore in positive \(AC\)-values.
-The magnitude of \(AC\)-values may be evaluated statistically, using a Wald test with \(AC=0\) as null hypothesis \citep{box15}.
+Positive \(AC\)-values occur when \(\theta\)-values of subsequent iterations are similar.
+When high \(\theta\)-values are followed by high \(\theta\)-values, and low \(\theta\)-values are followed by low \(\theta\)-values, we speak of recurrence.
+Recurrence within imputation chains indicates trending.
+Ergo, non-stationarity may be diagnosed when \(AC\) values are larger than zero.
 
@@
 
-Upon convergence, \(\widehat{R}=1\) and \(AC=0\), which are unlikely thresholds for MCMC algorithms, because of its convergence to a distribution.
-Even in the most converged state, the algorithm will show some signs of non-mixing and non-stationarity.
-Upon \emph{sufficient} convergence, the imputation chains will intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm (\(\widehat{R} > 1\)), and there may be some dependency between subsequent iterations of imputation chains (\(AC > 0\)).
-In practice, non-convergence is usually diagnosed when \(\widehat{R}\) \textgreater{} 1.2 or 1.1 or even 1.01.
-And a t-test is performed to assess whether \(AC\) is significantly different from zero.
+It is unlikely that an MCMC algorithm will achieve the ideal values of \(\widehat{R}=1\) and \(AC=0\).
+Because of its convergence to a distribution, the algorithm will show some signs of non-mixing and non-stationarity even in the most converged state.
+Upon approximate convergence, the imputation chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm (\(\widehat{R} > 1\)), and there is some dependency between subsequent iterations of imputation chains (\(AC > 0\)).
+In practice, non-convergence is therefore diagnosed when non-convergence diagnostics exeed a certain threshold.
+The conventional threshold to diagnose non-mixing is \(\widehat{R} > 1.2\) \citep{gelm92}.
+Another generally accepted threshold is \(\widehat{R} > 1.1\) \citep{gelm13}.
+And more recently, \citet{veht19} proposed an even more stringent threshold of \(\widehat{R} > 1.01\).
+The magnitude of \(AC\)-values may be evaluated statistically, using a Wald test with \(AC=0\) as null hypothesis \citep{box15}.
+\(AC\)-values that are significantly higher than zero indicate non-stationarity.
+\textbf{{[}think about one-sided test{]}}
 
@@
 
-\caption{Convergence diagnostics applied on the imputation algorithms of Figure 2. $\theta$ = chain mean in $y_{imp, m}$.}\label{fig:diagnostics}
+\caption{Convergence diagnostics applied on the imputation algorithms of Figure 2. $\theta$ = chain mean in $ \dot y_{\rm imp, \ell} $.}\label{fig:diagnostics}
 \end{figure}
@@
 
-\textbf{Describe panel C here and why it is not optimal{]}} The \(\widehat{R}\)-values in panel D do meet the requirements specified above.
+\textbf{Describe panel C here and why it is not optimal: NB.
+In panel C: De grotere variantie overall zorgt er kennelijk voor dat \(\widehat{R}\) groeit.
+De \(\widehat{R}\) geeft de foutieve boodschap dat convergentie steeds slechter gaat.
+Verklaring: de overdispersie assumptie klopt niet als er trend is.{]}} The \(\widehat{R}\)-values in panel D do meet the requirements specified above.
 \(\widehat{R}\) as computed conform \citet{veht19} does indeed indicate less signs of non-convergence as the number of iterations goes up.
@@
 
-The aim of the simulation study is to evaluate the impact of inducing non-convergence in the MICE algorithm on several quantities of scientific interest \(Q\).
+The aim of the simulation study is to evaluate the \textbf{impact of inducing non-convergence {[}``early stopping'' en ``proportion of complete cases''{]}} in the MICE algorithm on several quantities of scientific interest \(Q\).
 Subsequently we will evaluate how well \(\widehat{R}\) and \(AC\) perform in identifying the effects of non-convergence \textbf{{[}and performance of different thetas{]}}.
-And finally, we will formulate an informed advice on the requirements to safely assume sufficient convergence in practice.
-That is, we assume that the algorithm is sufficiently converged when each estimate \(\bar{Q}\) is an unbiased and confidence valid estimate of the corresponding estimand \(Q\).
+And finally, we will formulate an informed advice on the requirements to safely conclude sufficient convergence in practice.
+That is, we conclude that convergence is sufficient when each estimate \(\bar{Q}\) is an unbiased and confidence-valid estimate of the corresponding estimand \(Q\).
 \textbf{{[}and formulate advice on thetas?{]}}
 
-Non-convergence will be induced by: 1) increasing the missingness proportion \(p_{mis}\) in dataset \(y\) incrementally, and 2) terminating the imputation algorithm at a varying number of iterations \(T\).
+Non-convergence will be induced by: 1) increasing the missingness proportion \(p_{\rm mis}\) in dataset \(y\) incrementally, and 2) terminating the imputation algorithm at a varying number of iterations \(T\).
 The first set of simulation conditions---the missingness proportions---is chosen to reflect the difficulty of the missingness problem.
@@
 
-\textbf{{[}SEGWAY to this section{]} \(Q\)s are the targets that will be estimated in each simulation repetition, after first \emph{amputing} the complete data with varying missingness proportions, then imputing the missingness with a varying number of iterations, then estimating the \(Q\)s on the \(M\) completed datasets, and finally pooling the \(M\) estimates \(\hat{Q}\) to obtain a single \(\bar{Q}\) for each \(Q\).
+\textbf{{[}SEGWAY to this section{]} \(Q\)s are the targets that will be estimated in each simulation repetition, after first \emph{amputing} the complete data with varying missingness proportions, then imputing the missingness with a varying number of iterations, then estimating the \(Q\)s on the \(m\) completed datasets, and finally pooling the \(m\) estimates \(\hat{Q}_\ell\) to obtain a single \(\bar{Q}\) for each \(Q\).
 We try to predict the effect of non-convergence (i.e., the difference between \(Q\) and \(\bar{Q}\)) with the convergence diagnostics \(\widehat{R}\) and \(AC\), by applying these methods to several scalar summaries of the state space of the algorithm, \(\theta\)s.}
@@
 \item
-  We expect that simulation conditions with a high missingness proportion \(p_{mis}\) and a low number of iterations \(T\) will result in biased, invalid estimates of the quantities of scientific interest, \(Q\)s.
+  We expect that simulation conditions with a high missingness proportion \(p_{\rm mis}\) and a low number of iterations \(T\) will more often result in biased, invalid estimates of the quantities of scientific interest, \(Q\)s.
 \item
-  We hypothesize that \(\widehat{R}\) and \(AC\) will correctly identify signs of non-convergence in those simulation conditions where the \(\bar{Q}\)s are \emph{not} unbiased and confidence valid estimates of the \(Q\)s.
+  We hypothesize that \(\widehat{R}\) and \(AC\) will correctly identify signs of non-convergence in those simulation conditions where the \(\bar{Q}\)s are \emph{not} unbiased and confidence-valid estimates of the \(Q\)s.
 \item
@@
 The simulation set-up is summarized in the pseudo-code below.
-The complete \texttt{R} script of the simulation study is available from \href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy}{github.com/gerkovink/shinyMice}.
+The complete \texttt{R} script of the simulation study is available from \href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/}{github.com/gerkovink/shinyMice}.
 
@@
 A finite population of \(N=1000\) is simulated using the \texttt{mvtnorm} package \citep{mvtnorm}.
-Subsequently, a fourth variable is constructed as outcome \(Y\).
+Subsequently, a fourth variable is constructed as outcome variable \(Y\).
 For each unit \(i = 1, 2,..., N\), let
@@
 \]
-where \(\epsilon \sim \mathcal{N}(0, 100)\). From the complete set, we obtain the true values of several quantities of scientific interest, \(Q\)s.
+where \(\epsilon \sim \mathcal{N}(0, 100)\).
+This results in a complete set \(y\), with observed values for all units \(i\) on all variables \(j\) (\(j = Y, X_1, X_2, X_3\)).
+From the complete set we obtain the true values of the scientific estimands \(Q\).
 
@@
 And finally, the proportion of variance explained by the set of predictors: coefficient of determination \(r^2\) \textbf{(note that lower case \(r\) is used to avoid confusion with non-convergence diagnostic \(\widehat{R}\))}.
-The true values of these scientific estimands are calculated on the complete data with the \texttt{R} functions \texttt{base::mean()}, \texttt{stats::sd()}, and \texttt{stats::lm()} \citep{R}.
+The true values of these scientific estimands are calculated on the complete data.
 
@@
 In each simulation repetition, we \emph{ampute} the complete dataset five times to obtain five different incomplete datasets, \(y\)s.
-The ratio between \(y_{obs}\) and \(y_{mis}\) in each \(y\) depends on the missingness proportion: \(p_{mis} =.05,.25,.5,.75,.95\).
+The ratio between \(y_{\rm obs}\) and \(y_{\rm mis}\) in each \(y\) depends on the missingness proportion: \(p_{\rm mis} =.05,.25,.5,.75,.95\).
 We ampute the complete data with the \texttt{mice} package \citep[function \texttt{mice::ampute()};][]{mice}.
 We consider all possible univariate and multivariate patterns of missingness, conform a `missing completely at random' missingness mechanism \citep{rubin87}.
-I.e., the probability to be missing is the same for all \(N \times k\) cells in \(y\), conditional on the missingness proportion \(p_{mis}\).
+I.e., the probability to be missing is the same for all \(N \times k\) cells in \(y\), conditional on the missingness proportion \(p_{\rm mis}\).
 
 Missing datapoints in each incomplete dataset \(y\) are imputed with the \texttt{mice()} function \citep{mice}.
-All imputation procedures are performed with Bayesian linear regression imputation, and five imputation chains (\(M=5\)).
-Because \(M=5\), each run of the \texttt{mice()} function results in five sets of imputations: \(y_{imp,m}\), where \(m = 1, 2, ..., 5\).
+All imputation procedures are performed with Bayesian linear regression imputation, and five imputation chains (\(m=5\)).
+Because \(m=5\), each run of the \texttt{mice()} function results in five sets of imputations: \(\dot y_{\rm imp, \ell}\), where \(m = 1, 2, ..., 5\).
 The number of iterations in each run of the algorithm varies between simulation conditions (\(T=1,2,...,100\)).
 
-From the 5 imputations \(y_{imp,m}\) we obtain the chain means and chain variances for each variable \(j\).
+From the 5 imputations \(\dot y_{\rm imp, \ell}\) we obtain the chain means and chain variances for each variable \(j\).
 These will serve as scalar summary \(\theta\) to apply non-convergence diagnostics \(\widehat{R}\) and \(AC\) on.
-We then use \texttt{mice::complete()} to combine the imputed data \(y_{imp,m}\) with the observed data \(y_{obs}\), resulting in five completed datasets \{\(y_{obs}, y_{imp, m}\)\}.
+We then use \texttt{mice::complete()} to combine the imputed data \(\dot y_{\rm imp, \ell}\) with the observed data \(y_{\rm obs}\), resulting in five completed datasets \{\(y_{\rm obs}, y_{imp, m}\)\}.
 
-On the 5 completed datasets \{\(y_{obs}, y_{imp, m}\)\}, we:
+On the 5 completed datasets \{\(y_{\rm obs}, y_{imp, m}\)\}, we:
 
@@
 \item
-  Estimate \(\hat{Q}\)s for \(Q= \mu\) and \(Q=\sigma\) with \texttt{base::mean()}, and \texttt{stats::sd()} \citep{R}. We pool the \(\hat{Q}\)s with \texttt{base::mean()} to get \(\bar{Q}\)s.
+  Estimate \(\hat{Q}_\ell\)s for \(Q= \mu\) and \(Q=\sigma\) with \texttt{base::mean()}, and \texttt{stats::sd()} \citep{R}.
+We pool the \(\hat{Q}_\ell\)s with \texttt{base::mean()} to get \(\bar{Q}\)s.
 \item
-  Compute \(\lambda_1\), to use as \(\theta\) in the two convergence diagnostics.
-By definition, the first eigenvalue of the variance-covariance matrix, \(\lambda_1\), is equal to the variance of the principal component solution of each \{\(y_{obs}, y_{imp, m}\)\}, which we calculate with \texttt{stats::princomp()} \citep{R}.
+  Compute \(\lambda_{1, \ell}\), to use as \(\theta\) in the two convergence diagnostics.
+By definition, the first eigenvalue of the variance-covariance matrix, \(\lambda_{1, \ell}\), is equal to the variance of the principal component solution of each \{\(y_{\rm obs}, y_{imp, m}\)\}, which we calculate with \texttt{stats::princomp()} \citep{R}.
 \item
-  Perform multiple linear regression with \texttt{stats::lm()} \citep{R} to get the \(\hat{Q}\)s for \(Q=\beta\) and \(Q=r^2\).
-We pool the \(\hat{Q}\)s conform \citet{vink14} to get \(\bar{Q}\)s for the \(\beta\)s and \(r^2\) \textbf{{[}check finite pooling of \(r^2\), instead of \texttt{mice::pool.r.squared()}{]}}.
-Additionally, we use the \(\hat{Q}\)s of \(Q=\beta\) as \(\theta\) to apply the convergence diagnostics to.
+  Perform multiple linear regression with \texttt{stats::lm()} \citep{R} to get the \(\hat{Q}_\ell\)s for \(Q=\beta\) and \(Q=r^2\).
+We pool the \(\hat{Q}_\ell\)s conform \citet{vink14} to get \(\bar{Q}\)s for the \(\beta\)s and \(r^2\) \textbf{{[}check finite pooling of \(r^2\), instead of \texttt{mice::pool.r.squared()}{]}}.
+Additionally, we use the \(\hat{Q}_\ell\)s of \(Q=\beta\) as \(\theta\) to apply the convergence diagnostics to.
 \end{itemize}
@@
 
-We assess the impact of inducing non-convergence in the iterative imputation algorithm by comparing \(\bar{Q}\)s with \(Q\)s.
+We assess the performance of the iterative imputation algorithm in each simulation condition by comparing \(\bar{Q}\)s with \(Q\)s.
 For each \(Q\), we calculate bias as \(\bar{Q} - Q\).
@@
 \[\text{CI} = \bar{Q} \pm t_{(M-1)} \times SE_{\bar{Q}},\]
-where \(t_{(M-1)}\) is the quantile of a \(t\)-distribution with \(M-1\) degrees of freedom, and \(SE_{\bar{Q}}\) is the square root of the pooled variance estimate.
-If we obtain nominal coverage (CR = 95\%), we can conclude that the \(\bar{Q}\)s are confidence valid estimates of the \(Q\)s.
+where \(t_{(M-1)}\) is the quantile of a \(t\)-distribution with \(m-1\) degrees of freedom, and \(SE_{\bar{Q}}\) is the square root of the pooled variance estimate.
+If we obtain nominal coverage (CR = 95\%), we can conclude that the \(\bar{Q}\)s are confidence-valid estimates of the \(Q\)s.
 Finally, we inspect CI width (CIW): the difference between the lower and upper bound of the 95\% confidence interval around \(\bar{Q}\).
-CIW is of interest because too short CIs indicate under-estimation of the variance in \(\bar{Q}\), which may yield spurious inferences.
+CIW is of interest because it is a measure of efficiency.
+Under nominal coverage, short CIs are preferred.
+Too short CIs, however, indicate under-estimation of the variance in \(\bar{Q}\), \textbf{which may yield inferences with low statistical power if CIW is large}.
 
-With these performance measures, we also evaluate how well the non-convergence diagnostics \(\widehat{R}\) and \(AC\) can identify simulation conditions in which the inferences are affected by non-convergence, i.e., where \(\bar{Q}\) is \emph{not} an unbiased, confidence valid estimate of \(Q\).
+With these performance measures, we also evaluate how well the non-convergence diagnostics \(\widehat{R}\) and \(AC\) can identify simulation conditions in which the inferences are affected by non-convergence, i.e., where \(\bar{Q}\) is \emph{not} an unbiased, confidence-valid estimate of \(Q\).
 \textbf{{[}REPHRASE this:{]} For performance of univariate \(Q\)s, we apply the non-convergence diagnostics to univariate \(\theta\)s.
@@
 \item
-  For \(Q=\mu\), \(\theta\)s are chain means.
+  For \(Q=\mu\), \(\theta\)s are chain means \textbf{{[}ADD that these should be unbiased because MCAR!{]}}.
 \item
@@
 \item
-  For \(Q=r^2\), \(\theta\)s are the first eigenvalue of the variance-covariance matrix.
+  For \(Q=r^2\), \(\theta\)s are the first eigenvalue of the variance-covariance matrix, \(\lambda_{1,\ell}\).
 \end{itemize}
@@
 
-Our results show the impact of inducing non-convergence in an iterative imputation algorithm on several scientific estimates, and whether the non-convergence diagnostics \(\widehat{R}\) and \(AC\) identify the signs of non-convergence in the algorithm.
+Our results show the impact of \textbf{inducing non-convergence {[}``early stopping'' en ``proportion of complete cases'' {]}} in an iterative imputation algorithm on several scientific estimates, and whether the non-convergence diagnostics \(\widehat{R}\) and \(AC\) identify the signs of non-convergence in the algorithm.
 For reasons of brevity, we only discuss the non-convergence diagnostics for estimates with the worst performance in terms of bias.
@@
 
-\caption{Impact of non-convergence on statistical inferences.}\label{fig:Qs}
+\caption{Impact of non-convergence on statistical inferences.
+Depicted are the bias, coverage rate (CR) and confidence interval width (CIW) of the estimates $\bar{Q}$.
+We present the subset of $Q$s with the worst performance in terms of bias: the mean and standard deviation of $Y$, and the regression coefficient of $X_1$.}\label{fig:Qs}
 \end{figure}
 
-Figure \ref{fig:Qs} displays the influence of inducing non-convergence on the estimated quantities of scientific interest, \(Q\)s.
+Figure \ref{fig:Qs} displays the influence of \textbf{inducing non-convergence {[}``early stopping'' en ``proportion of complete cases''{]}} on the estimated quantities of scientific interest, \(Q\)s.
 \textbf{{[}Add panel labels, describe panels{]}} Simulation conditions that are impacted by non-convergence portray more extreme bias in the estimated \(Q\)s, and non-nominal coverages.
-Overall, we see that conditions with a higher proportion of missingness are affected more severely by non-convergence.
-Multivariate \(Q\)s are also substantially influenced by the number of iterations.
-This means that the algorithm has indeed not converged at \(t=1\), and converges gradually as \(t\) moves to \(T\).
-The point at which an additional iteration does not lead to an improvement in the estimates depends on quantity of scientific interest \(Q\) and the missingness proportion \(p_{mis}\).
-The highest number of iterations that is necessary to reach this point across all simulation conditions is \(T=7\).
-This implies that, under the current specifications, no more than seven iterations are necessary to obtain sufficient performance in terms of bias and confidence validity.
 
@@
 This implies that valid inferences may be obtained with as little as one iteration (\(T \geq 1\)).
-The magnitude of the bias in \(\bar{Q}\) depends solely on the missingness proportion \(p_{mis}\), with missingness proportions of 75\% and 95\% leading to more extreme biases than other missingness proportions.
-Unbiased estimates are obtained in all conditions where \(p_{mis}\leq.50\).
-Note however, that even in the conditions where \(p_{mis}\geq.75\), the bias is small: \(\mu_Y\) is maximally under-estimated by 0.02 units, while the true value of \(\mu_Y\) is 25.81 (\(\sigma_Y = 11.32\)).
+The magnitude of the bias in \(\bar{Q}\) depends solely on the missingness proportion \(p_{\rm mis}\), with missingness proportions of 75\% and 95\% leading to more extreme biases than other missingness proportions \textbf{{[}note that these are curious{]}}.
+Unbiased estimates are obtained in all conditions where \(p_{\rm mis}\leq.50\).
+Note however, that even in the conditions where \(p_{\rm mis}\geq.75\), the bias is small: \(\mu_Y\) is maximally under-estimated by 0.02 units, while the true value of \(\mu_Y\) is 25.81 (\(\sigma_Y = 11.32\)).
 We conclude that non-convergence does not substantially impact the validity of the inferences for \(Q=\mu\).
+\textbf{{[}To be expected, because MCAR{]}}
 
@@
 Interestingly, the bias in first iteration is actually \emph{less} extreme than for any other number of iterations.
-We may therefore conclude that no iteration at all is needed to obtain approximately unbiased estimates for \(Q=\sigma\).
-Completely unbiased estimates are only obtained in conditions where \(p_{mis}=.05\).
-Conditions where \(p_{mis}\geq.25\) are impacted by non-convergence in order of increasing missingness proportion.
-Similar to the bias for \(Q=\mu\) the magnitude of the bias in \(Q=\sigma\) is not substantial in any condition.
-We conclude that both of the univariate \(Q\)s are practically unaffected by non-convergence.
+\textbf{{[}Due to MCAR and randomly drawn initial values?{]}} We may therefore conclude that no iteration at all is needed to obtain approximately unbiased estimates for \(Q=\sigma\).
+Completely unbiased estimates are only obtained in conditions where \(p_{\rm mis}=.05\).
+Conditions where \(p_{\rm mis}\geq.25\) are impacted by non-convergence in order of increasing missingness proportion.
+Similar to the bias for \(Q=\mu\) the magnitude of the bias in \(Q=\sigma\) is negligible.
+We therefore conclude that neither of the univariate \(Q\)s is affected substantially by non-convergence.
 
@@
 The magnitude of the bias clearly follows the incremental order of the missingness proportions.
-We, therefore, observe approximately unbiased estimates at a different number of iterations for each missingness proportion.
-For \(p_{mis}\leq.25\), the minimum number of iterations is two (\(T\geq2\)), while \(p_{mis}=.50\) requires \(T\geq4\), and \(p_{mis}=.75\) requires \(T\geq5\).
-The highest number of iterations needed to obtain valid inferences is seven (\(p_{mis}=.95\); \(T\geq7\)).
-Unbiased estimates, however, may only be obtained if the missingness proportion is at most 25\%.
-\textbf{{[}interpret sign of bias{]}}
+Completely unbiased estimates may only be obtained if the missingness proportion is at most 25\%.
+Approximately unbiased estimates are observed for any missingness proportion.
+For \(p_{\rm mis}\leq.25\), the minimum number of iterations is two (\(T\geq2\)), for \(p_{\rm mis}=.50\) it is \(T\geq4\), and \(p_{\rm mis}=.75\) requires \(T\geq5\).
+The highest number of iterations needed to obtain valid inferences is seven (\(p_{\rm mis}=.95\); \(T\geq7\)).
+\textbf{{[}interpret sign of bias?
+ADD what this implies: dat - zelfs bij een gebiasede methode - je maar 7 iteraties nodig hebt om het eindresultaat te bereiken.{]}}
 
@@
 For the estimated regression coefficients, we consider bias, confidence interval width (CIW), and coverage rate (CR) as performance measures.
-Unbiased estimates are obtained in conditions where \(p_{mis}\leq.50\) and \(T\geq3\).
-Approximately confidence valid estimates are obtained in any condition where \(T\geq3\)---despite of persistant bias in conditions where \(p_{mis}\geq.75\).
-These approximately nominal coverages are explained by the CIWs: conditions with higher missingness proportions result in more uncertainty in the estimates (i.e., wider CIs), which diminish the effect of the bias.
-Confidence valid inferences may be obtained when \(T\geq3\), while we need at least seven iterations to reach valid inferences in terms of bias.
+\textbf{{[}Split panels{]}} Unbiased estimates are obtained in conditions where \(p_{\rm mis}\leq.50\) and \(T\geq3\).
+Approximately confidence-valid estimates are obtained in any condition where \(T\geq3\)---despite of persistant bias in conditions where \(p_{\rm mis}\geq.75\).
+These approximately nominal coverages are explained by the CIWs: as expected, conditions with higher missingness proportions result in more uncertainty in the estimates (i.e., wider CIs), which diminish the effect of the bias.
+confidence-valid inferences may be obtained when \(T\geq3\), while we need at least seven iterations to reach valid inferences in terms of bias.
 Continued iteration after \(T\geq7\) seems futile.
+\textbf{{[}add max T for other betas?{]}}
+
+\hypertarget{summary}{%
+\subsubsection{Summary}\label{summary}}
+
+Overall, we see that conditions with a higher proportion of missingness are affected more severely by non-convergence.
+Multivariate \(Q\)s are also substantially influenced by the number of iterations.
+This means that the algorithm has indeed not converged at \(t=1\), and converges gradually as \(t\) moves to \(T\).
+The point at which an additional iteration does not lead to an improvement in the estimates depends on quantity of scientific interest \(Q\) and the missingness proportion \(p_{\rm mis}\).
+The highest number of iterations that is necessary to reach this point across all simulation conditions is \(T=7\).
+This implies that, under the current specifications, no more than seven iterations are necessary to obtain sufficient performance in terms of bias and confidence-validity.
 
@@
 
-We evaluate the non-convergence diagnostics \(\widehat{R}\) and \(AC\) against the performance measures described above (i.e., bias and confidence validity of the quantities of scientific interest \(Q\)).
+We evaluate the non-convergence diagnostics \(\widehat{R}\) and \(AC\) against the performance measures described above (i.e., bias and confidence-validity of the quantities of scientific interest \(Q\)).
 The diagnostics should identify those simulation conditions that are negatively impacted by non-convergence.
-We expect the most signs of non-convergence in conditions where \emph{sufficient} estimates are not obtained (high missingness proportions, \(T<7\)), and at least some signs of non-convergence in conditions where estimates are not approximately unbiased (\(p_{mis}=.95\), any number of iterations).
+We expect the most signs of non-convergence in conditions where valid estimates are not obtained (high missingness proportions, \(T<7\)), and at least some signs of non-convergence in conditions where estimates are not approximately unbiased (\(p_{\rm mis}=.95\), any number of iterations).
 
 Figure \ref{fig:rhats-acs} displays the non-convergence diagnostics \(\widehat{R}\) and \(AC\) for each of the four \(\theta\)s under consideration.
-\textbf{{[}Add panel labels and explain how to read the plots{]}} In general, the diagnostics identify more signs of non-convergence in simulation conditions with a lower number of iterations.
-The missingness proportions, however, only seem to impact multivariate convergence.
-When \(\widehat{R}\) and \(AC\) are applied to univariate \(\theta\)s, they identify the same amount of non-convergence for any missingness proportion.
-When the diagnostics are applied to multivariate \(\theta\)s, they generally identify more signs of non-convergence for higher missingness proportions.
-
-As expected, none of the simulation conditions that we consider resulted in complete convergence of the algorithm (defined as \(\widehat{R}=1\) and \(AC=0\)).
-However, the thresholds used in practice to diagnose non-convergence seem too conservative for imputation purposes.
-\(\widehat{R}\) and \(AC\) identify signs of non-convergence in conditions where the \(\bar{Q}\)s are unbiased and confidence valid estimates of the \(Q\)s.
-The most recent recommended threshold for diagnosing non-convergence with \(\widehat{R}\) is never overcome in this study (all \(\widehat{R}\)\textgreater1.01).
-Apparently, this threshold is too stringent for iterative imputation algorithms, and will not be discussed any further.
+\textbf{{[}Add panel labels and explain how to read the plots{]}}
 
@@
 
+\textbf{{[}As expected, \(\widehat{R}\) lowers with the number of iterations.
+At a cut-point of 1.2 , we have convergence at about 8 iteration, whereas we need 12 and 25 iteration for more conservative cut-off 1.1 and 1.05.
+We do not reach the very strict 1.01.{]}}
+
 We evaluate the performance of the non-convergence diagnostics applied to chain means against the bias in \(Q=\mu\).
@@
 This does not correspond to the impact of non-convergence in the univariate \(Q\)s.
-\textbf{Compared to the multivariate \(Q\)s {[}add before that we evaluate the thetas against all Qs?{]}, \(\widehat{R}\) and \(AC\) applied to the chain means fail to identify the non-convergence in conditions where \(p_{mis}=.95\).}
+\textbf{Compared to the multivariate \(Q\)s {[}add before that we evaluate the thetas against all Qs?{]}, \(\widehat{R}\) and \(AC\) applied to the chain means fail to identify the non-convergence in conditions where \(p_{\rm mis}=.95\).}
 
@@
 
-\hypertarget{thetahatq.}{%
-\subsubsection{\texorpdfstring{\(\theta=\hat{Q}\).}{\textbackslash theta=\textbackslash hat\{Q\}.}}\label{thetahatq.}}
+\hypertarget{thetahatq_ell.}{%
+\subsubsection{\texorpdfstring{\(\theta=\hat{Q}_\ell\).}{\textbackslash theta=\textbackslash hat\{Q\}\_\textbackslash ell.}}\label{thetahatq_ell.}}
+
+\textbf{{[}We zien weer hetzelfde voor laag-medium, beetje meer voor extreme misisnh data.{]}}
 
@@
 The thresholds to diagnose non-convergence are overcome in different conditions, depending on the number of iterations and the missingness proportion.
-The most lenient threshold (\(\widehat{R}>1.2\)) is overcome at \(T\geq9\) for \(p_{mis} = 0.05\), and \(T\geq12\) for \(p_{mis} = 0.95\).
+The most lenient threshold (\(\widehat{R}>1.2\)) is overcome at \(T\geq9\) for \(p_{\rm mis} = 0.05\), and \(T\geq12\) for \(p_{\rm mis} = 0.95\).
 The conventional \(\widehat{R}>1.1\) is surpassed at \(T\geq12\) and \(T\geq24\), respectively.
 
-For the autocorrelations applied to the \(\hat{Q}\)s, we see an even stronger effect of the missingness proportions.
-Especially the conditions where \(p_{mis} = 0.75\) and \(p_{mis} = 0.95\) are identified as non-stationary.
+For the autocorrelations applied to the \(\hat{Q}_\ell\)s, we see an even stronger effect of the missingness proportions.
+Especially the conditions where \(p_{\rm mis} = 0.75\) and \(p_{\rm mis} = 0.95\) are identified as non-stationary.
 The threshold used to evaluate whether \(AC=0\), however, is only reached in conditions with at least ten iterations.
@@
 However, the critical value for diagnosing non-convergence with \(AC\) decreases as a function of the number of iterations.
-The conclusion to identify non-stationarity in conditions where \(T\geq10\) and \(p_{mis} = 0.95\) might therefore be spurious.
+The conclusion to identify non-stationarity in conditions where \(T\geq10\) and \(p_{\rm mis} = 0.95\) might therefore be spurious.
 
-Compared to the bias in \(\bar{Q}\), the diagnostics applied to the \(\hat{Q}\)s over-estimate the severity of the non-convergence.
+Compared to the bias in \(\bar{Q}\), the diagnostics applied to the \(\hat{Q}_\ell\)s over-estimate the severity of the non-convergence.
 
-\hypertarget{thetalambda_1.}{%
-\subsubsection{\texorpdfstring{\(\theta=\lambda_1\).}{\textbackslash theta=\textbackslash lambda\_1.}}\label{thetalambda_1.}}
+\hypertarget{summary-1}{%
+\subsubsection{Summary}\label{summary-1}}
+
+In general, the diagnostics identify more signs of non-convergence in simulation conditions with a lower number of iterations.
+The missingness proportions, however, only seem to impact multivariate convergence.
+When \(\widehat{R}\) and \(AC\) are applied to univariate \(\theta\)s, they identify the same amount of non-convergence for any missingness proportion.
+When the diagnostics are applied to multivariate \(\theta\)s, they generally identify more signs of non-convergence for higher missingness proportions.
+
+As expected, none of the simulation conditions that we consider resulted in complete convergence of the algorithm (defined as \(\widehat{R}=1\) and \(AC=0\)).
+However, the thresholds used in practice to diagnose non-convergence seem too conservative for imputation purposes.
+\(\widehat{R}\) and \(AC\) identify signs of non-convergence in conditions where the \(\bar{Q}\)s are unbiased and confidence-valid estimates of the \(Q\)s.
+The most recent recommended threshold for diagnosing non-convergence with \(\widehat{R}\) is never overcome in this study (all \(\widehat{R}\)\textgreater1.01).
+Apparently, this threshold is too stringent for iterative imputation algorithms, and will not be discussed any further.
+
+\hypertarget{thetalambda_1-ell.}{%
+\subsubsection{\texorpdfstring{\(\theta=\lambda_{1, \ell}\).}{\textbackslash theta=\textbackslash lambda\_\{1, \textbackslash ell\}.}}\label{thetalambda_1-ell.}}
 
@@
 Similar to the results the the regression coefficients as \(\theta\)s, we see effects of both the missingness proportion and number of iterations in each simulation condition.
-The most signs of non-convergence are identified in conditions where \(p_{mis}\) is large and \(T\) is small.
+The most signs of non-convergence are identified in conditions where \(p_{\rm mis}\) is large and \(T\) is small.
 
-Notwithstanding the similarities with \(\theta=\hat{Q}\), there is a big difference in \(\widehat{R}\)-values for conditions where \(T<4\).
+Notwithstanding the similarities with \(\theta=\hat{Q}_\ell\), there is a big difference in \(\widehat{R}\)-values for conditions where \(T<4\).
 In these conditions, \(\widehat{R}\) behaves oppositely of our expectation: conditions with the lowest missingness proportions yield the highest \(\widehat{R}\)-values.
 In conditions with a higher number of iterations, this artifact disappears.
-This suggests that the combination of \(\widehat{R}\) and \(\lambda_1\) is only appropriate for imputation algorithms where the number of iterations at least four.
+This suggests that the combination of \(\widehat{R}\) and \(\lambda_{1, \ell}\) is only appropriate for imputation algorithms where the number of iterations at least four.
 Interpreting the \(\widehat{R}\) values of conditions where \(T<3\) may lead to the incorrect conclusion that only conditions with \emph{low} missingness proportions show signs of non-convergence.
@@
 Using this cut-off we conveniently ignore the initial `dip' at \(T=3\) as well.
-The traditional threshold \(\widehat{R} > 1.2\) is overcome in conditions where \(p_{mis} = .95\) and \(T\geq10\), and for all other missingness proportions in conditions where \(T\geq8\).
+The traditional threshold \(\widehat{R} > 1.2\) is overcome in conditions where \(p_{\rm mis} = .95\) and \(T\geq10\), and for all other missingness proportions in conditions where \(T\geq8\).
 With the \(\widehat{R} > 1.1\) threshold, the number of iterations are \(T\geq20\) and \(T\geq16\), respectively.
@@
 The \(AC\)-values largely depend on the missingness proportion in each condition.
-Conditions where \(p_{mis}=.95\) show statistically significant autocorrelations for any number of iterations.
-The other extreme, conditions where \(p_{mis}=.05\), does not yield \(AC\) values above the threshold.
+Conditions where \(p_{\rm mis}=.95\) show statistically significant autocorrelations for any number of iterations.
+The other extreme, conditions where \(p_{\rm mis}=.05\), does not yield \(AC\) values above the threshold.
 The \(AC\) values with this missingness proportion are equivalent for all conditions where \(T\geq6\), which suggests that the algorithm is as stationary as it can be.
@@
 In comparison to the \(Q\)s, the diagnostics again over-estimate the signs of non-convergence.
-Conditions with low missingness proportions (e.g., 5\% or 25\% of the complete data) yield unbiased, confidence valid estimates with as little as two iterations.
+Conditions with low missingness proportions (e.g., 5\% or 25\% of the complete data) yield unbiased, confidence-valid estimates with as little as two iterations.
 The convergence diagnostics, however, indicate improved convergence up-to eight iterations according to \(AC\) and thirty to forty iterations according to \(\widehat{R}\).
-The model-independent \(\theta\) \(\lambda_1\) seems as well-equipped to diagnose signs of non-convergence as \(\theta=\hat{Q}\).
+The model-independent \(\theta\) \(\lambda_{1, \ell}\) seems as well-equipped to diagnose signs of non-convergence as \(\theta=\hat{Q}_\ell\).
 The univariate \(\theta\)s do not differentiate between bias induced by the missingness proportions.
@@
 
-In short, iterative imputation algorithms may yield approximately unbiased, confidence valid estimates under different levels of induced non-convergence.
+We arrive back at the original questions that we posed.
+
+\begin{itemize}
+\item
+  Non-convergence can be identified with non-convergence diagnostics \(\widehat{R}\) and \(AC\).
+But common thresholds are too conservative.
+Why?
+Part of the distribution is already determined by the observed data.
+in regular MCMC the entire distribution is unknown (\textbf{priors??}).
+In iterative imputation algorithms we have the same characteristics as MCMC in the limit, but because we combine the known distribution with the unknown distribution, convergence is reached much sooner.
+\item
+  Inferences are unbiased much quicker than the algorithm has converged.
+All things univariate are converged almost intantly after initiating the iterative algorithm with a valid imputation model.
+But less quick with multivariate parameters.
+\end{itemize}
+
+In short, iterative imputation algorithms may yield approximately unbiased, confidence-valid estimates under different levels of induced non-convergence.
 Univariate estimates seem robust against terminating the algorithm early: There is no clear effect of the number of iterations on the bias in estimated means and standard deviations.
-The bias in these \(Q\)s was only impacted by the missingness proporion \(p_{mis}\), with higher \(p_{mis}\) resulting in more bias.
-Bias in multivariate estimates depended on both \(p_{mis}\) and the number of iterations.
+The bias in these \(Q\)s was only impacted by the missingness proporion \(p_{\rm mis}\), with higher \(p_{\rm mis}\) resulting in more bias.
+Bias in multivariate estimates depended on both \(p_{\rm mis}\) and the number of iterations.
 However, the only effect of the number of iterations was observed when the algorithm was terminated at seven iterations or earlier.
@@
 Yet, the non-convergence diagnostics \(\widehat{R}\) and \(AC\) indicated that the iterative imputation algorithm did not reach a stable state before twenty to thirty iterations.
-\textbf{{[}MERGE or remove{]} \(\widehat{R}\) and autocorrelation indicate that there are signs of non-convergence in the algorithm up-to at least twenty iterations, while unbiased, confidence valid estimates may be obtained with as little as one iteration.
+\textbf{{[}MERGE or remove{]} \(\widehat{R}\) and autocorrelation indicate that there are signs of non-convergence in the algorithm up-to at least twenty iterations, while unbiased, confidence-valid estimates may be obtained with as little as one iteration.
 This observation is in agreement with }one of {[}still add the others!{]}** the simulation hypotheses: \(\widehat{R}\) over-estimates the severity of non-convergence when applied to MI procedures.**
@@
 \item
-  Is the point of \emph{sufficient} convergence (at which the \(\bar{Q}\)s are unbiased and confidence valid estimates of the \(Q\)s) the same point as identified by the non-convergence diagnostics?
+  Is the point of \emph{sufficient} convergence (at which the \(\bar{Q}\)s are unbiased and confidence-valid estimates of the \(Q\)s) the same point as identified by the non-convergence diagnostics?
 (NOPE, they are too conservative).
 \item
-  Is one of the thetas better? (YES, while it doesn't really matter as long as it's multivariate, \(\lambda_1\) is model-independent and therefore preferred).
+  Is one of the thetas better? (YES, while it doesn't really matter as long as it's multivariate, \(\lambda_{1, \ell}\) is model-independent and therefore preferred).
 \item
@@
 \item
-  Monitor non-convergence diagnostics \(\widehat{R}\) and \(AC\). Use the thresholds for \(\widehat{R}>1.2\) and autocorrelation \textbf{???}. Keep iterating until these thresholds are reached.
+  Monitor non-convergence diagnostics \(\widehat{R}\) and \(AC\).
+Do not use the traditional calculation for \(\widehat{R}\), or the \texttt{R} function \texttt{stats::acf()} to compute \(AC\).
+Instead, calculate \(\widehat{R}\) conform \citet{veht19} and compute autocorrelations manually (see e.g., \href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy}{github.com/gerkovink/shinyMice}).
 \item
@@
 \item
-  Do not use the traditional calculation for \(\widehat{R}\), or the \texttt{R} function \texttt{stats::acf()} to compute \(AC\).
-Instead, calculate \(\widehat{R}\) conform \citet{veht19} and compute autocorrelations manually (see e.g., \href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy}{github.com/gerkovink/shinyMice}).
+  Use the threshold \(\widehat{R}>1.2\) for mixing, and assess stationarity by plotting the autocorrelation over iterations.
+Keep iterating until the threshold for mixing is reached, and until a reasonably decreasing asympototic \(AC\) value is obtained.
+At that point, inferences are unlikely to improve by continued iteration.
 \end{enumerate}
@@
 To conclude, in the case of drawing inference from incomplete data, convergence of the iterative imputation algorithm is often a convenience but not a necessity.
+\textbf{{[}Our study found that - in the cases considered - inferential validity was achieved after 5-10 iterations, much earlier than indicated by the \(\widehat{R}\) and AC diagnostics.
+Of course, it never hurts to iterate longer, but such calculations hardly bring added value.{]}}
 

### 2020-05-07 3.Thesis/1.Execute.R

@@ -27,8 +27,8 @@ source("3.Thesis/Functions/AC_supplement.R")
 
 # simulation parameters
 populationsize <- 1000 #n of simulated dataset
-n.iter <- 100 #nr of iterations (varying 1:n.iter)
-n.sim <- 1000 #nr of simulations per iteration value
+n.iter <- 20 #nr of iterations (varying 1:n.iter)
+n.sim <- 10 #nr of simulations per iteration value
 p.miss <- c(.05, .25, .5, .75, .95)
 true.effect <- 2 #regression coefficient to be estimated
 

### 2020-05-09 3.Thesis/Writeup/2.Manuscript.Rmd
Missing only some parts of the methods and discussion

@@
 
-# Overview of sections
-
-Intro
-
-- This is missingness
-
-- This is how we solve missingness
-
-- For that we need an algorithm, but what if it has not converged?
-
-- That is what we solve in this paper
-
-Algorithmic convergence
-
-- What is algorithmic convergence?
-
-- When is it converged? Stationarity versus mixing.
-
-- How is convergence currently investigated? And how should it be investigated? Multivariate versus univariate.
-
-Methods
-
-- Which methods are there? And how do they relate to mixing and stationarity?
-
-- $\widehat{R}$
-
-- AC
-
-- We can use these on any $\theta$
-
-The Simulation study, Results, Discussion.
+<!-- # Overview of sections -->
+
+<!-- Intro -->
+
+<!-- - This is missingness -->
+
+<!-- - This is how we solve missingness -->
+
+<!-- - For that we need an algorithm, but what if it has not converged? -->
+
+<!-- - That is what we solve in this paper -->
+
+<!-- Algorithmic convergence -->
+
+<!-- - What is algorithmic convergence? -->
+
+<!-- - When is it converged? Stationarity versus mixing. -->
+
+<!-- - How is convergence currently investigated? And how should it be investigated? Multivariate versus univariate. -->
+
+<!-- Methods -->
+
+<!-- - Which methods are there? And how do they relate to mixing and stationarity? -->
+
+<!-- - $\widehat{R}$ -->
+
+<!-- - AC -->
+
+<!-- - We can use these on any $\theta$ -->
+
+<!-- Then Simulation study, Results, Discussion. -->
 
@@
 Missing data is not only ubiquitous across science, but treating it can also be a tedious task.
-If a dataset contains just one incomplete observation, calculations are not defined, **hence** statistical models cannot do fitted to the data.
+If a dataset contains just one incomplete observation, calculations are not defined, and statistical models cannot be fitted to the data.
 To circumvent this, many statistical packages employ list-wise deletion by default (i.e., ignoring incomplete observations).
@@
 Since $Q$ is estimated on each completed dataset, $m$ separate $\hat{Q}_\ell$-values are obtained.
-Finally, the $m$ $\hat{Q}_\ell$-values are combined into a single pooled estimate $\bar{Q}$.
-The premise of multiple imputation is that $\bar{Q}$ is an unbiased and confidence-valid estimate of the true---but unobserved---inference [@rubin96].
+Finally, the $\hat{Q}_\ell$-values are combined into a single pooled estimate $\bar{Q}$.
+The premise of multiple imputation is that $\bar{Q}$ is an unbiased and confidence-valid estimate of the true---but unobserved---scientific estimand $Q$ [@rubin96].
 
@@
 
-```{r diagram, out.width="\\linewidth", include=TRUE, fig.cap="Scheme of the main steps in multiple imputation (where $m=3$)---from an incomplete dataset, to $m$ multiply imputed datasets, to $m$ estimated quantities of scientific interest $\\hat{Q}$s, to a single pooled estimate $\\bar{Q}$."}
+```{r diagram, out.width="\\linewidth", include=TRUE, fig.cap="Scheme of the main steps in multiple imputation---from an incomplete dataset, to $m=3$ multiply imputed datasets, to $m=3$ estimated quantities of scientific interest $\\hat{Q}$s, to a single pooled estimate $\\bar{Q}$."}
 knitr::include_graphics("../Figures/DiagramMI.pdf") #adapted from [@buur18] par. 1.4.1.
@@
 
-A popular method to obtain imputations is to use the 'Multiple Imputation by Chained Equations' algorithm, shorthand 'MICE'[@mice].
-With MICE, imputed values $\dot y_{\rm imp, \ell}$ are drawn from the posterior predictive distribution of the missing values $y_{\rm mis}$.
+A popular method to obtain imputations  $\dot y_{\rm imp, \ell}$ is to use the 'Multiple Imputation by Chained Equations' algorithm, shorthand 'MICE'[@mice].
+With MICE, imputed values are drawn from the posterior predictive distribution of the missing values $y_{\rm mis}$.
 The algorithm is named after the iterative algorithmic procedure by which imputed values are generated: a multivariate distribution is obtained by iterating over a sequence of univariate imputations.
@@
 These remain open questions, since the convergence properties of iterative imputation algorithms have not been systematically studied [@buur18].
-There is no scientific consensus on how to evaluate convergence of imputation algorithms [@taka17].
-**Moreover, the convergence of chained equations algorithms such as MICE under certain default imputation models (e.g., 'predictive mean matching') is an entirely open question [@murr18].** Therefore, algorithmic convergence should be monitored carefully.
-
-Currently, the recommended practice for evaluating the convergence of iterative imputation algorithms is through visual inspection.
-After running the imputation algorithm for a certain number of iterations, its convergence is monitiored by plotting a scalar summary of the state space of the algorithm $\theta$ against the iteration number.
-Typically, the $\theta$s under evaluation are chain means and variances---the average and the variance of the imputed values per imputation.
-Monitoring convergence by inspecting traceplots of these $\theta$s may be undesirable for several reasons on two counts: 1) it may be challenging to the untrained eye, and 2) only severely pathological cases of non-convergence may be diagnosed [@buur18, $\S$ 6.5.2] and, 3) there is not an objective measure that quantifies convergence.
-Therefore, a quantitative, diagnostic evaluation of convergence would be preferred---although not straightforward.
+There is no scientific consensus on how to evaluate the convergence of imputation algorithms [@taka17].
+Moreover, the behavoir of such algorithms under certain default imputation models (e.g., 'predictive mean matching') is an entirely open question [@murr18].
+**Therefore, algorithmic convergence should be monitored carefully, **although not straightforward.
 Iterative imputation algorithms such as MICE are special cases of Markov chain Monte Carlo (MCMC) methods.
@@
 We define several diagnostics and evaluate how these diagnostics could be appropriate for iterative imputation applications.
-We then adress the impact of inducing non-convergence in iterative imputation algorithms by means of model-based simulation in `R` [@R].
-The aim of the simulation is to determine whether unbiased, confidence-valid inferences may be obtained under different levels of non-convergence (**and at which point convergence may safely be assumed**).
-Additionally, we will assess how well different non-convergence diagnostics perform in identifying signs of non-convergence (**and on which scalar summary of the state space of the algorithm these diagnostics should be applied**).
+We then adress the impact of inducing non-convergence in iterative imputation algorithms through model-based simulation in `R` [@R].
+The aim of the simulation study is to determine whether unbiased, confidence-valid inferences may be obtained if the algorithm has not (yet) converged.
+And additionally, to formulate an informed advice on when approxiamte convergence may be safely concluded.
+We will therefore assess the performance of different methods to identify non-convergence.
 We translate the results of the study into guidelines for applied researchers.
-**[add more societal relevance: what is the consequence of non-convergence?
-see example of non-convergence in Figure 2 (e.g., biased estimates, invalid inference, etc.)]** For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular `mice` package [@mice] in `R` [@R].
+**ADD that these guidelines may aid to draw valid inference from incomplete data?]** For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular `mice` package [@mice] in `R` [@R].
 
@@
 The data value of unit $i$ ($i = 1, 2, \dots, n$) on variable $j$ ($j = 1, 2, \dots, k$) may be either observed or missing.
-The number of units $i$ in dataset $y$ with at least one missing value, divided by the total number of units $n$, is called the missingness proportion $p_{\rm mis}$.
+The number of units $i$ with at least one missing value, divided by the total number of units $n$, is called the missingness proportion $p_{\rm mis}$ in dataset $y$.
 The collection of observed data values in $y$ is denoted by $y_{\rm obs}$; the missing part of $y$ is referred to as $y_{\rm mis}$.
 For each datapoint in $y_{\rm mis}$, we sample $m \times T$ plausible values, where $m$ is the number of imputations ($\ell = 1, 2, \dots, m$) and $T$ is the number of iterations ($t = 1, 2, \dots, T$) in the imputation algorithm.
-The collection of samples between the initial value (at $t=1$) and the final imputed value (at $t=T$) will be referred to as an 'imputation chain'.
+The state space of the algorithm at a certain iteration $t$ may be summarized by a scalar summary $\theta$ (e.g., **the average of the imputed values**).
+The collection of $\theta$-values between the **initial value** (at $t=1$) and the final imputed value (at $t=T$) will be referred to as an 'imputation chain'.
 <!-- **[If not defined elsewhere, add: $\theta$s are scalar summaries of the algorithm at a certain iteration (e.g., chain means; the average of the imputed values in each imputation chain).] ** -->
@@
 There are two requirements for convergence of iterative algorithms: mixing and stationarity [@gelm13].
-Without mixing, imputation chains do not intermingle nicely, indicating that the distribution of imputed values differs across imputations.
-Chains may be 'stuck' at a local optimum, instead of sampling imputed values from the entire predictive posterior distribution of the missing values.
-This may cause under-estimation of the variance within chains, which results in spurious, invalid inferences.
+In iterative imputation algorithms mixing implies that imputation chains intermingle nicely, and stationarity is characterized by **the absence of trending** within chains.
+If one of these two requirements is not met, there is non-convergence in the algorithm.
+Without mixing, chains may be 'stuck' at a local optimum, instead of sampling imputed values from the entire predictive posterior distribution of the missing values.
+The distribution of imputed values then differs across imputations.
+This may cause under-estimation of the variance between chains, which results in spurious, invalid inferences.
 Without stationarity, there is trending within imputation chains.
 Trending implies that further iterations would yield a systematically lower or higher set of imputations.
-Iterative imputation algorithms that have not yet reached stationarity, may thus yield biased estimates.
+Iterative imputation algorithms that have not (yet) reached stationarity, may thus yield biased estimates.
 
@@
 
-**[SEGWAY to this section]** To illustrate what non-convergence looks like in iterative imputation algorithms, we reproduce an example from van Buuren [-@buur18, $\S$ 6.5.2].
-Figure \ref{fig:non-conv} displays a subset of the example: a variable that we will call $j$.
-We see two traceplots of the chain means for $j$---i.e., the average of the imputed values for $j$ in each imputation $\dot y_{\rm imp, \ell}$, plotted against the iteration number.
-Each line portrays one imputation.
-The plot on the left-hand side of the figure shows typical convergence of an iterative imputation algorithm.
+To illustrate what non-mixing and non-stationarity looks like in iterative imputation algorithms, we reproduce an example from van Buuren [-@buur18, $\S$ 6.5.2].
+Figure \ref{fig:non-conv} displays two scenarios from the example.
+The graph on the left-hand side of the figure shows typical convergence of an iterative imputation algorithm.
 The right-hand side displays pathological non-convergence, induced by purposefully mis-specifying the imputation model.
-**In the typical convergence situation, the imputation chains intermingle nicely and there is little to no trending.
-In the non-convergence plot, there is a lot of trending and some chains do not intermingle.** Importantly, the chain means at the last iteration (the imputed value per imputation $\ell$) are very different between the two plots.
+Each line portrays one imputation chain: scalar summary $\theta$ plotted against the iteration number $t$.
+The $\theta$ in this example is the chain mean of a variable that we will call $j$ (i.e., the average of the imputed values for $j$ in each imputation $\dot y_{\rm imp, \ell}$).
+**In the typical convergence scenario, the imputation chains intermingle nicely and there is little to no trending.
+In the non-convergence scenario, there is a lot of trending and some chains do not intermingle.** Importantly, the chain means at the last iteration (the imputed value per imputation $\ell$) are very different between the two plots.
 The algorithm with the mis-specified model yields imputed values that are on average a factor two larger than those of the typically converged algorithm.
-That means that non-convergence has a impact on the mean of $j$ in $\dot y_{\rm imp, \ell}$.
-This difference (presumably) translates to bias in the $m$ completed data estimates $\hat{\mu}_{j,\ell}$, and consequently also to the pooled estimate $\bar{\mu}_j$.
-Non-convergence thus leads to biased estimates.
+Non-convergence thus influences the mean of $j$ in $\dot y_{\rm imp, \ell}$.
+This difference (presumably) translates to the $m$ completed data estimates $\hat{\mu}_{j,\ell}$, and consequently to the pooled estimate $\bar{\mu}_j$.
+Ergo, non-convergence leads to biased estimates.
 This shows the importance of reaching converged states in iterative imputation algorithms.
@@
 
-**[MERGE with section below]**.
-Inspecting non-convergence through traceplots of chain means and chain variances may be insufficient because they are univariate summaries of the state space of the algorithm, while MICE is not only concerned with a single column, but the entire multivariate distribution of $y_{\rm imp}$.
+**[MERGE with sections below]** Currently, the recommended practice for evaluating the convergence of the MICE algorithm is through visual inspection.
+After running the imputation algorithm for a certain number of iterations, its convergence is monitiored by plotting a scalar summary of the state space of the algorithm $\theta$ against the iteration number.
+Typically, the $\theta$s under evaluation are chain means and variances---the average and the variance of the imputed values per imputation.
+Monitoring convergence by inspecting traceplots of these $\theta$s may be undesirable for several reasons: 1) it may be challenging to the untrained eye, and 2) only severely pathological cases of non-convergence may be diagnosed [@buur18, $\S$ 6.5.2] and, 3) there is not an objective measure that quantifies convergence.
+
+**[MERGE with section above and below]** Inspecting non-convergence through traceplots of chain means and chain variances may be insufficient because they are univariate summaries of the state space of the algorithm, while MICE is not only concerned with a single column, but the entire multivariate distribution of $y_{\rm imp}$.
 Ideally, a multivariate $\theta$ should be monitored.
@@
 
-**[MERGE with section above]** Convergence has historically been inspected visually, by monitoring the imputation algorithm over iterations.
+**[MERGE with sections above]** Convergence has historically been inspected visually, by monitoring the imputation algorithm over iterations.
 This is typically done through traceplots.
@@
 version of $\widehat{R}$ may be suitable for iterative imputation.
-We therefore use their definition of the diagnostic.** Let $m$ be the total number of chains, $T$ the number of iterations per chain ($T\geq2$), and $\theta$ the scalar summary of interest (e.g., chain means).
+We therefore use their definition of the diagnostic.** Let $m$ be the total number of chains, $T$ the number of iterations per chain (where $T\geq2$), and $\theta$ the scalar summary of interest.
 For each chain ($\ell = 1, 2, \dots, m$), we estimate the variance of $\theta$, and average these to obtain within-chain variance $W$.
@@
 \begin{align*}
-W&=\frac{1}{m} \sum_{\ell=1}^{m} s_{j}^{2}, \text { where } s_{\ell}^{2}=\frac{1}{T-1} \sum_{t=1}^{T}\left(\theta^{(t \ell)}-\bar{\theta}^{(\cdot \ell)}\right)^{2}. 
+W&=\frac{1}{m} \sum_{\ell=1}^{m} s_{\ell}^{2}, \text { where } s_{\ell}^{2}=\frac{1}{T-1} \sum_{t=1}^{T}\left(\theta^{(t \ell)}-\bar{\theta}^{(\cdot \ell)}\right)^{2}. 
 \end{align*}
@@
 
-\noindent We then estimate between-chain variance $B$ as the variance of the collection of average $\theta$ per chain. **[Note that this is not the usual B in MI]**
+\noindent We then estimate between-chain variance $B$ as the variance of the collection of average $\theta$s per chain. **[Note that this is not the usual B in MI]**
 
@@
 
-\noindent From the between- and within-chain variances we compute a weighted average, $\widehat{\operatorname{var}}^{+}$, which approximates the total variance of $\theta$.
+\noindent From the between- and within-chain variances we compute a weighted average, $\widehat{\operatorname{var}}^{+}$, which **approximates** the total variance of $\theta$.
 $\widehat{R}$ is then obtained as a ratio between the total variance and the within-chain variance:
@@
 \widehat{R}=\sqrt{\frac{\widehat{\operatorname{var}}^{+}(\theta | y)}{W}},
-\text{ where } \widehat{\operatorname{var}}^{+}(\theta | y)=\frac{N-1}{N} W+\frac{1}{N} B.
+\text{ where } \widehat{\operatorname{var}}^{+}(\theta | y)=\frac{T-1}{T} W+\frac{1}{T} B.
 \end{equation*}
@@
 We can interpret $\widehat{R}$ as potential scale reduction factor since it indicates by how much the variance of $\theta$ could be shrunken down if an infinite number of iterations per chain would be run [@gelm92].
-This interpretation assumes that chains are 'over-dispersed' at $t=1$, and reach convergence as $T \to \infty$.
+The assumption underlying this interpretation is that chains are 'over-dispersed' at $t=1$, and reach convergence as $T \to \infty$.
 Over-dispersion implies that the initial values of the chains are 'far away' from the target distribution and each other.
 When the sampled values in each chain are independent of the chain's initial value, the mixing component of convergence is satisfied.
-The variance between chains is then equivalent to the variance within chains, and $\widehat{R}$-values will be close to one.
+The variance between chains $B$ is then equivalent to the variance within chains $W$, and $\widehat{R}$-values will be close to one.
 High $\widehat{R}$-values thus indicate non-convergence.
@@
 
-Following the same notation, we define autocorrelation as the correlation between two subsequent $\theta$-values within the same chain [@lync07, p.
+Autocorrelation is defined as the correlation between two subsequent $\theta$-values within the same chain [@lync07, p.
 147].
 In this study, we only consider $AC$ at lag 1, i.e., the correlation between the $t^{th}$ and $t+1^{th}$ iteration of the same chain.
+Following the same notation as for $\widehat{R}$,
 
@@
 
-We can interpret $AC$-values as a measure of stationarity.
-$AC$-values close to zero indicate  independence between subsequent $\theta$s in imputation chains.
-Negative $AC$-values show no threat to the stationarity component of convergence.
+We can interpret $AC$-values as a measure of non-stationarity.
+If there is dependence between subsequent $\theta$-values in imputation chains, $AC$-values are non-zero.
+Positive $AC$-values occur when $\theta$-values are recurring (i.e., high $\theta$-values are followed by high $\theta$-values, and low $\theta$-values are followed by low $\theta$-values).
+Recurrence within imputation chains may lead to trending.
+Negative $AC$-values occur when $\theta$-values of subsequent iterations are less similar, or diverge from one another.
+Divergence within imputation poses no threat to the convergence of the algorithm---it may even speed up convergence.
+**Complete stationarity is reached when $AC=0$.** As non-convergence diagnostic, our interest is in positive $AC$-values.
+
+<!-- close to zero indicate  independence between subsequent $\theta$s in imputation chains.  -->
+
+<!-- Negative $AC$-values show no threat to the stationarity component of convergence.
 On the contrary even---negative $AC$-values indicate that $\theta$-values of subsequent iterations diverge from one another, which may increase the variance of $\theta$ and speed up convergence.
 As convergence diagnostic, the interest is therefore in positive $AC$-values.
-Positive $AC$-values occur when $\theta$-values of subsequent iterations are similar.
-When high $\theta$-values are  followed by high $\theta$-values, and low $\theta$-values are followed by low $\theta$-values, we speak of recurrence.
+-->
+
+<!-- Positive $AC$-values occur when $\theta$-values of subsequent iterations are similar.
+When , we speak of recurrence.
 Recurrence within imputation chains indicates trending.
 Ergo, non-stationarity may be diagnosed when $AC$ values are larger than zero.
+-->
 
@@
 <!-- We assume that complete convergence, defined as $\widehat{R} = 1$ and $AC = 0$, will not be observed.  -->
-It is unlikely that an MCMC algorithm will achieve the ideal values of $\widehat{R}=1$ and $AC=0$.
+It is unlikely that iterative algorithms such as MICE will achieve the ideal values of $\widehat{R}=1$ and $AC=0$.
 Because of its convergence to a distribution, the algorithm will show some signs of non-mixing and non-stationarity even in the most converged state.
-Upon approximate convergence, the imputation chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm ($\widehat{R} > 1$), and there is some dependency between subsequent iterations of imputation chains ($AC > 0$).
-In practice, non-convergence is therefore diagnosed when non-convergence diagnostics exeed a certain threshold.
-The conventional threshold to diagnose non-mixing is $\widehat{R} > 1.2$ [@gelm92].
-Another generally accepted threshold is $\widehat{R} > 1.1$ [@gelm13].
-And more recently, @veht19 proposed an even more stringent threshold of $\widehat{R} > 1.01$.
+The aim is therefore to reach approximate convergence [@gelm13].
+Upon approximate convergence, the imputation chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm ($\widehat{R} > 1$), and there is little dependency between subsequent iterations of imputation chains ($AC > 0$).
+**Approximate convergence is violated when the non-convergence diagnostics exceed a certain threshold.
+[or] In practice, non-convergence is therefore diagnosed when non-convergence diagnostics exeed a certain threshold.** The conventional thresholds to diagnose non-mixing are $\widehat{R} > 1.2$ [@gelm92] or $\widehat{R} > 1.1$ [@gelm13].
+@veht19 proposed a much more stringent threshold of $\widehat{R} > 1.01$.
 The magnitude of $AC$-values may be evaluated statistically, using a Wald test with $AC=0$ as null hypothesis [@box15].
@@
 Before we evaluate the performance of the non-convergence diagnostics $\widehat{R}$ and $AC$ quantitatively through simulation, we apply them to the example of pathological non-convergence by @buur18 that we reproduced above.
-We want to be able to draw the same conclusions as we did from visually inspecting the two algorithms.
-But $\widehat{R}$ and $AC$ should at least be able to distinguish between the two algorithms plotted in Figure \ref{fig:non-conv}.
-From visual inspection, we know that the typically converged algorithm initially portrayed some signs of non-mixing (around $t=2$), but intermingled nicely overall.
-Additionally, there was very little trending.
-The algorithm with pathological non-convergence showed severe non-mixing, although this gradually improved (from $t=7$).
-Moreover, there was a lot of trending as well initially (up-to $t=6$), after which the chains reached a somewhat more stationary state.
-To assess whether $\widehat{R}$ and $AC$ may be appropriate non-convergence identifiers for iterative imputation algorithms, they should identify the same trends as the visual inspection.
-But at least, the following condition must hold: the methods should indicate worse performance for the mis-specified model with pathological non-convergence than the typically converged algorithm (i.e., higher $\widehat{R}$- and $AC$-values).
+We want to be able to draw the same conclusions as we did from visually inspecting the two scenarios in Figure \ref{fig:non-conv}.
+The methods should at least indicate worse performance for the scenario with pathological non-convergence compared to the typically converged algorithm (i.e., higher $\widehat{R}$- and $AC$-values).
 
@@
 
+**[REPHRASE as active about panel A]** From visual inspection, we know that the typically converged algorithm initially portrayed some signs of non-mixing (around $t=2$), but intermingled nicely overall.
+Additionally, there was very little trending.
+The algorithm with pathological non-convergence showed severe non-mixing, although this gradually improved (from $t=7$).
+In this scenario, there was a lot of trending as well initially (up-to $t=6$), after which the chains reached a somewhat more stationary state.
+
 When we look at panel B, we conclude something weird.
@@
 Moreover, the best convergence (as indicated by the lowest $AC$-value) is observed at $t=2$, but looking at the chain means in panel A, there should be some signs of trending up-to iteration number seven.
-After consulting the documentation on `stats::acf()` [@R], we conclude that this $AC$ function is not suited for iterative imputation algorithms.
+After consulting the documentation on `stats::acf()`, we conclude that this $AC$ function is not suited for iterative imputation algorithms.
 The function is optimized for performance when $t\geq50$ [@box15], while the default number of iterations in iterative imputation is often much lower.
@@
 **In conclusion, we will use $\widehat{R}$ conform @veht19 and $AC$ computed manually as non-convergence diagnostics in the simulation study.
-The diagnostics will be applied to four $\theta$s of interest: chain means, chain variances, a quantity of scientific interest, and the first eigenvalue of the variance-covariance matrix.**
+The diagnostics will be applied to four $\theta$s of interest: chain means, chain variances, a quantity of scientific interest, and the first eigenvalue of the variance-covariance matrix.
+Results in eight methods to identify non-convergence**
 
@@
 
-The aim of the simulation study is to evaluate the **impact of inducing non-convergence ["early stopping" en "proportion of complete cases"]** in the MICE algorithm on several quantities of scientific interest $Q$.
+The aim of the simulation study is to determine the **impact of  non-convergence ["early stopping" en "proportion of complete cases"]** in iterative imputation algorithms.
+Specifically, we are interested in the bias and confidence-validity of estimates obtained from incomplete data with the MICE algortihm.
+
+**[REMOVE?] Several quantities of scientific interest $Q$.
 Subsequently we will evaluate how well $\widehat{R}$ and $AC$ perform in identifying the effects of non-convergence **[and performance of different thetas]**.
 And finally, we will formulate an informed advice on the requirements to safely conclude sufficient convergence in practice.
-That is, we conclude that  convergence is sufficient when each estimate $\bar{Q}$ is an unbiased and confidence-valid estimate of the corresponding estimand $Q$.
-**[and formulate advice on thetas?]**
-
-Non-convergence will be induced by: 1) increasing the missingness proportion $p_{\rm mis}$ in dataset $y$ incrementally, and 2) terminating the imputation algorithm at a varying number of iterations $T$.
-The first set of simulation conditions---the missingness proportions---is chosen to reflect the difficulty of the missingness problem.
+That is, we conclude that  convergence is sufficient when each estimate $\bar{Q}$ is an unbiased and confidence-valid estimate of the corresponding estimand $Q$.**
+
+To induce non-convergence, we consider two sets of simulation conditions: we vary the missingness proportion $p_{\rm mis}$ in the incomplete data, and we vary the number of iterations $T$ in the imputation algorithm.
+The missingness conditions are chosen to reflect the difficulty of the missingness problem.
 The underlying assumption is that low missingness proportions lead to quick algorithmic convergence, since there is a lot of information in the observed data.
@@
 
-**[SEGWAY to this section] $Q$s are the targets that will be estimated in each simulation repetition, after first *amputing* the complete data with varying missingness proportions, then imputing the missingness with a varying number of iterations, then estimating the $Q$s on the $m$ completed datasets, and finally pooling the $m$ estimates $\hat{Q}_\ell$ to obtain a single $\bar{Q}$ for each $Q$.
+**[REMOVE?] $Q$s are the targets that will be estimated in each simulation repetition, after first *amputing* the complete data with varying missingness proportions, then imputing the missingness with a varying number of iterations, then estimating the $Q$s on the $m$ completed datasets, and finally pooling the $m$ estimates $\hat{Q}_\ell$ to obtain a single $\bar{Q}$ for each $Q$.
 We try to predict the effect of non-convergence (i.e., the difference between $Q$ and $\bar{Q}$) with the convergence diagnostics $\widehat{R}$ and $AC$, by applying these methods to several scalar summaries of the state space of the algorithm, $\theta$s.**
@@
 437].
-In e.g.
-`mice`, initial values are chosen randomly from the observed data, hence we cannot be certain of over-dispersion in the initial values.
+In  `mice`, initial values are chosen randomly from the observed data, hence we cannot be certain of over-dispersion in the initial values.
 In practice, we do not expect this to cause problems for identifying non-convergence with $\widehat{R}$.
+**[ADD limitation that other MI packages may have different way to obtain initial values!]**
 
@@
 We expect that high $AC$ values are implausible in iterative imputation algorithms with typical convergence.
-That is, after only a few iterations, the randomness induced by the algorithm should effectively mitigate the risk of dependency within chains.
+After only a few iterations, the randomness induced by the algorithm should effectively mitigate the risk of dependency within chains.
 
@@
 We investigate non-convergence of the MICE algorithm through model-based simulation in `R` [version 3.6.3; @R].
-The number of simulation repetitions is 1000.
-The simulation set-up is summarized in the pseudo-code below.
+The number of simulation repetitions is 1000, and the simulation set-up is summarized in the pseudo-code below.
 The complete `R` script of the simulation study is available from [github.com/gerkovink/shinyMice](https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/).
@@
 where $\epsilon \sim \mathcal{N}(0, 100)$.
-This results in a complete set $y$, with observed values for all units $i$ on all variables $j$ ($j =  Y, X_1, X_2, X_3$).
+This results in a complete set $y$, with observed values for all units $i$ on all variables $j$ (where $j =  Y, X_1, X_2, X_3$).
 From the complete set we obtain the true values of the scientific estimands $Q$.
@@
 We consider four types of $Q$s that are often of interest in empirical research.
-Namely, two descriptive statistics: the mean $\mu_j$ and standard deviation $\sigma_j$ of each variable $j = Y, X_1, X_2, X_3$.
+Namely, two descriptive statistics: the mean $\mu_j$ and standard deviation $\sigma_j$ of each variable $j$.
 We also consider the regression coefficients of the predictors: $\beta_1$, $\beta_2$, and $\beta_3$ in regression equation
@@
 And finally, the proportion of variance explained by the set of predictors: coefficient of determination $r^2$ **(note that lower case $r$ is used to avoid confusion with non-convergence diagnostic $\widehat{R}$)**.
-The true values of these scientific estimands are calculated on the complete data.
+These $Q$s are estimated by $\bar{Q}$ in each simulation condition.
+
 <!-- Each of these $Q$s is estimated with a corresponding $\bar{Q}$, after the complete dataset is first *amputed* (with a varying missingness proportion $p_{\rm mis}$) and subsequently imputed (with a varying number of iterations $T$).
@@
 
-### Methods.
+### Simulation conditions.
 
@@
 Let
-$$\text{CI} = \bar{Q} \pm t_{(M-1)} \times SE_{\bar{Q}},$$
-where $t_{(M-1)}$ is the quantile of a $t$-distribution with $m-1$ degrees of freedom, and $SE_{\bar{Q}}$ is the square root of the pooled variance estimate.
-If we obtain nominal coverage (CR = 95%), we can conclude that the $\bar{Q}$s are confidence-valid estimates of the $Q$s.
+$$\text{CI} = \bar{Q} \pm t_{(m-1)} \times \rm SE_{\bar{Q}},$$
+where $t_{(M-1)}$ is the quantile of a $t$-distribution with $m-1$ degrees of freedom, and $\rm SE_{\bar{Q}}$ is the square root of the pooled variance estimate.
+If we obtain nominal coverage (CR = 95%), we can conclude that $\bar{Q}$ is a confidence-valid estimate of $Q$.
 Finally, we inspect CI width (CIW): the difference between the lower and upper bound of the 95\% confidence interval around $\bar{Q}$.
 CIW is of interest because it is a measure of efficiency.
-Under nominal coverage, short CIs are preferred.
-Too short CIs, however, indicate under-estimation of the variance in $\bar{Q}$, **which may yield inferences with low statistical power if CIW is large**.
-
-With these performance measures, we also evaluate how well the non-convergence diagnostics $\widehat{R}$ and $AC$ can identify simulation conditions in which the inferences are affected by non-convergence, i.e., where $\bar{Q}$ is *not* an unbiased, confidence-valid estimate of $Q$.
-**[REPHRASE this:] For performance of univariate $Q$s, we apply the non-convergence diagnostics to univariate $\theta$s.
-When $Q$ is multivariate, we look at signs of non-convergence identified in multivariate $\theta$s.**
-
-- For $Q=\mu$, $\theta$s are chain means **[ADD that these should be unbiased because MCAR!]**.
-
-- For $Q=\sigma$, $\theta$s are chain variances.
-
-- For $Q=\beta$, $\theta$s are estimated $\beta$s per imputation.
-
-- For $Q=r^2$, $\theta$s are the first eigenvalue of the variance-covariance matrix, $\lambda_{1,\ell}$.
+Under nominal coverage, short CIs are preferred, since wider CIs indicate lower statistical power.
+
+We evaluate the methods to identify non-convergence against these performance measures.
+$\widehat{R}$ and $AC$ should identify non-convergence in simulation conditions, where $\bar{Q}$ is *not* an unbiased, confidence-valid estimate of $Q$.
 
@@
 
-Our results show the impact of **inducing non-convergence ["early stopping" en "proportion of complete cases" ]** in an iterative imputation algorithm on several scientific estimates, and whether the non-convergence diagnostics $\widehat{R}$ and $AC$ identify the signs of non-convergence in the algorithm.
-For reasons of brevity, we only discuss the non-convergence diagnostics for estimates with the worst performance in terms of bias.
-Bias in the descriptive statistics ($Q=\mu$ and $Q=\sigma$) is most pronounced for the outcome variable $Y$.
-Of the regression coefficients ($Q=\beta$) we observe the largest bias for $\beta_1$ (the effect of $X_1$ on $Y$).
-There is only one estimate for $Q=r^2$ to consider.
-In the figures, we present simulation conditions where the number of iterations is $1 \leq T\leq50$, since the results are more or less stable for conditions where $T \geq 30$.
+Our results show the performance of the MICE algorithm under different conditions of missingness and iteration length.
+For reasons of brevity, we only discuss results for the worst-performing estimates in terms of bias.
+For univariate scientific estimands ($Q=\mu$ and $Q=\sigma$) we observe the largest bias in the outcome variable $Y$.
+And for the regression coefficients ($Q=\beta$) the bias is most pronounced in $\beta_1$ (the effect of $X_1$ on $Y$).
+Since there is just one estimate for $Q=r^2$ to consider, we **end up with** $Q=\mu_Y, \sigma_Y,r^2, \beta_1$.
+In the figures, we present all missingness conditions, but only **iteration length conditions** $1 \leq T\leq50$, since the results are more or less stable for conditions where $T \geq 30$.
 Full results are available from [github.com/gerkovink/shinyMice](https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy).
@@
 The estimated univariate means seem unaffected by the number of iterations in the algorithm.
-This implies that valid inferences may be obtained with as little as one iteration ($T \geq 1$).
-The magnitude of the bias in $\bar{Q}$ depends solely on the missingness proportion $p_{\rm mis}$, with missingness proportions of 75% and 95% leading to more extreme biases than other missingness proportions **[note that these are curious]**.
-Unbiased estimates are obtained in all conditions where $p_{\rm mis}\leq.50$.
-Note however, that even in the conditions where $p_{\rm mis}\geq.75$, the bias is small: $\mu_Y$ is maximally under-estimated by 0.02 units, while the true value of $\mu_Y$ is 25.81 ($\sigma_Y = 11.32$).
-We conclude that non-convergence does not substantially impact the validity of the inferences for $Q=\mu$.
-**[To be expected, because MCAR]**
+This implies that approximately unbiased estimates may be obtained with as little as one iteration ($T \geq 1$).
+The magnitude of the bias in $\bar{Q}$ depends solely on the missingness proportion $p_{\rm mis}$, with missingness proportions of 75% and 95% leading to more extreme biases than other missingness proportions.
+Completely unbiased estimates are only obtained in conditions where $p_{\rm mis}\leq.50$.
+**This is curious because the missingness mechanism MCAR should yield unbiased univariate estimates without employing multiple imputation.** Note however, that the magnitude of the bias in these conditions is small: $\mu_Y$ is maximally under-estimated by 0.02 units, while the true value of $\mu_Y$ is 25.81 ($\sigma_Y = 11.32$).
+We, therefore, conclude that non-convergence does not substantially impact the validity of the inferences for $Q=\mu$.
 
 ### $Q=\sigma$.
-The estimated standard deviations are largely affected by the missingness proportions, but not to much by the number of iterations.
-Valid inferences are obtained in conditions with just two iterations.
-Interestingly, the bias in first iteration is actually *less* extreme than for any other number of iterations.
-**[Due to MCAR and randomly drawn initial values?]** We may therefore conclude that no iteration at all is needed to obtain approximately unbiased estimates for $Q=\sigma$.
+The estimated standard deviations not affected substantially by **early stopping** either.
+While stable estimates are obtained in conditions with two iterations, the bias in the first iteration is actually *less* severe than in other **iteration conditions**.
+Therefore, no iteration at all is needed to obtain approximately unbiased estimates for $Q=\sigma$.
 Completely unbiased estimates are only obtained in conditions where $p_{\rm mis}=.05$.
@@
 ### $Q=r^2$.
-Bias in the estimated variance explained depends both on missingness proportions and the number of iterations.
-The magnitude of the bias clearly follows the incremental order of the missingness proportions.
-Completely unbiased estimates may only be obtained if the missingness proportion is at most 25%.
+The estimated coefficient of determination is clearly affected by early stopping.
+As expected, we observe worse performance in conditions with a lower number of iterations.
+The magnitude of the bias, however, depends on the missingness proportion---following the incremental order in $p_{\rm mis}$.
+**And so does** the number of iterations necessary to reach stable, approximately unbiased estimates.
+In conditions where $p_{\rm mis}=.05$, estimates are unbiased after one iteration.
+We observe approximately unbiased estimates when $T\geq 2,4, \text{and }5$, in conditions where $p_{\rm mis} = .25, .50, \text{and }.75$, respectively.
+The highest number of iterations necessary to reach approximate unbiasedness is seven.
+Completely unbiased estimates are only obtained in conditions where $p_{\rm mis}\leq.25$.
+This implies that even for biased estimates, $T\geq7$ would suffice to reach a stable solution.
+
+<!-- portion is at most 25%.
 Approximately unbiased estimates are observed for any missingness proportion.
@@
 **[interpret sign of bias?
-ADD what this implies: dat - zelfs bij een gebiasede methode - je maar 7 iteraties nodig hebt om het eindresultaat te bereiken.]**
+ADD what this implies: dat - zelfs bij een gebiasede methode - je maar 7 iteraties nodig hebt om het eindresultaat te bereiken.]** -->
 
@@
 ### $Q=\beta$.
-For the estimated regression coefficients, we consider bias, confidence interval width (CIW), and coverage rate (CR) as performance measures.
+For the estimated regression coefficients, we first consider bias, before discussing CR and CIW.
+As Figure \ref{fig:Qs} shows, the bias in $\bar{Q}$ is affected by both the missingness proportion and the number of iterations.
+Approximately unbiased estimates are observed after one to seven iterations, depending on $p_{\rm mis}$.Completely unbiased estimates are only obtained in conditions where $T\geq3$ and $p_{\rm mis}\leq .50$.
+
+Despite persistant bias in conditions where $p_{\rm mis}\geq.75$, coverage rates indicate that all **missingness conditions** have nominal coverages after just three iterations.
+Confidence-valid estimates may be obtained in any condition where $T\geq3$.
+<!-- , whereas we need $T\geq7$ to reach approximate unbiasedness. -->
+
+The nominal coverages are explained by the CIWs.
+Conform the theoretical foundation of MI, CIs are wider in conditions with higher missingness proportions (i.e., there is less information in the data, and thus more uncertainty due to missingness).
+The wider CIs around $\bar{Q}$ in conditions with higher missingness proportions **results in the inclusion of** the true value of $Q$ despite of bias.
+
+We conclude that approximately unbiased estimates may be obtained in conditions where $T\geq7$, whereas confidence-valid estimates require at most three iterations.
+
+<!-- confidence interval width (CIW), and coverage rate (CR) as performance measures.
 **[Split panels]** Unbiased estimates are obtained in conditions where $p_{\rm mis}\leq.50$ and $T\geq3$.
@@
 Continued iteration after $T\geq7$ seems futile.
-**[add max T for other betas?]**
+**[add max T for other betas?]** -->
 <!-- One exception is the CIW of the most extreme condition ($p_{\rm mis}=.95$; $T=1$). -->
@@
 
-Overall, we see that conditions with a higher proportion of missingness are affected more severely by non-convergence.
+These results demonstrate that the estimates of univariate scientific estimands $Q$ are not impacted by early stopping of the MICE algorithm, irrespective of the missingness proportion in $y$.
+Unbiased estimates may be obtained after just one iteration.
+Multivariate estimates, by contrast, are affected by both the number of iterations and the missingness proportion.
+Completely unbiased estimates are only observed under low to moderate missingness ($p_{\rm mis}\leq.50$), after at most three iterations ($T\leq1,2,3$, depending on $p_{\rm mis}$).
+Approximately unbiased estimates are observed for any missingness condition, after at most seven iterations.
+This implies that the algorithm produces stable, unimproving estimates for any condition where $T\geq7$.
+
+<!-- Overall, we see that conditions with a higher proportion of missingness are affected more severely by non-convergence.
 Multivariate $Q$s are also substantially influenced by the number of iterations.
@@
 This implies that, under the current specifications, no more than seven iterations are necessary to obtain sufficient performance in terms of bias and confidence-validity.
+-->
 
@@
 
-We evaluate the non-convergence diagnostics $\widehat{R}$ and $AC$ against the performance measures described above (i.e., bias and confidence-validity of the quantities of scientific interest $Q$).
+Figure \ref{fig:rhats-acs} displays results for the two non-convergence diagnostics applied to the four $\theta$s.
+**[describe panels]** We evaluate the performance of these methods by establishing whether they correctly identify conditions in which $\bar{Q}$ is *not* an unbiased, confidence-valid estimate of $Q$.
+Biased estimates are observed in conditions where $T\leq7$.
+Non-nominal coverage rates are observed only when $T\leq3$.
+Additionally, there is persistant bias in conditions where $p_{\rm mis}\geq.75$, irrespective of the iteration length.
+**[appropriate methods thus identify these conditions]**
+
+<!-- We evaluate the non-convergence diagnostics $\widehat{R}$ and $AC$ against the performance measures described above (i.e., bias and confidence-validity of the quantities of scientific interest $Q$).
 The diagnostics should identify those simulation conditions that are negatively impacted by non-convergence.
 We expect the most signs of non-convergence in conditions where valid estimates are not obtained (high missingness proportions, $T<7$), and at least some signs of non-convergence in conditions where estimates are not approximately unbiased ($p_{\rm mis}=.95$, any number of iterations).
+-->
 
@@
 ### $\theta=$ chain mean.
-**[As expected, $\widehat{R}$ lowers with the number of iterations.
-At a cut-point of 1.2 , we have convergence at about 8 iteration, whereas we need 12 and 25 iteration for more conservative cut-off 1.1 and 1.05.
-We do not reach the very strict 1.01.]**
-
-We evaluate the performance of the non-convergence diagnostics applied to chain means against the bias in $Q=\mu$.
+As expected, $\widehat{R}$ lowers with the number of iterations.
+After an initial 'dip' in conditions where $T=3$, we observe a gradual decrease in $\widehat{R}$-values with every additional iteration.
+This implies improving convergence, until the decrease tapers off after 30 to 40 iterations.
+Using the threshold $\widehat{R}>1.2$, we diagnose non-convergence in conditions where $T\leq7$, whereas the threshold 1.1 is exceeded when $T\leq13$.
+The very strict threshold 1.01 is surpassed in all conditions $T\leq100$**, and therefore not informative in the current study**.
+With this method and $\theta$, we fail to identify simuation conditions with persistant bias due to high missingness proportions.
+
+Similarly, $AC$ decreases with higher $T$s.
+$AC$ indicates improving stationarity in conditions where $T\leq6$.
+The $AC$-values do not exceed the threshold **defined by statistically significant $AC$s**.
+Based on this threshold, we fail to not diagnose non-convergence in any condition.
+
+<!-- We evaluate the performance of the non-convergence diagnostics applied to chain means against the bias in $Q=\mu$.
 Based on the bias in $Q$, we concluded that one iteration is sufficient to obtain valid inferences, irrespective of the missingness proportion.
@@
 This does not correspond to the impact of non-convergence in the univariate $Q$s.
-**Compared to the multivariate $Q$s [add before that we evaluate the thetas against all Qs?], $\widehat{R}$ and $AC$ applied to the chain means fail to identify the non-convergence in conditions where $p_{\rm mis}=.95$.**
+**Compared to the multivariate $Q$s [add before that we evaluate the thetas against all Qs?], $\widehat{R}$ and $AC$ applied to the chain means fail to identify the non-convergence in conditions where $p_{\rm mis}=.95$.**  -->
 
@@
 ### $\theta=$ chain variance.
-The non-convergence diagnostics applied to the chain variances as $\theta$ yield highly similar results to the chain means.
-**[Segway to performance of $Q=\sigma$] While the bias in the estimated standard deviations mainly depends on the missingness proportion of each condition, $\widehat{R}$ and $AC$ are influenced mostly by the number of iterations.
-We thus conclude that there is no added value of evaluating non-convergence with chain variances as $\theta$ in the diagnostics.**
+As Figure \ref{fig:rhats-acs} shows, the results of the two diagnostics applied to the chain variances are highly similar to those on the chain means.
+This holds for both $\widehat{R}$ and $AC$.
+The only apparent difference is in the $\widehat{R}$ threshold 1.1.
+Instead of diagnosing non-convergence in conditions where $T\leq13$, we now diagnose this when $T\leq11$.
+All other observations are equivalent.
+
+<!-- We thus conclude that there is no added value of evaluating non-convergence with chain variances as $\theta$ in the diagnostics. -->
 
@@
 ### $\theta=\hat{Q}_\ell$.
-**[We zien weer hetzelfde voor laag-medium, beetje meer voor extreme misisnh data.]**
-
-The non-convergence diagnostics are applied to the estimated regression coefficients in each imputation, and evaluated against the bias in the corresponding pooled estimate.
-
-$\widehat{R}$ indicates signs of non-convergence across all missingness proportions and each number of iterations, in the expected order: simulation conditions with higher missingness proportions and less iterations are identified as the worst converged.
+The $\widehat{R}$-values show a similar trend across iterations as we observed for chain means and chain variances.
+Only conditions with very high missingness proportions ($p_{\rm mis}\geq.75$) diverge from the earlier observations.
+The $\widehat{R}$-values in these conditions do not taper off after 30 to 40 iterations, but rather between 40 to 50 iterations.
+The highest number of iterations at which non-convergence is still diagnosed according to the thresholds are: 11 for $\widehat{R}>1.2$, 23 for $\widehat{R}>1.1$, and 100 for $\widehat{R}>1.01$.
+
+The $AC$-values also follow earlier observations, with the exception of the same very high missingness proportions that **diverged** for $\widehat{R}$.
+To reach maximum stationarity, these conditions require not five but six or seven iterations.
+Moreover, the $AC$-values in these conditions exceed the threshold to diagnose non-convergence.
+Interestingly, this only occurs after ten or even thrity iterations, while we would expect more signs of trending early in the iterations.
+**Since the $AC$-values are not improving or worsening after seven iterations, we conclude that non-convergence is incorrectly diagnosed based on the threshold.**
+
+<!-- The non-convergence diagnostics are applied to the estimated regression coefficients in each imputation, and evaluated against the bias in the corresponding pooled estimate.  -->
+
+<!-- $\widehat{R}$ indicates signs of non-convergence across all missingness proportions and each number of iterations, in the expected order: simulation conditions with higher missingness proportions and less iterations are identified as the worst converged.
 The thresholds to diagnose non-convergence are overcome in different conditions, depending on the number of iterations and the missingness proportion.
@@
 The conventional $\widehat{R}>1.1$ is surpassed at $T\geq12$ and $T\geq24$, respectively.
-
-For the autocorrelations applied to the $\hat{Q}_\ell$s, we see an even stronger effect of the missingness proportions.
+-->
+
+<!-- For the autocorrelations applied to the $\hat{Q}_\ell$s, we see an even stronger effect of the missingness proportions.
 Especially the conditions where $p_{\rm mis} = 0.75$ and $p_{\rm mis} = 0.95$ are identified as non-stationary.
@@
 The conclusion to identify non-stationarity in conditions where $T\geq10$ and $p_{\rm mis} = 0.95$ might therefore be spurious.
+-->
 
@@
 
-Compared to the bias in $\bar{Q}$, the diagnostics applied to the $\hat{Q}_\ell$s over-estimate the severity of the non-convergence.
-
-### Summary
-
-In general, the diagnostics identify more signs of non-convergence in simulation conditions with a lower number of iterations.
-The missingness proportions, however, only seem to impact multivariate convergence.
-When $\widehat{R}$ and $AC$ are applied to  univariate $\theta$s, they identify the same amount of non-convergence for any missingness proportion.
-When the diagnostics are applied to multivariate $\theta$s, they generally identify more signs of non-convergence for higher missingness proportions.
-
-As expected, none of the simulation conditions that we consider resulted in complete convergence of the algorithm (defined as $\widehat{R}=1$ and $AC=0$).
-However, the thresholds used in practice to diagnose non-convergence seem too conservative for imputation purposes.
-$\widehat{R}$ and $AC$ identify signs of non-convergence in conditions where the $\bar{Q}$s are unbiased and confidence-valid estimates of the $Q$s.
-The most recent recommended threshold for diagnosing non-convergence with $\widehat{R}$ is never overcome in this study (all $\widehat{R}$>1.01).
-Apparently, this threshold is too stringent for iterative imputation algorithms, and will not be discussed any further.
-
-<!-- This is odd, since the $\theta$ and $Q$ are directly related.   -->
-
-```{r est, fig.height = 6.8, fig.cap="Convergence diagnostics regression coefficient."}
-# est_bias + est_Rh + est_AC + plot_layout(guides = "collect", ncol = 1) + plot_annotation(tag_levels = "A", tag_suffix = ".", title = bquote("Q = " ~ beta[1] ~ "= 2.06; " ~ theta ~ "= " ~ hat(Q) ~ "in {" ~ y[obs] ~ "," ~ y[imp][",m"] ~ "}"))
-```
+<!-- Compared to the bias in $\bar{Q}$, the diagnostics applied to the $\hat{Q}_\ell$s over-estimate the severity of the non-convergence.  -->
 
 ### $\theta=\lambda_{1, \ell}$.
-Finally, we evaluate the non-convergence diagnostics applied to the first eigenvalue of the variance-covariance matrices of the completed data.
+Once again, we observe $\widehat{R}$-values with a similar trend across iterations.
+For this $\theta$, the results differ only in the most extreme missingness condition.
+We diagnose non-convergence in conditions where $T\leq9$ according to $\widehat{R}>1.2$, and $T\leq19$ for $\widehat{R}>1.1$.
+The 'dip' in $\widehat{R}$-values at $T=3$ is even more pronounced than with other $\theta$s.
+Some $\widehat{R}$-values even overcome the threshold of 1.2.
+If we would terminate the algorithm after three iterations, we would incorrectly conclude sufficient convergence according to the $\widehat{R}>1.2$ threshold.
+
+$AC$-values are systematically higher for this $\theta$ than in the other results.
+The $AC$-values also taper off at a later point in the iterations, indicating improving stationarity up-to ten, or even thirty iterations, depending on the missingness proportion.
+Using the threashold, we diagnose non-convergence at some point for any missingness proportion.
+
+<!-- And especially in conditions with higher missingness proportions, we observe signs of increasing stationarity the decrease in the $AC$-values across iterations is also  -->
+
+<!-- Finally, we evaluate the non-convergence diagnostics applied to the first eigenvalue of the variance-covariance matrices of the completed data.
 Similar to the results the the regression coefficients as $\theta$s, we see effects of both the missingness proportion and number of iterations in each simulation condition.
 The most signs of non-convergence are identified in conditions where $p_{\rm mis}$ is large and $T$ is small.
-
-Notwithstanding the similarities with $\theta=\hat{Q}_\ell$, there is a big difference in $\widehat{R}$-values for conditions where $T<4$.
+-->
+
+<!-- Notwithstanding the similarities with $\theta=\hat{Q}_\ell$, there is a big difference in $\widehat{R}$-values for conditions where $T<4$.
 In these conditions, $\widehat{R}$ behaves oppositely of our expectation: conditions with the lowest missingness proportions yield the highest $\widehat{R}$-values.
@@
 With the $\widehat{R} > 1.1$ threshold, the number of iterations are $T\geq20$ and $T\geq16$, respectively.
-
-The $AC$-values largely depend on the missingness proportion in each condition.
+-->
+
+<!-- The $AC$-values largely depend on the missingness proportion in each condition.
 Conditions where $p_{\rm mis}=.95$ show statistically significant autocorrelations for any number of iterations.
@@
 A similar point is only observed after twenty iterations in conditions with higher missingness proportions.
-
-In comparison to the $Q$s, the diagnostics again over-estimate the signs of non-convergence.
+-->
+
+<!-- In comparison to the $Q$s, the diagnostics again over-estimate the signs of non-convergence.
 Conditions with low missingness proportions (e.g., 5% or 25% of the complete data) yield unbiased, confidence-valid estimates with as little as two iterations.
@@
 The univariate $\theta$s do not differentiate between bias induced by the missingness proportions.
+-->
+
+### Summary
+
+Overall, the methods to diagnose non-convergence perform as expected: they indicate more signs of non-convergence in conditions with worse performance in terms of bias and confidence-validity.
+Under low to moderate missingness, it does not seems to matter on which $\theta$ the diagnostics are applied.
+If we look at complete unbiasedness, however, we notice that univariate $\theta$s fail to diagnose the persistant bias in conditions where $p_{\rm mis}\geq.75$.
+The number of iterations necessary to obtain approximately unbiased, confidence-valid estimates corresponds to the $\widehat{R}$ threshold 1.2.
+The thresholds 1.1 and 1.01 seem to strict compared to the validity of the inferences.
+The $AC$ threshold defined by the statistical significance of the $AC$-values does not appear to be appropriate in the current set-up **[or iterative imputation in general??]**.
+Non-stationarity may be better diagnosed by evaluating the $AC$-values across iterations.
+If the values do not substantially decrease, stationarity may be concluded **[similar to the elbow heuristic in PCA]**.
+
+<!-- In general, the diagnostics identify more signs of non-convergence in simulation conditions with a lower number of iterations.
+The missingness proportions, however, only seem to impact multivariate convergence.
+When $\widehat{R}$ and $AC$ are applied to  univariate $\theta$s, they identify the same amount of non-convergence for any missingness proportion.
+When the diagnostics are applied to multivariate $\theta$s, they generally identify more signs of non-convergence for higher missingness proportions.
+-->
+
+<!-- As expected, none of the simulation conditions that we consider resulted in complete convergence of the algorithm (defined as $\widehat{R}=1$ and $AC=0$).
+However, the thresholds used in practice to diagnose non-convergence seem too conservative for imputation purposes.
+$\widehat{R}$ and $AC$ identify signs of non-convergence in conditions where the $\bar{Q}$s are unbiased and confidence-valid estimates of the $Q$s.
+The most recent recommended threshold for diagnosing non-convergence with $\widehat{R}$ is never overcome in this study (all $\widehat{R}$>1.01).
+Apparently, this threshold is too stringent for iterative imputation algorithms, and will not be discussed any further.
+-->
+
+<!-- This is odd, since the $\theta$ and $Q$ are directly related.   -->
+
+```{r est, fig.height = 6.8, fig.cap="Convergence diagnostics regression coefficient."}
+# est_bias + est_Rh + est_AC + plot_layout(guides = "collect", ncol = 1) + plot_annotation(tag_levels = "A", tag_suffix = ".", title = bquote("Q = " ~ beta[1] ~ "= 2.06; " ~ theta ~ "= " ~ hat(Q) ~ "in {" ~ y[obs] ~ "," ~ y[imp][",m"] ~ "}"))
+```
 
@@
 # Discussion
+
+H1: As hypothesized, biased, invalid estimates occur more often in simulation conditions with a higher missingness proportion and lower number of iterations.
+However, this only holds for multivariate scientific estimands.
+Univariate $Q$s are not substantially impacted by early stopping.
+
+H2: As expected, we observe higher $\widehat{R}$ and $AC$-values in conditions where $\bar{Q}$s were biased, invalid estimates of $Q$s.
+
+H3: The hypothesis that the recommended thresholds would be too strict is only patially supported.
+$\widehat{R}$ indicated improving convergence until 30 to 40 iterations, while approximately unbiased, confidence-valid estimates were obtained after just seven iterations.
+This rougly corresponds to the threshold of 1.2.
+The more stringent trheshold 1.1 seems plausible compared to the decrease in $\widehat{R}$ values across iterations.
+The most recent threshold 1.01 does indeed seem to over-estimate the signs of non-convergence.
+
+H4: The results of this simulation study do not support the hypothesis that ...
+**[insert hypothesis]**.
+We observed $AC$-values as high as theoretically possible $AC=$.
+The second part of the hypothesis is not refuted: the $AC$-values did decrease quicky as a function of $T$.
+
+H5: We did find some support for the final hypothesis. Multivariate $\theta$s indeed have superior performance under certain conditions (e.g., multivariate $Q$s and extreme missingness proportions).
+
+# Conlusion
 

### 2020-05-09 3.Thesis/Writeup/2.Manuscript.tex
Missing only some parts of the methods and discussion

@@
 
-\hypertarget{overview-of-sections}{%
-\section{Overview of sections}\label{overview-of-sections}}
-
-Intro
-
-\begin{itemize}
-\item
-  This is missingness
-\item
-  This is how we solve missingness
-\item
-  For that we need an algorithm, but what if it has not converged?
-\item
-  That is what we solve in this paper
-\end{itemize}
-
-Algorithmic convergence
-
-\begin{itemize}
-\item
-  What is algorithmic convergence?
-\item
-  When is it converged? Stationarity versus mixing.
-\item
-  How is convergence currently investigated? And how should it be investigated? Multivariate versus univariate.
-\end{itemize}
-
-Methods
-
-\begin{itemize}
-\item
-  Which methods are there? And how do they relate to mixing and stationarity?
-\item
-  \(\widehat{R}\)
-\item
-  AC
-\item
-  We can use these on any \(\theta\)
-\end{itemize}
-
-The Simulation study, Results, Discussion.
-
 \hypertarget{introduction}{%
@@
 Missing data is not only ubiquitous across science, but treating it can also be a tedious task.
-If a dataset contains just one incomplete observation, calculations are not defined, \textbf{hence} statistical models cannot do fitted to the data.
+If a dataset contains just one incomplete observation, calculations are not defined, and statistical models cannot be fitted to the data.
 To circumvent this, many statistical packages employ list-wise deletion by default (i.e., ignoring incomplete observations).
@@
 Since \(Q\) is estimated on each completed dataset, \(m\) separate \(\hat{Q}_\ell\)-values are obtained.
-Finally, the \(m\) \(\hat{Q}_\ell\)-values are combined into a single pooled estimate \(\bar{Q}\).
-The premise of multiple imputation is that \(\bar{Q}\) is an unbiased and confidence-valid estimate of the true---but unobserved---inference \citep{rubin96}.
+Finally, the \(\hat{Q}_\ell\)-values are combined into a single pooled estimate \(\bar{Q}\).
+The premise of multiple imputation is that \(\bar{Q}\) is an unbiased and confidence-valid estimate of the true---but unobserved---scientific estimand \(Q\) \citep{rubin96}.
 
@@
 
-\caption{Scheme of the main steps in multiple imputation (where $m=3$)---from an incomplete dataset, to $m$ multiply imputed datasets, to $m$ estimated quantities of scientific interest $\hat{Q}$s, to a single pooled estimate $\bar{Q}$.}\label{fig:diagram}
+\caption{Scheme of the main steps in multiple imputation---from an incomplete dataset, to $m=3$ multiply imputed datasets, to $m=3$ estimated quantities of scientific interest $\hat{Q}$s, to a single pooled estimate $\bar{Q}$.}\label{fig:diagram}
 \end{figure}
 
-A popular method to obtain imputations is to use the `Multiple Imputation by Chained Equations' algorithm, shorthand `MICE'\citep{mice}.
-With MICE, imputed values \(\dot y_{\rm imp, \ell}\) are drawn from the posterior predictive distribution of the missing values \(y_{\rm mis}\).
+A popular method to obtain imputations \(\dot y_{\rm imp, \ell}\) is to use the `Multiple Imputation by Chained Equations' algorithm, shorthand `MICE'\citep{mice}.
+With MICE, imputed values are drawn from the posterior predictive distribution of the missing values \(y_{\rm mis}\).
 The algorithm is named after the iterative algorithmic procedure by which imputed values are generated: a multivariate distribution is obtained by iterating over a sequence of univariate imputations.
@@
 These remain open questions, since the convergence properties of iterative imputation algorithms have not been systematically studied \citep{buur18}.
-There is no scientific consensus on how to evaluate convergence of imputation algorithms \citep{taka17}.
-\textbf{Moreover, the convergence of chained equations algorithms such as MICE under certain default imputation models (e.g., `predictive mean matching') is an entirely open question \citep{murr18}.} Therefore, algorithmic convergence should be monitored carefully.
-
-Currently, the recommended practice for evaluating the convergence of iterative imputation algorithms is through visual inspection.
-After running the imputation algorithm for a certain number of iterations, its convergence is monitiored by plotting a scalar summary of the state space of the algorithm \(\theta\) against the iteration number.
-Typically, the \(\theta\)s under evaluation are chain means and variances---the average and the variance of the imputed values per imputation.
-Monitoring convergence by inspecting traceplots of these \(\theta\)s may be undesirable for several reasons on two counts: 1) it may be challenging to the untrained eye, and 2) only severely pathological cases of non-convergence may be diagnosed \citep[\(\S\) 6.5.2]{buur18} and, 3) there is not an objective measure that quantifies convergence.
-Therefore, a quantitative, diagnostic evaluation of convergence would be preferred---although not straightforward.
+There is no scientific consensus on how to evaluate the convergence of imputation algorithms \citep{taka17}.
+Moreover, the behavoir of such algorithms under certain default imputation models (e.g., `predictive mean matching') is an entirely open question \citep{murr18}.
+\textbf{Therefore, algorithmic convergence should be monitored carefully, }although not straightforward.
 Iterative imputation algorithms such as MICE are special cases of Markov chain Monte Carlo (MCMC) methods.
@@
 We define several diagnostics and evaluate how these diagnostics could be appropriate for iterative imputation applications.
-We then adress the impact of inducing non-convergence in iterative imputation algorithms by means of model-based simulation in \texttt{R} \citep{R}.
-The aim of the simulation is to determine whether unbiased, confidence-valid inferences may be obtained under different levels of non-convergence (\textbf{and at which point convergence may safely be assumed}).
-Additionally, we will assess how well different non-convergence diagnostics perform in identifying signs of non-convergence (\textbf{and on which scalar summary of the state space of the algorithm these diagnostics should be applied}).
+We then adress the impact of inducing non-convergence in iterative imputation algorithms through model-based simulation in \texttt{R} \citep{R}.
+The aim of the simulation study is to determine whether unbiased, confidence-valid inferences may be obtained if the algorithm has not (yet) converged.
+And additionally, to formulate an informed advice on when approxiamte convergence may be safely concluded.
+We will therefore assess the performance of different methods to identify non-convergence.
 We translate the results of the study into guidelines for applied researchers.
-\textbf{{[}add more societal relevance: what is the consequence of non-convergence?
-see example of non-convergence in Figure 2 (e.g., biased estimates, invalid inference, etc.){]}} For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular \texttt{mice} package \citep{mice} in \texttt{R} \citep{R}.
+\textbf{ADD that these guidelines may aid to draw valid inference from incomplete data?{]}} For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular \texttt{mice} package \citep{mice} in \texttt{R} \citep{R}.
 
@@
 The data value of unit \(i\) (\(i = 1, 2, \dots, n\)) on variable \(j\) (\(j = 1, 2, \dots, k\)) may be either observed or missing.
-The number of units \(i\) in dataset \(y\) with at least one missing value, divided by the total number of units \(n\), is called the missingness proportion \(p_{\rm mis}\).
+The number of units \(i\) with at least one missing value, divided by the total number of units \(n\), is called the missingness proportion \(p_{\rm mis}\) in dataset \(y\).
 The collection of observed data values in \(y\) is denoted by \(y_{\rm obs}\); the missing part of \(y\) is referred to as \(y_{\rm mis}\).
 For each datapoint in \(y_{\rm mis}\), we sample \(m \times T\) plausible values, where \(m\) is the number of imputations (\(\ell = 1, 2, \dots, m\)) and \(T\) is the number of iterations (\(t = 1, 2, \dots, T\)) in the imputation algorithm.
-The collection of samples between the initial value (at \(t=1\)) and the final imputed value (at \(t=T\)) will be referred to as an `imputation chain'.
+The state space of the algorithm at a certain iteration \(t\) may be summarized by a scalar summary \(\theta\) (e.g., \textbf{the average of the imputed values}).
+The collection of \(\theta\)-values between the \textbf{initial value} (at \(t=1\)) and the final imputed value (at \(t=T\)) will be referred to as an `imputation chain'.
 
@@
 There are two requirements for convergence of iterative algorithms: mixing and stationarity \citep{gelm13}.
-Without mixing, imputation chains do not intermingle nicely, indicating that the distribution of imputed values differs across imputations.
-Chains may be `stuck' at a local optimum, instead of sampling imputed values from the entire predictive posterior distribution of the missing values.
-This may cause under-estimation of the variance within chains, which results in spurious, invalid inferences.
+In iterative imputation algorithms mixing implies that imputation chains intermingle nicely, and stationarity is characterized by \textbf{the absence of trending} within chains.
+If one of these two requirements is not met, there is non-convergence in the algorithm.
+Without mixing, chains may be `stuck' at a local optimum, instead of sampling imputed values from the entire predictive posterior distribution of the missing values.
+The distribution of imputed values then differs across imputations.
+This may cause under-estimation of the variance between chains, which results in spurious, invalid inferences.
 Without stationarity, there is trending within imputation chains.
 Trending implies that further iterations would yield a systematically lower or higher set of imputations.
-Iterative imputation algorithms that have not yet reached stationarity, may thus yield biased estimates.
+Iterative imputation algorithms that have not (yet) reached stationarity, may thus yield biased estimates.
 
-\textbf{{[}SEGWAY to this section{]}} To illustrate what non-convergence looks like in iterative imputation algorithms, we reproduce an example from van Buuren \citeyearpar[\(\S\) 6.5.2]{buur18}.
-Figure \ref{fig:non-conv} displays a subset of the example: a variable that we will call \(j\).
-We see two traceplots of the chain means for \(j\)---i.e., the average of the imputed values for \(j\) in each imputation \(\dot y_{\rm imp, \ell}\), plotted against the iteration number.
-Each line portrays one imputation.
-The plot on the left-hand side of the figure shows typical convergence of an iterative imputation algorithm.
+To illustrate what non-mixing and non-stationarity looks like in iterative imputation algorithms, we reproduce an example from van Buuren \citeyearpar[\(\S\) 6.5.2]{buur18}.
+Figure \ref{fig:non-conv} displays two scenarios from the example.
+The graph on the left-hand side of the figure shows typical convergence of an iterative imputation algorithm.
 The right-hand side displays pathological non-convergence, induced by purposefully mis-specifying the imputation model.
-\textbf{In the typical convergence situation, the imputation chains intermingle nicely and there is little to no trending.
-In the non-convergence plot, there is a lot of trending and some chains do not intermingle.} Importantly, the chain means at the last iteration (the imputed value per imputation \(\ell\)) are very different between the two plots.
+Each line portrays one imputation chain: scalar summary \(\theta\) plotted against the iteration number \(t\).
+The \(\theta\) in this example is the chain mean of a variable that we will call \(j\) (i.e., the average of the imputed values for \(j\) in each imputation \(\dot y_{\rm imp, \ell}\)).
+\textbf{In the typical convergence scenario, the imputation chains intermingle nicely and there is little to no trending.
+In the non-convergence scenario, there is a lot of trending and some chains do not intermingle.} Importantly, the chain means at the last iteration (the imputed value per imputation \(\ell\)) are very different between the two plots.
 The algorithm with the mis-specified model yields imputed values that are on average a factor two larger than those of the typically converged algorithm.
-That means that non-convergence has a impact on the mean of \(j\) in \(\dot y_{\rm imp, \ell}\).
-This difference (presumably) translates to bias in the \(m\) completed data estimates \(\hat{\mu}_{j,\ell}\), and consequently also to the pooled estimate \(\bar{\mu}_j\).
-Non-convergence thus leads to biased estimates.
+Non-convergence thus influences the mean of \(j\) in \(\dot y_{\rm imp, \ell}\).
+This difference (presumably) translates to the \(m\) completed data estimates \(\hat{\mu}_{j,\ell}\), and consequently to the pooled estimate \(\bar{\mu}_j\).
+Ergo, non-convergence leads to biased estimates.
 This shows the importance of reaching converged states in iterative imputation algorithms.
@@
 
-\textbf{{[}MERGE with section below{]}}.
-Inspecting non-convergence through traceplots of chain means and chain variances may be insufficient because they are univariate summaries of the state space of the algorithm, while MICE is not only concerned with a single column, but the entire multivariate distribution of \(y_{\rm imp}\).
+\textbf{{[}MERGE with sections below{]}} Currently, the recommended practice for evaluating the convergence of the MICE algorithm is through visual inspection.
+After running the imputation algorithm for a certain number of iterations, its convergence is monitiored by plotting a scalar summary of the state space of the algorithm \(\theta\) against the iteration number.
+Typically, the \(\theta\)s under evaluation are chain means and variances---the average and the variance of the imputed values per imputation.
+Monitoring convergence by inspecting traceplots of these \(\theta\)s may be undesirable for several reasons: 1) it may be challenging to the untrained eye, and 2) only severely pathological cases of non-convergence may be diagnosed \citep[\(\S\) 6.5.2]{buur18} and, 3) there is not an objective measure that quantifies convergence.
+
+\textbf{{[}MERGE with section above and below{]}} Inspecting non-convergence through traceplots of chain means and chain variances may be insufficient because they are univariate summaries of the state space of the algorithm, while MICE is not only concerned with a single column, but the entire multivariate distribution of \(y_{\rm imp}\).
 Ideally, a multivariate \(\theta\) should be monitored.
@@
 
-\textbf{{[}MERGE with section above{]}} Convergence has historically been inspected visually, by monitoring the imputation algorithm over iterations.
+\textbf{{[}MERGE with sections above{]}} Convergence has historically been inspected visually, by monitoring the imputation algorithm over iterations.
 This is typically done through traceplots.
@@
 The Vehtari et al.~version of \(\widehat{R}\) may be suitable for iterative imputation.
-We therefore use their definition of the diagnostic.} Let \(m\) be the total number of chains, \(T\) the number of iterations per chain (\(T\geq2\)), and \(\theta\) the scalar summary of interest (e.g., chain means).
+We therefore use their definition of the diagnostic.} Let \(m\) be the total number of chains, \(T\) the number of iterations per chain (where \(T\geq2\)), and \(\theta\) the scalar summary of interest.
 For each chain (\(\ell = 1, 2, \dots, m\)), we estimate the variance of \(\theta\), and average these to obtain within-chain variance \(W\).
@@
 \begin{align*}
-W&=\frac{1}{m} \sum_{\ell=1}^{m} s_{j}^{2}, \text { where } s_{\ell}^{2}=\frac{1}{T-1} \sum_{t=1}^{T}\left(\theta^{(t \ell)}-\bar{\theta}^{(\cdot \ell)}\right)^{2}. 
+W&=\frac{1}{m} \sum_{\ell=1}^{m} s_{\ell}^{2}, \text { where } s_{\ell}^{2}=\frac{1}{T-1} \sum_{t=1}^{T}\left(\theta^{(t \ell)}-\bar{\theta}^{(\cdot \ell)}\right)^{2}. 
 \end{align*}
 
-\noindent We then estimate between-chain variance \(B\) as the variance of the collection of average \(\theta\) per chain. \textbf{{[}Note that this is not the usual B in MI{]}}
+\noindent We then estimate between-chain variance \(B\) as the variance of the collection of average \(\theta\)s per chain. \textbf{{[}Note that this is not the usual B in MI{]}}
 
@@
 
-\noindent From the between- and within-chain variances we compute a weighted average, \(\widehat{\operatorname{var}}^{+}\), which approximates the total variance of \(\theta\).
+\noindent From the between- and within-chain variances we compute a weighted average, \(\widehat{\operatorname{var}}^{+}\), which \textbf{approximates} the total variance of \(\theta\).
 \(\widehat{R}\) is then obtained as a ratio between the total variance and the within-chain variance:
@@
 We can interpret \(\widehat{R}\) as potential scale reduction factor since it indicates by how much the variance of \(\theta\) could be shrunken down if an infinite number of iterations per chain would be run \citep{gelm92}.
-This interpretation assumes that chains are `over-dispersed' at \(t=1\), and reach convergence as \(T \to \infty\).
+The assumption underlying this interpretation is that chains are `over-dispersed' at \(t=1\), and reach convergence as \(T \to \infty\).
 Over-dispersion implies that the initial values of the chains are `far away' from the target distribution and each other.
 When the sampled values in each chain are independent of the chain's initial value, the mixing component of convergence is satisfied.
-The variance between chains is then equivalent to the variance within chains, and \(\widehat{R}\)-values will be close to one.
+The variance between chains \(B\) is then equivalent to the variance within chains \(W\), and \(\widehat{R}\)-values will be close to one.
 High \(\widehat{R}\)-values thus indicate non-convergence.
@@
 
-Following the same notation, we define autocorrelation as the correlation between two subsequent \(\theta\)-values within the same chain \citep[p.~147]{lync07}.
+Autocorrelation is defined as the correlation between two subsequent \(\theta\)-values within the same chain \citep[p.~147]{lync07}.
 In this study, we only consider \(AC\) at lag 1, i.e., the correlation between the \(t^{th}\) and \(t+1^{th}\) iteration of the same chain.
+Following the same notation as for \(\widehat{R}\),
 
@@
 
-We can interpret \(AC\)-values as a measure of stationarity.
-\(AC\)-values close to zero indicate independence between subsequent \(\theta\)s in imputation chains.
-Negative \(AC\)-values show no threat to the stationarity component of convergence.
-On the contrary even---negative \(AC\)-values indicate that \(\theta\)-values of subsequent iterations diverge from one another, which may increase the variance of \(\theta\) and speed up convergence.
-As convergence diagnostic, the interest is therefore in positive \(AC\)-values.
-Positive \(AC\)-values occur when \(\theta\)-values of subsequent iterations are similar.
-When high \(\theta\)-values are followed by high \(\theta\)-values, and low \(\theta\)-values are followed by low \(\theta\)-values, we speak of recurrence.
-Recurrence within imputation chains indicates trending.
-Ergo, non-stationarity may be diagnosed when \(AC\) values are larger than zero.
+We can interpret \(AC\)-values as a measure of non-stationarity.
+If there is dependence between subsequent \(\theta\)-values in imputation chains, \(AC\)-values are non-zero.
+Positive \(AC\)-values occur when \(\theta\)-values are recurring (i.e., high \(\theta\)-values are followed by high \(\theta\)-values, and low \(\theta\)-values are followed by low \(\theta\)-values).
+Recurrence within imputation chains may lead to trending.
+Negative \(AC\)-values occur when \(\theta\)-values of subsequent iterations are less similar, or diverge from one another.
+Divergence within imputation poses no threat to the convergence of the algorithm---it may even speed up convergence.
+\textbf{Complete stationarity is reached when \(AC=0\).} As non-convergence diagnostic, our interest is in positive \(AC\)-values.
 
@@
 
-It is unlikely that an MCMC algorithm will achieve the ideal values of \(\widehat{R}=1\) and \(AC=0\).
+It is unlikely that iterative algorithms such as MICE will achieve the ideal values of \(\widehat{R}=1\) and \(AC=0\).
 Because of its convergence to a distribution, the algorithm will show some signs of non-mixing and non-stationarity even in the most converged state.
-Upon approximate convergence, the imputation chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm (\(\widehat{R} > 1\)), and there is some dependency between subsequent iterations of imputation chains (\(AC > 0\)).
-In practice, non-convergence is therefore diagnosed when non-convergence diagnostics exeed a certain threshold.
-The conventional threshold to diagnose non-mixing is \(\widehat{R} > 1.2\) \citep{gelm92}.
-Another generally accepted threshold is \(\widehat{R} > 1.1\) \citep{gelm13}.
-And more recently, \citet{veht19} proposed an even more stringent threshold of \(\widehat{R} > 1.01\).
+The aim is therefore to reach approximate convergence \citep{gelm13}.
+Upon approximate convergence, the imputation chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm (\(\widehat{R} > 1\)), and there is little dependency between subsequent iterations of imputation chains (\(AC > 0\)).
+\textbf{Approximate convergence is violated when the non-convergence diagnostics exceed a certain threshold.
+{[}or{]} In practice, non-convergence is therefore diagnosed when non-convergence diagnostics exeed a certain threshold.} The conventional thresholds to diagnose non-mixing are \(\widehat{R} > 1.2\) \citep{gelm92} or \(\widehat{R} > 1.1\) \citep{gelm13}.
+\citet{veht19} proposed a much more stringent threshold of \(\widehat{R} > 1.01\).
 The magnitude of \(AC\)-values may be evaluated statistically, using a Wald test with \(AC=0\) as null hypothesis \citep{box15}.
@@
 Before we evaluate the performance of the non-convergence diagnostics \(\widehat{R}\) and \(AC\) quantitatively through simulation, we apply them to the example of pathological non-convergence by \citet{buur18} that we reproduced above.
-We want to be able to draw the same conclusions as we did from visually inspecting the two algorithms.
-But \(\widehat{R}\) and \(AC\) should at least be able to distinguish between the two algorithms plotted in Figure \ref{fig:non-conv}.
-From visual inspection, we know that the typically converged algorithm initially portrayed some signs of non-mixing (around \(t=2\)), but intermingled nicely overall.
-Additionally, there was very little trending.
-The algorithm with pathological non-convergence showed severe non-mixing, although this gradually improved (from \(t=7\)).
-Moreover, there was a lot of trending as well initially (up-to \(t=6\)), after which the chains reached a somewhat more stationary state.
-To assess whether \(\widehat{R}\) and \(AC\) may be appropriate non-convergence identifiers for iterative imputation algorithms, they should identify the same trends as the visual inspection.
-But at least, the following condition must hold: the methods should indicate worse performance for the mis-specified model with pathological non-convergence than the typically converged algorithm (i.e., higher \(\widehat{R}\)- and \(AC\)-values).
+We want to be able to draw the same conclusions as we did from visually inspecting the two scenarios in Figure \ref{fig:non-conv}.
+The methods should at least indicate worse performance for the scenario with pathological non-convergence compared to the typically converged algorithm (i.e., higher \(\widehat{R}\)- and \(AC\)-values).
 
@@
 
+\textbf{{[}REPHRASE as active about panel A{]}} From visual inspection, we know that the typically converged algorithm initially portrayed some signs of non-mixing (around \(t=2\)), but intermingled nicely overall.
+Additionally, there was very little trending.
+The algorithm with pathological non-convergence showed severe non-mixing, although this gradually improved (from \(t=7\)).
+In this scenario, there was a lot of trending as well initially (up-to \(t=6\)), after which the chains reached a somewhat more stationary state.
+
 When we look at panel B, we conclude something weird.
@@
 Moreover, the best convergence (as indicated by the lowest \(AC\)-value) is observed at \(t=2\), but looking at the chain means in panel A, there should be some signs of trending up-to iteration number seven.
-After consulting the documentation on \texttt{stats::acf()} \citep{R}, we conclude that this \(AC\) function is not suited for iterative imputation algorithms.
+After consulting the documentation on \texttt{stats::acf()}, we conclude that this \(AC\) function is not suited for iterative imputation algorithms.
 The function is optimized for performance when \(t\geq50\) \citep{box15}, while the default number of iterations in iterative imputation is often much lower.
@@
 \textbf{In conclusion, we will use \(\widehat{R}\) conform \citet{veht19} and \(AC\) computed manually as non-convergence diagnostics in the simulation study.
-The diagnostics will be applied to four \(\theta\)s of interest: chain means, chain variances, a quantity of scientific interest, and the first eigenvalue of the variance-covariance matrix.}
+The diagnostics will be applied to four \(\theta\)s of interest: chain means, chain variances, a quantity of scientific interest, and the first eigenvalue of the variance-covariance matrix.
+Results in eight methods to identify non-convergence}
 
@@
 
-The aim of the simulation study is to evaluate the \textbf{impact of inducing non-convergence {[}``early stopping'' en ``proportion of complete cases''{]}} in the MICE algorithm on several quantities of scientific interest \(Q\).
-Subsequently we will evaluate how well \(\widehat{R}\) and \(AC\) perform in identifying the effects of non-convergence \textbf{{[}and performance of different thetas{]}}.
+The aim of the simulation study is to determine the \textbf{impact of non-convergence {[}``early stopping'' en ``proportion of complete cases''{]}} in iterative imputation algorithms.
+Specifically, we are interested in the bias and confidence-validity of estimates obtained from incomplete data with the MICE algortihm.
+
+\textbf{{[}REMOVE?{]} Several quantities of scientific interest \(Q\).
+Subsequently we will evaluate how well \(\widehat{R}\) and \(AC\) perform in identifying the effects of non-convergence }{[}and performance of different thetas{]}\textbf{.
 And finally, we will formulate an informed advice on the requirements to safely conclude sufficient convergence in practice.
-That is, we conclude that convergence is sufficient when each estimate \(\bar{Q}\) is an unbiased and confidence-valid estimate of the corresponding estimand \(Q\).
-\textbf{{[}and formulate advice on thetas?{]}}
+That is, we conclude that convergence is sufficient when each estimate \(\bar{Q}\) is an unbiased and confidence-valid estimate of the corresponding estimand \(Q\).}
 
-Non-convergence will be induced by: 1) increasing the missingness proportion \(p_{\rm mis}\) in dataset \(y\) incrementally, and 2) terminating the imputation algorithm at a varying number of iterations \(T\).
-The first set of simulation conditions---the missingness proportions---is chosen to reflect the difficulty of the missingness problem.
+To induce non-convergence, we consider two sets of simulation conditions: we vary the missingness proportion \(p_{\rm mis}\) in the incomplete data, and we vary the number of iterations \(T\) in the imputation algorithm.
+The missingness conditions are chosen to reflect the difficulty of the missingness problem.
 The underlying assumption is that low missingness proportions lead to quick algorithmic convergence, since there is a lot of information in the observed data.
@@
 
-\textbf{{[}SEGWAY to this section{]} \(Q\)s are the targets that will be estimated in each simulation repetition, after first \emph{amputing} the complete data with varying missingness proportions, then imputing the missingness with a varying number of iterations, then estimating the \(Q\)s on the \(m\) completed datasets, and finally pooling the \(m\) estimates \(\hat{Q}_\ell\) to obtain a single \(\bar{Q}\) for each \(Q\).
+\textbf{{[}REMOVE?{]} \(Q\)s are the targets that will be estimated in each simulation repetition, after first \emph{amputing} the complete data with varying missingness proportions, then imputing the missingness with a varying number of iterations, then estimating the \(Q\)s on the \(m\) completed datasets, and finally pooling the \(m\) estimates \(\hat{Q}_\ell\) to obtain a single \(\bar{Q}\) for each \(Q\).
 We try to predict the effect of non-convergence (i.e., the difference between \(Q\) and \(\bar{Q}\)) with the convergence diagnostics \(\widehat{R}\) and \(AC\), by applying these methods to several scalar summaries of the state space of the algorithm, \(\theta\)s.}
@@
 In contrast to this, it may occur that signs of non-convergence are under-estimated by \(\widehat{R}\), in exceptional cases where the initial values of the algorithm are not appropriately over-dispersed \citep[p.~437]{broo98}.
-In e.g.~\texttt{mice}, initial values are chosen randomly from the observed data, hence we cannot be certain of over-dispersion in the initial values.
+In \texttt{mice}, initial values are chosen randomly from the observed data, hence we cannot be certain of over-dispersion in the initial values.
 In practice, we do not expect this to cause problems for identifying non-convergence with \(\widehat{R}\).
+\textbf{{[}ADD limitation that other MI packages may have different way to obtain initial values!{]}}
 \item
   We expect that high \(AC\) values are implausible in iterative imputation algorithms with typical convergence.
-That is, after only a few iterations, the randomness induced by the algorithm should effectively mitigate the risk of dependency within chains.
+After only a few iterations, the randomness induced by the algorithm should effectively mitigate the risk of dependency within chains.
 \item
@@
 We investigate non-convergence of the MICE algorithm through model-based simulation in \texttt{R} \citep[version 3.6.3;][]{R}.
-The number of simulation repetitions is 1000.
-The simulation set-up is summarized in the pseudo-code below.
+The number of simulation repetitions is 1000, and the simulation set-up is summarized in the pseudo-code below.
 The complete \texttt{R} script of the simulation study is available from \href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/}{github.com/gerkovink/shinyMice}.
@@
 where \(\epsilon \sim \mathcal{N}(0, 100)\).
-This results in a complete set \(y\), with observed values for all units \(i\) on all variables \(j\) (\(j = Y, X_1, X_2, X_3\)).
+This results in a complete set \(y\), with observed values for all units \(i\) on all variables \(j\) (where \(j = Y, X_1, X_2, X_3\)).
 From the complete set we obtain the true values of the scientific estimands \(Q\).
@@
 We consider four types of \(Q\)s that are often of interest in empirical research.
-Namely, two descriptive statistics: the mean \(\mu_j\) and standard deviation \(\sigma_j\) of each variable \(j = Y, X_1, X_2, X_3\).
+Namely, two descriptive statistics: the mean \(\mu_j\) and standard deviation \(\sigma_j\) of each variable \(j\).
 We also consider the regression coefficients of the predictors: \(\beta_1\), \(\beta_2\), and \(\beta_3\) in regression equation
@@
 And finally, the proportion of variance explained by the set of predictors: coefficient of determination \(r^2\) \textbf{(note that lower case \(r\) is used to avoid confusion with non-convergence diagnostic \(\widehat{R}\))}.
-The true values of these scientific estimands are calculated on the complete data.
+These \(Q\)s are estimated by \(\bar{Q}\) in each simulation condition.
 
-\hypertarget{methods.}{%
-\subsubsection{Methods.}\label{methods.}}
+\hypertarget{simulation-conditions.}{%
+\subsubsection{Simulation conditions.}\label{simulation-conditions.}}
 
@@
 Let
-\[\text{CI} = \bar{Q} \pm t_{(M-1)} \times SE_{\bar{Q}},\]
-where \(t_{(M-1)}\) is the quantile of a \(t\)-distribution with \(m-1\) degrees of freedom, and \(SE_{\bar{Q}}\) is the square root of the pooled variance estimate.
-If we obtain nominal coverage (CR = 95\%), we can conclude that the \(\bar{Q}\)s are confidence-valid estimates of the \(Q\)s.
+\[\text{CI} = \bar{Q} \pm t_{(m-1)} \times \rm SE_{\bar{Q}},\]
+where \(t_{(M-1)}\) is the quantile of a \(t\)-distribution with \(m-1\) degrees of freedom, and \(\rm SE_{\bar{Q}}\) is the square root of the pooled variance estimate.
+If we obtain nominal coverage (CR = 95\%), we can conclude that \(\bar{Q}\) is a confidence-valid estimate of \(Q\).
 Finally, we inspect CI width (CIW): the difference between the lower and upper bound of the 95\% confidence interval around \(\bar{Q}\).
 CIW is of interest because it is a measure of efficiency.
-Under nominal coverage, short CIs are preferred.
-Too short CIs, however, indicate under-estimation of the variance in \(\bar{Q}\), \textbf{which may yield inferences with low statistical power if CIW is large}.
+Under nominal coverage, short CIs are preferred, since wider CIs indicate lower statistical power.
 
-With these performance measures, we also evaluate how well the non-convergence diagnostics \(\widehat{R}\) and \(AC\) can identify simulation conditions in which the inferences are affected by non-convergence, i.e., where \(\bar{Q}\) is \emph{not} an unbiased, confidence-valid estimate of \(Q\).
-\textbf{{[}REPHRASE this:{]} For performance of univariate \(Q\)s, we apply the non-convergence diagnostics to univariate \(\theta\)s.
-When \(Q\) is multivariate, we look at signs of non-convergence identified in multivariate \(\theta\)s.}
-
-\begin{itemize}
-\item
-  For \(Q=\mu\), \(\theta\)s are chain means \textbf{{[}ADD that these should be unbiased because MCAR!{]}}.
-\item
-  For \(Q=\sigma\), \(\theta\)s are chain variances.
-\item
-  For \(Q=\beta\), \(\theta\)s are estimated \(\beta\)s per imputation.
-\item
-  For \(Q=r^2\), \(\theta\)s are the first eigenvalue of the variance-covariance matrix, \(\lambda_{1,\ell}\).
-\end{itemize}
+We evaluate the methods to identify non-convergence against these performance measures.
+\(\widehat{R}\) and \(AC\) should identify non-convergence in simulation conditions, where \(\bar{Q}\) is \emph{not} an unbiased, confidence-valid estimate of \(Q\).
 
@@
 
-Our results show the impact of \textbf{inducing non-convergence {[}``early stopping'' en ``proportion of complete cases'' {]}} in an iterative imputation algorithm on several scientific estimates, and whether the non-convergence diagnostics \(\widehat{R}\) and \(AC\) identify the signs of non-convergence in the algorithm.
-For reasons of brevity, we only discuss the non-convergence diagnostics for estimates with the worst performance in terms of bias.
-Bias in the descriptive statistics (\(Q=\mu\) and \(Q=\sigma\)) is most pronounced for the outcome variable \(Y\).
-Of the regression coefficients (\(Q=\beta\)) we observe the largest bias for \(\beta_1\) (the effect of \(X_1\) on \(Y\)).
-There is only one estimate for \(Q=r^2\) to consider.
-In the figures, we present simulation conditions where the number of iterations is \(1 \leq T\leq50\), since the results are more or less stable for conditions where \(T \geq 30\).
+Our results show the performance of the MICE algorithm under different conditions of missingness and iteration length.
+For reasons of brevity, we only discuss results for the worst-performing estimates in terms of bias.
+For univariate scientific estimands (\(Q=\mu\) and \(Q=\sigma\)) we observe the largest bias in the outcome variable \(Y\).
+And for the regression coefficients (\(Q=\beta\)) the bias is most pronounced in \(\beta_1\) (the effect of \(X_1\) on \(Y\)).
+Since there is just one estimate for \(Q=r^2\) to consider, we \textbf{end up with} \(Q=\mu_Y, \sigma_Y,r^2, \beta_1\).
+In the figures, we present all missingness conditions, but only \textbf{iteration length conditions} \(1 \leq T\leq50\), since the results are more or less stable for conditions where \(T \geq 30\).
 Full results are available from \href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy}{github.com/gerkovink/shinyMice}.
@@
 The estimated univariate means seem unaffected by the number of iterations in the algorithm.
-This implies that valid inferences may be obtained with as little as one iteration (\(T \geq 1\)).
-The magnitude of the bias in \(\bar{Q}\) depends solely on the missingness proportion \(p_{\rm mis}\), with missingness proportions of 75\% and 95\% leading to more extreme biases than other missingness proportions \textbf{{[}note that these are curious{]}}.
-Unbiased estimates are obtained in all conditions where \(p_{\rm mis}\leq.50\).
-Note however, that even in the conditions where \(p_{\rm mis}\geq.75\), the bias is small: \(\mu_Y\) is maximally under-estimated by 0.02 units, while the true value of \(\mu_Y\) is 25.81 (\(\sigma_Y = 11.32\)).
-We conclude that non-convergence does not substantially impact the validity of the inferences for \(Q=\mu\).
-\textbf{{[}To be expected, because MCAR{]}}
+This implies that approximately unbiased estimates may be obtained with as little as one iteration (\(T \geq 1\)).
+The magnitude of the bias in \(\bar{Q}\) depends solely on the missingness proportion \(p_{\rm mis}\), with missingness proportions of 75\% and 95\% leading to more extreme biases than other missingness proportions.
+Completely unbiased estimates are only obtained in conditions where \(p_{\rm mis}\leq.50\).
+\textbf{This is curious because the missingness mechanism MCAR should yield unbiased univariate estimates without employing multiple imputation.} Note however, that the magnitude of the bias in these conditions is small: \(\mu_Y\) is maximally under-estimated by 0.02 units, while the true value of \(\mu_Y\) is 25.81 (\(\sigma_Y = 11.32\)).
+We, therefore, conclude that non-convergence does not substantially impact the validity of the inferences for \(Q=\mu\).
 
@@
 
-The estimated standard deviations are largely affected by the missingness proportions, but not to much by the number of iterations.
-Valid inferences are obtained in conditions with just two iterations.
-Interestingly, the bias in first iteration is actually \emph{less} extreme than for any other number of iterations.
-\textbf{{[}Due to MCAR and randomly drawn initial values?{]}} We may therefore conclude that no iteration at all is needed to obtain approximately unbiased estimates for \(Q=\sigma\).
+The estimated standard deviations not affected substantially by \textbf{early stopping} either.
+While stable estimates are obtained in conditions with two iterations, the bias in the first iteration is actually \emph{less} severe than in other \textbf{iteration conditions}.
+Therefore, no iteration at all is needed to obtain approximately unbiased estimates for \(Q=\sigma\).
 Completely unbiased estimates are only obtained in conditions where \(p_{\rm mis}=.05\).
@@
 
-Bias in the estimated variance explained depends both on missingness proportions and the number of iterations.
-The magnitude of the bias clearly follows the incremental order of the missingness proportions.
-Completely unbiased estimates may only be obtained if the missingness proportion is at most 25\%.
-Approximately unbiased estimates are observed for any missingness proportion.
-For \(p_{\rm mis}\leq.25\), the minimum number of iterations is two (\(T\geq2\)), for \(p_{\rm mis}=.50\) it is \(T\geq4\), and \(p_{\rm mis}=.75\) requires \(T\geq5\).
-The highest number of iterations needed to obtain valid inferences is seven (\(p_{\rm mis}=.95\); \(T\geq7\)).
-\textbf{{[}interpret sign of bias?
-ADD what this implies: dat - zelfs bij een gebiasede methode - je maar 7 iteraties nodig hebt om het eindresultaat te bereiken.{]}}
+The estimated coefficient of determination is clearly affected by early stopping.
+As expected, we observe worse performance in conditions with a lower number of iterations.
+The magnitude of the bias, however, depends on the missingness proportion---following the incremental order in \(p_{\rm mis}\).
+\textbf{And so does} the number of iterations necessary to reach stable, approximately unbiased estimates.
+In conditions where \(p_{\rm mis}=.05\), estimates are unbiased after one iteration.
+We observe approximately unbiased estimates when \(T\geq 2,4, \text{and }5\), in conditions where \(p_{\rm mis} = .25, .50, \text{and }.75\), respectively.
+The highest number of iterations necessary to reach approximate unbiasedness is seven.
+Completely unbiased estimates are only obtained in conditions where \(p_{\rm mis}\leq.25\).
+This implies that even for biased estimates, \(T\geq7\) would suffice to reach a stable solution.
 
@@
 
-For the estimated regression coefficients, we consider bias, confidence interval width (CIW), and coverage rate (CR) as performance measures.
-\textbf{{[}Split panels{]}} Unbiased estimates are obtained in conditions where \(p_{\rm mis}\leq.50\) and \(T\geq3\).
-Approximately confidence-valid estimates are obtained in any condition where \(T\geq3\)---despite of persistant bias in conditions where \(p_{\rm mis}\geq.75\).
-These approximately nominal coverages are explained by the CIWs: as expected, conditions with higher missingness proportions result in more uncertainty in the estimates (i.e., wider CIs), which diminish the effect of the bias.
-confidence-valid inferences may be obtained when \(T\geq3\), while we need at least seven iterations to reach valid inferences in terms of bias.
-Continued iteration after \(T\geq7\) seems futile.
-\textbf{{[}add max T for other betas?{]}}
+For the estimated regression coefficients, we first consider bias, before discussing CR and CIW.
+As Figure \ref{fig:Qs} shows, the bias in \(\bar{Q}\) is affected by both the missingness proportion and the number of iterations.
+Approximately unbiased estimates are observed after one to seven iterations, depending on \(p_{\rm mis}\).Completely unbiased estimates are only obtained in conditions where \(T\geq3\) and \(p_{\rm mis}\leq .50\).
+
+Despite persistant bias in conditions where \(p_{\rm mis}\geq.75\), coverage rates indicate that all \textbf{missingness conditions} have nominal coverages after just three iterations.
+Confidence-valid estimates may be obtained in any condition where \(T\geq3\).
+
+The nominal coverages are explained by the CIWs.
+Conform the theoretical foundation of MI, CIs are wider in conditions with higher missingness proportions (i.e., there is less information in the data, and thus more uncertainty due to missingness).
+The wider CIs around \(\bar{Q}\) in conditions with higher missingness proportions \textbf{results in the inclusion of} the true value of \(Q\) despite of bias.
+
+We conclude that approximately unbiased estimates may be obtained in conditions where \(T\geq7\), whereas confidence-valid estimates require at most three iterations.
 
@@
 
-Overall, we see that conditions with a higher proportion of missingness are affected more severely by non-convergence.
-Multivariate \(Q\)s are also substantially influenced by the number of iterations.
-This means that the algorithm has indeed not converged at \(t=1\), and converges gradually as \(t\) moves to \(T\).
-The point at which an additional iteration does not lead to an improvement in the estimates depends on quantity of scientific interest \(Q\) and the missingness proportion \(p_{\rm mis}\).
-The highest number of iterations that is necessary to reach this point across all simulation conditions is \(T=7\).
-This implies that, under the current specifications, no more than seven iterations are necessary to obtain sufficient performance in terms of bias and confidence-validity.
+These results demonstrate that the estimates of univariate scientific estimands \(Q\) are not impacted by early stopping of the MICE algorithm, irrespective of the missingness proportion in \(y\).
+Unbiased estimates may be obtained after just one iteration.
+Multivariate estimates, by contrast, are affected by both the number of iterations and the missingness proportion.
+Completely unbiased estimates are only observed under low to moderate missingness (\(p_{\rm mis}\leq.50\)), after at most three iterations (\(T\leq1,2,3\), depending on \(p_{\rm mis}\)).
+Approximately unbiased estimates are observed for any missingness condition, after at most seven iterations.
+This implies that the algorithm produces stable, unimproving estimates for any condition where \(T\geq7\).
 
@@
 
-We evaluate the non-convergence diagnostics \(\widehat{R}\) and \(AC\) against the performance measures described above (i.e., bias and confidence-validity of the quantities of scientific interest \(Q\)).
-The diagnostics should identify those simulation conditions that are negatively impacted by non-convergence.
-We expect the most signs of non-convergence in conditions where valid estimates are not obtained (high missingness proportions, \(T<7\)), and at least some signs of non-convergence in conditions where estimates are not approximately unbiased (\(p_{\rm mis}=.95\), any number of iterations).
+Figure \ref{fig:rhats-acs} displays results for the two non-convergence diagnostics applied to the four \(\theta\)s.
+\textbf{{[}describe panels{]}} We evaluate the performance of these methods by establishing whether they correctly identify conditions in which \(\bar{Q}\) is \emph{not} an unbiased, confidence-valid estimate of \(Q\).
+Biased estimates are observed in conditions where \(T\leq7\).
+Non-nominal coverage rates are observed only when \(T\leq3\).
+Additionally, there is persistant bias in conditions where \(p_{\rm mis}\geq.75\), irrespective of the iteration length.
+\textbf{{[}appropriate methods thus identify these conditions{]}}
 
@@
 
-\textbf{{[}As expected, \(\widehat{R}\) lowers with the number of iterations.
-At a cut-point of 1.2 , we have convergence at about 8 iteration, whereas we need 12 and 25 iteration for more conservative cut-off 1.1 and 1.05.
-We do not reach the very strict 1.01.{]}}
+As expected, \(\widehat{R}\) lowers with the number of iterations.
+After an initial `dip' in conditions where \(T=3\), we observe a gradual decrease in \(\widehat{R}\)-values with every additional iteration.
+This implies improving convergence, until the decrease tapers off after 30 to 40 iterations.
+Using the threshold \(\widehat{R}>1.2\), we diagnose non-convergence in conditions where \(T\leq7\), whereas the threshold 1.1 is exceeded when \(T\leq13\).
+The very strict threshold 1.01 is surpassed in all conditions \(T\leq100\)\textbf{, and therefore not informative in the current study}.
+With this method and \(\theta\), we fail to identify simuation conditions with persistant bias due to high missingness proportions.
 
-We evaluate the performance of the non-convergence diagnostics applied to chain means against the bias in \(Q=\mu\).
-Based on the bias in \(Q\), we concluded that one iteration is sufficient to obtain valid inferences, irrespective of the missingness proportion.
-The non-convergence diagnostics, however, identify some signs of non-convergence in conditions with as many as thirty iterations.
-This does not correspond to the impact of non-convergence in the univariate \(Q\)s.
-\textbf{Compared to the multivariate \(Q\)s {[}add before that we evaluate the thetas against all Qs?{]}, \(\widehat{R}\) and \(AC\) applied to the chain means fail to identify the non-convergence in conditions where \(p_{\rm mis}=.95\).}
+Similarly, \(AC\) decreases with higher \(T\)s.
+\(AC\) indicates improving stationarity in conditions where \(T\leq6\).
+The \(AC\)-values do not exceed the threshold \textbf{defined by statistically significant \(AC\)s}.
+Based on this threshold, we fail to not diagnose non-convergence in any condition.
 
@@
 
-The non-convergence diagnostics applied to the chain variances as \(\theta\) yield highly similar results to the chain means.
-\textbf{{[}Segway to performance of \(Q=\sigma\){]} While the bias in the estimated standard deviations mainly depends on the missingness proportion of each condition, \(\widehat{R}\) and \(AC\) are influenced mostly by the number of iterations.
-We thus conclude that there is no added value of evaluating non-convergence with chain variances as \(\theta\) in the diagnostics.}
+As Figure \ref{fig:rhats-acs} shows, the results of the two diagnostics applied to the chain variances are highly similar to those on the chain means.
+This holds for both \(\widehat{R}\) and \(AC\).
+The only apparent difference is in the \(\widehat{R}\) threshold 1.1.
+Instead of diagnosing non-convergence in conditions where \(T\leq13\), we now diagnose this when \(T\leq11\).
+All other observations are equivalent.
 
@@
 
-\textbf{{[}We zien weer hetzelfde voor laag-medium, beetje meer voor extreme misisnh data.{]}}
+The \(\widehat{R}\)-values show a similar trend across iterations as we observed for chain means and chain variances.
+Only conditions with very high missingness proportions (\(p_{\rm mis}\geq.75\)) diverge from the earlier observations.
+The \(\widehat{R}\)-values in these conditions do not taper off after 30 to 40 iterations, but rather between 40 to 50 iterations.
+The highest number of iterations at which non-convergence is still diagnosed according to the thresholds are: 11 for \(\widehat{R}>1.2\), 23 for \(\widehat{R}>1.1\), and 100 for \(\widehat{R}>1.01\).
 
-The non-convergence diagnostics are applied to the estimated regression coefficients in each imputation, and evaluated against the bias in the corresponding pooled estimate.
+The \(AC\)-values also follow earlier observations, with the exception of the same very high missingness proportions that \textbf{diverged} for \(\widehat{R}\).
+To reach maximum stationarity, these conditions require not five but six or seven iterations.
+Moreover, the \(AC\)-values in these conditions exceed the threshold to diagnose non-convergence.
+Interestingly, this only occurs after ten or even thrity iterations, while we would expect more signs of trending early in the iterations.
+\textbf{Since the \(AC\)-values are not improving or worsening after seven iterations, we conclude that non-convergence is incorrectly diagnosed based on the threshold.}
 
-\(\widehat{R}\) indicates signs of non-convergence across all missingness proportions and each number of iterations, in the expected order: simulation conditions with higher missingness proportions and less iterations are identified as the worst converged.
-The thresholds to diagnose non-convergence are overcome in different conditions, depending on the number of iterations and the missingness proportion.
-The most lenient threshold (\(\widehat{R}>1.2\)) is overcome at \(T\geq9\) for \(p_{\rm mis} = 0.05\), and \(T\geq12\) for \(p_{\rm mis} = 0.95\).
-The conventional \(\widehat{R}>1.1\) is surpassed at \(T\geq12\) and \(T\geq24\), respectively.
+\hypertarget{thetalambda_1-ell.}{%
+\subsubsection{\texorpdfstring{\(\theta=\lambda_{1, \ell}\).}{\textbackslash theta=\textbackslash lambda\_\{1, \textbackslash ell\}.}}\label{thetalambda_1-ell.}}
 
-For the autocorrelations applied to the \(\hat{Q}_\ell\)s, we see an even stronger effect of the missingness proportions.
-Especially the conditions where \(p_{\rm mis} = 0.75\) and \(p_{\rm mis} = 0.95\) are identified as non-stationary.
-The threshold used to evaluate whether \(AC=0\), however, is only reached in conditions with at least ten iterations.
-This suggests that the default number of iterations may be too low to detect non-stationarity with \(AC\).
-However, the critical value for diagnosing non-convergence with \(AC\) decreases as a function of the number of iterations.
-The conclusion to identify non-stationarity in conditions where \(T\geq10\) and \(p_{\rm mis} = 0.95\) might therefore be spurious.
+Once again, we observe \(\widehat{R}\)-values with a similar trend across iterations.
+For this \(\theta\), the results differ only in the most extreme missingness condition.
+We diagnose non-convergence in conditions where \(T\leq9\) according to \(\widehat{R}>1.2\), and \(T\leq19\) for \(\widehat{R}>1.1\).
+The `dip' in \(\widehat{R}\)-values at \(T=3\) is even more pronounced than with other \(\theta\)s.
+Some \(\widehat{R}\)-values even overcome the threshold of 1.2.
+If we would terminate the algorithm after three iterations, we would incorrectly conclude sufficient convergence according to the \(\widehat{R}>1.2\) threshold.
 
-Compared to the bias in \(\bar{Q}\), the diagnostics applied to the \(\hat{Q}_\ell\)s over-estimate the severity of the non-convergence.
+\(AC\)-values are systematically higher for this \(\theta\) than in the other results.
+The \(AC\)-values also taper off at a later point in the iterations, indicating improving stationarity up-to ten, or even thirty iterations, depending on the missingness proportion.
+Using the threashold, we diagnose non-convergence at some point for any missingness proportion.
 
@@
 
-In general, the diagnostics identify more signs of non-convergence in simulation conditions with a lower number of iterations.
-The missingness proportions, however, only seem to impact multivariate convergence.
-When \(\widehat{R}\) and \(AC\) are applied to univariate \(\theta\)s, they identify the same amount of non-convergence for any missingness proportion.
-When the diagnostics are applied to multivariate \(\theta\)s, they generally identify more signs of non-convergence for higher missingness proportions.
-
-As expected, none of the simulation conditions that we consider resulted in complete convergence of the algorithm (defined as \(\widehat{R}=1\) and \(AC=0\)).
-However, the thresholds used in practice to diagnose non-convergence seem too conservative for imputation purposes.
-\(\widehat{R}\) and \(AC\) identify signs of non-convergence in conditions where the \(\bar{Q}\)s are unbiased and confidence-valid estimates of the \(Q\)s.
-The most recent recommended threshold for diagnosing non-convergence with \(\widehat{R}\) is never overcome in this study (all \(\widehat{R}\)\textgreater1.01).
-Apparently, this threshold is too stringent for iterative imputation algorithms, and will not be discussed any further.
-
-\hypertarget{thetalambda_1-ell.}{%
-\subsubsection{\texorpdfstring{\(\theta=\lambda_{1, \ell}\).}{\textbackslash theta=\textbackslash lambda\_\{1, \textbackslash ell\}.}}\label{thetalambda_1-ell.}}
-
-Finally, we evaluate the non-convergence diagnostics applied to the first eigenvalue of the variance-covariance matrices of the completed data.
-Similar to the results the the regression coefficients as \(\theta\)s, we see effects of both the missingness proportion and number of iterations in each simulation condition.
-The most signs of non-convergence are identified in conditions where \(p_{\rm mis}\) is large and \(T\) is small.
-
-Notwithstanding the similarities with \(\theta=\hat{Q}_\ell\), there is a big difference in \(\widehat{R}\)-values for conditions where \(T<4\).
-In these conditions, \(\widehat{R}\) behaves oppositely of our expectation: conditions with the lowest missingness proportions yield the highest \(\widehat{R}\)-values.
-In conditions with a higher number of iterations, this artifact disappears.
-This suggests that the combination of \(\widehat{R}\) and \(\lambda_{1, \ell}\) is only appropriate for imputation algorithms where the number of iterations at least four.
-Interpreting the \(\widehat{R}\) values of conditions where \(T<3\) may lead to the incorrect conclusion that only conditions with \emph{low} missingness proportions show signs of non-convergence.
-For applications as a model-independent multivariate \(\theta\), we should therefore only interpret the \(\widehat{R}\)-values when \(T\geq4\).
-Using this cut-off we conveniently ignore the initial `dip' at \(T=3\) as well.
-The traditional threshold \(\widehat{R} > 1.2\) is overcome in conditions where \(p_{\rm mis} = .95\) and \(T\geq10\), and for all other missingness proportions in conditions where \(T\geq8\).
-With the \(\widehat{R} > 1.1\) threshold, the number of iterations are \(T\geq20\) and \(T\geq16\), respectively.
-
-The \(AC\)-values largely depend on the missingness proportion in each condition.
-Conditions where \(p_{\rm mis}=.95\) show statistically significant autocorrelations for any number of iterations.
-The other extreme, conditions where \(p_{\rm mis}=.05\), does not yield \(AC\) values above the threshold.
-The \(AC\) values with this missingness proportion are equivalent for all conditions where \(T\geq6\), which suggests that the algorithm is as stationary as it can be.
-A similar point is only observed after twenty iterations in conditions with higher missingness proportions.
-
-In comparison to the \(Q\)s, the diagnostics again over-estimate the signs of non-convergence.
-Conditions with low missingness proportions (e.g., 5\% or 25\% of the complete data) yield unbiased, confidence-valid estimates with as little as two iterations.
-The convergence diagnostics, however, indicate improved convergence up-to eight iterations according to \(AC\) and thirty to forty iterations according to \(\widehat{R}\).
-The model-independent \(\theta\) \(\lambda_{1, \ell}\) seems as well-equipped to diagnose signs of non-convergence as \(\theta=\hat{Q}_\ell\).
-The univariate \(\theta\)s do not differentiate between bias induced by the missingness proportions.
+Overall, the methods to diagnose non-convergence perform as expected: they indicate more signs of non-convergence in conditions with worse performance in terms of bias and confidence-validity.
+Under low to moderate missingness, it does not seems to matter on which \(\theta\) the diagnostics are applied.
+If we look at complete unbiasedness, however, we notice that univariate \(\theta\)s fail to diagnose the persistant bias in conditions where \(p_{\rm mis}\geq.75\).
+The number of iterations necessary to obtain approximately unbiased, confidence-valid estimates corresponds to the \(\widehat{R}\) threshold 1.2.
+The thresholds 1.1 and 1.01 seem to strict compared to the validity of the inferences.
+The \(AC\) threshold defined by the statistical significance of the \(AC\)-values does not appear to be appropriate in the current set-up \textbf{{[}or iterative imputation in general??{]}}.
+Non-stationarity may be better diagnosed by evaluating the \(AC\)-values across iterations.
+If the values do not substantially decrease, stationarity may be concluded \textbf{{[}similar to the elbow heuristic in PCA{]}}.
 
@@
 \section{Discussion}\label{discussion}}
+
+H1: As hypothesized, biased, invalid estimates occur more often in simulation conditions with a higher missingness proportion and lower number of iterations.
+However, this only holds for multivariate scientific estimands.
+Univariate \(Q\)s are not substantially impacted by early stopping.
+
+H2: As expected, we observe higher \(\widehat{R}\) and \(AC\)-values in conditions where \(\bar{Q}\)s were biased, invalid estimates of \(Q\)s.
+
+H3: The hypothesis that the recommended thresholds would be too strict is only patially supported.
+\(\widehat{R}\) indicated improving convergence until 30 to 40 iterations, while approximately unbiased, confidence-valid estimates were obtained after just seven iterations.
+This rougly corresponds to the threshold of 1.2.
+The more stringent trheshold 1.1 seems plausible compared to the decrease in \(\widehat{R}\) values across iterations.
+The most recent threshold 1.01 does indeed seem to over-estimate the signs of non-convergence.
+
+H4: The results of this simulation study do not support the hypothesis that \ldots{} \textbf{{[}insert hypothesis{]}}.
+We observed \(AC\)-values as high as theoretically possible \(AC=\).
+The second part of the hypothesis is not refuted: the \(AC\)-values did decrease quicky as a function of \(T\).
+
+H5: We did find some support for the final hypothesis.
+Multivariate \(\theta\)s indeed have superior performance under certain conditions (e.g., multivariate \(Q\)s and extreme missingness proportions).
+
+\hypertarget{conlusion}{%
+\section{Conlusion}\label{conlusion}}
 

### 2020-05-09 3.Thesis/Writeup/2.Manuscript.Rmd
Last feedback before hand-in

@@
 Moreover, the behavoir of such algorithms under certain default imputation models (e.g., 'predictive mean matching') is an entirely open question [@murr18].
-**Therefore, algorithmic convergence should be monitored carefully, **although not straightforward.
+Therefore, algorithmic convergence should be monitored carefully, **but this is not straightforward**.
 Iterative imputation algorithms such as MICE are special cases of Markov chain Monte Carlo (MCMC) methods.
@@
 
-**[Include sub-questions and make structure of the paper clear!]** How can non-convergence be identified diagnostically?
+<!-- **[Include sub-questions and make structure of the paper clear!]** How can non-convergence be identified diagnostically?
 Are common MCMC non-convergence diagnostics appropriate for MICE?
@@
 Can the parameter of interest $Q$ be correct when the algorithm is not (yet) converged, and vice versa?
+-->
 
@@
 The $\theta$ in this example is the chain mean of a variable that we will call $j$ (i.e., the average of the imputed values for $j$ in each imputation $\dot y_{\rm imp, \ell}$).
-**In the typical convergence scenario, the imputation chains intermingle nicely and there is little to no trending.
-In the non-convergence scenario, there is a lot of trending and some chains do not intermingle.** Importantly, the chain means at the last iteration (the imputed value per imputation $\ell$) are very different between the two plots.
+In the typical convergence scenario, the imputation chains intermingle nicely and there is little to no trending.
+In the non-convergence scenario, there is a lot of trending and some chains do not intermingle.
+Importantly, the chain means at the last iteration (the imputed value per imputation $\ell$) are very different between the two plots.
 The algorithm with the mis-specified model yields imputed values that are on average a factor two larger than those of the typically converged algorithm.
@@
 $\lambda_{1, \ell}$ is measure that summarizes the covariances in the completed datasets.
-**The first eigenvalue has the appealing property that is not dependent on the substantive model of interest.
-Eigenvalue decomposition is inspired by @mack03.**
+The first eigenvalue has the appealing property that is not dependent on the substantive model of interest.
+Eigenvalue decomposition is inspired by @mack03.**[ADD definition variance in first PCA component is equal to the first eigenvalue of the variance-covariance matrix]**
 
@@
 Specifically, we are interested in the bias and confidence-validity of estimates obtained from incomplete data with the MICE algortihm.
-
-**[REMOVE?] Several quantities of scientific interest $Q$.
-Subsequently we will evaluate how well $\widehat{R}$ and $AC$ perform in identifying the effects of non-convergence **[and performance of different thetas]**.
+**[REMOVE?] Subsequently we will evaluate how well $\widehat{R}$ and $AC$ perform in identifying the effects of non-convergence **[and performance of different thetas]**.
 And finally, we will formulate an informed advice on the requirements to safely conclude sufficient convergence in practice.
@@
 
-**[REMOVE?] $Q$s are the targets that will be estimated in each simulation repetition, after first *amputing* the complete data with varying missingness proportions, then imputing the missingness with a varying number of iterations, then estimating the $Q$s on the $m$ completed datasets, and finally pooling the $m$ estimates $\hat{Q}_\ell$ to obtain a single $\bar{Q}$ for each $Q$.
-We try to predict the effect of non-convergence (i.e., the difference between $Q$ and $\bar{Q}$) with the convergence diagnostics $\widehat{R}$ and $AC$, by applying these methods to several scalar summaries of the state space of the algorithm, $\theta$s.**
+<!-- **[REMOVE?] $Q$s are the targets that will be estimated in each simulation repetition, after first *amputing* the complete data with varying missingness proportions, then imputing the missingness with a varying number of iterations, then estimating the $Q$s on the $m$ completed datasets, and finally pooling the $m$ estimates $\hat{Q}_\ell$ to obtain a single $\bar{Q}$ for each $Q$.
+We try to predict the effect of non-convergence (i.e., the difference between $Q$ and $\bar{Q}$) with the convergence diagnostics $\widehat{R}$ and $AC$, by applying these methods to several scalar summaries of the state space of the algorithm, $\theta$s.** -->
 
@@
 
-In each simulation repetition, we *ampute* the complete dataset five times to obtain five different incomplete datasets, $y$s.
+The complete dataset is *amputed* in each simulation repetition to create five incomplete datasets $y$ with different missingness proportions $p_{\rm mis}$.
+The $p_{\rm mis}$ varies between 5 and 95%, in incremental steps ($p_{\rm mis}$ = .05, .25, .50, .75, .95).
+We consider all possible univariate and multivariate patterns of missingness, conform a 'missing completely at random' missingness mechanism [@rubin87].
+I.e., the probability to be missing is the same for all $N \times k$ cells in $y$, conditional on the missingness proportion $p_{\rm mis}$.
+This is performed with the 'mice' package [function `mice::ampute()`; @mice].
+
+The second set of simulation conditions is the number of iterations in the imputation algorithm.
+We vary the number of iterations between one and one hundred ($T = 1,2,.,100$).
+All imputation procedures are performed with Bayesian linear regression imputation, and five imputation chains ($m=5$).
+We perform this with the `mice()` function [@mice].
+The result is five sets of imputations per simulation condition, $y_{\rm imp, \ell}$.
+
+Subsequently, we obtain the completed data {$y_{\rm obs}, y_{\rm imp, \ell}$} by combining the imputed data with the observed data [function `mice::complete()`; @mice].
+Univariate quantities of scientific interest $Q$ are estimated directly from the completed data {$y_{\rm obs}, y_{\rm imp, \ell}$}.
+Estimates of multivariate $Q$s are obtained after employing multiple linear regression [function `stats::lm()`, @R].
+From the estimates in each imputation, $\hat{Q}_\ell$, we obtain the pooled estimate $\bar{Q}$.
+For univariate estimates we pool $\hat{Q}_\ell$ as average.
+Multivariate $\hat{Q}_\ell$ are pooled conform @vink14.
+
+<!-- In each simulation repetition, we *ampute* the complete dataset five times to obtain five different incomplete datasets, $y$s.
 The ratio between $y_{\rm obs}$ and $y_{\rm mis}$ in each $y$ depends on the missingness proportion: $p_{\rm mis} =.05,.25,.5,.75,.95$.
@@
 I.e., the probability to be missing is the same for all $N \times k$ cells in $y$, conditional on the missingness proportion $p_{\rm mis}$.
-<!-- This note only considers a 'missing completely at random' (MCAR) mechanism, where the probability of being missing is equal for all $n \times k$ cells in $y$ [@rubin87].
+-->
+<!-- <!-- This note only considers a 'missing completely at random' (MCAR) mechanism, where the probability of being missing is equal for all $n \times k$ cells in $y$ [@rubin87].
 The probability to be missing is the same for all $n \times k$ cells in $y$ [a 'missing completely at random' missingness mechanism; @rubin87].
@@
 
-Missing datapoints in each incomplete dataset $y$ are imputed with the `mice()` function [@mice].
+<!-- Missing datapoints in each incomplete dataset $y$ are imputed with the `mice()` function [@mice].
 All imputation procedures are performed with Bayesian linear regression imputation, and five imputation chains ($m=5$).
@@
 The number of iterations in each run of the algorithm varies between simulation conditions ($T=1,2,...,100$).
-<!-- Each repetition of the simulation thus starts from 1 complete dataset, then splits into 5 missingness conditions ($p_{\rm mis} =.05,.25,.5,.75,.95$), each of the amputed datasets is imputed 5 times ($m = 1, 2,..., 5$) for each iteration condition ($t = 1, 2,..., 100$).
+-->
+<!-- <!-- Each repetition of the simulation thus starts from 1 complete dataset, then splits into 5 missingness conditions ($p_{\rm mis} =.05,.25,.5,.75,.95$), each of the amputed datasets is imputed 5 times ($m = 1, 2,..., 5$) for each iteration condition ($t = 1, 2,..., 100$).
 From these $5\times5\times100$ imputations, we extract the $\theta$s to apply $\widehat{R}$ and $AC$ on (**i.e., chain means, chain variances, $\beta$s per imputation, and the first eigenvalue of the variance-covariance matrix per imputation**).
@@
 
-From the 5 imputations $\dot y_{\rm imp, \ell}$ we obtain the chain means and chain variances for each variable $j$.
+<!-- From the 5 imputations $\dot y_{\rm imp, \ell}$ we obtain the chain means and chain variances for each variable $j$.
 These will serve as scalar summary $\theta$ to apply non-convergence diagnostics $\widehat{R}$ and $AC$ on.
 We then use `mice::complete()` to combine the imputed data $\dot y_{\rm imp, \ell}$ with the observed data $y_{\rm obs}$, resulting in five completed datasets {$y_{\rm obs}, y_{imp, m}$}.
-
-On the 5 completed datasets {$y_{\rm obs}, y_{imp, m}$}, we:
-
-- Estimate $\hat{Q}_\ell$s for $Q= \mu$ and $Q=\sigma$ with `base::mean()`, and `stats::sd()` [@R]. We pool the $\hat{Q}_\ell$s with `base::mean()` to get $\bar{Q}$s.
-
-- Compute $\lambda_{1, \ell}$, to use as $\theta$ in the two convergence diagnostics.
+-->
+
+<!-- On the 5 completed datasets {$y_{\rm obs}, y_{imp, m}$}, we:  -->
+
+<!-- - Estimate $\hat{Q}_\ell$s for $Q= \mu$ and $Q=\sigma$ with `base::mean()`, and `stats::sd()` [@R]. We pool the $\hat{Q}_\ell$s with `base::mean()` to get $\bar{Q}$s. -->
+
+<!-- - Compute $\lambda_{1, \ell}$, to use as $\theta$ in the two convergence diagnostics.
 By definition, the first eigenvalue of the variance-covariance matrix, $\lambda_{1, \ell}$, is equal to the variance of the principal component solution of each {$y_{\rm obs}, y_{imp, m}$}, which we calculate with `stats::princomp()` [@R].
-
-- Perform multiple linear regression with `stats::lm()` [@R] to get the $\hat{Q}_\ell$s for $Q=\beta$ and $Q=r^2$.
+-->
+
+<!-- - Perform multiple linear regression with `stats::lm()` [@R] to get the $\hat{Q}_\ell$s for $Q=\beta$ and $Q=r^2$.
 We pool the $\hat{Q}_\ell$s conform @vink14 to get $\bar{Q}$s for the $\beta$s and $r^2$ **[check  finite pooling of  $r^2$, instead of `mice::pool.r.squared()`]**.
 Additionally, we use the $\hat{Q}_\ell$s of $Q=\beta$ as $\theta$ to apply the convergence diagnostics to.
-
-<!-- After each run of the `mice` algorithm, the function `mice::complete()` is used to combine the imputed data $\dot y_{\rm imp, \ell}$ with the observed data $y_{\rm obs}$, resulting in five completed datasets {$y_{\rm obs}, y_{imp, m}$}.
--->
-
-<!-- The estimator for each $Q$ is $\bar{Q}$---the pooled aggregate of the $\hat{Q}_\ell$s across imputations.
+-->
+
+<!-- <!-- After each run of the `mice` algorithm, the function `mice::complete()` is used to combine the imputed data $\dot y_{\rm imp, \ell}$ with the observed data $y_{\rm obs}$, resulting in five completed datasets {$y_{\rm obs}, y_{imp, m}$}.
+-->
+
+<!-- <!-- The estimator for each $Q$ is $\bar{Q}$---the pooled aggregate of the $\hat{Q}_\ell$s across imputations.
 To calculate the $\bar{Q}$s, we combine the observed data $y_{\rm obs}$ and the imputed data for each imputation $\dot y_{\rm imp, \ell}$ with the function `mice::complete()`.
@@
 
-We apply non-convergence diagnostics $\widehat{R}$ and $AC$ to each scalar summary $\theta$.
+<!-- We apply non-convergence diagnostics $\widehat{R}$ and $AC$ to each scalar summary $\theta$.
 We calculate $\widehat{R}$ by implementing Vehtari et al.'s [-@veht19] recommendations, and $AC$ as the correlation between the $t^{th}$ and the $(t+1)^{th}$ iteration.
+-->
+
+### Methods.
+
+In total, we consider eight methods to identify non-convergence.
+Namely, we apply the two non-convergence diagnostics $\widehat{R}$ and $AC$ on four different $\theta$s.
+Univariate $\theta$s are obtained from $y_{\rm imp, \ell}$.
+The multivariate $\theta$ $\lambda_{1,\ell}$ is obtained from {$y_{\rm obs}, y_{\rm imp, \ell}$}.
+The other multivariate $\theta$ is equal to the estimated regression coefficient $\hat{Q}_\ell$ in {$y_{\rm obs}, y_{\rm imp, \ell}$}.
+We apply the two non-convergence diagnostics to each $\theta$ by implementing @veht19 's recommended $\widehat{R}$ and we compute $AC$ manually.
 
@@
 
-We assess the performance of the iterative imputation algorithm in each simulation condition by comparing $\bar{Q}$s with $Q$s.
-For each $Q$, we calculate bias as $\bar{Q} - Q$.
-For the regression coefficients $Q=\beta_{1,2,3}$, we also compute the empirical coverage rate (CR).
-CR is defined as the percentage of simulation repetitions in which the 95\% confidence interval (CI) around $\bar{Q}$ covers the true estimand $Q$.
-Let
+As recommended by @buur18 we evaluate the performance of the imputation algorithm with bias, coverage rate (CR), and confidence interval width (CIW).
+Bias is calculated for each $Q$.
+CR and CIW only for $Q = \beta$.
+We calculate bias as $\bar{Q} - Q$. CR is defined as the percentage of simulation repetitions in which the 95\% confidence interval (CI) around $\bar{Q}$ covers the true estimand $Q$. Let
 $$\text{CI} = \bar{Q} \pm t_{(m-1)} \times \rm SE_{\bar{Q}},$$
@@
 Similar to the bias for $Q=\mu$ the magnitude of the bias in $Q=\sigma$ is negligible.
-We therefore conclude that neither of the univariate $Q$s is affected substantially by non-convergence.
+We therefore conclude that neither one of the univariate $Q$s is affected substantially by non-convergence.
 <!-- in the first iteration.
@@
 In conditions where $p_{\rm mis}=.05$, estimates are unbiased after one iteration.
-We observe approximately unbiased estimates when $T\geq 2,4, \text{and }5$, in conditions where $p_{\rm mis} = .25, .50, \text{and }.75$, respectively.
+We observe approximately unbiased estimates when $T\geq 2,4, \text{and }5$, in conditions where $p_{\rm mis} = .25, .50, \text{and }.75$, respectively **[only report the worst?]**.
 The highest number of iterations necessary to reach approximate unbiasedness is seven.
@@
 The $\widehat{R}$-values in these conditions do not taper off after 30 to 40 iterations, but rather between 40 to 50 iterations.
-The highest number of iterations at which non-convergence is still diagnosed according to the thresholds are: 11 for $\widehat{R}>1.2$, 23 for $\widehat{R}>1.1$, and 100 for $\widehat{R}>1.01$.
+The highest number of iterations at which non-convergence is still diagnosed according to the thresholds are: 11 for $\widehat{R}>1.2$, 23 for $\widehat{R}>1.1$, and 100 for $\widehat{R}>1.01$ **[only report the worst?]**.
 
@@
 
-# Discussion
+# Conclusion
 
@@
 
-# Conlusion
-
-We arrive back at the original questions that we posed.
-
-- Non-convergence can be identified with non-convergence diagnostics $\widehat{R}$ and $AC$.
+# Discussion
+
+We have shown that non-convergence in iteratiive imputation algorithms leads to biased, invalid estimates of multivariate quantities of scientific interest $\theta$.
+Since the current practice of visually inspecting univariate $\theta$s does not suffice, we applied non-convergence diagnostics $\widehat{R}$ and $AC$ to both univariate and multivariate $\theta$s.
+Signs of non-convergence due to early stopping are identified with each of the methods---i.e., by applying $\widehat{R}$ and $AC$ to any $\theta$.
+Bias due to high missingness proportions, is only identified with multivariate $\theta$s.
+Under moderate missingness ($p_{\rm mis}\leq.50$) $\widehat{R}$-values decreased substantially until 30 to 40 iterations, which implies that mixing in the algorithm improved with each additional iteration.
+$AC$-values only improved until about six iterations, suggesting minimal improvement in trending is obtained beyond $T=6$.
+Since we obtained approximately unbiased, confidence valid estimates after at most seven iterations, we conclude that the $\widehat{R}$ threshold 1.2 is the most appropriate diagnostic cut-off.
+The threshold to diagnose non-stationarity with $AC$ does not seem to apply.
+The main finding of this study is that valid inferences may be obtained much quicker than approximate algorithmic convergence is reached.
+Under the current specifications, univariate $Q$s did not require algorithmic convergence at all.
+They are unbiased almost instantly after the algorithm is initiated.
+Approximately unbiased, confidence-valid estimates of multivariate $Q$s were obtained after a maximum of seven iterations.
+Continued iterations beyond $T = 7$ did not yield better estimates.
+
+## Implications
+
+Iterative imputation algorithms such as MICE have been known to yield valid results under severe missingness.
+With this study, we have shown that valid inferences can also be obtained under a combination of severe missingness and early stopping.
+Based on our results, the traditional threshold of $\widehat{R}> 1.2$ would be appropriate for diagnosing non-convergence in iterative imputation algorithms.
+This is not in line with the recent recommendations by veht19 to lower the threshold to 1.01.
+The discrepancy between our conclusion and @veht19's may be explained by the nature of iterative imputation algorithms.
+In the limit, iterative imputation algorithms have the same character as other MCM algorithms, but in imputation procedures a part of the distribution is already determined by the observed data, whereas the entire distribution is unknown in many other MCMC applications.
+Because we combine a known distribution with an unknown distribution, valid estimates may be reached much sooner.
+Convergence may therefore be diagnosed at a less stringent threshold.
+**[HERE were the two asterixes]**
+
+The threshold to diagnose non-convergence with $AC$ does not seem appropriate at all in iterative imputation algorithms.
+A better diagnostic cut-off would be the number of iterations at which an additional iteration does not substantially decrease $AC$.
+In practice, this implies that the default number of iterations in MICE is not sufficient.
+$AC$ can only be computed if $T\geq3$, while the current default in the `mice` package is five [@mice].
+Identifying a decrease in the $AC$-values across iterations (an 'elbow') requires more than three observations.
+We therefore suggest to increase the default number of iterations in MICE to 10.
+The computational cost of 5 additional iterations has become less of a burden since MICE was introduced in [-@mice].
+Moreover, an increased number of iterations would grant the opportunity to exclude the first iterations from evaluation with $\widehat{R}$.
+Excluding $T\leq3$ provides empirical researchers with a less ambiguous heuristic to diagnose non-convergence, because we expect the initial 'dip' in $\widehat{R}$-values to disappear.
+
+## Recommendations for empirical researchers
+
+Based on our results, we formulated several recommendations for empirical researchers who employ iterative imputation to draw inferences from incomplete data.
+We suggest that non-convergence may still be evaluated visually, but in addition to inspecting univariate summaries of the state space of the algorithm ($\theta$s, e.g., chain means), multivariate $\theta$s should be considered.
+We propose the following steps:
+
+1. First, check traceplots of the default $\theta$s (chain means and chain variances) for signs of pathological non-convergence. Adjust the imputation model if necessary.
+
+2.
+Track multivariate $\theta$s over iterations, e.g., the novel $\theta$ that is  scientific model-independent.
+Or specify your own scalar summary of interest (see e.g., @buur18).
+Monitor these $\theta$s across iterations through visual inspection, or using a non-convergence diagnostic.
+
+3.
+Compute non-convergence diagnostics $\widehat{R}$ and $AC$.
+Do not use the traditional calculation for $\widehat{R}$ [@gelm92], or the `R` function `stats::acf()` to compute $AC$ [@R].
+Instead, calculate $\widehat{R}$ conform @veht19 and compute autocorrelations manually (see e.g., [github.com/gerkovink/shinyMice](https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy)).
+
+4.
+Use the threshold $\widehat{R}>1.2$ to diagnose non-mixing, and assess stationarity by plotting the autocorrelation over iterations.
+Keep iterating until the threshold for mixing is overcome, and until reasonably decreasing asympototic $AC$-values are obtained.
+At that point, inferences are unlikely to improve by continued iteration.
+
+5. Profit.
+
+## Limitations
+
+Much remains unknown about non-convergence in iterative imputation algorithms.
+Even though we have demonstrated the appropriateness of $\widehat{R}$ and $AC$ as non-convergence diagnostics in this study, results may not extrapolate to other situations.
+The current simulation conditions were restricted to a single missingness mechanism.
+Proper performance under a 'missing completely at random' (MCAR) mechanism is a necessary condition for any missing data method.
+It doesn't, however, guarantee equal performance under different missingness mechanisms.
+Future research should determine the performance of $\widehat{R}$ and $AC$ under 'missing at random' and 'missing not at random' mechanisms.
+Another parameter to consider in future research is the choice of the imputation method.
+As @murr18 concluded, the behaviour of algorithms such as MICE under certain default imputation models is still an open research question.
+We have investigated the behaviour of MICE under only one type of model (Bayesian linear regression).
+Different imputations models might converge more poorly.
+Moreover, the imputation model may even be mis-specified un purpose to induce different levels of non-convergence.
+Qualitatively, we have shown that $\widehat{R}$ and $AC$ are appropriate under clear violation of convergence.
+Quantitatively, we have only demonstrated that they can identify signs of non-convergence under MCAR, when the imputation model is a correctly specified Bayesian linear regression model.
+\newline \newline
+
+<!-- # Old discussion -->
+
+<!-- We arrive back at the original questions that we posed. -->
+
+<!-- - Non-convergence can be identified with non-convergence diagnostics $\widehat{R}$ and $AC$.
 But common thresholds are too conservative.
@@
 In iterative imputation algorithms we have the same characteristics as MCMC in the limit, but because we combine the known distribution with the unknown distribution, convergence is reached much sooner.
-
-- Inferences are unbiased much quicker than the algorithm has converged.
+-->
+
+<!-- - Inferences are unbiased much quicker than the algorithm has converged.
 All things univariate are converged almost intantly after initiating the iterative algorithm with a valid imputation model.
 But less quick with multivariate parameters.
-
-In short, iterative imputation algorithms may yield approximately unbiased, confidence-valid estimates under different levels of induced non-convergence.
+-->
+
+<!-- In short, iterative imputation algorithms may yield approximately unbiased, confidence-valid estimates under different levels of induced non-convergence.
 Univariate estimates seem robust against terminating the algorithm early: There is no clear effect of the number of iterations on the bias in estimated means and standard deviations.
@@
 **[MERGE or remove] $\widehat{R}$ and autocorrelation indicate that there are signs of non-convergence in the algorithm up-to at least twenty iterations, while unbiased, confidence-valid estimates may be obtained with as little as one iteration.
-This observation is in agreement with **one of [still add the others!]** the simulation hypotheses: $\widehat{R}$ over-estimates the severity of non-convergence when applied to MI procedures.**
-
-Over-estimation of the signs of non-convergence may be due to the methods (and their thresholds) or due to the $Q$s (descriptive and multivariate linear regression, not higher dimensional/more complex $Q$s).
+This observation is in agreement with **one of [still add the others!]** the simulation hypotheses: $\widehat{R}$ over-estimates the severity of non-convergence when applied to MI procedures.**  -->
+
+<!-- Over-estimation of the signs of non-convergence may be due to the methods (and their thresholds) or due to the $Q$s (descriptive and multivariate linear regression, not higher dimensional/more complex $Q$s).
 More complicated $Q$s (e.g., higher-order effects or variance components) might show bias, under- or over-coverage at higher $T$.
 Which might be why $\widehat{R}$ and $AC$ identify non-convergence.
-
-```{r eval=FALSE, include=FALSE}
-# **[INCLUDE this] The highest $\widehat{R}$-values are obtained for the condition $T=2$.
+-->
+
+<!-- ```{r eval=FALSE, include=FALSE} -->
+<!-- # **[INCLUDE this] The highest $\widehat{R}$-values are obtained for the condition $T=2$.
 We observe the same 'dip' in $\widehat{R}$ as we did in Figure \ref{fig:diagnostics} for conditions with three iterations.
@@
 The common threshold $\widehat{R}>1.1$ is overcome in conditions where $T>14$.
-**
-#   
-#   **Autocorrelations indicate some signs of non-stationarity in all simulation conditions, but primarily in conditions where $T<6$.
+** -->
+<!-- #    -->
+<!-- #   **Autocorrelations indicate some signs of non-stationarity in all simulation conditions, but primarily in conditions where $T<6$.
 The threshold for diagnosing non-convergence is not reached in any condition.
 We would thus conclude that the algorithm is sufficiently converged regardless of the number of iterations or missingness proportion.
-**
-```
-
-Some other interesting findings are **[REPHRASE]**
-
-- Is there more bias in conditions where non-convergence was induced? (YES for multivar., NOPE for univariates).
-
-- Is the point of *sufficient* convergence (at which the $\bar{Q}$s are unbiased and confidence-valid estimates of the $Q$s) the same point as identified by the non-convergence diagnostics?
+** -->
+<!-- ``` -->
+
+<!-- Some other interesting findings are **[REPHRASE]** -->
+
+<!-- - Is there more bias in conditions where non-convergence was induced? (YES for multivar., NOPE for univariates). -->
+
+<!-- - Is the point of *sufficient* convergence (at which the $\bar{Q}$s are unbiased and confidence-valid estimates of the $Q$s) the same point as identified by the non-convergence diagnostics?
 (NOPE, they are too conservative).
-
-- Is one of the thetas better? (YES, while it doesn't really matter as long as it's multivariate, $\lambda_{1, \ell}$ is model-independent and therefore preferred).
-
-- What should the thresholds be? (much HIGHER than recommended for $\widehat{R}$, difficult to say for $AC$).
-
-- And the default nr of iterations? (MAYBE increase to 10 to detect significant $AC$ and be sure not to miss the dip in $\widehat{R}$?).
+-->
+
+<!-- - Is one of the thetas better? (YES, while it doesn't really matter as long as it's multivariate, $\lambda_{1, \ell}$ is model-independent and therefore preferred). -->
+
+<!-- - What should the thresholds be? (much HIGHER than recommended for $\widehat{R}$, difficult to say for $AC$).  -->
+
+<!-- - And the default nr of iterations? (MAYBE increase to 10 to detect significant $AC$ and be sure not to miss the dip in $\widehat{R}$?). -->
 
@@
 
-## Recommendations for empirical researchers
-
-After imputation it is always wise to take the following steps:
-
-1. First, check traceplots for signs of pathological non-convergence. Adjust imputation model if necessary.
-
-2.
-Monitor non-convergence diagnostics $\widehat{R}$ and $AC$.
-Do not use the traditional calculation for $\widehat{R}$, or the `R` function `stats::acf()` to compute $AC$.
-Instead, calculate $\widehat{R}$ conform @veht19 and compute autocorrelations manually (see e.g., [github.com/gerkovink/shinyMice](https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy)).
-
-3. Track $\theta$s over iterations, e.g., the novel $\theta$ that is 'substantive model-independent'.
-
-4. Track your own scalar summary of interest (see e.g., @buur18). Compute $\widehat{R}$ and autocorrelation values for this scalar summary.
-
-5.
-Use the threshold $\widehat{R}>1.2$ for mixing, and assess stationarity by plotting the autocorrelation over iterations.
-Keep iterating until the threshold for mixing is reached, and until a reasonably decreasing asympototic $AC$ value is obtained.
-At that point, inferences are unlikely to improve by continued iteration.
-
-## Recommendations for future research
-
-Much remains unknown.
+<!-- ## Recommendations for future research -->
+
+<!-- Much remains unknown.
 about non-convergence in iterative imputation algorithms.
@@
 dependency between predictors (predictors with very high correlations).
-For example:
-
-- Induce non-convergence with mis-specified imputation model.
-
-- Introduce more difficult missingness problems, i.e., M(N)AR instead of MCAR.
+For example: -->
+
+<!-- - Induce non-convergence with mis-specified imputation model.  -->
+
+<!-- - Introduce more difficult missingness problems, i.e., M(N)AR instead of MCAR.
 That is, proper performance under a 'missing completely at random' missingness mechanism is necessary to demonstrate the appropriateness of $\widehat{R}$ and $AC$ as non-convergence diagnostics.
 However, results may not be extrapolated to other missingness mechanisms.
-
-- Imputation technique with questionable convergence properties, e.g., PMM.
-
-- Apply recommendations to empirical data and write a vignette for applied researchers.
-
-- Implement recommendations in imputation software, e.g., evaluation environment for applied researchers.
-
-- Develop an additional convergence diagnostic unique for iterative imputation: $AC$ across imputations, instead of iterations.
-
-In short, we have shown that an iterative imputation algorithm can yield correct outcomes, even when a converged state has not yet formally been reached.
+-->
+
+<!-- - Imputation technique with questionable convergence properties, e.g., PMM. -->
+
+<!-- - Apply recommendations to empirical data and write a vignette for applied researchers. -->
+
+<!-- - Implement recommendations in imputation software, e.g., evaluation environment for applied researchers. -->
+
+<!-- - Develop an additional convergence diagnostic unique for iterative imputation: $AC$ across imputations, instead of iterations. -->
+
+\noindent In short, we have shown that an iterative imputation algorithm can yield correct outcomes, even when a converged state has not yet formally been reached.
 Any further iterations would then burn computational resources without improving the statistical inferences.

### 2020-05-09 3.Thesis/Writeup/2.Manuscript.tex
Last feedback before hand-in

@@
 Moreover, the behavoir of such algorithms under certain default imputation models (e.g., `predictive mean matching') is an entirely open question \citep{murr18}.
-\textbf{Therefore, algorithmic convergence should be monitored carefully, }although not straightforward.
+Therefore, algorithmic convergence should be monitored carefully, \textbf{but this is not straightforward}.
 Iterative imputation algorithms such as MICE are special cases of Markov chain Monte Carlo (MCMC) methods.
@@
 \textbf{ADD that these guidelines may aid to draw valid inference from incomplete data?{]}} For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular \texttt{mice} package \citep{mice} in \texttt{R} \citep{R}.
-
-\textbf{{[}Include sub-questions and make structure of the paper clear!{]}} How can non-convergence be identified diagnostically?
-Are common MCMC non-convergence diagnostics appropriate for MICE?
-And if so, which threshold should be used to diagnose non-convergence?
-How many iterations are sufficient/needed to be able to diagnose non-convergence?
-Are the default number of iterations sufficient (i.e., 5 in mice, 10 in SPSS and Stata, 30 in mi)?
-How severe is it when the algorithm has not converged?
-And what are guidelines for practice?
-Can the parameter of interest \(Q\) be correct when the algorithm is not (yet) converged, and vice versa?
 
@@
 The \(\theta\) in this example is the chain mean of a variable that we will call \(j\) (i.e., the average of the imputed values for \(j\) in each imputation \(\dot y_{\rm imp, \ell}\)).
-\textbf{In the typical convergence scenario, the imputation chains intermingle nicely and there is little to no trending.
-In the non-convergence scenario, there is a lot of trending and some chains do not intermingle.} Importantly, the chain means at the last iteration (the imputed value per imputation \(\ell\)) are very different between the two plots.
+In the typical convergence scenario, the imputation chains intermingle nicely and there is little to no trending.
+In the non-convergence scenario, there is a lot of trending and some chains do not intermingle.
+Importantly, the chain means at the last iteration (the imputed value per imputation \(\ell\)) are very different between the two plots.
 The algorithm with the mis-specified model yields imputed values that are on average a factor two larger than those of the typically converged algorithm.
@@
 \(\lambda_{1, \ell}\) is measure that summarizes the covariances in the completed datasets.
-\textbf{The first eigenvalue has the appealing property that is not dependent on the substantive model of interest.
-Eigenvalue decomposition is inspired by \citet{mack03}.}
+The first eigenvalue has the appealing property that is not dependent on the substantive model of interest.
+Eigenvalue decomposition is inspired by \citet{mack03}.\textbf{{[}ADD definition variance in first PCA component is equal to the first eigenvalue of the variance-covariance matrix{]}}
 
@@
 Specifically, we are interested in the bias and confidence-validity of estimates obtained from incomplete data with the MICE algortihm.
-
-\textbf{{[}REMOVE?{]} Several quantities of scientific interest \(Q\).
-Subsequently we will evaluate how well \(\widehat{R}\) and \(AC\) perform in identifying the effects of non-convergence }{[}and performance of different thetas{]}\textbf{.
+\textbf{{[}REMOVE?{]} Subsequently we will evaluate how well \(\widehat{R}\) and \(AC\) perform in identifying the effects of non-convergence }{[}and performance of different thetas{]}\textbf{.
 And finally, we will formulate an informed advice on the requirements to safely conclude sufficient convergence in practice.
@@
 We assume that we can induce non-convergence at least in conditions where \(T\) is smaller than the default number of iterations in \texttt{mice}, \(T=5\).
-
-\textbf{{[}REMOVE?{]} \(Q\)s are the targets that will be estimated in each simulation repetition, after first \emph{amputing} the complete data with varying missingness proportions, then imputing the missingness with a varying number of iterations, then estimating the \(Q\)s on the \(m\) completed datasets, and finally pooling the \(m\) estimates \(\hat{Q}_\ell\) to obtain a single \(\bar{Q}\) for each \(Q\).
-We try to predict the effect of non-convergence (i.e., the difference between \(Q\) and \(\bar{Q}\)) with the convergence diagnostics \(\widehat{R}\) and \(AC\), by applying these methods to several scalar summaries of the state space of the algorithm, \(\theta\)s.}
 
@@
 
-In each simulation repetition, we \emph{ampute} the complete dataset five times to obtain five different incomplete datasets, \(y\)s.
-The ratio between \(y_{\rm obs}\) and \(y_{\rm mis}\) in each \(y\) depends on the missingness proportion: \(p_{\rm mis} =.05,.25,.5,.75,.95\).
-We ampute the complete data with the \texttt{mice} package \citep[function \texttt{mice::ampute()};][]{mice}.
+The complete dataset is \emph{amputed} in each simulation repetition to create five incomplete datasets \(y\) with different missingness proportions \(p_{\rm mis}\).
+The \(p_{\rm mis}\) varies between 5 and 95\%, in incremental steps (\(p_{\rm mis}\) = .05, .25, .50, .75, .95).
 We consider all possible univariate and multivariate patterns of missingness, conform a `missing completely at random' missingness mechanism \citep{rubin87}.
 I.e., the probability to be missing is the same for all \(N \times k\) cells in \(y\), conditional on the missingness proportion \(p_{\rm mis}\).
+This is performed with the `mice' package \citep[function \texttt{mice::ampute()};][]{mice}.
 
-Missing datapoints in each incomplete dataset \(y\) are imputed with the \texttt{mice()} function \citep{mice}.
+The second set of simulation conditions is the number of iterations in the imputation algorithm.
+We vary the number of iterations between one and one hundred (\(T = 1,2,.,100\)).
 All imputation procedures are performed with Bayesian linear regression imputation, and five imputation chains (\(m=5\)).
-Because \(m=5\), each run of the \texttt{mice()} function results in five sets of imputations: \(\dot y_{\rm imp, \ell}\), where \(m = 1, 2, ..., 5\).
-The number of iterations in each run of the algorithm varies between simulation conditions (\(T=1,2,...,100\)).
+We perform this with the \texttt{mice()} function \citep{mice}.
+The result is five sets of imputations per simulation condition, \(y_{\rm imp, \ell}\).
 
-From the 5 imputations \(\dot y_{\rm imp, \ell}\) we obtain the chain means and chain variances for each variable \(j\).
-These will serve as scalar summary \(\theta\) to apply non-convergence diagnostics \(\widehat{R}\) and \(AC\) on.
-We then use \texttt{mice::complete()} to combine the imputed data \(\dot y_{\rm imp, \ell}\) with the observed data \(y_{\rm obs}\), resulting in five completed datasets \{\(y_{\rm obs}, y_{imp, m}\)\}.
+Subsequently, we obtain the completed data \{\(y_{\rm obs}, y_{\rm imp, \ell}\)\} by combining the imputed data with the observed data \citep[function \texttt{mice::complete()};][]{mice}.
+Univariate quantities of scientific interest \(Q\) are estimated directly from the completed data \{\(y_{\rm obs}, y_{\rm imp, \ell}\)\}.
+Estimates of multivariate \(Q\)s are obtained after employing multiple linear regression \citep[function \texttt{stats::lm()},][]{R}.
+From the estimates in each imputation, \(\hat{Q}_\ell\), we obtain the pooled estimate \(\bar{Q}\).
+For univariate estimates we pool \(\hat{Q}_\ell\) as average.
+Multivariate \(\hat{Q}_\ell\) are pooled conform \citet{vink14}.
 
-On the 5 completed datasets \{\(y_{\rm obs}, y_{imp, m}\)\}, we:
+\hypertarget{methods.}{%
+\subsubsection{Methods.}\label{methods.}}
 
-\begin{itemize}
-\item
-  Estimate \(\hat{Q}_\ell\)s for \(Q= \mu\) and \(Q=\sigma\) with \texttt{base::mean()}, and \texttt{stats::sd()} \citep{R}.
-We pool the \(\hat{Q}_\ell\)s with \texttt{base::mean()} to get \(\bar{Q}\)s.
-\item
-  Compute \(\lambda_{1, \ell}\), to use as \(\theta\) in the two convergence diagnostics.
-By definition, the first eigenvalue of the variance-covariance matrix, \(\lambda_{1, \ell}\), is equal to the variance of the principal component solution of each \{\(y_{\rm obs}, y_{imp, m}\)\}, which we calculate with \texttt{stats::princomp()} \citep{R}.
-\item
-  Perform multiple linear regression with \texttt{stats::lm()} \citep{R} to get the \(\hat{Q}_\ell\)s for \(Q=\beta\) and \(Q=r^2\).
-We pool the \(\hat{Q}_\ell\)s conform \citet{vink14} to get \(\bar{Q}\)s for the \(\beta\)s and \(r^2\) \textbf{{[}check finite pooling of \(r^2\), instead of \texttt{mice::pool.r.squared()}{]}}.
-Additionally, we use the \(\hat{Q}_\ell\)s of \(Q=\beta\) as \(\theta\) to apply the convergence diagnostics to.
-\end{itemize}
-
-We apply non-convergence diagnostics \(\widehat{R}\) and \(AC\) to each scalar summary \(\theta\).
-We calculate \(\widehat{R}\) by implementing Vehtari et al.'s \citeyearpar{veht19} recommendations, and \(AC\) as the correlation between the \(t^{th}\) and the \((t+1)^{th}\) iteration.
+In total, we consider eight methods to identify non-convergence.
+Namely, we apply the two non-convergence diagnostics \(\widehat{R}\) and \(AC\) on four different \(\theta\)s.
+Univariate \(\theta\)s are obtained from \(y_{\rm imp, \ell}\).
+The multivariate \(\theta\) \(\lambda_{1,\ell}\) is obtained from \{\(y_{\rm obs}, y_{\rm imp, \ell}\)\}.
+The other multivariate \(\theta\) is equal to the estimated regression coefficient \(\hat{Q}_\ell\) in \{\(y_{\rm obs}, y_{\rm imp, \ell}\)\}.
+We apply the two non-convergence diagnostics to each \(\theta\) by implementing \citet{veht19} 's recommended \(\widehat{R}\) and we compute \(AC\) manually.
 
@@
 
-We assess the performance of the iterative imputation algorithm in each simulation condition by comparing \(\bar{Q}\)s with \(Q\)s.
-For each \(Q\), we calculate bias as \(\bar{Q} - Q\).
-For the regression coefficients \(Q=\beta_{1,2,3}\), we also compute the empirical coverage rate (CR).
-CR is defined as the percentage of simulation repetitions in which the 95\% confidence interval (CI) around \(\bar{Q}\) covers the true estimand \(Q\).
-Let
+As recommended by \citet{buur18} we evaluate the performance of the imputation algorithm with bias, coverage rate (CR), and confidence interval width (CIW).
+Bias is calculated for each \(Q\).
+CR and CIW only for \(Q = \beta\).
+We calculate bias as \(\bar{Q} - Q\). CR is defined as the percentage of simulation repetitions in which the 95\% confidence interval (CI) around \(\bar{Q}\) covers the true estimand \(Q\). Let
 \[\text{CI} = \bar{Q} \pm t_{(m-1)} \times \rm SE_{\bar{Q}},\]
@@
 Similar to the bias for \(Q=\mu\) the magnitude of the bias in \(Q=\sigma\) is negligible.
-We therefore conclude that neither of the univariate \(Q\)s is affected substantially by non-convergence.
+We therefore conclude that neither one of the univariate \(Q\)s is affected substantially by non-convergence.
 
@@
 In conditions where \(p_{\rm mis}=.05\), estimates are unbiased after one iteration.
-We observe approximately unbiased estimates when \(T\geq 2,4, \text{and }5\), in conditions where \(p_{\rm mis} = .25, .50, \text{and }.75\), respectively.
+We observe approximately unbiased estimates when \(T\geq 2,4, \text{and }5\), in conditions where \(p_{\rm mis} = .25, .50, \text{and }.75\), respectively \textbf{{[}only report the worst?{]}}.
 The highest number of iterations necessary to reach approximate unbiasedness is seven.
@@
 The \(\widehat{R}\)-values in these conditions do not taper off after 30 to 40 iterations, but rather between 40 to 50 iterations.
-The highest number of iterations at which non-convergence is still diagnosed according to the thresholds are: 11 for \(\widehat{R}>1.2\), 23 for \(\widehat{R}>1.1\), and 100 for \(\widehat{R}>1.01\).
+The highest number of iterations at which non-convergence is still diagnosed according to the thresholds are: 11 for \(\widehat{R}>1.2\), 23 for \(\widehat{R}>1.1\), and 100 for \(\widehat{R}>1.01\) \textbf{{[}only report the worst?{]}}.
 
@@
 
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
+\hypertarget{conclusion}{%
+\section{Conclusion}\label{conclusion}}
 
@@
 
-\hypertarget{conlusion}{%
-\section{Conlusion}\label{conlusion}}
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
 
-We arrive back at the original questions that we posed.
+We have shown that non-convergence in iteratiive imputation algorithms leads to biased, invalid estimates of multivariate quantities of scientific interest \(\theta\).
+Since the current practice of visually inspecting univariate \(\theta\)s does not suffice, we applied non-convergence diagnostics \(\widehat{R}\) and \(AC\) to both univariate and multivariate \(\theta\)s.
+Signs of non-convergence due to early stopping are identified with each of the methods---i.e., by applying \(\widehat{R}\) and \(AC\) to any \(\theta\).
+Bias due to high missingness proportions, is only identified with multivariate \(\theta\)s.
+Under moderate missingness (\(p_{\rm mis}\leq.50\)) \(\widehat{R}\)-values decreased substantially until 30 to 40 iterations, which implies that mixing in the algorithm improved with each additional iteration.
+\(AC\)-values only improved until about six iterations, suggesting minimal improvement in trending is obtained beyond \(T=6\).
+Since we obtained approximately unbiased, confidence valid estimates after at most seven iterations, we conclude that the \(\widehat{R}\) threshold 1.2 is the most appropriate diagnostic cut-off.
+The threshold to diagnose non-stationarity with \(AC\) does not seem to apply.
+The main finding of this study is that valid inferences may be obtained much quicker than approximate algorithmic convergence is reached.
+Under the current specifications, univariate \(Q\)s did not require algorithmic convergence at all.
+They are unbiased almost instantly after the algorithm is initiated.
+Approximately unbiased, confidence-valid estimates of multivariate \(Q\)s were obtained after a maximum of seven iterations.
+Continued iterations beyond \(T = 7\) did not yield better estimates.
 
-\begin{itemize}
-\item
-  Non-convergence can be identified with non-convergence diagnostics \(\widehat{R}\) and \(AC\).
-But common thresholds are too conservative.
-Why?
-Part of the distribution is already determined by the observed data.
-in regular MCMC the entire distribution is unknown (\textbf{priors??}).
-In iterative imputation algorithms we have the same characteristics as MCMC in the limit, but because we combine the known distribution with the unknown distribution, convergence is reached much sooner.
-\item
-  Inferences are unbiased much quicker than the algorithm has converged.
-All things univariate are converged almost intantly after initiating the iterative algorithm with a valid imputation model.
-But less quick with multivariate parameters.
-\end{itemize}
+\hypertarget{implications}{%
+\subsection{Implications}\label{implications}}
 
-In short, iterative imputation algorithms may yield approximately unbiased, confidence-valid estimates under different levels of induced non-convergence.
-Univariate estimates seem robust against terminating the algorithm early: There is no clear effect of the number of iterations on the bias in estimated means and standard deviations.
-The bias in these \(Q\)s was only impacted by the missingness proporion \(p_{\rm mis}\), with higher \(p_{\rm mis}\) resulting in more bias.
-Bias in multivariate estimates depended on both \(p_{\rm mis}\) and the number of iterations.
-However, the only effect of the number of iterations was observed when the algorithm was terminated at seven iterations or earlier.
-This suggests that continued iteration beyond seven iterations does not yield better estimates.
-Yet, the non-convergence diagnostics \(\widehat{R}\) and \(AC\) indicated that the iterative imputation algorithm did not reach a stable state before twenty to thirty iterations.
-\textbf{{[}MERGE or remove{]} \(\widehat{R}\) and autocorrelation indicate that there are signs of non-convergence in the algorithm up-to at least twenty iterations, while unbiased, confidence-valid estimates may be obtained with as little as one iteration.
-This observation is in agreement with }one of {[}still add the others!{]}** the simulation hypotheses: \(\widehat{R}\) over-estimates the severity of non-convergence when applied to MI procedures.**
+Iterative imputation algorithms such as MICE have been known to yield valid results under severe missingness.
+With this study, we have shown that valid inferences can also be obtained under a combination of severe missingness and early stopping.
+Based on our results, the traditional threshold of \(\widehat{R}> 1.2\) would be appropriate for diagnosing non-convergence in iterative imputation algorithms.
+This is not in line with the recent recommendations by veht19 to lower the threshold to 1.01.
+The discrepancy between our conclusion and \citet{veht19}'s may be explained by the nature of iterative imputation algorithms.
+In the limit, iterative imputation algorithms have the same character as other MCM algorithms, but in imputation procedures a part of the distribution is already determined by the observed data, whereas the entire distribution is unknown in many other MCMC applications.
+Because we combine a known distribution with an unknown distribution, valid estimates may be reached much sooner.
+Convergence may therefore be diagnosed at a less stringent threshold.
+\textbf{{[}HERE were the two asterixes{]}}
 
-Over-estimation of the signs of non-convergence may be due to the methods (and their thresholds) or due to the \(Q\)s (descriptive and multivariate linear regression, not higher dimensional/more complex \(Q\)s).
-More complicated \(Q\)s (e.g., higher-order effects or variance components) might show bias, under- or over-coverage at higher \(T\).
-Which might be why \(\widehat{R}\) and \(AC\) identify non-convergence.
-
-Some other interesting findings are \textbf{{[}REPHRASE{]}}
-
-\begin{itemize}
-\item
-  Is there more bias in conditions where non-convergence was induced? (YES for multivar., NOPE for univariates).
-\item
-  Is the point of \emph{sufficient} convergence (at which the \(\bar{Q}\)s are unbiased and confidence-valid estimates of the \(Q\)s) the same point as identified by the non-convergence diagnostics?
-(NOPE, they are too conservative).
-\item
-  Is one of the thetas better? (YES, while it doesn't really matter as long as it's multivariate, \(\lambda_{1, \ell}\) is model-independent and therefore preferred).
-\item
-  What should the thresholds be? (much HIGHER than recommended for \(\widehat{R}\), difficult to say for \(AC\)).
-\item
-  And the default nr of iterations? (MAYBE increase to 10 to detect significant \(AC\) and be sure not to miss the dip in \(\widehat{R}\)?).
-\end{itemize}
+The threshold to diagnose non-convergence with \(AC\) does not seem appropriate at all in iterative imputation algorithms.
+A better diagnostic cut-off would be the number of iterations at which an additional iteration does not substantially decrease \(AC\).
+In practice, this implies that the default number of iterations in MICE is not sufficient.
+\(AC\) can only be computed if \(T\geq3\), while the current default in the \texttt{mice} package is five \citep{mice}.
+Identifying a decrease in the \(AC\)-values across iterations (an `elbow') requires more than three observations.
+We therefore suggest to increase the default number of iterations in MICE to 10.
+The computational cost of 5 additional iterations has become less of a burden since MICE was introduced in \citeyearpar{mice}.
+Moreover, an increased number of iterations would grant the opportunity to exclude the first iterations from evaluation with \(\widehat{R}\).
+Excluding \(T\leq3\) provides empirical researchers with a less ambiguous heuristic to diagnose non-convergence, because we expect the initial `dip' in \(\widehat{R}\)-values to disappear.
 
@@
 
-After imputation it is always wise to take the following steps:
+Based on our results, we formulated several recommendations for empirical researchers who employ iterative imputation to draw inferences from incomplete data.
+We suggest that non-convergence may still be evaluated visually, but in addition to inspecting univariate summaries of the state space of the algorithm (\(\theta\)s, e.g., chain means), multivariate \(\theta\)s should be considered.
+We propose the following steps:
 
@@
 \item
-  First, check traceplots for signs of pathological non-convergence. Adjust imputation model if necessary.
+  First, check traceplots of the default \(\theta\)s (chain means and chain variances) for signs of pathological non-convergence. Adjust the imputation model if necessary.
 \item
-  Monitor non-convergence diagnostics \(\widehat{R}\) and \(AC\).
-Do not use the traditional calculation for \(\widehat{R}\), or the \texttt{R} function \texttt{stats::acf()} to compute \(AC\).
+  Track multivariate \(\theta\)s over iterations, e.g., the novel \(\theta\) that is scientific model-independent.
+Or specify your own scalar summary of interest (see e.g., \citet{buur18}).
+Monitor these \(\theta\)s across iterations through visual inspection, or using a non-convergence diagnostic.
+\item
+  Compute non-convergence diagnostics \(\widehat{R}\) and \(AC\).
+Do not use the traditional calculation for \(\widehat{R}\) \citep{gelm92}, or the \texttt{R} function \texttt{stats::acf()} to compute \(AC\) \citep{R}.
 Instead, calculate \(\widehat{R}\) conform \citet{veht19} and compute autocorrelations manually (see e.g., \href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy}{github.com/gerkovink/shinyMice}).
 \item
-  Track \(\theta\)s over iterations, e.g., the novel \(\theta\) that is `substantive model-independent'.
+  Use the threshold \(\widehat{R}>1.2\) to diagnose non-mixing, and assess stationarity by plotting the autocorrelation over iterations.
+Keep iterating until the threshold for mixing is overcome, and until reasonably decreasing asympototic \(AC\)-values are obtained.
+At that point, inferences are unlikely to improve by continued iteration.
 \item
-  Track your own scalar summary of interest (see e.g., \citet{buur18}). Compute \(\widehat{R}\) and autocorrelation values for this scalar summary.
-\item
-  Use the threshold \(\widehat{R}>1.2\) for mixing, and assess stationarity by plotting the autocorrelation over iterations.
-Keep iterating until the threshold for mixing is reached, and until a reasonably decreasing asympototic \(AC\) value is obtained.
-At that point, inferences are unlikely to improve by continued iteration.
+  Profit.
 \end{enumerate}
 
-\hypertarget{recommendations-for-future-research}{%
-\subsection{Recommendations for future research}\label{recommendations-for-future-research}}
+\hypertarget{limitations}{%
+\subsection{Limitations}\label{limitations}}
 
-Much remains unknown.
-about non-convergence in iterative imputation algorithms.
-Further research is needed to investigate the performance of iterative imputation algorithms under clear violation of convergence, e.g.~dependency between predictors (predictors with very high correlations).
-For example:
+Much remains unknown about non-convergence in iterative imputation algorithms.
+Even though we have demonstrated the appropriateness of \(\widehat{R}\) and \(AC\) as non-convergence diagnostics in this study, results may not extrapolate to other situations.
+The current simulation conditions were restricted to a single missingness mechanism.
+Proper performance under a `missing completely at random' (MCAR) mechanism is a necessary condition for any missing data method.
+It doesn't, however, guarantee equal performance under different missingness mechanisms.
+Future research should determine the performance of \(\widehat{R}\) and \(AC\) under `missing at random' and `missing not at random' mechanisms.
+Another parameter to consider in future research is the choice of the imputation method.
+As \citet{murr18} concluded, the behaviour of algorithms such as MICE under certain default imputation models is still an open research question.
+We have investigated the behaviour of MICE under only one type of model (Bayesian linear regression).
+Different imputations models might converge more poorly.
+Moreover, the imputation model may even be mis-specified un purpose to induce different levels of non-convergence.
+Qualitatively, we have shown that \(\widehat{R}\) and \(AC\) are appropriate under clear violation of convergence.
+Quantitatively, we have only demonstrated that they can identify signs of non-convergence under MCAR, when the imputation model is a correctly specified Bayesian linear regression model.
+\newline \newline
 
-\begin{itemize}
-\item
-  Induce non-convergence with mis-specified imputation model.
-\item
-  Introduce more difficult missingness problems, i.e., M(N)AR instead of MCAR.
-That is, proper performance under a `missing completely at random' missingness mechanism is necessary to demonstrate the appropriateness of \(\widehat{R}\) and \(AC\) as non-convergence diagnostics.
-However, results may not be extrapolated to other missingness mechanisms.
-\item
-  Imputation technique with questionable convergence properties, e.g., PMM.
-\item
-  Apply recommendations to empirical data and write a vignette for applied researchers.
-\item
-  Implement recommendations in imputation software, e.g., evaluation environment for applied researchers.
-\item
-  Develop an additional convergence diagnostic unique for iterative imputation: \(AC\) across imputations, instead of iterations.
-\end{itemize}
-
-In short, we have shown that an iterative imputation algorithm can yield correct outcomes, even when a converged state has not yet formally been reached.
+\noindent In short, we have shown that an iterative imputation algorithm can yield correct outcomes, even when a converged state has not yet formally been reached.
 Any further iterations would then burn computational resources without improving the statistical inferences.

### 2020-05-10 3.Thesis/Writeup/2.Manuscript.Rmd
Until diagnostics

@@
 There is no consensus on how to evaluate the convergence properties of the method.
-This paper provides insight into identifying non-convergence of iterative imputation algorithms.
+This paper provides insight into identifying non-convergence in iterative imputation algorithms.
 We conclude that in the case of drawing inference from incomplete data, convergence of the iterative imputation algorithm is often a convenience but not a necessity."
-keywords: missing data, iterative imputation, non-convergence, mice
+keywords: missing data, iterative imputation, non-convergence, MICE
 classoption:
@@
 This technique is known as 'multiple imputation' [MI; @rubin76].
-MI has proven to be a powerful technique to draw valid inferences from incomplete data under many circumstances [@buur18].
+MI has proven to be a powerful tool to draw valid inferences from incomplete data under many circumstances [@buur18].
 
@@
 
-```{r diagram, out.width="\\linewidth", include=TRUE, fig.cap="Scheme of the main steps in multiple imputation---from an incomplete dataset, to $m=3$ multiply imputed datasets, to $m=3$ estimated quantities of scientific interest $\\hat{Q}$s, to a single pooled estimate $\\bar{Q}$."}
+```{r diagram, out.width="\\linewidth", include=TRUE, fig.cap="Scheme of the main steps in multiple imputation---from an incomplete dataset, to $m=3$ multiply imputed datasets, to $m=3$ estimated quantities of scientific interest $\\hat{Q}$, to a single pooled estimate $\\bar{Q}$."}
 knitr::include_graphics("../Figures/DiagramMI.pdf") #adapted from [@buur18] par. 1.4.1.
@@
 
-A popular method to obtain imputations  $\dot y_{\rm imp, \ell}$ is to use the 'Multiple Imputation by Chained Equations' algorithm, shorthand 'MICE'[@mice].
-With MICE, imputed values are drawn from the posterior predictive distribution of the missing values $y_{\rm mis}$.
+A popular method to obtain imputations is to use the 'Multiple Imputation by Chained Equations' algorithm, shorthand 'MICE'[@mice].
+With MICE, imputed values are drawn from the posterior predictive distribution of the missing values.
 The algorithm is named after the iterative algorithmic procedure by which imputed values are generated: a multivariate distribution is obtained by iterating over a sequence of univariate imputations.
@@
 And can we rely on the inference obtained on the completed data?
+<!-- $\dot y_{\rm imp, \ell}$ $y_{\rm mis}$ -->
 
@@
 The values generated by the algorithm (e.g., imputed values) will vary even after convergence [@gelm13].
-Since MCMC algorithms do not reach a unique point at which convergence is established, diagnostics  may only identify signs of *non*-convergence [@hoff09].
-Several of such non-convergence diagnostics exist, but it is not known whether these are appropriate for iterative imputation algorithms.
+Since MCMC algorithms do not reach a unique point at which convergence is established, **methods** may only identify signs of *non*-convergence [@hoff09].
+Several non-convergence diagnostics exist, but it is not known whether these are appropriate for iterative imputation algorithms.
 <!-- "a weak diagnostics is better than no diagnostic at all" [@cowl96]. -->
@@
 The aim of the simulation study is to determine whether unbiased, confidence-valid inferences may be obtained if the algorithm has not (yet) converged.
-And additionally, to formulate an informed advice on when approxiamte convergence may be safely concluded.
+And additionally, to formulate an informed advice on when to safely conclude approximate convergence.
 We will therefore assess the performance of different methods to identify non-convergence.
@@
 There are two requirements for convergence of iterative algorithms: mixing and stationarity [@gelm13].
-In iterative imputation algorithms mixing implies that imputation chains intermingle nicely, and stationarity is characterized by **the absence of trending** within chains.
-If one of these two requirements is not met, there is non-convergence in the algorithm.
+In iterative imputation algorithms, mixing implies that imputation chains intermingle nicely, and stationarity is characterized by **the absence of trending** within chains.
+If one of these two requirements is not met, we speak of non-convergence.
 Without mixing, chains may be 'stuck' at a local optimum, instead of sampling imputed values from the entire predictive posterior distribution of the missing values.
@@
 
+## Non-convergence in MICE
+
 To illustrate what non-mixing and non-stationarity looks like in iterative imputation algorithms, we reproduce an example from van Buuren [-@buur18, $\S$ 6.5.2].
@@
 Each line portrays one imputation chain: scalar summary $\theta$ plotted against the iteration number $t$.
-The $\theta$ in this example is the chain mean of a variable that we will call $j$ (i.e., the average of the imputed values for $j$ in each imputation $\dot y_{\rm imp, \ell}$).
+The $\theta$ in this example is the average of the imputed values for variable $j$ in each imputation $\dot y_{\rm imp, \ell}$ (also called the 'chain mean' of $j$).
 In the typical convergence scenario, the imputation chains intermingle nicely and there is little to no trending.
 In the non-convergence scenario, there is a lot of trending and some chains do not intermingle.
-Importantly, the chain means at the last iteration (the imputed value per imputation $\ell$) are very different between the two plots.
+Importantly, the chain means at the last iteration (the imputed value per imputation $\ell$) are very different between the two scenarios.
 The algorithm with the mis-specified model yields imputed values that are on average a factor two larger than those of the typically converged algorithm.
-Non-convergence thus influences the mean of $j$ in $\dot y_{\rm imp, \ell}$.
-This difference (presumably) translates to the $m$ completed data estimates $\hat{\mu}_{j,\ell}$, and consequently to the pooled estimate $\bar{\mu}_j$.
-Ergo, non-convergence leads to biased estimates.
-This shows the importance of reaching converged states in iterative imputation algorithms.
+It is obvious that non-convergence impacts the distribution of the imputed values in this example.
+This effect may translate into the distribution of the completed data, which consequently affects the estimated quantity of scientific interest $\bar{Q}$.
+$\bar{Q}$ may be biased due to systematically higher or lower imputed values, and invalid because the variance within imputation chains is under-estimated.
+Therefore, it is important to reach algorithmic convergence in iterative imputation algorithms.
+**[ADD t and l to each theta??]**
 
@@
 
-**[MERGE with sections below]** Currently, the recommended practice for evaluating the convergence of the MICE algorithm is through visual inspection.
-After running the imputation algorithm for a certain number of iterations, its convergence is monitiored by plotting a scalar summary of the state space of the algorithm $\theta$ against the iteration number.
-Typically, the $\theta$s under evaluation are chain means and variances---the average and the variance of the imputed values per imputation.
-Monitoring convergence by inspecting traceplots of these $\theta$s may be undesirable for several reasons: 1) it may be challenging to the untrained eye, and 2) only severely pathological cases of non-convergence may be diagnosed [@buur18, $\S$ 6.5.2] and, 3) there is not an objective measure that quantifies convergence.
-
-**[MERGE with section above and below]** Inspecting non-convergence through traceplots of chain means and chain variances may be insufficient because they are univariate summaries of the state space of the algorithm, while MICE is not only concerned with a single column, but the entire multivariate distribution of $y_{\rm imp}$.
+## Identifying non-convergence
+
+Currently, the recommended practice for evaluating the convergence of the MICE algorithm is through visual inspection.
+After running the imputation algorithm for a certain number of iterations, researchers are encouraged to produce traceplots.
+In a traceplot, a scalar summary of the state space of the algorithm $\theta$ is plotted against the iteration number, as depicted in Figure \ref{fig:non-conv}.
+Non-convergence is diagnosed if the imputation chains are not freely intermingled with one another, or if the chains show definite trends [@buur18].
+In `mice`, the default scalar summaries to inspect are chain means and chain variances.
+
+Identifying non-convergence by inspecting traceplots may be undesirable for several reasons: 1) it may be challenging to the untrained eye, 2) only severely pathological cases of non-convergence may be diagnosed, and 3) there is not an objective measure that quantifies convergence [@buur18, $\S$ 6.5.2].
+Moreover, traceplots typically only display chain means and chain variances, which are univariate scalar summaries of the state space of the algorithm.
+Monitoring univariate $\theta$s may be insufficient because MICE is not only concerned with a single column, but the entire multivariate distribution of $\dot y_{\rm imp, \ell}$.
 Ideally, a multivariate $\theta$ should be monitored.
+
 A suggestion by @buur18 for a multivariate $\theta$ to monitor is the quantity of scientific interest $Q$ (e.g., a regression coefficient), which usually is multivariate in nature.
 Implementing this, however, might be somewhat too technical for empirical researchers.
-Moreover, this scalar summary is not model-independent, i.e., it only applies to one model of scientific interest.
-Yet, one of the advantages of MI is that the missing data problem and scientific problem are solved independently.
+Moreover, this scalar summary is not model-independent, i.e., it is not universal to all complete data problems.
+Focusing on the convergence of outcome parameters may influence the iterative imputation procedure in the sense that the model of evaluation favors the model of interest.
+Additionally, one of the advantages of iterative imputation is that the missing data problem and scientific problem are solved independently.
 On these grounds, such a $\theta$ can be considered insufficient too.
-Van Buuren [-@buur18, $\S$ 4.5.2] also proposed multivariable evaluation of the MICE algorithm through eigenvalue decomposition, building on the work of @mack03.
-Eigenvalues of a variance-covariance matrix are a measure of the data's total covariance.
-Such a $\theta$ would be a model-independent, multivariate scalar summary to monitor, but it is not implemented in `mice`.
+
+As alternative, van Buuren [-@buur18, $\S$ 4.5.2] proposed multivariate evaluation of the MICE algorithm through eigenvalue decomposition, building on the work of @mack03.
+This technique may yield a model-independent, multivariate scalar summary to monitor, but it is not implemented in `mice`.
+
 <!-- Recommendation: "the different streams should be freely intermingled with one another, without showing any definite trends". -->
 
-**[MERGE with sections above]** Convergence has historically been inspected visually, by monitoring the imputation algorithm over iterations.
+<!-- **[MERGE with sections above]** Convergence has historically been inspected visually, by monitoring the imputation algorithm over iterations.
 This is typically done through traceplots.
@@
 $\theta$s are scalar summaries of the state space of the algorithm at a specific iteration.
-The default scalar summaries in `mice` are chain means and chain variances.
+-->
+
+<!-- The default scalar summaries in `mice` are chain means and chain variances.
 Additionally, researchers may "monitor some statistic of scientific interest" [@buur18, $\S$ 6.5.2].
@@
 Such a user-defined scalar summary, however, may be somewhat advanced for empirical researchers.
-And moreover, it is not universal to all complete data problems.
-Focusing on the convergence of outcome parameters may influence the iterative imputation procedure in the sense that the model of evaluation favors the model of interest.
-The downside to these two approaches is that they either focus on the univariate state space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest.
-Ideally, one would like to evaluate a model-independent parameter that summarizes the multivariate nature of the data.
-Therefore, we propose a $\theta$ that summarizes the multivariate state space of the algorithm, but is independent from the model of scientific interest.
+-->
+
+## A novel scalar summary
+
+The downside to the current $\theta$s to evaluate is that they either focus on the univariate state space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest.
+Ideally, one would like to evaluate a model-independent $\theta$ that summarizes the multivariate nature of the data.
+
 We propose $\lambda_{1, \ell}$ as such a scalar summary.

### 2020-05-10 3.Thesis/Writeup/2.Manuscript.tex
Until diagnostics

@@
 There is no consensus on how to evaluate the convergence properties of the method.
-This paper provides insight into identifying non-convergence of iterative imputation algorithms.
+This paper provides insight into identifying non-convergence in iterative imputation algorithms.
 We conclude that in the case of drawing inference from incomplete data, convergence of the iterative imputation algorithm is often a convenience but not a necessity.
@@
 
-\keywords{missing data, iterative imputation, non-convergence, mice}
+\keywords{missing data, iterative imputation, non-convergence, MICE}
 
@@
 This technique is known as `multiple imputation' \citep[MI;][]{rubin76}.
-MI has proven to be a powerful technique to draw valid inferences from incomplete data under many circumstances \citep{buur18}.
+MI has proven to be a powerful tool to draw valid inferences from incomplete data under many circumstances \citep{buur18}.
 
@@
 
-\caption{Scheme of the main steps in multiple imputation---from an incomplete dataset, to $m=3$ multiply imputed datasets, to $m=3$ estimated quantities of scientific interest $\hat{Q}$s, to a single pooled estimate $\bar{Q}$.}\label{fig:diagram}
+\caption{Scheme of the main steps in multiple imputation---from an incomplete dataset, to $m=3$ multiply imputed datasets, to $m=3$ estimated quantities of scientific interest $\hat{Q}$, to a single pooled estimate $\bar{Q}$.}\label{fig:diagram}
 \end{figure}
 
-A popular method to obtain imputations \(\dot y_{\rm imp, \ell}\) is to use the `Multiple Imputation by Chained Equations' algorithm, shorthand `MICE'\citep{mice}.
-With MICE, imputed values are drawn from the posterior predictive distribution of the missing values \(y_{\rm mis}\).
+A popular method to obtain imputations is to use the `Multiple Imputation by Chained Equations' algorithm, shorthand `MICE'\citep{mice}.
+With MICE, imputed values are drawn from the posterior predictive distribution of the missing values.
 The algorithm is named after the iterative algorithmic procedure by which imputed values are generated: a multivariate distribution is obtained by iterating over a sequence of univariate imputations.
@@
 The values generated by the algorithm (e.g., imputed values) will vary even after convergence \citep{gelm13}.
-Since MCMC algorithms do not reach a unique point at which convergence is established, diagnostics may only identify signs of \emph{non}-convergence \citep{hoff09}.
-Several of such non-convergence diagnostics exist, but it is not known whether these are appropriate for iterative imputation algorithms.
+Since MCMC algorithms do not reach a unique point at which convergence is established, \textbf{methods} may only identify signs of \emph{non}-convergence \citep{hoff09}.
+Several non-convergence diagnostics exist, but it is not known whether these are appropriate for iterative imputation algorithms.
 
@@
 There are two requirements for convergence of iterative algorithms: mixing and stationarity \citep{gelm13}.
-In iterative imputation algorithms mixing implies that imputation chains intermingle nicely, and stationarity is characterized by \textbf{the absence of trending} within chains.
-If one of these two requirements is not met, there is non-convergence in the algorithm.
+In iterative imputation algorithms, mixing implies that imputation chains intermingle nicely, and stationarity is characterized by \textbf{the absence of trending} within chains.
+If one of these two requirements is not met, we speak of non-convergence.
 Without mixing, chains may be `stuck' at a local optimum, instead of sampling imputed values from the entire predictive posterior distribution of the missing values.
@@
 
+\hypertarget{non-convergence-in-mice}{%
+\subsection{Non-convergence in MICE}\label{non-convergence-in-mice}}
+
 To illustrate what non-mixing and non-stationarity looks like in iterative imputation algorithms, we reproduce an example from van Buuren \citeyearpar[\(\S\) 6.5.2]{buur18}.
@@
 Each line portrays one imputation chain: scalar summary \(\theta\) plotted against the iteration number \(t\).
-The \(\theta\) in this example is the chain mean of a variable that we will call \(j\) (i.e., the average of the imputed values for \(j\) in each imputation \(\dot y_{\rm imp, \ell}\)).
+The \(\theta\) in this example is the average of the imputed values for variable \(j\) in each imputation \(\dot y_{\rm imp, \ell}\) (also called the `chain mean' of \(j\)).
 In the typical convergence scenario, the imputation chains intermingle nicely and there is little to no trending.
 In the non-convergence scenario, there is a lot of trending and some chains do not intermingle.
-Importantly, the chain means at the last iteration (the imputed value per imputation \(\ell\)) are very different between the two plots.
+Importantly, the chain means at the last iteration (the imputed value per imputation \(\ell\)) are very different between the two scenarios.
 The algorithm with the mis-specified model yields imputed values that are on average a factor two larger than those of the typically converged algorithm.
-Non-convergence thus influences the mean of \(j\) in \(\dot y_{\rm imp, \ell}\).
-This difference (presumably) translates to the \(m\) completed data estimates \(\hat{\mu}_{j,\ell}\), and consequently to the pooled estimate \(\bar{\mu}_j\).
-Ergo, non-convergence leads to biased estimates.
-This shows the importance of reaching converged states in iterative imputation algorithms.
+It is obvious that non-convergence impacts the distribution of the imputed values in this example.
+This effect may translate into the distribution of the completed data, which consequently affects the estimated quantity of scientific interest \(\bar{Q}\).
+\(\bar{Q}\) may be biased due to systematically higher or lower imputed values, and invalid because the variance within imputation chains is under-estimated.
+Therefore, it is important to reach algorithmic convergence in iterative imputation algorithms.
+\textbf{{[}ADD t and l to each theta??{]}}
 
@@
 
-\textbf{{[}MERGE with sections below{]}} Currently, the recommended practice for evaluating the convergence of the MICE algorithm is through visual inspection.
-After running the imputation algorithm for a certain number of iterations, its convergence is monitiored by plotting a scalar summary of the state space of the algorithm \(\theta\) against the iteration number.
-Typically, the \(\theta\)s under evaluation are chain means and variances---the average and the variance of the imputed values per imputation.
-Monitoring convergence by inspecting traceplots of these \(\theta\)s may be undesirable for several reasons: 1) it may be challenging to the untrained eye, and 2) only severely pathological cases of non-convergence may be diagnosed \citep[\(\S\) 6.5.2]{buur18} and, 3) there is not an objective measure that quantifies convergence.
+\hypertarget{identifying-non-convergence}{%
+\subsection{Identifying non-convergence}\label{identifying-non-convergence}}
 
-\textbf{{[}MERGE with section above and below{]}} Inspecting non-convergence through traceplots of chain means and chain variances may be insufficient because they are univariate summaries of the state space of the algorithm, while MICE is not only concerned with a single column, but the entire multivariate distribution of \(y_{\rm imp}\).
+Currently, the recommended practice for evaluating the convergence of the MICE algorithm is through visual inspection.
+After running the imputation algorithm for a certain number of iterations, researchers are encouraged to produce traceplots.
+In a traceplot, a scalar summary of the state space of the algorithm \(\theta\) is plotted against the iteration number, as depicted in Figure \ref{fig:non-conv}.
+Non-convergence is diagnosed if the imputation chains are not freely intermingled with one another, or if the chains show definite trends \citep{buur18}.
+In \texttt{mice}, the default scalar summaries to inspect are chain means and chain variances.
+
+Identifying non-convergence by inspecting traceplots may be undesirable for several reasons: 1) it may be challenging to the untrained eye, 2) only severely pathological cases of non-convergence may be diagnosed, and 3) there is not an objective measure that quantifies convergence \citep[\(\S\) 6.5.2]{buur18}.
+Moreover, traceplots typically only display chain means and chain variances, which are univariate scalar summaries of the state space of the algorithm.
+Monitoring univariate \(\theta\)s may be insufficient because MICE is not only concerned with a single column, but the entire multivariate distribution of \(\dot y_{\rm imp, \ell}\).
 Ideally, a multivariate \(\theta\) should be monitored.
+
 A suggestion by \citet{buur18} for a multivariate \(\theta\) to monitor is the quantity of scientific interest \(Q\) (e.g., a regression coefficient), which usually is multivariate in nature.
 Implementing this, however, might be somewhat too technical for empirical researchers.
-Moreover, this scalar summary is not model-independent, i.e., it only applies to one model of scientific interest.
-Yet, one of the advantages of MI is that the missing data problem and scientific problem are solved independently.
+Moreover, this scalar summary is not model-independent, i.e., it is not universal to all complete data problems.
+Focusing on the convergence of outcome parameters may influence the iterative imputation procedure in the sense that the model of evaluation favors the model of interest.
+Additionally, one of the advantages of iterative imputation is that the missing data problem and scientific problem are solved independently.
 On these grounds, such a \(\theta\) can be considered insufficient too.
-Van Buuren \citeyearpar[\(\S\) 4.5.2]{buur18} also proposed multivariable evaluation of the MICE algorithm through eigenvalue decomposition, building on the work of \citet{mack03}.
-Eigenvalues of a variance-covariance matrix are a measure of the data's total covariance.
-Such a \(\theta\) would be a model-independent, multivariate scalar summary to monitor, but it is not implemented in \texttt{mice}.
 
-\textbf{{[}MERGE with sections above{]}} Convergence has historically been inspected visually, by monitoring the imputation algorithm over iterations.
-This is typically done through traceplots.
-In a traceplot, the iteration number is plotted against a certain \(\theta\).
-\(\theta\)s are scalar summaries of the state space of the algorithm at a specific iteration.
-The default scalar summaries in \texttt{mice} are chain means and chain variances.
-Additionally, researchers may ``monitor some statistic of scientific interest'' \citep[\(\S\) 6.5.2]{buur18}.
-As \citet{buur18} describes, researchers can specify their quantity of scientific interest \(Q\) (e.g., a regression coefficient) as scalar summary \(\theta\).
-Such a user-defined scalar summary, however, may be somewhat advanced for empirical researchers.
-And moreover, it is not universal to all complete data problems.
-Focusing on the convergence of outcome parameters may influence the iterative imputation procedure in the sense that the model of evaluation favors the model of interest.
-The downside to these two approaches is that they either focus on the univariate state space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest.
-Ideally, one would like to evaluate a model-independent parameter that summarizes the multivariate nature of the data.
-Therefore, we propose a \(\theta\) that summarizes the multivariate state space of the algorithm, but is independent from the model of scientific interest.
+As alternative, van Buuren \citeyearpar[\(\S\) 4.5.2]{buur18} proposed multivariate evaluation of the MICE algorithm through eigenvalue decomposition, building on the work of \citet{mack03}.
+This technique may yield a model-independent, multivariate scalar summary to monitor, but it is not implemented in \texttt{mice}.
+
+\hypertarget{a-novel-scalar-summary}{%
+\subsection{A novel scalar summary}\label{a-novel-scalar-summary}}
+
+The downside to the current \(\theta\)s to evaluate is that they either focus on the univariate state space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest.
+Ideally, one would like to evaluate a model-independent \(\theta\) that summarizes the multivariate nature of the data.
+
 We propose \(\lambda_{1, \ell}\) as such a scalar summary.

### 2020-05-10 3.Thesis/Writeup/2.Manuscript.Rmd
Until p. 9

@@
 
-These remain open questions, since the convergence properties of iterative imputation algorithms have not been systematically studied [@buur18].
+These remain open questions, since the convergence properties of iterative imputation algorithms have not been systematically studied [@zhu15; @buur18].
 There is no scientific consensus on how to evaluate the convergence of imputation algorithms [@taka17].
@@
 We then adress the impact of inducing non-convergence in iterative imputation algorithms through model-based simulation in `R` [@R].
+For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular `mice` package [@mice] in `R`.
 The aim of the simulation study is to determine whether unbiased, confidence-valid inferences may be obtained if the algorithm has not (yet) converged.
-And additionally, to formulate an informed advice on when to safely conclude approximate convergence.
-We will therefore assess the performance of different methods to identify non-convergence.
+And additionally, to evaluate the behavior and performance of several diagnostic methods to identify non-convergence.
+With that, we formulate an informed advice on when approximate convergence, necessary for valid inferences, may be safely concluded.
 We translate the results of the study into guidelines for applied researchers.
-**ADD that these guidelines may aid to draw valid inference from incomplete data?]** For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular `mice` package [@mice] in `R` [@R].
+**ADD that these guidelines may aid to draw valid inference from incomplete data?]**
 
@@
 For each datapoint in $y_{\rm mis}$, we sample $m \times T$ plausible values, where $m$ is the number of imputations ($\ell = 1, 2, \dots, m$) and $T$ is the number of iterations ($t = 1, 2, \dots, T$) in the imputation algorithm.
-The state space of the algorithm at a certain iteration $t$ may be summarized by a scalar summary $\theta$ (e.g., **the average of the imputed values**).
-The collection of $\theta$-values between the **initial value** (at $t=1$) and the final imputed value (at $t=T$) will be referred to as an 'imputation chain'.
+The state space of the algorithm at a certain iteration $t$ may be summarized by a scalar summary $\theta$ (e.g., the average of the **imputed values**).
+The collection of $\theta$-values between $t=1$ (summarizing the state space of the algorithm at initialization) and $t=T$ (summarizing the state space for the imputed values) will be referred to as an 'imputation chain'.
 <!-- **[If not defined elsewhere, add: $\theta$s are scalar summaries of the algorithm at a certain iteration (e.g., chain means; the average of the imputed values in each imputation chain).] ** -->
@@
 There are two requirements for convergence of iterative algorithms: mixing and stationarity [@gelm13].
-In iterative imputation algorithms, mixing implies that imputation chains intermingle nicely, and stationarity is characterized by **the absence of trending** within chains.
-If one of these two requirements is not met, we speak of non-convergence.
+In iterative imputation algorithms, mixing implies that imputation chains intermingle nicely, and stationarity is characterized by the absence of trending across iterations.
+If one of the two requirements is not met, we speak of non-convergence.
 Without mixing, chains may be 'stuck' at a local optimum, instead of sampling imputed values from the entire predictive posterior distribution of the missing values.
@@
 
-## Non-convergence in MICE
-
 To illustrate what non-mixing and non-stationarity looks like in iterative imputation algorithms, we reproduce an example from van Buuren [-@buur18, $\S$ 6.5.2].
@@
 The right-hand side displays pathological non-convergence, induced by purposefully mis-specifying the imputation model.
-Each line portrays one imputation chain: scalar summary $\theta$ plotted against the iteration number $t$.
-The $\theta$ in this example is the average of the imputed values for variable $j$ in each imputation $\dot y_{\rm imp, \ell}$ (also called the 'chain mean' of $j$).
+Each line portrays one imputation chain (i.e., the values of a scalar summary $\theta$ across iterations).
+The $\theta$ depicted here is the 'chain mean' of variable $j$, which is defined as the average of variable $j$ in $\dot y_{\rm imp, \ell}$.
+
 In the typical convergence scenario, the imputation chains intermingle nicely and there is little to no trending.
@@
 The algorithm with the mis-specified model yields imputed values that are on average a factor two larger than those of the typically converged algorithm.
-It is obvious that non-convergence impacts the distribution of the imputed values in this example.
-This effect may translate into the distribution of the completed data, which consequently affects the estimated quantity of scientific interest $\bar{Q}$.
-$\bar{Q}$ may be biased due to systematically higher or lower imputed values, and invalid because the variance within imputation chains is under-estimated.
+It is obvious that non-convergence impacts the distribution of the imputed values in this example, $\dot y_{\rm imp, \ell}$.
+This effect may translate into the distribution of the completed data {$y_{\rm obs}, \dot y_{\rm imp, \ell}$}, which consequently affects the estimated quantities of scientific interest $\hat{Q}_{\ell}$ and the pooled estimate $\bar{Q}$.
+$\bar{Q}$ is then a  biased, invalid estimate of $Q$, since systematically higher imputed values lead to bias, and under-estimation of the variance within imputation chains poses a threat to confidence-validity.
 Therefore, it is important to reach algorithmic convergence in iterative imputation algorithms.
@@
 ```{r non-conv, out.width='.49\\linewidth', fig.width=3.5, fig.height=3,fig.show='hold',fig.cap="Typical convergence versus pathological non-convergence.
-Please note that this is the same data with a different imputation model, leading to different imputations (e.g., see the units on the y-axis)."}
+Depicted are are chain means of variable $j$ in each imputation ($m=5$; $T=10$).
+Please note that any difference between the two scenarios (see e.g., the magnitude of the chain means on the y-axes) stems from the imputation algorithms---the two scenarios have equal incomplete data $y$."}
 par(mar = c(2, 2, 0.1, 0.1))
@@
 In a traceplot, a scalar summary of the state space of the algorithm $\theta$ is plotted against the iteration number, as depicted in Figure \ref{fig:non-conv}.
+In `mice`, the default scalar summaries to inspect are chain means and chain variances.
 Non-convergence is diagnosed if the imputation chains are not freely intermingled with one another, or if the chains show definite trends [@buur18].
-In `mice`, the default scalar summaries to inspect are chain means and chain variances.
 
 Identifying non-convergence by inspecting traceplots may be undesirable for several reasons: 1) it may be challenging to the untrained eye, 2) only severely pathological cases of non-convergence may be diagnosed, and 3) there is not an objective measure that quantifies convergence [@buur18, $\S$ 6.5.2].
-Moreover, traceplots typically only display chain means and chain variances, which are univariate scalar summaries of the state space of the algorithm.
-Monitoring univariate $\theta$s may be insufficient because MICE is not only concerned with a single column, but the entire multivariate distribution of $\dot y_{\rm imp, \ell}$.
-Ideally, a multivariate $\theta$ should be monitored.
-
-A suggestion by @buur18 for a multivariate $\theta$ to monitor is the quantity of scientific interest $Q$ (e.g., a regression coefficient), which usually is multivariate in nature.
+Moreover, the typical traceplots to inspect only **consider** univariate $\theta$s.
+Monitoring univariate summaries of the state space of the algorithm may be insufficient because MICE is not only concerned with a single column, but the entire multivariate distribution of the imputations.
+Ideally, we would monitor a multivariate $\theta$.
+
+A suggestion by @buur18 for a multivariate $\theta$ to monitor is the estimated quantity of scientific interest $\hat{Q}$, which usually is multivariate in nature.
+This $\theta$ is computed as the estimate in each imputation $\ell$, $\hat{Q}_\ell$, across iterations.
 Implementing this, however, might be somewhat too technical for empirical researchers.
-Moreover, this scalar summary is not model-independent, i.e., it is not universal to all complete data problems.
+Besides, this scalar summary is not model-independent.
+That is, this $\theta$ is not universal to all complete data problems, while one of the advantages of iterative imputation is that the missing data problem and scientific problem are solved independently.
 Focusing on the convergence of outcome parameters may influence the iterative imputation procedure in the sense that the model of evaluation favors the model of interest.
-Additionally, one of the advantages of iterative imputation is that the missing data problem and scientific problem are solved independently.
 On these grounds, such a $\theta$ can be considered insufficient too.
@@
 
-The downside to the current $\theta$s to evaluate is that they either focus on the univariate state space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest.
+To monitor non-convergence in iterative imputation algorithms, we may  summarize the state space of the algorithm with several $\theta$s.
+The downside to the current $\theta$s is that they either focus on the univariate state space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest.
 Ideally, one would like to evaluate a model-independent $\theta$ that summarizes the multivariate nature of the data.
 
-We propose $\lambda_{1, \ell}$ as such a scalar summary.
-We define $\lambda_{1, \ell}$ as the first eigenvalue of the variance-covariance matrix of the completed data.
+We propose $\lambda_{1}$ as such a scalar summary.
+We define $\lambda_{1}$ as the first eigenvalue of the variance-covariance matrix in the $m$ completed datasets.
+The eigenvalues of a variance-covariance matrix $S$ summarize the total set of covariances in the data.
 Let $\lambda_{1, \ell} \geq \lambda_{2, \ell} \geq ...
-\geq \lambda_{j, \ell}$ be the eigenvalues of $\Sigma_\ell$ in each of the $m$ completed datasets {$y_{\rm obs}, \dot y_{\rm imp, \ell}$}.
-$\lambda_{1, \ell}$ is measure that summarizes the covariances in the completed datasets.
-The first eigenvalue has the appealing property that is not dependent on the substantive model of interest.
-Eigenvalue decomposition is inspired by @mack03.**[ADD definition variance in first PCA component is equal to the first eigenvalue of the variance-covariance matrix]**
+\geq \lambda_{j, \ell}$ be the eigenvalues of $S_\ell$ in each imputation $\ell$.
+$\lambda_{1,\ell}$ then summarizes the largest possible amount of covariance in the $m$ completed datasets {$y_{\rm obs}, \dot y_{\rm imp, \ell}$}.
+By definition, each $\lambda_{1, \ell}$-value is equal to the variance of the first component that would be obtained by performing principal component analysis on {$y_{\rm obs}, \dot y_{\rm imp, \ell}$}.
+As scalar summary, $\lambda_{1}$ has the appealing property that it is not dependent on the substantive model of interest, yet still summarizes the multivariate state space of the algorithm.
+\newline
 
@@
 
+\noindent In this study, we consider each of the four $\theta$s that were discussed above.
+Namely, two univariate scalar summaries (chain mean and chain variance), a model-dependent multivariate summary $\hat{Q}$, and the novel model-independent multivariate summary $\lambda_1$.
+
 # Non-convergence diagnostics
@@
 
-Non-convergence diagnostics for MCMC algorithms typically identify problems in either one of the two requirements for convergence: mixing or stationarity.
-Non-stationarity within chains may be diagnosed with e.g., autocorrelation [$AC$; @scha97; @gelm13], numeric standard error ['MC error'; @gewe92], or Raftery and Lewis's [-@raft91] procedure to determine the effect of trending on the precision of estimates.
-A widely used diagnostic to monitor mixing between chains is the potential scale reduction factor $\widehat{R}$ ['Gelman-Rubin statistic'; @gelm92].
+There are many diagnostic tools to identify non-convergence in iterative (MCMC) algorithms [@broo98; @elad06].
+We consider only two of them that may be appropriate for imputation algorithms---one to monitor signs of non-mixing, and one for non-stationarity.
+As recommended by e.g.
+@cowl96 we will use the potential scale reduction factor $\widehat{R}$ ['Gelman-Rubin statistic'; @gelm92] to evaluate mixing, and autocorrelation [$AC$; @scha97; @gelm13] to diagnose trending.
 With a recently proposed adaptation, $\widehat{R}$ might also serve to diagnose non-stationarity, but this has not yet been thoroughly investigated [@veht19].
+Therefore, we will evaluate the appropriateness of both $\widehat{R}$ versions, in addition to $AC$.
+Other methods are outside the scope of this study, because they e.g.
+assume that values within chains represent independent samples, whereas MI only uses the final iteration to impute.
+
+<!-- Non-stationarity within chains may be diagnosed with e.g., autocorrelation [$AC$; @scha97; @gelm13], numeric standard error ['MC error'; @gewe92], or Raftery and Lewis's [-@raft91] procedure to determine the effect of trending on the precision of estimates.
+-->
+
+<!-- A widely used diagnostic to monitor mixing between chains is the potential scale reduction factor $\widehat{R}$ ['Gelman-Rubin statistic'; @gelm92].
 Therefore, we use $\widehat{R}$ **[WHICH?
 Explain that we'll use both]** and $AC$ to evaluate mixing and stationarity separately, as recommended by e.g., @cowl96.
-<!-- @cowl96, p. 898. -->
-
-**[Add: we don't use effective sample size and MC error because they assume independent samples *within* chains.
+-->
+
+<!-- **[Add: we don't use effective sample size and MC error because they assume independent samples *within* chains.
 In MI, we only use the final estimate.
 Also, computing n_eff assumes infinitely many iterations, which is fine in Bayesian analyses where T is often at least one thousand.
-But the default in iterative imp is often substantially lower.]**
+But the default in iterative imp is often substantially lower.]** -->
 
@@
 
-The potential scale reduction factor $\widehat{R}$ was coined by @gelm92.
-An updated version of $\widehat{R}$ has been proposed by Vehtari et al.
-[-@veht19, p.
-5].
-**This version is better suited to detect non-convergence in the tails of distributions, by using three transformations on the scalar summary that it is applied to.
+In 2019, Vehtari et al.
+proposed an updated version of the potential scale reduction factor $\widehat{R}$, originally coined in 1992.
+The adapted version would be better suited to detect non-mixing in the tails of distributions, and even identify non-stationarity.
+This version uses three transformations on the scalar summary $\theta$, before computing $\widehat{R}$-values.
 Namely, rank-normalization, folding, and localization.
-The Vehtari et al.
-version of $\widehat{R}$ may be suitable for iterative imputation.
-We therefore use their definition of the diagnostic.** Let $m$ be the total number of chains, $T$ the number of iterations per chain (where $T\geq2$), and $\theta$ the scalar summary of interest.
+The definition of the diagnostic itself is equal to the original $\widehat{R}$.
+We follow Vehtari et al.'s formulation (2019, p.
+5).
+Let $m$ be the total number of chains, $T$ the number of iterations per chain (where $T\geq2$), and $\theta$ the scalar summary of interest.
 For each chain ($\ell = 1, 2, \dots, m$), we estimate the variance of $\theta$, and average these to obtain within-chain variance $W$.
@@
 
-\noindent We then estimate between-chain variance $B$ as the variance of the collection of average $\theta$s per chain. **[Note that this is not the usual B in MI]**
+\noindent We then estimate between-chain variance $B$ as the variance of the collection of average $\theta$s per chain.
+**Note that $B$ refers to the variance in $\theta$, instead of the typical notation in MI where $B$ denotes the variance between estimated quantities of scientific interest $\hat{Q}_{\ell}$.**
 
@@
 147].
-In this study, we only consider $AC$ at lag 1, i.e., the correlation between the $t^{th}$ and $t+1^{th}$ iteration of the same chain.
+In this study, we only consider $AC$ at lag 1, i.e., the correlation between the $t^{th}$ and $(t+1)^{th}$ iteration of the same chain.
 Following the same notation as for $\widehat{R}$,
@@
 Divergence within imputation poses no threat to the convergence of the algorithm---it may even speed up convergence.
-**Complete stationarity is reached when $AC=0$.** As non-convergence diagnostic, our interest is in positive $AC$-values.
+Complete stationarity is reached when $AC=0$.
+As non-convergence diagnostic, our interest is in positive $AC$-values.
 
@@
 Because of its convergence to a distribution, the algorithm will show some signs of non-mixing and non-stationarity even in the most converged state.
-The aim is therefore to reach approximate convergence [@gelm13].
+The aim, therefore, is to reach approximate convergence [@gelm13].
 Upon approximate convergence, the imputation chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm ($\widehat{R} > 1$), and there is little dependency between subsequent iterations of imputation chains ($AC > 0$).
-**Approximate convergence is violated when the non-convergence diagnostics exceed a certain threshold.
-[or] In practice, non-convergence is therefore diagnosed when non-convergence diagnostics exeed a certain threshold.** The conventional thresholds to diagnose non-mixing are $\widehat{R} > 1.2$ [@gelm92] or $\widehat{R} > 1.1$ [@gelm13].
+In practice, we diagnose non-convergence when approximate convergence is violated, i.e., when $\widehat{R}$ and $AC$ exceed a certain threshold.
+The conventional thresholds to diagnose non-mixing are $\widehat{R} > 1.2$ [@gelm92] or $\widehat{R} > 1.1$ [@gelm13].
 @veht19 proposed a much more stringent threshold of $\widehat{R} > 1.01$.
@@
 $AC$-values that are significantly higher than zero indicate non-stationarity.
-**[think about one-sided test]**
 
@@
 
-Before we evaluate the performance of the non-convergence diagnostics $\widehat{R}$ and $AC$ quantitatively through simulation, we apply them to the example of pathological non-convergence by @buur18 that we reproduced above.
-We want to be able to draw the same conclusions as we did from visually inspecting the two scenarios in Figure \ref{fig:non-conv}.
-The methods should at least indicate worse performance for the scenario with pathological non-convergence compared to the typically converged algorithm (i.e., higher $\widehat{R}$- and $AC$-values).
-
-In Figure \ref{fig:diagnostics}A, the chain means from Figure \ref{fig:non-conv} are plotted again---now together.
-Panel B shows two versions of calculating $AC$s applied to the $\theta$s of panel A: the default calculation with `R` function `stats::acf()` [@R], and manual calculation as the correlation between $\theta$ in iteration $t$ and $\theta$ in iteration $t+1$.
-Panel C shows the traditional computation of $\widehat{R}$, and panel D shows $\widehat{R}$ as computed by implementing @veht19 's recommendations.
+Before we evaluate the performance of the non-convergence diagnostics $\widehat{R}$ and $AC$ quantitatively through simulation, we first assess their appropriateness qualitatively.
+We do this by applying them to the example of pathological non-convergence that we reproduced form @buur18.
+Ideally, the diagnostics are as informative as, or better than visual inspection of the traceplots.
+The methods should at least indicate worse performance (higher $\widehat{R}$- and $AC$-values) for the scenario with pathological non-convergence, compared to the typically converged algorithm.
+
+Each panel in Figure \ref{fig:diagnostics} depicts one method to identify non-convergence, applied to the two scenarios from Figure \ref{fig:non-conv}.
+Panel A consists of the two traceplots that may be evaluated through visual inspection.
+Panel B shows two versions of the $AC$: the default calculation with `R` function `stats::acf()` [@R], and manual calculation as the correlation between $\theta$ in iteration $t$ and $\theta$ in iteration $t+1$.
+Panel C displays the traditional computation of $\widehat{R}$ conform @gelm92, and in panel D we see $\widehat{R}$ as computed by implementing @veht19 's recommendations.
 **[Add labels for the two types of Rhat to the figure!]**

### 2020-05-10 3.Thesis/Writeup/2.Manuscript.tex
Until p. 9

@@
 
-These remain open questions, since the convergence properties of iterative imputation algorithms have not been systematically studied \citep{buur18}.
+These remain open questions, since the convergence properties of iterative imputation algorithms have not been systematically studied \citep{zhu15, buur18}.
 There is no scientific consensus on how to evaluate the convergence of imputation algorithms \citep{taka17}.
@@
 We then adress the impact of inducing non-convergence in iterative imputation algorithms through model-based simulation in \texttt{R} \citep{R}.
+For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular \texttt{mice} package \citep{mice} in \texttt{R}.
 The aim of the simulation study is to determine whether unbiased, confidence-valid inferences may be obtained if the algorithm has not (yet) converged.
-And additionally, to formulate an informed advice on when to safely conclude approximate convergence.
-We will therefore assess the performance of different methods to identify non-convergence.
+And additionally, to evaluate the behavior and performance of several diagnostic methods to identify non-convergence.
+With that, we formulate an informed advice on when approximate convergence, necessary for valid inferences, may be safely concluded.
 We translate the results of the study into guidelines for applied researchers.
-\textbf{ADD that these guidelines may aid to draw valid inference from incomplete data?{]}} For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular \texttt{mice} package \citep{mice} in \texttt{R} \citep{R}.
+\textbf{ADD that these guidelines may aid to draw valid inference from incomplete data?{]}}
 
@@
 For each datapoint in \(y_{\rm mis}\), we sample \(m \times T\) plausible values, where \(m\) is the number of imputations (\(\ell = 1, 2, \dots, m\)) and \(T\) is the number of iterations (\(t = 1, 2, \dots, T\)) in the imputation algorithm.
-The state space of the algorithm at a certain iteration \(t\) may be summarized by a scalar summary \(\theta\) (e.g., \textbf{the average of the imputed values}).
-The collection of \(\theta\)-values between the \textbf{initial value} (at \(t=1\)) and the final imputed value (at \(t=T\)) will be referred to as an `imputation chain'.
+The state space of the algorithm at a certain iteration \(t\) may be summarized by a scalar summary \(\theta\) (e.g., the average of the \textbf{imputed values}).
+The collection of \(\theta\)-values between \(t=1\) (summarizing the state space of the algorithm at initialization) and \(t=T\) (summarizing the state space for the imputed values) will be referred to as an `imputation chain'.
 
@@
 There are two requirements for convergence of iterative algorithms: mixing and stationarity \citep{gelm13}.
-In iterative imputation algorithms, mixing implies that imputation chains intermingle nicely, and stationarity is characterized by \textbf{the absence of trending} within chains.
-If one of these two requirements is not met, we speak of non-convergence.
+In iterative imputation algorithms, mixing implies that imputation chains intermingle nicely, and stationarity is characterized by the absence of trending across iterations.
+If one of the two requirements is not met, we speak of non-convergence.
 Without mixing, chains may be `stuck' at a local optimum, instead of sampling imputed values from the entire predictive posterior distribution of the missing values.
@@
 
-\hypertarget{non-convergence-in-mice}{%
-\subsection{Non-convergence in MICE}\label{non-convergence-in-mice}}
-
 To illustrate what non-mixing and non-stationarity looks like in iterative imputation algorithms, we reproduce an example from van Buuren \citeyearpar[\(\S\) 6.5.2]{buur18}.
@@
 The right-hand side displays pathological non-convergence, induced by purposefully mis-specifying the imputation model.
-Each line portrays one imputation chain: scalar summary \(\theta\) plotted against the iteration number \(t\).
-The \(\theta\) in this example is the average of the imputed values for variable \(j\) in each imputation \(\dot y_{\rm imp, \ell}\) (also called the `chain mean' of \(j\)).
+Each line portrays one imputation chain (i.e., the values of a scalar summary \(\theta\) across iterations).
+The \(\theta\) depicted here is the `chain mean' of variable \(j\), which is defined as the average of variable \(j\) in \(\dot y_{\rm imp, \ell}\).
+
 In the typical convergence scenario, the imputation chains intermingle nicely and there is little to no trending.
@@
 The algorithm with the mis-specified model yields imputed values that are on average a factor two larger than those of the typically converged algorithm.
-It is obvious that non-convergence impacts the distribution of the imputed values in this example.
-This effect may translate into the distribution of the completed data, which consequently affects the estimated quantity of scientific interest \(\bar{Q}\).
-\(\bar{Q}\) may be biased due to systematically higher or lower imputed values, and invalid because the variance within imputation chains is under-estimated.
+It is obvious that non-convergence impacts the distribution of the imputed values in this example, \(\dot y_{\rm imp, \ell}\).
+This effect may translate into the distribution of the completed data \{\(y_{\rm obs}, \dot y_{\rm imp, \ell}\)\}, which consequently affects the estimated quantities of scientific interest \(\hat{Q}_{\ell}\) and the pooled estimate \(\bar{Q}\).
+\(\bar{Q}\) is then a biased, invalid estimate of \(Q\), since systematically higher imputed values lead to bias, and under-estimation of the variance within imputation chains poses a threat to confidence-validity.
 Therefore, it is important to reach algorithmic convergence in iterative imputation algorithms.
@@
 \caption{Typical convergence versus pathological non-convergence.
-Please note that this is the same data with a different imputation model, leading to different imputations (e.g., see the units on the y-axis).}\label{fig:non-conv}
+Depicted are are chain means of variable $j$ in each imputation ($m=5$; $T=10$).
+Please note that any difference between the two scenarios (see e.g., the magnitude of the chain means on the y-axes) stems from the imputation algorithms---the two scenarios have equal incomplete data $y$.}\label{fig:non-conv}
 \end{figure}
@@
 In a traceplot, a scalar summary of the state space of the algorithm \(\theta\) is plotted against the iteration number, as depicted in Figure \ref{fig:non-conv}.
+In \texttt{mice}, the default scalar summaries to inspect are chain means and chain variances.
 Non-convergence is diagnosed if the imputation chains are not freely intermingled with one another, or if the chains show definite trends \citep{buur18}.
-In \texttt{mice}, the default scalar summaries to inspect are chain means and chain variances.
 
 Identifying non-convergence by inspecting traceplots may be undesirable for several reasons: 1) it may be challenging to the untrained eye, 2) only severely pathological cases of non-convergence may be diagnosed, and 3) there is not an objective measure that quantifies convergence \citep[\(\S\) 6.5.2]{buur18}.
-Moreover, traceplots typically only display chain means and chain variances, which are univariate scalar summaries of the state space of the algorithm.
-Monitoring univariate \(\theta\)s may be insufficient because MICE is not only concerned with a single column, but the entire multivariate distribution of \(\dot y_{\rm imp, \ell}\).
-Ideally, a multivariate \(\theta\) should be monitored.
+Moreover, the typical traceplots to inspect only \textbf{consider} univariate \(\theta\)s.
+Monitoring univariate summaries of the state space of the algorithm may be insufficient because MICE is not only concerned with a single column, but the entire multivariate distribution of the imputations.
+Ideally, we would monitor a multivariate \(\theta\).
 
-A suggestion by \citet{buur18} for a multivariate \(\theta\) to monitor is the quantity of scientific interest \(Q\) (e.g., a regression coefficient), which usually is multivariate in nature.
+A suggestion by \citet{buur18} for a multivariate \(\theta\) to monitor is the estimated quantity of scientific interest \(\hat{Q}\), which usually is multivariate in nature.
+This \(\theta\) is computed as the estimate in each imputation \(\ell\), \(\hat{Q}_\ell\), across iterations.
 Implementing this, however, might be somewhat too technical for empirical researchers.
-Moreover, this scalar summary is not model-independent, i.e., it is not universal to all complete data problems.
+Besides, this scalar summary is not model-independent.
+That is, this \(\theta\) is not universal to all complete data problems, while one of the advantages of iterative imputation is that the missing data problem and scientific problem are solved independently.
 Focusing on the convergence of outcome parameters may influence the iterative imputation procedure in the sense that the model of evaluation favors the model of interest.
-Additionally, one of the advantages of iterative imputation is that the missing data problem and scientific problem are solved independently.
 On these grounds, such a \(\theta\) can be considered insufficient too.
@@
 
-The downside to the current \(\theta\)s to evaluate is that they either focus on the univariate state space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest.
+To monitor non-convergence in iterative imputation algorithms, we may summarize the state space of the algorithm with several \(\theta\)s.
+The downside to the current \(\theta\)s is that they either focus on the univariate state space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest.
 Ideally, one would like to evaluate a model-independent \(\theta\) that summarizes the multivariate nature of the data.
 
-We propose \(\lambda_{1, \ell}\) as such a scalar summary.
-We define \(\lambda_{1, \ell}\) as the first eigenvalue of the variance-covariance matrix of the completed data.
+We propose \(\lambda_{1}\) as such a scalar summary.
+We define \(\lambda_{1}\) as the first eigenvalue of the variance-covariance matrix in the \(m\) completed datasets.
+The eigenvalues of a variance-covariance matrix \(S\) summarize the total set of covariances in the data.
 Let \(\lambda_{1, \ell} \geq \lambda_{2, \ell} \geq ...
-\geq \lambda_{j, \ell}\) be the eigenvalues of \(\Sigma_\ell\) in each of the \(m\) completed datasets \{\(y_{\rm obs}, \dot y_{\rm imp, \ell}\)\}.
-\(\lambda_{1, \ell}\) is measure that summarizes the covariances in the completed datasets.
-The first eigenvalue has the appealing property that is not dependent on the substantive model of interest.
-Eigenvalue decomposition is inspired by \citet{mack03}.\textbf{{[}ADD definition variance in first PCA component is equal to the first eigenvalue of the variance-covariance matrix{]}}
+\geq \lambda_{j, \ell}\) be the eigenvalues of \(S_\ell\) in each imputation \(\ell\).
+\(\lambda_{1,\ell}\) then summarizes the largest possible amount of covariance in the \(m\) completed datasets \{\(y_{\rm obs}, \dot y_{\rm imp, \ell}\)\}.
+By definition, each \(\lambda_{1, \ell}\)-value is equal to the variance of the first component that would be obtained by performing principal component analysis on \{\(y_{\rm obs}, \dot y_{\rm imp, \ell}\)\}.
+As scalar summary, \(\lambda_{1}\) has the appealing property that it is not dependent on the substantive model of interest, yet still summarizes the multivariate state space of the algorithm.
+\newline
+
+\noindent In this study, we consider each of the four \(\theta\)s that were discussed above.
+Namely, two univariate scalar summaries (chain mean and chain variance), a model-dependent multivariate summary \(\hat{Q}\), and the novel model-independent multivariate summary \(\lambda_1\).
 
@@
 
-Non-convergence diagnostics for MCMC algorithms typically identify problems in either one of the two requirements for convergence: mixing or stationarity.
-Non-stationarity within chains may be diagnosed with e.g., autocorrelation \citep[\(AC\);][]{scha97, gelm13}, numeric standard error \citep[`MC error';][]{gewe92}, or Raftery and Lewis's \citeyearpar{raft91} procedure to determine the effect of trending on the precision of estimates.
-A widely used diagnostic to monitor mixing between chains is the potential scale reduction factor \(\widehat{R}\) \citep[`Gelman-Rubin statistic';][]{gelm92}.
+There are many diagnostic tools to identify non-convergence in iterative (MCMC) algorithms \citep{broo98, elad06}.
+We consider only two of them that may be appropriate for imputation algorithms---one to monitor signs of non-mixing, and one for non-stationarity.
+As recommended by e.g.~\citet{cowl96} we will use the potential scale reduction factor \(\widehat{R}\) \citep[`Gelman-Rubin statistic';][]{gelm92} to evaluate mixing, and autocorrelation \citep[\(AC\);][]{scha97, gelm13} to diagnose trending.
 With a recently proposed adaptation, \(\widehat{R}\) might also serve to diagnose non-stationarity, but this has not yet been thoroughly investigated \citep{veht19}.
-Therefore, we use \(\widehat{R}\) \textbf{{[}WHICH?
-Explain that we'll use both{]}} and \(AC\) to evaluate mixing and stationarity separately, as recommended by e.g., \citet{cowl96}.
-
-\textbf{{[}Add: we don't use effective sample size and MC error because they assume independent samples \emph{within} chains.
-In MI, we only use the final estimate.
-Also, computing n\_eff assumes infinitely many iterations, which is fine in Bayesian analyses where T is often at least one thousand.
-But the default in iterative imp is often substantially lower.{]}}
+Therefore, we will evaluate the appropriateness of both \(\widehat{R}\) versions, in addition to \(AC\).
+Other methods are outside the scope of this study, because they e.g.~assume that values within chains represent independent samples, whereas MI only uses the final iteration to impute.
 
@@
 
-The potential scale reduction factor \(\widehat{R}\) was coined by \citet{gelm92}.
-An updated version of \(\widehat{R}\) has been proposed by Vehtari et al.~\citeyearpar[p.~5]{veht19}.
-\textbf{This version is better suited to detect non-convergence in the tails of distributions, by using three transformations on the scalar summary that it is applied to.
+In 2019, Vehtari et al.~proposed an updated version of the potential scale reduction factor \(\widehat{R}\), originally coined in 1992.
+The adapted version would be better suited to detect non-mixing in the tails of distributions, and even identify non-stationarity.
+This version uses three transformations on the scalar summary \(\theta\), before computing \(\widehat{R}\)-values.
 Namely, rank-normalization, folding, and localization.
-The Vehtari et al.~version of \(\widehat{R}\) may be suitable for iterative imputation.
-We therefore use their definition of the diagnostic.} Let \(m\) be the total number of chains, \(T\) the number of iterations per chain (where \(T\geq2\)), and \(\theta\) the scalar summary of interest.
+The definition of the diagnostic itself is equal to the original \(\widehat{R}\).
+We follow Vehtari et al.'s formulation (2019, p.~5).
+Let \(m\) be the total number of chains, \(T\) the number of iterations per chain (where \(T\geq2\)), and \(\theta\) the scalar summary of interest.
 For each chain (\(\ell = 1, 2, \dots, m\)), we estimate the variance of \(\theta\), and average these to obtain within-chain variance \(W\).
@@
 
-\noindent We then estimate between-chain variance \(B\) as the variance of the collection of average \(\theta\)s per chain. \textbf{{[}Note that this is not the usual B in MI{]}}
+\noindent We then estimate between-chain variance \(B\) as the variance of the collection of average \(\theta\)s per chain.
+\textbf{Note that \(B\) refers to the variance in \(\theta\), instead of the typical notation in MI where \(B\) denotes the variance between estimated quantities of scientific interest \(\hat{Q}_{\ell}\).}
 
@@
 Autocorrelation is defined as the correlation between two subsequent \(\theta\)-values within the same chain \citep[p.~147]{lync07}.
-In this study, we only consider \(AC\) at lag 1, i.e., the correlation between the \(t^{th}\) and \(t+1^{th}\) iteration of the same chain.
+In this study, we only consider \(AC\) at lag 1, i.e., the correlation between the \(t^{th}\) and \((t+1)^{th}\) iteration of the same chain.
 Following the same notation as for \(\widehat{R}\),
@@
 Divergence within imputation poses no threat to the convergence of the algorithm---it may even speed up convergence.
-\textbf{Complete stationarity is reached when \(AC=0\).} As non-convergence diagnostic, our interest is in positive \(AC\)-values.
+Complete stationarity is reached when \(AC=0\).
+As non-convergence diagnostic, our interest is in positive \(AC\)-values.
 
@@
 Because of its convergence to a distribution, the algorithm will show some signs of non-mixing and non-stationarity even in the most converged state.
-The aim is therefore to reach approximate convergence \citep{gelm13}.
+The aim, therefore, is to reach approximate convergence \citep{gelm13}.
 Upon approximate convergence, the imputation chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm (\(\widehat{R} > 1\)), and there is little dependency between subsequent iterations of imputation chains (\(AC > 0\)).
-\textbf{Approximate convergence is violated when the non-convergence diagnostics exceed a certain threshold.
-{[}or{]} In practice, non-convergence is therefore diagnosed when non-convergence diagnostics exeed a certain threshold.} The conventional thresholds to diagnose non-mixing are \(\widehat{R} > 1.2\) \citep{gelm92} or \(\widehat{R} > 1.1\) \citep{gelm13}.
+In practice, we diagnose non-convergence when approximate convergence is violated, i.e., when \(\widehat{R}\) and \(AC\) exceed a certain threshold.
+The conventional thresholds to diagnose non-mixing are \(\widehat{R} > 1.2\) \citep{gelm92} or \(\widehat{R} > 1.1\) \citep{gelm13}.
 \citet{veht19} proposed a much more stringent threshold of \(\widehat{R} > 1.01\).
@@
 \(AC\)-values that are significantly higher than zero indicate non-stationarity.
-\textbf{{[}think about one-sided test{]}}
 
@@
 
-Before we evaluate the performance of the non-convergence diagnostics \(\widehat{R}\) and \(AC\) quantitatively through simulation, we apply them to the example of pathological non-convergence by \citet{buur18} that we reproduced above.
-We want to be able to draw the same conclusions as we did from visually inspecting the two scenarios in Figure \ref{fig:non-conv}.
-The methods should at least indicate worse performance for the scenario with pathological non-convergence compared to the typically converged algorithm (i.e., higher \(\widehat{R}\)- and \(AC\)-values).
+Before we evaluate the performance of the non-convergence diagnostics \(\widehat{R}\) and \(AC\) quantitatively through simulation, we first assess their appropriateness qualitatively.
+We do this by applying them to the example of pathological non-convergence that we reproduced form \citet{buur18}.
+Ideally, the diagnostics are as informative as, or better than visual inspection of the traceplots.
+The methods should at least indicate worse performance (higher \(\widehat{R}\)- and \(AC\)-values) for the scenario with pathological non-convergence, compared to the typically converged algorithm.
 
-In Figure \ref{fig:diagnostics}A, the chain means from Figure \ref{fig:non-conv} are plotted again---now together.
-Panel B shows two versions of calculating \(AC\)s applied to the \(\theta\)s of panel A: the default calculation with \texttt{R} function \texttt{stats::acf()} \citep{R}, and manual calculation as the correlation between \(\theta\) in iteration \(t\) and \(\theta\) in iteration \(t+1\).
-Panel C shows the traditional computation of \(\widehat{R}\), and panel D shows \(\widehat{R}\) as computed by implementing \citet{veht19} 's recommendations.
+Each panel in Figure \ref{fig:diagnostics} depicts one method to identify non-convergence, applied to the two scenarios from Figure \ref{fig:non-conv}.
+Panel A consists of the two traceplots that may be evaluated through visual inspection.
+Panel B shows two versions of the \(AC\): the default calculation with \texttt{R} function \texttt{stats::acf()} \citep{R}, and manual calculation as the correlation between \(\theta\) in iteration \(t\) and \(\theta\) in iteration \(t+1\).
+Panel C displays the traditional computation of \(\widehat{R}\) conform \citet{gelm92}, and in panel D we see \(\widehat{R}\) as computed by implementing \citet{veht19} 's recommendations.
 \textbf{{[}Add labels for the two types of Rhat to the figure!{]}}

### 2020-05-10 3.Thesis/Writeup/2.Manuscript.Rmd
After reading through everything

@@
 There is no scientific consensus on how to evaluate the convergence of imputation algorithms [@taka17].
-Moreover, the behavoir of such algorithms under certain default imputation models (e.g., 'predictive mean matching') is an entirely open question [@murr18].
+Moreover, the behavior of such algorithms under certain default imputation models (e.g., 'predictive mean matching') is an entirely open question [@murr18].
 Therefore, algorithmic convergence should be monitored carefully, **but this is not straightforward**.
@@
 The values generated by the algorithm (e.g., imputed values) will vary even after convergence [@gelm13].
-Since MCMC algorithms do not reach a unique point at which convergence is established, **methods** may only identify signs of *non*-convergence [@hoff09].
+Since MCMC algorithms do not reach a unique point at which convergence is established, diagnostic methods may only identify signs of *non*-convergence [@hoff09].
 Several non-convergence diagnostics exist, but it is not known whether these are appropriate for iterative imputation algorithms.
@@
 We define several diagnostics and evaluate how these diagnostics could be appropriate for iterative imputation applications.
-We then adress the impact of inducing non-convergence in iterative imputation algorithms through model-based simulation in `R` [@R].
+We then address the impact of inducing non-convergence in iterative imputation algorithms through model-based simulation in `R` [@R].
 For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular `mice` package [@mice] in `R`.
@@
 And additionally, to evaluate the behavior and performance of several diagnostic methods to identify non-convergence.
-With that, we formulate an informed advice on when approximate convergence, necessary for valid inferences, may be safely concluded.
+With that, we formulate an informed advice on when it is safe to conclude that the algorithm is converged *enough* for valid inferences.
 We translate the results of the study into guidelines for applied researchers.
@@
 The data value of unit $i$ ($i = 1, 2, \dots, n$) on variable $j$ ($j = 1, 2, \dots, k$) may be either observed or missing.
-The number of units $i$ with at least one missing value, divided by the total number of units $n$, is called the missingness proportion $p_{\rm mis}$ in dataset $y$.
+The number of units $i$ with at least one missing value, divided by the total number of units $n$, is called the proportion of incomplete cases $p_{\rm inc}$ in dataset $y$.
 The collection of observed data values in $y$ is denoted by $y_{\rm obs}$; the missing part of $y$ is referred to as $y_{\rm mis}$.
@@
 
-\noindent We then estimate between-chain variance $B$ as the variance of the collection of average $\theta$s per chain.
-**Note that $B$ refers to the variance in $\theta$, instead of the typical notation in MI where $B$ denotes the variance between estimated quantities of scientific interest $\hat{Q}_{\ell}$.**
+\noindent We then estimate between-chain variance $B$ (note that we diverge from the typical notation in MI, where $B$ denotes the variance between the estimated quantities of scientific interest $\hat{Q}_{\ell}$).
+$B$ is defined as the variance of the collection of average $\theta$s per chain:
 
@@
 When we look at panel B, we conclude something weird.
-The $AC$-values calculated with the default function indicate equal performance for the typical convergence and the pathological non-convergence (up-to $t=5$), while there is obvious trending in the $\theta$s of the latter.
-Moreover, the best convergence (as indicated by the lowest $AC$-value) is observed at $t=2$, but looking at the chain means in panel A, there should be some signs of trending up-to iteration number seven.
+The $AC$-values calculated with the default function indicate equal performance for the typical convergence and the pathological non-convergence scenarios (up-to $t=5$), while there is obvious trending in the $\theta$s of the latter.
+Moreover, the best convergence (as indicated by the lowest $AC$-value) is observed at $t=2$ for both scenarios.
+However, if we look at the chain means of the non-convergence scenario, there should be signs of trending  up-to iteration number seven.
 After consulting the documentation on `stats::acf()`, we conclude that this $AC$ function is not suited for iterative imputation algorithms.
@@
 Therefore, we compute $AC$ manually, see the solid line.
-These $AC$ values indicate non-stationarity as expected.
-We therefore calculate $AC$ manually.
-
-**Describe panel C here and why it is not optimal: NB.
-In panel C: De grotere variantie overall zorgt er kennelijk voor dat $\widehat{R}$ groeit.
+These $AC$ values behave as expected from visual inspection.
+We will, therefore, only consider manually calculated $AC$-values as non-convergence diagnostic.
+
+**Describe panel C here and why it is not optimal: De grotere variantie overall zorgt er kennelijk voor dat $\widehat{R}$ groeit.
 De $\widehat{R}$ geeft de foutieve boodschap dat convergentie steeds slechter gaat.
-Verklaring: de overdispersie assumptie klopt niet als er trend is.]** The $\widehat{R}$-values in panel D do meet the requirements specified above.
-$\widehat{R}$ as computed conform @veht19 does indeed indicate less signs of non-convergence as the number of iterations goes up.
+Verklaring: de overdispersie assumptie klopt niet als er trend is.]** The $\widehat{R}$-values in panel D are conform our expectations.
+The adapted version of $\widehat{R}$ proposed by @veht19 does indeed indicate less signs of non-convergence as the number of iterations goes up.
 **[explain the dip in Rhat values at t=2.
 Namely, because we can only use 2 of the 3 tricks by @veht19 if the number of iterations is very low (t<4).
-That's why the $\widehat{R}$s are more similar to the traditional GR.]**
+That's why the $\widehat{R}$s are more similar to the traditional GR.]** \newline
 <!-- However, the superior performance of the typical convergence over the pathological non-convergence is less prominent, even flipped for $t<4$.  -->
 
-**In conclusion, we will use $\widehat{R}$ conform @veht19 and $AC$ computed manually as non-convergence diagnostics in the simulation study.
-The diagnostics will be applied to four $\theta$s of interest: chain means, chain variances, a quantity of scientific interest, and the first eigenvalue of the variance-covariance matrix.
-Results in eight methods to identify non-convergence**
+\noindent In short, both non-convergence diagnostics have a version that may be appropriate for iterative imputation algorithms.
+We will compute $\widehat{R}$ conform @veht19, and calculate $AC$ manually.
+The diagnostics will be applied to the four types of scalar summaries $\theta$ described above: chain means, chain variances, a quantity of scientific interest, and the first eigenvalue of the variance-covariance matrix.
+We evaluate the performance of the eight sets of diagnostic methods (two diagnostics; four $\theta$s) through simulation.
 
@@
 
-The aim of the simulation study is to determine the **impact of  non-convergence ["early stopping" en "proportion of complete cases"]** in iterative imputation algorithms.
-Specifically, we are interested in the bias and confidence-validity of estimates obtained from incomplete data with the MICE algortihm.
-**[REMOVE?] Subsequently we will evaluate how well $\widehat{R}$ and $AC$ perform in identifying the effects of non-convergence **[and performance of different thetas]**.
+The aim of the simulation study is to determine to what extend non-convergence affects iterative imputation algorithms.
+Specifically, we are interested in the validity of inferences drawn from incomplete data using the MICE algorithm.
+And, whether invalid inferences due to non-convergence may be detected using eight sets of diagnostic methods.
+We will conclude that the algorithm is sufficiently converged for practical purposes when the estimated quantities of scientific interest $\bar{Q}$ are unbiased and confidence-valid estimates of ${Q}$.
+In contrast to this, we defined *approximate* convergence as the most converged state that the algorithm can obtain.
+Continued iteration after reaching approximate convergence does not yield better mixing or stationarity (indicated by non-improving $\widehat{R}$- and $AC$-values).
+Continued iteration after obtaining valid inferences may lead to a more converged state of the algorithm, but not better estimates.
+
+<!-- is an unbiased and confidence-valid**[REMOVE?] Subsequently we will evaluate how well $\widehat{R}$ and $AC$ perform in identifying the effects of non-convergence **[and performance of different thetas]**.
 And finally, we will formulate an informed advice on the requirements to safely conclude sufficient convergence in practice.
-That is, we conclude that  convergence is sufficient when each estimate $\bar{Q}$ is an unbiased and confidence-valid estimate of the corresponding estimand $Q$.**
-
-To induce non-convergence, we consider two sets of simulation conditions: we vary the missingness proportion $p_{\rm mis}$ in the incomplete data, and we vary the number of iterations $T$ in the imputation algorithm.
-The missingness conditions are chosen to reflect the difficulty of the missingness problem.
-The underlying assumption is that low missingness proportions lead to quick algorithmic convergence, since there is a lot of information in the observed data.
-Higher missingness proportions should yield slower convergence.
-Unless, however, the fraction of missing information is so high that the random component in the imputation algorithm outweighs the information in the observed data.
-Then, convergence to a stable but highly variable state may be reached instantaneously.
-We expect that the incremental missingness proportions in our study will result in a corresponding increase in signs of non-convergence.
-
-The assumption inherent to the second set of simulation conditions---the number of iterations---is that terminating the imputation algorithm too early causes non-convergence.
-Generally, the algorithm will not reach convergence if $T=1$, because the imputed values in the first iteration (at $t=1$) depend on the initial values of the algorithm (which are sampled randomly from the set of observed datapoints).
-As the number of iterations increases, the imputation chains should become independent of the initial values, until the point at which adding an extra iteration does not lead to a more converged state.
-We assume that we can induce non-convergence at least in conditions where $T$ is smaller than the default number of iterations in `mice`, $T=5$.
+That is, we conclude that  convergence is sufficient when each estimate $\bar{Q}$ is an unbiased and confidence-valid estimate of the corresponding estimand $Q$.** -->
+
+To induce non-convergence, we consider two sets of simulation conditions: 'incompleteness' and 'early stopping'.
+Incompleteness refers to the  proportion of incomplete cases in dataset $y$, $p_{\rm inc}$.
+Early stopping is defined by the number of iterations in the imputation algorithm, $T$.
+
+<!-- **["early stopping" en "proportion of complete cases"]** :  -->
+
+<!-- we vary the missingness proportion $p_{\rm inc}$ in the incomplete data, and we vary the number of iterations $T$ in the imputation algorithm.  -->
+
+The incompleteness conditions are chosen to reflect the difficulty of the missingness problem.
+The underlying assumption is that a low proportion of incomplete cases $p_{\rm inc}$ leads to quick algorithmic convergence, since there is a lot of information in the observed data.
+Higher $p_{\rm inc}$ yields slower convergence---unless there is so little information in $y_{\rm obs}$ that this is outweighed by the random component in the imputation algorithm.
+Then, convergence to a stable but highly variable state may be reached instantly.
+**Still, we expect that the incompleteness conditions will result in more signs of non-convergence.**
+
+The assumption inherent to the second set of simulation conditions, early stopping, is that terminating the imputation algorithm too early causes non-convergence.
+Generally, the algorithm will not reach convergence if $T=1$, because the imputed values in the first iteration (at $t=1$) depend on the initial values of the algorithm [which are sampled randomly from the set of observed datapoints; @buur18].
+As the number of iterations increases, the imputation chains should become independent of the initial values, until approximate convergence is reached (i.e., when an additional iteration does not lead to a more converged state).
+We expect that we can induce non-convergence by early stopping in at least those conditions where $T$ is smaller than the default number of iterations in `mice`, $T=5$ [@mice].
 
@@
 1.
-We expect that simulation conditions with a high missingness proportion $p_{\rm mis}$ and a low number of iterations $T$ will more often result in biased, invalid estimates of the quantities of scientific interest, $Q$s.
+We expect that simulation conditions with a high proportion of incomplete cases $p_{\rm inc}$ and a low number of iterations $T$ will more often result in biased, invalid estimates of the quantities of scientific interest $Q$.
 
 2.
-We hypothesize that $\widehat{R}$ and $AC$ will correctly identify signs of non-convergence in those simulation conditions where the $\bar{Q}$s are *not* unbiased and confidence-valid estimates of the $Q$s.
+We hypothesize that $\widehat{R}$ and $AC$ will correctly identify signs of non-convergence in those simulation conditions where any $\bar{Q}$ is *not* an unbiased and confidence-valid estimate of $Q$.
 
@@
 Yet, scientific estimates were insensitive to continued iteration from $t=5$ onward [@lace07].
-We therefore suspect that $\widehat{R}$ may over-estimate signs of non-convergence in iterative imputation algorithms, compared to the validity of estimates.
+We, therefore, suspect that $\widehat{R}$ may over-estimate signs of non-convergence in iterative imputation algorithms, compared to the validity of estimates.
 In contrast to this, it may occur that signs of non-convergence are under-estimated by $\widehat{R}$, in exceptional cases where the initial values of the algorithm are not appropriately over-dispersed [@broo98, p.
@@
 In practice, we do not expect this to cause problems for identifying non-convergence with $\widehat{R}$.
-**[ADD limitation that other MI packages may have different way to obtain initial values!]**
 
@@
 where $\epsilon \sim \mathcal{N}(0, 100)$.
-This results in a complete set $y$, with observed values for all units $i$ on all variables $j$ (where $j =  Y, X_1, X_2, X_3$).
-From the complete set we obtain the true values of the scientific estimands $Q$.
+This results in a dataset $y$, where $y_{\rm mis} = \emptyset$.
+In other words, $y$ contains the complete set of observations for all units $i$ on all variables $j$ (where $j =  Y, X_1, X_2, X_3$).
+From the complete dataset we obtain the true values of the scientific estimands $Q$.
 
@@
 where $Y'$ is the expected value of the outcome.
-And finally, the proportion of variance explained by the set of predictors: coefficient of determination $r^2$ **(note that lower case $r$ is used to avoid confusion with non-convergence diagnostic $\widehat{R}$)**.
-These $Q$s are estimated by $\bar{Q}$ in each simulation condition.
-
-<!-- Each of these $Q$s is estimated with a corresponding $\bar{Q}$, after the complete dataset is first *amputed* (with a varying missingness proportion $p_{\rm mis}$) and subsequently imputed (with a varying number of iterations $T$).
+The final estimand is the proportion of variance explained by the set of predictors: coefficient of determination $r^2$ (note that lower case $r$ is used to avoid confusion with non-convergence diagnostic $\widehat{R}$).
+The scientific estimands are estimated by $\bar{Q}$ in each simulation condition.
+
+<!-- Each of these $Q$s is estimated with a corresponding $\bar{Q}$, after the complete dataset is first *amputed* (with a varying missingness proportion $p_{\rm inc}$) and subsequently imputed (with a varying number of iterations $T$).
 -->
@@
 
-The complete dataset is *amputed* in each simulation repetition to create five incomplete datasets $y$ with different missingness proportions $p_{\rm mis}$.
-The $p_{\rm mis}$ varies between 5 and 95%, in incremental steps ($p_{\rm mis}$ = .05, .25, .50, .75, .95).
-We consider all possible univariate and multivariate patterns of missingness, conform a 'missing completely at random' missingness mechanism [@rubin87].
-I.e., the probability to be missing is the same for all $N \times k$ cells in $y$, conditional on the missingness proportion $p_{\rm mis}$.
-This is performed with the 'mice' package [function `mice::ampute()`; @mice].
-
-The second set of simulation conditions is the number of iterations in the imputation algorithm.
-We vary the number of iterations between one and one hundred ($T = 1,2,.,100$).
-All imputation procedures are performed with Bayesian linear regression imputation, and five imputation chains ($m=5$).
-We perform this with the `mice()` function [@mice].
-The result is five sets of imputations per simulation condition, $y_{\rm imp, \ell}$.
-
-Subsequently, we obtain the completed data {$y_{\rm obs}, y_{\rm imp, \ell}$} by combining the imputed data with the observed data [function `mice::complete()`; @mice].
-Univariate quantities of scientific interest $Q$ are estimated directly from the completed data {$y_{\rm obs}, y_{\rm imp, \ell}$}.
-Estimates of multivariate $Q$s are obtained after employing multiple linear regression [function `stats::lm()`, @R].
-From the estimates in each imputation, $\hat{Q}_\ell$, we obtain the pooled estimate $\bar{Q}$.
-For univariate estimates we pool $\hat{Q}_\ell$ as average.
-Multivariate $\hat{Q}_\ell$ are pooled conform @vink14.
+In each simulation repetition, we *ampute* the complete dataset conform the 'incompleteness' conditions.
+The proportion of incomplete cases $p_{\rm inc}$ varies between 5% and 95%, in incremental steps ($p_{\rm inc}$ = .05, .25, .50, .75, .95).
+We consider all possible univariate and multivariate patterns of missingness, conform a 'missing completely at random' missingness mechanism [MCAR; @rubin87].
+I.e., the probability to be missing is the same for all $N \times k$ cells in $y$, conditional on $p_{\rm inc}$.
+We use the 'mice' package [function `mice::ampute()`; @mice] to obtain an incomplete dataset {$y_{\rm obs}, y_{\rm mis}$} for each of the five 'incompleteness' conditions.
+
+We impute the missing values in {$y_{\rm obs}, y_{\rm mis}$} according to the 'early stopping' conditions.
+That is, we vary the number of iterations in the imputation algorithm between one and one hundred ($T = 1,2,.,100$).
+All imputation procedures are performed using `mice` [function `mice()`; @mice], with Bayesian linear regression imputation, and five imputation chains ($m=5$).
+This results in five sets of imputations for each of the one hundred 'early stopping' conditions, $y_{\rm imp, \ell}$.
+
+Subsequently, we obtain the completed data per imputation, {$y_{\rm obs}, y_{\rm imp, \ell}$}, by combining the imputed data with the observed data [function `mice::complete()`; @mice].
+We estimate univariate $Q$s directly from the completed data, by averaging across imputations to obtain $\bar{Q}$.
+Estimates of multivariate $Q$s are obtained after performing multiple linear regression [function `stats::lm()`, @R].
+We pool the resulting estimates, $\hat{Q}_\ell$, into a single $\bar{Q}$ conform @vink14.
 
 <!-- In each simulation repetition, we *ampute* the complete dataset five times to obtain five different incomplete datasets, $y$s.
-The ratio between $y_{\rm obs}$ and $y_{\rm mis}$ in each $y$ depends on the missingness proportion: $p_{\rm mis} =.05,.25,.5,.75,.95$.
+The ratio between $y_{\rm obs}$ and $y_{\rm mis}$ in each $y$ depends on the missingness proportion: $p_{\rm inc} =.05,.25,.5,.75,.95$.
 We ampute the complete data with the `mice` package [function `mice::ampute()`; @mice].
 We consider all possible univariate and multivariate patterns of missingness, conform a 'missing completely at random' missingness mechanism [@rubin87].
-I.e., the probability to be missing is the same for all $N \times k$ cells in $y$, conditional on the missingness proportion $p_{\rm mis}$.
+I.e., the probability to be missing is the same for all $N \times k$ cells in $y$, conditional on the missingness proportion $p_{\rm inc}$.
 -->
@@
 -->
-<!-- <!-- Each repetition of the simulation thus starts from 1 complete dataset, then splits into 5 missingness conditions ($p_{\rm mis} =.05,.25,.5,.75,.95$), each of the amputed datasets is imputed 5 times ($m = 1, 2,..., 5$) for each iteration condition ($t = 1, 2,..., 100$).
+<!-- <!-- Each repetition of the simulation thus starts from 1 complete dataset, then splits into 5 missingness conditions ($p_{\rm inc} =.05,.25,.5,.75,.95$), each of the amputed datasets is imputed 5 times ($m = 1, 2,..., 5$) for each iteration condition ($t = 1, 2,..., 100$).
 From these $5\times5\times100$ imputations, we extract the $\theta$s to apply $\widehat{R}$ and $AC$ on (**i.e., chain means, chain variances, $\beta$s per imputation, and the first eigenvalue of the variance-covariance matrix per imputation**).
@@
 
-In total, we consider eight methods to identify non-convergence.
-Namely, we apply the two non-convergence diagnostics $\widehat{R}$ and $AC$ on four different $\theta$s.
+We consider eight sets of diagnostic methods to identify non-convergence, by applying $\widehat{R}$ and $AC$ on four different $\theta$s.
 Univariate $\theta$s are obtained from $y_{\rm imp, \ell}$.
-The multivariate $\theta$ $\lambda_{1,\ell}$ is obtained from {$y_{\rm obs}, y_{\rm imp, \ell}$}.
-The other multivariate $\theta$ is equal to the estimated regression coefficient $\hat{Q}_\ell$ in {$y_{\rm obs}, y_{\rm imp, \ell}$}.
-We apply the two non-convergence diagnostics to each $\theta$ by implementing @veht19 's recommended $\widehat{R}$ and we compute $AC$ manually.
+The model-independent multivariate $\theta$, $\lambda_{1}$, is obtained from {$y_{\rm obs}, y_{\rm imp, \ell}$}.
+The model-dependent multivariate $\theta$ is one of the quantities of scientific interest: the estimated regression coefficients in {$y_{\rm obs}, y_{\rm imp, \ell}$}.
+We apply the two non-convergence diagnostics on the $\theta$s by implementing the adapted version of $\widehat{R}$ and by manually programming the $AC$ function.
 
@@
 
-As recommended by @buur18 we evaluate the performance of the imputation algorithm with bias, coverage rate (CR), and confidence interval width (CIW).
-Bias is calculated for each $Q$.
-CR and CIW only for $Q = \beta$.
-We calculate bias as $\bar{Q} - Q$. CR is defined as the percentage of simulation repetitions in which the 95\% confidence interval (CI) around $\bar{Q}$ covers the true estimand $Q$. Let
+As recommended by @buur18, we evaluate the performance of the imputation algorithm with bias, coverage rate (CR), and confidence interval width (CIW).
+Bias is calculated for each $Q$, whereas CR and CIW are only obtained for $Q = \beta$.
+We calculate bias as $\bar{Q} - Q$.
+CR is defined as the percentage of simulation repetitions in which the 95\% confidence interval (CI) around $\bar{Q}$ covers the true estimand $Q$.
+Let
 $$\text{CI} = \bar{Q} \pm t_{(m-1)} \times \rm SE_{\bar{Q}},$$
@@
 
-We evaluate the methods to identify non-convergence against these performance measures.
+We evaluate the diagnostic methods to identify non-convergence against these performance measures.
 $\widehat{R}$ and $AC$ should identify non-convergence in simulation conditions, where $\bar{Q}$ is *not* an unbiased, confidence-valid estimate of $Q$.
@@
 More specifically, we are interested in the bias of $\bar{Q}$ for all $Q$s and the confidence-validity of $\bar{Q}$ for $Q=\beta$.
-Therefore, we evaluate these performance measures against the simulation conditions (the number of iterations $T$ and the missingness proportion $p_{\rm mis}$).
+Therefore, we evaluate these performance measures against the simulation conditions (the number of iterations $T$ and the missingness proportion $p_{\rm inc}$).
 We look at averages across simulation repetitions.
@@
 
-**[Add more info about figure legends and axes]**
-
-Our results show the performance of the MICE algorithm under different conditions of missingness and iteration length.
+Our results show the performance of the MICE algorithm under different conditions of 'incompleteness' and 'early stopping'.
 For reasons of brevity, we only discuss results for the worst-performing estimates in terms of bias.
@@
 Since there is just one estimate for $Q=r^2$ to consider, we **end up with** $Q=\mu_Y, \sigma_Y,r^2, \beta_1$.
-In the figures, we present all missingness conditions, but only **iteration length conditions** $1 \leq T\leq50$, since the results are more or less stable for conditions where $T \geq 30$.
-Full results are available from [github.com/gerkovink/shinyMice](https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy).
+In the figures, we present all 'incompleteness' conditions, but only 'early stopping' conditions where $1 \leq T\leq50$.
+That is, the results are more or less stable for conditions where $T \geq 30$.
+Full results are available from [github.com/gerkovink/shinyMice](https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/).
 
@@
 
-Figure \ref{fig:Qs} displays the influence of **inducing non-convergence ["early stopping" en "proportion of complete cases"]** on the estimated quantities of scientific interest, $Q$s.
+Figure \ref{fig:Qs} displays the influence of incompleteness and early stopping on the estimated quantities of scientific interest, $Q$s.
 **[Add panel labels, describe panels]** Simulation conditions that are impacted by non-convergence portray more extreme bias in the estimated $Q$s, and non-nominal coverages.
@@
 ### $Q=\mu$.
-The estimated univariate means seem unaffected by the number of iterations in the algorithm.
+The estimated univariate means seem unaffected by early stopping.
 This implies that approximately unbiased estimates may be obtained with as little as one iteration ($T \geq 1$).
-The magnitude of the bias in $\bar{Q}$ depends solely on the missingness proportion $p_{\rm mis}$, with missingness proportions of 75% and 95% leading to more extreme biases than other missingness proportions.
-Completely unbiased estimates are only obtained in conditions where $p_{\rm mis}\leq.50$.
-**This is curious because the missingness mechanism MCAR should yield unbiased univariate estimates without employing multiple imputation.** Note however, that the magnitude of the bias in these conditions is small: $\mu_Y$ is maximally under-estimated by 0.02 units, while the true value of $\mu_Y$ is 25.81 ($\sigma_Y = 11.32$).
+The bias in $\bar{Q}$ depends solely on the proportion of incomplete cases, with $p_{\rm inc}\geq.75$ leading to more extreme biases than other conditions.
+Completely unbiased estimates are only obtained in conditions where $p_{\rm inc}\leq.50$.
+**This is curious, since the MCAR missingness mechanism should yield unbiased univariate estimates without employing multiple imputation.** Note however, that the magnitude of the bias in these conditions is small: $\mu_Y$ is maximally under-estimated by 0.02 units, while the true value of $\mu_Y$ is 25.81 ($\sigma_Y = 11.32$).
 We, therefore, conclude that non-convergence does not substantially impact the validity of the inferences for $Q=\mu$.
@@
 ### $Q=\sigma$.
-The estimated standard deviations not affected substantially by **early stopping** either.
-While stable estimates are obtained in conditions with two iterations, the bias in the first iteration is actually *less* severe than in other **iteration conditions**.
+The estimated standard deviations are not substantially impacted by early stopping either.
+While the estimates for $T=1$ are not equal to other conditions, the bias is actually *less* severe than for $T\geq2$.
 Therefore, no iteration at all is needed to obtain approximately unbiased estimates for $Q=\sigma$.
-Completely unbiased estimates are only obtained in conditions where $p_{\rm mis}=.05$.
-Conditions where $p_{\rm mis}\geq.25$ are impacted by non-convergence in order of increasing missingness proportion.
-Similar to the bias for $Q=\mu$ the magnitude of the bias in $Q=\sigma$ is negligible.
+Completely unbiased estimates are only obtained in conditions where $p_{\rm inc}=.05$.
+The other incompleteness conditions are impacted by non-convergence in the order of increasing $p_{\rm inc}$.
+Similar to $Q=\mu$, the magnitude of the bias in $Q=\sigma$ is negligible.
 We therefore conclude that neither one of the univariate $Q$s is affected substantially by non-convergence.
@@
 Interestingly, the bias in conditions where $T=1$ is *less* severe than after continued iteration.
-For the least possible bias, we may use any number of iterations $T\geq1$, but only $p_{\rm mis} = 0.05$ results in completely unbiased $\bar{Q}$s.
+For the least possible bias, we may use any number of iterations $T\geq1$, but only $p_{\rm inc} = 0.05$ results in completely unbiased $\bar{Q}$s.
 -->
@@
 ### $Q=r^2$.
-The estimated coefficient of determination is clearly affected by early stopping.
-As expected, we observe worse performance in conditions with a lower number of iterations.
-The magnitude of the bias, however, depends on the missingness proportion---following the incremental order in $p_{\rm mis}$.
-**And so does** the number of iterations necessary to reach stable, approximately unbiased estimates.
-In conditions where $p_{\rm mis}=.05$, estimates are unbiased after one iteration.
-We observe approximately unbiased estimates when $T\geq 2,4, \text{and }5$, in conditions where $p_{\rm mis} = .25, .50, \text{and }.75$, respectively **[only report the worst?]**.
-The highest number of iterations necessary to reach approximate unbiasedness is seven.
-Completely unbiased estimates are only obtained in conditions where $p_{\rm mis}\leq.25$.
+In contrast to the univariate estimates, the estimated coefficient of determination is clearly affected by early stopping.
+That is, the bias in $Q$ does not only follow the incremental order of the incompleteness conditions, but also decreases with each additional iteration.
+The number of iterations necessary to reach stable, non-improving estimates depends on $p_{\rm inc}$.
+In conditions where $p_{\rm inc}=.05$, estimates are unbiased after one iteration.
+The highest number of iterations necessary to reach approximate unbiasedness is seven ($p_{\rm inc}=.95$; $T\geq7$).
+Completely unbiased estimates are only obtained in conditions where $p_{\rm inc}\leq.25$.
 This implies that even for biased estimates, $T\geq7$ would suffice to reach a stable solution.
@@
 Approximately unbiased estimates are observed for any missingness proportion.
-For $p_{\rm mis}\leq.25$, the minimum number of iterations is two ($T\geq2$),  for $p_{\rm mis}=.50$ it is $T\geq4$, and $p_{\rm mis}=.75$ requires $T\geq5$.
-The highest number of iterations needed to obtain valid inferences is seven ($p_{\rm mis}=.95$; $T\geq7$).
+For $p_{\rm inc}\leq.25$, the minimum number of iterations is two ($T\geq2$),  for $p_{\rm inc}=.50$ it is $T\geq4$, and $p_{\rm inc}=.75$ requires $T\geq5$.
+The highest number of iterations needed to obtain valid inferences is seven ($p_{\rm inc}=.95$; $T\geq7$).
 **[interpret sign of bias?
@@
 <!-- while the best performance occurs with less missingness and more iterations.
-Only conditions with the lowest missingness proportion ($p_{\rm mis}=0.05$) results in unbiased estimates for every number of iterations.
+Only conditions with the lowest missingness proportion ($p_{\rm inc}=0.05$) results in unbiased estimates for every number of iterations.
 In other conditions there is at least some bias irrespective of $T$.
@@
 For the estimated regression coefficients, we first consider bias, before discussing CR and CIW.
-As Figure \ref{fig:Qs} shows, the bias in $\bar{Q}$ is affected by both the missingness proportion and the number of iterations.
-Approximately unbiased estimates are observed after one to seven iterations, depending on $p_{\rm mis}$.Completely unbiased estimates are only obtained in conditions where $T\geq3$ and $p_{\rm mis}\leq .50$.
-
-Despite persistant bias in conditions where $p_{\rm mis}\geq.75$, coverage rates indicate that all **missingness conditions** have nominal coverages after just three iterations.
-Confidence-valid estimates may be obtained in any condition where $T\geq3$.
+As Figure \ref{fig:Qs}D shows, the bias in $\bar{Q}$ is affected by both the proportion of incomplete cases and the number of iterations.
+Similar to $Q=r^2$, we observe approximately unbiased estimates after at least one, and at most seven iterations, depending on $p_{\rm inc}$.
+Completely unbiased estimates are only obtained in conditions where $T\geq3$ and $p_{\rm inc}\leq .50$.
+Despite persistent bias in conditions where $p_{\rm inc}\geq.75$, the coverage rates are non-improving after after just three iterations, irrespective of $p_{\rm inc}$.
+We conclude that any condition where $T\geq3$ yields approximately nominal coverage rates, and thus confidence-valid estimates.
 <!-- , whereas we need $T\geq7$ to reach approximate unbiasedness. -->
-
-The nominal coverages are explained by the CIWs.
-Conform the theoretical foundation of MI, CIs are wider in conditions with higher missingness proportions (i.e., there is less information in the data, and thus more uncertainty due to missingness).
-The wider CIs around $\bar{Q}$ in conditions with higher missingness proportions **results in the inclusion of** the true value of $Q$ despite of bias.
-
-We conclude that approximately unbiased estimates may be obtained in conditions where $T\geq7$, whereas confidence-valid estimates require at most three iterations.
+The nominal coverages can be explained by the CIWs.
+The CIWs depicted in Figure \ref{fig:Qs}F are conform the theoretical foundation of MI: CIs are wider in conditions with higher proportions of incomplete cases (i.e., there is less information in the data, and thus more uncertainty due to missingness).
+Since conditions with higher $p_{\rm inc}$ result in both wider CIs and more severe bias, the true value of $Q$ may be included in the CI, despite of the bias in $\bar{Q}$.
+<!-- around $\bar{Q}$ in conditions with higher missingness proportions **results in the inclusion of** the true value of $Q$ despite of bias. -->
+We conclude that approximately unbiased estimates may be obtained in conditions where $T\geq7$, whereas confidence-valid estimates require at most three iterations. \newline
 
 <!-- confidence interval width (CIW), and coverage rate (CR) as performance measures.
-**[Split panels]** Unbiased estimates are obtained in conditions where $p_{\rm mis}\leq.50$ and $T\geq3$.
-Approximately confidence-valid estimates are obtained in any condition where $T\geq3$---despite of persistant bias in conditions where $p_{\rm mis}\geq.75$.
+**[Split panels]** Unbiased estimates are obtained in conditions where $p_{\rm inc}\leq.50$ and $T\geq3$.
+Approximately confidence-valid estimates are obtained in any condition where $T\geq3$---despite of persistent bias in conditions where $p_{\rm inc}\geq.75$.
 These approximately nominal coverages are explained by the CIWs: as expected, conditions with higher missingness proportions result in more uncertainty in the estimates (i.e., wider CIs), which diminish the effect of the bias.
@@
 **[add max T for other betas?]** -->
-<!-- One exception is the CIW of the most extreme condition ($p_{\rm mis}=.95$; $T=1$). -->
-
-### Summary
-
-These results demonstrate that the estimates of univariate scientific estimands $Q$ are not impacted by early stopping of the MICE algorithm, irrespective of the missingness proportion in $y$.
+<!-- One exception is the CIW of the most extreme condition ($p_{\rm inc}=.95$; $T=1$). -->
+
+\noindent These results demonstrate that the estimates of univariate scientific estimands $Q$ are not impacted by early stopping of the MICE algorithm or the proportion of incomplete cases in $y$.
 Unbiased estimates may be obtained after just one iteration.
-Multivariate estimates, by contrast, are affected by both the number of iterations and the missingness proportion.
-Completely unbiased estimates are only observed under low to moderate missingness ($p_{\rm mis}\leq.50$), after at most three iterations ($T\leq1,2,3$, depending on $p_{\rm mis}$).
-Approximately unbiased estimates are observed for any missingness condition, after at most seven iterations.
-This implies that the algorithm produces stable, unimproving estimates for any condition where $T\geq7$.
+Multivariate estimates, by contrast, are affected by both the number of iterations and the proportion of missing cases.
+Completely unbiased estimates are only obtained under low to moderate missingness ($p_{\rm inc}\leq.50$), after at most three iterations.
+We observe approximately unbiased estimates after at most seven iterations (for any $p_{\rm inc}$ considered).
+This implies that the algorithm produces stable, non-improving estimates when $T\geq7$.
 
@@
 This means that the algorithm has indeed not converged at $t=1$, and converges gradually as $t$ moves to $T$.
-The point at which an additional iteration does not lead to an improvement in the estimates depends on quantity of scientific interest $Q$ and the missingness proportion $p_{\rm mis}$.
+The point at which an additional iteration does not lead to an improvement in the estimates depends on quantity of scientific interest $Q$ and the missingness proportion $p_{\rm inc}$.
 The highest number of iterations that is necessary to reach this point across all simulation conditions is $T=7$.
@@
 This is especially pronounced in conditions with high missingness proportions: with 95% missingness we need at least seven iterations before an additional iteration does not lead to improved estimates.
-Conditions with a lower missingness proportion (e.g., $p_{\rm mis} = {0.5}$), produce *optimal* estimates with as many as two iterations.
-Approximately unbiased estimates are obtained for all conditions where $p_{\rm mis}\leq0.75$ and $T\leq6$.
+Conditions with a lower missingness proportion (e.g., $p_{\rm inc} = {0.5}$), produce *optimal* estimates with as many as two iterations.
+Approximately unbiased estimates are obtained for all conditions where $p_{\rm inc}\leq0.75$ and $T\leq6$.
 **[Add CIW and CR]** -->
@@
 
-Figure \ref{fig:rhats-acs} displays results for the two non-convergence diagnostics applied to the four $\theta$s.
-**[describe panels]** We evaluate the performance of these methods by establishing whether they correctly identify conditions in which $\bar{Q}$ is *not* an unbiased, confidence-valid estimate of $Q$.
-Biased estimates are observed in conditions where $T\leq7$.
-Non-nominal coverage rates are observed only when $T\leq3$.
-Additionally, there is persistant bias in conditions where $p_{\rm mis}\geq.75$, irrespective of the iteration length.
-**[appropriate methods thus identify these conditions]**
+Figure \ref{fig:rhats-acs} displays results for the eight sets of diagnostic methods.
+The two columns in the figure represent the  non-convergence diagnostics $\widehat{R}$ and $AC$; the four rows depict the four scalar summaries $\theta$ under consideration.
+**[describe panels??]** We evaluate the performance of the methods by establishing whether they correctly identify conditions in which $\bar{Q}$ is *not* an unbiased, confidence-valid estimate of $Q$.
+As we concluded in the last section, substantially biased estimates were only obtained in conditions where $T\leq6$, and we observed non-nominal coverage only when $T\leq3$.
+However, there was some persistent bias in conditions where $p_{\rm inc}\geq.75$, irrespective of the number of iterations.
+If the diagnostic methods are appropriate for MI, we will detect non-convergence in these conditions.
+
+<!-- are appropriate,  **[appropriate methods thus identify these conditions]** -->
 
@@
 The diagnostics should identify those simulation conditions that are negatively impacted by non-convergence.
-We expect the most signs of non-convergence in conditions where valid estimates are not obtained (high missingness proportions, $T<7$), and at least some signs of non-convergence in conditions where estimates are not approximately unbiased ($p_{\rm mis}=.95$, any number of iterations).
--->
-
-<!-- As we have observed, the affected conditions are $p_{\rm mis}=.95$ for all all $Q$s are impacted by the missingness proportions, but only the multivariate quantities of scientific interest are affected by the number of iterations.
--->
-
-Figure \ref{fig:rhats-acs} displays the non-convergence diagnostics $\widehat{R}$ and $AC$ for each of the four $\theta$s under consideration. **[Add panel labels and explain how to read the plots]**
+We expect the most signs of non-convergence in conditions where valid estimates are not obtained (high missingness proportions, $T<7$), and at least some signs of non-convergence in conditions where estimates are not approximately unbiased ($p_{\rm inc}=.95$, any number of iterations).
+-->
+
+<!-- As we have observed, the affected conditions are $p_{\rm inc}=.95$ for all all $Q$s are impacted by the missingness proportions, but only the multivariate quantities of scientific interest are affected by the number of iterations.
+-->
+
+<!-- Figure \ref{fig:rhats-acs} displays the non-convergence diagnostics $\widehat{R}$ and $AC$ for each of the four $\theta$s under consideration.
+**[Add panel labels and explain how to read the plots]**  -->
 
@@
 <!-- **Message: after one iteration ($T$=2), the algorithm is as well converged as the missingness proportion allows.
-Sooner even with lower missingness proportions---$T$=1 might be sufficient for $p_{\rm mis}$ = 5, 25, and 50.** Number of iterations has little influence on bias in $\hat{Q}_\ell$, only in first iteration there is some bias in missingness conditions of 25 and up.
+Sooner even with lower missingness proportions---$T$=1 might be sufficient for $p_{\rm inc}$ = 5, 25, and 50.** Number of iterations has little influence on bias in $\hat{Q}_\ell$, only in first iteration there is some bias in missingness conditions of 25 and up.
 There is a clear different between the missingness conditions.
@@
 For this $Q$, t=2 would be sufficient.
-And only for $p_{\rm mis}$ = 5 the $\bar{Q}$s are unbiased.
-All other $p_{\rm mis}$ conditions have bias irrespective of the number of iterations.
+And only for $p_{\rm inc}$ = 5 the $\bar{Q}$s are unbiased.
+All other $p_{\rm inc}$ conditions have bias irrespective of the number of iterations.
 -->
@@
 Using the threshold $\widehat{R}>1.2$, we diagnose non-convergence in conditions where $T\leq7$, whereas the threshold 1.1 is exceeded when $T\leq13$.
-The very strict threshold 1.01 is surpassed in all conditions $T\leq100$**, and therefore not informative in the current study**.
-With this method and $\theta$, we fail to identify simuation conditions with persistant bias due to high missingness proportions.
+The very strict threshold 1.01 is surpassed in all conditions $T\leq100$ (not shown), and therefore not informative under the current specifications.
+With this set of methods, we fail to identify simulation conditions with persistent bias due to high missingness proportions.
 
@@
 $AC$ indicates improving stationarity in conditions where $T\leq6$.
-The $AC$-values do not exceed the threshold **defined by statistically significant $AC$s**.
-Based on this threshold, we fail to not diagnose non-convergence in any condition.
+The $AC$-values do not exceed the threshold defined by statistical significance of the $AC$-values.
+Based on this threshold, we fail to diagnose non-convergence in any condition.
 
@@
 This does not correspond to the impact of non-convergence in the univariate $Q$s.
-**Compared to the multivariate $Q$s [add before that we evaluate the thetas against all Qs?], $\widehat{R}$ and $AC$ applied to the chain means fail to identify the non-convergence in conditions where $p_{\rm mis}=.95$.**  -->
+**Compared to the multivariate $Q$s [add before that we evaluate the thetas against all Qs?], $\widehat{R}$ and $AC$ applied to the chain means fail to identify the non-convergence in conditions where $p_{\rm inc}=.95$.**  -->
 
@@
 ### $\theta=$ chain variance.
-As Figure \ref{fig:rhats-acs} shows, the results of the two diagnostics applied to the chain variances are highly similar to those on the chain means.
+As panel C and D in Figure \ref{fig:rhats-acs} show, the results of this set of methods are highly similar to those with chain means as $\theta$.
 This holds for both $\widehat{R}$ and $AC$.
@@
 
-### $\theta=\hat{Q}_\ell$.
-The $\widehat{R}$-values show a similar trend across iterations as we observed for chain means and chain variances.
-Only conditions with very high missingness proportions ($p_{\rm mis}\geq.75$) diverge from the earlier observations.
+### $\theta=\hat{Q}$.
+The $\widehat{R}$-values in Figure \ref{fig:rhats-acs}E show a similar trend across iterations as we observed for chain means and chain variances.
+Only conditions with very high missingness proportions ($p_{\rm inc}\geq.75$) diverge from the earlier observations.
 The $\widehat{R}$-values in these conditions do not taper off after 30 to 40 iterations, but rather between 40 to 50 iterations.
-The highest number of iterations at which non-convergence is still diagnosed according to the thresholds are: 11 for $\widehat{R}>1.2$, 23 for $\widehat{R}>1.1$, and 100 for $\widehat{R}>1.01$ **[only report the worst?]**.
-
-The $AC$-values also follow earlier observations, with the exception of the same very high missingness proportions that **diverged** for $\widehat{R}$.
-To reach maximum stationarity, these conditions require not five but six or seven iterations.
+The highest number of iterations at which non-convergence is still diagnosed according to the thresholds are: 11 for $\widehat{R}>1.2$, 23 for $\widehat{R}>1.1$, and 100 for $\widehat{R}>1.01$ (not shown).
+
+The $AC$-values also follow earlier observations, with the same exception as for $\widehat{R}$ (when $p_{\rm inc}\geq.75$).
+To reach maximum stationarity, these conditions require at most seven iterations instead of five.
 Moreover, the $AC$-values in these conditions exceed the threshold to diagnose non-convergence.
-Interestingly, this only occurs after ten or even thrity iterations, while we would expect more signs of trending early in the iterations.
-**Since the $AC$-values are not improving or worsening after seven iterations, we conclude that non-convergence is incorrectly diagnosed based on the threshold.**
+Interestingly, this only occurs after ten or even thirty iterations, while we would expect more signs of trending early in the iterations.
+Since the $AC$-values are not improving or worsening when $T\geq7$, we conclude that non-convergence is incorrectly diagnosed based on the threshold.
 
@@
 The thresholds to diagnose non-convergence are overcome in different conditions, depending on the number of iterations and the missingness proportion.
-The most lenient threshold ($\widehat{R}>1.2$) is overcome at $T\geq9$ for $p_{\rm mis} = 0.05$, and $T\geq12$ for $p_{\rm mis} = 0.95$.
+The most lenient threshold ($\widehat{R}>1.2$) is overcome at $T\geq9$ for $p_{\rm inc} = 0.05$, and $T\geq12$ for $p_{\rm inc} = 0.95$.
 The conventional $\widehat{R}>1.1$ is surpassed at $T\geq12$ and $T\geq24$, respectively.
@@
 <!-- For the autocorrelations applied to the $\hat{Q}_\ell$s, we see an even stronger effect of the missingness proportions.
-Especially the conditions where $p_{\rm mis} = 0.75$ and $p_{\rm mis} = 0.95$ are identified as non-stationary.
+Especially the conditions where $p_{\rm inc} = 0.75$ and $p_{\rm inc} = 0.95$ are identified as non-stationary.
 The threshold used to evaluate whether $AC=0$, however, is only reached in conditions with at least ten iterations.
@@
 However, the critical value for diagnosing non-convergence with $AC$ decreases as a function of the number of iterations.
-The conclusion to identify non-stationarity in conditions where $T\geq10$ and $p_{\rm mis} = 0.95$ might therefore be spurious.
+The conclusion to identify non-stationarity in conditions where $T\geq10$ and $p_{\rm inc} = 0.95$ might therefore be spurious.
 -->
@@
 
-### $\theta=\lambda_{1, \ell}$.
+### $\theta=\lambda_{1}$.
 Once again, we observe $\widehat{R}$-values with a similar trend across iterations.
 For this $\theta$, the results differ only in the most extreme missingness condition.
-We diagnose non-convergence in conditions where $T\leq9$ according to $\widehat{R}>1.2$, and $T\leq19$ for $\widehat{R}>1.1$.
+When $p_{\rm inc}=.95$, we diagnose non-convergence at $T\leq9$ according to $\widehat{R}>1.2$, $T\leq19$ for $\widehat{R}>1.1$, and 100 for $\widehat{R}>1.01$ (not shown).
 The 'dip' in $\widehat{R}$-values at $T=3$ is even more pronounced than with other $\theta$s.
 Some $\widehat{R}$-values even overcome the threshold of 1.2.
-If we would terminate the algorithm after three iterations, we would incorrectly conclude sufficient convergence according to the $\widehat{R}>1.2$ threshold.
-
-$AC$-values are systematically higher for this $\theta$ than in the other results.
-The $AC$-values also taper off at a later point in the iterations, indicating improving stationarity up-to ten, or even thirty iterations, depending on the missingness proportion.
-Using the threashold, we diagnose non-convergence at some point for any missingness proportion.
+If we would terminate the algorithm at $T=3$, we would incorrectly conclude that the algorithm reached approximate convergence, according to this threshold.
+
+$AC$-values are systematically higher for this $\theta$ than for other scalar summaries.
+The $AC$-values also taper off at a later point in the iterations.
+This suggests improving stationarity up-to ten, or even thirty iterations, depending on the missingness proportion.
+Using the threshold, we diagnose non-convergence at some point in all incompleteness conditions.
+However, this occurs only after reasonably decreasing $AC$-values are obtained---again suggesting that non-convergence is incorrectly diagnosed.
+\newline
 
@@
 Similar to the results the the regression coefficients as $\theta$s, we see effects of both the missingness proportion and number of iterations in each simulation condition.
-The most signs of non-convergence are identified in conditions where $p_{\rm mis}$ is large and $T$ is small.
+The most signs of non-convergence are identified in conditions where $p_{\rm inc}$ is large and $T$ is small.
 -->
@@
 Using this cut-off we conveniently ignore the initial 'dip' at $T=3$ as well.
-The traditional threshold $\widehat{R} > 1.2$ is overcome in conditions where $p_{\rm mis} = .95$ and $T\geq10$, and for all other missingness proportions in conditions where $T\geq8$.
+The traditional threshold $\widehat{R} > 1.2$ is overcome in conditions where $p_{\rm inc} = .95$ and $T\geq10$, and for all other missingness proportions in conditions where $T\geq8$.
 With the $\widehat{R} > 1.1$ threshold, the number of iterations are $T\geq20$ and $T\geq16$, respectively.
@@
 <!-- The $AC$-values largely depend on the missingness proportion in each condition.
-Conditions where $p_{\rm mis}=.95$ show statistically significant autocorrelations for any number of iterations.
-The other extreme, conditions where $p_{\rm mis}=.05$, does not yield $AC$ values above the threshold.
+Conditions where $p_{\rm inc}=.95$ show statistically significant autocorrelations for any number of iterations.
+The other extreme, conditions where $p_{\rm inc}=.05$, does not yield $AC$ values above the threshold.
 The $AC$ values with this missingness proportion are equivalent for all conditions where $T\geq6$, which suggests that the algorithm is as stationary as it can be.
@@
 
-### Summary
-
-Overall, the methods to diagnose non-convergence perform as expected: they indicate more signs of non-convergence in conditions with worse performance in terms of bias and confidence-validity.
-Under low to moderate missingness, it does not seems to matter on which $\theta$ the diagnostics are applied.
-If we look at complete unbiasedness, however, we notice that univariate $\theta$s fail to diagnose the persistant bias in conditions where $p_{\rm mis}\geq.75$.
+\noindent Overall, the methods to diagnose non-convergence perform as expected: they indicate more signs of non-convergence in conditions with worse performance in terms of bias and confidence-validity.
+Under low to moderate missingness, it does not seems to matter on which $\theta$ the non-convergence diagnostics are applied.
+If we look at complete unbiasedness, however, we notice that univariate $\theta$s fail to diagnose the persistent bias in conditions where $p_{\rm inc}\geq.75$.
+
 The number of iterations necessary to obtain approximately unbiased, confidence-valid estimates corresponds to the $\widehat{R}$ threshold 1.2.
-The thresholds 1.1 and 1.01 seem to strict compared to the validity of the inferences.
-The $AC$ threshold defined by the statistical significance of the $AC$-values does not appear to be appropriate in the current set-up **[or iterative imputation in general??]**.
-Non-stationarity may be better diagnosed by evaluating the $AC$-values across iterations.
-If the values do not substantially decrease, stationarity may be concluded **[similar to the elbow heuristic in PCA]**.
+The thresholds 1.1 and 1.01 seem too strict compared to the validity of the inferences.
+
+The threshold to diagnose significant $AC$-values does not appear to be appropriate in at least the current set-up, and perhaps in iterative imputation in general.
+A better heuristic to diagnose non-stationarity with $AC$ may be through evaluation of the $AC$-values across iterations: If the values do not substantially decrease with $T$, approximate stationarity may be concluded.
+<!-- **[similar to the elbow heuristic in PCA]**. -->
 
@@
 
-# Conclusion
-
-H1: As hypothesized, biased, invalid estimates occur more often in simulation conditions with a higher missingness proportion and lower number of iterations.
+# Discussion
+
+Before drawing any further conclusions, let us first return to the hypotheses that we formulated for the simulation study.
+
+1.
+As hypothesized, biased, invalid estimates of quantities of scientific interest occurred more often in simulation conditions with a high proportion of incomplete cases $p_{\rm inc}$ and a low number of iterations $T$.
 However, this only holds for multivariate scientific estimands.
@@
 
-H2: As expected, we observe higher $\widehat{R}$ and $AC$-values in conditions where $\bar{Q}$s were biased, invalid estimates of $Q$s.
-
-H3: The hypothesis that the recommended thresholds would be too strict is only patially supported.
+2.
+Our results also support the hyothesis that $\widehat{R}$ and $AC$ would correctly identify signs of non-convergence.
+We observed higher $\widehat{R}$ and $AC$-values in conditions where $\bar{Q}$s were biased, invalid estimates of $Q$s.
+
+3.
+We hypothesized that the recommended thresholds to diagnose non-convergence with $\widehat{R}$ would be too stringent for iterative imputation applications.
+This hypothesis is only partially supported.
 $\widehat{R}$ indicated improving convergence until 30 to 40 iterations, while approximately unbiased, confidence-valid estimates were obtained after just seven iterations.
-This rougly corresponds to the threshold of 1.2.
-The more stringent trheshold 1.1 seems plausible compared to the decrease in $\widehat{R}$ values across iterations.
+This roughly corresponds to the threshold of 1.2.
+The more stringent threshold 1.1 seems plausible compared to the decrease in $\widehat{R}$ values across iterations.
 The most recent threshold 1.01 does indeed seem to over-estimate the signs of non-convergence.
 
-H4: The results of this simulation study do not support the hypothesis that ...
-**[insert hypothesis]**.
-We observed $AC$-values as high as theoretically possible $AC=$.
-The second part of the hypothesis is not refuted: the $AC$-values did decrease quicky as a function of $T$.
-
-H5: We did find some support for the final hypothesis. Multivariate $\theta$s indeed have superior performance under certain conditions (e.g., multivariate $Q$s and extreme missingness proportions).
-
-# Discussion
-
-We have shown that non-convergence in iteratiive imputation algorithms leads to biased, invalid estimates of multivariate quantities of scientific interest $\theta$.
-Since the current practice of visually inspecting univariate $\theta$s does not suffice, we applied non-convergence diagnostics $\widehat{R}$ and $AC$ to both univariate and multivariate $\theta$s.
-Signs of non-convergence due to early stopping are identified with each of the methods---i.e., by applying $\widehat{R}$ and $AC$ to any $\theta$.
-Bias due to high missingness proportions, is only identified with multivariate $\theta$s.
-Under moderate missingness ($p_{\rm mis}\leq.50$) $\widehat{R}$-values decreased substantially until 30 to 40 iterations, which implies that mixing in the algorithm improved with each additional iteration.
-$AC$-values only improved until about six iterations, suggesting minimal improvement in trending is obtained beyond $T=6$.
+4.
+The results of this simulation study do not support the hypothesis that high $AC$ values are implausible in iterative imputation algorithms with typical convergence.
+We expected that the randomness induced by the algorithm would effectively mitigate the risk of dependency within chains after a few iterations.
+In this study, however, we observed $AC$-values as high as theoretically possible ($AC=1$).
+The second part of the hypothesis is not refuted: the $AC$-values did decrease quickly as a function of $T$.
+
+5.
+Our final hypothesis was that multivariate $\theta$s would be better at detecting non-convergence than univariate $\theta$s.
+Multivariate $\theta$s did indeed show superior performance under certain conditions (e.g., multivariate $Q$s and extreme missingness proportions).
+
+## Conclusion
+
+We have shown that non-convergence in iterative imputation algorithms goes hand in hand with biased, invalid estimates---at least for the multivariate quantities of scientific interest considered in this study.
+Identifying non-convergence may be a crucial aspect of drawing valid statistical inferences from incomplete data.
+We concluded that the current practice of visually inspecting non-convergence through traceplots of univariate scalar summaries of the state space of the algorithm does not suffice.
+We, therefore, considered eight sets of diagnostic methods to identify non-convergence: two non-convergence diagnostics ($\widehat{R}$ and $AC$), two univariate $\theta$s, a model-dependent multivariate $\theta$, and a novel, model-independent multivariate $\theta$.
+Signs of non-convergence due to early stopping were identified with each of the methods.
+Bias due to high proportions of incomplete cases were only identified with multivariate $\theta$s.
+
 Since we obtained approximately unbiased, confidence valid estimates after at most seven iterations, we conclude that the $\widehat{R}$ threshold 1.2 is the most appropriate diagnostic cut-off.
 The threshold to diagnose non-stationarity with $AC$ does not seem to apply.
+However, some signs of non-convergence were detected in simulation conditions many more iterations.
+Under moderate missingness ($p_{\rm inc}\leq.50$) $\widehat{R}$-values decreased substantially until 30 to 40 iterations, which implies that mixing in the algorithm still improved with each additional iteration.
+$AC$-values only improved until about six iterations, suggesting minimal improvement in trending was obtained beyond $T=6$ in the current simulation set-up.
+
 The main finding of this study is that valid inferences may be obtained much quicker than approximate algorithmic convergence is reached.
@@
 
-Iterative imputation algorithms such as MICE have been known to yield valid results under severe missingness.
+Iterative imputation algorithms such as MICE are known to yield valid results under severe missingness.
 With this study, we have shown that valid inferences can also be obtained under a combination of severe missingness and early stopping.
 Based on our results, the traditional threshold of $\widehat{R}> 1.2$ would be appropriate for diagnosing non-convergence in iterative imputation algorithms.
-This is not in line with the recent recommendations by veht19 to lower the threshold to 1.01.
-The discrepancy between our conclusion and @veht19's may be explained by the nature of iterative imputation algorithms.
-In the limit, iterative imputation algorithms have the same character as other MCM algorithms, but in imputation procedures a part of the distribution is already determined by the observed data, whereas the entire distribution is unknown in many other MCMC applications.
-Because we combine a known distribution with an unknown distribution, valid estimates may be reached much sooner.
-Convergence may therefore be diagnosed at a less stringent threshold.
-**[HERE were the two asterixes]**
+This is, however, not in line with the recent recommendations by @veht19 to lower the threshold to 1.01.
+The discrepancy between our conclusion and Vehtari et al.
+may be explained by the nature of iterative imputation algorithms.
+In the limit, iterative imputation algorithms have the same characteristics as other MCMC algorithms.
+But, in imputation procedures, a part of the distribution is already determined by the observed data, whereas the entire distribution is unknown in many other MCMC applications.
+Since we combine a known distribution with an unknown distribution, valid estimates may be reached much sooner.
+Convergence may, therefore, be diagnosed at a less stringent threshold.
 
 The threshold to diagnose non-convergence with $AC$ does not seem appropriate at all in iterative imputation algorithms.
-A better diagnostic cut-off would be the number of iterations at which an additional iteration does not substantially decrease $AC$.
+A better diagnostic cut-off would be the number of iterations at which an additional iteration does not substantially decrease the $AC$-values.
 In practice, this implies that the default number of iterations in MICE is not sufficient.
 $AC$ can only be computed if $T\geq3$, while the current default in the `mice` package is five [@mice].
-Identifying a decrease in the $AC$-values across iterations (an 'elbow') requires more than three observations.
-We therefore suggest to increase the default number of iterations in MICE to 10.
-The computational cost of 5 additional iterations has become less of a burden since MICE was introduced in [-@mice].
+Identifying a reasonable decrease in the $AC$-values across iterations (an 'elbow') requires more than three observations.
+We, therefore, suggest to increase the default number of iterations in MICE to ten.
+The computational cost of five additional iterations has become less of a burden since MICE was introduced [-@mice].
 Moreover, an increased number of iterations would grant the opportunity to exclude the first iterations from evaluation with $\widehat{R}$.
@@
 
-Based on our results, we formulated several recommendations for empirical researchers who employ iterative imputation to draw inferences from incomplete data.
+Based on our results, we formulate several recommendations for empirical researchers who employ iterative imputation to draw inferences from incomplete data.
 We suggest that non-convergence may still be evaluated visually, but in addition to inspecting univariate summaries of the state space of the algorithm ($\theta$s, e.g., chain means), multivariate $\theta$s should be considered.
@@
 Use the threshold $\widehat{R}>1.2$ to diagnose non-mixing, and assess stationarity by plotting the autocorrelation over iterations.
-Keep iterating until the threshold for mixing is overcome, and until reasonably decreasing asympototic $AC$-values are obtained.
+Keep iterating until the threshold for mixing is overcome, and until reasonably decreasing asymptotic $AC$-values are obtained.
 At that point, inferences are unlikely to improve by continued iteration.
-
-5. Profit.
 
@@
 Proper performance under a 'missing completely at random' (MCAR) mechanism is a necessary condition for any missing data method.
-It doesn't, however, guarantee equal performance under different missingness mechanisms.
+It does not't, however, guarantee equal performance under different missingness mechanisms.
 Future research should determine the performance of $\widehat{R}$ and $AC$ under 'missing at random' and 'missing not at random' mechanisms.
 Another parameter to consider in future research is the choice of the imputation method.
-As @murr18 concluded, the behaviour of algorithms such as MICE under certain default imputation models is still an open research question.
-We have investigated the behaviour of MICE under only one type of model (Bayesian linear regression).
+As @murr18 concluded, the behavior of algorithms such as MICE under certain default imputation models is still an open research question.
+We have investigated the behavior of MICE under only one type of model (Bayesian linear regression).
 Different imputations models might converge more poorly.
-Moreover, the imputation model may even be mis-specified un purpose to induce different levels of non-convergence.
+Moreover, the imputation model may even be mis-specified on purpose to induce different levels of non-convergence.
 Qualitatively, we have shown that $\widehat{R}$ and $AC$ are appropriate under clear violation of convergence.
 Quantitatively, we have only demonstrated that they can identify signs of non-convergence under MCAR, when the imputation model is a correctly specified Bayesian linear regression model.
-\newline \newline
+**[ADD limitation that other MI packages may have different way to obtain initial values!]** \newline \newline
 
@@
 <!-- - Inferences are unbiased much quicker than the algorithm has converged.
-All things univariate are converged almost intantly after initiating the iterative algorithm with a valid imputation model.
+All things univariate are converged almost instantly after initiating the iterative algorithm with a valid imputation model.
 But less quick with multivariate parameters.
@@
 Univariate estimates seem robust against terminating the algorithm early: There is no clear effect of the number of iterations on the bias in estimated means and standard deviations.
-The bias in these $Q$s was only impacted by the missingness proporion $p_{\rm mis}$, with higher $p_{\rm mis}$ resulting in more bias.
-Bias in multivariate estimates depended on both $p_{\rm mis}$ and the number of iterations.
+The bias in these $Q$s was only impacted by the missingness proportion $p_{\rm inc}$, with higher $p_{\rm inc}$ resulting in more bias.
+Bias in multivariate estimates depended on both $p_{\rm inc}$ and the number of iterations.
 However, the only effect of the number of iterations was observed when the algorithm was terminated at seven iterations or earlier.
@@
 
-\noindent In short, we have shown that an iterative imputation algorithm can yield correct outcomes, even when a converged state has not yet formally been reached.
+\noindent In short, we have shown that iterative imputation algorithms can yield correct outcomes, even when a converged state has not yet formally been reached.
 Any further iterations would then burn computational resources without improving the statistical inferences.

### 2020-05-10 3.Thesis/Writeup/2.Manuscript.tex
After reading through everything

@@
 There is no scientific consensus on how to evaluate the convergence of imputation algorithms \citep{taka17}.
-Moreover, the behavoir of such algorithms under certain default imputation models (e.g., `predictive mean matching') is an entirely open question \citep{murr18}.
+Moreover, the behavior of such algorithms under certain default imputation models (e.g., `predictive mean matching') is an entirely open question \citep{murr18}.
 Therefore, algorithmic convergence should be monitored carefully, \textbf{but this is not straightforward}.
@@
 The values generated by the algorithm (e.g., imputed values) will vary even after convergence \citep{gelm13}.
-Since MCMC algorithms do not reach a unique point at which convergence is established, \textbf{methods} may only identify signs of \emph{non}-convergence \citep{hoff09}.
+Since MCMC algorithms do not reach a unique point at which convergence is established, diagnostic methods may only identify signs of \emph{non}-convergence \citep{hoff09}.
 Several non-convergence diagnostics exist, but it is not known whether these are appropriate for iterative imputation algorithms.
@@
 We define several diagnostics and evaluate how these diagnostics could be appropriate for iterative imputation applications.
-We then adress the impact of inducing non-convergence in iterative imputation algorithms through model-based simulation in \texttt{R} \citep{R}.
+We then address the impact of inducing non-convergence in iterative imputation algorithms through model-based simulation in \texttt{R} \citep{R}.
 For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular \texttt{mice} package \citep{mice} in \texttt{R}.
@@
 And additionally, to evaluate the behavior and performance of several diagnostic methods to identify non-convergence.
-With that, we formulate an informed advice on when approximate convergence, necessary for valid inferences, may be safely concluded.
+With that, we formulate an informed advice on when it is safe to conclude that the algorithm is converged \emph{enough} for valid inferences.
 We translate the results of the study into guidelines for applied researchers.
@@
 The data value of unit \(i\) (\(i = 1, 2, \dots, n\)) on variable \(j\) (\(j = 1, 2, \dots, k\)) may be either observed or missing.
-The number of units \(i\) with at least one missing value, divided by the total number of units \(n\), is called the missingness proportion \(p_{\rm mis}\) in dataset \(y\).
+The number of units \(i\) with at least one missing value, divided by the total number of units \(n\), is called the proportion of incomplete cases \(p_{\rm inc}\) in dataset \(y\).
 The collection of observed data values in \(y\) is denoted by \(y_{\rm obs}\); the missing part of \(y\) is referred to as \(y_{\rm mis}\).
@@
 
-\noindent We then estimate between-chain variance \(B\) as the variance of the collection of average \(\theta\)s per chain.
-\textbf{Note that \(B\) refers to the variance in \(\theta\), instead of the typical notation in MI where \(B\) denotes the variance between estimated quantities of scientific interest \(\hat{Q}_{\ell}\).}
+\noindent We then estimate between-chain variance \(B\) (note that we diverge from the typical notation in MI, where \(B\) denotes the variance between the estimated quantities of scientific interest \(\hat{Q}_{\ell}\)).
+\(B\) is defined as the variance of the collection of average \(\theta\)s per chain:
 
@@
 When we look at panel B, we conclude something weird.
-The \(AC\)-values calculated with the default function indicate equal performance for the typical convergence and the pathological non-convergence (up-to \(t=5\)), while there is obvious trending in the \(\theta\)s of the latter.
-Moreover, the best convergence (as indicated by the lowest \(AC\)-value) is observed at \(t=2\), but looking at the chain means in panel A, there should be some signs of trending up-to iteration number seven.
+The \(AC\)-values calculated with the default function indicate equal performance for the typical convergence and the pathological non-convergence scenarios (up-to \(t=5\)), while there is obvious trending in the \(\theta\)s of the latter.
+Moreover, the best convergence (as indicated by the lowest \(AC\)-value) is observed at \(t=2\) for both scenarios.
+However, if we look at the chain means of the non-convergence scenario, there should be signs of trending up-to iteration number seven.
 After consulting the documentation on \texttt{stats::acf()}, we conclude that this \(AC\) function is not suited for iterative imputation algorithms.
@@
 Therefore, we compute \(AC\) manually, see the solid line.
-These \(AC\) values indicate non-stationarity as expected.
-We therefore calculate \(AC\) manually.
+These \(AC\) values behave as expected from visual inspection.
+We will, therefore, only consider manually calculated \(AC\)-values as non-convergence diagnostic.
 
-\textbf{Describe panel C here and why it is not optimal: NB.
-In panel C: De grotere variantie overall zorgt er kennelijk voor dat \(\widehat{R}\) groeit.
+\textbf{Describe panel C here and why it is not optimal: De grotere variantie overall zorgt er kennelijk voor dat \(\widehat{R}\) groeit.
 De \(\widehat{R}\) geeft de foutieve boodschap dat convergentie steeds slechter gaat.
-Verklaring: de overdispersie assumptie klopt niet als er trend is.{]}} The \(\widehat{R}\)-values in panel D do meet the requirements specified above.
-\(\widehat{R}\) as computed conform \citet{veht19} does indeed indicate less signs of non-convergence as the number of iterations goes up.
+Verklaring: de overdispersie assumptie klopt niet als er trend is.{]}} The \(\widehat{R}\)-values in panel D are conform our expectations.
+The adapted version of \(\widehat{R}\) proposed by \citet{veht19} does indeed indicate less signs of non-convergence as the number of iterations goes up.
 \textbf{\citep[explain the dip in Rhat values at t=2.
 Namely, because we can only use 2 of the 3 tricks by][ if the number of iterations is very low (t\textless4).
-That's why the \(\widehat{R}\)s are more similar to the traditional GR.]{veht19}}
+That's why the \(\widehat{R}\)s are more similar to the traditional GR.]{veht19}} \newline
 
-\textbf{In conclusion, we will use \(\widehat{R}\) conform \citet{veht19} and \(AC\) computed manually as non-convergence diagnostics in the simulation study.
-The diagnostics will be applied to four \(\theta\)s of interest: chain means, chain variances, a quantity of scientific interest, and the first eigenvalue of the variance-covariance matrix.
-Results in eight methods to identify non-convergence}
+\noindent In short, both non-convergence diagnostics have a version that may be appropriate for iterative imputation algorithms.
+We will compute \(\widehat{R}\) conform \citet{veht19}, and calculate \(AC\) manually.
+The diagnostics will be applied to the four types of scalar summaries \(\theta\) described above: chain means, chain variances, a quantity of scientific interest, and the first eigenvalue of the variance-covariance matrix.
+We evaluate the performance of the eight sets of diagnostic methods (two diagnostics; four \(\theta\)s) through simulation.
 
@@
 
-The aim of the simulation study is to determine the \textbf{impact of non-convergence {[}``early stopping'' en ``proportion of complete cases''{]}} in iterative imputation algorithms.
-Specifically, we are interested in the bias and confidence-validity of estimates obtained from incomplete data with the MICE algortihm.
-\textbf{{[}REMOVE?{]} Subsequently we will evaluate how well \(\widehat{R}\) and \(AC\) perform in identifying the effects of non-convergence }{[}and performance of different thetas{]}\textbf{.
-And finally, we will formulate an informed advice on the requirements to safely conclude sufficient convergence in practice.
-That is, we conclude that convergence is sufficient when each estimate \(\bar{Q}\) is an unbiased and confidence-valid estimate of the corresponding estimand \(Q\).}
+The aim of the simulation study is to determine to what extend non-convergence affects iterative imputation algorithms.
+Specifically, we are interested in the validity of inferences drawn from incomplete data using the MICE algorithm.
+And, whether invalid inferences due to non-convergence may be detected using eight sets of diagnostic methods.
+We will conclude that the algorithm is sufficiently converged for practical purposes when the estimated quantities of scientific interest \(\bar{Q}\) are unbiased and confidence-valid estimates of \({Q}\).
+In contrast to this, we defined \emph{approximate} convergence as the most converged state that the algorithm can obtain.
+Continued iteration after reaching approximate convergence does not yield better mixing or stationarity (indicated by non-improving \(\widehat{R}\)- and \(AC\)-values).
+Continued iteration after obtaining valid inferences may lead to a more converged state of the algorithm, but not better estimates.
 
-To induce non-convergence, we consider two sets of simulation conditions: we vary the missingness proportion \(p_{\rm mis}\) in the incomplete data, and we vary the number of iterations \(T\) in the imputation algorithm.
-The missingness conditions are chosen to reflect the difficulty of the missingness problem.
-The underlying assumption is that low missingness proportions lead to quick algorithmic convergence, since there is a lot of information in the observed data.
-Higher missingness proportions should yield slower convergence.
-Unless, however, the fraction of missing information is so high that the random component in the imputation algorithm outweighs the information in the observed data.
-Then, convergence to a stable but highly variable state may be reached instantaneously.
-We expect that the incremental missingness proportions in our study will result in a corresponding increase in signs of non-convergence.
+To induce non-convergence, we consider two sets of simulation conditions: `incompleteness' and `early stopping'.
+Incompleteness refers to the proportion of incomplete cases in dataset \(y\), \(p_{\rm inc}\).
+Early stopping is defined by the number of iterations in the imputation algorithm, \(T\).
 
-The assumption inherent to the second set of simulation conditions---the number of iterations---is that terminating the imputation algorithm too early causes non-convergence.
-Generally, the algorithm will not reach convergence if \(T=1\), because the imputed values in the first iteration (at \(t=1\)) depend on the initial values of the algorithm (which are sampled randomly from the set of observed datapoints).
-As the number of iterations increases, the imputation chains should become independent of the initial values, until the point at which adding an extra iteration does not lead to a more converged state.
-We assume that we can induce non-convergence at least in conditions where \(T\) is smaller than the default number of iterations in \texttt{mice}, \(T=5\).
+The incompleteness conditions are chosen to reflect the difficulty of the missingness problem.
+The underlying assumption is that a low proportion of incomplete cases \(p_{\rm inc}\) leads to quick algorithmic convergence, since there is a lot of information in the observed data.
+Higher \(p_{\rm inc}\) yields slower convergence---unless there is so little information in \(y_{\rm obs}\) that this is outweighed by the random component in the imputation algorithm.
+Then, convergence to a stable but highly variable state may be reached instantly.
+\textbf{Still, we expect that the incompleteness conditions will result in more signs of non-convergence.}
+
+The assumption inherent to the second set of simulation conditions, early stopping, is that terminating the imputation algorithm too early causes non-convergence.
+Generally, the algorithm will not reach convergence if \(T=1\), because the imputed values in the first iteration (at \(t=1\)) depend on the initial values of the algorithm \citep[which are sampled randomly from the set of observed datapoints;][]{buur18}.
+As the number of iterations increases, the imputation chains should become independent of the initial values, until approximate convergence is reached (i.e., when an additional iteration does not lead to a more converged state).
+We expect that we can induce non-convergence by early stopping in at least those conditions where \(T\) is smaller than the default number of iterations in \texttt{mice}, \(T=5\) \citep{mice}.
 
@@
 \item
-  We expect that simulation conditions with a high missingness proportion \(p_{\rm mis}\) and a low number of iterations \(T\) will more often result in biased, invalid estimates of the quantities of scientific interest, \(Q\)s.
+  We expect that simulation conditions with a high proportion of incomplete cases \(p_{\rm inc}\) and a low number of iterations \(T\) will more often result in biased, invalid estimates of the quantities of scientific interest \(Q\).
 \item
-  We hypothesize that \(\widehat{R}\) and \(AC\) will correctly identify signs of non-convergence in those simulation conditions where the \(\bar{Q}\)s are \emph{not} unbiased and confidence-valid estimates of the \(Q\)s.
+  We hypothesize that \(\widehat{R}\) and \(AC\) will correctly identify signs of non-convergence in those simulation conditions where any \(\bar{Q}\) is \emph{not} an unbiased and confidence-valid estimate of \(Q\).
 \item
@@
 Yet, scientific estimates were insensitive to continued iteration from \(t=5\) onward \citep{lace07}.
-We therefore suspect that \(\widehat{R}\) may over-estimate signs of non-convergence in iterative imputation algorithms, compared to the validity of estimates.
+We, therefore, suspect that \(\widehat{R}\) may over-estimate signs of non-convergence in iterative imputation algorithms, compared to the validity of estimates.
 In contrast to this, it may occur that signs of non-convergence are under-estimated by \(\widehat{R}\), in exceptional cases where the initial values of the algorithm are not appropriately over-dispersed \citep[p.~437]{broo98}.
@@
 In practice, we do not expect this to cause problems for identifying non-convergence with \(\widehat{R}\).
-\textbf{{[}ADD limitation that other MI packages may have different way to obtain initial values!{]}}
 \item
@@
 where \(\epsilon \sim \mathcal{N}(0, 100)\).
-This results in a complete set \(y\), with observed values for all units \(i\) on all variables \(j\) (where \(j = Y, X_1, X_2, X_3\)).
-From the complete set we obtain the true values of the scientific estimands \(Q\).
+This results in a dataset \(y\), where \(y_{\rm mis} = \emptyset\).
+In other words, \(y\) contains the complete set of observations for all units \(i\) on all variables \(j\) (where \(j = Y, X_1, X_2, X_3\)).
+From the complete dataset we obtain the true values of the scientific estimands \(Q\).
 
@@
 where \(Y'\) is the expected value of the outcome.
-And finally, the proportion of variance explained by the set of predictors: coefficient of determination \(r^2\) \textbf{(note that lower case \(r\) is used to avoid confusion with non-convergence diagnostic \(\widehat{R}\))}.
-These \(Q\)s are estimated by \(\bar{Q}\) in each simulation condition.
+The final estimand is the proportion of variance explained by the set of predictors: coefficient of determination \(r^2\) (note that lower case \(r\) is used to avoid confusion with non-convergence diagnostic \(\widehat{R}\)).
+The scientific estimands are estimated by \(\bar{Q}\) in each simulation condition.
 
@@
 
-The complete dataset is \emph{amputed} in each simulation repetition to create five incomplete datasets \(y\) with different missingness proportions \(p_{\rm mis}\).
-The \(p_{\rm mis}\) varies between 5 and 95\%, in incremental steps (\(p_{\rm mis}\) = .05, .25, .50, .75, .95).
-We consider all possible univariate and multivariate patterns of missingness, conform a `missing completely at random' missingness mechanism \citep{rubin87}.
-I.e., the probability to be missing is the same for all \(N \times k\) cells in \(y\), conditional on the missingness proportion \(p_{\rm mis}\).
-This is performed with the `mice' package \citep[function \texttt{mice::ampute()};][]{mice}.
+In each simulation repetition, we \emph{ampute} the complete dataset conform the `incompleteness' conditions.
+The proportion of incomplete cases \(p_{\rm inc}\) varies between 5\% and 95\%, in incremental steps (\(p_{\rm inc}\) = .05, .25, .50, .75, .95).
+We consider all possible univariate and multivariate patterns of missingness, conform a `missing completely at random' missingness mechanism \citep[MCAR;][]{rubin87}.
+I.e., the probability to be missing is the same for all \(N \times k\) cells in \(y\), conditional on \(p_{\rm inc}\).
+We use the `mice' package \citep[function \texttt{mice::ampute()};][]{mice} to obtain an incomplete dataset \{\(y_{\rm obs}, y_{\rm mis}\)\} for each of the five `incompleteness' conditions.
 
-The second set of simulation conditions is the number of iterations in the imputation algorithm.
-We vary the number of iterations between one and one hundred (\(T = 1,2,.,100\)).
-All imputation procedures are performed with Bayesian linear regression imputation, and five imputation chains (\(m=5\)).
-We perform this with the \texttt{mice()} function \citep{mice}.
-The result is five sets of imputations per simulation condition, \(y_{\rm imp, \ell}\).
+We impute the missing values in \{\(y_{\rm obs}, y_{\rm mis}\)\} according to the `early stopping' conditions.
+That is, we vary the number of iterations in the imputation algorithm between one and one hundred (\(T = 1,2,.,100\)).
+All imputation procedures are performed using \texttt{mice} \citep[function \texttt{mice()};][]{mice}, with Bayesian linear regression imputation, and five imputation chains (\(m=5\)).
+This results in five sets of imputations for each of the one hundred `early stopping' conditions, \(y_{\rm imp, \ell}\).
 
-Subsequently, we obtain the completed data \{\(y_{\rm obs}, y_{\rm imp, \ell}\)\} by combining the imputed data with the observed data \citep[function \texttt{mice::complete()};][]{mice}.
-Univariate quantities of scientific interest \(Q\) are estimated directly from the completed data \{\(y_{\rm obs}, y_{\rm imp, \ell}\)\}.
-Estimates of multivariate \(Q\)s are obtained after employing multiple linear regression \citep[function \texttt{stats::lm()},][]{R}.
-From the estimates in each imputation, \(\hat{Q}_\ell\), we obtain the pooled estimate \(\bar{Q}\).
-For univariate estimates we pool \(\hat{Q}_\ell\) as average.
-Multivariate \(\hat{Q}_\ell\) are pooled conform \citet{vink14}.
+Subsequently, we obtain the completed data per imputation, \{\(y_{\rm obs}, y_{\rm imp, \ell}\)\}, by combining the imputed data with the observed data \citep[function \texttt{mice::complete()};][]{mice}.
+We estimate univariate \(Q\)s directly from the completed data, by averaging across imputations to obtain \(\bar{Q}\).
+Estimates of multivariate \(Q\)s are obtained after performing multiple linear regression \citep[function \texttt{stats::lm()},][]{R}.
+We pool the resulting estimates, \(\hat{Q}_\ell\), into a single \(\bar{Q}\) conform \citet{vink14}.
 
@@
 
-In total, we consider eight methods to identify non-convergence.
-Namely, we apply the two non-convergence diagnostics \(\widehat{R}\) and \(AC\) on four different \(\theta\)s.
+We consider eight sets of diagnostic methods to identify non-convergence, by applying \(\widehat{R}\) and \(AC\) on four different \(\theta\)s.
 Univariate \(\theta\)s are obtained from \(y_{\rm imp, \ell}\).
-The multivariate \(\theta\) \(\lambda_{1,\ell}\) is obtained from \{\(y_{\rm obs}, y_{\rm imp, \ell}\)\}.
-The other multivariate \(\theta\) is equal to the estimated regression coefficient \(\hat{Q}_\ell\) in \{\(y_{\rm obs}, y_{\rm imp, \ell}\)\}.
-We apply the two non-convergence diagnostics to each \(\theta\) by implementing \citet{veht19} 's recommended \(\widehat{R}\) and we compute \(AC\) manually.
+The model-independent multivariate \(\theta\), \(\lambda_{1}\), is obtained from \{\(y_{\rm obs}, y_{\rm imp, \ell}\)\}.
+The model-dependent multivariate \(\theta\) is one of the quantities of scientific interest: the estimated regression coefficients in \{\(y_{\rm obs}, y_{\rm imp, \ell}\)\}.
+We apply the two non-convergence diagnostics on the \(\theta\)s by implementing the adapted version of \(\widehat{R}\) and by manually programming the \(AC\) function.
 
@@
 
-As recommended by \citet{buur18} we evaluate the performance of the imputation algorithm with bias, coverage rate (CR), and confidence interval width (CIW).
-Bias is calculated for each \(Q\).
-CR and CIW only for \(Q = \beta\).
-We calculate bias as \(\bar{Q} - Q\). CR is defined as the percentage of simulation repetitions in which the 95\% confidence interval (CI) around \(\bar{Q}\) covers the true estimand \(Q\). Let
+As recommended by \citet{buur18}, we evaluate the performance of the imputation algorithm with bias, coverage rate (CR), and confidence interval width (CIW).
+Bias is calculated for each \(Q\), whereas CR and CIW are only obtained for \(Q = \beta\).
+We calculate bias as \(\bar{Q} - Q\).
+CR is defined as the percentage of simulation repetitions in which the 95\% confidence interval (CI) around \(\bar{Q}\) covers the true estimand \(Q\).
+Let
 \[\text{CI} = \bar{Q} \pm t_{(m-1)} \times \rm SE_{\bar{Q}},\]
@@
 
-We evaluate the methods to identify non-convergence against these performance measures.
+We evaluate the diagnostic methods to identify non-convergence against these performance measures.
 \(\widehat{R}\) and \(AC\) should identify non-convergence in simulation conditions, where \(\bar{Q}\) is \emph{not} an unbiased, confidence-valid estimate of \(Q\).
@@
 
-\textbf{{[}Add more info about figure legends and axes{]}}
-
-Our results show the performance of the MICE algorithm under different conditions of missingness and iteration length.
+Our results show the performance of the MICE algorithm under different conditions of `incompleteness' and `early stopping'.
 For reasons of brevity, we only discuss results for the worst-performing estimates in terms of bias.
@@
 Since there is just one estimate for \(Q=r^2\) to consider, we \textbf{end up with} \(Q=\mu_Y, \sigma_Y,r^2, \beta_1\).
-In the figures, we present all missingness conditions, but only \textbf{iteration length conditions} \(1 \leq T\leq50\), since the results are more or less stable for conditions where \(T \geq 30\).
-Full results are available from \href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy}{github.com/gerkovink/shinyMice}.
+In the figures, we present all `incompleteness' conditions, but only `early stopping' conditions where \(1 \leq T\leq50\).
+That is, the results are more or less stable for conditions where \(T \geq 30\).
+Full results are available from \href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/}{github.com/gerkovink/shinyMice}.
 
@@
 
-Figure \ref{fig:Qs} displays the influence of \textbf{inducing non-convergence {[}``early stopping'' en ``proportion of complete cases''{]}} on the estimated quantities of scientific interest, \(Q\)s.
+Figure \ref{fig:Qs} displays the influence of incompleteness and early stopping on the estimated quantities of scientific interest, \(Q\)s.
 \textbf{{[}Add panel labels, describe panels{]}} Simulation conditions that are impacted by non-convergence portray more extreme bias in the estimated \(Q\)s, and non-nominal coverages.
@@
 
-The estimated univariate means seem unaffected by the number of iterations in the algorithm.
+The estimated univariate means seem unaffected by early stopping.
 This implies that approximately unbiased estimates may be obtained with as little as one iteration (\(T \geq 1\)).
-The magnitude of the bias in \(\bar{Q}\) depends solely on the missingness proportion \(p_{\rm mis}\), with missingness proportions of 75\% and 95\% leading to more extreme biases than other missingness proportions.
-Completely unbiased estimates are only obtained in conditions where \(p_{\rm mis}\leq.50\).
-\textbf{This is curious because the missingness mechanism MCAR should yield unbiased univariate estimates without employing multiple imputation.} Note however, that the magnitude of the bias in these conditions is small: \(\mu_Y\) is maximally under-estimated by 0.02 units, while the true value of \(\mu_Y\) is 25.81 (\(\sigma_Y = 11.32\)).
+The bias in \(\bar{Q}\) depends solely on the proportion of incomplete cases, with \(p_{\rm inc}\geq.75\) leading to more extreme biases than other conditions.
+Completely unbiased estimates are only obtained in conditions where \(p_{\rm inc}\leq.50\).
+\textbf{This is curious, since the MCAR missingness mechanism should yield unbiased univariate estimates without employing multiple imputation.} Note however, that the magnitude of the bias in these conditions is small: \(\mu_Y\) is maximally under-estimated by 0.02 units, while the true value of \(\mu_Y\) is 25.81 (\(\sigma_Y = 11.32\)).
 We, therefore, conclude that non-convergence does not substantially impact the validity of the inferences for \(Q=\mu\).
@@
 
-The estimated standard deviations not affected substantially by \textbf{early stopping} either.
-While stable estimates are obtained in conditions with two iterations, the bias in the first iteration is actually \emph{less} severe than in other \textbf{iteration conditions}.
+The estimated standard deviations are not substantially impacted by early stopping either.
+While the estimates for \(T=1\) are not equal to other conditions, the bias is actually \emph{less} severe than for \(T\geq2\).
 Therefore, no iteration at all is needed to obtain approximately unbiased estimates for \(Q=\sigma\).
-Completely unbiased estimates are only obtained in conditions where \(p_{\rm mis}=.05\).
-Conditions where \(p_{\rm mis}\geq.25\) are impacted by non-convergence in order of increasing missingness proportion.
-Similar to the bias for \(Q=\mu\) the magnitude of the bias in \(Q=\sigma\) is negligible.
+Completely unbiased estimates are only obtained in conditions where \(p_{\rm inc}=.05\).
+The other incompleteness conditions are impacted by non-convergence in the order of increasing \(p_{\rm inc}\).
+Similar to \(Q=\mu\), the magnitude of the bias in \(Q=\sigma\) is negligible.
 We therefore conclude that neither one of the univariate \(Q\)s is affected substantially by non-convergence.
@@
 
-The estimated coefficient of determination is clearly affected by early stopping.
-As expected, we observe worse performance in conditions with a lower number of iterations.
-The magnitude of the bias, however, depends on the missingness proportion---following the incremental order in \(p_{\rm mis}\).
-\textbf{And so does} the number of iterations necessary to reach stable, approximately unbiased estimates.
-In conditions where \(p_{\rm mis}=.05\), estimates are unbiased after one iteration.
-We observe approximately unbiased estimates when \(T\geq 2,4, \text{and }5\), in conditions where \(p_{\rm mis} = .25, .50, \text{and }.75\), respectively \textbf{{[}only report the worst?{]}}.
-The highest number of iterations necessary to reach approximate unbiasedness is seven.
-Completely unbiased estimates are only obtained in conditions where \(p_{\rm mis}\leq.25\).
+In contrast to the univariate estimates, the estimated coefficient of determination is clearly affected by early stopping.
+That is, the bias in \(Q\) does not only follow the incremental order of the incompleteness conditions, but also decreases with each additional iteration.
+The number of iterations necessary to reach stable, non-improving estimates depends on \(p_{\rm inc}\).
+In conditions where \(p_{\rm inc}=.05\), estimates are unbiased after one iteration.
+The highest number of iterations necessary to reach approximate unbiasedness is seven (\(p_{\rm inc}=.95\); \(T\geq7\)).
+Completely unbiased estimates are only obtained in conditions where \(p_{\rm inc}\leq.25\).
 This implies that even for biased estimates, \(T\geq7\) would suffice to reach a stable solution.
@@
 For the estimated regression coefficients, we first consider bias, before discussing CR and CIW.
-As Figure \ref{fig:Qs} shows, the bias in \(\bar{Q}\) is affected by both the missingness proportion and the number of iterations.
-Approximately unbiased estimates are observed after one to seven iterations, depending on \(p_{\rm mis}\).Completely unbiased estimates are only obtained in conditions where \(T\geq3\) and \(p_{\rm mis}\leq .50\).
+As Figure \ref{fig:Qs}D shows, the bias in \(\bar{Q}\) is affected by both the proportion of incomplete cases and the number of iterations.
+Similar to \(Q=r^2\), we observe approximately unbiased estimates after at least one, and at most seven iterations, depending on \(p_{\rm inc}\).
+Completely unbiased estimates are only obtained in conditions where \(T\geq3\) and \(p_{\rm inc}\leq .50\).
+Despite persistent bias in conditions where \(p_{\rm inc}\geq.75\), the coverage rates are non-improving after after just three iterations, irrespective of \(p_{\rm inc}\).
+We conclude that any condition where \(T\geq3\) yields approximately nominal coverage rates, and thus confidence-valid estimates.
+The nominal coverages can be explained by the CIWs.
+The CIWs depicted in Figure \ref{fig:Qs}F are conform the theoretical foundation of MI: CIs are wider in conditions with higher proportions of incomplete cases (i.e., there is less information in the data, and thus more uncertainty due to missingness).
+Since conditions with higher \(p_{\rm inc}\) result in both wider CIs and more severe bias, the true value of \(Q\) may be included in the CI, despite of the bias in \(\bar{Q}\).
+We conclude that approximately unbiased estimates may be obtained in conditions where \(T\geq7\), whereas confidence-valid estimates require at most three iterations. \newline
 
-Despite persistant bias in conditions where \(p_{\rm mis}\geq.75\), coverage rates indicate that all \textbf{missingness conditions} have nominal coverages after just three iterations.
-Confidence-valid estimates may be obtained in any condition where \(T\geq3\).
-
-The nominal coverages are explained by the CIWs.
-Conform the theoretical foundation of MI, CIs are wider in conditions with higher missingness proportions (i.e., there is less information in the data, and thus more uncertainty due to missingness).
-The wider CIs around \(\bar{Q}\) in conditions with higher missingness proportions \textbf{results in the inclusion of} the true value of \(Q\) despite of bias.
-
-We conclude that approximately unbiased estimates may be obtained in conditions where \(T\geq7\), whereas confidence-valid estimates require at most three iterations.
-
-\hypertarget{summary}{%
-\subsubsection{Summary}\label{summary}}
-
-These results demonstrate that the estimates of univariate scientific estimands \(Q\) are not impacted by early stopping of the MICE algorithm, irrespective of the missingness proportion in \(y\).
+\noindent These results demonstrate that the estimates of univariate scientific estimands \(Q\) are not impacted by early stopping of the MICE algorithm or the proportion of incomplete cases in \(y\).
 Unbiased estimates may be obtained after just one iteration.
-Multivariate estimates, by contrast, are affected by both the number of iterations and the missingness proportion.
-Completely unbiased estimates are only observed under low to moderate missingness (\(p_{\rm mis}\leq.50\)), after at most three iterations (\(T\leq1,2,3\), depending on \(p_{\rm mis}\)).
-Approximately unbiased estimates are observed for any missingness condition, after at most seven iterations.
-This implies that the algorithm produces stable, unimproving estimates for any condition where \(T\geq7\).
+Multivariate estimates, by contrast, are affected by both the number of iterations and the proportion of missing cases.
+Completely unbiased estimates are only obtained under low to moderate missingness (\(p_{\rm inc}\leq.50\)), after at most three iterations.
+We observe approximately unbiased estimates after at most seven iterations (for any \(p_{\rm inc}\) considered).
+This implies that the algorithm produces stable, non-improving estimates when \(T\geq7\).
 
@@
 
-Figure \ref{fig:rhats-acs} displays results for the two non-convergence diagnostics applied to the four \(\theta\)s.
-\textbf{{[}describe panels{]}} We evaluate the performance of these methods by establishing whether they correctly identify conditions in which \(\bar{Q}\) is \emph{not} an unbiased, confidence-valid estimate of \(Q\).
-Biased estimates are observed in conditions where \(T\leq7\).
-Non-nominal coverage rates are observed only when \(T\leq3\).
-Additionally, there is persistant bias in conditions where \(p_{\rm mis}\geq.75\), irrespective of the iteration length.
-\textbf{{[}appropriate methods thus identify these conditions{]}}
-
-Figure \ref{fig:rhats-acs} displays the non-convergence diagnostics \(\widehat{R}\) and \(AC\) for each of the four \(\theta\)s under consideration.
-\textbf{{[}Add panel labels and explain how to read the plots{]}}
+Figure \ref{fig:rhats-acs} displays results for the eight sets of diagnostic methods.
+The two columns in the figure represent the non-convergence diagnostics \(\widehat{R}\) and \(AC\); the four rows depict the four scalar summaries \(\theta\) under consideration.
+\textbf{{[}describe panels??{]}} We evaluate the performance of the methods by establishing whether they correctly identify conditions in which \(\bar{Q}\) is \emph{not} an unbiased, confidence-valid estimate of \(Q\).
+As we concluded in the last section, substantially biased estimates were only obtained in conditions where \(T\leq6\), and we observed non-nominal coverage only when \(T\leq3\).
+However, there was some persistent bias in conditions where \(p_{\rm inc}\geq.75\), irrespective of the number of iterations.
+If the diagnostic methods are appropriate for MI, we will detect non-convergence in these conditions.
 
@@
 Using the threshold \(\widehat{R}>1.2\), we diagnose non-convergence in conditions where \(T\leq7\), whereas the threshold 1.1 is exceeded when \(T\leq13\).
-The very strict threshold 1.01 is surpassed in all conditions \(T\leq100\)\textbf{, and therefore not informative in the current study}.
-With this method and \(\theta\), we fail to identify simuation conditions with persistant bias due to high missingness proportions.
+The very strict threshold 1.01 is surpassed in all conditions \(T\leq100\) (not shown), and therefore not informative under the current specifications.
+With this set of methods, we fail to identify simulation conditions with persistent bias due to high missingness proportions.
 
@@
 \(AC\) indicates improving stationarity in conditions where \(T\leq6\).
-The \(AC\)-values do not exceed the threshold \textbf{defined by statistically significant \(AC\)s}.
-Based on this threshold, we fail to not diagnose non-convergence in any condition.
+The \(AC\)-values do not exceed the threshold defined by statistical significance of the \(AC\)-values.
+Based on this threshold, we fail to diagnose non-convergence in any condition.
 
@@
 
-As Figure \ref{fig:rhats-acs} shows, the results of the two diagnostics applied to the chain variances are highly similar to those on the chain means.
+As panel C and D in Figure \ref{fig:rhats-acs} show, the results of this set of methods are highly similar to those with chain means as \(\theta\).
 This holds for both \(\widehat{R}\) and \(AC\).
@@
 
-\hypertarget{thetahatq_ell.}{%
-\subsubsection{\texorpdfstring{\(\theta=\hat{Q}_\ell\).}{\textbackslash theta=\textbackslash hat\{Q\}\_\textbackslash ell.}}\label{thetahatq_ell.}}
+\hypertarget{thetahatq.}{%
+\subsubsection{\texorpdfstring{\(\theta=\hat{Q}\).}{\textbackslash theta=\textbackslash hat\{Q\}.}}\label{thetahatq.}}
 
-The \(\widehat{R}\)-values show a similar trend across iterations as we observed for chain means and chain variances.
-Only conditions with very high missingness proportions (\(p_{\rm mis}\geq.75\)) diverge from the earlier observations.
+The \(\widehat{R}\)-values in Figure \ref{fig:rhats-acs}E show a similar trend across iterations as we observed for chain means and chain variances.
+Only conditions with very high missingness proportions (\(p_{\rm inc}\geq.75\)) diverge from the earlier observations.
 The \(\widehat{R}\)-values in these conditions do not taper off after 30 to 40 iterations, but rather between 40 to 50 iterations.
-The highest number of iterations at which non-convergence is still diagnosed according to the thresholds are: 11 for \(\widehat{R}>1.2\), 23 for \(\widehat{R}>1.1\), and 100 for \(\widehat{R}>1.01\) \textbf{{[}only report the worst?{]}}.
+The highest number of iterations at which non-convergence is still diagnosed according to the thresholds are: 11 for \(\widehat{R}>1.2\), 23 for \(\widehat{R}>1.1\), and 100 for \(\widehat{R}>1.01\) (not shown).
 
-The \(AC\)-values also follow earlier observations, with the exception of the same very high missingness proportions that \textbf{diverged} for \(\widehat{R}\).
-To reach maximum stationarity, these conditions require not five but six or seven iterations.
+The \(AC\)-values also follow earlier observations, with the same exception as for \(\widehat{R}\) (when \(p_{\rm inc}\geq.75\)).
+To reach maximum stationarity, these conditions require at most seven iterations instead of five.
 Moreover, the \(AC\)-values in these conditions exceed the threshold to diagnose non-convergence.
-Interestingly, this only occurs after ten or even thrity iterations, while we would expect more signs of trending early in the iterations.
-\textbf{Since the \(AC\)-values are not improving or worsening after seven iterations, we conclude that non-convergence is incorrectly diagnosed based on the threshold.}
+Interestingly, this only occurs after ten or even thirty iterations, while we would expect more signs of trending early in the iterations.
+Since the \(AC\)-values are not improving or worsening when \(T\geq7\), we conclude that non-convergence is incorrectly diagnosed based on the threshold.
 
-\hypertarget{thetalambda_1-ell.}{%
-\subsubsection{\texorpdfstring{\(\theta=\lambda_{1, \ell}\).}{\textbackslash theta=\textbackslash lambda\_\{1, \textbackslash ell\}.}}\label{thetalambda_1-ell.}}
+\hypertarget{thetalambda_1.}{%
+\subsubsection{\texorpdfstring{\(\theta=\lambda_{1}\).}{\textbackslash theta=\textbackslash lambda\_\{1\}.}}\label{thetalambda_1.}}
 
@@
 For this \(\theta\), the results differ only in the most extreme missingness condition.
-We diagnose non-convergence in conditions where \(T\leq9\) according to \(\widehat{R}>1.2\), and \(T\leq19\) for \(\widehat{R}>1.1\).
+When \(p_{\rm inc}=.95\), we diagnose non-convergence at \(T\leq9\) according to \(\widehat{R}>1.2\), \(T\leq19\) for \(\widehat{R}>1.1\), and 100 for \(\widehat{R}>1.01\) (not shown).
 The `dip' in \(\widehat{R}\)-values at \(T=3\) is even more pronounced than with other \(\theta\)s.
 Some \(\widehat{R}\)-values even overcome the threshold of 1.2.
-If we would terminate the algorithm after three iterations, we would incorrectly conclude sufficient convergence according to the \(\widehat{R}>1.2\) threshold.
+If we would terminate the algorithm at \(T=3\), we would incorrectly conclude that the algorithm reached approximate convergence, according to this threshold.
 
-\(AC\)-values are systematically higher for this \(\theta\) than in the other results.
-The \(AC\)-values also taper off at a later point in the iterations, indicating improving stationarity up-to ten, or even thirty iterations, depending on the missingness proportion.
-Using the threashold, we diagnose non-convergence at some point for any missingness proportion.
+\(AC\)-values are systematically higher for this \(\theta\) than for other scalar summaries.
+The \(AC\)-values also taper off at a later point in the iterations.
+This suggests improving stationarity up-to ten, or even thirty iterations, depending on the missingness proportion.
+Using the threshold, we diagnose non-convergence at some point in all incompleteness conditions.
+However, this occurs only after reasonably decreasing \(AC\)-values are obtained---again suggesting that non-convergence is incorrectly diagnosed.
+\newline
 
-\hypertarget{summary-1}{%
-\subsubsection{Summary}\label{summary-1}}
+\noindent Overall, the methods to diagnose non-convergence perform as expected: they indicate more signs of non-convergence in conditions with worse performance in terms of bias and confidence-validity.
+Under low to moderate missingness, it does not seems to matter on which \(\theta\) the non-convergence diagnostics are applied.
+If we look at complete unbiasedness, however, we notice that univariate \(\theta\)s fail to diagnose the persistent bias in conditions where \(p_{\rm inc}\geq.75\).
 
-Overall, the methods to diagnose non-convergence perform as expected: they indicate more signs of non-convergence in conditions with worse performance in terms of bias and confidence-validity.
-Under low to moderate missingness, it does not seems to matter on which \(\theta\) the diagnostics are applied.
-If we look at complete unbiasedness, however, we notice that univariate \(\theta\)s fail to diagnose the persistant bias in conditions where \(p_{\rm mis}\geq.75\).
 The number of iterations necessary to obtain approximately unbiased, confidence-valid estimates corresponds to the \(\widehat{R}\) threshold 1.2.
-The thresholds 1.1 and 1.01 seem to strict compared to the validity of the inferences.
-The \(AC\) threshold defined by the statistical significance of the \(AC\)-values does not appear to be appropriate in the current set-up \textbf{{[}or iterative imputation in general??{]}}.
-Non-stationarity may be better diagnosed by evaluating the \(AC\)-values across iterations.
-If the values do not substantially decrease, stationarity may be concluded \textbf{{[}similar to the elbow heuristic in PCA{]}}.
+The thresholds 1.1 and 1.01 seem too strict compared to the validity of the inferences.
 
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
-
-H1: As hypothesized, biased, invalid estimates occur more often in simulation conditions with a higher missingness proportion and lower number of iterations.
-However, this only holds for multivariate scientific estimands.
-Univariate \(Q\)s are not substantially impacted by early stopping.
-
-H2: As expected, we observe higher \(\widehat{R}\) and \(AC\)-values in conditions where \(\bar{Q}\)s were biased, invalid estimates of \(Q\)s.
-
-H3: The hypothesis that the recommended thresholds would be too strict is only patially supported.
-\(\widehat{R}\) indicated improving convergence until 30 to 40 iterations, while approximately unbiased, confidence-valid estimates were obtained after just seven iterations.
-This rougly corresponds to the threshold of 1.2.
-The more stringent trheshold 1.1 seems plausible compared to the decrease in \(\widehat{R}\) values across iterations.
-The most recent threshold 1.01 does indeed seem to over-estimate the signs of non-convergence.
-
-H4: The results of this simulation study do not support the hypothesis that \ldots{} \textbf{{[}insert hypothesis{]}}.
-We observed \(AC\)-values as high as theoretically possible \(AC=\).
-The second part of the hypothesis is not refuted: the \(AC\)-values did decrease quicky as a function of \(T\).
-
-H5: We did find some support for the final hypothesis.
-Multivariate \(\theta\)s indeed have superior performance under certain conditions (e.g., multivariate \(Q\)s and extreme missingness proportions).
+The threshold to diagnose significant \(AC\)-values does not appear to be appropriate in at least the current set-up, and perhaps in iterative imputation in general.
+A better heuristic to diagnose non-stationarity with \(AC\) may be through evaluation of the \(AC\)-values across iterations: If the values do not substantially decrease with \(T\), approximate stationarity may be concluded.
 
@@
 
-We have shown that non-convergence in iteratiive imputation algorithms leads to biased, invalid estimates of multivariate quantities of scientific interest \(\theta\).
-Since the current practice of visually inspecting univariate \(\theta\)s does not suffice, we applied non-convergence diagnostics \(\widehat{R}\) and \(AC\) to both univariate and multivariate \(\theta\)s.
-Signs of non-convergence due to early stopping are identified with each of the methods---i.e., by applying \(\widehat{R}\) and \(AC\) to any \(\theta\).
-Bias due to high missingness proportions, is only identified with multivariate \(\theta\)s.
-Under moderate missingness (\(p_{\rm mis}\leq.50\)) \(\widehat{R}\)-values decreased substantially until 30 to 40 iterations, which implies that mixing in the algorithm improved with each additional iteration.
-\(AC\)-values only improved until about six iterations, suggesting minimal improvement in trending is obtained beyond \(T=6\).
+Before drawing any further conclusions, let us first return to the hypotheses that we formulated for the simulation study.
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\item
+  As hypothesized, biased, invalid estimates of quantities of scientific interest occurred more often in simulation conditions with a high proportion of incomplete cases \(p_{\rm inc}\) and a low number of iterations \(T\).
+However, this only holds for multivariate scientific estimands.
+Univariate \(Q\)s are not substantially impacted by early stopping.
+\item
+  Our results also support the hyothesis that \(\widehat{R}\) and \(AC\) would correctly identify signs of non-convergence.
+We observed higher \(\widehat{R}\) and \(AC\)-values in conditions where \(\bar{Q}\)s were biased, invalid estimates of \(Q\)s.
+\item
+  We hypothesized that the recommended thresholds to diagnose non-convergence with \(\widehat{R}\) would be too stringent for iterative imputation applications.
+This hypothesis is only partially supported.
+\(\widehat{R}\) indicated improving convergence until 30 to 40 iterations, while approximately unbiased, confidence-valid estimates were obtained after just seven iterations.
+This roughly corresponds to the threshold of 1.2.
+The more stringent threshold 1.1 seems plausible compared to the decrease in \(\widehat{R}\) values across iterations.
+The most recent threshold 1.01 does indeed seem to over-estimate the signs of non-convergence.
+\item
+  The results of this simulation study do not support the hypothesis that high \(AC\) values are implausible in iterative imputation algorithms with typical convergence.
+We expected that the randomness induced by the algorithm would effectively mitigate the risk of dependency within chains after a few iterations.
+In this study, however, we observed \(AC\)-values as high as theoretically possible (\(AC=1\)).
+The second part of the hypothesis is not refuted: the \(AC\)-values did decrease quickly as a function of \(T\).
+\item
+  Our final hypothesis was that multivariate \(\theta\)s would be better at detecting non-convergence than univariate \(\theta\)s.
+Multivariate \(\theta\)s did indeed show superior performance under certain conditions (e.g., multivariate \(Q\)s and extreme missingness proportions).
+\end{enumerate}
+
+\hypertarget{conclusion}{%
+\subsection{Conclusion}\label{conclusion}}
+
+We have shown that non-convergence in iterative imputation algorithms goes hand in hand with biased, invalid estimates---at least for the multivariate quantities of scientific interest considered in this study.
+Identifying non-convergence may be a crucial aspect of drawing valid statistical inferences from incomplete data.
+We concluded that the current practice of visually inspecting non-convergence through traceplots of univariate scalar summaries of the state space of the algorithm does not suffice.
+We, therefore, considered eight sets of diagnostic methods to identify non-convergence: two non-convergence diagnostics (\(\widehat{R}\) and \(AC\)), two univariate \(\theta\)s, a model-dependent multivariate \(\theta\), and a novel, model-independent multivariate \(\theta\).
+Signs of non-convergence due to early stopping were identified with each of the methods.
+Bias due to high proportions of incomplete cases were only identified with multivariate \(\theta\)s.
+
 Since we obtained approximately unbiased, confidence valid estimates after at most seven iterations, we conclude that the \(\widehat{R}\) threshold 1.2 is the most appropriate diagnostic cut-off.
 The threshold to diagnose non-stationarity with \(AC\) does not seem to apply.
+However, some signs of non-convergence were detected in simulation conditions many more iterations.
+Under moderate missingness (\(p_{\rm inc}\leq.50\)) \(\widehat{R}\)-values decreased substantially until 30 to 40 iterations, which implies that mixing in the algorithm still improved with each additional iteration.
+\(AC\)-values only improved until about six iterations, suggesting minimal improvement in trending was obtained beyond \(T=6\) in the current simulation set-up.
+
 The main finding of this study is that valid inferences may be obtained much quicker than approximate algorithmic convergence is reached.
@@
 
-Iterative imputation algorithms such as MICE have been known to yield valid results under severe missingness.
+Iterative imputation algorithms such as MICE are known to yield valid results under severe missingness.
 With this study, we have shown that valid inferences can also be obtained under a combination of severe missingness and early stopping.
 Based on our results, the traditional threshold of \(\widehat{R}> 1.2\) would be appropriate for diagnosing non-convergence in iterative imputation algorithms.
-This is not in line with the recent recommendations by veht19 to lower the threshold to 1.01.
-The discrepancy between our conclusion and \citet{veht19}'s may be explained by the nature of iterative imputation algorithms.
-In the limit, iterative imputation algorithms have the same character as other MCM algorithms, but in imputation procedures a part of the distribution is already determined by the observed data, whereas the entire distribution is unknown in many other MCMC applications.
-Because we combine a known distribution with an unknown distribution, valid estimates may be reached much sooner.
-Convergence may therefore be diagnosed at a less stringent threshold.
-\textbf{{[}HERE were the two asterixes{]}}
+This is, however, not in line with the recent recommendations by \citet{veht19} to lower the threshold to 1.01.
+The discrepancy between our conclusion and Vehtari et al.~may be explained by the nature of iterative imputation algorithms.
+In the limit, iterative imputation algorithms have the same characteristics as other MCMC algorithms.
+But, in imputation procedures, a part of the distribution is already determined by the observed data, whereas the entire distribution is unknown in many other MCMC applications.
+Since we combine a known distribution with an unknown distribution, valid estimates may be reached much sooner.
+Convergence may, therefore, be diagnosed at a less stringent threshold.
 
 The threshold to diagnose non-convergence with \(AC\) does not seem appropriate at all in iterative imputation algorithms.
-A better diagnostic cut-off would be the number of iterations at which an additional iteration does not substantially decrease \(AC\).
+A better diagnostic cut-off would be the number of iterations at which an additional iteration does not substantially decrease the \(AC\)-values.
 In practice, this implies that the default number of iterations in MICE is not sufficient.
 \(AC\) can only be computed if \(T\geq3\), while the current default in the \texttt{mice} package is five \citep{mice}.
-Identifying a decrease in the \(AC\)-values across iterations (an `elbow') requires more than three observations.
-We therefore suggest to increase the default number of iterations in MICE to 10.
-The computational cost of 5 additional iterations has become less of a burden since MICE was introduced in \citeyearpar{mice}.
+Identifying a reasonable decrease in the \(AC\)-values across iterations (an `elbow') requires more than three observations.
+We, therefore, suggest to increase the default number of iterations in MICE to ten.
+The computational cost of five additional iterations has become less of a burden since MICE was introduced \citeyearpar{mice}.
 Moreover, an increased number of iterations would grant the opportunity to exclude the first iterations from evaluation with \(\widehat{R}\).
@@
 
-Based on our results, we formulated several recommendations for empirical researchers who employ iterative imputation to draw inferences from incomplete data.
+Based on our results, we formulate several recommendations for empirical researchers who employ iterative imputation to draw inferences from incomplete data.
 We suggest that non-convergence may still be evaluated visually, but in addition to inspecting univariate summaries of the state space of the algorithm (\(\theta\)s, e.g., chain means), multivariate \(\theta\)s should be considered.
@@
   Use the threshold \(\widehat{R}>1.2\) to diagnose non-mixing, and assess stationarity by plotting the autocorrelation over iterations.
-Keep iterating until the threshold for mixing is overcome, and until reasonably decreasing asympototic \(AC\)-values are obtained.
+Keep iterating until the threshold for mixing is overcome, and until reasonably decreasing asymptotic \(AC\)-values are obtained.
 At that point, inferences are unlikely to improve by continued iteration.
-\item
-  Profit.
 \end{enumerate}
@@
 Proper performance under a `missing completely at random' (MCAR) mechanism is a necessary condition for any missing data method.
-It doesn't, however, guarantee equal performance under different missingness mechanisms.
+It does not't, however, guarantee equal performance under different missingness mechanisms.
 Future research should determine the performance of \(\widehat{R}\) and \(AC\) under `missing at random' and `missing not at random' mechanisms.
 Another parameter to consider in future research is the choice of the imputation method.
-As \citet{murr18} concluded, the behaviour of algorithms such as MICE under certain default imputation models is still an open research question.
-We have investigated the behaviour of MICE under only one type of model (Bayesian linear regression).
+As \citet{murr18} concluded, the behavior of algorithms such as MICE under certain default imputation models is still an open research question.
+We have investigated the behavior of MICE under only one type of model (Bayesian linear regression).
 Different imputations models might converge more poorly.
-Moreover, the imputation model may even be mis-specified un purpose to induce different levels of non-convergence.
+Moreover, the imputation model may even be mis-specified on purpose to induce different levels of non-convergence.
 Qualitatively, we have shown that \(\widehat{R}\) and \(AC\) are appropriate under clear violation of convergence.
 Quantitatively, we have only demonstrated that they can identify signs of non-convergence under MCAR, when the imputation model is a correctly specified Bayesian linear regression model.
-\newline \newline
+\textbf{{[}ADD limitation that other MI packages may have different way to obtain initial values!{]}} \newline \newline
 
-\noindent In short, we have shown that an iterative imputation algorithm can yield correct outcomes, even when a converged state has not yet formally been reached.
+\noindent In short, we have shown that iterative imputation algorithms can yield correct outcomes, even when a converged state has not yet formally been reached.
 Any further iterations would then burn computational resources without improving the statistical inferences.

### 2020-05-10 3.Thesis/Writeup/2.Manuscript.Rmd
After weird default latex layout occurred

@@
 
-In this paper, we study different methods for assessing non-convergence in iterative imputation algorithms.
+In this paper we study different methods for assessing non-convergence in iterative imputation algorithms.
 We define several diagnostics and evaluate how these diagnostics could be appropriate for iterative imputation applications.
@@
 ```{r non-conv, out.width='.49\\linewidth', fig.width=3.5, fig.height=3,fig.show='hold',fig.cap="Typical convergence versus pathological non-convergence.
-Depicted are are chain means of variable $j$ in each imputation ($m=5$; $T=10$).
+Depicted are chain means of variable $j$ in each imputation ($m=5$; $T=10$).
 Please note that any difference between the two scenarios (see e.g., the magnitude of the chain means on the y-axes) stems from the imputation algorithms---the two scenarios have equal incomplete data $y$."}
@@
 In a traceplot, a scalar summary of the state space of the algorithm $\theta$ is plotted against the iteration number, as depicted in Figure \ref{fig:non-conv}.
-In `mice`, the default scalar summaries to inspect are chain means and chain variances.
+The default scalar summaries to inspect for the MICE algorithm are chain means and chain variances.
 Non-convergence is diagnosed if the imputation chains are not freely intermingled with one another, or if the chains show definite trends [@buur18].
@@
 As alternative, van Buuren [-@buur18, $\S$ 4.5.2] proposed multivariate evaluation of the MICE algorithm through eigenvalue decomposition, building on the work of @mack03.
-This technique may yield a model-independent, multivariate scalar summary to monitor, but it is not implemented in `mice`.
+This technique may yield a model-independent, multivariate scalar summary to monitor, but it is not implemented in the  `mice` package [@mice].
 
@@
 Recurrence within imputation chains indicates trending.
-Ergo, non-stationarity may be diagnosed when $AC$ values are larger than zero.
+Ergo, non-stationarity may be diagnosed when $AC$-values are larger than zero.
 -->
@@
 Therefore, we compute $AC$ manually, see the solid line.
-These $AC$ values behave as expected from visual inspection.
+These $AC$-values behave as expected from visual inspection.
 We will, therefore, only consider manually calculated $AC$-values as non-convergence diagnostic.
@@
 
-The assumption inherent to the second set of simulation conditions, early stopping, is that terminating the imputation algorithm too early causes non-convergence.
+The assumption inherent to the early stopping conditions is that terminating the imputation algorithm too early causes non-convergence.
 Generally, the algorithm will not reach convergence if $T=1$, because the imputed values in the first iteration (at $t=1$) depend on the initial values of the algorithm [which are sampled randomly from the set of observed datapoints; @buur18].
@@
 4.
-We expect that high $AC$ values are implausible in iterative imputation algorithms with typical convergence.
+We expect that high $AC$-values are implausible in iterative imputation algorithms with typical convergence.
 After only a few iterations, the randomness induced by the algorithm should effectively mitigate the risk of dependency within chains.
@@
 
-In each simulation repetition, we *ampute* the complete dataset conform the 'incompleteness' conditions.
+In each simulation repetition, we *ampute* the complete dataset conform the incompleteness conditions.
 The proportion of incomplete cases $p_{\rm inc}$ varies between 5% and 95%, in incremental steps ($p_{\rm inc}$ = .05, .25, .50, .75, .95).
@@
 I.e., the probability to be missing is the same for all $N \times k$ cells in $y$, conditional on $p_{\rm inc}$.
-We use the 'mice' package [function `mice::ampute()`; @mice] to obtain an incomplete dataset {$y_{\rm obs}, y_{\rm mis}$} for each of the five 'incompleteness' conditions.
-
-We impute the missing values in {$y_{\rm obs}, y_{\rm mis}$} according to the 'early stopping' conditions.
+We use the 'mice' package [function `mice::ampute()`; @mice] to obtain an incomplete dataset {$y_{\rm obs}, y_{\rm mis}$} for each of the five incompleteness conditions.
+
+We impute the missing values in {$y_{\rm obs}, y_{\rm mis}$} according to the early stopping conditions.
 That is, we vary the number of iterations in the imputation algorithm between one and one hundred ($T = 1,2,.,100$).
 All imputation procedures are performed using `mice` [function `mice()`; @mice], with Bayesian linear regression imputation, and five imputation chains ($m=5$).
-This results in five sets of imputations for each of the one hundred 'early stopping' conditions, $y_{\rm imp, \ell}$.
+This results in five sets of imputations for each of the one hundred early stopping conditions, $y_{\rm imp, \ell}$.
 
@@
 
-Our results show the performance of the MICE algorithm under different conditions of 'incompleteness' and 'early stopping'.
+Our results show the performance of the MICE algorithm under different conditions of incompleteness and early stopping.
 For reasons of brevity, we only discuss results for the worst-performing estimates in terms of bias.
@@
 Since there is just one estimate for $Q=r^2$ to consider, we **end up with** $Q=\mu_Y, \sigma_Y,r^2, \beta_1$.
-In the figures, we present all 'incompleteness' conditions, but only 'early stopping' conditions where $1 \leq T\leq50$.
+In the figures, we present all incompleteness conditions, but only early stopping conditions where $1 \leq T\leq50$.
 That is, the results are more or less stable for conditions where $T \geq 30$.
@@
 This suggests that the combination of $\widehat{R}$ and $\lambda_{1, \ell}$ is only appropriate for imputation algorithms where the number of iterations at least four.
-Interpreting the $\widehat{R}$ values of conditions where $T<3$ may lead to the incorrect conclusion that only conditions with *low* missingness proportions show signs of non-convergence.
+Interpreting the $\widehat{R}$-values of conditions where $T<3$ may lead to the incorrect conclusion that only conditions with *low* missingness proportions show signs of non-convergence.
 For applications as a model-independent multivariate $\theta$, we should therefore only interpret the $\widehat{R}$-values when $T\geq4$.
@@
 Conditions where $p_{\rm inc}=.95$ show statistically significant autocorrelations for any number of iterations.
-The other extreme, conditions where $p_{\rm inc}=.05$, does not yield $AC$ values above the threshold.
-The $AC$ values with this missingness proportion are equivalent for all conditions where $T\geq6$, which suggests that the algorithm is as stationary as it can be.
+The other extreme, conditions where $p_{\rm inc}=.05$, does not yield $AC$-values above the threshold.
+The $AC$-values with this missingness proportion are equivalent for all conditions where $T\geq6$, which suggests that the algorithm is as stationary as it can be.
 A similar point is only observed after twenty iterations in conditions with higher missingness proportions.
@@
 \noindent Overall, the methods to diagnose non-convergence perform as expected: they indicate more signs of non-convergence in conditions with worse performance in terms of bias and confidence-validity.
-Under low to moderate missingness, it does not seems to matter on which $\theta$ the non-convergence diagnostics are applied.
+Under low to moderate missingness, it does not seem to matter on which $\theta$ the non-convergence diagnostics are applied.
 If we look at complete unbiasedness, however, we notice that univariate $\theta$s fail to diagnose the persistent bias in conditions where $p_{\rm inc}\geq.75$.
@@
 **This is only for 75% missingness:** The conventional $\widehat{R}$ threshold of 1.2 is reached in simulation conditions $T = 3$ and $T > 6$.
-(With the default number of iterations (maxit = 5), this dip in $\widehat{R}$ values would be spotted, so it is no problem.) The point at which an extra iteration does not seem to improve the $\widehat{R}$ value is around $T=30$.
+(With the default number of iterations (maxit = 5), this dip in $\widehat{R}$-values would be spotted, so it is no problem.) The point at which an extra iteration does not seem to improve the $\widehat{R}$ value is around $T=30$.
 -->
@@
 
-<!-- The $\widehat{R}$ and $AC$ values for the imputation chain variances show equal trends to the chain means and are therefore not discussed separately.
+<!-- The $\widehat{R}$ and $AC$-values for the imputation chain variances show equal trends to the chain means and are therefore not discussed separately.
 Taken together, univariate estimates seem robust across simulation conditions.
@@
 This roughly corresponds to the threshold of 1.2.
-The more stringent threshold 1.1 seems plausible compared to the decrease in $\widehat{R}$ values across iterations.
+The more stringent threshold 1.1 seems plausible compared to the decrease in $\widehat{R}$-values across iterations.
 The most recent threshold 1.01 does indeed seem to over-estimate the signs of non-convergence.
@@
 4.
-The results of this simulation study do not support the hypothesis that high $AC$ values are implausible in iterative imputation algorithms with typical convergence.
+The results of this simulation study do not support the hypothesis that high $AC$-values are implausible in iterative imputation algorithms with typical convergence.
 We expected that the randomness induced by the algorithm would effectively mitigate the risk of dependency within chains after a few iterations.
@@
 $\widehat{R}$ <1.2 is not stringent enough, because conditions in which $\widehat{R}$ is smaller than 1.2 do not all yield unbiased estimates of the $Q$s.
-Moreover, there is a dip in the $\widehat{R}$ values, after which $\widehat{R}$ > 1.2 again.
-If the algorithms would be terminated at the iteration where $\widehat{R}$ < 1.2 occurs for the first time, the increase in $\widehat{R}$ values might be missed.
+Moreover, there is a dip in the $\widehat{R}$-values, after which $\widehat{R}$ > 1.2 again.
+If the algorithms would be terminated at the iteration where $\widehat{R}$ < 1.2 occurs for the first time, the increase in $\widehat{R}$-values might be missed.
 The threshold $\widehat{R}$ < 1.1 is somewhat conservative in comparison with the performance measures.

### 2020-05-10 3.Thesis/Writeup/2.Manuscript.tex
After weird default latex layout occurred

@@
 \caption{Typical convergence versus pathological non-convergence.
-Depicted are are chain means of variable $j$ in each imputation ($m=5$; $T=10$).
+Depicted are chain means of variable $j$ in each imputation ($m=5$; $T=10$).
 Please note that any difference between the two scenarios (see e.g., the magnitude of the chain means on the y-axes) stems from the imputation algorithms---the two scenarios have equal incomplete data $y$.}\label{fig:non-conv}
@@
 In a traceplot, a scalar summary of the state space of the algorithm \(\theta\) is plotted against the iteration number, as depicted in Figure \ref{fig:non-conv}.
-In \texttt{mice}, the default scalar summaries to inspect are chain means and chain variances.
+The default scalar summaries to inspect for the MICE algorithm are chain means and chain variances.
 Non-convergence is diagnosed if the imputation chains are not freely intermingled with one another, or if the chains show definite trends \citep{buur18}.
@@
 As alternative, van Buuren \citeyearpar[\(\S\) 4.5.2]{buur18} proposed multivariate evaluation of the MICE algorithm through eigenvalue decomposition, building on the work of \citet{mack03}.
-This technique may yield a model-independent, multivariate scalar summary to monitor, but it is not implemented in \texttt{mice}.
+This technique may yield a model-independent, multivariate scalar summary to monitor, but it is not implemented in the \texttt{mice} package \citep{mice}.
 
@@
 Therefore, we compute \(AC\) manually, see the solid line.
-These \(AC\) values behave as expected from visual inspection.
+These \(AC\)-values behave as expected from visual inspection.
 We will, therefore, only consider manually calculated \(AC\)-values as non-convergence diagnostic.
@@
 
-The assumption inherent to the second set of simulation conditions, early stopping, is that terminating the imputation algorithm too early causes non-convergence.
+The assumption inherent to the early stopping conditions is that terminating the imputation algorithm too early causes non-convergence.
 Generally, the algorithm will not reach convergence if \(T=1\), because the imputed values in the first iteration (at \(t=1\)) depend on the initial values of the algorithm \citep[which are sampled randomly from the set of observed datapoints;][]{buur18}.
@@
 \item
-  We expect that high \(AC\) values are implausible in iterative imputation algorithms with typical convergence.
+  We expect that high \(AC\)-values are implausible in iterative imputation algorithms with typical convergence.
 After only a few iterations, the randomness induced by the algorithm should effectively mitigate the risk of dependency within chains.
@@
 
-In each simulation repetition, we \emph{ampute} the complete dataset conform the `incompleteness' conditions.
+In each simulation repetition, we \emph{ampute} the complete dataset conform the incompleteness conditions.
 The proportion of incomplete cases \(p_{\rm inc}\) varies between 5\% and 95\%, in incremental steps (\(p_{\rm inc}\) = .05, .25, .50, .75, .95).
@@
 I.e., the probability to be missing is the same for all \(N \times k\) cells in \(y\), conditional on \(p_{\rm inc}\).
-We use the `mice' package \citep[function \texttt{mice::ampute()};][]{mice} to obtain an incomplete dataset \{\(y_{\rm obs}, y_{\rm mis}\)\} for each of the five `incompleteness' conditions.
+We use the `mice' package \citep[function \texttt{mice::ampute()};][]{mice} to obtain an incomplete dataset \{\(y_{\rm obs}, y_{\rm mis}\)\} for each of the five incompleteness conditions.
 
-We impute the missing values in \{\(y_{\rm obs}, y_{\rm mis}\)\} according to the `early stopping' conditions.
+We impute the missing values in \{\(y_{\rm obs}, y_{\rm mis}\)\} according to the early stopping conditions.
 That is, we vary the number of iterations in the imputation algorithm between one and one hundred (\(T = 1,2,.,100\)).
 All imputation procedures are performed using \texttt{mice} \citep[function \texttt{mice()};][]{mice}, with Bayesian linear regression imputation, and five imputation chains (\(m=5\)).
-This results in five sets of imputations for each of the one hundred `early stopping' conditions, \(y_{\rm imp, \ell}\).
+This results in five sets of imputations for each of the one hundred early stopping conditions, \(y_{\rm imp, \ell}\).
 
@@
 
-Our results show the performance of the MICE algorithm under different conditions of `incompleteness' and `early stopping'.
+Our results show the performance of the MICE algorithm under different conditions of incompleteness and early stopping.
 For reasons of brevity, we only discuss results for the worst-performing estimates in terms of bias.
@@
 Since there is just one estimate for \(Q=r^2\) to consider, we \textbf{end up with} \(Q=\mu_Y, \sigma_Y,r^2, \beta_1\).
-In the figures, we present all `incompleteness' conditions, but only `early stopping' conditions where \(1 \leq T\leq50\).
+In the figures, we present all incompleteness conditions, but only early stopping conditions where \(1 \leq T\leq50\).
 That is, the results are more or less stable for conditions where \(T \geq 30\).
@@
 \noindent Overall, the methods to diagnose non-convergence perform as expected: they indicate more signs of non-convergence in conditions with worse performance in terms of bias and confidence-validity.
-Under low to moderate missingness, it does not seems to matter on which \(\theta\) the non-convergence diagnostics are applied.
+Under low to moderate missingness, it does not seem to matter on which \(\theta\) the non-convergence diagnostics are applied.
 If we look at complete unbiasedness, however, we notice that univariate \(\theta\)s fail to diagnose the persistent bias in conditions where \(p_{\rm inc}\geq.75\).
@@
 This roughly corresponds to the threshold of 1.2.
-The more stringent threshold 1.1 seems plausible compared to the decrease in \(\widehat{R}\) values across iterations.
+The more stringent threshold 1.1 seems plausible compared to the decrease in \(\widehat{R}\)-values across iterations.
 The most recent threshold 1.01 does indeed seem to over-estimate the signs of non-convergence.
 \item
-  The results of this simulation study do not support the hypothesis that high \(AC\) values are implausible in iterative imputation algorithms with typical convergence.
+  The results of this simulation study do not support the hypothesis that high \(AC\)-values are implausible in iterative imputation algorithms with typical convergence.
 We expected that the randomness induced by the algorithm would effectively mitigate the risk of dependency within chains after a few iterations.

### 2020-05-11 3.Thesis/Writeup/2.Manuscript.Rmd
All but last section DONE

@@
 Moreover, the behavior of such algorithms under certain default imputation models (e.g., 'predictive mean matching') is an entirely open question [@murr18].
-Therefore, algorithmic convergence should be monitored carefully, **but this is not straightforward**.
+Therefore, algorithmic convergence should be monitored carefully, but this is not straightforward.
 Iterative imputation algorithms such as MICE are special cases of Markov chain Monte Carlo (MCMC) methods.
@@
 With that, we formulate an informed advice on when it is safe to conclude that the algorithm is converged *enough* for valid inferences.
-We translate the results of the study into guidelines for applied researchers.
-**ADD that these guidelines may aid to draw valid inference from incomplete data?]**
+We translate the results of the study into guidelines for applied researchers, which may facilitate drawing valid inferences from incomplete data.
 
@@
 For each datapoint in $y_{\rm mis}$, we sample $m \times T$ plausible values, where $m$ is the number of imputations ($\ell = 1, 2, \dots, m$) and $T$ is the number of iterations ($t = 1, 2, \dots, T$) in the imputation algorithm.
-The state space of the algorithm at a certain iteration $t$ may be summarized by a scalar summary $\theta$ (e.g., the average of the **imputed values**).
+The state space of the algorithm at a certain iteration $t$ may be summarized by a scalar summary $\theta$ (e.g., the average of the imputed values).
 The collection of $\theta$-values between $t=1$ (summarizing the state space of the algorithm at initialization) and $t=T$ (summarizing the state space for the imputed values) will be referred to as an 'imputation chain'.
@@
 In the non-convergence scenario, there is a lot of trending and some chains do not intermingle.
-Importantly, the chain means at the last iteration (the imputed value per imputation $\ell$) are very different between the two scenarios.
+Importantly, the chain means at the last iteration (the imputed values per imputation $\ell$) are very different between the two scenarios.
 The algorithm with the mis-specified model yields imputed values that are on average a factor two larger than those of the typically converged algorithm.
@@
 Therefore, it is important to reach algorithmic convergence in iterative imputation algorithms.
-**[ADD t and l to each theta??]**
-
-```{r non-conv, out.width='.49\\linewidth', fig.width=3.5, fig.height=3,fig.show='hold',fig.cap="Typical convergence versus pathological non-convergence.
+
+```{r non-conv, out.width='.49\\linewidth', fig.width=3.5, fig.height=3,fig.show='hold',fig.cap="Typical convergence versus pathological non-convergence (reproduced from van Buuren; 2018).
 Depicted are chain means of variable $j$ in each imputation ($m=5$; $T=10$).
@@
 Identifying non-convergence by inspecting traceplots may be undesirable for several reasons: 1) it may be challenging to the untrained eye, 2) only severely pathological cases of non-convergence may be diagnosed, and 3) there is not an objective measure that quantifies convergence [@buur18, $\S$ 6.5.2].
-Moreover, the typical traceplots to inspect only **consider** univariate $\theta$s.
+Moreover, traceplots are typically only used to inspect  univariate $\theta$s.
 Monitoring univariate summaries of the state space of the algorithm may be insufficient because MICE is not only concerned with a single column, but the entire multivariate distribution of the imputations.
@@
 Panel B shows two versions of the $AC$: the default calculation with `R` function `stats::acf()` [@R], and manual calculation as the correlation between $\theta$ in iteration $t$ and $\theta$ in iteration $t+1$.
-Panel C displays the traditional computation of $\widehat{R}$ conform @gelm92, and in panel D we see $\widehat{R}$ as computed by implementing @veht19 's recommendations.
-**[Add labels for the two types of Rhat to the figure!]**
-
-```{r diagnostics, fig.cap="Convergence diagnostics applied on the imputation algorithms of Figure 2. $\\theta$ = chain mean in $ \\dot y_{\\rm imp, \\ell} $."}
+Panel C displays the traditional computation of $\widehat{R}$ conform @gelm92 (original), and in panel D we see $\widehat{R}$ as computed by implementing Vehtari et al.'s recommendations (adapted).
+
+```{r diagnostics, fig.cap="Convergence diagnostics applied on the two scenarios from Figure 2 ($\\theta$ = chain mean of variable $j$; $m=5$; $T=10$).
+Depicted are the imputation chains in panel A, two versions of the autocorrelation in panel B, and two versions of $\\widehat{R}$ in panels C and D."}
 # load("./images/diagnostics_plot.Rdata") #bquote(theta ~ "= chain mean in "~ y[imp][",m"])
@@
 
-**[REPHRASE as active about panel A]** From visual inspection, we know that the typically converged algorithm initially portrayed some signs of non-mixing (around $t=2$), but intermingled nicely overall.
-Additionally, there was very little trending.
-The algorithm with pathological non-convergence showed severe non-mixing, although this gradually improved (from $t=7$).
-In this scenario, there was a lot of trending as well initially (up-to $t=6$), after which the chains reached a somewhat more stationary state.
+By visually inspecting the imputation chains in panel A, we conclude that the two scenarios show very different convergence.
+The typically converged algorithm initially portrayes some signs of non-mixing (around $t=2$), but intermingles nicely overall.
+Additionally, there is very little trending in this scenario.
+The algorithm with pathological non-convergence shows severe non-mixing, although this gradually improves (beyond $t=7$).
+In this scenario, we see a lot of trending initially (up-to $t=6$), after which the chains reached a somewhat more stationary state.
 
@@
 
-**Describe panel C here and why it is not optimal: De grotere variantie overall zorgt er kennelijk voor dat $\widehat{R}$ groeit.
-De $\widehat{R}$ geeft de foutieve boodschap dat convergentie steeds slechter gaat.
-Verklaring: de overdispersie assumptie klopt niet als er trend is.]** The $\widehat{R}$-values in panel D are conform our expectations.
-The adapted version of $\widehat{R}$ proposed by @veht19 does indeed indicate less signs of non-convergence as the number of iterations goes up.
-**[explain the dip in Rhat values at t=2.
-Namely, because we can only use 2 of the 3 tricks by @veht19 if the number of iterations is very low (t<4).
-That's why the $\widehat{R}$s are more similar to the traditional GR.]** \newline
+In panel C, we do not see a lot of difference between the two scenarios.
+The $\widehat{R}$-values are even somewhat lower for the non-convergence scenario than for the typical convergence.
+And apparently, the large variance causes an increase in $\widehat{R}$-values.
+Convergence seems to worsen with every additional iteration.
+This paints the wrong picture.
+An explanation for this behavior may be that under severe trending, the assumption of over-dispersion is violated.
+Taken together, this version of $\widehat{R}$ is sub-optimal for assessing non-convergence in MI.
+
+The $\widehat{R}$-values in panel D are conform our expectations.
+The adapted version of $\widehat{R}$ depicted here does indicate less signs of non-convergence as the number of iterations goes up.
+An exception to this general trend is the 'dip' around $t=3$.
+As diagnostic tool, this may lead to the incorrect conclusion that the non-convergence scenario has superior convergence compared to the typical scenario.
+We can explain this as a downside of the low number of iteration in iterative imputation: the adapted version of $\widehat{R}$ can only be completely employed if the number of iterations is at least four.
+Otherwise, it is not possible to perform all three transformations to the chains of $\theta$-values.
+For low values of $t$, the $\widehat{R}$-values conform Vehtari et al.
+are, therefore, more similar to the original $\widehat{R}$ by Gelman and Rubin.
+\newline
 <!-- However, the superior performance of the typical convergence over the pathological non-convergence is less prominent, even flipped for $t<4$.  -->
@@
 Then, convergence to a stable but highly variable state may be reached instantly.
-**Still, we expect that the incompleteness conditions will result in more signs of non-convergence.**
+Still, we expect that incompleteness conditions with higher $p_{\rm inc}$ will result in more signs of non-convergence.
 
@@
 And for the regression coefficients ($Q=\beta$) the bias is most pronounced in $\beta_1$ (the effect of $X_1$ on $Y$).
-Since there is just one estimate for $Q=r^2$ to consider, we **end up with** $Q=\mu_Y, \sigma_Y,r^2, \beta_1$.
+Since there is just one estimate for $Q=r^2$ to evaluate, we consider $Q=\mu_Y, \sigma_Y,r^2, \beta_1$.
 In the figures, we present all incompleteness conditions, but only early stopping conditions where $1 \leq T\leq50$.
@@
 We present the subset of $Q$s with the worst performance in terms of bias: the mean and standard deviation of $Y$, and the regression coefficient of $X_1$."}
-mean_bias + sd_bias + Rsq_bias + est_bias + est_cov + est_ciw + plot_layout(ncol = 2, guides = "collect")
+mean_bias + sd_bias + Rsq_bias + est_bias + est_cov + est_ciw + plot_layout(ncol = 2, guides = "collect") + plot_annotation(tag_levels = "A", tag_suffix = ".")
 ```
@@
 Figure \ref{fig:Qs} displays the influence of incompleteness and early stopping on the estimated quantities of scientific interest, $Q$s.
-**[Add panel labels, describe panels]** Simulation conditions that are impacted by non-convergence portray more extreme bias in the estimated $Q$s, and non-nominal coverages.
+In the first row of the figure, we see the bias in the estimated descriptive statistics ($Q=\mu_Y$ and $Q=\sigma_Y$).
+The second row shows the bias in the multivariate scientific estimands ($Q=r^2$ and $Q=\beta_1$).
+The third row consists of the coverage rate (CR) and confidence interval width (CIW) of the estimated regression coefficient ($Q=\beta_1$).
+
+<!-- Simulation conditions that are impacted by non-convergence portray more extreme bias in the estimated $Q$s, and non-nominal coverages.  -->
 
@@
 Completely unbiased estimates are only obtained in conditions where $p_{\rm inc}\leq.50$.
-**This is curious, since the MCAR missingness mechanism should yield unbiased univariate estimates without employing multiple imputation.** Note however, that the magnitude of the bias in these conditions is small: $\mu_Y$ is maximally under-estimated by 0.02 units, while the true value of $\mu_Y$ is 25.81 ($\sigma_Y = 11.32$).
+This is curious, since the MCAR missingness mechanism should yield unbiased univariate estimates without employing multiple imputation.
+Note however, that the magnitude of the bias in these conditions is small: $\mu_Y$ is maximally under-estimated by 0.02 units, while the true value of $\mu_Y$ is 25.81 ($\sigma_Y = 11.32$).
 We, therefore, conclude that non-convergence does not substantially impact the validity of the inferences for $Q=\mu$.
@@
 ### $Q=r^2$.
-In contrast to the univariate estimates, the estimated coefficient of determination is clearly affected by early stopping.
+In contrast to the univariate estimates, the estimated coefficient of determination is clearly affected by early stopping, see Figure \ref{fig:Qs}C.
 That is, the bias in $Q$ does not only follow the incremental order of the incompleteness conditions, but also decreases with each additional iteration.
@@
 
-Figure \ref{fig:rhats-acs} displays results for the eight sets of diagnostic methods.
-The two columns in the figure represent the  non-convergence diagnostics $\widehat{R}$ and $AC$; the four rows depict the four scalar summaries $\theta$ under consideration.
-**[describe panels??]** We evaluate the performance of the methods by establishing whether they correctly identify conditions in which $\bar{Q}$ is *not* an unbiased, confidence-valid estimate of $Q$.
+Figure \ref{fig:rhats-acs} displays results of the eight sets of diagnostic methods.
+The two columns in the figure represent the  non-convergence diagnostics $\widehat{R}$ and $AC$; the rows depict the four scalar summaries $\theta$ under consideration.
+The first row in Figure \ref{fig:rhats-acs} thus shows $\widehat{R}$ and $AC$ applied on the chain means.
+In the second row we see the diagnostics applied on the chain variances.
+The third row depicts the same for $\theta=\hat{Q}$.
+And in the last row we see the diagnostics applied on the novel $\theta$ to consider: $\lambda_1$.
+We evaluate the performance of the methods by establishing whether they correctly identify conditions in which $\bar{Q}$ is *not* an unbiased, confidence-valid estimate of $Q$.
 As we concluded in the last section, substantially biased estimates were only obtained in conditions where $T\leq6$, and we observed non-nominal coverage only when $T\leq3$.
@@
 
-```{r rhats-acs, fig.height = 6.8, fig.cap="Non-convergence diagnostics. [Split up into 2 or 4 figures??]"}
-mean_Rh + mean_AC + var_Rh + var_AC + est_Rh + est_AC + PCA_Rh + PCA_AC + plot_layout(guides = "collect", ncol = 2)
+```{r rhats-acs, fig.height = 6.8, fig.cap="Non-convergence diagnostics $\\widehat{R}$ and $AC$ applied on several $\\theta$s.
+The left-hand side of the figure contains $\\widehat{R}$-values, and the right-hand side contains $AC$-values.
+Depicted in the rows are the scalar summaries $\\theta$: chain mean, chain variance, the quantity of scientific interest $\\hat{Q}$, and the first eigenvalue of the variance-covariance matrix $\\lambda_1$."}
+mean_Rh + mean_AC + var_Rh + var_AC + est_Rh + est_AC + PCA_Rh + PCA_AC + plot_layout(guides = "collect", ncol = 2) + plot_annotation(tag_levels = "A", tag_suffix = ".")
 ```
@@
 The thresholds 1.1 and 1.01 seem too strict compared to the validity of the inferences.
-
 The threshold to diagnose significant $AC$-values does not appear to be appropriate in at least the current set-up, and perhaps in iterative imputation in general.
@@
 
-1. First, check traceplots of the default $\theta$s (chain means and chain variances) for signs of pathological non-convergence. Adjust the imputation model if necessary.
+1. First, check traceplots of the default $\theta$s (e.g., chain means and chain variances) for signs of pathological non-convergence. Adjust the imputation model if necessary.
 
 2.
-Track multivariate $\theta$s over iterations, e.g., the novel $\theta$ that is  scientific model-independent.
-Or specify your own scalar summary of interest (see e.g., @buur18).
-Monitor these $\theta$s across iterations through visual inspection, or using a non-convergence diagnostic.
+Subsequently decide which multivariate $\theta$s to track across iterations.
+The novel $\theta$ that we propose in this study, $\lambda_1$ is scientific model-independent and may thus always be employed.
+Alternatively, specify your own scalar summary of interest [see e.g., @buur18].
+Monitor these $\theta$s through visual inspection, or using a non-convergence diagnostic.
 
@@
 Compute non-convergence diagnostics $\widehat{R}$ and $AC$.
-Do not use the traditional calculation for $\widehat{R}$ [@gelm92], or the `R` function `stats::acf()` to compute $AC$ [@R].
-Instead, calculate $\widehat{R}$ conform @veht19 and compute autocorrelations manually (see e.g., [github.com/gerkovink/shinyMice](https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy)).
+Do not use the original implementation to calculate $\widehat{R}$ [@gelm92], or the `R` function `stats::acf()` to compute $AC$ [@R].
+Instead, calculate $\widehat{R}$ conform @veht19 and compute autocorrelations manually (see e.g., [github.com/gerkovink/shinyMice](https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/)).
 

### 2020-05-11 3.Thesis/Writeup/2.Manuscript.tex
All but last section DONE

@@
 Moreover, the behavior of such algorithms under certain default imputation models (e.g., `predictive mean matching') is an entirely open question \citep{murr18}.
-Therefore, algorithmic convergence should be monitored carefully, \textbf{but this is not straightforward}.
+Therefore, algorithmic convergence should be monitored carefully, but this is not straightforward.
 Iterative imputation algorithms such as MICE are special cases of Markov chain Monte Carlo (MCMC) methods.
@@
 With that, we formulate an informed advice on when it is safe to conclude that the algorithm is converged \emph{enough} for valid inferences.
-We translate the results of the study into guidelines for applied researchers.
-\textbf{ADD that these guidelines may aid to draw valid inference from incomplete data?{]}}
+We translate the results of the study into guidelines for applied researchers, which may facilitate drawing valid inferences from incomplete data.
 
@@
 For each datapoint in \(y_{\rm mis}\), we sample \(m \times T\) plausible values, where \(m\) is the number of imputations (\(\ell = 1, 2, \dots, m\)) and \(T\) is the number of iterations (\(t = 1, 2, \dots, T\)) in the imputation algorithm.
-The state space of the algorithm at a certain iteration \(t\) may be summarized by a scalar summary \(\theta\) (e.g., the average of the \textbf{imputed values}).
+The state space of the algorithm at a certain iteration \(t\) may be summarized by a scalar summary \(\theta\) (e.g., the average of the imputed values).
 The collection of \(\theta\)-values between \(t=1\) (summarizing the state space of the algorithm at initialization) and \(t=T\) (summarizing the state space for the imputed values) will be referred to as an `imputation chain'.
@@
 In the non-convergence scenario, there is a lot of trending and some chains do not intermingle.
-Importantly, the chain means at the last iteration (the imputed value per imputation \(\ell\)) are very different between the two scenarios.
+Importantly, the chain means at the last iteration (the imputed values per imputation \(\ell\)) are very different between the two scenarios.
 The algorithm with the mis-specified model yields imputed values that are on average a factor two larger than those of the typically converged algorithm.
@@
 Therefore, it is important to reach algorithmic convergence in iterative imputation algorithms.
-\textbf{{[}ADD t and l to each theta??{]}}
 
@@
 
-\caption{Typical convergence versus pathological non-convergence.
+\caption{Typical convergence versus pathological non-convergence (reproduced from van Buuren; 2018).
 Depicted are chain means of variable $j$ in each imputation ($m=5$; $T=10$).
@@
 Identifying non-convergence by inspecting traceplots may be undesirable for several reasons: 1) it may be challenging to the untrained eye, 2) only severely pathological cases of non-convergence may be diagnosed, and 3) there is not an objective measure that quantifies convergence \citep[\(\S\) 6.5.2]{buur18}.
-Moreover, the typical traceplots to inspect only \textbf{consider} univariate \(\theta\)s.
+Moreover, traceplots are typically only used to inspect univariate \(\theta\)s.
 Monitoring univariate summaries of the state space of the algorithm may be insufficient because MICE is not only concerned with a single column, but the entire multivariate distribution of the imputations.
@@
 Panel B shows two versions of the \(AC\): the default calculation with \texttt{R} function \texttt{stats::acf()} \citep{R}, and manual calculation as the correlation between \(\theta\) in iteration \(t\) and \(\theta\) in iteration \(t+1\).
-Panel C displays the traditional computation of \(\widehat{R}\) conform \citet{gelm92}, and in panel D we see \(\widehat{R}\) as computed by implementing \citet{veht19} 's recommendations.
-\textbf{{[}Add labels for the two types of Rhat to the figure!{]}}
+Panel C displays the traditional computation of \(\widehat{R}\) conform \citet{gelm92} (original), and in panel D we see \(\widehat{R}\) as computed by implementing Vehtari et al.'s recommendations (adapted).
 
@@
 
-\caption{Convergence diagnostics applied on the imputation algorithms of Figure 2. $\theta$ = chain mean in $ \dot y_{\rm imp, \ell} $.}\label{fig:diagnostics}
+\caption{Convergence diagnostics applied on the two scenarios from Figure 2 ($\theta$ = chain mean of variable $j$; $m=5$; $T=10$).
+Depicted are the imputation chains in panel A, two versions of the autocorrelation in panel B, and two versions of $\widehat{R}$ in panels C and D.}\label{fig:diagnostics}
 \end{figure}
 
-\textbf{{[}REPHRASE as active about panel A{]}} From visual inspection, we know that the typically converged algorithm initially portrayed some signs of non-mixing (around \(t=2\)), but intermingled nicely overall.
-Additionally, there was very little trending.
-The algorithm with pathological non-convergence showed severe non-mixing, although this gradually improved (from \(t=7\)).
-In this scenario, there was a lot of trending as well initially (up-to \(t=6\)), after which the chains reached a somewhat more stationary state.
+By visually inspecting the imputation chains in panel A, we conclude that the two scenarios show very different convergence.
+The typically converged algorithm initially portrayes some signs of non-mixing (around \(t=2\)), but intermingles nicely overall.
+Additionally, there is very little trending in this scenario.
+The algorithm with pathological non-convergence shows severe non-mixing, although this gradually improves (beyond \(t=7\)).
+In this scenario, we see a lot of trending initially (up-to \(t=6\)), after which the chains reached a somewhat more stationary state.
 
@@
 
-\textbf{Describe panel C here and why it is not optimal: De grotere variantie overall zorgt er kennelijk voor dat \(\widehat{R}\) groeit.
-De \(\widehat{R}\) geeft de foutieve boodschap dat convergentie steeds slechter gaat.
-Verklaring: de overdispersie assumptie klopt niet als er trend is.{]}} The \(\widehat{R}\)-values in panel D are conform our expectations.
-The adapted version of \(\widehat{R}\) proposed by \citet{veht19} does indeed indicate less signs of non-convergence as the number of iterations goes up.
-\textbf{\citep[explain the dip in Rhat values at t=2.
-Namely, because we can only use 2 of the 3 tricks by][ if the number of iterations is very low (t\textless4).
-That's why the \(\widehat{R}\)s are more similar to the traditional GR.]{veht19}} \newline
+In panel C, we do not see a lot of difference between the two scenarios.
+The \(\widehat{R}\)-values are even somewhat lower for the non-convergence scenario than for the typical convergence.
+And apparently, the large variance causes an increase in \(\widehat{R}\)-values.
+Convergence seems to worsen with every additional iteration.
+This paints the wrong picture.
+An explanation for this behavior may be that under severe trending, the assumption of over-dispersion is violated.
+Taken together, this version of \(\widehat{R}\) is sub-optimal for assessing non-convergence in MI.
+
+The \(\widehat{R}\)-values in panel D are conform our expectations.
+The adapted version of \(\widehat{R}\) depicted here does indicate less signs of non-convergence as the number of iterations goes up.
+An exception to this general trend is the `dip' around \(t=3\).
+As diagnostic tool, this may lead to the incorrect conclusion that the non-convergence scenario has superior convergence compared to the typical scenario.
+We can explain this as a downside of the low number of iteration in iterative imputation: the adapted version of \(\widehat{R}\) can only be completely employed if the number of iterations is at least four.
+Otherwise, it is not possible to perform all three transformations to the chains of \(\theta\)-values.
+For low values of \(t\), the \(\widehat{R}\)-values conform Vehtari et al.~are, therefore, more similar to the original \(\widehat{R}\) by Gelman and Rubin.
+\newline
 
@@
 Then, convergence to a stable but highly variable state may be reached instantly.
-\textbf{Still, we expect that the incompleteness conditions will result in more signs of non-convergence.}
+Still, we expect that incompleteness conditions with higher \(p_{\rm inc}\) will result in more signs of non-convergence.
 
@@
 And for the regression coefficients (\(Q=\beta\)) the bias is most pronounced in \(\beta_1\) (the effect of \(X_1\) on \(Y\)).
-Since there is just one estimate for \(Q=r^2\) to consider, we \textbf{end up with} \(Q=\mu_Y, \sigma_Y,r^2, \beta_1\).
+Since there is just one estimate for \(Q=r^2\) to evaluate, we consider \(Q=\mu_Y, \sigma_Y,r^2, \beta_1\).
 In the figures, we present all incompleteness conditions, but only early stopping conditions where \(1 \leq T\leq50\).
@@
 Figure \ref{fig:Qs} displays the influence of incompleteness and early stopping on the estimated quantities of scientific interest, \(Q\)s.
-\textbf{{[}Add panel labels, describe panels{]}} Simulation conditions that are impacted by non-convergence portray more extreme bias in the estimated \(Q\)s, and non-nominal coverages.
+In the first row of the figure, we see the bias in the estimated descriptive statistics (\(Q=\mu_Y\) and \(Q=\sigma_Y\)).
+The second row shows the bias in the multivariate scientific estimands (\(Q=r^2\) and \(Q=\beta_1\)).
+The third row consists of the coverage rate (CR) and confidence interval width (CIW) of the estimated regression coefficient (\(Q=\beta_1\)).
 
@@
 Completely unbiased estimates are only obtained in conditions where \(p_{\rm inc}\leq.50\).
-\textbf{This is curious, since the MCAR missingness mechanism should yield unbiased univariate estimates without employing multiple imputation.} Note however, that the magnitude of the bias in these conditions is small: \(\mu_Y\) is maximally under-estimated by 0.02 units, while the true value of \(\mu_Y\) is 25.81 (\(\sigma_Y = 11.32\)).
+This is curious, since the MCAR missingness mechanism should yield unbiased univariate estimates without employing multiple imputation.
+Note however, that the magnitude of the bias in these conditions is small: \(\mu_Y\) is maximally under-estimated by 0.02 units, while the true value of \(\mu_Y\) is 25.81 (\(\sigma_Y = 11.32\)).
 We, therefore, conclude that non-convergence does not substantially impact the validity of the inferences for \(Q=\mu\).
@@
 
-In contrast to the univariate estimates, the estimated coefficient of determination is clearly affected by early stopping.
+In contrast to the univariate estimates, the estimated coefficient of determination is clearly affected by early stopping, see Figure \ref{fig:Qs}C.
 That is, the bias in \(Q\) does not only follow the incremental order of the incompleteness conditions, but also decreases with each additional iteration.
@@
 
-Figure \ref{fig:rhats-acs} displays results for the eight sets of diagnostic methods.
-The two columns in the figure represent the non-convergence diagnostics \(\widehat{R}\) and \(AC\); the four rows depict the four scalar summaries \(\theta\) under consideration.
-\textbf{{[}describe panels??{]}} We evaluate the performance of the methods by establishing whether they correctly identify conditions in which \(\bar{Q}\) is \emph{not} an unbiased, confidence-valid estimate of \(Q\).
+Figure \ref{fig:rhats-acs} displays results of the eight sets of diagnostic methods.
+The two columns in the figure represent the non-convergence diagnostics \(\widehat{R}\) and \(AC\); the rows depict the four scalar summaries \(\theta\) under consideration.
+The first row in Figure \ref{fig:rhats-acs} thus shows \(\widehat{R}\) and \(AC\) applied on the chain means.
+In the second row we see the diagnostics applied on the chain variances.
+The third row depicts the same for \(\theta=\hat{Q}\).
+And in the last row we see the diagnostics applied on the novel \(\theta\) to consider: \(\lambda_1\).
+We evaluate the performance of the methods by establishing whether they correctly identify conditions in which \(\bar{Q}\) is \emph{not} an unbiased, confidence-valid estimate of \(Q\).
 As we concluded in the last section, substantially biased estimates were only obtained in conditions where \(T\leq6\), and we observed non-nominal coverage only when \(T\leq3\).
@@
 
-\caption{Non-convergence diagnostics. [Split up into 2 or 4 figures??]}\label{fig:rhats-acs}
+\caption{Non-convergence diagnostics $\widehat{R}$ and $AC$ applied on several $\theta$s.
+The left-hand side of the figure contains $\widehat{R}$-values, and the right-hand side contains $AC$-values.
+Depicted in the rows are the scalar summaries $\theta$: chain mean, chain variance, the quantity of scientific interest $\hat{Q}$, and the first eigenvalue of the variance-covariance matrix $\lambda_1$.}\label{fig:rhats-acs}
 \end{figure}
@@
 The thresholds 1.1 and 1.01 seem too strict compared to the validity of the inferences.
-
 The threshold to diagnose significant \(AC\)-values does not appear to be appropriate in at least the current set-up, and perhaps in iterative imputation in general.
@@
 \item
-  First, check traceplots of the default \(\theta\)s (chain means and chain variances) for signs of pathological non-convergence. Adjust the imputation model if necessary.
+  First, check traceplots of the default \(\theta\)s (e.g., chain means and chain variances) for signs of pathological non-convergence. Adjust the imputation model if necessary.
 \item
-  Track multivariate \(\theta\)s over iterations, e.g., the novel \(\theta\) that is scientific model-independent.
-Or specify your own scalar summary of interest (see e.g., \citet{buur18}).
-Monitor these \(\theta\)s across iterations through visual inspection, or using a non-convergence diagnostic.
+  Subsequently decide which multivariate \(\theta\)s to track across iterations.
+The novel \(\theta\) that we propose in this study, \(\lambda_1\) is scientific model-independent and may thus always be employed.
+Alternatively, specify your own scalar summary of interest \citep[see e.g.,][]{buur18}.
+Monitor these \(\theta\)s through visual inspection, or using a non-convergence diagnostic.
 \item
   Compute non-convergence diagnostics \(\widehat{R}\) and \(AC\).
-Do not use the traditional calculation for \(\widehat{R}\) \citep{gelm92}, or the \texttt{R} function \texttt{stats::acf()} to compute \(AC\) \citep{R}.
-Instead, calculate \(\widehat{R}\) conform \citet{veht19} and compute autocorrelations manually (see e.g., \href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy}{github.com/gerkovink/shinyMice}).
+Do not use the original implementation to calculate \(\widehat{R}\) \citep{gelm92}, or the \texttt{R} function \texttt{stats::acf()} to compute \(AC\) \citep{R}.
+Instead, calculate \(\widehat{R}\) conform \citet{veht19} and compute autocorrelations manually (see e.g., \href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/}{github.com/gerkovink/shinyMice}).
 \item

### 2020-05-11 3.Thesis/Writeup/2.Manuscript.Rmd
Update reference list

@@
 By visually inspecting the imputation chains in panel A, we conclude that the two scenarios show very different convergence.
-The typically converged algorithm initially portrayes some signs of non-mixing (around $t=2$), but intermingles nicely overall.
+The typically converged algorithm initially portrays some signs of non-mixing (around $t=2$), but intermingles nicely overall.
 Additionally, there is very little trending in this scenario.
@@
 2.
-Our results also support the hyothesis that $\widehat{R}$ and $AC$ would correctly identify signs of non-convergence.
+Our results also support the hypothesis that $\widehat{R}$ and $AC$ would correctly identify signs of non-convergence.
 We observed higher $\widehat{R}$ and $AC$-values in conditions where $\bar{Q}$s were biased, invalid estimates of $Q$s.
@@
 Proper performance under a 'missing completely at random' (MCAR) mechanism is a necessary condition for any missing data method.
-It does not't, however, guarantee equal performance under different missingness mechanisms.
+It does not, however, guarantee equal performance under different missingness mechanisms.
 Future research should determine the performance of $\widehat{R}$ and $AC$ under 'missing at random' and 'missing not at random' mechanisms.

### 2020-05-11 3.Thesis/Writeup/2.Manuscript.tex
Update reference list

@@
 By visually inspecting the imputation chains in panel A, we conclude that the two scenarios show very different convergence.
-The typically converged algorithm initially portrayes some signs of non-mixing (around \(t=2\)), but intermingles nicely overall.
+The typically converged algorithm initially portrays some signs of non-mixing (around \(t=2\)), but intermingles nicely overall.
 Additionally, there is very little trending in this scenario.
@@
 \item
-  Our results also support the hyothesis that \(\widehat{R}\) and \(AC\) would correctly identify signs of non-convergence.
+  Our results also support the hypothesis that \(\widehat{R}\) and \(AC\) would correctly identify signs of non-convergence.
 We observed higher \(\widehat{R}\) and \(AC\)-values in conditions where \(\bar{Q}\)s were biased, invalid estimates of \(Q\)s.
@@
 Proper performance under a `missing completely at random' (MCAR) mechanism is a necessary condition for any missing data method.
-It does not't, however, guarantee equal performance under different missingness mechanisms.
+It does not, however, guarantee equal performance under different missingness mechanisms.
 Future research should determine the performance of \(\widehat{R}\) and \(AC\) under `missing at random' and `missing not at random' mechanisms.

### 2020-05-11 3.Thesis/Writeup/2.Manuscript.Rmd
Final version before forking

@@
 This paper provides insight into identifying non-convergence in iterative imputation algorithms.
-We conclude that in the case of drawing inference from incomplete data, convergence of the iterative imputation algorithm is often a convenience but not a necessity."
+Our study found that---in the cases considered---inferential validity was achieved after five to ten iterations, much earlier than indicated by the $\\widehat{R}$ and AC diagnostics.
+Of course, it never hurts to iterate longer, but such calculations hardly bring added value."
 keywords: missing data, iterative imputation, non-convergence, MICE
@@
 
-These remain open questions, since the convergence properties of iterative imputation algorithms have not been systematically studied [@zhu15; @buur18].
-There is no scientific consensus on how to evaluate the convergence of imputation algorithms [@taka17].
+These remain open questions since the convergence properties of iterative imputation algorithms have not been systematically studied [@buur18].
+There is no scientific consensus on how to evaluate the convergence of imputation algorithms [@zhu15; @taka17].
 Moreover, the behavior of such algorithms under certain default imputation models (e.g., 'predictive mean matching') is an entirely open question [@murr18].
@@
 For each datapoint in $y_{\rm mis}$, we sample $m \times T$ plausible values, where $m$ is the number of imputations ($\ell = 1, 2, \dots, m$) and $T$ is the number of iterations ($t = 1, 2, \dots, T$) in the imputation algorithm.
-The state space of the algorithm at a certain iteration $t$ may be summarized by a scalar summary $\theta$ (e.g., the average of the imputed values).
-The collection of $\theta$-values between $t=1$ (summarizing the state space of the algorithm at initialization) and $t=T$ (summarizing the state space for the imputed values) will be referred to as an 'imputation chain'.
+The state-space of the algorithm at a certain iteration $t$ may be summarized by a scalar summary $\theta$ (e.g., the average of the imputed values).
+The collection of $\theta$-values between $t=1$ (summarizing the state-space of the algorithm at initialization) and $t=T$ (summarizing the state-space for the imputed values) will be referred to as an 'imputation chain'.
 <!-- **[If not defined elsewhere, add: $\theta$s are scalar summaries of the algorithm at a certain iteration (e.g., chain means; the average of the imputed values in each imputation chain).] ** -->
@@
 Traditionally, the inspection of the multivariate convergence is done by visually inspecting the univariate chain means and variances over the iterations.
-The downside to these approaches is that they either focus on the univariate state space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest.
+The downside to these approaches is that they either focus on the univariate state-space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest.
 Focusing on the convergence of these outcome parameters may influence this procedure in the sense that the model of evaluation favors the model of interest.
@@
 
-To illustrate what non-mixing and non-stationarity looks like in iterative imputation algorithms, we reproduce an example from van Buuren [-@buur18, $\S$ 6.5.2].
+To illustrate what non-mixing and non-stationarity look like in iterative imputation algorithms, we reproduce an example from van Buuren [-@buur18, $\S$ 6.5.2].
 Figure \ref{fig:non-conv} displays two scenarios from the example.
@@
 After running the imputation algorithm for a certain number of iterations, researchers are encouraged to produce traceplots.
-In a traceplot, a scalar summary of the state space of the algorithm $\theta$ is plotted against the iteration number, as depicted in Figure \ref{fig:non-conv}.
+In a traceplot, a scalar summary of the state-space of the algorithm $\theta$ is plotted against the iteration number, as depicted in Figure \ref{fig:non-conv}.
 The default scalar summaries to inspect for the MICE algorithm are chain means and chain variances.
@@
 Moreover, traceplots are typically only used to inspect  univariate $\theta$s.
-Monitoring univariate summaries of the state space of the algorithm may be insufficient because MICE is not only concerned with a single column, but the entire multivariate distribution of the imputations.
+Monitoring univariate summaries of the state-space of the algorithm may be insufficient because MICE is not only concerned with a single column, but the entire multivariate distribution of the imputations.
 Ideally, we would monitor a multivariate $\theta$.
@@
 In a traceplot, the iteration number is plotted against a certain $\theta$.
-$\theta$s are scalar summaries of the state space of the algorithm at a specific iteration.
+$\theta$s are scalar summaries of the state-space of the algorithm at a specific iteration.
 -->
@@
 
-To monitor non-convergence in iterative imputation algorithms, we may  summarize the state space of the algorithm with several $\theta$s.
-The downside to the current $\theta$s is that they either focus on the univariate state space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest.
+To monitor non-convergence in iterative imputation algorithms, we may  summarize the state-space of the algorithm with several $\theta$s.
+The downside to the current $\theta$s is that they either focus on the univariate state-space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest.
 Ideally, one would like to evaluate a model-independent $\theta$ that summarizes the multivariate nature of the data.
@@
 $\lambda_{1,\ell}$ then summarizes the largest possible amount of covariance in the $m$ completed datasets {$y_{\rm obs}, \dot y_{\rm imp, \ell}$}.
-By definition, each $\lambda_{1, \ell}$-value is equal to the variance of the first component that would be obtained by performing principal component analysis on {$y_{\rm obs}, \dot y_{\rm imp, \ell}$}.
-As scalar summary, $\lambda_{1}$ has the appealing property that it is not dependent on the substantive model of interest, yet still summarizes the multivariate state space of the algorithm.
-\newline
+By definition, each $\lambda_{1, \ell}$-value is equal to the variance of the first component that would be obtained by performing principal component analysis (PCA) on {$y_{\rm obs}, \dot y_{\rm imp, \ell}$}.
+As scalar summary, $\lambda_{1}$ has the appealing property that it is not dependent on the substantive model of interest, yet still summarizes the multivariate state-space of the algorithm.
 
 <!-- **Since convergence of iterative imputation algorithms is in distribution, there is not a unique point at which the algorithm reaches a converged state.
-We can only evaluate scalar summaries of the multivariate state space of the algorithm, $\theta$s.
+We can only evaluate scalar summaries of the multivariate state-space of the algorithm, $\theta$s.
 The recommended scalar summaries to evaluate are chain means and chain variances.
@@
 
-\noindent In this study, we consider each of the four $\theta$s that were discussed above.
+In this study, we consider each of the four $\theta$s that were discussed above.
 Namely, two univariate scalar summaries (chain mean and chain variance), a model-dependent multivariate summary $\hat{Q}$, and the novel model-independent multivariate summary $\lambda_1$.
@@
 Therefore, we will evaluate the appropriateness of both $\widehat{R}$ versions, in addition to $AC$.
-Other methods are outside the scope of this study, because they e.g.
+Other methods are outside the scope of this study because they e.g.
 assume that values within chains represent independent samples, whereas MI only uses the final iteration to impute.
@@
 
-\noindent From the between- and within-chain variances we compute a weighted average, $\widehat{\operatorname{var}}^{+}$, which **approximates** the total variance of $\theta$.
-$\widehat{R}$ is then obtained as a ratio between the total variance and the within-chain variance:
+\noindent From the between- and within-chain variances we compute a weighted average, $\widehat{\operatorname{var}}^{+}$, which over-estimates the total variance of $\theta$.
+$\widehat{R}$ is then obtained as a ratio between this total variance and the within-chain variance:
 
@@
 are, therefore, more similar to the original $\widehat{R}$ by Gelman and Rubin.
-\newline
 <!-- However, the superior performance of the typical convergence over the pathological non-convergence is less prominent, even flipped for $t<4$.  -->
 
-\noindent In short, both non-convergence diagnostics have a version that may be appropriate for iterative imputation algorithms.
+In short, both non-convergence diagnostics have a version that may be appropriate for iterative imputation algorithms.
 We will compute $\widehat{R}$ conform @veht19, and calculate $AC$ manually.
@@
 
-The aim of the simulation study is to determine to what extend non-convergence affects iterative imputation algorithms.
+The simulation study aims to determine to what extend non-convergence affects iterative imputation algorithms.
 Specifically, we are interested in the validity of inferences drawn from incomplete data using the MICE algorithm.
@@
 
-To induce non-convergence, we consider two sets of simulation conditions: 'incompleteness' and 'early stopping'.
-Incompleteness refers to the  proportion of incomplete cases in dataset $y$, $p_{\rm inc}$.
+To induce non-convergence, we consider two sets of simulation conditions: 'missingness' and 'early stopping'.
+Missingness refers to the severity of the missing data problem, which we determine by the proportion of incomplete cases in dataset $y$, $p_{\rm inc}$.
 Early stopping is defined by the number of iterations in the imputation algorithm, $T$.
@@
 
-The incompleteness conditions are chosen to reflect the difficulty of the missingness problem.
+The missingness conditions are chosen to reflect the difficulty of the missingness problem.
 The underlying assumption is that a low proportion of incomplete cases $p_{\rm inc}$ leads to quick algorithmic convergence, since there is a lot of information in the observed data.
@@
 Then, convergence to a stable but highly variable state may be reached instantly.
-Still, we expect that incompleteness conditions with higher $p_{\rm inc}$ will result in more signs of non-convergence.
+Still, we expect that missingness conditions with higher $p_{\rm inc}$ will result in more signs of non-convergence.
 
@@
 <!-- **[REMOVE?] $Q$s are the targets that will be estimated in each simulation repetition, after first *amputing* the complete data with varying missingness proportions, then imputing the missingness with a varying number of iterations, then estimating the $Q$s on the $m$ completed datasets, and finally pooling the $m$ estimates $\hat{Q}_\ell$ to obtain a single $\bar{Q}$ for each $Q$.
-We try to predict the effect of non-convergence (i.e., the difference between $Q$ and $\bar{Q}$) with the convergence diagnostics $\widehat{R}$ and $AC$, by applying these methods to several scalar summaries of the state space of the algorithm, $\theta$s.** -->
+We try to predict the effect of non-convergence (i.e., the difference between $Q$ and $\bar{Q}$) with the convergence diagnostics $\widehat{R}$ and $AC$, by applying these methods to several scalar summaries of the state-space of the algorithm, $\theta$s.** -->
 
@@
 We, therefore, suspect that $\widehat{R}$ may over-estimate signs of non-convergence in iterative imputation algorithms, compared to the validity of estimates.
-In contrast to this, it may occur that signs of non-convergence are under-estimated by $\widehat{R}$, in exceptional cases where the initial values of the algorithm are not appropriately over-dispersed [@broo98, p.
+In contrast to this, signs of non-convergence can be under-estimated by $\widehat{R}$, in exceptional cases where the initial values of the algorithm are not appropriately over-dispersed [@broo98, p.
 437].
@@
 
-In each simulation repetition, we *ampute* the complete dataset conform the incompleteness conditions.
+In each simulation repetition, we *ampute* the complete dataset conform the missingness conditions.
 The proportion of incomplete cases $p_{\rm inc}$ varies between 5% and 95%, in incremental steps ($p_{\rm inc}$ = .05, .25, .50, .75, .95).
@@
 I.e., the probability to be missing is the same for all $N \times k$ cells in $y$, conditional on $p_{\rm inc}$.
-We use the 'mice' package [function `mice::ampute()`; @mice] to obtain an incomplete dataset {$y_{\rm obs}, y_{\rm mis}$} for each of the five incompleteness conditions.
-
-We impute the missing values in {$y_{\rm obs}, y_{\rm mis}$} according to the early stopping conditions.
+We use the 'mice' package [function `mice::ampute()`; @mice] to obtain an incomplete dataset {$y_{\rm obs}, y_{\rm mis}$} for each of the five missingness conditions.
+
+We impute the missing values in {$y_{\rm obs}, \dot y_{\rm mis}$} according to the early stopping conditions.
 That is, we vary the number of iterations in the imputation algorithm between one and one hundred ($T = 1,2,.,100$).
 All imputation procedures are performed using `mice` [function `mice()`; @mice], with Bayesian linear regression imputation, and five imputation chains ($m=5$).
-This results in five sets of imputations for each of the one hundred early stopping conditions, $y_{\rm imp, \ell}$.
-
-Subsequently, we obtain the completed data per imputation, {$y_{\rm obs}, y_{\rm imp, \ell}$}, by combining the imputed data with the observed data [function `mice::complete()`; @mice].
+This results in $m=5$ sets of imputations for each of the one hundred early stopping conditions, $\dot y_{\rm imp, \ell}$.
+
+Subsequently, we obtain the completed data per imputation, {$y_{\rm obs}, \dot y_{\rm imp, \ell}$}, by combining the imputed data with the observed data [function `mice::complete()`; @mice].
 We estimate univariate $Q$s directly from the completed data, by averaging across imputations to obtain $\bar{Q}$.
@@
 We consider eight sets of diagnostic methods to identify non-convergence, by applying $\widehat{R}$ and $AC$ on four different $\theta$s.
-Univariate $\theta$s are obtained from $y_{\rm imp, \ell}$.
-The model-independent multivariate $\theta$, $\lambda_{1}$, is obtained from {$y_{\rm obs}, y_{\rm imp, \ell}$}.
-The model-dependent multivariate $\theta$ is one of the quantities of scientific interest: the estimated regression coefficients in {$y_{\rm obs}, y_{\rm imp, \ell}$}.
+Univariate $\theta$s are obtained from $\dot y_{\rm imp, \ell}$.
+The model-independent multivariate $\theta$, $\lambda_{1}$, is obtained from {$y_{\rm obs}, \dot y_{\rm imp, \ell}$} as the variance in the first PCA component [function `stats::princomp`; R].
+The model-dependent multivariate $\theta$ is one of the quantities of scientific interest: the estimated regression coefficients in {$y_{\rm obs}, \dot y_{\rm imp, \ell}$}.
 We apply the two non-convergence diagnostics on the $\theta$s by implementing the adapted version of $\widehat{R}$ and by manually programming the $AC$ function.
@@
 
-Our results show the performance of the MICE algorithm under different conditions of incompleteness and early stopping.
+Our results show the performance of the MICE algorithm under different conditions of missingness and early stopping.
 For reasons of brevity, we only discuss results for the worst-performing estimates in terms of bias.
@@
 Since there is just one estimate for $Q=r^2$ to evaluate, we consider $Q=\mu_Y, \sigma_Y,r^2, \beta_1$.
-In the figures, we present all incompleteness conditions, but only early stopping conditions where $1 \leq T\leq50$.
+In the figures, we present all missingness conditions, but only early stopping conditions where $1 \leq T\leq50$.
 That is, the results are more or less stable for conditions where $T \geq 30$.
@@
 
-Figure \ref{fig:Qs} displays the influence of incompleteness and early stopping on the estimated quantities of scientific interest, $Q$s.
-In the first row of the figure, we see the bias in the estimated descriptive statistics ($Q=\mu_Y$ and $Q=\sigma_Y$).
-The second row shows the bias in the multivariate scientific estimands ($Q=r^2$ and $Q=\beta_1$).
-The third row consists of the coverage rate (CR) and confidence interval width (CIW) of the estimated regression coefficient ($Q=\beta_1$).
+Figure \ref{fig:Qs} displays the influence of missingness and early stopping on the estimated quantities of scientific interest, $Q$s.
+In the first row of the figure, we see the bias in the estimated descriptive statistics ($Q=\mu_Y$ and $Q=\sigma_Y$; panels A and B).
+The second row shows the bias in the multivariate scientific estimands ($Q=r^2$ and $Q=\beta_1$; C and D).
+The third row consists of the coverage rate (CR) and confidence interval width (CIW) of the estimated regression coefficient ($Q=\beta_1$; E and F).
 
@@
 Completely unbiased estimates are only obtained in conditions where $p_{\rm inc}\leq.50$.
-This is curious, since the MCAR missingness mechanism should yield unbiased univariate estimates without employing multiple imputation.
+This is curious since the MCAR missingness mechanism should yield unbiased univariate estimates without employing multiple imputation.
 Note however, that the magnitude of the bias in these conditions is small: $\mu_Y$ is maximally under-estimated by 0.02 units, while the true value of $\mu_Y$ is 25.81 ($\sigma_Y = 11.32$).
@@
 Completely unbiased estimates are only obtained in conditions where $p_{\rm inc}=.05$.
-The other incompleteness conditions are impacted by non-convergence in the order of increasing $p_{\rm inc}$.
+The other missingness conditions are impacted by non-convergence in the order of increasing $p_{\rm inc}$.
 Similar to $Q=\mu$, the magnitude of the bias in $Q=\sigma$ is negligible.
-We therefore conclude that neither one of the univariate $Q$s is affected substantially by non-convergence.
+We, therefore, conclude that neither one of the univariate $Q$s is affected substantially by non-convergence.
 <!-- in the first iteration.
@@
 In contrast to the univariate estimates, the estimated coefficient of determination is clearly affected by early stopping, see Figure \ref{fig:Qs}C.
-That is, the bias in $Q$ does not only follow the incremental order of the incompleteness conditions, but also decreases with each additional iteration.
-The number of iterations necessary to reach stable, non-improving estimates depends on $p_{\rm inc}$.
+The bias in $Q$ appears to decrease with each additional iteration, while the magnitude of the bias still depends on the proportion of incomplete cases.
+Therefore, we see that the number of iterations that is necessary to reach stable, non-improving estimates also differs across missingness conditions.
 In conditions where $p_{\rm inc}=.05$, estimates are unbiased after one iteration.
@@
 The highest number of iterations needed to obtain valid inferences is seven ($p_{\rm inc}=.95$; $T\geq7$).
-**[interpret sign of bias?
-ADD what this implies: dat - zelfs bij een gebiasede methode - je maar 7 iteraties nodig hebt om het eindresultaat te bereiken.]** -->
-
-<!-- while the best performance occurs with less missingness and more iterations.
+while the best performance occurs with less missingness and more iterations.
 Only conditions with the lowest missingness proportion ($p_{\rm inc}=0.05$) results in unbiased estimates for every number of iterations.
@@
 Completely unbiased estimates are only obtained in conditions where $T\geq3$ and $p_{\rm inc}\leq .50$.
-Despite persistent bias in conditions where $p_{\rm inc}\geq.75$, the coverage rates are non-improving after after just three iterations, irrespective of $p_{\rm inc}$.
+Despite persistent bias in conditions where $p_{\rm inc}\geq.75$, the coverage rates are non-improving after just three iterations, irrespective of $p_{\rm inc}$.
 We conclude that any condition where $T\geq3$ yields approximately nominal coverage rates, and thus confidence-valid estimates.
@@
 <!-- around $\bar{Q}$ in conditions with higher missingness proportions **results in the inclusion of** the true value of $Q$ despite of bias. -->
-We conclude that approximately unbiased estimates may be obtained in conditions where $T\geq7$, whereas confidence-valid estimates require at most three iterations. \newline
+We conclude that approximately unbiased estimates may be obtained in conditions where $T\geq7$, whereas confidence-valid estimates require at most three iterations.
 
@@
 
-\noindent These results demonstrate that the estimates of univariate scientific estimands $Q$ are not impacted by early stopping of the MICE algorithm or the proportion of incomplete cases in $y$.
+### Summary. 
+These results demonstrate that the estimates of univariate scientific estimands $Q$ are not impacted by early stopping of the MICE algorithm or the proportion of incomplete cases in $y$.
 Unbiased estimates may be obtained after just one iteration.
@@
 The two columns in the figure represent the  non-convergence diagnostics $\widehat{R}$ and $AC$; the rows depict the four scalar summaries $\theta$ under consideration.
-The first row in Figure \ref{fig:rhats-acs} thus shows $\widehat{R}$ and $AC$ applied on the chain means.
-In the second row we see the diagnostics applied on the chain variances.
-The third row depicts the same for $\theta=\hat{Q}$.
-And in the last row we see the diagnostics applied on the novel $\theta$ to consider: $\lambda_1$.
+The first row in Figure \ref{fig:rhats-acs} thus shows $\widehat{R}$ and $AC$ applied on the chain means (panels A and B).
+In the second row, we see the diagnostics applied on the chain variances (C and D).
+The third row depicts the same for $\theta=\hat{Q}$ (E and F).
+And in the last row, we see the diagnostics applied on the novel $\theta$ to consider: $\lambda_1$ (G and H).
 We evaluate the performance of the methods by establishing whether they correctly identify conditions in which $\bar{Q}$ is *not* an unbiased, confidence-valid estimate of $Q$.
@@
 The $\widehat{R}$-values in these conditions do not taper off after 30 to 40 iterations, but rather between 40 to 50 iterations.
-The highest number of iterations at which non-convergence is still diagnosed according to the thresholds are: 11 for $\widehat{R}>1.2$, 23 for $\widehat{R}>1.1$, and 100 for $\widehat{R}>1.01$ (not shown).
+The highest number of iterations at which non-convergence is still diagnosed according to the thresholds are 11 for $\widehat{R}>1.2$, 23 for $\widehat{R}>1.1$, and 100 for $\widehat{R}>1.01$ (not shown).
 
@@
 This suggests improving stationarity up-to ten, or even thirty iterations, depending on the missingness proportion.
-Using the threshold, we diagnose non-convergence at some point in all incompleteness conditions.
+Using the threshold, we diagnose non-convergence at some point in all missingness conditions.
 However, this occurs only after reasonably decreasing $AC$-values are obtained---again suggesting that non-convergence is incorrectly diagnosed.
-\newline
 
@@
 
-\noindent Overall, the methods to diagnose non-convergence perform as expected: they indicate more signs of non-convergence in conditions with worse performance in terms of bias and confidence-validity.
+### Summary.
+Overall, the methods to diagnose non-convergence perform as expected: they indicate more signs of non-convergence in conditions with worse performance in terms of bias and confidence-validity.
 Under low to moderate missingness, it does not seem to matter on which $\theta$ the non-convergence diagnostics are applied.
@@
 Identifying non-convergence may be a crucial aspect of drawing valid statistical inferences from incomplete data.
-We concluded that the current practice of visually inspecting non-convergence through traceplots of univariate scalar summaries of the state space of the algorithm does not suffice.
+We concluded that the current practice of visually inspecting non-convergence through traceplots of univariate scalar summaries of the state-space of the algorithm does not suffice.
 We, therefore, considered eight sets of diagnostic methods to identify non-convergence: two non-convergence diagnostics ($\widehat{R}$ and $AC$), two univariate $\theta$s, a model-dependent multivariate $\theta$, and a novel, model-independent multivariate $\theta$.
 Signs of non-convergence due to early stopping were identified with each of the methods.
-Bias due to high proportions of incomplete cases were only identified with multivariate $\theta$s.
-
-Since we obtained approximately unbiased, confidence valid estimates after at most seven iterations, we conclude that the $\widehat{R}$ threshold 1.2 is the most appropriate diagnostic cut-off.
+Bias due to high proportions of incomplete cases was only identified with multivariate $\theta$s.
+
+Since we obtained approximately unbiased, confidence-valid estimates after at most seven iterations, we conclude that the $\widehat{R}$ threshold 1.2 is the most appropriate diagnostic cut-off.
 The threshold to diagnose non-stationarity with $AC$ does not seem to apply.
@@
 Based on our results, we formulate several recommendations for empirical researchers who employ iterative imputation to draw inferences from incomplete data.
-We suggest that non-convergence may still be evaluated visually, but in addition to inspecting univariate summaries of the state space of the algorithm ($\theta$s, e.g., chain means), multivariate $\theta$s should be considered.
+We suggest that non-convergence may still be evaluated visually, but in addition to inspecting univariate summaries of the state-space of the algorithm ($\theta$s, e.g., chain means), multivariate $\theta$s should be considered.
 We propose the following steps:
@@
 2.
-Subsequently decide which multivariate $\theta$s to track across iterations.
+Subsequently, decide which multivariate $\theta$s to track across iterations.
 The novel $\theta$ that we propose in this study, $\lambda_1$ is scientific model-independent and may thus always be employed.
@@
 Compute non-convergence diagnostics $\widehat{R}$ and $AC$.
-Do not use the original implementation to calculate $\widehat{R}$ [@gelm92], or the `R` function `stats::acf()` to compute $AC$ [@R].
+Do *not* use the original implementation to calculate $\widehat{R}$ [@gelm92], or the `R` function `stats::acf()` to compute $AC$ [@R].
 Instead, calculate $\widehat{R}$ conform @veht19 and compute autocorrelations manually (see e.g., [github.com/gerkovink/shinyMice](https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/)).
@@
 4.
-Use the threshold $\widehat{R}>1.2$ to diagnose non-mixing, and assess stationarity by plotting the autocorrelation over iterations.
+And finally, use the threshold $\widehat{R}>1.2$ to diagnose non-mixing, and assess stationarity by plotting the autocorrelation over iterations.
 Keep iterating until the threshold for mixing is overcome, and until reasonably decreasing asymptotic $AC$-values are obtained.
@@
 Even though we have demonstrated the appropriateness of $\widehat{R}$ and $AC$ as non-convergence diagnostics in this study, results may not extrapolate to other situations.
-The current simulation conditions were restricted to a single missingness mechanism.
+In this study, we only considered the iterative imputation algorithm implemented in the `mice` package.
+The performance of the non-convergence diagnostics may depend on the characteristics of this specific algorithm.
+A potential threat to the ability to detect non-mixing with $\widehat{R}$, for example, is the assumption of over-dispersed initial states.
+In `mice`, the algorithm is initiated by sampling starting values at random from the observed data, whereas other imputation software may use the average of the observed cases.
+The latter could lead to a systematical under-estimation of the variability in the first few iterations and violate the assumption of over-dispersion.
+In those cases, it would be more difficult to detect non-convergence with $\widehat{R}$.
+Therefore, the diagnostic might not be  appropriate outside of `mice`.
+
+Moreover, the current simulation conditions were restricted to a single missingness mechanism.
 Proper performance under a 'missing completely at random' (MCAR) mechanism is a necessary condition for any missing data method.
@@
 Different imputations models might converge more poorly.
-Moreover, the imputation model may even be mis-specified on purpose to induce different levels of non-convergence.
+
+The imputation model may even be mis-specified on purpose, to induce non-convergence of different severity levels.
 Qualitatively, we have shown that $\widehat{R}$ and $AC$ are appropriate under clear violation of convergence.
 Quantitatively, we have only demonstrated that they can identify signs of non-convergence under MCAR, when the imputation model is a correctly specified Bayesian linear regression model.
-**[ADD limitation that other MI packages may have different way to obtain initial values!]** \newline \newline
+\newline \newline
+
+<!-- **[ADD limitation that other MI packages may have different way to obtain initial values!]** -->
 
@@
 Any further iterations would then burn computational resources without improving the statistical inferences.
-To conclude, in the case of drawing inference from incomplete data, convergence of the iterative imputation algorithm is often a convenience but not a necessity.
-**[Our study found that - in the cases considered - inferential validity was achieved after 5-10 iterations, much earlier than indicated by the $\widehat{R}$ and AC diagnostics.
-Of course, it never hurts to iterate longer, but such calculations hardly bring added value.]**
+Our study found that---in the cases considered---inferential validity was achieved after five to ten iterations, much earlier than indicated by the $\widehat{R}$ and AC diagnostics.
+Of course, it never hurts to iterate longer, but such calculations hardly bring added value.
 

### 2020-05-11 3.Thesis/Writeup/2.Manuscript.tex

@@
 This paper provides insight into identifying non-convergence in iterative imputation algorithms.
-We conclude that in the case of drawing inference from incomplete data, convergence of the iterative imputation algorithm is often a convenience but not a necessity.
+Our study found that---in the cases considered---inferential validity was achieved after five to ten iterations, much earlier than indicated by the \(\widehat{R}\) and AC diagnostics.
+Of course, it never hurts to iterate longer, but such calculations hardly bring added value.
 \end{abstract}
@@
 
-These remain open questions, since the convergence properties of iterative imputation algorithms have not been systematically studied \citep{zhu15, buur18}.
-There is no scientific consensus on how to evaluate the convergence of imputation algorithms \citep{taka17}.
+These remain open questions since the convergence properties of iterative imputation algorithms have not been systematically studied \citep{buur18}.
+There is no scientific consensus on how to evaluate the convergence of imputation algorithms \citep{zhu15, taka17}.
 Moreover, the behavior of such algorithms under certain default imputation models (e.g., `predictive mean matching') is an entirely open question \citep{murr18}.
@@
 For each datapoint in \(y_{\rm mis}\), we sample \(m \times T\) plausible values, where \(m\) is the number of imputations (\(\ell = 1, 2, \dots, m\)) and \(T\) is the number of iterations (\(t = 1, 2, \dots, T\)) in the imputation algorithm.
-The state space of the algorithm at a certain iteration \(t\) may be summarized by a scalar summary \(\theta\) (e.g., the average of the imputed values).
-The collection of \(\theta\)-values between \(t=1\) (summarizing the state space of the algorithm at initialization) and \(t=T\) (summarizing the state space for the imputed values) will be referred to as an `imputation chain'.
+The state-space of the algorithm at a certain iteration \(t\) may be summarized by a scalar summary \(\theta\) (e.g., the average of the imputed values).
+The collection of \(\theta\)-values between \(t=1\) (summarizing the state-space of the algorithm at initialization) and \(t=T\) (summarizing the state-space for the imputed values) will be referred to as an `imputation chain'.
 
@@
 
-To illustrate what non-mixing and non-stationarity looks like in iterative imputation algorithms, we reproduce an example from van Buuren \citeyearpar[\(\S\) 6.5.2]{buur18}.
+To illustrate what non-mixing and non-stationarity look like in iterative imputation algorithms, we reproduce an example from van Buuren \citeyearpar[\(\S\) 6.5.2]{buur18}.
 Figure \ref{fig:non-conv} displays two scenarios from the example.
@@
 After running the imputation algorithm for a certain number of iterations, researchers are encouraged to produce traceplots.
-In a traceplot, a scalar summary of the state space of the algorithm \(\theta\) is plotted against the iteration number, as depicted in Figure \ref{fig:non-conv}.
+In a traceplot, a scalar summary of the state-space of the algorithm \(\theta\) is plotted against the iteration number, as depicted in Figure \ref{fig:non-conv}.
 The default scalar summaries to inspect for the MICE algorithm are chain means and chain variances.
@@
 Moreover, traceplots are typically only used to inspect univariate \(\theta\)s.
-Monitoring univariate summaries of the state space of the algorithm may be insufficient because MICE is not only concerned with a single column, but the entire multivariate distribution of the imputations.
+Monitoring univariate summaries of the state-space of the algorithm may be insufficient because MICE is not only concerned with a single column, but the entire multivariate distribution of the imputations.
 Ideally, we would monitor a multivariate \(\theta\).
@@
 
-To monitor non-convergence in iterative imputation algorithms, we may summarize the state space of the algorithm with several \(\theta\)s.
-The downside to the current \(\theta\)s is that they either focus on the univariate state space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest.
+To monitor non-convergence in iterative imputation algorithms, we may summarize the state-space of the algorithm with several \(\theta\)s.
+The downside to the current \(\theta\)s is that they either focus on the univariate state-space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest.
 Ideally, one would like to evaluate a model-independent \(\theta\) that summarizes the multivariate nature of the data.
@@
 \(\lambda_{1,\ell}\) then summarizes the largest possible amount of covariance in the \(m\) completed datasets \{\(y_{\rm obs}, \dot y_{\rm imp, \ell}\)\}.
-By definition, each \(\lambda_{1, \ell}\)-value is equal to the variance of the first component that would be obtained by performing principal component analysis on \{\(y_{\rm obs}, \dot y_{\rm imp, \ell}\)\}.
-As scalar summary, \(\lambda_{1}\) has the appealing property that it is not dependent on the substantive model of interest, yet still summarizes the multivariate state space of the algorithm.
-\newline
+By definition, each \(\lambda_{1, \ell}\)-value is equal to the variance of the first component that would be obtained by performing principal component analysis (PCA) on \{\(y_{\rm obs}, \dot y_{\rm imp, \ell}\)\}.
+As scalar summary, \(\lambda_{1}\) has the appealing property that it is not dependent on the substantive model of interest, yet still summarizes the multivariate state-space of the algorithm.
 
-\noindent In this study, we consider each of the four \(\theta\)s that were discussed above.
+In this study, we consider each of the four \(\theta\)s that were discussed above.
 Namely, two univariate scalar summaries (chain mean and chain variance), a model-dependent multivariate summary \(\hat{Q}\), and the novel model-independent multivariate summary \(\lambda_1\).
@@
 Therefore, we will evaluate the appropriateness of both \(\widehat{R}\) versions, in addition to \(AC\).
-Other methods are outside the scope of this study, because they e.g.~assume that values within chains represent independent samples, whereas MI only uses the final iteration to impute.
+Other methods are outside the scope of this study because they e.g.~assume that values within chains represent independent samples, whereas MI only uses the final iteration to impute.
 
@@
 
-\noindent From the between- and within-chain variances we compute a weighted average, \(\widehat{\operatorname{var}}^{+}\), which \textbf{approximates} the total variance of \(\theta\).
-\(\widehat{R}\) is then obtained as a ratio between the total variance and the within-chain variance:
+\noindent From the between- and within-chain variances we compute a weighted average, \(\widehat{\operatorname{var}}^{+}\), which over-estimates the total variance of \(\theta\).
+\(\widehat{R}\) is then obtained as a ratio between this total variance and the within-chain variance:
 
@@
 For low values of \(t\), the \(\widehat{R}\)-values conform Vehtari et al.~are, therefore, more similar to the original \(\widehat{R}\) by Gelman and Rubin.
-\newline
 
-\noindent In short, both non-convergence diagnostics have a version that may be appropriate for iterative imputation algorithms.
+In short, both non-convergence diagnostics have a version that may be appropriate for iterative imputation algorithms.
 We will compute \(\widehat{R}\) conform \citet{veht19}, and calculate \(AC\) manually.
@@
 
-To induce non-convergence, we consider two sets of simulation conditions: `incompleteness' and `early stopping'.
-Incompleteness refers to the proportion of incomplete cases in dataset \(y\), \(p_{\rm inc}\).
+To induce non-convergence, we consider two sets of simulation conditions: `missingness' and `early stopping'.
+Missingness refers to the severity of the missing data problem, which we determine by the proportion of incomplete cases in dataset \(y\), \(p_{\rm inc}\).
 Early stopping is defined by the number of iterations in the imputation algorithm, \(T\).
 
-The incompleteness conditions are chosen to reflect the difficulty of the missingness problem.
+The missingness conditions are chosen to reflect the difficulty of the missingness problem.
 The underlying assumption is that a low proportion of incomplete cases \(p_{\rm inc}\) leads to quick algorithmic convergence, since there is a lot of information in the observed data.
@@
 Then, convergence to a stable but highly variable state may be reached instantly.
-Still, we expect that incompleteness conditions with higher \(p_{\rm inc}\) will result in more signs of non-convergence.
+Still, we expect that missingness conditions with higher \(p_{\rm inc}\) will result in more signs of non-convergence.
 
@@
 We, therefore, suspect that \(\widehat{R}\) may over-estimate signs of non-convergence in iterative imputation algorithms, compared to the validity of estimates.
-In contrast to this, it may occur that signs of non-convergence are under-estimated by \(\widehat{R}\), in exceptional cases where the initial values of the algorithm are not appropriately over-dispersed \citep[p.~437]{broo98}.
+In contrast to this, signs of non-convergence can be under-estimated by \(\widehat{R}\), in exceptional cases where the initial values of the algorithm are not appropriately over-dispersed \citep[p.~437]{broo98}.
 In \texttt{mice}, initial values are chosen randomly from the observed data, hence we cannot be certain of over-dispersion in the initial values.
@@
 
-In each simulation repetition, we \emph{ampute} the complete dataset conform the incompleteness conditions.
+In each simulation repetition, we \emph{ampute} the complete dataset conform the missingness conditions.
 The proportion of incomplete cases \(p_{\rm inc}\) varies between 5\% and 95\%, in incremental steps (\(p_{\rm inc}\) = .05, .25, .50, .75, .95).
@@
 I.e., the probability to be missing is the same for all \(N \times k\) cells in \(y\), conditional on \(p_{\rm inc}\).
-We use the `mice' package \citep[function \texttt{mice::ampute()};][]{mice} to obtain an incomplete dataset \{\(y_{\rm obs}, y_{\rm mis}\)\} for each of the five incompleteness conditions.
+We use the `mice' package \citep[function \texttt{mice::ampute()};][]{mice} to obtain an incomplete dataset \{\(y_{\rm obs}, y_{\rm mis}\)\} for each of the five missingness conditions.
 
-We impute the missing values in \{\(y_{\rm obs}, y_{\rm mis}\)\} according to the early stopping conditions.
+We impute the missing values in \{\(y_{\rm obs}, \dot y_{\rm mis}\)\} according to the early stopping conditions.
 That is, we vary the number of iterations in the imputation algorithm between one and one hundred (\(T = 1,2,.,100\)).
 All imputation procedures are performed using \texttt{mice} \citep[function \texttt{mice()};][]{mice}, with Bayesian linear regression imputation, and five imputation chains (\(m=5\)).
-This results in five sets of imputations for each of the one hundred early stopping conditions, \(y_{\rm imp, \ell}\).
+This results in \(m=5\) sets of imputations for each of the one hundred early stopping conditions, \(\dot y_{\rm imp, \ell}\).
 
-Subsequently, we obtain the completed data per imputation, \{\(y_{\rm obs}, y_{\rm imp, \ell}\)\}, by combining the imputed data with the observed data \citep[function \texttt{mice::complete()};][]{mice}.
+Subsequently, we obtain the completed data per imputation, \{\(y_{\rm obs}, \dot y_{\rm imp, \ell}\)\}, by combining the imputed data with the observed data \citep[function \texttt{mice::complete()};][]{mice}.
 We estimate univariate \(Q\)s directly from the completed data, by averaging across imputations to obtain \(\bar{Q}\).
@@
 We consider eight sets of diagnostic methods to identify non-convergence, by applying \(\widehat{R}\) and \(AC\) on four different \(\theta\)s.
-Univariate \(\theta\)s are obtained from \(y_{\rm imp, \ell}\).
-The model-independent multivariate \(\theta\), \(\lambda_{1}\), is obtained from \{\(y_{\rm obs}, y_{\rm imp, \ell}\)\}.
-The model-dependent multivariate \(\theta\) is one of the quantities of scientific interest: the estimated regression coefficients in \{\(y_{\rm obs}, y_{\rm imp, \ell}\)\}.
+Univariate \(\theta\)s are obtained from \(\dot y_{\rm imp, \ell}\).
+The model-independent multivariate \(\theta\), \(\lambda_{1}\), is obtained from \{\(y_{\rm obs}, \dot y_{\rm imp, \ell}\)\} as the variance in the first PCA component {[}function \texttt{stats::princomp}; R{]}.
+The model-dependent multivariate \(\theta\) is one of the quantities of scientific interest: the estimated regression coefficients in \{\(y_{\rm obs}, \dot y_{\rm imp, \ell}\)\}.
 We apply the two non-convergence diagnostics on the \(\theta\)s by implementing the adapted version of \(\widehat{R}\) and by manually programming the \(AC\) function.
@@
 Since there is just one estimate for \(Q=r^2\) to evaluate, we consider \(Q=\mu_Y, \sigma_Y,r^2, \beta_1\).
-In the figures, we present all incompleteness conditions, but only early stopping conditions where \(1 \leq T\leq50\).
+In the figures, we present all missingness conditions, but only early stopping conditions where \(1 \leq T\leq50\).
 That is, the results are more or less stable for conditions where \(T \geq 30\).
@@
 
-Figure \ref{fig:Qs} displays the influence of incompleteness and early stopping on the estimated quantities of scientific interest, \(Q\)s.
-In the first row of the figure, we see the bias in the estimated descriptive statistics (\(Q=\mu_Y\) and \(Q=\sigma_Y\)).
-The second row shows the bias in the multivariate scientific estimands (\(Q=r^2\) and \(Q=\beta_1\)).
-The third row consists of the coverage rate (CR) and confidence interval width (CIW) of the estimated regression coefficient (\(Q=\beta_1\)).
+Figure \ref{fig:Qs} displays the influence of missingness and early stopping on the estimated quantities of scientific interest, \(Q\)s.
+In the first row of the figure, we see the bias in the estimated descriptive statistics (\(Q=\mu_Y\) and \(Q=\sigma_Y\); panels A and B).
+The second row shows the bias in the multivariate scientific estimands (\(Q=r^2\) and \(Q=\beta_1\); C and D).
+The third row consists of the coverage rate (CR) and confidence interval width (CIW) of the estimated regression coefficient (\(Q=\beta_1\); E and F).
 
@@
 Completely unbiased estimates are only obtained in conditions where \(p_{\rm inc}\leq.50\).
-This is curious, since the MCAR missingness mechanism should yield unbiased univariate estimates without employing multiple imputation.
+This is curious since the MCAR missingness mechanism should yield unbiased univariate estimates without employing multiple imputation.
 Note however, that the magnitude of the bias in these conditions is small: \(\mu_Y\) is maximally under-estimated by 0.02 units, while the true value of \(\mu_Y\) is 25.81 (\(\sigma_Y = 11.32\)).
@@
 Completely unbiased estimates are only obtained in conditions where \(p_{\rm inc}=.05\).
-The other incompleteness conditions are impacted by non-convergence in the order of increasing \(p_{\rm inc}\).
+The other missingness conditions are impacted by non-convergence in the order of increasing \(p_{\rm inc}\).
 Similar to \(Q=\mu\), the magnitude of the bias in \(Q=\sigma\) is negligible.
-We therefore conclude that neither one of the univariate \(Q\)s is affected substantially by non-convergence.
+We, therefore, conclude that neither one of the univariate \(Q\)s is affected substantially by non-convergence.
 
@@
 In contrast to the univariate estimates, the estimated coefficient of determination is clearly affected by early stopping, see Figure \ref{fig:Qs}C.
-That is, the bias in \(Q\) does not only follow the incremental order of the incompleteness conditions, but also decreases with each additional iteration.
-The number of iterations necessary to reach stable, non-improving estimates depends on \(p_{\rm inc}\).
+The bias in \(Q\) appears to decrease with each additional iteration, while the magnitude of the bias still depends on the proportion of incomplete cases.
+Therefore, we see that the number of iterations that is necessary to reach stable, non-improving estimates also differs across missingness conditions.
 In conditions where \(p_{\rm inc}=.05\), estimates are unbiased after one iteration.
@@
 Completely unbiased estimates are only obtained in conditions where \(T\geq3\) and \(p_{\rm inc}\leq .50\).
-Despite persistent bias in conditions where \(p_{\rm inc}\geq.75\), the coverage rates are non-improving after after just three iterations, irrespective of \(p_{\rm inc}\).
+Despite persistent bias in conditions where \(p_{\rm inc}\geq.75\), the coverage rates are non-improving after just three iterations, irrespective of \(p_{\rm inc}\).
 We conclude that any condition where \(T\geq3\) yields approximately nominal coverage rates, and thus confidence-valid estimates.
@@
 Since conditions with higher \(p_{\rm inc}\) result in both wider CIs and more severe bias, the true value of \(Q\) may be included in the CI, despite of the bias in \(\bar{Q}\).
-We conclude that approximately unbiased estimates may be obtained in conditions where \(T\geq7\), whereas confidence-valid estimates require at most three iterations. \newline
+We conclude that approximately unbiased estimates may be obtained in conditions where \(T\geq7\), whereas confidence-valid estimates require at most three iterations.
 
-\noindent These results demonstrate that the estimates of univariate scientific estimands \(Q\) are not impacted by early stopping of the MICE algorithm or the proportion of incomplete cases in \(y\).
+\hypertarget{summary.}{%
+\subsubsection{Summary.}\label{summary.}}
+
+These results demonstrate that the estimates of univariate scientific estimands \(Q\) are not impacted by early stopping of the MICE algorithm or the proportion of incomplete cases in \(y\).
 Unbiased estimates may be obtained after just one iteration.
@@
 The two columns in the figure represent the non-convergence diagnostics \(\widehat{R}\) and \(AC\); the rows depict the four scalar summaries \(\theta\) under consideration.
-The first row in Figure \ref{fig:rhats-acs} thus shows \(\widehat{R}\) and \(AC\) applied on the chain means.
-In the second row we see the diagnostics applied on the chain variances.
-The third row depicts the same for \(\theta=\hat{Q}\).
-And in the last row we see the diagnostics applied on the novel \(\theta\) to consider: \(\lambda_1\).
+The first row in Figure \ref{fig:rhats-acs} thus shows \(\widehat{R}\) and \(AC\) applied on the chain means (panels A and B).
+In the second row, we see the diagnostics applied on the chain variances (C and D).
+The third row depicts the same for \(\theta=\hat{Q}\) (E and F).
+And in the last row, we see the diagnostics applied on the novel \(\theta\) to consider: \(\lambda_1\) (G and H).
 We evaluate the performance of the methods by establishing whether they correctly identify conditions in which \(\bar{Q}\) is \emph{not} an unbiased, confidence-valid estimate of \(Q\).
@@
 The \(\widehat{R}\)-values in these conditions do not taper off after 30 to 40 iterations, but rather between 40 to 50 iterations.
-The highest number of iterations at which non-convergence is still diagnosed according to the thresholds are: 11 for \(\widehat{R}>1.2\), 23 for \(\widehat{R}>1.1\), and 100 for \(\widehat{R}>1.01\) (not shown).
+The highest number of iterations at which non-convergence is still diagnosed according to the thresholds are 11 for \(\widehat{R}>1.2\), 23 for \(\widehat{R}>1.1\), and 100 for \(\widehat{R}>1.01\) (not shown).
 
@@
 This suggests improving stationarity up-to ten, or even thirty iterations, depending on the missingness proportion.
-Using the threshold, we diagnose non-convergence at some point in all incompleteness conditions.
+Using the threshold, we diagnose non-convergence at some point in all missingness conditions.
 However, this occurs only after reasonably decreasing \(AC\)-values are obtained---again suggesting that non-convergence is incorrectly diagnosed.
-\newline
 
-\noindent Overall, the methods to diagnose non-convergence perform as expected: they indicate more signs of non-convergence in conditions with worse performance in terms of bias and confidence-validity.
+\hypertarget{summary.-overall-the-methods-to-diagnose-non-convergence-perform-as-expected-they-indicate-more-signs-of-non-convergence-in-conditions-with-worse-performance-in-terms-of-bias-and-confidence-validity.-under-low-to-moderate-missingness-it-does-not-seem-to-matter-on-which-theta-the-non-convergence-diagnostics-are-applied.-if-we-look-at-complete-unbiasedness-however-we-notice-that-univariate-thetas-fail-to-diagnose-the-persistent-bias-in-conditions-where-p_rm-incgeq.75.}{%
+\subsubsection{\texorpdfstring{Summary.
+Overall, the methods to diagnose non-convergence perform as expected: they indicate more signs of non-convergence in conditions with worse performance in terms of bias and confidence-validity.
 Under low to moderate missingness, it does not seem to matter on which \(\theta\) the non-convergence diagnostics are applied.
-If we look at complete unbiasedness, however, we notice that univariate \(\theta\)s fail to diagnose the persistent bias in conditions where \(p_{\rm inc}\geq.75\).
+If we look at complete unbiasedness, however, we notice that univariate \(\theta\)s fail to diagnose the persistent bias in conditions where \(p_{\rm inc}\geq.75\).}{Summary.
+Overall, the methods to diagnose non-convergence perform as expected: they indicate more signs of non-convergence in conditions with worse performance in terms of bias and confidence-validity.
+Under low to moderate missingness, it does not seem to matter on which \textbackslash theta the non-convergence diagnostics are applied.
+If we look at complete unbiasedness, however, we notice that univariate \textbackslash thetas fail to diagnose the persistent bias in conditions where p\_\{\textbackslash rm inc\}\textbackslash geq.75.}}\label{summary.-overall-the-methods-to-diagnose-non-convergence-perform-as-expected-they-indicate-more-signs-of-non-convergence-in-conditions-with-worse-performance-in-terms-of-bias-and-confidence-validity.-under-low-to-moderate-missingness-it-does-not-seem-to-matter-on-which-theta-the-non-convergence-diagnostics-are-applied.-if-we-look-at-complete-unbiasedness-however-we-notice-that-univariate-thetas-fail-to-diagnose-the-persistent-bias-in-conditions-where-p_rm-incgeq.75.}}
 
@@
 Identifying non-convergence may be a crucial aspect of drawing valid statistical inferences from incomplete data.
-We concluded that the current practice of visually inspecting non-convergence through traceplots of univariate scalar summaries of the state space of the algorithm does not suffice.
+We concluded that the current practice of visually inspecting non-convergence through traceplots of univariate scalar summaries of the state-space of the algorithm does not suffice.
 We, therefore, considered eight sets of diagnostic methods to identify non-convergence: two non-convergence diagnostics (\(\widehat{R}\) and \(AC\)), two univariate \(\theta\)s, a model-dependent multivariate \(\theta\), and a novel, model-independent multivariate \(\theta\).
 Signs of non-convergence due to early stopping were identified with each of the methods.
-Bias due to high proportions of incomplete cases were only identified with multivariate \(\theta\)s.
+Bias due to high proportions of incomplete cases was only identified with multivariate \(\theta\)s.
 
-Since we obtained approximately unbiased, confidence valid estimates after at most seven iterations, we conclude that the \(\widehat{R}\) threshold 1.2 is the most appropriate diagnostic cut-off.
+Since we obtained approximately unbiased, confidence-valid estimates after at most seven iterations, we conclude that the \(\widehat{R}\) threshold 1.2 is the most appropriate diagnostic cut-off.
 The threshold to diagnose non-stationarity with \(AC\) does not seem to apply.
@@
 Based on our results, we formulate several recommendations for empirical researchers who employ iterative imputation to draw inferences from incomplete data.
-We suggest that non-convergence may still be evaluated visually, but in addition to inspecting univariate summaries of the state space of the algorithm (\(\theta\)s, e.g., chain means), multivariate \(\theta\)s should be considered.
+We suggest that non-convergence may still be evaluated visually, but in addition to inspecting univariate summaries of the state-space of the algorithm (\(\theta\)s, e.g., chain means), multivariate \(\theta\)s should be considered.
 We propose the following steps:
@@
 \item
-  Subsequently decide which multivariate \(\theta\)s to track across iterations.
+  Subsequently, decide which multivariate \(\theta\)s to track across iterations.
 The novel \(\theta\) that we propose in this study, \(\lambda_1\) is scientific model-independent and may thus always be employed.
@@
   Compute non-convergence diagnostics \(\widehat{R}\) and \(AC\).
-Do not use the original implementation to calculate \(\widehat{R}\) \citep{gelm92}, or the \texttt{R} function \texttt{stats::acf()} to compute \(AC\) \citep{R}.
+Do \emph{not} use the original implementation to calculate \(\widehat{R}\) \citep{gelm92}, or the \texttt{R} function \texttt{stats::acf()} to compute \(AC\) \citep{R}.
 Instead, calculate \(\widehat{R}\) conform \citet{veht19} and compute autocorrelations manually (see e.g., \href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/}{github.com/gerkovink/shinyMice}).
 \item
-  Use the threshold \(\widehat{R}>1.2\) to diagnose non-mixing, and assess stationarity by plotting the autocorrelation over iterations.
+  And finally, use the threshold \(\widehat{R}>1.2\) to diagnose non-mixing, and assess stationarity by plotting the autocorrelation over iterations.
 Keep iterating until the threshold for mixing is overcome, and until reasonably decreasing asymptotic \(AC\)-values are obtained.
@@
 Even though we have demonstrated the appropriateness of \(\widehat{R}\) and \(AC\) as non-convergence diagnostics in this study, results may not extrapolate to other situations.
-The current simulation conditions were restricted to a single missingness mechanism.
+In this study, we only considered the iterative imputation algorithm implemented in the \texttt{mice} package.
+The performance of the non-convergence diagnostics may depend on the characteristics of this specific algorithm.
+A potential threat to the ability to detect non-mixing with \(\widehat{R}\), for example, is the assumption of over-dispersed initial states.
+In \texttt{mice}, the algorithm is initiated by sampling starting values at random from the observed data, whereas other imputation software may use the average of the observed cases.
+The latter could lead to a systematical under-estimation of the variability in the first few iterations and violate the assumption of over-dispersion.
+In those cases, it would be more difficult to detect non-convergence with \(\widehat{R}\).
+Therefore, the diagnostic might not be appropriate outside of \texttt{mice}.
+
+Moreover, the current simulation conditions were restricted to a single missingness mechanism.
 Proper performance under a `missing completely at random' (MCAR) mechanism is a necessary condition for any missing data method.
@@
 Different imputations models might converge more poorly.
-Moreover, the imputation model may even be mis-specified on purpose to induce different levels of non-convergence.
+
+The imputation model may even be mis-specified on purpose, to induce non-convergence of different severity levels.
 Qualitatively, we have shown that \(\widehat{R}\) and \(AC\) are appropriate under clear violation of convergence.
 Quantitatively, we have only demonstrated that they can identify signs of non-convergence under MCAR, when the imputation model is a correctly specified Bayesian linear regression model.
-\textbf{{[}ADD limitation that other MI packages may have different way to obtain initial values!{]}} \newline \newline
+\newline \newline
 
@@
 Any further iterations would then burn computational resources without improving the statistical inferences.
-To conclude, in the case of drawing inference from incomplete data, convergence of the iterative imputation algorithm is often a convenience but not a necessity.
-\textbf{{[}Our study found that - in the cases considered - inferential validity was achieved after 5-10 iterations, much earlier than indicated by the \(\widehat{R}\) and AC diagnostics.
-Of course, it never hurts to iterate longer, but such calculations hardly bring added value.{]}}
+Our study found that---in the cases considered---inferential validity was achieved after five to ten iterations, much earlier than indicated by the \(\widehat{R}\) and AC diagnostics.
+Of course, it never hurts to iterate longer, but such calculations hardly bring added value.
 

### 2020-05-11 3.Thesis/Writeup/2.Manuscript.Rmd
Final version thesis

@@
   num: 1
-  org: Department of Methodology and Statistics, Utrecht University, Utrecht, The Netherlands
-corrauth: "Hanne Oberman, Sjoerd Groenman building, Utrecht Science Park, Utrecht, The Netherlands."
+  org: Department of Methodology and Statistics, Utrecht University, Utrecht, The Netherlands.
+corrauth: "Hanne Oberman, Sjoerd Groenman building, Padualaan 14, 3584 CH, Utrecht, The Netherlands."
 email: h.i.oberman@uu.nl
@@
 This paper provides insight into identifying non-convergence in iterative imputation algorithms.
-Our study found that---in the cases considered---inferential validity was achieved after five to ten iterations, much earlier than indicated by the $\\widehat{R}$ and AC diagnostics.
-Of course, it never hurts to iterate longer, but such calculations hardly bring added value."
+Our study found that---in the cases considered---inferential validity was achieved after five to ten iterations, much earlier than indicated by the $\\widehat{R}$ and $AC$ diagnostics.
+We conclude that it never hurts to iterate longer, but such calculations hardly bring added value."
 keywords: missing data, iterative imputation, non-convergence, MICE
@@
 With MICE, imputed values are drawn from the posterior predictive distribution of the missing values.
-The algorithm is named after the iterative algorithmic procedure by which imputed values are generated: a multivariate distribution is obtained by iterating over a sequence of univariate imputations.
-The iterative nature of algorithms like MICE introduces a potential threat to the validity of the imputations: What if the algorithm has not converged?
+The algorithm is named after its iterative nature: a multivariate distribution is obtained by iterating over a sequence of univariate imputations.
+Iteration, however, also introduces a potential threat to the validity of the imputations: What if the algorithm has not converged?
 Are the imputations then to be trusted?
@@
 Moreover, the behavior of such algorithms under certain default imputation models (e.g., 'predictive mean matching') is an entirely open question [@murr18].
-Therefore, algorithmic convergence should be monitored carefully, but this is not straightforward.
+Therefore, algorithmic convergence should be monitored carefully---although this is not straightforward.
 Iterative imputation algorithms such as MICE are special cases of Markov chain Monte Carlo (MCMC) methods.
@@
 Since MCMC algorithms do not reach a unique point at which convergence is established, diagnostic methods may only identify signs of *non*-convergence [@hoff09].
-Several non-convergence diagnostics exist, but it is not known whether these are appropriate for iterative imputation algorithms.
+Several non-convergence diagnostics exist, but it is not known whether these are appropriate within the imputation framework.
 <!-- "a weak diagnostics is better than no diagnostic at all" [@cowl96]. -->
@@
 We then address the impact of inducing non-convergence in iterative imputation algorithms through model-based simulation in `R` [@R].
-For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular `mice` package [@mice] in `R`.
+For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular `mice` package in `R` [@mice].
 The aim of the simulation study is to determine whether unbiased, confidence-valid inferences may be obtained if the algorithm has not (yet) converged.
@@
 The collection of observed data values in $y$ is denoted by $y_{\rm obs}$; the missing part of $y$ is referred to as $y_{\rm mis}$.
-For each datapoint in $y_{\rm mis}$, we sample $m \times T$ plausible values, where $m$ is the number of imputations ($\ell = 1, 2, \dots, m$) and $T$ is the number of iterations ($t = 1, 2, \dots, T$) in the imputation algorithm.
+For each datapoint in $y_{\rm mis}$, we sample $m \times T$ plausible values, where $m$ is the number of imputations ($\ell = 1, 2, \dots, m$) and $T$ is the number of iterations in the imputation algorithm ($t = 1, 2, \dots, T$).
 The state-space of the algorithm at a certain iteration $t$ may be summarized by a scalar summary $\theta$ (e.g., the average of the imputed values).
@@
 Figure \ref{fig:non-conv} displays two scenarios from the example.
-The graph on the left-hand side of the figure shows typical convergence of an iterative imputation algorithm.
+The panel on the left-hand side of the figure shows typical convergence of an iterative imputation algorithm.
 The right-hand side displays pathological non-convergence, induced by purposefully mis-specifying the imputation model.
 Each line portrays one imputation chain (i.e., the values of a scalar summary $\theta$ across iterations).
-The $\theta$ depicted here is the 'chain mean' of variable $j$, which is defined as the average of variable $j$ in $\dot y_{\rm imp, \ell}$.
+The $\theta$ depicted here is the 'chain mean' of variable $j$, which is defined as the average of variable $j$ in the $m$ sets of imputations $\dot y_{\rm imp, \ell}$.
 
@@
 The algorithm with the mis-specified model yields imputed values that are on average a factor two larger than those of the typically converged algorithm.
-It is obvious that non-convergence impacts the distribution of the imputed values in this example, $\dot y_{\rm imp, \ell}$.
-This effect may translate into the distribution of the completed data {$y_{\rm obs}, \dot y_{\rm imp, \ell}$}, which consequently affects the estimated quantities of scientific interest $\hat{Q}_{\ell}$ and the pooled estimate $\bar{Q}$.
-$\bar{Q}$ is then a  biased, invalid estimate of $Q$, since systematically higher imputed values lead to bias, and under-estimation of the variance within imputation chains poses a threat to confidence-validity.
+It is obvious that non-convergence in this example impacts the distribution of the imputed values per imputation $\dot y_{\rm imp, \ell}$.
+This effect may translate into the distribution of the $m$ sets of completed data {$y_{\rm obs}, \dot y_{\rm imp, \ell}$}, which consequently affects the estimated quantities of scientific interest $\hat{Q}_{\ell}$, and finally the pooled estimate $\bar{Q}$.
+$\bar{Q}$ is then a  biased, invalid estimate of $Q$.
 Therefore, it is important to reach algorithmic convergence in iterative imputation algorithms.
+
+<!-- , since systematically higher imputed values lead to bias, and under-estimation of the variance within imputation chains poses a threat to confidence-validity. -->
 
@@
 The default scalar summaries to inspect for the MICE algorithm are chain means and chain variances.
-Non-convergence is diagnosed if the imputation chains are not freely intermingled with one another, or if the chains show definite trends [@buur18].
+Non-convergence is diagnosed if the imputation chains are not freely intermingled with one another, or if the chains show definite trends [@buur18, $\S$ 6.5.2].
 
@@
 A suggestion by @buur18 for a multivariate $\theta$ to monitor is the estimated quantity of scientific interest $\hat{Q}$, which usually is multivariate in nature.
-This $\theta$ is computed as the estimate in each imputation $\ell$, $\hat{Q}_\ell$, across iterations.
+This $\theta$ is computed as the estimate in each imputation, $\hat{Q}_\ell$, across iterations.
 Implementing this, however, might be somewhat too technical for empirical researchers.
 Besides, this scalar summary is not model-independent.
-That is, this $\theta$ is not universal to all complete data problems, while one of the advantages of iterative imputation is that the missing data problem and scientific problem are solved independently.
+That is, $\theta=\hat{Q}$ is not universal to all complete data problems, while one of the advantages of iterative imputation is that the missing data problem and scientific problem are solved independently.
 Focusing on the convergence of outcome parameters may influence the iterative imputation procedure in the sense that the model of evaluation favors the model of interest.
@@
 Let $\lambda_{1, \ell} \geq \lambda_{2, \ell} \geq ...
-\geq \lambda_{j, \ell}$ be the eigenvalues of $S_\ell$ in each imputation $\ell$.
-$\lambda_{1,\ell}$ then summarizes the largest possible amount of covariance in the $m$ completed datasets {$y_{\rm obs}, \dot y_{\rm imp, \ell}$}.
-By definition, each $\lambda_{1, \ell}$-value is equal to the variance of the first component that would be obtained by performing principal component analysis (PCA) on {$y_{\rm obs}, \dot y_{\rm imp, \ell}$}.
-As scalar summary, $\lambda_{1}$ has the appealing property that it is not dependent on the substantive model of interest, yet still summarizes the multivariate state-space of the algorithm.
+\geq \lambda_{j, \ell}$ be the eigenvalues of $S_\ell$ in each imputation.
+The first eigenvalue, $\lambda_{1,\ell}$, then summarizes the largest possible amount of covariance in each completed dataset {$y_{\rm obs}, \dot y_{\rm imp, \ell}$}.
+By definition, $\lambda_{1}$-values are equal to the variance of the first component that would be obtained by performing principal component analysis (PCA) on the completed data.
+As scalar summary, $\lambda_{1}$ has the appealing property that it is not dependent on the model of scientific interest, yet still summarizes the multivariate state-space of the algorithm.
 
@@
 
-In this study, we consider each of the four $\theta$s that were discussed above.
+In this study, we consider each of the four $\theta$s that we discussed.
 Namely, two univariate scalar summaries (chain mean and chain variance), a model-dependent multivariate summary $\hat{Q}$, and the novel model-independent multivariate summary $\lambda_1$.
@@
 As recommended by e.g.
-@cowl96 we will use the potential scale reduction factor $\widehat{R}$ ['Gelman-Rubin statistic'; @gelm92] to evaluate mixing, and autocorrelation [$AC$; @scha97; @gelm13] to diagnose trending.
+@cowl96 we will use the potential scale reduction factor $\widehat{R}$ to evaluate mixing ['Gelman-Rubin statistic'; @gelm92], and autocorrelation to diagnose trending [$AC$; @scha97; @gelm13].
 With a recently proposed adaptation, $\widehat{R}$ might also serve to diagnose non-stationarity, but this has not yet been thoroughly investigated [@veht19].
@@
 Other methods are outside the scope of this study because they e.g.
-assume that values within chains represent independent samples, whereas MI only uses the final iteration to impute.
+assume that values within chains represent independent samples, whereas the MICE algorithm only uses the final iteration to produce imputations.
 
@@
 The definition of the diagnostic itself is equal to the original $\widehat{R}$.
-We follow Vehtari et al.'s formulation (2019, p.
-5).
+Therefore, we may follow Vehtari et al.'s formulation (2019, p.
+5) to define both $\widehat{R}$ versions.
 Let $m$ be the total number of chains, $T$ the number of iterations per chain (where $T\geq2$), and $\theta$ the scalar summary of interest.
@@
 When the sampled values in each chain are independent of the chain's initial value, the mixing component of convergence is satisfied.
-The variance between chains $B$ is then equivalent to the variance within chains $W$, and $\widehat{R}$-values will be close to one.
+The variance between chains, $B$, is then equivalent to the variance within chains, $W$, and $\widehat{R}$-values will be close to one.
 High $\widehat{R}$-values thus indicate non-convergence.
@@
 Panel B shows two versions of the $AC$: the default calculation with `R` function `stats::acf()` [@R], and manual calculation as the correlation between $\theta$ in iteration $t$ and $\theta$ in iteration $t+1$.
-Panel C displays the traditional computation of $\widehat{R}$ conform @gelm92 (original), and in panel D we see $\widehat{R}$ as computed by implementing Vehtari et al.'s recommendations (adapted).
+Panel C displays the traditional computation of $\widehat{R}$ conform Gelman and Rubin (original), and in panel D we see $\widehat{R}$ as computed by implementing Vehtari et al.'s recommendations (adapted).
 
@@
 The function is optimized for performance when $t\geq50$ [@box15], while the default number of iterations in iterative imputation is often much lower.
-Therefore, we compute $AC$ manually, see the solid line.
+Therefore, we compute $AC$ manually (presented with solid lines in panel B).
 These $AC$-values behave as expected from visual inspection.
@@
 
-In panel C, we do not see a lot of difference between the two scenarios.
+In panel C, we do not see a lot of difference between the two scenarios either.
 The $\widehat{R}$-values are even somewhat lower for the non-convergence scenario than for the typical convergence.
@@
 As diagnostic tool, this may lead to the incorrect conclusion that the non-convergence scenario has superior convergence compared to the typical scenario.
-We can explain this as a downside of the low number of iteration in iterative imputation: the adapted version of $\widehat{R}$ can only be completely employed if the number of iterations is at least four.
+We can explain this as a downside of the low number of iterations in iterative imputation: the adapted version of $\widehat{R}$ can only be completely employed if the number of iterations is at least four.
 Otherwise, it is not possible to perform all three transformations to the chains of $\theta$-values.
@@
 We will compute $\widehat{R}$ conform @veht19, and calculate $AC$ manually.
-The diagnostics will be applied to the four types of scalar summaries $\theta$ described above: chain means, chain variances, a quantity of scientific interest, and the first eigenvalue of the variance-covariance matrix.
-We evaluate the performance of the eight sets of diagnostic methods (two diagnostics; four $\theta$s) through simulation.
+The diagnostics will be applied to the four types of scalar summaries $\theta$ described earlier: chain means, chain variances, a quantity of scientific interest, and the first eigenvalue of the variance-covariance matrix.
+We evaluate the performance of these eight sets of diagnostic methods (two diagnostics; four $\theta$s) through simulation.
 
@@
 And, whether invalid inferences due to non-convergence may be detected using eight sets of diagnostic methods.
-We will conclude that the algorithm is sufficiently converged for practical purposes when the estimated quantities of scientific interest $\bar{Q}$ are unbiased and confidence-valid estimates of ${Q}$.
+If the estimated quantities of scientific interest $\bar{Q}$ are unbiased and confidence-valid estimates of ${Q}$, we will conclude that the algorithm is sufficiently converged for practical purposes.
 In contrast to this, we defined *approximate* convergence as the most converged state that the algorithm can obtain.
@@
 The underlying assumption is that a low proportion of incomplete cases $p_{\rm inc}$ leads to quick algorithmic convergence, since there is a lot of information in the observed data.
-Higher $p_{\rm inc}$ yields slower convergence---unless there is so little information in $y_{\rm obs}$ that this is outweighed by the random component in the imputation algorithm.
+A higher $p_{\rm inc}$ yields slower convergence---unless there is so little information in $y_{\rm obs}$ that this is outweighed by the random component in the imputation algorithm.
 Then, convergence to a stable but highly variable state may be reached instantly.
@@
 The assumption inherent to the early stopping conditions is that terminating the imputation algorithm too early causes non-convergence.
-Generally, the algorithm will not reach convergence if $T=1$, because the imputed values in the first iteration (at $t=1$) depend on the initial values of the algorithm [which are sampled randomly from the set of observed datapoints; @buur18].
+Generally, the algorithm will not reach convergence if $T=1$, because the imputed values in the first iteration (at $t=1$) depend on the starting values [which are sampled randomly from the set of observed datapoints; @buur18].
 As the number of iterations increases, the imputation chains should become independent of the initial values, until approximate convergence is reached (i.e., when an additional iteration does not lead to a more converged state).
@@
 The number of simulation repetitions is 1000, and the simulation set-up is summarized in the pseudo-code below.
-The complete `R` script of the simulation study is available from [github.com/gerkovink/shinyMice](https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/).
+The complete `R` script of the simulation study is available from [github.com/hanneoberman/MissingThePoint](https://github.com/hanneoberman/MissingThePoint).
 
@@
 That is, the results are more or less stable for conditions where $T \geq 30$.
-Full results are available from [github.com/gerkovink/shinyMice](https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/).
+Full results are available from [github.com/hanneoberman/MissingThePoint](https://github.com/hanneoberman/MissingThePoint).
 
@@
 Do *not* use the original implementation to calculate $\widehat{R}$ [@gelm92], or the `R` function `stats::acf()` to compute $AC$ [@R].
-Instead, calculate $\widehat{R}$ conform @veht19 and compute autocorrelations manually (see e.g., [github.com/gerkovink/shinyMice](https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/)).
+Instead, calculate $\widehat{R}$ conform @veht19 and compute autocorrelations manually (see e.g., [github.com/hanneoberman/MissingThePoint](https://github.com/hanneoberman/MissingThePoint)).
 

### 2020-05-11 3.Thesis/Writeup/2.Manuscript.tex
Final version thesis

@@
 
-\affiliation{\affilnum{1}{Department of Methodology and Statistics, Utrecht University, Utrecht, The Netherlands}}
+\affiliation{\affilnum{1}{Department of Methodology and Statistics, Utrecht University, Utrecht, The Netherlands.}}
 
-\corrauth{Hanne Oberman, Sjoerd Groenman building, Utrecht Science Park, Utrecht, The Netherlands.}
+\corrauth{Hanne Oberman, Sjoerd Groenman building, Padualaan 14, 3584 CH, Utrecht, The Netherlands.}
 
@@
 This paper provides insight into identifying non-convergence in iterative imputation algorithms.
-Our study found that---in the cases considered---inferential validity was achieved after five to ten iterations, much earlier than indicated by the \(\widehat{R}\) and AC diagnostics.
-Of course, it never hurts to iterate longer, but such calculations hardly bring added value.
+Our study found that---in the cases considered---inferential validity was achieved after five to ten iterations, much earlier than indicated by the \(\widehat{R}\) and \(AC\) diagnostics.
+We conclude that it never hurts to iterate longer, but such calculations hardly bring added value.
 \end{abstract}
@@
 Moreover, the behavior of such algorithms under certain default imputation models (e.g., `predictive mean matching') is an entirely open question \citep{murr18}.
-Therefore, algorithmic convergence should be monitored carefully, but this is not straightforward.
+Therefore, algorithmic convergence should be monitored carefully---although this is not straightforward.
 Iterative imputation algorithms such as MICE are special cases of Markov chain Monte Carlo (MCMC) methods.
@@
 Since MCMC algorithms do not reach a unique point at which convergence is established, diagnostic methods may only identify signs of \emph{non}-convergence \citep{hoff09}.
-Several non-convergence diagnostics exist, but it is not known whether these are appropriate for iterative imputation algorithms.
+Several non-convergence diagnostics exist, but it is not known whether these are appropriate within the imputation framework.
 
@@
 We then address the impact of inducing non-convergence in iterative imputation algorithms through model-based simulation in \texttt{R} \citep{R}.
-For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular \texttt{mice} package \citep{mice} in \texttt{R}.
+For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular \texttt{mice} package in \texttt{R} \citep{mice}.
 The aim of the simulation study is to determine whether unbiased, confidence-valid inferences may be obtained if the algorithm has not (yet) converged.
@@
 The collection of observed data values in \(y\) is denoted by \(y_{\rm obs}\); the missing part of \(y\) is referred to as \(y_{\rm mis}\).
-For each datapoint in \(y_{\rm mis}\), we sample \(m \times T\) plausible values, where \(m\) is the number of imputations (\(\ell = 1, 2, \dots, m\)) and \(T\) is the number of iterations (\(t = 1, 2, \dots, T\)) in the imputation algorithm.
+For each datapoint in \(y_{\rm mis}\), we sample \(m \times T\) plausible values, where \(m\) is the number of imputations (\(\ell = 1, 2, \dots, m\)) and \(T\) is the number of iterations in the imputation algorithm (\(t = 1, 2, \dots, T\)).
 The state-space of the algorithm at a certain iteration \(t\) may be summarized by a scalar summary \(\theta\) (e.g., the average of the imputed values).
@@
 Figure \ref{fig:non-conv} displays two scenarios from the example.
-The graph on the left-hand side of the figure shows typical convergence of an iterative imputation algorithm.
+The panel on the left-hand side of the figure shows typical convergence of an iterative imputation algorithm.
 The right-hand side displays pathological non-convergence, induced by purposefully mis-specifying the imputation model.
 Each line portrays one imputation chain (i.e., the values of a scalar summary \(\theta\) across iterations).
-The \(\theta\) depicted here is the `chain mean' of variable \(j\), which is defined as the average of variable \(j\) in \(\dot y_{\rm imp, \ell}\).
+The \(\theta\) depicted here is the `chain mean' of variable \(j\), which is defined as the average of variable \(j\) in the \(m\) sets of imputations \(\dot y_{\rm imp, \ell}\).
 
@@
 The algorithm with the mis-specified model yields imputed values that are on average a factor two larger than those of the typically converged algorithm.
-It is obvious that non-convergence impacts the distribution of the imputed values in this example, \(\dot y_{\rm imp, \ell}\).
-This effect may translate into the distribution of the completed data \{\(y_{\rm obs}, \dot y_{\rm imp, \ell}\)\}, which consequently affects the estimated quantities of scientific interest \(\hat{Q}_{\ell}\) and the pooled estimate \(\bar{Q}\).
-\(\bar{Q}\) is then a biased, invalid estimate of \(Q\), since systematically higher imputed values lead to bias, and under-estimation of the variance within imputation chains poses a threat to confidence-validity.
+It is obvious that non-convergence in this example impacts the distribution of the imputed values per imputation \(\dot y_{\rm imp, \ell}\).
+This effect may translate into the distribution of the \(m\) sets of completed data \{\(y_{\rm obs}, \dot y_{\rm imp, \ell}\)\}, which consequently affects the estimated quantities of scientific interest \(\hat{Q}_{\ell}\), and finally the pooled estimate \(\bar{Q}\).
+\(\bar{Q}\) is then a biased, invalid estimate of \(Q\).
 Therefore, it is important to reach algorithmic convergence in iterative imputation algorithms.
@@
 The default scalar summaries to inspect for the MICE algorithm are chain means and chain variances.
-Non-convergence is diagnosed if the imputation chains are not freely intermingled with one another, or if the chains show definite trends \citep{buur18}.
+Non-convergence is diagnosed if the imputation chains are not freely intermingled with one another, or if the chains show definite trends \citep[\(\S\) 6.5.2]{buur18}.
 
@@
 A suggestion by \citet{buur18} for a multivariate \(\theta\) to monitor is the estimated quantity of scientific interest \(\hat{Q}\), which usually is multivariate in nature.
-This \(\theta\) is computed as the estimate in each imputation \(\ell\), \(\hat{Q}_\ell\), across iterations.
+This \(\theta\) is computed as the estimate in each imputation, \(\hat{Q}_\ell\), across iterations.
 Implementing this, however, might be somewhat too technical for empirical researchers.
 Besides, this scalar summary is not model-independent.
-That is, this \(\theta\) is not universal to all complete data problems, while one of the advantages of iterative imputation is that the missing data problem and scientific problem are solved independently.
+That is, \(\theta=\hat{Q}\) is not universal to all complete data problems, while one of the advantages of iterative imputation is that the missing data problem and scientific problem are solved independently.
 Focusing on the convergence of outcome parameters may influence the iterative imputation procedure in the sense that the model of evaluation favors the model of interest.
@@
 Let \(\lambda_{1, \ell} \geq \lambda_{2, \ell} \geq ...
-\geq \lambda_{j, \ell}\) be the eigenvalues of \(S_\ell\) in each imputation \(\ell\).
-\(\lambda_{1,\ell}\) then summarizes the largest possible amount of covariance in the \(m\) completed datasets \{\(y_{\rm obs}, \dot y_{\rm imp, \ell}\)\}.
-By definition, each \(\lambda_{1, \ell}\)-value is equal to the variance of the first component that would be obtained by performing principal component analysis (PCA) on \{\(y_{\rm obs}, \dot y_{\rm imp, \ell}\)\}.
-As scalar summary, \(\lambda_{1}\) has the appealing property that it is not dependent on the substantive model of interest, yet still summarizes the multivariate state-space of the algorithm.
+\geq \lambda_{j, \ell}\) be the eigenvalues of \(S_\ell\) in each imputation.
+The first eigenvalue, \(\lambda_{1,\ell}\), then summarizes the largest possible amount of covariance in each completed dataset \{\(y_{\rm obs}, \dot y_{\rm imp, \ell}\)\}.
+By definition, \(\lambda_{1}\)-values are equal to the variance of the first component that would be obtained by performing principal component analysis (PCA) on the completed data.
+As scalar summary, \(\lambda_{1}\) has the appealing property that it is not dependent on the model of scientific interest, yet still summarizes the multivariate state-space of the algorithm.
 
-In this study, we consider each of the four \(\theta\)s that were discussed above.
+In this study, we consider each of the four \(\theta\)s that we discussed.
 Namely, two univariate scalar summaries (chain mean and chain variance), a model-dependent multivariate summary \(\hat{Q}\), and the novel model-independent multivariate summary \(\lambda_1\).
@@
 We consider only two of them that may be appropriate for imputation algorithms---one to monitor signs of non-mixing, and one for non-stationarity.
-As recommended by e.g.~\citet{cowl96} we will use the potential scale reduction factor \(\widehat{R}\) \citep[`Gelman-Rubin statistic';][]{gelm92} to evaluate mixing, and autocorrelation \citep[\(AC\);][]{scha97, gelm13} to diagnose trending.
+As recommended by e.g.~\citet{cowl96} we will use the potential scale reduction factor \(\widehat{R}\) to evaluate mixing \citep[`Gelman-Rubin statistic';][]{gelm92}, and autocorrelation to diagnose trending \citep[\(AC\);][]{scha97, gelm13}.
 With a recently proposed adaptation, \(\widehat{R}\) might also serve to diagnose non-stationarity, but this has not yet been thoroughly investigated \citep{veht19}.
 Therefore, we will evaluate the appropriateness of both \(\widehat{R}\) versions, in addition to \(AC\).
-Other methods are outside the scope of this study because they e.g.~assume that values within chains represent independent samples, whereas MI only uses the final iteration to impute.
+Other methods are outside the scope of this study because they e.g.~assume that values within chains represent independent samples, whereas the MICE algorithm only uses the final iteration to produce imputations.
 
@@
 The definition of the diagnostic itself is equal to the original \(\widehat{R}\).
-We follow Vehtari et al.'s formulation (2019, p.~5).
+Therefore, we may follow Vehtari et al.'s formulation (2019, p.~5) to define both \(\widehat{R}\) versions.
 Let \(m\) be the total number of chains, \(T\) the number of iterations per chain (where \(T\geq2\)), and \(\theta\) the scalar summary of interest.
@@
 When the sampled values in each chain are independent of the chain's initial value, the mixing component of convergence is satisfied.
-The variance between chains \(B\) is then equivalent to the variance within chains \(W\), and \(\widehat{R}\)-values will be close to one.
+The variance between chains, \(B\), is then equivalent to the variance within chains, \(W\), and \(\widehat{R}\)-values will be close to one.
 High \(\widehat{R}\)-values thus indicate non-convergence.
@@
 Panel B shows two versions of the \(AC\): the default calculation with \texttt{R} function \texttt{stats::acf()} \citep{R}, and manual calculation as the correlation between \(\theta\) in iteration \(t\) and \(\theta\) in iteration \(t+1\).
-Panel C displays the traditional computation of \(\widehat{R}\) conform \citet{gelm92} (original), and in panel D we see \(\widehat{R}\) as computed by implementing Vehtari et al.'s recommendations (adapted).
+Panel C displays the traditional computation of \(\widehat{R}\) conform Gelman and Rubin (original), and in panel D we see \(\widehat{R}\) as computed by implementing Vehtari et al.'s recommendations (adapted).
 
@@
 The function is optimized for performance when \(t\geq50\) \citep{box15}, while the default number of iterations in iterative imputation is often much lower.
-Therefore, we compute \(AC\) manually, see the solid line.
+Therefore, we compute \(AC\) manually (presented with solid lines in panel B).
 These \(AC\)-values behave as expected from visual inspection.
@@
 
-In panel C, we do not see a lot of difference between the two scenarios.
+In panel C, we do not see a lot of difference between the two scenarios either.
 The \(\widehat{R}\)-values are even somewhat lower for the non-convergence scenario than for the typical convergence.
@@
 As diagnostic tool, this may lead to the incorrect conclusion that the non-convergence scenario has superior convergence compared to the typical scenario.
-We can explain this as a downside of the low number of iteration in iterative imputation: the adapted version of \(\widehat{R}\) can only be completely employed if the number of iterations is at least four.
+We can explain this as a downside of the low number of iterations in iterative imputation: the adapted version of \(\widehat{R}\) can only be completely employed if the number of iterations is at least four.
 Otherwise, it is not possible to perform all three transformations to the chains of \(\theta\)-values.
@@
 We will compute \(\widehat{R}\) conform \citet{veht19}, and calculate \(AC\) manually.
-The diagnostics will be applied to the four types of scalar summaries \(\theta\) described above: chain means, chain variances, a quantity of scientific interest, and the first eigenvalue of the variance-covariance matrix.
-We evaluate the performance of the eight sets of diagnostic methods (two diagnostics; four \(\theta\)s) through simulation.
+The diagnostics will be applied to the four types of scalar summaries \(\theta\) described earlier: chain means, chain variances, a quantity of scientific interest, and the first eigenvalue of the variance-covariance matrix.
+We evaluate the performance of these eight sets of diagnostic methods (two diagnostics; four \(\theta\)s) through simulation.
 
@@
 And, whether invalid inferences due to non-convergence may be detected using eight sets of diagnostic methods.
-We will conclude that the algorithm is sufficiently converged for practical purposes when the estimated quantities of scientific interest \(\bar{Q}\) are unbiased and confidence-valid estimates of \({Q}\).
+If the estimated quantities of scientific interest \(\bar{Q}\) are unbiased and confidence-valid estimates of \({Q}\), we will conclude that the algorithm is sufficiently converged for practical purposes.
 In contrast to this, we defined \emph{approximate} convergence as the most converged state that the algorithm can obtain.
@@
 The underlying assumption is that a low proportion of incomplete cases \(p_{\rm inc}\) leads to quick algorithmic convergence, since there is a lot of information in the observed data.
-Higher \(p_{\rm inc}\) yields slower convergence---unless there is so little information in \(y_{\rm obs}\) that this is outweighed by the random component in the imputation algorithm.
+A higher \(p_{\rm inc}\) yields slower convergence---unless there is so little information in \(y_{\rm obs}\) that this is outweighed by the random component in the imputation algorithm.
 Then, convergence to a stable but highly variable state may be reached instantly.
@@
 The assumption inherent to the early stopping conditions is that terminating the imputation algorithm too early causes non-convergence.
-Generally, the algorithm will not reach convergence if \(T=1\), because the imputed values in the first iteration (at \(t=1\)) depend on the initial values of the algorithm \citep[which are sampled randomly from the set of observed datapoints;][]{buur18}.
+Generally, the algorithm will not reach convergence if \(T=1\), because the imputed values in the first iteration (at \(t=1\)) depend on the starting values \citep[which are sampled randomly from the set of observed datapoints;][]{buur18}.
 As the number of iterations increases, the imputation chains should become independent of the initial values, until approximate convergence is reached (i.e., when an additional iteration does not lead to a more converged state).
@@
 The number of simulation repetitions is 1000, and the simulation set-up is summarized in the pseudo-code below.
-The complete \texttt{R} script of the simulation study is available from \href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/}{github.com/gerkovink/shinyMice}.
+The complete \texttt{R} script of the simulation study is available from \href{https://github.com/hanneoberman/MissingThePoint}{github.com/hanneoberman/MissingThePoint}.
 
@@
 That is, the results are more or less stable for conditions where \(T \geq 30\).
-Full results are available from \href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/}{github.com/gerkovink/shinyMice}.
+Full results are available from \href{https://github.com/hanneoberman/MissingThePoint}{github.com/hanneoberman/MissingThePoint}.
 
@@
 Do \emph{not} use the original implementation to calculate \(\widehat{R}\) \citep{gelm92}, or the \texttt{R} function \texttt{stats::acf()} to compute \(AC\) \citep{R}.
-Instead, calculate \(\widehat{R}\) conform \citet{veht19} and compute autocorrelations manually (see e.g., \href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/}{github.com/gerkovink/shinyMice}).
+Instead, calculate \(\widehat{R}\) conform \citet{veht19} and compute autocorrelations manually (see e.g., \href{https://github.com/hanneoberman/MissingThePoint}{github.com/hanneoberman/MissingThePoint}).
 \item

### 2020-05-11 3.Thesis/Writeup/2.Manuscript.tex
Now with correct layout

@@
 
-\hypertarget{summary.-overall-the-methods-to-diagnose-non-convergence-perform-as-expected-they-indicate-more-signs-of-non-convergence-in-conditions-with-worse-performance-in-terms-of-bias-and-confidence-validity.-under-low-to-moderate-missingness-it-does-not-seem-to-matter-on-which-theta-the-non-convergence-diagnostics-are-applied.-if-we-look-at-complete-unbiasedness-however-we-notice-that-univariate-thetas-fail-to-diagnose-the-persistent-bias-in-conditions-where-p_rm-incgeq.75.}{%
-\subsubsection{\texorpdfstring{Summary.
+\hypertarget{summary.-1}{%
+\subsubsection{Summary.}\label{summary.-1}}
+
 Overall, the methods to diagnose non-convergence perform as expected: they indicate more signs of non-convergence in conditions with worse performance in terms of bias and confidence-validity.
 Under low to moderate missingness, it does not seem to matter on which \(\theta\) the non-convergence diagnostics are applied.
-If we look at complete unbiasedness, however, we notice that univariate \(\theta\)s fail to diagnose the persistent bias in conditions where \(p_{\rm inc}\geq.75\).}{Summary.
-Overall, the methods to diagnose non-convergence perform as expected: they indicate more signs of non-convergence in conditions with worse performance in terms of bias and confidence-validity.
-Under low to moderate missingness, it does not seem to matter on which \textbackslash theta the non-convergence diagnostics are applied.
-If we look at complete unbiasedness, however, we notice that univariate \textbackslash thetas fail to diagnose the persistent bias in conditions where p\_\{\textbackslash rm inc\}\textbackslash geq.75.}}\label{summary.-overall-the-methods-to-diagnose-non-convergence-perform-as-expected-they-indicate-more-signs-of-non-convergence-in-conditions-with-worse-performance-in-terms-of-bias-and-confidence-validity.-under-low-to-moderate-missingness-it-does-not-seem-to-matter-on-which-theta-the-non-convergence-diagnostics-are-applied.-if-we-look-at-complete-unbiasedness-however-we-notice-that-univariate-thetas-fail-to-diagnose-the-persistent-bias-in-conditions-where-p_rm-incgeq.75.}}
+If we look at complete unbiasedness, however, we notice that univariate \(\theta\)s fail to diagnose the persistent bias in conditions where \(p_{\rm inc}\geq.75\).
 

### 2020-05-11 Writeup/2.Manuscript.Rmd
Update research archive conform requirements

--- 2020-05-11 3.Thesis/Writeup/2.Manuscript.Rmd
@@
 ```{r setup, include=FALSE}
+# latex options
 knitr::opts_chunk$set(fig.width = 4.9, echo = FALSE, fig.align = "center")
-# knitr::opts_chunk$set(fig.width = 4.8, fig.height = 3.6, echo = FALSE, fig.align = "center")
-
-# packages
-library(mice)
-library(dplyr)
-library(ggplot2)
-library(patchwork)
+
+# load packages
+library(mice) #version 3.8.0
+library(dplyr) #version 0.8.5
+library(ggplot2) #version 3.3.0
+library(patchwork) #version 1.0.0
 
@@
 load("../Results/example_diagnostics.Rdata")
-source("../3.PlotExampleNon-convergence.R")
+source("../Figures/PlotExample.R")
 
@@
 results <- results %>% filter(t<51)
-source("../4.PlotResults.R")
+source("../Figures/PlotResults.R")
 
@@
  legend.margin = margin(0,0,0,0),
- # adjust text size
  text = element_text(size = 8),
@@
 ```
-
-<!-- # Overview of sections -->
-
-<!-- Intro -->
-
-<!-- - This is missingness -->
-
-<!-- - This is how we solve missingness -->
-
-<!-- - For that we need an algorithm, but what if it has not converged? -->
-
-<!-- - That is what we solve in this paper -->
-
-<!-- Algorithmic convergence -->
-
-<!-- - What is algorithmic convergence? -->
-
-<!-- - When is it converged? Stationarity versus mixing. -->
-
-<!-- - How is convergence currently investigated? And how should it be investigated? Multivariate versus univariate. -->
-
-<!-- Methods -->
-
-<!-- - Which methods are there? And how do they relate to mixing and stationarity? -->
-
-<!-- - $\widehat{R}$ -->
-
-<!-- - AC -->
-
-<!-- - We can use these on any $\theta$ -->
-
-<!-- Then Simulation study, Results, Discussion. -->
 
@@
 The premise of multiple imputation is that $\bar{Q}$ is an unbiased and confidence-valid estimate of the true---but unobserved---scientific estimand $Q$ [@rubin96].
-
-<!-- $Q$ is quantity of scientific interest: "A scientific estimand $Q$ is a quantity of scientific interest that we can calculate if we would observe the entire population" [@buur18, par 2.3.1] -->
+<!-- "A scientific estimand $Q$ is a quantity of scientific interest that we can calculate if we would observe the entire population" [@buur18, par 2.3.1] -->
 
 ```{r diagram, out.width="\\linewidth", include=TRUE, fig.cap="Scheme of the main steps in multiple imputation---from an incomplete dataset, to $m=3$ multiply imputed datasets, to $m=3$ estimated quantities of scientific interest $\\hat{Q}$, to a single pooled estimate $\\bar{Q}$."}
-knitr::include_graphics("../Figures/DiagramMI.pdf") #adapted from [@buur18] par. 1.4.1.
+knitr::include_graphics("../Figures/DiagramMI.pdf") #adapted from [@buur18] section 1.4.1.
 ```
@@
 And can we rely on the inference obtained on the completed data?
-<!-- $\dot y_{\rm imp, \ell}$ $y_{\rm mis}$ -->
 
@@
 Several non-convergence diagnostics exist, but it is not known whether these are appropriate within the imputation framework.
-<!-- "a weak diagnostics is better than no diagnostic at all" [@cowl96]. -->
+<!-- "a weak diagnostics is better than no diagnostic at all" [@cowl96].
+-->
 
@@
 
-<!-- The results of this study are guidelines for identifying non-convergence in iterative imputation algorithms, which will aid applied researchers in drawing valid inference from incomplete datasets.
--->
-
-<!-- **[Include sub-questions and make structure of the paper clear!]** How can non-convergence be identified diagnostically?
-Are common MCMC non-convergence diagnostics appropriate for MICE?
-And if so, which threshold should be used to diagnose non-convergence?
-How many iterations are sufficient/needed to be able to diagnose non-convergence?
-Are the default number of iterations sufficient (i.e., 5 in mice, 10 in SPSS and Stata, 30 in mi)?
-How severe is it when the algorithm has not converged?
-And what are guidelines for practice?
-Can the parameter of interest $Q$ be correct when the algorithm is not (yet) converged, and vice versa?
--->
-
-<!-- The defaults of common MI software: SPSS = 10 [(link)](https://www.ibm.com/support/knowledgecenter/SSLVMB_24.0.0/spss/mva/syn_multiple_imputation_impute.html), Mplus = 100??
+<!-- The default number of iterations in common MI software: SPSS = 10 [(link)](https://www.ibm.com/support/knowledgecenter/SSLVMB_24.0.0/spss/mva/syn_multiple_imputation_impute.html), Mplus = 100??
 [(link)](https://pdfs.semanticscholar.org/e20e/29e008592cbfbaa567931f74cdfdb5451405.pdf?_ga=2.55354671.54033656.1584698748-527613517.1584698748), Stata = 10 [(link)](https://www.stata.com/manuals13/mi.pdf, p.
 139), Amelia = NA, because re-sampling, not convergence [(link)](https://cran.r-project.org/web/packages/Amelia/Amelia.pdf), en MI = 30 [(link)](https://cran.r-project.org/web/packages/mi/mi.pdf).
--->
-
-<!-- The goal of this research project is to develop novel methodology and guidelines for evaluating MI methods.
-These tools will subsequently be implemented in an interactive evaluation framework for multiple imputation, which will aid applied researchers in drawing valid inference from incomplete datasets.
 -->
@@
 The collection of $\theta$-values between $t=1$ (summarizing the state-space of the algorithm at initialization) and $t=T$ (summarizing the state-space for the imputed values) will be referred to as an 'imputation chain'.
-<!-- **[If not defined elsewhere, add: $\theta$s are scalar summaries of the algorithm at a certain iteration (e.g., chain means; the average of the imputed values in each imputation chain).] ** -->
 
@@
 
-<!-- When we declare approximate convergence, we are actually concluding that each individual sequence appears stationary and that the observed sequences -->
-<!-- have mixed well with each other. @gelm13, p. 288-->
-
-<!-- **The goal of imputation through e.g.
-MICE is to obtain a multivariate distribution by iterating over a sequence of univariate imputations.
-Traditionally, the inspection of the multivariate convergence is done by visually inspecting the univariate chain means and variances over the iterations.
-The downside to these approaches is that they either focus on the univariate state-space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest.
-Focusing on the convergence of these outcome parameters may influence this procedure in the sense that the model of evaluation favors the model of interest.
-Ideally, one would like to evaluate a model-independent parameter that summarizes the multivariate nature of the data.
-We propose eigenvalue as such a parameter, because it....
-** -->
-
 To illustrate what non-mixing and non-stationarity look like in iterative imputation algorithms, we reproduce an example from van Buuren [-@buur18, $\S$ 6.5.2].
@@
 
-<!-- , since systematically higher imputed values lead to bias, and under-estimation of the variance within imputation chains poses a threat to confidence-validity. -->
-
 ```{r non-conv, out.width='.49\\linewidth', fig.width=3.5, fig.height=3,fig.show='hold',fig.cap="Typical convergence versus pathological non-convergence (reproduced from van Buuren; 2018).
@@
 Non-convergence is diagnosed if the imputation chains are not freely intermingled with one another, or if the chains show definite trends [@buur18, $\S$ 6.5.2].
+<!-- Recommendation @buur18: "the different streams should be freely intermingled with one another, without showing any definite trends". -->
 
@@
 On these grounds, such a $\theta$ can be considered insufficient too.
+<!-- Researchers may "monitor some statistic of scientific interest" [@buur18, $\S$ 6.5.2].  -->
 
@@
 This technique may yield a model-independent, multivariate scalar summary to monitor, but it is not implemented in the  `mice` package [@mice].
-
-<!-- Recommendation: "the different streams should be freely intermingled with one another, without showing any definite trends". -->
-
-<!-- **[MERGE with sections above]** Convergence has historically been inspected visually, by monitoring the imputation algorithm over iterations.
-This is typically done through traceplots.
-In a traceplot, the iteration number is plotted against a certain $\theta$.
-$\theta$s are scalar summaries of the state-space of the algorithm at a specific iteration.
--->
-
-<!-- The default scalar summaries in `mice` are chain means and chain variances.
-Additionally, researchers may "monitor some statistic of scientific interest" [@buur18, $\S$ 6.5.2].
-As @buur18 describes, researchers can specify their quantity of scientific interest $Q$ (e.g., a regression coefficient) as scalar summary $\theta$.
-Such a user-defined scalar summary, however, may be somewhat advanced for empirical researchers.
--->
 
@@
 
-<!-- **Since convergence of iterative imputation algorithms is in distribution, there is not a unique point at which the algorithm reaches a converged state.
-We can only evaluate scalar summaries of the multivariate state-space of the algorithm, $\theta$s.
-The recommended scalar summaries to evaluate are chain means and chain variances.
-Additionally, researchers may "monitor some statistic of scientific interest" [@buur18, $\S$ 6.5.2].
-Because of the issues mentioned above, we propose a novel $\theta$ to monitor: the first eigenvalue of the variance-covariance matrix of the completed data.
-Let $\lambda_{1, \ell} \geq \lambda_2 \geq ...
-\geq \lambda_j$ be the eigenvalues of $\Sigma$ in each of the $m$ completed datasets ${y_{\rm obs}, y_{\rm imp}}$.
-$\lambda_{1, \ell}$ is measure that summarizes the covariances in the completed datasets.**  -->
-
 In this study, we consider each of the four $\theta$s that we discussed.
@@
 # Non-convergence diagnostics
-
-<!-- Requirements for diagnostics to identify non-convergence in iterative imputation are: 1) work for a relatively low number of iterations (defaults in popular iterative imputation software between 5 and 30); 2) may be applied to different $\theta$s, including model-independent $\theta$s (e.g., chain mean, or eigenvalue).
--->
 
@@
 
-<!-- A widely used diagnostic to monitor mixing between chains is the potential scale reduction factor $\widehat{R}$ ['Gelman-Rubin statistic'; @gelm92].
-Therefore, we use $\widehat{R}$ **[WHICH?
-Explain that we'll use both]** and $AC$ to evaluate mixing and stationarity separately, as recommended by e.g., @cowl96.
--->
-
-<!-- **[Add: we don't use effective sample size and MC error because they assume independent samples *within* chains.
+<!-- "We can use effective sample size ^neff to give us a sense of the precision obtained from our simulations.
+As we have discussed in Section 10.5, for many purposes it should suffice to have 100 or even 10 independent simulation draws.
+(If neff = 10, the simulation standard error is increased by p 1 + 1/10 = 1.05).
+As a default rule, we suggest running the simulation until ^neff is at least 5m, that is, until there are the equivalent of at least 10 independent draws per sequence (recall that m is twice the number of sequences, as we have split each sequence into two parts so that Rb can assess stationarity as well as mixing).
+Having an effective sample size of 10 per sequence should typically correspond to stability of all the simulated sequences."  @gelm13, p.
+287 -->
+
+<!-- We don't use effective sample size and MC error because they assume independent samples *within* chains.
 In MI, we only use the final estimate.
 Also, computing n_eff assumes infinitely many iterations, which is fine in Bayesian analyses where T is often at least one thousand.
-But the default in iterative imp is often substantially lower.]** -->
+But the default in iterative imp is often substantially lower.
+-->
 
@@
 \end{align*}
-<!-- %\text{ \cite[p.~5]{veht19}} -->
 
@@
 
-<!-- close to zero indicate  independence between subsequent $\theta$s in imputation chains.  -->
-
-<!-- Negative $AC$-values show no threat to the stationarity component of convergence.
-On the contrary even---negative $AC$-values indicate that $\theta$-values of subsequent iterations diverge from one another, which may increase the variance of $\theta$ and speed up convergence.
-As convergence diagnostic, the interest is therefore in positive $AC$-values.
--->
-
-<!-- Positive $AC$-values occur when $\theta$-values of subsequent iterations are similar.
-When , we speak of recurrence.
-Recurrence within imputation chains indicates trending.
-Ergo, non-stationarity may be diagnosed when $AC$-values are larger than zero.
--->
-
-<!-- Negative $AC$-values indicate divergence within imputation chains.
-**Subsequent sampled values within each imputation chain are less alike.** Positive $AC$-values indicate recurrence.
-If $\theta$-values of subsequent iterations are similar, trending may occur.
--->
-
 ## Thresholds
 
-<!-- We assume that complete convergence, defined as $\widehat{R} = 1$ and $AC = 0$, will not be observed.  -->
 It is unlikely that iterative algorithms such as MICE will achieve the ideal values of $\widehat{R}=1$ and $AC=0$.
@@
 $AC$-values that are significantly higher than zero indicate non-stationarity.
-
-<!-- We assess discrepancy from convergence via practical significance (or some conventional version thereof, such as Rh > 1.1) @gelm13, p. 288 -->
-
-<!-- We can use effective sample size ^neff to give us a sense of the precision obtained from our simulations.
-As we have discussed in Section 10.5, for many purposes it should suffice to have 100 or even 10 independent simulation draws.
-(If neff = 10, the simulation standard error is increased by p 1 + 1/10 = 1.05).
-As a default rule, we suggest running the simulation until ^neff is at least 5m, that is, until there are the equivalent of at least 10 independent draws per sequence (recall that m is twice the number of sequences, as we have split each sequence into two parts so that Rb can assess stationarity as well as mixing).
-Having an effective sample size of 10 per sequence should typically correspond to stability of all the simulated sequences.
-@gelm13, p.
-287 -->
+<!-- Threshold Rhat > 1.1 @gelm13, p. 288 -->
 
@@
 Depicted are the imputation chains in panel A, two versions of the autocorrelation in panel B, and two versions of $\\widehat{R}$ in panels C and D."}
-# load("./images/diagnostics_plot.Rdata") #bquote(theta ~ "= chain mean in "~ y[imp][",m"])
-(theta + plot_layout(guides = "collect", ncol = 2)) + (ac_both + plot_layout(guides = "keep")) + old_rhat + new_rhat +  plot_annotation(tag_levels = "A", tag_suffix = ".") # add 'ncol=1' for plots under each other
+(theta + plot_layout(guides = "collect", ncol = 2)) + (ac_both + plot_layout(guides = "keep")) + old_rhat + new_rhat +  plot_annotation(tag_levels = "A", tag_suffix = ".") 
 ```
@@
 are, therefore, more similar to the original $\widehat{R}$ by Gelman and Rubin.
-<!-- However, the superior performance of the typical convergence over the pathological non-convergence is less prominent, even flipped for $t<4$.  -->
 
@@
 
-<!-- is an unbiased and confidence-valid**[REMOVE?] Subsequently we will evaluate how well $\widehat{R}$ and $AC$ perform in identifying the effects of non-convergence **[and performance of different thetas]**.
-And finally, we will formulate an informed advice on the requirements to safely conclude sufficient convergence in practice.
-That is, we conclude that  convergence is sufficient when each estimate $\bar{Q}$ is an unbiased and confidence-valid estimate of the corresponding estimand $Q$.** -->
-
 To induce non-convergence, we consider two sets of simulation conditions: 'missingness' and 'early stopping'.
@@
 Early stopping is defined by the number of iterations in the imputation algorithm, $T$.
-
-<!-- **["early stopping" en "proportion of complete cases"]** :  -->
-
-<!-- we vary the missingness proportion $p_{\rm inc}$ in the incomplete data, and we vary the number of iterations $T$ in the imputation algorithm.  -->
 
@@
 We expect that we can induce non-convergence by early stopping in at least those conditions where $T$ is smaller than the default number of iterations in `mice`, $T=5$ [@mice].
-
-<!-- **[REMOVE?] $Q$s are the targets that will be estimated in each simulation repetition, after first *amputing* the complete data with varying missingness proportions, then imputing the missingness with a varying number of iterations, then estimating the $Q$s on the $m$ completed datasets, and finally pooling the $m$ estimates $\hat{Q}_\ell$ to obtain a single $\bar{Q}$ for each $Q$.
-We try to predict the effect of non-convergence (i.e., the difference between $Q$ and $\bar{Q}$) with the convergence diagnostics $\widehat{R}$ and $AC$, by applying these methods to several scalar summaries of the state-space of the algorithm, $\theta$s.** -->
 
@@
 5. We further hypothesize that multivariate $\theta$s are better at detecting non-convergence than univariate $\theta$s.
-
-<!-- **(bruggetje hier)** For multiple imputation algorithms, it holds that convergence is reached when there is no dependency between subsequent iterations of imputation chains ($AC = 0$), and chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm ($\widehat{R} = 1$).
-**We expect that a completely converged state will not be reached.**  -->
-
-<!-- This study evaluates whether $\widehat{R}$ and $AC$ could diagnose convergence of multiple imputation algorithms.
-We assess the performance of the two convergence diagnostics against the recommended evaluation criteria for MI methods [i.e., average bias, average confidence interval width, and empirical coverage rate across simulations; @buur18, $\S$ 2.5.2].
-**That is, there is no baseline measure available to evaluate performance against.** -->
-
-<!-- We could also look at distributional characteristics, and plausibility of imputed values, see @vinknd. For now, this is outside of the scope of this study. -->
-
-<!-- Based on an empirical finding [@lace07], we hypothesize that $\widehat{R}$ will over-estimate non-convergence of MI algorithms **explain why?**.
-Lacerda concluded that rhat dropped below 1.2 after at least 50 iterations.
-Therefore we hypothesize that Vehtari's proposed threshold of $\widehat{R} < 1.01$ will be too stringent for diagnosing convergence in iterative imputation.
-This over-estimation may, however, be diminished because $\widehat{R}$ can falsely diagnose convergence if initial values of the algorithm are not appropriately over-dispersed [@broo98, p.
-437].
-In e.g.
-`mice`, initial values are chosen randomly from the observed data.
-Therefore, we cannot be certain that the initial values are over-dispersed.
-We expect this to have little effect on the hypothesized performance of $\widehat{R}$.
-**(Add actual hypothesis: we thought that mice would converge sooner than $\widehat{R}$ would indicate that it did).
-On the one hand, $\widehat{R}$ is known to under-estimate signs of non-convergence when the initial values of the algorithm are not appropriately over-dispersed [@broo98, p.
-437], which may be the case for iterative imputation algorithms.
-In e.g.
-`mice`, initial values are chosen randomly from the observed data.
-Therefore, we cannot be certain that the initial values are sufficiently over-dispersed.
-**  -->
 
@@
 
-<!-- Each of these $Q$s is estimated with a corresponding $\bar{Q}$, after the complete dataset is first *amputed* (with a varying missingness proportion $p_{\rm inc}$) and subsequently imputed (with a varying number of iterations $T$).
--->
-
 ### Simulation conditions.
@@
 We pool the resulting estimates, $\hat{Q}_\ell$, into a single $\bar{Q}$ conform @vink14.
-
-<!-- In each simulation repetition, we *ampute* the complete dataset five times to obtain five different incomplete datasets, $y$s.
-The ratio between $y_{\rm obs}$ and $y_{\rm mis}$ in each $y$ depends on the missingness proportion: $p_{\rm inc} =.05,.25,.5,.75,.95$.
-We ampute the complete data with the `mice` package [function `mice::ampute()`; @mice].
-We consider all possible univariate and multivariate patterns of missingness, conform a 'missing completely at random' missingness mechanism [@rubin87].
-I.e., the probability to be missing is the same for all $N \times k$ cells in $y$, conditional on the missingness proportion $p_{\rm inc}$.
--->
-<!-- <!-- This note only considers a 'missing completely at random' (MCAR) mechanism, where the probability of being missing is equal for all $n \times k$ cells in $y$ [@rubin87].
-The probability to be missing is the same for all $n \times k$ cells in $y$ [a 'missing completely at random' missingness mechanism; @rubin87].
--->
-
-<!-- Missing datapoints in each incomplete dataset $y$ are imputed with the `mice()` function [@mice].
-All imputation procedures are performed with Bayesian linear regression imputation, and five imputation chains ($m=5$).
-Because $m=5$, each run of the `mice()` function results in five sets of imputations: $\dot y_{\rm imp, \ell}$, where $m = 1, 2, ..., 5$.
-The number of iterations in each run of the algorithm varies between simulation conditions ($T=1,2,...,100$).
--->
-<!-- <!-- Each repetition of the simulation thus starts from 1 complete dataset, then splits into 5 missingness conditions ($p_{\rm inc} =.05,.25,.5,.75,.95$), each of the amputed datasets is imputed 5 times ($m = 1, 2,..., 5$) for each iteration condition ($t = 1, 2,..., 100$).
-From these $5\times5\times100$ imputations, we extract the $\theta$s to apply $\widehat{R}$ and $AC$ on (**i.e., chain means, chain variances, $\beta$s per imputation, and the first eigenvalue of the variance-covariance matrix per imputation**).
--->
-
-<!-- From the 5 imputations $\dot y_{\rm imp, \ell}$ we obtain the chain means and chain variances for each variable $j$.
-These will serve as scalar summary $\theta$ to apply non-convergence diagnostics $\widehat{R}$ and $AC$ on.
-We then use `mice::complete()` to combine the imputed data $\dot y_{\rm imp, \ell}$ with the observed data $y_{\rm obs}$, resulting in five completed datasets {$y_{\rm obs}, y_{imp, m}$}.
--->
-
-<!-- On the 5 completed datasets {$y_{\rm obs}, y_{imp, m}$}, we:  -->
-
-<!-- - Estimate $\hat{Q}_\ell$s for $Q= \mu$ and $Q=\sigma$ with `base::mean()`, and `stats::sd()` [@R]. We pool the $\hat{Q}_\ell$s with `base::mean()` to get $\bar{Q}$s. -->
-
-<!-- - Compute $\lambda_{1, \ell}$, to use as $\theta$ in the two convergence diagnostics.
-By definition, the first eigenvalue of the variance-covariance matrix, $\lambda_{1, \ell}$, is equal to the variance of the principal component solution of each {$y_{\rm obs}, y_{imp, m}$}, which we calculate with `stats::princomp()` [@R].
--->
-
-<!-- - Perform multiple linear regression with `stats::lm()` [@R] to get the $\hat{Q}_\ell$s for $Q=\beta$ and $Q=r^2$.
-We pool the $\hat{Q}_\ell$s conform @vink14 to get $\bar{Q}$s for the $\beta$s and $r^2$ **[check  finite pooling of  $r^2$, instead of `mice::pool.r.squared()`]**.
-Additionally, we use the $\hat{Q}_\ell$s of $Q=\beta$ as $\theta$ to apply the convergence diagnostics to.
--->
-
-<!-- <!-- After each run of the `mice` algorithm, the function `mice::complete()` is used to combine the imputed data $\dot y_{\rm imp, \ell}$ with the observed data $y_{\rm obs}$, resulting in five completed datasets {$y_{\rm obs}, y_{imp, m}$}.
--->
-
-<!-- <!-- The estimator for each $Q$ is $\bar{Q}$---the pooled aggregate of the $\hat{Q}_\ell$s across imputations.
-To calculate the $\bar{Q}$s, we combine the observed data $y_{\rm obs}$ and the imputed data for each imputation $\dot y_{\rm imp, \ell}$ with the function `mice::complete()`.
-Descriptive statistics are computed as the average across imputations for each variable ($\mu_j$ and $\sigma_j$, where $j = Y, X_1, X_2, X_3$).
-Estimated regression coefficients are obtained with the function `stats::lm()` for each imputation, and then pooled conform @vink14.
-The coefficient of determination is estimated for each imputation, and pooled using `mice::pool.r.squared()`.
--->
-
-<!-- We apply non-convergence diagnostics $\widehat{R}$ and $AC$ to each scalar summary $\theta$.
-We calculate $\widehat{R}$ by implementing Vehtari et al.'s [-@veht19] recommendations, and $AC$ as the correlation between the $t^{th}$ and the $(t+1)^{th}$ iteration.
--->
 
@@
 
-<!-- We measure the performance of $\widehat{R}$ and $AC$ by the four scalar summaries $\theta$: chain means (i.e., the mean in each $y_{imp, m, j}$) versus $Q=\mu$, chain variances (i.e., the variance in each $y_{imp, m, j}$) versus $Q=\sigma$, the estimated regression coefficients $\beta$ in each {$y_{\rm obs}, y_{imp, m}$} (i.e., a user-defined statistic of scientific interest $\hat{Q}_\ell$) against $Q=\beta$, and the first eigenvalue of the variance-covariance matrix in each {$y_{\rm obs}, y_{imp, m}$}, $\lambda_{1, \ell}$ versus $Q=r^2$.** -->
-
-<!-- **[Remove this?] Primarily, we want to know what the effect is of inducing non-convergence on the estimated $Q$s.
-More specifically, we are interested in the bias of $\bar{Q}$ for all $Q$s and the confidence-validity of $\bar{Q}$ for $Q=\beta$.
-Therefore, we evaluate these performance measures against the simulation conditions (the number of iterations $T$ and the missingness proportion $p_{\rm inc}$).
-We look at averages across simulation repetitions.
-Secondly, we want to know if conditions in which the $Q$s are affected by non-convergence are also identified by the non-convergence diagnostics $\widehat{R}$ and $AC$.
-We do this, again, by looking at the performance measures for the $Q$s, but now we evaluate these against what $\widehat{R}$ and $AC$ indicate about signs of non-convergence in the $\theta$s.** -->
+<!-- Performance measures @buur18, $\S$ 2.5.2 -->
+
+<!-- We could also look at distributional characteristics, and plausibility of imputed values, see @vinknd. For now, this is outside of the scope of this study. -->
 
@@
 
-<!-- Simulation conditions that are impacted by non-convergence portray more extreme bias in the estimated $Q$s, and non-nominal coverages.  -->
-
 ### $Q=\mu$.
@@
 We, therefore, conclude that neither one of the univariate $Q$s is affected substantially by non-convergence.
-<!-- in the first iteration.
-how impact of non-convergence similar to the means.
-The magnitude of the bias $\bar{Q}$ follows the incremental order of the missingness proportions.
-The number of iterations in the algorithm has little effect on the bias.
-Only the estimates in simulation conditions with just one iteration ($T=1$) differ from other conditions that have the same missingness proportion but a higher number of iterations.
-Interestingly, the bias in conditions where $T=1$ is *less* severe than after continued iteration.
-For the least possible bias, we may use any number of iterations $T\geq1$, but only $p_{\rm inc} = 0.05$ results in completely unbiased $\bar{Q}$s.
--->
 
@@
 
-<!-- portion is at most 25%.
-Approximately unbiased estimates are observed for any missingness proportion.
-For $p_{\rm inc}\leq.25$, the minimum number of iterations is two ($T\geq2$),  for $p_{\rm inc}=.50$ it is $T\geq4$, and $p_{\rm inc}=.75$ requires $T\geq5$.
-The highest number of iterations needed to obtain valid inferences is seven ($p_{\rm inc}=.95$; $T\geq7$).
-while the best performance occurs with less missingness and more iterations.
-Only conditions with the lowest missingness proportion ($p_{\rm inc}=0.05$) results in unbiased estimates for every number of iterations.
-In other conditions there is at least some bias irrespective of $T$.
-The inferences do not improve in any condition where $T \geq 6$.
-This suggests that the algorithm is as converged as necessary to obtain valid inferences.
--->
-
 ### $Q=\beta$.
@@
 We conclude that any condition where $T\geq3$ yields approximately nominal coverage rates, and thus confidence-valid estimates.
-<!-- , whereas we need $T\geq7$ to reach approximate unbiasedness. -->
 The nominal coverages can be explained by the CIWs.
@@
 Since conditions with higher $p_{\rm inc}$ result in both wider CIs and more severe bias, the true value of $Q$ may be included in the CI, despite of the bias in $\bar{Q}$.
-<!-- around $\bar{Q}$ in conditions with higher missingness proportions **results in the inclusion of** the true value of $Q$ despite of bias. -->
 We conclude that approximately unbiased estimates may be obtained in conditions where $T\geq7$, whereas confidence-valid estimates require at most three iterations.
-
-<!-- confidence interval width (CIW), and coverage rate (CR) as performance measures.
-**[Split panels]** Unbiased estimates are obtained in conditions where $p_{\rm inc}\leq.50$ and $T\geq3$.
-Approximately confidence-valid estimates are obtained in any condition where $T\geq3$---despite of persistent bias in conditions where $p_{\rm inc}\geq.75$.
-These approximately nominal coverages are explained by the CIWs: as expected, conditions with higher missingness proportions result in more uncertainty in the estimates (i.e., wider CIs), which diminish the effect of the bias.
-confidence-valid inferences may be obtained when $T\geq3$, while we need at least seven iterations to reach valid inferences in terms of bias.
-Continued iteration after $T\geq7$ seems futile.
-**[add max T for other betas?]** -->
-<!-- One exception is the CIW of the most extreme condition ($p_{\rm inc}=.95$; $T=1$). -->
 
@@
 This implies that the algorithm produces stable, non-improving estimates when $T\geq7$.
-
-<!-- Overall, we see that conditions with a higher proportion of missingness are affected more severely by non-convergence.
-Multivariate $Q$s are also substantially influenced by the number of iterations.
-This means that the algorithm has indeed not converged at $t=1$, and converges gradually as $t$ moves to $T$.
-The point at which an additional iteration does not lead to an improvement in the estimates depends on quantity of scientific interest $Q$ and the missingness proportion $p_{\rm inc}$.
-The highest number of iterations that is necessary to reach this point across all simulation conditions is $T=7$.
-This implies that, under the current specifications, no more than seven iterations are necessary to obtain sufficient performance in terms of bias and confidence-validity.
--->
-
-<!-- Conditions with a lower number of iterations yield more bias in $\bar{Q}$.
-This is especially pronounced in conditions with high missingness proportions: with 95% missingness we need at least seven iterations before an additional iteration does not lead to improved estimates.
-Conditions with a lower missingness proportion (e.g., $p_{\rm inc} = {0.5}$), produce *optimal* estimates with as many as two iterations.
-Approximately unbiased estimates are obtained for all conditions where $p_{\rm inc}\leq0.75$ and $T\leq6$.
-**[Add CIW and CR]** -->
-
-<!-- The sign of the bias is logical: when there is less info in the data, the relations between the predictors and the outcome is under-estimated. -->
 
@@
 
-<!-- are appropriate,  **[appropriate methods thus identify these conditions]** -->
-
-<!-- We evaluate the non-convergence diagnostics $\widehat{R}$ and $AC$ against the performance measures described above (i.e., bias and confidence-validity of the quantities of scientific interest $Q$).
-The diagnostics should identify those simulation conditions that are negatively impacted by non-convergence.
-We expect the most signs of non-convergence in conditions where valid estimates are not obtained (high missingness proportions, $T<7$), and at least some signs of non-convergence in conditions where estimates are not approximately unbiased ($p_{\rm inc}=.95$, any number of iterations).
--->
-
-<!-- As we have observed, the affected conditions are $p_{\rm inc}=.95$ for all all $Q$s are impacted by the missingness proportions, but only the multivariate quantities of scientific interest are affected by the number of iterations.
--->
-
-<!-- Figure \ref{fig:rhats-acs} displays the non-convergence diagnostics $\widehat{R}$ and $AC$ for each of the four $\theta$s under consideration.
-**[Add panel labels and explain how to read the plots]**  -->
-
 ```{r rhats-acs, fig.height = 6.8, fig.cap="Non-convergence diagnostics $\\widehat{R}$ and $AC$ applied on several $\\theta$s.
@@
 ```
-
-<!-- **Message: after one iteration ($T$=2), the algorithm is as well converged as the missingness proportion allows.
-Sooner even with lower missingness proportions---$T$=1 might be sufficient for $p_{\rm inc}$ = 5, 25, and 50.** Number of iterations has little influence on bias in $\hat{Q}_\ell$, only in first iteration there is some bias in missingness conditions of 25 and up.
-There is a clear different between the missingness conditions.
-The higher the missingness prop, the larger the magnitude of the bias.
-For this $Q$, t=2 would be sufficient.
-And only for $p_{\rm inc}$ = 5 the $\bar{Q}$s are unbiased.
-All other $p_{\rm inc}$ conditions have bias irrespective of the number of iterations.
--->
 
@@
 
-<!-- We evaluate the performance of the non-convergence diagnostics applied to chain means against the bias in $Q=\mu$.
-Based on the bias in $Q$, we concluded that one iteration is sufficient to obtain valid inferences, irrespective of the missingness proportion.
-The non-convergence diagnostics, however, identify some signs of non-convergence in conditions with as many as thirty iterations.
-This does not correspond to the impact of non-convergence in the univariate $Q$s.
-**Compared to the multivariate $Q$s [add before that we evaluate the thetas against all Qs?], $\widehat{R}$ and $AC$ applied to the chain means fail to identify the non-convergence in conditions where $p_{\rm inc}=.95$.**  -->
-
-```{r mean, fig.height = 6.8, fig.cap="Convergence diagnostics chain mean."}
-# mean_bias + mean_Rh + mean_AC + plot_layout(guides = "collect", ncol = 1) + plot_annotation(tag_levels = "A", tag_suffix = ".", title = bquote("Q = " ~ mu[Y] ~ "= 25.81; " ~ theta ~ "= chain mean of Y in" ~ y[imp][",m"])) #& theme(legend.position = "bottom")
-```
-
 ### $\theta=$ chain variance.
@@
 
-<!-- We thus conclude that there is no added value of evaluating non-convergence with chain variances as $\theta$ in the diagnostics. -->
-
-```{r sd, fig.height = 6.8, fig.cap="Convergence diagnostics chain variance."}
-# sd_bias + var_Rh + var_AC + plot_layout(guides = "collect", ncol = 1) + plot_annotation(tag_levels = "A", tag_suffix = ".", title = bquote("Q = " ~ sigma[Y] ~ "= 11.32; " ~ theta ~ "= chain variance of Y in" ~ y[imp][",m"]))
-```
-
 ### $\theta=\hat{Q}$.
@@
 Since the $AC$-values are not improving or worsening when $T\geq7$, we conclude that non-convergence is incorrectly diagnosed based on the threshold.
-
-<!-- The non-convergence diagnostics are applied to the estimated regression coefficients in each imputation, and evaluated against the bias in the corresponding pooled estimate.  -->
-
-<!-- $\widehat{R}$ indicates signs of non-convergence across all missingness proportions and each number of iterations, in the expected order: simulation conditions with higher missingness proportions and less iterations are identified as the worst converged.
-The thresholds to diagnose non-convergence are overcome in different conditions, depending on the number of iterations and the missingness proportion.
-The most lenient threshold ($\widehat{R}>1.2$) is overcome at $T\geq9$ for $p_{\rm inc} = 0.05$, and $T\geq12$ for $p_{\rm inc} = 0.95$.
-The conventional $\widehat{R}>1.1$ is surpassed at $T\geq12$ and $T\geq24$, respectively.
--->
-
-<!-- For the autocorrelations applied to the $\hat{Q}_\ell$s, we see an even stronger effect of the missingness proportions.
-Especially the conditions where $p_{\rm inc} = 0.75$ and $p_{\rm inc} = 0.95$ are identified as non-stationary.
-The threshold used to evaluate whether $AC=0$, however, is only reached in conditions with at least ten iterations.
-This suggests that the default number of iterations may be too low to detect non-stationarity with $AC$.
-However, the critical value for diagnosing non-convergence with $AC$ decreases as a function of the number of iterations.
-The conclusion to identify non-stationarity in conditions where $T\geq10$ and $p_{\rm inc} = 0.95$ might therefore be spurious.
--->
-
-<!-- **[Also, the $AC$ across missingness proportions shows the opposite of what you would expect: shouldn't there be higher autocorrelations in conditions with comparatively more observed data in the completed datasets?
-I don't understand this.]** -->
-
-<!-- Compared to the bias in $\bar{Q}$, the diagnostics applied to the $\hat{Q}_\ell$s over-estimate the severity of the non-convergence.  -->
 
@@
 
-<!-- And especially in conditions with higher missingness proportions, we observe signs of increasing stationarity the decrease in the $AC$-values across iterations is also  -->
-
-<!-- Finally, we evaluate the non-convergence diagnostics applied to the first eigenvalue of the variance-covariance matrices of the completed data.
-Similar to the results the the regression coefficients as $\theta$s, we see effects of both the missingness proportion and number of iterations in each simulation condition.
-The most signs of non-convergence are identified in conditions where $p_{\rm inc}$ is large and $T$ is small.
--->
-
-<!-- Notwithstanding the similarities with $\theta=\hat{Q}_\ell$, there is a big difference in $\widehat{R}$-values for conditions where $T<4$.
-In these conditions, $\widehat{R}$ behaves oppositely of our expectation: conditions with the lowest missingness proportions yield the highest $\widehat{R}$-values.
-In conditions with a higher number of iterations, this artifact disappears.
-This suggests that the combination of $\widehat{R}$ and $\lambda_{1, \ell}$ is only appropriate for imputation algorithms where the number of iterations at least four.
-Interpreting the $\widehat{R}$-values of conditions where $T<3$ may lead to the incorrect conclusion that only conditions with *low* missingness proportions show signs of non-convergence.
-For applications as a model-independent multivariate $\theta$, we should therefore only interpret the $\widehat{R}$-values when $T\geq4$.
-Using this cut-off we conveniently ignore the initial 'dip' at $T=3$ as well.
-The traditional threshold $\widehat{R} > 1.2$ is overcome in conditions where $p_{\rm inc} = .95$ and $T\geq10$, and for all other missingness proportions in conditions where $T\geq8$.
-With the $\widehat{R} > 1.1$ threshold, the number of iterations are $T\geq20$ and $T\geq16$, respectively.
--->
-
-<!-- The $AC$-values largely depend on the missingness proportion in each condition.
-Conditions where $p_{\rm inc}=.95$ show statistically significant autocorrelations for any number of iterations.
-The other extreme, conditions where $p_{\rm inc}=.05$, does not yield $AC$-values above the threshold.
-The $AC$-values with this missingness proportion are equivalent for all conditions where $T\geq6$, which suggests that the algorithm is as stationary as it can be.
-A similar point is only observed after twenty iterations in conditions with higher missingness proportions.
--->
-
-<!-- In comparison to the $Q$s, the diagnostics again over-estimate the signs of non-convergence.
-Conditions with low missingness proportions (e.g., 5% or 25% of the complete data) yield unbiased, confidence-valid estimates with as little as two iterations.
-The convergence diagnostics, however, indicate improved convergence up-to eight iterations according to $AC$ and thirty to forty iterations according to $\widehat{R}$.
-The model-independent $\theta$ $\lambda_{1, \ell}$ seems as well-equipped to diagnose signs of non-convergence as $\theta=\hat{Q}_\ell$.
-The univariate $\theta$s do not differentiate between bias induced by the missingness proportions.
--->
-
 ### Summary. 
@@
 A better heuristic to diagnose non-stationarity with $AC$ may be through evaluation of the $AC$-values across iterations: If the values do not substantially decrease with $T$, approximate stationarity may be concluded.
-<!-- **[similar to the elbow heuristic in PCA]**. -->
-
-<!-- In general, the diagnostics identify more signs of non-convergence in simulation conditions with a lower number of iterations.
-The missingness proportions, however, only seem to impact multivariate convergence.
-When $\widehat{R}$ and $AC$ are applied to  univariate $\theta$s, they identify the same amount of non-convergence for any missingness proportion.
-When the diagnostics are applied to multivariate $\theta$s, they generally identify more signs of non-convergence for higher missingness proportions.
--->
-
-<!-- As expected, none of the simulation conditions that we consider resulted in complete convergence of the algorithm (defined as $\widehat{R}=1$ and $AC=0$).
-However, the thresholds used in practice to diagnose non-convergence seem too conservative for imputation purposes.
-$\widehat{R}$ and $AC$ identify signs of non-convergence in conditions where the $\bar{Q}$s are unbiased and confidence-valid estimates of the $Q$s.
-The most recent recommended threshold for diagnosing non-convergence with $\widehat{R}$ is never overcome in this study (all $\widehat{R}$>1.01).
-Apparently, this threshold is too stringent for iterative imputation algorithms, and will not be discussed any further.
--->
-
-<!-- This is odd, since the $\theta$ and $Q$ are directly related.   -->
-
-```{r est, fig.height = 6.8, fig.cap="Convergence diagnostics regression coefficient."}
-# est_bias + est_Rh + est_AC + plot_layout(guides = "collect", ncol = 1) + plot_annotation(tag_levels = "A", tag_suffix = ".", title = bquote("Q = " ~ beta[1] ~ "= 2.06; " ~ theta ~ "= " ~ hat(Q) ~ "in {" ~ y[obs] ~ "," ~ y[imp][",m"] ~ "}"))
-```
-
-```{r pred, fig.height = 6.8, fig.cap="Convergence diagnostics regression coefficient."}
-# Rsq_bias + PCA_Rh + PCA_AC + plot_layout(guides = "collect", ncol = 1) + plot_annotation(tag_levels = "A", tag_suffix = ".", title = bquote("Q = " ~ r^2 ~ "\u00D7 100 = 19.25; " ~ theta ~ "= " ~ lambda[1] ~ "in {" ~ y[obs] ~ "," ~ y[imp][",m"] ~ "}")) #first eigenvalue of" ~ Sigma ~
-```
-
-<!-- ## Univariate estimates and convergence diagnostics  -->
-
-<!-- The bias in the estimates of the variable means shows little to no difference between simulation conditions.
-It doesn't seem to matter how many iterations you use in the mice algorithm, the estimates are unbiased.
-Similarly, the bias in the estimated variances is more or less stable across simulation conditions.
-These univariate quantities appear to be unaffected by the number of iterations.
--->
-
-<!-- When applied to the imputation chain means, $\widehat{R}$ indicates that the mice algorithm does not reach a converged state ($\widehat{R}$ = 1) in any of the simulation conditions.
-Neither is the most recent recommended threshold reached ($\widehat{R}$ < 1.01).
-**This is only for 75% missingness:** The conventional $\widehat{R}$ threshold of 1.2 is reached in simulation conditions $T = 3$ and $T > 6$.
-(With the default number of iterations (maxit = 5), this dip in $\widehat{R}$-values would be spotted, so it is no problem.) The point at which an extra iteration does not seem to improve the $\widehat{R}$ value is around $T=30$.
--->
-
-<!-- Autocorrelations indicate no sign of trending **(NOT correct anymore, fluke of the stats::acf() function!!)** within imputation chains.
-In most simulation conditions $AC$ is smaller than or about equal to zero.
-Across simulation conditions, however, the autocorrelation curve does not trend towards 0.
-Autocorrelation values plateau off at a value of around.1.
-This is a small positive autocorrelation, which would indicate some trending within chains.
--->
-
-<!-- The $\widehat{R}$ and $AC$-values for the imputation chain variances show equal trends to the chain means and are therefore not discussed separately.
-Taken together, univariate estimates seem robust across simulation conditions.
-There is no clear effect of the number of iterations on the bias in these estimates, while the convergence diagnostics indicate that the algorithm did not reach a completely converged state (yet).
--->
-
-<!-- ## Multivariate estimates  -->
-
-<!-- **This is only true for 75% missingness:** There is a clear bias in the regression estimates in simulation conditions where the number of iterations is smaller than four.
-In simulation conditions where $T > 5$ there is little to no bias in the estimated regression coefficients.
-In most simulation conditions nominal coverage is obtained (i.e., coverage rates of.95) for the confidence intervals around the regression coefficients.
-Conditions with only one or two iterations show some under-coverage.
-Since confidence interval width is stable across conditions, the under-coverage may be attributed to the bias in the estimated regression coefficients.
--->
-
-<!-- **This is only true for 75% missingness:** If we look at the estimated proportion of explained variance in outcome variable Y we see that the coefficient of determination is underestimate estimated in conditions where the number of iterations is equal to two or less, and slightly overestimated in conditions where the number of iterations is equal to three or more.
--->
-
-```{r thresh-Rh, include=FALSE}
-# # chain means
-# dat[-1, 6:9] %>% mutate(max_Rh = apply(., 1, max)) %>%.$max_Rh<1.2
-# 
-# # chain variances
-# dat[-1, 26:29] %>% mutate(max_Rh = apply(., 1, max)) %>%.$max_Rh<1.2
-
-# qnorm((1 + .95) / 2) / sqrt(t)
-
-```
+<!-- [similar to the elbow heuristic in PCA]. -->
 
@@
 
-<!-- **[ADD limitation that other MI packages may have different way to obtain initial values!]** -->
-
-<!-- # Old discussion -->
-
-<!-- We arrive back at the original questions that we posed. -->
-
-<!-- - Non-convergence can be identified with non-convergence diagnostics $\widehat{R}$ and $AC$.
-But common thresholds are too conservative.
-Why?
-Part of the distribution is already determined by the observed data.
-in regular MCMC the entire distribution is unknown (**priors??**).
-In iterative imputation algorithms we have the same characteristics as MCMC in the limit, but because we combine the known distribution with the unknown distribution, convergence is reached much sooner.
--->
-
-<!-- - Inferences are unbiased much quicker than the algorithm has converged.
-All things univariate are converged almost instantly after initiating the iterative algorithm with a valid imputation model.
-But less quick with multivariate parameters.
--->
-
-<!-- In short, iterative imputation algorithms may yield approximately unbiased, confidence-valid estimates under different levels of induced non-convergence.
-Univariate estimates seem robust against terminating the algorithm early: There is no clear effect of the number of iterations on the bias in estimated means and standard deviations.
-The bias in these $Q$s was only impacted by the missingness proportion $p_{\rm inc}$, with higher $p_{\rm inc}$ resulting in more bias.
-Bias in multivariate estimates depended on both $p_{\rm inc}$ and the number of iterations.
-However, the only effect of the number of iterations was observed when the algorithm was terminated at seven iterations or earlier.
-This suggests that continued iteration beyond seven iterations does not yield better estimates.
-Yet, the non-convergence diagnostics $\widehat{R}$ and $AC$ indicated that the iterative imputation algorithm did not reach a stable state before twenty to thirty iterations.
-**[MERGE or remove] $\widehat{R}$ and autocorrelation indicate that there are signs of non-convergence in the algorithm up-to at least twenty iterations, while unbiased, confidence-valid estimates may be obtained with as little as one iteration.
-This observation is in agreement with **one of [still add the others!]** the simulation hypotheses: $\widehat{R}$ over-estimates the severity of non-convergence when applied to MI procedures.**  -->
-
-<!-- Over-estimation of the signs of non-convergence may be due to the methods (and their thresholds) or due to the $Q$s (descriptive and multivariate linear regression, not higher dimensional/more complex $Q$s).
-More complicated $Q$s (e.g., higher-order effects or variance components) might show bias, under- or over-coverage at higher $T$.
-Which might be why $\widehat{R}$ and $AC$ identify non-convergence.
--->
-
-<!-- ```{r eval=FALSE, include=FALSE} -->
-<!-- # **[INCLUDE this] The highest $\widehat{R}$-values are obtained for the condition $T=2$.
-We observe the same 'dip' in $\widehat{R}$ as we did in Figure \ref{fig:diagnostics} for conditions with three iterations.
-$\widehat{R}$ gradually decreases as the number of iterations increases across conditions.
-By the traditional threshold  $\widehat{R}>1.2$ we should conclude that there is non-convergence in conditions where $T<8$.
-The common threshold $\widehat{R}>1.1$ is overcome in conditions where $T>14$.
-** -->
-<!-- #    -->
-<!-- #   **Autocorrelations indicate some signs of non-stationarity in all simulation conditions, but primarily in conditions where $T<6$.
-The threshold for diagnosing non-convergence is not reached in any condition.
-We would thus conclude that the algorithm is sufficiently converged regardless of the number of iterations or missingness proportion.
-** -->
-<!-- ``` -->
-
-<!-- Some other interesting findings are **[REPHRASE]** -->
-
-<!-- - Is there more bias in conditions where non-convergence was induced? (YES for multivar., NOPE for univariates). -->
-
-<!-- - Is the point of *sufficient* convergence (at which the $\bar{Q}$s are unbiased and confidence-valid estimates of the $Q$s) the same point as identified by the non-convergence diagnostics?
-(NOPE, they are too conservative).
--->
-
-<!-- - Is one of the thetas better? (YES, while it doesn't really matter as long as it's multivariate, $\lambda_{1, \ell}$ is model-independent and therefore preferred). -->
-
-<!-- - What should the thresholds be? (much HIGHER than recommended for $\widehat{R}$, difficult to say for $AC$).  -->
-
-<!-- - And the default nr of iterations? (MAYBE increase to 10 to detect significant $AC$ and be sure not to miss the dip in $\widehat{R}$?). -->
-
-<!-- As expected, complete convergence ($\widehat{R}=1$ and $AC=0$) is never observed.
-But what about the thresholds for practice?
-$\widehat{R}$ <1.2 is not stringent enough, because conditions in which $\widehat{R}$ is smaller than 1.2 do not all yield unbiased estimates of the $Q$s.
-Moreover, there is a dip in the $\widehat{R}$-values, after which $\widehat{R}$ > 1.2 again.
-If the algorithms would be terminated at the iteration where $\widehat{R}$ < 1.2 occurs for the first time, the increase in $\widehat{R}$-values might be missed.
-The threshold $\widehat{R}$ < 1.1 is somewhat conservative in comparison with the performance measures.
-Overall it seems OK.
-$\widehat{R}$ < 1.01 is too stringent compared to the performance measures.
-And it is not obtained for any number of iterations (**but may be necessary in more complex complete data models???**).
--->
-
-<!-- According to this simulation study, the recently proposed threshold of $\widehat{R}<1.01$ may be too stringent for MI algorithms.
-Under the relatively easy missing data problem of the current study, the threshold was not reached.
-The other extreme of the $\widehat{R}$-thresholds, the conventionally acceptable $\widehat{R} <1.2$, may be too lenient for MI procedures.
-Applying this threshold to the current data, lead to falsely diagnosing convergence at $T = 3$ (**because it goes up after, not because it is not converged enough**).
-It appears that the widely used threshold of $\widehat{R} < 1.1$ suits MI algorithms the best.
-We might, however, also formulate a new threshold, specifically for the evaluation of MI algorithms.
-The current study suggests that $\widehat{R} < 1.05$ may be implemented, since that is the level at which the $\widehat{R}$ stabilize (around $T = 20$) (**Not necessary for this Q, but maybe for more complicated $Q$s**).
--->
-
-<!-- The negative $AC$-values obtained in this study show no threat of non-stationarity.
-However, the initial dip in $AC$-values **(which disappeared!)** may have implications for the default number of iterations in `mice` (`maxit = 5`).
-Terminating the algorithm at $T=5$ may not be the most appropriate, since this lead to the worst convergence (**nope, only for Rh, not $AC$**), as indicated by $\widehat{R}$ and $AC$.
-Under the current specifications, $T>20$ would be more appropriate.
--->
-
-<!-- ## Recommendations for future research -->
-
-<!-- Much remains unknown.
-about non-convergence in iterative imputation algorithms.
-Further research is needed to investigate the performance of iterative imputation algorithms under clear violation of convergence, e.g.
-dependency between predictors (predictors with very high correlations).
-For example: -->
-
-<!-- - Induce non-convergence with mis-specified imputation model.  -->
-
-<!-- - Introduce more difficult missingness problems, i.e., M(N)AR instead of MCAR.
-That is, proper performance under a 'missing completely at random' missingness mechanism is necessary to demonstrate the appropriateness of $\widehat{R}$ and $AC$ as non-convergence diagnostics.
-However, results may not be extrapolated to other missingness mechanisms.
--->
-
-<!-- - Imputation technique with questionable convergence properties, e.g., PMM. -->
-
-<!-- - Apply recommendations to empirical data and write a vignette for applied researchers. -->
-
-<!-- - Implement recommendations in imputation software, e.g., evaluation environment for applied researchers. -->
-
-<!-- - Develop an additional convergence diagnostic unique for iterative imputation: $AC$ across imputations, instead of iterations. -->
-
 \noindent In short, we have shown that iterative imputation algorithms can yield correct outcomes, even when a converged state has not yet formally been reached.
@@
 Of course, it never hurts to iterate longer, but such calculations hardly bring added value.
-
-<!-- # Grading -->
-
-<!-- - Scientific contribution  -->
-<!-- (clear research question, convincing argumentation that the study adds to the scientific literature) -->
-
-<!-- - Theory -->
-<!-- (use of theory, clear hypotheses, relevant literature) -->
-
-<!-- - Method -->
-<!-- (correct data collection/handling, measures, choice and description of method of analysis) -->
-
-<!-- - Results -->
-<!-- (correct interpretation of findings, clear presentation) -->
-
-<!-- - Discussion and conclusions -->
-<!-- (summary, broader implications, limitations, future research) -->
-
-<!-- - Written presentation -->
-<!-- (writing, structure, tables and figures, references) -->
-
-<!-- # References -->

### 2020-05-11 3.GenerateSupplementaryFigures.R

@@ -1,0 +1,38 @@
+# generate supplementary figures that display complete results
+# (we only display conditions where T<51 in the manuscript)
+# requires the results file 'complete.Rdata'
+
+# load packages
+library(dplyr) #version 0.8.5
+library(ggplot2) #version 3.3.0
+library(patchwork) #version 1.0.0
+
+# set default graphing behavior
+theme_update(
+  plot.title = element_text(hjust = 0.5),
+  plot.subtitle = element_text(hjust = 0.5),
+  panel.border = element_blank(),
+  panel.grid.major = element_blank(),
+  panel.grid.minor = element_blank(),
+  panel.background = element_blank(),
+  axis.line = element_line(colour = "black"),
+  legend.key = element_blank(),
+  legend.position = "bottom",
+  legend.margin = margin(0,0,0,0),
+  text = element_text(size = 7),
+  title = element_text(size = 8),
+  legend.title = element_text(size = 7)
+)
+
+# load results and figures
+load("Results/complete.Rdata") 
+source("Figures/PlotResults.R")
+
+# plot the performance measures
+estimates <- mean_bias + sd_bias + Rsq_bias + est_bias + est_cov + est_ciw + plot_layout(ncol = 2, guides = "collect") + plot_annotation(tag_levels = "A", tag_suffix = ".", title = "Impact of non-convergence on statistical inferences (see thesis manuscript for definitions)")
+ggsave("Figures/ResultsEstimates.png", estimates, width = 8.3, height =  5.8)
+
+# plot the non-convergence diagnostics
+diagnostics <- mean_Rh + mean_AC + var_Rh + var_AC + est_Rh + est_AC + PCA_Rh + PCA_AC + plot_layout(guides = "collect", ncol = 2) + plot_annotation(tag_levels = "A", tag_suffix = ".", title = "Non-convergence identified by diagnostics (see thesis manuscript for definitions)")
+ggsave("Figures/ResultsDiagnostics.png", diagnostics, width = 8.3, height =  5.8)
+

### 2020-05-11 Functions/CreateData.R

@@ -1,0 +1,35 @@
+# function to create data and ampute it
+# requires the packages 'mice' and 'mvtnorm'
+
+data.simulation <- function(n = populationsize, true_effect = 2) {
+  # simulate multivariate normal predictors
+  means <- c(12, 3, 0.5) #predictor means
+  vars  <- c(4, 16, 9) #predictor variances
+  R <- matrix(numeric(3 * 3), nrow = 3) #correlation matrix
+  diag(R) <- 1 #set diagonal to 1
+  R[upper.tri(R)] <-
+    R[lower.tri(R)] <- c(.5, .3, .4) #set bivariate correlations
+  sigma <-
+    diag(sqrt(vars)) %*% R %*% diag(sqrt(vars)) #variance-covariance matrix
+  simdata <-
+    as.data.frame(rmvnorm(n = populationsize, mean = means, sigma = sigma)) #create data
+  colnames(simdata) <- c("X1", "X2", "X3") #set predictors names
+  
+  # compute  outcome variable from predictors
+  simdata$Y <-
+    1 + true_effect * simdata$X1 + 0.5 * simdata$X2 - 1 * simdata$X3 + rnorm(n = populationsize, sd = 10)
+  
+  # estimate comlpete data parameters Q
+  true.effect <<-
+    lm(Y ~ X1 + X2 + X3, data = simdata)[["coefficients"]]
+  true.R.sq <<-
+    lm(Y ~ X1 + X2 + X3, data = simdata) %>% summary() %>% .$r.squared
+  true.mean <<- apply(simdata, 2, mean)
+  true.sd <<- apply(simdata, 2, sd)
+  true.sigma <<-
+    lm(Y ~ X1 + X2 + X3, data = simdata)[["residuals"]] %>% var()
+  
+  # output
+  return(simdata)
+  
+}

### 2020-06-02 ICML/example_paper.tex
Add ICML folder and style

%%%%%%%% ICML 2020 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%

\documentclass{article}

% Recommended, but optional, packages for figures and better typesetting:
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{booktabs} % for professional tables

% hyperref makes hyperlinks in the resulting PDF.
% If your build breaks (sometimes temporarily if a hyperlink spans a page)
% please comment out the following usepackage line and replace
% \usepackage{icml2020} with \usepackage[nohyperref]{icml2020} above.
\usepackage{hyperref}

% Attempt to make hyperref and algorithmic work together better:
\newcommand{\theHalgorithm}{\arabic{algorithm}}

% Use the following line for the initial blind version submitted for review:
\usepackage{icml2020}

% If accepted, instead use the following line for the camera-ready submission:
%\usepackage[accepted]{icml2020}

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{Submission and Formatting Instructions for ICML 2020}

\begin{document}

\twocolumn[
\icmltitle{Submission and Formatting Instructions for \\
           International Conference on Machine Learning (ICML 2020)}

% It is OKAY to include author information, even for blind
% submissions: the style file will automatically remove it for you
% unless you've provided the [accepted] option to the icml2020
% package.

% List of affiliations: The first argument should be a (short)
% identifier you will use later to specify author affiliations
% Academic affiliations should list Department, University, City, Region, Country
% Industry affiliations should list Company, City, Region, Country

% You can specify symbols, otherwise they are numbered in order.
% Ideally, you should not use this facility. Affiliations will be numbered
% in order of appearance and this is the preferred way.
\icmlsetsymbol{equal}{*}

\begin{icmlauthorlist}
\icmlauthor{Aeiau Zzzz}{equal,to}
\icmlauthor{Bauiu C.~Yyyy}{equal,to,goo}
\icmlauthor{Cieua Vvvvv}{goo}
\icmlauthor{Iaesut Saoeu}{ed}
\icmlauthor{Fiuea Rrrr}{to}
\icmlauthor{Tateu H.~Yasehe}{ed,to,goo}
\icmlauthor{Aaoeu Iasoh}{goo}
\icmlauthor{Buiui Eueu}{ed}
\icmlauthor{Aeuia Zzzz}{ed}
\icmlauthor{Bieea C.~Yyyy}{to,goo}
\icmlauthor{Teoau Xxxx}{ed}
\icmlauthor{Eee Pppp}{ed}
\end{icmlauthorlist}

\icmlaffiliation{to}{Department of Computation, University of Torontoland, Torontoland, Canada}
\icmlaffiliation{goo}{Googol ShallowMind, New London, Michigan, USA}
\icmlaffiliation{ed}{School of Computation, University of Edenborrow, Edenborrow, United Kingdom}

\icmlcorrespondingauthor{Cieua Vvvvv}{c.vvvvv@googol.com}
\icmlcorrespondingauthor{Eee Pppp}{ep@eden.co.uk}

% You may provide any keywords that you
% find helpful for describing your paper; these are used to populate
% the "keywords" metadata in the PDF but will not be shown in the document
\icmlkeywords{Machine Learning, ICML}

\vskip 0.3in
]

% this must go after the closing bracket ] following \twocolumn[ ...

% This command actually creates the footnote in the first column
% listing the affiliations and the copyright notice.
% The command takes one argument, which is text to display at the start of the footnote.
% The \icmlEqualContribution command is standard text for equal contribution.
% Remove it (just {}) if you do not need this facility.

%\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.

\begin{abstract}
This document provides a basic paper template and submission guidelines.
Abstracts must be a single paragraph, ideally between 4--6 sentences long.
Gross violations will trigger corrections at the camera-ready phase.
\end{abstract}

\section{Electronic Submission}
\label{submission}

Submission to ICML 2020 will be entirely electronic, via a web site
(not email). Information about the submission process and \LaTeX\ templates
are available on the conference web site at:
\begin{center}
\textbf{\texttt{http://icml.cc/}}
\end{center}

The guidelines below will be enforced for initial submissions and
camera-ready copies. Here is a brief summary:
\begin{itemize}
\item Submissions must be in PDF\@.
\item Submitted papers can be up to eight pages long, not including references, plus unlimited space for references. Accepted papers can be up to nine pages long, not including references, to allow authors to address reviewer comments. Any paper exceeding this length will automatically be rejected. 
\item \textbf{Do not include author information or acknowledgements} in your
    initial submission.
\item Your paper should be in \textbf{10 point Times font}.
\item Make sure your PDF file only uses Type-1 fonts.
\item Place figure captions \emph{under} the figure (and omit titles from inside
    the graphic file itself). Place table captions \emph{over} the table.
\item References must include page numbers whenever possible and be as complete
    as possible. Place multiple citations in chronological order.
\item Do not alter the style template; in particular, do not compress the paper
    format by reducing the vertical spaces.
\item Keep your abstract brief and self-contained, one paragraph and roughly
    4--6 sentences. Gross violations will require correction at the
    camera-ready phase. The title should have content words capitalized.
\end{itemize}

\subsection{Submitting Papers}

\textbf{Paper Deadline:} The deadline for paper submission that is
advertised on the conference website is strict. If your full,
anonymized, submission does not reach us on time, it will not be
considered for publication.

\textbf{Anonymous Submission:} ICML uses double-blind review: no identifying
author information may appear on the title page or in the paper
itself. Section~\ref{author info} gives further details.

\textbf{Simultaneous Submission:} ICML will not accept any paper which,
at the time of submission, is under review for another conference or
has already been published. This policy also applies to papers that
overlap substantially in technical content with conference papers
under review or previously published. ICML submissions must not be
submitted to other conferences during ICML's review period. Authors
may submit to ICML substantially different versions of journal papers
that are currently under review by the journal, but not yet accepted
at the time of submission. Informal publications, such as technical
reports or papers in workshop proceedings which do not appear in
print, do not fall under these restrictions.

\medskip

Authors must provide their manuscripts in \textbf{PDF} format.
Furthermore, please make sure that files contain only embedded Type-1 fonts
(e.g.,~using the program \texttt{pdffonts} in linux or using
File/DocumentProperties/Fonts in Acrobat). Other fonts (like Type-3)
might come from graphics files imported into the document.

Authors using \textbf{Word} must convert their document to PDF\@. Most
of the latest versions of Word have the facility to do this
automatically. Submissions will not be accepted in Word format or any
format other than PDF\@. Really. We're not joking. Don't send Word.

Those who use \textbf{\LaTeX} should avoid including Type-3 fonts.
Those using \texttt{latex} and \texttt{dvips} may need the following
two commands:

{\footnotesize
\begin{verbatim}
dvips -Ppdf -tletter -G0 -o paper.ps paper.dvi
ps2pdf paper.ps
\end{verbatim}}
It is a zero following the ``-G'', which tells dvips to use
the config.pdf file. Newer \TeX\ distributions don't always need this
option.

Using \texttt{pdflatex} rather than \texttt{latex}, often gives better
results. This program avoids the Type-3 font problem, and supports more
advanced features in the \texttt{microtype} package.

\textbf{Graphics files} should be a reasonable size, and included from
an appropriate format. Use vector formats (.eps/.pdf) for plots,
lossless bitmap formats (.png) for raster graphics with sharp lines, and
jpeg for photo-like images.

The style file uses the \texttt{hyperref} package to make clickable
links in documents. If this causes problems for you, add
\texttt{nohyperref} as one of the options to the \texttt{icml2020}
usepackage statement.

\subsection{Submitting Final Camera-Ready Copy}

The final versions of papers accepted for publication should follow the
same format and naming convention as initial submissions, except that
author information (names and affiliations) should be given. See
Section~\ref{final author} for formatting instructions.

The footnote, ``Preliminary work. Under review by the International
Conference on Machine Learning (ICML). Do not distribute.'' must be
modified to ``\textit{Proceedings of the
$\mathit{37}^{th}$ International Conference on Machine Learning},
Vienna, Austria, 2020.
Copyright 2020 by the author(s).''

For those using the \textbf{\LaTeX} style file, this change (and others) is
handled automatically by simply changing
$\mathtt{\backslash usepackage\{icml2020\}}$ to
$$\mathtt{\backslash usepackage[accepted]\{icml2020\}}$$
Authors using \textbf{Word} must edit the
footnote on the first page of the document themselves.

Camera-ready copies should have the title of the paper as running head
on each page except the first one. The running title consists of a
single line centered above a horizontal rule which is $1$~point thick.
The running head should be centered, bold and in $9$~point type. The
rule should be $10$~points above the main text. For those using the
\textbf{\LaTeX} style file, the original title is automatically set as running
head using the \texttt{fancyhdr} package which is included in the ICML
2020 style file package. In case that the original title exceeds the
size restrictions, a shorter form can be supplied by using

\verb|\icmltitlerunning{...}|

just before $\mathtt{\backslash begin\{document\}}$.
Authors using \textbf{Word} must edit the header of the document themselves.

\section{Format of the Paper}

All submissions must follow the specified format.

\subsection{Dimensions}

The text of the paper should be formatted in two columns, with an
overall width of 6.75~inches, height of 9.0~inches, and 0.25~inches
between the columns. The left margin should be 0.75~inches and the top
margin 1.0~inch (2.54~cm). The right and bottom margins will depend on
whether you print on US letter or A4 paper, but all final versions
must be produced for US letter size.

The paper body should be set in 10~point type with a vertical spacing
of 11~points. Please use Times typeface throughout the text.

\subsection{Title}

The paper title should be set in 14~point bold type and centered
between two horizontal rules that are 1~point thick, with 1.0~inch
between the top rule and the top edge of the page. Capitalize the
first letter of content words and put the rest of the title in lower
case.

\subsection{Author Information for Submission}
\label{author info}

ICML uses double-blind review, so author information must not appear. If
you are using \LaTeX\/ and the \texttt{icml2020.sty} file, use
\verb+\icmlauthor{...}+ to specify authors and \verb+\icmlaffiliation{...}+ to specify affiliations. (Read the TeX code used to produce this document for an example usage.) The author information
will not be printed unless \texttt{accepted} is passed as an argument to the
style file.
Submissions that include the author information will not
be reviewed.

\subsubsection{Self-Citations}

If you are citing published papers for which you are an author, refer
to yourself in the third person. In particular, do not use phrases
that reveal your identity (e.g., ``in previous work \cite{langley00}, we
have shown \ldots'').

Do not anonymize citations in the reference section. The only exception are manuscripts that are
not yet published (e.g., under submission). If you choose to refer to
such unpublished manuscripts \cite{anonymous}, anonymized copies have
to be submitted
as Supplementary Material via CMT\@. However, keep in mind that an ICML
paper should be self contained and should contain sufficient detail
for the reviewers to evaluate the work. In particular, reviewers are
not required to look at the Supplementary Material when writing their
review.

\subsubsection{Camera-Ready Author Information}
\label{final author}

If a paper is accepted, a final camera-ready copy must be prepared.
%
For camera-ready papers, author information should start 0.3~inches below the
bottom rule surrounding the title. The authors' names should appear in 10~point
bold type, in a row, separated by white space, and centered. Author names should
not be broken across lines. Unbolded superscripted numbers, starting 1, should
be used to refer to affiliations.

Affiliations should be numbered in the order of appearance. A single footnote
block of text should be used to list all the affiliations. (Academic
affiliations should list Department, University, City, State/Region, Country.
Similarly for industrial affiliations.)

Each distinct affiliations should be listed once. If an author has multiple
affiliations, multiple superscripts should be placed after the name, separated
by thin spaces. If the authors would like to highlight equal contribution by
multiple first authors, those authors should have an asterisk placed after their
name in superscript, and the term ``\textsuperscript{*}Equal contribution"
should be placed in the footnote block ahead of the list of affiliations. A
list of corresponding authors and their emails (in the format Full Name
\textless{}email@domain.com\textgreater{}) can follow the list of affiliations.
Ideally only one or two names should be listed.

A sample file with author names is included in the ICML2020 style file
package. Turn on the \texttt{[accepted]} option to the stylefile to
see the names rendered. All of the guidelines above are implemented
by the \LaTeX\ style file.

\subsection{Abstract}

The paper abstract should begin in the left column, 0.4~inches below the final
address. The heading `Abstract' should be centered, bold, and in 11~point type.
The abstract body should use 10~point type, with a vertical spacing of
11~points, and should be indented 0.25~inches more than normal on left-hand and
right-hand margins. Insert 0.4~inches of blank space after the body. Keep your
abstract brief and self-contained, limiting it to one paragraph and roughly 4--6
sentences. Gross violations will require correction at the camera-ready phase.

\subsection{Partitioning the Text}

You should organize your paper into sections and paragraphs to help
readers place a structure on the material and understand its
contributions.

\subsubsection{Sections and Subsections}

Section headings should be numbered, flush left, and set in 11~pt bold
type with the content words capitalized. Leave 0.25~inches of space
before the heading and 0.15~inches after the heading.

Similarly, subsection headings should be numbered, flush left, and set
in 10~pt bold type with the content words capitalized. Leave
0.2~inches of space before the heading and 0.13~inches afterward.

Finally, subsubsection headings should be numbered, flush left, and
set in 10~pt small caps with the content words capitalized. Leave
0.18~inches of space before the heading and 0.1~inches after the
heading.

Please use no more than three levels of headings.

\subsubsection{Paragraphs and Footnotes}

Within each section or subsection, you should further partition the
paper into paragraphs. Do not indent the first line of a given
paragraph, but insert a blank line between succeeding ones.

You can use footnotes\footnote{Footnotes
should be complete sentences.} to provide readers with additional
information about a topic without interrupting the flow of the paper.
Indicate footnotes with a number in the text where the point is most
relevant. Place the footnote in 9~point type at the bottom of the
column in which it appears. Precede the first footnote in a column
with a horizontal rule of 0.8~inches.\footnote{Multiple footnotes can
appear in each column, in the same order as they appear in the text,
but spread them across columns and pages if possible.}

\begin{figure}[ht]
\vskip 0.2in
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{icml_numpapers}}
\caption{Historical locations and number of accepted papers for International
Machine Learning Conferences (ICML 1993 -- ICML 2008) and International
Workshops on Machine Learning (ML 1988 -- ML 1992). At the time this figure was
produced, the number of accepted papers for ICML 2008 was unknown and instead
estimated.}
\label{icml-historical}
\end{center}
\vskip -0.2in
\end{figure}

\subsection{Figures}

You may want to include figures in the paper to illustrate
your approach and results. Such artwork should be centered,
legible, and separated from the text. Lines should be dark and at
least 0.5~points thick for purposes of reproduction, and text should
not appear on a gray background.

Label all distinct components of each figure. If the figure takes the
form of a graph, then give a name for each axis and include a legend
that briefly describes each curve. Do not include a title inside the
figure; instead, the caption should serve this function.

Number figures sequentially, placing the figure number and caption
\emph{after} the graphics, with at least 0.1~inches of space before
the caption and 0.1~inches after it, as in
Figure~\ref{icml-historical}. The figure caption should be set in
9~point type and centered unless it runs two or more lines, in which
case it should be flush left. You may float figures to the top or
bottom of a column, and you may set wide figures across both columns
(use the environment \texttt{figure*} in \LaTeX). Always place
two-column figures at the top or bottom of the page.

\subsection{Algorithms}

If you are using \LaTeX, please use the ``algorithm'' and ``algorithmic''
environments to format pseudocode. These require
the corresponding stylefiles, algorithm.sty and
algorithmic.sty, which are supplied with this package.
Algorithm~\ref{alg:example} shows an example.

\begin{algorithm}[tb]
   \caption{Bubble Sort}
   \label{alg:example}
\begin{algorithmic}
   \STATE {\bfseries Input:} data $x_i$, size $m$
   \REPEAT
   \STATE Initialize $noChange = true$.
   \FOR{$i=1$ {\bfseries to} $m-1$}
   \IF{$x_i > x_{i+1}$}
   \STATE Swap $x_i$ and $x_{i+1}$
   \STATE $noChange = false$
   \ENDIF
   \ENDFOR
   \UNTIL{$noChange$ is $true$}
\end{algorithmic}
\end{algorithm}

\subsection{Tables}

You may also want to include tables that summarize material. Like
figures, these should be centered, legible, and numbered consecutively.
However, place the title \emph{above} the table with at least
0.1~inches of space before the title and the same after it, as in
Table~\ref{sample-table}. The table title should be set in 9~point
type and centered unless it runs two or more lines, in which case it
should be flush left.

% Note use of \abovespace and \belowspace to get reasonable spacing
% above and below tabular lines.

\begin{table}[t]
\caption{Classification accuracies for naive Bayes and flexible
Bayes on various data sets.}
\label{sample-table}
\vskip 0.15in
\begin{center}
\begin{small}
\begin{sc}
\begin{tabular}{lcccr}
\toprule
Data set & Naive & Flexible & Better? \\
\midrule
Breast    & 95.9$\pm$ 0.2& 96.7$\pm$ 0.2& $\surd$ \\
Cleveland & 83.3$\pm$ 0.6& 80.0$\pm$ 0.6& $\times$\\
Glass2    & 61.9$\pm$ 1.4& 83.8$\pm$ 0.7& $\surd$ \\
Credit    & 74.8$\pm$ 0.5& 78.3$\pm$ 0.6&         \\
Horse     & 73.3$\pm$ 0.9& 69.7$\pm$ 1.0& $\times$\\
Meta      & 67.1$\pm$ 0.6& 76.5$\pm$ 0.5& $\surd$ \\
Pima      & 75.1$\pm$ 0.6& 73.9$\pm$ 0.5&         \\
Vehicle   & 44.9$\pm$ 0.6& 61.5$\pm$ 0.4& $\surd$ \\
\bottomrule
\end{tabular}
\end{sc}
\end{small}
\end{center}
\vskip -0.1in
\end{table}

Tables contain textual material, whereas figures contain graphical material.
Specify the contents of each row and column in the table's topmost
row. Again, you may float tables to a column's top or bottom, and set
wide tables across both columns. Place two-column tables at the
top or bottom of the page.

\subsection{Citations and References}

Please use APA reference format regardless of your formatter
or word processor. If you rely on the \LaTeX\/ bibliographic
facility, use \texttt{natbib.sty} and \texttt{icml2020.bst}
included in the style-file package to obtain this format.

Citations within the text should include the authors' last names and
year. If the authors' names are included in the sentence, place only
the year in parentheses, for example when referencing Arthur Samuel's
pioneering work \yrcite{Samuel59}. Otherwise place the entire
reference in parentheses with the authors and year separated by a
comma \cite{Samuel59}. List multiple references separated by
semicolons \cite{kearns89,Samuel59,mitchell80}. Use the `et~al.'
construct only for citations with three or more authors or after
listing all authors to a publication in an earlier reference \cite{MachineLearningI}.

Authors should cite their own work in the third person
in the initial version of their paper submitted for blind review.
Please refer to Section~\ref{author info} for detailed instructions on how to
cite your own papers.

Use an unnumbered first-level section heading for the references, and use a
hanging indent style, with the first line of the reference flush against the
left margin and subsequent lines indented by 10 points. The references at the
end of this document give examples for journal articles \cite{Samuel59},
conference publications \cite{langley00}, book chapters \cite{Newell81}, books
\cite{DudaHart2nd}, edited volumes \cite{MachineLearningI}, technical reports
\cite{mitchell80}, and dissertations \cite{kearns89}.

Alphabetize references by the surnames of the first authors, with
single author entries preceding multiple author entries. Order
references for the same authors by year of publication, with the
earliest first. Make sure that each reference includes all relevant
information (e.g., page numbers).

Please put some effort into making references complete, presentable, and
consistent. If using bibtex, please protect capital letters of names and
abbreviations in titles, for example, use \{B\}ayesian or \{L\}ipschitz
in your .bib file.

\section*{Software and Data}

If a paper is accepted, we strongly encourage the publication of software and data with the
camera-ready version of the paper whenever appropriate. This can be
done by including a URL in the camera-ready copy. However, \textbf{do not}
include URLs that reveal your institution or identity in your
submission for review. Instead, provide an anonymous URL or upload
the material as ``Supplementary Material'' into the CMT reviewing
system. Note that reviewers are not required to look at this material
when writing their review.

% Acknowledgements should only appear in the accepted version.
\section*{Acknowledgements}

\textbf{Do not} include acknowledgements in the initial version of
the paper submitted for blind review.

If a paper is accepted, the final camera-ready version can (and
probably should) include acknowledgements. In this case, please
place such acknowledgements in an unnumbered section at the
end of the paper. Typically, this will include thanks to reviewers
who gave useful comments, to colleagues who contributed to the ideas,
and to funding agencies and corporate sponsors that provided financial
support.

% In the unusual situation where you want a paper to appear in the
% references without citing it in the main text, use \nocite
\nocite{langley00}

\bibliography{example_paper}
\bibliographystyle{icml2020}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% DELETE THIS PART. DO NOT PLACE CONTENT AFTER THE REFERENCES!
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\appendix
\section{Do \emph{not} have an appendix here}

\textbf{\emph{Do not put content after the references.}}
%
Put anything that you might normally include after the references in a separate
supplementary file.

We recommend that you build supplementary material in a separate document.
If you must create one PDF and cut it up, please be careful to use a tool that
doesn't alter the margins, and that doesn't aggressively rewrite the PDF file.
pdftk usually works fine.

\textbf{Please do not use Apple's preview to cut off supplementary material.} In
previous years it has altered margins, and created headaches at the camera-ready
stage. 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

% This document was modified from the file originally made available by
% Pat Langley and Andrea Danyluk for ICML-2K. This version was created
% by Iain Murray in 2018, and modified by Alexandre Bouchard in
% 2019 and 2020. Previous contributors include Dan Roy, Lise Getoor and Tobias
% Scheffer, which was slightly modified from the 2010 version by
% Thorsten Joachims & Johannes Fuernkranz, slightly modified from the
% 2009 version by Kiri Wagstaff and Sam Roweis's 2008 version, which is
% slightly modified from Prasad Tadepalli's 2007 version which is a
% lightly changed version of the previous year's version by Andrew
% Moore, which was in turn edited from those of Kristian Kersting and
% Codrina Lauth. Alex Smola contributed to the algorithmic style files.

### 2020-06-03 ICML_paper.Rmd
Create ICML_paper.Rmd

---
title: "ICML submission"
output: html_notebook
---

## In general

There is clear problem, existing methodology, a new proposed theta for identifying non-convergence, and the simulation approach results in a clear investigation into the convergence properties of iterative imputation algorithms. We propose guidelines for practice, identify pitfalls that arise with contemporary evaluation functions, and detail how to avoid such problems.

## Aim

Obtain valid inferences from incomplete data with statistical analyses or (un)supervised machine learning techniques. E.g., convey uncertainty due to missingness in potential outcomes framework for causal inference. Convergence of the algorithm used to solve the missing data problem is a topic that has not received much attention but is ever so important, as most imputation software packages draw inference from iterative imputation procedures.

## Missing data method
MI is...

## Problem

The validity of the inference all depends on the state space of the algorithm at the final iteration. If the algorithm does not reach convergence, "the imputed datasets will not be truly independent and the variability among them may understate the true levels of missing-data uncertainty" (Schafer and Olsen, p. 556). But when is this the case? Current practice of visually inspecting imputations is not sufficient because 1) no objective point at which convergence is established, 2) only severely pathological cases of non-convergence may be identified, and 3) diagnosing non-convergence can be challenging to the untrained eye. On top of that, the statistics to inspect are either univariate or dependent on the substantive model of interest. Converged univariate statistics do not guarantee multivariate convergence, and with a model-dependent statistic there is no guarantee that the algorithm is converged *enough* for other models of scientific interest. This negates the advantages of MI: solving the missing data - and scientific problem separately.

## Solution

Use quantitative diagnostic methods to identify non-convergence in the iterative imputation algorithm. We propose ...

## Simulation study

We simulate and investigate the plausibility of identifiers for non-convergence for iterative imputation algorithms ...

## Results

Valid inferences may be obtained when there is still non-convergence in the algorithm.

## Discussion

Does this also work for other situations? ...

### 2020-06-03 ICML_paper.Rmd
Add call for papers

@@
 ---
+
+# Proposal
 
@@
 Does this also work for other situations? ...
+
+# Call for papers (for reference)
+
+Analysis of large amounts of data offers new opportunities to understand many processes better.
+Yet, data accumulation often implies relaxing acquisition procedures or compounding diverse sources, leading to many observations with missing features.
+From questionnaires to collaborative filtering, from electronic health records to single-cell analysis, missingness is everywhere at play and is rather the norm than the exception.
+Even "clean" data sets are often barely "cleaned" versions of incomplete data sets—with all the unfortunate biases this cleaning process may have created.
+
+Despite this ubiquity, tackling missing values is often overlooked.
+Handling missing values poses many challenges, and there is a vast literature in the statistical community, with many implementations available.
+Yet, there are still many open issues and the need to design new methods or to introduce new point of views: for missing values in a supervised-learning setting, in deep learning architectures, to adapt available methods for high dimensional observed data with different type of missing values, deal with feature mismatch and distribution mismatch.
+Missing data is one of the eight pillars of causal wisdom for Judea Pearl who brought graphical model reasoning to tackle some missing not at random values.
+
+The goal of the Art of Learning with Missing Values (ARTEMISS) workshop is to give more momentum and exposition to research on missing values, both theoretical, methodological, and applied, and emphasize the connections with other areas of machine learning (e.g.
+causal inference, semi-supervised learning, generative modelling, uncertainty quantification, transfer learning, distributional shift, etc.).
+We will also attach importance to discussing the reproducibility problems that can be caused by missing data, the danger of forgetting the missing values issues and the importance of providing sound implementations.
+
+https://artemiss-workshop.github.io/

### 2020-06-04 ICML/ARTEMISS_proposal.tex
ICML proposal in correct format, minor changes to thesis manuscript (typo, figure label, etc.)

--- 2020-06-02 ICML/example_paper.tex
@@
 % Therefore, a short form for the running title is supplied here:
-\icmltitlerunning{Submission and Formatting Instructions for ICML 2020}
+\icmltitlerunning{Missing the Point}
 
@@
 \twocolumn[
-\icmltitle{Submission and Formatting Instructions for \\
-           International Conference on Machine Learning (ICML 2020)}
+\icmltitle{Missing the Point: Non-Convergence in Iterative Imputation Algorithms}
 
@@
 \begin{icmlauthorlist}
-\icmlauthor{Aeiau Zzzz}{equal,to}
-\icmlauthor{Bauiu C.~Yyyy}{equal,to,goo}
-\icmlauthor{Cieua Vvvvv}{goo}
-\icmlauthor{Iaesut Saoeu}{ed}
-\icmlauthor{Fiuea Rrrr}{to}
-\icmlauthor{Tateu H.~Yasehe}{ed,to,goo}
-\icmlauthor{Aaoeu Iasoh}{goo}
-\icmlauthor{Buiui Eueu}{ed}
-\icmlauthor{Aeuia Zzzz}{ed}
-\icmlauthor{Bieea C.~Yyyy}{to,goo}
-\icmlauthor{Teoau Xxxx}{ed}
-\icmlauthor{Eee Pppp}{ed}
+\icmlauthor{Hanne I.~Oberman}{UU}
+\icmlauthor{Stef van~Buuren}{UU}
+\icmlauthor{Gerko Vink}{UU}
 \end{icmlauthorlist}
 
-\icmlaffiliation{to}{Department of Computation, University of Torontoland, Torontoland, Canada}
-\icmlaffiliation{goo}{Googol ShallowMind, New London, Michigan, USA}
-\icmlaffiliation{ed}{School of Computation, University of Edenborrow, Edenborrow, United Kingdom}
+\icmlaffiliation{UU}{Department of Methodology and Statistics, Utrecht University, Utrecht, The Netherlands}
 
-\icmlcorrespondingauthor{Cieua Vvvvv}{c.vvvvv@googol.com}
-\icmlcorrespondingauthor{Eee Pppp}{ep@eden.co.uk}
+\icmlcorrespondingauthor{Hanne Oberman}{h.i.oberman@uu.nl}
 
@@
 % the "keywords" metadata in the PDF but will not be shown in the document
-\icmlkeywords{Machine Learning, ICML}
+\icmlkeywords{Iterative imputation, MICE, non-convergence}
 
@@
 \begin{abstract}
-This document provides a basic paper template and submission guidelines.
-Abstracts must be a single paragraph, ideally between 4--6 sentences long.
-Gross violations will trigger corrections at the camera-ready phase.
+Iterative imputation is a popular tool to accommodate missing data.
+While it is widely accepted that valid inferences can be obtained with this technique, these inferences all rely on algorithmic convergence.
+There is no consensus on how to evaluate the convergence properties of the method.
+This paper provides insight into identifying non-convergence in iterative imputation algorithms.
+Our study found that---in the cases considered---inferential validity was achieved after five to ten iterations, much earlier than indicated by diagnostic methods.
+We conclude that it never hurts to iterate longer, but such calculations hardly bring added value.
 \end{abstract}
 
-\section{Electronic Submission}
-\label{submission}
+\section{Introduction}
+\label{intro}
 
-Submission to ICML 2020 will be entirely electronic, via a web site
-(not email). Information about the submission process and \LaTeX\ templates
-are available on the conference web site at:
-\begin{center}
-\textbf{\texttt{http://icml.cc/}}
-\end{center}
+A popular method to accommodate missing data is to \textit{impute} (i.e., `fill in') the missing values in an incomplete dataset.
+It is widely accepted that imputation techniques such as multiple imputation (MI; @rubin76) can yield statistically valid inferences.
+The validity of these inferences relies on the method through which imputations are obtained---often iterative algorithms.
+Iterative imputation algorithms are employed in e.g., $\mathtt{SPSS}$, $\mathtt{Stata}$, and the $\mathtt{R}$ packages $\mathtt{MI}$, and $\mathtt{mice}$.
+\textbf{Or use:} Convergence of the algorithm used to solve the missing data problem is a topic that has not received much attention but is ever so important, as most imputation software packages draw inference from iterative imputation procedures.
 
-The guidelines below will be enforced for initial submissions and
-camera-ready copies. Here is a brief summary:
-\begin{itemize}
-\item Submissions must be in PDF\@.
-\item Submitted papers can be up to eight pages long, not including references, plus unlimited space for references.
-Accepted papers can be up to nine pages long, not including references, to allow authors to address reviewer comments.
-Any paper exceeding this length will automatically be rejected.
-\item \textbf{Do not include author information or acknowledgements} in your
-    initial submission.
-\item Your paper should be in \textbf{10 point Times font}.
-\item Make sure your PDF file only uses Type-1 fonts.
-\item Place figure captions \emph{under} the figure (and omit titles from inside
-    the graphic file itself). Place table captions \emph{over} the table.
-\item References must include page numbers whenever possible and be as complete
-    as possible. Place multiple citations in chronological order.
-\item Do not alter the style template; in particular, do not compress the paper
-    format by reducing the vertical spaces.
-\item Keep your abstract brief and self-contained, one paragraph and roughly
-    4--6 sentences. Gross violations will require correction at the
-    camera-ready phase. The title should have content words capitalized.
-\end{itemize}
+The validity of the inference all depends on the state space of the algorithm at the final iteration.
+If the algorithm does not reach convergence, ``the imputed datasets will not be truly independent and the variability among them may understate the true levels of missing-data uncertainty'' (Schafer and Olsen, p.
+556).
 
-\subsection{Submitting Papers}
+But when is this the case?
+Current practice of visually inspecting imputations is not sufficient because 1) no objective point at which convergence is established, 2) only severely pathological cases of non-convergence may be identified, and 3) diagnosing non-convergence can be challenging to the untrained eye.
+On top of that, the statistics to inspect are either univariate or dependent on the substantive model of interest.
+Converged univariate statistics do not guarantee multivariate convergence, and with a model-dependent statistic there is no guarantee that the algorithm is converged \textit{enough} for other models of scientific interest.
+This negates the advantages of MI: solving the missing data problem and complete data problem separately.
 
-\textbf{Paper Deadline:} The deadline for paper submission that is
-advertised on the conference website is strict. If your full,
-anonymized, submission does not reach us on time, it will not be
-considered for publication.
+Therefore, we propose a novel parameter to inspect. We use this parameter to identify non-convergence with two quantitative diagnostic methods: Rhat and AC.
 
-\textbf{Anonymous Submission:} ICML uses double-blind review: no identifying
-author information may appear on the title page or in the paper
-itself. Section~\ref{author info} gives further details.
+We simulate and investigate the plausibility of identifiers for non-convergence for iterative imputation algorithms ...
 
-\textbf{Simultaneous Submission:} ICML will not accept any paper which,
-at the time of submission, is under review for another conference or
-has already been published. This policy also applies to papers that
-overlap substantially in technical content with conference papers
-under review or previously published. ICML submissions must not be
-submitted to other conferences during ICML's review period. Authors
-may submit to ICML substantially different versions of journal papers
-that are currently under review by the journal, but not yet accepted
-at the time of submission. Informal publications, such as technical
-reports or papers in workshop proceedings which do not appear in
-print, do not fall under these restrictions.
+\section{Discussion}
+\label{discussion}
 
-\medskip
-
-Authors must provide their manuscripts in \textbf{PDF} format.
-Furthermore, please make sure that files contain only embedded Type-1 fonts
-(e.g.,~using the program \texttt{pdffonts} in linux or using
-File/DocumentProperties/Fonts in Acrobat). Other fonts (like Type-3)
-might come from graphics files imported into the document.
-
-Authors using \textbf{Word} must convert their document to PDF\@. Most
-of the latest versions of Word have the facility to do this
-automatically. Submissions will not be accepted in Word format or any
-format other than PDF\@. Really. We're not joking. Don't send Word.
-
-Those who use \textbf{\LaTeX} should avoid including Type-3 fonts.
-Those using \texttt{latex} and \texttt{dvips} may need the following
-two commands:
-
-{\footnotesize
-\begin{verbatim}
-dvips -Ppdf -tletter -G0 -o paper.ps paper.dvi
-ps2pdf paper.ps
-\end{verbatim}}
-It is a zero following the ``-G'', which tells dvips to use
-the config.pdf file. Newer \TeX\ distributions don't always need this
-option.
-
-Using \texttt{pdflatex} rather than \texttt{latex}, often gives better
-results. This program avoids the Type-3 font problem, and supports more
-advanced features in the \texttt{microtype} package.
-
-\textbf{Graphics files} should be a reasonable size, and included from
-an appropriate format. Use vector formats (.eps/.pdf) for plots,
-lossless bitmap formats (.png) for raster graphics with sharp lines, and
-jpeg for photo-like images.
-
-The style file uses the \texttt{hyperref} package to make clickable
-links in documents. If this causes problems for you, add
-\texttt{nohyperref} as one of the options to the \texttt{icml2020}
-usepackage statement.
-
-\subsection{Submitting Final Camera-Ready Copy}
-
-The final versions of papers accepted for publication should follow the
-same format and naming convention as initial submissions, except that
-author information (names and affiliations) should be given. See
-Section~\ref{final author} for formatting instructions.
-
-The footnote, ``Preliminary work. Under review by the International
-Conference on Machine Learning (ICML). Do not distribute.'' must be
-modified to ``\textit{Proceedings of the
-$\mathit{37}^{th}$ International Conference on Machine Learning},
-Vienna, Austria, 2020.
-Copyright 2020 by the author(s).''
-
-For those using the \textbf{\LaTeX} style file, this change (and others) is
-handled automatically by simply changing
-$\mathtt{\backslash usepackage\{icml2020\}}$ to
-$$\mathtt{\backslash usepackage[accepted]\{icml2020\}}$$
-Authors using \textbf{Word} must edit the
-footnote on the first page of the document themselves.
-
-Camera-ready copies should have the title of the paper as running head
-on each page except the first one. The running title consists of a
-single line centered above a horizontal rule which is $1$~point thick.
-The running head should be centered, bold and in $9$~point type. The
-rule should be $10$~points above the main text. For those using the
-\textbf{\LaTeX} style file, the original title is automatically set as running
-head using the \texttt{fancyhdr} package which is included in the ICML
-2020 style file package. In case that the original title exceeds the
-size restrictions, a shorter form can be supplied by using
-
-\verb|\icmltitlerunning{...}|
-
-just before $\mathtt{\backslash begin\{document\}}$.
-Authors using \textbf{Word} must edit the header of the document themselves.
-
-\section{Format of the Paper}
-
-All submissions must follow the specified format.
-
-\subsection{Dimensions}
-
-The text of the paper should be formatted in two columns, with an
-overall width of 6.75~inches, height of 9.0~inches, and 0.25~inches
-between the columns. The left margin should be 0.75~inches and the top
-margin 1.0~inch (2.54~cm). The right and bottom margins will depend on
-whether you print on US letter or A4 paper, but all final versions
-must be produced for US letter size.
-
-The paper body should be set in 10~point type with a vertical spacing
-of 11~points. Please use Times typeface throughout the text.
-
-\subsection{Title}
-
-The paper title should be set in 14~point bold type and centered
-between two horizontal rules that are 1~point thick, with 1.0~inch
-between the top rule and the top edge of the page. Capitalize the
-first letter of content words and put the rest of the title in lower
-case.
-
-\subsection{Author Information for Submission}
-\label{author info}
-
-ICML uses double-blind review, so author information must not appear. If
-you are using \LaTeX\/ and the \texttt{icml2020.sty} file, use
-\verb+\icmlauthor{...}+ to specify authors and \verb+\icmlaffiliation{...}+ to specify affiliations. (Read the TeX code used to produce this document for an example usage.) The author information
-will not be printed unless \texttt{accepted} is passed as an argument to the
-style file.
-Submissions that include the author information will not
-be reviewed.
-
-\subsubsection{Self-Citations}
-
-If you are citing published papers for which you are an author, refer
-to yourself in the third person. In particular, do not use phrases
-that reveal your identity (e.g., ``in previous work \cite{langley00}, we
-have shown \ldots'').
-
-Do not anonymize citations in the reference section. The only exception are manuscripts that are
-not yet published (e.g., under submission). If you choose to refer to
-such unpublished manuscripts \cite{anonymous}, anonymized copies have
-to be submitted
-as Supplementary Material via CMT\@. However, keep in mind that an ICML
-paper should be self contained and should contain sufficient detail
-for the reviewers to evaluate the work. In particular, reviewers are
-not required to look at the Supplementary Material when writing their
-review.
-
-\subsubsection{Camera-Ready Author Information}
-\label{final author}
-
-If a paper is accepted, a final camera-ready copy must be prepared.
-%
-For camera-ready papers, author information should start 0.3~inches below the
-bottom rule surrounding the title. The authors' names should appear in 10~point
-bold type, in a row, separated by white space, and centered. Author names should
-not be broken across lines. Unbolded superscripted numbers, starting 1, should
-be used to refer to affiliations.
-
-Affiliations should be numbered in the order of appearance. A single footnote
-block of text should be used to list all the affiliations. (Academic
-affiliations should list Department, University, City, State/Region, Country.
-Similarly for industrial affiliations.)
-
-Each distinct affiliations should be listed once. If an author has multiple
-affiliations, multiple superscripts should be placed after the name, separated
-by thin spaces. If the authors would like to highlight equal contribution by
-multiple first authors, those authors should have an asterisk placed after their
-name in superscript, and the term ``\textsuperscript{*}Equal contribution"
-should be placed in the footnote block ahead of the list of affiliations. A
-list of corresponding authors and their emails (in the format Full Name
-\textless{}email@domain.com\textgreater{}) can follow the list of affiliations.
-Ideally only one or two names should be listed.
-
-A sample file with author names is included in the ICML2020 style file
-package. Turn on the \texttt{[accepted]} option to the stylefile to
-see the names rendered. All of the guidelines above are implemented
-by the \LaTeX\ style file.
-
-\subsection{Abstract}
-
-The paper abstract should begin in the left column, 0.4~inches below the final
-address. The heading `Abstract' should be centered, bold, and in 11~point type.
-The abstract body should use 10~point type, with a vertical spacing of
-11~points, and should be indented 0.25~inches more than normal on left-hand and
-right-hand margins. Insert 0.4~inches of blank space after the body. Keep your
-abstract brief and self-contained, limiting it to one paragraph and roughly 4--6
-sentences. Gross violations will require correction at the camera-ready phase.
-
-\subsection{Partitioning the Text}
-
-You should organize your paper into sections and paragraphs to help
-readers place a structure on the material and understand its
-contributions.
-
-\subsubsection{Sections and Subsections}
-
-Section headings should be numbered, flush left, and set in 11~pt bold
-type with the content words capitalized. Leave 0.25~inches of space
-before the heading and 0.15~inches after the heading.
-
-Similarly, subsection headings should be numbered, flush left, and set
-in 10~pt bold type with the content words capitalized. Leave
-0.2~inches of space before the heading and 0.13~inches afterward.
-
-Finally, subsubsection headings should be numbered, flush left, and
-set in 10~pt small caps with the content words capitalized. Leave
-0.18~inches of space before the heading and 0.1~inches after the
-heading.
-
-Please use no more than three levels of headings.
-
-\subsubsection{Paragraphs and Footnotes}
-
-Within each section or subsection, you should further partition the
-paper into paragraphs. Do not indent the first line of a given
-paragraph, but insert a blank line between succeeding ones.
-
-You can use footnotes\footnote{Footnotes
-should be complete sentences.} to provide readers with additional
-information about a topic without interrupting the flow of the paper.
-Indicate footnotes with a number in the text where the point is most
-relevant. Place the footnote in 9~point type at the bottom of the
-column in which it appears. Precede the first footnote in a column
-with a horizontal rule of 0.8~inches.\footnote{Multiple footnotes can
-appear in each column, in the same order as they appear in the text,
-but spread them across columns and pages if possible.}
-
-\begin{figure}[ht]
-\vskip 0.2in
-\begin{center}
-\centerline{\includegraphics[width=\columnwidth]{icml_numpapers}}
-\caption{Historical locations and number of accepted papers for International
-Machine Learning Conferences (ICML 1993 -- ICML 2008) and International
-Workshops on Machine Learning (ML 1988 -- ML 1992). At the time this figure was
-produced, the number of accepted papers for ICML 2008 was unknown and instead
-estimated.}
-\label{icml-historical}
-\end{center}
-\vskip -0.2in
-\end{figure}
-
-\subsection{Figures}
-
-You may want to include figures in the paper to illustrate
-your approach and results. Such artwork should be centered,
-legible, and separated from the text. Lines should be dark and at
-least 0.5~points thick for purposes of reproduction, and text should
-not appear on a gray background.
-
-Label all distinct components of each figure. If the figure takes the
-form of a graph, then give a name for each axis and include a legend
-that briefly describes each curve. Do not include a title inside the
-figure; instead, the caption should serve this function.
-
-Number figures sequentially, placing the figure number and caption
-\emph{after} the graphics, with at least 0.1~inches of space before
-the caption and 0.1~inches after it, as in
-Figure~\ref{icml-historical}. The figure caption should be set in
-9~point type and centered unless it runs two or more lines, in which
-case it should be flush left. You may float figures to the top or
-bottom of a column, and you may set wide figures across both columns
-(use the environment \texttt{figure*} in \LaTeX). Always place
-two-column figures at the top or bottom of the page.
-
-\subsection{Algorithms}
-
-If you are using \LaTeX, please use the ``algorithm'' and ``algorithmic''
-environments to format pseudocode. These require
-the corresponding stylefiles, algorithm.sty and
-algorithmic.sty, which are supplied with this package.
-Algorithm~\ref{alg:example} shows an example.
-
-\begin{algorithm}[tb]
-   \caption{Bubble Sort}
-   \label{alg:example}
-\begin{algorithmic}
-   \STATE {\bfseries Input:} data $x_i$, size $m$
-   \REPEAT
-   \STATE Initialize $noChange = true$.
-   \FOR{$i=1$ {\bfseries to} $m-1$}
-   \IF{$x_i > x_{i+1}$}
-   \STATE Swap $x_i$ and $x_{i+1}$
-   \STATE $noChange = false$
-   \ENDIF
-   \ENDFOR
-   \UNTIL{$noChange$ is $true$}
-\end{algorithmic}
-\end{algorithm}
-
-\subsection{Tables}
-
-You may also want to include tables that summarize material. Like
-figures, these should be centered, legible, and numbered consecutively.
-However, place the title \emph{above} the table with at least
-0.1~inches of space before the title and the same after it, as in
-Table~\ref{sample-table}. The table title should be set in 9~point
-type and centered unless it runs two or more lines, in which case it
-should be flush left.
-
-% Note use of \abovespace and \belowspace to get reasonable spacing
-% above and below tabular lines.
-
-\begin{table}[t]
-\caption{Classification accuracies for naive Bayes and flexible
-Bayes on various data sets.}
-\label{sample-table}
-\vskip 0.15in
-\begin{center}
-\begin{small}
-\begin{sc}
-\begin{tabular}{lcccr}
-\toprule
-Data set & Naive & Flexible & Better? \\
-\midrule
-Breast    & 95.9$\pm$ 0.2& 96.7$\pm$ 0.2& $\surd$ \\
-Cleveland & 83.3$\pm$ 0.6& 80.0$\pm$ 0.6& $\times$\\
-Glass2    & 61.9$\pm$ 1.4& 83.8$\pm$ 0.7& $\surd$ \\
-Credit    & 74.8$\pm$ 0.5& 78.3$\pm$ 0.6&         \\
-Horse     & 73.3$\pm$ 0.9& 69.7$\pm$ 1.0& $\times$\\
-Meta      & 67.1$\pm$ 0.6& 76.5$\pm$ 0.5& $\surd$ \\
-Pima      & 75.1$\pm$ 0.6& 73.9$\pm$ 0.5&         \\
-Vehicle   & 44.9$\pm$ 0.6& 61.5$\pm$ 0.4& $\surd$ \\
-\bottomrule
-\end{tabular}
-\end{sc}
-\end{small}
-\end{center}
-\vskip -0.1in
-\end{table}
-
-Tables contain textual material, whereas figures contain graphical material.
-Specify the contents of each row and column in the table's topmost
-row. Again, you may float tables to a column's top or bottom, and set
-wide tables across both columns. Place two-column tables at the
-top or bottom of the page.
-
-\subsection{Citations and References}
-
-Please use APA reference format regardless of your formatter
-or word processor. If you rely on the \LaTeX\/ bibliographic
-facility, use \texttt{natbib.sty} and \texttt{icml2020.bst}
-included in the style-file package to obtain this format.
-
-Citations within the text should include the authors' last names and
-year. If the authors' names are included in the sentence, place only
-the year in parentheses, for example when referencing Arthur Samuel's
-pioneering work \yrcite{Samuel59}. Otherwise place the entire
-reference in parentheses with the authors and year separated by a
-comma \cite{Samuel59}. List multiple references separated by
-semicolons \cite{kearns89,Samuel59,mitchell80}. Use the `et~al.'
-construct only for citations with three or more authors or after
-listing all authors to a publication in an earlier reference \cite{MachineLearningI}.
-
-Authors should cite their own work in the third person
-in the initial version of their paper submitted for blind review.
-Please refer to Section~\ref{author info} for detailed instructions on how to
-cite your own papers.
-
-Use an unnumbered first-level section heading for the references, and use a
-hanging indent style, with the first line of the reference flush against the
-left margin and subsequent lines indented by 10 points. The references at the
-end of this document give examples for journal articles \cite{Samuel59},
-conference publications \cite{langley00}, book chapters \cite{Newell81}, books
-\cite{DudaHart2nd}, edited volumes \cite{MachineLearningI}, technical reports
-\cite{mitchell80}, and dissertations \cite{kearns89}.
-
-Alphabetize references by the surnames of the first authors, with
-single author entries preceding multiple author entries. Order
-references for the same authors by year of publication, with the
-earliest first. Make sure that each reference includes all relevant
-information (e.g., page numbers).
-
-Please put some effort into making references complete, presentable, and
-consistent. If using bibtex, please protect capital letters of names and
-abbreviations in titles, for example, use \{B\}ayesian or \{L\}ipschitz
-in your .bib file.
-
-\section*{Software and Data}
-
-If a paper is accepted, we strongly encourage the publication of software and data with the
-camera-ready version of the paper whenever appropriate. This can be
-done by including a URL in the camera-ready copy. However, \textbf{do not}
-include URLs that reveal your institution or identity in your
-submission for review. Instead, provide an anonymous URL or upload
-the material as ``Supplementary Material'' into the CMT reviewing
-system. Note that reviewers are not required to look at this material
-when writing their review.
-
-% Acknowledgements should only appear in the accepted version.
-\section*{Acknowledgements}
-
-\textbf{Do not} include acknowledgements in the initial version of
-the paper submitted for blind review.
-
-If a paper is accepted, the final camera-ready version can (and
-probably should) include acknowledgements. In this case, please
-place such acknowledgements in an unnumbered section at the
-end of the paper. Typically, this will include thanks to reviewers
-who gave useful comments, to colleagues who contributed to the ideas,
-and to funding agencies and corporate sponsors that provided financial
-support.
+Valid inferences may be obtained when there is still non-convergence in the algorithm.
 
@@
 % references without citing it in the main text, use \nocite
-\nocite{langley00}
 
@@
 \bibliographystyle{icml2020}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% DELETE THIS PART. DO NOT PLACE CONTENT AFTER THE REFERENCES!
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\appendix
-\section{Do \emph{not} have an appendix here}
-
-\textbf{\emph{Do not put content after the references.}}
-%
-Put anything that you might normally include after the references in a separate
-supplementary file.
-
-We recommend that you build supplementary material in a separate document.
-If you must create one PDF and cut it up, please be careful to use a tool that
-doesn't alter the margins, and that doesn't aggressively rewrite the PDF file.
-pdftk usually works fine.
-
-\textbf{Please do not use Apple's preview to cut off supplementary material.} In
-previous years it has altered margins, and created headaches at the camera-ready
-stage. 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2020-06-04 ICML_paper.Rmd
ICML proposal in correct format, minor changes to thesis manuscript (typo, figure label, etc.)

@@
 title: "ICML submission"
-output: html_notebook
+output:
+  html_document: 
+    theme: journal
+    number_sections: yes
 ---
-
-# Proposal
-
-## In general
-
-There is clear problem, existing methodology, a new proposed theta for identifying non-convergence, and the simulation approach results in a clear investigation into the convergence properties of iterative imputation algorithms.
-We propose guidelines for practice, identify pitfalls that arise with contemporary evaluation functions, and detail how to avoid such problems.
-
-## Aim
-
-Obtain valid inferences from incomplete data with statistical analyses or (un)supervised machine learning techniques.
-E.g., convey uncertainty due to missingness in potential outcomes framework for causal inference.
-Convergence of the algorithm used to solve the missing data problem is a topic that has not received much attention but is ever so important, as most imputation software packages draw inference from iterative imputation procedures.
-
-## Missing data method
-MI is...
-
-## Problem
-
-The validity of the inference all depends on the state space of the algorithm at the final iteration.
-If the algorithm does not reach convergence, "the imputed datasets will not be truly independent and the variability among them may understate the true levels of missing-data uncertainty" (Schafer and Olsen, p.
-556).
-But when is this the case?
-Current practice of visually inspecting imputations is not sufficient because 1) no objective point at which convergence is established, 2) only severely pathological cases of non-convergence may be identified, and 3) diagnosing non-convergence can be challenging to the untrained eye.
-On top of that, the statistics to inspect are either univariate or dependent on the substantive model of interest.
-Converged univariate statistics do not guarantee multivariate convergence, and with a model-dependent statistic there is no guarantee that the algorithm is converged *enough* for other models of scientific interest.
-This negates the advantages of MI: solving the missing data - and scientific problem separately.
-
-## Solution
-
-Use quantitative diagnostic methods to identify non-convergence in the iterative imputation algorithm. We propose ...
-
-## Simulation study
-
-We simulate and investigate the plausibility of identifiers for non-convergence for iterative imputation algorithms ...
-
-## Results
-
-Valid inferences may be obtained when there is still non-convergence in the algorithm.
-
-## Discussion
-
-Does this also work for other situations? ...
 
@@
 The goal of the Art of Learning with Missing Values (ARTEMISS) workshop is to give more momentum and exposition to research on missing values, both theoretical, methodological, and applied, and emphasize the connections with other areas of machine learning (e.g.
-causal inference, semi-supervised learning, generative modelling, uncertainty quantification, transfer learning, distributional shift, etc.).
+causal inference, semi-supervised learning, generative modeling, uncertainty quantification, transfer learning, distributional shift, etc.).
 We will also attach importance to discussing the reproducibility problems that can be caused by missing data, the danger of forgetting the missing values issues and the importance of providing sound implementations.
@@
 https://artemiss-workshop.github.io/
+
+# Proposal
+
+## In general
+
+There is clear problem (how to diagnose non-convergence of iterative imputation algorithms), existing methodology (diagnostic methods to identify non-convergence in MCMC algorithms), and a novel contribution (a new proposed parameter to inspect).
+The simulation approach results in a clear investigation into the convergence properties of iterative imputation algorithms.
+We propose guidelines for practice, identify pitfalls that arise with contemporary evaluation functions, and detail how to avoid such problems.
+We conclude that iterative imputation algorithms may yield statistically valid inferences despite signs of non-convergence in the algorithm.
+
+## Aim of the method
+
+Obtain valid inferences from incomplete data with statistical analyses or (un)supervised machine learning techniques.
+E.g., convey uncertainty due to missingness in potential outcomes framework for causal inference.
+Convergence of the algorithm used to solve the missing data problem is a topic that has not received much attention but is ever so important, as most imputation software packages draw inference from iterative imputation procedures.
+
+## Missing data methodology
+MI is...
+
+## Problem with the method
+
+The validity of the inference all depends on the state space of the algorithm at the final iteration.
+If the algorithm does not reach convergence, "the imputed datasets will not be truly independent and the variability among them may understate the true levels of missing-data uncertainty" (Schafer and Olsen, p.
+556).
+But when is this the case?
+Current practice of visually inspecting imputations is not sufficient because 1) no objective point at which convergence is established, 2) only severely pathological cases of non-convergence may be identified, and 3) diagnosing non-convergence can be challenging to the untrained eye.
+On top of that, the statistics to inspect are either univariate or dependent on the substantive model of interest.
+Converged univariate statistics do not guarantee multivariate convergence, and with a model-dependent statistic there is no guarantee that the algorithm is converged *enough* for other models of scientific interest.
+This negates the advantages of MI: solving the missing data - and scientific problem separately.
+
+## Solution to this problem
+
+Use quantitative diagnostic methods to identify non-convergence in the iterative imputation algorithm. We propose ...
+
+## Simulation study to test solution
+
+We simulate and investigate the plausibility of identifiers for non-convergence for iterative imputation algorithms ...
+
+## Results of simulation
+
+Valid inferences may be obtained when there is still non-convergence in the algorithm.
+
+## Discussion of results
+
+Does this also work for other situations? ...

### 2020-06-04 Writeup/2.Manuscript.Rmd
ICML proposal in correct format, minor changes to thesis manuscript (typo, figure label, etc.)

@@
 for (number of simulation runs from 1 to 1000)
- for (missingness proportions 5%, 25%, 50%, 75% and 95%)
+ for (proportion of missing cases between 5% and 95%)
   2. create missingness
@@
    7. compute performance measures
- 8. combine outcomes of all missingness proportions
-9. aggregate outcomes of all simulation runs
+ 8. combine outcomes of all conditions
+9. aggregate outcomes across simulation runs
 
@@
 Depicted are the bias, coverage rate (CR) and confidence interval width (CIW) of the estimates $\\bar{Q}$.
-We present the subset of $Q$s with the worst performance in terms of bias: the mean and standard deviation of $Y$, and the regression coefficient of $X_1$."}
+We present the subset of $Q$s with the worst performance in terms of bias: the mean and standard deviation of $Y$, and the regression coefficient of $X_1$.
+The solid gray line depicts the targets (unbiased estimates and nominal coverage rate)."}
 mean_bias + sd_bias + Rsq_bias + est_bias + est_cov + est_ciw + plot_layout(ncol = 2, guides = "collect") + plot_annotation(tag_levels = "A", tag_suffix = ".")
@@
 The left-hand side of the figure contains $\\widehat{R}$-values, and the right-hand side contains $AC$-values.
-Depicted in the rows are the scalar summaries $\\theta$: chain mean, chain variance, the quantity of scientific interest $\\hat{Q}$, and the first eigenvalue of the variance-covariance matrix $\\lambda_1$."}
+Depicted in the rows are the scalar summaries $\\theta$: chain mean, chain variance, the quantity of scientific interest $\\hat{Q}$, and the first eigenvalue of the variance-covariance matrix $\\lambda_1$.
+The dashed gray lines depict the different thresholds to diagnose non-convergence."}
 mean_Rh + mean_AC + var_Rh + var_AC + est_Rh + est_AC + PCA_Rh + PCA_AC + plot_layout(guides = "collect", ncol = 2) + plot_annotation(tag_levels = "A", tag_suffix = ".")
@@
 The very strict threshold 1.01 is surpassed in all conditions $T\leq100$ (not shown), and therefore not informative under the current specifications.
-With this set of methods, we fail to identify simulation conditions with persistent bias due to high missingness proportions.
+With this set of methods, we fail to identify non-convergence in conditions with persistent bias due to missingness.
 
@@
 The $\widehat{R}$-values in Figure \ref{fig:rhats-acs}E show a similar trend across iterations as we observed for chain means and chain variances.
-Only conditions with very high missingness proportions ($p_{\rm inc}\geq.75$) diverge from the earlier observations.
+Only conditions with a very high  proportion of missing cases ($p_{\rm inc}\geq.75$) diverge from the earlier observations.
 The $\widehat{R}$-values in these conditions do not taper off after 30 to 40 iterations, but rather between 40 to 50 iterations.
@@
 The $AC$-values also taper off at a later point in the iterations.
-This suggests improving stationarity up-to ten, or even thirty iterations, depending on the missingness proportion.
-Using the threshold, we diagnose non-convergence at some point in all missingness conditions.
+This suggests improving stationarity up-to ten, or even thirty iterations, depending on the missingness condition.
+According to the threshold, we should diagnose non-convergence at some point in any missingness condition.
 However, this occurs only after reasonably decreasing $AC$-values are obtained---again suggesting that non-convergence is incorrectly diagnosed.
@@
 3.
-We hypothesized that the recommended thresholds to diagnose non-convergence with $\widehat{R}$ would be too stringent for iterative imputation applications.
-This hypothesis is only partially supported.
-$\widehat{R}$ indicated improving convergence until 30 to 40 iterations, while approximately unbiased, confidence-valid estimates were obtained after just seven iterations.
-This roughly corresponds to the threshold of 1.2.
-The more stringent threshold 1.1 seems plausible compared to the decrease in $\widehat{R}$-values across iterations.
-The most recent threshold 1.01 does indeed seem to over-estimate the signs of non-convergence.
+We hypothesized that the recommended thresholds to diagnose non-convergence with $\widehat{R}$ would be too stringent for iterative imputation applications, compared to the number of iterations necessary for statistically valid inferences.
+This hypothesis is partially supported: On the one hand, the 1.01 and 1.1 thresholds indeed seem to over-estimate the signs of non-convergence.
+On the other hand, however, we observed that the traditional threshold of 1.2 roughly corresponds to the number of iterations required to obtain approximately unbiased, confidence-valid estimates.
 
@@
 Our final hypothesis was that multivariate $\theta$s would be better at detecting non-convergence than univariate $\theta$s.
-Multivariate $\theta$s did indeed show superior performance under certain conditions (e.g., multivariate $Q$s and extreme missingness proportions).
+Multivariate $\theta$s did indeed show superior performance under certain conditions (e.g., multivariate $Q$s and extreme missingness).
 

### 2020-06-04 ICML/ARTEMISS_proposal.tex
Fill in ICML format from thesis manuscript
Also change label x axis in plot from missingness proportion to proportion of missing cases, and some other things.

@@
 A popular method to accommodate missing data is to \textit{impute} (i.e., `fill in') the missing values in an incomplete dataset.
-It is widely accepted that imputation techniques such as multiple imputation (MI; @rubin76) can yield statistically valid inferences.
+It is widely accepted that imputation techniques such as multiple imputation \citep[MI;][]{rubin76} can yield statistically valid inferences.
 The validity of these inferences relies on the method through which imputations are obtained---often iterative algorithms.
 Iterative imputation algorithms are employed in e.g., $\mathtt{SPSS}$, $\mathtt{Stata}$, and the $\mathtt{R}$ packages $\mathtt{MI}$, and $\mathtt{mice}$.
-\textbf{Or use:} Convergence of the algorithm used to solve the missing data problem is a topic that has not received much attention but is ever so important, as most imputation software packages draw inference from iterative imputation procedures.
+\textit{[Or use:} Convergence of the algorithm used to solve the missing data problem is a topic that has not received much attention but is ever so important, as most imputation software packages draw inference from iterative imputation procedures.]
 
-The validity of the inference all depends on the state space of the algorithm at the final iteration.
-If the algorithm does not reach convergence, ``the imputed datasets will not be truly independent and the variability among them may understate the true levels of missing-data uncertainty'' (Schafer and Olsen, p.
-556).
+With iterative imputation, the validity of the inference depends on the state-space of the algorithm at the final iteration.
+This introduces a potential threat to the validity of the imputations: What if the algorithm has not converged?
+Are the imputations then to be trusted?
+And can we rely on the inference obtained on the completed data?
 
-But when is this the case?
-Current practice of visually inspecting imputations is not sufficient because 1) no objective point at which convergence is established, 2) only severely pathological cases of non-convergence may be identified, and 3) diagnosing non-convergence can be challenging to the untrained eye.
-On top of that, the statistics to inspect are either univariate or dependent on the substantive model of interest.
-Converged univariate statistics do not guarantee multivariate convergence, and with a model-dependent statistic there is no guarantee that the algorithm is converged \textit{enough} for other models of scientific interest.
-This negates the advantages of MI: solving the missing data problem and complete data problem separately.
+These remain open questions since the convergence properties of iterative imputation algorithms have not been systematically studied \citep{buur18}.\footnote{If the algorithm does not reach convergence, ``the imputed datasets will not be truly independent and the variability among them may understate the true levels of missing-data uncertainty'' (Schafer and Olsen, p.
+556).} There is no scientific consensus on how to evaluate the convergence of imputation algorithms \citep{zhu15, taka17}.
+Moreover, the behavior of such algorithms under certain default imputation models (e.g., `predictive mean matching') is an entirely open question \citep{murr18}.
+Therefore, algorithmic convergence should be monitored carefully.
 
-Therefore, we propose a novel parameter to inspect. We use this parameter to identify non-convergence with two quantitative diagnostic methods: Rhat and AC.
+The current practice of visually inspecting imputations may be undesirable for several reasons: 1) it may be challenging to the untrained eye, 2) only severely pathological cases of non-convergence may be diagnosed, and 3) there is not an objective measure that quantifies convergence \citep[\(\S\) 6.5.2]{buur18}.
+On top of that, the recommended parameters to inspect may be insufficient.
+The parameters are scalar summaries of the state-space of the algorithm.
+They are either univariate, or depend on the substantive model of interest.
+Apparent convergence of a univariate parameter does not guarantee convergence of the multivariate state-space of the algorithm.
+And with a model-dependent statistic there is no guarantee that the algorithm is converged \textit{enough} for other models of scientific interest.
+Additionally, model-dependency negates an important advantage of MI: solving the missing data problem separately from the scientific problem.
 
-We simulate and investigate the plausibility of identifiers for non-convergence for iterative imputation algorithms ...
+Therefore, we propose a novel parameter to inspect.
+We use this parameter to identify non-convergence with two quantitative diagnostic methods.
+We use model-based simulation to investigate the plausibility of identifiers for non-convergence for iterative imputation algorithms and provide guidelines for empirical research.
+For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular \texttt{mice} package in \texttt{R} \citep{mice, R}.
+\textit{[Provide notation here?
+E.g., the number of imputations, iterations, etc.]}
+
+\section{Methods}
+\label{methods}
+
+\subsection{Diagnostic Methods}
+We consider several sets of diagnostic methods to identify non-convergence in the iterative imputation algorithm.
+Each method is a combination of a parameter that summarizes the state-space of the algorithm, $\theta$, and an identifier.
+As $\theta$s we use the default parameters of the \texttt{MICE} algorithm (i.e., the univariate $\theta$s chain means and chain variances), a model-dependent user-specified multivariate parameter $Q$, and a novel parameter: $\lambda_1$.
+$\lambda_{1}$ has the appealing property that it is not dependent on the model of scientific interest, yet still summarizes the multivariate state-space of the algorithm.
+
+We define $\lambda_{1}$ as the first eigenvalue of the variance-covariance matrix of the completed data.\footnote{By definition, this is equal to the variance of the first component of the principal component analylis (PCA) solution on the same data.} For each imputation $\ell$, let $\lambda_{1, \ell} \geq \lambda_{2, \ell} \geq ...
+\geq \lambda_{j, \ell}$ be the eigenvalues of the variance-covariance matrix $S_{\ell}$, where $\ell = 1, ..., m$.
+Since the eigenvalues of a variance-covariance matrix summarize the total set of covariances in the data, $\lambda_1$ captures the multivariate nature of the completed data without imposing a substantive model of interest.
+
+The identifiers are as recommended by e.g.
+\cite{cowl96}, namely the potential scale reduction factor $\widehat{R}$ and autocorrelation.
+$\widehat{R}$ is also known as the `Gelman-Rubin statistic' after the original paper \cite{gelm92}.
+We use a recently proposed adapted version of $\widehat{R}$ \citep{veht19}.
+We diagnose non-convergence if $\widehat{R}$-values exceed a certain diagnostic threshold.
+The thresholds under consideration are the traditional threshold $\widehat{R}=1.2$ \citep{gelm92}, the common $\widehat{R}=1.1$ \citep{gelm13}, and the recent $\widehat{R}=1.01$ \citep{veht19}.
+
+We diagnose non-convergence with autocorrelation when the correlation between two subsequent $\theta$-values within the same chain is significantly different from zero \citep[p.~147, p.~42]{lync07, hynd18}.
+
+\subsection{Simulation Study}
+We investigate non-convergence of the iterative imputation algorithm implemented in the $\mathtt{R}$ package $\mathtt{mice}$ through model-based simulation.
+The simulation set-up is summarized in pseudo-code (see Algorithm \ref{alg:setup}).
+The complete script and technical details are available from \href{https://github.com/hanneoberman/MissingThePoint}{github.com/hanneoberman/MissingThePoint}.
+\textit{[Define conditions here somewhere, and $p_{\rm inc}$and $T$.]}
+
+\begin{algorithm}
+   \caption{Simulation set-up}
+   \label{alg:setup}
+\begin{algorithmic}
+\STATE Simulate data 
+\REPEAT 
+\FOR {$p_{\rm inc} = .05, .25, .50, .75, .95$ (missingness conditions)}
+  \STATE Create missingness
+  \FOR {$T = 1, 2, ..., 100$ (early stopping conditions)}
+   \STATE Impute missingness
+   \STATE Perform analysis of scientific interest
+   \STATE Compute non-convergence diagnostics 
+   \STATE Pool results across imputations
+   \STATE Compute performance measures
+   \ENDFOR
+ \ENDFOR 
+ \STATE Combine outcomes of all conditions
+\UNTIL{$n_{\rm sim} = 1000$ (simulation repetitions)}
+\STATE Aggregate outcomes across simulation runs
+\end{algorithmic}
+\end{algorithm}
+
+The aim of the simulation study is twofold: 1) to determine the effect of early stopping and missingness conditions on the validity of statistical inferences, 2) to investigate the plausibility of identifiers for non-convergence.
+
+We assess the validity of statistical inferences with bias, coverage rate and confidence interval width as performance measures (see @buur18 for definitions).
+And we evaluate the plausibility of the identifiers using the diagnostic thresholds for $\widehat{R}$ and autocorrelation (i.e., $\widehat{R}=1.2,1.1,1.01$ and the critical value for AC = 0).
+
+\textit{[Very incomplete from here!]} We consider several quantities of scientific interest.
+
+The statistical inference of interest is multivariate linear regression.
+
+The data generating mechanism of the predictor space is a multivariate normal distribution
+
+\section{Results}
+\label{results}
+
+\begin{figure*}
+{\centering \includegraphics{Figures/estimates} 
+}
+\caption{Impact of non-convergence on statistical inferences.
+Depicted are the bias, coverage rate (CR) and confidence interval width (CIW) of the worst-performing quantities of scientific interest $Q$ in terms of bias.
+The gray lines represent the objectives: unbiased estimates with nominal coverage.}\label{fig:est}
+\end{figure*}
+
+Results are displayed in Figure \ref{fig:est} and \ref{fig:diag}.
+Simulation conditions with a high proportion of missing cases and a low number of iterations most often result in biased estimates and non-nominal coverages.
+
+\textit{[Copy-paste from here!]} These results demonstrate that the estimates of univariate scientific estimands \(Q\) are not impacted by early stopping of the MICE algorithm or the proportion of incomplete cases in \(y\).
+Unbiased estimates may be obtained after just one iteration.
+Multivariate estimates, by contrast, are affected by both the number of iterations and the proportion of missing cases.
+Completely unbiased estimates are only obtained under low to moderate missingness (\(p_{\rm inc}\leq.50\)), after at most three iterations.
+We observe approximately unbiased estimates after at most seven iterations (for any \(p_{\rm inc}\) considered).
+This implies that the algorithm produces stable, non-improving estimates when \(T\geq7\).
+
+\begin{figure*}
+{\centering \includegraphics{Figures/diagnostics} 
+}
+\caption{Non-convergence identified by diagnostic methods: $\widehat{R}$ and $AC$ applied on several $\theta$s.
+The left-hand side of the figure contains $\widehat{R}$-values, and the right-hand side contains $AC$-values.
+Depicted in the rows are the scalar summaries $\theta$: chain mean, chain variance, the quantity of scientific interest $\hat{Q}$, and the first eigenvalue of the variance-covariance matrix $\lambda_1$.}\label{fig:diag}
+\end{figure*}
+
+Overall, the methods to diagnose non-convergence perform as expected: they indicate more signs of non-convergence in conditions with worse performance in terms of bias and confidence-validity.
+Under low to moderate missingness, it does not seem to matter on which \(\theta\) the non-convergence diagnostics are applied.
+If we look at complete unbiasedness, however, we notice that univariate \(\theta\)s fail to diagnose the persistent bias in conditions where \(p_{\rm inc}\geq.75\).
+
+The number of iterations necessary to obtain approximately unbiased, confidence-valid estimates corresponds to the \(\widehat{R}\) threshold 1.2.
+The thresholds 1.1 and 1.01 seem too strict compared to the validity of the inferences.
+The threshold to diagnose significant \(AC\)-values does not appear to be appropriate in at least the current set-up, and perhaps in iterative imputation in general.
+A better heuristic to diagnose non-stationarity with \(AC\) may be through evaluation of the \(AC\)-values across iterations: If the values do not substantially decrease with \(T\), approximate stationarity may be concluded.
 
@@
 
-Valid inferences may be obtained when there is still non-convergence in the algorithm.
+We have shown that iterative imputation algorithms can yield correct outcomes, even when a converged state has not yet formally been reached.
+Any further iterations would then burn computational resources without improving the statistical inferences.
+Our study found that---in the cases considered---inferential validity was achieved after five to ten iterations, much earlier than indicated by the \(\widehat{R}\) and AC diagnostics.
+Of course, it never hurts to iterate longer, but such calculations hardly bring added value.
 
@@
 
-\bibliography{example_paper}
+\bibliography{missing_the_point}
 \bibliographystyle{icml2020}

### 2020-06-04 Writeup/2.Manuscript.Rmd
Fill in ICML format from thesis manuscript
Also change label x axis in plot from missingness proportion to proportion of missing cases, and some other things.

@@
 The solid gray line depicts the targets (unbiased estimates and nominal coverage rate)."}
-mean_bias + sd_bias + Rsq_bias + est_bias + est_cov + est_ciw + plot_layout(ncol = 2, guides = "collect") + plot_annotation(tag_levels = "A", tag_suffix = ".")
+mean_bias + sd_bias + Rsq_bias + est_bias + est_cov + est_ciw + plot_annotation(tag_levels = "A", tag_suffix = ".") + plot_layout(ncol = 2, guides = "collect") 
 ```

### 2020-06-04 ICML/Figures/create_figures.R

@@ -1,0 +1,39 @@
+# create ICML figures, save in size 5.5x8.5in to fit half a page
+# requires the results file 'complete.Rdata'
+
+# load packages
+library(dplyr) #version 0.8.5
+library(ggplot2) #version 3.3.0
+library(patchwork) #version 1.0.0
+
+# set default graphing behavior
+theme_update(
+  plot.title = element_text(hjust = 0.5),
+  plot.subtitle = element_text(hjust = 0.5),
+  panel.border = element_blank(),
+  panel.grid.major = element_blank(),
+  panel.grid.minor = element_blank(),
+  panel.background = element_blank(),
+  axis.line = element_line(colour = "black"),
+  legend.key = element_blank(),
+  legend.position = "bottom",
+  legend.margin = margin(0,0,0,0),
+  text = element_text(size = 7),
+  title = element_text(size = 7),
+  legend.title = element_text(size = 7),
+  axis.text=element_text(size=7),
+  axis.title=element_text(size=7)
+)
+
+# load results and figures
+load("Results/complete.Rdata") 
+results <- results %>% filter(t<51)
+source("Figures/PlotResults.R")
+
+# plot estimates, save 6.75x4.5in or 5.5
+mean_bias + est_bias + est_cov + est_ciw + plot_annotation(tag_levels = "A", tag_suffix = ".") + plot_layout(ncol = 1, guides = "collect") 
+
+
+# plot diagnostics, save 6.75x6.5in or 8.0
+mean_Rh + mean_AC + est_Rh + est_AC + PCA_Rh + PCA_AC + plot_layout(guides = "collect", ncol = 2) + plot_annotation(tag_levels = "A", tag_suffix = ".")
+

### 2020-06-08 ICML/ManuscriptWithCommentsJP.Rmd
Comments on thesis manuscript

--- 2020-05-11 Writeup/2.Manuscript.Rmd
@@
 
+**Comments Gerko: "The thesis by Hanne covers algorithmic convergence in iterative imputation; a topic that has not received much attention but is ever so important as most imputation software packages draw inference from iterative imputation procedures.
+The validity of this inference all depends on the state space of the algorithm at the final iteration.
+Hanne examined the relevant streams of literature; identified the specific components and substantive model dependent and substantive model-independent evaluation measures that are necessary to identify non-convergence and assess the validity of obtained inferences and was able to simulate and investigate the plausibility of identifiers for non-convergence for iterative imputation algorithms.
+Her work results in a guide- line for practice.
+...
+There is clear problem, existing methodology, a new proposed theta for identifying non-convergence, and the simulation approach results in a clear investigation into the convergence properties of iterative imputation algorithms.
+Hanne proposes guidelines for practice and identifies pitfalls that arise with contemporary evaluation functions and details how to avoid such problems.
+Her work results in a thorough insight into the convergence properties of iterative imputation algorithms."**
+
 Anyone who analyzes person-data may run into a missing data problem.
@@
 And can we rely on the inference obtained on the completed data?
+**Comment jp: "it is not stressed that convergence issues apply to/stem from the Markov chain(s) that is produced by the algorithm."**
 
@@
 Several non-convergence diagnostics exist, but it is not known whether these are appropriate within the imputation framework.
-<!-- "a weak diagnostics is better than no diagnostic at all" [@cowl96].
--->
+**Comment JP: difference from Schafer 2002: "From a theoretical standpoint, this technique is problematic, because the sequence of regression models might not be consistent with a true joint distribution.
+Technically speaking, these iterative algorithms may never “converge” because the joint distribution to which they may converge does not exist.
+Nevertheless, simulation work (Brand, 1999) suggests that in some practical applications the method can indeed work well despite the theoretical problems" (p.
+169).
+Also, from Schafer and Olsen (p.
+556): "There is no danger in using too large a value of k, because once DA has converged the process remains stationary.
+On the other hand, ifk is too small, the imputed datasets will not be truly independent and the variability among them may understate the true levels of missing-data uncertainty.
+Time and computational resources allowing, it is therefore advisable to choose k to be somewhat larger than necessary, to create a margin of safety and ensure that the multiple imputations are truly independent." **
+
+<!-- https://www-tandfonline-com.proxy.library.uu.nl/doi/pdf/10.1207/s15327906mbr3304_5?needAccess=true -->
+
+<!-- "a weak diagnostics is better than no diagnostic at all" [@cowl96]. -->
 
@@
 Currently, the recommended practice for evaluating the convergence of the MICE algorithm is through visual inspection.
-After running the imputation algorithm for a certain number of iterations, researchers are encouraged to produce traceplots.
+**Comment JP: not possible to identify visually the point in an MCMC chain where convergence has been reached."** After running the imputation algorithm for a certain number of iterations, researchers are encouraged to produce traceplots.
 In a traceplot, a scalar summary of the state-space of the algorithm $\theta$ is plotted against the iteration number, as depicted in Figure \ref{fig:non-conv}.
@@
 By definition, $\lambda_{1}$-values are equal to the variance of the first component that would be obtained by performing principal component analysis (PCA) on the completed data.
-As scalar summary, $\lambda_{1}$ has the appealing property that it is not dependent on the model of scientific interest, yet still summarizes the multivariate state-space of the algorithm.
+As scalar summary, $\lambda_{1}$ has the appealing property that it is not dependent on the model of scientific interest, yet still summarizes the multivariate state-space of the algorithm.**comment JP: "It is unclear if the lambda_1 is meant for the algorithm in total or is it possible to monitor convergence of single observations or only for the multivariate set of observations?"**
 
@@
 assume that values within chains represent independent samples, whereas the MICE algorithm only uses the final iteration to produce imputations.
+**Comment JP: The main point is usually to identify the area where it can be assumed that the chain has converged, it is usually not the question if the chain converges or not."**
 
@@
 $\widehat{R}$ and $AC$ should identify non-convergence in simulation conditions, where $\bar{Q}$ is *not* an unbiased, confidence-valid estimate of $Q$.
+**Comment JP: "It is not clear if the object is to do more iterations if the diagnostics has not reached convergence."**
 

### 2020-06-08 ICML/ARTEMISS_proposal.tex
Remove redundant info in proposal ICML (it doesn't need to be a full summary of my thesis).

--- 2020-06-04 ICML/ARTEMISS_proposal.tex
@@
 % Use the following line for the initial blind version submitted for review:
-\usepackage{icml2020}
+\usepackage[accepted]{icml2020}
 
@@
 \icmlauthor{Hanne I.~Oberman}{UU}
-\icmlauthor{Stef van~Buuren}{UU}
+\icmlauthor{Stef van~Buuren}{UU,TNO}
 \icmlauthor{Gerko Vink}{UU}
@@
 \icmlaffiliation{UU}{Department of Methodology and Statistics, Utrecht University, Utrecht, The Netherlands}
+\icmlaffiliation{TNO}{Netherlands Organisation for Applied Scientific Research TNO, Leiden, The Netherlands}
 
@@
 
-\section{Introduction}
+\section{Iterative Imputation}
 \label{intro}
 
-A popular method to accommodate missing data is to \textit{impute} (i.e., `fill in') the missing values in an incomplete dataset.
-It is widely accepted that imputation techniques such as multiple imputation (MI; @rubin76) can yield statistically valid inferences.
-The validity of these inferences relies on the method through which imputations are obtained---often iterative algorithms.
-Iterative imputation algorithms are employed in e.g., $\mathtt{SPSS}$, $\mathtt{Stata}$, and the $\mathtt{R}$ packages $\mathtt{MI}$, and $\mathtt{mice}$.
-\textbf{Or use:} Convergence of the algorithm used to solve the missing data problem is a topic that has not received much attention but is ever so important, as most imputation software packages draw inference from iterative imputation procedures.
+Most imputation software packages draw inference from iterative imputation procedures.
+With iterative imputation, the validity of the inference depends on the state-space of the algorithm at the final iteration.
+This introduces a potential threat to the validity of the imputations: What if the algorithm has not converged?
+Are the imputations then to be trusted?
+And can we rely on the inference obtained on the completed data?
 
-The validity of the inference all depends on the state space of the algorithm at the final iteration.
-If the algorithm does not reach convergence, ``the imputed datasets will not be truly independent and the variability among them may understate the true levels of missing-data uncertainty'' (Schafer and Olsen, p.
-556).
+These remain open questions since the convergence properties of iterative imputation algorithms have not been systematically studied \citep[\(\S\) 6.5.2]{buur18}.
+There is no scientific consensus on how to evaluate the convergence of imputation algorithms \citep{ragh07, zhu15, taka17}.
+The current recommended practice is to visually inspect imputations for signs of non-convergence.
+This approach may be undesirable for several reasons: 1) it may be challenging to the untrained eye, 2) only severely pathological cases of non-convergence may be diagnosed, and 3) there is not an objective measure that quantifies convergence \citep[\(\S\) 6.5.2]{buur18}.
+Therefore, a quantitative, diagnostic method to identify non-convergence would be preferred.
 
-But when is this the case?
-Current practice of visually inspecting imputations is not sufficient because 1) no objective point at which convergence is established, 2) only severely pathological cases of non-convergence may be identified, and 3) diagnosing non-convergence can be challenging to the untrained eye.
-On top of that, the statistics to inspect are either univariate or dependent on the substantive model of interest.
-Converged univariate statistics do not guarantee multivariate convergence, and with a model-dependent statistic there is no guarantee that the algorithm is converged \textit{enough} for other models of scientific interest.
-This negates the advantages of MI: solving the missing data problem and complete data problem separately.
+\section{Identifying Non-Convergence}
+\label{methods}
 
-Therefore, we propose a novel parameter to inspect. We use this parameter to identify non-convergence with two quantitative diagnostic methods: Rhat and AC.
+In our study, we consider two identifiers for non-convergence in iterative algorithms: autocorrelation and potential scale reduction factor $\widehat{R}$ \citep[as~recommended~by~e.g.][]{cowl96}.
+We follow \citep[p.~147]{lync07} to calculate autocorrelation, and use the recently proposed adapted version of $\widehat{R}$ by \citep[p.~5]{veht19}.
+Aside from the usual parameters to monitor through visual inspection (i.e., chain means and chain variances), we also look at multivariate parameters to summarize the state-space of the algorithm.
+We investigate convergence of the parameter of scientific interest, and propose a novel parameter: the first eigenvalue of the variance-covariance matrix after imputation.
+This novel parameter has the appealing quality that it is not dependent on the model of scientific interest.
+With that, it suits one of the main advantages of imputation techniques---solving the missing data problem and the substantive scientific problem separately.
 
-We simulate and investigate the plausibility of identifiers for non-convergence for iterative imputation algorithms ...
+% Moreover, the behavior of such algorithms under certain default imputation models (e.g., `predictive mean matching') is an entirely open question \citep{murr18}.
+
+% \footnote{If the algorithm does not reach convergence, ``the imputed datasets will not be truly independent and the variability among them may understate the true levels of missing-data uncertainty'' (Schafer and Olsen, p.
+556).}
+
+% 
+% 
+% Therefore, we propose a novel parameter to inspect.
+We use this parameter to identify non-convergence with two quantitative diagnostic methods.
+We use model-based simulation to investigate the plausibility of identifiers for non-convergence for iterative imputation algorithms and provide guidelines for empirical research.
+For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular \texttt{mice} package in \texttt{R} \citep{mice, R}.
+\textit{[Provide notation here?
+E.g., the number of imputations, iterations, etc.]}
+
+We evaluate the performance and plausibility of these methods to identify non-convergence through model-based simulation in $\mathtt{R}$ \cite{R}.
+For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular $\mathtt{mice}$ package in $\mathtt{R}$ \citep{mice}.
+
+% \section{Methods}
+% \label{methods}
+% 
+% \subsection{Diagnostic Methods}
+% We consider several sets of diagnostic methods to identify non-convergence in the iterative imputation algorithm.
+Each method is a combination of a parameter that summarizes the state-space of the algorithm, $\theta$, and an identifier.
+As $\theta$s we use the default parameters of the \texttt{MICE} algorithm (i.e., the univariate $\theta$s chain means and chain variances), a model-dependent user-specified multivariate parameter $Q$, and a novel parameter: $\lambda_1$.
+$\lambda_{1}$ has the appealing property that it is not dependent on the model of scientific interest, yet still summarizes the multivariate state-space of the algorithm.
+% 
+% We define $\lambda_{1}$ as the first eigenvalue of the variance-covariance matrix of the completed data.\footnote{By definition, this is equal to the variance of the first component of the principal component analylis (PCA) solution on the same data.} For each imputation $\ell$, let $\lambda_{1, \ell} \geq \lambda_{2, \ell} \geq ...
+\geq \lambda_{j, \ell}$ be the eigenvalues of the variance-covariance matrix $S_{\ell}$, where $\ell = 1, ..., m$.
+Since the eigenvalues of a variance-covariance matrix summarize the total set of covariances in the data, $\lambda_1$ captures the multivariate nature of the completed data without imposing a substantive model of interest.
+% 
+% The identifiers are as recommended by e.g.
+\cite{cowl96}, namely the potential scale reduction factor $\widehat{R}$ and autocorrelation.
+$\widehat{R}$ is also known as the `Gelman-Rubin statistic' after the original paper \cite{gelm92}.
+We use a recently proposed adapted version of $\widehat{R}$ \citep{veht19}.
+We diagnose non-convergence if $\widehat{R}$-values exceed a certain diagnostic threshold.
+The thresholds under consideration are the traditional threshold $\widehat{R}=1.2$ \citep{gelm92}, the common $\widehat{R}=1.1$ \citep{gelm13}, and the recent $\widehat{R}=1.01$ \citep{veht19}.
+% 
+% We diagnose non-convergence with autocorrelation when the correlation between two subsequent $\theta$-values within the same chain is significantly different from zero \citep[p.~147, p.~42]{lync07, hynd18}.
+
+\section{Simulation Study}
+
+We induce non-convergence in the imputation algorithm using two sets of simulation conditions: early stopping and  missingness severity.
+Early stopping means that we terminate the imputation algorithm after a different number of iterations in each condition ($T = 1, 2, ..., 100$).
+The severity of the missingness is determined by the proportion of missing cases ($p_{\rm inc} = .05, .25, .50, .75, .95$)\footnote{We only consider a `missing completely at random' missingness mechanism \citep{rubin76}.}.
+For each simulation condition, we evaluate the validity of several statistical inferences, where inferential validity is defined as unbiased estimates and nominal coverage rates across simulation repetitions ($n_{\rm sim}=1000$).
+
+The simulation set-up is summarized in pseudo-code (see Algorithm \ref{alg:setup}).
+The complete script and technical details are available from \href{https://github.com/hanneoberman/MissingThePoint}{github.com/hanneoberman/MissingThePoint}.
+
+\begin{algorithm}
+   \caption{Simulation set-up}
+   \label{alg:setup}
+\begin{algorithmic}
+\STATE Simulate data 
+\REPEAT 
+% \FOR {$p_{\rm inc} = .05, .25, .50, .75, .95$ (missingness conditions)}
+\FOR {all missingness conditions}
+  \STATE Create missingness
+  % \FOR {$T = 1, 2, ..., 100$ (early stopping conditions)}
+  \FOR {all early stopping conditions}
+   \STATE Impute missingness
+   \STATE Perform analysis of scientific interest
+   \STATE Compute non-convergence diagnostics 
+   \STATE Pool results across imputations
+   \STATE Compute performance measures
+   \ENDFOR
+ \ENDFOR 
+ \STATE Combine outcomes of all conditions
+\UNTIL{all simulation repetitions are completed}
+\STATE Aggregate outcomes across simulation runs
+\end{algorithmic}
+\end{algorithm}
+
+% The aim of the simulation study is twofold: 1) to determine the effect of early stopping and missingness conditions on the validity of statistical inferences, 2) to investigate the plausibility of identifiers for non-convergence.
+% 
+% We assess the validity of statistical inferences with bias, coverage rate and confidence interval width as performance measures (see @buur18 for definitions).
+And we evaluate the plausibility of the identifiers using the diagnostic thresholds for $\widehat{R}$ and autocorrelation (i.e., $\widehat{R}=1.2,1.1,1.01$ and the critical value for AC = 0).
+% 
+% \textit{[Very incomplete from here!]} We consider several quantities of scientific interest.
+% 
+% The statistical inference of interest is multivariate linear regression.
+% 
+% The data generating mechanism of the predictor space is a multivariate normal distribution
+%
+
+\section{Results}
+\label{results}
+
+Our results demonstrate that inferential validity was achieved after five to ten iterations---much earlier than indicated by the diagnostic methods. For example,
+
+\begin{figure*}
+{\centering \includegraphics{Figures/estimates} 
+}
+\caption{Impact of non-convergence on statistical inferences.
+Depicted are the bias, coverage rate (CR) and confidence interval width (CIW) of the worst-performing quantities of scientific interest $Q$ in terms of bias.
+The gray lines represent the objectives: unbiased estimates with nominal coverage.}\label{fig:est}
+\end{figure*}
+
+% Results are displayed in Figure \ref{fig:est} and \ref{fig:diag}.
+Simulation conditions with a high proportion of missing cases and a low number of iterations most often result in biased estimates and non-nominal coverages.
+% 
+% \textit{[Copy-paste from here!]} These results demonstrate that the estimates of univariate scientific estimands \(Q\) are not impacted by early stopping of the MICE algorithm or the proportion of incomplete cases in \(y\).
+Unbiased estimates may be obtained after just one iteration.
+Multivariate estimates, by contrast, are affected by both the number of iterations and the proportion of missing cases.
+Completely unbiased estimates are only obtained under low to moderate missingness (\(p_{\rm inc}\leq.50\)), after at most three iterations.
+We observe approximately unbiased estimates after at most seven iterations (for any \(p_{\rm inc}\) considered).
+This implies that the algorithm produces stable, non-improving estimates when \(T\geq7\).
+
+\begin{figure*}
+{\centering \includegraphics{Figures/diagnostics} 
+}
+\caption{Non-convergence identified by diagnostic methods: $\widehat{R}$ and $AC$ applied on several $\theta$s.
+The left-hand side of the figure contains $\widehat{R}$-values, and the right-hand side contains $AC$-values.
+Depicted in the rows are the scalar summaries $\theta$: chain mean, chain variance, the quantity of scientific interest $\hat{Q}$, and the first eigenvalue of the variance-covariance matrix $\lambda_1$.}\label{fig:diag}
+\end{figure*}
+
+% Overall, the methods to diagnose non-convergence perform as expected: they indicate more signs of non-convergence in conditions with worse performance in terms of bias and confidence-validity.
+Under low to moderate missingness, it does not seem to matter on which \(\theta\) the non-convergence diagnostics are applied.
+If we look at complete unbiasedness, however, we notice that univariate \(\theta\)s fail to diagnose the persistent bias in conditions where \(p_{\rm inc}\geq.75\).
+% 
+% The number of iterations necessary to obtain approximately unbiased, confidence-valid estimates corresponds to the \(\widehat{R}\) threshold 1.2.
+The thresholds 1.1 and 1.01 seem too strict compared to the validity of the inferences.
+The threshold to diagnose significant \(AC\)-values does not appear to be appropriate in at least the current set-up, and perhaps in iterative imputation in general.
+A better heuristic to diagnose non-stationarity with \(AC\) may be through evaluation of the \(AC\)-values across iterations: If the values do not substantially decrease with \(T\), approximate stationarity may be concluded.
+%
 

### 2020-06-08 ICML/ARTEMISS_proposal.tex
ICML proposal with less info (removed excess figures, summarized results). Not yet with description of new figure.

@@
 
-Most imputation software packages draw inference from iterative imputation procedures.
+To draw inference from incomplete data, most imputation software packages use iterative imputation procedures.
 With iterative imputation, the validity of the inference depends on the state-space of the algorithm at the final iteration.
@@
 These remain open questions since the convergence properties of iterative imputation algorithms have not been systematically studied \citep[\(\S\) 6.5.2]{buur18}.
-There is no scientific consensus on how to evaluate the convergence of imputation algorithms \citep{ragh07, zhu15, taka17}.
-The current recommended practice is to visually inspect imputations for signs of non-convergence.
+While there is no scientific consensus on how to evaluate the convergence of imputation algorithms \citep{zhu15, taka17}, the current practice is to visually inspect imputations for signs of non-convergence.
 This approach may be undesirable for several reasons: 1) it may be challenging to the untrained eye, 2) only severely pathological cases of non-convergence may be diagnosed, and 3) there is not an objective measure that quantifies convergence \citep[\(\S\) 6.5.2]{buur18}.
@@
 
-In our study, we consider two identifiers for non-convergence in iterative algorithms: autocorrelation and potential scale reduction factor $\widehat{R}$ \citep[as~recommended~by~e.g.][]{cowl96}.
-We follow \citep[p.~147]{lync07} to calculate autocorrelation, and use the recently proposed adapted version of $\widehat{R}$ by \citep[p.~5]{veht19}.
-Aside from the usual parameters to monitor through visual inspection (i.e., chain means and chain variances), we also look at multivariate parameters to summarize the state-space of the algorithm.
-We investigate convergence of the parameter of scientific interest, and propose a novel parameter: the first eigenvalue of the variance-covariance matrix after imputation.
+In our study, we consider two non-convergence identifiers:\footnote{\citep[as recommended~by~e.g.][]{cowl96}} autocorrelation \citep[conform][p.~147]{lync07} and potential scale reduction factor $\widehat{R}$ \citep[conform][p.~5]{veht19}.
+Aside from the usual parameters to monitor---chain means and chain variances---we also investigate convergence of the multivariate state-space of the algorithm.
+
+We implement \cite{buur18}'s recommendation to track the parameter of scientific interest, and henceforth propose a novel parameter: the first eigenvalue of the variance-covariance matrix that is obtained after imputing the missing data.
 This novel parameter has the appealing quality that it is not dependent on the model of scientific interest.
@@
 
-We evaluate the performance and plausibility of these methods to identify non-convergence through model-based simulation in $\mathtt{R}$ \cite{R}.
+We evaluate the performance and plausibility of these diagnostic methods through model-based simulation in $\mathtt{R}$ \cite{R}.
 For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular $\mathtt{mice}$ package in $\mathtt{R}$ \citep{mice}.
@@
 
-We induce non-convergence in the imputation algorithm using two sets of simulation conditions: early stopping and  missingness severity.
-Early stopping means that we terminate the imputation algorithm after a different number of iterations in each condition ($T = 1, 2, ..., 100$).
-The severity of the missingness is determined by the proportion of missing cases ($p_{\rm inc} = .05, .25, .50, .75, .95$)\footnote{We only consider a `missing completely at random' missingness mechanism \citep{rubin76}.}.
-For each simulation condition, we evaluate the validity of several statistical inferences, where inferential validity is defined as unbiased estimates and nominal coverage rates across simulation repetitions ($n_{\rm sim}=1000$).
+The aim of the simulation study is to determine the impact of non-convergence on the validity of statistical inferences, and to assess whether non-convergence may be detected using several diagnostic methods.
+Inferential validity is reached when estimates are both unbiased and have nominal coverage across simulation repetitions ($n_{\rm sim}=1000$).
 
+We induce non-convergence in the imputation algorithm with two sets of simulation conditions: early stopping and missingness severity.
+Early stopping implies that we vary the number of iterations before terminating the imputation algorithm (between 1 and 100 iterations).
+And the missingness severity is determined by the proportion of incomplete cases (between 5 and 95\%)\footnote{We only consider a `missing completely at random' missingness mechanism \citep{rubin76}.}.
 The simulation set-up is summarized in pseudo-code (see Algorithm \ref{alg:setup}).
@@
 
-Our results demonstrate that inferential validity was achieved after five to ten iterations---much earlier than indicated by the diagnostic methods. For example,
+Our results demonstrate that inferential validity is achieved after five to ten iterations---even in the worst missingness conditions (i.e., up-to 95\% of cases having missing values).
+Conditions with a lower proportion of incomplete cases yield valid inferences almost instantly (after just two iterations).
+This is in grave contrast to the non-convergence identified by the diagnostic methods.
+Autocorrelation- and $\widehat{R}$-values still indicate improving convergence after 20 to 30 iterations.
 
 \begin{figure*}
-{\centering \includegraphics{Figures/estimates} 
+{\centering \includegraphics{Figures/resultssquare} 
 }
-\caption{Impact of non-convergence on statistical inferences.
-Depicted are the bias, coverage rate (CR) and confidence interval width (CIW) of the worst-performing quantities of scientific interest $Q$ in terms of bias.
-The gray lines represent the objectives: unbiased estimates with nominal coverage.}\label{fig:est}
+\caption{Subset of simulation results (truncated at 50 iterations).
+Depicted are performance measures bias and coverage rate, and non-convergence identifiers autocorrelation and $\widehat{R}$ for the same scientific estimate.
+The solid gray lines represent inferential validity of the estimate (i.e., unbiasedness and nominal coverage), whereas the dashed gray lines depict the common diagnostic thresholds of the identifiers.}\label{fig:res}
 \end{figure*}
+
+For example, in Figure \ref{fig:res} we see that...
+The novel parameter has equal performance to the quantity of scientific interest as parameter ($r=.975$), while having the advantage of being model-independent.
 
@@
 
-\begin{figure*}
-{\centering \includegraphics{Figures/diagnostics} 
-}
-\caption{Non-convergence identified by diagnostic methods: $\widehat{R}$ and $AC$ applied on several $\theta$s.
+% \begin{figure*}
+% {\centering \includegraphics{Figures/Rplot} 
+% }
+% \caption{Non-convergence identified by diagnostic methods: $\widehat{R}$ and $AC$ applied on several $\theta$s.
 The left-hand side of the figure contains $\widehat{R}$-values, and the right-hand side contains $AC$-values.
 Depicted in the rows are the scalar summaries $\theta$: chain mean, chain variance, the quantity of scientific interest $\hat{Q}$, and the first eigenvalue of the variance-covariance matrix $\lambda_1$.}\label{fig:diag}
-\end{figure*}
+% \end{figure*}
+%
 
@@
 
-We have shown that iterative imputation algorithms can yield correct outcomes, even when a converged state has not yet formally been reached.
+With this study, we show that iterative imputation algorithms can yield correct outcomes, even when a converged state has not yet formally been reached.
 Any further iterations would then burn computational resources without improving the statistical inferences.
-Our study found that---in the cases considered---inferential validity was achieved after five to ten iterations, much earlier than indicated by the \(\widehat{R}\) and AC diagnostics.
+Our study found that---in the cases considered---inferential validity was achieved after five to ten iterations, much earlier than indicated by the autocorrelation and \(\widehat{R}\) diagnostics.
 Of course, it never hurts to iterate longer, but such calculations hardly bring added value.

### 2020-06-08 ICML/Figures/create_figures.R

@@ -27,13 +27,85 @@ theme_update(
 
 # load results and figures
 load("Results/complete.Rdata") 
+# test correlation between two thetas
+r <- cor(results$r.hat.max.beta, results$r.hat.max.pca, use = "complete.obs")
+
+# pre-processing
 results <- results %>% filter(t<51)
-source("Figures/PlotResults.R")
+results <- results %>% mutate(bias.est.X1 = bias.est.X1/.0206)
+
+# create figures
+# source("Figures/PlotResults.R")
+
+# add diagnostic thresholds
+results <- results %>% mutate(crit = qnorm((1 + .95) / 2) / sqrt(t), thresh1.01 = 1.01, thresh1.1 = 1.1, thresh1.2 = 1.2)
+results$crit[results$crit>1] <- NA
+
+# define colorblind friendly colors
+paint5 <- c('#228833', '#66CCEE', '#CCBB44','#EE6677', '#AA3377')
+
+# create plots
+est_bias <- results %>% ggplot(aes(x = t, y = bias.est.X1, color = as.factor(p*100))) +
+  geom_hline(yintercept = 0,
+             color = "grey",
+             lwd = 1) +
+  geom_point(size = .25, na.rm = TRUE) +
+  geom_line(size = .25, na.rm = TRUE) +
+  scale_colour_manual(values=paint5) +
+  xlab("") +
+  ylab("Bias (%)") +
+  # ylab(bquote("Bias (Q = " ~ beta[1] ~ "= 2.06)")) +
+  labs(colour = "Proportion of missing cases (%)")
+
+est_Rh <- results %>% ggplot(aes(x = t, y = r.hat.max.beta, color = as.factor(p*100))) +
+  geom_hline(yintercept = 1,
+             color = "grey",
+             lwd = 1) +
+  geom_point(size = .25, na.rm = TRUE) +
+  geom_line(size = .25, na.rm = TRUE) +
+  geom_line(aes(x=t, y=thresh1.2), color = "grey", linetype = "dashed", size = .25, na.rm = TRUE) +
+  geom_line(aes(x=t, y=thresh1.1), color = "grey", linetype = "dashed", size = .25, na.rm = TRUE) +
+  geom_line(aes(x=t, y=thresh1.01), color = "grey", linetype = "dashed", size = .25, na.rm = TRUE) +
+  scale_colour_manual(values=paint5) +
[…]

### 2020-06-08 ICML/ARTEMISS_proposal.tex
Version for feedback Gerko.

@@
 Are the imputations then to be trusted?
-And can we rely on the inference obtained on the completed data?
+And can we rely on the inference obtained using the imputed data?
 
@@
 
-Our results demonstrate that inferential validity is achieved after five to ten iterations---even in the worst missingness conditions (i.e., up-to 95\% of cases having missing values).
-Conditions with a lower proportion of incomplete cases yield valid inferences almost instantly (after just two iterations).
+Our results demonstrate that inferential validity is achieved after five to ten iterations, even under severe missingness conditions (i.e., up-to 95\% of cases having missing values).
+Conditions with a lower proportion of incomplete cases yield valid inferences almost instantly---after just two iterations.
 This is in grave contrast to the non-convergence identified by the diagnostic methods.
@@
 
-For example, in Figure \ref{fig:res} we see that...
-The novel parameter has equal performance to the quantity of scientific interest as parameter ($r=.975$), while having the advantage of being model-independent.
+For example, in Figure \ref{fig:res} we see...
+
+The novel parameter has equal performance to the quantity of scientific interest as parameter (Spearman $r=.978$), while having the advantage of being model-independent.
 

### 2020-06-09 ICML/ARTEMISS_proposal.tex
Saving two adapted versions of my simulation (repo is a mess because I left everything in the main folder)

@@
 
+\setcounter{footnote}{2}
+
 \section{Iterative Imputation}
@@
 
-In our study, we consider two non-convergence identifiers:\footnote{\citep[as recommended~by~e.g.][]{cowl96}} autocorrelation \citep[conform][p.~147]{lync07} and potential scale reduction factor $\widehat{R}$ \citep[conform][p.~5]{veht19}.
-Aside from the usual parameters to monitor---chain means and chain variances---we also investigate convergence of the multivariate state-space of the algorithm.
+In our study, we consider two non-convergence identifiers: autocorrelation \citep[conform][p.~147]{lync07} and potential scale reduction factor $\widehat{R}$ \citep[conform][p.~5]{veht19}.\footnote{As recommended by e.g.
+\citet{cowl96}.} Aside from the usual parameters to monitor---chain means and chain variances \citep[\(\S\) 4.5.6]{buur18}---we also investigate convergence in the multivariate state-space of the algorithm.
 
-We implement \cite{buur18}'s recommendation to track the parameter of scientific interest, and henceforth propose a novel parameter: the first eigenvalue of the variance-covariance matrix that is obtained after imputing the missing data.
+We implement \citet[\(\S\) 6.5.1]{buur18}'s recommendation to track the parameter of scientific interest, and henceforth propose a novel parameter: the first eigenvalue of the variance-covariance matrix that is obtained after imputing the missing data.
 This novel parameter has the appealing quality that it is not dependent on the model of scientific interest.
@@
 
-We evaluate the performance and plausibility of these diagnostic methods through model-based simulation in $\mathtt{R}$ \cite{R}.
+We evaluate the performance and plausibility of these diagnostic methods through model-based simulation in $\mathtt{R}$ \citep{R}.
 For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular $\mathtt{mice}$ package in $\mathtt{R}$ \citep{mice}.
@@
 Early stopping implies that we vary the number of iterations before terminating the imputation algorithm (between 1 and 100 iterations).
-And the missingness severity is determined by the proportion of incomplete cases (between 5 and 95\%)\footnote{We only consider a `missing completely at random' missingness mechanism \citep{rubin76}.}.
-The simulation set-up is summarized in pseudo-code (see Algorithm \ref{alg:setup}).
-The complete script and technical details are available from \href{https://github.com/hanneoberman/MissingThePoint}{github.com/hanneoberman/MissingThePoint}.
+The missingness severity is determined\footnote{I.e., we only consider a `missing completely at random' missingness mechanism \citep{rubin76}.} by the proportion of incomplete cases (between 5 and 95\%).
+The simulation set-up is summarized in pseudo-code (see Algorithm \ref{alg:setup}), while the complete script and technical details are available from \href{https://github.com/hanneoberman/MissingThePoint}{github.com/hanneoberman/MissingThePoint}.
 
@@
 \begin{figure*}
-{\centering \includegraphics{Figures/resultssquare} 
+{\centering \includegraphics{Figures/results} 
 }
 \caption{Subset of simulation results (truncated at 50 iterations).
-Depicted are performance measures bias and coverage rate, and non-convergence identifiers autocorrelation and $\widehat{R}$ for the same scientific estimate.
-The solid gray lines represent inferential validity of the estimate (i.e., unbiasedness and nominal coverage), whereas the dashed gray lines depict the common diagnostic thresholds of the identifiers.}\label{fig:res}
+Depicted are performance measures percentage bias and coverage rate, and non-convergence identifiers autocorrelation and $\widehat{R}$ for a regression estimate.
+The solid gray lines represent inferential validity of the estimate (i.e., unbiasedness and nominal coverage), whereas the dashed gray lines depict common diagnostic thresholds of the identifiers.}\label{fig:res}
 \end{figure*}
 
-For example, in Figure \ref{fig:res} we see...
-
-The novel parameter has equal performance to the quantity of scientific interest as parameter (Spearman $r=.978$), while having the advantage of being model-independent.
+For example, in Figure \ref{fig:res} we show some results for a regression estimate.
+Depicted are percentage bias, coverage rate,  autocorrelation and $\widehat{R}$ of the estimate.
+From this, we see that within a few iterations the percentage bias approaches zero and nominal coverage is quickly reached---even when autocorrelation and $\widehat{R}$ would still signal non-convergence.
+This also holds for scenarios with severe missingness.
+Moreover, results seem invariant to the choice of parameter for the non-convergence diagnostics.
+The novel parameter that we propose has equal performance to the scientific estimate presented in Figure \ref{fig:res},\footnote{The Spearman correlation between the autocorrelation values for the two parameters is $r=.982$, and $r=.978$ between the $\widehat{R}$-values.} while having the advantage of being independent from the scientific model of interest.
 
@@
 Any further iterations would then burn computational resources without improving the statistical inferences.
+Preliminary findings suggest that these results also hold for more challenging scenarios, e.g.
+under different missingness mechanisms or imputation models.
+
 Our study found that---in the cases considered---inferential validity was achieved after five to ten iterations, much earlier than indicated by the autocorrelation and \(\widehat{R}\) diagnostics.

### 2020-06-09 1.Execute.R

@@ -29,8 +29,8 @@ source("Functions/EvaluateResults.R")
 
 # simulation parameters
 populationsize <- 1000 #nr of observations in simulated dataset
-n.iter <- 100 #nr of iterations (varying 1:n.iter)
-n.sim <- 1000 #nr of simulations per iteration value
+n.iter <- 10 #nr of iterations (varying 1:n.iter)
+n.sim <- 100 #nr of simulations per iteration value
 p.miss <- c(.05, .25, .5, .75, .95) #proportion of incomplete cases
 true.effect <- 2 #regression coefficient to be estimated
 
@@ -63,23 +63,23 @@ out <-
 ########################
 
 # summarize raw results
-results_without_CI <- evaluate.sim(sims = out)
+results <- evaluate.sim(sims = out)
 
-# calculate empirical CI
-CI_lower <-
-  evaluate.sim(sims = out, mean_or_SE = "lower")
-CI_upper <-
-  evaluate.sim(sims = out, mean_or_SE = "upper")
-
-# combine with other results
-results <-
-  results_without_CI %>%
-  left_join(CI_lower,
-            by = c("t", "p"),
-            suffix = c("", ".LL")) %>%
-  left_join(CI_upper,
-            by = c("t", "p"),
-            suffix = c("", ".UL"))
+# # calculate empirical CI
+# CI_lower <-
+#   evaluate.sim(sims = out, mean_or_SE = "lower")
+# CI_upper <-
+#   evaluate.sim(sims = out, mean_or_SE = "upper")
+# 
+# # combine with other results
+# results <-
+#   results %>%
+#   left_join(CI_lower,
+#             by = c("t", "p"),
+#             suffix = c("", ".LL")) %>%
+#   left_join(CI_upper,
+#             by = c("t", "p"),
+#             suffix = c("", ".UL"))
 
 # # if necessary, calculate MCMC SEs
 # MCMCSE <- evaluate.sim(sims = out, mean_or_SE = "se")
@@ -87,7 +87,7 @@ results <-
 # results <- left_join(results, MCMCSE, by = c("t", "p"), suffix = c("", ".SE"))
 
 # save results
-save(results, file = "Results/results.Rdata")
+save(results, file = "Results/data_Y_first.Rdata")
 
 # # if necessary, save raw results (too large for Github)
 # save(out, file = "Results/raw.Rdata")

### 2020-06-09 Functions/CreateData.R

@@ -16,8 +16,9 @@ data.simulation <- function(n = populationsize, true_effect = 2) {
   colnames(simdata) <- c("X1", "X2", "X3") #set predictors names
   
   # compute  outcome variable from predictors
-  simdata$Y <-
+  Y <-
     1 + true_effect * simdata$X1 + 0.5 * simdata$X2 - 1 * simdata$X3 + rnorm(n = populationsize, sd = 10)
+  simdata <- cbind(Y, simdata)
   
   # estimate comlpete data parameters Q
   true.effect <<-

### 2020-06-09 ICML/ARTEMISS_proposal.tex
After feedback Stef on ICML proposal

@@
 
-We implement \citet[\(\S\) 6.5.1]{buur18}'s recommendation to track the parameter of scientific interest, and henceforth propose a novel parameter: the first eigenvalue of the variance-covariance matrix that is obtained after imputing the missing data.
-This novel parameter has the appealing quality that it is not dependent on the model of scientific interest.
+We implement \citet[\(\S\) 6.5.1]{buur18}'s recommendation to track the parameter of scientific interest, and subsequently propose a novel parameter: the first eigenvalue of the variance-covariance matrix that is obtained after imputing the missing data.
+Compared to monitoring the estimate of scientific interest, this novel parameter has the appealing quality that it is not dependent on the complete-data inference.
 With that, it suits one of the main advantages of imputation techniques---solving the missing data problem and the substantive scientific problem separately.
@@
 
-Our results demonstrate that inferential validity is achieved after five to ten iterations, even under severe missingness conditions (i.e., up-to 95\% of cases having missing values).
+Our results indicate that inferential validity is achieved after five to ten iterations, even under severe missingness conditions (i.e., up-to 95\% of cases having missing values).
 Conditions with a lower proportion of incomplete cases yield valid inferences almost instantly---after just two iterations.

### 2020-06-17 ICML/ARTEMISS_proposal_Oberman_vanBuuren_Vink.tex

--- 2020-06-09 ICML/ARTEMISS_proposal.tex
@@
 There is no consensus on how to evaluate the convergence properties of the method.
-This paper provides insight into identifying non-convergence in iterative imputation algorithms.
-Our study found that---in the cases considered---inferential validity was achieved after five to ten iterations, much earlier than indicated by diagnostic methods.
+Our study provides insight into identifying non-convergence in iterative imputation algorithms.
+We found that---in the cases considered---inferential validity was achieved after five to ten iterations, much earlier than indicated by diagnostic methods.
 We conclude that it never hurts to iterate longer, but such calculations hardly bring added value.
@@
 
-We induce non-convergence in the imputation algorithm with two sets of simulation conditions: early stopping and missingness severity.
+To induce non-convergence in the imputation algorithm we use  two sets of simulation conditions: early stopping and missingness severity.
 Early stopping implies that we vary the number of iterations before terminating the imputation algorithm (between 1 and 100 iterations).
 The missingness severity is determined\footnote{I.e., we only consider a `missing completely at random' missingness mechanism \citep{rubin76}.} by the proportion of incomplete cases (between 5 and 95\%).
-The simulation set-up is summarized in pseudo-code (see Algorithm \ref{alg:setup}), while the complete script and technical details are available from \href{https://github.com/hanneoberman/MissingThePoint}{github.com/hanneoberman/MissingThePoint}.
+We provide a summary of the simulation set-up in Algorithm \ref{alg:setup}, whereas the complete script and technical details are available from \href{https://github.com/hanneoberman/MissingThePoint}{github.com/hanneoberman/MissingThePoint}.
 

### 2020-06-17 Functions/ComputeAC.R
Clean-up repo and update readme

@@ -1,43 +1,0 @@
-# extract convergence diagnostic autocorrelation from iteration object (e.g., chain means)
-# requires packages 'purrr' and 'magrittr'/'dplyr', and the functions in 'SupplementAC.R'
-
-# load supplementary file
-source('Functions/SupplementAC.R')
-
-# function for lag 1 AC for one or more imputation chains
-ac_adapted <- function(x, ac_function = "ac") {
-  # input: object with theta values (rows are iterations, columns are imputations),
-  # and preferred way of calculating ac values
-  # output: autocorrelations at each iteration in x, computed per imputation chain
-  # and with max and mean across imputations
-  
-  # set function to apply
-  if (ac_function == "acf") {
-    function_to_apply <- acf_lag1
-    min_it <- 2
-  } else {
-    function_to_apply <- ac_lag1
-    min_it <- 3
-  }
-  
-  # make input ready for mapping
-  if (is.numeric(x)) {
-    x <- as.data.frame(x)
-  }
-  t <- dim(x)[1] #nr of iterations
-  M <- dim(x)[2] #nr of imputations
-  
-  # compute ac for each iteration (by definition only possible for t > 2)
-  if (t < min_it) {
-    ac <-
-      matrix(NA, t, M + 2) %>% as.data.frame %>% set_names(paste0(ac_function, ".", c("mean", "max", names(x))))
-  } else {
-    ac <- map_dfr(min_it:t, function(it) {
-      # compute ac
-      x[1:it, ] %>% function_to_apply() %>% t() %>% data.frame(mean(.), max(.), .) %>% set_names(paste0(ac_function, ".", c("mean", "max", 1:M)))
-    }) %>% rbind(matrix(NA, min_it-1, M+2, dimnames = list(NULL, names(.))), .) %>% cbind(iteration = 1:t, .)
-  }
-  
-  # output
-  return(ac)
-}

### 2020-06-17 Functions/ComputeDiagnostics.R

@@ -1,16 +1,0 @@
-# function to compute Rhat conform Vehtari et al (2019) and calculate AC manually
-# requires 'dplyr', and the functions in 'ComputeAC.R', and 'ComputeRhat.R'
-
-# load functions
-source('Functions/ComputeRhat.R')
-source('Functions/ComputeAC.R')
-
-# combine the two diagnostics
-convergence <- function(x, include_acf = FALSE) {
-  # input: object with imputation chain(s), e.g. chain mean, choice of calculating default AC
-  # output: AC values at each iteration, computed manually, if necessary also using stats:acf() 
-  out <-  left_join(rhat_adapted(x), ac_adapted(x), by = "iteration") 
-  if(include_acf){ out <- left_join(out, ac_adapted(x, "acf"), by = "iteration")
-  }
-  return(out)
-}

### 2020-06-17 Functions/SupplementRhat.R

@@ -1,65 +1,0 @@
-# Supplementary functions for Rhat function (see Rhat.R)
-# To get folded, split-half, rank-normalized Rhat
-
-# fold chains for rhat of tails
-fold_sims <- function(sims) {
-  # fold Markov chain for rhat of tails, adapted from rstan
-  abs(sims - median(sims))
-}
-
-# split chains with maxit > 4 to detect trending
-split_chains <- function(sims) {
-  # split Markov chains, adapted from rstan
-  
-  # number of iterations
-  t <- dim(sims)[1]
-  
-  # do not split if result will be chains of length 1
-  if (t < 4)
-    return(sims)
-  else {
-    # split each chain to get 2m chains
-    lower <- 1:floor(t / 2)
-    upper <- ceiling((t / 2) + 1):t
-    splits <- base::cbind(sims[lower, ], sims[upper, ])
-    return(splits)
-  }
-}
-
-# rank-normalize chains because Gelman says so
-z_scale <- function(x) {
-  # rank-normalize Markov chain, copied from rstan
-  t <- length(x)
-  r <- rank(x, ties.method = 'average')
-  z <- qnorm((r - 1 / 2) / t)
-  
-  # output
-  if (!is.null(dim(x))) {
-    # output should have the input dimensions
-    z <- array(z, dim = dim(x), dimnames = dimnames(x))
-  }
-  return(z)
-}
-
-# compute rhat
-get.rhat <- function(sims) {
-  # compute potential scale reduction factor (rhat) for each variable in mids object
-  # equations adapted from Vehtari et al. (2019)
-  
-  # number of iterations
-  t <- length(sims)
-  
-  # between chain variance
-  var_between <-
-    t * var(apply(sims, 2, mean))
-  
-  # within chain variance
-  var_within <- mean(apply(sims, 2, var))
-  
-  # rhat
-  rhat <-
-    sqrt((var_between / var_within + t - 1) / t)
-  
-  # output
-  return(rhat)
-}

### 2020-06-17 Thesis/1.Execute.R

@@ -1,0 +1,93 @@
+# script to perform simulation study of the MSc thesis 'Missing the Point: 
+  # Non-Convergence in Iterative Imputation Algorithms'
+# results in the thesis were obtained on 03-05-2010, using R version 3.6.3, 
+  # on a Windows 10 personal computer (package versions and random seed can 
+  # be found in the script)
+# pseudo-code describing the simulation can be found in the thesis manuscript
+# author: Hanne Oberman
+
+########################
+# set-up environment ###
+########################
+
+# load required packages
+library(mice) #version 3.8.0
+library(miceadds) #version 3.8-9
+library(mvtnorm) #version 1.1-0
+library(data.table) #version 1.12.8
+library(tidyverse) #version 1.3.0
+# # if required, just load the necessary tidyverse packages, instead of all
+# library(dplyr)
+# library(ggplot2)
+# library(purrr)
+# library(magrittr)
+
+# load simulation/evaluation functions
+source("Functions/CreateData.R")
+source("Functions/RunSimulationOnce.R")
+source("Functions/EvaluateResults.R")
+
+# simulation parameters
+populationsize <- 1000 #nr of observations in simulated dataset
+n.iter <- 100 #nr of iterations (varying 1:n.iter)
+n.sim <- 1000 #nr of simulations per iteration value
+p.miss <- c(.05, .25, .5, .75, .95) #proportion of incomplete cases
+true.effect <- 2 #regression coefficient to be estimated
+
+
+########################
+# start simulation study
+########################
+
+# set random seed
+set.seed(1111)
+
+# simulate data once
+data <- data.simulation(n = populationsize, true.effect)
+
+# create patterns to ampute the data with multivariate missingness
+amp.pat <-
+  expand.grid(c(0, 1), c(0, 1), c(0, 1), c(0, 1)) %>% #define all possible combinations of univariate and multivariate missingness
+  .[c(-1,-16), ] #remove the completely (un)observed cases  
+names(amp.pat) <- ampute(data)$patterns %>% names() #obtain correct names of patterns
+
+# run simulation n.sim times
+out <-
+  replicate(n = n.sim,
+            expr = simulate(complete_data = data),
+            simplify = FALSE)
+
+
+########################
+# obtain and save result
+########################
+
+# summarize raw results
+results <- evaluate.sim(sims = out)
+
[…]

### 2020-06-17 Thesis/3.GenerateSupplementaryFigures.R

@@ -1,0 +1,63 @@
+# generate supplementary figures that display complete results
+# (we only display conditions where T<51 in the manuscript)
+# requires the results file 'complete.Rdata'
+
+# load packages
+library(dplyr) #version 0.8.5
+library(ggplot2) #version 3.3.0
+library(patchwork) #version 1.0.0
+
+# set default graphing behavior
+theme_update(
+  plot.title = element_text(hjust = 0.5),
+  plot.subtitle = element_text(hjust = 0.5),
+  panel.border = element_blank(),
+  panel.grid.major = element_blank(),
+  panel.grid.minor = element_blank(),
+  panel.background = element_blank(),
+  axis.line = element_line(colour = "black"),
+  legend.key = element_blank(),
+  legend.position = "bottom",
+  legend.margin = margin(0,0,0,0),
+  text = element_text(size = 7),
+  title = element_text(size = 8),
+  legend.title = element_text(size = 7),
+  axis.text=element_text(size=5),
+  axis.title=element_text(size=5)
+)
+
+# load results and figures
+load("Results/complete.Rdata") 
+source("Figures/PlotResults.R")
+
+
+## ALL EARLY STOPPING CONDITIONS ##
+
+# plot the performance measures
+estimates <- mean_bias + sd_bias + Rsq_bias + est_bias + est_cov + est_ciw + plot_layout(ncol = 2, guides = "collect") + plot_annotation(tag_levels = "A", tag_suffix = ".", title = "Impact of non-convergence on statistical inferences (see thesis manuscript for definitions)")
+ggsave("Figures/ResultsEstimates.png", estimates, width = 8.3, height =  5.8)
+
+# plot the non-convergence diagnostics
+diagnostics <- mean_Rh + mean_AC + var_Rh + var_AC + est_Rh + est_AC + PCA_Rh + PCA_AC + plot_layout(guides = "collect", ncol = 2) + plot_annotation(tag_levels = "A", tag_suffix = ".", title = "Non-convergence identified by diagnostics (see thesis manuscript for definitions)")
+ggsave("Figures/ResultsDiagnostics.png", diagnostics, width = 8.3, height =  5.8)
+
+
+## ALL ESTIMATES OF Q AND THETA ##
+
+# plot performance for all Q=mu
[…]

### 2020-06-17 Writeup/3.CombineTitlePageWithManuscript.R

@@ -1,2 +1,0 @@
-pdftools::pdf_combine(c("./3.Thesis/Writeup/1.TitlePage.pdf", "./3.Thesis/Writeup/2.Manuscript.pdf"), 
-            output = "./3.Thesis/Oberman,H.I.-MSBBSS.pdf")

### 2020-07-13 ICML/MissingThePoint.Rmd
Create presentation for ARTEMISS workshop

---
title: "Missing The Point:"
subtitle: "Non-Convergence in Iterative Imputation Algorithms"
author: "Hanne I. Oberman, Stef van Buuren, Gerko Vink"
institute: "Utrecht University"
date: "17-07-2020" #(updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    css: [default, default-fonts]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

# Algorithmic Non-Convergence

With iterative imputation, the validity of the inferences relies on algorithmic convergence.

```{r non-conv, echo=FALSE,  message=FALSE, warning=FALSE}
# set-up env
library(dplyr)
library(ggplot2)
library(plotly)
load("example_diagnostics.Rdata")

# define colorblind friendly colors
paint2 <- c('#66CCEE', '#EE6677')

# plot chains
trace <- ggplot(diagnostics) +
  geom_point(aes(x = iteration, y = `Chain 1`, color = patho),
             size = .25,
             na.rm = TRUE) +
  geom_point(aes(x = iteration, y = `Chain 2`, color = patho),
             size = .25,
             na.rm = TRUE) +
  geom_point(aes(x = iteration, y = `Chain 3`, color = patho),
             size = .25,
             na.rm = TRUE) +
  geom_point(aes(x = iteration, y = `Chain 4`, color = patho),
             size = .25,
             na.rm = TRUE) +
  geom_point(aes(x = iteration, y = `Chain 5`, color = patho),
             size = .25,
             na.rm = TRUE) +
  geom_line(aes(x = iteration, y = `Chain 1`, color = patho), size = .25, na.rm = TRUE) +
  geom_line(aes(x = iteration, y = `Chain 2`, color = patho), size = .25, na.rm = TRUE) +
  geom_line(aes(x = iteration, y = `Chain 3`, color = patho), size = .25, na.rm = TRUE) +
  geom_line(aes(x = iteration, y = `Chain 4`, color = patho), size = .25, na.rm = TRUE) +
  geom_line(aes(x = iteration, y = `Chain 5`, color = patho), size = .25, na.rm = TRUE) +
  scale_colour_manual(values=paint2) +
  scale_x_continuous(breaks = 1:10) +
  xlab("Iteration") +
  ylab("Chain mean") + #bquote(theta)) +
  labs(color = "") + 
  theme_classic() +
  theme(legend.position = "bottom")

ggplotly(trace)

# alternatively, use the actual data
# run 1.Execute.R until imputing anything, than run RunSimulationOnce.R until after amputing the data
# with the 'amps' object
load("../traceplotdata.Rdata")
set.seed(12)
d75 <- amps[[4]] %>% mice(maxit=20)
plot(d75)
gg.mids(d4, x="Y")[[1]]
```

---

# Results

```{r results, echo=FALSE, fig.width=9.8, message=FALSE, warning=FALSE}
load("complete.Rdata") 
#library(patchwork)

# pre-processing
results <- results %>% filter(t<51) %>% mutate(bias.est.X1 = bias.est.X1/.0206) %>% mutate(crit = qnorm((1 + .95) / 2) / sqrt(t), thresh1.01 = 1.01, thresh1.1 = 1.1, thresh1.2 = 1.2)
results$crit[results$crit>1] <- NA

# define colorblind friendly colors
paint5 <- c('#228833', '#66CCEE', '#CCBB44','#EE6677', '#AA3377')

# create plots
est_bias <- results %>% ggplot(aes(x = t, y = bias.est.X1, color = as.factor(p*100))) +
  geom_hline(yintercept = 0,
             color = "grey",
             lwd = 1) +
  geom_point(size = .25, na.rm = TRUE) +
  geom_line(size = .25, na.rm = TRUE) +
  scale_colour_manual(values=paint5) +
  xlab("") +
  ylab("Bias (%)") +
  labs(colour = "Proportion of missing cases (%)") +
  theme_classic() +
  theme(legend.position = "bottom")

est_Rh <- results %>% ggplot(aes(x = t, y = r.hat.max.beta, color = as.factor(p*100))) +
  geom_hline(yintercept = 1,
             color = "grey",
             lwd = 
               1) +
  geom_point(size = .25, na.rm = TRUE) +
  geom_line(size = .25, na.rm = TRUE) +
  geom_line(aes(x=t, y=thresh1.2), color = "grey", linetype = "dashed", size = .25, na.rm = TRUE) +
  geom_line(aes(x=t, y=thresh1.1), color = "grey", linetype = "dashed", size = .25, na.rm = TRUE) +
  geom_line(aes(x=t, y=thresh1.01), color = "grey", linetype = "dashed", size = .25, na.rm = TRUE) +
  scale_colour_manual(values=paint5) +
  scale_y_continuous(limits = c(1,1.53), breaks = c(1,1.1,1.2,1.3,1.4,1.5)) +
  xlab("Number of iterations") +
  ylab(bquote(widehat(R))) +
  labs(colour = "Proportion of missing cases (%)") +
  theme_classic() +
  theme(legend.position = "")

est_AC <- results %>% ggplot(aes(x = t, y = ac.max.beta, color = as.factor(p*100))) +
  geom_hline(yintercept = 0,
             color = "grey",
             lwd = 1) +
  geom_line(aes(x = t, y = crit), color = "grey", linetype = "dashed", size = .25, na.rm = TRUE) + 
  geom_point(size = .25, na.rm = TRUE) +
  geom_line(size = .25, na.rm = TRUE) +
  scale_colour_manual(values=paint5) +
  xlab("Number of iterations") +
  ylab("Autocorrelation") + 
  labs(colour = "Proportion of missing cases (%)") +
  theme_classic() +
  theme(legend.position = "")

est_cov <- results %>% ggplot(aes(x = t, y = cov.est.X1*100, color = as.factor(p*100))) +
  geom_hline(yintercept = 95,
             color = "grey",
             lwd = 1) +
  geom_point(size = .25, na.rm = TRUE) +
  geom_line(size = .25, na.rm = TRUE) +
  scale_colour_manual(values=paint5) +
  xlab("") +
  ylab("Coverage rate (%)") +
  labs(colour = "Proportion of missing cases (%)") +
  theme_classic() +
  theme(legend.position = "")

# plot estimates, save 6.75x4.5in or 5.5
# est_bias + est_cov + est_AC + est_Rh + plot_annotation(tag_levels = "A", tag_suffix = ".") + plot_layout(ncol = 2, guides = "collect")

ggplotly(est_bias)

```

---

# Thanks!

Slides created via the R package [**xaringan**](https://github.com/yihui/xaringan).

### 2020-07-14 ICML/MissingThePoint.Rmd
Add results figures to ARTEMISS presentation

@@
 
-```{r non-conv, echo=FALSE,  message=FALSE, warning=FALSE}
+```{r non-conv, echo=FALSE, fig.width=10, fig.height = 6.2, message=FALSE, warning=FALSE}
 # set-up env
@@
 # define colorblind friendly colors
-paint2 <- c('#66CCEE', '#EE6677')
+paint5 <- c('#228833', '#66CCEE', '#CCBB44','#EE6677', '#AA3377')
+# paint2 <- c('#66CCEE', '#EE6677')
+# 
+# # plot chain
+# trace <- ggplot(diagnostics) +
+#   geom_point(aes(x = iteration, y = `Chain 1`, color = patho),
+#              size = .5,
+#              na.rm = TRUE) +
+#   geom_point(aes(x = iteration, y = `Chain 2`, color = patho),
+#              size = .5,
+#              na.rm = TRUE) +
+#   geom_point(aes(x = iteration, y = `Chain 3`, color = patho),
+#              size = .5,
+#              na.rm = TRUE) +
+#   geom_point(aes(x = iteration, y = `Chain 4`, color = patho),
+#              size = .5,
+#              na.rm = TRUE) +
+#   geom_point(aes(x = iteration, y = `Chain 5`, color = patho),
+#              size = .5,
+#              na.rm = TRUE) +
+#   geom_line(aes(x = iteration, y = `Chain 2`, color = patho), size = .5, na.rm = TRUE) +
+#   geom_line(aes(x = iteration, y = `Chain 3`, color = patho), size = .5, na.rm = TRUE) +
+#   geom_line(aes(x = iteration, y = `Chain 1`, color = patho), size = .5, na.rm = TRUE) +
+#   geom_line(aes(x = iteration, y = `Chain 4`, color = patho), size = .5, na.rm = TRUE) +
+#   geom_line(aes(x = iteration, y = `Chain 5`, color = patho), size = .5, na.rm = TRUE) +
+#   scale_colour_manual(values=paint2) +
+#   scale_x_continuous(breaks = 1:10) +
+#   xlab("Iteration") +
+#   ylab("Chain mean") + #bquote(theta)) +
+#   labs(color = "") + 
+#   theme_classic() +
+#   theme(legend.position = "bottom")
 
-# plot chains
-trace <- ggplot(diagnostics) +
-  geom_point(aes(x = iteration, y = `Chain 1`, color = patho),
-             size = .25,
-             na.rm = TRUE) +
-  geom_point(aes(x = iteration, y = `Chain 2`, color = patho),
-             size = .25,
-             na.rm = TRUE) +
-  geom_point(aes(x = iteration, y = `Chain 3`, color = patho),
-             size = .25,
-             na.rm = TRUE) +
-  geom_point(aes(x = iteration, y = `Chain 4`, color = patho),
-             size = .25,
-             na.rm = TRUE) +
-  geom_point(aes(x = iteration, y = `Chain 5`, color = patho),
-             size = .25,
-             na.rm = TRUE) +
-  geom_line(aes(x = iteration, y = `Chain 1`, color = patho), size = .25, na.rm = TRUE) +
-  geom_line(aes(x = iteration, y = `Chain 2`, color = patho), size = .25, na.rm = TRUE) +
-  geom_line(aes(x = iteration, y = `Chain 3`, color = patho), size = .25, na.rm = TRUE) +
-  geom_line(aes(x = iteration, y = `Chain 4`, color = patho), size = .25, na.rm = TRUE) +
-  geom_line(aes(x = iteration, y = `Chain 5`, color = patho), size = .25, na.rm = TRUE) +
-  scale_colour_manual(values=paint2) +
-  scale_x_continuous(breaks = 1:10) +
-  xlab("Iteration") +
-  ylab("Chain mean") + #bquote(theta)) +
-  labs(color = "") + 
-  theme_classic() +
-  theme(legend.position = "bottom")
-
-ggplotly(trace)
+#ggplotly(trace)
 
@@
 # with the 'amps' object
+library(mice)
 load("../traceplotdata.Rdata")
 set.seed(12)
-d75 <- amps[[4]] %>% mice(maxit=20)
-plot(d75)
-gg.mids(d4, x="Y")[[1]]
+d75 <- amps[[4]] %>% mice(maxit=10, printFlag = FALSE)
+#plot(d75)
+p <- gg.mids(d75, x="Y")[[1]] 
+p <- p + 
+  scale_color_manual(values = paint5) + 
+  scale_x_continuous(breaks = 1:10) + 
+  labs(
+    #title = "Traceplot", 
+    y = "Chain mean", 
+    colour = "Imp.")
+
+p %>% ggplotly()
 ```
@@
 
-# Results
+# Simulation Study
+
+From an incomplete multivariate normal set ( $n_{\rm obs}=1000, n_{\rm sim} = 1000$ ), we obtain valid regression estimates after 5 to 10 iterations.
+<!-- Multivariate normal set ( $n = 1000$ ), 'amputed' with a varying proportion of incomplete cases, and 'imputed' with a varying number of iterations.
+Depicted are the bias and coverage rate of an estimated regression coefficient ( $n_{\rm sim} = 1000$ ).
+-->
+
+<br>
 
@@
 
-# define colorblind friendly colors
-paint5 <- c('#228833', '#66CCEE', '#CCBB44','#EE6677', '#AA3377')
-
 # create plots
@@
              lwd = 1) +
-  geom_point(size = .25, na.rm = TRUE) +
-  geom_line(size = .25, na.rm = TRUE) +
+  geom_point(size = .5, na.rm = TRUE) +
+  geom_line(size = .5, na.rm = TRUE) +
   scale_colour_manual(values=paint5) +
-  xlab("") +
-  ylab("Bias (%)") +
-  labs(colour = "Proportion of missing cases (%)") +
+  labs(
+    #title = "Average bias in regression estimate", 
+    #subtitle = "Multivariate normal data (n = 1000, nsim = 1000)", 
+    x = "Number of iterations", 
+    y = "Bias (%)", 
+    colour = "Miss. (%)") +
   theme_classic() +
-  theme(legend.position = "bottom")
+  theme(legend.position = "top")
 
@@
                1) +
-  geom_point(size = .25, na.rm = TRUE) +
-  geom_line(size = .25, na.rm = TRUE) +
+  geom_point(size = .5, na.rm = TRUE) +
+  geom_line(size = .5, na.rm = TRUE) +
   geom_line(aes(x=t, y=thresh1.2), color = "grey", linetype = "dashed", size = .25, na.rm = TRUE) +
@@
   scale_y_continuous(limits = c(1,1.53), breaks = c(1,1.1,1.2,1.3,1.4,1.5)) +
-  xlab("Number of iterations") +
-  ylab(bquote(widehat(R))) +
-  labs(colour = "Proportion of missing cases (%)") +
+  labs(x = "Number of iterations", 
+       y = "$\\widehat{R}$",
+       colour = "Miss. (%)") +
   theme_classic() +
-  theme(legend.position = "")
+  theme(legend.position = "bottom")
 
@@
   geom_line(aes(x = t, y = crit), color = "grey", linetype = "dashed", size = .25, na.rm = TRUE) + 
-  geom_point(size = .25, na.rm = TRUE) +
-  geom_line(size = .25, na.rm = TRUE) +
+  geom_point(size = .5, na.rm = TRUE) +
+  geom_line(size = .5, na.rm = TRUE) +
   scale_colour_manual(values=paint5) +
-  xlab("Number of iterations") +
-  ylab("Autocorrelation") + 
-  labs(colour = "Proportion of missing cases (%)") +
+  labs(x = "Number of iterations",
+       y = "Auto-correlation",
+       colour = "Miss. (%)") +
   theme_classic() +
-  theme(legend.position = "")
+  theme(legend.position = "bottom")
 
@@
              lwd = 1) +
-  geom_point(size = .25, na.rm = TRUE) +
-  geom_line(size = .25, na.rm = TRUE) +
+  geom_point(size = .5, na.rm = TRUE) +
+  geom_line(size = .5, na.rm = TRUE) +
   scale_colour_manual(values=paint5) +
-  xlab("") +
+  xlab("Number of iterations") +
   ylab("Coverage rate (%)") +
-  labs(colour = "Proportion of missing cases (%)") +
+  labs(colour = "Miss. (%)") +
   theme_classic() +
@@
 
+```
+
+.pull-left[
+```{r echo=FALSE, fig.width=5.1, fig.height=5.4, message=FALSE, warning=FALSE}
 ggplotly(est_bias)
+```
+]
 
+.pull-right[
+```{r echo=FALSE, fig.width=5.1, fig.height=5.4, message=FALSE, warning=FALSE}
+ggplotly(est_cov)
 ```
+]
 
@@
 
+# Simulation Study (2)
+
+Whereas non-convergence diagnostics $\widehat{R}$ and auto-correlation identify signs of non-convergence up-to 30 to 50 iterations.
+
+<br>
+
+.pull-left[
+```{r echo=FALSE, fig.width=5.1, fig.height=5.4, message=FALSE, warning=FALSE}
+ggplotly(est_Rh)
+```
+]
+
+.pull-right[
+```{r echo=FALSE, fig.width=5.1, fig.height=5.4, message=FALSE, warning=FALSE}
+ggplotly(est_AC)
+```
+]
+
+---
 # Thanks!
 
-Slides created via the R package [**xaringan**](https://github.com/yihui/xaringan).
+.footnote[Slides created via the R package [**xaringan**](https://github.com/yihui/xaringan).]

### 2020-07-14 ICML/example_paper.tex
ICML proposal in correct format, minor changes to thesis manuscript (typo, figure label, etc.)
Recompile ATREMISS submission with correct style file

@@
-%%%%%%%% ICML 2020 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%
+%%%%%%%% Artemiss 2020 ICML EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%
 
@@
 % please comment out the following usepackage line and replace
-% \usepackage{icml2020} with \usepackage[nohyperref]{icml2020} above.
+% \usepackage{artemiss2020icml} with \usepackage[nohyperref]{artemiss2020icml} above.
 \usepackage{hyperref}
@@
 % Use the following line for the initial blind version submitted for review:
-\usepackage{icml2020}
+%\usepackage{artemiss2020icml}
 
 % If accepted, instead use the following line for the camera-ready submission:
-%\usepackage[accepted]{icml2020}
+\usepackage[accepted]{artemiss2020icml}
 
@@
 $\mathit{37}^{th}$ International Conference on Machine Learning},
-Vienna, Austria, 2020.
+Vienna, Austria, PMLR 119, 2020.
 Copyright 2020 by the author(s).''

### 2020-07-14 ICML/MissingThePoint.Rmd
Add ARTEMISS style files
Added links etc. to ARTEMISS presentation

@@
 
-# Algorithmic Non-Convergence
+# Algorithmic Convergence
 
@@
 
-From an incomplete multivariate normal set ( $n_{\rm obs}=1000, n_{\rm sim} = 1000$ ), we obtain valid regression estimates after 5 to 10 iterations.
+When imputing an incomplete multivariate normal set ( $n_{\rm obs}=1000$, $m=5$, $n_{\rm sim} = 1000$), we obtain valid regression estimates after 5 to 10 iterations.
 <!-- Multivariate normal set ( $n = 1000$ ), 'amputed' with a varying proportion of incomplete cases, and 'imputed' with a varying number of iterations.
@@
   labs(x = "Number of iterations", 
-       y = "$\\widehat{R}$",
+       y = "Ř", #r'$\\widehat{R}$'
        colour = "Miss. (%)") +
@@
 
+Read more: [github.com/hanneoberman/MissingThePoint](https://github.com/hanneoberman/MissingThePoint).
+
+Or follow my updates on Twitter: [@hioberman](https://twitter.com/hioberman)
+
 .footnote[Slides created via the R package [**xaringan**](https://github.com/yihui/xaringan).]

### 2020-07-15 ICML/MissingThePoint.Rmd
Add take-away slide to ARTEMISS presentation

@@
 author: "Hanne I. Oberman, Stef van Buuren, Gerko Vink"
-institute: "Utrecht University"
+institute: "ICML ARTEMISS workshop 2020"
 date: "17-07-2020" #(updated: `r Sys.Date()`)"
@@
 
-With iterative imputation, the validity of the inferences relies on algorithmic convergence.
+With iterative imputation, the validity of inferences relies on algorithmic convergence. Signs of non-convergence (i.e., non-mixing, trending) are typically identified through visual inspection.
 
-```{r non-conv, echo=FALSE, fig.width=10, fig.height = 6.2, message=FALSE, warning=FALSE}
+<br>
+
+<!-- .footer[Missing The Point | Oberman, van Buuren, Vink | 2020 | github.com/hanneoberman/MissingThePoint] -->
+
+```{r non-conv, echo=FALSE, fig.width=10, fig.height = 5.2, message=FALSE, warning=FALSE}
 # set-up env
@@
 library(plotly)
-load("example_diagnostics.Rdata")
+library(mice)
+#load("example_diagnostics.Rdata")
+load("../traceplotdata.Rdata")
 
@@
 # with the 'amps' object
-library(mice)
-load("../traceplotdata.Rdata")
 set.seed(12)
@@
     #title = "Traceplot", 
+    x = "Iteration",
     y = "Chain mean", 
@@
 
-p %>% ggplotly()
+ggplotly(p)
+
 ```
-
 ---
@@
 
-When imputing an incomplete multivariate normal set ( $n_{\rm obs}=1000$, $m=5$, $n_{\rm sim} = 1000$), we obtain valid regression estimates after 5 to 10 iterations.
+When imputing an incomplete multivariate normal set ( $n_{\rm obs}=1000$, $n_{\rm sim} = 1000$), we obtain valid regression estimates after 5 to 10 iterations.
 <!-- Multivariate normal set ( $n = 1000$ ), 'amputed' with a varying proportion of incomplete cases, and 'imputed' with a varying number of iterations.
@@
 ---
-# Thanks!
+# Take-away
 
-Read more: [github.com/hanneoberman/MissingThePoint](https://github.com/hanneoberman/MissingThePoint).
+We conclude that&mdash;in the cases considered&mdash;it never hurts to iterate longer, but such calculations hardly bring added value.
 
-Or follow my updates on Twitter: [@hioberman](https://twitter.com/hioberman)
+<br>
 
-.footnote[Slides created via the R package [**xaringan**](https://github.com/yihui/xaringan).]
+Read more on [github.com/hanneoberman/MissingThePoint](https://github.com/hanneoberman/MissingThePoint). Or follow my updates through Twitter [@hioberman](https://twitter.com/hioberman).
+
+.pull-left[
+<br> <br> <br>
+<br> <br> <br>
+<br> <br> <br>
+Upcoming...
+]
+
+.pull-right[
+<img src="https://raw.githubusercontent.com/gerkovink/shinyMice/2dae400d9d23ede7bec031c1501acdc681eab13c/ThesisProposal/Figures/logo.png" width="420">
+]
+
+<!-- .footnote[Slides created via the R package [**xaringan**](https://github.com/yihui/xaringan).] -->

### 2020-07-15 ICML/MissingThePoint.Rmd
Crop shinymice hex image for presentation

@@
   xaringan::moon_reader:
+    #seal: false
+    self_contained: true
     css: [default, default-fonts]
@@
       countIncrementalSlides: false
+      ratio: "16:9"
 ---
+
+```{css echo=FALSE}
+.right-column{
+  padding-top: 0;
+}
+```
 
@@
 options(htmltools.dir.version = FALSE)
+
+# set-up env
+library(dplyr)
+library(ggplot2)
+library(plotly)
+library(mice)
+#load("example_diagnostics.Rdata")
+# load("../traceplotdata.Rdata")
+# set.seed(12)
+# d75 <- amps[[4]] %>% mice(maxit=10, printFlag = FALSE)
+# save(d75, file = "imps.Rdata")
+load("imps.Rdata")
+
+# define colorblind friendly colors
+paint5 <- c('#228833', '#66CCEE', '#CCBB44','#EE6677', '#AA3377')
+# paint2 <- c('#66CCEE', '#EE6677')
 ```
 
+.left-column[ 
+```{r echo=TRUE, eval=FALSE}
+
+Missing The Point
+
+Non-Convergence 
+In Iterative 
+Imputation 
+Algorithms
+
+Hanne I. Oberman, 
+Stef van Buuren, 
+& Gerko Vink
+
+ARTEMISS workshop
+ICML 2020
+
+```
+
+]
+
+.right-column[ 
 # Algorithmic Convergence
@@
 ```{r non-conv, echo=FALSE, fig.width=10, fig.height = 5.2, message=FALSE, warning=FALSE}
-# set-up env
-library(dplyr)
-library(ggplot2)
-library(plotly)
-library(mice)
-#load("example_diagnostics.Rdata")
-load("../traceplotdata.Rdata")
-
-# define colorblind friendly colors
-paint5 <- c('#228833', '#66CCEE', '#CCBB44','#EE6677', '#AA3377')
-# paint2 <- c('#66CCEE', '#EE6677')
-# 
-# # plot chain
-# trace <- ggplot(diagnostics) +
-#   geom_point(aes(x = iteration, y = `Chain 1`, color = patho),
-#              size = .5,
-#              na.rm = TRUE) +
-#   geom_point(aes(x = iteration, y = `Chain 2`, color = patho),
-#              size = .5,
-#              na.rm = TRUE) +
-#   geom_point(aes(x = iteration, y = `Chain 3`, color = patho),
-#              size = .5,
-#              na.rm = TRUE) +
-#   geom_point(aes(x = iteration, y = `Chain 4`, color = patho),
-#              size = .5,
-#              na.rm = TRUE) +
-#   geom_point(aes(x = iteration, y = `Chain 5`, color = patho),
-#              size = .5,
-#              na.rm = TRUE) +
-#   geom_line(aes(x = iteration, y = `Chain 2`, color = patho), size = .5, na.rm = TRUE) +
-#   geom_line(aes(x = iteration, y = `Chain 3`, color = patho), size = .5, na.rm = TRUE) +
-#   geom_line(aes(x = iteration, y = `Chain 1`, color = patho), size = .5, na.rm = TRUE) +
-#   geom_line(aes(x = iteration, y = `Chain 4`, color = patho), size = .5, na.rm = TRUE) +
-#   geom_line(aes(x = iteration, y = `Chain 5`, color = patho), size = .5, na.rm = TRUE) +
-#   scale_colour_manual(values=paint2) +
-#   scale_x_continuous(breaks = 1:10) +
-#   xlab("Iteration") +
-#   ylab("Chain mean") + #bquote(theta)) +
-#   labs(color = "") + 
-#   theme_classic() +
-#   theme(legend.position = "bottom")
-
-#ggplotly(trace)
-
-# alternatively, use the actual data
-# run 1.Execute.R until imputing anything, than run RunSimulationOnce.R until after amputing the data
-# with the 'amps' object
-set.seed(12)
-d75 <- amps[[4]] %>% mice(maxit=10, printFlag = FALSE)
-#plot(d75)
 p <- gg.mids(d75, x="Y")[[1]] 
-p <- p + 
+p_up <- p + 
   scale_color_manual(values = paint5) + 
@@
 
-ggplotly(p)
+```
+
+```{r echo=FALSE, fig.width=10, fig.height = 5.2, message=FALSE, warning=FALSE}
+ggplotly(p_up)
+```
+]
+---
+
+.left-column[ 
+```{r echo=TRUE, eval=FALSE}
+
+Missing The Point
+
+Non-Convergence 
+In Iterative 
+Imputation 
+Algorithms
+
+Hanne I. Oberman, 
+Stef van Buuren, 
+& Gerko Vink
+
+ARTEMISS workshop
+ICML 2020
 
 ```
----
 
+]
+
+.right-column[ 
 # Simulation Study
 
-When imputing an incomplete multivariate normal set ( $n_{\rm obs}=1000$, $n_{\rm sim} = 1000$), we obtain valid regression estimates after 5 to 10 iterations.
+When imputing an incomplete multivariate normal set ( $n_{\rm obs}=$ 1000, $n_{\rm sim}=$ 1000), we obtain valid regression estimates after 5 to 10 iterations.
 <!-- Multivariate normal set ( $n = 1000$ ), 'amputed' with a varying proportion of incomplete cases, and 'imputed' with a varying number of iterations.
@@
   labs(x = "Number of iterations", 
-       y = "Ř", #r'$\\widehat{R}$'
+       y = "Potential scale reduction factor", #r'$\\widehat{R}$'
        colour = "Miss. (%)") +
@@
 ]
-
+]
 ---
 
+.left-column[ 
+```{r echo=TRUE, eval=FALSE}
+
+Missing The Point
+
+Non-Convergence 
+In Iterative 
+Imputation 
+Algorithms
+
+Hanne I. Oberman, 
+Stef van Buuren, 
+& Gerko Vink
+
+ARTEMISS workshop
+ICML 2020
+
+```
+
+]
+
+.right-column[ 
 # Simulation Study (2)
 
-Whereas non-convergence diagnostics $\widehat{R}$ and auto-correlation identify signs of non-convergence up-to 30 to 50 iterations.
+Whereas non-convergence diagnostics (potential scale reduction factor, auto-correlation) identify signs of non-convergence up-to 30 to 50 iterations.
 
@@
 ]
+]
+---
 
----
+.left-column[ 
+```{r echo=TRUE, eval=FALSE}
+
+Missing The Point
+
+Non-Convergence 
+In Iterative 
+Imputation 
+Algorithms
+
+Hanne I. Oberman, 
+Stef van Buuren, 
+& Gerko Vink
+
+ARTEMISS workshop
+ICML 2020
+
+```
+
+]
+
+.right-column[ 
 # Take-away
@@
 We conclude that&mdash;in the cases considered&mdash;it never hurts to iterate longer, but such calculations hardly bring added value.
-
-<br>
 
@@
 <br> <br> <br>
+<br> <br> <br>
 Upcoming...
@@
 .pull-right[
-<img src="https://raw.githubusercontent.com/gerkovink/shinyMice/2dae400d9d23ede7bec031c1501acdc681eab13c/ThesisProposal/Figures/logo.png" width="420">
+<img src="https://raw.githubusercontent.com/gerkovink/shinyMice/2dae400d9d23ede7bec031c1501acdc681eab13c/ThesisProposal/Figures/logo.png" width="595">
 ]
-
+]
 <!-- .footnote[Slides created via the R package [**xaringan**](https://github.com/yihui/xaringan).] -->

### 2020-07-15 ICML/MissingThePoint.Rmd
Save pdf version of presentation

@@
   xaringan::moon_reader:
-    #seal: false
+    seal: false
     self_contained: true
@@
 .pull-right[
-<img src="https://raw.githubusercontent.com/gerkovink/shinyMice/2dae400d9d23ede7bec031c1501acdc681eab13c/ThesisProposal/Figures/logo.png" width="595">
+<img src="https://github.com/gerkovink/shinyMice/blob/edits/ICML/shinymicehex.png?raw=true" width="255">
 ]

### 2020-07-15 ICML/MissingThePoint.Rmd
Add pdf version to upload

@@
     seal: false
-    self_contained: true
+    self_contained: false
     css: [default, default-fonts]
@@
 load("imps.Rdata")
+load("complete.Rdata")
 
@@
 # paint2 <- c('#66CCEE', '#EE6677')
-```
 
-.left-column[ 
-```{r echo=TRUE, eval=FALSE}
-
-Missing The Point
-
-Non-Convergence 
-In Iterative 
-Imputation 
-Algorithms
-
-Hanne I. Oberman, 
-Stef van Buuren, 
-& Gerko Vink
-
-ARTEMISS workshop
-ICML 2020
-
-```
-
-]
-
-.right-column[ 
-# Algorithmic Convergence
-
-With iterative imputation, the validity of inferences relies on algorithmic convergence. Signs of non-convergence (i.e., non-mixing, trending) are typically identified through visual inspection.
-
-<br>
-
-<!-- .footer[Missing The Point | Oberman, van Buuren, Vink | 2020 | github.com/hanneoberman/MissingThePoint] -->
-
-```{r non-conv, echo=FALSE, fig.width=10, fig.height = 5.2, message=FALSE, warning=FALSE}
+# traceplot
 p <- gg.mids(d75, x="Y")[[1]] 
@@
     colour = "Imp.")
-
-```
-
-```{r echo=FALSE, fig.width=10, fig.height = 5.2, message=FALSE, warning=FALSE}
-ggplotly(p_up)
-```
-]
----
-
-.left-column[ 
-```{r echo=TRUE, eval=FALSE}
-
-Missing The Point
-
-Non-Convergence 
-In Iterative 
-Imputation 
-Algorithms
-
-Hanne I. Oberman, 
-Stef van Buuren, 
-& Gerko Vink
-
-ARTEMISS workshop
-ICML 2020
-
-```
-
-]
-
-.right-column[ 
-# Simulation Study
-
-When imputing an incomplete multivariate normal set ( $n_{\rm obs}=$ 1000, $n_{\rm sim}=$ 1000), we obtain valid regression estimates after 5 to 10 iterations.
-<!-- Multivariate normal set ( $n = 1000$ ), 'amputed' with a varying proportion of incomplete cases, and 'imputed' with a varying number of iterations.
-Depicted are the bias and coverage rate of an estimated regression coefficient ( $n_{\rm sim} = 1000$ ).
--->
-
-<br>
-
-```{r results, echo=FALSE, fig.width=9.8, message=FALSE, warning=FALSE}
-load("complete.Rdata") 
-#library(patchwork)
 
@@
   theme_classic() +
-  theme(legend.position = "top")
+  theme(legend.position = "bottom")
 
@@
   theme_classic() +
-  theme(legend.position = "")
-
-# plot estimates, save 6.75x4.5in or 5.5
-# est_bias + est_cov + est_AC + est_Rh + plot_annotation(tag_levels = "A", tag_suffix = ".") + plot_layout(ncol = 2, guides = "collect")
+  theme(legend.position = "bottom")
 
@@
 
+.left-column[ 
+```{r echo=TRUE, eval=FALSE}
+Missing The Point
+
+Non-Convergence 
+In Iterative 
+Imputation 
+Algorithms
+
+Hanne I. Oberman, 
+Stef van Buuren, 
+& Gerko Vink
+
+ARTEMISS workshop
+ICML 2020
+
+```
+
+]
+
+.right-column[
+
+# Algorithmic Convergence
+
+With iterative imputation, the validity of inferences relies on algorithmic convergence. Signs of non-convergence (i.e., non-mixing, trending) are typically identified through visual inspection.
+
+<br>
+
+```{r echo=FALSE, fig.width=10, fig.height = 5.2, message=FALSE, warning=FALSE}
+ggplotly(p_up)
+```
+]
+
+---
+
+.left-column[ 
+```{r echo=TRUE, eval=FALSE}
+
+Missing The Point
+
+Non-Convergence 
+In Iterative 
+Imputation 
+Algorithms
+
+Hanne I. Oberman, 
+Stef van Buuren, 
+& Gerko Vink
+
+ARTEMISS workshop
+ICML 2020
+
+```
+]
+
+.right-column[ 
+# Simulation Study
+
+When imputing an incomplete multivariate normal set ( $n_{\rm obs}=$ 1000, $n_{\rm sim}=$ 1000), we obtain valid regression estimates after 5 to 10 iterations.
+
+<br>
+
 .pull-left[
-```{r echo=FALSE, fig.width=5.1, fig.height=5.4, message=FALSE, warning=FALSE}
+```{r echo=FALSE, fig.height=5.4, fig.width=5.1, message=FALSE, warning=FALSE}
 ggplotly(est_bias)
@@
 ]
+
 ---
@@
 .right-column[
+
 # Simulation Study (2)
 
-Whereas non-convergence diagnostics (potential scale reduction factor, auto-correlation) identify signs of non-convergence up-to 30 to 50 iterations.
+Whereas non-convergence diagnostics (potential scale reduction factor and auto-correlation) identify signs of non-convergence up-to 30 to 50 iterations.
 
@@
 .pull-right[
-<img src="https://github.com/gerkovink/shinyMice/blob/edits/ICML/shinymicehex.png?raw=true" width="255">
+<img src="https://raw.githubusercontent.com/gerkovink/shinyMice/edits/shinymicehex.png" width="255">
 ]

### 2020-07-16 ICML/MissingThePoint.Rmd
Updated presentation with double legend removed

@@
   padding-top: 0;
+}
+.remark-slide-number {
+  font-size: 10pt;
+  margin-bottom: -11.6px;
+  margin-right: 10px;
+  color: #FFFFFF; /* white */
+  opacity: 1; /* default: 0.5 */
 }
@@
 ]
-
 ---
-
+---
 .left-column[ 
@@
 ]
-<!-- .footnote[Slides created via the R package [**xaringan**](https://github.com/yihui/xaringan).] -->

### 2020-07-20 shinymice/about.Rmd
Initial version shinymice app

---
  title: "About"
output: html_document
---

This Shiny app has the sole purpose of showing that I'm able to program an app in `R Shiny`. However, if all goes well, I might use this as the foundation of the interactive evaluation suite for evaluating multiply imputed data that I'm developing as Research Master's thesis project. 'ShinyMICE' will be integrated into the `mice()` environment in `R`.

Watch the current state of the project on [github.com/gerkovink/ShinyMICE](https://github.com/gerkovink/ShinyMICE/).

### 2020-08-03 shinymice/www/about.Rmd
Fix light/dark themes
Show dataset name instead of variables in Impute tab

--- 2020-07-20 shinymice/about.Rmd
@@
 However, if all goes well, I might use this as the foundation of the interactive evaluation suite for evaluating multiply imputed data that I'm developing as Research Master's thesis project.
-'ShinyMICE' will be integrated into the `mice()` environment in `R`.
+`shinymice` will be integrated into the `mice` environment in `R`.
 
-Watch the current state of the project on [github.com/gerkovink/ShinyMICE](https://github.com/gerkovink/ShinyMICE/).
+Watch the current state of the project on [github.com/gerkovink/shinyMice](https://github.com/gerkovink/shinyMice/).

### 2020-08-03 shinymice/ui.R

@@ -119,8 +119,7 @@ shinyUI(
                 icon = icon("calculator"),
                 h2("Impute missing data using `mice`"),
                 tags$b("Dataset to impute"),
-                verbatimTextOutput("names"),
-                helpText("Showing only the first 5 variable names to check what data is used."),
+                verbatimTextOutput("datname"),
                 numericInput(
                     "m",
                     label = "Number of imputations",
@@ -142,7 +141,7 @@ shinyUI(
                 actionButton("mice", "Impute", icon = icon("hourglass-start")),
                 helpText("This may take a minute."),
                 br(),
-                verbatimTextOutput("done") #%>% withSpinner(color="#0dc5c1")
+                verbatimTextOutput("done") 
                 #plotOutput("traceplot")
             ),
             
@@ -161,7 +160,6 @@ shinyUI(
                 tabPanel(
                     "Convergence",
                     icon = icon("chart-line"),
-                    #icon("list-alt")
                     h2("Evaluate convergence"),
                     varSelectInput("varnr", "Choose a variable:", data = mice::boys),
                     plotOutput("traceplot"),
@@ -231,10 +229,10 @@ shinyUI(
                     "About",
                     icon = icon("info-circle"),
                     h2("About this app"),
-                    fluidRow(column(6,
+                    fluidRow(column(8,
                                     includeMarkdown("www/about.Rmd")),
                              column(
-                                 3,
+                                 4,
                                  img(
                                      class = "img-polaroid",
                                      src = paste0(
@@ -245,7 +243,7 @@ shinyUI(
                                  tags$small(
                                      "Impression of the hex sticker ",
                                      "for the interactive evaulation ",
-                                     "suite for multiple imputation 'ShinyMICE'",
+                                     "suite for multiple imputation `shinymice`",
                                      a(href = "https://github.com/gerkovink/shinyMice")
                                  )
                              ))

### 2020-09-02 R/descriptive.statistics.R
Use psych::describe() instead of custom functions for descriptives #47

@@ -1,50 +1,50 @@
 # create descriptive stats table
 
-num_desc <- function(d, v) {
-  m <- mean(d[[v]], na.rm = TRUE)
-  s <- sd(d[[v]], na.rm = TRUE)
-  na <- sum(is.na(d[[v]]))
-  return(data.frame(
-    mean = round(m, 2),
-    sd = round(s, 2),
-    mode = "",
-    freq = "",
-    nmiss = na
-  )) #%>% round(.,2))
-}
-
-cat_desc <- function(d, v) {
-  t <- table(d[[v]])
-  m <- names(which.max(t))
-  p <- t[[m]]#/sum(t)*100
-  na <- sum(is.na(d[[v]]))
-  
-  return(data.frame(
-    mean = "",
-    sd = "",
-    mode = m,
-    freq = round(p, 2),
-    nmiss = na
-  ))
-}
-
-descriptives <- function(data) {
-  nums <- data %>%
-    select(where(is.numeric)) %>%
-    names(.) %>%
-    purrr::map_dfr( ~ {
-      num_desc(d = data, v = .x) %>%
-        cbind(column = .x, .)
-    })
-  cats <- data %>%
-    select(!(where(is.numeric))) %>%
-    names(.) %>%
-    purrr::map_dfr( ~ {
-      cat_desc(d = data, v = .x) %>%
-        cbind(column = .x, .)
-    })
-  return(rbind(nums, cats))
-}
+# num_desc <- function(d, v) {
+#   m <- mean(d[[v]], na.rm = TRUE)
+#   s <- sd(d[[v]], na.rm = TRUE)
+#   na <- sum(is.na(d[[v]]))
+#   return(data.frame(
+#     mean = round(m, 2),
+#     sd = round(s, 2),
+#     mode = "",
+#     freq = "",
+#     nmiss = na
+#   )) #%>% round(.,2))
+# }
+# 
+# cat_desc <- function(d, v) {
+#   t <- table(d[[v]])
+#   m <- names(which.max(t))
+#   p <- t[[m]]#/sum(t)*100
+#   na <- sum(is.na(d[[v]]))
+#   
+#   return(data.frame(
+#     mean = "",
+#     sd = "",
+#     mode = m,
+#     freq = round(p, 2),
+#     nmiss = na
+#   ))
+# }
+# 
+# descriptives <- function(data) {
+#   nums <- data %>%
+#     select(where(is.numeric)) %>%
+#     names(.) %>%
+#     purrr::map_dfr( ~ {
[…]

### 2020-10-23 shinymice/dev/02_dev.R
Initialize {golem} framework for more robust shinymice

@@ -1,0 +1,59 @@
+# Building a Prod-Ready, Robust Shiny Application.
+# 
+# README: each step of the dev files is optional, and you don't have to 
+# fill every dev scripts before getting started. 
+# 01_start.R should be filled at start. 
+# 02_dev.R should be used to keep track of your development during the project.
+# 03_deploy.R should be used once you need to deploy your app.
+# 
+# 
+###################################
+#### CURRENT FILE: DEV SCRIPT #####
+###################################
+
+# Engineering
+
+## Dependencies ----
+## Add one line by package you want to add as dependency
+usethis::use_package( "mice" )
+
+## Add modules ----
+## Create a module infrastructure in R/
+golem::add_module( name = "missingness") # Name of the module
+golem::add_module( name = "home" ) # Name of the module
+
+## Add helper functions ----
+## Creates ftc_* and utils_*
+golem::add_fct( "helpers" ) 
+golem::add_utils( "helpers" )
+
+## External resources
+## Creates .js and .css files at inst/app/www
+golem::add_js_file( "script" )
+golem::add_js_handler( "handlers" )
+golem::add_css_file( "custom" )
+
+## Add internal datasets ----
+## If you have data in your package
+usethis::use_data_raw( name = "my_dataset", open = FALSE ) 
+
+## Tests ----
+## Add one line by test you want to create
+usethis::use_test( "app" )
+
+# Documentation
+
+## Vignette ----
+usethis::use_vignette("shinymice")
+devtools::build_vignettes()
+
+## Code coverage ----
+## (You'll need GitHub there)
+usethis::use_github()
+usethis::use_travis()
+usethis::use_appveyor()
+
+# You're now set! ----
+# go to dev/03_deploy.R
+rstudioapi::navigateToFile("dev/03_deploy.R")
+

### 2020-10-26 shinymice/README.Rmd
Add logo's

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# shinymice

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
<!-- badges: end -->

The goal of shinymice is to ...

## Installation

You can install the released version of shinymice from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("shinymice")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("amices/shinyMice")
```
## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(shinymice)
## basic example code
```

What is special about using `README.Rmd` instead of just `README.md`? You can include R chunks like so:

```{r cars}
summary(cars)
```

You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date.

You can also embed plots, for example:

```{r pressure, echo = FALSE}
plot(pressure)
```

In that case, don't forget to commit and push the resulting figure files, so they display on GitHub!

### 2020-10-26 shinymice/R/app_ui.R

@@ -9,7 +9,9 @@ app_ui <- function(request) {
     golem_add_external_resources(),
     # List the first level UI elements here
     fluidPage(navbarPage(
-      title = "shinymice",
+      windowTitle = "shinymice",
+      title = div(img(src="www/logo.png", width = 120)),
+      header = br(),
       tabPanel("Home",
                mod_home_ui("home_ui_1")),
       tabPanel("Missing data",
@@ -18,7 +20,7 @@ app_ui <- function(request) {
                mod_imputationmodel_ui("imputationmodel_ui_1")),
       tabPanel("Imputed data",
                mod_imputeddata_ui("imputeddata_ui_1")),
-      tabPanel("Analyze the imputed data",
+      tabPanel("Scientific model",
                mod_scientificmodel_ui("scientificmodel_ui_1")),
       tabPanel("Exit",
                mod_save_ui("save_ui_1"))
@@ -38,7 +40,7 @@ golem_add_external_resources <- function(){
     'www', app_sys('app/www')
   )
   tags$head(
-    favicon(),
+    favicon(ext = 'png'),
     bundle_resources(
       path = app_sys('app/www'),
       app_title = 'shinymice'

### 2020-10-29 shinymice/R/mod_missingness.R
Start vignette for ggmice functions

@@ -56,10 +56,20 @@ mod_missingness_ui <- function(id) {
 #' @noRd
 mod_missingness_server <- function(id) {
   moduleServer(id, function(input, output, session) {
+    p <- dummy_plot()
     output$na_tab <- DT::renderDT(cbind(mice::boys, mice::boys))
     output$na_desc <- renderTable(mis_descr(mice::boys))
-    output$md_pat <- renderPlot(dummy_plot())
-    output$na_plot <- renderPlot(dummy_plot() + list(ggplot2::labs(title = paste0("We'll eventually plot variables '", input$var1, "' and '", input$var2, "'"))))
+    output$md_pat <- renderPlot(p)
+    output$na_plot <- renderPlot(p +
+                                   list(ggplot2::labs(
+                                     title = paste0(
+                                       "We'll eventually plot variables '",
+                                       input$var1,
+                                       "' and '",
+                                       input$var2,
+                                       "'"
+                                     )
+                                   )))
   })
 }
 

### 2020-11-02 shinymice/dev/02_dev.R
Get demo deployed

@@ -18,6 +18,7 @@
 usethis::use_package("mice")
 usethis::use_package("ggplot2")
 usethis::use_package("dplyr")
+usethis::use_package("psych")
 usethis::use_pipe()
 
 ## Add modules ----

### 2020-11-03 shinymice/R/mod_imputationmodel.R
Add fluxplot and hacky version of NA plot to demo shinymice

@@ -24,23 +24,24 @@ mod_imputationmodel_ui <- function(id) {
       "3. And the number of iterations (maxit)",
       set_number("maxit", val = 10),
       no_br(),
-      "4. Check (and modify) the predictor matrix.",
+      "4. Evaluate the in and outflux of the model.",
       br(),
-      "5. Run `mice()`.",
+      "5. Check (and modify) the predictor matrix.",
+      br(),
+      "6. Run `mice()`.",
       br(),
       actionButton(ns("run_mice"), label = "Impute", width = 100),
       br(),
-      "6. Evaluate the in and outflux of the model.",
-      br(),
-      "6. Monitor potential non-convergence through visual inspection.",
+      "7. Monitor potential non-convergence through visual inspection.",
       verbatimTextOutput(ns("micecall"))
     ),
     column(9,
            tabsetPanel(
+             tabPanel("Fluxplot",
+                      #plotly::plotlyOutput(
+                        plotOutput(ns("flux_plot"))),
              tabPanel("Predictor matrix",
                       plotOutput(ns("pred_plot"))),
-             tabPanel("Fluxplot",
-                      plotOutput(ns("flux_plot"))),
              tabPanel(
                "Traceplot",
                select_var(ns("var1")),
@@ -57,8 +58,12 @@ mod_imputationmodel_server <- function(id) {
   moduleServer(id, function(input, output, session) {
     ns <- session$ns
     p <- dummy_plot()
+    #mids <- mice::mice(mice::boys, maxit = 1)
     output$pred_plot <- renderPlot(p)
-    output$flux_plot <- renderPlot(p)
+    output$flux_plot <- renderPlot({#plotly::renderPlotly({
+      plot_flux(mice::boys) 
+      #return(plotly::ggplotly())
+      })
     output$trace_plot <- renderPlot(p)
   })
 }

### 2020-11-03 shinymice/R/mod_missingness.R
Add conditional density/barplots

@@ -28,26 +28,37 @@ mod_missingness_ui <- function(id) {
       br(),
       "3. Browse the dataset to view missing data points.",
       br(),
-      "4. Look at the missing data pattern for an overview of the missingness.",
+      "4. Evaluate the distribution of variables conditional on missingness in another variable.",
       br(),
       "5. Evaluate the bivariate relations in the incomplete data.",
+      br(),
+      "6. Look at the missing data pattern for an overview of the missingness.",
       br()
     ),
-    column(9,
-           tabsetPanel(
-             tabPanel("Descriptives",
-                      tableOutput(ns("na_desc"))),
-             tabPanel("Browse",
-                      DT::DTOutput(ns("na_tab"))),
-             tabPanel("Missingness pattern",
-                      plotOutput(ns("md_pat"))),
-             tabPanel(
-               "Distributions",
-               select_var(ns("var1")),
-               select_var(ns("var2")),
-               plotOutput(ns("na_plot"))
-             )
-           ))
+    column(
+      9,
+      tabsetPanel(
+        tabPanel("Descriptives",
+                 tableOutput(ns("na_desc"))),
+        tabPanel("Browse",
+                 DT::DTOutput(ns("na_tab"))),
+        
+        tabPanel(
+          "Conditional distributions",
+          select_var(ns("var1")),
+          select_var(ns("var2")),
+          plotOutput(ns("cond_plot"))
+        ),
+        tabPanel(
+          "Bivariate distributions",
+          select_var(ns("var1")),
+          select_var(ns("var2")),
+          plotOutput(ns("na_plot"))
+        ),
+        tabPanel("Missingness pattern",
+                 plotOutput(ns("md_pat")))
+      )
+    )
   ))
 }
 
@@ -60,16 +71,23 @@ mod_missingness_server <- function(id) {
     output$na_tab <- DT::renderDT(cbind(mice::boys, mice::boys))
     output$na_desc <- renderTable(mis_descr(mice::boys))
     output$md_pat <- renderPlot(p)
-    output$na_plot <- renderPlot({plot_NA_margins(data = mice::boys, y = "hc", x = "age") +
-                                   list(ggplot2::labs(
-                                     title = paste0(
-                                       "We'll eventually plot variables '",
-                                       input$var1,
-                                       "' and '",
-                                       input$var2,
-                                       "' (select above)."
-                                     )
-                                   ))})
+    output$cond_plot <-
[…]

### 2020-11-03 shinymice/R/mod_missingness_utils_server.R

@@ -1,8 +1,39 @@
 # descriptive statistics for incomplete data
+
+# show NAs in tables
 options("DT.TOJSON_ARGS" = list(na = "string"))
-mis_descr <- function(d){
-  tab <- psych::describe(d)[,c("n", "mean", "sd", "min", "max", "median")] %>% 
+
+# descriptive table
+mis_descr <- function(d) {
+  tab <-
+    psych::describe(d)[, c("n", "mean", "sd", "min", "max", "median")] %>%
     cbind(variable = rownames(.), ., n_missing = as.integer(nrow(d) - .$n))
-  tab[,2] <- as.integer(tab[,2])
+  tab[, 2] <- as.integer(tab[, 2])
   return(tab)
+}
+
+# conditional distribution plot
+plot_conditional <- function(dat, x, z) {
+  if (is.numeric(dat[[x]])) {
+    geom <-
+      ggplot2::geom_density(ggplot2::aes(x = .data[[x]], y = ..count.., color = conditional))
+  } else {
+    geom <-
+      ggplot2::geom_bar(
+        ggplot2::aes(x = .data[[x]], color = conditional),
+        position = ggplot2::position_dodge(),
+        fill = "white"
+      )
+  }
+  dat %>%
+    dplyr::mutate(conditional = factor(
+      is.na(.data[[z]]),
+      levels = c(TRUE, FALSE),
+      labels = c("Missing", "Observed")
+    )) %>%
+    ggplot2::remove_missing(., vars = x) %>%
+    ggplot2::ggplot() +
+    geom +
+    theme_mice() +
+    ggplot2::labs(color = paste0("Variable '", z, "' is:"))
 }

### 2020-11-04 shinymice/R/utils_server.R
Develop traceplot function

@@ -17,6 +17,7 @@ dummy_plot <- function() {
     ggplot2::ggtitle("Dummy plot (to be replaced)")
 }
 
+# hide some plotly options, see https://plotly-r.com/control-modebar.html
 clean_plotly <- function(pp, ...) {
   plotly::config(
     p = pp,

### 2020-11-11 shinymice/R/app_ui.R
Test histogram/barplot with new preprocessing

@@ -36,7 +36,7 @@ app_ui <- function(request) {
         # tabPanel("Scientific model",
         #          icon = icon("calculator"),
         #          mod_scientificmodel_ui("scientificmodel_ui_1")),
-        tabPanel("Exit",
+        tabPanel("Save",
                  icon = icon("file-download"),
                  mod_save_ui("save_ui_1"))
       )

### 2020-11-17 shinymice/R/mod_missingness_utils_server.R
Add plotting functions to utils files

@@ -12,110 +12,143 @@ mis_descr <- function(d) {
   return(tab)
 }
 
-# conditional distribution plot
-plot_conditional <- function(dat, x, z) {
-  # define geom for numerical vs. categorical data
+# histogram/bar plot conditional on missingness in another variable
+plot_NA_cond <- function(dat, x, z) {
+  # define graphing elements to add to plot
   if (is.numeric(dat[[x]])) {
-    geom <-
-      ggplot2::geom_density(ggplot2::aes(x = .data[[x]], y = ..count.., color = conditional), size = 1)
+    # for continuous variables
+    geom <- ggplot2::geom_histogram()
   } else {
-    geom <-
-      ggplot2::geom_bar(
-        ggplot2::aes(x = .data[[x]], color = conditional),
-        position = ggplot2::position_dodge(),
-        fill = "white",
-        size = 1
-      )
+    # for categorical variables
+    geom <- ggplot2::geom_bar()
   }
-  # preprocess and plot the data
-  dat %>%
+  # create facet labels
+  facet_labs <- c(paste(z, "observed"), paste(z, "missing")) %>%
+    setNames(c("observed", "missing"))
+  # preprocess the data
+  dat[!is.na(dat[[x]]), ] %>%
     dplyr::mutate(conditional = factor(
       is.na(.data[[z]]),
-      levels = c(TRUE, FALSE),
-      labels = c("Missing", "Observed")
+      levels = c(FALSE, TRUE),
+      labels = c("observed", "missing")
     )) %>%
-    ggplot2::remove_missing(., vars = x) %>%
-    ggplot2::ggplot() +
+    # plot
+    ggplot2::ggplot(ggplot2::aes(x = .data[[x]],
+                                 color = conditional,
+                                 fill = "white")) +
     geom +
+    # split by conditional variable
+    ggplot2::facet_wrap(
+      ~ conditional,
+      ncol = 1,
+      scales = "free_y",
+      labeller = ggplot2::labeller(conditional = facet_labs)
+    ) +
+    # style
     theme_mice() +
-    ggplot2::labs(color = paste0("Variable '", z, "' is:"))
+    theme(legend.position = "none",
[…]

### 2020-11-17 shinymice/R/utils_server.R

@@ -7,6 +7,11 @@
 #' @return The return value, if any, from executing the utility.
 #'
 #' @export
+
+# not in
+`%nin%` <- Negate(`%in%`)
+
+# dummy plot
 dummy_plot <- function() {
   ggplot2::ggplot(data = data.frame(
     a = c(0, 1),
@@ -15,6 +20,49 @@ dummy_plot <- function() {
   )) +
     ggplot2::geom_text(ggplot2::aes(x = a, y = b, label = c), size = 5) +
     ggplot2::ggtitle("Dummy plot (to be replaced)")
+}
+
+# mice theme
+theme_mice <- function() {
+  # change default plotting options (move this outside of the function to unnecessary avoid re-running?)
+  ggplot2::update_geom_defaults("point", list(
+    shape = 21,
+    size = 1.9,
+    stroke = 1.1,
+    alpha = 0.5
+  ))
+  ggplot2::update_geom_defaults("boxplot", list(size = 1))
+  ggplot2::update_geom_defaults("line", list(size = 1))
+  # mice style settings
+  theme <- list(
+    ggplot2::theme_classic(),
+    ggplot2::theme(legend.position = "top"),
+    ggplot2::scale_color_manual(
+      values = c(
+        "observed" = mice:::mdc(1),
+        "missing" = mice:::mdc(2),
+        "imputed" = mice:::mdc(2)
+      )
+    ),
+    ggplot2::scale_fill_manual(
+      values = c(
+        "observed" = mice:::mdc(1),
+        "missing" = mice:::mdc(2),
+        "imputed" = mice:::mdc(2)
+      )
+    ),
+    ggplot2::scale_size_manual(values = c(
+      "observed" = 1,
+      "imputed" = 0.5
+    )),
+    ggplot2::labs(
+      color = NULL,
+      fill = NULL,
+      size = NULL
+    )
+  )
+  # output
+  return(theme)
 }
 
 # hide some plotly options, see https://plotly-r.com/control-modebar.html

### 2020-11-17 shinymice/R/mod_missingness_utils_server.R
Add plotting functions to shinymice demo

@@ -46,8 +46,8 @@ plot_NA_cond <- function(dat, x, z) {
     ) +
     # style
     theme_mice() +
-    theme(legend.position = "none",
-          strip.background = element_rect(size = 0.5))
+    ggplot2::theme(legend.position = "none",
+          strip.background = ggplot2::element_rect(size = 0.5))
 }
 
 # scatterplot with NAs

### 2020-11-17 shinymice/R/mod_imputationmodel_utils_server.R
Make plot on incomplete data tab interactive (select variables)

@@ -15,6 +15,7 @@ plot_md_pattern <- function(dat) {
   n_pat_obs <- as.numeric(rownames(pat))[-nrow(pat)]
   n_vrb_inc <- as.numeric(pat[, ncol(pat)])[-nrow(pat)]
   n_val_mis <- as.numeric(pat[nrow(pat), ])[-ncol(pat)]
+  n_mis_tot <- pat[nrow(pat), ncol(pat)]
   # make the pattern tidy
   long_pat <- pat[-nrow(pat), ] %>%
     cbind(., n_pat_obs, pat_nr = 1:nrow(.)) %>%
@@ -42,11 +43,12 @@ plot_md_pattern <- function(dat) {
       labels = as.character(n_pat_obs),
       expand = c(0.01, 0.01),
       sec.axis = ggplot2::dup_axis(labels = as.character(n_vrb_inc),
-                                   name = "Incomplete variables (per pattern)")
+                                   name = "Number of missing entries per pattern")
     ) +
     # add labels
-    ggplot2::labs(x = "Incomplete cases (per variable)",
-                  y = "Pattern frequency") +
+    ggplot2::labs(x = "Number of missing entries per variable",
+                  y = "Pattern frequency",
+                  title = paste0("Missing data pattern (total number of missing cells = ", n_mis_tot, ")\n")) +
     ggplot2::geom_text(ggplot2::aes(x = vrb, y = -Inf, label = vrb),
                        data = long_pat,
                        vjust = -0.5) +
@@ -54,7 +56,7 @@ plot_md_pattern <- function(dat) {
     # add styling
     ggplot2::theme(
       legend.position = "none",
-      plot.margin = ggplot2::margin(t = 25, l = 10, b = 10, r = 10, unit = "pt"),
+      plot.margin = ggplot2::margin(t = 20, l = 10, b = 10, r = 10, unit = "pt"),
       axis.title.y.right = ggplot2::element_text(margin = ggplot2::margin(l = 10)),
       panel.grid = ggplot2::element_blank()
     ) +

### 2020-11-23 shinymice/R/mod_missingness_utils_server.R
Add feedback to user to select variable(s)

@@ -28,7 +28,7 @@ mis_descr <- function(d) {
 #' @examples
 plot_NA_cond <- function(dat, x, z) {
   # escape function if no variable is selected
-  if (x == "Select a variable" | z == "Select a variable") {return(NULL)}
+  if (x == "Select a variable" | z == "Select a variable") {return(ggplot2::ggplot(dat) + ggplot2::ggtitle("Please select variable(s)"))}
   # define graphing elements to add to plot
   if (is.numeric(dat[[x]])) {
     # for continuous variables
@@ -78,7 +78,7 @@ plot_NA_cond <- function(dat, x, z) {
 #' @examples
 plot_NA_scatter <- function(dat, x, y) {
   # escape function if no variable is selected
-  if (x == "Select a variable" | y == "Select a variable") {return(NULL)}
+  if (x == "Select a variable" | y == "Select a variable") {return(ggplot2::ggplot(dat) + ggplot2::ggtitle("Please select variable(s)"))}
   # define graphing elements to add to plot later
   # set NA value and scale for variable x
   if (is.numeric(dat[[x]])) {

### 2020-11-23 shinymice/R/mod_imputeddata.R
Add faceted bar plot for imputed categorical variables

@@ -27,22 +27,22 @@ mod_imputeddata_ui <- function(id) {
         tabPanel("Descriptives",
                  tableOutput(ns("imp_desc"))),
         tabPanel(
-          "Stripplot",
+          "Data points",
           select_var(ns("var1")),
           plotOutput(ns("strip_plot"))
         ),
         tabPanel(
-          "Boxplot",
+          "Distribution",
           select_var(ns("var2")),
           plotOutput(ns("bw_plot"))
         ),
         tabPanel(
-          "Densityplot",
+          "Density",
            select_var(ns("var3")),
            plotOutput(ns("dens_plot"))
         ),
         tabPanel(
-          "Scatterplot",
+          "Scatter plot",
           select_var(ns("var4")),
           select_var(ns("var5")),
           plotOutput(ns("xy_plot"))

### 2020-11-24 shinymice/R/mod_imputationmodel_utils_server.R
Adjust predictor matrix plot function to use predictro matrix as input (instead of just data or mids objects)

@@ -104,24 +104,23 @@ plot_flux <- function(dat) {
 #' @export
 #'
 #' @examples
-plot_pred_matrix <- function(dat) {
+plot_pred_matrix <- function(d) {
   # parse input
-  if (mice::is.mids(dat)) {
-    imp <- dat
+  if (is(d, "mids")) {
+    pred <- d$predictorMatrix
+  } else if (is(d, "matrix")) {
+    pred <- d
   } else {
-    imp <- mice::mice(dat, maxit = 0)
+    pred <- mice::mice(d, maxit = 0) %>% .$predictorMatrix
   }
-  # get the predictor matrix and add the rownames as variable
-  pred <- imp$predictorMatrix %>%
-    as.data.frame() %>%
-    cbind(vrb_to_imp = rownames(.), .)
   # make the data tidy to plot
-  long_pred <-
+  long_pred <- pred %>%
+    as.data.frame() %>%
+    cbind(vrb_to_imp = rownames(.), .) %>% 
     tidyr::pivot_longer(
-      pred,
-      cols = names(pred)[-1],
+      cols = names(.)[-1],
       names_to = "predictor",
-      names_transform = list(predictord = as.factor)
+      names_transform = list(predictor = as.factor)
     )
   # plot the predictor matrix
   p <- long_pred %>%

### 2020-11-30 shinymice/R/mod_imputationmodel.R
Add rhat plot to traceplot tab

@@ -61,10 +61,14 @@ mod_imputationmodel_ui <- function(id) {
                tags$b("Interpretation:"),
                "Each row in the predictor matrix identifies which predictors are to be used for the variable in the row name."
              ),
-             tabPanel("Traceplot",
-                      h6("Please make sure to impute the incomplete data first (see left-hand side)"),
-                      select_var(ns("var1")),
-                      plotOutput(ns("trace_plot")))
+             tabPanel(
+               "Traceplot",
+               h6("Please make sure to impute the incomplete data first (see left-hand side)"),
+               select_var(ns("var1")),
+               plotOutput(ns("trace_plot")),
+               "Convergence diagnostic",
+               plotOutput(ns("rhat_plot"))
+             )
            ))
   ))
 }
@@ -93,7 +97,8 @@ mod_imputationmodel_server <- function(id, data) {
     imp <- eventReactive(input$run_mice, mice::mice(data(), seed = input$seed, m = input$m, maxit = input$maxit))
     chains <- reactive(preprocess_thetas(imp()))
     output$trace_plot <- renderPlot(trace_one_variable(chains(), x = input$var1))
-  return(reactive(imp()))
+    output$rhat_plot <- renderPlot(plot_rhat(imp(), x = input$var1, theta = "both"))
+    return(reactive(imp()))
     })
 }
 

### 2020-12-16 shinymice/R/mod_imputationmodel_utils_server.R
Clean up all utils files

@@ -1,374 +1,363 @@
-# plot md pattern of incomplete data
-#' Title
-#'
-#' @param dat An incomplete dataset of class dataframe
-#'
-#' @return
-#' @export
-plot_md_pattern <- function(dat) {
-  # escape function if dataset is complete
-  if (!any(is.na(dat))) {
-    return(plot_a_mouse())
-  }
-  # get md pattern and store additional info
-  pat <- mice::md.pattern(dat, plot = FALSE)
-  vrb <- colnames(pat)[-ncol(pat)]
-  colnames(pat) <- c(vrb, "NA_per_pat")
-  pat_freq <- as.numeric(rownames(pat))[-nrow(pat)]
-  NA_per_pat <- as.numeric(pat[, ncol(pat)])[-nrow(pat)]
-  NA_per_vrb <- as.numeric(pat[nrow(pat),])[-ncol(pat)]
-  NA_total <- pat[nrow(pat), ncol(pat)]
-  # make the pattern tidy
-  long_pat <- pat[-nrow(pat),] %>%
-    cbind(., pat_freq, pat_nr = 1:nrow(.)) %>%
-    as.data.frame() %>%
-    tidyr::pivot_longer(cols = all_of(vrb),
-                        names_to = "vrb",
-                        values_to = "obs") %>%
-    cbind(., NA_per_vrb)
-  # plot the md pattern
-  p <- long_pat %>%
-    ggplot2::ggplot() +
-    ggplot2::geom_tile(ggplot2::aes(
-      x = vrb,
-      y = pat_nr,
-      fill = as.factor(obs),
-      group = NA_per_pat#,
-      # text = paste('pat_freq: ', pat_freq,
-      #              '</br>NA_per_vrb: ', NA_per_vrb)
-    ),
-    color = "black") +
-    # set axes
-    ggplot2::scale_x_discrete(
-      limits = vrb,
-      position = "bottom",
-      labels = as.character(NA_per_vrb),
-      expand = c(0, 0)
-    ) +
-    ggplot2::scale_y_reverse(
-      breaks = 1:max(long_pat$pat_nr),
-      labels = as.character(pat_freq),
-      expand = c(0, 0),
-      sec.axis = ggplot2::dup_axis(labels = as.character(NA_per_pat),
-                                   name = "Number of missing entries per pattern")
-    ) +
-    # add labels
-    ggplot2::labs(
-      x = "Number of missing entries per variable",
-      y = "Pattern frequency",
[…]

### 2020-12-16 shinymice/dev/02_dev.R

@@ -41,9 +41,9 @@ golem::add_utils("ui")
 golem::add_utils("server")
 # golem::add_utils("00_home")
 golem::add_utils("01_data")
-# golem::add_utils("02_model")
-# golem::add_utils("03_convergence")
-# golem::add_utils("04_imputations")
+golem::add_utils("02_model")
+golem::add_utils("03_convergence")
+golem::add_utils("04_imputations")
 # golem::add_utils("05_save")
 
 # #golem::add_utils("ui", module = "missingness")

### 2021-12-16 00_TODO/ProportionEffect/test_prop_of_inc_cases.R
Clean-up repo

@@ -1,124 +1,0 @@
-# to do: korte note schrijven over hoe de hoeveelheid missingness regelateerd is aan de estimates --> bias trend boven 50% miss
-# does the miss prop shift the bias towards a particular direction?
-
-# set-up environment
-library(dplyr)
-library(mice)
-library(purrr)
-library(ggplot2)
-
-# set random seed
-set.seed(1111)
-
-# set proportions
-# p.inc <- c(.05, .25, .5, .75, .95)
-p.inc <- seq(0.05, 0.95, by = 0.05)
-
-# set number of simulations
-n.sims <- 1000
-
-# generate some data and get true values
-# data generating mechanism = multivariate normal distribution
-means <- c(12, 3, 0.5) #means
-vars  <- c(4, 16, 9) #variances
-R <- matrix(numeric(3 * 3), nrow = 3) #correlation matrix
-diag(R) <- 1 #set diagonal to 1
-R[upper.tri(R)] <-
-  R[lower.tri(R)] <- c(.5, .3, .4) #set bivariate correlations
-sigma <-
-  diag(sqrt(vars)) %*% R %*% diag(sqrt(vars)) #variance-covariance matrix
-dat <-
-  as.data.frame(mvtnorm::rmvnorm(n = 100, mean = means, sigma = sigma)) #create data
-colnames(dat) <- c("X1", "X2", "X3") #set predictors names
-
-# true estimates
-means <- colMeans(dat)
-# betas <- lm(X1~., data = dat)$coefficients
-
-# create patterns to ampute the data with multivariate missingness
-amp.pat <-
-  expand.grid(c(0, 1), c(0, 1), c(0, 1)) %>% #define all possible combinations of univariate and multivariate missingness
-  .[c(-1, -8),] #remove the completely (un)observed cases
-names(amp.pat) <-
-  ampute(dat)$patterns %>% names() #obtain correct names of patterns
-
-simulation <- function(dataset, proportions, pattern) {
-  # ampute data with different missingness
-  purrr::map_dfr(p.inc, function(p) {
-    mice::ampute(
-      data = dat,
-      patterns = pattern,
-      prop = p,
-      mech = "MCAR"
-    )$amp %>%
-      mice::mice(., print = FALSE) %>%
-      mice::complete(., "all") %>% purrr::map_df(., ~ {
-        colMeans(.)
-      }) %>%
-      colMeans() %>% .[2] %>% 
-      data.frame(est = ., prop = p)
-  })
-}
-
[…]

### 2021-12-16 shinymice/R/mod_01_data.R

@@ -1,103 +1,0 @@
-#' 01_data UI Function
-#'
-#' @description A shiny Module.
-#'
-#' @param id,input,output,session Internal parameters for {shiny}.
-#'
-#' @noRd 
-#'
-#' @importFrom shiny NS tagList 
-mod_01_data_ui <- function(id){
-  ns <- NS(id)
-  fluidPage(fluidRow(
-    column(
-      4,
-      tags$b("Explore the missingness"),
-      br(),
-      br(),
-      "1. Load the incomplete data.",
-      fileInput(
-        ns("dat"),
-        label = NULL,
-        buttonLabel = icon("search"),
-        placeholder = ".csv/.Rdata/.rda",
-        # add sav
-        accept = c(".rdata", ".rda", ".csv", "text", ".txt", ".tsv")
-      ),
-      no_br(),
-      "2. Check how much missingness there is in each variable.",
-      br(),
-      "3. Browse the dataset to view missing data points.",
-      br(),
-      "4. Evaluate the bivariate relations in the incomplete data.",
-      br(),
-      "5. Evaluate the distribution of variables conditional on missingness in another variable.",
-      br()
-    ),
-    column(8,
-           tabsetPanel(
-             tabPanel(
-               "Descriptives",
-               h6(
-                 "Note that variables with an asteriks ('*') are categorical, so the mean and standard deviation may not be meaningful."
-               ),
-               DT::DTOutput(ns("na_desc"))
-             ),
-             tabPanel("Browse",
-                      DT::DTOutput(ns("na_tab"))),
-             tabPanel(
-               "Scatter plot",
-               select_var(ns("var1")),
-               select_var(ns("var2")),
-               plotOutput(ns("na_plot"))
-             ),
-             tabPanel(
-               "Conditional distribution",
-               select_var(ns("var3")),
-               select_var(ns("var4")),
-               plotOutput(ns("cond_plot"))
-             )
-           ))
-  ))
-}
-    
-#' 01_data Server Functions
-#'
-#' @noRd 
-mod_01_data_server <- function(id){
-  moduleServer( id, function(input, output, session){
-    ns <- session$ns
-    data <- reactive({
-      d <- read_data(file = input$dat)
-      validate(
-        need(
-          is.data.frame(d),
-          "Sorry, {shinymice}  cannot process this dataset. Please check if you specified the right file."
-        )
-      )
[…]

### 2021-12-16 shinymice/R/utils_04_imputations.R

@@ -1,170 +1,0 @@
-#' Title Compute descriptive statistics for imputed data
-#'
-#' @param imp A multiply imputed data set (`mids`) object
-#'
-#' @return A datatable with descriptives
-imp_descr <- function(imp) {
-  # compute descriptives
-  tab <- imp %>% mice::complete("all") %>%
-    purrr::map_df(., ~ {
-      psych::describe(.)[, c("n", "mean", "sd", "min", "max", "median")] %>%
-        cbind(., variable = rownames(.))
-    }) %>%
-    aggregate(. ~ variable, data = ., FUN = "mean") %>%
-    dplyr::mutate(
-      n = as.integer(n),
-      mean = round(mean, 2),
-      sd = round(sd, 2),
-      n_imputed = as.integer(imp$nmis)
-    )
-  #tab[, 2] <- as.integer(tab[, 2])
-  # create datatable
-  dt <- DT::datatable(tab, rownames = FALSE, options = list(dom = 't'), selection = 'none') %>%
-    DT::formatStyle(
-      "n_imputed",
-      color = DT::styleInterval(cuts = 0, values = c("black", "#B61A51")),
-      fontWeight = "bold"
-    )
-  return(dt)
-}
-
-#' Title Preprocess imputations and initialize plot
-#'
-#' @param imp A multiply imputed data set (`mids`) object
-#' @param x A variable to plot
-#' @param y An optional second variable to plot
-#'
-#' @return A ggplot object with imputed data
-plot_imps <- function(imp, x, y = NULL) {
-  # parse inputs
-  if (is.null(y)) {
-    y <- x
-  }
-  # combine observed and imputed data
-  xy_obs <- imp$data %>%
-    cbind(datapoint = "observed",
-          .imp = 0,
-          .id = 1:nrow(.),
-          .) %>%
-    .[!is.na(imp$data[[x]]) &
-        !is.na(imp$data[[y]]),]
-  xy_imps <- imp %>%
-    mice::complete("long") %>%
-    cbind(datapoint = "imputed", .) %>%
-    .[.$.id %nin% xy_obs$.id, ]
-  xy_dat <- rbind(xy_obs, xy_imps) %>%
-    dplyr::mutate(datapoint = factor(datapoint, levels = c("observed", "imputed")))
-  # initialize plot
-  p <- xy_dat %>%
-    ggplot2::ggplot() +
-    theme_mice()
-  # output
-  return(p)
-}
-
-#' Title Plot boxplot/faceted barplot of imputed data
-#'
[…]

### 2021-12-16 shinymice/R/utils_05_save.R

@@ -1,17 +1,0 @@
-#' Title Save imputations in SPSS readable format
-#'
-#' @param imp A multiply imputed data set (mids) object
-#' @param file Temporary file location needed for shiny::downloadHandler()
-#'
-#' @return A SPPS readable multiply imputed data set
-mids2spss <- function(imp, file) {
-  .id <- NULL # avoid empty global variable binding
-  #extract a completed dataset (long format - all imputations stacked)
-  #rename the .imp variable to imputation_, such that SPSS can identify a multiply imputed dataset
-  out <-
-    imp %>%
-    mice::complete(action = "long", include = TRUE) %>%
-    dplyr::select(-.id) %>%
-    dplyr::rename("Imputation_" = ".imp")
-  haven::write_sav(out, file)
-} # what to do after export? See https://bookdown.org/mwheymans/bookmi/multiple-imputation.html#multiple-imputation-in-spss

### 2021-12-16 shinymice/R/utils_server.R

@@ -1,86 +1,0 @@
-#' server utility
-#'
-#' @description A shiny utility
-#'
-#' @param Internal parameters for the utility.
-#'
-#' @return The return value, if any, from executing the utility.
-#'
-#' @export
-
-# not in
-`%nin%` <- Negate(`%in%`)
-
-# mice theme
-theme_mice <- function() {
-  # change default plotting options (move this outside of the function to unnecessary avoid re-running?)
-  ggplot2::update_geom_defaults("point", list(
-    shape = 21,
-    size = 1.9,
-    stroke = 1.1,
-    alpha = 0.5
-  ))
-  ggplot2::update_geom_defaults("boxplot", list(size = 1))
-  ggplot2::update_geom_defaults("line", list(size = 1))
-  # mice style settings
-  theme <- list(
-    ggplot2::theme_classic(),
-    ggplot2::theme(legend.position = "top"),
-    ggplot2::scale_color_manual(
-      values = c(
-        "observed" = mice:::mdc(1),
-        "missing" = mice:::mdc(2),
-        "imputed" = mice:::mdc(2)
-      )
-    ),
-    ggplot2::scale_fill_manual(
-      values = c(
-        "observed" = mice:::mdc(1),
-        "missing" = mice:::mdc(2),
-        "imputed" = mice:::mdc(2)
-      )
-    ),
-    ggplot2::scale_size_manual(values = c(
-      "observed" = 1,
-      "imputed" = 0.5
-    )),
-    ggplot2::labs(
-      color = NULL,
-      fill = NULL,
-      size = NULL
-    )
-  )
-  # output
-  return(theme)
-}
-
-# hide some plotly options, see https://plotly-r.com/control-modebar.html
-clean_plotly <- function(pp, ...) {
-  plotly::config(
-    p = pp,
-    displaylogo = FALSE,
-    modeBarButtonsToRemove = c(
-      "zoomIn2d",
-      "zoomOut2d",
-      "autoScale2d",
-      "pan2d",
-      "lasso2d",
-      "select2d",
-      "toggleSpikelines",
-      "hoverCompareCartesian",
-      "hoverClosestCartesian"
-    )
-  )
-}
-
-# mock plot for complete datasets
-plot_a_mouse <- function() {
-  ggplot2::ggplot(mouse, ggplot2::aes(x = X, y = Y)) +
-    ggplot2::geom_point(size = 5,
-                        shape = 20,
[…]

### 2021-12-16 shinymice/app.R

@@ -1,7 +1,0 @@
-# Launch the ShinyApp (Do not remove this comment)
-# To deploy, run: rsconnect::deployApp()
-# Or use the blue button on top of this file
-
-pkgload::load_all(export_all = FALSE,helpers = FALSE,attach_testthat = FALSE)
-options( "golem.app.prod" = TRUE)
-shinymice::run_app() # add parameters here (if any)

### 2021-12-16 shinymice/dev/02_dev.R

@@ -1,80 +1,0 @@
-# Building a Prod-Ready, Robust Shiny Application.
-#
-# README: each step of the dev files is optional, and you don't have to
-# fill every dev scripts before getting started.
-# 01_start.R should be filled at start.
-# 02_dev.R should be used to keep track of your development during the project.
-# 03_deploy.R should be used once you need to deploy your app.
-#
-#
-###################################
-#### CURRENT FILE: DEV SCRIPT #####
-###################################
-
-# Engineering
-
-## Dependencies ----
-## Add one line by package you want to add as dependency
-usethis::use_package("mice")
-usethis::use_package("ggplot2")
-usethis::use_package("dplyr")
-usethis::use_package("plotly")
-usethis::use_package("psych")
-usethis::use_package("haven")
-usethis::use_package("shinyjs")
-usethis::use_package("waiter")
-usethis::use_pipe()
-
-## Add modules ----
-## Create a module infrastructure in R/
-golem::add_module(name = "00_home") 
-golem::add_module(name = "01_data") 
-golem::add_module(name = "02_model") 
-golem::add_module(name = "03_convergence") 
-golem::add_module(name = "04_imputations") 
-golem::add_module(name = "05_save") 
-
-## Add helper functions ----
-## Creates ftc_* and utils_*
-golem::add_utils("ui")
-golem::add_utils("server")
-# golem::add_utils("00_home")
-golem::add_utils("01_data")
-golem::add_utils("02_model")
-golem::add_utils("03_convergence")
-golem::add_utils("04_imputations")
-golem::add_utils("05_save")
-
-## External resources
-## Creates .js and .css files at inst/app/www
-golem::add_js_file("script")
-golem::add_js_handler("handlers")
-golem::add_css_file("custom")
-
-## Add internal datasets ----
-## If you have data in your package
-# usethis::use_data_raw(name = "my_dataset", open = FALSE)
-# usethis::use_data(mouse)
-
-## Tests ----
-## Add one line by test you want to create
-usethis::use_test("plot_md_pattern")
-usethis::use_test("plot_flux")
-usethis::use_test("plot_pred_matrix")
-
-# Documentation
-
-## Vignette ----
-usethis::use_vignette("shinymice")
-usethis::use_vignette("ggmice")
-devtools::build_vignettes()
-
-## Code coverage ----
-## (You'll need GitHub there)
-usethis::use_github()
-usethis::use_travis()
[…]

### 2021-12-16 README.Rmd
Remove duplicate info from the repo

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# shinymice

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
<!-- badges: end -->

This repository serves the development of `{shinymice}`, an interactive evaluation device for missing data by Hanne Oberman, guided by Gerko Vink and Stef van Buuren.

![shinymice hex sticker](inst/app/www/logo.png)

For a demo version of the app, see https://hanneoberman.shinyapps.io/shinymice-demo/.

-Hanne (16-12-2021)

<!-- The goal of shinymice is to ... -->

<!-- ## Installation -->

<!-- You can install the released version of shinymice from [CRAN](https://CRAN.R-project.org) with: -->

<!-- ``` r -->
<!-- install.packages("shinymice") -->
<!-- ``` -->

<!-- And the development version from [GitHub](https://github.com/) with: -->

<!-- ``` r -->
<!-- # install.packages("devtools") -->
<!-- devtools::install_github("amices/shinyMice") -->
<!-- ``` -->
<!-- ## Example -->

<!-- This is a basic example which shows you how to solve a common problem: -->

<!-- ```{r example} -->
<!-- library(shinymice) -->
<!-- ## basic example code -->
<!-- ``` -->

<!-- What is special about using `README.Rmd` instead of just `README.md`? You can include R chunks like so: -->

<!-- ```{r cars} -->
<!-- summary(cars) -->
<!-- ``` -->

<!-- You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date. -->

<!-- You can also embed plots, for example: -->

<!-- ```{r pressure, echo = FALSE} -->
<!-- plot(pressure) -->
<!-- ``` -->

<!-- In that case, don't forget to commit and push the resulting figure files, so they display on GitHub! -->
